### 2022-03-03 code/translocation_survival_analysis.Rmd
First commit

---
title: "Predictors of frog survival following translocations - analyses"
author: "Roland Knapp"
output: html_notebook
---

The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package, brms.

### 2022-03-03 code/translocation_survival_createdata.Rmd
First commit

---
title: "Predictors of frog survival following translocations - create dataset"
author: "Roland Knapp"
output: html_notebook
---

The code in this notebook creates a dataset for use in an analysis to identify predictors of frog survival following translocation (translocation_survival_analysis.Rmd). The code is written in distinct chunks to allow review of the output from one chunk before running the next. Users without credentials to access the database will not be able to create the raw data files, but can use the saved files to run the remainder of the code. Frog translocation and survival data were generated using code in the [SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repository.

### Load packages

```{r load_packages}
library(RPostgreSQL)
library(tidyverse) 
library(lubridate)
```

## Create site characteristics dataset

### Connect to database

```{r db_connect1}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve site_ids and site characteristics from database

```{sql retrieve_site_data, connection=con, output.var="translocation_sites"}
select
  id as site_id,
  elevation
from site
where id in (70134, 70370, 70413, 70414, 70449, 70505, 70550, 70556, 70619, 70628, 70641, 74976)
order by site_id
```

### Disconnect from database

```{r db_disconnect1}
dbDisconnect(con)
rm(con)
```

### Add additional site characteristics

```{r sites_add_shore}
translocation_sites <- translocation_sites %>% 
  mutate(shore = if_else(site_id %in% c(70370, 70413, 70550, 70556, 70619, 70628, 74976), 1, 0),
         site_id = as.character(site_id))
```

### Save site characteristics dataset

```{r save_site_data}
translocation_sites %>% write_csv(here::here("data", "clean", "translocation_sites.csv"))
# translocation_sites <- read_csv(here::here("data", "translocation_sites.csv"))
```

## Create translocated frogs dataset

### Combine translocation files from cmr-analysis repo

```{r retrieve_translocated_frogs_data }
files <- fs::dir_ls("../../cmr-analysis/data/clean", glob = "*translocation.csv")
frog_translocation <- read_csv(files, id = "path") %>% 
  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  
  filter(release_date < "2021-01-01") %>% 
  rename(pit_tag_ref = pit_tag_id) %>% 
  select(-path)
rm(files)
```

### Save raw translocated frog dataset

```{r save_raw_translocated_frogs_datra}
frog_translocation %>% write_csv(here::here("data", "raw", "frog_translocation_raw.csv"))
```

### Add frog characteristics from database

#### Connect to database

```{r db_connect2}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

#### Retrieve frog characteristics from database

```{sql retrieve_translocation_data, connection=con, output.var="frog_characteristics"}
select
  relocate.id as relocate_id,
  release_siteid1,
  release_date,
  collect_siteid,
  type,
  pit_tag_ref,
  sex,
  length,
  weight,
  swab_id,
  bd_load
from relocate
inner join relocate_frog on relocate.id = relocate_id
left join bd_load on swab_id = sample_id  
where release_siteid1 in (70134, 70279, 70370, 70413, 70414, 70449, 70505, 70550, 70556, 70619, 70628, 70641, 74976) and
  (replicate = 1 or replicate is null)
order by release_siteid1
```

#### Disconnect from database

```{r db_disconnect2}
dbDisconnect(con)
rm(con)
```

### Save raw frog characteristics dataset

```{r save_frog_characteristics_data}
frog_characteristics %>% write_csv(here::here("data", "raw", "frog_characteristics_raw.csv"))
# translocation_sites <- read_csv(here::here("data", "translocation_sites.csv"))
```

## Create frog survival dataset

### Combine survival files from cmr-analysis repo

```{r retrieve_frog_survival_data}
files <- fs::dir_ls("../../cmr-analysis/out/tables", glob = "*ind.csv")
frog_survival <- read_csv(files, id = "path") %>% 
  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  
  rename(years_since_intro = years_since_introduction,
         pit_tag_ref = pit_tag_id) %>% 
  filter(years_since_intro == 1) %>% 
  select(site_id, release_date, years_since_intro, pit_tag_ref, median_survival)
rm(files)
```

### Save raw frog survival dataset

```{r save_raw_frog_survival_data}
frog_survival %>% write_csv(here::here("data", "raw", "frog_survival_raw.csv"))
```

### Data checks - first pass

```{r data_checking}
# Check for duplicate pit_tag_refs
frog_translocation %>% count(pit_tag_ref) %>% filter(n > 1)
frog_characteristics %>% count(pit_tag_ref) %>% filter(n > 1)
frog_survival %>% count(pit_tag_ref) %>% filter(n > 1)
```

### Clean data - first pass

```{r data_cleaning_step1}
# Duplicate pit_tag_ref found in frog_translocation, frog_characteristics, and frog survival datasets.
# Shown as released at 70641 and 74976, but subsequently only recaptured at 74976 so that record is incorrect.
# Recode as pit_tag_ref not found in database
frog_translocation <- frog_translocation %>% mutate(
  pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
frog_characteristics <- frog_characteristics %>% mutate(
  pit_tag_ref = replace(pit_tag_ref, release_siteid1 == 70641 & pit_tag_ref == "900067000117396", "900067000117397"))
frog_survival <-frog_survival %>% mutate(
  pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
```

### Add frog and site characteristics to frog translocation dataset
```{r join_frog_data}
frog_characteristics <- frog_characteristics %>% 
  mutate(release_siteid1 = as.character(release_siteid1)) %>% 
  select(-relocate_id, -release_siteid1, -release_date, -type)
  as_tibble()

frog_survival <- frog_survival %>% 
  select(pit_tag_ref, median_survival)

frog_translocation <- frog_translocation %>% 
  left_join(frog_characteristics, by = "pit_tag_ref") %>% 
  left_join(frog_survival, by = "pit_tag_ref") %>% 
  left_join(translocation_sites, by = "site_id")
```

### Check data - second pass
```{r}
# Check for null values
frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))

# Check that site_ids match expectation
frog_translocation %>% distinct(site_id)
frog_translocation %>% distinct(site_id, collect_siteid)

# Check if sex at collection matches sex on recapture

# Plot length x weight to check for outliers

```

### Clean data - second pass

```{r}
# Missing frog characteristics associated with pit_tag_ref = 900067000114499. 
# Replace null values (except bd_load) with values from capture at release site early the next summer. 
# Bd_load not replaced because load could have changed between release and recapture.

# Missing frog characteristics associated with pit_tag_ref = 900118001027669 
# Replace null values with values from capture at release site 1-2 weeks later. 
# Bd_load replaced because load likely would not have changed much between release and recapture.

frog_translocation <- frog_translocation %>% 
  mutate(
    collect_siteid = replace(collect_siteid, is.na(collect_siteid) & site_id == "70641" & pit_tag_ref == "900067000114499", "72996"),
    sex = replace(sex, is.na(sex) & site_id == "70641" & pit_tag_ref == "900067000114499", "f"),
    length = replace(length, is.na(length) & site_id == "70641" & pit_tag_ref == "900067000114499", 52),
    weight = replace(weight, is.na(weight) & site_id == "70641" & pit_tag_ref == "900067000114499", 12),
    length = replace(length, is.na(length) & site_id == "70413" & pit_tag_ref == "900118001027669", 52),
    weight = replace(weight, is.na(weight) & site_id == "70413" & pit_tag_ref == "900118001027669", 16),
    swab_id = replace(swab_id, swab_id == "" & site_id == "70413" & pit_tag_ref == "900118001027669", "YFS364"),
    bd_load = replace(bd_load, is.na(bd_load) & site_id == "70413" & pit_tag_ref == "900118001027669", 0))

# Collect_siteid 72974 merged with adjacent larger 70459
frog_translocation <- frog_translocation %>% 
  mutate(collect_siteid = replace(collect_siteid, collect_siteid == "72974" & site_id == "70556", "70459"))
```

### Save clean frog translocation dataset

```{r save_clean_frog_translocation_data}
frog_translocation %>% write_csv(here::here("data", "clean", "frog_translocation.csv"))
```

### 2022-03-03 doc/notebook/notebook.Rmd
First commit

---
title: "Notebook"
author: "Roland Knapp"
date: "`r Sys.Date()`"
editor_options:
  markdown:
    mode: gfm
---

## Predictors of frog survival following translocation

### Dataset structure

-   site_id: site to which frogs were translocated
-   elevation
-   release_date
-   release_day
-   release_year
-   siteid_donor
-   order: first translocation to site = 0, subsequent translocation = 1
-   type: translocation conducted on foot versus by helicopter
-   shore: suitability of shore/bank habitat for frog overwintering
-   pit_tag_ref
-   survival: individual-level estimated survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation
-   sex
-   length: at release
-   weight: at release
-   condition: index based on length and weight
-   swab_id
-   bd_load: at release

Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.

### Model structure

survival \~ elevation + siteid_donor + release_day + release_year + order + shore + sex + length + condition + log(bdload) \| site_id

Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include release_year, perhaps nested within site_id? Given that some sites only have a single release_year, will this cause problems?

### 2022-03-03 doc/manuscript/pnas_demo/pnas_demo.Rmd
Add PNAS manuscript template

---
title: Reintroduction of disease-resistant individuals allows recovery of an endangered frog in the presence of a lethal fungal disease

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Roland A. Knapp
    affiliation: a,b,1
  - name: Thomas C. Smith
    affiliation: a,b
  - name: Mark Q. Wilber
    affiliation: c
  - name: Andrew P. Rothstein
    affiliation: d,e,2
  - name: Allison Q. Byrne
    affiliation: d,e,2
  - name: Maxwell B. Joseph
    affiliation: f

address:
  - code: a
    address: Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA 93546
  - code: b
    address: Earth Research Institute, University of California-Santa Barbara, Santa Barbara, CA 93106
  - code: c
    address: Department of Forestry, Wildlife, and Fisheries, University of Tennessee, Knoxville, TN, xxxxx
  - code: d
    address: Department of Environmental Science, Policy, and Management, University of California-Berkeley, Berkeley, CA, 94720
  - code: e
    address: Museum of Vertebrate Zoology, University of California-Berkeley, Berkeley, CA, 94720
  - code: f
    address: Earth Lab, University of Colorado, Boulder, CO, 80303

corresponding_author:
  code: 1
  text: "To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu"

# For footer text
lead_author_surname: Knapp

## Remove this if not required    
equal_authors:
  code: 2
  text: "A.P.R. and A.Q.B. contributed equally to this work."

author_contributions: |
  R.A.K, T.C.S, M.Q.W, A.P.R., A.Q.B, and M.B.J. designed research; R.A.K, T.C.S, M.Q.W, A.P.R., and A.Q.B performed research; R.A.K, M.Q.W, A.P.R., A.Q.B, and M.B.J. analyzed data; M.B.J. contributed software; and R.A.K, T.C.S, M.Q.W, A.P.R., A.Q.B, and M.B.J. wrote the paper.

## Remove this if not required
conflict_of_interest: |
  The authors declare no conflicts of interest.

abstract: |
  Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.

significance: |
  Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers.

acknowledgements: |
  Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.

keywords:
  - population recovery
  - amphibian declines
  - emerging infectious disease
  - reintroduction
  - Batrachochytrium dendrobatis

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
pnas_type: pnasresearcharticle

bibliography: pnas.bib
csl: pnas.csl

## change to true to add optional line numbering
lineno: false

output: rticles::pnas_article
---

This PNAS journal template is provided to help you write your work in the
correct journal format. Instructions for use are provided below.

Note: please start your introduction without including the word
"Introduction" as a section heading (except for math articles in the
Physical Sciences section); this heading is implied in the first
paragraphs.

Results {.unnumbered}
========================================

Evolution of Reduced Disease Susceptibility {#results-subheading1 .unnumbered}
-------------------

Reintroduction of Disease-resistant Frogs {#results-subheading2 .unnumbered}
-------------------

Viability of Reestablished Populations {#results-subheading3 .unnumbered}
-------------------

Conceptual Model of Population Establishment {#results-subheading4 .unnumbered}
-------------------

Discussion {.unnumbered}
========================================

Conclusions {#discussion-subheading1 .unnumbered}
-------------------

Materials and Methods {.unnumbered}
========================================

Statistical Analyses {#methods-subheading1 .unnumbered}
-------------------

Data Availability {#methods-subheading2 .unnumbered}
-------------------

Guide to using this template {.unnumbered}
========================================

Please note that whilst this template provides a preview of the typeset
manuscript for submission, to help in this preparation, it will not
necessarily be the final publication layout. For more detailed
information please see the [PNAS Information for
Authors](http://www.pnas.org/site/authors/format.xhtml).

Author Affiliations {#author-affiliations .unnumbered}
-------------------

Include department, institution, and complete address, with the
ZIP/postal code, for each author. Use lower case letters to match
authors with institutions, as shown in the example. Authors with an
ORCID ID may supply this information at submission.

Submitting Manuscripts {#submitting-manuscripts .unnumbered}
----------------------

All authors must submit their articles at
[PNAScentral](http://www.pnascentral.org/cgi-bin/main.plex). If you are
using Overleaf to write your article, you can use the "Submit to PNAS"
option in the top bar of the editor window.

Format {#format .unnumbered}
------

Many authors find it useful to organize their manuscripts with the
following order of sections; Title, Author Affiliation, Keywords,
Abstract, Significance Statement, Results, Discussion, Materials and
methods, Acknowledgments, and References. Other orders and headings are
permitted.

Manuscript Length {#manuscript-length .unnumbered}
-----------------

PNAS generally uses a two-column format averaging 67 characters,
including spaces, per line. The maximum length of a Direct Submission
research article is six pages and a PNAS PLUS research article is ten
pages including all text, spaces, and the number of characters displaced
by figures, tables, and equations. When submitting tables, figures,
and/or equations in addition to text, keep the text for your manuscript
under 39,000 characters (including spaces) for Direct Submissions and
72,000 characters (including spaces) for PNAS PLUS.

References {#references .unnumbered}
----------

References should be cited in numerical order as they appear in text;
this will be done automatically via bibtex, e.g. @belkin2002using and
@berard1994embedding [@coifman2005geometric]. All references, including
for the SI, should be included in the main manuscript file. References
appearing in both sections should not be duplicated. SI references
included in tables should be included with the main reference section.

Data Archival {#data-archival .unnumbered}
-------------

PNAS must be able to archive the data essential to a published article.
Where such archiving is not possible, deposition of data in public
databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata,
and others outlined in the Information for Authors, is acceptable.

Language-Editing Services {#language-editing-services .unnumbered}
-------------------------

Prior to submission, authors who believe their manuscripts would benefit
from professional editing are encouraged to use a language-editing
service (see list at www.pnas.org/site/authors/language-editing.xhtml).
PNAS does not take responsibility for or endorse these services, and
their use has no bearing on acceptance of a manuscript for publication.

![Placeholder image of a frog with a long example caption to show
justification setting.<span data-label="fig:frog"></span>](frog.png)

Digital Figures {#sec:figures .unnumbered}
---------------

Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for
figures that will appear in the main text, and images must be final
size. Authors may submit U3D or PRC files for 3D images; these must be
accompanied by 2D representations in TIFF, EPS, or high-resolution PDF
format. Color images must be in RGB (red, green, blue) mode. Include the
font files for any text.

Figures and Tables should be labelled and referenced in the standard way
using the `\label{}` and `\ref{}` commands.

Figure \[fig:frog\] shows an example of how to insert a column-wide
figure. To insert a figure wider than one column, please use the
`\begin{figure*}...\end{figure*}` environment. Figures wider than one
column should be sized to 11.4 cm or 17.8 cm wide.

Single column equations {#single-column-equations .unnumbered}
-----------------------

Authors may use 1- or 2-column equations in their article, according to
their preference.

To allow an equation to span both columns, options are to use the
`\begin{figure*}...\end{figure*}` environment mentioned above for
figures, or to use the `\begin{widetext}...\end{widetext}` environment
as shown in equation \[eqn:example\] below.

Please note that this option may run into problems with floats and
footnotes, as mentioned in the [cuted package
documentation](http://texdoc.net/pkg/cuted). In the case of problems
with footnotes, it may be possible to correct the situation using
commands `\footnotemark` and `\footnotetext`.

$$\begin{aligned}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \label{eqn:example} \\
       &=x^3+3x^2y+3xy^3+x^3. 
\end{aligned}$$

<!-- pandoc writes all tables using longtable, which fails in 2-column mode

  Species                    CBS     CV     G3
  ----------------------- ------ ------ ------
  1\. Acetaldehyde           0.0    0.0    0.0
  2\. Vinyl alcohol          9.1    9.6   13.5
  3\. Hydroxyethylidene     50.8   51.2   54.0

  : Comparison of the fitted potential energy surfaces and ab initio
  benchmark electronic energy calculations

-->

Supporting Information (SI) {#supporting-information-si .unnumbered}
---------------------------

The main text of the paper must stand on its own without the SI. Refer
to SI in the manuscript at an appropriate point in the text. Number
supporting figures and tables starting with S1, S2, etc. Authors are
limited to no more than 10 SI files, not including movie files. Authors
who place detailed materials and methods in SI must provide sufficient
detail in the main text methods to enable a reader to follow the logic
of the procedures and results and also must reference the online
methods. If a paper is fundamentally a study of a new method or
technique, then the methods must be described completely in the main
text. Because PNAS edits SI and composes it into a single PDF, authors
must provide the following file formats only.

### SI Text {#si-text .unnumbered}

Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a
PDF with the same file name for visual reference).

### SI Figures {#si-figures .unnumbered}

Provide a brief legend for each supporting figure after the supporting
text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or
GIF format; figures may not be embedded in manuscript text. When saving
TIFF files, use only LZW compression; do not use JPEG compression. Do
not save figure numbers, legends, or author names as part of the image.
Composite figures must be pre-assembled.

### 3D Figures {#d-figures .unnumbered}

Supply a composable U3D or PRC file so that it may be edited and
composed. Authors may submit a PDF file but please note it will be
published in raw format and will not be edited or composed.

### SI Tables {#si-tables .unnumbered}

Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a
PDF with the same file name for visual reference); include only one
table per file. Do not use tabs or spaces to separate columns in Word
tables.

### SI Datasets {#si-datasets .unnumbered}

Supply Excel (.xls), RTF, or PDF files. This file type will be published
in raw format and will not be edited or composed.

### SI Movies {#si-movies .unnumbered}

Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media
(wmv), animated GIF (gif), or MPEG files and submit a brief legend for
each movie in a Word or RTF file. All movies should be submitted at the
desired reproduction size and length. Movies should be no more than 10
MB in size.

### Still images {#still-images .unnumbered}

Authors must provide a still image from each video file. Supply TIFF,
EPS, high-resolution PDF, JPEG, or GIF files.

### Appendices {#appendices .unnumbered}

PNAS prefers that authors submit individual source files to ensure
readability. If this is not possible, supply a single PDF file that
contains all of the SI associated with the paper. This file type will be
published in raw format and will not be edited or composed.

<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
\showmatmethods
\showacknow
\pnasbreak

### 2022-03-03 doc/manuscript/pnas_demo/pnas_demo.Rmd
Update template to latest Rmarkdown format

@@
 title: Reintroduction of disease-resistant individuals allows recovery of an endangered frog in the presence of a lethal fungal disease
-
-# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 author:
@@
     affiliation: f
+address:
 
-address:
   - code: a
@@
     address: Earth Lab, University of Colorado, Boulder, CO, 80303
-
 corresponding_author:
@@
   text: "To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu"
-
-# For footer text
 lead_author_surname: Knapp
-
-## Remove this if not required    
 equal_authors:
@@
   text: "A.P.R. and A.Q.B. contributed equally to this work."
-
 author_contributions: |
@@
 wrote the paper.
-
-## Remove this if not required
 conflict_of_interest: |
   The authors declare no conflicts of interest.
-
 abstract: |
@@
 References in the abstract must be cited in full within the abstract itself and cited in the text.
-
 significance: |
@@
 The Significance Statement appears in the paper itself and is required for all research papers.
-
 acknowledgements: |
   Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.
-
 keywords:
@@
   - Batrachochytrium dendrobatis
-
 ## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
 pnas_type: pnasresearcharticle
-
 bibliography: pnas.bib
 csl: pnas.csl
-
-## change to true to add optional line numbering
 lineno: false
-
 output: rticles::pnas_article
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
 
-This PNAS journal template is provided to help you write your work in the
-correct journal format. Instructions for use are provided below.
+This PNAS journal template is provided to help you write your work in
+the correct journal format. Instructions for use are provided below.
 
@@
 
-Results {.unnumbered}
-========================================
+# Results {.unnumbered}
 
-Evolution of Reduced Disease Susceptibility {#results-subheading1 .unnumbered}
--------------------
+## Evolution of Reduced Disease Susceptibility {#results-subheading1 .unnumbered}
 
-Reintroduction of Disease-resistant Frogs {#results-subheading2 .unnumbered}
--------------------
+## Reintroduction of Disease-resistant Frogs {#results-subheading2 .unnumbered}
 
-Viability of Reestablished Populations {#results-subheading3 .unnumbered}
--------------------
+## Viability of Reestablished Populations {#results-subheading3 .unnumbered}
 
-Conceptual Model of Population Establishment {#results-subheading4 .unnumbered}
--------------------
+## Conceptual Model of Population Establishment {#results-subheading4 .unnumbered}
 
-Discussion {.unnumbered}
-========================================
+# Discussion {.unnumbered}
 
-Conclusions {#discussion-subheading1 .unnumbered}
--------------------
+## Conclusions {#discussion-subheading1 .unnumbered}
 
-Materials and Methods {.unnumbered}
-========================================
+# Materials and Methods {.unnumbered}
 
-Statistical Analyses {#methods-subheading1 .unnumbered}
--------------------
+## Statistical Analyses {#methods-subheading1 .unnumbered}
 
-Data Availability {#methods-subheading2 .unnumbered}
--------------------
+## Data Availability {#methods-subheading2 .unnumbered}
 
-Guide to using this template {.unnumbered}
-========================================
+# Guide to using this template {.unnumbered}
 
@@
 
-Author Affiliations {#author-affiliations .unnumbered}
--------------------
+## Author Affiliations {#author-affiliations .unnumbered}
 
@@
 
-Submitting Manuscripts {#submitting-manuscripts .unnumbered}
-----------------------
+## Submitting Manuscripts {#submitting-manuscripts .unnumbered}
 
@@
 
-Format {#format .unnumbered}
-------
+## Format {#format .unnumbered}
 
@@
 
-Manuscript Length {#manuscript-length .unnumbered}
------------------
+## Manuscript Length {#manuscript-length .unnumbered}
 
@@
 
-References {#references .unnumbered}
-----------
+## References {#references .unnumbered}
 
@@
 this will be done automatically via bibtex, e.g. @belkin2002using and
-@berard1994embedding [@coifman2005geometric]. All references, including
+[@berard1994embedding; @coifman2005geometric]. All references, including
 for the SI, should be included in the main manuscript file. References
@@
 
-Data Archival {#data-archival .unnumbered}
--------------
+## Data Archival {#data-archival .unnumbered}
 
@@
 
-Language-Editing Services {#language-editing-services .unnumbered}
--------------------------
+## Language-Editing Services {#language-editing-services .unnumbered}
 
@@
 ![Placeholder image of a frog with a long example caption to show
-justification setting.<span data-label="fig:frog"></span>](frog.png)
+justification setting.](frog.png)
 
-Digital Figures {#sec:figures .unnumbered}
----------------
+## Digital Figures {#sec:figures .unnumbered}
 
@@
 
-Figure \[fig:frog\] shows an example of how to insert a column-wide
+Figure $$fig:frog$$ shows an example of how to insert a column-wide
 figure. To insert a figure wider than one column, please use the
@@
 
-Single column equations {#single-column-equations .unnumbered}
------------------------
+## Single column equations {#single-column-equations .unnumbered}
 
@@
 figures, or to use the `\begin{widetext}...\end{widetext}` environment
-as shown in equation \[eqn:example\] below.
+as shown in equation $$eqn:example$$ below.
 
@@
 
+```{=html}
 <!-- pandoc writes all tables using longtable, which fails in 2-column mode
@@
 -->
-
-Supporting Information (SI) {#supporting-information-si .unnumbered}
----------------------------
+```
+## Supporting Information (SI) {#supporting-information-si .unnumbered}
 
@@
 <!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
+
+```{=tex}
 \showmatmethods
@@
 \pnasbreak
+```

### 2022-03-04 code/translocation_survival_createdata.Rmd
Add additional dataset creation steps

@@
 # Check that site_ids match expectation
-frog_translocation %>% distinct(site_id)
 frog_translocation %>% distinct(site_id, collect_siteid)
-
-# Check if sex at collection matches sex on recapture
-
-# Plot length x weight to check for outliers
-
 ```
@@
 
+### Check data - third pass
+```{r}
+#### Connect to database
+con = dbConnect(dbDriver("PostgreSQL"),
+                host = Sys.getenv("host"), 
+                port = Sys.getenv("port"), 
+                dbname = Sys.getenv("dbname"),
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password")
+               )
+```
+```{sql retrieve_capture_data, connection=con, output.var="frog_captures"}
+select
+	site_id,
+	visit_date,
+	survey_type,
+	capture_life_stage,
+	capture_animal_state,
+	pit_tag_ref, 
+	sex 
+from visit
+inner join survey on visit.id = visit_id
+inner join capture_survey on survey.id = survey_id
+inner join surveyor on surveyor.id = surveyor_id
+where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
+	survey_type = 'cmr' and
+	capture_animal_state != 'dead' 
+```
+```{r}
+dbDisconnect(con)
+rm(con)
+```
+```{r}
+# Check if sex at collection matches sex on recapture
+frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1)
+frog_captures %>% distinct(pit_tag_ref, sex) %>% 
+  rename(sexcapture = sex) %>% 
+  inner_join(frog_translocation, by = "pit_tag_ref") %>% 
+  filter(sex != sexcapture)
+```
+```{r}
+# Plot length x weight to check for outliers
+frog_translocation %>% ggplot(aes(x = length, y = weight)) +
+  geom_point() +
+  geom_smooth(method = lm)
+
+frog_translocation %>% ggplot(aes(x = log10(length), y = log10(weight))) +
+  geom_point() +
+  geom_smooth(method = gam)
+
+```
+No obvious outliers.
+
+### Clean data - third pass
+
+```{r}
+# Change sex at time of translocation based on subsequent recaptures
+frog_translocation <- frog_translocation %>% 
+  mutate(sex = replace(sex, pit_tag_ref == "900067000117464" & sex == "f", "m"),
+         sex = replace(sex, pit_tag_ref == "982000365440205" & sex == "f", "m"),
+         sex = replace(sex, pit_tag_ref == "900067000114152" & sex == "m", "f"))
+```
+
+### Add translocation order
+
+```{r}
+frog_translocation %>% 
+  distinct(site_id, release_date) %>% 
+  group_by(site_id) %>% 
+  mutate(order = seq_len(n())) %>% 
+  arrange(site_id, release_date) %>% 
+  inner_join(frog_translocation, by = c("site_id", "release_date")) 
+  ungroup() -> frog_translocation
+```
+
+### Extract day-of-year and year from date
+
+```{r}
+frog_translocation <- frog_translocation %>% 
+  mutate(day = yday(release_date),
+         year = year(release_date))
+```
+
+### Check for null values
+
+```{r}
+# Check for null values
+frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+### Create final dataset
+
+```{r}
+frog_translocation <- frog_translocation %>% 
+  rename(survival = median_survival,
+         donor = collect_siteid,
+         date = release_date) %>% 
+  select(site_id, elevation, shore, date, day, year, order, pit_tag_ref, donor, sex, length, weight, bd_load, survival) 
+```
+
 ### Save clean frog translocation dataset
@@
 ```{r save_clean_frog_translocation_data}
-frog_translocation %>% write_csv(here::here("data", "clean", "frog_translocation.csv"))
+frog_translocation %>% 
+   write_csv(here::here("data", "clean", "frog_translocation.csv"))
 ```

### 2022-03-04 doc/notebook/notebook.Rmd
Change output format to github_document
Add notes about frog_translocation dataset

@@
 date: "`r Sys.Date()`"
-editor_options:
-  markdown:
-    mode: gfm
+output:
+  github_document
 ---
+
+```{r}
+library(tidyverse)
+```
 
@@
 -   elevation
--   release_date
--   release_day
--   release_year
--   siteid_donor
+-   date
+-   day
+-   year
+-   donor
 -   order: first translocation to site = 0, subsequent translocation = 1
--   type: translocation conducted on foot versus by helicopter
+-   type: translocation conducted on foot versus by helicopter (not included)
 -   shore: suitability of shore/bank habitat for frog overwintering
@@
 
+Frog condition might be problematic because size and residuals are correlated.
+Specifically, the largest frogs have the largest positive residuals from a weight:length regression line, even after log10 transformation.
+As such, length and condition are probably not independent.
+
+```{r, fig.width = 3, fig.align = "center"}
+read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 
+ggplot(aes(x = log10(length), y = log10(weight))) +
+  geom_point() +
+  geom_smooth(method = lm)
+```
+
+Another consideration when designing the analysis is that there are 12 records in which bd_load is null.
+Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival.
+39 frogs were translocated on that date, so should probably drop the 9 records where bd_load = NA prior to running the analysis.
+If bd_load is not an important predictor, can run analysis using the full dataset.
+
 ### Model structure
 
-survival \~ elevation + siteid_donor + release_day + release_year + order + shore + sex + length + condition + log(bdload) \| site_id
+survival \~ elevation + shore + day + year + order + donor + sex + length + log(bdload) \| site_id
 
@@
 Given that some sites only have a single release_year, will this cause problems?
+
+Final steps before building model (in analysis Rmd):
+
+-   Change to factor: order, day?, year?
+-   Standardize variables?
+-   Survival: Create 0/1 variable from probabilities. Change to factor?

### 2022-03-04 doc/notebook/notebook.Rmd
Edited code for frog length:weight plot

@@
   geom_smooth(method = lm)
+plot
 ```

### 2022-03-04 doc/notebook/notebook.Rmd
Change size of frog length:weight plot

--- 2022-03-04 doc/notebook/notebook.Rmd
@@
 
-```{r, fig.width = 3, fig.align = "center"}
+```{r, fig.width = 6, fig.align = "center"}
 read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 

### 2022-03-05 doc/notebook/notebook.Rmd
Change format to html_notebook
Hide code for frog length:weight plot

--- 2022-03-04 doc/notebook/notebook.Rmd
@@
 output:
-  github_document
+  html_notebook
 ---
 
-```{r}
+```{r message=FALSE}
 library(tidyverse)
@@
 
-```{r, fig.width = 3, fig.align = "center"}
+```{r frog-lenth-weight-plot, echo=FALSE}
 read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 

### 2022-03-05 code/translocation_survival_createdata.Rmd
Clean up all code chunks

--- 2022-03-03 code/translocation_survival_createdata.Rmd
@@
 author: "Roland Knapp"
+date: "`r Sys.Date()`"
 output: html_notebook
@@
 
-The code in this notebook creates a dataset for use in an analysis to identify predictors of frog survival following translocation (translocation_survival_analysis.Rmd).
+The code in this notebook creates a dataset for use in analyses to identify predictors of frog survival following translocation (using translocation_survival_analysis.Rmd).
 The code is written in distinct chunks to allow review of the output from one chunk before running the next.
-Users without credentials to access the database will not be able to create the raw data files, but can use the saved files to run the remainder of the code.
+Users without credentials to access the database will not be able to create several of the raw data files, but can use the saved raw data files to run the remainder of the code.
 Frog translocation and survival data were generated using code in the [SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repository.
 
-### Load packages
+## Load packages
 
-```{r load_packages}
+```{r load-packages}
 library(RPostgreSQL)
@@
 
-```{r db_connect1}
+```{r db-connect1}
 con = dbConnect(dbDriver("PostgreSQL"),
@@
 
-### Retrieve site_ids and site characteristics from database
+### Retrieve site-level data from database
 
-```{sql retrieve_site_data, connection=con, output.var="translocation_sites"}
+```{sql retrieve-site-data, connection=con, output.var="translocation_sites"}
 select
@@
 
-```{r db_disconnect1}
+```{r db-disconnect1}
 dbDisconnect(con)
@@
 
-```{r sites_add_shore}
+```{r sites-add-shore}
 translocation_sites <- translocation_sites %>% 
@@
 
-```{r save_site_data}
+```{r save-site-data}
 translocation_sites %>% write_csv(here::here("data", "clean", "translocation_sites.csv"))
-# translocation_sites <- read_csv(here::here("data", "translocation_sites.csv"))
+# translocation_sites <- read_csv(here::here("data", "clean", "translocation_sites.csv"))
 ```
@@
 
-```{r retrieve_translocated_frogs_data }
+The files accessed by this code chunk are generated by code in the xxxxx_createdata_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo.
+To avoid copying those files to this project and duplicating files as a result, this code chunk accesses those files directly.
+As such, the file path used will likely need to be changed to reflect local directory structure.
+
+```{r retrieve-translocated-frogs-data }
 files <- fs::dir_ls("../../cmr-analysis/data/clean", glob = "*translocation.csv")
 frog_translocation <- read_csv(files, id = "path") %>% 
-  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  
+  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts numeric string from file path
   filter(release_date < "2021-01-01") %>% 
@@
 
-```{r save_raw_translocated_frogs_datra}
+```{r save-raw-translocated-frogs-data}
 frog_translocation %>% write_csv(here::here("data", "raw", "frog_translocation_raw.csv"))
+# frog_translocation <- read_csv(here::here("data", "raw", "frog_translocation_raw.csv"))
 ```
@@
 
-```{r db_connect2}
+```{r db-connect2}
 con = dbConnect(dbDriver("PostgreSQL"),
@@
 
-```{sql retrieve_translocation_data, connection=con, output.var="frog_characteristics"}
+```{sql retrieve-translocation-data, connection=con, output.var="frog_characteristics"}
 select
@@
 
-```{r db_disconnect2}
+```{r db-disconnect2}
 dbDisconnect(con)
@@
 
-```{r save_frog_characteristics_data}
+```{r save-frog-characteristics-data}
 frog_characteristics %>% write_csv(here::here("data", "raw", "frog_characteristics_raw.csv"))
-# translocation_sites <- read_csv(here::here("data", "translocation_sites.csv"))
+# frog_characteristics <- read_csv(here::here("data", "raw", "frog_characteristics_raw.csv"))
 ```
@@
 
-### Combine survival files from cmr-analysis repo
+### Combine frog survival files from cmr-analysis repo
 
-```{r retrieve_frog_survival_data}
+The files accessed by this code chunk are generated by code in the xxxxx_analysis_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo.
+To avoid copying those files to this project and duplicating files as a result, this code chunk accesses those files directly.
+As such, the file path used will likely need to be changed to reflect local directory structure.
+
+```{r retrieve-frog-survival-data}
 files <- fs::dir_ls("../../cmr-analysis/out/tables", glob = "*ind.csv")
 frog_survival <- read_csv(files, id = "path") %>% 
-  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  
+  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts numeric string from file path
   rename(years_since_intro = years_since_introduction,
@@
 
-```{r save_raw_frog_survival_data}
+```{r save-raw-frog-survival-data}
 frog_survival %>% write_csv(here::here("data", "raw", "frog_survival_raw.csv"))
+# frog_survival <- read_csv(here::here("data", "raw", "frog_survival_raw.csv"))
 ```
 
-### Data checks - first pass
+### Data checks #1
 
-```{r data_checking}
-# Check for duplicate pit_tag_refs
+Check for duplicate pit_tag_refs that could affect table joins.
+
+```{r data-check1}
 frog_translocation %>% count(pit_tag_ref) %>% filter(n > 1)
@@
 
-### Clean data - first pass
+### Clean data #1
 
-```{r data_cleaning_step1}
-# Duplicate pit_tag_ref found in frog_translocation, frog_characteristics, and frog survival datasets.
-# Shown as released at 70641 and 74976, but subsequently only recaptured at 74976 so that record is incorrect.
-# Recode as pit_tag_ref not found in database
+The same duplicate pit_tag_ref was found in frog_translocation, frog_characteristics, and frog survival datasets.
+That tag was recorded from frogs released at 70641 and 74976, but tag was subsequently only recaptured at 74976: tag associated with 70641 is incorrect.
+Recode erroneous tag as a pit_tag_ref not present in database to avoid errors during table joins.
+
+```{r data-clean1}
 frog_translocation <- frog_translocation %>% mutate(
@@
 
-### Add frog and site characteristics to frog translocation dataset
-```{r join_frog_data}
+### Add frog, translocation, and site characteristics to frog translocation dataset
+
+```{r join-frog-data}
 frog_characteristics <- frog_characteristics %>% 
@@
   select(-relocate_id, -release_siteid1, -release_date, -type)
-  as_tibble()
 
@@
   left_join(translocation_sites, by = "site_id")
+
+# Add translocation order
+frog_translocation <- frog_translocation %>%  
+  distinct(site_id, release_date) %>% 
+  group_by(site_id) %>% 
+  mutate(order = seq_len(n())) %>% 
+  ungroup() %>% 
+  inner_join(frog_translocation, by = c("site_id", "release_date"))
 ```
 
-### Check data - second pass
-```{r}
-# Check for null values
+### Check data #2
+
+Check for null values and hat site_ids match expection
+
+```{r data-check2}
 frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
 
-# Check that site_ids match expectation
-frog_translocation %>% distinct(site_id)
 frog_translocation %>% distinct(site_id, collect_siteid)
-
-# Check if sex at collection matches sex on recapture
-
-# Plot length x weight to check for outliers
-
 ```
 
-### Clean data - second pass
+### Clean data #2
 
-```{r}
-# Missing frog characteristics associated with pit_tag_ref = 900067000114499. 
-# Replace null values (except bd_load) with values from capture at release site early the next summer. 
-# Bd_load not replaced because load could have changed between release and recapture.
+Missing frog characteristics associated with pit_tag_ref = 900067000114499.
+Replace null values (except bd_load) with values from capture at release site early the next summer.
+Bd_load not replaced because load could have changed during the \>9 months between release and recapture.
 
-# Missing frog characteristics associated with pit_tag_ref = 900118001027669 
-# Replace null values with values from capture at release site 1-2 weeks later. 
-# Bd_load replaced because load likely would not have changed much between release and recapture.
+Missing frog characteristics associated with pit_tag_ref = 900118001027669.
+Replace null values with values from capture at release site 1-2 weeks later.
+Bd_load replaced because load likely would not have changed much in the short time period between release and recapture.
 
+Collection (= donor) site_id 72974 recorded as a distinct site despite being closely associated with 70459.
+Therefore, merged 72974 into the larger site.
+
+```{r data-clean2}
 frog_translocation <- frog_translocation %>% 
@@
 
-# Collect_siteid 72974 merged with adjacent larger 70459
 frog_translocation <- frog_translocation %>% 
@@
 
+### Check data #3
+
+#### Check if sex is correct.
+
+Sex was recorded for frogs immediately prior to their translocation and should match the sex recorded on recapture at the recipient site.
+Mismatches are often due to frog sex being mis-identified at the donor site due to typically small sizes of frogs at some of the donor sites.
+
+```{r db-connect3}
+#### Connect to database
+con = dbConnect(dbDriver("PostgreSQL"),
+                host = Sys.getenv("host"), 
+                port = Sys.getenv("port"), 
+                dbname = Sys.getenv("dbname"),
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password")
+               )
+```
+
+```{sql retrieve_capture_data, connection=con, output.var="frog_captures"}
+select
+	site_id,
+	visit_date,
+	survey_type,
+	capture_life_stage,
+	capture_animal_state,
+	pit_tag_ref, 
+	sex 
+from visit
+inner join survey on visit.id = visit_id
+inner join capture_survey on survey.id = survey_id
+inner join surveyor on surveyor.id = surveyor_id
+where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
+	survey_type = 'cmr' and
+	capture_animal_state != 'dead' 
+```
+
+```{r db-disconnect3}
+dbDisconnect(con)
+rm(con)
+```
+This code chunk indicates that there were likely no errors in sex recorded at recipient site when frogs were recaptured. 
+Therefore, it is a useful check to compare sex recorded at translocation to sex recorded at recipient site. 
+Three mismatches were detected. 
+```{r check-sex}
+frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1) # make sure that only one sex recorded per tag on recapture
+frog_captures %>% distinct(pit_tag_ref, sex) %>% 
+  rename(sexcapture = sex) %>% 
+  inner_join(frog_translocation, by = "pit_tag_ref") %>% 
+  filter(sex != sexcapture)
+```
+
+#### Check for possible errors in frog length and weight
+
+Plot frog length x weight to check for outliers - no extreme residuals observed
+
+```{r check-length-weight}
+frog_translocation %>% ggplot(aes(x = length, y = weight)) +
+  geom_point() +
+  geom_smooth(method = "lm")
+
+frog_translocation %>% ggplot(aes(x = log10(length), y = log10(weight))) +
+  geom_point() +
+  geom_smooth(method = "lm")
+```
+
+### Clean data #3
+
+Change sex at time of translocation based on subsequent recaptures at recipient site.
+
+```{r update-sex}
+frog_translocation <- frog_translocation %>% 
+  mutate(sex = replace(sex, pit_tag_ref == "900067000117464" & sex == "f", "m"),
+         sex = replace(sex, pit_tag_ref == "982000365440205" & sex == "f", "m"),
+         sex = replace(sex, pit_tag_ref == "900067000114152" & sex == "m", "f"))
+```
+
+### Create day-of-year and year columns from date
+
+```{r create-day-year}
+frog_translocation <- frog_translocation %>% 
+  mutate(day = yday(release_date),
+         year = year(release_date))
+```
+
+### Final data check
+
+#### Check for null values
+
+Results as expected.
+
+```{r check-null}
+frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+### Create final dataset
+
+```{r final-translocation-dataset}
+frog_translocation <- frog_translocation %>% 
+  rename(survival = median_survival,
+         donor = collect_siteid,
+         date = release_date) %>% 
+  select(site_id, elevation, shore, date, day, year, order, pit_tag_ref, donor, sex, length, weight, bd_load, survival) 
+```
+
 ### Save clean frog translocation dataset
 
-```{r save_clean_frog_translocation_data}
-frog_translocation %>% write_csv(here::here("data", "clean", "frog_translocation.csv"))
+```{r save_clean_translocation_data}
+frog_translocation %>% 
+   write_csv(here::here("data", "clean", "frog_translocation.csv"))
 ```

### 2022-03-05 doc/notebook/notebook.Rmd
Minor edit to code for length:weight plot

@@
   geom_point() +
-  geom_smooth(method = lm)
+  geom_smooth(method = "lm")
 plot

### 2022-03-06 code/translocation_survival_createdata.Rmd
Edit code to access cmr-analysis files locally

@@
 
-The files accessed by this code chunk are generated by code in the xxxxx_createdata_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo.
-To avoid copying those files to this project and duplicating files as a result, this code chunk accesses those files directly.
-As such, the file path used will likely need to be changed to reflect local directory structure.
+The files accessed by this code chunk are generated by code in the xxxxx_createdata_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021).
 
 ```{r retrieve-translocated-frogs-data }
-files <- fs::dir_ls("../../cmr-analysis/data/clean", glob = "*translocation.csv")
+files <- fs::dir_ls("../data/raw/cmr-analysis/translocation")  
 frog_translocation <- read_csv(files, id = "path") %>% 
@@
 
-The files accessed by this code chunk are generated by code in the xxxxx_analysis_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo.
-To avoid copying those files to this project and duplicating files as a result, this code chunk accesses those files directly.
-As such, the file path used will likely need to be changed to reflect local directory structure.
+The files accessed by this code chunk are generated by code in the xxxxx_analysis_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021).
 
 ```{r retrieve-frog-survival-data}
-files <- fs::dir_ls("../../cmr-analysis/out/tables", glob = "*ind.csv")
+files <- fs::dir_ls("../data/raw/cmr-analysis/survival")
 frog_survival <- read_csv(files, id = "path") %>% 

### 2022-03-06 doc/notebook/notebook_translocation.Rmd
Change notebook format to html_document

--- 2022-03-04 doc/notebook/notebook.Rmd
@@
 ---
-title: "Notebook"
+title: "Translocation notebook"
 author: "Roland Knapp"
@@
 output:
-  github_document
+  html_document
 ---
+
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, results='hide')
+```
 
@@
 
-```{r, fig.width = 6, fig.align = "center"}
+```{r frog-length-weight-plot}
 read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 

### 2022-03-07 doc/notebook/notebook_translocation.Rmd
Edits related to switch to github_document format

@@
 ---
-title: "Translocation notebook"
+title: "Notebook: Predictors of frog survival following translocation"
 author: "Roland Knapp"
@@
 output:
-  html_document
+  github_document: 
+    toc: true
 ---
@@
 
-## Predictors of frog survival following translocation
-
-### Dataset structure
+## Dataset structure
 
@@
 
-### Model structure
+## Model structure
 
@@
 -   Survival: Create 0/1 variable from probabilities. Change to factor?
+
+## Results
+
+To be added soon.

### 2022-03-09 doc/manuscript/pnas_demo/pnas_demo.Rmd
Update manuscript demo files

@@
   - code: c
-    address: Department of Forestry, Wildlife, and Fisheries, University of Tennessee, Knoxville, TN, xxxxx
+    address: Department of Forestry, Wildlife, and Fisheries, University of Tennessee, Knoxville, TN 37996
   - code: d
-    address: Department of Environmental Science, Policy, and Management, University of California-Berkeley, Berkeley, CA, 94720
+    address: Department of Environmental Science, Policy, and Management, University of California-Berkeley, Berkeley, CA 94720
   - code: e
-    address: Museum of Vertebrate Zoology, University of California-Berkeley, Berkeley, CA, 94720
+    address: Museum of Vertebrate Zoology, University of California-Berkeley, Berkeley, CA 94720
   - code: f
-    address: Earth Lab, University of Colorado, Boulder, CO, 80303
+    address: Earth Lab, University of Colorado, Boulder, CO 80303
 corresponding_author:
@@
 abstract: |
-  Please provide an abstract of no more than 250 words in a single paragraph.
-Abstracts should explain to the general reader the major contributions of the article.
-References in the abstract must be cited in full within the abstract itself and cited in the text.
+  <!-- <=250 words in a single paragraph--> Pharetra pharetra massa massa ultricies mi quis hendrerit dolor magna eget est lorem ipsum dolor sit amet consectetur adipiscing elit pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas integer eget aliquet nibh praesent tristique magna sit amet purus gravida quis blandit turpis cursus in hac habitasse platea dictumst quisque sagittis purus sit amet volutpat consequat mauris nunc congue nisi vitae suscipit tellus mauris a diam maecenas sed enim ut sem viverra aliquet eget sit amet tellus cras adipiscing enim eu turpis egestas pretium aenean pharetra magna ac placerat vestibulum lectus mauris ultrices eros in cursus turpis massa tincidunt dui ut ornare lectus sit amet est placerat in egestas erat imperdiet sed euismod nisi porta lorem mollis aliquam ut porttitor leo a diam sollicitudin tempor id eu nisl nunc mi ipsum faucibus vitae aliquet nec ullamcorper sit amet risus nullam eget felis eget nunc lobortis mattis aliquam faucibus purus in massa tempor nec feugiat nisl pretium fusce id velit ut tortor pretium viverra suspendisse potenti nullam ac tortor vitae purus faucibus ornare suspendisse sed nisi lacus sed viverra tellus in hac habitasse platea dictumst vestibulum rhoncus est pellentesque elit ullamcorper dignissim cras tincidunt lobortis feugiat vivamus at augue eget arcu dictum varius duis at consectetur lorem donec massa sapien faucibus et molestie ac feugiat sed lectus vestibulum mattis ullamcorper velit sed ullamcorper morbi tincidunt ornare massa eget egestas purus viverra accumsan in nisl nisi scelerisque eu ultrices vitae auctor eu augue ut lectus arcu bibendum at.
 significance: |
-  Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.
-The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
-The Significance Statement appears in the paper itself and is required for all research papers.
+  <!-- <=120 words --> Enim blandit volutpat maecenas volutpat blandit aliquam etiam erat velit scelerisque in dictum non consectetur a erat nam at lectus urna duis convallis convallis tellus id interdum velit laoreet id donec ultrices tincidunt arcu non sodales neque sodales ut etiam sit amet nisl purus in mollis nunc sed id semper risus in hendrerit gravida rutrum quisque non tellus orci ac auctor augue mauris augue neque gravida in fermentum et sollicitudin ac orci phasellus egestas tellus rutrum tellus pellentesque eu tincidunt tortor aliquam nulla facilisi cras fermentum odio eu feugiat pretium nibh ipsum consequat nisl vel pretium lectus quam id leo in vitae turpis massa sed elementum tempus egestas sed sed risus pretium quam vulputate dignissim suspendisse in est ante in.
 acknowledgements: |
-  Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.
+  Funding for this research was provided by the National Science Foundation, Yosemite National Park, Yosemite Conservancy, and U.S.
+Geological Survey NRPP program.
+Erica Hegeman and A.
+Lindauer were instrumental in assisting with field work, logistics, and permitting.
+Many summer field assistants also contributed importantly.
 keywords:
@@
   markdown: 
-    wrap: 72
+    wrap: sentence
 ---
 
-This PNAS journal template is provided to help you write your work in
-the correct journal format. Instructions for use are provided below.
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, results='hide')
+```
 
-Note: please start your introduction without including the word
-"Introduction" as a section heading (except for math articles in the
-Physical Sciences section); this heading is implied in the first
-paragraphs.
+```{r load-packages}
+library(tidyverse)
+```
+<!-- text formatting & citations -->
+The amphibian disease chytridiomycosis, caused by the chytrid fungus
+*Batrachochytrium dendrobatidis* ("Bd"), is one of the most destructive
+wildlife diseases in recorded history. This recently-emerged fungus
+[@ohanlon2018recent] is highly pathogenic to a wide range of amphibian
+taxa and, by one estimate, has caused the severe decline or extinction
+of at least 500 amphibian species [@scheele2019amphibian], with many
+more predicted to be at risk [@rodder2009global]. The pathogenesis of
+chytridiomycosis is the result of Bd infection disrupting cutaneous
+osmoregulatory function, leading to electrolyte imbalance and death
+[@voyles2009pathogenesis]. Bd infection is transferred between hosts via
+an aquatic flagellated zoospore stage [@longcore1999batrachochytrium]
+pulvinar neque laoreet suspendisse interdum consectetur libero id
+faucibus nisl tincidunt eget nullam non nisi est sit amet facilisis
+magna etiam tempor orci eu lobortis elementum nibh tellus molestie nunc
+non blandit massa enim nec dui nunc mattis enim ut tellus elementum
+sagittis vitae et leo duis ut diam quam nulla porttitor massa id neque
+aliquam vestibulum morbi blandit cursus risus at ultrices mi tempus
+imperdiet nulla malesuada pellentesque elit eget gravida cum sociis
+natoque penatibus et magnis dis parturient montes nascetur ridiculus mus
+mauris vitae ultricies leo integer malesuada nunc vel risus commodo
+viverra maecenas accumsan lacus vel facilisis volutpat est velit egestas
+dui id ornare arcu odio ut sem nulla pharetra diam sit amet nisl
+suscipit adipiscing bibendum est ultricies integer quis auctor elit sed
+vulputate mi sit.
+
+amet mauris commodo quis imperdiet massa tincidunt nunc pulvinar sapien
+et ligula ullamcorper malesuada proin libero nunc consequat interdum
+varius sit amet mattis vulputate enim nulla aliquet porttitor lacus
+luctus accumsan tortor posuere ac ut consequat semper viverra nam libero
+justo laoreet sit amet cursus sit amet dictum sit amet justo donec enim
+diam vulputate ut pharetra sit amet aliquam id diam maecenas ultricies
+mi eget mauris pharetra et ultrices neque ornare aenean euismod
+elementum nisi quis eleifend quam adipiscing vitae proin sagittis nisl
+rhoncus mattis rhoncus urna neque viverra justo nec ultrices dui sapien
+eget mi proin sed libero enim sed faucibus turpis in eu mi bibendum
+neque egestas congue quisque egestas diam in arcu cursus euismod quis
+viverra. <!-- Inline code & embedded plots -->
 
 # Results {.unnumbered}
+
+```{r inline-values, include=FALSE}
+frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 
+  mutate(site_id = as.character(site_id)) 
+count_translocation <- frog_translocation %>% distinct(site_id, date) %>% count()
+count_site <- frog_translocation %>% distinct(site_id) %>% count()
+min_elev <- min(frog_translocation$elevation)
+max_elev <- max(frog_translocation$elevation)
+bdload_groupmed <- 
+  frog_translocation %>% group_by(site_id, date) %>% summarize(median_bdload = median(bd_load, na.rm = TRUE)) 
+bd_min <- min(bdload_groupmed$median_bdload) %>% round()
+bd_max <- max(bdload_groupmed$median_bdload) %>% round() %>% format(scientific = FALSE)
+```
+
+```{r lw-plot, include=FALSE, dev="png", fig.width=3.3, fig.height=3.3, dpi=300, fig.show="hide"}
+lw_plot <- frog_translocation %>% ggplot(aes(x = length, y = weight)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  scale_x_log10() +
+  scale_y_log10() +
+  labs(x = "Length (mm)", 
+       y = "Weight (g)") +
+  theme_classic()       
+plot(lw_plot)
+
+```
+
+During the 2006--2020 period covered by this study, we conducted
+`r count_translocation$n` translocations to `r count_site$n` sites.
+Sites ranged in elevation from `r min(frog_translocation$elevation)` to
+`r max(frog_translocation$elevation)` m, and median Bd loads of
+translocated cohorts ranged from `r bd_min` to `r bd_max` ITS copies per
+skin swab. pulvinar neque laoreet suspendisse interdum consectetur
+libero id faucibus nisl tincidunt eget nullam non nisi est sit amet
+facilisis magna etiam tempor orci eu lobortis elementum nibh tellus
+molestie nunc non blandit massa enim nec dui nunc mattis enim ut tellus
+elementum sagittis vitae et leo duis ut diam quam nulla porttitor massa
+id neque aliquam vestibulum morbi blandit cursus risus at ultrices mi
+tempus imperdiet nulla malesuada pellentesque elit eget gravida cum
+sociis natoque penatibus et magnis dis parturient montes nascetur
+ridiculus mus mauris vitae ultricies leo integer malesuada nunc vel
+risus commodo viverra maecenas accumsan lacus vel facilisis volutpat est
+velit egestas dui id ornare arcu odio ut sem nulla pharetra diam sit
+amet nisl suscipit adipiscing bibendum est ultricies integer quis auctor
+elit sed vulputate mi sit.
+
+amet mauris commodo quis imperdiet massa tincidunt nunc pulvinar sapien
+et ligula ullamcorper malesuada proin libero nunc consequat interdum
+varius sit amet mattis vulputate enim nulla aliquet porttitor lacus
+luctus accumsan tortor posuere ac ut consequat semper viverra nam libero
+justo laoreet sit amet cursus sit amet dictum sit amet justo donec enim
+diam vulputate ut pharetra sit amet aliquam id diam maecenas ultricies
+mi eget mauris pharetra et ultrices neque ornare aenean euismod
+elementum nisi quis eleifend quam adipiscing vitae proin sagittis nisl
+rhoncus mattis rhoncus urna neque viverra justo nec ultrices dui sapien
+eget mi proin sed libero enim sed faucibus turpis in eu mi bibendum
+neque egestas congue quisque egestas diam in arcu cursus euismod quis
+viverra. Following log~10~-transformation, frog weight was linearly
+related to frog length across most of the range of frog lengths.
+However, for lengths \> 56 mm nearly all residuals from the
+length:weight regression line were mostly positive.
+
+![](`r knitr::fig_chunk('lw-plot', 'png')`)
+
+Figure 1. For translocated mountain yellow-legged frogs, the relationship between length and weight (on a log~10~-log~10~
+scale).
 
@@
 
-# Materials and Methods {.unnumbered}
+# Materials and Methods {#matmethods .unnumbered}
 
-## Statistical Analyses {#methods-subheading1 .unnumbered}
-
-## Data Availability {#methods-subheading2 .unnumbered}
-
-# Guide to using this template {.unnumbered}
-
-Please note that whilst this template provides a preview of the typeset
-manuscript for submission, to help in this preparation, it will not
-necessarily be the final publication layout. For more detailed
-information please see the [PNAS Information for
-Authors](http://www.pnas.org/site/authors/format.xhtml).
-
-## Author Affiliations {#author-affiliations .unnumbered}
-
-Include department, institution, and complete address, with the
-ZIP/postal code, for each author. Use lower case letters to match
-authors with institutions, as shown in the example. Authors with an
-ORCID ID may supply this information at submission.
-
-## Submitting Manuscripts {#submitting-manuscripts .unnumbered}
-
-All authors must submit their articles at
-[PNAScentral](http://www.pnascentral.org/cgi-bin/main.plex). If you are
-using Overleaf to write your article, you can use the "Submit to PNAS"
-option in the top bar of the editor window.
-
-## Format {#format .unnumbered}
-
-Many authors find it useful to organize their manuscripts with the
-following order of sections; Title, Author Affiliation, Keywords,
-Abstract, Significance Statement, Results, Discussion, Materials and
-methods, Acknowledgments, and References. Other orders and headings are
-permitted.
-
-## Manuscript Length {#manuscript-length .unnumbered}
-
-PNAS generally uses a two-column format averaging 67 characters,
-including spaces, per line. The maximum length of a Direct Submission
-research article is six pages and a PNAS PLUS research article is ten
-pages including all text, spaces, and the number of characters displaced
-by figures, tables, and equations. When submitting tables, figures,
-and/or equations in addition to text, keep the text for your manuscript
-under 39,000 characters (including spaces) for Direct Submissions and
-72,000 characters (including spaces) for PNAS PLUS.
-
-## References {#references .unnumbered}
-
-References should be cited in numerical order as they appear in text;
-this will be done automatically via bibtex, e.g. @belkin2002using and
-[@berard1994embedding; @coifman2005geometric]. All references, including
-for the SI, should be included in the main manuscript file. References
-appearing in both sections should not be duplicated. SI references
-included in tables should be included with the main reference section.
-
-## Data Archival {#data-archival .unnumbered}
-
-PNAS must be able to archive the data essential to a published article.
-Where such archiving is not possible, deposition of data in public
-databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata,
-and others outlined in the Information for Authors, is acceptable.
-
-## Language-Editing Services {#language-editing-services .unnumbered}
-
-Prior to submission, authors who believe their manuscripts would benefit
-from professional editing are encouraged to use a language-editing
-service (see list at www.pnas.org/site/authors/language-editing.xhtml).
-PNAS does not take responsibility for or endorse these services, and
-their use has no bearing on acceptance of a manuscript for publication.
-
-![Placeholder image of a frog with a long example caption to show
-justification setting.](frog.png)
-
-## Digital Figures {#sec:figures .unnumbered}
-
-Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for
-figures that will appear in the main text, and images must be final
-size. Authors may submit U3D or PRC files for 3D images; these must be
-accompanied by 2D representations in TIFF, EPS, or high-resolution PDF
-format. Color images must be in RGB (red, green, blue) mode. Include the
-font files for any text.
-
-Figures and Tables should be labelled and referenced in the standard way
-using the `\label{}` and `\ref{}` commands.
-
-Figure $$fig:frog$$ shows an example of how to insert a column-wide
-figure. To insert a figure wider than one column, please use the
-`\begin{figure*}...\end{figure*}` environment. Figures wider than one
-column should be sized to 11.4 cm or 17.8 cm wide.
-
-## Single column equations {#single-column-equations .unnumbered}
-
-Authors may use 1- or 2-column equations in their article, according to
-their preference.
-
-To allow an equation to span both columns, options are to use the
-`\begin{figure*}...\end{figure*}` environment mentioned above for
-figures, or to use the `\begin{widetext}...\end{widetext}` environment
-as shown in equation $$eqn:example$$ below.
-
-Please note that this option may run into problems with floats and
-footnotes, as mentioned in the [cuted package
-documentation](http://texdoc.net/pkg/cuted). In the case of problems
-with footnotes, it may be possible to correct the situation using
-commands `\footnotemark` and `\footnotetext`.
-
-$$\begin{aligned}
-(x+y)^3&=(x+y)(x+y)^2\\
-       &=(x+y)(x^2+2xy+y^2) \label{eqn:example} \\
-       &=x^3+3x^2y+3xy^3+x^3. 
-\end{aligned}$$
-
-```{=html}
-<!-- pandoc writes all tables using longtable, which fails in 2-column mode
-
-  Species                    CBS     CV     G3
-  ----------------------- ------ ------ ------
-  1\. Acetaldehyde           0.0    0.0    0.0
-  2\. Vinyl alcohol          9.1    9.6   13.5
-  3\. Hydroxyethylidene     50.8   51.2   54.0
-
-  : Comparison of the fitted potential energy surfaces and ab initio
-  benchmark electronic energy calculations
-
--->
-```
-## Supporting Information (SI) {#supporting-information-si .unnumbered}
-
-The main text of the paper must stand on its own without the SI. Refer
-to SI in the manuscript at an appropriate point in the text. Number
-supporting figures and tables starting with S1, S2, etc. Authors are
-limited to no more than 10 SI files, not including movie files. Authors
-who place detailed materials and methods in SI must provide sufficient
-detail in the main text methods to enable a reader to follow the logic
-of the procedures and results and also must reference the online
-methods. If a paper is fundamentally a study of a new method or
-technique, then the methods must be described completely in the main
-text. Because PNAS edits SI and composes it into a single PDF, authors
-must provide the following file formats only.
-
-### SI Text {#si-text .unnumbered}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a
-PDF with the same file name for visual reference).
-
-### SI Figures {#si-figures .unnumbered}
-
-Provide a brief legend for each supporting figure after the supporting
-text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or
-GIF format; figures may not be embedded in manuscript text. When saving
-TIFF files, use only LZW compression; do not use JPEG compression. Do
-not save figure numbers, legends, or author names as part of the image.
-Composite figures must be pre-assembled.
-
-### 3D Figures {#d-figures .unnumbered}
-
-Supply a composable U3D or PRC file so that it may be edited and
-composed. Authors may submit a PDF file but please note it will be
-published in raw format and will not be edited or composed.
-
-### SI Tables {#si-tables .unnumbered}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a
-PDF with the same file name for visual reference); include only one
-table per file. Do not use tabs or spaces to separate columns in Word
-tables.
-
-### SI Datasets {#si-datasets .unnumbered}
-
-Supply Excel (.xls), RTF, or PDF files. This file type will be published
-in raw format and will not be edited or composed.
-
-### SI Movies {#si-movies .unnumbered}
-
-Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media
-(wmv), animated GIF (gif), or MPEG files and submit a brief legend for
-each movie in a Word or RTF file. All movies should be submitted at the
-desired reproduction size and length. Movies should be no more than 10
-MB in size.
-
-### Still images {#still-images .unnumbered}
-
-Authors must provide a still image from each video file. Supply TIFF,
-EPS, high-resolution PDF, JPEG, or GIF files.
-
-### Appendices {#appendices .unnumbered}
-
-PNAS prefers that authors submit individual source files to ensure
-readability. If this is not possible, supply a single PDF file that
-contains all of the SI associated with the paper. This file type will be
-published in raw format and will not be edited or composed.
-
+amet mauris commodo quis imperdiet massa tincidunt nunc pulvinar sapien
+et ligula ullamcorper malesuada proin libero nunc consequat interdum
+varius sit amet mattis vulputate enim nulla aliquet porttitor lacus
+luctus accumsan tortor posuere ac ut consequat semper viverra nam libero
+justo laoreet sit amet cursus sit amet dictum sit amet justo donec enim
+diam vulputate ut pharetra sit amet aliquam id diam maecenas ultricies
+mi eget mauris pharetra et ultrices neque ornare aenean euismod
+elementum nisi quis eleifend quam adipiscing vitae proin sagittis nisl
+rhoncus mattis rhoncus urna neque viverra justo nec ultrices dui sapien
+eget mi proin sed libero enim sed faucibus turpis in eu mi bibendum
+neque egestas congue quisque egestas diam in arcu cursus euismod quis
+viverra. Following log\~10-transformation, frog weight was linearly
+related to frog length across most of the range of frog lengths.
+However, for lengths \> 56 mm nearly all residuals from the
+length:weight regression line were mostly. positive.
 <!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
@@
 ```
+\bigskip

### 2022-03-12 code/translocation_survival_createdata.Rmd
Replace code to connect to database with scripts

@@
 
-## Create site characteristics dataset
+## Create raw sites, frog characteristics, and frog captures data files
 
 ### Connect to database
-
-```{r db-connect1}
-con = dbConnect(dbDriver("PostgreSQL"),
-                host = Sys.getenv("host"), 
-                port = Sys.getenv("port"), 
-                dbname = Sys.getenv("dbname"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password")
-               )
+```{r db-connect}
+source("db_connect.R")
 ```
 
-### Retrieve site-level data from database
+### Create site characteristics dataset
+
+#### Retrieve site-level data from database
 
@@
 
-### Disconnect from database
-
-```{r db-disconnect1}
-dbDisconnect(con)
-rm(con)
-```
-
-### Add additional site characteristics
+#### Add additional site characteristics
 
@@
 
-### Save site characteristics dataset
+#### Save site characteristics dataset
 
 ```{r save-site-data}
-translocation_sites %>% write_csv(here::here("data", "clean", "translocation_sites.csv"))
-# translocation_sites <- read_csv(here::here("data", "clean", "translocation_sites.csv"))
+translocation_sites %>% write_csv(here::here("data", "raw", "translocation_sites.csv"))
+# translocation_sites <- read_csv(here::here("data", "raw", "translocation_sites.csv"))
 ```
 
-## Create translocated frogs dataset
-
-### Combine translocation files from cmr-analysis repo
-
-The files accessed by this code chunk are generated by code in the xxxxx_createdata_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021).
-
-```{r retrieve-translocated-frogs-data }
-files <- fs::dir_ls("../data/raw/cmr-analysis/translocation")  
-frog_translocation <- read_csv(files, id = "path") %>% 
-  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts numeric string from file path
-  filter(release_date < "2021-01-01") %>% 
-  rename(pit_tag_ref = pit_tag_id) %>% 
-  select(-path)
-rm(files)
-```
-
-### Save raw translocated frog dataset
-
-```{r save-raw-translocated-frogs-data}
-frog_translocation %>% write_csv(here::here("data", "raw", "frog_translocation_raw.csv"))
-# frog_translocation <- read_csv(here::here("data", "raw", "frog_translocation_raw.csv"))
-```
-
-### Add frog characteristics from database
-
-#### Connect to database
-
-```{r db-connect2}
-con = dbConnect(dbDriver("PostgreSQL"),
-                host = Sys.getenv("host"), 
-                port = Sys.getenv("port"), 
-                dbname = Sys.getenv("dbname"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password")
-               )
-```
+### Create frog characteristics dataset
 
@@
 
-#### Disconnect from database
+#### Save raw frog characteristics dataset
 
-```{r db-disconnect2}
-dbDisconnect(con)
-rm(con)
+```{r save-frog-characteristics-data}
+frog_characteristics %>% write_csv(here::here("data", "raw", "frog_characteristics.csv"))
+# frog_characteristics <- read_csv(here::here("data", "raw", "frog_characteristics.csv"))
 ```
 
-### Save raw frog characteristics dataset
+### Create frog captures dataset for error checking step ("check-sex")
 
-```{r save-frog-characteristics-data}
-frog_characteristics %>% write_csv(here::here("data", "raw", "frog_characteristics_raw.csv"))
-# frog_characteristics <- read_csv(here::here("data", "raw", "frog_characteristics_raw.csv"))
+#### Retrieve frog captures from database
+
+```{sql retrieve_capture_data, connection=con, output.var="frog_captures"}
+select
+	site_id,
+	visit_date,
+	survey_type,
+	capture_life_stage,
+	capture_animal_state,
+	pit_tag_ref, 
+	sex 
+from visit
+inner join survey on visit.id = visit_id
+inner join capture_survey on survey.id = survey_id
+inner join surveyor on surveyor.id = surveyor_id
+where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
+	survey_type = 'cmr' and
+	capture_animal_state != 'dead' 
+```
+
+#### Save frog captures dataset
+
+```{r save-frog-captures-data}
+frog_captures %>% write_csv(here::here("data", "raw", "frog_captures.csv"))
+# frog_captures <- read_csv(here::here("data", "raw", "frog_captures.csv"))
+```
+
+### Disconnect from database
+
+```{r db-disconnect}
+source("db_disconnect.R")
+```
+
+## Create translocated frogs dataset
+
+### Combine translocation files
+
+The files accessed by this code chunk are generated by code in the xxxxx_createdata_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021), and copies are added to the /data/raw/cmr-analysis/translocation/ directory in this (i.e., translocation) repo.
+
+```{r retrieve-translocated-frogs-data }
+files <- fs::dir_ls("../data/raw/cmr-analysis/translocation")  
+frog_translocation <- read_csv(files, id = "path") %>% 
+  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts site_id as numeric string from file path
+  filter(release_date < "2021-01-01") %>%  # only include frogs translocated before 2021
+  rename(pit_tag_ref = pit_tag_id) %>% 
+  select(-path)
+rm(files)
+```
+
+### Save raw translocated frog dataset
+
+```{r save-raw-translocated-frogs-data}
+frog_translocation %>% write_csv(here::here("data", "raw", "frog_translocation.csv"))
+# frog_translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv"))
 ```
@@
 
-### Combine frog survival files from cmr-analysis repo
+### Combine frog survival files
 
-The files accessed by this code chunk are generated by code in the xxxxx_analysis_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021).
+The files accessed by this code chunk are generated by code in the xxxxx_analysis_mrmr.Rmd files in the [cmr-analysis](https://github.com/SNARL1/cmr-analysis) GitHub repo (version 2021), and copies are added to the /data/raw/cmr-analysis/survival/ directory in this (i.e., translocation) repo.
 
@@
 frog_survival <- read_csv(files, id = "path") %>% 
-  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts numeric string from file path
+  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts site_id as numeric string from file path
   rename(years_since_intro = years_since_introduction,
@@
 ```{r save-raw-frog-survival-data}
-frog_survival %>% write_csv(here::here("data", "raw", "frog_survival_raw.csv"))
-# frog_survival <- read_csv(here::here("data", "raw", "frog_survival_raw.csv"))
+frog_survival %>% write_csv(here::here("data", "raw", "frog_survival.csv"))
+# frog_survival <- read_csv(here::here("data", "raw", "frog_survival.csv"))
 ```
@@
 ```{r data-clean1}
-frog_translocation <- frog_translocation %>% mutate(
-  pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
-frog_characteristics <- frog_characteristics %>% mutate(
-  pit_tag_ref = replace(pit_tag_ref, release_siteid1 == 70641 & pit_tag_ref == "900067000117396", "900067000117397"))
-frog_survival <-frog_survival %>% mutate(
-  pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
+frog_translocation <- frog_translocation %>% 
+  mutate(pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
+frog_characteristics <- frog_characteristics %>% 
+  mutate(pit_tag_ref = replace(pit_tag_ref, release_siteid1 == 70641 & pit_tag_ref == "900067000117396", "900067000117397"))
+frog_survival <-frog_survival %>% 
+  mutate(pit_tag_ref = replace(pit_tag_ref, site_id == "70641" & pit_tag_ref == "900067000117396", "900067000117397"))
 ```
@@
 
-Check for null values and hat site_ids match expection
+Check for null values and that donor-recipient site_ids match expectation
 
@@
 
-```{r db-connect3}
-#### Connect to database
-con = dbConnect(dbDriver("PostgreSQL"),
-                host = Sys.getenv("host"), 
-                port = Sys.getenv("port"), 
-                dbname = Sys.getenv("dbname"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password")
-               )
-```
+The result from the first line of this code chunk indicates that there were likely no errors in sex recorded at recipient site when frogs were recaptured. 
+Therefore, it is a useful check to compare sex recorded at translocation to sex recorded at recipient site (next 4 lines). 
+Three mismatches were detected.
 
-```{sql retrieve_capture_data, connection=con, output.var="frog_captures"}
-select
-	site_id,
-	visit_date,
-	survey_type,
-	capture_life_stage,
-	capture_animal_state,
-	pit_tag_ref, 
-	sex 
-from visit
-inner join survey on visit.id = visit_id
-inner join capture_survey on survey.id = survey_id
-inner join surveyor on surveyor.id = surveyor_id
-where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
-	survey_type = 'cmr' and
-	capture_animal_state != 'dead' 
-```
+```{r check-sex}
+frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1) # make sure that only one sex recorded per tag at recipient site
 
-```{r db-disconnect3}
-dbDisconnect(con)
-rm(con)
-```
-This code chunk indicates that there were likely no errors in sex recorded at recipient site when frogs were recaptured. 
-Therefore, it is a useful check to compare sex recorded at translocation to sex recorded at recipient site. 
-Three mismatches were detected. 
-```{r check-sex}
-frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1) # make sure that only one sex recorded per tag on recapture
 frog_captures %>% distinct(pit_tag_ref, sex) %>% 
@@
 
-Change sex at time of translocation based on subsequent recaptures at recipient site.
+Change sex recorded at donor site at time of translocation, based on subsequent recaptures at recipient site.
 
@@
 ```{r save_clean_translocation_data}
-frog_translocation %>% 
-   write_csv(here::here("data", "clean", "frog_translocation.csv"))
+frog_translocation %>% write_csv(here::here("data", "clean", "frog_translocation_final.csv"))
+# frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
 ```

### 2022-03-12 code/db_connect.R
Add scripts to connect to database

@@ -1,0 +1,10 @@
+# Connect to PostgreSQL database
+
+con = dbConnect(dbDriver("PostgreSQL"),
+                host = Sys.getenv("host"), 
+                port = Sys.getenv("port"), 
+                dbname = Sys.getenv("dbname"),
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password")
+                )
+

### 2022-03-12 code/db_disconnect.R

@@ -1,0 +1,4 @@
+# Disconnect from PostgreSQL database
+
+dbDisconnect(con)
+rm(con)

### 2022-03-13 code/translocation_survival_createdata.Rmd
Update nomenclature of relative paths

@@
 ```{r retrieve-translocated-frogs-data }
-files <- fs::dir_ls("../data/raw/cmr-analysis/translocation")  
+files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "translocation"))  
 frog_translocation <- read_csv(files, id = "path") %>% 
@@
 ```{r retrieve-frog-survival-data}
-files <- fs::dir_ls("../data/raw/cmr-analysis/survival")
+files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survival"), glob = "*_ind.csv")
 frog_survival <- read_csv(files, id = "path") %>% 

### 2022-03-13 code/translocation_survival_exploratory.Rmd
Add Rmd file for exploratory data analysis

---
title: "Predictors of frog survival following translocations - analyses"
author: "Roland Knapp"
output: html_notebook
---

This notebook contains exploratory analyses to aid in the design of the final analyses of factors influencing post-translocation survival (conducted in translocation_survival_analysis.Rmd).

## Load packages

```{r load-packages}
library(tidyverse)
library(lubridate)
```

## Read in frog translocation dataset

```{r read-translocation-data}
frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
  mutate(site_id = as.character(site_id),
         donor = as.character(donor),
         surv = as.integer(if_else(survival < 0.5, 0, 1)))
```

## Check for null values

```{r check-null}
frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
```
* 12 records were bd_load is NA, no other NAs in dataset

## Retrieve cohort-level frog survival data

```{r retrieve-frog-survival-data}
files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survival"), glob = "*_cohort.csv")
cohort_survival <- read_csv(files, id = "path") %>% 
  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts site_id as numeric string from file path
  rename(years_since_intro = years_since_introduction) %>% 
  filter(years_since_intro == 1) %>% 
  select(site_id, release_date, years_since_intro, median_survival)
rm(files)
```

## Plot of frog survival for each site

```{r frog-survival-by-site}
cohort_survival %>% 
  mutate(release_year = year(release_date)) %>% 
  group_by(site_id) %>% 
  summarize(plot_survival = mean(median_survival)) %>%  # create variable by which to order sites for plotting
  arrange(plot_survival) %>% 
  mutate(siteid_order = row_number()) %>% 
  inner_join(cohort_survival, by = "site_id") %>% 
  select(-plot_survival) %>% 
    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival)) +  # reorder site_id by siteid_order)
      geom_point(size = 3) +
      labs(x = "Recipient site", y = "1-year frog survival")
```
* Strong effect of site on frog survival.
* Survival is highly variable across sites, but highly repeatable within a site. 
* Indicates need to include site_id as a group-level effect.

## Relationship between bd_load at release and 1-year survival
```{r bdload-frogsurvival-plot}
# Plot distribution of bd_load
frog_translocation %>% 
  drop_na(bd_load) %>% 
  mutate(lbd_load = (log10(bd_load + 1))) %>% 
  ggplot(aes(lbd_load)) +
    geom_histogram()

# Plot frog survival as a function of bd_load
frog_translocation %>% 
  select(bd_load, surv) %>% 
  drop_na(bd_load) %>% 
  mutate(lbd_load = (log10(bd_load + 1)),
         bd_interval = cut_interval(lbd_load, n = 4)) %>% # create 4 bins from lbd_load range 
  group_by(bd_interval, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>%  # contingency table of survival proportions, sum calculated for each interval
  filter(surv == 1) %>% 
  ggplot(aes(x = bd_interval, y = propsurv)) +
    geom_bar(stat = "identity") +
    labs(x = "Bd load interval ((log10 + 1) transformed)",
         y = "Proportion that survived at least 1 year")
```
* Suggests that frogs that had the highest loads at the time of translocation had relatively low survival.

## Relationship between shore variable and 1-year survival

```{r shore-frogsurvival-plot}
frog_translocation %>% 
  mutate(shore = as.factor(shore)) %>% 
  group_by(shore, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>%  # contingency table of survival proportions
  filter(surv == 1) %>% 
  ggplot(aes(x = shore, y = propsurv)) +
    geom_bar(stat = "identity") +
    labs(x = "Shore habitat",
         y = "Proportion that survived at least 1 year")

```
* The presence of shore habitat that is accessible to frogs appears to have a strong effect on survival. 
* Is this a direct effect, or an effect mediated through a correlated variable? For example, lakes lacking shore habitat are often those whose surface elevation decreases during the summer, leaving the shore habitat dry and inaccessible. Lower elevation lakes are more likely to experience lake elevation reductions, and are also likely to contain abundant frog predators. 
* Will be important to include both shore habitat and elevation in model.

## Relationship between translocation order and 1-year survival

```{r order-frogsurvival-plot}
frog_translocation %>% 
  mutate(order = as.factor(replace(order, order > 1, 2))) %>% 
  group_by(order, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>%  # contingency table of survival proportions
  filter(surv == 1) %>% 
  ggplot(aes(x = order, y = propsurv)) +
    geom_bar(stat = "identity") +
    labs(x = "Translocation order",
         y = "Proportion that survived at least 1 year")

cohort_survival %>% 
  count(site_id) %>% 
  inner_join(cohort_survival, by = "site_id") %>% 
  arrange(site_id, release_date) %>% 
  group_by(site_id) %>% 
  mutate(order = row_number()) %>% 
  filter(n > 1) %>% 
    ggplot(aes(x = order, y = median_survival, group = site_id, color = site_id)) + 
      geom_line() +
      geom_point()
```
* Contingency table results suggest that translocation order does not affect survival of translocated cohort. However, not a particularly rigorous comparison because order is grouped within sites. 
* Line plot includes all sites that received two or more translocated cohorts. At 6 of 8 of these sites, cohorts after the first cohort (i.e., 2-4) had lower survival than the first cohort. At 2 of 8 sites, the second cohort had higher survival.
* Appears that translocation order has a weak negative effect on survival.

## Relationship between donor site and 1-year survival

```{r donor-frogsurvival-plot}
frog_translocation %>% 
  group_by(donor, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>%  # contingency table of survival proportions
  filter(surv == 1) %>% 
  ggplot(aes(x = donor, y = propsurv)) +
    geom_bar(stat = "identity") +
    labs(x = "Donor site",
         y = "Proportion that survived at least 1 year")
```
* Substantial differences between donor sites in the survival of translocated frogs.

## Does bd_load of translocated cohorts differ between donor sites?

```{r donor-bdload-plot}
recipient_donor <- frog_translocation %>% 
  distinct(site_id, date, donor)

frog_translocation %>% 
  drop_na(bd_load) %>% 
  mutate(lbdload = log10(bd_load + 1)) %>% 
  group_by(site_id, date) %>% 
  summarize(lbdload_med = median(lbdload)) %>% 
  inner_join(recipient_donor, by = c("site_id", "date")) %>% 
    ggplot(aes(x = donor, y = lbdload_med)) + 
      geom_point(size = 3) +
      labs(x = "Donor site", y = "Median Bd load (log10 + 1 transformed)")
```
* The 3 donor sites differ somewhat in bd_load of translocated cohorts. 
* Among donor sites, survival of translocated cohorts is inversely related to median bd_load of cohorts.

## Relationship between sex and 1-year survival

```{r sex-frogsurvival-plot}
frog_translocation %>% 
  group_by(sex, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>%  # contingency table of survival proportions
  filter(surv == 1) %>% 
  ggplot(aes(x = sex, y = propsurv)) +
    geom_bar(stat = "identity") +
    labs(x = "Frog sex",
         y = "Proportion that survived at least 1 year")
```
* No obvious difference in survival between females and males.

## Relationship between elevation and frog survival

```{r elevation-frogsurvival-plot}
frog_translocation %>% 
  mutate(elev_interval = cut_interval(elevation, n = 3)) %>%
  group_by(elev_interval, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>% 
  filter(surv == 1) %>% 
    ggplot(aes(x = elev_interval, y = propsurv)) +
      geom_bar(stat = "identity") +
      labs(x = "Elevation", y = "Proportion that survived at least 1 year")
```
* Only 12 elevations represented, so binned by 3 to ensure that 0s and 1s were represented in bin (not with case with 4 bins). 
* Elevation has positive effect on frog survival.

## Relationship between day and frog survival

```{r day-frogsurival-plot}
frog_translocation %>% 
  mutate(day_interval = cut_interval(day, n = 3)) %>%
  group_by(day_interval, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>% 
  filter(surv == 1) %>% 
    ggplot(aes(x = day_interval, y = propsurv)) +
      geom_bar(stat = "identity") +
      labs(x = "Day of the year", y = "Proportion that survived at least 1 year")
```
* No obvious effect of day of the year on frog survival

## Relationship between year and frog survival

```{r year-frogsurvival-plot}
frog_translocation %>% 
  mutate(year_interval = cut_interval(year, n = 3)) %>%
  group_by(year_interval, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>% 
  filter(surv == 1) %>% 
    ggplot(aes(x = year_interval, y = propsurv)) +
      geom_bar(stat = "identity") +
      labs(x = "Year", y = "Proportion that survived at least 1 year")
```

* Frog survival has increased with year, perhaps as a result of improved ability to select sites with high quality frog habitat.

## Relationship between length and frog survival

```{r length-frogsurvival-plot}
frog_translocation %>% 
  mutate(svl_interval = cut_interval(length, n = 5)) %>%
  group_by(svl_interval, surv) %>% 
  summarize(n = n()) %>% 
  mutate(propsurv = n/sum(n)) %>% 
  filter(surv == 1) %>% 
    ggplot(aes(x = svl_interval, y = propsurv)) +
      geom_bar(stat = "identity") +
      labs(x = "Frog length", y = "Proportion that survived at least 1 year")
```
* Frog size at the time of translocation has a strong positive effect on survival. 
* This effect could explain, at least in part, the differences in survival between donor sites. 72996 contains the smallest frogs and has the lowest survival of translocated frogs. In theory, the size effect could be caused by a positive relationship between frog size and bd_load. However, the plot below suggests that bd_load is unaffected by frog size. As such, effect of frog size appears independent of an effect of bd_load.

## Effect of frog size on bd_load

```{r}
frog_translocation %>% 
  drop_na(bd_load) %>% 
  mutate(lbdload = log10(bd_load + 1)) %>% 
  ggplot(aes(x = length, y = lbdload)) +
    geom_point() +
    geom_smooth(method = "lm")

```

### 2022-03-16 code/snowwc_check.R
Add scripts to check/clean snow survey data

@@ -1,0 +1,21 @@
+# Dana Meadows snow survey data
+# Data-checking script for dataset created by snowwc_clean.R
+
+library(readr)
+library(dplyr)
+
+# Create dataset
+source(here::here("code", "snowwc_clean.R"))
+
+# Check that there are no missing years
+yr <- snowwc_dan %>% pull(year)
+min <- min(yr)
+max <- max(yr)
+seq_yr <- seq(min, max, by = 1)
+identical(yr, seq_yr)
+rm(yr, min, max, seq_yr)
+
+# Check that snowwc_pave are within acceptable range and without any NAs
+range(snowwc_dan$snowwc_pave)
+snowwc_dan %>% filter(is.na(snowwc_pave)) %>% count()
+

### 2022-03-16 code/snowwc_clean.R

@@ -1,0 +1,18 @@
+# Dana Meadows snow survey data
+# Data-cleaning script
+# Data downloaded as csv from http://cdec.water.ca.gov/dynamicapp/wsSensorData
+# Query parameters: station_id = dan, sensor # = 3, start_date = 1900-01-01, end_date = 2021-12-31
+# View as csv, save as text file, open with LibreOffice Calc, save as csv. 
+
+library(tidyverse)
+
+# Read raw data and clean
+snowwc_dan <- read_csv(here::here("data", "raw", "danameadows_snowsurvey.csv")) %>% 
+  rename_with(tolower) %>% 
+  rename(date = "date time") %>% 
+  filter(str_detect(date, "04-01$")) %>% 
+  mutate(snowwc_cm = (as.numeric(value) * 2.54),
+         year = year(date),
+         station_id = tolower(station_id),
+         snowwc_pave = (snowwc_cm/mean(snowwc_cm)) * 100) %>%  # expressed as percent of average snowwc_cm as measured over entire time series
+  select(station_id, date, year, snowwc_cm, snowwc_pave)

### 2022-03-16 code/translocation_survival_createdata.Rmd
Add snow_wc variable to translocation dataset

@@
 
-```{r create-day-year}
+```{r add-day-year}
 frog_translocation <- frog_translocation %>% 
@@
          year = year(release_date))
+```
+
+### Add snowpack data for winter of and winter following translocation
+
+```{r add-snowpack}
+
+source("snowwc_clean.R")
+
+frog_translocation <- snowwc_dan %>% 
+  mutate(year_t = year, 
+         year_t1 = (year + 1)) %>% 
+  rename(snow_t = snowwc_cm) %>% 
+  select(year, year_t, year_t1, snow_t) %>% 
+  inner_join(snowwc_dan, by = c("year_t1" = "year")) %>% 
+  rename(snow_t1 = snowwc_cm) %>% 
+  select(-station_id, -date, -snowwc_pave) %>% 
+  inner_join(frog_translocation, by = "year")
 ```
@@
          date = release_date) %>% 
-  select(site_id, elevation, shore, date, day, year, order, pit_tag_ref, donor, sex, length, weight, bd_load, survival) 
+  select(site_id, elevation, shore, snow_t, snow_t1, date, year, day, order, pit_tag_ref, donor, sex, length, weight, bd_load, survival) 
 ```

### 2022-03-16 code/translocation_survival_exploratory.Rmd
Add additional exploratory analyses

@@
 ```
-* 12 records were bd_load is NA, no other NAs in dataset
+* 12 records where bd_load is NA, no other NAs in dataset
 
@@
 
-## Plot of frog survival for each site
+# define color palette
+```{r define-color-palette}
+palette_cbb <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+```
+
+## Plot of frog survival for each translocation - grouped by site
 
@@
 Lower elevation lakes are more likely to experience lake elevation reductions, and are also likely to contain abundant frog predators.
-* Will be important to include both shore habitat and elevation in model.
+* Due to potential dependence between elevation and shore variables, unclear if it will be possible to include both shore habitat and elevation in model.
+
+## Relationship between shore habitat and elevation
+
+```{r shorehabitat-elevation-plot}
+frog_translocation %>% 
+  distinct(site_id, shore, elevation) %>% 
+  mutate(shore = as_factor(shore)) %>% 
+  ggplot(aes(x = shore, y = elevation)) +
+    geom_boxplot() +
+    labs(x = "Shore habitat availability", y = "Elevation")
+```
 
@@
       geom_line() +
-      geom_point()
+      geom_point() +
+      scale_color_manual(values = palette_cbb)
 ```
@@
     ggplot(aes(x = donor, y = lbdload_med)) + 
-      geom_point(size = 3) +
+      geom_boxplot() +
       labs(x = "Donor site", y = "Median Bd load (log10 + 1 transformed)")
+
+frog_translocation %>% 
+  drop_na(bd_load) %>% 
+  mutate(lbdload = log10(bd_load + 1)) %>% 
+  group_by(site_id, date) %>% 
+  summarize(lbdload_med = median(lbdload)) %>% 
+  inner_join(recipient_donor, by = c("site_id", "date")) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = lbdload_med, y = median_survival)) + 
+      geom_point() +
+      geom_smooth(method = "lm") +
+      labs(x = "Cohort-level median Bd load (log10 + 1 transformed)", y = "Median cohort-level frog survival")
+
+frog_translocation %>% 
+  drop_na(bd_load) %>% 
+  mutate(lbdload = log10(bd_load + 1)) %>% 
+  group_by(site_id, date) %>% 
+  summarize(lbdload_med = median(lbdload)) %>% 
+  inner_join(recipient_donor, by = c("site_id", "date")) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = lbdload_med, y = median_survival, color = donor, group = donor)) + 
+      geom_point() +
+      geom_smooth(method = "lm", se = FALSE) +
+      labs(x = "Cohort-level median Bd load (log10 + 1 transformed)", y = "Median cohort-level frog survival") +
+      scale_color_manual(values = palette_cbb)
 ```
-* The 3 donor sites differ somewhat in bd_load of translocated cohorts. 
-* Among donor sites, survival of translocated cohorts is inversely related to median bd_load of cohorts.
+* Across all translocated frogs, the 3 donor sites have similar bd_loads. A plot of The 3 donor sites differ somewhat in bd_load of translocated cohorts. 
+* Across all donor sites and translocations, survival of translocated cohorts may be weakly inversely related to median bd_load of cohorts.
+However, grouping by donor site shows that each donor site has its own bd_load-frog survival relationship.
+* bd_load unrelated, or perhaps weakly related to differences in frog survival between donor sites.
 
@@
 
-## Relationship between length and frog survival
+## Relationship between frog size and survival
 
@@
 
-## Effect of frog size on bd_load
+## Relationship between frog size and bd_load
 
-```{r}
+```{r frogsize-bdload-plot}
 frog_translocation %>% 
@@
 ```
+
+## Relationship between donor site and frog length
+
+```{r site-frogsize-plot}
+frog_translocation %>% 
+  ggplot(aes(x = donor, y = length)) +
+    geom_boxplot()
+```
+* The relationship between donor site and frog length suggests that frog length is only partly responsible for the relationship between donor site and frog survival (frog survival: 70459 > 70567 > 72996).
+Perhaps some combination of frog size, bd_load, and genetic factors is associated with frog susceptibility?
+
+## Relationship between winter severity in year of translocation on frog survival
+
+```{r}
+frog_translocation %>% 
+  mutate(snow_t_int = cut_interval(snow_t, n = 3)) %>% # create 3 bins from snow_t range  
+  group_by(snow_t_int, surv) %>% 
+  summarize(n = n()) %>% 
+  mutate(propsurv = n/sum(n)) %>% 
+  filter(surv == 1) %>% 
+  ggplot(aes(x = snow_t_int, y = propsurv)) + 
+    geom_bar(stat = "identity") +
+    labs(x = "Winter severity in year of translocation", y = "Proportion that survived at least 1 year")
+
+frog_translocation %>% 
+  distinct(site_id, date, donor, snow_t) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = snow_t, y = median_survival)) + 
+      geom_point() +
+      geom_smooth(method = "lm") +
+      labs(x = "Winter severity in year of translocation", y = "Median cohort-level frog survival")
+
+frog_translocation %>% 
+  distinct(site_id, date, donor, snow_t) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = snow_t, y = median_survival, color = donor)) + 
+      geom_point() +
+      geom_smooth(method = "lm") +
+      labs(x = "Winter severity in year of translocation", y = "Median cohort-level frog survival") +
+      scale_color_manual(values = palette_cbb)
+```
+* Hypothesized mechanism: Severity of winter preceding translocation could affect frog condition and subsequent survival of translocated frogs.
+For example, frogs that just experienced a long winter would be in relatively poor condition and that could reduce survival following translocation.
+* No obvious effect of winter severity in year of translocation and frog survival.
+
+## Relationship between winter severity in year following translocation on frog survival
+
+```{r}
+frog_translocation %>% 
+  mutate(snow_t1_int = cut_interval(snow_t1, n = 3)) %>% # create 4 bins from snow_t range  
+  group_by(snow_t1_int, surv) %>% 
+  summarize(n = n()) %>% 
+  mutate(propsurv = n/sum(n)) %>% 
+  filter(surv == 1) %>% 
+  ggplot(aes(x = snow_t1_int, y = propsurv)) + 
+    geom_bar(stat = "identity") +
+    labs(x = "Winter severity in year following translocation", y = "Proportion that survived at least 1 year")
+
+frog_translocation %>% 
+  distinct(site_id, date, donor, snow_t1) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = snow_t1, y = median_survival)) + 
+      geom_point() +
+      geom_smooth(method = "lm") +
+      labs(x = "Winter severity in year following translocation", y = "Median cohort-level frog survival")
+
+frog_translocation %>% 
+  distinct(site_id, date, donor, snow_t1) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = snow_t1, y = median_survival, color = donor)) + 
+      geom_point() +
+      geom_smooth(method = "lm") +
+      labs(x = "Winter severity in year following translocation", y = "Median cohort-level frog survival") +
+      scale_color_manual(values = palette_cbb)
+```
+
+* Hypothesized mechanism: Severity of winter follow translocation could affect frog condition and subsequent survival of translocated frogs in the first year following translocation.
+For example, a long winter following translocation could reduce survival in the first year.
+* No obvious effect of winter severity in year following translocation and frog survival.

### 2022-03-16 doc/notebook/notebook_translocation.Rmd
Add summary of results from exploratory analyses

@@
 
--   site_id: site to which frogs were translocated
--   elevation
--   date
--   day
--   year
--   donor
--   order: first translocation to site = 0, subsequent translocation = 1
--   type: translocation conducted on foot versus by helicopter (not included)
--   shore: suitability of shore/bank habitat for frog overwintering
--   pit_tag_ref
--   survival: individual-level estimated survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation
--   sex
--   length: at release
--   weight: at release
--   condition: index based on length and weight
--   swab_id
--   bd_load: at release
+* site_id: site to which frogs were translocated  
+* elevation: water body surface elevation above sea level, in meters  
+* shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)  
+* snow_t: snow water content (measured on April 1) in year of translocation  
+* snow_t1: snow water content (measured on April) in year following translocation  
+* date: date frogs released at recipient site  
+* year: year frogs released at recipient site  
+* day: day of the year on which frogs were released  
+* order: order of translocations to a site, from 1-4. Recoded prior to analysis ("order_f") as 0/1: 0 = first translocation to site; 1 = subsequent translocations  
+* donor: site_id from which frogs were collected for translocation  
+* pit_tag_ref: pit tag id  
+* sex: m or f, recoded for analysis as 0 for males and 1 for females  
+* length: frog snout-vent length at release (in mm)  
+* weight: frog weight at release (in grams)  
+* bd_load: number of ITS copies at release, measured from swab swab collected on capture (as with length and weight)  
+* survival: individual-level estimated survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation.
+Prior to analysis, recoded ("surv") as 0/1 (0: survival < 0.5; 1: survival $\geq$ 0.5)
 
@@
 ```{r frog-length-weight-plot}
-read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 
-ggplot(aes(x = log10(length), y = log10(weight))) +
+read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+ggplot(aes(x = length, y = weight)) +
   geom_point() +
-  geom_smooth(method = "lm")
-plot
+  geom_smooth(method = "lm") +
+  scale_x_log10() +
+  scale_y_log10() +
+  labs(x = "Length (mm)", 
+       y = "Weight (g)")
 ```
@@
 Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival.
-39 frogs were translocated on that date, so should probably drop the 9 records where bd_load = NA prior to running the analysis.
-If bd_load is not an important predictor, can run analysis using the full dataset.
+39 frogs were translocated on that date, so plenty of swab results are available to represent that site.
+Prior to running the analyses, should drop the 12 records where bd_load = NA.
+If bd_load is not an important predictor, can run analysis (without the bd_load predictor) using the full dataset.
+
+## Exploratory analyses
+
+Exploratory data analyses conducted in translocation_survival_exploratory.Rmd provide several notable insights.
+
+### Donor site
+* 1-year frog survival is highly variable across sites, but highly repeatable within a site.
+This indicates the importance of including site_id as a group-level effect.
+It remains unclear whether also including a group effect (year) that identifies each translocation might also be needed (or feasible given relatively small number of translocations per site).
+* Donor site has a strong effect on frog survival, with 70459 > 70567 > 72996.
+Explored possible effects of bd_load and frog size as drivers of this pattern.
+Differences in bd_load between donor sites seems unrelated to survival, but appears closely associated with frog size (see sub-bullets below).
+Whether frog size is the sole driver or is a results of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) remains to be seen.
+  * Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival. 
+  * Strong effect of frog size on 1-year survival, with frogs /> 55 mm having substantially higher 1-year survival. Frogs />60 mm had by far the highest survival. 
+  * The effect of frog size on survival was unrelated to bd_load: No relationship between the two variables. 
+  * Frog size differs between donor site: 72996 < 70459 = 70567. This partially mirrors the observed differences in survival between sites.
+
+### Bd load
+* Frogs with the highest loads at the time of translocation had the lowest survival.
+The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies.
+Notable result given that the threshold level of 56,000 is /~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.
+
+### Recipient site characteristics
+* **Shore habitat**: The presence of shore habitat substantially increased frog survival.
+Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature.
+Association with shore habitat presence/absence and elevation will make it difficult to disentangle these effects.
+It may only be possible to include one of these two variables in the model.
+If so, elevation would make the most sense to include.
+* **Elevation**: Strong positive effect on frog survival.
+
+### Translocation timing
+* Day of the year showed no obvious relationship to frog survival. 
+* Year was positively associated with survival.
+This likely is not a direct effect of year, but instead reflects better-informed selection of recipient sites over time that contained high-quality frog habitat.
+As such, consider not including year in the model.
+
+### Winter severity
+* No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival.
+
+### Translocation order
+* Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect was not particularly strong. 
+* It remains unclear what the mechanism is of the slightly higher mortality of later cohorts.
+
+### Frog characteristics
+* Frog size had strong effect on 1-year survival. 
+* Minimal differences in survival between males and females (slightly higher in males).
 
 ## Model structure
+As originally conceived: survival \~ elevation + shore + snow_t + snow_t1 + year + day + order + donor + sex + length + log(bdload) \| site_id
 
-survival \~ elevation + shore + day + year + order + donor + sex + length + log(bdload) \| site_id
+Based on exploratory analyses: survival \~ elevation + snow_t + snow_t1 + day + order + donor + sex + length + log(bdload) \| site_id + year
 
-Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include release_year, perhaps nested within site_id?
-Given that some sites only have a single release_year, will this cause problems?
+Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include year of translocation?
+Given that some sites only received a single translocation, will this cause problems?
 
@@
 
--   Change to factor: order, day?, year?
--   Standardize variables?
--   Survival: Create 0/1 variable from probabilities. Change to factor?
+* Recode order as 0/1 (instead of 1--4).  
+* Recode survival as 0/1 (instead of continuous probality).  
+* Recode binary variables as 0/1 (or format as factor): sex (see Gelman reference below).  
+* Format some variables as factor: surv, shore, order, donor, sex, site_id, year.  
+* Standardize continuous variables: elevation, snow_t, snow_t1, day, length, log(bd_load).
+Gelman recommends standardizing by dividing by 2 standard deviations.
+Standardizing allows assessment of importance of predictors by comparing regressions coefficients.
+Dividing by 2 SD puts continuous variables on approximately the same scale as binary (0/1) variables.
+This is true only if binary predictors have similar number of 0s and 1s.
+Sex: female = 428, male = 351; order: 0 = 507, 1 = 272.
+
+Relevant references are as follows:
+
+* Gelman 2008: https://doi.org/10.1002/sim.3107  
+* Schielzeth 2010: https://doi.org/10.1111/j.2041-210X.2010.00012.x
 

### 2022-03-16 code/translocation_survival_exploratory.Rmd
Minor edit to add reference for color palette

@@
 
-# define color palette
+# define color palette (palette_OkabeIto_black)
 ```{r define-color-palette}

### 2022-03-19 doc/notebook/notebook_translocation.Rmd
Add notes on structure of dataset and model

@@
 Explored possible effects of bd_load and frog size as drivers of this pattern.
-Differences in bd_load between donor sites seems unrelated to survival, but appears closely associated with frog size (see sub-bullets below).
+Differences in survival between donor sites seem unrelated to bd_load, but appear closely associated with frog size (see sub-bullets below).
 Whether frog size is the sole driver or is a results of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) remains to be seen.
@@
 ## Model structure
+
 As originally conceived: survival \~ elevation + shore + snow_t + snow_t1 + year + day + order + donor + sex + length + log(bdload) \| site_id
 
-Based on exploratory analyses: survival \~ elevation + snow_t + snow_t1 + day + order + donor + sex + length + log(bdload) \| site_id + year
+Based on exploratory analyses: survival \~ elevation + snow_t + snow_t1 + day + order + donor + sex + length + log(bdload) \| site_id
 
-Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include year of translocation?
-Given that some sites only received a single translocation, will this cause problems?
+Make pairs plot to identify correlations between predictors (ggpairs from GGally package).
 
-Final steps before building model (in analysis Rmd):
+It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across sites, causing dependency between donor and length.
+As such, these two variables have the potential to compete for the same variation.
+Need to look at this more closely, perhaps by plotting frog lengths at each donor site.
+Also, plot length against survival, grouped by donor.
+Are these distributions as donor-specific as they apear to be?
+Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.
+
+Also need to disentangle elevation from site_id - currently only one elevation per site_id.
+Try breaking continuous elevation into categories, with > 2 site_ids per category.
+Hopefully, that will make elevation variable less dependent on site_id.
+
+Shore and elevation may also be closely associated. Are all sites with shore = 0 at low elevation? If so, may need to only include elevation in model.
+
+Group-level effects: Numerous levels of structure inherent in the data, including that caused by year, site_id, donor, and translocation id.
+However, most of these are not sufficiently hierarchical to allow inclusion as a group-level effect.
+For example, 70413 only received frogs collected from 70567.
+For year, only a small subset of site ids are represented in each year and the subset changes dramaticall from year to year.
+Donor and translocation id have even sparser matrices with site_id.
+As such, site_id is the only reasonable group-level effect to consider for inclusion in model.
+Related to this, general guidance is that 5 or more levels are generally necessary for a grouping variable to be considered for inclusion as a group-level effect.
+Models with fewer levels may converge but the group-level variable may have little or no effect.
+
+Interactions: General guidance is that interactions between the most important predictors should be included in model.
+
+## Final steps before building model (in analysis Rmd):
 
 * Recode order as 0/1 (instead of 1--4).  
-* Recode survival as 0/1 (instead of continuous probality).  
-* Recode binary variables as 0/1 (or format as factor): sex (see Gelman reference below).  
+* Recode survival as 0/1 (instead of continuous probability).  
+* Recode binary variables as 0/1 indicators: sex (see Gelman reference below).  
 * Format some variables as factor: surv, shore, order, donor, sex, site_id, year.  
-* Standardize continuous variables: elevation, snow_t, snow_t1, day, length, log(bd_load).
-Gelman recommends standardizing by dividing by 2 standard deviations.
+* Standardize continuous variables: elevation, snow_t, snow_t1, day, length.
+Possibly standardize log(bd_load) also, but see Gelman for recommendations.
+Gelman recommends centering by subtracting the mean and standardizing dividing by 2 standard deviations.
+Centering to a mean of 0 makes interactions more interpretable.
 Standardizing allows assessment of importance of predictors by comparing regressions coefficients.
@@
 This is true only if binary predictors have similar number of 0s and 1s.
+In the current dataset, there is some deviation from 1:1.
 Sex: female = 428, male = 351; order: 0 = 507, 1 = 272.
+Binary variables should be centered but not standardized.
+Variables in need of log-transformation should be transformed and then rescaled, although rescaling will not always be necessary.
 
@@
 
+## Priors
+
+* For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.
+https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations.
+* Other recommendations in:
+  * https://doi.org/10.1111/oik.0598.  
+  * Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution.  https://doi.org/10.1002/ecs2.3739
+
 ## Results
 
-To be added soon.
+* Figures showing estimated coefficients from models with and without site_id as group-level effect.
+
+## Next steps
+
+* Site_id group-level effect largely accounted for differences between donor sites in frog survival, frog size, etc.
+To provide another assessment of the effect of frog size on survival, create a second model that includes only donor = 72996.
+Given that there is considerable variation in frog sizes translocated from this site, this might allow an assessment of the effect of frog size without the group-level effect (because there probably aren't enough sites to which frogs from 72996 were translocated) to make site_id a useful group-level effect.
+* To assess within-site association between survival for the first and second translocated cohorts, build another dataset that includes all of those recipient sites for which > 1 translocation was conducted.
+Include survival from the first translocation as the response variable and survival from the second translocation as a predictor variable.
+Other predictors could also be included.

### 2022-05-11 code/translocation_survival_exploratory.Rmd
Add additional plots and other explorations

@@
 library(lubridate)
+library (GGally)
 ```
@@
 
-## Plot of frog survival for each translocation - grouped by site
+## Pairs plot to evaluate between-variable correlations
+```{r pairs-plots}
+# Translocation-specific variables
+frog_translocation %>% 
+  distinct(site_id, date, .keep_all = TRUE) %>% 
+  select(elevation, shore, order, snow_t, snow_t1, year, day) %>% 
+  mutate(across(c("shore", "order"), as.factor)) %>% 
+  ggpairs()
+
+# Frog-specific variables
+frog_translocation %>% 
+  mutate(bdload_l = log10(bd_load + 1),
+         across(c("donor" , "sex"), as.factor)) %>% 
+  select(donor, sex, length, weight, bdload_l) %>% 
+  ggpairs()
+```
+* Elevations for shore = 0 and shore = 1 do not overlap at all, so there is no ability to disentangle these two variables.
+Include only elevation in model.
+* snow_t and day are positively correlated, reflecting the fact that in summers following severe winters, tranlocations were conducted later in the summer.
+Correlation is relatively weak so may not cause issues in analysis.
+* Few translocations with order > 2, combine group = 2-4 into a single group. 
+* Frogs from the 72996 donor population were smaller and had somewhat higher bd loads than frogs from the other two donor populations.  
+* Frog length and weight were highly correlated, as expected.
+
+## Association between site_id grouping variable and some predictors
+* If site_id is closely associated with a predictor, will adding site_id as a group-level effect largely account for the effect of the predictor?
+```{r siteid-predictor-count}
+frog_translocation %>% 
+  distinct(site_id, date, .keep_all = TRUE) %>% 
+  distinct(site_id, elevation) %>% 
+  count(site_id)
+
+frog_translocation %>% 
+  distinct(site_id, date, .keep_all = TRUE) %>% 
+  mutate(first = if_else(order == 1, 1, 0)) %>% 
+  distinct(site_id, order) %>% 
+  count(site_id)
+
+frog_translocation %>% 
+  distinct(site_id, date, .keep_all = TRUE) %>% 
+  distinct(site_id, snow_t) %>% 
+  count(site_id)
+```
+* Only a single value of elevation per site_id (obviously), so could bin elevation (low elevation, high elevation) to weaken that association.  
+* Multiple values of winter severity and order per site_id, so those two variables are only loosely associated with site_id.
+
+```{r siteid-elevation-binned}
+frog_translocation %>% 
+  mutate(elev_b = cut_number(elevation, n = 2)) %>% 
+  distinct(site_id, elevation, elev_b) %>% 
+  count(elev_b)
+```
+
+## Plot of frog survival for each translocation - grouped by site and arranged by site-specific median survival
 
@@
 ```
-* Strong effect of site on frog survival.
+* Strong effect of site/translocation on frog survival.
 * Survival is highly variable across sites, but highly repeatable within a site. 
-* Indicates need to include site_id as a group-level effect.
+* Indicates need to include site_id and/or translocation_id as group-level effect(s).
+
+## Plot of frog survival for each translocation - grouped by elevation
+
+```{r frog-survival-by-elevation}
+frog_translocation %>%
+  distinct(site_id, date, elevation) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = elevation, y = median_survival)) + 
+      geom_vline(aes(xintercept = elevation), linetype = "dotted", color = "darkgray") +
+      geom_point(size = 3) +
+      geom_text(aes(label = site_id, x = elevation, y = 1.05), angle = 90) +
+      coord_cartesian(ylim = c(0, 1.08)) +
+      labs(x = "Elevation of recipient site", y = "1-year frog survival")
+```
+Elevation has a clear effect on frog survival, but there is a substantial amount of variation that remains unaccounted for.
 
@@
 ```
-* Across all translocated frogs, the 3 donor sites have similar bd_loads. A plot of The 3 donor sites differ somewhat in bd_load of translocated cohorts. 
+* Across all translocated frogs, the 3 donor sites have similar bd_loads. A plot of the 3 donor sites indicate slight differences in bd_load of translocated cohorts. 
 * Across all donor sites and translocations, survival of translocated cohorts may be weakly inversely related to median bd_load of cohorts.
 However, grouping by donor site shows that each donor site has its own bd_load-frog survival relationship.
-* bd_load unrelated, or perhaps weakly related to differences in frog survival between donor sites.
+* bd_load unrelated, or perhaps weakly related, to differences in frog survival between donor sites.
 
@@
   ggplot(aes(x = donor, y = length)) +
-    geom_boxplot()
+    geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.4, fill = "white", stackratio = 0.7)
 ```

### 2022-05-12 code/classification_summary.R
Add classification_summary function

@@ -1,0 +1,70 @@
+#' Posterior Classification Summaries
+#'
+#' Given a set of observed data including a binary response variable y 
+#' and an rstanreg model of y, 
+#' this function returns summaries of the model's posterior classification quality.
+#' These summaries include a confusion matrix as well as estimates of the model's
+#' sensitivity, specificity, and overall accuracy.
+#' 
+#' @param model an rstanreg model object with binary y
+#' @param data data frame including the variables in the model, both response y and predictors x
+#' @param cutoff probability cutoff to classify a new case as positive (0.5 is the default)
+#'
+#' @return a list
+#' @export
+#' @import janitor dplyr
+#' @importFrom rstanarm posterior_predict
+#'
+#' @examples
+#' x <- rnorm(20)
+#' z <- 3*x
+#' prob <- 1/(1+exp(-z))
+#' y <- rbinom(20, 1, prob)
+#' example_data <- data.frame(x = x, y = y)
+#' example_model <- rstanarm::stan_glm(y ~ x, data = example_data, family = binomial)
+#' classification_summary(model = example_model, data = example_data, cutoff = 0.5)                   
+classification_summary <- function(model, data, cutoff = 0.5){
+  # This function summarizes the classifications across all cases
+  if(!("stanreg" %in% class(model))){ stop("the model must be a stanreg object.")}
+  
+  # Calculate probability posterior predictions
+  predictions <- posterior_predict(model, 
+                                   newdata = data)
+  
+  # Turn the predictions into classifications
+  if("lmerMod" %in% class(model)){
+    y <- as.data.frame(data %>% dplyr::select(as.character(model$formula)[2]))[,1]
+  }
+  else{
+    y <- as.data.frame(data %>% dplyr::select(model$terms[[2]]))[,1]
+  }
+  
+  classifications <- data.frame(proportion = colMeans(predictions)) %>% 
+    mutate(classification = as.numeric(proportion >= cutoff)) %>% 
+    mutate(y = y)
+  
+  # Confusion matrix
+  confusion_matrix <- classifications %>% 
+    tabyl(y, classification)
+  if(ncol(confusion_matrix) == 2){
+    if("1" %in% names(confusion_matrix)){
+      
+      confusion_matrix <- confusion_matrix %>% 
+        mutate("0" = rep(0,nrow(.)))
+    }
[…]

### 2022-05-12 code/translocation_survival_analysis.Rmd
Add analysis and model checking code

@@
 
-The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package, brms.
+The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package rstanarm.
+
+## Load packages
+
+```{r load-packages}
+library(tidyverse)
+# library(brms)
+library(rstanarm)
+library(bayesplot)
+library(tidybayes)
+library(loo)
+library(broom.mixed)
+library(janitor)
+```
+
+## Read in frog translocation dataset  
+* Read in data, create variables for analysis, standardize continuous variables. 
+* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+```{r read-translocation-data}
+d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  drop_na(bd_load) %>%  # drop xx records from 70xxx where bd_load is NA
+  mutate(
+    first = if_else(order == 1, 1, 0),
+    surv = if_else(survival < 0.5, 0, 1),
+    male = if_else(sex == "m", 1, 0),
+    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
+    elev_z = arm::rescale(elevation),
+    snowt_z = arm::rescale(snow_t),
+    snowt1_z = arm::rescale(snow_t1),
+    day_z = arm::rescale(day),
+    length_z = arm::rescale(length),
+    bdload_l = log10(bd_load + 1),
+    bdload_z = arm::rescale(bdload_l),
+    shore_c = shore - mean(shore),
+    male_c = male - mean(male),
+    elevlo_c = elev_lo - mean(elev_lo),
+    first_c = first - mean(first),
+    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
+    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d1 <- d1 %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup() %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  select(-transno) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d1, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id)
+```
+
+## Null values - check
+
+```{r check-null}
+d1 %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+## Model using non-standardized predictors
+### Model specification
+```{r}
+m1a <- stan_glm(
+  surv ~ length + snow_t + snow_t1 + day + bdload_l + elevation + first + male + donor,
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors, chain diagnostics, posterior predictive check
+```{r}
+prior_summary(m1a)
+
+mcmc_trace(m1a, size = 0.1)
+mcmc_dens_overlay(m1a)
+
+summary(m1a)
+
+pp_check(m1a, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+- mcmc diagnostics all look good.
+- Posterior predictive distribution implied by regression model closely matches original data (based on mean frog survival).
+
+### Posterior analysis
+```{r}
+tidy(m1a, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_intervals(m1a, point_est = "median", prob = 0.80, prob_outer = 0.9) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  # add 0-line to accentuate default line
+  ggtitle("Model with non-standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mcmc_areas(m1a, area_method = "equal height", prob = 0.90) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with non-standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1a))
+```
+- Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.  
+- Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+
+## Model using standardized predictors
+### Model specification
+```{r}
+m1b <- stan_glm(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor,
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4, 
+  seed = 84735)
+```
+
+### Priors, chain diagnostics, posterior predictive check
+```{r}
+prior_summary(m1b)
+
+mcmc_trace(m1b, size = 0.1)
+mcmc_dens_overlay(m1a)
+
+summary(m1b)
+
+pp_check(m1b, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+- mcmc diagnostics all look good.
+- Posterior predictive distribution implied by regression model closely matches original data (based on mean frog survival).
+
+### Posterior analysis
+```{r}
+tidy(m1b, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1b, area_method = "equal height", prob = 0.90) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1b))
+```
+- As expected, using standardized predictors allows easy comparison across predictors. Important predictors are unchanged. 
+- Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+
+## Possible grouping structures to add to model
+The dataset has three groups: site_id, translocation_id, and donor.
+Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful.
+Site_id and translocation_id contain 12 and 24 levels, respectively.
+Because translocation_id identifies every translocation and site_id contains a collection of 1-4 translocations, translocation_id may provide the best model fit.
+Including site_id and translocation_id as separate grouping variables in a single model is also worth considering.
+However, this would require dropping all site ids (4) from the dataset at which only 1 translocation has been conducted (to ensure within-group variation).
+This would not take full advantage of the dataset so was excluded from further consideration.
+Going forward, will build 2 models with identical predictors and either site_id or translocation_id as a grouping variables and compare fit using loo.
+
+## Model using standardized predictors and site_id grouping variable
+### Model specification
+```{r}
+m1f <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor + (1 | site_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1f)
+
+b1 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_trace(m1f, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1f, pars = b1)
+
+summary(m1f)
+```
+
+### Posterior predictive check
+```{r}
+loo_m1f <- loo(m1f, cores = 4, save_psis = TRUE)
+print(loo_m1f)
+
+pp_check(m1f, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Posterior analysis
+```{r}
+tidy(m1f, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1f, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & site_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1f))
+
+# Frog survival by site_id
+mcmc_areas_ridges(m1f, regex_pars = "site_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each recipient site_id")
+```
+- Based on 90% uncertainty intervals, important predictors reduced to elevation and male.
+
+## Model using standardized predictors and translocation_id grouping variable)
+### Model specification
+```{r}
+m1g <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor + (1 | translocation_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1g)
+
+b1 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_trace(m1g, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1g, pars = b1)
+
+summary(m1g)
+```
+
+### Posterior predictive check
+```{r}
+loo_m1g <- loo(m1g, cores = 4, save_psis = TRUE)
+print(loo_m1g)
+
+pp_check(m1g, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Posterior analysis
+```{r}
+tidy(m1g, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1g, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1g))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m1g, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+- Based on loo, model with translocation_id as grouping variable provides slightly better fit than model with site_id as grouping variable. 
+- Based on 90% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
+
+## Calculate posterior classification accuracy
+```{r}
+# Posterior predictive models for each frog in dataset
+set.seed(84735)
+survival_predict_1 <- posterior_predict(m1g, newdata = d1)
+
+# Classify frog survival based on mean predicted survival probability
+survival_predict_2 <- d1 %>% 
+  mutate(survival_prob = colMeans(survival_predict_1),
+         survival_class = as.numeric(survival_prob >= 0.5)) %>% 
+  select(site_id, translocation_id, year, snowt1_z, elev_z, male_c, donor, surv, survival_prob, survival_class)
+
+# Generate confusion matrix and calculate accuracy
+survival_predict_2 %>% 
+  tabyl(surv, survival_class) %>% 
+  adorn_totals(c("row", "col"))
+
+(471 + 138)/767 # overall_accuracy 
+471/533 # true negative rate
+138/234 # true positive rate
+
+# Generate confusion matrix using function
+set.seed(84735)
+classification_summary(m1g, data = d1, cutoff = 0.5)
+```
+- Specificity or "true negative rate" measures the proportion of Y = 0 observations that are accurately classified (e.g., proportion of frogs correctly identified as not surviving).
+- Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.
+
+## Posterior predictive models
+### Create reduced model containing only important predictors
+- Reduced model is helpful because it reduces the number of values that need to be generated for use in predictive models. 
+```{r}
+m1g_reduce <- stan_glmer(
+  surv ~ snowt1_z + elev_z + male_c + donor + (1 | translocation_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Simulate and plot posterior predictive models & observations
+```{r}
+m1g_group_means <- d1 %>% 
+  group_by(translocation_id) %>% 
+  summarize(count = n(), psurv = mean(surv), elev_z = mean(elev_z), snowt1_z = mean(snowt1_z)) %>% 
+  mutate(
+    male_c = as.factor(0.548), 
+    donor = case_when(
+      str_detect(translocation_id, "74976") | str_detect(translocation_id, "70556") ~ "70459",
+      str_detect(translocation_id, "70413") ~ "70567",
+      TRUE ~ "72996")) %>% 
+  arrange(psurv)  # to arrange plot values/labels by psurv
+
+predictions_m1g <- posterior_predict(m1g_reduce, newdata = m1g_group_means)
+
+ppc_intervals(
+  y = m1g_group_means$psurv,
+  yrep = predictions_m1g, 
+  prob = 0.5, 
+  prob_outer = 0.8) +
+ggplot2::scale_x_continuous(
+  labels = m1g_group_means$translocation_id,
+  breaks = 1:nrow(m1g_group_means)) +
+  xaxis_text(angle = 90, vjust = 0.5) +
+xlab("Translocation_id")
+```
+- Uncertainty intervals appear not correct for binomial model, but not sure how to adjust. 
+- It appears that the model is doing a good job of predicting survival based on site-level and individual-level characteristics.  
+- What explains the low predicted survival at 70370 (both translocations), and high versus low survival at 70550 for 1st and 2nd translocations, respectively?
+Both sites seem to contain high quality habitat (i.e., high elevation).
+- Estimated survival only for males. Run another set of models for females?
+
+## TO DO
+- Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).  
+- Create plots to show effect of all important variables.  
+- When model-building and testing is complete, renumber models in sequence, rerun all code chunks.  
+- Create analysis of survival from first translocation at a site as a predictor of survival from subsequent translocations.
+For this analysis, will need to drop all sites at which only a single translocation was conducted.

### 2022-05-30 code/translocation_survival_analysis.Rmd
Minor code updates and typo fixes

--- 2022-03-03 code/translocation_survival_analysis.Rmd
@@
 
-The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package, brms.
+The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package rstanarm.
+
+## Load packages
+
+```{r load-packages}
+library(tidyverse)
+library(brms)
+library(rstanarm)
+library(bayesplot)
+library(tidybayes)
+library(loo)
+library(broom.mixed)
+library(janitor)
+
+source("classification_summary.R")
+```
+
+## Read in frog translocation dataset  
+* Read in data, create variables for analysis, standardize continuous variables. 
+* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+```{r read-translocation-data}
+d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  drop_na(bd_load) %>%  # drop xx records from 70xxx where bd_load is NA
+  mutate(
+    first = if_else(order == 1, 1, 0),
+    surv = if_else(survival < 0.5, 0, 1),
+    male = if_else(sex == "m", 1, 0),
+    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
+    elev_z = arm::rescale(elevation),
+    snowt_z = arm::rescale(snow_t),
+    snowt1_z = arm::rescale(snow_t1),
+    day_z = arm::rescale(day),
+    length_z = arm::rescale(length),
+    bdload_l = log10(bd_load + 1),
+    bdload_z = arm::rescale(bdload_l),
+    shore_c = shore - mean(shore),
+    male_c = male - mean(male),
+    elevlo_c = elev_lo - mean(elev_lo),
+    first_c = first - mean(first),
+    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
+    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d1 <- d1 %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id) %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  select(-transno) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d1, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id)
+```
+
+## Null values - check
+
+```{r check-null}
+d1 %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+## Model using non-standardized predictors
+### Model specification
+```{r}
+m1a <- stan_glm(
+  surv ~ length + snow_t + snow_t1 + day + bdload_l + elevation + first + male + donor,
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1a)
+
+mcmc_trace(m1a, size = 0.1)
+mcmc_dens_overlay(m1a)
+
+summary(m1a)
+```
+- mcmc diagnostics all look good.
+
+### List model variable names
+```{r}
+get_variables(m1a)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m1a, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+- Posterior predictive distribution implied by regression model closely matches original data (based on mean frog survival).
+
+### Calculate leave-one-out cross-validation (loo)
+```{r}
+loo_m1a <- loo(m1a, cores = 4, save_psis = TRUE)
+print(loo_m1a)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1a, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_intervals(m1a, point_est = "median", prob = 0.80, prob_outer = 0.9) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  # add 0-line to accentuate default line
+  ggtitle("Model with non-standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mcmc_areas(m1a, area_method = "equal height", prob = 0.90) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with non-standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1a))
+```
+- Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.  
+- Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+
+## Model using standardized predictors
+### Model specification
+```{r}
+m1b <- stan_glm(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor,
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4, 
+  seed = 84735)
+```
+
+### Priors, chain diagnostics, posterior predictive check
+```{r}
+prior_summary(m1b)
+
+mcmc_trace(m1b, size = 0.1)
+mcmc_dens_overlay(m1a)
+
+summary(m1b)
+```
+- mcmc diagnostics all look good.
+
+### List model variable names
+```{r}
+get_variables(m1b)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m1b, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+- Posterior predictive distribution implied by regression model closely matches original data (based on mean frog survival).
+
+### Calculate loo
+```{r}
+loo_m1b <- loo(m1b, cores = 4, save_psis = TRUE)
+print(loo_m1b)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1b, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1b, area_method = "equal height", prob = 0.90) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors only",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1b))
+```
+- As expected, using standardized predictors allows easy comparison across predictors. Important predictors are unchanged. 
+- Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+
+## Compare models
+```{r}
+loo_compare(loo_m1a, loo_m1b)
+```
+- As expected, models with non-standardized and standardized predictor variables have identical predictive accuracy.
+
+## Possible grouping structures to add to model
+The dataset has three groups: site_id, translocation_id, and donor site.
+Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful.
+Site_id and translocation_id contain 12 and 24 levels, respectively, making them suitable as grouping variables.
+Including both site_id and translocation_id as nested grouping variables (i.e., translocations nested within site ids) also seems justified.
+Having more levels within a group would seem to increase the power to detect effects of predictors.
+With fewer levels/less variability, effect may be accounted for primarily by grouping variable itself instead of predictor.
+With only 1-4 translocations per site and only 12 site ids total, it is possible that model with site_id as sole grouping variable and model with site_id and translocation_id as nested grouping variables will attribute more of the variation in frog survival to the grouping variables and less to the predictors.
+A priori, that would suggest that the model with translocation_id as the grouping variable might be the most useful model structure (assuming that all models have similar posterior predictive accuracy).
+Lacking any strong rationale for  one model structure over another, built 3 models and compared their posterior predictive accuracy using loo.
+
+## Model using standardized predictors and site_id grouping variable
+### Model specification
+```{r}
+m1c <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor + (1 | site_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1c)
+
+b1 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_trace(m1c, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1c, pars = b1)
+
+summary(m1c)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m1c, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m1c <- loo(m1c, cores = 4, save_psis = TRUE)
+print(loo_m1c)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1c, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1c, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & site_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1c))
+
+# Frog survival by site_id
+mcmc_areas_ridges(m1c, regex_pars = "site_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each recipient site_id")
+```
+- Based on 90% uncertainty intervals, important predictors reduced to elevation and male.
+
+## Model using standardized predictors and translocation_id grouping variable)
+### Model specification
+```{r}
+m1d <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor + (1 | translocation_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1d)
+
+b1 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_trace(m1d, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1d, pars = b1)
+
+summary(m1d)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m1d, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m1d <- loo(m1d, cores = 4, save_psis = TRUE)
+print(loo_m1d)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1d, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1d, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1d))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m1d, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+- Based on loo, model with translocation_id as grouping variable provides slightly better fit than model with site_id as grouping variable. 
+- Based on 90% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
+
+## Model using standardized predictors and 2 grouping variables: translocation_id nested within site_id)
+### Model specification
+```{r}
+m1e <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + donor + (1 | translocation_id) + (1 | site_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m1e)
+
+b1 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_trace(m1e, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1e, pars = b1)
+
+summary(m1e)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m1e, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m1e <- loo(m1e, cores = 4, save_psis = TRUE)
+print(loo_m1e)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1e, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1e, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1e))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m1e, regex_pars = ("translocation_id")) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+
+# Frog survival by site_id
+mcmc_areas_ridges(m1e, regex_pars = ("site_id")) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+
+## Compare 4 models with different grouping structures
+```{r}
+loo_compare(loo_m1b, loo_m1c, loo_m1d, loo_m1e)
+```
+- Based on expected log-predictive densities (ELPD), 3 models with grouping variables all have substantially higher posterior predictive accuracy than model without any grouping variables (elpd_diff of m1b is more than 2 SE lower than elpd_diff of m1c: elpd_diff = -18.2 $\pm$ 12.6, 2 SE interval does not overlap 0: -5.6 -- -30.8).
+- 3 models with 1 or more grouping variables are equivalent and have similar posterior predictive accuracy (elpd_diff of m1e and m1d are within 2 SE of elpd_diff of m1c).  
+- m1d seems to provide more insight into the effect of the predictors, so chose this model for additional analysis.
+
+## Additional analysis of model m1d
+### Posterior classification accuracy
+```{r}
+# Posterior predictive models for each frog in dataset
+set.seed(84735)
+survival_predict_1 <- posterior_predict(m1d, newdata = d1)
+
+# Classify frog survival based on mean predicted survival probability
+survival_predict_2 <- d1 %>% 
+  mutate(survival_prob = colMeans(survival_predict_1),
+         survival_class = as.numeric(survival_prob >= 0.5)) %>% 
+  select(site_id, translocation_id, year, snowt1_z, elev_z, male_c, donor, surv, survival_prob, survival_class)
+
+# Generate confusion matrix and calculate accuracy
+survival_predict_2 %>% 
+  tabyl(surv, survival_class) %>% 
+  adorn_totals(c("row", "col"))
+
+(471 + 138)/767 # overall_accuracy 
+471/533 # true negative rate
+138/234 # true positive rate
+
+# Generate confusion matrix using function
+set.seed(84735)
+classification_summary(m1d, data = d1, cutoff = 0.5)
+```
+- Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.  
+- Specificity or "true negative rate" measures the proportion of Y = 0 observations that are accurately classified (i.e., proportion of frogs correctly identified as not surviving).
+
+### Create reduced model containing only important predictors for subsequent analyses
+- Reduced model is helpful because it reduces the number of values used to calculate predictive models. 
+```{r}
+m1d_reduce <- stan_glmer(
+  surv ~ snowt1_z + elev_z + male_c + donor + (1 | translocation_id),
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Confirm fit of reduced model is comparable to full model
+```{r}
+loo_m1d_reduce <- loo(m1d, cores = 4, save_psis = TRUE)
+print(loo_m1d)
+
+loo_compare(loo_m1d, loo_m1d_reduce)
+```
+
+### Simulate and plot posterior predictive models & observations
+```{r}
+m1d_group_means <- d1 %>% 
+  group_by(translocation_id) %>% 
+  summarize(count = n(), psurv = mean(surv), elev_z = mean(elev_z), snowt1_z = mean(snowt1_z)) %>% 
+  mutate(
+    male_c = as.factor(0.548), 
+    donor = case_when(
+      str_detect(translocation_id, "74976") | str_detect(translocation_id, "70556") ~ "70459",
+      str_detect(translocation_id, "70413") ~ "70567",
+      TRUE ~ "72996")) %>% 
+  arrange(psurv)  # to arrange plot values/labels by psurv
+
+predictions_m1d <- posterior_predict(m1d_reduce, newdata = m1d_group_means)
+
+ppc_intervals(
+  y = m1d_group_means$psurv,
+  yrep = predictions_m1d, 
+  prob = 0.5, 
+  prob_outer = 0.8) +
+ggplot2::scale_x_continuous(
+  labels = m1d_group_means$translocation_id,
+  breaks = 1:nrow(m1d_group_means)) +
+  xaxis_text(angle = 90, vjust = 0.5) +
+xlab("Translocation_id")
+```
+- Uncertainty intervals appear incorrect for binomial model, but not sure how to adjust.  
+- Based on mean values, model does an adequate job of predicting survival based on group-level and individual-level characteristics.  
+- What explains the low predicted survival at 70370 (both translocations), and high versus low predicted survival at 70550 for 1st and 2nd translocations, respectively?
+Both sites seem to contain high quality habitat (i.e., high elevation).
+- The predicted survival is based on assumption that only males are included. Run another set of models for females?
+
+### Plot posterior plausible relationships between each predictor and the probability of frog survival (conditional effects)
+- As described in https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on.
+By default, the mean is used for continuous variables and the reference category is used for factors, but you may change these values via argument conditions.
+This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."
+- Using rstanarm, I was unable to do this as I expected to: Start with data = d1 and using "newdata = " to specify that predictions should be based on data in d1_elevz (for example), in which non-plotted continuous predictors are set to the mean value.
+Not sure if my alternative approach is correct.
+- Approach used: Instead of starting the routine with d1, I started with the new dataset in which all variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).
+This produced plots/results that were very similar to those produced in brms using the conditional_effects function.
+
+#### Effect of elevation
+```{r}
+d1_elevz <- d1 %>% 
+  mutate(snowt1_z = mean(snowt1_z), 
+         male_c = recode(male_c, "-0.452" = "0.548"),
+         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
+
+d1_elevz %>% distinct(snowt1_z, male_c, donor) # ensure variables were set to correct values
+
+d1_elevz %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = elev_z, y = surv)) +
+    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+#    geom_point(aes(y = .epred)) +
+  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
+
+# d1 %>%  # Test of add_predicted_draws used for similar purpose - unsuccessful
+#   select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+#   add_predicted_draws(m1d_reduce, ndraws = 100, re_formula = NA, newdata = d1_elevz) %>% View()
+#   ggplot(aes(x = elev_z, y = surv)) +
+#     geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+#     geom_point(aes(y = .epred)) +
+#   labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
+```
+
+#### Effect of winter severity in year t1
+```{r}
+d1_snowt1z <- d1 %>% 
+  mutate(elev_z = mean(elev_z), 
+         male_c = recode(male_c, "-0.452" = "0.548"),
+         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
+
+d1_snowt1z %>% distinct(elev_z, male_c, donor) # ensure variables were set to correct values
+
+d1_snowt1z %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = snowt1_z, y = surv)) +
+    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+#    geom_point(aes(y = .epred)) +
+  labs(x = "Winter severity in year following translocation (standardized)", y = "Probability of 1-year frog survival") 
+```
+
+#### Effect of sex
+```{r}
+d1_malec <- d1 %>% 
+  mutate(elev_z = mean(elev_z), 
+         snowt1_z = mean(snowt1_z),
+         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
+
+d1_malec %>% distinct(elev_z, snowt1_z, donor) # ensure variables were set to correct values
+
+d1_malec %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = male_c, y = surv)) +
+    geom_boxplot(aes(y = .epred, x = male_c)) +
+#    geom_point(aes(y = .epred)) +
+    labs(x = "Sex (centered)", y = "Probability of 1-year frog survival")
+```
+
+#### Effect of donor
+```{r}
+d1_donor <- d1 %>% 
+  mutate(elev_z = mean(elev_z), 
+         snowt1_z = mean(snowt1_z),
+         male_c = recode(male_c, "-0.452" = "0.548"))
+
+d1_donor %>% distinct(elev_z, snowt1_z, male_c) # ensure variables were set to correct values
+
+d1_donor %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = donor, y = surv)) +
+    geom_boxplot(aes(y = .epred, x = donor)) +
+#    geom_point(aes(y = .epred)) +
+    labs(x = "Donor population", y = "Probability of 1-year frog survival")
+```
+
+#### Joint effects of elevation and sex
+```{r}
+d1_elevz_malec <- d1 %>% 
+  mutate(snowt1_z = mean(snowt1_z), 
+         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
+
+d1_elevz_malec %>% distinct(snowt1_z, donor) # ensure variables were set to correct values
+
+pal <- c("#0d0887", "#f89540")
+d1_elevz_malec %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = elev_z, y = surv, color = male_c)) +
+    geom_line(aes(y = .epred, group = paste(male_c, .draw)), size = 0.4, alpha = 0.5) +
+#    geom_point(aes(y = .epred)) +
+    labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival") +
+    scale_color_manual(values = pal)  + 
+    theme_classic()
+```
+
+## For comparison, fit model using brms and conditional_effects function to plot variable-specific effects
+### Fit model
+```{r}
+m1d_reduce_brms <- brm(
+  surv ~ snowt1_z + elev_z + male_c + donor + (1 | translocation_id),
+  data = d1,
+  family = bernoulli(),
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Plot conditional effects
+```{r}
+plot(conditional_effects(m1d_reduce_brms, spaghetti = TRUE, mean = TRUE, ndraws = 100), ask = FALSE)
+```
+
+## TO DO
+- Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).  
+- Check approach to plotting conditional effects.  
+- Create analysis in which survival from first translocation to a site is a predictor of survival from subsequent translocations.
+For this analysis, will need to drop all sites at which only a single translocation was conducted.
+- Try un-standardizing predictor values in conditional-effects plots (and other plots?) to produce more intuitive visualizations.

### 2022-05-30 doc/notebook/notebook_translocation.Rmd
Update text to reflect current status of analysis

--- 2022-03-07 doc/notebook/notebook_translocation.Rmd
@@
 
--   site_id: site to which frogs were translocated
--   elevation
--   date
--   day
--   year
--   donor
--   order: first translocation to site = 0, subsequent translocation = 1
--   type: translocation conducted on foot versus by helicopter (not included)
--   shore: suitability of shore/bank habitat for frog overwintering
--   pit_tag_ref
--   survival: individual-level estimated survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation
--   sex
--   length: at release
--   weight: at release
--   condition: index based on length and weight
--   swab_id
--   bd_load: at release
+* site_id: site to which frogs were translocated  
+* elevation: water body surface elevation above sea level, in meters  
+* shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)  
+* snow_t: snow water content (measured on April 1) in year of translocation  
+* snow_t1: snow water content (measured on April 1) in year following translocation  
+* date: date frogs released at recipient site  
+* year: year frogs released at recipient site  
+* day: day of the year on which frogs were released  
+* order: order of translocations to a site, from 1-4. Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation  
+* donor: site_id from which frogs were collected for translocation  
+* pit_tag_ref: pit tag id  
+* sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female   
+* length: frog snout-vent length at capture at donor site (in mm)  
+* weight: frog weight at capture at donor site (in grams)  
+* bd_load: number of ITS copies at release, measured from swab collected at capture at donor site  
+* survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation.
+Re-coded as indicator variable (variable name = "surv"): 0 = survival probability < 0.5; 1 = survival probability $\geq$ 0.5)
 
-Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.
-
-Frog condition might be problematic because size and residuals are correlated.
-Specifically, the largest frogs have the largest positive residuals from a weight:length regression line, even after log10 transformation.
-As such, length and condition are probably not independent.
+Other variables
+* Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.
+* Frog condition is problematic because size and residuals from length:weight regression line are correlated.
+Specifically, the largest frogs have the largest positive residuals from a length:weight regression line, even after log10 transformation.
+As such, length and condition are not independent.
 
 ```{r frog-length-weight-plot}
-read_csv(here::here("data", "clean", "frog_translocation.csv")) %>% 
-ggplot(aes(x = log10(length), y = log10(weight))) +
+read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+ggplot(aes(x = length, y = weight)) +
   geom_point() +
-  geom_smooth(method = "lm")
-plot
+  geom_smooth(method = "lm") +
+  scale_x_log10() +
+  scale_y_log10() +
+  labs(x = "Length (mm)", 
+       y = "Weight (g)")
 ```
 
-Another consideration when designing the analysis is that there are 12 records in which bd_load is null.
+* Another consideration when designing the analysis is that there are 12 records in which bd_load is null.
 Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival.
-39 frogs were translocated on that date, so should probably drop the 9 records where bd_load = NA prior to running the analysis.
-If bd_load is not an important predictor, can run analysis using the full dataset.
+39 frogs were translocated on that date, so plenty of swab results are available to represent that site.
+Prior to running the analyses, should drop the 12 records where bd_load = NA.
+If bd_load is not an important predictor, can run analysis (without the bd_load predictor) using the full dataset.
+
+## Exploratory analyses
+
+Exploratory data analyses conducted in translocation_survival_exploratory.Rmd provided several notable insights.
+
+### Frog survival across recipient sites
+* 1-year frog survival is highly variable across sites, but highly repeatable within a site.
+This indicates the importance of including site_id or translocation_id as group-level effects.
+It remains unclear whether including a translocation_id nested within site_id should be considered (or if this nested group-level effect is even feasible given the relatively small number of translocations per site).
+
+### Donor site
+* Donor site has a strong association with frog survival, with survival at 70459 > 70567 > 72996.
+Explored possible effects of bd_load and frog size as drivers of this pattern.
+Differences in survival between donor sites seem unrelated to bd_load, but appear closely associated with frog size.
+Whether frog size is the sole driver or is a result of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) is unknown.
+  * Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.  
+  * Strong effect of frog size on 1-year survival, with frogs > 55 mm having substantially higher 1-year survival. Frogs >60 mm had by far the highest survival.  
+  * The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.  
+  * Frog size differs between donor site: size at 72996 < 70459 = 70567.
+This partially mirrors the observed differences in survival between sites.
+Sufficient overlap in frog sizes between donor sites that it might be possible to include both frog length and donor in the model.
+
+### Bd load
+* Frogs with the highest loads at the time of translocation had the lowest survival.
+The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies.
+Notable result given that the threshold level of 56,000 copies is /~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.
+
+### Recipient site characteristics
+* **Shore habitat**: The presence of shore habitat is associated with increased frog survival.
+Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature.
+Tight association between shore habitat presence/absence and elevation will make it impossible to disentangle these effects, include only elevation in the model.
+* **Elevation**: Strong positive effect on frog survival.
+
+### Translocation timing
+* Day of the year showed no obvious relationship to frog survival.  
+* Year was positively associated with survival.
+This is likely not a direct effect of year, but instead reflects better-informed selection of recipient sites with high-quality frog habitat over time.
+As such, consider not including year in the model.
+
+### Winter severity
+* No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.
+
+### Translocation order
+* Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.  
+* It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.
+
+### Frog characteristics
+* Frog size had strong effect on 1-year survival.  
+* Minimal differences in survival between males and females (slightly higher in males).
 
@@
 
-survival \~ elevation + shore + day + year + order + donor + sex + length + log(bdload) \| site_id
+As originally conceived: survival \~ elevation + shore + snow_t + snow_t1 + year + day + order + donor + sex + length + log(bdload) \| site_id
 
-Group-level effects: In addition to the inclusion of site_id as a group-level effect, may need to include release_year, perhaps nested within site_id?
-Given that some sites only have a single release_year, will this cause problems?
+Based on exploratory analyses: survival \~ elevation + snow_t + snow_t1 + day + order + donor + sex + length + log(bdload) \| site_id
 
-Final steps before building model (in analysis Rmd):
+Pairs plot to identify correlations between predictors (ggpairs from GGally package):
+* It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length.
+As such, these two variables have the potential to compete for the same variation.
+Need to look at this more closely, perhaps by plotting frog lengths at each donor site.
+Also, plot length against survival, grouped by donor.
+Are these distributions as donor-specific as they appear to be?
+Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.
+* Also need to disentangle elevation from site_id - currently only one elevation per site_id.
+Try breaking continuous elevation into categories, with > 2 site_ids per category.
+Hopefully, that will make elevation variable less dependent on site_id.
+A similar situation may exist for the two winter severity variables and site_id.
+* Shore and elevation may also be closely associated. Are all sites with shore = 0 at low elevation? If so, may need to only include elevation in model.
 
--   Change to factor: order, day?, year?
--   Standardize variables?
--   Survival: Create 0/1 variable from probabilities. Change to factor?
+Group-level effects: 
+The dataset has three sources of hierarchical structure: site_id, translocation_id, and donor site.
+Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful.
+Site_id and translocation_id contain 12 and 24 levels, respectively, making them suitable as grouping variables.
+Including both site_id and translocation_id as nested grouping variables (i.e., translocations nested within site ids) also seems justified.
+Having more levels within a group would seem to increase the power to detect effects of predictors.
+With fewer levels/less variability, effect may be accounted for primarily by grouping variable itself instead of an associated predictor (e.g., elevation is closely associated with both site_id and translocation_id).
+With only 1-4 translocations per site and only 12 site ids total, it is possible that a model with site_id as sole grouping variable or a model with site_id and translocation_id as nested grouping variables will both attribute more of the variation in frog survival to the grouping variables and less to the predictors.
+A priori, that would suggest that the model with translocation_id as the grouping variable might be the most useful model structure (assuming that all models have similar posterior predictive accuracy).
+Lacking any strong rationale for one model structure over another, build 3 models and compare their posterior predictive accuracy using loo.
 
-## Results
+Interactions between predictor variables: General guidance is that interactions between the most important predictors should be included in model.
 
-To be added soon.
+## Final steps before building model (in analysis Rmd):
+
+* Bin elevation into "low" and "high" groups.  
+* Recode order as 0/1 (instead of 1--4): "first".  
+* Recode survival as 0/1 (instead of continuous probability): "surv".  
+* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below).  
+* Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.  
+* Standardize continuous variables: elevation, snow_t, snow_t1, day, length.
+Possibly standardize log(bd_load) also, but see Gelman for recommendations.
+Gelman recommends centering by subtracting the mean and standardizing by dividing by 2 standard deviations.
+Centering to a mean of 0 makes interactions more interpretable.
+Standardizing allows assessment of importance of predictors by comparing regression coefficients.
+Dividing by 2 SD puts continuous variables on approximately the same scale as binary (0/1) variables.
+This is true only if binary predictors have similar number of 0s and 1s.
+In the current dataset, there is some deviation from 1:1.
+Sex: female = 428, male = 351; order: 0 = 507, 1 = 272.
+Binary variables should be centered but not standardized.
+Variables in need of log-transformation should be transformed and then rescaled, although rescaling will not always be necessary.
+
+Relevant references:  
+* Gelman 2008: https://doi.org/10.1002/sim.3107  
+* Schielzeth 2010: https://doi.org/10.1111/j.2041-210X.2010.00012.x
+
+## Priors
+* For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.
+https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations.
+* Other recommendations in:  
+  * Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: https://doi.org/10.1111/oik.05985.  
+  * Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: https://doi.org/10.1002/ecs2.3739.
+
+## Next steps
+* Frog size had a marginally important effect on frog survival.
+Given association between frog size and donor varibles, create a second model that includes only donor = 72996.
+Given that there is considerable variation in frog sizes translocated from this site, this might allow an assessment of the effect of frog size independent of any effect of donor.
+* To assess within-site association between survival for the first and subsequent translocated cohorts, build another dataset that includes all recipient sites for which > 1 translocation was conducted.
+Include survival from the first translocation as the response variable and survival from the subsequent translocations as a predictor variable.
+Other predictors could also be included.

### 2022-06-07 doc/notebook/notebook_translocation.Rmd
Minor edits to fix typos, add Results header

@@
 
+## Results
+* Important predictor variables in all models are relatively consistent (see `translocation_analysis_survival.Rmd` for details).
+
 ## Next steps
 * Frog size had a marginally important effect on frog survival.
-Given association between frog size and donor varibles, create a second model that includes only donor = 72996.
+Given association between frog size and donor variables, create a second model that includes only donor = 72996.
 Given that there is considerable variation in frog sizes translocated from this site, this might allow an assessment of the effect of frog size independent of any effect of donor.
@@
 Include survival from the first translocation as the response variable and survival from the subsequent translocations as a predictor variable.
-Other predictors could also be included.
+Other predictors could also be included (e.g., year).

### 2022-06-09 code/translocation_survival_analysis.Rmd
Add model predicting survival based on 1st translocation

@@
 The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package rstanarm.
+
+# Model 1: Site-level and frog-level predictors
 
@@
 
-## Read in frog translocation dataset  
+## Create frog translocation dataset for analysis  
 * Read in data, create variables for analysis, standardize continuous variables. 
@@
 d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  drop_na(bd_load) %>%  # drop xx records from 70xxx where bd_load is NA
+  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
   mutate(
@@
 (471 + 138)/767 # overall_accuracy 
-471/533 # true negative rate
-138/234 # true positive rate
+471/533 # true negative rate (specificity)
+138/234 # true positive rate (sensitivity)
 
@@
 
+# Model 2: Survival of frogs from first translocation as predictor of survival in subsequent translocations
+## Load packages
+
+```{r}
+library(tidyverse)
+library(brms)
+library(rstanarm)
+library(bayesplot)
+library(tidybayes)
+library(loo)
+library(broom.mixed)
+library(janitor)
+
+source("classification_summary.R")
+```
+
+## Create frog translocation dataset for analysis  
+* Read in data, create variables for analysis, standardize continuous variables. 
+* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+```{r}
+d2 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  select(site_id, year, pit_tag_ref, survival) %>% 
+  mutate(
+    surv = if_else(survival < 0.5, 0, 1), 
+    across(c(site_id, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d2 <- d2 %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id) %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d2, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id) %>% 
+  relocate(transno, .after = translocation_id)
+
+```
+
+## Restructure dataset
+```{r}
+# Remove sites that received only 1 translocation
+d2 <- d2 %>% 
+  distinct(site_id, transno) %>% 
+  group_by(site_id) %>% 
+  mutate(trans_max = max(transno)) %>% 
+  ungroup(site_id) %>% 
+  select(-transno) %>% 
+  filter(trans_max > 1) %>% 
+  distinct(site_id, trans_max) %>% 
+  inner_join(d2, by = c("site_id"))
+
+# Create predictor from survival of first translocated cohort
+d2 <- d2 %>% 
+  filter(transno == 1) %>% 
+  group_by(site_id) %>% 
+  mutate(survival_trans1 = mean(survival)) %>%    # use mean or median?
+  ungroup(site_id) %>% 
+  distinct(site_id, survival_trans1) %>% 
+  mutate(across(c(survival_trans1), round, 3)) %>% 
+  inner_join(d2, by = c("site_id")) %>% 
+  filter(transno != 1) %>% 
+  relocate(translocation_id, .after = site_id) %>% 
+  mutate(survival_trans1_z = arm::rescale(survival_trans1), .after = survival_trans1) %>% 
+  select(-transno, -trans_max) 
+```
+
+## Model with non-standardized predictor, no grouping variable
+### Model specification
+```{r}
+m2a <- stan_glm(
+  surv ~ survival_trans1,
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### List model variable names
+```{r}
+get_variables(m2a)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m2a)
+
+b2 <- c("(Intercept)", "survival_trans1")
+
+mcmc_trace(m2a, size = 0.1, pars = b2)
+mcmc_dens_overlay(m2a, pars = b2)
+
+summary(m2a)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m2a, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m2a <- loo(m2a, cores = 4, save_psis = TRUE)
+print(loo_m2a)
+```
+
+### Posterior analysis
+```{r}
+tidy(m2a, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2a, area_method = "equal height", prob = 0.90, pars = b2) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with unstandardized predictor & no grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m2a))
+```
+
+## Model with standardized predictor, no grouping variable
+### Model specification
+```{r}
+m2b <- stan_glm(
+  surv ~ survival_trans1_z,
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### List model variable names
+```{r}
+get_variables(m2b)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m2b)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2b, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2b, pars = b3)
+
+summary(m2b)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m2b, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m2b <- loo(m2b, cores = 4, save_psis = TRUE)
+print(loo_m2b)
+```
+
+### Posterior analysis
+```{r}
+tidy(m2b, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2b, area_method = "equal height", prob = 0.90, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictor & no grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m2b))
+```
+
+## Model with standardized predictor, translocation_id grouping variable
+### Model specification
+```{r}
+m2c <- stan_glmer(
+  surv ~ survival_trans1_z + (1 | translocation_id),
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m2c)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2c, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2c, pars = b3)
+
+summary(m2c)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m2c, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m2c <- loo(m2c, cores = 4, save_psis = TRUE)
+print(loo_m2c)
+```
+
+### Posterior analysis
+```{r}
+tidy(m2c, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2c, area_method = "equal height", prob = 0.90, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictor & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m2c))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m2c, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+
+## Model with standardized predictor, site_id grouping variable
+### Model specification
+```{r}
+m2d <- stan_glmer(
+  surv ~ survival_trans1_z + (1 | site_id),
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+```{r}
+prior_summary(m2d)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2d, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2d, pars = b3)
+
+summary(m2d)
+```
+
+### Posterior predictive check
+```{r}
+pp_check(m2d, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Calculate loo
+```{r}
+loo_m2d <- loo(m2d, cores = 4, save_psis = TRUE)
+print(loo_m2d)
+```
+
+### Posterior analysis
+```{r}
+tidy(m2d, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2d, area_method = "equal height", prob = 0.90, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictor & site_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m2d))
+
+# Frog survival by site_id
+mcmc_areas_ridges(m2d, regex_pars = "site_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each site_id")
+```
+
+### Compare 3 models with different grouping structures
+```{r}
+loo_compare(loo_m2b, loo_m2c, loo_m2d)
+```
+- Based on expected log-predictive densities (ELPD), models with and without grouping variables have similar posterior predictive accuracy (elpd_diff of m2b is less than 2 SE lower than elpd_diff of m2d: elpd_diff = -3.1 $\pm$ 5.8, 2 SE interval overlaps 0: 2.7 -- -8.9).
+- Conclusion from all 3 models is the same: Average frog survival of first translocation is an important predictor of survival of individual frogs in subsequent translocations.
+
+## Posterior classification accuracy of best models
+```{r}
+# Generate confusion matrix for m2b
+set.seed(84735)
+classification_summary(m2b, data = d2, cutoff = 0.5)
+
+# Generate confusion matrix for m2d
+set.seed(84735)
+classification_summary(m2d, data = d2, cutoff = 0.5)
+```
+- As expected from loo results, models with and without grouping variable have the same posterior predictive accuracy.
+
 ## TO DO
@@
 - Check approach to plotting conditional effects.  
-- Create analysis in which survival from first translocation to a site is a predictor of survival from subsequent translocations.
-For this analysis, will need to drop all sites at which only a single translocation was conducted.
 - Try un-standardizing predictor values in conditional-effects plots (and other plots?) to produce more intuitive visualizations.

### 2022-06-10 code/translocation_survival_analysis_suppl.Rmd
Add supplementary analyses

---
title: "Predictors of frog survival following translocations - supplemental analyses"
author: "Roland Knapp"
output: html_notebook
---

The supplemental analyses in this notebook provide additional details to analyses in `translocation_survival_analysis.Rmd`.

## Load packages

```{r load-packages}
library(tidyverse)
library(brms)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(loo)
library(broom.mixed)
library(janitor)

source("classification_summary.R")
```

## Evaluate effect of frog size independent of donor population: Model m1d from `translocation_survival_analysis.Rmd` but using data subset where donor = 72996

### Read in frog translocation dataset  
* Read in data, create variables for analysis, standardize continuous variables. 
* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
```{r read-translocation-data}
d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
  mutate(
    first = if_else(order == 1, 1, 0),
    surv = if_else(survival < 0.5, 0, 1),
    male = if_else(sex == "m", 1, 0),
    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
    elev_z = arm::rescale(elevation),
    snowt_z = arm::rescale(snow_t),
    snowt1_z = arm::rescale(snow_t1),
    day_z = arm::rescale(day),
    length_z = arm::rescale(length),
    bdload_l = log10(bd_load + 1),
    bdload_z = arm::rescale(bdload_l),
    shore_c = shore - mean(shore),
    male_c = male - mean(male),
    elevlo_c = elev_lo - mean(elev_lo),
    first_c = first - mean(first),
    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
    surv = as.integer(surv))

# Add a descriptive translocation_id
d1 <- d1 %>% 
  distinct(site_id, year) %>% 
  arrange(site_id, year) %>% 
  group_by(site_id) %>% 
  mutate(transno = seq_len(n())) %>% 
  ungroup(site_id) %>% 
  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
  select(-transno) %>% 
  mutate(translocation_id = as.factor(translocation_id)) %>% 
  inner_join(d1, by = c("site_id", "year")) %>% 
  relocate(translocation_id, .after = site_id)
```

### Model specification
```{r}
d1_72996 <- d1 %>% 
  filter(donor == "72996" & !is.na(bd_load))

m1d_72996 <- stan_glmer(
  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + (1 | translocation_id),
  data = d1_72996,
  family = binomial,
  chains = 4, 
  iter = 5000*2, 
  cores = 4,
  seed = 84735)
```

### Priors & chain diagnostics
```{r}
prior_summary(m1d_72996)

b1_72996 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548")

mcmc_trace(m1d_72996, size = 0.1, pars = b1_72996)
mcmc_dens_overlay(m1d_72996, pars = b1_72996)

summary(m1d_72996)
```

### Posterior predictive check
```{r}
pp_check(m1d_72996, plotfun = "stat") +
    xlab("Frog survival rate")
```

### Posterior analysis
```{r}
tidy(m1d_72996, conf.int = TRUE, conf.level = 0.90)

mcmc_areas(m1d_72996, area_method = "equal height", prob = 0.90, pars = b1_72996) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
          "Posterior distributions with medians & 90% uncertainty intervals")

mean(bayes_R2(m1d_72996))

# Frog survival by translocation_id
mcmc_areas_ridges(m1d_72996, regex_pars = "translocation_id") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
  ggtitle("Frog survival for each translocation_id")
```
- Using dataset containing only frogs translocated from 72996 had relatively little effect on the predictors of frog survival. Specific to the objective of this analysis, frog size remained unimportant as a predictor of survival. However, effect of Bd load became marginally unimportant instead of unimportant (i.e., uncertainty interval includes 0 but barely).

## Assess effect on m1d of including all records, including where bd_load = NA

### Read in frog translocation dataset  
* Read in data, create variables for analysis, standardize continuous variables. 
* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
```{r read-translocation-data-including-na}
d1_allna <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
#  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
  mutate(
    first = if_else(order == 1, 1, 0),
    surv = if_else(survival < 0.5, 0, 1),
    male = if_else(sex == "m", 1, 0),
    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
    elev_z = arm::rescale(elevation),
    snowt_z = arm::rescale(snow_t),
    snowt1_z = arm::rescale(snow_t1),
    day_z = arm::rescale(day),
    length_z = arm::rescale(length),
    bdload_l = log10(bd_load + 1),
    bdload_z = arm::rescale(bdload_l),
    shore_c = shore - mean(shore),
    male_c = male - mean(male),
    elevlo_c = elev_lo - mean(elev_lo),
    first_c = first - mean(first),
    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
    surv = as.integer(surv))

# Add a descriptive translocation_id
d1_allna <- d1_allna %>% 
  distinct(site_id, year) %>% 
  arrange(site_id, year) %>% 
  group_by(site_id) %>% 
  mutate(transno = seq_len(n())) %>% 
  ungroup(site_id) %>% 
  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
  select(-transno) %>% 
  mutate(translocation_id = as.factor(translocation_id)) %>% 
  inner_join(d1_allna, by = c("site_id", "year")) %>% 
  relocate(translocation_id, .after = site_id)
```

### Model specification
```{r}
m1d_allna <- stan_glmer(
  surv ~ length_z + snowt_z + snowt1_z + day_z + elev_z + first_c + male_c + donor + (1 | translocation_id),
  data = d1_allna,
  family = binomial,
  chains = 4, 
  iter = 5000*2, 
  cores = 4,
  seed = 84735)
```

### List model variable names
```{r}
get_variables(m1d_allna)
```

### Priors & chain diagnostics
```{r}
prior_summary(m1d_allna)

b1_allna <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "elev_z", "first_c0.349", "male_c0.549", "donor70567", "donor72996")

mcmc_trace(m1d_allna, size = 0.1, pars = b1_allna)
mcmc_dens_overlay(m1d_allna, pars = b1_allna)

summary(m1d_allna)
```

### Posterior predictive check
```{r}
pp_check(m1d_allna, plotfun = "stat") +
    xlab("Frog survival rate")
```

### Posterior analysis
```{r}
tidy(m1d_allna, conf.int = TRUE, conf.level = 0.90)

mcmc_areas(m1d_allna, area_method = "equal height", prob = 0.90, pars = b1_allna) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
          "Posterior distributions with medians & 90% uncertainty intervals")

mean(bayes_R2(m1d_allna))

# Frog survival by translocation_id
mcmc_areas_ridges(m1d_allna, regex_pars = "translocation_id") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
  ggtitle("Frog survival for each translocation_id")
```

- Inclusion of all records (i.e., 12 records where bd_load = NA that were excluded in primary analysis) did not change the results.

### 2022-06-10 doc/notebook/notebook_translocation.Rmd
Add details for primary, secondary, and supplementary analyses

--- 2022-05-30 doc/notebook/notebook_translocation.Rmd
@@
 
-## Model structure
+## Analyses
+
+### Model structure
 
@@
 
-## Final steps before building model (in analysis Rmd):
+### Dataset: Final steps before building model
 
@@
 * Recode survival as 0/1 (instead of continuous probability): "surv".  
-* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below).  
+* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below.  
 * Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.  
@@
 
-## Priors
+### Priors
 * For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.
@@
 
-## Next steps
-* Frog size had a marginally important effect on frog survival.
-Given association between frog size and donor varibles, create a second model that includes only donor = 72996.
-Given that there is considerable variation in frog sizes translocated from this site, this might allow an assessment of the effect of frog size independent of any effect of donor.
-* To assess within-site association between survival for the first and subsequent translocated cohorts, build another dataset that includes all recipient sites for which > 1 translocation was conducted.
-Include survival from the first translocation as the response variable and survival from the subsequent translocations as a predictor variable.
-Other predictors could also be included.
+## Results
+- Important predictors of the 1-year survival of translocated frogs were quite consistent across all models (see `translocation_analysis_survival.Rmd` for details).  
+- In the "best" model that included standardized predictors and translocation_id as a grouping variable, important predictors included elevation (+), winter severity in the year following translocation (-), and donor[72996] (-).
+Model R2 was 0.32, so a substantial amount of variation remained unaccounted for.
+- Comparison of different grouping structures (translocation_id, site_id, translocation_id nested within site_id) indicated that all had similar posterior predictive accuracy.
+However, using translocation_id as the grouping variable allowed greater insights into the effects of the predictor variables, presumably due to the greater variation associated with the translocation_id grouping variable (n = 24) compared to the site_id grouping variables (n = 12).
+- Posterior predictive accuracy of "best" model was 0.79, and the true negative rate (proportion of frogs that did not survive that were accurately classified as not surviving: 0.88) was higher than the true positive rate (proportion of frogs that survived that were accurately classified as surviving: 0.59).
+Indicates that with the available site-level and frog-level predictors, frog non-survival is more accurately predicted than frog survival.
+This is borne out by sites such as 70370 which has attributes suggesting high frog survival (e.g., high elevation, available bank habitat, few predators), but in fact frogs from both translocations experienced very low survival.
+In contrast, no examples exist of sites where frog survival was (retrospectively) predicted to be low, but actual survival was high (although 70449 comes closest to providing such an example).
+- Two supplementary analyses were conducted to gain additional insights into the 1-year survival model.  
+  - Because the three donor populations differed in the size of collected/translocated frogs, frog size was potentially associated with the important effect of donor population.
+Therefore, I reran the analysis using only frogs translocated from 72996.
+Despite substantial variation in the size of frogs collected/translocated from this site, results indicated that frog size was unrelated to frog survival.
+This suggests that the effect of donor population is due to an unmeasured population-level attribute, e.g., a characteristic associated with frog resistance/tolerance of Bd infection.
+The existence of such an attribute is suggested by the consistently high survival resulting from the four translocations conducted to date from 70459 to two very different sites, and the fact that in both recipient sites that have received frogs from 70459 tadpole mouthparts are fully pigmented.
+That is not the case with tadpoles at sites that received frogs from 72996.
+  - The frog translocation dataset contained 12 records for which bd_load = NA.
+To allow inclusion of bd_load in the model, these 12 records were dropped.
+Given that bd_load was not an important predictor of frog survival, I reran the model using the complete dataset and dropping bd_load from the model.
+Dropping/including those 12 records had no effect on model results.
+- The results from a secondary analysis to complement the 1-yr survival model demonstrated that the average survival in the first translocation was an important predictor of frog survival in subsequent translocations.
+This demonstrates the importance of recipient site attributes (elevation, habitat, predators, etc.) in determining frog survival.

### 2022-10-04 doc/manuscript/translocation.qmd
Add notes about possible journal outlets

---
format: docx
editor: visual
csl: science.csl
bibliography: translocation.bib
reference-doc: translocation.docx
---

## Title: Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### Authors:

Roland A. Knapp^1,2^\*, Allison Q. Byrne^3,4^, Mark Q. Wilber^5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^3,4^, Rob L. Grasso^8^, Erica Bree Rosenblum^3,4^ (exact order to be determined)

### Affiliations:

^1^Sierra Nevada Aquatic Research Laboratory, University of California; Mammoth Lakes, CA 93546, USA

^2^Earth Research Institute, University of California; Santa Barbara, CA, 93106, USA

^3^Department of Environmental Science, Policy, and Management, University of California; Berkeley, CA 94720, USA

^4^Museum of Vertebrate Zoology, University of California; Berkeley, CA 94720, USA

^5^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture; Knoxville, TN 37996, USA

^6^Earth Lab, University of Colorado; Boulder, CO 80303, USA

^7^NCX; San Francisco, CA, XXXXX, USA

^8^Resources Management and Science, Yosemite National Park; El Portal, CA 95318, USA

\*Corresponding author: roland.knapp\@ucsb.edu

### Abstract:

### One Sentence Summary

### Main Text:

### References and Notes:

### Acknowledgements:

\#### Funding

\#### Author contributions:

\#### Competing interests:

\#### Data and materials availability

### Supplementary Materials:

### Fig. #.

### Table #.

### Notes (author use only - not part of manuscript)

Science:

-   Research articles

    -   Present a major advance.

    -   Up to \~4500 words, including references, notes and captions.
        Corresponds to approximately 5 printed pages in journal.

    -   Up to 6 figures or tables, and about 40 references.

-   Reports

    -   New research results of broad significance.

    -   Up to \~2500 words, including references, notes and captions.
        Corresponds to approximately 3 printed pages in journal.

    -   Up to 4 figures or tables, and about 30 references.

Nature

-   Articles

    -   Original reports whose conclusions represent a substantial advance in understanding of an important problem and have immediate, far-reaching implications.

    -   Biological sciences papers do not normally exceed 8 pages.
        This would be approximately 4300 words and 5-6 modest display items.

PNAS

-   Research reports

    -   Original research of exceptional importance.

    -   Preferred length is 6 pages, but up to 12 pages is allowed.
        6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code.
You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-10-07 doc/manuscript/translocation.qmd
Add text to frame introduction

@@
 bibliography: translocation.bib
-reference-doc: translocation.docx
+reference-doc: translocation_style_ref.docx
 ---
@@
 ### Main Text:
+
+Severe effects of disease on global biodiversity (bats, primates, amphibians, corals, etc.)
+
+Host evolution and population persistence/recovery
+
+-   Tasmanian devils: Epstein et al. (2016)
+
+-   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020)
+
+-   Frogs in Panama: Voyles et al. (2018)
+
+-   Frogs in the Sierra Nevada: Knapp et al. (2016)
+
+Population reestablishment using reintroduction of resistant hosts
+
+-   Mendelson et al. (2019)
+
+-   Scheele et al. (2021)
+
+-   Joseph and Knapp (2018)
 

### 2022-10-10 doc/manuscript/translocation.qmd
Added some potential knowledge gaps to introduction framing

@@
 
-Severe effects of disease on global biodiversity (bats, primates, amphibians, corals, etc.)
+Severe effects of disease on global biodiversity (bats, primates, amphibians, corals, sea stars, etc.)
 
@@
 
--   Tasmanian devils: Epstein et al. (2016)
+-   Species recovery in the face of disease will depend the ability of a species to move or change.
 
--   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020)
+-   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)
+
+-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018).
+
+-   Tasmanian devils: Epstein et al. (2016), Wells et al., Ecology (2019)
+
+-   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., Biological Conservation (2014)
 
@@
 
+-   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
+
+-   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely [never?] been documented in wildlife populations.
+
 Population reestablishment using reintroduction of resistant hosts
+
+-   "Rescued" populations provide a source to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
 
@@
 -   Joseph and Knapp (2018)
+
+-   **Potential knowledge gap/opportunity/selling point**: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+A unique attribute of this study is that by combining repeated translocations of "rescued" populations, longitudinal surveys, mark-recapture analysis, and population modeling we demonstrate the capacity of "rescued" populations to persist across a landscape of disease.
+Thus, ER combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
 

### 2022-10-10 doc/manuscript/translocation.qmd
Add additional references and comments to framing

@@
 bibliography: translocation.bib
-reference-doc: translocation_style_ref.docx
 ---
@@
 
--   Species recovery in the face of disease will depend the ability of a species to move or change.
+-   Species recovery in the face of disease will depend on the ability of a species to move or change.
 
--   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)
+-   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021).
+    (Not sure if Allie wants to go the ER direction)
 
@@
 
--   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
+        PNAS (2018)
 
--   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely [never?] been documented in wildlife populations.
+-   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\]
+    been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
 
@@
     Thus, ER combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+
+[This is excellent. Thanks for adding it.]: #
 

### 2022-10-12 doc/manuscript/translocation.qmd
Add comment to framing text
Add field methods text, minor edits to framing text

@@
 
-    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., Biological Conservation (2014)
+    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., (2014)
 
@@
 -   **Potential knowledge gap/opportunity/selling point**: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
-    A unique attribute of this study is that by combining repeated translocations of "rescued" populations, longitudinal surveys, mark-recapture analysis, and population modeling we demonstrate the capacity of "rescued" populations to persist across a landscape of disease.
+    A unique attribute of this study is that we describe the genetic changes associated with evolutionary rescue, and combine repeated translocations of "rescued" populations, longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate the capacity of "rescued" populations to persist across a landscape of disease.
     Thus, ER combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+    These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
 
-[This is excellent. Thanks for adding it.]: #
+Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+#### Frog evolution in response to Bd
+
+#### Frog translocations
+
+#### Population viability modeling
+
+### Methods
+
+#### Frog evolution in response to Bd
+
+#### Frog translocations
+
+Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al. 2016).
+Results indicated that these populations contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
+Bd prevalence donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
+Recipient sites to which frogs were translocated were chosen based on previous *R.
+sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species (Knapp 2005).
+
+Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 13 recipient sites (Table S1).
+In each translocation, adult frogs ($\geq$ 40 mm snout-vent length) were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
+Frogs were transported to the recipient site either on foot or via helicopter.
+Following release, translocated populations were visited one to six times per summer (calculate min, max, mean) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
+CMR surveys allowed quantification of adult frog survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
+
+<!--# Create table of translocation sites, years, frog numbers. Could also create a map if necessary -->
+
+During 2006-2012, CMR surveys were conducted conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
+Frog were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
+During 2013-2022, we used a robust design in which all habitats were searched during several consecutive days (secondary periods - calculate min, max, mean), and frogs were processed as described above.
+Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and identified only during subsequent captures.
+Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods as described in Joseph and Knapp (2018).
+
+During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
+These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
+
+#### Frog evolution in response to Bd
+
+We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
+Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
+These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
+Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing.
+
+<!--# Mention declining populations also? -->
+
+DNA samples were collected via buccal swabbing of xx-xx frogs from eight Bd-naive populations and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
+
+<!--# Create map of sample sites -->
+
+#### Population viability modeling
 
@@
 
-    -   Up to \~4500 words, including references, notes and captions.
-        Corresponds to approximately 5 printed pages in journal.
+    -   Up to approximately 4500 words, including references, notes and captions.
+        Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
 
@@
     -   Up to \~2500 words, including references, notes and captions.
-        Corresponds to approximately 3 printed pages in journal.
+        Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
 

### 2022-10-13 doc/manuscript/translocation.qmd
Describe requirements for Nature Communications journal
Minor edits to methods text

@@
 
--   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021).
-    (Not sure if Allie wants to go the ER direction)
+-   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)
 
@@
 
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
-        PNAS (2018)
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
 
--   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\]
-    been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
+-   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
 
@@
 
-Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al. 2016).
-Results indicated that these populations contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
+Translocations of *R.
+sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al.
+2016).
+Results indicated that these populations contained several hundred *R.
+sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
 Bd prevalence donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
@@
 During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
-These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R.
+sierrae* tadpoles and subadults.
+These *R.
+sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
 
@@
 
-DNA samples were collected via buccal swabbing of xx-xx frogs from eight Bd-naive populations and seven recovering populations.
+DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size? --> and seven recovering populations.
 These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
@@
 
-    -   Up to approximately 4500 words, including references, notes and captions.
-        Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
+    -   Up to approximately 4500 words, including references, notes and captions. Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
 
@@
 
-    -   Up to \~2500 words, including references, notes and captions.
-        Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
+    -   Up to \~2500 words, including references, notes and captions. Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
 
@@
 
-    -   Biological sciences papers do not normally exceed 8 pages.
-        This would be approximately 4300 words and 5-6 modest display items.
+    -   Biological sciences papers do not normally exceed 8 pages. This would be approximately 4300 words and 5-6 modest display items.
 
@@
 
-    -   Preferred length is 6 pages, but up to 12 pages is allowed.
-        6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
+    -   Preferred length is 6 pages, but up to 12 pages is allowed. 6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
+
+Nature Communications
+
+-   Article
+
+    -   Papers published by the journal represent important advances of significance to specialists within each field.
+
+    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words. The maximum title length is 15 words.
 
@@
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code.
-You can embed code like this:
+When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
 

### 2022-10-13 code/translocation_survival_exploratory.Rmd
Fix typo

@@
 
-* Hypothesized mechanism: Severity of winter follow translocation could affect frog condition and subsequent survival of translocated frogs in the first year following translocation.
+* Hypothesized mechanism: Severity of winter following translocation could affect frog condition and subsequent survival of translocated frogs in the first year following translocation.
 For example, a long winter following translocation could reduce survival in the first year.

### 2022-10-13 code/translocation_survival_analysis.Rmd
Add text describing motivation for Model 2

@@
 # Model 2: Survival of frogs from first translocation as predictor of survival in subsequent translocations
+
+This analysis explores the results shown in translocation_survival_exploratory.Rmd (code chunk name = "frog-survival-by-site").
+This plot shows frog survival of all translocations grouped by site, and indicates that survival is highly variable across sites, but repeatable within a site.
+This analysis quantifies the association between survival in the first translocation and survival in subsequent translocations.
+
 ## Load packages

### 2022-10-13 doc/manuscript/translocation.qmd
Add additional details to journal summary
Add methods for population viability modeling

@@
 
-DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size? --> and seven recovering populations.
+DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size?
+--> and seven recovering populations.
 These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
@@
 #### Population viability modeling
+
+**Model description**
+
+To determine the implications of acquired Bd resistance on long-term population viability, we developed a dynamic, population model for MYLF.
+Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
+We developed a model that tracked seven state variables of an MYLF population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
+We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.
+
+We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year. The dynamics are given by
+
+$$
+\begin{bmatrix}
+L_1 \\
+L_2 \\
+L_3 \\
+J_1 \\
+J_2 \\
+A_R \\ 
+A_T
+\end{bmatrix}(t + 1) = 
+\begin{bmatrix}
+  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
+  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
+  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
+  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
+  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
+\end{bmatrix} \begin{bmatrix}
+L_1 \\
+L_2 \\
+L_3 \\
+J_1 \\
+J_2 \\
+A_R \\
+A_T \\
+\end{bmatrix}(t)
+$$
+
+The parameters in this model are yearly survival probability $\sigma_{*}$, where the "\*" indicates a particular state variable, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
+The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
+
+In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
+We also did not directly consider the dynamics of Bd in this model.
+We made this decision because i) all translocated populations are infected with Bd at high prevalence ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system.
+Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
+
+**Model analysis**
+
+**Incorporating yearly variability in vital rates**
+
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the mark-recapture model described elsewhere.
+While we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability.
+Thus, we did not include yearly within-population variability in adult survival in this analysis.
+
+However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability about a mean value.
+Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (more details below).
+
+**Estimating model parameters**
+
+The baseline parameter values for the model and how they were estimated are given in Table 1.
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our mark-recapture model described elsewhere.
+For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$.
+For populations where we had not pit-tagged any naturally recruited adults, we could not directly estimate $\sigma_{A_R}$ and set $\sigma_{A_R} = \sigma_{A_T}$.
+
+**Model analysis and simulation**
+
+We performed four analyses on our model.
+First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$ and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
+We then fixed $\omega = 0.1$ and calculated the predicted growth rate of our six populations.
+Second, we performed a local sensitivity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
+
+Third, we defined version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among year variability in $\omega$.
+We used this model to simulate a one time introduction of 40 translocated adult frogs.
+We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
+We varied the mean recruitment probability $\omega$ from 0.05 and 0.3 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+
+Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
+We focused on population 70550 because this was our longest mark-recapture time series for a translocated population and because this population shows evidence of recovery.
+We simulated our model for 16 years, repeating the simulation 50,000 times.
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$).
+Using Approximate Bayesian Computing and rejection sampling (Kosmala et al.
+2015, Ecology), we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
+The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution (Beaumont 2010).
+Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
 
@@
 
-Science:
+Science (impact factor = 63.714)
 
@@
 
-Nature
+Nature (impact factor = 69.8)
 
@@
 
-PNAS
+PNAS (impact factor = 12.77)
 
@@
 
-Nature Communications
+Nature Communications (impact factor = 17.694)
 
@@
 
-    -   Papers published by the journal represent important advances of significance to specialists within each field.
+    -   Papers published by the journal represent important advances of significance to specialists within each field. Open access.
 
     -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words. The maximum title length is 15 words.
+
+Science Advances (impact factor = 14.957)
+
+-   Publishes impactful research papers and reviews in any area of science.
+Journal extends the capacity of *Science* magazine to identify and promote significant advances in science and engineering.
+Open access.
+
+-   Research article
+
+    -   Should present a major advance.
+
+    -   Articles can be short of long format, up to 15,000 words, with up to 10 figures and/or tables.
 

### 2022-10-14 doc/manuscript/translocation.qmd
Quarto reformatting
Add manuscript formatting guidance

@@
 ## Title: Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
+
+<!--# 75 characters. Suggestions to make this more compelling? Use "Evolutionary rescue"? -->
 
@@
 
-### Abstract:
+### Abstract
 
-### One Sentence Summary
+<!--# No more than 200 words, follow structure described in instructions for authors.
+2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
+ and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
+See [annotated example](https://www.nature.com/documents/nature-summary-paragraph.pdf).
+-->
 
-### Main Text:
+### Main
+
+<!--# subheadings can be used, can be "teasers" of results (up to 40 characters) -->
 
@@
 
--   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)
+-   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021).
+    (Not sure if Allie wants to go the ER direction)
 
@@
 
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
+        PNAS (2018)
 
--   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
+-   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\]
+    been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
 
@@
 
+### Figures and figure legends
+
+<!--# In initial submission, include figure legends with their corresponding figures. -->
+
 ### Methods
+
+<!--# Typically less than 3000 words -->
 
@@
 
+<!--# subheadings can be used, up to 40 characters -->
+
+We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
+Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
+These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
+Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing.
+
+<!--# Mention sampled declining populations also? -->
+
+DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size? --> and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
+
+<!--# Create map of sample sites -->
+
 #### Frog translocations
 
-Translocations of *R.
-sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al.
-2016).
-Results indicated that these populations contained several hundred *R.
-sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
+Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al. 2016).
+Results indicated that these populations contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
 Bd prevalence donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
@@
 During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R.
-sierrae* tadpoles and subadults.
-These *R.
-sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
-
-#### Frog evolution in response to Bd
-
-We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
-Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
-These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing.
-
-<!--# Mention declining populations also? -->
-
-DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size?
---> and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
-
-<!--# Create map of sample sites -->
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
+These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
 
@@
 
-We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year. The dynamics are given by
+We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
+The dynamics are given by
 
@@
 
-### References and Notes:
+### References
 
-### Acknowledgements:
+### Acknowledgements
 
-\#### Funding
+### Author contributions
 
-\#### Author contributions:
+### Competing interests statement
 
-\#### Competing interests:
+### Additional information
 
-\#### Data and materials availability
+Supplementary Information is available for this paper.
 
-### Supplementary Materials:
+Correspondence and requests for materials should be addressed to Roland Knapp.
 
-### Fig. #.
+Peer review information.
 
-### Table #.
+Reprints and permissions information is available at www.nature.com/reprints.
+
+### Supplementary Information
 
@@
 
-    -   Up to approximately 4500 words, including references, notes and captions. Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
+    -   Up to approximately 4500 words, including references, notes and captions.
+        Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
 
@@
 
-    -   Up to \~2500 words, including references, notes and captions. Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
+    -   Up to \~2500 words, including references, notes and captions.
+        Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
 
@@
 
-    -   Biological sciences papers do not normally exceed 8 pages. This would be approximately 4300 words and 5-6 modest display items.
+    -   Biological sciences papers do not normally exceed 8 pages.
+        This would be approximately 4300 words and 5-6 modest display items.
 
@@
 
-    -   Preferred length is 6 pages, but up to 12 pages is allowed. 6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
+    -   Preferred length is 6 pages, but up to 12 pages is allowed.
+        6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
 
@@
 
-    -   Papers published by the journal represent important advances of significance to specialists within each field. Open access.
+    -   Papers published by the journal represent important advances of significance to specialists within each field.
+        Open access.
 
-    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words. The maximum title length is 15 words.
+    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words.
+        The maximum title length is 15 words.
 
@@
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+When you click the **Render** button a document will be generated that includes both content and the output of embedded code.
+You can embed code like this:
 

### 2022-10-14 doc/manuscript/translocation.qmd
Format manuscript, add .csl and .bib files

@@
 editor: visual
-csl: science.csl
+csl: nature.csl
 bibliography: translocation.bib
@@
 
-Roland A.
-Knapp^1,2^\*, Allison Q.
-Byrne^3,4^, Mark Q.
-Wilber^5^, Maxwell B.
-Joseph^6,7^, Thomas C.
-Smith^1,2^, Andrew P.
-Rothstein^3,4^, Rob L.
-Grasso^8^, Erica Bree Rosenblum^3,4^ (exact order to be determined)
+Roland A. Knapp^1,2^\*, Allison Q. Byrne^3,4^, Mark Q. Wilber^5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^3,4^, Rob L. Grasso^8^, Erica Bree Rosenblum^3,4^
+
+<!--# Exact author order to be determined) -->
 
@@
 
-^1^Sierra Nevada Aquatic Research Laboratory, University of California; Mammoth Lakes, CA 93546, USA
+^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, USA
 
-^2^Earth Research Institute, University of California; Santa Barbara, CA, 93106, USA
+^2^Earth Research Institute, University of California, Santa Barbara, CA, USA
 
-^3^Department of Environmental Science, Policy, and Management, University of California; Berkeley, CA 94720, USA
+^3^Department of Environmental Science, Policy, and Management, University of California.
+Berkeley, Berkeley, CA, USA
 
-^4^Museum of Vertebrate Zoology, University of California; Berkeley, CA 94720, USA
+^4^Museum of Vertebrate Zoology, University of California, Berkeley, Berkeley, CA, USA
 
-^5^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture; Knoxville, TN 37996, USA
+^5^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture, Knoxville, TN, USA
 
-^6^Earth Lab, University of Colorado; Boulder, CO 80303, USA
+^6^Earth Lab, University of Colorado, Boulder, CO, USA
 
-^7^NCX; San Francisco, CA, XXXXX, USA
+^7^NCX, San Francisco, CA, USA
 
-^8^Resources Management and Science, Yosemite National Park; El Portal, CA 95318, USA
+^8^Resources Management and Science, Yosemite National Park, El Portal, CA , USA
 
@@
 
+```{=html}
 <!--# No more than 200 words, follow structure described in instructions for authors.
@@
 -->
-
+```
 ### Main
@@
 
-    -   Frogs in the Sierra Nevada: Knapp et al. (2016)
+    -   Frogs in the Sierra Nevada: Knapp et al. (2016) @knapp2016
 
@@
 <!--# In initial submission, include figure legends with their corresponding figures. -->
+
+### References
+
+::: {#refs}
+:::
 
@@
 
+### Data availability
+
 ### References
@@
 
-### Competing interests statement
+### Competing interests
+
+The authors declare no competing interests.
 

### 2022-10-18 code/translocation_survival_analysis.Rmd
Update code for simulating predictive models

@@
 
-The goal of the analyses in this notebook is to identify predictors of frog survival following translocation. The analyses are conducted in a Bayesian framework using the R package rstanarm.
+The goal of the analyses in this notebook is to identify predictors of frog survival following translocation.
+The analyses are conducted in a Bayesian framework using the R package rstanarm.
 
-# Model 1: Site-level and frog-level predictors
+# Analysis 1: Site-level and frog-level predictors
 
@@
 ## Create frog translocation dataset for analysis
-* Read in data, create variables for analysis, standardize continuous variables. 
-* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+-   Read in data, create variables for analysis, standardize continuous variables.
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
 ```{r read-translocation-data}
@@
 ## Model using non-standardized predictors
+
 ### Model specification
+
 ```{r}
@@
 ### Priors & chain diagnostics
+
 ```{r}
@@
 ```
+
 -   mcmc diagnostics all look good.
@@
 ### List model variable names
+
 ```{r}
@@
 ### Posterior predictive check
+
 ```{r}
@@
 ```
+
 -   Posterior predictive distribution implied by regression model closely matches original data (based on mean frog survival).
@@
 ### Calculate leave-one-out cross-validation (loo)
+
 ```{r}
@@
 ### Posterior analysis
+
 ```{r}
@@
 ```
-- Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.  
+
+-   Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.\
 -   Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
@@
 ## Model using standardized predictors
+
 ### Model specification
+
 ```{r}
@@
 ### Priors, chain diagnostics, posterior predictive check
+
 ```{r}
@@
 mcmc_trace(m1b, size = 0.1)
-mcmc_dens_overlay(m1a)
+mcmc_dens_overlay(m1b)
 
@@
 ### Calculate loo
+
 ```{r}
@@
 ```
+
 -   As expected, using standardized predictors allows easy comparison across predictors. Important predictors are unchanged.
@@
 ## Compare models
+
 ```{r}
@@
 ```
+
 -   As expected, models with non-standardized and standardized predictor variables have identical predictive accuracy.
@@
 ## Possible grouping structures to add to model
+
 The dataset has three groups: site_id, translocation_id, and donor site.
@@
 ## Model using standardized predictors and site_id grouping variable
+
 ### Model specification
+
 ```{r}
@@
 ```
+
 -   Based on 90% uncertainty intervals, important predictors reduced to elevation and male.
@@
 ## Model using standardized predictors and translocation_id grouping variable)
+
 ### Model specification
+
 ```{r}
@@
 ```
+
 -   Based on loo, model with translocation_id as grouping variable provides slightly better fit than model with site_id as grouping variable.
@@
 ## Model using standardized predictors and 2 grouping variables: translocation_id nested within site_id)
+
 ### Model specification
+
 ```{r}
@@
 ## Compare 4 models with different grouping structures
+
 ```{r}
@@
 ```
+
 -   Based on expected log-predictive densities (ELPD), 3 models with grouping variables all have substantially higher posterior predictive accuracy than model without any grouping variables (elpd_diff of m1b is more than 2 SE lower than elpd_diff of m1c: elpd_diff = -18.2 $\pm$ 12.6, 2 SE interval does not overlap 0: -5.6 -- -30.8).
-- 3 models with 1 or more grouping variables are equivalent and have similar posterior predictive accuracy (elpd_diff of m1e and m1d are within 2 SE of elpd_diff of m1c).  
+-   3 models with 1 or more grouping variables are equivalent and have similar posterior predictive accuracy (elpd_diff of m1e and m1d are within 2 SE of elpd_diff of m1c).\
 -   m1d seems to provide more insight into the effect of the predictors, so chose this model for additional analysis.
@@
 ## Additional analysis of model m1d
+
 ### Posterior classification accuracy
+
 ```{r}
@@
 ```
-- Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.  
+
+-   Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.\
 -   Specificity or "true negative rate" measures the proportion of Y = 0 observations that are accurately classified (i.e., proportion of frogs correctly identified as not surviving).
@@
 ### Create reduced model containing only important predictors for subsequent analyses
+
 -   Reduced model is helpful because it reduces the number of values used to calculate predictive models.
+
 ```{r}
@@
 ### Confirm fit of reduced model is comparable to full model
+
 ```{r}
@@
 ### Simulate and plot posterior predictive models & observations
+#### Reduced model with translocation_id as grouping variable
 ```{r}
@@
 
-predictions_m1d <- posterior_predict(m1d_reduce, newdata = m1d_group_means)
+predictions_m1d <- posterior_epred(m1d_reduce, newdata = m1d_group_means)
+# predictions_m1d <- posterior_predict(m1d_reduce, newdata = m1d_group_means) # original code that produced incorrect uncertainty estimates
 
@@
   yrep = predictions_m1d, 
-  prob = 0.5, 
-  prob_outer = 0.8) +
+  prob = 0.8, 
+  prob_outer = 0.9) +
 ggplot2::scale_x_continuous(
@@
 ```
-- Uncertainty intervals appear incorrect for binomial model, but not sure how to adjust.  
-- Based on mean values, model does an adequate job of predicting survival based on group-level and individual-level characteristics.  
+
+-   Model does a good job of predicting survival based on group-level and individual-level characteristics.\
 -   What explains the low predicted survival at 70370 (both translocations), and high versus low predicted survival at 70550 for 1st and 2nd translocations, respectively?
-Both sites seem to contain high quality habitat (i.e., high elevation).
+Both sites seem to contain high quality habitat (i.e., high elevation).\
 -   The predicted survival is based on assumption that only males are included. Run another set of models for females?
 
+#### Reduced model without grouping variable
+To explore how addition of grouping variable affected predictions, especially that for 70370 and 70550. 
+```{r}
+m1d_reduce_nogroup <- stan_glm(
+  surv ~ snowt1_z + elev_z + male_c + donor,
+  data = d1,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+
+m1d_group_means <- d1 %>% 
+  group_by(translocation_id) %>% 
+  summarize(count = n(), psurv = mean(surv), elev_z = mean(elev_z), snowt1_z = mean(snowt1_z)) %>% 
+  mutate(
+    male_c = as.factor(0.548), 
+    donor = case_when(
+      str_detect(translocation_id, "74976") | str_detect(translocation_id, "70556") ~ "70459",
+      str_detect(translocation_id, "70413") ~ "70567",
+      TRUE ~ "72996")) %>% 
+  arrange(psurv)  # to arrange plot values/labels by psurv
+
+predictions_m1d_nogroup <- posterior_epred(m1d_reduce_nogroup, newdata = m1d_group_means)
+
+ppc_intervals(
+  y = m1d_group_means$psurv,
+  yrep = predictions_m1d_nogroup, 
+  prob = 0.8, 
+  prob_outer = 0.9) +
+ggplot2::scale_x_continuous(
+  labels = m1d_group_means$translocation_id,
+  breaks = 1:nrow(m1d_group_means)) +
+  xaxis_text(angle = 90, vjust = 0.5) +
+xlab("Translocation_id")
+```
+
 ### Plot posterior plausible relationships between each predictor and the probability of frog survival (conditional effects)
-- As described in https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on.
+
+-   As described in <https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html>, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on.
 By default, the mean is used for continuous variables and the reference category is used for factors, but you may change these values via argument conditions.
-This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."
-- Using rstanarm, I was unable to do this as I expected to: Start with data = d1 and using "newdata = " to specify that predictions should be based on data in d1_elevz (for example), in which non-plotted continuous predictors are set to the mean value.
-Not sure if my alternative approach is correct.
+This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."\
+-   Using rstanarm, I was unable to do this as I expected to: Start with data = d1 and using "newdata =" to specify that predictions should be based on data in d1_elevz (for example), in which non-plotted continuous predictors are set to the mean value.
+Not sure if my alternative approach is correct.\
 -   Approach used: Instead of starting the routine with d1, I started with the new dataset in which all variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).
@@
 #### Effect of elevation
+
 ```{r}
@@
 #### Effect of winter severity in year t1
+
 ```{r}
@@
 #### Effect of sex
+
 ```{r}
@@
 #### Effect of donor
+
 ```{r}
@@
 #### Joint effects of elevation and sex
+
 ```{r}
@@
 ## For comparison, fit model using brms and conditional_effects function to plot variable-specific effects
+
 ### Fit model
+
 ```{r}
@@
 ### Plot conditional effects
+
 ```{r}
@@
 
-# Model 2: Survival of frogs from first translocation as predictor of survival in subsequent translocations
+# Analysis 2: Survival of frogs from first translocation as predictor of survival in subsequent translocations
 
 This analysis explores the results shown in translocation_survival_exploratory.Rmd (code chunk name = "frog-survival-by-site").
-This plot shows frog survival of all translocations grouped by site, and indicates that survival is highly variable across sites, but repeatable within a site.
+This plot shows frog survival of all translocations grouped by site, and indicates that survival is highly variable across sites, but repeatable within sites.
 This analysis quantifies the association between survival in the first translocation and survival in subsequent translocations.
@@
 ## Create frog translocation dataset for analysis
-* Read in data, create variables for analysis, standardize continuous variables. 
-* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+-   Read in data, create variables for analysis, standardize continuous variables.
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
 ```{r}
@@
 ## Restructure dataset
+
 ```{r}
@@
 ## Model with non-standardized predictor, no grouping variable
+
 ### Model specification
+
 ```{r}
@@
 ## Model with standardized predictor, no grouping variable
+
 ### Model specification
+
 ```{r}
@@
 ## Model with standardized predictor, translocation_id grouping variable
+
 ### Model specification
+
 ```{r}
@@
 ## Model with standardized predictor, site_id grouping variable
+
 ### Model specification
+
 ```{r}
@@
 ### Compare 3 models with different grouping structures
+
 ```{r}
@@
 ```
-- Based on expected log-predictive densities (ELPD), models with and without grouping variables have similar posterior predictive accuracy (elpd_diff of m2b is less than 2 SE lower than elpd_diff of m2d: elpd_diff = -3.1 $\pm$ 5.8, 2 SE interval overlaps 0: 2.7 -- -8.9).
+
+-   Based on expected log-predictive densities (ELPD), models with and without grouping variables have similar posterior predictive accuracy (elpd_diff of m2b is less than 2 SE lower than elpd_diff of m2d: elpd_diff = -3.1 $\pm$ 5.8, 2 SE interval overlaps 0: 2.7 -- -8.9).\
 -   Conclusion from all 3 models is the same: Average frog survival of first translocation is an important predictor of survival of individual frogs in subsequent translocations.
@@
 ## Posterior classification accuracy of best models
+
 ```{r}
@@
 ```
+
 -   As expected from loo results, models with and without grouping variable have the same posterior predictive accuracy.
@@
 ## TO DO
-- Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).  
-- Check approach to plotting conditional effects.  
+
+-   Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).\
+-   Check approach to plotting conditional effects.\
 -   Try un-standardizing predictor values in conditional-effects plots (and other plots?) to produce more intuitive visualizations.

### 2022-10-23 doc/manuscript/translocation.qmd
Undo hard wrap in manuscript
Add a white-nose reference for the introduction
Add link to example Nature article
Add results and figure for population viability analysis

@@
 
-^3^Department of Environmental Science, Policy, and Management, University of California.
-Berkeley, Berkeley, CA, USA
+^3^Department of Environmental Science, Policy, and Management, University of California. Berkeley, Berkeley, CA, USA
 
@@
 
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
-        PNAS (2018)
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
 
--   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\]
-    been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
+    -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.
+
+-   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
 
@@
 
+Given observed yearly adult survival probabilities, our stage-structured model predicted that six of our twelve translocated population should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (Fig.
+XA; Fig.
+SXA; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
+When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (Fig.
+XA).
+In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our modeled predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult lifestage was near 0.4 (Fig.
+XA).
+A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (Fig.
+SXA).
+
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50 year extinction probabilities of less 1 (Fig.
+XB) -- in other words, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
+For these six populations, as $\omega$ increased to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (Fig.
+XB).
+In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (Fig.
+SXB; population 70550, surveyed for 16 years).
+
 ### Figures and figure legends
@@
 <!--# In initial submission, include figure legends with their corresponding figures. -->
+
+<!--# Discuss referencing and in-text generation of figures.
+We don't want to re-run the population viability modeling figures each time we compile the manuscript.
+A makefile could be a good solution.
+-->
+
+![Figure X: A.
+The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability or recruitment $\omega$, given the parameterized, deterministic model for translocated $R.
+sierrae$.
+Color points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment to adults is $\omega = 0.1$ (given by the dashed line).
+B.
+The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$ and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.1$, for comparison to A.](figures/pop_viability_figures_for_manuscript.jpg)
 
@@
 
-We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
-The dynamics are given by
+We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year. The dynamics are given by
 
@@
 
-    -   Biological sciences papers do not normally exceed 8 pages.
-        This would be approximately 4300 words and 5-6 modest display items.
+    -   Biological sciences papers do not normally exceed 8 pages. This would be approximately 4300 words and 5-6 modest display items.
+
+    -   This [article](https://www.nature.com/articles/s41586-022-04824-9) provides a potentially useful (e.g., recent) example of formatting and structure for a paper focused on ecology/evolution
 
@@
 
-    -   Preferred length is 6 pages, but up to 12 pages is allowed.
-        6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
+    -   Preferred length is 6 pages, but up to 12 pages is allowed. 6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
 
@@
 
-    -   Papers published by the journal represent important advances of significance to specialists within each field.
-        Open access.
+    -   Papers published by the journal represent important advances of significance to specialists within each field. Open access.
 
-    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words.
-        The maximum title length is 15 words.
+    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words. The maximum title length is 15 words.
 

### 2022-10-23 doc/manuscript/translocation.qmd
Fix #17, change MYLF to R. sierrae

@@
 
-To determine the implications of acquired Bd resistance on long-term population viability, we developed a dynamic, population model for MYLF.
+To determine the implications of acquired Bd resistance on long-term population viability, we developed a dynamic, population model for *R.
+sierrae*.
 Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of an MYLF population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
+We developed a model that tracked seven state variables of a *R.
+sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
 We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.

### 2022-10-24 code/translocation_survival_analysis.Rmd
Address #10, demonstrate how to unstandardize effect plots

@@
 
+Demonstrate how to unstandardize effect of elevation. This can be applied to all other effect plots.
+
+```{r}
+d1_elevz <- d1 %>% 
+  mutate(snowt1_z = mean(snowt1_z), 
+         male_c = recode(male_c, "-0.452" = "0.548"),
+         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
+
+d1_elevz %>% distinct(snowt1_z, male_c, donor) # ensure variables were set to correct values
+
+### Try unstandardizing two ways and confirm both give the same answer
+
+# First: Just replace elev_z with elevation
+set.seed(12) # set seed for comparison between the two
+d1_elevz %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv, elevation) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = elevation, y = surv)) +
+    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+#    geom_point(aes(y = .epred)) +
+  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
+
+# Second: Untransform elev_z based on mean and sd
+set.seed(12) # Same seed for comparison
+
+# Calculate mean and sd for elevation covariate
+mean_elev = mean(d1$elevation)
+sd_elev = sd(d1$elevation)
+
+d1_elevz %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv, elevation) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  ggplot(aes(x = (elev_z * 2*sd_elev) + mean_elev, y = surv)) +
+    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+#    geom_point(aes(y = .epred)) +
+  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
+
+```
+
+These two approaches give the same answer and can be applied to the code below.
+
 #### Effect of winter severity in year t1

### 2022-10-25 code/translocation_survival_analysis.Rmd
Address #4, provide an example of a flat prior on lines 90-110

@@
 
+## Explore a flat prior
+
+We can explore a flat prior on the intercept using the code below.
+
+```{r}
+m1_flat <- stan_glm(
+  surv ~ length + snow_t + snow_t1 + day + bdload_l + elevation + first + male + donor,
+  data = d1,
+  family = binomial,
+  prior_intercept=NULL,
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+
+# Confirm our prior is flat
+prior_summary(m1_flat)
+
+# Notice that the estimate of the intercept is unchanged between the two models
+m1_flat
+m1a
+```
+
 ### Priors & chain diagnostics

### 2022-10-27 code/translocation_survival_createdata.Rmd
Add column descriptions

@@
 ```
+
+### Description of dataset columns
+- site_id: Unique 5-digit site identifier.
+- elevation: Elevation of site above sea level (in meters). Obtained from 7.5' USGS topographic maps. 
+- shore: Complex shoreline habitats (boulders, overhanging banks, rock crevices) available to frogs during winter, or not present and/or unavailable due to reduction in water level (1 or 0, respectively).
+- snow_t: April 1 snow water equivalent (SWE) in year of translocation, measured at Dana Meadows (DAN station).
+- snow_t1: April 1 snow water equivalent (SWE) in year following translocation, measured at Dana Meadows (DAN station).
+- date: Date of translocation (yyyy-mm-dd).
+- year: Year of translocation.
+- day: Day of year on which translocation occurred, expressed as number of days since 01-January.  
+- order: Order of translocation at the site. 1 = first translocation, 2 = second translocation, etc.
+- pit_tag_ref: Unique 9 or 15-digit PIT tag identifier.  
+- donor: Unique 5-digit identifier of donor site at which frogs were collected. 
+- sex: Sex of translocated frog. 
+- length: Snout-vent length of translocated frog, measured at donor site at time of capture and $\leq$ 24 hours before translocation (in millimeters).
+- weight: Weight of translocated frog, measured at donor site at time of capture and $\leq$ 24 hours before translocation (in grams).
+- bd_load: Bd infection intensity, measured at donor site at time of capture and $\leq$ 24 hours before translocation (ITS1 copies * 80).
+- survival: Probability of surviving one year after translocation.

### 2022-10-29 code/translocation_survival_analysis_suppl.Rmd
Add dataset creation for analysis of bd load changes

@@
 - Inclusion of all records (i.e., 12 records where bd_load = NA that were excluded in primary analysis) did not change the results.
+
+## Analysis of changes in Bd load following translocation
+
+-  Rescale bd_load, center period variable. 
+-  Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years > 1.
+-  Possible models: 
+  -  bd_load ~ period + (1 | pit_tag_ref) + (1 | translocation_id)
+  -  bd_load ~ period + (1 | pit_tag_ref) + (1 | site_id)
+
+### Load additional libraries
+```{r}
+library(RPostgreSQL)
+library(lubridate)
+```
+
+### Read in raw translocation and capture datasets, bind them
+```{r read-data}
+translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv")) %>% 
+  relocate(visit_date = release_date, .before = type) %>% 
+  rename(survey_type = type) %>% 
+  mutate(pit_tag_ref = replace(pit_tag_ref, pit_tag_ref == "900067000117396" & site_id == 70641, "111111111111111")) %>% # remove incorrect pit_tag_ref identified in translocation_survival_createdata.Rmd
+  filter(pit_tag_ref != "111111111111111")
+
+capture <- read_csv(here::here("data", "raw", "frog_captures.csv")) %>% 
+  select(-capture_life_stage, -capture_animal_state, -sex)
+
+# select only translocated frogs (exclude naturally-recruited frogs and translocated frogs that were never recaptured after translocation)
+capture <- translocation %>% 
+  select(pit_tag_ref) %>% 
+  inner_join(capture, by = c("pit_tag_ref")) %>% 
+  relocate(pit_tag_ref, .after = last_col())
+
+frogs_all <- bind_rows(translocation, capture)
+```
+
+### Connect to PostgreSQL database `amphibians`
+```{r}
+con = dbConnect(dbDriver("PostgreSQL"),
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password"),
+                host = "frogdb.eri.ucsb.edu",
+                port = 5432,
+                dbname = "amphibians")
+
+```
+
+### Retrieve Bd data for translocated frogs from database
+```{sql retrieve-translocation-bd-data, connection=con, output.var="translocate_bd"}
+select
+  release_date as visit_date,
+  pit_tag_ref,
+  swab_id,
+  bd_load
+from relocate
+inner join relocate_frog on relocate.id = relocate_id
+inner join bd_load on swab_id = sample_id  
+where release_siteid1 in (70134, 70279, 70370, 70413, 70414, 70449, 70505, 70550, 70556, 70619, 70628, 70641, 74976) and
+  (replicate = 1 or replicate is null)
+```
+
+### Retrieve Bd data for captured frogs from database
+```{sql retrieve-capture-bd-data, connection=con, output.var="capture_bd"}
+select
+	visit_date,
+	pit_tag_ref, 
+	swab_id,
+	bd_load
+from visit
+inner join survey on visit.id = visit_id
+inner join capture_survey on survey.id = survey_id
+inner join bd_load on swab_id = sample_id  
+where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
+	survey_type = 'cmr' and
+	capture_animal_state != 'dead' and
+	(replicate = 1 or replicate is null)
+```
+
+### Disconnect from database
+```{r}
+dbDisconnect(con) 
+rm("con")
+```
+
+### Combine Bd data from translocate and capture datasets
+```{r}
+bd_add <- bind_rows(translocate_bd, capture_bd)
+```
+
+### Add Bd data to frog data, restructure dataset
+```{r}
+frogsbd <- frogs_all %>% 
+  inner_join(bd_add, by = c("visit_date", "pit_tag_ref")) %>% 
+  mutate(period = if_else(survey_type == "cmr", "after", "before")) %>% # use 0/1 instead?
+  relocate(period, .after = survey_type) %>% 
+  arrange(site_id, visit_date)
+```
+
+### Add years since translocation
+```{r}
+frogsbd <- translocation %>% 
+  mutate(year_translocate = year(visit_date)) %>% 
+  select(year_translocate, pit_tag_ref) %>% 
+  inner_join(frogsbd, by = c("pit_tag_ref")) %>% 
+  relocate(year_translocate, .after = visit_date) %>% 
+  relocate(pit_tag_ref, .after = period) %>% 
+  mutate(year_capture = year(visit_date), .after = year_translocate) %>% 
+  mutate(interval = (year_capture - year_translocate), .after = year_capture) %>% 
+  arrange(site_id, visit_date, pit_tag_ref)
+```
+
+### Update site ids to standardize usage
+```{r}
+frogsbd <- frogsbd %>% 
+  mutate(site_id1 = site_id, .after = site_id,
+         site_id1 = replace(site_id1, site_id == 70034, 70449),
+         site_id1 = replace(site_id1, site_id == 70175, 70413),
+         site_id1 = replace(site_id1, site_id == 70279, 70413),
+         site_id1 = replace(site_id1, site_id == 72093, 70413),
+         site_id1 = replace(site_id1, site_id == 72442, 70413),
+         site_id1 = replace(site_id1, site_id == 70688, 70134),
+         site_id1 = replace(site_id1, site_id == 72092, 70505),
+         site_id1 = replace(site_id1, site_id == 72973, 74976))
+```
+
+### Add translocation_id
+```{r}
+frogsbd <- frogsbd %>% 
+  distinct(site_id1, year_translocate) %>% 
+  arrange(site_id1, year_translocate) %>% 
+  group_by(site_id1) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id1) %>% 
+  unite(translocation_id, c("site_id1", "transno"), remove = FALSE) %>% 
+  select(-transno) %>% 
+  inner_join(frogsbd, by = c("site_id1", "year_translocate")) %>% 
+  select(site_id, site_id1, translocation_id, visit_date, survey_type, period, year_translocate, year_capture, interval, pit_tag_ref, swab_id, bd_load) %>% 
+  arrange(site_id1, visit_date)
+```
+
+### Save frog-Bd dataset
+```{r}
+frogsbd %>% write_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv"))
+```

### 2022-11-01 code/translocation_survival_analysis_suppl.Rmd
Add analysis code

@@
 ## Analysis of changes in Bd load following translocation
+-  Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years > 1. Only frogs that were recaptured at least once following translocation are included.
 
--  Rescale bd_load, center period variable. 
--  Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years > 1.
--  Possible models: 
-  -  bd_load ~ period + (1 | pit_tag_ref) + (1 | translocation_id)
-  -  bd_load ~ period + (1 | pit_tag_ref) + (1 | site_id)
+### Create dataset
 
@@
 ```
+
+## Analyze dataset
+### Read in data, standardize variables
+```{r}
+d1 <- read_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv")) %>% 
+  mutate(after = if_else(period == "after", 1, 0),
+         bd_load = as.integer(round(bd_load)),  # if using negative binomial distribution
+         after_c = after - mean(after))
+```
+
+### Null values - check
+```{r check-null}
+d1 %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+### Include only CMR years 0 and 1
+```{r}
+d2 <- d1 %>% 
+  filter(interval < 2)
+```
+
+### Plot distribution of bd_load response variable
+```{r}
+d2 %>% 
+  ggplot(aes(x = bd_load)) +
+    geom_histogram()
+
+d2 %>% 
+  ggplot(aes(sample = bd_load)) +
+    geom_qq() +
+    geom_qq_line()
+```
+* Data are strongly left-skewed, possibly zero-inflated. Model using negative binomial family.
+
+### Model without group-level effects
+#### Model specification
+```{r}
+m1a <- stan_glm(
+  bd_load ~ after_c,
+  data = d2,
+  family = neg_binomial_2(),
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+#### Priors and chain diagnostics
+```{r}
+prior_summary(m1a)
+
+mcmc_trace(m1a, size = 0.1)
+mcmc_dens_overlay(m1a)
+
+summary(m1a)
+```
+
+#### Posterior predictive check
+```{r}
+pp_check(m1a, plotfun = "stat") +
+    xlab("Bd load")
+```
+
+-   Posterior predictive distribution implied by regression model closely matches original data (based on mean Bd load).
+
+#### Calculate loo
+```{r}
+loo_m1a <- loo(m1a, cores = 4, save_psis = TRUE)
+print(loo_m1a)
+```
+
+#### Posterior analysis
+```{r}
+tidy(m1a, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1a, area_method = "equal height", prob = 0.90) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors, no group-level effects",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+# mean(bayes_R2(m1a))   # bayes_R2 not available for negative binomials models in stanreg. 
+```
+
+### Model with 1 grouping variable: pit_tag_ref
+#### Model specification
+```{r}
+m1b <- stan_glmer(
+  bd_load ~ after_c + (1 | pit_tag_ref),
+  data = d2,
+  family = neg_binomial_2(),
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+#### List model variable names
+
+```{r}
+get_variables(m1b)
+```
+
+#### Priors and chain diagnostics
+```{r}
+prior_summary(m1b)
+
+b1 <- c("(Intercept)", "after_c")
+
+mcmc_trace(m1b, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1b, pars = b1)
+
+summary(m1b)
+```
+
+#### Posterior predictive check
+```{r}
+pp_check(m1b, plotfun = "stat") +
+    xlab("Bd load")
+```
+
+-   Posterior predictive distribution implied by regression model shows some mismatch between actual data and model results (based on mean Bd load).
+
+#### Calculate loo
+```{r}
+loo_m1b <- loo(m1b, cores = 4, save_psis = TRUE)
+print(loo_m1b)
+```
+
+#### Posterior analysis
+```{r}
+tidy(m1b, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1b, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with pit_tag_ref as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+# mean(bayes_R2(m1b))
+```
+
+### Model with 2 grouping variables: pit_tag_ref nested within translocation_id
+#### Model specification
+```{r}
+m1c <- stan_glmer(
+  bd_load ~ after_c + (1 | pit_tag_ref) + (1 | translocation_id),
+  data = d2,
+  family = neg_binomial_2(),
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+#### Priors and chain diagnostics
+```{r}
+prior_summary(m1c)
+
+b1 <- c("(Intercept)", "after_c")
+
+mcmc_trace(m1c, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1c, pars = b1)
+
+summary(m1c)
+```
+
+#### Posterior predictive check
+```{r}
+pp_check(m1c, plotfun = "stat") +
+    xlab("Bd load")
+```
+
+-   Posterior predictive distribution implied by regression model closely matches original data (based on mean Bd load).
+
+#### Calculate loo
+```{r}
+loo_m1c <- loo(m1c, cores = 4, save_psis = TRUE)
+print(loo_m1c)
+```
+
+#### Posterior analysis
+```{r}
+tidy(m1c, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1c, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with pit_tag_ref and translocation_id as grouping variables",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+# mean(bayes_R2(m1c))
+```
+
+### Model with 2 grouping variables: pit_tag_ref nested within site_id1
+#### Model specification
+```{r}
+m1d <- stan_glmer(
+  bd_load ~ after_c + (1 | pit_tag_ref) + (1 | site_id1),
+  data = d2,
+  family = neg_binomial_2(),
+  chains = 4, 
+  iter = 5000*2,
+  cores = 4,
+  seed = 84735)
+```
+
+#### Priors and chain diagnostics
+```{r}
+prior_summary(m1d)
+
+b1 <- c("(Intercept)", "after_c")
+
+mcmc_trace(m1d, size = 0.1, pars = b1)
+mcmc_dens_overlay(m1d, pars = b1)
+
+summary(m1d)
+```
+
+#### Posterior predictive check
+```{r}
+pp_check(m1d, plotfun = "stat") +
+    xlab("Bd load")
+```
+
+-   Posterior predictive distribution implied by regression model closely matches original data (based on mean Bd load).
+
+#### Calculate loo
+```{r}
+loo_m1d <- loo(m1d, cores = 4, save_psis = TRUE)
+print(loo_m1d)
+```
+
+#### Posterior analysis
+```{r}
+tidy(m1d, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1d, area_method = "equal height", prob = 0.90, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with pit_tag_ref and site_id1 as grouping variables",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+# mean(bayes_R2(m1d))
+```
+
+## Compare 4 models with different grouping structures
+```{r}
+loo_compare(loo_m1a, loo_m1b, loo_m1c, loo_m1d)
+```

### 2022-11-01 code/translocation_survival_analysis.Rmd
Minor code edit to ensure correct rounding

@@
     across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
-    surv = as.integer(surv))
+    surv = as.integer(round(surv)))
 

### 2022-11-01 code/translocation_survival_analysis_suppl.Rmd
Add description of dataset columns

@@
 
+### Description of dataset columns 
+-  site_id: Unique 5-digit site identifier.  
+-  site_id1: Assigned 5-digit site identifier for a collection of nearby sites utilized by a frog population.
+-  translocation_id: Unique identifier for each translocation.  
+-  visit_date: Date on which translocation/survey was conducted.  
+-  survey_type: "translocation" or "cmr".  
+-  period: "before" or "after" translocation.
+-  year_translocate: Year in which translocation was conducted.  
+-  year_capture: Year in which frog was captured.  
+-  interval: Number of years between translocation and frog capture.  
+-  pit_tag_ref: Unique PIT tag identifier.  
+-  swab_id: Unique identifier for skin swab sample.  
+-  bd_load: Bd infection intensity determined from skin swab sample, units are number of ITS1 copies * 80.
+
 ## Analyze dataset
@@
   mutate(after = if_else(period == "after", 1, 0),
-         bd_load = as.integer(round(bd_load)),  # if using negative binomial distribution
+         bd_load = as.integer(round(bd_load)), 
          after_c = after - mean(after))

### 2022-11-01 code/translocation_survival_analysis_suppl.Rmd
Fix typos

@@
 -  site_id: Unique 5-digit site identifier.  
--  site_id1: Assigned 5-digit site identifier for a collection of nearby sites utilized by a frog population.
+-  site_id1: Assigned 5-digit site identifier for a collection of adjacent sites that are utilized by a frog population.
 -  translocation_id: Unique identifier for each translocation.  
--  visit_date: Date on which translocation/survey was conducted.  
+-  visit_date: Date on which translocation or survey was conducted.  
 -  survey_type: "translocation" or "cmr".  
@@
 -  swab_id: Unique identifier for skin swab sample.  
--  bd_load: Bd infection intensity determined from skin swab sample, units are number of ITS1 copies * 80.
+-  bd_load: Bd infection intensity determined from skin swab samples, units are number of ITS1 copies * 80.
 

### 2022-11-01 code/translocation_survival_analysis.Rmd
Use unstandardized data in conditional effects plots

@@
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+  ggtitle("Model with standardized predictors & translocation_id/site_id as grouping variables",
           "Posterior distributions with medians & 90% uncertainty intervals")
@@
   seed = 84735)
+```
+
+### Posterior analysis
+```{r}
+tidy(m1d_reduce, conf.int = TRUE, conf.level = 0.90)
+
+b2 <- c("(Intercept)", "snowt1_z", "elev_z", "male_c0.548", "donor70567", "donor72996")
+
+mcmc_areas(m1d_reduce, area_method = "equal height", prob = 0.90, pars = b2) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Reduced model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
 ```
@@
 
--   Model does a good job of predicting survival based on group-level and individual-level characteristics.\
--   What explains the low predicted survival at 70370 (both translocations), and high versus low predicted survival at 70550 for 1st and 2nd translocations, respectively?
-Both sites seem to contain high quality habitat (i.e., high elevation).\
+-   Model does a good job of predicting survival based on group-level and individual-level characteristics.  
 -   The predicted survival is based on assumption that only males are included. Run another set of models for females?
@@
 
+### Values of important predictors for translocations of interest
+```{r}
+d1 %>% 
+  filter(site_id == "70370" | site_id == "70550") %>% 
+  distinct(translocation_id, elevation, snow_t1, donor)
+
+```
+
+-   This helps to explain the low predicted survival at 70370 (both translocations), and high versus low predicted survival at 70550 for 1st and 2nd translocations, respectively.  
+  -   Without translocation_id as group-level effect, survival from two translocations at 70550 are not predicted well and both are predicted to have similar (and moderate) frog survival (based on high elevation, low snow_t1, and donor = 72996 for both translocations).
+Therefore, the group-level effect is responsible for good prediction of survival at this site.
+  -   Low survival at 70370 in both translocations is predicted relatively well by models with and without group-level effect.
+Low predicted survival can be attributed to high snow_t1 and donor = 72996 in both translocations (although high elevation of site should have positive effect).
+
 ### Plot posterior plausible relationships between each predictor and the probability of frog survival (conditional effects)
@@
 
+# use untransformed predictor (elevation) to make plot easier to interpret  
 d1_elevz %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, elevation, male_c, donor, surv) %>% 
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = elev_z, y = surv)) +
+  ggplot(aes(x = elevation, y = surv)) +  #   "ggplot(aes(x = (elev_z * 2*sd_elev) + mean_elev, y = surv)) +" gives identical result
     geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
     # geom_point(aes(y = .epred)) +
-  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
-
-# d1 %>%  # Test of add_predicted_draws used for similar purpose - unsuccessful
-#   select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
-#   add_predicted_draws(m1d_reduce, ndraws = 100, re_formula = NA, newdata = d1_elevz) %>% View()
-#   ggplot(aes(x = elev_z, y = surv)) +
-#     geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
-#     geom_point(aes(y = .epred)) +
-#   labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
-```
-
-Demonstrate how to unstandardize effect of elevation. This can be applied to all other effect plots.
-
-```{r}
-d1_elevz <- d1 %>% 
-  mutate(snowt1_z = mean(snowt1_z), 
-         male_c = recode(male_c, "-0.452" = "0.548"),
-         donor = recode(donor, "70567" = "70459", "72996" = "70459"))
-
-d1_elevz %>% distinct(snowt1_z, male_c, donor) # ensure variables were set to correct values
-
-### Try unstandardizing two ways and confirm both give the same answer
-
-# First: Just replace elev_z with elevation
-set.seed(12) # set seed for comparison between the two
-d1_elevz %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv, elevation) %>% 
-  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = elevation, y = surv)) +
-    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
-#    geom_point(aes(y = .epred)) +
-  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
-
-# Second: Untransform elev_z based on mean and sd
-set.seed(12) # Same seed for comparison
-
-# Calculate mean and sd for elevation covariate
-mean_elev = mean(d1$elevation)
-sd_elev = sd(d1$elevation)
-
-d1_elevz %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv, elevation) %>% 
-  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = (elev_z * 2*sd_elev) + mean_elev, y = surv)) +
-    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
-#    geom_point(aes(y = .epred)) +
-  labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival")
-
-```
-
-These two approaches give the same answer and can be applied to the code below.
+  labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival")
+```
 
@@
 d1_snowt1z %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  select(site_id, translocation_id, snowt1_z, snow_t1, elev_z, male_c, donor, surv) %>% 
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = snowt1_z, y = surv)) +
+  ggplot(aes(x = snow_t1, y = surv)) +
     geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
 #    geom_point(aes(y = .epred)) +
-  labs(x = "Winter severity in year following translocation (standardized)", y = "Probability of 1-year frog survival") 
+  labs(x = "Winter severity in year following translocation (% of average snow depth)", y = "Probability of 1-year frog survival") 
 ```
@@
 d1_malec %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, male_c, sex, donor, surv) %>% 
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = male_c, y = surv)) +
-    geom_boxplot(aes(y = .epred, x = male_c)) +
+  ggplot(aes(x = male, y = surv)) +
+    geom_boxplot(aes(y = .epred, x = sex)) +
 #    geom_point(aes(y = .epred)) +
-    labs(x = "Sex (centered)", y = "Probability of 1-year frog survival")
+    labs(x = "Sex", y = "Probability of 1-year frog survival")
 ```
@@
 d1_elevz_malec %>% 
-  select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, elevation, male_c, sex, donor, surv) %>% 
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = elev_z, y = surv, color = male_c)) +
-    geom_line(aes(y = .epred, group = paste(male_c, .draw)), size = 0.4, alpha = 0.5) +
+  ggplot(aes(x = elevation, y = surv, color = sex)) +
+    geom_line(aes(y = .epred, group = paste(sex, .draw)), size = 0.4, alpha = 0.5) +
 #    geom_point(aes(y = .epred)) +
-    labs(x = "Elevation of recipient site (standardized)", y = "Probability of 1-year frog survival") +
+    labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
     scale_color_manual(values = pal) + 
@@
 This plot shows frog survival of all translocations grouped by site, and indicates that survival is highly variable across sites, but repeatable within sites.
-This analysis quantifies the association between survival in the first translocation and survival in subsequent translocations.
+The following analysis quantifies the association between survival in the first translocation and survival in subsequent translocations.
 

### 2022-11-02 doc/manuscript/translocation.qmd
Small edits to manuscript

@@
 
-We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
+We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
 Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
@@
 sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
-Bd prevalence donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
+Bd prevalence in donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 We also did not directly consider the dynamics of Bd in this model.
-We made this decision because i) all translocated populations are infected with Bd at high prevalence ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system.
+We made this decision because i) all translocated populations are infected with Bd at high prevalence ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system (Wilber et al.
+2022).
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.

### 2022-11-04 code/maps.qmd
Initial commit

---
title: "maps"
format: html
editor: visual
---
## Load packages
```{r}
library(sf)
library(tmap)
library(stars)
library(tidyverse)
```

## Map of sample sites for exome capture analysis
### Import data
```{r}
nps_boundary <- st_read("/home/roland/pCloudDrive/MLRG/gis/Coverages/NPS_boundary/yose_kica_sequ_boundaries.shp", stringsAsFactors=F) #read in shapefile like a dataframe using sf package
class(nps_boundary)
str(nps_boundary)
st_crs(nps_boundary) #determine CRS
utm11nCRS <- st_crs(nps_boundary)

# California Color Hillshade (90m) from koordinates.com
hillshade <- read_stars("/home/roland/Downloads/hillshade-90m-GTiff-UTMzone11/california-color-hillshade-90m.tif")
hillshade

```

### Check if shapefile polygons are valid
```{r}
st_is_valid(nps_boundary) # all 3 are invalid

nps_boundary <- st_make_valid(nps_boundary$geometry)
# tmap_options(check.and.fix = TRUE) # Don't know how to use this option, or whether it would fix the issue

st_is_valid(nps_boundary) # fixed
```

### Create map
```{r}
tm_shape(hillshade) + 
  tm_raster(style = "cont") + 
  tm_shape(nps_boundary) +
  tm_polygons()

tm_shape(nps_boundary) +
  tm_polygons()

```

### 2022-11-05 code/maps.qmd
Use tm_rgb to map hillshade raster

@@
 ---
+
 ## Load packages
+
 ```{r}
@@
 ## Map of sample sites for exome capture analysis
+
 ### Import data
+
 ```{r}
@@
 st_crs(nps_boundary) #determine CRS
-utm11nCRS <- st_crs(nps_boundary)
+#utm11nCRS <- st_crs(nps_boundary)
 
@@
 hillshade
+str_rgb
+st_crs(hillshade)
 
@@
 ### Check if shapefile polygons are valid
+
 ```{r}
@@
 
+### Set projection
+
+```{r}
+crs(hillshade) <- crs(nps_boundary)
+```
+
+### Crop raster with bounding box
+
+```{r}
+# as example
+b <- as(extent(6, 6.4, 49.75, 50), 'SpatialPolygons')
+crs(b) <- crs(r)
+rb <- crop(r, b)
+```
+
 ### Create map
+
 ```{r}
 tm_shape(hillshade) + 
-  tm_raster(style = "cont") + 
+  tm_rgb(r = 1, g = 2, b = 3) +
   tm_shape(nps_boundary) + 
-  tm_polygons()
+  tm_borders(col = "black")
 
-tm_shape(nps_boundary) +
-  tm_polygons()
+# Need to crop raster to smaller size: https://semba-blog.netlify.app/12/14/2020/manipulate-and-visualize-raster-data-with-leaflet-and-tmap-in-r/#crop-raster
+
+# Might be easier to re-export smaller shape
 
 ```
+
+Add study site points, north arrow, distance bar

### 2022-11-08 code/maps.qmd
Finish code for map of exome sampling points

--- 2022-11-04 code/maps.qmd
@@
 library(tidyverse)
+library(RPostgreSQL)
 ```
@@
 ## Map of sample sites for exome capture analysis
-### Import data
+
+Run the following 5 chunks only if you have access to the project database.
+For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data" chunk.
+
+### Create dataset of sample collection points
 ```{r}
-nps_boundary <- st_read("/home/roland/pCloudDrive/MLRG/gis/Coverages/NPS_boundary/yose_kica_sequ_boundaries.shp", stringsAsFactors=F) #read in shapefile like a dataframe using sf package
-class(nps_boundary)
-str(nps_boundary)
-st_crs(nps_boundary) #determine CRS
-utm11nCRS <- st_crs(nps_boundary)
+exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
+```
+#### Connect to database
+
+```{r}
+source("db_connect.R")
+```
+
+#### Retrieve data
+
+```{sql retrieve-site-xy, connection=con, output.var="exome_sites_db"}
+select
+  id as site_id,
+  utme,
+  utmn
+from site
+where id in (74281, 50731, 50307, 52127, 10220, 10490, 50837, 20169, 72996, 72808, 70567, 54188, 10055, 10593, 50783)
+```
+
+#### Disconnect from database
+
+```{r}
+source("db_disconnect.R")
+```
+
+#### Create points layer
+
+```{r}
+exome_sites_xy <- 
+  exome_sites %>% 
+  inner_join(exome_sites_db, by = "site_id")
+
+# Add x-y random noise to obscure exact site locations
+exome_sites_xy <- exome_sites_xy %>% 
+  add_column(x = sample(-2000:2000, size = 15),
+             y = sample(-2000:2000, size = 15))
+
+exome_sites_xy <- exome_sites_xy %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+
+# Save file 
+write_csv(exome_sites_xy, here::here("data", "maps", "exome_sites_xy.csv"))
+```
+
+### Import map layers
+
+```{r import-map-data}
+# Read file, if lacking database access
+exome_sites_xy <- read_csv(here::here("data", "maps", "exome_sites_xy.csv"))
+exome_points = st_as_sf(exome_sites_xy, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
+
+nps_boundaries <- st_read(here::here(
+  "data", "maps", "nps-boundaries", "yose_kica_sequ_boundaries.shp"))
+st_crs(nps_boundaries) #determine coordinate reference system (CRS)
 
 # California Color Hillshade (90m) from koordinates.com
-hillshade <- read_stars("/home/roland/Downloads/hillshade-90m-GTiff-UTMzone11/california-color-hillshade-90m.tif")
-hillshade
-
+hillshade <- read_stars(here::here(
+  "data", "maps", "hillshade-90m", "california-color-hillshade-90m.tif"))
+st_crs(hillshade)
 ```
 
-### Check if shapefile polygons are valid
+### Check & fix shapefile polygons
+
 ```{r}
-st_is_valid(nps_boundary) # all 3 are invalid
+st_is_valid(nps_boundaries) # all 3 are invalid
 
-nps_boundary <- st_make_valid(nps_boundary$geometry)
-# tmap_options(check.and.fix = TRUE) # Don't know how to use this option, or whether it would fix the issue
+nps_boundaries <- st_make_valid(nps_boundaries$geometry)
 
-st_is_valid(nps_boundary) # fixed
+st_is_valid(nps_boundaries) # fix polygons
+```
+
+### Transform CRS of nps_boundary
+
+```{r}
+nps_boundaries <- st_transform(nps_boundaries, 26911)
+```
+
+### Crop raster with bounding box
+
+```{r}
+bbox <- st_bbox(c(xmin = 238000, ymax = 4235000, xmax = 403000, ymin = 4011000), crs = st_crs(nps_boundaries))
+hillshade_crop <- st_crop(hillshade, bbox)
+hillshade_crop <- st_transform(hillshade_crop, 26911)
 ```
@@
 ### Create map
+
 ```{r}
-tm_shape(hillshade) + 
-  tm_raster(style = "cont") + 
-  tm_shape(nps_boundary) +
-  tm_polygons()
+exome_points_map <- tm_shape(hillshade_crop) + 
+  tm_rgb(r = 1, g = 2, b = 3) +
+  tm_shape(nps_boundaries) + 
+  tm_borders(col = "black") +
+  tm_shape(exome_points) + 
+  tm_symbols(size = 0.5, 
+             col = "bd_status", 
+             palette = c("#FFEC8B", "#87CEFA"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
+             title.col = "Bd status") +
+  tm_layout(legend.position = c("right", "top")) +
+  tm_compass(type = "4star",
+             position = c("left", "bottom"),
+             size = 2) +
+  tm_scale_bar(breaks = c(0, 25, 50), 
+               text.size = 0.7, 
+               position = c("left", "bottom"))
 
-tm_shape(nps_boundary) +
-  tm_polygons()
+exome_points_map
+```
 
+### Save map
+```{r}
+tmap_save(exome_points_map, here::here("out", "map_exome_points.png"))
 ```

### 2022-11-09 code/translocation_survival_exploratory.Rmd
Color-code translocations by donor site

@@
 
+-  Points colored by donor site.
+
 ```{r frog-survival-by-site}
+palette1 <- c("green4", "blue", "black")
+
+donors <- frog_translocation %>% 
+  distinct(site_id, donor)
+
 cohort_survival %>% 
@@
   inner_join(cohort_survival, by = "site_id") %>% 
+  inner_join(donors, by = "site_id") %>% 
   select(-plot_survival) %>% 
-    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival)) +  # reorder site_id by siteid_order)
-      geom_point(size = 3) +
+    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, group = donor)) +  # reorder site_id by siteid_order)
+      geom_point(aes(color = donor), size = 3) +
+      scale_color_manual(values = palette1) + 
       labs(x = "Recipient site", y = "1-year frog survival")

### 2022-11-09 code/translocation_survival_analysis.Rmd
Address #10, make effect plots with confidence bands

@@
   labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival")
+
+# Plot with confidence bands
+alpha = 1 - 0.95
+d1_elevz %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, elevation, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 1000, re_formula = NA) %>% 
+  group_by(elevation) %>%
+  summarise(lower=quantile(.epred, alpha), 
+            upper=quantile(.epred, 1 - alpha),  
+            med=quantile(.epred, 0.5)) %>%
+  ggplot(aes(x=elevation)) + 
+          geom_line(aes(y = med), color = "blue", size = 1, alpha = 0.8) +
+          geom_ribbon(aes(ymin=lower, ymax=upper), fill="blue", alpha=0.1) +
+          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
+          theme_classic()
 ```

### 2022-11-09 code/maps.qmd
Minor code edits

@@
 ### Create dataset of sample collection points
+
 ```{r}
@@
 ```
+
 #### Connect to database
@@
   tm_shape(exome_points) + 
-  tm_symbols(size = 0.5, 
+  tm_symbols(size = 0.5, # increase point size, use darker blue
              col = "bd_status", 
@@
 ### Save map
+
 ```{r}

### 2022-11-09 code/translocation_survival_analysis.Rmd
Minor code edits

@@
 ```{r}
-tidy(m1d, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m1d, area_method = "equal height", prob = 0.90, pars = b1) +
+tidy(m1d, conf.int = TRUE, conf.level = 0.95)
+
+mcmc_areas(m1d, area_method = "equal height", prob = 0.95, pars = b1) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictors & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 

### 2022-11-09 doc/manuscript/translocation.qmd
Add figures to translocation section

@@
 
+Results text will be added soon. Figure XX is likely to be included in the main manuscript, and Figure XXX will probably be moved to the Supplementary Information.
+
+![Figure XX.
+Average survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
+Sites are arranged along the x-axis using the average of the median survival per translocation.
+Frog survival for the one-year period following each transloation was estimated using capture-mark-recapture methods.
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png)
+
+![Figure XXX.
+For the best model predicting frog survival following translocations, 95% uncertainty intervals for the coefficients of each predictor variable.
+Uncertainty intervals are shown as shaded areas under the estimated posterior density curve.
+Posterior medians are indicated by the verical line in each curve.
+Important predictors of frog survival are those for which the uncertainty interval does not include zero (vertical red dashed line).
+Bd load at the time of translocation did not affect subsequent frog survival.](figures/mcmc_areas_m1d.png)
+
 #### Population viability modeling
@@
 We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
-Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
+Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering").
 These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
@@
 
-DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size?
---> and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
+DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure SXXXX), and the latitudinal range of the two population types was similar.
 
-<!--# Create map of sample sites -->
+![Figure SXXXX.
+Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia national parks (from top to bottom).](figures/map_exome_points.png)
 
@@
 sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
+
+<!--# Add modeling methods -->
 

### 2022-11-09 code/maps.qmd
Change point size and color

@@
   tm_shape(exome_points) + 
-  tm_symbols(size = 0.5, # increase point size, use darker blue
+  tm_symbols(size = 0.7, # increase point size, use darker blue
              col = "bd_status", 
-             palette = c("#FFEC8B", "#87CEFA"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
+             palette = c("#FFEC8B", "#00B2EE"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
              title.col = "Bd status") +

### 2022-11-23 code/maps.qmd
Add inset map to map of study area

@@
 library(stars)
+library(grid)
 library(tidyverse)
@@
   "data", "maps", "nps-boundaries", "yose_kica_sequ_boundaries.shp"))
-st_crs(nps_boundaries) #determine coordinate reference system (CRS)
+st_crs(nps_boundaries) # determine coordinate reference system (CRS)
+nps_boundaries <- st_transform(nps_boundaries, 26911) # transform to NAD83 UTM11
 
 # California Color Hillshade (90m) from koordinates.com
-hillshade <- read_stars(here::here(
+hillshade_90m <- read_stars(here::here(
   "data", "maps", "hillshade-90m", "california-color-hillshade-90m.tif"))
-st_crs(hillshade)
+st_crs(hillshade_90m)
+
+hillshade_300m <- read_stars(here::here(
+  "data", "maps", "hillshade-300m", "california-color-hillshade-300m.tif"))
 ```
@@
 
-### Transform CRS of nps_boundary
-
-```{r}
-nps_boundaries <- st_transform(nps_boundaries, 26911)
-```
-
 ### Crop raster with bounding box
@@
 ```{r}
-bbox <- st_bbox(c(xmin = 238000, ymax = 4235000, xmax = 403000, ymin = 4011000), crs = st_crs(nps_boundaries))
-hillshade_crop <- st_crop(hillshade, bbox)
+bbox <- st_bbox(c(xmin = 238000, ymax = 4235000, xmax = 403000, ymin = 4011000), 
+                crs = st_crs(nps_boundaries)) %>%  
+  st_as_sfc()
+hillshade_crop <- st_crop(hillshade_90m, bbox)
 hillshade_crop <- st_transform(hillshade_crop, 26911)
@@
 
-### Create map
+### Create study area map
 
@@
 
+### Create inset map of California
+
+```{r}
+inset_map <- tm_shape(hillshade_300m) +
+  tm_rgb(r = 1, g = 2, b = 3) +
+  tm_shape(bbox) + 
+  tm_borders(col = "black")
+
+inset_map
+```
+
+### Study area map with inset map
+
+```{r}
+exome_points_map
+print(inset_map, vp = viewport(0.35, 0.37, width = 0.2, height = 0.2))
+```
+
 ### Save map
@@
 ```{r}
-tmap_save(exome_points_map, here::here("out", "map_exome_points.png"))
+tmap_save(exome_points_map, 
+          here::here("out", "map_exome_points.png"),
+          insets_tm = inset_map,
+          insets_vp = viewport(0.14, 0.24, width = 0.2, height = 0.2))
 ```
+
+#### To Do
+
+-   Exported CA tif has no whitespace around state, has black border. How to remove border and add whitespace?

### 2022-11-25 code/maps.qmd
Create Yosemite map, minor edits to exome map

@@
 
-Run the following 5 chunks only if you have access to the project database.
-For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data" chunk.
+Run the following 4 chunks only if you have access to the project database.
+For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data1" chunk.
 
@@
 
-#### Retrieve data
+#### Retrieve data & create points layer
 
-```{sql retrieve-site-xy, connection=con, output.var="exome_sites_db"}
-select
-  id as site_id,
-  utme,
-  utmn
-from site
-where id in (74281, 50731, 50307, 52127, 10220, 10490, 50837, 20169, 72996, 72808, 70567, 54188, 10055, 10593, 50783)
+```{r}
+n <- nrow(exome_sites)
+
+exome_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(exome_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+
+write_csv(exome_points, here::here("data", "maps", "exome_points.csv"))
 ```
@@
 
-#### Create points layer
-
-```{r}
-exome_sites_xy <- 
-  exome_sites %>% 
-  inner_join(exome_sites_db, by = "site_id")
-
-# Add x-y random noise to obscure exact site locations
-exome_sites_xy <- exome_sites_xy %>% 
-  add_column(x = sample(-2000:2000, size = 15),
-             y = sample(-2000:2000, size = 15))
-
-exome_sites_xy <- exome_sites_xy %>% 
-  mutate(utme_noise = utme + x,
-         utmn_noise = utmn + y) %>% 
-  select(-utme, -utmn, -x, -y)
-
-# Save file 
-write_csv(exome_sites_xy, here::here("data", "maps", "exome_sites_xy.csv"))
-```
-
 ### Import map layers
 
-```{r import-map-data}
+```{r import-map-data1}
 # Read file, if lacking database access
-exome_sites_xy <- read_csv(here::here("data", "maps", "exome_sites_xy.csv"))
-exome_points = st_as_sf(exome_sites_xy, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
+exome_points <- read_csv(here::here("data", "maps", "exome_points.csv"))
+exome_points <- st_as_sf(exome_points, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
 
@@
 tmap_save(exome_points_map, 
-          here::here("out", "map_exome_points.png"),
+          here::here("doc", "manuscript", "figures",  "map_exome_points.png"),
           insets_tm = inset_map,
@@
 
-#### To Do
+### To Do
 
--   Exported CA tif has no whitespace around state, has black border. How to remove border and add whitespace?
+-   Exported California hillshade tif has no whitespace around edges, has black border. How to remove border and add whitespace?
+
+## Map of Yosemite translocation & donor sites
+
+Run the following 4 chunks only if you have access to the project database.
+For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data2" chunk.
+
+### Create dataset of site points
+
+```{r}
+donor_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(donor) %>% 
+  rename(site_id = donor) %>% 
+  add_column(type = "donor")
+
+translocation_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(site_id) %>% 
+  add_column(type = "translocation") %>% 
+  bind_rows(donor_sites)
+
+rm(donor_sites)
+```
+
+#### Connect to database
+
+```{r}
+source("db_connect.R")
+```
+
+#### Retrieve data & create points layer
+
+```{r}
+n <- nrow(translocation_sites)
+
+translocation_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(translocation_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+
+write_csv(translocation_points, here::here("data", "maps", "translocation_points.csv"))
+```
+
+#### Disconnect from database
+
+```{r}
+source("db_disconnect.R")
+```
+
+### Import map layers
+
+```{r import-map-data2}
+translocation_points <- read_csv(here::here("data", "maps", "translocation_points.csv"))
+translocation_points <- st_as_sf(translocation_points, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
+
+yose_boundary <- st_read(here::here(
+  "data", "maps", "nps-boundaries", "yose_boundary.shp"))
+st_crs(yose_boundary) # determine coordinate reference system (CRS)
+yose_boundary <- st_transform(yose_boundary, 26911) # transform to NAD83 UTM11
+
+# California Color Hillshade (90m) from koordinates.com
+hillshade_90m <- read_stars(here::here(
+  "data", "maps", "hillshade-90m", "california-color-hillshade-90m.tif"))
+st_crs(hillshade_90m)
+```
+
+### Check & fix shapefile polygons
+
+```{r}
+st_is_valid(yose_boundary) # invalid polygon
+
+yose_boundary <- st_make_valid(yose_boundary$geometry)
+
+st_is_valid(yose_boundary) # confirm that polygon is valid
+```
+
+### Crop raster with bounding box
+
+```{r}
+# For printing on screen: xmin = 244000, ymax = 4242500, xmax = 309000, ymin = 4142000
+bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 308000, ymin = 4151000), 
+                crs = st_crs(yose_boundary)) %>%  
+  st_as_sfc()
+hillshade_yose_crop <- st_crop(hillshade_90m, bbox)
+hillshade_yose_crop <- st_transform(hillshade_yose_crop, 26911)
+```
+
+### Create study area map
+
+```{r}
+translocation_points_map <- tm_shape(hillshade_yose_crop) + 
+  tm_rgb(r = 1, g = 2, b = 3) +
+  tm_shape(yose_boundary) + 
+  tm_borders(col = "black") +
+  tm_shape(translocation_points) + 
+  tm_symbols(size = 0.7, # increase point size, use darker blue
+             col = "type", 
+             palette = c("#008B00", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
+             title.col = "Population type") +
+  tm_layout(legend.position = c("right", "top"),
+            legend.bg.color = "white") +
+  tm_compass(type = "4star",
+             position = c(0.80, 0.10),
+             size = 2) +
+  tm_scale_bar(breaks = c(0, 10, 20, 30), 
+               text.size = 0.7, 
+               position = c("right", "bottom"))
+
+translocation_points_map
+```
+
+### Save map
+
+```{r}
+tmap_save(translocation_points_map, 
+          here::here("doc", "manuscript", "figures", "map_translocation_points.png"))
+```

### 2022-11-25 doc/manuscript/translocation.qmd
Add exome and Yosemite maps

@@
 Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
+Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia national parks (from top to bottom).
 To obscure the exact locations of populations, random noise was added to all point coordinates.
-Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia national parks (from top to bottom).](figures/map_exome_points.png)
+Inset map shows the location of the study area with California.
+In both maps, elevation is indicated with the colored hillshade layer (dark green = lowest elevation, white - highest elevation)](figures/map_exome_points.png)
 
@@
 
-Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 13 recipient sites (Table S1).
+Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 12 recipient sites (Table S1, Figure SXXXX).
 In each translocation, adult frogs ($\geq$ 40 mm snout-vent length) were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
@@
 
-<!--# Create table of translocation sites, years, frog numbers. Could also create a map if necessary -->
+<!--# Create table of translocation sites, years, frog numbers. -->
+
+![Figure SXXXX.
+Map showing the locations of translocated and donor populations in Yosemite National Park (park border indicated by black polygon).
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Elevations are indicated as in Figure SXXXX.
+See inset map in Figure SXXXX to locate Yosemite National Park within California.](figures/map_translocation_points.png)
 

### 2022-11-26 code/translocation_survival_analysis_suppl.Rmd
Add box plots to analysis of changes in Bd load

--- 2022-11-01 code/translocation_survival_analysis_suppl.Rmd
@@
 
-## Evaluate effect of frog size independent of donor population: Model m1d from `translocation_survival_analysis.Rmd` but using data subset where donor = 72996
+## Evaluate effect of frog size independent of donor populationRead in frog translocation dataset
 
-### Read in frog translocation dataset  
-* Read in data, create variables for analysis, standardize continuous variables. 
-* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+-   Model m1d from `translocation_survival_analysis.Rmd` but using data subset where donor = 72996
+-   Read in data, create variables for analysis, standardize continuous variables.
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
 ```{r read-translocation-data}
@@
 ### Model specification
+
 ```{r}
@@
 ```
+
 -   Using dataset containing only frogs translocated from 72996 had relatively little effect on the predictors of frog survival.
@@
 ### Read in frog translocation dataset
-* Read in data, create variables for analysis, standardize continuous variables. 
-* arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+-   Read in data, create variables for analysis, standardize continuous variables.
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
 ```{r read-translocation-data-including-na}
@@
 ## Analysis of changes in Bd load following translocation
--  Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years > 1. Only frogs that were recaptured at least once following translocation are included.
+
+-   Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years \> 1. Only frogs that were recaptured at least once following translocation are included.
 
@@
 ### Load additional libraries
+
 ```{r}
@@
 ### Read in raw translocation and capture datasets, bind them
+
 ```{r read-data}
@@
 ### Connect to PostgreSQL database `amphibians`
+
 ```{r}
@@
 ### Retrieve Bd data for translocated frogs from database
+
 ```{sql retrieve-translocation-bd-data, connection=con, output.var="translocate_bd"}
@@
 ### Retrieve Bd data for captured frogs from database
+
 ```{sql retrieve-capture-bd-data, connection=con, output.var="capture_bd"}
@@
 ### Disconnect from database
+
 ```{r}
@@
 ### Combine Bd data from translocate and capture datasets
+
 ```{r}
@@
 ### Add Bd data to frog data, restructure dataset
+
 ```{r}
@@
 ### Add years since translocation
+
 ```{r}
@@
 ### Update site ids to standardize usage
+
 ```{r}
@@
 ### Add translocation_id
+
 ```{r}
@@
 ### Save frog-Bd dataset
+
 ```{r}
@@
 
-## Analyze dataset
-### Read in data, standardize variables
+### Description of dataset columns
+
+-   site_id: Unique 5-digit site identifier.\
+-   site_id1: Assigned 5-digit site identifier for a collection of adjacent sites that are utilized by a frog population.
+-   translocation_id: Unique identifier for each translocation.\
+-   visit_date: Date on which translocation or survey was conducted.\
+-   survey_type: "translocation" or "cmr".\
+-   period: "before" or "after" translocation.
+-   year_translocate: Year in which translocation was conducted.\
+-   year_capture: Year in which frog was captured.\
+-   interval: Number of years between translocation and frog capture.\
+-   pit_tag_ref: Unique PIT tag identifier.\
+-   swab_id: Unique identifier for skin swab sample.\
+-   bd_load: Bd infection intensity determined from skin swab samples, units are number of ITS1 copies \* 80.
+
+### Analyze dataset
+
+#### Read in data
+
 ```{r}
-d1 <- read_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv")) %>% 
+d1 <- read_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv")) 
+```
+
+#### Data exploration
+
+```{r}
+d1 %>% 
+  mutate(period = factor(period, levels = c("before", "after"))) %>% 
+  ggplot(aes(x = period, y = log10(bd_load + 1))) +
+          geom_boxplot()
+
+d1 %>% 
+  mutate(period = factor(period, levels = c("before", "after"))) %>% 
+  ggplot(aes(x = period, y = log10(bd_load + 1))) +
+    geom_boxplot() + 
+    facet_grid(. ~ site_id1) +
+    theme(axis.text.x = element_text(angle = 90))
+```
+
+-   In most cases, relatively small increases in load following translocation.
+
+#### Prepare variables for analysis
+
+```{r}
+d1 <- d1 %>% 
   mutate(after = if_else(period == "after", 1, 0),
-         bd_load = as.integer(round(bd_load)),  # if using negative binomial distribution
+         bd_load = as.integer(round(bd_load)), 
          after_c = after - mean(after)) 
@@
 
-### Null values - check
+#### Null values - check
+
 ```{r check-null}
@@
 
-### Include only CMR years 0 and 1
+#### Include only CMR years 0 and 1
+
 ```{r}
@@
 
-### Plot distribution of bd_load response variable
+#### Plot distribution of bd_load response variable
+
 ```{r}
@@
 ```
-* Data are strongly left-skewed, possibly zero-inflated. Model using negative binomial family.
+
+-   Data are strongly left-skewed, possibly zero-inflated. Model using negative binomial family.
 
 ### Model without group-level effects
+
 #### Model specification
+
 ```{r}
@@
 #### Priors and chain diagnostics
+
 ```{r}
@@
 #### Posterior predictive check
+
 ```{r}
@@
 #### Calculate loo
+
 ```{r}
@@
 #### Posterior analysis
+
 ```{r}
@@
 ### Model with 1 grouping variable: pit_tag_ref
+
 #### Model specification
+
 ```{r}
@@
 ### Model with 2 grouping variables: pit_tag_ref nested within translocation_id
+
 #### Model specification
+
 ```{r}
@@
 ### Model with 2 grouping variables: pit_tag_ref nested within site_id1
+
 #### Model specification
+
 ```{r}

### 2022-11-29 doc/manuscript/translocation.qmd
Add results text and figures, clean up formatting

--- 2022-11-02 doc/manuscript/translocation.qmd
@@
 
-^8^Resources Management and Science, Yosemite National Park, El Portal, CA , USA
+^8^Resources Management and Science, Yosemite National Park, El Portal, CA, USA
 
@@
 
-```{=html}
 <!--# No more than 200 words, follow structure described in instructions for authors.
-2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
- and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
+2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
 See [annotated example](https://www.nature.com/documents/nature-summary-paragraph.pdf).
 -->
-```
+
 ### Main
@@
 
--   **Potential knowledge gap**: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
+-   Potential knowledge gap: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
 
@@
 
--   "Rescued" populations provide a source to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
+-   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
 
@@
 
--   **Potential knowledge gap/opportunity/selling point**: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+-   Potential knowledge gap/opportunity/selling point: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
 A unique attribute of this study is that we describe the genetic changes associated with evolutionary rescue, and combine repeated translocations of "rescued" populations, longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate the capacity of "rescued" populations to persist across a landscape of disease.
@@
 
+@fig-exomemap
+
 #### Frog translocations
+
+```{r, include=FALSE}
+library(readr)
+library(dplyr)
+library(magrittr)
+d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
+
+transno_min <- d1 %>% 
+  distinct(site_id, date) %>% 
+  group_by(site_id) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(min(n))
+
+transno_max <- d1 %>% 
+  distinct(site_id, date) %>% 
+  group_by(site_id) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(max(n))
+
+transfrogs_max <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(max(n))
+
+transfrogs_min <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(min(n))
+
+```
+
+During the period `r min(d1$year)`-`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`-`r transno_max` translocations per site; @fig-yosemap) <!--# Add table of all translocations: site_id, date, frog #s, donor site, etc.?
+-->.
+Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
+<!--# #fig-yosemap will likely be moved to supplement -->Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (when multiple translocations were made to individual sites; @fig-translocation-survival).
+This suggests an important effect of site characteristics on frog survival.
+
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation (bdload_z) was not among them (@fig-survival-postdens).
+This result is consistent with the exome capture results suggesting that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
+Important predictors included winter severity in the year following translocation (snowt1_z; negative effect), site elevation (elev_z; positive effect), and donor population (donor70567, donor72996; Frogs from 72996 had lower survival than those from either 70459 or 70567).
+Males had higher survival than females, but this effect was only marginally important.
+Effects of frog size (length_z), winter severity in the year of translocation (snowt_z), day of year on which translocation was conducted (day_z), and translocation order (first_c0.353) were also unimportant.
+
+<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals? Simplify names of predictor variables? -->
+
+<!--# Present results indicating that Bd loads on frogs changed little following translocation? -->
 
@@
 
-Given observed yearly adult survival probabilities, our stage-structured model predicted that six of our twelve translocated population should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (Fig.
-XA; Fig.
-SXA; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
-When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (Fig.
-XA).
-In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our modeled predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult lifestage was near 0.4 (Fig.
-XA).
-A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (Fig.
-SXA).
+Given observed yearly adult survival probabilities, our stage-structured model predicted that six of our twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
+When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (@fig-viability A).
+In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
+A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50 year extinction probabilities of less 1 (Fig.
-XB) -- in other words, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
-For these six populations, as $\omega$ increased to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (Fig.
-XB).
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B) -- in other words, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
+For these six populations, as $\omega$ increased to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (Fig.
-SXB; population 70550, surveyed for 16 years).
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (@fig-viability-supp B; population 70550, surveyed for 16 years).
 
@@
 A makefile could be a good solution.
--->
+Caching figures could be another option (https://quarto.org/docs/computations/caching.html) -->
 
-![Figure X: A.
-The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability or recruitment $\omega$, given the parameterized, deterministic model for translocated $R.
-sierrae$.
+![Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
+Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia national parks (from top to bottom).
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Inset map shows the location of the study area within California.
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation)](figures/map_exome_points.png){#fig-exomemap}
+
+![Map showing the locations of translocated and donor populations in Yosemite National Park (park border indicated by black polygon).
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Elevations are indicated as in @fig-exomemap.
+See inset map in @fig-exomemap to locate Yosemite National Park within California.](figures/map_translocation_points.png){#fig-yosemap}
+
+![Average survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
+Sites are arranged along the x-axis using the average of the median survival per translocation.
+Frog survival for the one-year period following each translocation was estimated using capture-mark-recapture methods.
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival}
+
+<!--# Remove gray background and grid. -->
+
+![From the best model identifying predictors of 1-year frog survival following translocations, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
+
+<!--# Remove text header -->
+
+![A.
+The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability or recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
+sierrae*.
 Color points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment to adults is $\omega = 0.1$ (given by the dashed line).
@@
 The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$ and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.1$, for comparison to A.](figures/pop_viability_figures_for_manuscript.jpg)
+The dashed line shows where $\omega = 0.1$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v1.jpg){#fig-viability}
+
+<!-- Use figure v1 or v2? If v1, sort legend labels by sigma-AR-->
+
+![A.
+The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F*.
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.1.
+B.
+The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (colored line).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+
+<!-- In legend for (B), change "lake" to "site".
+Switch color of 70550 in A so that color of line in B can be easily referenced in text, e.g., "orange".
+Also, use colorblind-friendly palette for figure, if possible.
+-->
 
@@
 
-<!--# Typically less than 3000 words -->
+<!-- Typically less than 3000 words -->
 
@@
 
-<!--# subheadings can be used, up to 40 characters -->
+<!-- subheadings can be used, up to 40 characters -->
 
 We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
-Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering).
+Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering").
 These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
@@
 
-DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations <!--# Dusy 10220 not included due to small sample size?
---> and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada from northern Yosemite National Park to south of Sequoia National Park (Figure Sx), and the latitudinal range of the two population types was similar.
-
-<!--# Create map of sample sites -->
+DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-exomemap).
+The latitudinal range of the two population types was similar.
 
@@
 sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-We identified donor populations from which adult frogs could be collected using several years of visual encounter surveys (VES) and skin swab collections (Knapp et al.
-2016).
+We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016].
 Results indicated that these populations contained several hundred *R.
-sierrae* adults and hundreds or thousands of tadpoles, and had recovered from previous Bd-caused declines.
-Bd prevalence in donor populations is high and Bd infection intensity (load) is below the level at which frog mortality is predicted (calculate mean prevalence and load).
+sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines.
+Bd prevalence in donor populations was high and Bd infection intensity (load) was below the level at which frog mortality is predicted <!--Calculate median prevalence and load -->.
 Recipient sites to which frogs were translocated were chosen based on previous *R.
-sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species (Knapp 2005).
+sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
 
-Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 13 recipient sites (Table S1).
+Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 12 recipient sites (@fig-yosemap).
 In each translocation, adult frogs ($\geq$ 40 mm snout-vent length) were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
@@
 
-<!--# Create table of translocation sites, years, frog numbers. Could also create a map if necessary -->
+<!--# Create table of translocation sites, years, frog numbers? -->
 
-During 2006-2012, CMR surveys were conducted conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
+During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
 Frog were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we used a robust design in which all habitats were searched during several consecutive days (secondary periods - calculate min, max, mean), and frogs were processed as described above.
+During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (secondary periods - calculate min, max, mean), and frogs were processed as described above.
+However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
 Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and identified only during subsequent captures.
-Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods as described in Joseph and Knapp (2018).
+Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
+For details of survey and swab analysis methods, see @joseph2018.
 
@@
 These *R.
-sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance.
+sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance <!-- reference -->.
+
+<!--# Add description of CMR analysis using mrmr package -->
+
+To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
+We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
 
@@
 For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling (Kosmala et al.
-2015, Ecology), we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution (Beaumont 2010).
+Using Approximate Bayesian Computing and rejection sampling (@kosmala2016), we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
+The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
@@
 ### References
+
+<!--# References for Methods section only -->
 
@@
 
-## Running Code
-
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
-
-```{r}
-1 + 1
-```
-
-You can add options to executable code like this
-
-```{r}
-#| echo: false
-2 * 2
-```
-
-The `echo: false` option disables the printing of code (only output is displayed).
+-

### 2022-12-07 code/translocation_survival_analysis.Rmd
Update translocation figures to improve readability

--- 2022-11-01 code/translocation_survival_analysis.Rmd
@@
 library(janitor)
+library(patchwork)
 
@@
 ```{r}
-tidy(m1d, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m1d, area_method = "equal height", prob = 0.90, pars = b1) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+tidy(m1d, conf.int = TRUE, conf.level = 0.95)
+
+labels_alt <- c("donor_72996", "donor_70567", "sex_male", "order_first", "elevation", "bd_load", "day", "snow_t1", "snow_t", "length", "(Intercept)")
+mcmc_areas(m1d, area_method = "equal height", prob = 0.95, pars = b1) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +
+  scale_y_discrete(labels = labels_alt, limits = rev) +
+  theme(plot.background = element_rect(fill = "white", color = "white"))
 
@@
 -   Based on 90% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
+
+### Save plot
+
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m1d.png"), width = 5, height = 5)
+```
 
@@
 
+## Reduced model from m1d
+
 ### Create reduced model containing only important predictors for subsequent analyses
 
--   Reduced model is helpful because it reduces the number of values used to calculate predictive models.
+- Reduced model is helpful because it reduces the number of values used to calculate probabilities from predictive models.
 
@@
 ```{r}
-tidy(m1d_reduce, conf.int = TRUE, conf.level = 0.90)
+tidy(m1d_reduce, conf.int = TRUE, conf.level = 0.95)
 
@@
 
-mcmc_areas(m1d_reduce, area_method = "equal height", prob = 0.90, pars = b2) +
+mcmc_areas(m1d_reduce, area_method = "equal height", prob = 0.95, pars = b2) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Reduced model with standardized predictors & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
-```
-
-### Confirm fit of reduced model is comparable to full model
+          "Posterior distributions with medians & 95% uncertainty intervals")
+```
+
+### Confirm that fit of reduced model is comparable to full model
 
@@
 - Model does a good job of predicting survival based on group-level and individual-level characteristics.  
--   The predicted survival is based on assumption that only males are included. Run another set of models for females?
+- Predicted survival is based on assumption that only males are included. Run another set of models for females?
 
@@
   ggplot(aes(x = elevation, y = surv)) +  #   "ggplot(aes(x = (elev_z * 2*sd_elev) + mean_elev, y = surv)) +" gives identical result
-    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
+          geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 1) +
         # geom_point(aes(y = .epred)) +
           labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival")
+
+# Plot with confidence bands
+alpha = 1 - 0.95
+
+p_elev <- d1_elevz %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, elevation, male_c, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 1000, re_formula = NA) %>% 
+  group_by(elevation) %>%
+  summarise(lower = quantile(.epred, alpha), 
+            upper = quantile(.epred, 1 - alpha),  
+            median = quantile(.epred, 0.5)) %>%
+  ggplot(aes(x = elevation)) + 
+          geom_ribbon(aes(ymin = lower, ymax = upper), fill= "lightgray", alpha = 1) +
+          geom_line(aes(y = median), color = "blue", size = 1, alpha = 1) +
+          labs(x = "Elevation of recipient site (meters)", y = "P(1-year frog survival)") +
+          annotate(geom = "text", x = 2500, y = 0.94, label = "B", size = 4) +
+          theme_classic()
+
+p_elev
 ```
@@
 
-d1_snowt1z %>% 
+alpha = 1 - 0.95
+
+p_snowt1 <- d1_snowt1z %>% 
   select(site_id, translocation_id, snowt1_z, snow_t1, elev_z, male_c, donor, surv) %>% 
-  add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
-  ggplot(aes(x = snow_t1, y = surv)) +
-    geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 0.5) +
-#    geom_point(aes(y = .epred)) +
-  labs(x = "Winter severity in year following translocation (% of average snow depth)", y = "Probability of 1-year frog survival") 
+  add_epred_draws(m1d_reduce, ndraws = 1000, re_formula = NA) %>% 
+  group_by(snow_t1) %>%
+  summarise(lower = quantile(.epred, alpha), 
+            upper = quantile(.epred, 1 - alpha),  
+            median = quantile(.epred, 0.5)) %>%
+  ggplot(aes(x = snow_t1)) +
+          geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightgray", alpha = 1) +
+          geom_line(aes(y = median), color = "blue", size = 1, alpha = 1) +
+          labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year frog survival)") +
+          annotate(geom = "text", x = 157, y = 0.92, label = "A", size = 4) +
+          theme_classic()
+
+p_snowt1
 ```
@@
 
-d1_malec %>% 
+p_sex <- d1_malec %>% 
   select(site_id, translocation_id, snowt1_z, elev_z, male_c, sex, donor, surv) %>% 
@@
   ggplot(aes(x = male, y = surv)) +
-    geom_boxplot(aes(y = .epred, x = sex)) +
-#    geom_point(aes(y = .epred)) +
-    labs(x = "Sex", y = "Probability of 1-year frog survival")
+    geom_boxplot(aes(y = .epred, x = sex), fill = "lightgray") +
+    labs(x = "Sex", y = "P(1-year frog survival)") +
+    annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "C", size = 4) +
+    theme_classic()
+
+p_sex
 ```
@@
 
-d1_donor %>% 
+p_donor <- d1_donor %>% 
   select(site_id, translocation_id, snowt1_z, elev_z, male_c, donor, surv) %>% 
@@
   ggplot(aes(x = donor, y = surv)) +
-    geom_boxplot(aes(y = .epred, x = donor)) +
-#    geom_point(aes(y = .epred)) +
-    labs(x = "Donor population", y = "Probability of 1-year frog survival")
+    geom_boxplot(aes(y = .epred, x = donor), fill = "lightgray") +
+    labs(x = "Donor population", y = "P(1-year frog survival)") +
+    annotate(geom = "text", x = 3.3, y = 0.83, hjust = 0, label = "D", size = 4) +
+    theme_classic()
+
+p_donor
 ```
@@
 
+# plot individual draws grouped by sex 
 pal <- c("#0d0887", "#f89540")
@@
           geom_line(aes(y = .epred, group = paste(sex, .draw)), size = 0.4, alpha = 0.5) +
-#    geom_point(aes(y = .epred)) +
           labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
@@
           theme_classic()
+
+# plot medians and uncertainty ribbons grouped by sex
+p_elevsex <- d1_elevz_malec %>% 
+  select(site_id, translocation_id, snowt1_z, elev_z, elevation, male_c, sex, donor, surv) %>% 
+  add_epred_draws(m1d_reduce, ndraws = 1000, re_formula = NA) %>% 
+  group_by(elevation, sex) %>% 
+  summarise(lower = quantile(.epred, alpha), 
+            upper = quantile(.epred, 1 - alpha),  
+            median = quantile(.epred, 0.5)) %>%
+  ggplot(aes(x = elevation, group = sex)) +
+          geom_ribbon(aes(ymin = lower, ymax = upper, fill = sex), alpha = 0.2, linetype = 0) +
+          geom_line(aes(y = median, color = sex), size = 1, alpha = 1) +
+          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
+          scale_color_manual(values = pal) + 
+          scale_fill_manual(values = pal) +
+          theme_classic()
+
+p_elevsex
+```
+
+### Create multiplot conditional effects figure
+
+```{r}
+cond_effects_plot <- (p_snowt1 | p_elev) / (p_sex | p_donor)
+```
+
+### Save plot
+
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "cond_effects_plot.png"), width = 7, height = 4)
 ```

### 2022-12-07 code/translocation_survival_exploratory.Rmd
Update translocation figures to improve readability

@@
 
-cohort_survival %>% 
+cohort_survival_plot <- cohort_survival %>% 
   mutate(release_year = year(release_date)) %>% 
@@
       scale_color_manual(values = palette1) + 
-      labs(x = "Recipient site", y = "1-year frog survival")
+      labs(x = "Recipient site", y = "1-year frog survival") +
+      theme_classic() +
+      theme(
+        legend.background = element_rect(color = "black", size = 0.3),
+        legend.position = c(0.2, 0.8),
+        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
+
+cohort_survival_plot
 ```
@@
 * Indicates need to include site_id and/or translocation_id as group-level effect(s).
+
+### Save plot
+
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "translocation_survival_bysiteid.png"), width = 5, height = 5) 
+```
 

### 2022-12-07 doc/manuscript/translocation.qmd
Update translocation figures to improve readability

@@
 This suggests an important effect of site characteristics on frog survival.
+<!--# Add analysis results: translocation_survival_analysis, analysis 2 -->
 
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation (bdload_z) was not among them (@fig-survival-postdens).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation (bdload) was not among them (@fig-survival-postdens).
 This result is consistent with the exome capture results suggesting that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
-Important predictors included winter severity in the year following translocation (snowt1_z; negative effect), site elevation (elev_z; positive effect), and donor population (donor70567, donor72996; Frogs from 72996 had lower survival than those from either 70459 or 70567).
-Males had higher survival than females, but this effect was only marginally important.
-Effects of frog size (length_z), winter severity in the year of translocation (snowt_z), day of year on which translocation was conducted (day_z), and translocation order (first_c0.353) were also unimportant.
+Important predictors are shown in @fig-cond-effects, and included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996).
+Males had higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
+In addition to Bd load, effects of frog size (length_z), winter severity in the year of translocation (snow_t), day of year on which translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
 
@@
 
-![Average survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
-Sites are arranged along the x-axis using the average of the median survival per translocation.
+![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
+Sites are arranged along the x-axis using the average by site of the median survival per translocation.
 Frog survival for the one-year period following each translocation was estimated using capture-mark-recapture methods.
@@
 
-<!--# Remove gray background and grid. -->
-
-![From the best model identifying predictors of 1-year frog survival following translocations, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
+![For the best model identifying predictors of 1-year frog survival following translocations, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
 Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
 
-<!--# Remove text header -->
+![Conditional effects of the important predictors of 1-year frog survival: (A) elevation, (B) winter severity in the year following the translocation, (C) sex, and (D) donor population.
+In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
 

### 2022-12-07 doc/manuscript/translocation.qmd
Update methods and references

@@
 
-<!--# Exact author order to be determined) -->
+<!--# Exact author order to be determined -->
 
@@
 
-#### Frog translocations
+#### Frog translocations result in population reestablishment despite ongoing Bd infection
 
@@
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-<!--# #fig-yosemap will likely be moved to supplement -->Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (when multiple translocations were made to individual sites; @fig-translocation-survival).
+<!--# #fig-yosemap will likely be moved to supplement -->Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
 This suggests an important effect of site characteristics on frog survival.
@@
 
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation (bdload) was not among them (@fig-survival-postdens).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
 This result is consistent with the exome capture results suggesting that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
@@
 Males had higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
-In addition to Bd load, effects of frog size (length_z), winter severity in the year of translocation (snow_t), day of year on which translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
+In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
 
@@
 
-#### Population viability modeling
+#### Many reestablished populations are likely to persist over the long term
 
-Given observed yearly adult survival probabilities, our stage-structured model predicted that six of our twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
+Given observed yearly adult survival probabilities, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
 When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (@fig-viability A).
@@
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B) -- in other words, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
-For these six populations, as $\omega$ increased to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
+For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (@fig-viability-supp B; population 70550, surveyed for 16 years).
+<!--# Add information about natural recruitment probability in the study populations?
+Can we even calculate recruitment probability?
+-->
 
@@
 ![Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
-Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia national parks (from top to bottom).
+Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia National Parks (from top to bottom).
 To obscure the exact locations of populations, random noise was added to all point coordinates.
@@
 
-![Map showing the locations of translocated and donor populations in Yosemite National Park (park border indicated by black polygon).
+![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
 To obscure the exact locations of populations, random noise was added to all point coordinates.
 Elevations are indicated as in @fig-exomemap.
-See inset map in @fig-exomemap to locate Yosemite National Park within California.](figures/map_translocation_points.png){#fig-yosemap}
+See inset map in @fig-exomemap to locate Yosemite National Park within California.
+<!--# Add park boundaries to inset map?
+-->](figures/map_translocation_points.png){#fig-yosemap}
 
 ![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
-Sites are arranged along the x-axis using the average by site of the median survival per translocation.
+Sites are arranged along the x-axis using the average of the median survival per translocation at each site.
 Frog survival for the one-year period following each translocation was estimated using capture-mark-recapture methods.
@@
 
-![For the best model identifying predictors of 1-year frog survival following translocations, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
+![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
 Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
 
-![Conditional effects of the important predictors of 1-year frog survival: (A) elevation, (B) winter severity in the year following the translocation, (C) sex, and (D) donor population.
+![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) sex, and (D) donor population.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
@@
 ![A.
-The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability or recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
+The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
 sierrae*.
-Color points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment to adults is $\omega = 0.1$ (given by the dashed line).
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.1$ (given by the dashed line).
 B.
-The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$ and different mean values of $\omega$.
+The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
 The dashed line shows where $\omega = 0.1$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v1.jpg){#fig-viability}
 
-<!-- Use figure v1 or v2? If v1, sort legend labels by sigma-AR-->
+<!-- Use figure v1 or v2 (omega = 0.3)? If v1, sort legend labels by sigma-AR-->
 
 ![A.
-The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F*.
+The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.1.
@@
 
-<!-- In legend for (B), change "lake" to "site".
-Switch color of 70550 in A so that color of line in B can be easily referenced in text, e.g., "orange".
+<!-- In legend for (A), change "lake" to "site".
+Switch color of 70550 in A so that color of associated line in B can be easily referenced in text, e.g., "orange".
 Also, use colorblind-friendly palette for figure, if possible.
@@
 
-We used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure, in an effort to identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects.
-Specifically, we compared genomes of frogs that have never experienced a Bd-caused epizootic ("Bd-naive") versus populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering").
-These populations persist in an enzootic state, characterized by high Bd prevalence and Bd infection intensities (load) in adults that are well below the level expected to cause frog mortality.
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing.
+To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010dynamics].
+These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010dynamics].
+Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
 
@@
 
-DNA samples were collected via buccal swabbing of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
+DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
 These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-exomemap).
-The latitudinal range of the two population types was similar.
+To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
 
 #### Frog translocations
+
+##### Field methods
 
@@
 Results indicated that these populations contained several hundred *R.
-sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines.
+sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines [@knapp2016].
 Bd prevalence in donor populations was high and Bd infection intensity (load) was below the level at which frog mortality is predicted <!--Calculate median prevalence and load -->.
@@
 
-Between 2006 and 2020 we conducted 1-4 translocations of adult frogs to 12 recipient sites (@fig-yosemap).
-In each translocation, adult frogs ($\geq$ 40 mm snout-vent length) were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
+Between 2006 and 2020, we conducted 1-4 translocations of adult frogs to 12 recipient sites (@fig-yosemap).
+In each translocation, adult frogs were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited one to six times per summer (calculate min, max, mean) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
-CMR surveys allowed quantification of adult frog survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
+Following release, translocated populations were visited one to six times per summer <!--# calculate min, max, mean --> and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
+CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
 
@@
 Frog were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (secondary periods - calculate min, max, mean), and frogs were processed as described above.
+During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (secondary periods <!--# calculate min, max, mean-->), and frogs were processed as described above.
 However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
-Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and identified only during subsequent captures.
+Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and during subsequent captures were only identified and released.
 Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
@@
 These *R.
-sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance <!-- reference -->.
+sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
+
+##### Estimation of adult survival and population size
 
@@
 
+##### Predictors of post-translocation frog survival
+
 To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
 We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
+<!--# Add details and citations, from PeerJ paper?
+-->
 
@@
 
-**Model description**
+##### Model description
 
@@
 Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of a *R.
+We developed a model that tracked seven state variables of an *R.
 sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
@@
 
-The parameters in this model are yearly survival probability $\sigma_{*}$, where the "\*" indicates a particular state variable, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
+The parameters in this model are yearly survival probability $\sigma_{*}$, where the "\*" indicates a particular state variable<!--# "*" subscript is not translating correctly in .docx.
+Shows as upsidedown question mark -->, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
 The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
@@
 We also did not directly consider the dynamics of Bd in this model.
-We made this decision because i) all translocated populations are infected with Bd at high prevalence ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system (Wilber et al.
-2022).
+We made this decision because i) all translocated populations are infected with Bd at high prevalence, and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
 
-**Model analysis**
+##### Model analysis
 
-**Incorporating yearly variability in vital rates**
+###### Incorporating yearly variability in vital rates
 
-We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the mark-recapture model described elsewhere.
-While we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability.
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
+Although we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability.
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
@@
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability about a mean value.
-Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (more details below).
+Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
 
-**Estimating model parameters**
+###### Estimating model parameters
 
-The baseline parameter values for the model and how they were estimated are given in Table 1.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our mark-recapture model described elsewhere.
+The baseline parameter values for the model and how they were estimated are given in Table 1 <!--# Create table --> .
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described elsewhere.
 For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$.
-For populations where we had not pit-tagged any naturally recruited adults, we could not directly estimate $\sigma_{A_R}$ and set $\sigma_{A_R} = \sigma_{A_T}$.
+For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
 
-**Model analysis and simulation**
+###### Model analysis and simulation
 
 We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$ and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
+First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
 We then fixed $\omega = 0.1$ and calculated the predicted growth rate of our six populations.
-Second, we performed a local sensitivity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
 
-Third, we defined version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among year variability in $\omega$.
-We used this model to simulate a one time introduction of 40 translocated adult frogs.
+Second, we performed a local sensitivity analysis <!--# referred to elsewhere as "elasticity" -->on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
+
+Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
+We used this model to simulate a one-time introduction of 40 translocated adult frogs.
 We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
@@
 Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
-We focused on population 70550 because this was our longest mark-recapture time series for a translocated population and because this population shows evidence of recovery.
+We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
 We simulated our model for 16 years, repeating the simulation 50,000 times.

### 2022-12-08 doc/manuscript/translocation.qmd
Test push with fixed typo
Add analysis methods to translocation section

@@
 
-During the period `r min(d1$year)`-`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`-`r transno_max` translocations per site; @fig-yosemap) <!--# Add table of all translocations: site_id, date, frog #s, donor site, etc.?
--->.
+During the period `r min(d1$year)`-`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`-`r transno_max` translocations per site; @fig-yosemap).
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-<!--# #fig-yosemap will likely be moved to supplement -->Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-This suggests an important effect of site characteristics on frog survival.
-<!--# Add analysis results: translocation_survival_analysis, analysis 2 -->
+Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details).
+The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival.
+<!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd?
+-->
 
@@
 In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-
-<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals? Simplify names of predictor variables? -->
 
@@
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (@fig-viability-supp B; population 70550, surveyed for 16 years).
-<!--# Add information about natural recruitment probability in the study populations?
-Can we even calculate recruitment probability?
+<!--# To help constrain the possible values of omega, add information about recruitment probability  observed in the study populations?
+Because we don't have accurate counts of subadults, can we even calculate recruitment probability?
 -->
@@
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).
+<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals?
+Simplify names of predictor variables?
+-->](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
 
@@
 
-<!--# Mention sampled declining populations also? -->
+<!--# Mention that we sampled declining populations also? Will need to update map and text to show only the samples that were sequenced. -->
 
@@
 
-##### Estimation of adult survival and population size
+##### Estimation of adult survival and population size from CMR surveys
 
 <!--# Add description of CMR analysis using mrmr package -->
+
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; \@rsoftware2022].
+The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
+The possible states of individuals include "not recruited", "alive", and "dead".
+The possible observations of individuals include "detected" and "not detected".
+We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
+For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
+We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
+Code used to construct CMR datasets and fit models for each translocated population is available from a GitHub repository ([https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.)).
 
@@
 We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
-<!--# Add details and citations, from PeerJ paper?
--->
+We started with models that included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
+Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
+Both approaches utilized the loo package.
+
+Of the two analyses we conducted, one focused on identifying predictors of post-translocation frog survival from a wide range of variables describing characteristics of frogs, habitat, and translocations \[@fig-survival-postdens\].
+The other analysis focused specifically on whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
+In both analyses, the response variable, probability of frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
+For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
+We checked convergence of the MCMC algorithm using trace plots and Rhat.
+Code to create datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
 

### 2022-12-11 doc/manuscript/translocation.qmd
Add inline calculations to translocation methods

@@
 
-#### Frog translocations result in population reestablishment despite ongoing Bd infection
+#### Translocations allow population reestablishment
 
-```{r, include=FALSE}
+```{r}
+#| include: false
+
 library(readr)
@@
 library(magrittr)
+library(stringr)
+library(lubridate)
 d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
@@
 
+bdload_min <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  group_by(site_id) %>% 
+  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
+  ungroup() %>% 
+  summarize(min(bdload_ave))
+
+bdload_max <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  group_by(site_id) %>% 
+  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
+  ungroup() %>% 
+  summarize(max(bdload_ave))
+
+bdprev_max <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
+  group_by(site_id) %>% 
+  summarize(prev = sum(bdload01)/n()) %>% 
+  ungroup() %>% select(prev) %>% 
+  slice_max(prev, n = 1, with_ties = FALSE)
+
+bdprev_min <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
+  group_by(site_id) %>% 
+  summarize(prev = sum(bdload01)/n()) %>% 
+  ungroup() %>% select(prev) %>% 
+  slice_min(prev, n = 1, with_ties = FALSE)
+
+files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
+surveys <- read_csv(files, id = "path") %>% 
+  filter(survey_date < "2023-01-01") %>% 
+  select(-path)
+rm(files)
+
+surveys_yr_mean <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(mean(n))
+
+surveys_yr_max <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(max(n))
+
+surveys_yr_min <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(min(n))
+
+robust_year_mean <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(mean(max_2period))
+
+robust_year_min <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(min(max_2period))
+
+robust_year_max <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(max(max_2period))
 ```
 
-During the period `r min(d1$year)`-`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`-`r transno_max` translocations per site; @fig-yosemap).
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+<!--# Create table of translocation sites, years, frog numbers?
+--> Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
 As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details).
 The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival.
-<!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd?
+<!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)?
 -->
@@
 
-#### Many reestablished populations are likely to persist over the long term
+#### Long-term persistence of reestablished populations
 
@@
 <!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations?
-Because we don't have accurate counts of subadults, can we even calculate recruitment probability?
+Because we don't have accurate counts of subadults, can we even provide rough estimate of recruitment probability?
 -->
@@
 ### Figures and figure legends
-
-<!--# In initial submission, include figure legends with their corresponding figures. -->
 
@@
 Inset map shows the location of the study area within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation)](figures/map_exome_points.png){#fig-exomemap}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation)](figures/map_exome_points.png){#fig-exomemap width="450"}
 
@@
 <!--# Add park boundaries to inset map?
--->](figures/map_translocation_points.png){#fig-yosemap}
+Also, for this and previous figure, is basemap necessary/helpful -->](figures/map_translocation_points.png){#fig-yosemap width="450"}
 
@@
 Frog survival for the one-year period following each translocation was estimated using capture-mark-recapture methods.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival}
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).
-<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals?
-Simplify names of predictor variables?
--->](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+
+<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals? -->
 
@@
 To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010dynamics].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010dynamics].
+Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
+These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
 Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
@@
 sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines [@knapp2016].
-Bd prevalence in donor populations was high and Bd infection intensity (load) was below the level at which frog mortality is predicted <!--Calculate median prevalence and load -->.
+As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd infection intensity ("load", expressed as median by site: `r round(bdload_min, 2)`--`r round(bdload_max, 2)`) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 ITS copies per swab)[@vredenburg2010; @joseph2018].
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 
-Between 2006 and 2020, we conducted 1-4 translocations of adult frogs to 12 recipient sites (@fig-yosemap).
-In each translocation, adult frogs were collected from one of three donor populations and measured, weighed, swabbed, and PIT tagged.
+During each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited one to six times per summer <!--# calculate min, max, mean --> and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
+Following release, translocated populations were visited approximately once per month during the summer active season (mean number of visits per summer = `r round(surveys_yr_mean)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
 CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
-
-<!--# Create table of translocation sites, years, frog numbers? -->
-
 During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
-Frog were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (secondary periods <!--# calculate min, max, mean-->), and frogs were processed as described above.
+Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
+During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (mean number of secondary periods per primary period = `r round(robust_year_mean)`; range = `r robust_year_min`--`r robust_year_max`)<!--# calculate min, max, median-->), and frogs were processed as described above.
 However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
@@
 
-##### Estimation of adult survival and population size from CMR surveys
-
-<!--# Add description of CMR analysis using mrmr package -->
+##### Estimation of frog survival and abundance
 
@@
 
-Of the two analyses we conducted, one focused on identifying predictors of post-translocation frog survival from a wide range of variables describing characteristics of frogs, habitat, and translocations \[@fig-survival-postdens\].
+Of the two analyses we conducted, one focused on identifying predictors of post-translocation frog survival from a wide range of variables describing characteristics of frogs, habitat, and translocations [@fig-survival-postdens].
 The other analysis focused specifically on whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.

### 2022-12-11 doc/manuscript/translocation.qmd
Address comments in manuscript, add table, update figure

@@
 sierrae*.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.1$ (given by the dashed line).
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
 B.
 The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.1$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v1.jpg){#fig-viability}
-
-<!-- Use figure v1 or v2 (omega = 0.3)? If v1, sort legend labels by sigma-AR-->
+The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
 
@@
 The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.1.
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
 B.
-The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (colored line).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
 
-<!-- In legend for (A), change "lake" to "site".
-Switch color of 70550 in A so that color of associated line in B can be easily referenced in text, e.g., "orange".
-Also, use colorblind-friendly palette for figure, if possible.
--->
+<!-- Left color palette as is as 12-color color-blind friendly is not the most friendly color palette to look at.  We can discuss. -->
 
@@
 
-The parameters in this model are yearly survival probability $\sigma_{*}$, where the "\*" indicates a particular state variable<!--# "*" subscript is not translating correctly in .docx.
-Shows as upsidedown question mark -->, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$, where the "$\cdot$" indicates a particular state variable, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
 The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
@@
 
-The baseline parameter values for the model and how they were estimated are given in Table 1 <!--# Create table --> .
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described elsewhere.
+The baseline parameter values for the model and how they were estimated are given in Table 1.
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
 For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$.
 For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+
+Table 1: Description and values of parameters used in the model.
+All survival probabilities are in the presence of the fungal pathogen Bd.
+<!--# Roland, please add any additional details needed to the "Source" column.
+We ideally want to keep this succinct.
+-->
+
+| **Parameter**                                                             | **Value**            | **Source**                                                             |
+|:----------------------------|:------------------|:-----------------------|
+| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge     |
+| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, and natural history knowledge |
+| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, and natural history knowledge |
+| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge     |
+| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge     |
+| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                                  |
+| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                                  |
+| $p_{L_1}$, Probability of a year 1 larvae remaining as a larvae           | 1                    | Estimated from field data, observations, and natural history knowledge |
+| $p_{L_2}$, Probability of a year 2 larvae remaining as a larvae           | 0.25                 | Estimated from field data, observations, and natural history knowledge |
+| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, and natural history knowledge |
+| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                     |
+| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                     |
+| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                            |
 
@@
 First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
-We then fixed $\omega = 0.1$ and calculated the predicted growth rate of our six populations.
+We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
 
-Second, we performed a local sensitivity analysis <!--# referred to elsewhere as "elasticity" -->on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
 
@@
 We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
-We varied the mean recruitment probability $\omega$ from 0.05 and 0.3 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
 

### 2022-12-13 code/translocation_survival_analysis_suppl.Rmd
Add calculations of viability model parameters

@@
 
-## Evaluate effect of frog size independent of donor populationRead in frog translocation dataset
+## Evaluate effect of frog size independent of donor population
 
@@
 -   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+### Read in frog translocation dataset
 
@@
 ```
+
+## Calculate parameters for viability analysis
+
+### Load packages
+
+```{r}
+library(RPostgreSQL)
+library(lubridate)
+library(dplyr)
+library(magrittr)
+library(readr)
+```
+
+### Estimate subadult survival
+
+#### Connect to database
+
+-   Use 2nd chunk when connecting via SSH.
+
+-   Users without credentials to access database can start with `subadult_survival.csv` saved below.
+
+```{r}
+con = dbConnect(dbDriver("PostgreSQL"),
+                host = Sys.getenv("host"), 
+                port = Sys.getenv("port"), 
+                dbname = Sys.getenv("dbname"),
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password")
+               )
+```
+
+```{r}
+con = dbConnect(dbDriver("PostgreSQL"),
+                host = "localhost",
+                port = 15432,
+                dbname = Sys.getenv("dbname"), 
+                user = rstudioapi::askForPassword("user name"),
+                password = rstudioapi::askForPassword("password")
+                )
+```
+
+#### Retrieve Bd data for translocated frogs from database
+
+-   Used 70550 because of its long and relatively intensive survey history.
+
+```{sql, connection=con, output.var="surv"}
+select
+  site_id,
+  visit_date,
+  species,
+  visual_life_stage,
+  visual_animal_state,
+  count
+from visit
+inner join survey on visit_id = visit.id
+inner join visual_survey on survey_id = survey.id  
+where site_id in (70550) and
+  species = 'ramu' and
+  visual_life_stage = 'subadult' and
+  visual_animal_state = 'alive'
+order by site_id, visit_date
+```
+
+#### Disconnect from database
+
+```{r}
+dbDisconnect(con) 
+rm("con")
+```
+
+#### Save raw dataset
+
+```{r}
+surv %>% write_csv(here::here("data", "raw", "subadult_survival.csv"))
+```
+
+#### Calculate maximum annual counts
+
+```{r}
+surv <- surv %>% 
+  group_by(visit_date) %>% 
+  summarize(count_sum = sum(count)) %>% 
+  ungroup() %>% 
+  mutate(year = year(visit_date)) %>% 
+  group_by(year) %>% 
+  summarize(count_max = max(count_sum))
+```
+
+#### Extract number of new recruits from mrmr model
+
+```{r}
+model <- read_rds(here::here("data", "clean", "70550_model.rds"))
+
+get_numeric_indices <- function(string) {
+  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
+  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
+  tibble::as_tibble(idx_mat) %>%
+    dplyr::mutate_all(readr::parse_integer)
+}
+
+survey_prim_periods <- model$data$surveys %>%
+  group_by(.data$primary_period) %>%
+  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
+  ungroup()
+
+p <- model$m_fit$draws("B", format = "draws_df") %>%
+  tidyr::pivot_longer(tidyselect::starts_with("B")) %>%
+  suppressWarnings() %>%
+  mutate(
+    get_numeric_indices(.data$name),
+    primary_period = .data$index_1 + 1
+  ) %>%
+  group_by(.data$primary_period) %>%
+  summarize(lo = quantile(.data$value, .025),
+            med = median(.data$value),
+            hi = quantile(.data$value, .975),
+            .groups = "drop") %>%
+  left_join(survey_prim_periods, by = "primary_period")
+
+recruit <- p %>% 
+  group_by(year) %>% 
+  summarize(recruit_sum = sum(med))
+```
+
+#### Create final dataset
+
+```{r}
+subadult_survival <- surv %>% 
+  inner_join(recruit, by = c("year"))
+```
+
+#### Calculate subadult survival
+
+-   Assumes that p(detection) for subadults is 0.25.
+
+```{r}
+subadult_sum <- subadult_survival %>% 
+  filter(between(year, 2008, 2020)) %>% 
+  summarize(sum(count_max))
+
+recruit_sum <- subadult_survival %>% 
+  filter(between(year, 2009, 2021)) %>% 
+  summarize(sum(recruit_sum))
+
+survival_sub <- recruit_sum / (subadult_sum * 4) 
+```
+
+### Estimate 1st-year tadpole survival
+
+-   Used 74976 due to its shallow habitats that allow accurate tadpole counts.
+
+-   1st-year tadpole counts are from field notebooks.
+
+```{r}
+surv_tad <- tibble(site_id = 74976, year = 2019:2021, count = c(250, 300, 305))
+```
+
+#### Extract adult abundance from mrmr model
+
+-   Used minimum survey date per year to estimate number of females during breeding season.
+
+```{r}
+model <- read_rds(here::here("data", "clean", "74976_model.rds"))
+
+get_numeric_indices <- function(string) {
+  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
+  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
+  tibble::as_tibble(idx_mat) %>%
+    dplyr::mutate_all(readr::parse_integer)
+}
+
+survey_prim_periods <- model$data$surveys %>%
+  group_by(.data$primary_period) %>%
+  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
+  ungroup()
+
+p <- model$m_fit$draws("N", format = "draws_df") %>%
+        tidyr::pivot_longer(tidyselect::starts_with("N")) %>%
+        suppressWarnings() %>%
+        mutate(
+          get_numeric_indices(.data$name),
+          primary_period = .data$index_1 + 1
+        ) %>%
+        group_by(.data$primary_period) %>%
+        summarize(lo = quantile(.data$value, .025),
+                  med = median(.data$value),
+                  hi = quantile(.data$value, .975),
+                  .groups = "drop") %>%
+        left_join(survey_prim_periods, by = "primary_period")
+
+abund_ad <- p %>% 
+  group_by(year) %>% 
+  summarize(survey_date = min(survey_date)) %>%
+  select(survey_date) %>% 
+  inner_join(p, by = c("survey_date")) %>% 
+  select(year, med)
+```
+
+#### Create final dataset
+
+```{r}
+surv_tad <- surv_tad %>% 
+  inner_join(abund_ad, by = "year") %>% 
+  rename(adults = med)
+```
+
+#### Calculate tadpole survival
+
+-   Assumptions:
+
+    -   p(detection) for 1st-year tadpoles is 0.5.
+
+    -   \# females is half of total adult abundance
+
+    -   female fecundity = 100 eggs
+
+-   Assume that
+
+```{r}
+surv_tad <- surv_tad %>% 
+  mutate(surv_1styr = (count * 2) / ((adults / 2) * 100)) %>% 
+  summarize(mean(surv_1styr))
+```

### 2022-12-13 doc/manuscript/translocation.qmd
Minor edits to table of parameter values

@@
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, and natural history knowledge |
-| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, and natural history knowledge |
+| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
 | $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
@@
 | $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
-| $p_{L_1}$, Probability of a year 1 larvae remaining as a larvae           | 1                    | Estimated from field data, observations, and natural history knowledge |
-| $p_{L_2}$, Probability of a year 2 larvae remaining as a larvae           | 0.25                 | Estimated from field data, observations, and natural history knowledge |
-| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, and natural history knowledge |
+| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
+| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
+| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
 | $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |

### 2022-12-13 doc/manuscript/translocation.qmd
Quarto auto-format to table of parameter values

@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:------------------|:-----------------------|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-01-17 doc/manuscript/translocation.qmd
Add text describing evolutionary rescue expectations

--- 2022-12-13 doc/manuscript/translocation.qmd
@@
 Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+Describe expectations if frog populations are recovering due to evolutionary rescue.
+
+-   Differences in allele frequencies between naive and recovering populations
+
+-   Shared selective response across recovering populations
+
+-   Changes in population growth rate extend beyond the natal environment
+
+-   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
+
+-   Population viability: Any expectation here?
+
+Then show how each result is consistent with those expectations.
 
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; \@rsoftware2022].
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software \[v4.4.4; \@rsoftware2022\].
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:------------------|:-----------------------|
+|:----------------------------|:-----------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-01-17 doc/manuscript/translocation.qmd
Improve text describing evolutionary rescue expectations

@@
 
--   Differences in allele frequencies between naive and recovering populations
+-   Differences in allele frequencies exist between Bd-naive and recovering populations
 
@@
 
--   Changes in population growth rate extend beyond the natal environment
+-   Changes in population growth rate due to evolution extend beyond the natal environment
 

### 2023-01-18 doc/manuscript/translocation.qmd
Update knowledge gaps using selection vocabulary

@@
 -   Potential knowledge gap/opportunity/selling point: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
-A unique attribute of this study is that we describe the genetic changes associated with evolutionary rescue, and combine repeated translocations of "rescued" populations, longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate the capacity of "rescued" populations to persist across a landscape of disease.
-Thus, ER combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion has occurred.
+We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
+Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
 These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.

### 2023-01-18 doc/manuscript/translocation.qmd
Add population viability expectation

@@
 
--   Population viability: Any expectation here?
+-   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
 
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software \[v4.4.4; \@rsoftware2022\].
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; \@rsoftware2022].
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:-----------------|:-------------------------|
+|:---------------------------|:-----------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-01-19 doc/manuscript/translocation.qmd
Mark's minor text edits

--- 2023-01-18 doc/manuscript/translocation.qmd
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:-----------------|:-------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-01-19 doc/manuscript/translocation.qmd
Changes to intro text to de-emaphize evo rescue

@@
 
--   Species recovery in the face of disease will depend on the ability of a species to move or change.
+-   Species recovery in the face of severe disease (population growth rate \< 1 ) will depend on the ability of a species to move or change.
 
--   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)
+-   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al.
+2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
+Restrict to species of conservation concern?).
 
--   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018).
+-   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease.
+For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
+These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+2019, Joseph and Knapp 2018).
+
+Population reestablishment using reintroduction of resistant hosts
+
+-   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
+
+-   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+
+-   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
+We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
+Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
+
+Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+<!--# I'm not sure the following text still makes sense to include.
+If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper.
+The following is my best attempt at doing so.
+I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript.
+-->
+
+We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence.
+These hypotheses generate the following expectations:
+
+-   Differences in allele frequencies exist between Bd-naive and recovering populations
+
+-   Shared selective response across recovering populations
+
+-   Changes in population growth rate due to evolution extend beyond the natal environment
+
+-   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
+
+-   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
+
+Then show how each result is consistent with those expectations.
+
+<!--# Extra ER text for use in manuscript -->
+
+-   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al.
+2014), can reverse ongoing declines in some species (Golas et al.
+2021; Searle et al.
+2021).
+
+-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al.
+2016, Schiebelhut et al.
+2018).
+This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
+(2013).
 
@@
     -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.
-
--   Potential knowledge gap: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).
-
-Population reestablishment using reintroduction of resistant hosts
-
--   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
-
--   Mendelson et al. (2019)
-
--   Scheele et al. (2021)
-
--   Joseph and Knapp (2018)
-
--   Potential knowledge gap/opportunity/selling point: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
-In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion has occurred.
-We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
-Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
-These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
-
-Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
-
-Describe expectations if frog populations are recovering due to evolutionary rescue.
-
--   Differences in allele frequencies exist between Bd-naive and recovering populations
-
--   Shared selective response across recovering populations
-
--   Changes in population growth rate due to evolution extend beyond the natal environment
-
--   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
-
--   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
-
-Then show how each result is consistent with those expectations.
 

### 2023-01-20 doc/manuscript/translocation.qmd
Update introduction outline with 5 paragraph flow

@@
 
-Severe effects of disease on global biodiversity (bats, primates, amphibians, corals, sea stars, etc.)
+P1: Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.
 
-Host evolution and population persistence/recovery
+-   In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa \[Scheele, Avain Malaria, Black-footed ferrets, sea stars, etc.\].
 
--   Species recovery in the face of severe disease (population growth rate \< 1 ) will depend on the ability of a species to move or change.
+-   While the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating negative disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).
 
--   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al.
-2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
-Restrict to species of conservation concern?).
+-   However, given the nascent nature of wildlife recoveries in the presence of disease we know far less about the mechanisms driving population recovery and how these mechanisms can be leveraged to re-establish host populations across large spatial scales.
 
--   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease.
-For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
-These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+-   In particular for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology (but see XX).
+
+P2: Host evolution following disease-induced declines is one mechanisms through which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).
+
+-   Evolutionary rescue is a type of natural host evolution where rapid evolutionary changes originating from variation within a host population reverse ongoing population declines (Carlson).
+
+-   In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes (Golas et al.
+2021).
+
+-   Important for conservation action, evolutionary rescue not only promotes recoveries of a focal populations, but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
 2019, Joseph and Knapp 2018).
 
-Population reestablishment using reintroduction of resistant hosts
+P3: For resistant populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
 
--   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
+-   Translocations of individuals from putatively resistant populations to formerly-occupied-but-now-extirpated habitats provides a way to disentangle the contributes of natal environmental and host genetics on host persistence recovery.
 
--   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+-   Something here about previous work with translocations and disease.
 
--   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
-We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
-Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+-   Simply put, resistant hosts that demonstrate disease resistance, population growth, and long-term viability in non-natal environments provide a means promote species recovery at a landscape-scale.
+
+P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+P5: Here, we combine X translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
+
+-   We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection and argue evolutionary rescue is a strong candidate mechanism for emergence of this resistance based on five lines of evidence.
+
+1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
+2.  Differences in allele frequencies exist between Bd-naive and recovering populations
+3.  There is a shared selective response across recovering populations.
+4.  Changes in population growth rate due to evolution extend beyond the natal environment
+5.
+In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+
+-   As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
 These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
-
-Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
-
-<!--# I'm not sure the following text still makes sense to include.
-If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper.
-The following is my best attempt at doing so.
-I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript.
--->
-
-We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence.
-These hypotheses generate the following expectations:
-
--   Differences in allele frequencies exist between Bd-naive and recovering populations
-
--   Shared selective response across recovering populations
-
--   Changes in population growth rate due to evolution extend beyond the natal environment
-
--   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
-
--   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
-
-Then show how each result is consistent with those expectations.
-
-<!--# Extra ER text for use in manuscript -->
-
--   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al.
-2014), can reverse ongoing declines in some species (Golas et al.
-2021; Searle et al.
-2021).
-
--   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al.
-2016, Schiebelhut et al.
-2018).
-This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
-(2013).
-
-    -   Tasmanian devils: Epstein et al. (2016), Wells et al., Ecology (2019)
-
-    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., (2014)
-
-    -   Frogs in Panama: Voyles et al. (2018)
-
-    -   Frogs in the Sierra Nevada: Knapp et al. (2016) @knapp2016
-
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
-
-    -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.
 
@@
 
--
+## Other introduction material for quick reference
+
+P5:
+
+we show using amphibian-pathogen system to demonstrate that resistance first demonstrate that first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
+We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
+Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
+
+Host evolution and population persistence/recovery
+
+-   Species recovery in the face of severe disease (population growth rate \> 1 ) will depend on the ability of a species to move or change.
+
+-   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al.
+2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
+Restrict to species of conservation concern?).
+
+-   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease.
+For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
+These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+2019, Joseph and Knapp 2018).
+
+Population reestablishment using reintroduction of resistant hosts
+
+-   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
+
+-   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+
+-   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
+We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
+Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
+
+Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+<!--# I'm not sure the following text still makes sense to include.
+If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper.
+The following is my best attempt at doing so.
+I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript.
+-->
+
+We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence.
+These hypotheses generate the following expectations:
+
+-   Differences in allele frequencies exist between Bd-naive and recovering populations
+
+-   Shared selective response across recovering populations
+
+-   Changes in population growth rate due to evolution extend beyond the natal environment
+
+-   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
+
+-   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
+
+Then show how each result is consistent with those expectations.
+
+<!--# Extra ER text for use in manuscript -->
+
+-   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al.
+2014), can reverse ongoing declines in some species (Golas et al.
+2021; Searle et al.
+2021).
+
+-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al.
+2016, Schiebelhut et al.
+2018).
+This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
+(2013).
+
+    -   Tasmanian devils: Epstein et al. (2016), Wells et al., Ecology (2019)
+
+    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., (2014)
+
+    -   Frogs in Panama: Voyles et al. (2018)
+
+    -   Frogs in the Sierra Nevada: Knapp et al. (2016) @knapp2016
+
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
+
+    -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.

### 2023-01-27 doc/manuscript/translocation.qmd
Minor edits to introduction outline

--- 2023-01-19 doc/manuscript/translocation.qmd
@@
 
-Severe effects of disease on global biodiversity (bats, primates, amphibians, corals, sea stars, etc.)
+P1: Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.
 
-Host evolution and population persistence/recovery
+-   In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa (Scheele, avian malaria, black-footed ferrets, sea stars, etc.).
 
--   Species recovery in the face of severe disease (population growth rate \< 1 ) will depend on the ability of a species to move or change.
+-   While the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).
 
--   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al.
-2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
-Restrict to species of conservation concern?).
+-   However, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery and how these mechanisms can be leveraged to re-establish host populations across large spatial scales.
 
--   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease.
-For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
-These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+-   In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology (but see XX).
+
+P2: Host evolution following disease-induced declines is one mechanisms through which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).
+
+-   Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines (Carlson).
+
+-   In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes (Golas et al.
+2021).
+
+-   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations, but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
 2019, Joseph and Knapp 2018).
 
-Population reestablishment using reintroduction of resistant hosts
+P3: For resistant populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environments.
 
--   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
+-   Translocations of individuals from putatively resistant populations to formerly-occupied-but-now-extirpated habitats provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
 
--   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
+-   Something here about previous work with translocations and disease.
 
--   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
-We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
-Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+-   Simply put, resistant hosts that demonstrate disease resistance, population growth, and long-term viability in non-natal environments provide a means to recover disease-impacted species at a landscape-scale.
+
+P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+
+P5: Here, we combine X translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
+
+-   We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection and provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
+
+1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
+2.  Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
+3.  There is a shared selective response across recovering populations.
+4.  Changes in population growth rate due to evolution extend beyond the natal environment.
+5.
+In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+
+-   As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
 These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
-
-Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
-
-<!--# I'm not sure the following text still makes sense to include.
-If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper.
-The following is my best attempt at doing so.
-I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript.
--->
-
-We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence.
-These hypotheses generate the following expectations:
-
--   Differences in allele frequencies exist between Bd-naive and recovering populations
-
--   Shared selective response across recovering populations
-
--   Changes in population growth rate due to evolution extend beyond the natal environment
-
--   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
-
--   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
-
-Then show how each result is consistent with those expectations.
-
-<!--# Extra ER text for use in manuscript -->
-
--   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al.
-2014), can reverse ongoing declines in some species (Golas et al.
-2021; Searle et al.
-2021).
-
--   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al.
-2016, Schiebelhut et al.
-2018).
-This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
-(2013).
-
-    -   Tasmanian devils: Epstein et al. (2016), Wells et al., Ecology (2019)
-
-    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., (2014)
-
-    -   Frogs in Panama: Voyles et al. (2018)
-
-    -   Frogs in the Sierra Nevada: Knapp et al. (2016) @knapp2016
-
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
-
-    -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.
 
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software \[v4.4.4; \@rsoftware2022\].
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; @rsoftware2022].
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
+|:--------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-02-07 doc/manuscript/translocation.qmd
Add selection figures and make minor text updates

@@
 editor: visual
-csl: nature.csl
+csl: pnas.csl
 bibliography: translocation.bib
@@
 
+<!--# Intended outlet is PNAS.
+Preferred length of research reports is 6 pages, but up to 12 pages is allowed.
+A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables) -->
+
 ## Title: Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
 
-<!--# 75 characters. Suggestions to make this more compelling? Use "Evolutionary rescue"? -->
+<!--# Title should be less than 135 characters -->
 
@@
 
-Roland A. Knapp^1,2^\*, Allison Q. Byrne^3,4^, Mark Q. Wilber^5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^3,4^, Rob L. Grasso^8^, Erica Bree Rosenblum^3,4^
+Roland A. Knapp^1,2^\*, Mark Q. Wilber^3^, Allison Q. Byrne^4,5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^3,4^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
 
@@
 
-^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, USA
+^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
 
-^2^Earth Research Institute, University of California, Santa Barbara, CA, USA
+^2^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
 
-^3^Department of Environmental Science, Policy, and Management, University of California. Berkeley, Berkeley, CA, USA
+^3^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
 
-^4^Museum of Vertebrate Zoology, University of California, Berkeley, Berkeley, CA, USA
+^4^Department of Environmental Science, Policy, and Management, University of California. Berkeley, Berkeley, CA, 94720-3114
 
-^5^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture, Knoxville, TN, USA
+^5^Museum of Vertebrate Zoology, University of California, Berkeley, Berkeley, CA, 94720-3160
 
-^6^Earth Lab, University of Colorado, Boulder, CO, USA
+^6^Earth Lab, University of Colorado, Boulder, CO, 80303
 
-^7^NCX, San Francisco, CA, USA
+^7^NCX, San Francisco, CA, 94115
 
-^8^Resources Management and Science, Yosemite National Park, El Portal, CA, USA
+^8^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
 
-\*Corresponding author: roland.knapp\@ucsb.edu
+\*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu
+
+**Preprint Server:**
+
+**Classification:** Biological Sciences - Ecology
 
@@
 
-<!--# No more than 200 words, follow structure described in instructions for authors.
-2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
-See [annotated example](https://www.nature.com/documents/nature-summary-paragraph.pdf).
--->
+<!--# No more than 250 words -->
 
-### Main
+### Significance Statement
 
-<!--# subheadings can be used, can be "teasers" of results (up to 40 characters) -->
+<!--# Less than 120 words. Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty.  -->
+
+### Introduction
 
@@
 
--   While the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).
+-   Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).
 
@@
 
-P2: Host evolution following disease-induced declines is one mechanisms through which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).
+P2: Host evolution following disease-induced declines is one mechanism by which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).
 
@@
 
--   Translocations of individuals from putatively resistant populations to formerly-occupied-but-now-extirpated habitats provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
+-   Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
 
@@
 P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
+Cite Knapp et al.
+(2016) to emphasize that frogs in recovering populations are more resistant to Bd than those in Bd-naive populations.
 
-P5: Here, we combine X translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
+P5: Here, we combine 24 translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
 
--   We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection and provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
+We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection and provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
 
@@
 
--   As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+<!--# These bullet points might be better addressed in the Discussion than in the Introduction. -->
+
+As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
 These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
 
-#### Frog evolution in response to Bd
-
-@fig-exomemap
+### Results
 
@@
 
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
+<!--# Likely need an introductory few sentences to put the translocations in the proper context.
+-->During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-<!--# Create table of translocation sites, years, frog numbers?
---> Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
 As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details).
@@
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
-This result is consistent with the exome capture results suggesting that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
+This result is consistent with our hypothesis that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
 Important predictors are shown in @fig-cond-effects, and included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996).
-Males had higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
-In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
+Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
+In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
+
+<!--# Add concluding 1-2 sentences to emphasize take-home message. -->
 
@@
 
+<!--# Present results suggesting what donor-specific characteristics are responsible for important effect of donor population? -->
+
+<!--# Mention the improved survival in recent compared with early translocations resulting from our better-informed site selection?
+Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations.
+-->
+
 #### Long-term persistence of reestablished populations
 
-Given observed yearly adult survival probabilities, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
+<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.
+-->Given observed yearly adult survival probabilities, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
 When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (@fig-viability A).
@@
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
 Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
@@
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest surveillance history (@fig-viability-supp B; population 70550, surveyed for 16 years).
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
 <!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations?
@@
 -->
+
+#### Frog evolution in response to Bd
+
+@fig-exomemap
+
+@fig-selectionresults
+
+@fig-allelefrequencies
 
@@
 
-![Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
-Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia National Parks (from top to bottom).
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of the study area within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation)](figures/map_exome_points.png){#fig-exomemap width="450"}
-
 ![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
@@
 See inset map in @fig-exomemap to locate Yosemite National Park within California.
-<!--# Add park boundaries to inset map?
-Also, for this and previous figure, is basemap necessary/helpful -->](figures/map_translocation_points.png){#fig-yosemap width="450"}
+<!--# Move to Supporting Information.
+Add YOSE boundary to inset map of fig-exomemap?
+Also, is basemap necessary/helpful?
+Finally, better to add color coding showing which donor was used at which sites?-->](figures/map_translocation_points.png){#fig-yosemap width="450"}
 
 ![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
-Sites are arranged along the x-axis using the average of the median survival per translocation at each site.
-Frog survival for the one-year period following each translocation was estimated using capture-mark-recapture methods.
+Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
+Frog survival was estimated using capture-mark-recapture methods.
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
@@
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
-
-<!--# Change @fig-survival-postdens from mcmc_areas to mcmc_intervals? -->
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).
+<!--# Move to Supporting Information.
+Change @fig-survival-postdens from mcmc_areas to mcmc_intervals?
+-->](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
@@
 
-![A.
+![<!--# Add first sentence that provides a brief overview of the entire figure.
+--> A.
 The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
@@
 
-![A.
+![<!--# Add first sentence that provides a brief overview of the entire figure.
+--> A.
 The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
@@
 
-<!-- Left color palette as is as 12-color color-blind friendly is not the most friendly color palette to look at.  We can discuss. -->
+<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.  We can discuss. -->
+
+![Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
+Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia National Parks (from top to bottom).
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Inset map shows the location of the study area within California.
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
+<!--# Remove this map.
+Map in Allie's fig-selectionresults is likely sufficient.
+-->](figures/map_exome_points.png){#fig-exomemap width="450"}
+
+![<!--# Add first sentence that provides a brief overview of the entire figure.
+--> (a) Map of sampling locations showing Bd-naive populations in red and persistent populations in blue.
+(b) PCA calculated from binary SNPs showing genomic relationship of samples.
+Labels match sampling location in (a) and are colored as in (a).
+(c) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
+(d) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
+SNPs are sorted by genomic position and Contigs are sorted by size.
+Red line shows Bonferroni corrected p-value for threshold of 0.01.
+Outlier SNPs above this threshold are highlighted and labeled.
+Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
+
+![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R.
+sierrae* populations.
+The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery.
+](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+
+### Tables
+
+<!--# Include table of translocation sites, years, frog numbers? -->
+
+<!--# Include Allie's table describing outlier SNPs? -->
 
@@
 
-### Methods
-
-<!-- Typically less than 3000 words -->
-
-#### Frog evolution in response to Bd
-
-<!-- subheadings can be used, up to 40 characters -->
-
-To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
-
-<!--# Mention that we sampled declining populations also? Will need to update map and text to show only the samples that were sequenced. -->
-
-DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-exomemap).
-To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
+### Materials and methods
 
@@
 
-### Data availability
+#### Frog evolution in response to Bd
+
+To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
+These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
+
+DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-exomemap).
+To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
+
+### Acknowledgements and funding sources
 
@@
 
-<!--# References for Methods section only -->
+### Figure legends
 
-### Acknowledgements
-
-### Author contributions
-
-### Competing interests
-
-The authors declare no competing interests.
-
-### Additional information
-
-Supplementary Information is available for this paper.
-
-Correspondence and requests for materials should be addressed to Roland Knapp.
-
-Peer review information.
-
-Reprints and permissions information is available at www.nature.com/reprints.
-
-### Supplementary Information
+### Supporting Information
 

### 2023-02-07 doc/manuscript/translocation.qmd
Make suggestions on wording in results to improve between section flow

@@
 
-<!--# Likely need an introductory few sentences to put the translocations in the proper context.
--->During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
+<!--# Likely need an introductory few sentences to put the translocations in the proper context. -->
+
+<!--# Mark W: Pulling from the methods, we say: "Translocations of [putatively resistant] *R.
+sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species."  Could be a good starting sentence.
+Also, as I am reading through the results, our sections do not elude to the Allie's section.
+We should discuss how our sections can set-up Allie's section with a few well-chosen words throughout.
+-->
+
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
@@
 
+<!--# Suggested text to wrap this section up and build into the next two sections: Overall, our analyses show that putatively resistant frogs can maintain the benefits of resistance (i.e., minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
+Moreover, in three locations where longer mark-recapture time series allow us to assess the survival of adults recruited to the population due to the breeding of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults.
+These results show that putative resistance in non-natal habitats is maintained across generations of frogs, suggesting that acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations.
+We test these claims in the following sections.
+-->
+
+<!--# Mark W: Regarding the above paragraph, what evidence do we have of "tolerance"?
+Are Bd loads significantly reduced in populations compared to epizootic levels?
+Can we convincingly state that the likely mechanism is more consistent with resistance or tolerance?
+I ask because it would be nice to just use one of these terms in the paper and not get caught up in any resistance vs.
+tolerance discussions with reviewers.
+-->
+
 <!--# Add concluding 1-2 sentences to emphasize take-home message. -->
@@
 Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations.
+Mark W: Yes, I think this is important.
+I think we want to drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd.
 -->
@@
 
-<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.
--->Given observed yearly adult survival probabilities, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations).
-When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (@fig-viability A).
+<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.  Mark W: I have made an attempt at this. -->
+
+To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values).
+Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
 In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
@@
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
+Overall, our model demonstrates that given observed yearly adult survival probabilities of putatively resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
+
 <!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations?
 Because we don't have accurate counts of subadults, can we even provide rough estimate of recruitment probability?
--->
+Mark W here: This would be great.
+If we have even ballpark estimates of \omega it would be useful to inclue.-->
 
 #### Frog evolution in response to Bd
+
+#### Conclusions
+
+<!--# Placeholder for our conclusions section -->
 
@@
 ![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
-To obscure the exact locations of populations, random noise was added to all point coordinates.
+To obscure the exact locations of populations, a small amount of random noise was added to all point coordinates.
 Elevations are indicated as in @fig-exomemap.
@@
 Also, is basemap necessary/helpful?
-Finally, better to add color coding showing which donor was used at which sites?-->](figures/map_translocation_points.png){#fig-yosemap width="450"}
+Finally, better to add color coding showing which donor was used at which sites?
+Mark W: I like the basemap.
+-->](figures/map_translocation_points.png){#fig-yosemap width="450"}
 
@@
 Frog survival was estimated using capture-mark-recapture methods.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.<!--# Mark W: Do we want to include error bars on these median estimates?
+-->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 
@@
 
-![<!--# Add first sentence that provides a brief overview of the entire figure.
---> A.
+![A population viability analysis of translocated *R.
+sierrae* that quantifies the long-run growth rate and extinction probability of frog populations.
+A.
 The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
@@
 
-![<!--# Add first sentence that provides a brief overview of the entire figure.
---> A.
+![Sensitivity analysis and empirical validation of the stage-structured *R.
+sierrae* model.
+A.
 The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$.
 B.
@@
 sierrae* populations.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery.
-](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
 
@@
 
+<!-- Mark W: We should specify whether the default priors were uninformative, slightly informative, etc. -->
+
 #### Population viability modeling
@@
 
-To determine the implications of acquired Bd resistance on long-term population viability, we developed a dynamic, population model for *R.
+To determine the implications of observed yearly adult survival on the long-term population viability of putatively resistant frogs in non-natal habitats, we developed a population model for *R.
 sierrae*.
@@
 We also did not directly consider the dynamics of Bd in this model.
-We made this decision because i) all translocated populations are infected with Bd at high prevalence, and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
+We made this decision because i) translocated populations are infected with Bd at high prevalence [@joseph2018], and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
@@
 
-The baseline parameter values for the model and how they were estimated are given in Table 1.
+The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
 Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
@@
 For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-Table 1: Description and values of parameters used in the model.
-All survival probabilities are in the presence of the fungal pathogen Bd.
-<!--# Roland, please add any additional details needed to the "Source" column.
-We ideally want to keep this succinct.
--->
 
@@
 | $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
+
+: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
 

### 2023-02-08 doc/manuscript/translocation.qmd
Minor additions to previously-added comments

@@
 sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species."  Could be a good starting sentence.
-Also, as I am reading through the results, our sections do not elude to the Allie's section.
+Also, as I am reading through the results, our sections do not allude to the Allie's section.
 We should discuss how our sections can set-up Allie's section with a few well-chosen words throughout.
@@
 tolerance discussions with reviewers.
+Roland: Have a look at this paragraph of Knapp et al.
+PeerJ 2022: https://peerj.com/articles/12712/#p-61.
+We argue that there is evidence for resistance and tolerance.
 -->
@@
 Mark W here: This would be great.
-If we have even ballpark estimates of \omega it would be useful to inclue.-->
+If we have even ballpark estimates of \omega it would be useful to include.-->
 
 #### Frog evolution in response to Bd
-
-#### Conclusions
-
-<!--# Placeholder for our conclusions section -->
 
@@
 @fig-allelefrequencies
+
+#### Conclusions
+
+<!--# Placeholder for our conclusions section -->
 

### 2023-02-10 doc/manuscript/translocation.qmd
Fix author affiliation superscript
Edits to outline of introduction

@@
 
-Roland A. Knapp^1,2^\*, Mark Q. Wilber^3^, Allison Q. Byrne^4,5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^3,4^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
+Roland A. Knapp^1,2^\*, Mark Q. Wilber^3^, Allison Q. Byrne^4,5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
 
@@
 
--   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations, but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+-   Evolutionary rescue be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al.
+2021, Lilley et al.
+2020).
+<!--# Maybe move this to the Discussion?
+-->
+
+-   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations (Gignous-Wolfsohn et al.
+2021), but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
 2019, Joseph and Knapp 2018).
@@
 
--   Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
+-   Translocation of individuals from putatively resistant <!--# Can we use "reduced susceptibilty" strategically to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlyies lower Bd loads and higher frog survival?
+--> populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
 
@@
 -   Simply put, resistant hosts that demonstrate disease resistance, population growth, and long-term viability in non-natal environments provide a means to recover disease-impacted species at a landscape-scale.
+
+-   <!--# Zooming out here to make a broader point --> Few, if any, studies of disease-impacted species have provided comprehensive evidence that (1) evolutionary rescue can increase host resistance, (2) increased host resistance contributes to higher individual fitness (e.g., survival) in the presence of disease and allows population growth, (3) higher individual fitness and positive population growth applies beyond the natal environment, (4) individuals from higher-fitness populations can be used in translocations to reestablish extirpated populations, and (5) many reestablished populations have a high probability of persisting over the long term in a landscape of disease.
+<!--# I don't know how many of these boxes we can check, but they seem like important points to help us lay the foundation for our study (which checks more of these boxes than do most studies).
+I actually wonder if taking the broader approach outlined in this bullet would make for a more compelling argument than one focused more narrowly on issue #3, as most of the bullets in this paragraph currently do.
+That issue, although important, may distract from the broader perspective that I tried to highlight.
+-->
 
@@
 
-We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection and provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
-
-1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
-2.  Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
-3.  There is a shared selective response across recovering populations.
-4.  Changes in population growth rate due to evolution extend beyond the natal environment.
-5.
-In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-
-<!--# These bullet points might be better addressed in the Discussion than in the Introduction. -->
+We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection.
 
@@
 <!--# Placeholder for our conclusions section -->
+
+We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
+<!--# Consider broadening the scope of this sentence and the paragraph to be developed from it to encompass all of our findings -->
+
+1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
+2.  Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
+3.  There is a shared selective response across recovering populations.
+4.  Changes in population growth rate due to evolution extend beyond the natal environment.
+5.
+In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
 

### 2023-02-10 doc/manuscript/translocation.qmd
Additional edits to comments in introduction

@@
 -   <!--# Zooming out here to make a broader point --> Few, if any, studies of disease-impacted species have provided comprehensive evidence that (1) evolutionary rescue can increase host resistance, (2) increased host resistance contributes to higher individual fitness (e.g., survival) in the presence of disease and allows population growth, (3) higher individual fitness and positive population growth applies beyond the natal environment, (4) individuals from higher-fitness populations can be used in translocations to reestablish extirpated populations, and (5) many reestablished populations have a high probability of persisting over the long term in a landscape of disease.
-<!--# I don't know how many of these boxes we can check, but they seem like important points to help us lay the foundation for our study (which checks more of these boxes than do most studies).
+<!--# I don't know how many of these boxes we can check convincingly, but they seem like important points to help us lay the foundation for our study (which checks more of these boxes than do most studies).
 I actually wonder if taking the broader approach outlined in this bullet would make for a more compelling argument than one focused more narrowly on issue #3, as most of the bullets in this paragraph currently do.
@@
 Cite Knapp et al.
-(2016) to emphasize that frogs in recovering populations are more resistant to Bd than those in Bd-naive populations.
+(2016) to emphasize that frogs in recovering populations are less susceptible to Bd than those in Bd-naive populations.
 

### 2023-02-11 doc/manuscript/translocation.qmd
Add text to introduction

--- 2023-02-08 doc/manuscript/translocation.qmd
@@
 
--   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations, but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
+-   Evolutionary rescue be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al.
+2021, Lilley et al.
+2020).
+<!--# Maybe move this to the Discussion?
+-->
+
+-   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations (Gignous-Wolfsohn et al.
+2021), but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
 2019, Joseph and Knapp 2018).
 
-P3: For resistant populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environments.
+P3: <!--# In this paragraph, I took a broader approach than what was originally outlined (which focused on fitness benefits extending beyond natal environment).
+I'm happy to revisit this approach as necessary.
+-->
+
+<!--# Can we use "reduced susceptibilty" strategically to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlies lower Bd loads and higher frog survival?
+-->
 
@@
 
--   Something here about previous work with translocations and disease.
-
--   Simply put, resistant hosts that demonstrate disease resistance, population growth, and long-term viability in non-natal environments provide a means to recover disease-impacted species at a landscape-scale.
+Given the dramatic impact of disease on global wildlife populations, reintroductions will often be necessary to reestablish extirpated populations at a landscape scale.
+However, there are several conditions that will determine the feasibility of this much-needed recovery approach.
+First, populations that survived following disease arrival have evolved increased resistance.
+Second, increased resistance contributes to higher individual fitness in the presence of disease, and allows positive population growth.
+Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments.
+And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease.
+To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea (@scheele2021).
+Furthermore, a lack of relevant information about disease or a lack of suitable options to overcome the threat of disease has resulted in the failure of numerous reintroduction efforts (@scheele2021, @fellers2007).
+With the increase in emerging infectious diseases in recent decades (@daszak2000), these failures highlight the critical need to develop effective tools to recover disease-impacted biodiversity worldwide.
+<!--# Add text mentioning that most discussion of reintroductions/disease has focused on protecting wild populations from diseases vectored by reintroduced individuals?
+(Viggers et al.
+1993, Ewan et al.
+2012, Hartley and Sainsbury 2017)  -->
 
@@
 Cite Knapp et al.
-(2016) to emphasize that frogs in recovering populations are more resistant to Bd than those in Bd-naive populations.
+(2016) to emphasize that frogs in recovering populations are less susceptible to Bd than those in Bd-naive populations.
+
+The mountain yellow-legged frog, comprised of the sister species *Rana muscosa* and *Rana sierrae*, is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; @scheele2019).
+Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; @grinnell1924), this frog has disappeared from more than 90% of its historical range during the past century.
+Due to the severity of its decline and the increasing probability of extinction, both species as now listed as "endangered" under the U.S.
+Endangered Species Act.
+Although the introduction of non-native trout into the once-fishless portion of the Sierra Nevada initiated this decline (@bradford1989; @knapp2000), the arrival of *B.
+dendrobatidis* (Bd) in the 1960s and its subsequent spread (@vredenburg2019) caused additional widespread population extirpations (@vredenburg2010, @rachowicz2006).
+These Bd-caused declines are fundamentally different than the fish-caused declines.
+The removal of fish populations is feasible and results in the rapid recovery of frogs.
+In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts (reference), and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.
 

### 2023-02-11 doc/manuscript/translocation.qmd
Add additional paragraphs to introduction

--- 2023-02-10 doc/manuscript/translocation.qmd
@@
 
--   Evolutionary rescue be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al.
+-   Evolutionary rescue can be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al.
 2021, Lilley et al.
@@
 
-P3: For resistant populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environments.
-
--   Translocation of individuals from putatively resistant <!--# Can we use "reduced susceptibilty" strategically to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlyies lower Bd loads and higher frog survival?
---> populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
-
--   Something here about previous work with translocations and disease.
-
--   Simply put, resistant hosts that demonstrate disease resistance, population growth, and long-term viability in non-natal environments provide a means to recover disease-impacted species at a landscape-scale.
-
--   <!--# Zooming out here to make a broader point --> Few, if any, studies of disease-impacted species have provided comprehensive evidence that (1) evolutionary rescue can increase host resistance, (2) increased host resistance contributes to higher individual fitness (e.g., survival) in the presence of disease and allows population growth, (3) higher individual fitness and positive population growth applies beyond the natal environment, (4) individuals from higher-fitness populations can be used in translocations to reestablish extirpated populations, and (5) many reestablished populations have a high probability of persisting over the long term in a landscape of disease.
-<!--# I don't know how many of these boxes we can check, but they seem like important points to help us lay the foundation for our study (which checks more of these boxes than do most studies).
-I actually wonder if taking the broader approach outlined in this bullet would make for a more compelling argument than one focused more narrowly on issue #3, as most of the bullets in this paragraph currently do.
-That issue, although important, may distract from the broader perspective that I tried to highlight.
+<!--# P3: In this paragraph, I took a broader approach than what was originally outlined (which focused on fitness benefits extending beyond natal environment).
+I'm happy to revisit this approach as necessary.
+Also, can we use "reduced susceptibilty" strategically (especially in Results) to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlies lower Bd loads and higher frog survival?
 -->
 
-P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
-Cite Knapp et al.
-(2016) to emphasize that frogs in recovering populations are more resistant to Bd than those in Bd-naive populations.
+<!--# The following point was made in the original outline but is not included in P3, P4, or P5.
+Could be useful to include when introducing translocations in Results: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
+-->
 
-P5: Here, we combine 24 translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
+Given the dramatic impact of disease on global wildlife populations, reintroductions will often be necessary to reestablish extirpated populations at a landscape scale.
+However, there are several conditions that will often determine the feasibility of this much-needed recovery approach.
+First, populations that survived following disease arrival have evolved increased resistance.
+Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth.
+Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments.
+And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease.
+To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021].
+Furthermore, a lack of relevant information about disease or a lack of suitable options to overcome the threat of disease has resulted in the failure of numerous reintroduction efforts [@scheele2021; @fellers2007].
+With the increase in emerging infectious diseases in recent decades [@daszak2000], these failures highlight the critical need to develop effective tools to recover disease-impacted biodiversity worldwide.
+<!--# Need to better tie these two sentences for the conditins listed above.
+-->
 
+<!--# P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery. -->
+
+The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; [@scheele2019]).
+Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; [@grinnell1924]), this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
+Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
+Endangered Species Act.
+Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B.
+dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
+These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs.
+In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.
+
+<!--# P5: Here, we combine 24 translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
 We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection.
+As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
+-->
 
-As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
-These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted following epizootics [@briggs2010] and are now recovering [@knapp2016].
+Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
+Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
+Here, we report the results of a greatly expanded MYL frog reintroduction effort and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
+Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
+These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
 

### 2023-02-12 doc/manuscript/translocation.qmd
Draft paragraph 1 and 2 of introduction

@@
 
-P1: Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.
+Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.
+In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
+Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts [@langwig2015], populations previously decimated by disease are also beginning to recover [@newell2013; @voyles2018].
+Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
+In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology [@venesky2012; @mendelson2019].
 
--   In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa (Scheele, avian malaria, black-footed ferrets, sea stars, etc.).
-
--   Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).
-
--   However, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery and how these mechanisms can be leveraged to re-establish host populations across large spatial scales.
-
--   In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology (but see XX).
-
-P2: Host evolution following disease-induced declines is one mechanism by which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).
-
--   Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines (Carlson).
-
--   In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes (Golas et al.
-2021).
-
--   Evolutionary rescue can be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al.
-2021, Lilley et al.
-2020).
-<!--# Maybe move this to the Discussion?
--->
-
--   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations (Gignous-Wolfsohn et al.
-2021), but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
-2019, Joseph and Knapp 2018).
+Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
+Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014].
+In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
+Evolutionary changes following disease-induced declines have been observed in black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016].
+However, linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
+Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence [@joseph2018; @mendelson2019].
+These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate other extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.
 
@@
 -->
+
+<!--#  Suggested start to this paragraph to link with the previous paragraph: "There are four conditions that will often..." -->
 

### 2023-02-13 doc/manuscript/translocation.qmd
Remove specific examples from intro paragraph 2

@@
 In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-Evolutionary changes following disease-induced declines have been observed in black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016].
-However, linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
+Evolutionary changes following disease-induced declines have been repeatedly observed in wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
 Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence [@joseph2018; @mendelson2019].
-These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate other extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.
+These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.
+
+<!--# Example text with more specific examples: black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016]-->
 

### 2023-02-14 doc/manuscript/translocation.qmd
Moved 2 figures to supplement and fixed typos

@@
 Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts [@langwig2015], populations previously decimated by disease are also beginning to recover [@newell2013; @voyles2018].
-Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
+Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
 In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology [@venesky2012; @mendelson2019].
@@
 
-Given the dramatic impact of disease on global wildlife populations, reintroductions will often be necessary to reestablish extirpated populations at a landscape scale.
-However, there are several conditions that will often determine the feasibility of this much-needed recovery approach.
+There are four conditions that will often determine the feasibility of this much-needed recovery approach.
 First, populations that survived following disease arrival have evolved increased resistance.
@@
 With the increase in emerging infectious diseases in recent decades [@daszak2000], these failures highlight the critical need to develop effective tools to recover disease-impacted biodiversity worldwide.
-<!--# Need to better tie these two sentences for the conditins listed above.
+<!--# Need to better tie these two sentences to the conditions listed above.
 -->
@@
 
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap).
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
+-->.
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
@@
 
-@fig-exomemap
-
 @fig-selectionresults
@@
 
-![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
-To obscure the exact locations of populations, a small amount of random noise was added to all point coordinates.
-Elevations are indicated as in @fig-exomemap.
-See inset map in @fig-exomemap to locate Yosemite National Park within California.
-<!--# Move to Supporting Information.
-Add YOSE boundary to inset map of fig-exomemap?
-Also, is basemap necessary/helpful?
-Finally, better to add color coding showing which donor was used at which sites?
-Mark W: I like the basemap.
--->](figures/map_translocation_points.png){#fig-yosemap width="450"}
-
 ![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
@@
 Frog survival was estimated using capture-mark-recapture methods.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.<!--# Mark W: Do we want to include error bars on these median estimates?
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
+<!--# Include error bars on these median estimates.
 -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
-
-![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).
-<!--# Move to Supporting Information.
-Change @fig-survival-postdens from mcmc_areas to mcmc_intervals?
--->](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
@@
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
+<!--# Switch C and D panels.
+-->](figures/cond_effects_plot.png){#fig-cond-effects}
 
@@
 
-![Map showing the locations of Bd-naive or recovering frog populations from which DNA samples were collected for the exome capture analyses.
-Black polygons indicate the boundaries of Yosemite, Kings Canyon, and Sequoia National Parks (from top to bottom).
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of the study area within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Remove this map.
-Map in Allie's fig-selectionresults is likely sufficient.
--->](figures/map_exome_points.png){#fig-exomemap width="450"}
-
 ![<!--# Add first sentence that provides a brief overview of the entire figure.
@@
 sierrae* populations.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
 
@@
 DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-exomemap).
+These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-selectionresults).
 To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
@@
 ### Supporting Information
+
+![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Inset map shows the location of the study area within California.
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
+<!--# Add color coding showing which donor was used at which sites.
+Add inset map locating YOSE within California.
+Zoom in on eastern YOSE only instead of entire park.
+-->](figures/map_translocation_points.png){#fig-yosemap width="450"}
+
+![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 

### 2023-02-14 doc/manuscript/translocation.qmd
Added inset map to Yosemite map and changed symbology

@@
 ![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
+Symbol shapes indicate the donor population used for each translocation site.
 To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of the study area within California.
+Inset map shows the location of Yosemite within California.
 In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Add color coding showing which donor was used at which sites.
-Add inset map locating YOSE within California.
-Zoom in on eastern YOSE only instead of entire park.
--->](figures/map_translocation_points.png){#fig-yosemap width="450"}
+](figures/map_translocation_points.png){#fig-yosemap width="450"}
 

### 2023-02-16 doc/manuscript/translocation.qmd
Minor edits to comments

@@
 These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.
+<!--# Roland says: Modestly speaking, Joseph and Knapp (2018) does provide an example, and one that predates Mendelson (2019) paper.
+-->
 
@@
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-](figures/map_translocation_points.png){#fig-yosemap width="450"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="450"}
 

### 2023-02-16 doc/manuscript/translocation.qmd
Add transition between Introduction and Results sections

@@
 
-<!--# Mark W: Pulling from the methods, we say: "Translocations of [putatively resistant] *R.
-sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species."  Could be a good starting sentence.
-Also, as I am reading through the results, our sections do not allude to the Allie's section.
+<!--# Mark W: As I am reading through the results, our sections do not allude to the Allie's section.
 We should discuss how our sections can set-up Allie's section with a few well-chosen words throughout.
@@
 
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
+Reintroductions of *R.
+sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+These reintroductions involved the collection of adult frogs from nearby large donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to recipient sites.
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
 -->.

### 2023-02-21 code/translocation_survival_analysis_suppl.Rmd
Revise axis labels of bd load figure

@@
     geom_boxplot() + 
+    labs(x = "Period", y = "Bd load (log10(bd_load + 1))") +
     facet_grid(. ~ site_id1) +
     theme(axis.text.x = element_text(angle = 90))
+ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 5, height = 3)
 ```

### 2023-02-21 doc/manuscript/translocation.qmd
Edits throughout introduction and results

@@
 
-Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.
-In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts [@langwig2015], populations previously decimated by disease are also beginning to recover [@newell2013; @voyles2018].
-Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
-In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology [@venesky2012; @mendelson2019].
+Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations.
+In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
+Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007].
+Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
+Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
+At a time when novel infectious diseases are emerging at an unprecedented rate, this knowledge gap has far-reaching consequences for global biodiversity.
 
@@
 In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-Evolutionary changes following disease-induced declines have been repeatedly observed in wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
-Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence [@joseph2018; @mendelson2019].
-These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.
-<!--# Roland says: Modestly speaking, Joseph and Knapp (2018) does provide an example, and one that predates Mendelson (2019) paper.
+Evolutionary changes following disease-induced declines are known from a diversity wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
+Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
+To date, this potentially important species recovery strategy is rarely utilized in wildlife conservation efforts [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
+<!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example.
 -->
 
-<!--# Example text with more specific examples: black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016]-->
-
-<!--# P3: In this paragraph, I took a broader approach than what was originally outlined (which focused on fitness benefits extending beyond natal environment).
-I'm happy to revisit this approach as necessary.
-Also, can we use "reduced susceptibilty" strategically (especially in Results) to avoid potential pitfalls associated with resistance, tolerance, and our lack of understanding of what actual mechanism underlies lower Bd loads and higher frog survival?
--->
-
-<!--# The following point was made in the original outline but is not included in P3, P4, or P5.
-Could be useful to include when introducing translocations in Results: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
--->
-
-<!--#  Suggested start to this paragraph to link with the previous paragraph: "There are four conditions that will often..." -->
-
-There are four conditions that will often determine the feasibility of this much-needed recovery approach.
+There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
 First, populations that survived following disease arrival have evolved increased resistance.
@@
 To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021].
-Furthermore, a lack of relevant information about disease or a lack of suitable options to overcome the threat of disease has resulted in the failure of numerous reintroduction efforts [@scheele2021; @fellers2007].
-With the increase in emerging infectious diseases in recent decades [@daszak2000], these failures highlight the critical need to develop effective tools to recover disease-impacted biodiversity worldwide.
-<!--# Need to better tie these two sentences to the conditions listed above.
--->
-
-<!--# P4: Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery. -->
 
@@
 dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
-These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs.
+These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
 In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.
 
-<!--# P5: Here, we combine 24 translocations of resistant MYL populations, X longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that resistant populations of an endangered amphibian species can recover and persist across a landscape of disease.
-We then use genetic evidence to demonstrate that observed disease resistance is likely a result of selection.
-As such, we conclude that natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
--->
-
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted following epizootics [@briggs2010] and are now recovering [@knapp2016].
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
@@
 
+<!--# The following point was made in the original outline but is not included in P3, P4, or P5.
+Could be useful to include when introducing translocations in Results: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
+-->
+
 ### Results
 
-#### Translocations allow population reestablishment
+<!--# Let's revisit whether the selection results should go here or at the end of the results. Putting those results here would reduce the need to describe populations as "putatively resistant". -->
+
+#### Reintroductions allow population reestablishment
 
@@
 
-<!--# Likely need an introductory few sentences to put the translocations in the proper context. -->
-
 <!--# Mark W: As I am reading through the results, our sections do not allude to the Allie's section.
@@
 sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-These reintroductions involved the collection of adult frogs from nearby large donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to recipient sites.
+These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations <!--# Add figure to SI showing Bd loads before and after translocations -->that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
+We refer to these reintroductions as "translocations".
 During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
@@
 Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details).
-The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival.
-<!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)?
--->
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28 <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)?
+--> ; see Supporting Information for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
+For frogs translocated to each site, Bd load changed little between the periods immediately prior to translocation versus during the one year after translocation (see SI for analysis details).
 
+To identify drivers of post-translocation frog survival, we developed a hierarchical model in a Bayesian framework (see SI for details).
+Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
-This result is consistent with our hypothesis that frogs in recovering populations have evolved increased resistance and/or tolerance to Bd infection.
-Important predictors are shown in @fig-cond-effects, and included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996).
+Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
 Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
 In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-
-<!--# Suggested text to wrap this section up and build into the next two sections: Overall, our analyses show that putatively resistant frogs can maintain the benefits of resistance (i.e., minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
-Moreover, in three locations where longer mark-recapture time series allow us to assess the survival of adults recruited to the population due to the breeding of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults.
-These results show that putative resistance in non-natal habitats is maintained across generations of frogs, suggesting that acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations.
-We test these claims in the following sections.
--->
+In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance and/or tolerance to Bd infection.
+In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
+-->.
+Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.
 
@@
 
-<!--# Add concluding 1-2 sentences to emphasize take-home message. -->
-
-<!--# Present results indicating that Bd loads on frogs changed little following translocation? -->
-
 <!--# Present results suggesting what donor-specific characteristics are responsible for important effect of donor population? -->
@@
 #### Long-term persistence of reestablished populations
-
-<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.  Mark W: I have made an attempt at this. -->
 
@@
 
+![For frogs translocated to each recipient site, change in Bd load between the periods immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+
 ### Notes (author use only - not part of manuscript)

### 2023-02-22 doc/manuscript/translocation.qmd
Add fisher citation and adjust intro wording

@@
 
-^3^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
+^3^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
 
@@
 Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
-At a time when novel infectious diseases are emerging at an unprecedented rate, this knowledge gap has far-reaching consequences for global biodiversity.
+At a time when novel infectious diseases are emerging at an unprecedented rate [@fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.
 
@@
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of a greatly expanded MYL frog reintroduction effort and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
+Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability modeling, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
 Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
@@
 
-<!--# Let's revisit whether the selection results should go here or at the end of the results. Putting those results here would reduce the need to describe populations as "putatively resistant". -->
+<!--# Let's revisit whether the selection results should go here or at the end of the results.
+Putting those results here would reduce the need to describe populations as "putatively resistant".
+Mark W: Agreed.
+-->
 
@@
 In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance and/or tolerance to Bd infection.
+In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance to Bd infection.
 In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
@@
 We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values).
-Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
 When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
@@
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
-Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Therefore, even under conservative assumptions about recruitment probability $\omega$ (a parameter where ), these populations had some probability of persistence over 50 years following a single introduction.
 For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
@@
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
-Overall, our model demonstrates that given observed yearly adult survival probabilities of putatively resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
+Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
 

### 2023-02-22 code/translocation_survival_analysis.Rmd
Change plot order of conditional effects figure
Edit figures, add those and other figures to manuscript

@@
   ggtitle("Model with non-standardized predictors only",
-          "Posterior distributions with medians & 90% uncertainty intervals")
-
-mcmc_areas(m1a, area_method = "equal height", prob = 0.90) +
+          "Posterior distributions with medians & 95% uncertainty intervals")
+
+mcmc_areas(m1a, area_method = "equal height", prob = 0.95) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with non-standardized predictors only",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 -   Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.\
--   Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+-   Based on 95% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
 
@@
 
-mcmc_areas(m1b, area_method = "equal height", prob = 0.90) +
+mcmc_areas(m1b, area_method = "equal height", prob = 0.95) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictors only",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 -   As expected, using standardized predictors allows easy comparison across predictors. Important predictors are unchanged.
--   Based on 90% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
+-   Based on 95% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
 
@@
 
-mcmc_areas(m1c, area_method = "equal height", prob = 0.90, pars = b1) +
+mcmc_areas(m1c, area_method = "equal height", prob = 0.95, pars = b1) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictors & site_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 
--   Based on 90% uncertainty intervals, important predictors reduced to elevation and male.
+-   Based on 95% uncertainty intervals, important predictors reduced to elevation and male.
 
@@
 -   Based on loo, model with translocation_id as grouping variable provides slightly better fit than model with site_id as grouping variable.
--   Based on 90% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
+-   Based on 95% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
 
@@
 
-mcmc_areas(m1e, area_method = "equal height", prob = 0.90, pars = b1) +
+mcmc_areas(m1e, area_method = "equal height", prob = 0.95, pars = b1) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictors & translocation_id/site_id as grouping variables",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
     labs(x = "Sex", y = "P(1-year frog survival)") +
-    annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "C", size = 4) +
+    annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "D", size = 4) +
     theme_classic()
@@
     labs(x = "Donor population", y = "P(1-year frog survival)") +
-    annotate(geom = "text", x = 3.3, y = 0.83, hjust = 0, label = "D", size = 4) +
+    annotate(geom = "text", x = 3.3, y = 0.94, hjust = 0, label = "C", size = 4) +
     theme_classic()
@@
 ```{r}
-cond_effects_plot <- (p_snowt1 | p_elev) / (p_sex | p_donor)
+cond_effects_plot <- (p_snowt1 | p_elev) / (p_donor | p_sex)
 ```
@@
 
-mcmc_areas(m2a, area_method = "equal height", prob = 0.90, pars = b2) +
+mcmc_areas(m2a, area_method = "equal height", prob = 0.95, pars = b2) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with unstandardized predictor & no grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 
-## Model with standardized predictor, no grouping variable
+## Model with standardized predictors, no grouping variable
 
@@
 
-mcmc_areas(m2b, area_method = "equal height", prob = 0.90, pars = b3) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictor & no grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+labels_alt <- c("survival_trans1", "(Intercept)")
+mcmc_areas(m2b, area_method = "equal height", prob = 0.95, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +
+  scale_y_discrete(labels = labels_alt, limits = rev)
+#  ggtitle("Model with standardized predictor & no grouping variable",
+#          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 
-## Model with standardized predictor, translocation_id grouping variable
+### Save plot
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m2b.png"), bg = "white", width = 7, height = 4)
+```
+
+## Model with standardized predictors, translocation_id grouping variable
 
@@
 
-mcmc_areas(m2c, area_method = "equal height", prob = 0.90, pars = b3) +
+mcmc_areas(m2c, area_method = "equal height", prob = 0.95, pars = b3) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictor & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 
@@
 
-## Model with standardized predictor, site_id grouping variable
+## Model with standardized predictors, site_id grouping variable
 
@@
 
-mcmc_areas(m2d, area_method = "equal height", prob = 0.90, pars = b3) +
+mcmc_areas(m2d, area_method = "equal height", prob = 0.95, pars = b3) +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
   ggtitle("Model with standardized predictor & site_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
+          "Posterior distributions with medians & 95% uncertainty intervals")
 

### 2023-02-22 doc/manuscript/translocation.qmd
Edit figures, add those and other figures to manuscript

--- 2023-02-21 doc/manuscript/translocation.qmd
@@
 sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations <!--# Add figure to SI showing Bd loads before and after translocations -->that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
+These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
 We refer to these reintroductions as "translocations".
@@
 Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28 <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)?
---> ; see Supporting Information for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-For frogs translocated to each site, Bd load changed little between the periods immediately prior to translocation versus during the one year after translocation (see SI for analysis details).
+Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
+For translocated frogs, Bd load changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details).
+<!--# Add text describing that loads were below lethal threshold -->
 
-To identify drivers of post-translocation frog survival, we developed a hierarchical model in a Bayesian framework (see SI for details).
+To identify drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
 Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
@@
 In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance and/or tolerance to Bd infection.
-In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
+In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection.
+In addition, the small changes in Bd load following translocation and relatively high survival of frogds in the majority of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bd loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
 Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
@@
 
-![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) sex, and (D) donor population.
+![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-<!--# Switch C and D panels.
--->](figures/cond_effects_plot.png){#fig-cond-effects}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
 
@@
 
-### Materials and methods
+### Materials and Methods
 
@@
 
+<!--# Add text and results describing Bd load before versus after translocation. -->
+
 ![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
@@
 
-![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for all predictor variables.
+![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
 Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
-![For frogs translocated to each recipient site, change in Bd load between the periods immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+
+![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
+A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
 

### 2023-02-22 doc/manuscript/translocation.qmd
Push edit to table header

@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------|:------------------|:-------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-02-23 doc/manuscript/translocation.qmd
Changed formatting option to wrap text by sentences
Add table of viability for early-late translocations

--- 2023-02-22 doc/manuscript/translocation.qmd
@@
 
-^4^Department of Environmental Science, Policy, and Management, University of California. Berkeley, Berkeley, CA, 94720-3114
+^4^Department of Environmental Science, Policy, and Management, University of California.
+Berkeley, Berkeley, CA, 94720-3114
 
@@
 To date, this potentially important species recovery strategy is rarely utilized in wildlife conservation efforts [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
-<!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example.
--->
+<!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example. -->
 
@@
 
-Reintroductions of *R.
-sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
 These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
@@
 Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-For translocated frogs, Bd load changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details).
-<!--# Add text describing that loads were below lethal threshold -->
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details).
+This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (\@tbl-viability-earlylate; Table 2) <!--# Why isn't cross-referencing working for this table?
+-->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+sierrae*.
+Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018).
+<!--# Add median Bd load, proportion below 5.8 copies threshold?
+The text repeats in part similar results in Methods -- Translocaiton.
+-->
 
-To identify drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
+To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
 Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
@@
 In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection.
-In addition, the small changes in Bd load following translocation and relatively high survival of frogds in the majority of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bd loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats.
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
+In addition, the small changes in Bd load following translocation and relatively high survival of frogs in a substantial number of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats (i.e., low-moderate Bd loads, minimal Bd-induced mortality).
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
 -->.
-Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.
+This result shows that frog resistance in non-natal habitats is maintained across generations of frogs.
 
@@
 
-<!--# Present results suggesting what donor-specific characteristics are responsible for important effect of donor population? -->
-
-<!--# Mention the improved survival in recent compared with early translocations resulting from our better-informed site selection?
-Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations.
-Mark W: Yes, I think this is important.
-I think we want to drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd.
+<!--# Drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd.
 -->
@@
 
-#### Conclusions
-
-<!--# Placeholder for our conclusions section -->
+#### Discussion
 
@@
 In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+
+<!--# Discuss effect of donor population on survival of translocated frogs. Differences in Bd resistance? -->
 
@@
 
-Translocations of *R.
-sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
 We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016].
@@
 During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R.
-sierrae* tadpoles and subadults.
-These *R.
-sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
+These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
 
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to construct CMR datasets and fit models for each translocated population is available from a GitHub repository ([https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.)).
+Code used to construct CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
 
@@
 We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
-We started with models that included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
+We used default, weakly informative priors.
+Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
 Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
@@
 Code to create datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-
-<!-- Mark W: We should specify whether the default priors were uninformative, slightly informative, etc. -->
 
@@
 
+![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
+A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
+
+![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
+<!--# Remove gray background and gridlines -->](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
@@
 
-![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+```{r}
+#| label: tbl-viability-earlylate
+#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013), respectively) for which the probability of extinction over 50 years was 1 or < 1 (viability = 0 or 1, respectively)."
+#| echo: FALSE
+#| message: FALSE
 
-![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
+viability_table <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(site_id, year) %>% 
+  group_by(site_id) %>% 
+  filter(year == min(year)) %>% 
+  mutate(site_id = as.character(site_id),
+         viability = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), 0, 1),
+         period = if_else(year < 2013, "early", "late")) %>% 
+  group_by(period) %>% 
+  count(period, viability) %>% 
+  mutate(n_period = sum(n), 
+         prop = (n / n_period)) %>% 
+  mutate(across(prop, round, 3)) %>% 
+  arrange(viability, period)
+
+  knitr::kable(viability_table, "pipe", align = "lcccr", caption = )
+
+```
 
@@
 2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
-Restrict to species of conservation concern?).
+    Restrict to species of conservation concern?)
+    .
 
@@
     For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
-These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al.
-2019, Joseph and Knapp 2018).
+    These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al. 2019, Joseph and Knapp 2018).
 
@@
 
--   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al.
-2016, Schiebelhut et al.
-2018).
+-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018).
     This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
@@
 
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al. PNAS (2018)
+    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
+        PNAS (2018)
 

### 2023-02-23 code/translocation_survival_analysis_suppl.Rmd
Add error bars and remove background from figures

@@
     facet_grid(. ~ site_id1) +
-    theme(axis.text.x = element_text(angle = 90))
-ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 5, height = 3)
+    theme_bw() +
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
+          strip.text.x = element_text(size = 8),
+          panel.grid = element_blank())
+ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 6, height = 3)
 ```

### 2023-02-23 code/translocation_survival_exploratory.Rmd
Add error bars and remove background from figures

--- 2022-11-09 code/translocation_survival_exploratory.Rmd
@@
   filter(years_since_intro == 1) %>% 
-  select(site_id, release_date, years_since_intro, median_survival)
+  select(site_id, release_date, years_since_intro, median_survival, lo_survival, hi_survival)
 rm(files)
@@
   select(-plot_survival) %>% 
-    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, group = donor)) +  # reorder site_id by siteid_order)
-      geom_point(aes(color = donor), size = 3) +
+    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor)) +  # reorder site_id by siteid_order)
+      geom_errorbar(aes(ymin = lo_survival, ymax = hi_survival), color = "black", width = 0.15, position = position_dodge2(width = 0.3)) +
+      geom_point(size = 3) +
       scale_color_manual(values = palette1) + 
-      labs(x = "Recipient site", y = "1-year frog survival")
+      labs(x = "Recipient site", y = "1-year frog survival") +
+      theme_classic() +
+      theme(
+        legend.background = element_rect(color = "black", size = 0.3),
+        legend.position = c(0.2, 0.8),
+        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
+
+cohort_survival_plot
+
+# Unable to dodge points and error bars together using position_dodge and group = site_id. Why?
 ```

### 2023-02-23 doc/manuscript/translocation.qmd
Add error bars and remove background from figures

@@
 
-^4^Department of Environmental Science, Policy, and Management, University of California.
-Berkeley, Berkeley, CA, 94720-3114
+^4^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
 
-^5^Museum of Vertebrate Zoology, University of California, Berkeley, Berkeley, CA, 94720-3160
+^5^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
 
@@
 
-<!--# The following point was made in the original outline but is not included in P3, P4, or P5.
-Could be useful to include when introducing translocations in Results: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
--->
-
 ### Results
 
-<!--# Let's revisit whether the selection results should go here or at the end of the results.
-Putting those results here would reduce the need to describe populations as "putatively resistant".
-Mark W: Agreed.
--->
+<!--# Add Allie's selection results here -->
 
@@
 
-<!--# Mark W: As I am reading through the results, our sections do not allude to the Allie's section.
-We should discuss how our sections can set-up Allie's section with a few well-chosen words throughout.
--->
-
 Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
@@
 Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018).
-<!--# Add median Bd load, proportion below 5.8 copies threshold?
-The text repeats in part similar results in Methods -- Translocaiton.
--->
+<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocaiton. -->
 
@@
 Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-
-<!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations?
-Because we don't have accurate counts of subadults, can we even provide rough estimate of recruitment probability?
-Mark W here: This would be great.
-If we have even ballpark estimates of \omega it would be useful to include.-->
+<!--# Add text to the end of this paragraph that refers to higher survival of cohorts translocated later in the 2006-2020 study period (Table 2), to make the case that more than 50% of future cohorts are expected to have high survival.
+-->
 
@@
 
+<!--# The following point was made in the original Introduction outline but ultimately not addressed there: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
+-->
+
 ### Figures and figure legends
@@
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-<!--# Include error bars on these median estimates.
--->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
+<!--# Unable to dodge points and associated error bars together using group.
+Any help with this?
+Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 

### 2023-02-24 doc/manuscript/translocation.qmd
Edits to formatting and text throughout

@@
 editor: visual
+bibliography: translocation.bib
 csl: pnas.csl
-bibliography: translocation.bib
 ---
@@
 Evolutionary changes following disease-induced declines are known from a diversity wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
-Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
-To date, this potentially important species recovery strategy is rarely utilized in wildlife conservation efforts [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
-<!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example. -->
+Important for conservation action, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
+To date, the use of resistant individuals for species recovery is rare [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
+<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->
 
@@
 Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments.
+Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of populations reintroduced into novel environments.
 And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease.
@@
 
-The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; [@scheele2019]).
-Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; [@grinnell1924]), this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
+The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
+Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
 Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
@@
 These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
-In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.
+In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
 
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability modeling, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
+Here, we report the results of (i) a greatly expanded MYL frog reintroduction effort, (ii) population viability modeling, and (iii) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
 Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
@@
 
-<!--# Add Allie's selection results here -->
+#### Frog evolution in response to Bd
 
-#### Reintroductions allow population reestablishment
+To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+<!--# Modify text/remove text here/elsewhere to avoid repetition of this.
+--> Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
+The comparison of frog genomes between naive and recovering populations is not as powerful as comparing samples collected from populations before and after Bd exposure, but because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable, no such samples are available and are unlikely to be available in the future.
+As such, the current study design is the best possible under the circumstances.
+To detect regions of the genome under selection, we used a multivariate linear mixed model to compare genomes of frogs from naive and recovering populations, and created "liberal" and "stringent" sets of outlier variants using a Bonferroni-corrected significance level of 0.05 and 0.01, respectively.
+
+The samples clustered into three separate groups in principal component space, each containing at least one population from both the naïve and persistent groups (@fig-selectionresults B: SI).
+The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
+One gene in this group does not have an associated annotation.
+No biological functions, molecular functions, or cellular processes were significantly overrepresented in the set of 29 annotated outlier genes.
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI).
+For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
+Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
+However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
+<!--# Allie should add a brief description of the function of RIN3 here. Also, emphasize key messages related to selection, resistance. And, of course, rework any of the text as necessary. -->
+
+<!--# Figure edits:  selectionresults: Correctly locate population 7, capitalize panel labels, label liberal and stringent outlier sets.
+outlier map: Rearrange panels to put RIN3 last?
+Relocate pie charts to avoid the perception that all recovering populations are west of naive populations, add A, B, C labels to panels.
+-->
+
+#### Frog translocations
 
@@
 Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
+These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
 We refer to these reintroductions as "translocations".
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI) <!--# How to restart numbering for figures in SI, add "S" prefix?
 -->.
-Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
+Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
 Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details).
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (\@tbl-viability-earlylate; Table 2) <!--# Why isn't cross-referencing working for this table?
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Should be Table 1, not Table 2.
+Rearrange table order in manuscript.
 -->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
-Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018).
-<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocaiton. -->
+Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
+<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocation. -->
 
-To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
+To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework.
 Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
 Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
 Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
-In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection.
-In addition, the small changes in Bd load following translocation and relatively high survival of frogs in a substantial number of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats (i.e., low-moderate Bd loads, minimal Bd-induced mortality).
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
+In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
+Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
+In addition, the low-moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results before "Collectively" above.
+Summarize result here, e.g., with the next sentence.
 -->.
-This result shows that frog resistance in non-natal habitats is maintained across generations of frogs.
+Therefore, frog resistance in non-natal habitats is maintained across generations.
 
@@
 
-#### Long-term persistence of reestablished populations
+#### Population viability modeling
 
@@
 
-#### Frog evolution in response to Bd
-
-@fig-selectionresults
-
-@fig-allelefrequencies
-
-#### Discussion
+### Discussion
 
@@
 
-### Figures and figure legends
+### Materials and Methods
 
-<!--# Discuss referencing and in-text generation of figures.
-We don't want to re-run the population viability modeling figures each time we compile the manuscript.
-A makefile could be a good solution.
-Caching figures could be another option (https://quarto.org/docs/computations/caching.html) -->
+#### Frog evolution in response to Bd
 
-![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
-Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Frog survival was estimated using capture-mark-recapture methods.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
-<!--# Unable to dodge points and associated error bars together using group.
-Any help with this?
-Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
+These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
 
-![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) donor population, and (D) sex.
-In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
-
-![A population viability analysis of translocated *R.
-sierrae* that quantifies the long-run growth rate and extinction probability of frog populations.
-A.
-The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
-sierrae*.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
-B.
-The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
-
-![Sensitivity analysis and empirical validation of the stage-structured *R.
-sierrae* model.
-A.
-The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$.
-B.
-The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
-
-<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.  We can discuss. -->
-
-![<!--# Add first sentence that provides a brief overview of the entire figure.
---> (a) Map of sampling locations showing Bd-naive populations in red and persistent populations in blue.
-(b) PCA calculated from binary SNPs showing genomic relationship of samples.
-Labels match sampling location in (a) and are colored as in (a).
-(c) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
-Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
-(d) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
-SNPs are sorted by genomic position and Contigs are sorted by size.
-Red line shows Bonferroni corrected p-value for threshold of 0.01.
-Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
-
-![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R.
-sierrae* populations.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
-
-### Tables
-
-<!--# Include table of translocation sites, years, frog numbers? -->
-
-<!--# Include Allie's table describing outlier SNPs? -->
-
-### References
-
-::: {#refs}
-:::
-
-### Materials and Methods
+DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-selectionresults).
+To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------|:------------------|:-------------------------|
+|:---------------------------|:-----------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
@@
 
-#### Frog evolution in response to Bd
-
-To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
-
-DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-selectionresults).
-To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
-
 ### Acknowledgements and funding sources
@@
 
-### Figure legends
+::: {#refs}
+:::
 
-### Supporting Information
+### Figures
+
+<!--# Discuss referencing and in-text generation of figures.
+We don't want to re-run the population viability modeling figures each time we compile the manuscript.
+A makefile could be a good solution.
+Caching figures could be another option (https://quarto.org/docs/computations/caching.html) -->
+
+![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R.
+sierrae* populations.
+The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+
+![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
+Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
+Frog survival was estimated using capture-mark-recapture methods.
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
+Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
+<!--# Unable to dodge points and associated error bars together using group.
+Any help with this?
+Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+
+![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) donor population, and (D) sex.
+In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
+
+![A population viability analysis of translocated *R.
+sierrae* that quantifies the long-run growth rate and extinction probability of frog populations.
+A.
+The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
+sierrae*.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
+B.
+The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
+
+![Sensitivity analysis and empirical validation of the stage-structured *R.
+sierrae* model.
+A.
+The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$.
+B.
+The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+
+<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.  We can discuss. -->
+
+## Supporting Information
+
+### References
+
+### Figures
 
 <!--# Add text and results describing Bd load before versus after translocation. -->
+
+![<!--# Add first sentence that provides a brief overview of the entire figure.
+--> (a) Map of sampling locations showing Bd-naive populations in red and persistent populations in blue.
+(b) PCA calculated from binary SNPs showing genomic relationship of samples.
+Labels match sampling location in (a) and are colored as in (a).
+(c) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
+(d) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
+SNPs are sorted by genomic position and Contigs are sorted by size.
+Red line shows Bonferroni corrected p-value for threshold of 0.01.
+Outlier SNPs above this threshold are highlighted and labeled.
+Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
 
@@
 
-![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-<!--# Remove gray background and gridlines -->](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
 
@@
 Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+
+### Tables
 
@@
 
-### Notes (author use only - not part of manuscript)
+<!--# Include table of translocation sites, years, frog numbers? -->
 
-Science (impact factor = 63.714)
+<!--# Include Allie's table describing outlier SNPs? -->
 
--   Research articles
-
-    -   Present a major advance.
-
-    -   Up to approximately 4500 words, including references, notes and captions.
-        Corresponds to approximately 5 printed pages in journal (approxiumately 11 manuscript pages).
-
-    -   Up to 6 figures or tables, and about 40 references.
-
--   Reports
-
-    -   New research results of broad significance.
-
-    -   Up to \~2500 words, including references, notes and captions.
-        Corresponds to approximately 3 printed pages in journal (approximately 6 manuscript pages).
-
-    -   Up to 4 figures or tables, and about 30 references.
-
-Nature (impact factor = 69.8)
-
--   Articles
-
-    -   Original reports whose conclusions represent a substantial advance in understanding of an important problem and have immediate, far-reaching implications.
-
-    -   Biological sciences papers do not normally exceed 8 pages.
-        This would be approximately 4300 words and 5-6 modest display items.
-
-    -   This [article](https://www.nature.com/articles/s41586-022-04824-9) provides a potentially useful (e.g., recent) example of formatting and structure for a paper focused on ecology/evolution
-
-PNAS (impact factor = 12.77)
-
--   Research reports
-
-    -   Original research of exceptional importance.
-
-    -   Preferred length is 6 pages, but up to 12 pages is allowed.
-        6 page article is approximately 4,000 words, 50 references, and 4 medium-size figures or tables.
-
-Nature Communications (impact factor = 17.694)
-
--   Article
-
-    -   Papers published by the journal represent important advances of significance to specialists within each field.
-        Open access.
-
-    -   The main text (not including abstract, Methods, References and figure legends) is limited to 5,000 words.
-        The maximum title length is 15 words.
-
-Science Advances (impact factor = 14.957)
-
--   Publishes impactful research papers and reviews in any area of science.
-    Journal extends the capacity of *Science* magazine to identify and promote significant advances in science and engineering.
-    Open access.
-
--   Research article
-
-    -   Should present a major advance.
-
-    -   Articles can be short of long format, up to 15,000 words, with up to 10 figures and/or tables.
-
-## Other introduction material for quick reference
-
-P5:
-
-we show using amphibian-pathogen system to demonstrate that resistance first demonstrate that first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
-We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
-Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
-These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
-
-Host evolution and population persistence/recovery
-
--   Species recovery in the face of severe disease (population growth rate \> 1 ) will depend on the ability of a species to move or change.
-
--   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al.
-2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
-    Restrict to species of conservation concern?)
-    .
-
--   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease.
-    For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue.
-    These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al. 2019, Joseph and Knapp 2018).
-
-Population reestablishment using reintroduction of resistant hosts
-
--   "Rescued" populations provide a source of animals from which to repatriate extirpated populations and disentangle the effects of the local environment and genetics on host recoveries.
-
--   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.
-
--   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred.
-    We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease.
-    Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction.
-    These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.
-
-Description of MYL frog species complex, impacts of Bd, and evidence of recent recovery.
-
-<!--# I'm not sure the following text still makes sense to include.
-If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper.
-The following is my best attempt at doing so.
-I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript.
--->
-
-We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence.
-These hypotheses generate the following expectations:
-
--   Differences in allele frequencies exist between Bd-naive and recovering populations
-
--   Shared selective response across recovering populations
-
--   Changes in population growth rate due to evolution extend beyond the natal environment
-
--   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics)
-
--   Population viability: Yearly population growth rate of translocated populations in similar environments to the natal environment should be greater than 1, indicating the capacity of the selected frogs to recover in the presence of the pathogen.
-
-Then show how each result is consistent with those expectations.
-
-<!--# Extra ER text for use in manuscript -->
-
--   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al.
-2014), can reverse ongoing declines in some species (Golas et al.
-2021; Searle et al.
-2021).
-
--   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018).
-    This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al.
-(2013).
-
-    -   Tasmanian devils: Epstein et al. (2016), Wells et al., Ecology (2019)
-
-    -   Frogs in Australia: Grogan et al. (2018), McKnight et al. (2020), Newell et al. (2013), Scheele et al., (2014)
-
-    -   Frogs in Panama: Voyles et al. (2018)
-
-    -   Frogs in the Sierra Nevada: Knapp et al. (2016) @knapp2016
-
-    -   Sea Star Wasting Disease: Menge et al., Plos One (2016), Schiebelhut et al.
-        PNAS (2018)
-
-    -   White-nose syndrome: Gignoux-Wolfsohn et al. (2021) Molecular ecology.
+###

### 2023-02-24 code/translocation_survival_exploratory.Rmd
Change errorbar to linerange

@@
   ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor)) +  # reorder site_id by siteid_order)
-      geom_errorbar(aes(ymin = lo_survival, ymax = hi_survival), color = "black", width = 0.15, position = position_dodge2(width = 0.3)) +
-      geom_point(size = 3) +
+      geom_linerange(aes(ymin = lo_survival, ymax = hi_survival), color = "black", width = 0.5, position = position_dodge2(width=0.5)) +
+      geom_point(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor), size = 3, position = position_dodge2(width=0.5)) + 
       scale_color_manual(values = palette1) + 
@@
 ```
+
 * Strong effect of site/translocation on frog survival.

### 2023-02-24 doc/manuscript/translocation.qmd
Add figure to compare survival probabilities

--- 2023-02-23 doc/manuscript/translocation.qmd
@@
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability modeling, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
+Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability model, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
 Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
@@
 Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018).
-<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocaiton. -->
+<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocation. -->
 
@@
 In addition, the small changes in Bd load following translocation and relatively high survival of frogs in a substantial number of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats (i.e., low-moderate Bd loads, minimal Bd-induced mortality).
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
--->.
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
 This result shows that frog resistance in non-natal habitats is maintained across generations of frogs.
@@
 
-To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+To test whether the observed yearly survival probabilities of resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
 We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values).
@@
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
-Therefore, even under conservative assumptions about recruitment probability $\omega$ (a parameter where ), these populations had some probability of persistence over 50 years following a single introduction.
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
 For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
@@
 Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-<!--# Add text to the end of this paragraph that refers to higher survival of cohorts translocated later in the 2006-2020 study period (Table 2), to make the case that more than 50% of future cohorts are expected to have high survival.
+Moreover, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
+Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
+<!--# Mark here: Just figured I would point this out.
+This is trending in the right direction, but there isn't strong statistical evidence that these two proportions differ.
+I tried to phrase this carefully.
 -->
@@
 Any help with this?
-Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+Code is in translocation_survival_exploratory.Rmd:  Mark: My suggested fix is to use geom_linerange instead of geom_errorbar.
+-->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 
@@
 
+![A comparison of the average yearly adult survival probabilities for translocated adults and adults that were naturally recruited to a population as a result of reproduction by translocated frogs.
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but averaged survival probabilities across multiple years.
+Points are median estimates and error bars give the 90% credible intervals around the estimates.
+All estimates are derived from the mrmr package [see @joseph2018 for details].](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
+
 ```{r}

### 2023-02-24 doc/manuscript/notes_unusedtext.qmd

--- 2023-02-23 doc/manuscript/translocation.qmd
@@
 ---
+title: "Notes and other text"
 format: docx
 editor: visual
-csl: pnas.csl
-bibliography: translocation.bib
 ---
 
-<!--# Intended outlet is PNAS.
-Preferred length of research reports is 6 pages, but up to 12 pages is allowed.
-A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables) -->
-
-## Title: Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
-
-<!--# Title should be less than 135 characters -->
-
-### Authors:
-
-Roland A. Knapp^1,2^\*, Mark Q. Wilber^3^, Allison Q. Byrne^4,5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
-
-<!--# Exact author order to be determined -->
-
-### Affiliations:
-
-^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
-
-^2^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
-
-^3^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
-
-^4^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
-
-^5^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
-
-^6^Earth Lab, University of Colorado, Boulder, CO, 80303
-
-^7^NCX, San Francisco, CA, 94115
-
-^8^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
-
-\*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu
-
-**Preprint Server:**
-
-**Classification:** Biological Sciences - Ecology
-
-### Abstract
-
-<!--# No more than 250 words -->
-
-### Significance Statement
-
-<!--# Less than 120 words. Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty.  -->
-
-### Introduction
-
-Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations.
-In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007].
-Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
-Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
-At a time when novel infectious diseases are emerging at an unprecedented rate [@fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.
-
-Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014].
-In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-Evolutionary changes following disease-induced declines are known from a diversity wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
-Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
-To date, this potentially important species recovery strategy is rarely utilized in wildlife conservation efforts [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
-<!--# Roland says: Modestly speaking, when combined with the selection results in the current paper, Joseph and Knapp (2018) does provide an example. -->
-
-There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
-First, populations that survived following disease arrival have evolved increased resistance.
-Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment, allowing for the establishment of populations reintroduced into novel environments.
-And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease.
-To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021].
-
-The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; [@scheele2019]).
-Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains (California, USA; [@grinnell1924]), this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
-Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
-Endangered Species Act.
-Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B.
-dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
-These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
-In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish persistent populations.
-
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
-Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability modeling, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
-Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
-These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
-
-### Results
-
-<!--# Add Allie's selection results here -->
-
-#### Reintroductions allow population reestablishment
-
-```{r}
-#| include: false
-
-library(readr)
-library(dplyr)
-library(magrittr)
-library(stringr)
-library(lubridate)
-d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
-
-transno_min <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-transno_max <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_max <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_min <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-bdload_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(min(bdload_ave))
-
-bdload_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(max(bdload_ave))
-
-bdprev_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_max(prev, n = 1, with_ties = FALSE)
-
-bdprev_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_min(prev, n = 1, with_ties = FALSE)
-
-files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
-surveys <- read_csv(files, id = "path") %>% 
-  filter(survey_date < "2023-01-01") %>% 
-  select(-path)
-rm(files)
-
-surveys_yr_mean <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(mean(n))
-
-surveys_yr_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(max(n))
-
-surveys_yr_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(min(n))
-
-robust_year_mean <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(mean(max_2period))
-
-robust_year_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(min(max_2period))
-
-robust_year_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(max(max_2period))
-```
-
-Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
-We refer to these reintroductions as "translocations".
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix?
--->.
-Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
-Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details).
-This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (\@tbl-viability-earlylate; Table 2) <!--# Why isn't cross-referencing working for this table?
--->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
-sierrae*.
-Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018).
-<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocaiton. -->
-
-To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
-Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
-Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
-In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
-In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection.
-In addition, the small changes in Bd load following translocation and relatively high survival of frogs in a substantial number of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats (i.e., low-moderate Bd loads, minimal Bd-induced mortality).
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results.
--->.
-This result shows that frog resistance in non-natal habitats is maintained across generations of frogs.
-
-<!--# Mark W: Regarding the above paragraph, what evidence do we have of "tolerance"?
-Are Bd loads significantly reduced in populations compared to epizootic levels?
-Can we convincingly state that the likely mechanism is more consistent with resistance or tolerance?
-I ask because it would be nice to just use one of these terms in the paper and not get caught up in any resistance vs.
-tolerance discussions with reviewers.
-Roland: Have a look at this paragraph of Knapp et al.
-PeerJ 2022: https://peerj.com/articles/12712/#p-61.
-We argue that there is evidence for resistance and tolerance.
--->
-
-<!--# Drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd.
--->
-
-#### Long-term persistence of reestablished populations
-
-To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values).
-Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
-When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
-In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
-A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
-
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
-Therefore, even under conservative assumptions about recruitment probability $\omega$ (a parameter where ), these populations had some probability of persistence over 50 years following a single introduction.
-For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
-In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
-Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-<!--# Add text to the end of this paragraph that refers to higher survival of cohorts translocated later in the 2006-2020 study period (Table 2), to make the case that more than 50% of future cohorts are expected to have high survival.
--->
-
-#### Frog evolution in response to Bd
-
-@fig-selectionresults
-
-@fig-allelefrequencies
-
-#### Discussion
-
-We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
-<!--# Consider broadening the scope of this sentence and the paragraph to be developed from it to encompass all of our findings -->
-
-1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
-2.  Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
-3.  There is a shared selective response across recovering populations.
-4.  Changes in population growth rate due to evolution extend beyond the natal environment.
-5.
-In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-
-<!--# Discuss effect of donor population on survival of translocated frogs. Differences in Bd resistance? -->
-
-<!--# The following point was made in the original Introduction outline but ultimately not addressed there: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
--->
-
-### Figures and figure legends
-
-<!--# Discuss referencing and in-text generation of figures.
-We don't want to re-run the population viability modeling figures each time we compile the manuscript.
-A makefile could be a good solution.
-Caching figures could be another option (https://quarto.org/docs/computations/caching.html) -->
-
-![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
-Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Frog survival was estimated using capture-mark-recapture methods.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
-<!--# Unable to dodge points and associated error bars together using group.
-Any help with this?
-Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
-
-![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) donor population, and (D) sex.
-In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
-
-![A population viability analysis of translocated *R.
-sierrae* that quantifies the long-run growth rate and extinction probability of frog populations.
-A.
-The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
-sierrae*.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
-B.
-The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
-
-![Sensitivity analysis and empirical validation of the stage-structured *R.
-sierrae* model.
-A.
-The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$.
-B.
-The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
-
-<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.  We can discuss. -->
-
-![<!--# Add first sentence that provides a brief overview of the entire figure.
---> (a) Map of sampling locations showing Bd-naive populations in red and persistent populations in blue.
-(b) PCA calculated from binary SNPs showing genomic relationship of samples.
-Labels match sampling location in (a) and are colored as in (a).
-(c) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
-Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
-(d) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
-SNPs are sorted by genomic position and Contigs are sorted by size.
-Red line shows Bonferroni corrected p-value for threshold of 0.01.
-Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
-
-![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R.
-sierrae* populations.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
-
-### Tables
-
-<!--# Include table of translocation sites, years, frog numbers? -->
-
-<!--# Include Allie's table describing outlier SNPs? -->
-
-### References
-
-::: {#refs}
-:::
-
-### Materials and Methods
-
-#### Frog translocations
-
-##### Field methods
-
-Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016].
-Results indicated that these populations contained several hundred *R.
-sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd infection intensity ("load", expressed as median by site: `r round(bdload_min, 2)`--`r round(bdload_max, 2)`) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 ITS copies per swab)[@vredenburg2010; @joseph2018].
-Recipient sites to which frogs were translocated were chosen based on previous *R.
-sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
-
-During each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
-Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (mean number of visits per summer = `r round(surveys_yr_mean)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
-CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
-During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
-Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (mean number of secondary periods per primary period = `r round(robust_year_mean)`; range = `r robust_year_min`--`r robust_year_max`)<!--# calculate min, max, median-->), and frogs were processed as described above.
-However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
-Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and during subsequent captures were only identified and released.
-Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
-For details of survey and swab analysis methods, see @joseph2018.
-
-During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
-These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
-
-##### Estimation of frog survival and abundance
-
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; @rsoftware2022].
-The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
-The possible states of individuals include "not recruited", "alive", and "dead".
-The possible observations of individuals include "detected" and "not detected".
-We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
-For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
-We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to construct CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
-
-##### Predictors of post-translocation frog survival
-
-To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
-We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
-We used default, weakly informative priors.
-Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
-Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
-Both approaches utilized the loo package.
-
-Of the two analyses we conducted, one focused on identifying predictors of post-translocation frog survival from a wide range of variables describing characteristics of frogs, habitat, and translocations [@fig-survival-postdens].
-The other analysis focused specifically on whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
-In both analyses, the response variable, probability of frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
-For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked convergence of the MCMC algorithm using trace plots and Rhat.
-Code to create datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-
-#### Population viability modeling
-
-##### Model description
-
-To determine the implications of observed yearly adult survival on the long-term population viability of putatively resistant frogs in non-natal habitats, we developed a population model for *R.
-sierrae*.
-Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of an *R.
-sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.
-
-We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
-The dynamics are given by
-
-$$
-\begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\ 
-A_T
-\end{bmatrix}(t + 1) = 
-\begin{bmatrix}
-  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
-  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
-  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
-  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
-  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
-\end{bmatrix} \begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\
-A_T \\
-\end{bmatrix}(t) 
-$$
-
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$, where the "$\cdot$" indicates a particular state variable, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
-The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
-
-In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
-We also did not directly consider the dynamics of Bd in this model.
-We made this decision because i) translocated populations are infected with Bd at high prevalence [@joseph2018], and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
-Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-
-##### Model analysis
-
-###### Incorporating yearly variability in vital rates
-
-We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
-Although we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability.
-Thus, we did not include yearly within-population variability in adult survival in this analysis.
-
-However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability about a mean value.
-Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
-
-###### Estimating model parameters
-
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
-For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$.
-For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-| **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------|:------------------|:-------------------------|
-| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
-| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
-| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
-| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
-| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
-| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
-
-: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
-
-###### Model analysis and simulation
-
-We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
-We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
-
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
-
-Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
-We used this model to simulate a one-time introduction of 40 translocated adult frogs.
-We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
-We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
-
-Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
-We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
-We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling (@kosmala2016), we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
-Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
-
-#### Frog evolution in response to Bd
-
-To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
-
-DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-selectionresults).
-To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
-
-### Acknowledgements and funding sources
-
-### References
-
-### Figure legends
-
-### Supporting Information
-
-<!--# Add text and results describing Bd load before versus after translocation. -->
-
-![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
-Symbol shapes indicate the donor population used for each translocation site.
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="450"}
-
-![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
-
-![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-<!--# Remove gray background and gridlines -->](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
-
-![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
-
-```{r}
-#| label: tbl-viability-earlylate
-#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013), respectively) for which the probability of extinction over 50 years was 1 or < 1 (viability = 0 or 1, respectively)."
-#| echo: FALSE
-#| message: FALSE
-
-viability_table <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  distinct(site_id, year) %>% 
-  group_by(site_id) %>% 
-  filter(year == min(year)) %>% 
-  mutate(site_id = as.character(site_id),
-         viability = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), 0, 1),
-         period = if_else(year < 2013, "early", "late")) %>% 
-  group_by(period) %>% 
-  count(period, viability) %>% 
-  mutate(n_period = sum(n), 
-         prop = (n / n_period)) %>% 
-  mutate(across(prop, round, 3)) %>% 
-  arrange(viability, period)
-
-  knitr::kable(viability_table, "pipe", align = "lcccr", caption = )
-
-```
-
-### Notes (author use only - not part of manuscript)
+### Notes and other text (for possible use in manuscript)
 
@@
 
-## Other introduction material for quick reference
+### Other introduction material for quick reference
 

### 2023-02-24 doc/manuscript/translocation.qmd

--- 2023-02-24 doc/manuscript/translocation.qmd
@@
 In addition, the low-moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results before "Collectively" above.
-Summarize result here, e.g., with the next sentence.
--->.
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
 Therefore, frog resistance in non-natal habitats is maintained across generations.
@@
 
+![A comparison of the average yearly adult survival probabilities for translocated adults and adults that were naturally recruited to a population as a result of reproduction by translocated frogs.
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but averaged survival probabilities across multiple years.
+Points are median estimates and error bars give the 90% credible intervals around the estimates.
+All estimates are derived from the mrmr package [see @joseph2018 for details].](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
+
 ### Tables

### 2023-03-01 doc/manuscript/translocation.qmd
Edits throughout manuscript

@@
 #| include: false
+#| code-fold: true
 
@@
 These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
-We refer to these reintroductions as "translocations".
+We refer to these reintroductions as "translocations" [@seddon2014].
 During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI) <!--# How to restart numbering for figures in SI, add "S" prefix?
@@
 sierrae*.
-Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
-<!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocation. -->
+Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
+<!--# Add median Bd load, proportion below 5.8 copies threshold?
+--> <!--# The Methods -- Translocation section contains text that repeats similar results.
+Edit to reduce repetition.
+--> <!--# To allow comparison with other translocation studies, consider adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
+-->
 
-To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework.
+To identify site and frog-specific drivers of post-translocation survival, we developed a multilevel model in a Bayesian framework.
 Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
@@
 In addition, the low-moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
+Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
 Therefore, frog resistance in non-natal habitats is maintained across generations.
@@
 To test whether the observed yearly survival probabilities of resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values).
+We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values SI).
 Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
@@
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
 Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
@@
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
-Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
+Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
 Moreover, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
@@
 
+<!--# Consider broadening the scope of this sentence and the paragraph to be developed from it to encompass all of our findings -->
+
+TO BE WRITTEN.
+
 We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
-<!--# Consider broadening the scope of this sentence and the paragraph to be developed from it to encompass all of our findings -->
 
@@
 
+<!--# Translocation "success" often defined as evidence of reproduction by translocated individuals (see IUCN and related references in notes).
+We used a much higher and more conservative standard to assess success, i.e.,  -->
+
 ### Materials and Methods
@@
 To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering"; @knapp2016, @vredenburg2010].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities (load) in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering": @knapp2016; @vredenburg2010].
+These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection loads in adults that are typically well below the level expected to cause frog mortality [@knapp2011; @vredenburg2010].
 Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
 
-DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from seven Bd-naive populations and seven recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to south of Sequoia National Park (@fig-selectionresults).
-To minimize potential biases associated with different latitudinal ranges of the two population types, we selected populations to represent similar latitudinal ranges.
+DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from four Bd-naive populations and five recovering populations.
+These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-selectionresults A).
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
 
@@
 Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016].
+We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], as well as landscape genetic information [@poorten2017].
 Results indicated that these populations contained several hundred *R.
@@
 Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
-Both approaches utilized the loo package.
+Both approaches utilized the loo package [@vehtari2022].
+Group-level effects included translocation_id,
 
-Of the two analyses we conducted, one focused on identifying predictors of post-translocation frog survival from a wide range of variables describing characteristics of frogs, habitat, and translocations [@fig-survival-postdens].
-The other analysis focused specifically on whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
-In both analyses, the response variable, probability of frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
+We conducted two analyses.
+In the first, we addressed the question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations [@fig-survival-postdens].
+In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
 For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
 We checked convergence of the MCMC algorithm using trace plots and Rhat.
-Code to create datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+
+##### Changes in Bd load following translocations
+
+To assess the magnitude of changes in Bd load on frogs following translocations, we compared Bd loads measured before versus after translocation.
+Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
+After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
+Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
+
+To quantify changes in Bd load, we used a workflow similar to that described above (in "Predictors of post-translocation frog survival").
+Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
+We compared models with and without group-level effects, and group-level effects included (i) pit tag id, (ii) pit tag id nested within site id, and (iii) pit tag id nested within translocation id.
+All models used a negative binomial family.
+Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
 
@@
 
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
+The baseline parameter values for the model and how they were estimated are given in @tbl-param_values : SI.
 Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
@@
 For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-| **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:---------------------------|:-----------------|:-------------------------|
-| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
-| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
-| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
-| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
-| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
-| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
-| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
-
-: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
 
@@
 Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
-<!--# Unable to dodge points and associated error bars together using group.
-Any help with this?
-Code is in translocation_survival_exploratory.Rmd -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+When multiple points exist for a site, points/error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 
@@
 
-<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.  We can discuss. -->
+<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.
+We can discuss.
+Roland says: Maybe try a palette that you like better and for those colors that will be hard for eyeballs like mine to distinguish, we can either tweak the colors or add diagnonal lines (ggpattern::geom_bar_pattern).
+-->
 
@@
 ### Figures
-
-<!--# Add text and results describing Bd load before versus after translocation. -->
 
@@
 
-![A comparison of the average yearly adult survival probabilities for translocated adults and adults that were naturally recruited to a population as a result of reproduction by translocated frogs.
+![A comparison of the average yearly adult survival probabilities for translocated adults versus adults that were naturally recruited to a population as a result of reproduction by translocated frogs.
 In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but averaged survival probabilities across multiple years.
-Points are median estimates and error bars give the 90% credible intervals around the estimates.
-All estimates are derived from the mrmr package [see @joseph2018 for details].](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
+Points are median estimates and error bars give the 90% uncertainty intervals around the estimates.
+All estimates are derived from the mrmr package [see @joseph2018 for details].
+<!--# Should we use 95% intervals to be consistent with other analyses?
+-->](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
 
@@
   knitr::kable(viability_table, "pipe", align = "lcccr", caption = ) 
+```
 
-```
+| **Parameter**                                                             | **Value**            | **Source**                                                         |
+|:----------------------------|:-----------------|:-------------------------|
+| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
+| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
+| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
+| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
+| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
+| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
+| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
+| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
+
+: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
 

### 2023-03-01 doc/manuscript/translocation.qmd
Minor edits to translocation methods

@@
 
-surveys_yr_mean <- surveys %>% 
+surveys_yr_median <- surveys %>% 
   mutate(yr = year(ymd(survey_date))) %>%
@@
   count(yr) %>% ungroup() %>% 
-  summarize(mean(n))
+  summarize(median(n))
 
@@
 
-robust_year_mean <- surveys %>% 
+robust_year_median <- surveys %>% 
   mutate(yr = year(ymd(survey_date))) %>%
@@
   summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(mean(max_2period))
+  summarize(median(max_2period))
 
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (mean number of visits per summer = `r round(surveys_yr_mean)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
+Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
 CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
@@
 Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (mean number of secondary periods per primary period = `r round(robust_year_mean)`; range = `r robust_year_min`--`r robust_year_max`)<!--# calculate min, max, median-->), and frogs were processed as described above.
+During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
 However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; @rsoftware2022].
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
-We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
+We performed all analyses using the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022] .
 We used default, weakly informative priors.
@@
 Both approaches utilized the loo package [@vehtari2022].
-Group-level effects included translocation_id,
 
 We conducted two analyses.
-In the first, we addressed the question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations [@fig-survival-postdens].
+In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
+Group-level effects included site_id or translocation_id.
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations [@fig-survival-postdens] SI.
+Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
 In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
@@
 
-##### Changes in Bd load following translocations
+##### Changes in Bd load following translocation
 
-To assess the magnitude of changes in Bd load on frogs following translocations, we compared Bd loads measured before versus after translocation.
+To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
 Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
@@
 Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
-We compared models with and without group-level effects, and group-level effects included (i) pit tag id, (ii) pit tag id nested within site id, and (iii) pit tag id nested within translocation id.
+We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
 All models used a negative binomial family.
+Priors, chains, iterations, and checks were as described above.
 Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:-----------------|:-------------------------|
+|:---------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-03-03 doc/manuscript/translocation.qmd
Substantive edits throughout manuscript

@@
 ---
+bibliography: translocation.bib
+csl: pnas.csl
 format: docx
 editor: visual
-bibliography: translocation.bib
-csl: pnas.csl
 ---
@@
 Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014].
+"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014].
 In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-Evolutionary changes following disease-induced declines are known from a diversity wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare.
-Important for conservation action, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
-To date, the use of resistant individuals for species recovery is rare [see @joseph2018 for a notable example], and studies that evaluate its potential utility are critically needed.
+Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
+In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
+To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
 <!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->
@@
 Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of populations reintroduced into novel environments.
-And fourth, reintroduced populations show long-term growth and persistence in a landscape of disease.
-To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021].
+Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
+And fourth, reintroduced populations show long-term growth and persistence.
+To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].
 
@@
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of (i) a greatly expanded MYL frog reintroduction effort, (ii) population viability modeling, and (iii) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd.
+Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
 Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
@@
 <!--# Modify text/remove text here/elsewhere to avoid repetition of this.
---> Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
-The comparison of frog genomes between naive and recovering populations is not as powerful as comparing samples collected from populations before and after Bd exposure, but because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable, no such samples are available and are unlikely to be available in the future.
+--> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
+Our comparison of frog genomes between naive and recovering populations is not as powerful an approach as comparing samples collected from populations before and after Bd exposure, but because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable, no such samples are available and are unlikely to be available in the future.
 As such, the current study design is the best possible under the circumstances.
-To detect regions of the genome under selection, we used a multivariate linear mixed model to compare genomes of frogs from naive and recovering populations, and created "liberal" and "stringent" sets of outlier variants using a Bonferroni-corrected significance level of 0.05 and 0.01, respectively.
+To detect regions of the genome under selection, we compared genomes of frogs from naive and recovering populations using a multivariate linear mixed model, and created "liberal" and "stringent" sets of outlier variants using a Bonferroni-corrected significance level of 0.05 and 0.01, respectively.
 
-The samples clustered into three separate groups in principal component space, each containing at least one population from both the naïve and persistent groups (@fig-selectionresults B: SI).
+The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and persistent groups (@fig-selectionresults B: SI).
 The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
@@
   summarize(min(n))
+
+transfrogs_median <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(median(n))
 
@@
 We refer to these reintroductions as "translocations" [@seddon2014].
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI) <!--# How to restart numbering for figures in SI, add "S" prefix?
--->.
-Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
+During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI).
+Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`) collected from 1 of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
 Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
@@
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Should be Table 1, not Table 2.
-Rearrange table order in manuscript.
--->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
@@
 <!--# Add median Bd load, proportion below 5.8 copies threshold?
---> <!--# The Methods -- Translocation section contains text that repeats similar results.
+Mention results from statistical model more explicitly?
+--> <!--# The Methods-Translocation section contains text that repeats similar results.
 Edit to reduce repetition.
@@
 Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
+Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
 In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
 Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
-In addition, the low-moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+In addition, the moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
+Finally, in three locations where longer capture-mark-recapture (CMR) time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
 Therefore, frog resistance in non-natal habitats is maintained across generations.
@@
 
-To identify putative genes under selection that are linked to reduced frog susceptibility to Bd-caused lethal and sub-lethal effects, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs from populations that have never experienced a Bd-caused epizootic ("Bd-naive") versus frogs from populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ["recovering": @knapp2016; @vredenburg2010].
-These latter populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection loads in adults that are typically well below the level expected to cause frog mortality [@knapp2011; @vredenburg2010].
-Bd-exposure histories of these populations are based on a decade or more of data from VES and Bd surveillance using skin swabbing [e.g., @knapp2016].
+##### Sampling and Sequencing
 
-DNA samples were collected via buccal swabbing [@broquet2007] of xx-xx frogs from four Bd-naive populations and five recovering populations.
-These populations represented a wide latitudinal range in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-selectionresults A).
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four Bd-naive populations, and 29 from five persistent/recovering populations.
+These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-selectionresults A).
+Sampling was even across the sites, with 5-6 samples per population.
 To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
+DNA was extracted following Qiagen DNEasy manufacturer's protocols.
+We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
+Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
+Capture baits were designed based on the coding regions of the *Rana muscosa* transcriptome (GenBank accession GKCT00000000).
+Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
+
+##### Data Pre-processing and Cleaning
+
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]) with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Exact PCR Duplicates were marked using Picard.
+Variants were then called following GATK best practices (v.XX, GATK ref).
+Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
+Next, genotypes were jointly called using GenotypeGVCFs.
+Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
+This initial filter resulted in 1,595,206 variant sites across 53 individuals.
+
+We then further filtered our dataset at the individual and variant level.
+First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
+This filter resulted in 427,038 sites, of which 353,172 were single nucleotide polymorphisms (SNPs) and 73,866 were insertions or deletions (INDELS).
+Finally, we trimmed samples with an average depth across filtered sites \< 7x.
+Our final dataset included 50 samples (23 from naive and 27 from persistent/recovering populations) with an average depth of 16.7x (range = 7.4x -- 26.1x).
+
+##### Data Analysis
+
+To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet package [@jombart2008].
+Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014].
+GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
+For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
+GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
+We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
+Results were visualized using a Manhattan plot and qqplot.
+
+Using the liberal set of outliers, we performed a statistical overrepresentation test using PANTHER.
+First, we retrieved the UnitProtKB ID for each gene in the *Xenopus laevis* genome, Next, we compared the 29 unique outlier genes from the stringent set to the complete set of genes in the *X.
+laevis* genome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+
+Finally, for the stringent set of outliers we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+First, we extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
+Next, we reannotated each sequence using blast to get the predicted gene location based on the closest annotated reference [@altschul1997].
+We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
+After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
 
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal capture-mark-recapture (CMR) surveys and VES.
+Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
 CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
@@
 However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
-Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed only during the first capture, and during subsequent captures were only identified and released.
+Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
 Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
@@
 Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations [@fig-survival-postdens] SI.
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations ([@fig-survival-postdens] : SI).
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
@@
 Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect. That is the model that is most analogous to a paired t-test. -->
 
@@
 
-To determine the implications of observed yearly adult survival on the long-term population viability of putatively resistant frogs in non-natal habitats, we developed a population model for *R.
-sierrae*.
+To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for *R. sierrae*.
 Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
@@
 sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.
+We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] : SI).
 
@@
 We also did not directly consider the dynamics of Bd in this model.
-We made this decision because i) translocated populations are infected with Bd at high prevalence [@joseph2018], and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
+We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
@@
 We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
-Although we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability.
+Although we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability ([@fig-translocation-survival]).
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
-
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability about a mean value.
@@
 Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
-For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$.
+For populations where we had a sufficient number of pit-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
 For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
@@
 We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling (@kosmala2016), we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$) <!--# Why 2 distributions?
+When was one used versus the other?
+Or are these 2 distributions equivalent?
+-->.
+Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
 The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
@@
 
-### Acknowledgements and funding sources
+### Acknowledgments and funding sources
 
@@
 
-![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R.
+<!--# Standardize panel labels. I suggest capital letters with no punctuation, e.g., "A", not "A.". -->
+
+![Allele frequencies of all outlier SNPs in the stringent set for each of the naive and recovering *R.
 sierrae* populations.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.
+<!--# Allie: Rearrange panels and add panel identifiers (A, B, C) as shown here.
+-->](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
 
-![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study.
+![Median 1-year survival of each cohort of translocated frogs at the 12 recipient sites included in this study.
 Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
@@
 Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
-When multiple points exist for a site, points/error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
 
-![Conditional effects of the important predictors of 1-year frog survival: (A) elevation of recipient site, (B) winter severity in the year following translocation, (C) donor population, and (D) sex.
+![Conditional effects of the important predictors of 1-year frog survival.
+(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
@@
 ![A population viability analysis of translocated *R.
-sierrae* that quantifies the long-run growth rate and extinction probability of frog populations.
-A.
-The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
+sierrae* that quantifies the long-run growth rate and extinction probability of frog populations: (A) The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
 sierrae*.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
-B.
-The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
+(B) The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.3$, for comparison to A.
+<!--# Mention that there are six lines at p(extinction) = 1?
+-->](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
 
@@
 sierrae* model.
-A.
-The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$.
-B.
-The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+(A) The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
+(B) The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
 
@@
 
-![<!--# Add first sentence that provides a brief overview of the entire figure.
---> (a) Map of sampling locations showing Bd-naive populations in red and persistent populations in blue.
-(b) PCA calculated from binary SNPs showing genomic relationship of samples.
-Labels match sampling location in (a) and are colored as in (a).
-(c) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+![Results from the exome capture selection study.
+(A) Map of sampling locations showing Bd-naive populations in red and persistent/recovering populations in blue.
+(B) PCA calculated from binary SNPs showing genomic relationship of samples.
+Numeric labels match those in (A) and are colored as in (A).
+(C) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
 Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
-(d) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
-SNPs are sorted by genomic position and Contigs are sorted by size.
+(D) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
+SNPs are sorted by genomic position and contigs are sorted by size.
 Red line shows Bonferroni corrected p-value for threshold of 0.01.
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
+Bold labels indicate at least one non-synonymous SNP is present in that gene.
+<!--# Is panel (C) important to include here?
+Or should it be moved to the SI?
+-->](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
 
@@
 
-![A comparison of the average yearly adult survival probabilities for translocated adults versus adults that were naturally recruited to a population as a result of reproduction by translocated frogs.
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but averaged survival probabilities across multiple years.
+![A comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years.
 Points are median estimates and error bars give the 90% uncertainty intervals around the estimates.
-All estimates are derived from the mrmr package [see @joseph2018 for details].
-<!--# Should we use 95% intervals to be consistent with other analyses?
+All estimates are derived from the mrmr package [@joseph2018].
+<!--# Use 95% intervals to be consistent with other analyses?
 -->](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
@@
 #| label: tbl-viability-earlylate
-#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013), respectively) for which the probability of extinction over 50 years was 1 or < 1 (viability = 0 or 1, respectively)."
+#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of extinction over 50 years was 1 or < 1 (i.e., viability = 0 or 1, respectively)."
 #| echo: FALSE
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:---------------------------|:------------------|:-------------------------|
+|:--------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-03-04 code/extinction_and_growth_rate_analysis.R
Build Makefile and incorporate viability analysis into repo

@@ -1,0 +1,517 @@
+library(data.table)
+source("population_viability_functions.R")
+library(ggplot2)
+library(patchwork)
+
+#### Data loading and preliminary analysis ######
+
+# Load Stan fits and lake ids.  The stan fits contain a model that has analyzed
+# The MRMR results of yearly adult survival probability.
+fit = readRDS(file.path("..", "out", "adult_survival_probability_model.rds"))
+keep_lakes = readRDS(file.path("..", "out", "lakes_to_use_in_analysis.rds"))
+
+colors = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f',
+					 '#ff7f00','#cab2d6','#6a3d9a','#ffff99','#b15928')
+
+# Color blind-friendly palette
+# colors = c("#000000","#004949","#009292","#ff6db6","#ffb6db",
+#  "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
+#  "#920000","#924900","#db6d00","#24ff24","#ffff6d")
+
+# Extract lake-specific probabilities of translocated survival
+beta0 = rstan::extract(fit, "beta0")$beta0
+alpha = rstan::extract(fit, "alpha")$alpha
+mus = apply(alpha, 2, function(x) 1 / (1 + exp(-(beta0 + x))))
+colnames(mus) = keep_lakes
+mu_dist = melt(data.table(mus), variable.name = "lake_id", value.name="surv_est")
+mean_mu_dist = mu_dist[, .(mean_est=mean(surv_est)), by=.(lake_id)]
+
+phi = exp(rstan::extract(fit, "logphi")$logphi)
+phi = 10000 # Set to be large to ignore inter-annual variability in survival
+
+# Load in survival estimates from all individuals
+all_surv = fread(file.path("..",
+						   "out", 
+						   "yearly_survival_estimates_all_individuals.csv")
+				)[order(lake_id)]
+
+# Load survival probabilities estimated for recruited individuals
+recruited_surv = fread(file.path("..",
+						   "out", 
+						   "yearly_survival_estimates_only_recruited_individuals.csv")
+				)[order(lake_id)]
+
+#############################################################
+###########  Analysis 1: Lambda calculation #################
+#############################################################
+
+# Calculate deterministic growth rates for site
+
+lake_growth_rates = list()
[…]

### 2023-03-04 code/extract_abundance_trajectories.R

@@ -1,0 +1,38 @@
+## Extract abundance and recruitment trajectories for different lakes
+
+library(cmdstanr)
+library(data.table)
+library(magrittr)
+source("population_viability_functions.R")
+
+# Lakes to include
+lake_ids = c(70414, 70370, 70134, 70505, 70550, 70641, 70619, 70449, 70413,
+			 70628, 70556, 74976)
+thin_values = rep(10, length(lake_ids))
+
+for(l in 1:length(lake_ids)){
+
+	lake_number = lake_ids[l]
+	cat("Working on lake", lake_number, "\n")
+	cat("INFO: Assuming cmr-analysis folder is stored in the same directory as translocation folder", "\n")
+	modelpath = file.path("..", "..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path("..", "..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+
+	model = readRDS(modelpath)
+	abundance_table = abundances_from_model(model, "abundance")
+	recruitment_table = abundances_from_model(model, "recruitment")
+
+	# Save tables to disk
+	dpath = file.path("..", "data", "clean", "abundance_and_recruitment")
+	if(!dir.exists(dpath)){
+		dir.create(dpath)
+	}
+	fwrite(abundance_table, file.path("..", "data", "clean", "abundance_and_recruitment", paste0(lake_number, "_abundance.csv")))
+	fwrite(recruitment_table, file.path("..", "data", "clean", "abundance_and_recruitment", paste0(lake_number, "_recruitment.csv")))
+}
+
+
+
+
+
+

### 2023-03-04 code/extract_survival_estimates.R

@@ -1,0 +1,136 @@
+## Script to extract all individuals using the mrmr output
+## Here we extract the time varying survival estimates,
+## calculate the time period over which these phi estimates were
+## estimated, and then calculate a weighted mean of the survival rate.
+## This survival estimate includes the survival probability for all individuals
+## including translocated individuals and survival estiamtes for non-translocated 
+## individuals
+
+library(cmdstanr)
+library(data.table)
+library(magrittr)
+source("survival_table_nontranslocated.R")
+
+# Lakes to include
+lake_ids = c(70414, 70370, 70134, 70505, 70550, 70641, 70619, 70449, 70413,
+			 70628, 70556, 74976)
+
+thin_values = rep(10, length(lake_ids))
+
+all_surv_ests = list()
+only_recruited_surv_ests = list()
+
+count = 1
+for(l in 1:length(lake_ids)){
+
+	lake_number = lake_ids[l]
+	cat("Working on lake", lake_number, "\n")
+	basepath = file.path("", "Users", "mqwilber", "Repos")
+	modelpath = file.path(basepath, "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path(basepath, "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+
+	model = readRDS(modelpath)
+	survey_dat = fread(datapath)
+
+	### Extract survival of only non-translocated individuals ###
+	res = survival_table(model, thin=thin_values[l])
+
+	# Only execute if there are non-translocated pittag animals
+	if(res$num_pittags > 0) { 
+
+		surv_probs = res$surv_probs
+
+		# Find the first stretch of primary periods where survival is 1 and remove them
+		# Doing this because individuals were not actually observed during this period
+		# and it is always conditional on recruitment.  Thus, survival has to be 1 
+		# as we know they were alive later.
+
+		nm_append = c("lower", "med", "upper")
+		rle_res = rle(surv_probs[["med"]] == 1)
+		cut = ifelse(rle_res$values[1] == TRUE, rle_res$lengths[1], 0)
+		surv_probs = surv_probs[(cut + 1):nrow(surv_probs)]
+		surv_list = list()
+		for(nm in nm_append){
+
[…]

### 2023-03-04 code/extract_translocation_data.R

@@ -1,0 +1,211 @@
+library(data.table)
+library(ggplot2)
+library(rstan)
+library(rethinking)
+library(mrmr)
+library(lubridate)
+
+## Extract and summarize relevant survival analysis data for MYL Frogs
+## Fit a Stan survival model to the analyses
+## Extract empirical patterns of recruitment
+
+# The base path of the CMR analysis.  Assumes the working directory is code
+# and that CMR analyses are stored two directories up.
+home_path = file.path("..", "..")
+
+#####################################
+#### Extract survival estimates #####
+#####################################
+
+# Extract survival estimates for all translocated lakes
+table_path = file.path(home_path, "cmr-analysis", "out", "tables")
+table_files = Sys.glob(file.path(table_path, "*survival_cohort.csv"))
+
+# Loop through each file and load table
+yearly_surv_probs = list()
+count = 1
+for(file in table_files){
+
+	lake_id = strsplit(basename(file), "_")[[1]][1]
+	surv_table = fread(file)
+	surv_table[, release_date:=as.character(release_date)]
+
+	release_dates = unique(surv_table$release_date)
+	for(rl in release_dates){
+
+		# Focus on a specific release date
+		ttab = surv_table[release_date == rl & years_since_introduction > 0][order(years_since_introduction)]
+
+		if(nrow(ttab) > 0){
+			med_surv = ttab$median_survival
+			med_surv = c(1, med_surv)
+			yearly_probs = med_surv[2:length(med_surv)] / med_surv[1:(length(med_surv) - 1)]
+			new_dat = data.frame(surv_prob=yearly_probs, 
+								 lake_id=lake_id, 
+								 release_date=rl, 
+								 total_surv=med_surv[1:(length(med_surv) - 1)])
+			yearly_surv_probs[[count]] = new_dat
+			count = count + 1
+		}
+	}
+}
+
+
+yearly_surv_probs_dt = data.table(do.call(rbind, yearly_surv_probs))[!is.na(surv_prob)]
+
+# Look at yearly survival probabilities only where TOTAL survival probabilities where
+# > 0.4.  Did this to remove high variability estimates as cohort population size dwindled
+ggplot(yearly_surv_probs_dt[total_surv > .4]) + geom_boxplot(aes(x=lake_id, y=surv_prob))
+
[…]

### 2023-03-04 code/install_R_packages.R

@@ -1,0 +1,21 @@
+# Install the packages necessary for R analyses
+
+list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
+					  "rstan", "lubridate", "tidyverse", 
+					  "coda","mvtnorm","devtools","loo","dagitty")
+new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+# Load Github packages
+github_packages = c("rethinking", "mrmr")
+
+for(package in github_packages){
+
+	if(!(package %in% installed.packages()[,"Package"])){
+		if(package == "rethinking"){
+			devtools::install_github("rmcelreath/rethinking")
+		} else if(package == "mrmr"){
+			devtools::install_github("SNARL1/mrmr")
+		} 
+	}
+}

### 2023-03-04 code/survival_table_nontranslocated.R

@@ -1,0 +1,142 @@
+library(tidyverse)
+
+# Code from MRMR package with slight modifications
+
+get_numeric_indices <- function(string) {
+  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
+  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
+  tibble::as_tibble(idx_mat) %>%
+    dplyr::mutate_all(readr::parse_integer)
+}
+
+mean_na = function(x){
+
+  if(all(is.na(x))) {
+    return(NA)
+  }else{
+    return(mean(x, na.rm=TRUE))
+  }
+}
+
+get_surv_vector = function(primary_period, value){
+
+  val1 = value[primary_period < max(primary_period)]
+  val2 = value[primary_period > min(primary_period)]
+
+  # Survived 
+  died = (val1 == 2) & (val2 == 3) # Died
+  survived =  (val1 == 2) & (val2 == 2) # Survived
+  died_vect = array(NA, length(died))
+  died_vect[died] = 0
+  died_vect[survived] = 1
+  return(as.list(died_vect))
+
+}
+
+convert_to_death_rates = function(surv_probs, deltat){
+  # Convert survival probabilities to death rates given
+  # deltat.  Assuming geometric survival proability
+  #
+  # Parameters
+  # ----------
+  # surv_probs : array-like
+  #   Survival probabilities
+  # deltat : array-like
+  #   Time period over which survival probabiliites are calculated
+  #
+  # Returns
+  # -------
+  # : death ratess
+
+  rates = -log(surv_probs) / deltat
+  return(rates)
+
+}
+
+survival_table <- function(model, thin=1, upperlower=c(0.25, 0.75)) {
+  # Take in MRMR model and return survival probabilities for pit-tagged 
+  # non-translocated individuals.
+  #
+  # Parameters
+  # ----------
+  # model : MRMR fitted model
+  # thin : int, thin the bayesian chains
+  #
+  # Returns
+  # -------
+  # : list 
+  #   number of pit tags in population, survival probabilities of non-translocated
+
+  any_translocations <- 'data.frame' %in% class(model$data$translocations)
+
+  if (!any_translocations) {
+    stop(paste("No translocation data are present, so a cohort survival",
+               "table cannot be created."))
+  }
+
+   primary_period_dates <- model$data$surveys %>%
+    group_by(.data$primary_period) %>%
+    summarize(date = min(.data$survey_date),
[…]

### 2023-03-05 code/install_R_packages.R
Add cran link to install_R_packages.R

@@ -4,7 +4,7 @@ list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magritt
 					  "rstan", "lubridate", "tidyverse", 
 					  "coda","mvtnorm","devtools","loo","dagitty")
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
+if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
 
 # Load Github packages
 github_packages = c("rethinking", "mrmr")

### 2023-03-10 code/extract_survival_estimates.R
Remove absolute path

@@ -25,9 +25,8 @@ for(l in 1:length(lake_ids)){
 
 	lake_number = lake_ids[l]
 	cat("Working on lake", lake_number, "\n")
-	basepath = file.path("", "Users", "mqwilber", "Repos")
-	modelpath = file.path(basepath, "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
-	datapath = file.path(basepath, "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+	modelpath = file.path("..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path("..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
 
 	model = readRDS(modelpath)
 	survey_dat = fread(datapath)

### 2023-03-10 code/translocation_survival_analysis.Rmd
Update and reorganize all analyses in Rmd files

@@
 ---
-title: "Predictors of frog survival following translocations - analyses"
+title: "Predictors of frog survival following translocations - primary analyses"
 author: "Roland Knapp"
@@
 
-The goal of the analyses in this notebook is to identify predictors of frog survival following translocation.
+The goal of these analyses is to identify predictors of frog survival following translocation.
 The analyses are conducted in a Bayesian framework using the R package rstanarm.
 
-# Analysis 1: Site-level and frog-level predictors
-
-## Load packages
-
-```{r load-packages}
+## ANALYSIS 1: Effect of site on survival of all translocated cohorts
+
+### Visualize survival across sites
+
+#### Read in frog translocation dataset
+
+```{r}
+frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  mutate(site_id = as.character(site_id),
+         donor = as.character(donor),
+         surv = as.integer(if_else(survival < 0.5, 0, 1)))
+```
+
+#### Check for null values
+
+```{r}
+frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
+```
+
+-   12 records where bd_load is NA, no other NAs in dataset
+
+#### Retrieve cohort-level frog survival data
+
+```{r}
+files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survival"), glob = "*_cohort.csv")
+cohort_survival <- read_csv(files, id = "path") %>% 
+  mutate(site_id = str_extract(path, "(\\d)+"), .after = path) %>%  # extracts site_id as numeric string from file path
+  rename(years_since_intro = years_since_introduction) %>% 
+  filter(years_since_intro == 1) %>% 
+  select(site_id, release_date, years_since_intro, median_survival, lo_survival, hi_survival)
+rm(files)
+```
+
+#### Plot of frog survival for each translocation - grouped by site and arranged by site-specific median survival
+
+```{r}
+palette1 <- c("green4", "blue", "black")
+
+donors <- frog_translocation %>% 
+  distinct(site_id, donor)
+
+cohort_survival_plot <- cohort_survival %>% 
+  mutate(release_year = year(release_date)) %>% 
+  group_by(site_id) %>% 
+  summarize(plot_survival = mean(median_survival)) %>%  # create variable by which to order sites for plotting
+  arrange(plot_survival) %>% 
+  mutate(siteid_order = row_number()) %>% 
+  inner_join(cohort_survival, by = "site_id") %>% 
+  inner_join(donors, by = "site_id") %>% 
+  select(-plot_survival) %>%
+  ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor)) +  # reorder site_id by siteid_order)
+      geom_linerange(aes(ymin = lo_survival, ymax = hi_survival), color = "black", width = 0.5, position = position_dodge2(width=0.5)) +
+      geom_point(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor), size = 3, position = position_dodge2(width=0.5)) + 
+      scale_color_manual(values = palette1) + 
+      labs(x = "Recipient site", y = "1-year frog survival") +
+      theme_classic() +
+      theme(
+        legend.background = element_rect(color = "black", size = 0.3),
+        legend.position = c(0.2, 0.8),
+        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
+
+cohort_survival_plot
+```
+
+-   Strong effect of site/translocation on frog survival.
+-   Survival is highly variable across sites, but relatively invariant within a site.
+-   Indicates need to include site_id and/or translocation_id as group-level effect(s).
+
+#### Save plot
+
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "translocation_survival_bysiteid.png"), width = 5, height = 5) 
+```
+
+#### Plot of frog survival for each translocation - grouped by elevation
+
+```{r}
+frog_translocation %>%
+  distinct(site_id, date, elevation) %>% 
+  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
+    ggplot(aes(x = elevation, y = median_survival)) + 
+      geom_vline(aes(xintercept = elevation), linetype = "dotted", color = "darkgray") +
+      geom_point(size = 3) +
+      geom_text(aes(label = site_id, x = elevation, y = 1.05), angle = 90) +
+      coord_cartesian(ylim = c(0, 1.08)) +
+      labs(x = "Elevation of recipient site", y = "1-year frog survival")
+```
+
+-   Elevation has a clear effect on frog survival, but there is a substantial amount of variation that remains unaccounted for.
+
+### Model survival of frogs from first translocation as predictor of survival in subsequent translocations
+
+-   This analysis explores the results shown in the plot created above (frog survival for all translocations grouped by site).
+
+#### Load packages
+
+```{r}
+library(tidyverse)
+library(brms)
+library(rstanarm)
+library(bayesplot)
+library(tidybayes)
+library(loo)
+library(broom.mixed)
+library(janitor)
+
+source("classification_summary.R")
+```
+
+#### Create frog translocation dataset for analysis
+
+-   Read in data, create variables for analysis, standardize continuous variables.
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+```{r}
+d2 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  select(site_id, year, pit_tag_ref, survival) %>% 
+  mutate(
+    surv = if_else(survival < 0.5, 0, 1), 
+    across(c(site_id, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d2 <- d2 %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id) %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d2, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id) %>% 
+  relocate(transno, .after = translocation_id)
+
+```
+
+#### Restructure dataset
+
+```{r}
+# Remove sites that received only 1 translocation
+d2 <- d2 %>% 
+  distinct(site_id, transno) %>% 
+  group_by(site_id) %>% 
+  mutate(trans_max = max(transno)) %>% 
+  ungroup(site_id) %>% 
+  select(-transno) %>% 
+  filter(trans_max > 1) %>% 
+  distinct(site_id, trans_max) %>% 
+  inner_join(d2, by = c("site_id"))
+
+# Create predictor from survival of first translocated cohort
+d2 <- d2 %>% 
+  filter(transno == 1) %>% 
+  group_by(site_id) %>% 
+  mutate(survival_trans1 = mean(survival)) %>%    # use mean or median?
+  ungroup(site_id) %>% 
+  distinct(site_id, survival_trans1) %>% 
+  mutate(across(c(survival_trans1), round, 3)) %>% 
+  inner_join(d2, by = c("site_id")) %>% 
+  filter(transno != 1) %>% 
+  relocate(translocation_id, .after = site_id) %>% 
+  mutate(survival_trans1_z = arm::rescale(survival_trans1), .after = survival_trans1) %>% 
+  select(-transno, -trans_max) 
+```
+
+#### Model with non-standardized predictor, no grouping variable
+
+##### Model specification
+
+```{r}
+m2a <- stan_glm(
+  surv ~ survival_trans1,
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+##### List model variable names
+
+```{r}
+get_variables(m2a)
+```
+
+##### Priors & chain diagnostics
+
+```{r}
+prior_summary(m2a)
+
+b2 <- c("(Intercept)", "survival_trans1")
+
+mcmc_trace(m2a, size = 0.1, pars = b2)
+mcmc_dens_overlay(m2a, pars = b2)
+
+summary(m2a)
+```
+
+##### Posterior predictive check
+
+```{r}
+pp_check(m2a, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+##### Calculate loo
+
+```{r}
+loo_m2a <- loo(m2a, cores = 4, save_psis = TRUE)
+print(loo_m2a)
+```
+
+##### Posterior analysis
+
+```{r}
+tidy(m2a, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2a, area_method = "equal height", prob = 0.95, pars = b2) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with unstandardized predictor & no grouping variable",
+          "Posterior distributions with medians & 95% uncertainty intervals")
+
+mean(bayes_R2(m2a))
+```
+
+#### Model with standardized predictors, no grouping variable
+
+##### Model specification
+
+```{r}
+m2b <- stan_glm(
+  surv ~ survival_trans1_z,
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+##### List model variable names
+
+```{r}
+get_variables(m2b)
+```
+
+##### Priors & chain diagnostics
+
+```{r}
+prior_summary(m2b)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2b, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2b, pars = b3)
+
+summary(m2b)
+```
+
+##### Posterior predictive check
+
+```{r}
+pp_check(m2b, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+##### Calculate loo
+
+```{r}
+loo_m2b <- loo(m2b, cores = 4, save_psis = TRUE)
+print(loo_m2b)
+```
+
+##### Posterior analysis
+
+```{r}
+tidy(m2b, conf.int = TRUE, conf.level = 0.90)
+
+labels_alt <- c("survival_trans1", "(Intercept)")
+mcmc_areas(m2b, area_method = "equal height", prob = 0.95, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +
+  scale_y_discrete(labels = labels_alt, limits = rev)
+#  ggtitle("Model with standardized predictor & no grouping variable",
+#          "Posterior distributions with medians & 95% uncertainty intervals")
+
+mean(bayes_R2(m2b))
+```
+
+##### Save plot
+
+```{r}
+ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m2b.png"), bg = "white", width = 7, height = 4)
+```
+
+#### Model with standardized predictors, translocation_id grouping variable
+
+##### Model specification
+
+```{r}
+m2c <- stan_glmer(
+  surv ~ survival_trans1_z + (1 | translocation_id),
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+##### Priors & chain diagnostics
+
+```{r}
+prior_summary(m2c)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2c, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2c, pars = b3)
+
+summary(m2c)
+```
+
+##### Posterior predictive check
+
+```{r}
+pp_check(m2c, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+##### Calculate loo
+
+```{r}
+loo_m2c <- loo(m2c, cores = 4, save_psis = TRUE)
+print(loo_m2c)
+```
+
+##### Posterior analysis
+
+```{r}
+tidy(m2c, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2c, area_method = "equal height", prob = 0.95, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictor & translocation_id as grouping variable",
+          "Posterior distributions with medians & 95% uncertainty intervals")
+
+mean(bayes_R2(m2c))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m2c, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+
+#### Model with standardized predictors, site_id grouping variable
+
+##### Model specification
+
+```{r}
+m2d <- stan_glmer(
+  surv ~ survival_trans1_z + (1 | site_id),
+  data = d2,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+##### Priors & chain diagnostics
+
+```{r}
+prior_summary(m2d)
+
+b3 <- c("(Intercept)", "survival_trans1_z")
+
+mcmc_trace(m2d, size = 0.1, pars = b3)
+mcmc_dens_overlay(m2d, pars = b3)
+
+summary(m2d)
+```
+
+##### Posterior predictive check
+
+```{r}
+pp_check(m2d, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+##### Calculate loo
+
+```{r}
+loo_m2d <- loo(m2d, cores = 4, save_psis = TRUE)
+print(loo_m2d)
+```
+
+##### Posterior analysis
+
+```{r}
+tidy(m2d, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m2d, area_method = "equal height", prob = 0.95, pars = b3) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictor & site_id as grouping variable",
+          "Posterior distributions with medians & 95% uncertainty intervals")
+
+mean(bayes_R2(m2d))
+
+# Frog survival by site_id
+mcmc_areas_ridges(m2d, regex_pars = "site_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each site_id")
+```
+
+#### Compare 3 models with different grouping structures
+
+```{r}
+loo_compare(loo_m2b, loo_m2c, loo_m2d)
+```
+
+-   Based on expected log-predictive densities (ELPD), models with and without grouping variables have similar posterior predictive accuracy (elpd_diff of m2b is less than 2 SE lower than elpd_diff of m2d: elpd_diff = -3.1 $\pm$ 5.8, 2 SE interval overlaps 0: 2.7 -- -8.9).
+-   Conclusion from all 3 models is the same: Average frog survival of first translocation is an important predictor of survival of individual frogs in subsequent translocations.
+
+#### Posterior classification accuracy of best models
+
+```{r}
+# Generate confusion matrix for m2b
+set.seed(84735)
+classification_summary(m2b, data = d2, cutoff = 0.5)
+
+# Generate confusion matrix for m2d
+set.seed(84735)
+classification_summary(m2d, data = d2, cutoff = 0.5)
+```
+
+-   As expected from loo results, models with and without grouping variable have the same posterior predictive accuracy.
+
+### TO DO
+
+-   Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).
+
+## ANALYSIS 2: Model of site-level and frog-level predictors
+
+### Load packages
+
+```{r}
 library(tidyverse)
@@
 
-## Create frog translocation dataset for analysis
+### Create frog translocation dataset for analysis
 
@@
 
-```{r read-translocation-data}
+```{r}
 d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
@@
 
-## Null values - check
-
-```{r check-null}
+### Check for null values
+
+```{r}
 d1 %>% summarize(across(everything(), ~ sum(is.na(.))))
@@
 
-## Model using non-standardized predictors
-
-### Model specification
+### Model using non-standardized predictors
+
+#### Model specification
 
@@
 
-## Explore a flat prior
-
-We can explore a flat prior on the intercept using the code below.
+#### Explore a flat prior on intercept
 
@@
 
-### Priors & chain diagnostics
+#### Priors & chain diagnostics
 
@@
 
-### List model variable names
+#### List model variable names
 
@@
 
-### Posterior predictive check
+#### Posterior predictive check
 
@@
 
-### Calculate leave-one-out cross-validation (loo)
+#### Calculate leave-one-out cross-validation (loo)
 
@@
 
-### Posterior analysis
+#### Posterior analysis
 
@@
 
--   Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.\
+-   Using non-standardized predictors makes plot hard to read due to vastly different scales of predictors.
 -   Based on 95% uncertainty intervals, important predictors are snow_t1, elevation, first, donor70567, and donor72996.
 
-## Model using standardized predictors
-
-### Model specification
+### Model using standardized predictors
+
+#### Model specification
 
@@
 
-### Priors, chain diagnostics, posterior predictive check
+#### Priors, chain diagnostics, posterior predictive check
 
@@
 
-### Calculate loo
+#### Calculate loo
 
@@
 
-## Compare models
+#### Compare models
 
@@
 
-## Possible grouping structures to add to model
-
-The dataset has three groups: site_id, translocation_id, and donor site.
-Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful.
-Site_id and translocation_id contain 12 and 24 levels, respectively, making them suitable as grouping variables.
-Including both site_id and translocation_id as nested grouping variables (i.e., translocations nested within site ids) also seems justified.
-Having more levels within a group would seem to increase the power to detect effects of predictors.
-With fewer levels/less variability, effect may be accounted for primarily by grouping variable itself instead of predictor.
-With only 1-4 translocations per site and only 12 site ids total, it is possible that model with site_id as sole grouping variable and model with site_id and translocation_id as nested grouping variables will attribute more of the variation in frog survival to the grouping variables and less to the predictors.
-A priori, that would suggest that the model with translocation_id as the grouping variable might be the most useful model structure (assuming that all models have similar posterior predictive accuracy).
-Lacking any strong rationale for one model structure over another, built 3 models and compared their posterior predictive accuracy using loo.
-
-## Model using standardized predictors and site_id grouping variable
-
-### Model specification
+### Model using standardized predictors and site_id grouping variable
+
+#### Model specification
 
@@
 
-## Model using standardized predictors and translocation_id grouping variable)
-
-### Model specification
+### Model using standardized predictors and translocation_id grouping variable)
+
+#### Model specification
 
@@
 
-### Save plot
+#### Save plot
 
@@
 
-## Model using standardized predictors and 2 grouping variables: translocation_id nested within site_id)
-
-### Model specification
+### Model using standardized predictors and 2 grouping variables: translocation_id nested within site_id)
+
+#### Model specification
 
@@
 
-## Compare 4 models with different grouping structures
+### Compare 4 models with different grouping structures
 
@@
 -   Based on expected log-predictive densities (ELPD), 3 models with grouping variables all have substantially higher posterior predictive accuracy than model without any grouping variables (elpd_diff of m1b is more than 2 SE lower than elpd_diff of m1c: elpd_diff = -18.2 $\pm$ 12.6, 2 SE interval does not overlap 0: -5.6 -- -30.8).
--   3 models with 1 or more grouping variables are equivalent and have similar posterior predictive accuracy (elpd_diff of m1e and m1d are within 2 SE of elpd_diff of m1c).\
+-   3 models with 1 or more grouping variables are equivalent and have similar posterior predictive accuracy (elpd_diff of m1e and m1d are within 2 SE of elpd_diff of m1c).
 -   m1d seems to provide more insight into the effect of the predictors, so chose this model for additional analysis.
 
-## Additional analysis of model m1d
-
-### Posterior classification accuracy
+### Additional analysis of model m1d
+
+#### Posterior classification accuracy
 
@@
 
--   Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.\
+-   Sensitivity or "true positive rate" measures the proportion of Y = 1 observations that are accurately classified.
 -   Specificity or "true negative rate" measures the proportion of Y = 0 observations that are accurately classified (i.e., proportion of frogs correctly identified as not surviving).
 
-## Reduced model from m1d
-
-### Create reduced model containing only important predictors for subsequent analyses
-
-- Reduced model is helpful because it reduces the number of values used to calculate probabilities from predictive models.
+### Reduced model from m1d
+
+#### Create reduced model containing only important predictors for subsequent analyses
 
@@
 
-### Posterior analysis
+#### Posterior analysis
+
 ```{r}
@@
 
-### Confirm that fit of reduced model is comparable to full model
+#### Confirm that fit of reduced model is comparable to full model
 
@@
 
-### Simulate and plot posterior predictive models & observations
-#### Reduced model with translocation_id as grouping variable
+#### Simulate and plot posterior predictive models & observations
+
+##### Reduced model with translocation_id as grouping variable
+
 ```{r}
@@
 
-#### Reduced model without grouping variable
+##### Reduced model without grouping variable
+
 To explore how addition of grouping variable affected predictions, especially that for 70370 and 70550.
+
 ```{r}
@@
 
-### Values of important predictors for translocations of interest
+##### Values of important predictors for translocations of interest
+
 ```{r}
@@
 
-### Plot posterior plausible relationships between each predictor and the probability of frog survival (conditional effects)
-
-- As described in <https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html>, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on.
-By default, the mean is used for continuous variables and the reference category is used for factors, but you may change these values via argument conditions.
-This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."\
--   Using rstanarm, I was unable to do this as I expected to: Start with data = d1 and using "newdata =" to specify that predictions should be based on data in d1_elevz (for example), in which non-plotted continuous predictors are set to the mean value.
-Not sure if my alternative approach is correct.\
--   Approach used: Instead of starting the routine with d1, I started with the new dataset in which all variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).
-This produced plots/results that were very similar to those produced in brms using the conditional_effects function.
-
-#### Effect of elevation
+#### Plot posterior plausible relationships between each predictor and the probability of frog survival (conditional effects)
+
+-   All variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).
+
+##### Effect of elevation
 
@@
 
-#### Effect of winter severity in year t1
+##### Effect of winter severity in year t1
 
@@
 
-#### Effect of sex
+##### Effect of sex
 
@@
 
-#### Effect of donor
+##### Effect of donor
 
@@
 
-#### Joint effects of elevation and sex
+##### Joint effects of elevation and sex
 
@@
 
-### Create multiplot conditional effects figure
+##### Create multi-panel conditional effects figure
 
@@
 
-### Save plot
+##### Save plot
 
@@
 
-## For comparison, fit model using brms and conditional_effects function to plot variable-specific effects
-
-### Fit model
+#### For comparison, fit model using brms and conditional_effects function to plot variable-specific effects
+
+##### Fit model
 
@@
 
-### Plot conditional effects
+##### Plot conditional effects
 
@@
 ```
-
-# Analysis 2: Survival of frogs from first translocation as predictor of survival in subsequent translocations
-
-This analysis explores the results shown in translocation_survival_exploratory.Rmd (code chunk name = "frog-survival-by-site").
-This plot shows frog survival of all translocations grouped by site, and indicates that survival is highly variable across sites, but repeatable within sites.
-The following analysis quantifies the association between survival in the first translocation and survival in subsequent translocations.
-
-## Load packages
-
-```{r}
-library(tidyverse)
-library(brms)
-library(rstanarm)
-library(bayesplot)
-library(tidybayes)
-library(loo)
-library(broom.mixed)
-library(janitor)
-
-source("classification_summary.R")
-```
-
-## Create frog translocation dataset for analysis
-
--   Read in data, create variables for analysis, standardize continuous variables.
--   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
-
-```{r}
-d2 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  select(site_id, year, pit_tag_ref, survival) %>% 
-  mutate(
-    surv = if_else(survival < 0.5, 0, 1), 
-    across(c(site_id, year), as.factor),
-    surv = as.integer(surv))
-
-# Add a descriptive translocation_id
-d2 <- d2 %>% 
-  distinct(site_id, year) %>% 
-  arrange(site_id, year) %>% 
-  group_by(site_id) %>% 
-  mutate(transno = seq_len(n())) %>% 
-  ungroup(site_id) %>% 
-  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
-  mutate(translocation_id = as.factor(translocation_id)) %>% 
-  inner_join(d2, by = c("site_id", "year")) %>% 
-  relocate(translocation_id, .after = site_id) %>% 
-  relocate(transno, .after = translocation_id)
-
-```
-
-## Restructure dataset
-
-```{r}
-# Remove sites that received only 1 translocation
-d2 <- d2 %>% 
-  distinct(site_id, transno) %>% 
-  group_by(site_id) %>% 
-  mutate(trans_max = max(transno)) %>% 
-  ungroup(site_id) %>% 
-  select(-transno) %>% 
-  filter(trans_max > 1) %>% 
-  distinct(site_id, trans_max) %>% 
-  inner_join(d2, by = c("site_id"))
-
-# Create predictor from survival of first translocated cohort
-d2 <- d2 %>% 
-  filter(transno == 1) %>% 
-  group_by(site_id) %>% 
-  mutate(survival_trans1 = mean(survival)) %>%    # use mean or median?
-  ungroup(site_id) %>% 
-  distinct(site_id, survival_trans1) %>% 
-  mutate(across(c(survival_trans1), round, 3)) %>% 
-  inner_join(d2, by = c("site_id")) %>% 
-  filter(transno != 1) %>% 
-  relocate(translocation_id, .after = site_id) %>% 
-  mutate(survival_trans1_z = arm::rescale(survival_trans1), .after = survival_trans1) %>% 
-  select(-transno, -trans_max) 
-```
-
-## Model with non-standardized predictor, no grouping variable
-
-### Model specification
-
-```{r}
-m2a <- stan_glm(
-  surv ~ survival_trans1,
-  data = d2,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### List model variable names
-
-```{r}
-get_variables(m2a)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m2a)
-
-b2 <- c("(Intercept)", "survival_trans1")
-
-mcmc_trace(m2a, size = 0.1, pars = b2)
-mcmc_dens_overlay(m2a, pars = b2)
-
-summary(m2a)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m2a, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Calculate loo
-
-```{r}
-loo_m2a <- loo(m2a, cores = 4, save_psis = TRUE)
-print(loo_m2a)
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m2a, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m2a, area_method = "equal height", prob = 0.95, pars = b2) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with unstandardized predictor & no grouping variable",
-          "Posterior distributions with medians & 95% uncertainty intervals")
-
-mean(bayes_R2(m2a))
-```
-
-## Model with standardized predictors, no grouping variable
-
-### Model specification
-
-```{r}
-m2b <- stan_glm(
-  surv ~ survival_trans1_z,
-  data = d2,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### List model variable names
-
-```{r}
-get_variables(m2b)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m2b)
-
-b3 <- c("(Intercept)", "survival_trans1_z")
-
-mcmc_trace(m2b, size = 0.1, pars = b3)
-mcmc_dens_overlay(m2b, pars = b3)
-
-summary(m2b)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m2b, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Calculate loo
-
-```{r}
-loo_m2b <- loo(m2b, cores = 4, save_psis = TRUE)
-print(loo_m2b)
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m2b, conf.int = TRUE, conf.level = 0.90)
-
-labels_alt <- c("survival_trans1", "(Intercept)")
-mcmc_areas(m2b, area_method = "equal height", prob = 0.95, pars = b3) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +
-  scale_y_discrete(labels = labels_alt, limits = rev)
-#  ggtitle("Model with standardized predictor & no grouping variable",
-#          "Posterior distributions with medians & 95% uncertainty intervals")
-
-mean(bayes_R2(m2b))
-```
-
-### Save plot
-```{r}
-ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m2b.png"), bg = "white", width = 7, height = 4)
-```
-
-## Model with standardized predictors, translocation_id grouping variable
-
-### Model specification
-
-```{r}
-m2c <- stan_glmer(
-  surv ~ survival_trans1_z + (1 | translocation_id),
-  data = d2,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m2c)
-
-b3 <- c("(Intercept)", "survival_trans1_z")
-
-mcmc_trace(m2c, size = 0.1, pars = b3)
-mcmc_dens_overlay(m2c, pars = b3)
-
-summary(m2c)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m2c, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Calculate loo
-
-```{r}
-loo_m2c <- loo(m2c, cores = 4, save_psis = TRUE)
-print(loo_m2c)
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m2c, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m2c, area_method = "equal height", prob = 0.95, pars = b3) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictor & translocation_id as grouping variable",
-          "Posterior distributions with medians & 95% uncertainty intervals")
-
-mean(bayes_R2(m2c))
-
-# Frog survival by translocation_id
-mcmc_areas_ridges(m2c, regex_pars = "translocation_id") +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Frog survival for each translocation_id")
-```
-
-## Model with standardized predictors, site_id grouping variable
-
-### Model specification
-
-```{r}
-m2d <- stan_glmer(
-  surv ~ survival_trans1_z + (1 | site_id),
-  data = d2,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m2d)
-
-b3 <- c("(Intercept)", "survival_trans1_z")
-
-mcmc_trace(m2d, size = 0.1, pars = b3)
-mcmc_dens_overlay(m2d, pars = b3)
-
-summary(m2d)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m2d, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Calculate loo
-
-```{r}
-loo_m2d <- loo(m2d, cores = 4, save_psis = TRUE)
-print(loo_m2d)
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m2d, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m2d, area_method = "equal height", prob = 0.95, pars = b3) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictor & site_id as grouping variable",
-          "Posterior distributions with medians & 95% uncertainty intervals")
-
-mean(bayes_R2(m2d))
-
-# Frog survival by site_id
-mcmc_areas_ridges(m2d, regex_pars = "site_id") +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Frog survival for each site_id")
-```
-
-### Compare 3 models with different grouping structures
-
-```{r}
-loo_compare(loo_m2b, loo_m2c, loo_m2d)
-```
-
--   Based on expected log-predictive densities (ELPD), models with and without grouping variables have similar posterior predictive accuracy (elpd_diff of m2b is less than 2 SE lower than elpd_diff of m2d: elpd_diff = -3.1 $\pm$ 5.8, 2 SE interval overlaps 0: 2.7 -- -8.9).\
--   Conclusion from all 3 models is the same: Average frog survival of first translocation is an important predictor of survival of individual frogs in subsequent translocations.
-
-## Posterior classification accuracy of best models
-
-```{r}
-# Generate confusion matrix for m2b
-set.seed(84735)
-classification_summary(m2b, data = d2, cutoff = 0.5)
-
-# Generate confusion matrix for m2d
-set.seed(84735)
-classification_summary(m2d, data = d2, cutoff = 0.5)
-```
-
--   As expected from loo results, models with and without grouping variable have the same posterior predictive accuracy.
-
-## TO DO
-
--   Validate priors, tune as necessary. Add final priors to model specification (in case the defaults change).\
--   Check approach to plotting conditional effects.\
--   Try un-standardizing predictor values in conditional-effects plots (and other plots?) to produce more intuitive visualizations.

### 2023-03-10 code/translocation_survival_analysis_suppl.Rmd
Update and reorganize all analyses in Rmd files

@@
 
-The supplemental analyses in this notebook provide additional details to analyses in `translocation_survival_analysis.Rmd`.
+The following are supplemental analyses, some related to analyses in `translocation_survival_analysis.Rmd` and some unrelated.
 
@@
 
-```{r load-packages}
+```{r}
 library(tidyverse)
@@
 
-## Evaluate effect of frog size independent of donor population
-
--   Model m1d from `translocation_survival_analysis.Rmd` but using data subset where donor = 72996
--   Read in data, create variables for analysis, standardize continuous variables.
--   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
-
-### Read in frog translocation dataset
-
-```{r read-translocation-data}
-d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
-  mutate(
-    first = if_else(order == 1, 1, 0),
-    surv = if_else(survival < 0.5, 0, 1),
-    male = if_else(sex == "m", 1, 0),
-    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
-    elev_z = arm::rescale(elevation),
-    snowt_z = arm::rescale(snow_t),
-    snowt1_z = arm::rescale(snow_t1),
-    day_z = arm::rescale(day),
-    length_z = arm::rescale(length),
-    bdload_l = log10(bd_load + 1),
-    bdload_z = arm::rescale(bdload_l),
-    shore_c = shore - mean(shore),
-    male_c = male - mean(male),
-    elevlo_c = elev_lo - mean(elev_lo),
-    first_c = first - mean(first),
-    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
-    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
-    surv = as.integer(surv))
-
-# Add a descriptive translocation_id
-d1 <- d1 %>% 
-  distinct(site_id, year) %>% 
-  arrange(site_id, year) %>% 
-  group_by(site_id) %>% 
-  mutate(transno = seq_len(n())) %>% 
-  ungroup(site_id) %>% 
-  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
-  select(-transno) %>% 
-  mutate(translocation_id = as.factor(translocation_id)) %>% 
-  inner_join(d1, by = c("site_id", "year")) %>% 
-  relocate(translocation_id, .after = site_id)
-```
-
-### Model specification
-
-```{r}
-d1_72996 <- d1 %>% 
-  filter(donor == "72996" & !is.na(bd_load))
-
-m1d_72996 <- stan_glmer(
-  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + (1 | translocation_id),
-  data = d1_72996,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m1d_72996)
-
-b1_72996 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548")
-
-mcmc_trace(m1d_72996, size = 0.1, pars = b1_72996)
-mcmc_dens_overlay(m1d_72996, pars = b1_72996)
-
-summary(m1d_72996)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m1d_72996, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m1d_72996, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m1d_72996, area_method = "equal height", prob = 0.90, pars = b1_72996) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
-
-mean(bayes_R2(m1d_72996))
-
-# Frog survival by translocation_id
-mcmc_areas_ridges(m1d_72996, regex_pars = "translocation_id") +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Frog survival for each translocation_id")
-```
-
--   Using dataset containing only frogs translocated from 72996 had relatively little effect on the predictors of frog survival.
-Specific to the objective of this analysis, frog size remained unimportant as a predictor of survival.
-However, effect of Bd load became marginally unimportant instead of unimportant (i.e., uncertainty interval includes 0 but barely).
-
-## Assess effect on m1d of including all records, including where bd_load = NA
-
-### Read in frog translocation dataset
-
--   Read in data, create variables for analysis, standardize continuous variables.
--   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
-
-```{r read-translocation-data-including-na}
-d1_allna <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-#  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
-  mutate(
-    first = if_else(order == 1, 1, 0),
-    surv = if_else(survival < 0.5, 0, 1),
-    male = if_else(sex == "m", 1, 0),
-    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
-    elev_z = arm::rescale(elevation),
-    snowt_z = arm::rescale(snow_t),
-    snowt1_z = arm::rescale(snow_t1),
-    day_z = arm::rescale(day),
-    length_z = arm::rescale(length),
-    bdload_l = log10(bd_load + 1),
-    bdload_z = arm::rescale(bdload_l),
-    shore_c = shore - mean(shore),
-    male_c = male - mean(male),
-    elevlo_c = elev_lo - mean(elev_lo),
-    first_c = first - mean(first),
-    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
-    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
-    surv = as.integer(surv))
-
-# Add a descriptive translocation_id
-d1_allna <- d1_allna %>% 
-  distinct(site_id, year) %>% 
-  arrange(site_id, year) %>% 
-  group_by(site_id) %>% 
-  mutate(transno = seq_len(n())) %>% 
-  ungroup(site_id) %>% 
-  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
-  select(-transno) %>% 
-  mutate(translocation_id = as.factor(translocation_id)) %>% 
-  inner_join(d1_allna, by = c("site_id", "year")) %>% 
-  relocate(translocation_id, .after = site_id)
-```
-
-### Model specification
-
-```{r}
-m1d_allna <- stan_glmer(
-  surv ~ length_z + snowt_z + snowt1_z + day_z + elev_z + first_c + male_c + donor + (1 | translocation_id),
-  data = d1_allna,
-  family = binomial,
-  chains = 4, 
-  iter = 5000*2, 
-  cores = 4,
-  seed = 84735)
-```
-
-### List model variable names
-
-```{r}
-get_variables(m1d_allna)
-```
-
-### Priors & chain diagnostics
-
-```{r}
-prior_summary(m1d_allna)
-
-b1_allna <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "elev_z", "first_c0.349", "male_c0.549", "donor70567", "donor72996")
-
-mcmc_trace(m1d_allna, size = 0.1, pars = b1_allna)
-mcmc_dens_overlay(m1d_allna, pars = b1_allna)
-
-summary(m1d_allna)
-```
-
-### Posterior predictive check
-
-```{r}
-pp_check(m1d_allna, plotfun = "stat") +
-    xlab("Frog survival rate")
-```
-
-### Posterior analysis
-
-```{r}
-tidy(m1d_allna, conf.int = TRUE, conf.level = 0.90)
-
-mcmc_areas(m1d_allna, area_method = "equal height", prob = 0.90, pars = b1_allna) +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
-          "Posterior distributions with medians & 90% uncertainty intervals")
-
-mean(bayes_R2(m1d_allna))
-
-# Frog survival by translocation_id
-mcmc_areas_ridges(m1d_allna, regex_pars = "translocation_id") +
-  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
-  ggtitle("Frog survival for each translocation_id")
-```
-
--   Inclusion of all records (i.e., 12 records where bd_load = NA that were excluded in primary analysis) did not change the results.
-
-## Analysis of changes in Bd load following translocation
+## ANALYSIS 1: Analysis of changes in Bd load following translocation
 
@@
 
-```{r read-data}
+```{r}
 translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv")) %>% 
@@
 
--   site_id: Unique 5-digit site identifier.\
+-   site_id: Unique 5-digit site identifier.
 -   site_id1: Assigned 5-digit site identifier for a collection of adjacent sites that are utilized by a frog population.
--   translocation_id: Unique identifier for each translocation.\
--   visit_date: Date on which translocation or survey was conducted.\
--   survey_type: "translocation" or "cmr".\
+-   translocation_id: Unique identifier for each translocation.
+-   visit_date: Date on which translocation or survey was conducted.
+-   survey_type: "translocation" or "cmr".
 -   period: "before" or "after" translocation.
--   year_translocate: Year in which translocation was conducted.\
--   year_capture: Year in which frog was captured.\
--   interval: Number of years between translocation and frog capture.\
--   pit_tag_ref: Unique PIT tag identifier.\
--   swab_id: Unique identifier for skin swab sample.\
+-   year_translocate: Year in which translocation was conducted.
+-   year_capture: Year in which frog was captured.
+-   interval: Number of years between translocation and frog capture.
+-   pit_tag_ref: Unique PIT tag identifier.
+-   swab_id: Unique identifier for skin swab sample.
 -   bd_load: Bd infection intensity determined from skin swab samples, units are number of ITS1 copies \* 80.
@@
 ### Analyze dataset
+
+-   Model with pit_tag_ref grouping variable is the model of primary interest, roughly analogous to paired t-test. Included other models to allow a better understanding of the data.
 
@@
 
-#### Null values - check
+#### Check for null values
 
@@
 
-## Calculate parameters for viability analysis
+-   All models are equivalent, so pairing before-after by pit_tag_ref is unimportant.
+
+## ANALYSIS 2: Evaluate effect of frog size independent of donor population
+
+-   Model m1d from `translocation_survival_analysis.Rmd` but using data subset where donor = 72996 only.
+
+### Read in frog translocation dataset
+
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+```{r read-translocation-data}
+d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
+  mutate(
+    first = if_else(order == 1, 1, 0),
+    surv = if_else(survival < 0.5, 0, 1),
+    male = if_else(sex == "m", 1, 0),
+    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
+    elev_z = arm::rescale(elevation),
+    snowt_z = arm::rescale(snow_t),
+    snowt1_z = arm::rescale(snow_t1),
+    day_z = arm::rescale(day),
+    length_z = arm::rescale(length),
+    bdload_l = log10(bd_load + 1),
+    bdload_z = arm::rescale(bdload_l),
+    shore_c = shore - mean(shore),
+    male_c = male - mean(male),
+    elevlo_c = elev_lo - mean(elev_lo),
+    first_c = first - mean(first),
+    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
+    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d1 <- d1 %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id) %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  select(-transno) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d1, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id)
+```
+
+### Model specification
+
+```{r}
+d1_72996 <- d1 %>% 
+  filter(donor == "72996" & !is.na(bd_load))
+
+m1d_72996 <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + bdload_z + elev_z + first_c + male_c + (1 | translocation_id),
+  data = d1_72996,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### Priors & chain diagnostics
+
+```{r}
+prior_summary(m1d_72996)
+
+b1_72996 <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "bdload_z", "elev_z", "first_c0.353", "male_c0.548")
+
+mcmc_trace(m1d_72996, size = 0.1, pars = b1_72996)
+mcmc_dens_overlay(m1d_72996, pars = b1_72996)
+
+summary(m1d_72996)
+```
+
+### Posterior predictive check
+
+```{r}
+pp_check(m1d_72996, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Posterior analysis
+
+```{r}
+tidy(m1d_72996, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1d_72996, area_method = "equal height", prob = 0.90, pars = b1_72996) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1d_72996))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m1d_72996, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+
+-   Using dataset containing only frogs translocated from 72996 had relatively little effect on the predictors of frog survival.
+Specific to the objective of this analysis, frog size remained unimportant as a predictor of survival.
+However, effect of Bd load became marginally unimportant instead of unimportant (i.e., uncertainty interval includes 0 but barely).
+
+## ANALYSIS 3: Assess effect on m1d of including all records, including where bd_load = NA
+
+-   Additional analysis of model m1d from `translocation_survival_analysis.Rmd`.
+
+### Read in frog translocation dataset
+
+-   arm package used to rescale continuous variables (substract mean, divide by 2 SD), manual centering of binary variables.
+
+```{r read-translocation-data-including-na}
+d1_allna <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+#  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
+  mutate(
+    first = if_else(order == 1, 1, 0),
+    surv = if_else(survival < 0.5, 0, 1),
+    male = if_else(sex == "m", 1, 0),
+    elev_lo = if_else(elevation < 3020, 1, 0), # using interval defined by cut_interval, n = 2
+    elev_z = arm::rescale(elevation),
+    snowt_z = arm::rescale(snow_t),
+    snowt1_z = arm::rescale(snow_t1),
+    day_z = arm::rescale(day),
+    length_z = arm::rescale(length),
+    bdload_l = log10(bd_load + 1),
+    bdload_z = arm::rescale(bdload_l),
+    shore_c = shore - mean(shore),
+    male_c = male - mean(male),
+    elevlo_c = elev_lo - mean(elev_lo),
+    first_c = first - mean(first),
+    across(c(shore_c, male_c, elevlo_c, first_c), round, 3), 
+    across(c(site_id, shore_c, first_c, donor, male_c, elevlo_c, year), as.factor),
+    surv = as.integer(surv))
+
+# Add a descriptive translocation_id
+d1_allna <- d1_allna %>% 
+  distinct(site_id, year) %>% 
+  arrange(site_id, year) %>% 
+  group_by(site_id) %>% 
+  mutate(transno = seq_len(n())) %>% 
+  ungroup(site_id) %>% 
+  unite(translocation_id, c("site_id", "transno"), remove = FALSE) %>% 
+  select(-transno) %>% 
+  mutate(translocation_id = as.factor(translocation_id)) %>% 
+  inner_join(d1_allna, by = c("site_id", "year")) %>% 
+  relocate(translocation_id, .after = site_id)
+```
+
+### Model specification
+
+```{r}
+m1d_allna <- stan_glmer(
+  surv ~ length_z + snowt_z + snowt1_z + day_z + elev_z + first_c + male_c + donor + (1 | translocation_id),
+  data = d1_allna,
+  family = binomial,
+  chains = 4, 
+  iter = 5000*2, 
+  cores = 4,
+  seed = 84735)
+```
+
+### List model variable names
+
+```{r}
+get_variables(m1d_allna)
+```
+
+### Priors & chain diagnostics
+
+```{r}
+prior_summary(m1d_allna)
+
+b1_allna <- c("(Intercept)", "length_z", "snowt_z", "snowt1_z", "day_z", "elev_z", "first_c0.349", "male_c0.549", "donor70567", "donor72996")
+
+mcmc_trace(m1d_allna, size = 0.1, pars = b1_allna)
+mcmc_dens_overlay(m1d_allna, pars = b1_allna)
+
+summary(m1d_allna)
+```
+
+### Posterior predictive check
+
+```{r}
+pp_check(m1d_allna, plotfun = "stat") +
+    xlab("Frog survival rate")
+```
+
+### Posterior analysis
+
+```{r}
+tidy(m1d_allna, conf.int = TRUE, conf.level = 0.90)
+
+mcmc_areas(m1d_allna, area_method = "equal height", prob = 0.90, pars = b1_allna) +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Model with standardized predictors & translocation_id as grouping variable",
+          "Posterior distributions with medians & 90% uncertainty intervals")
+
+mean(bayes_R2(m1d_allna))
+
+# Frog survival by translocation_id
+mcmc_areas_ridges(m1d_allna, regex_pars = "translocation_id") +
+  geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +  
+  ggtitle("Frog survival for each translocation_id")
+```
+
+-   Inclusion of all records (i.e., 12 records where bd_load = NA that were excluded in primary analysis) did not change the results.
+
+## ANALYSIS 4: Calculate parameters for viability analysis
 
@@
 
--   Assumptions:
+Assumptions:
 
@@
 
--   Assume that
-
 ```{r}

### 2023-03-10 code/extract_survival_estimates.R
Update relative file path

@@ -25,8 +25,8 @@ for(l in 1:length(lake_ids)){
 
 	lake_number = lake_ids[l]
 	cat("Working on lake", lake_number, "\n")
-	modelpath = file.path("..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
-	datapath = file.path("..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+	modelpath = file.path("..", "..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path("..", "..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
 
 	model = readRDS(modelpath)
 	survey_dat = fread(datapath)

### 2023-03-14 code/translocation_survival_analysis.Rmd
Rearrange load-libraries code chunks
Minor updates to analysis code

@@
 The analyses are conducted in a Bayesian framework using the R package rstanarm.
+
+## Load packages
+
+```{r}
+library(tidyverse)
+library(brms)
+library(rstanarm)
+library(bayesplot)
+library(tidybayes)
+library(loo)
+library(broom.mixed)
+library(janitor)
+library(patchwork)
+library(lubridate)
+
+source("classification_summary.R")
+```
 
@@
 ## ANALYSIS 2: Model of site-level and frog-level predictors
-
-### Load packages
-
-```{r}
-library(tidyverse)
-library(brms)
-library(rstanarm)
-library(bayesplot)
-library(tidybayes)
-library(loo)
-library(broom.mixed)
-library(janitor)
-library(patchwork)
-
-source("classification_summary.R")
-```
 
@@
 cond_effects_plot <- (p_snowt1 | p_elev) / (p_donor | p_sex)
+cond_effects_plot
 ```

### 2023-03-14 code/translocation_survival_analysis_suppl.Rmd
Minor updates to analysis code

@@
 ```{r}
+library(RPostgreSQL)
 library(tidyverse)
+library(lubridate)
 library(brms)
@@
 ### Create dataset
-
-### Load additional libraries
-
-```{r}
-library(RPostgreSQL)
-library(lubridate)
-```
 
@@
 
-### Load packages
-
-```{r}
-library(RPostgreSQL)
-library(lubridate)
-library(dplyr)
-library(magrittr)
-library(readr)
-```
-
 ### Estimate subadult survival
@@
 #### Connect to database
-
--   Use 2nd chunk when connecting via SSH.
 
@@
                 port = Sys.getenv("port"), 
-                dbname = Sys.getenv("dbname"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password")
-               )
-```
-
-```{r}
-con = dbConnect(dbDriver("PostgreSQL"),
-                host = "localhost",
-                port = 15432,
                 dbname = Sys.getenv("dbname"),

### 2023-03-14 code/translocation_survival_analysis.Rmd
Quarto auto-formatting

--- 2023-02-22 code/translocation_survival_analysis.Rmd
@@
 ### Simulate and plot posterior predictive models & observations
+
 #### Reduced model with translocation_id as grouping variable
+
 ```{r}
@@
 
-- Model does a good job of predicting survival based on group-level and individual-level characteristics.  
+-   Model does a good job of predicting survival based on group-level and individual-level characteristics.\
 -   Predicted survival is based on assumption that only males are included. Run another set of models for females?
@@
 #### Reduced model without grouping variable
+
 To explore how addition of grouping variable affected predictions, especially that for 70370 and 70550.
+
 ```{r}
@@
 ### Values of important predictors for translocations of interest
+
 ```{r}
@@
 ### Save plot
+
 ```{r}

### 2023-03-15 code/translocation_survival_analysis_suppl.Rmd
Divide code for dataset creation and analysis

@@
 -   Compare loads at time of translocation to those from CMR years 0 and 1. Exclude years \> 1. Only frogs that were recaptured at least once following translocation are included.
+-   Model with pit_tag_ref as grouping variable is the model of primary interest, roughly analogous to paired t-test. Included other models to allow a better understanding of the data.
 
-### Create dataset
-
-### Read in raw translocation and capture datasets, bind them
+### Read in frog-Bd dataset
 
 ```{r}
-translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv")) %>% 
-  relocate(visit_date = release_date, .before = type) %>% 
-  rename(survey_type = type) %>% 
-  mutate(pit_tag_ref = replace(pit_tag_ref, pit_tag_ref == "900067000117396" & site_id == 70641, "111111111111111")) %>% # remove incorrect pit_tag_ref identified in translocation_survival_createdata.Rmd
-  filter(pit_tag_ref != "111111111111111")
-
-capture <- read_csv(here::here("data", "raw", "frog_captures.csv")) %>% 
-  select(-capture_life_stage, -capture_animal_state, -sex)
-
-# select only translocated frogs (exclude naturally-recruited frogs and translocated frogs that were never recaptured after translocation)
-capture <- translocation %>% 
-  select(pit_tag_ref) %>% 
-  inner_join(capture, by = c("pit_tag_ref")) %>% 
-  relocate(pit_tag_ref, .after = last_col())
-
-frogs_all <- bind_rows(translocation, capture)
+d1 <- read_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv"))
 ```
-
-### Connect to PostgreSQL database `amphibians`
-
-```{r}
-con = dbConnect(dbDriver("PostgreSQL"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password"),
-                host = "frogdb.eri.ucsb.edu",
-                port = 5432,
-                dbname = "amphibians")
-
-```
-
-### Retrieve Bd data for translocated frogs from database
-
-```{sql retrieve-translocation-bd-data, connection=con, output.var="translocate_bd"}
-select
-  release_date as visit_date,
-  pit_tag_ref,
-  swab_id,
-  bd_load
-from relocate
-inner join relocate_frog on relocate.id = relocate_id
-inner join bd_load on swab_id = sample_id  
-where release_siteid1 in (70134, 70279, 70370, 70413, 70414, 70449, 70505, 70550, 70556, 70619, 70628, 70641, 74976) and
-  (replicate = 1 or replicate is null)
-```
-
-### Retrieve Bd data for captured frogs from database
-
-```{sql retrieve-capture-bd-data, connection=con, output.var="capture_bd"}
-select
-	visit_date,
-	pit_tag_ref, 
-	swab_id,
-	bd_load
-from visit
-inner join survey on visit.id = visit_id
-inner join capture_survey on survey.id = survey_id
-inner join bd_load on swab_id = sample_id  
-where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and
-	survey_type = 'cmr' and
-	capture_animal_state != 'dead' and
-	(replicate = 1 or replicate is null)
-```
-
-### Disconnect from database
-
-```{r}
-dbDisconnect(con) 
-rm("con")
-```
-
-### Combine Bd data from translocate and capture datasets
-
-```{r}
-bd_add <- bind_rows(translocate_bd, capture_bd)
-```
-
-### Add Bd data to frog data, restructure dataset
-
-```{r}
-frogsbd <- frogs_all %>% 
-  inner_join(bd_add, by = c("visit_date", "pit_tag_ref")) %>% 
-  mutate(period = if_else(survey_type == "cmr", "after", "before")) %>% # use 0/1 instead?
-  relocate(period, .after = survey_type) %>% 
-  arrange(site_id, visit_date)
-```
-
-### Add years since translocation
-
-```{r}
-frogsbd <- translocation %>% 
-  mutate(year_translocate = year(visit_date)) %>% 
-  select(year_translocate, pit_tag_ref) %>% 
-  inner_join(frogsbd, by = c("pit_tag_ref")) %>% 
-  relocate(year_translocate, .after = visit_date) %>% 
-  relocate(pit_tag_ref, .after = period) %>% 
-  mutate(year_capture = year(visit_date), .after = year_translocate) %>% 
-  mutate(interval = (year_capture - year_translocate), .after = year_capture) %>% 
-  arrange(site_id, visit_date, pit_tag_ref)
-```
-
-### Update site ids to standardize usage
-
-```{r}
-frogsbd <- frogsbd %>% 
-  mutate(site_id1 = site_id, .after = site_id,
-         site_id1 = replace(site_id1, site_id == 70034, 70449),
-         site_id1 = replace(site_id1, site_id == 70175, 70413),
-         site_id1 = replace(site_id1, site_id == 70279, 70413),
-         site_id1 = replace(site_id1, site_id == 72093, 70413),
-         site_id1 = replace(site_id1, site_id == 72442, 70413),
-         site_id1 = replace(site_id1, site_id == 70688, 70134),
-         site_id1 = replace(site_id1, site_id == 72092, 70505),
-         site_id1 = replace(site_id1, site_id == 72973, 74976))
-```
-
-### Add translocation_id
-
-```{r}
-frogsbd <- frogsbd %>% 
-  distinct(site_id1, year_translocate) %>% 
-  arrange(site_id1, year_translocate) %>% 
-  group_by(site_id1) %>% 
-  mutate(transno = seq_len(n())) %>% 
-  ungroup(site_id1) %>% 
-  unite(translocation_id, c("site_id1", "transno"), remove = FALSE) %>% 
-  select(-transno) %>% 
-  inner_join(frogsbd, by = c("site_id1", "year_translocate")) %>% 
-  select(site_id, site_id1, translocation_id, visit_date, survey_type, period, year_translocate, year_capture, interval, pit_tag_ref, swab_id, bd_load) %>% 
-  arrange(site_id1, visit_date)
-```
-
-### Save frog-Bd dataset
-
-```{r}
-frogsbd %>% write_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv"))
-```
-
-### Description of dataset columns
 
@@
 
-### Analyze dataset
-
--   Model with pit_tag_ref grouping variable is the model of primary interest, roughly analogous to paired t-test. Included other models to allow a better understanding of the data.
-
-#### Read in data
-
-```{r}
-d1 <- read_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv")) 
-```
-
-#### Data exploration
+### Data exploration
 
@@
 
-#### Prepare variables for analysis
+### Prepare variables for analysis
 
@@
 
-#### Check for null values
+### Check for null values
 
@@
 
-#### Include only CMR years 0 and 1
+### Include only CMR years 0 and 1
 
@@
 
-#### Plot distribution of bd_load response variable
+### Plot distribution of bd_load response variable
 
@@
 d1_allna <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-#  drop_na(bd_load) %>%  # drop 12 records where bd_load is NA (9 are from 70413)
   mutate(
@@
 -   Inclusion of all records (i.e., 12 records where bd_load = NA that were excluded in primary analysis) did not change the results.
-
-## ANALYSIS 4: Calculate parameters for viability analysis
-
-### Estimate subadult survival
-
-#### Connect to database
-
--   Users without credentials to access database can start with `subadult_survival.csv` saved below.
-
-```{r}
-con = dbConnect(dbDriver("PostgreSQL"),
-                host = Sys.getenv("host"), 
-                port = Sys.getenv("port"), 
-                dbname = Sys.getenv("dbname"),
-                user = rstudioapi::askForPassword("user name"),
-                password = rstudioapi::askForPassword("password")
-               )
-```
-
-#### Retrieve Bd data for translocated frogs from database
-
--   Used 70550 because of its long and relatively intensive survey history.
-
-```{sql, connection=con, output.var="surv"}
-select
-  site_id,
-  visit_date,
-  species,
-  visual_life_stage,
-  visual_animal_state,
-  count
-from visit
-inner join survey on visit_id = visit.id
-inner join visual_survey on survey_id = survey.id  
-where site_id in (70550) and
-  species = 'ramu' and
-  visual_life_stage = 'subadult' and
-  visual_animal_state = 'alive'
-order by site_id, visit_date
-```
-
-#### Disconnect from database
-
-```{r}
-dbDisconnect(con) 
-rm("con")
-```
-
-#### Save raw dataset
-
-```{r}
-surv %>% write_csv(here::here("data", "raw", "subadult_survival.csv"))
-```
-
-#### Calculate maximum annual counts
-
-```{r}
-surv <- surv %>% 
-  group_by(visit_date) %>% 
-  summarize(count_sum = sum(count)) %>% 
-  ungroup() %>% 
-  mutate(year = year(visit_date)) %>% 
-  group_by(year) %>% 
-  summarize(count_max = max(count_sum))
-```
-
-#### Extract number of new recruits from mrmr model
-
-```{r}
-model <- read_rds(here::here("data", "clean", "70550_model.rds"))
-
-get_numeric_indices <- function(string) {
-  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
-  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
-  tibble::as_tibble(idx_mat) %>%
-    dplyr::mutate_all(readr::parse_integer)
-}
-
-survey_prim_periods <- model$data$surveys %>%
-  group_by(.data$primary_period) %>%
-  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
-  ungroup()
-
-p <- model$m_fit$draws("B", format = "draws_df") %>%
-  tidyr::pivot_longer(tidyselect::starts_with("B")) %>%
-  suppressWarnings() %>%
-  mutate(
-    get_numeric_indices(.data$name),
-    primary_period = .data$index_1 + 1
-  ) %>%
-  group_by(.data$primary_period) %>%
-  summarize(lo = quantile(.data$value, .025),
-            med = median(.data$value),
-            hi = quantile(.data$value, .975),
-            .groups = "drop") %>%
-  left_join(survey_prim_periods, by = "primary_period")
-
-recruit <- p %>% 
-  group_by(year) %>% 
-  summarize(recruit_sum = sum(med))
-```
-
-#### Create final dataset
-
-```{r}
-subadult_survival <- surv %>% 
-  inner_join(recruit, by = c("year"))
-```
-
-#### Calculate subadult survival
-
--   Assumes that p(detection) for subadults is 0.25.
-
-```{r}
-subadult_sum <- subadult_survival %>% 
-  filter(between(year, 2008, 2020)) %>% 
-  summarize(sum(count_max))
-
-recruit_sum <- subadult_survival %>% 
-  filter(between(year, 2009, 2021)) %>% 
-  summarize(sum(recruit_sum))
-
-survival_sub <- recruit_sum / (subadult_sum * 4) 
-```
-
-### Estimate 1st-year tadpole survival
-
--   Used 74976 due to its shallow habitats that allow accurate tadpole counts.
-
--   1st-year tadpole counts are from field notebooks.
-
-```{r}
-surv_tad <- tibble(site_id = 74976, year = 2019:2021, count = c(250, 300, 305))
-```
-
-#### Extract adult abundance from mrmr model
-
--   Used minimum survey date per year to estimate number of females during breeding season.
-
-```{r}
-model <- read_rds(here::here("data", "clean", "74976_model.rds"))
-
-get_numeric_indices <- function(string) {
-  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
-  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
-  tibble::as_tibble(idx_mat) %>%
-    dplyr::mutate_all(readr::parse_integer)
-}
-
-survey_prim_periods <- model$data$surveys %>%
-  group_by(.data$primary_period) %>%
-  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
-  ungroup()
-
-p <- model$m_fit$draws("N", format = "draws_df") %>%
-        tidyr::pivot_longer(tidyselect::starts_with("N")) %>%
-        suppressWarnings() %>%
-        mutate(
-          get_numeric_indices(.data$name),
-          primary_period = .data$index_1 + 1
-        ) %>%
-        group_by(.data$primary_period) %>%
-        summarize(lo = quantile(.data$value, .025),
-                  med = median(.data$value),
-                  hi = quantile(.data$value, .975),
-                  .groups = "drop") %>%
-        left_join(survey_prim_periods, by = "primary_period")
-
-abund_ad <- p %>% 
-  group_by(year) %>% 
-  summarize(survey_date = min(survey_date)) %>%
-  select(survey_date) %>% 
-  inner_join(p, by = c("survey_date")) %>% 
-  select(year, med)
-```
-
-#### Create final dataset
-
-```{r}
-surv_tad <- surv_tad %>% 
-  inner_join(abund_ad, by = "year") %>% 
-  rename(adults = med)
-```
-
-#### Calculate tadpole survival
-
-Assumptions:
-
--   p(detection) for 1st-year tadpoles is 0.5.
-
--   \# females is half of total adult abundance
-
--   female fecundity = 100 eggs
-
-```{r}
-surv_tad <- surv_tad %>% 
-  mutate(surv_1styr = (count * 2) / ((adults / 2) * 100)) %>% 
-  summarize(mean(surv_1styr))
-```

### 2023-03-15 code/extract_abundance_trajectories.R
Update cmr-analysis paths to reflect directory change

@@ -5,6 +5,8 @@ library(data.table)
 library(magrittr)
 source("population_viability_functions.R")
 
+cmr_path = file.path("..", "data", "raw", "cmr-analysis")
+
 # Lakes to include
 lake_ids = c(70414, 70370, 70134, 70505, 70550, 70641, 70619, 70449, 70413,
 			 70628, 70556, 74976)
@@ -14,9 +16,8 @@ for(l in 1:length(lake_ids)){
 
 	lake_number = lake_ids[l]
 	cat("Working on lake", lake_number, "\n")
-	cat("INFO: Assuming cmr-analysis folder is stored in the same directory as translocation folder", "\n")
-	modelpath = file.path("..", "..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
-	datapath = file.path("..", "..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+	modelpath = file.path(cmr_path, "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path(cmr_path, "survey", paste0(lake_number, "_survey.csv"))
 
 	model = readRDS(modelpath)
 	abundance_table = abundances_from_model(model, "abundance")

### 2023-03-15 code/extract_survival_estimates.R

@@ -11,6 +11,9 @@ library(data.table)
 library(magrittr)
 source("survival_table_nontranslocated.R")
 
+
+cmr_path = file.path("..", "data", "raw", "cmr-analysis")
+
 # Lakes to include
 lake_ids = c(70414, 70370, 70134, 70505, 70550, 70641, 70619, 70449, 70413,
 			 70628, 70556, 74976)
@@ -25,8 +28,8 @@ for(l in 1:length(lake_ids)){
 
 	lake_number = lake_ids[l]
 	cat("Working on lake", lake_number, "\n")
-	modelpath = file.path("..", "..", "cmr-analysis", "out", "model", paste0(lake_number, "_model.rds"))
-	datapath = file.path("..", "..", "cmr-analysis", "data", "clean", paste0(lake_number, "_survey.csv"))
+	modelpath = file.path(cmr_path, "model", paste0(lake_number, "_model.rds"))
+	datapath = file.path(cmr_path, "survey", paste0(lake_number, "_survey.csv"))
 
 	model = readRDS(modelpath)
 	survey_dat = fread(datapath)

### 2023-03-15 code/extract_translocation_data.R

@@ -11,15 +11,14 @@ library(lubridate)
 
 # The base path of the CMR analysis.  Assumes the working directory is code
 # and that CMR analyses are stored two directories up.
-home_path = file.path("..", "..")
+cmr_path = file.path("..", "data", "raw", "cmr-analysis")
 
 #####################################
 #### Extract survival estimates #####
 #####################################
 
 # Extract survival estimates for all translocated lakes
-table_path = file.path(home_path, "cmr-analysis", "out", "tables")
-table_files = Sys.glob(file.path(table_path, "*survival_cohort.csv"))
+table_files = Sys.glob(file.path(cmr_path, "survival", "*survival_cohort.csv"))
 
 # Loop through each file and load table
 yearly_surv_probs = list()
@@ -163,15 +162,15 @@ keep_lakes = readRDS(file.path("..", "out", "lakes_to_use_in_analysis.rds"))
 all_recruit = list()
 for(lake in keep_lakes){
 
-	file = file.path(home_path, "cmr-analysis", "out", "model", paste0(lake, "_model.rds"))
+	file = file.path(cmr_path, "model", paste0(lake, "_model.rds"))
 	fit = readRDS(file)
 	recruitment = data.table(fit$m_fit$summary(variables="B"))
 	pop_size = data.table(fit$m_fit$summary(variables="N"))
 
 	# Build the stan data so we know when translocations occurred
-	captures = readr::read_csv(file.path(home_path, "cmr-analysis", "data", "clean", paste0(lake, "_capture.csv")))
-	surveys = readr::read_csv(file.path(home_path, "cmr-analysis","data", "clean", paste0(lake, "_survey.csv")))
-	translocations = readr::read_csv(file.path(home_path, "cmr-analysis","data", "clean", paste0(lake, "_translocation.csv")))
+	captures = readr::read_csv(file.path(cmr_path, "capture" paste0(lake, "_capture.csv")))
+	surveys = readr::read_csv(file.path(cmr_path, "survey", paste0(lake, "_survey.csv")))
+	translocations = readr::read_csv(file.path(cmr_path, translocation, paste0(lake, "_translocation.csv")))
 	mrmr_dat = clean_data(captures = captures, surveys = surveys, translocations = translocations)
 
 	# Extract year of primary period

### 2023-03-21 code/maps.qmd

--- 2022-11-23 code/maps.qmd
@@
 ---
-title: "maps"
+title: "Maps for translocation paper"
 format: html
@@
 library(grid)
-library(tidyverse)
-library(RPostgreSQL)
+library(readr)
 ```
 
-## Map of sample sites for exome capture analysis
-
-Run the following 5 chunks only if you have access to the project database.
-For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data" chunk.
-
-### Create dataset of sample collection points
-
-```{r}
-exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
-```
-
-#### Connect to database
-
-```{r}
-source("db_connect.R")
-```
-
-#### Retrieve data
-
-```{sql retrieve-site-xy, connection=con, output.var="exome_sites_db"}
-select
-  id as site_id,
-  utme,
-  utmn
-from site
-where id in (74281, 50731, 50307, 52127, 10220, 10490, 50837, 20169, 72996, 72808, 70567, 54188, 10055, 10593, 50783)
-```
-
-#### Disconnect from database
-
-```{r}
-source("db_disconnect.R")
-```
-
-#### Create points layer
-
-```{r}
-exome_sites_xy <- 
-  exome_sites %>% 
-  inner_join(exome_sites_db, by = "site_id")
-
-# Add x-y random noise to obscure exact site locations
-exome_sites_xy <- exome_sites_xy %>% 
-  add_column(x = sample(-2000:2000, size = 15),
-             y = sample(-2000:2000, size = 15))
-
-exome_sites_xy <- exome_sites_xy %>% 
-  mutate(utme_noise = utme + x,
-         utmn_noise = utmn + y) %>% 
-  select(-utme, -utmn, -x, -y)
-
-# Save file 
-write_csv(exome_sites_xy, here::here("data", "maps", "exome_sites_xy.csv"))
-```
+## Map of exome capture sample sites
 
@@
 
-```{r import-map-data}
+```{r import-map-data1}
 # Read file, if lacking database access
-exome_sites_xy <- read_csv(here::here("data", "maps", "exome_sites_xy.csv"))
-exome_points = st_as_sf(exome_sites_xy, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
+exome_points <- read_csv(here::here("data", "maps", "exome_points.csv"))
+exome_points <- st_as_sf(exome_points, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
 
@@
 
-#### To Do
+### To Do
 
--   Exported CA tif has no whitespace around state, has black border. How to remove border and add whitespace?
+-   Exported California hillshade tif has no whitespace around edges, has black border. How to remove border and add whitespace?
+
+## Map of Yosemite translocation & donor sites
+
+### Import map layers
+
+```{r import-map-data2}
+translocation_points <- read_csv(here::here("data", "maps", "translocation_points.csv")) %>% 
+  mutate(donor_id = as.factor(donor_id))
+
+translocation_points <- st_as_sf(translocation_points, coords = c("utme_noise", "utmn_noise"), crs = st_crs(26911))
+
+yose_boundary <- st_read(here::here(
+  "data", "maps", "nps-boundaries", "yose_boundary.shp"))
+st_crs(yose_boundary) # determine coordinate reference system (CRS)
+yose_boundary <- st_transform(yose_boundary, 26911) # transform to NAD83 UTM11
+
+# California Color Hillshade (90m & 300m) from koordinates.com
+hillshade_90m <- read_stars(here::here(
+  "data", "maps", "hillshade-90m", "california-color-hillshade-90m.tif"))
+st_crs(hillshade_90m)
+
+hillshade_300m <- read_stars(here::here(
+  "data", "maps", "hillshade-300m", "california-color-hillshade-300m.tif"))
+```
+
+### Check & fix shapefile polygons
+
+```{r}
+st_is_valid(yose_boundary) # invalid polygon
+
+yose_boundary <- st_make_valid(yose_boundary$geometry)
+
+st_is_valid(yose_boundary) # confirm that polygon is valid
+```
+
+### Crop raster with bounding box
+
+```{r}
+# For printing on screen: xmin = 244000, ymax = 4242500, xmax = 309000, ymin = 4142000
+bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 308000, ymin = 4151000), 
+                crs = st_crs(yose_boundary)) %>%  
+  st_as_sfc()
+hillshade_yose_crop <- st_crop(hillshade_90m, bbox)
+hillshade_yose_crop <- st_transform(hillshade_yose_crop, 26911)
+```
+
+### Create study area map
+
+```{r}
+translocation_points_map <- tm_shape(hillshade_yose_crop) + 
+  tm_rgb(r = 1, g = 2, b = 3) +
+  tm_shape(yose_boundary) + 
+  tm_borders(col = "black", lwd = 2) +
+  tm_shape(translocation_points) + 
+  tm_symbols(size = 0.7, 
+             shape = "donor_id",
+             col = "type", 
+             legend.shape.show = FALSE, 
+             palette = c("#0000FF", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
+             title.col = "Population type") +
+  tm_layout(legend.position = c("right", "top"),
+            legend.bg.color = "white",
+            legend.title.size = 1.3,
+            legend.text.size = 1.0) +
+  tm_compass(type = "4star",
+             position = c(0.80, 0.11),
+             size = 3) +
+  tm_scale_bar(breaks = c(0, 10, 20), 
+               text.size = 1.0, 
+               position = c("right", "bottom"))
+
+translocation_points_map
+```
+
+### Create inset map of California
+
+```{r}
+inset_map <- tm_shape(hillshade_300m) +
+  tm_rgb(r = 1, g = 2, b = 3) +
+  tm_shape(yose_boundary) +
+  tm_borders(col = "black")
+inset_map
+```
+
+### Study area map with inset map
+
+```{r}
+translocation_points_map
+print(inset_map, vp = viewport(0.34, 0.20, width = 0.2, height = 0.2))
+```
+
+### Save map
+
+```{r}
+tmap_save(translocation_points_map, 
+          insets_tm = inset_map,
+          insets_vp = viewport(0.17, 0.15, width = 0.25, height = 0.25),
+          here::here("doc", "manuscript", "figures", "map_translocation_points.png"))
+```

### 2023-03-21 code/translocation_survival_createdata_supp.Rmd

---
title: "Create datasets for supplemental analyses"
author: "Roland Knapp"
date: "`r Sys.Date()`"
output: html_notebook
---

## Load packages

```{r load-packages}
library(RPostgreSQL)
library(dplyr) 
library(readr)
library(tidyr)
library(lubridate)
```

-   Users without credentials to access the database can utilize .csv files saved below when running code in the `translocation_survival_supp.Rmd` and `maps.qmd` files.

## DATASET 1: Bd load before versus after translocation

-   Loads at time of translocation versus from CMR years 0 and 1. Exclude years \> 1. Only frogs that were recaptured at least once following translocation are included.

### Read in raw translocation and capture datasets, bind them

```{r}
translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv")) %>% 
  relocate(visit_date = release_date, .before = type) %>% 
  rename(survey_type = type) %>% 
  mutate(pit_tag_ref = replace(pit_tag_ref, pit_tag_ref == "900067000117396" & site_id == 70641, "111111111111111")) %>% # remove incorrect pit_tag_ref identified in translocation_survival_createdata.Rmd
  filter(pit_tag_ref != "111111111111111")

capture <- read_csv(here::here("data", "raw", "frog_captures.csv")) %>% 
  select(-capture_life_stage, -capture_animal_state, -sex)

# select only translocated frogs (exclude naturally-recruited frogs and translocated frogs that were never recaptured after translocation)
capture <- translocation %>% 
  select(pit_tag_ref) %>% 
  inner_join(capture, by = c("pit_tag_ref")) %>% 
  relocate(pit_tag_ref, .after = last_col())

frogs_all <- bind_rows(translocation, capture)
```

### Connect to PostgreSQL database `amphibians`

```{r}
source("db_connect.R")
```

### Retrieve Bd data for translocated frogs

```{sql retrieve-translocation-bd-data, connection=con, output.var="translocate_bd"}
select
  release_date as visit_date,
  pit_tag_ref,
  swab_id,
  bd_load
from relocate
inner join relocate_frog on relocate.id = relocate_id
inner join bd_load on swab_id = sample_id  
where release_siteid1 in (70134, 70279, 70370, 70413, 70414, 70449, 70505, 70550, 70556, 70619, 70628, 70641, 74976) and
  (replicate = 1 or replicate is null)
```

### Retrieve Bd data for captured frogs

```{sql retrieve-capture-bd-data, connection=con, output.var="capture_bd"}
select
	visit_date,
	pit_tag_ref, 
	swab_id,
	bd_load
from visit
inner join survey on visit.id = visit_id
inner join capture_survey on survey.id = survey_id
inner join bd_load on swab_id = sample_id  
where site_id in (70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973) and 
	survey_type = 'cmr' and
	capture_animal_state != 'dead' and
	(replicate = 1 or replicate is null)
```

### Disconnect from database

```{r}
source("db_disconnect.R")
```

### Combine Bd data from translocate and capture datasets

```{r}
bd_add <- bind_rows(translocate_bd, capture_bd)
```

### Add Bd data to frog data, restructure dataset

```{r}
frogsbd <- frogs_all %>% 
  inner_join(bd_add, by = c("visit_date", "pit_tag_ref")) %>% 
  mutate(period = if_else(survey_type == "cmr", "after", "before")) %>% # use 0/1 instead?
  relocate(period, .after = survey_type) %>% 
  arrange(site_id, visit_date)
```

### Add years since translocation

```{r}
frogsbd <- translocation %>% 
  mutate(year_translocate = year(visit_date)) %>% 
  select(year_translocate, pit_tag_ref) %>% 
  inner_join(frogsbd, by = c("pit_tag_ref")) %>% 
  relocate(year_translocate, .after = visit_date) %>% 
  relocate(pit_tag_ref, .after = period) %>% 
  mutate(year_capture = year(visit_date), .after = year_translocate) %>% 
  mutate(interval = (year_capture - year_translocate), .after = year_capture) %>% 
  arrange(site_id, visit_date, pit_tag_ref)
```

### Update site ids to standardize usage

```{r}
frogsbd <- frogsbd %>% 
  mutate(site_id1 = site_id, .after = site_id,
         site_id1 = replace(site_id1, site_id == 70034, 70449),
         site_id1 = replace(site_id1, site_id == 70175, 70413),
         site_id1 = replace(site_id1, site_id == 70279, 70413),
         site_id1 = replace(site_id1, site_id == 72093, 70413),
         site_id1 = replace(site_id1, site_id == 72442, 70413),
         site_id1 = replace(site_id1, site_id == 70688, 70134),
         site_id1 = replace(site_id1, site_id == 72092, 70505),
         site_id1 = replace(site_id1, site_id == 72973, 74976))
```

### Add translocation_id

```{r}
frogsbd <- frogsbd %>% 
  distinct(site_id1, year_translocate) %>% 
  arrange(site_id1, year_translocate) %>% 
  group_by(site_id1) %>% 
  mutate(transno = seq_len(n())) %>% 
  ungroup(site_id1) %>% 
  unite(translocation_id, c("site_id1", "transno"), remove = FALSE) %>% 
  select(-transno) %>% 
  inner_join(frogsbd, by = c("site_id1", "year_translocate")) %>% 
  select(site_id, site_id1, translocation_id, visit_date, survey_type, period, year_translocate, year_capture, interval, pit_tag_ref, swab_id, bd_load) %>% 
  arrange(site_id1, visit_date)
```

### Save frog-Bd dataset

```{r}
frogsbd %>% write_csv(here::here("data", "clean", "bd_beforeafter_translocation.csv"))
```

### Description of dataset columns

-   site_id: Unique 5-digit site identifier.
-   site_id1: Assigned 5-digit site identifier for a collection of adjacent sites that are utilized by a frog population.
-   translocation_id: Unique identifier for each translocation.
-   visit_date: Date on which translocation or survey was conducted.
-   survey_type: "translocation" or "cmr".
-   period: "before" or "after" translocation.
-   year_translocate: Year in which translocation was conducted.
-   year_capture: Year in which frog was captured.
-   interval: Number of years between translocation and frog capture.
-   pit_tag_ref: Unique PIT tag identifier.
-   swab_id: Unique identifier for skin swab sample.
-   bd_load: Bd infection intensity determined from skin swab samples, units are number of ITS1 copies \* 80.

## DATASETS 2: Create datasets for use in estimating selected parameters for viability analysis

### Estimate subadult survival
-   Used 70550 because of its long and relatively intensive survey history.

#### Connect to database

```{r}
source("db_connect.R")
```

#### Retrieve subadult count data

```{sql, connection=con, output.var="surv"}
select
  site_id,
  visit_date,
  species,
  visual_life_stage,
  visual_animal_state,
  count
from visit
inner join survey on visit_id = visit.id
inner join visual_survey on survey_id = survey.id  
where site_id in (70550) and
  species = 'ramu' and
  visual_life_stage = 'subadult' and
  visual_animal_state = 'alive'
order by site_id, visit_date
```

#### Disconnect from database

```{r}
source("db_disconnect.R")
```

#### Save raw dataset

```{r}
surv %>% write_csv(here::here("data", "raw", "subadult_survival.csv"))
```

#### Calculate maximum annual counts

```{r}
surv <- surv %>% 
  group_by(visit_date) %>% 
  summarize(count_sum = sum(count)) %>% 
  ungroup() %>% 
  mutate(year = year(visit_date)) %>% 
  group_by(year) %>% 
  summarize(count_max = max(count_sum))
```

#### Extract number of new recruits from mrmr model

```{r}
model <- read_rds(here::here("data", "clean", "70550_model.rds"))

get_numeric_indices <- function(string) {
  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
  tibble::as_tibble(idx_mat) %>%
    dplyr::mutate_all(readr::parse_integer)
}

survey_prim_periods <- model$data$surveys %>%
  group_by(.data$primary_period) %>%
  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
  ungroup()

p <- model$m_fit$draws("B", format = "draws_df") %>%
  tidyr::pivot_longer(tidyselect::starts_with("B")) %>%
  suppressWarnings() %>%
  mutate(
    get_numeric_indices(.data$name),
    primary_period = .data$index_1 + 1
  ) %>%
  group_by(.data$primary_period) %>%
  summarize(lo = quantile(.data$value, .025),
            med = median(.data$value),
            hi = quantile(.data$value, .975),
            .groups = "drop") %>%
  left_join(survey_prim_periods, by = "primary_period")

recruit <- p %>% 
  group_by(year) %>% 
  summarize(recruit_sum = sum(med))
```

#### Create final dataset

```{r}
subadult_survival <- surv %>% 
  inner_join(recruit, by = c("year"))

subadult_survival %>% write_csv(here::here("data", "clean", "subadult_survival.csv"))
```

#### Calculate subadult survival

-   Assumes that p(detection) for subadults is 0.25.

```{r}
subadult_sum <- subadult_survival %>% 
  filter(between(year, 2008, 2020)) %>% 
  summarize(sum(count_max))

recruit_sum <- subadult_survival %>% 
  filter(between(year, 2009, 2021)) %>% 
  summarize(sum(recruit_sum))

surv_sub <- recruit_sum / (subadult_sum * 4) 
```

### Estimate 1st-year tadpole survival

-   Used 74976 due to its shallow habitats that allow accurate tadpole counts.

-   1st-year tadpole counts are from Knapp field notebooks - VES counts in database do not distinguish between tadpole age-classes.

```{r}
surv_tad <- tibble(site_id = 74976, year = 2019:2021, count = c(250, 300, 305))
```

#### Extract adult abundance from mrmr model

-   Used minimum survey date per year to estimate number of females during breeding season.

```{r}
model <- read_rds(here::here("data", "clean", "74976_model.rds"))

get_numeric_indices <- function(string) {
  idx_mat <- stringr::str_extract_all(string, "[0-9]+", simplify = TRUE)
  colnames(idx_mat) <- paste0("index_", seq_len(ncol(idx_mat)))
  tibble::as_tibble(idx_mat) %>%
    dplyr::mutate_all(readr::parse_integer)
}

survey_prim_periods <- model$data$surveys %>%
  group_by(.data$primary_period) %>%
  filter(.data$secondary_period == min(.data$secondary_period)) %>% 
  ungroup()

p <- model$m_fit$draws("N", format = "draws_df") %>%
        tidyr::pivot_longer(tidyselect::starts_with("N")) %>%
        suppressWarnings() %>%
        mutate(
          get_numeric_indices(.data$name),
          primary_period = .data$index_1 + 1
        ) %>%
        group_by(.data$primary_period) %>%
        summarize(lo = quantile(.data$value, .025),
                  med = median(.data$value),
                  hi = quantile(.data$value, .975),
                  .groups = "drop") %>%
        left_join(survey_prim_periods, by = "primary_period")

abund_ad <- p %>% 
  group_by(year) %>% 
  summarize(survey_date = min(survey_date)) %>%
  select(survey_date) %>% 
  inner_join(p, by = c("survey_date")) %>% 
  select(year, med)
```

#### Create final dataset

```{r}
surv_tad <- surv_tad %>% 
  inner_join(abund_ad, by = "year") %>% 
  rename(adults = med)

surv_tad %>% write_csv(here::here("data", "clean", "tadpole_survival.csv"))
```

#### Calculate tadpole survival

Assumptions:

-   p(detection) for 1st-year tadpoles is 0.5.

-   \# females is half of total adult abundance

-   female fecundity = 100 eggs

```{r}
surv_tad <- surv_tad %>% 
  mutate(surv_1styr = (count * 2) / ((adults / 2) * 100)) %>% 
  summarize(mean(surv_1styr))
```

## DATASETS 3: Create datasets for study site maps

### Map of sample sites for exome capture analysis

#### Create dataset of sample collection points

```{r}
exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
```

#### Retrieve data from database & create points layer

```{r}
source("db_connect.R")

n <- nrow(exome_sites)

exome_points <- dbReadTable(con, c("site")) %>% 
  as_tibble() %>% 
  rename(site_id = id) %>% 
  select(site_id, utme, utmn) %>% 
  inner_join(exome_sites, by = "site_id") %>% 
  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
             y = sample(-2000:2000, size = n)) %>% 
  mutate(utme_noise = utme + x,
         utmn_noise = utmn + y) %>% 
  select(-utme, -utmn, -x, -y)

source("db_disconnect.R")

write_csv(exome_points, here::here("data", "maps", "exome_points.csv"))
```

### Map of Yosemite translocation & donor sites

#### Create dataset of site points

```{r}
donor_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
  distinct(donor) %>% 
  mutate(site_id = donor, .before = donor) %>% 
  rename(donor_id = donor) %>% 
  add_column(type = "donor")

translocation_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
  distinct(site_id, donor) %>% 
  rename(donor_id = donor) %>% 
  add_column(type = "translocation") %>% 
  bind_rows(donor_sites)

rm(donor_sites)
```

#### Retrieve data from database & create points layer

```{r}
source("db_connect.R")

n <- nrow(translocation_sites)

translocation_points <- dbReadTable(con, c("site")) %>% 
  as_tibble() %>% 
  rename(site_id = id) %>% 
  select(site_id, utme, utmn) %>% 
  inner_join(translocation_sites, by = "site_id") %>% 
  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
             y = sample(-2000:2000, size = n)) %>% 
  mutate(utme_noise = utme + x,
         utmn_noise = utmn + y) %>% 
  select(-utme, -utmn, -x, -y)

source("db_disconnect.R")

write_csv(translocation_points, here::here("data", "maps", "translocation_points.csv"))
```

### 2023-03-21 code/maps.qmd
Revert to commit 04afdf914d146eb948b43412668012bbc116cd82.

@@
 library(grid)
-library(readr)
+library(tidyverse)
+library(RPostgreSQL)
 ```
 
-## Map of exome capture sample sites
+## Map of sample sites for exome capture analysis
+
+Run the following 4 chunks only if you have access to the project database.
+For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data1" chunk.
+
+### Create dataset of sample collection points
+
+```{r}
+exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
+```
+
+#### Connect to database
+
+```{r}
+source("db_connect.R")
+```
+
+#### Retrieve data & create points layer
+
+```{r}
+n <- nrow(exome_sites)
+
+exome_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(exome_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+
+write_csv(exome_points, here::here("data", "maps", "exome_points.csv"))
+```
+
+#### Disconnect from database
+
+```{r}
+source("db_disconnect.R")
+```
 
@@
 
+Run the following 4 chunks only if you have access to the project database.
+For those without the necessary credentials, use the saved output file in data/maps/ that is available via the "import-map-data2" chunk.
+
+### Create dataset of site points
+
+```{r}
+donor_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(donor) %>% 
+  mutate(site_id = donor, .before = donor) %>% 
+  rename(donor_id = donor) %>% 
+  add_column(type = "donor")
+
+translocation_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(site_id, donor) %>% 
+  rename(donor_id = donor) %>% 
+  add_column(type = "translocation") %>% 
+  bind_rows(donor_sites)
+
+rm(donor_sites)
+```
+
+#### Connect to database
+
+```{r}
+source("db_connect.R")
+```
+
+#### Retrieve data & create points layer
+
+```{r}
+n <- nrow(translocation_sites)
+
+translocation_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(translocation_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+
+write_csv(translocation_points, here::here("data", "maps", "translocation_points.csv"))
+```
+
+#### Disconnect from database
+
+```{r}
+source("db_disconnect.R")
+```
+
 ### Import map layers

### 2023-03-21 code/translocation_survival_createdata_supp.Rmd
Revert to commit 04afdf914d146eb948b43412668012bbc116cd82.

@@
 
--   Users without credentials to access the database can utilize .csv files saved below when running code in the `translocation_survival_supp.Rmd` and `maps.qmd` files.
+-   Users without credentials to access database can utilize `subadult_survival.csv` and `tadpole_survival.csv` saved below.
 
@@
 ```
-
-## DATASETS 3: Create datasets for study site maps
-
-### Map of sample sites for exome capture analysis
-
-#### Create dataset of sample collection points
-
-```{r}
-exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
-```
-
-#### Retrieve data from database & create points layer
-
-```{r}
-source("db_connect.R")
-
-n <- nrow(exome_sites)
-
-exome_points <- dbReadTable(con, c("site")) %>% 
-  as_tibble() %>% 
-  rename(site_id = id) %>% 
-  select(site_id, utme, utmn) %>% 
-  inner_join(exome_sites, by = "site_id") %>% 
-  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
-             y = sample(-2000:2000, size = n)) %>% 
-  mutate(utme_noise = utme + x,
-         utmn_noise = utmn + y) %>% 
-  select(-utme, -utmn, -x, -y)
-
-source("db_disconnect.R")
-
-write_csv(exome_points, here::here("data", "maps", "exome_points.csv"))
-```
-
-### Map of Yosemite translocation & donor sites
-
-#### Create dataset of site points
-
-```{r}
-donor_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  distinct(donor) %>% 
-  mutate(site_id = donor, .before = donor) %>% 
-  rename(donor_id = donor) %>% 
-  add_column(type = "donor")
-
-translocation_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  distinct(site_id, donor) %>% 
-  rename(donor_id = donor) %>% 
-  add_column(type = "translocation") %>% 
-  bind_rows(donor_sites)
-
-rm(donor_sites)
-```
-
-#### Retrieve data from database & create points layer
-
-```{r}
-source("db_connect.R")
-
-n <- nrow(translocation_sites)
-
-translocation_points <- dbReadTable(con, c("site")) %>% 
-  as_tibble() %>% 
-  rename(site_id = id) %>% 
-  select(site_id, utme, utmn) %>% 
-  inner_join(translocation_sites, by = "site_id") %>% 
-  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
-             y = sample(-2000:2000, size = n)) %>% 
-  mutate(utme_noise = utme + x,
-         utmn_noise = utmn + y) %>% 
-  select(-utme, -utmn, -x, -y)
-
-source("db_disconnect.R")
-
-write_csv(translocation_points, here::here("data", "maps", "translocation_points.csv"))
-```

### 2023-03-22 code/create_R_scripts.R
Initial commit

@@ -1,0 +1,4 @@
+# Create scripts necessary for translocation analyses
+
+knitr::purl("code/translocation_survival_analysis.Rmd")
+file.rename("translocation_survival_analysis.R", "code/translocation_survival_analysis.R")

### 2023-03-23 code/create_R_scripts.R
Add rules to makefile

@@ -2,3 +2,9 @@
 
 knitr::purl("code/translocation_survival_analysis.Rmd")
 file.rename("translocation_survival_analysis.R", "code/translocation_survival_analysis.R")
+
+knitr::purl("code/translocation_survival_analysis_supp.Rmd")
+file.rename("translocation_survival_analysis_supp.R", "code/translocation_survival_analysis_supp.R")
+
+knitr::purl("code/maps.qmd")
+file.rename("maps.R", "code/maps.R")

### 2023-03-23 code/translocation_survival_createdata_supp.Rmd
Move dataset creation code chunks out of maps.qmd

@@
 library(readr)
+library(tibble)
 library(tidyr)
@@
 ```
+
+## DATASETS 3: Create datasets for maps
+
+### Dataset for map of sample sites for exome capture analysis
+
+Users without the necessary credentials to access the database can access the final dataset at data/maps/exome_points.csv.
+
+```{r}
+exome_sites <- read_csv(here::here("data", "maps", "exome_sample_sites.csv"))
+n <- nrow(exome_sites)
+
+source("db_connect.R")
+exome_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(exome_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+source("db_disconnect.R")
+
+write_csv(exome_points, here::here("data", "maps", "exome_points.csv"))
+```
+
+## Dataset for map of Yosemite translocation & donor sites
+
+Users without the necessary credentials to access the database can access the final dataset at data/maps/translocation_points.csv.
+
+### Retrieve data & create points layer
+
+```{r}
+donor_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(donor) %>% 
+  mutate(site_id = donor, .before = donor) %>% 
+  rename(donor_id = donor) %>% 
+  add_column(type = "donor")
+
+translocation_sites <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
+  distinct(site_id, donor) %>% 
+  rename(donor_id = donor) %>% 
+  add_column(type = "translocation") %>% 
+  bind_rows(donor_sites)
+
+n <- nrow(translocation_sites)
+
+rm(donor_sites)
+
+source("db_connect.R")
+translocation_points <- dbReadTable(con, c("site")) %>% 
+  as_tibble() %>% 
+  rename(site_id = id) %>% 
+  select(site_id, utme, utmn) %>% 
+  inner_join(translocation_sites, by = "site_id") %>% 
+  add_column(x = sample(-2000:2000, size = n), # Add x-y random noise to obscure exact site locations
+             y = sample(-2000:2000, size = n)) %>% 
+  mutate(utme_noise = utme + x,
+         utmn_noise = utmn + y) %>% 
+  select(-utme, -utmn, -x, -y)
+source("db_disconnect.R")
+
+write_csv(translocation_points, here::here("data", "maps", "translocation_points.csv"))
+```

### 2023-03-23 code/translocation_survival_analysis.Rmd
Update file path for sourced script
Revert "Update file path for sourced script"
This reverts commit 03a94dfefffa460c0f93a3dd46880bf9252ab56a.
Update ggplot code to replace deprecated aesthetic

--- 2023-03-14 code/translocation_survival_analysis.Rmd
@@
 
-source("classification_summary.R")
+source("code/classification_summary.R")
 ```
@@
       theme(
-        legend.background = element_rect(color = "black", size = 0.3),
+        legend.background = element_rect(color = "black", linewidth = 0.3),
         legend.position = c(0.2, 0.8),
@@
 
-#### Load packages
-
-```{r}
-library(tidyverse)
-library(brms)
-library(rstanarm)
-library(bayesplot)
-library(tidybayes)
-library(loo)
-library(broom.mixed)
-library(janitor)
-
-source("classification_summary.R")
-```
-
-#### Create frog translocation dataset for analysis
+#### Frog translocation dataset for analysis
 
@@
   ggplot(aes(x = elevation, y = surv)) +  #   "ggplot(aes(x = (elev_z * 2*sd_elev) + mean_elev, y = surv)) +" gives identical result
-          geom_line(aes(y = .epred, group = .draw), color = "blue", size = 0.2, alpha = 1) +
+          geom_line(aes(y = .epred, group = .draw), color = "blue", linewidth = 0.2, alpha = 1) +
         # geom_point(aes(y = .epred)) +
@@
           geom_ribbon(aes(ymin = lower, ymax = upper), fill= "lightgray", alpha = 1) +
-          geom_line(aes(y = median), color = "blue", size = 1, alpha = 1) +
+          geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
           labs(x = "Elevation of recipient site (meters)", y = "P(1-year frog survival)") +
@@
           geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightgray", alpha = 1) +
-          geom_line(aes(y = median), color = "blue", size = 1, alpha = 1) +
+          geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
           labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year frog survival)") +
@@
   ggplot(aes(x = elevation, y = surv, color = sex)) +
-          geom_line(aes(y = .epred, group = paste(sex, .draw)), size = 0.4, alpha = 0.5) +
+          geom_line(aes(y = .epred, group = paste(sex, .draw)), linewidth = 0.4, alpha = 0.5) +
           labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
@@
           geom_ribbon(aes(ymin = lower, ymax = upper, fill = sex), alpha = 0.2, linetype = 0) +
-          geom_line(aes(y = median, color = sex), size = 1, alpha = 1) +
+          geom_line(aes(y = median, color = sex), linewidth = 1, alpha = 1) +
           labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +

### 2023-03-24 code/translocation_survival_analysis_supp.Rmd
Edit file path for sourced script
Replace file with edited version from other branch

--- 2023-03-15 code/translocation_survival_analysis_suppl.Rmd
@@
 ```{r}
-library(RPostgreSQL)
 library(tidyverse)
-library(lubridate)
-library(brms)
 library(rstanarm)
@@
 library(janitor)
-
-source("classification_summary.R")
 ```

### 2023-03-29 code/install_R_packages.R
Add tmap as dependency

@@ -2,7 +2,7 @@
 
 list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
 					  "rstan", "lubridate", "tidyverse", 
-					  "coda","mvtnorm","devtools","loo","dagitty")
+					  "coda","mvtnorm","devtools","loo","dagitty", "tmap")
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
 

### 2023-03-29 code/install_R_packages.R
Add additional packages for makefile

@@ -2,7 +2,8 @@
 
 list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
 					  "rstan", "lubridate", "tidyverse", 
-					  "coda","mvtnorm","devtools","loo","dagitty", "tmap")
+					  "coda","mvtnorm","devtools","loo","dagitty", "tmap",
+					  "RPostgreSQL", "DBI", "here")
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
 

### 2023-03-29 code/translocation_survival_exploratory.Rmd

--- 2022-12-07 code/translocation_survival_exploratory.Rmd
@@
 ---
-title: "Predictors of frog survival following translocations - analyses"
+title: "Predictors of frog survival following translocations - exploratory analyses"
 author: "Roland Knapp"
@@
 
-This notebook contains exploratory analyses to aid in the design of the final analyses of factors influencing post-translocation survival (conducted in translocation_survival_analysis.Rmd).
+This notebook contains exploratory analyses to aid in the design of the final analyses of factors influencing post-translocation survival (in translocation_survival_analysis.Rmd and translocation_suvival_analysis_supp.Rmd).
 
@@
 
-```{r read-translocation-data}
+```{r}
 frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
@@
 
-```{r check-null}
+```{r}
 frog_translocation %>% summarize(across(everything(), ~ sum(is.na(.))))
 ```
-* 12 records where bd_load is NA, no other NAs in dataset
+
+-   12 records where bd_load is NA, no other NAs in dataset
 
@@
 
-```{r retrieve-frog-survival-data}
+```{r}
 files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survival"), glob = "*_cohort.csv")
@@
 
-# define color palette (palette_OkabeIto_black)
-```{r define-color-palette}
+## Define color palette (palette_OkabeIto_black)
+
+```{r}
 palette_cbb <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@
 ## Pairs plot to evaluate between-variable correlations
-```{r pairs-plots}
+
+```{r}
 # Translocation-specific variables
@@
 ```
-* Elevations for shore = 0 and shore = 1 do not overlap at all, so there is no ability to disentangle these two variables.
-Include only elevation in model.
-* snow_t and day are positively correlated, reflecting the fact that in summers following severe winters, tranlocations were conducted later in the summer.
+
+-   Elevations for shore = 0 and shore = 1 do not overlap at all, so there is no ability to disentangle these two variables. Include only elevation in model.
+-   snow_t and day are positively correlated, reflecting the fact that in summers following severe winters, translocations were conducted later in the summer.
 Correlation is relatively weak so may not cause issues in analysis.
-* Few translocations with order > 2, combine group = 2-4 into a single group. 
-* Frogs from the 72996 donor population were smaller and had somewhat higher bd loads than frogs from the other two donor populations.  
-* Frog length and weight were highly correlated, as expected.
+-   Few translocations with order \> 2, combine group = 2-4 into a single group.
+-   Frogs from the 72996 donor population were smaller and had somewhat higher Bd loads than frogs from the other two donor populations.
+-   Frog length and weight were highly correlated, as expected.
 
 ## Association between site_id grouping variable and some predictors
-* If site_id is closely associated with a predictor, will adding site_id as a group-level effect largely account for the effect of the predictor?
-```{r siteid-predictor-count}
+
+-   If site_id is closely associated with a predictor, will adding site_id as a group-level effect largely account for the effect of the predictor?
+
+```{r}
 frog_translocation %>% 
@@
 ```
-* Only a single value of elevation per site_id (obviously), so could bin elevation (low elevation, high elevation) to weaken that association.  
-* Multiple values of winter severity and order per site_id, so those two variables are only loosely associated with site_id.
 
-```{r siteid-elevation-binned}
+-   Only a single value of elevation per site_id (obviously), so could bin elevation (low elevation, high elevation) to weaken that association.
+-   Multiple values of winter severity and order per site_id, so those two variables are only loosely associated with site_id.
+
+### Create binary elevation variable
+
+```{r}
 frog_translocation %>% 
@@
 
-## Plot of frog survival for each translocation - grouped by site and arranged by site-specific median survival
-
--  Points colored by donor site.
-
-```{r frog-survival-by-site}
-palette1 <- c("green4", "blue", "black")
-
-donors <- frog_translocation %>% 
-  distinct(site_id, donor)
-
-cohort_survival_plot <- cohort_survival %>% 
-  mutate(release_year = year(release_date)) %>% 
-  group_by(site_id) %>% 
-  summarize(plot_survival = mean(median_survival)) %>%  # create variable by which to order sites for plotting
-  arrange(plot_survival) %>% 
-  mutate(siteid_order = row_number()) %>% 
-  inner_join(cohort_survival, by = "site_id") %>% 
-  inner_join(donors, by = "site_id") %>% 
-  select(-plot_survival) %>% 
-    ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, group = donor)) +  # reorder site_id by siteid_order)
-      geom_point(aes(color = donor), size = 3) +
-      scale_color_manual(values = palette1) + 
-      labs(x = "Recipient site", y = "1-year frog survival") +
-      theme_classic() +
-      theme(
-        legend.background = element_rect(color = "black", size = 0.3),
-        legend.position = c(0.2, 0.8),
-        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
-
-cohort_survival_plot
-```
-* Strong effect of site/translocation on frog survival.
-* Survival is highly variable across sites, but highly repeatable within a site. 
-* Indicates need to include site_id and/or translocation_id as group-level effect(s).
-
-### Save plot
+## Relationship between bd_load at release and 1-year survival
 
 ```{r}
-ggsave(here::here("doc", "manuscript", "figures", "translocation_survival_bysiteid.png"), width = 5, height = 5) 
-```
-
-## Plot of frog survival for each translocation - grouped by elevation
-
-```{r frog-survival-by-elevation}
-frog_translocation %>%
-  distinct(site_id, date, elevation) %>% 
-  inner_join(cohort_survival, by = c("site_id", "date" = "release_date")) %>% 
-    ggplot(aes(x = elevation, y = median_survival)) + 
-      geom_vline(aes(xintercept = elevation), linetype = "dotted", color = "darkgray") +
-      geom_point(size = 3) +
-      geom_text(aes(label = site_id, x = elevation, y = 1.05), angle = 90) +
-      coord_cartesian(ylim = c(0, 1.08)) +
-      labs(x = "Elevation of recipient site", y = "1-year frog survival")
-```
-Elevation has a clear effect on frog survival, but there is a substantial amount of variation that remains unaccounted for.
-
-## Relationship between bd_load at release and 1-year survival
-```{r bdload-frogsurvival-plot}
 # Plot distribution of bd_load
@@
 ```
-* Suggests that frogs that had the highest loads at the time of translocation had relatively low survival.
+
+-   Suggests that frogs that had the highest loads at the time of translocation had relatively low survival.
 
@@
 
-```{r shore-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* The presence of shore habitat that is accessible to frogs appears to have a strong effect on survival. 
-* Is this a direct effect, or an effect mediated through a correlated variable?
+
+-   The presence of shore habitat that is accessible to frogs appears to have a strong effect on survival.
+-   Is this a direct effect, or an effect mediated through a correlated variable?
 For example, lakes lacking shore habitat are often those whose surface elevation decreases during the summer, leaving the shore habitat dry and inaccessible.
 Lower elevation lakes are more likely to experience lake elevation reductions, and are also likely to contain abundant frog predators.
-* Due to potential dependence between elevation and shore variables, unclear if it will be possible to include both shore habitat and elevation in model.
+-   Due to potential dependence between elevation and shore variables, unclear if it will be possible to include both shore habitat and elevation in model.
 
@@
 
-```{r shorehabitat-elevation-plot}
+```{r}
 frog_translocation %>% 
@@
 
-```{r order-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* Contingency table results suggest that translocation order does not affect survival of translocated cohort. However, not a particularly rigorous comparison because order is grouped within sites. 
-* Line plot includes all sites that received two or more translocated cohorts.
+
+-   Contingency table results suggest that translocation order does not affect survival of translocated cohort. However, not a particularly rigorous comparison because order is grouped within sites.
+-   Line plot includes all sites that received two or more translocated cohorts.
 At 6 of 8 of these sites, cohorts after the first cohort (i.e., 2-4) had lower survival than the first cohort.
 At 2 of 8 sites, the second cohort had higher survival.
-* Appears that translocation order has a weak negative effect on survival.
+-   Appears that translocation order has a weak negative effect on survival.
 
@@
 
-```{r donor-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* Substantial differences between donor sites in the survival of translocated frogs.
+
+-   Substantial differences between donor sites in the survival of translocated frogs.
 
@@
 
-```{r donor-bdload-plot}
+```{r}
 recipient_donor <- frog_translocation %>% 
@@
 ```
-* Across all translocated frogs, the 3 donor sites have similar bd_loads. A plot of the 3 donor sites indicate slight differences in bd_load of translocated cohorts. 
-* Across all donor sites and translocations, survival of translocated cohorts may be weakly inversely related to median bd_load of cohorts.
+
+-   Across all translocated frogs, the 3 donor sites have similar bd_loads. A plot of the 3 donor sites indicate slight differences in bd_load of translocated cohorts.
+-   Across all donor sites and translocations, survival of translocated cohorts may be weakly inversely related to median bd_load of cohorts.
 However, grouping by donor site shows that each donor site has its own bd_load-frog survival relationship.
-* bd_load unrelated, or perhaps weakly related, to differences in frog survival between donor sites.
+-   bd_load unrelated, or perhaps weakly related, to differences in frog survival between donor sites.
 
@@
 
-```{r sex-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* No obvious difference in survival between females and males.
+
+-   No obvious difference in survival between females and males.
 
@@
 
-```{r elevation-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* Only 12 elevations represented, so binned by 3 to ensure that 0s and 1s were represented in bin (not with case with 4 bins). 
-* Elevation has positive effect on frog survival.
+
+-   Only 12 elevations represented, so binned by 3 to ensure that 0s and 1s were represented in bin (not with case with 4 bins).
+-   Elevation has positive effect on frog survival.
 
@@
 
-```{r day-frogsurival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* No obvious effect of day of the year on frog survival
+
+-   No obvious effect of day of the year on frog survival
 
@@
 
-```{r year-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 
-* Frog survival has increased with year, perhaps as a result of improved ability to select sites with high quality frog habitat.
+-   Frog survival has increased with year, perhaps as a result of improved ability to select sites with high quality frog habitat.
 
@@
 
-```{r length-frogsurvival-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* Frog size at the time of translocation has a strong positive effect on survival. 
-* This effect could explain, at least in part, the differences in survival between donor sites.
+
+-   Frog size at the time of translocation has a strong positive effect on survival.
+-   This effect could explain, at least in part, the differences in survival between donor sites.
 72996 contains the smallest frogs and has the lowest survival of translocated frogs.
@@
 
-```{r frogsize-bdload-plot}
+```{r}
 frog_translocation %>% 
@@
 
-```{r site-frogsize-plot}
+```{r}
 frog_translocation %>% 
@@
 ```
-* The relationship between donor site and frog length suggests that frog length is only partly responsible for the relationship between donor site and frog survival (frog survival: 70459 > 70567 > 72996).
+
+-   The relationship between donor site and frog length suggests that frog length is only partly responsible for the relationship between donor site and frog survival (frog survival: 70459 \> 70567 \> 72996).
 Perhaps some combination of frog size, bd_load, and genetic factors is associated with frog susceptibility?
@@
 ```
-* Hypothesized mechanism: Severity of winter preceding translocation could affect frog condition and subsequent survival of translocated frogs.
+
+-   Hypothesized mechanism: Severity of winter preceding translocation could affect frog condition and subsequent survival of translocated frogs.
 For example, frogs that just experienced a long winter would be in relatively poor condition and that could reduce survival following translocation.
-* No obvious effect of winter severity in year of translocation and frog survival.
+-   No obvious effect of winter severity in year of translocation and frog survival.
 
@@
 
-* Hypothesized mechanism: Severity of winter following translocation could affect frog condition and subsequent survival of translocated frogs in the first year following translocation.
+-   Hypothesized mechanism: Severity of winter following translocation could affect frog condition and subsequent survival of translocated frogs in the first year following translocation.
     For example, a long winter following translocation could reduce survival in the first year.
-* No obvious effect of winter severity in year following translocation and frog survival.
+
+-   No obvious association between winter severity in year following translocation and frog survival.

### 2023-03-29 doc/notebook/notebook_translocation.Rmd

@@
 
-* site_id: site to which frogs were translocated  
-* elevation: water body surface elevation above sea level, in meters  
-* shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)  
-* snow_t: snow water content (measured on April 1) in year of translocation  
-* snow_t1: snow water content (measured on April 1) in year following translocation  
-* date: date frogs released at recipient site  
-* year: year frogs released at recipient site  
-* day: day of the year on which frogs were released  
-* order: order of translocations to a site, from 1-4. Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation  
-* donor: site_id from which frogs were collected for translocation  
-* pit_tag_ref: pit tag id  
-* sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female   
-* length: frog snout-vent length at capture at donor site (in mm)  
-* weight: frog weight at capture at donor site (in grams)  
-* bd_load: number of ITS copies at release, measured from swab collected at capture at donor site  
-* survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation.
-Re-coded as indicator variable (variable name = "surv"): 0 = survival probability < 0.5; 1 = survival probability $\geq$ 0.5)
+-   site_id: site to which frogs were translocated
 
-Other variables
-* Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.
-* Frog condition is problematic because size and residuals from length:weight regression line are correlated.
+-   elevation: water body surface elevation above sea level, in meters
+
+-   shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)
+
+-   snow_t: snow water content (measured on April 1) in year of translocation
+
+-   snow_t1: snow water content (measured on April 1) in year following translocation
+
+-   date: date frogs released at recipient site
+
+-   year: year frogs released at recipient site
+
+-   day: day of the year on which frogs were released
+
+-   order: order of translocations to a site, from 1-4.
+    Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation
+
+-   donor: site_id from which frogs were collected for translocation
+
+-   pit_tag_ref: pit tag id
+
+-   sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female
+
+-   length: frog snout-vent length at capture at donor site (in mm)
+
+-   weight: frog weight at capture at donor site (in grams)
+
+-   bd_load: number of ITS copies at release, measured from swab collected at capture at donor site
+
+-   survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation.
+    Re-coded as indicator variable (variable name = "surv"): 0 = survival probability \< 0.5; 1 = survival probability $\geq$ 0.5)
+
+Other variables:
+
+-   Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.
+
+-    Frog condition is problematic because size and residuals from length:weight regression line are correlated.
     Specifically, the largest frogs have the largest positive residuals from a length:weight regression line, even after log10 transformation.
@@
 
-* Another consideration when designing the analysis is that there are 12 records in which bd_load is null.
+-   Another consideration when designing the analysis is that there are 12 records in which bd_load is null.
 Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival.
@@
 
-Exploratory data analyses conducted in translocation_survival_exploratory.Rmd provided several notable insights.
+Exploratory data analyses of univariate associations between predictors and frog survival provided several insights.
 
 ### Frog survival across recipient sites
-* 1-year frog survival is highly variable across sites, but highly repeatable within a site.
+
+-   1-year frog survival is highly variable across sites, but highly repeatable within a site.
 This indicates the importance of including site_id or translocation_id as group-level effects.
@@
 ### Donor site
-* Donor site has a strong association with frog survival, with survival at 70459 > 70567 > 72996.
+
+-   Donor site has a strong association with frog survival, with survival at 70459 \> 70567 \> 72996.
 Explored possible effects of bd_load and frog size as drivers of this pattern.
@@
 Whether frog size is the sole driver or is a result of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) is unknown.
-  * Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.  
-  * Strong effect of frog size on 1-year survival, with frogs > 55 mm having substantially higher 1-year survival. Frogs >60 mm had by far the highest survival.  
-  * The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.  
-  * Frog size differs between donor site: size at 72996 < 70459 = 70567.
+    -   Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.
+    -   Strong effect of frog size on 1-year survival, with frogs \> 55 mm having substantially higher 1-year survival. Frogs \>60 mm had by far the highest survival.
+    -   The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.
+    -   Frog size differs between donor site: size at 72996 \< 70459 = 70567.
 This partially mirrors the observed differences in survival between sites.
@@
 ### Bd load
-* Frogs with the highest loads at the time of translocation had the lowest survival.
+
+-   Frogs with the highest loads at the time of translocation had the lowest survival.
 The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies.
-Notable result given that the threshold level of 56,000 copies is /~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.
+Notable result given that the threshold level of 56,000 copies is /\~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.
 
 ### Recipient site characteristics
-* **Shore habitat**: The presence of shore habitat is associated with increased frog survival.
+
+-   Shore habitat: The presence of shore habitat is associated with increased frog survival.
 Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature.
 Tight association between shore habitat presence/absence and elevation will make it impossible to disentangle these effects, include only elevation in the model.
-* **Elevation**: Strong positive effect on frog survival.
+-   Elevation: Strong positive effect on frog survival.
 
 ### Translocation timing
-* Day of the year showed no obvious relationship to frog survival.  
-* Year was positively associated with survival.
+
+-   Day of the year showed no obvious relationship to frog survival.\
+-   Year was positively associated with survival.
 This is likely not a direct effect of year, but instead reflects better-informed selection of recipient sites with high-quality frog habitat over time.
@@
 ### Winter severity
-* No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.
+
+-   No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.
 
 ### Translocation order
-* Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.  
-* It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.
+
+-   Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.
+-   It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.
 
 ### Frog characteristics
-* Frog size had strong effect on 1-year survival.  
-* Minimal differences in survival between males and females (slightly higher in males).
+
+-   Frog size is associated with 1-year survival.
+-   Minimal differences in survival between males and females (slightly higher in males).
 
@@
 Pairs plot to identify correlations between predictors (ggpairs from GGally package):
-* It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length.
+
+-   It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length.
     As such, these two variables have the potential to compete for the same variation.
@@
     Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.
-* Also need to disentangle elevation from site_id - currently only one elevation per site_id.
-Try breaking continuous elevation into categories, with > 2 site_ids per category.
+
+-   Also need to disentangle elevation from site_id - currently only one elevation per site_id.
+    Try breaking continuous elevation into categories, with \> 2 site_ids per category.
     Hopefully, that will make elevation variable less dependent on site_id.
     A similar situation may exist for the two winter severity variables and site_id.
-* Shore and elevation may also be closely associated. Are all sites with shore = 0 at low elevation? If so, may need to only include elevation in model.
 
-Group-level effects: 
-The dataset has three sources of hierarchical structure: site_id, translocation_id, and donor site.
+-   Shore and elevation are also closely associated.
+    Are all sites with shore = 0 at low elevation?
+    If so, may need to only include elevation in model.
+
+Group-level effects: The dataset has three sources of hierarchical structure: site_id, translocation_id, and donor site.
 Donor site is not suitable as a grouping variable because there are only 3 levels (i.e., site ids), not enough to make this useful.
@@
 
-* Bin elevation into "low" and "high" groups.  
-* Recode order as 0/1 (instead of 1--4): "first".  
-* Recode survival as 0/1 (instead of continuous probability): "surv".  
-* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below.  
-* Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.  
-* Standardize continuous variables: elevation, snow_t, snow_t1, day, length.
+-   Bin elevation into "low" and "high" groups.\
+-   Recode order as 0/1 (instead of 1--4): "first".\
+-   Recode survival as 0/1 (instead of continuous probability): "surv".\
+-   Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below.\
+-   Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.\
+-   Standardize continuous variables: elevation, snow_t, snow_t1, day, length.
 Possibly standardize log(bd_load) also, but see Gelman for recommendations.
@@
 
-Relevant references:  
-* Gelman 2008: https://doi.org/10.1002/sim.3107  
-* Schielzeth 2010: https://doi.org/10.1111/j.2041-210X.2010.00012.x
+Relevant references:\
+\* Gelman 2008: <https://doi.org/10.1002/sim.3107>\
+\* Schielzeth 2010: <https://doi.org/10.1111/j.2041-210X.2010.00012.x>
 
 ### Priors
-* For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.
-https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations.
-* Other recommendations in:  
-  * Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: https://doi.org/10.1111/oik.05985.  
-  * Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: https://doi.org/10.1002/ecs2.3739.
+
+-   For regression coefficients in logistic regression: A recommended weakly informative prior is beta \~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3\<nu\<7.
+<https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations>.\
+-   Other recommendations in:
+    -   Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: <https://doi.org/10.1111/oik.05985>.\
+    -   Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: <https://doi.org/10.1002/ecs2.3739>.
 
 ## Results
-- Important predictors of the 1-year survival of translocated frogs were quite consistent across all models (see `translocation_analysis_survival.Rmd` for details).  
+
+-   Important predictors of the 1-year survival of translocated frogs were quite consistent across all models (see `translocation_analysis_survival.Rmd` for details).\
 -   In the "best" model that included standardized predictors and translocation_id as a grouping variable, important predictors included elevation (+), winter severity in the year following translocation (-), and donor[72996] (-).
-Model R2 was 0.32, so a substantial amount of variation remained unaccounted for.
+Model R2 was 0.32, so a substantial amount of variation remained unaccounted for.\
 -   Comparison of different grouping structures (translocation_id, site_id, translocation_id nested within site_id) indicated that all had similar posterior predictive accuracy.
-However, using translocation_id as the grouping variable allowed greater insights into the effects of the predictor variables, presumably due to the greater variation associated with the translocation_id grouping variable (n = 24) compared to the site_id grouping variables (n = 12).
+However, using translocation_id as the grouping variable allowed greater insights into the effects of the predictor variables, presumably due to the greater variation associated with the translocation_id grouping variable (n = 24) compared to the site_id grouping variables (n = 12).\
 -   Posterior predictive accuracy of "best" model was 0.79, and the true negative rate (proportion of frogs that did not survive that were accurately classified as not surviving: 0.88) was higher than the true positive rate (proportion of frogs that survived that were accurately classified as surviving: 0.59).
@@
 This is borne out by sites such as 70370 which has attributes suggesting high frog survival (e.g., high elevation, available bank habitat, few predators), but in fact frogs from both translocations experienced very low survival.
-In contrast, no examples exist of sites where frog survival was (retrospectively) predicted to be low, but actual survival was high (although 70449 comes closest to providing such an example).
+In contrast, no examples exist of sites where frog survival was (retrospectively) predicted to be low, but actual survival was high (although 70449 comes closest to providing such an example).\
 -   Two supplementary analyses were conducted to gain additional insights into the 1-year survival model.
@@
 The existence of such an attribute is suggested by the consistently high survival resulting from the four translocations conducted to date from 70459 to two very different sites, and the fact that in both recipient sites that have received frogs from 70459 tadpole mouthparts are fully pigmented.
-That is not the case with tadpoles at sites that received frogs from 72996.
+That is not the case with tadpoles at sites that received frogs from 72996.\
     -   The frog translocation dataset contained 12 records for which bd_load = NA.
@@
 Given that bd_load was not an important predictor of frog survival, I reran the model using the complete dataset and dropping bd_load from the model.
-Dropping/including those 12 records had no effect on model results.
+Dropping/including those 12 records had no effect on model results.\
 -   The results from a secondary analysis to complement the 1-yr survival model demonstrated that the average survival in the first translocation was an important predictor of frog survival in subsequent translocations.
 This demonstrates the importance of recipient site attributes (elevation, habitat, predators, etc.) in determining frog survival.
+
+### Visualize conditional effects
+
+-   As described in <https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html>, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on.
+By default, the mean is used for continuous variables and the reference category is used for factors, but you may change these values via argument conditions.
+This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."
+-   Created a new dataset for each predictor in which all variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).

### 2023-03-29 code/translocation_survival_analysis.Rmd
Updated with new figures and edits/comments from Allie/Bree

@@
   geom_vline(xintercept = 0, color = "red", linetype = "dashed", lwd = 0.5) +
-  scale_y_discrete(labels = labels_alt, limits = rev)
-#  ggtitle("Model with standardized predictor & no grouping variable",
-#          "Posterior distributions with medians & 95% uncertainty intervals")
+  scale_y_discrete(labels = labels_alt, limits = rev) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.line = element_line(color = "black"))
+
+ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m2b.png"), bg = "white", width = 7, height = 4)
 
 mean(bayes_R2(m2b))
-```
-
-##### Save plot
-
-```{r}
-ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m2b.png"), bg = "white", width = 7, height = 4)
 ```
@@
   scale_y_discrete(labels = labels_alt, limits = rev) +
-  theme(plot.background = element_rect(fill = "white", color = "white"))
+  theme(panel.background = element_rect(fill = "white"),
+        axis.line = element_line(color = "black"))
+
+ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m1d.png"), width = 5, height = 5)
 
@@
 -   Based on 95% uncertainty intervals, important predictors are similar to those in model without grouping variable: snowt1_z, elev_z, male_c0.548, donor72996.
-
-#### Save plot
-
-```{r}
-ggsave(here::here("doc", "manuscript", "figures", "mcmc_areas_m1d.png"), width = 5, height = 5)
-```
 

### 2023-03-29 doc/manuscript/translocation.qmd
Updated with new figures and edits/comments from Allie/Bree

@@
 Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014].
-In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
+"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
+In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]<!--# The previous sentence might fit better in the Discussion when we take a deeper dive into evolutionary rescue -->.
 Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
@@
 These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
+<!--# Bree comment: All of this was written before we knew the genomics had something cool to say.
+But now that the genomics feel a bit more elevated, I think this paragraph should be more bold.
+For example, we might want more here or in the Discussion to make it clear how the results of each of these sections are super important on their own.
+And that combining them is really amazing.
+I think the current tone is appropriately factual, but at some point we need to be louder about the novelty involved and the incredible amount of work that this represents.
+This version feels like, we did a,b,c and so MYLF meet the conditions for feasibility (which makes it sound like we've only just gotten to the *starting* point of the study.
+I think we need to go further than saying they meet the four conditions of a specific conservation action.
+This is where beefing up the conclusions about evolutionary rescue seem appropriate.
+Also, genomic analyses typically come last in our papers as a dive into possible mechanistic explanations for observed patterns in the field.
+I think the order of the primary sections makes sense given the way you set up the argument about evolutionary rescue.
+That said, the order combined with the tone feels like it's contributing to the sense that the genomics are a box to be checked rather than contributing something exciting to the overall story.
+Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->
 
@@
 --> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
-Our comparison of frog genomes between naive and recovering populations is not as powerful an approach as comparing samples collected from populations before and after Bd exposure, but because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable, no such samples are available and are unlikely to be available in the future.
-As such, the current study design is the best possible under the circumstances.
-To detect regions of the genome under selection, we compared genomes of frogs from naive and recovering populations using a multivariate linear mixed model, and created "liberal" and "stringent" sets of outlier variants using a Bonferroni-corrected significance level of 0.05 and 0.01, respectively.
+Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
+The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
+As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
+-->.
+We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
 
-The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and persistent groups (@fig-selectionresults B: SI).
+The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
 The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
@@
 No biological functions, molecular functions, or cellular processes were significantly overrepresented in the set of 29 annotated outlier genes.
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI).
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
 For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
@@
 However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
-<!--# Allie should add a brief description of the function of RIN3 here. Also, emphasize key messages related to selection, resistance. And, of course, rework any of the text as necessary. -->
 
-<!--# Figure edits:  selectionresults: Correctly locate population 7, capitalize panel labels, label liberal and stringent outlier sets.
+In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
+-->.
+We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers in both analyses and two of these shared regions also contained outlier SNPs.
+A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (Table 3: SI <!--# add this table -->).
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
+Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) ([#fig-spline-manhattan](#fig-spline-manhattan) D).
+This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
+In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+
+<!--# Figure edits:  selectionresults: Capitalize panel labels, label liberal and stringent outlier sets.
 outlier map: Rearrange panels to put RIN3 last?
@@
 
+<!--# Bree comment: Here or elsewhere - it feels like we need a more specific bridge from the genomics section into the translocation section.
+I think we need to be explicit about how the populations we chose for the genomics section relate to the populations used in translocations.
+Even if they aren't matched in any way, we should reiterate something about how the experimental designs comport.
+For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations.
+-->
+
+<!--# Roland comment: 2 of the 3 donor populations are included in the selection study. -->
+
 Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
@@
 Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
+<!--# Allie comment: This phrasing is a bit confusing for me.
+--> The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
 Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
@@
 Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
-In addition, the moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
+<!--# Allie comment: Maybe tolerance is a better word here?
+Since they were infected.
+Or maybe we need to explicitly define what we mean with resistance vs tolerance.
+Bree comment: I noticed this also – in the Intro, you do say that you are lumping both resistance and tolerance into the phrase "resistance" but you are talking about multiple things (being able to avoid infection, being able to fight infection, being able to tolerate infection, etc).
+Maybe at some point we want to address specifically that the genomic changes we are observing could contribute to multiple facets of resistance / tolerance.
+And that we are lumping those together in an "evolved resistance" bin.
+--> In addition, the moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
 Finally, in three locations where longer capture-mark-recapture (CMR) time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
@@
 Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
+Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
 This initial filter resulted in 1,595,206 variant sites across 53 individuals.
@@
 
-To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet package [@jombart2008].
-Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014].
+To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
+Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014] <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
+-->.
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
@@
 
-Using the liberal set of outliers, we performed a statistical overrepresentation test using PANTHER.
-First, we retrieved the UnitProtKB ID for each gene in the *Xenopus laevis* genome, Next, we compared the 29 unique outlier genes from the stringent set to the complete set of genes in the *X.
-laevis* genome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+To detect regions of the genome under selection <!--# Improve wording -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
+First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
+This method also calculates a W-statistic allowing for outlier identification.
+We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
+Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
+We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.
 
-Finally, for the stringent set of outliers we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+Using the liberal set of variant outliers, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] .
+First, we retrieved the Blast hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 29 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+For the stringent set of SNP outliers, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
 First, we extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
@@
 
-### References
-
-::: {#refs}
-:::
-
 ### Figures
 
-<!--# Discuss referencing and in-text generation of figures.
-We don't want to re-run the population viability modeling figures each time we compile the manuscript.
-A makefile could be a good solution.
-Caching figures could be another option (https://quarto.org/docs/computations/caching.html) -->
+<!--# Bree comment: At some point it might be good to talk about what figures go in main text and which stay in supplement (or if we might make some multi-panel figures so that we can fit more in the main text).
+One thing I'm struck by in the figures is that each of the three main sections (genetics, translocations, PVA) are each so information rich but also somewhat disconnected.
+Not sure if we want to create some kind of synthetic or hybrid conceptual/data figure that pulls the main point of each analysis out and puts it in visual/conceptual relationship with the others.
+The results are so cool and we don't want readers to gloss over the details or have to work too hard to put it all back together again.
+-->
 
@@
 ![Allele frequencies of all outlier SNPs in the stringent set for each of the naive and recovering *R.
-sierrae* populations.
+sierrae* populations superimposed on a map of the Sierra Nevada study area.
+Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) are shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada is shown in light gray.
 The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.
-<!--# Allie: Rearrange panels and add panel identifiers (A, B, C) as shown here.
--->](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}
+<!--# Allie: Add panel identifiers (A, B, C)?.
+Put RIN3 panel last?
+Add compass rose and format distance bar to match that in Yosemite map--> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful -->](figures/allele_maps.jpg){#fig-allelefrequencies}
+
+![Results from the splined window analyses showing outlier windows for $\pi_{diff}$ (a) and F~ST~ (b).
+Outlier windows in only one analysis are shown in blue and overlapping outlier windows are red.
+(c) Magnified Contig19 (from (a)) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(d) Magnified Contig8 from (b) showing F~ST~ outlier window that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 10 annotated genes known to occur in the extended MHC Class I and II regions.](figures/spline_manhattan_figure.jpg){#fig-spline-manhattan}
 
@@
 
-<!-- Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.
+<!--# Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.
 We can discuss.
@@
 
-### References
-
 ### Figures
@@
 ![Results from the exome capture selection study.
-(A) Map of sampling locations showing Bd-naive populations in red and persistent/recovering populations in blue.
+(A) Map of sampling locations in the Sierra Nevada study area showing Bd-naive populations in red and persistent/recovering populations in blue.
+Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) are shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada is shown in light gray.
 (B) PCA calculated from binary SNPs showing genomic relationship of samples.
@@
 Bold labels indicate at least one non-synonymous SNP is present in that gene.
-<!--# Is panel (C) important to include here?
-Or should it be moved to the SI?
--->](figures/mylf_selection_figure_5Dec2022.png){#fig-selectionresults}
+<!--# Map is duplicated here and in the mapped allele frequency figure.
+Remove from this figure?
+-->](figures/map_qqplot_manhattan.jpg){#fig-selectionresults}
 
@@
 Points are median estimates and error bars give the 90% uncertainty intervals around the estimates.
-All estimates are derived from the mrmr package [@joseph2018].
-<!--# Use 95% intervals to be consistent with other analyses?
+All estimates are derived from the mrmr package.
+<!--# Use 95% uncertainty intervals for consistency?
 -->](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
@@
 
-  knitr::kable(viability_table, "pipe", align = "lcccr", caption = ) 
+  knitr::kable(viability_table, "pipe", align = "lcccr", col.names = c("Period", "Viability category", "Number of translocations", "Total translocations", "Proportion of translocations")) 
 ```
@@
 
+#### Table files
+
+-   gemma_outliers.csv
+
+-   spline_window_shared_outliers.csv
+
 <!--# Include table of translocation sites, years, frog numbers? -->
 
-<!--# Include Allie's table describing outlier SNPs? -->
+### References
 
-###
+##

### 2023-03-29 doc/manuscript/translocation.qmd
Add key points to Discussion

@@
 
-<!--# Consider broadening the scope of this sentence and the paragraph to be developed from it to encompass all of our findings -->
+1.
+Summary of all of the results in a way that integrates the different components/sections (selection, translocations, viability), and makes the pitch for how unique the results are (as per comment from Bree):
+    -   Evidence for evolutionary rescue in a conservation-relevant species.
+    -   Successful reestablishment of extirpated populations via translocation despite ongoing Bd infection.
+    -   Evidence for long-term viability of reestablished populations.
 
-TO BE WRITTEN.
+```{=html}
+<!-- -->
+```
+2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.
+    -   Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
+    -   Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
+    -   There is a shared selective response across recovering populations.
+    -   Changes in population growth rate due to evolution extend beyond the natal environment.
+    -   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
 
-We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for emergence of this resistance.
+Could include mention here of why MYL frogs show evidence of evolutionary rescue in response to Bd exposure when so few other amphibians do.
+Text from Introduction describing the conditions under which evolutionary rescue is most likely?
 
-1.  Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
-2.  Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
-3.  There is a shared selective response across recovering populations.
-4.  Changes in population growth rate due to evolution extend beyond the natal environment.
-5.
-In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-
-<!--# Discuss effect of donor population on survival of translocated frogs. Differences in Bd resistance? -->
-
-<!--# The following point was made in the original Introduction outline but ultimately not addressed there: Translocation of individuals from putatively resistant populations to sites that were formerly occupied but from which populations were extirpated by disease provides a way to disentangle the contributions of natal environment and host genetics on host persistence and recovery.
--->
-
-<!--# Translocation "success" often defined as evidence of reproduction by translocated individuals (see IUCN and related references in notes).
-We used a much higher and more conservative standard to assess success, i.e.,  -->
+3.  Translocations (including viability results?):
+    -   Recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
+Hundreds of declining species, but mostly failed attempts at recovery.
+Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
+In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
+    -   Observed reduced susceptibility is likely due to evolutionary rescue, and this outcome is likely facilitating both natural recovery (Knapp et al.
+2016) and success of translocation efforts described in this study.
+MYL frogs provide a rare example of highly susceptible species recovering in the presence of pathogen/disease.
+Recovery would likely not be possible without evolutionary rescue, as suggested by numerous failed attempts to reestablish populations of Bd-impacted amphibians.
+    -   IUCN guidance for reintroduction/translocation success is the presence of reproduction.
+Based on this standard, X of our 12 translocation efforts would be deemed successful.
+We used a higher standard - 1 year survival and probability of extinction over 50 years.
+More relevant standard because of the bottleneck to population establishment posed by subadult survival which despite evolutionary rescue is strongly influenced by disease.
+    -   This success in the face of a highly virulent pathogen is all the more surprising given the generally low rate of translocation/reintroduction success (several review papers).
+Despite relatively high success rate in MYL frogs, it is apparent that not all such recovery efforts will succeed.
+In addition to ongoing effects of Bd, several deterministic and stochastic factors likely also play an important role: (1) effects of elevation, winter severity, and donor population on frog survival, and (2) variability in recruitment driven by Bd-caused mortality of subadults.
+    -   Now have a means of reestablishing MYL frog populations across their historical range, including in areas in which natural recolonization is precluded by dispersal limitation.
+Nonetheless, recovery across this large landscape will require substantial resources over many decades.
+    -   Big picture: May also have implications for the recovery of the many Bd-impacted amphibians worldwide.
+Remains to be seen how many other amphibians respond to Bd exposure via evolutionary rescue.
 
@@
 Bold labels indicate at least one non-synonymous SNP is present in that gene.
-<!--# Map is duplicated here and in the mapped allele frequency figure.
+<!--# Map (panel A) is duplicated here and in the mapped allele frequency figure.
 Remove from this figure?
@@
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="450"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
+<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="450"}
 

### 2023-03-29 code/install_R_packages.R
Make a master Makefile with translocation and viability analyses

@@ -3,7 +3,8 @@
 list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
 					  "rstan", "lubridate", "tidyverse", 
 					  "coda","mvtnorm","devtools","loo","dagitty", "tmap",
-					  "RPostgreSQL", "DBI", "here")
+					  "RPostgreSQL", "DBI", "here", "brms", "rstanarm", 
+					  "tidybayes", "broom.mixed", "janitor", "arm")
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
 

### 2023-03-29 doc/manuscript/translocation.qmd
Minor edits and add several comments

@@
 
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+<!--# Roland will address some of Bree's points here -->Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
@@
 Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+
+<!--# Need summary of importance of genetic results, also need better integration of different selection  results -->
 
@@
 
-<!--# Roland comment: 2 of the 3 donor populations are included in the selection study. -->
+<!--# Roland comment: 2 of the 3 donor populations are included in the selection study. Add 2 sentences that summarize selection results and bridge to translocation section. -->
 
@@
 To identify site and frog-specific drivers of post-translocation survival, we developed a multilevel model in a Bayesian framework.
-Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
+Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
 <!--# Allie comment: This phrasing is a bit confusing for me.
@@
 Therefore, frog resistance in non-natal habitats is maintained across generations.
-
-<!--# Mark W: Regarding the above paragraph, what evidence do we have of "tolerance"?
-Are Bd loads significantly reduced in populations compared to epizootic levels?
-Can we convincingly state that the likely mechanism is more consistent with resistance or tolerance?
-I ask because it would be nice to just use one of these terms in the paper and not get caught up in any resistance vs.
-tolerance discussions with reviewers.
-Roland: Have a look at this paragraph of Knapp et al.
-PeerJ 2022: https://peerj.com/articles/12712/#p-61.
-We argue that there is evidence for resistance and tolerance.
--->
 

### 2023-03-29 code/extract_translocation_data.R
Add removed comma

@@ -168,7 +168,7 @@ for(lake in keep_lakes){
 	pop_size = data.table(fit$m_fit$summary(variables="N"))
 
 	# Build the stan data so we know when translocations occurred
-	captures = readr::read_csv(file.path(cmr_path, "capture" paste0(lake, "_capture.csv")))
+	captures = readr::read_csv(file.path(cmr_path, "capture", paste0(lake, "_capture.csv")))
 	surveys = readr::read_csv(file.path(cmr_path, "survey", paste0(lake, "_survey.csv")))
 	translocations = readr::read_csv(file.path(cmr_path, translocation, paste0(lake, "_translocation.csv")))
 	mrmr_dat = clean_data(captures = captures, surveys = surveys, translocations = translocations)

### 2023-03-30 code/extract_translocation_data.R
Test merged makefile and add environment.yml file

@@ -170,7 +170,7 @@ for(lake in keep_lakes){
 	# Build the stan data so we know when translocations occurred
 	captures = readr::read_csv(file.path(cmr_path, "capture", paste0(lake, "_capture.csv")))
 	surveys = readr::read_csv(file.path(cmr_path, "survey", paste0(lake, "_survey.csv")))
-	translocations = readr::read_csv(file.path(cmr_path, translocation, paste0(lake, "_translocation.csv")))
+	translocations = readr::read_csv(file.path(cmr_path, "translocation", paste0(lake, "_translocation.csv")))
 	mrmr_dat = clean_data(captures = captures, surveys = surveys, translocations = translocations)
 
 	# Extract year of primary period

### 2023-03-30 code/install_R_packages.R

@@ -1,23 +1,18 @@
 # Install the packages necessary for R analyses
 
-list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
-					  "rstan", "lubridate", "tidyverse", 
-					  "coda","mvtnorm","devtools","loo","dagitty", "tmap",
-					  "RPostgreSQL", "DBI", "here", "brms", "rstanarm", 
-					  "tidybayes", "broom.mixed", "janitor", "arm")
-new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
+# List of CRAN packages needed for the analysis.  Uncomment and run to load packages in your 
+# native R environment. Otherwise, packages are assumed to be loaded through conda.
+# list.of.packages <- c("data.table", "ggplot2", "patchwork", "cmdstanr", "magrittr",
+# 					  "rstan", "lubridate", "tidyverse", 
+# 					  "coda","mvtnorm","devtools","loo","dagitty", "tmap",
+# 					  "RPostgreSQL", "DBI", "here", "brms", "rstanarm", 
+# 					  "tidybayes", "broom.mixed", "janitor", "arm", "stars")
+# new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
+# if(length(new.packages)) install.packages(new.packages, repos='http://cran.us.r-project.org')
 
 # Load Github packages
-github_packages = c("rethinking", "mrmr")
+github_packages = c("SNARL1/mrmr", "rmcelreath/rethinking")
 
 for(package in github_packages){
-
-	if(!(package %in% installed.packages()[,"Package"])){
-		if(package == "rethinking"){
-			devtools::install_github("rmcelreath/rethinking")
-		} else if(package == "mrmr"){
-			devtools::install_github("SNARL1/mrmr")
-		} 
-	}
+	devtools::install_github(package)
 }

### 2023-03-30 doc/manuscript/translocation.qmd
Add text to Intro emphasizing study importance

@@
 csl: pnas.csl
-format: docx
+format: pdf
 editor: visual
@@
 First, populations that survived following disease arrival have evolved increased resistance.
-Second, increased resistance contributes to higher individual fitness in the presence of disease and allows positive population growth.
+Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"?
+Resistance could convey complete prevention of disease, so I often struggle with which word to use -->and allows positive population growth.
 Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
@@
 
-<!--# Roland will address some of Bree's points here -->Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
-Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
-These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
+
+In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
+-->.
+We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
+Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
+
 <!--# Bree comment: All of this was written before we knew the genomics had something cool to say.
@@
 Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->
+
+<!--# Original pre-edit text from the above paragraph: Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
+Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
+These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
+-->
 
@@
 
-```{=html}
 <!-- -->
-```
+
 2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.

### 2023-03-30 doc/manuscript/translocation.qmd
Initial comments on genomics analysis

--- 2023-03-29 doc/manuscript/translocation.qmd
@@
 
-<!--# Roland will address some of Bree's points here -->Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+<!--# Roland will address some of Bree's points here -->
+
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
@@
 
-To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+To assess whether frog populations display an evolutionary response to Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Our goal was to identify putative genes under selection and potentially related to increased frog resistance to Bd.
 Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
@@
 Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014] <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
--->.
+For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
+The first approach A, did X, Y, and Z but had C limitations.
+The second approach B, did X, Y, and Z and D limitations."-->.
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
 For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
-GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
+GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?
+Probably obvious for genomics folks, but not totally clear to me.
+I am guessing we are testing whether the `population` coefficient (naive vs.
+persistent) is significant for describing variability in a SNP?
+--> for the set of 148,307 high quality SNPs across 127 contigs.
 We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
-Results were visualized using a Manhattan plot and qqplot.
+<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs.
+persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.
 
-To detect regions of the genome under selection <!--# Improve wording -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
+To detect regions of the genome under selection <!--# Improve wording.
+Also, can we contrast this with the previous approach?
+What is this analysis doing that the other one didn't?
+-->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
 First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].

### 2023-03-31 doc/manuscript/translocation.qmd
Numerous edits throughout Methods and Results

--- 2023-03-30 doc/manuscript/translocation.qmd
@@
 -->.
-We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+We then reintroduced these resistant frogs to 24 sites and described the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
 These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
@@
 As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
--->.
-We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
+To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window analysis to evaluate larger genomic regions showing differentiation between population types.
+For single SNPs, we identified outlier variants using a Bonferroni-corrected significance level (liberal = 0.05, stringent = 0.01), and for the splined window analysis we used a W-statistic of 4 standard deviations from the mean.
+<!--# This is different from the definition in the Methods: 4 * standard deviation -->
 
@@
 One gene in this group does not have an associated annotation.
-No biological functions, molecular functions, or cellular processes were significantly overrepresented in the set of 29 annotated outlier genes.
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
-For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
+No biological functions, molecular functions, or cellular processes <!--# Are these the "GO terms" referred to in the next paragraph?
+If so, add here paranthetically.
+--> were significantly overrepresented in the set of 29 annotated outlier genes.
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
+For these SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
 Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
@@
 
-In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
+In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both *F~ST~* and $\pi_{diff}$ ("shared" regions) and 2 of the 9 shared regions also contained the outlier SNPs described above <!--# SNPs from the liberal or strict set?
+Also "windows" and "regions" are used interchangeably in this paragraph.
+To avoid confusion, use "windows" throughout?
 -->.
-We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers in both analyses and two of these shared regions also contained outlier SNPs.
-A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (Table 3: SI <!--# add this table -->).
+A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
 The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
@@
 Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) ([#fig-spline-manhattan](#fig-spline-manhattan) D).
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
 This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
@@
 Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+<!--# Based on Fig.
+2, there are numerous outlier windows.
+How were the specific windows mentioned above chosen?
+Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria.
+-->
 
-<!--# Need summary of importance of genetic results, also need better integration of different selection  results -->
-
-<!--# Figure edits:  selectionresults: Capitalize panel labels, label liberal and stringent outlier sets.
-outlier map: Rearrange panels to put RIN3 last?
-Relocate pie charts to avoid the perception that all recovering populations are west of naive populations, add A, B, C labels to panels.
+<!--# In reading through these results, I am not seeing a compelling theme that emerges from all of the different results, both within and between the SNP and splined window sections.
+The results show differences in various metrics between population types, but are these changes large or small?
+I lack the context to understand that.
+Similarly, is the magnitude of differences consistent with strong or weak selection?
+In addition to the above, this section needs a concluding summary of the key genetic results.
 -->
@@
 
-<!--# Roland comment: 2 of the 3 donor populations are included in the selection study. Add 2 sentences that summarize selection results and bridge to translocation section. -->
-
-Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies, consistent with strong selection following Bd exposure.
+Here we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog genome study (@fig-selectionresults: population 1 and 4).
+<!--# Check this --> Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
 These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
@@
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Mention of viability is confusing because this issue isn't dealt with until the third section.
+Perhaps make the categorization based on frog survival?
+-->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
@@
 <!--# Add median Bd load, proportion below 5.8 copies threshold?
-Mention results from statistical model more explicitly?
 --> <!--# The Methods-Translocation section contains text that repeats similar results.
 Edit to reduce repetition.
---> <!--# To allow comparison with other translocation studies, consider adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
+--> <!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
 -->
 
-To identify site and frog-specific drivers of post-translocation survival, we developed a multilevel model in a Bayesian framework.
+To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
 Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
-<!--# Allie comment: This phrasing is a bit confusing for me.
---> The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
 Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
@@
 And that we are lumping those together in an "evolved resistance" bin.
---> In addition, the moderate Bd loads that characterized frogs following translocation and relatively high survival of frogs in the majority of more-recently translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in three locations where longer capture-mark-recapture (CMR) time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+--> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
+Finally, in three locations where longer CMR time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
 Therefore, frog resistance in non-natal habitats is maintained across generations.
-
-<!--# Drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd.
--->
 
@@
 
-To test whether the observed yearly survival probabilities of resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values SI).
-Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
+Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
+To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
+Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+These six populations all had observed yearly adult survival greater than 0.5.
 When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
-In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
+In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
 A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
+<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focursed on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., sensitivity analysis, etc.).
+It is my understanding that you are generally assuming that these two probabilities are equivalent and interchangable, but additional clarification would be helpful.
+Might also require some clarification in Table 2.-->
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
+Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model <!--# Is this a different model, now with demographic stochasiticity and environmental variability in juvenile recruitment, or the same model used to estimate lamda in the previous paragraph?
+Clarify.
+-->predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
 Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
-For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
+<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less then 1", "some probability of persistence".
+In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are much less than 1.
+Consider tweaking wording to clarify this.
+I'd argue for a relatively strong initial statement that we then fine tune only slightly, instead of a weak statement that we then strengthen a lot.
+Initial impressions can be hard to change.
+--> For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5 <!--# Based on Fig.
+5, looks like extinction propability is < 0.5 for all six populations only for omega > ~0.22.
+Reword as necessary.
+-->, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 I tried to phrase this carefully.
+-->
+
+<!--# Is it worth mentioning that predictions may be conservative because yearly survival probability of naturally recruited animals may be higher than yearly survival probability of the translocated animals that was used to parameterize model (Figure 12).
 -->
@@
 Recovery would likely not be possible without evolutionary rescue, as suggested by numerous failed attempts to reestablish populations of Bd-impacted amphibians.
+Although evolution of resistance in donor populations facilitates the long-term viability of translocated populations, survival is still highly dependent on environment independent of Bd.
     -   IUCN guidance for reintroduction/translocation success is the presence of reproduction.
@@
 
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four Bd-naive populations, and 29 from five persistent/recovering populations.
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four Bd-naive populations, and 29 from five recovering populations.
 These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-selectionresults A).
-Sampling was even across the sites, with 5-6 samples per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
+Samples were collected from 5-6 frogs per population.
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges (@fig-selectionresults A).
 DNA was extracted following Qiagen DNEasy manufacturer's protocols.
@@
 
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]) with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome [NCBI SRA: SRS6533475, reference = @hon2020] with repetitive elements masked using bwa ("mem" mode) [@li2013].
 Exact PCR Duplicates were marked using Picard.
@@
 Finally, we trimmed samples with an average depth across filtered sites \< 7x.
-Our final dataset included 50 samples (23 from naive and 27 from persistent/recovering populations) with an average depth of 16.7x (range = 7.4x -- 26.1x).
+Our final dataset included 50 samples (23 from naive and 27 from recovering populations) with an average depth of 16.7x (range = 7.4x -- 26.1x).
 
@@
 To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
-Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014] <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
--->.
+Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
+The multivariate linear mixed model was implemented in the program GEMMA [@zhou2014].
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
-For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
+For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
 GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
-We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
+We calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
 Results were visualized using a Manhattan plot and qqplot.
 
-To detect regions of the genome under selection <!--# Improve wording -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
-First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We conducted the splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
+First, we calculated per-site *F~ST~* between individuals from naïve and recovering populations for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
 We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
@@
 For the stringent set of SNP outliers, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-First, we extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
-Next, we reannotated each sequence using blast to get the predicted gene location based on the closest annotated reference [@altschul1997].
+To do this, we first extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
+Next, we re-annotated each sequence using blast to get the predicted gene location based on the closest annotated reference [@altschul1997].
 We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
@@
 Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+<!--# Mention that in both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available.
+-->
 
@@
 
-![Allele frequencies of all outlier SNPs in the stringent set for each of the naive and recovering *R.
-sierrae* populations superimposed on a map of the Sierra Nevada study area.
-Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) are shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada is shown in light gray.
-The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests a role for RIN3 in frog resistance against Bd infection and associated population recovery.
-<!--# Allie: Add panel identifiers (A, B, C)?.
-Put RIN3 panel last?
-Add compass rose and format distance bar to match that in Yosemite map--> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful -->](figures/allele_maps.jpg){#fig-allelefrequencies}
+![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the stringent set of 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
+The inset map locates the study area in California.
+<!--# Allie: Add compass rose and format distance bar to match that in Yosemite map.
+Relocate pie charts to avoid the perception that all recovering populations are west of naive populations.
+In legend, provide interpretation of "Homozygous reference", "Homozygous alternate", and "Heterozygous".
+Try moving inset map to lower left portion of panel - more room there to enlarge inset?
+How about identifying each population as R.
+muscosa or R.
+sierrae using squares and dots?
+--> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful.
+Roland response:  -->](figures/allele_maps.jpg){#fig-allelefrequencies}
 
-![Results from the splined window analyses showing outlier windows for $\pi_{diff}$ (a) and F~ST~ (b).
-Outlier windows in only one analysis are shown in blue and overlapping outlier windows are red.
-(c) Magnified Contig19 (from (a)) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(d) Magnified Contig8 from (b) showing F~ST~ outlier window that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 10 annotated genes known to occur in the extended MHC Class I and II regions.](figures/spline_manhattan_figure.jpg){#fig-spline-manhattan}
+![Results from the splined window analyses showing outlier windows for $\pi_{diff}$ (A) and *F~ST~* (B).
+Outlier windows for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier windows for both $\pi_{diff}$ and *F~ST~* are shown in red.
+(C) Magnified Contig19 from A showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from B showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 10 annotated genes known to occur in the extended MHC Class I and II regions.
+<!--# As I commented in the Results text, why where the regions highlighted in C and D chosen?
+-->](figures/spline_manhattan_figure.jpg){#fig-spline-manhattan}
 
@@
 Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Frog survival was estimated using capture-mark-recapture methods.
+Frog survival was estimated using CMR methods.
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
@@
 The dashed line shows where $\omega = 0.3$, for comparison to A.
-<!--# Mention that there are six lines at p(extinction) = 1?
--->](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
 
@@
 We can discuss.
-Roland says: Maybe try a palette that you like better and for those colors that will be hard for eyeballs like mine to distinguish, we can either tweak the colors or add diagnonal lines (ggpattern::geom_bar_pattern).
+Roland says: Maybe try a palette that you like better and mention that order of bars (left to right) is the same as in the legend (top to bottom).
+In addition, if you think it is necessary, for those colors that will be hard for eyeballs like mine to distinguish, we can either tweak the colors or add diagnonal lines (ggpattern::geom_bar_pattern).
 -->
@@
 ![Results from the exome capture selection study.
-(A) Map of sampling locations in the Sierra Nevada study area showing Bd-naive populations in red and persistent/recovering populations in blue.
+(A) Map of sampling locations in the Sierra Nevada study area showing Bd-naive populations in red and recovering populations in blue.
 Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) are shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada is shown in light gray.
@@
 (C) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
-Dashed line shows Bonferroni corrected p-value for threshold of 0.01.
+Dashed line shows Bonferroni corrected p-value for threshold of 0.05 (liberal) and 0.01 (stringent) outliers <!--# Add line for 0.05, if that distinction would be useful.
+-->.
 (D) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
@@
 Red line shows Bonferroni corrected p-value for threshold of 0.01.
-Outlier SNPs above this threshold are highlighted and labeled.
+<!--# Show line for threshold of 0.05 also?
+-->Outlier SNPs above this threshold are highlighted and labeled.
 Bold labels indicate at least one non-synonymous SNP is present in that gene.
 <!--# Map (panel A) is duplicated here and in the mapped allele frequency figure.
-Remove from this figure?
+I think it can be removed from this figure.
+Capitalize panel labels.
+Label/identify both the liberal and stringent outlier sets?
 -->](figures/map_qqplot_manhattan.jpg){#fig-selectionresults}
@@
 
-![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
 
@@
 
-#### Table files
+### Datasets
 
--   gemma_outliers.csv
+Dataset 1: gemma_outliers.csv
 
--   spline_window_shared_outliers.csv
-
-<!--# Include table of translocation sites, years, frog numbers? -->
+Dataset 2: spline_window_shared_outliers.csv
 

### 2023-03-31 doc/manuscript/translocation.qmd
Comment on and clarify genomics results

--- 2023-03-30 doc/manuscript/translocation.qmd
@@
 
-To assess whether frog populations display an evolutionary response to Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Our goal was to identify putative genes under selection and potentially related to increased frog resistance to Bd.
-Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
-These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+To assess whether frog populations have displayed an evolutionary response following Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
+Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("persistent"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
 <!--# Modify text/remove text here/elsewhere to avoid repetition of this.
@@
 As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
+
+To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
+Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
+Mark here:  I am actually liking the ordering of the objectives now.
 -->.
-We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
+Finally, we assessed whether putative genes or genomics regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
+<!--# Could we move the following sentence to the methods?
+--> We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
 
-The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
+Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
+The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Is this true?
+-->.
+We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
 The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
-One gene in this group does not have an associated annotation.
-No biological functions, molecular functions, or cellular processes were significantly overrepresented in the set of 29 annotated outlier genes.
 The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
@@
 However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
+No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.
 
-In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
+Our second analysis used a splined window to identify regions of the genome under selection.
+We found regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
 -->.
 We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers in both analyses and two of these shared regions also contained outlier SNPs.
+Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
 A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (Table 3: SI <!--# add this table -->).
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
+Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) ([#fig-spline-manhattan](#fig-spline-manhattan) D).
@@
 In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.
+Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
 
-<!--# Need summary of importance of genetic results, also need better integration of different selection  results -->
+Together, our genomics results show that persistent populations have strong signatures of selection on multiple genes and regions of the genome.
+While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in persistent MYL frog populations.
+
+<!--# Need summary of importance of genetic results, also need better integration of different selection  results.
+Mark W: I took a crack at this to test my naive understanding of the genetic results.
+Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly.
+-->
 
@@
 Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
-<!--# Mark here: Just figured I would point this out.
-This is trending in the right direction, but there isn't strong statistical evidence that these two proportions differ.
-I tried to phrase this carefully.
--->
 
@@
 We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$) <!--# Why 2 distributions?
-When was one used versus the other?
-Or are these 2 distributions equivalent?
--->.
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
 Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.

### 2023-03-31 doc/manuscript/translocation.qmd
Expand on key points in the discussion

@@
 "Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
-In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]<!--# The previous sentence might fit better in the Discussion when we take a deeper dive into evolutionary rescue -->.
 Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
@@
 
-In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
+In the current study, we combine results from novel <!--# better adjective than "novel".
+Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
 Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
+Mark here:  This transition reads fine to me.
 -->.
 We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
-These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
+We use these data to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
 Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
@@
 Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
-The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Is this true?
--->.
+The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Is this true? -->.
 We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
@@
 
-1.
-Summary of all of the results in a way that integrates the different components/sections (selection, translocations, viability), and makes the pitch for how unique the results are (as per comment from Bree):
-    -   Evidence for evolutionary rescue in a conservation-relevant species.
-    -   Successful reestablishment of extirpated populations via translocation despite ongoing Bd infection.
-    -   Evidence for long-term viability of reestablished populations.
+Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need across systems.
+One such strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
+While this strategy has been previously proposed, it has never been successfully implemented.
+Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of some populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
+Leveraging individuals from these resistant and rescued populations, we use 15 years of translocation effort and population viability modeling to demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
+These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be leveraged to successfully recover a species in the presence of disease.
 
-<!-- -->
+While conclusively demonstrating that evolutionary rescue is driving population recoveries is challenging.
+we provide four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
+First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
+At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain long-lasting immunity to Bd and are susceptible to re-infection and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
+Thus, MYL frog-Bd systems have the epidemiological characteristics necessary for evolutionary rescue to occur.
+Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
+These differences are consistent with strong signatures of selection and regions of the genome under selection are associated with known mechanisms of resistance to Bd infection.
+Third, our translocations allowed us to conclude that observed resistance and recovery of resistant populations is not strictly due to local environmental conditions (though environment does play a role).
+Changes in population growth rate due to evolved resistance extend beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
+Finally, evolved resistance has a quantifiable phenotype.
+In persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), our study is unique in that we show that evolved resistance is associated with population recovery in both natal and non-natal habitats, providing an actionable strategy for promoting the long-term viability of a species.
 
-2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.
-    -   Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
-    -   Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
-    -   There is a shared selective response across recovering populations.
-    -   Changes in population growth rate due to evolution extend beyond the natal environment.
-    -   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
+Hundreds of declining species, but mostly failed attempts at recovery.
+Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
+In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
 

### 2023-03-31 doc/manuscript/translocation.qmd

--- 2023-03-31 doc/manuscript/translocation.qmd
@@
 
-In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
+In the current study, we combine results from novel <!--# better adjective than "novel".
+Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
 Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
+Mark here:  This transition reads fine to me.
 -->.
-We then reintroduced these resistant frogs to 24 sites and described the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
 These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
@@
 
-To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
-These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+To assess whether frog populations have displayed an evolutionary response following Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
+Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("persistent"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
 <!--# Modify text/remove text here/elsewhere to avoid repetition of this.
@@
 As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window analysis to evaluate larger genomic regions showing differentiation between population types.
-For single SNPs, we identified outlier variants using a Bonferroni-corrected significance level (liberal = 0.05, stringent = 0.01), and for the splined window analysis we used a W-statistic of 4 standard deviations from the mean.
-<!--# This is different from the definition in the Methods: 4 * standard deviation -->
 
-The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
+To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
+Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
+Mark here:  I am actually liking the ordering of the objectives now.
+-->.
+Finally, we assessed whether putative genes or genomics regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
+<!--# Could we move the following sentence to the methods?
+--> We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
+
+Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
+The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark wrote this...confirm validity -->.
+We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
 The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
-One gene in this group does not have an associated annotation.
-No biological functions, molecular functions, or cellular processes <!--# Are these the "GO terms" referred to in the next paragraph?
-If so, add here paranthetically.
---> were significantly overrepresented in the set of 29 annotated outlier genes.
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
-For these SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
+For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
 Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
 However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
+No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.
 
-In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both *F~ST~* and $\pi_{diff}$ ("shared" regions) and 2 of the 9 shared regions also contained the outlier SNPs described above <!--# SNPs from the liberal or strict set?
+Our second analysis used a splined window to identify regions of the genome under selection.
+We found regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
+-->.
+We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
+<!--# SNPs from the liberal or strict set?
 Also "windows" and "regions" are used interchangeably in this paragraph.
@@
 A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
+Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
@@
 In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.
+Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+
+Together, our genomics results show that persistent populations have strong signatures of selection on multiple genes and regions of the genome.
+While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in persistent MYL frog populations.
+
+<!--# Need summary of importance of genetic results, also need better integration of different selection  results.
+Mark W: I took a crack at this to test my naive understanding of the genetic results.
+Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly.
+-->
+
 <!--# Based on Fig.
@@
 Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
-The multivariate linear mixed model was implemented in the program GEMMA [@zhou2014].
+ <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
+For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
+The first approach A, did X, Y, and Z but had C limitations.
+The second approach B, did X, Y, and Z and D limitations."-->.
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
-For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
-GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
-We calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
-Results were visualized using a Manhattan plot and qqplot.
+For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
+GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?
+Probably obvious for genomics folks, but not totally clear to me.
+I am guessing we are testing whether the `population` coefficient (naive vs.
+persistent) is significant for describing variability in a SNP?
+--> for the set of 148,307 high quality SNPs across 127 contigs.
+We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
+<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs.
+persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.
 
-We conducted the splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
-First, we calculated per-site *F~ST~* between individuals from naïve and recovering populations for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+To detect regions of the genome under selection <!--# Improve wording.
+Also, can we contrast this with the previous approach?
+What is this analysis doing that the other one didn't?
+-->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
+First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+
 We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.

### 2023-03-31 doc/manuscript/translocation.qmd
Additional edits to Methods and figure legends

@@
 "Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
+In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]<!--# The previous sentence might fit better in the Discussion when we take a deeper dive into evolutionary rescue -->.
 Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
@@
 
-In the current study, we combine results from novel <!--# better adjective than "novel".
-Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
+In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
 Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
-Mark here:  This transition reads fine to me.
 -->.
-We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+We then reintroduced these resistant frogs to 24 sites and described the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
 These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
@@
 
-To assess whether frog populations have displayed an evolutionary response following Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
-Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("persistent"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
-These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
+To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
 <!--# Modify text/remove text here/elsewhere to avoid repetition of this.
@@
 As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window analysis to evaluate larger genomic regions showing differentiation between population types.
+For single SNPs, we identified outlier variants using a Bonferroni-corrected significance level (liberal = 0.05, stringent = 0.01), and for the splined window analysis we used a W-statistic of 4 standard deviations from the mean.
+<!--# This is different from the definition in the Methods: 4 * standard deviation -->
 
-To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
-Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
-Mark here:  I am actually liking the ordering of the objectives now.
--->.
-Finally, we assessed whether putative genes or genomics regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
-<!--# Could we move the following sentence to the methods?
---> We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
-
-Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
-The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark wrote this...confirm validity -->.
-We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
+The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
 The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
-For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
+One gene in this group does not have an associated annotation.
+No biological functions, molecular functions, or cellular processes <!--# Are these the "GO terms" referred to in the next paragraph?
+If so, add here paranthetically.
+--> were significantly overrepresented in the set of 29 annotated outlier genes.
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
+For these SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
 Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
 However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
-No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.
 
-Our second analysis used a splined window to identify regions of the genome under selection.
-We found regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
--->.
-We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
-<!--# SNPs from the liberal or strict set?
+In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both *F~ST~* and $\pi_{diff}$ ("shared" regions) and 2 of the 9 shared regions also contained the outlier SNPs described above <!--# SNPs from the liberal or strict set?
 Also "windows" and "regions" are used interchangeably in this paragraph.
@@
 A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
+Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
@@
 In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
-<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.
-Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
-
-Together, our genomics results show that persistent populations have strong signatures of selection on multiple genes and regions of the genome.
-While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in persistent MYL frog populations.
-
-<!--# Need summary of importance of genetic results, also need better integration of different selection  results.
-Mark W: I took a crack at this to test my naive understanding of the genetic results.
-Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly.
--->
-
+Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
 <!--# Based on Fig.
@@
 A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
-<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focursed on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., sensitivity analysis, etc.).
+<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focused on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., in the sensitivity analysis, etc.).
 It is my understanding that you are generally assuming that these two probabilities are equivalent and interchangable, but additional clarification would be helpful.
-Might also require some clarification in Table 2.-->
+Might also require some clarification in Table 2.
+You describe this in more detail in the Methods, but it seems that some additional clarification might be helpful here.
+-->
 
@@
 Clarify.
+I know from the Methods that you developed different versions of the model, some deterministic and some stochastic, but some clarification might be useful here too.
 -->predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
 Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
-<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less then 1", "some probability of persistence".
-In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are much less than 1.
+<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less than 1", "some probability of persistence".
+In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are substantially less than 1.
 Consider tweaking wording to clarify this.
@@
 Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
+<!--# Mark here: Just figured I would point this out.
+This is trending in the right direction, but there isn't strong statistical evidence that these two proportions differ.
+I tried to phrase this carefully.
+-->
 
@@
 
-Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need across systems.
-One such strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
-While this strategy has been previously proposed, it has never been successfully implemented.
-Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of some populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
-Leveraging individuals from these resistant and rescued populations, we use 15 years of translocation effort and population viability modeling to demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
-These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be leveraged to successfully recover a species in the presence of disease.
+1.
+Summary of all of the results in a way that integrates the different components/sections (selection, translocations, viability), and makes the pitch for how unique the results are (as per comment from Bree):
+    -   Evidence for evolutionary rescue in a conservation-relevant species.
+    -   Successful reestablishment of extirpated populations via translocation despite ongoing Bd infection.
+    -   Evidence for long-term viability of reestablished populations.
 
-While conclusively demonstrating that evolutionary rescue is driving population recoveries is challenging.
-we provide four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
-First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain long-lasting immunity to Bd and are susceptible to re-infection and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
-Thus, MYL frog-Bd systems have the epidemiological characteristics necessary for evolutionary rescue to occur.
-Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
-These differences are consistent with strong signatures of selection and regions of the genome under selection are associated with known mechanisms of resistance to Bd infection.
-Third, our translocations allowed us to conclude that observed resistance and recovery of resistant populations is not strictly due to local environmental conditions (though environment does play a role).
-Changes in population growth rate due to evolved resistance extend beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
-Finally, evolved resistance has a quantifiable phenotype.
-In persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), our study is unique in that we show that evolved resistance is associated with population recovery in both natal and non-natal habitats, providing an actionable strategy for promoting the long-term viability of a species.
+<!-- -->
 
-Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
-Hundreds of declining species, but mostly failed attempts at recovery.
-Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
-In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
+2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.
+    -   Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
+    -   Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
+    -   There is a shared selective response across recovering populations.
+    -   Changes in population growth rate due to evolution extend beyond the natal environment.
+    -   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
 
@@
 Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
- <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
-For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
-The first approach A, did X, Y, and Z but had C limitations.
-The second approach B, did X, Y, and Z and D limitations."-->.
+The multivariate linear mixed model was implemented in the program GEMMA [@zhou2014].
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
-For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
-GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?
-Probably obvious for genomics folks, but not totally clear to me.
-I am guessing we are testing whether the `population` coefficient (naive vs.
-persistent) is significant for describing variability in a SNP?
---> for the set of 148,307 high quality SNPs across 127 contigs.
-We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
-<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs.
-persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.
+For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
+GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
+We calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
+Results were visualized using a Manhattan plot and qqplot.
 
-To detect regions of the genome under selection <!--# Improve wording.
-Also, can we contrast this with the previous approach?
-What is this analysis doing that the other one didn't?
--->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
-First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-
+We conducted the splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
+First, we calculated per-site *F~ST~* between individuals from naïve and recovering populations for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
 We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
@@
 
-Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
+Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species (@fig-yosemap).
 We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], as well as landscape genetic information [@poorten2017].
@@
 sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd infection intensity ("load", expressed as median by site: `r round(bdload_min, 2)`--`r round(bdload_max, 2)`) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 ITS copies per swab)[@vredenburg2010; @joseph2018].
+As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd infection intensity ("load", expressed as median log~10~(load) by site: `r round(bdload_min, 2)`--`r round(bdload_max, 2)`) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 ITS copies per swab)[@vredenburg2010; @joseph2018].
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
-However, when the number of frogs detected on the first survey day was low, we conducted only a single-day CMR survey.
-Within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
+However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
+When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
 Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to construct CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
+Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
 
@@
 To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
-We performed all analyses using the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022] .
-We used default, weakly informative priors.
+We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
 Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
@@
 
-To quantify changes in Bd load, we used a workflow similar to that described above (in "Predictors of post-translocation frog survival").
+To quantify changes in Bd load, we used an analysis workflow similar to that described above.
 Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
@@
 
-To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for *R. sierrae*.
-Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of an *R.
-sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$).
+To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for MYL frogs.
+Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
+We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
 We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] : SI).
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$, where the "$\cdot$" indicates a particular state variable, the probability that a female frog reproduces in a given year $p_F$, the number of eggs produced by a female frog in a year that successfully hatch $F$, the probability of a first year tadpole remaining as a tadpole $p_{L1}$, the probability of a second year tadpole remaining as a tadpole $p_{L2}$, and the probability of a first year juvenile remaining as a juvenile $p_{J1}$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ ("$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
 The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
+<!--# Can the omega parameter be added to the list of other parameters: "..., probability of a first-year juvenile remaining as a juvenile pj1, and probability of a juvenile successfully recruiting to an adult omega." Listing omega separately seems unnecessary, at least to me.
+-->
 
@@
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
-However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability about a mean value.
+However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
 Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
@@
 Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
-For populations where we had a sufficient number of pit-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
-For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
+For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
 
@@
 We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or a beta distribution with mean 0.5 and $\phi = 2$) <!--# Why 2 distributions?
+When was one used versus the other?
+Or are these 2 distributions equivalent?
+-->.
 Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
@@
 In legend, provide interpretation of "Homozygous reference", "Homozygous alternate", and "Heterozygous".
-Try moving inset map to lower left portion of panel - more room there to enlarge inset?
 How about identifying each population as R.
 muscosa or R.
-sierrae using squares and dots?
+sierrae using squares and dots, respectively?
+The current version of the figure ignores the species identities, but including them might help with interpretation of results.
+Try moving inset map to lower left portion of panel - more room there to enlarge inset?
+Remove distance scale from inset map.
 --> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful.
@@
 Outlier windows for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier windows for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from A showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from B showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
+(C) Magnified Contig19 from (A) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
 This region of the genome contains 10 annotated genes known to occur in the extended MHC Class I and II regions.
-<!--# As I commented in the Results text, why where the regions highlighted in C and D chosen?
+<!--# As I commented in the Results text, numerous outlier regions were identified, so how and why were the regions highlighted in C and D chosen?
+Legend text that describes the outlier categories needs additional editing to clarify the categories.
+My attempt feels only partially successful.
+Capitalize plot identifiers (A, B, C, D).
 -->](figures/spline_manhattan_figure.jpg){#fig-spline-manhattan}
@@
 
-![A population viability analysis of translocated *R.
-sierrae* that quantifies the long-run growth rate and extinction probability of frog populations: (A) The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R.
-sierrae*.
+![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated frogs.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
-(B) The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to A.
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.3$, for comparison to (A).
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
+<!--# Remove period after plot labels, e.g., "A", not "A.".
+-->](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
 
-![Sensitivity analysis and empirical validation of the stage-structured *R.
-sierrae* model.
-(A) The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+![Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
+(A) Elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch) <!--# Necessary to define F (when sigma and omega are not defined)?
+Maybe instead just cite "Table 2 SI"?) -->.
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
-(B) The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+
+<!--# Remove period after plot labels -->
 
@@
 (B) PCA calculated from binary SNPs showing genomic relationship of samples.
-Numeric labels match those in (A) and are colored as in (A).
+Numeric labels and colors match those in (A).
 (C) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
@@
 
-![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon).
+![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by black polygon).
 Symbol shapes indicate the donor population used for each translocation site.
@@
 ![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
+A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
 
-![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
+Box plot details are as in @fig-cond-effects.
+](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
 
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+Variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
-![A comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
+![Comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
 In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years.
@@
 <!--# Use 95% uncertainty intervals for consistency?
+Also, use dark blue instead of gray for naturally-recruited?
+Might show up better.
 -->](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}

### 2023-03-31 doc/manuscript/translocation.qmd
Explore structures for opening paragraphs in discussion

--- 2023-03-31 doc/manuscript/translocation.qmd
@@
 Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
-<!--# SNPs from the liberal or strict set?
-Also "windows" and "regions" are used interchangeably in this paragraph.
-To avoid confusion, use "windows" throughout?
--->.
+<!--# SNPs from the liberal or strict set? Also "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? -->.
 A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
@@
 
-While conclusively demonstrating that evolutionary rescue is driving population recoveries is challenging.
-we provide four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
+While evolutionary rescue is not the only mechanism that can lead to host recovery in the presence of disease (Brannely), there are four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
 First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
@@
 Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
-Hundreds of declining species, but mostly failed attempts at recovery.
+
+-   Hundreds of declining species, but mostly failed attempts at recovery.
     Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.

### 2023-03-31 doc/manuscript/translocation.qmd
Add ideas to discussion

@@
 Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
-The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark wrote this...confirm validity -->.
+The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark W wrote this...confirm validity -->.
 We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
@@
 A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
-<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focused on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., in the sensitivity analysis, etc.).
-It is my understanding that you are generally assuming that these two probabilities are equivalent and interchangable, but additional clarification would be helpful.
-Might also require some clarification in Table 2.
-You describe this in more detail in the Methods, but it seems that some additional clarification might be helpful here.
--->
+Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \>5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
 
-Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model <!--# Is this a different model, now with demographic stochasiticity and environmental variability in juvenile recruitment, or the same model used to estimate lamda in the previous paragraph?
-Clarify.
-I know from the Methods that you developed different versions of the model, some deterministic and some stochastic, but some clarification might be useful here too.
--->predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
-Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
-<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less than 1", "some probability of persistence".
-In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are substantially less than 1.
-Consider tweaking wording to clarify this.
-I'd argue for a relatively strong initial statement that we then fine tune only slightly, instead of a weak statement that we then strengthen a lot.
-Initial impressions can be hard to change.
---> For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5 <!--# Based on Fig.
-5, looks like extinction propability is < 0.5 for all six populations only for omega > ~0.22.
-Reword as necessary.
--->, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
+We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
+Our model predicted that, following a single introduction of 40 adult individuals into a population, three of six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 with our most conservative estimate of $\omega = 0.15$ (@fig-viability B).
+As $\omega$ increased from 0.15 to 0.3 <!--# 0.3 is the vertical line on the figure, so I think it makes sense to keep this upper bound in the statement.
+-->, 50 year extinction probability was less 0.5 for all six populations indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-Moreover, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
+This is likely a conservative estimate because there is evidence that naturally recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
+Finally, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
 Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
-
-<!--# Is it worth mentioning that predictions may be conservative because yearly survival probability of naturally recruited animals may be higher than yearly survival probability of the translocated animals that was used to parameterize model (Figure 12).
--->
 
@@
 
-Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need across systems.
-One such strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
-While this strategy has been previously proposed, it has never been successfully implemented.
-Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of some populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
-Leveraging individuals from these resistant and rescued populations, we use 15 years of translocation effort and population viability modeling to demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
-These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be leveraged to successfully recover a species in the presence of disease.
+Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need.
+One potential strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
+While previously proposed (Mendelson), this strategy has never been successfully implemented in wildlife disease systems.
+Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
+Leveraging individuals from these resistant, rescued populations, 15 years of translocation effort and associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
+These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be used to successfully recover a species in the presence of disease.
 
-While evolutionary rescue is not the only mechanism that can lead to host recovery in the presence of disease (Brannely), there are four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
-First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain long-lasting immunity to Bd and are susceptible to re-infection and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
-Thus, MYL frog-Bd systems have the epidemiological characteristics necessary for evolutionary rescue to occur.
+Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is the mechanism of recovery.
+In MYL frog systems there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
+First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
+At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ) and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
 Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
-These differences are consistent with strong signatures of selection and regions of the genome under selection are associated with known mechanisms of resistance to Bd infection.
-Third, our translocations allowed us to conclude that observed resistance and recovery of resistant populations is not strictly due to local environmental conditions (though environment does play a role).
-Changes in population growth rate due to evolved resistance extend beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
-Finally, evolved resistance has a quantifiable phenotype.
-In persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), our study is unique in that we show that evolved resistance is associated with population recovery in both natal and non-natal habitats, providing an actionable strategy for promoting the long-term viability of a species.
+These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses \[vague\].
+Third, our translocations and long-term monitoring have allowed us to demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
+Finally, resistance can be statistically quantified --- in persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
+While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host defenses and subsequent host recovery.
+Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
 
-Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
+The ability to harness evolutionary rescue to promote species recovery across a landscape of disease has critical implications for the conservation for MYL frogs.
 
--   Hundreds of declining species, but mostly failed attempts at recovery.
-    Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
-    In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
+-   \[Elaborate on this a bit more\]
+
+-   Despite relatively high success rate in MYL frogs, it is apparent that not all such recovery efforts will succeed.
+    In addition to ongoing effects of Bd, several deterministic and stochastic factors likely also play an important role: (1) effects of elevation, winter severity, and donor population on frog survival, and (2) variability in recruitment driven by Bd-caused mortality of subadults.
+
+-   Now have a means of reestablishing MYL frog populations across their historical range, including in areas in which natural recolonization is precluded by dispersal limitation.
+    Nonetheless, recovery across this large landscape will require substantial resources over many decades.
+
+The recovery of other amphibian species impacted by Bd will continue to be challenging.
+
+-   For example, despite hundreds of declining amphibian species, most attempts at treatments and recovery in the presence of Bd have failed \[citations\].
+
+-   *In-situ* treatments such as the application of fungicides, probiotics, or habitat manipulations (Bosch), have shown some case-by-case successes, but generally lack the necessary effectiveness and scalability to promote large-scale recovery.
+
+-   Natural recoveries are occurring in the presence of Bd in other amphibian species (Newell, Tobin, Scheele, Voyles) and assisting natural recoveries may provide the most effective means for large-scale reocvery efforts.
+
+-   However, recoveries in other systems are not necessarily driven by the evolution of host defense (Brannley) and underlying mechanisms of recovery in other amphibian-Bd systems may be less translatable from a single population to the landscape-level than we observe in MYL frogs.
+    For example, if the local biotic (e.g., community interactions) and abiotic (e.g., temperature) environment strongly affect recovery dynamics and there is strong heterogeneity in these conditions across habitat patches on a landscape, then recovery in a natal habitat will be less likely to translate into recovery in a non-natal habitat.
+
+-   The MYL frog system is unique in that low productivity due to short summers provide a homogenizing effect across habitat patches, potentially minimizing the effect of the local environment on landscape-level recovery compared to what we might expect in higher productivity systems (for example, tropical amphibian assemblages) <!--# Roland, I am just brainstorming here and this might not make sense.
+We can delete all of this if necessary.
+--> .
+    That being said, even in this low productivity system we still observed a strong site and environmental effect on translocated adult survival.
+
+-   Thus, the local environment will be a key consideration when assessing the feasibility of assisted recovery in other amphibian systems.
+
+What does this study tell us more broadly about disease-induced recovery in other systems and the conservation of biodiversity?
+
+-   Identifying the mechanisms of recovery in the face of disease is huge knowledge gap.
+
+<!--# Previous text from Roland below -->
 
@@
 ![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-Box plot details are as in @fig-cond-effects.
-](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
 

### 2023-04-01 doc/manuscript/translocation.qmd
Edits to clean up recent merges

@@
 
-In the current study, we combine results from novel <!--# better adjective than "novel".
-Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
+In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
 Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
@@
 
-To assess whether frog populations have displayed an evolutionary response following Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+To assess whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
 Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
-Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("persistent"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
 These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
-<!--# Modify text/remove text here/elsewhere to avoid repetition of this.
---> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
+Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
 Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
@@
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
-Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives?
-Mark here:  I am actually liking the ordering of the objectives now.
+We identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
+Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
+In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) <!--# This is a different definition than the one given in the Methods: "4 * standard deviation".
 -->.
-Finally, we assessed whether putative genes or genomics regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
-<!--# Could we move the following sentence to the methods?
---> We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).
+Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
 
-Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
-The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark W wrote this...confirm validity -->.
-We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
-The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
-For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
-Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
-However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
+Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI) and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
+Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
+The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
+A single gene in both groups did not have an associated annotation.
+For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion?
+:  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
+-->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
 No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.
 
-Our second analysis used a splined window to identify regions of the genome under selection.
-We found regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff?
--->.
+Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
 We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
@@
 A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
@@
 
-Together, our genomics results show that persistent populations have strong signatures of selection on multiple genes and regions of the genome.
-While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in persistent MYL frog populations.
+Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
+While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in recovering MYL frog populations.
 
@@
 Our model predicted that, following a single introduction of 40 adult individuals into a population, three of six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 with our most conservative estimate of $\omega = 0.15$ (@fig-viability B).
-As $\omega$ increased from 0.15 to 0.3 <!--# 0.3 is the vertical line on the figure, so I think it makes sense to keep this upper bound in the statement.
--->, 50 year extinction probability was less 0.5 for all six populations indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
+As $\omega$ increased from 0.15 to 0.3, 50 year extinction probability was less 0.5 for all six populations, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 
-Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need.
+Disease-induced declines are a central conservation challenge for wildlife, and identifying biologically grounded strategies for assisted recovery is a critical need.
 One potential strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
@@
 <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
-For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
+For example, for us genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
 The first approach A, did X, Y, and Z but had C limitations.
@@
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
-For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
+For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
 GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?
@@
 
-To detect regions of the genome under selection <!--# Improve wording.
+To detect regions of the genome under selection <!--# Improve wording to indicate that this is the second analysis that we conducted to detect regions of the genome under selection.
 Also, can we contrast this with the previous approach?
@@
 -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
-First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+First, we calculated per-site *F~ST~* between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
 Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
 
-We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+We then used the package WINGEN R package [@bishop2023] <!--# Correct citation?
+No citation was provided in the original text from Allie so I guessed at what a suitable citation would be.
+--> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
 This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
@@
 
-<!--# Standardize panel labels. I suggest capital letters with no punctuation, e.g., "A", not "A.". -->
-
 ![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the stringent set of 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.

### 2023-04-02 doc/manuscript/translocation.qmd
Minor edits throughout Intro and Results

@@
 csl: pnas.csl
-format: pdf
+format: docx
 editor: visual
@@
 Endangered Species Act.
-Although this decline was initiated by the introduction of non-native trout into the once-fishless portion of the Sierra Nevada [@bradford1989; @knapp2000], the arrival of *B.
+In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B.
 dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
@@
 -->.
-We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
 These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
@@
 
-To assess whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
+To determine whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
 Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
@@
 To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
-We identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
+For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
 Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
@@
 -->.
-Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
+Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.
 
-Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI) and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
+<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here.
+-->Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
 Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
@@
 This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
-No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.
+No GO terms were significantly over-represented in the set of 29 annotated outlier genes.
 
 Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.
-<!--# SNPs from the liberal or strict set? Also "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? -->.
-A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph.
+To avoid confusion, use "windows" throughout?
+--> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9-->) and no GO terms were overrepresented within this set of genes.
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
@@
 In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
 <!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.
-Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove.
+Among susceptible species, survivors of epizootics bear alleles encoding these conformations."?
+What about MHC Class II characteristics linked to resistance by Bataille et al.
+(2015)?
+Savage and Zamudio (2011, 2016), Kosch et al.
+(2018)?-->
 
 Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
-While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in recovering MYL frog populations.
+While the functions of all genes and regions showing signatures of selection were not clearly interpretable, regions of the genome associated with resistant phenotypes in other amphibian systems such as the MHC complex, showed strong evidence of selection in recovering MYL frog populations.
 
@@
 2, there are numerous outlier windows.
-How were the specific windows mentioned above chosen?
+How were the specific windows mentioned in above and in Fig.
+2 C, D chosen?
 Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria.
@@
 Similarly, is the magnitude of differences consistent with strong or weak selection?
-In addition to the above, this section needs a concluding summary of the key genetic results.
 -->
@@
 
-Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies, consistent with strong selection following Bd exposure.
-Here we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog genome study (@fig-selectionresults: population 1 and 4).
-<!--# Check this --> Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
-These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
+Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
+In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
+<!--# Check this --> Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
+These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
 We refer to these reintroductions as "translocations" [@seddon2014].
@@
 We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
@@
 A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
-Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \>5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
+Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
 
 We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, three of six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 with our most conservative estimate of $\omega = 0.15$ (@fig-viability B).
-As $\omega$ increased from 0.15 to 0.3, 50 year extinction probability was less 0.5 for all six populations, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
+<!--# In regard to the previous sentence, is it worth stating that the inclusion of demographic stocahsticity and environmental variability will not change estimates of lambda?
+That might not be apparent.
+--><!--# In the following sentences, I edited the text to present the results based on omega = 0.3 first and then the results with reduced omega of 0.15.
+This matches the order in which we present the results in the previous paragraph.
+-->Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, all six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 (<!--# provide actual range here -->@fig-viability B).
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
+When we decreased $\omega$ from 0.3 to our most conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.5 to 0.X <!--# give actual range here -->for three additional populations (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
-Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
+In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
+This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
 Finally, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
@@
 
-Disease-induced declines are a central conservation challenge for wildlife, and identifying biologically grounded strategies for assisted recovery is a critical need.
-One potential strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
-While previously proposed (Mendelson), this strategy has never been successfully implemented in wildlife disease systems.
-Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
-Leveraging individuals from these resistant, rescued populations, 15 years of translocation effort and associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
-These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be used to successfully recover a species in the presence of disease.
+Disease-induced declines are a central conservation challenge for wildlife, and identifying strategies to recover affected species is a critical need.
+One potential strategy, and the one we use in the current study, leverages disease-induced genetic changes in host populations and evolutionary rescue to recover species at the landscape scale.
+Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
+Here, we provide strong evidence that natural selection for resistance and the associated evolutionary rescue has facilitated recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
+The use of individuals from these resistant, rescued populations in a 15 year translocation effort plus associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
+These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how evolutionary rescue can be used to successfully recover a species in the presence of disease.
 
-Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is the mechanism of recovery.
-In MYL frog systems there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
+Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
+In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
 First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ) and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
-Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
-These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses \[vague\].
-Third, our translocations and long-term monitoring have allowed us to demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
-Finally, resistance can be statistically quantified --- in persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
-While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host defenses and subsequent host recovery.
+At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped.
+Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery.
+-->.
+Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
+These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions.
+-->.
+Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
+Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence.
+-->.
+Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
 Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
@@
 
--   Natural recoveries are occurring in the presence of Bd in other amphibian species (Newell, Tobin, Scheele, Voyles) and assisting natural recoveries may provide the most effective means for large-scale reocvery efforts.
+-   Natural recoveries are occurring in the presence of Bd in other amphibian species (Newell, Tobin, Scheele, Voyles) and assisting natural recoveries may provide the most effective means for large-scale recovery efforts.
 
@@
     That being said, even in this low productivity system we still observed a strong site and environmental effect on translocated adult survival.
+    <!--# This is an interesting topic, but despite low productivity MYL frog habitats are highly variable: ponds, streams, meadows, alpine lakes, forested lakes, etc.
+I think we should drop this bullet.
+-->
 

### 2023-04-06 doc/manuscript/translocation.qmd
Add text and comments from Allie and Bree

@@
 Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
+"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Allie comment.
+I believe evolutionary rescue also encompasses the introduction of new variation through gene flow.
+See box 1 in ref 16 for a helpful discussion of genetic vs evolutionary rescue.
+Could delete "Originating from variation within a host population" here and just add more information about what make evolutionary rescue special - that is the increase in locally adapted phenotypes.
+Bree comment.
+Minor – there are a lot of clauses in this sentence.
+Wonder if setting this as " – as necessary for evolution rescue – " might make the sentence more readable-->.
 Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
-In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
+In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals <!--# Bree comment: Again minor, but seems like this could be slimmed (as the sentence repeats the idea of promoting recovery twice and the "provide sources of individuals" seems awkward).
+It could read something like "In addition to promoting recovery of remnant populations after disease-induced declines, natural selection can also lead to increased resistance and/or tolerance…." -->with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
 To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
@@
 There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
-First, populations that survived following disease arrival have evolved increased resistance.
+First, populations that survived <!--# Bree comment.
+Worth keeping tense consistent so that the four conditions feel parallel?
+This sentence could read "First, populations that survive following disease arrival evolve increased resistance" -->following disease arrival have evolved increased resistance.
 Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"?
@@
 Endangered Species Act.
-In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B.
+In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] <!--# Bree comment: Very minor but there are lots of places with long structurally complex sentences that seem just as good as broken up shorter sentences -->, and the arrival of *B.
 dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
 These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
-In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
+In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
 
@@
 For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
-Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
-In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) <!--# This is a different definition than the one given in the Methods: "4 * standard deviation".
+Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
+pi-diff.
 -->.
+In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overalapped for *F~ST~* and $\pi$ <!--# This is a different definition than the one given in the Methods: "4 * standard deviation".
+--> <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
+pi - diff -->.
 Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.
 
-<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here.
--->Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
+<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->
+
+Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
 Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
@@
 To avoid confusion, use "windows" throughout?
---> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9-->) and no GO terms were overrepresented within this set of genes.
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+--> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).
+-->) and no GO terms were overrepresented within this set of genes.
+The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.
+-->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1 <!--# Allie comment: This gene is even more interesting now that I've done a lit review.
+More in the discussion!
+-->).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
 Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.
+-->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
 This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
@@
 For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations.
+Roland response: I added what I hope are bridging sentences between all three sections.
 -->
@@
 And that we are lumping those together in an "evolved resistance" bin.
+Roland comment: We reworded this sentence, but probably have not fully dealt with resistance/tolerance issue.
 --> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
@@
 
-Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
+<!--# Bree comment: It seems to me that this paragraph might be better as a shorter high level compelling summary and then the subsequent sections of the Discussion go into each of the four lines of evidence in more detail.
+Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once.
+-->Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
 In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
@@
 
+<!--# The next 2 paragraphs are new text from Allie -->Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
+Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
+uninfected *Xenopus tropicalis* [@rosenblum2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
+These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
+
+Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
+Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
+Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
+Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
+Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
+But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
+If it feels like we can't justify discussing it we can cut it and just focus on the other two.
+-->.
+This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
+Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
+
 -   \[Elaborate on this a bit more\]
+
+-   <!--# Allie text -->One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
+    For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al.
+2012) however many studies have found that immune overactivation is linked to susceptibility for Bd.
+    This could be different for populations with different histories of Bd exposure.
+    Perhaps Rana muscosa has a more recent history with Bd (maybe with another lineage) than for instance Atelopus.
+    All of this has led to me think this whole question deserves another paper integrating the Atelopus and MYLF datasets, so maybe we save all this for that?
+    Open to whatever you all think is relevant to this story.
 
@@
 Exact PCR Duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.XX, GATK ref).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
+Variants were then called following GATK best practices (v.4.2.0.0, @van2020). Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
 Next, genotypes were jointly called using GenotypeGVCFs.
@@
 The first approach A, did X, Y, and Z but had C limitations.
-The second approach B, did X, Y, and Z and D limitations."-->.
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
+The second approach B, did X, Y, and Z and D limitations."--> GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
 For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
@@
 This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
+We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for F~ST~ or above/below the mean for $\pi_{diff}$.
 Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
@@
 ### References
-
-##

### 2023-04-11 doc/manuscript/translocation.qmd
Draft first three paragraphs of discussion

@@
 Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once.
--->Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
-In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
-First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
-At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped.
-Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery.
--->.
-Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
-These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions.
--->.
-Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
-Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence.
--->.
-Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
-Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
+Mark here: I have broken this paragraph up in to two and tried to keep descriptions pretty short.
+-->
 
-The ability to harness evolutionary rescue to promote species recovery across a landscape of disease has critical implications for the conservation for MYL frogs.
+We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
+First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> populations, consistent with strong selection.
+Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008a; @ellison2014] (for more information, see SI X).
+For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering population and contains eight genes associated with either the MHC Class I or Class III regions.
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
+<!--# For brevity in the discussion, I tried to a pull a few key bullets out of Allie's text. Allie and Bree: Edit and expand as you see fit. -->
 
-<!--# The next 2 paragraphs are new text from Allie -->Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
-Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
-These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
-For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
-uninfected *Xenopus tropicalis* [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
-These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
+The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistance phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
+Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
+Translocated populations maintain infection intensities below this threshold of infection intensity such infection intensity is not a significant predictor of adult survival following translocations to new environments.
+Third and finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
+Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
 
-Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
-Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
-Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
-Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
-But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
-If it feels like we can't justify discussing it we can cut it and just focus on the other two.
--->.
-This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
-Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
+Implications for recovery of Mountain yellow legged frogs
 
@@
 
--   <!--# Allie text -->One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
+-   <!--# Allie text -->
+
+    One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
     For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al.
@@
 I think we should drop this bullet.
--->
+Mark here:  -->
 
@@
 -   Identifying the mechanisms of recovery in the face of disease is huge knowledge gap.
+
+<!--# Previous text from Allie below -->
+
+<!--# The next 2 paragraphs are new text from Allie -->
+
+Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
+Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
+uninfected *Xenopus tropicalis* [@rosenblum2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
+These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
+
+Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
+Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
+Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
+Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
+Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
+But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
+If it feels like we can't justify discussing it we can cut it and just focus on the other two.
+-->.
+This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
+Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
 
@@
 Exact PCR Duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0, @van2020). Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
+Variants were then called following GATK best practices (v.4.2.0.0, @van2020).
+Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
 Next, genotypes were jointly called using GenotypeGVCFs.

### 2023-04-13 doc/manuscript/translocation.qmd
Minor edits to Discussion text

@@
 
+<!--# If we switch the order of sections in the Results, we might want to mirror that order here too: translocations > selection results & evolutionary rescue > viability modeling -->
+
 Disease-induced declines are a central conservation challenge for wildlife, and identifying strategies to recover affected species is a critical need.
@@
 We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
-First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> populations, consistent with strong selection.
+First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
 Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008a; @ellison2014] (for more information, see SI X).
-For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering population and contains eight genes associated with either the MHC Class I or Class III regions.
-These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
+These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
 Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
@@
 
-The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistance phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
+The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistant phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
 Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
-Translocated populations maintain infection intensities below this threshold of infection intensity such infection intensity is not a significant predictor of adult survival following translocations to new environments.
-Third and finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
+Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
+Third, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
 Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

### 2023-04-14 doc/manuscript/translocation.qmd
Addiitional minor edits throughout

@@
 
-<!--# Bree comment: Here or elsewhere - it feels like we need a more specific bridge from the genomics section into the translocation section.
-I think we need to be explicit about how the populations we chose for the genomics section relate to the populations used in translocations.
-Even if they aren't matched in any way, we should reiterate something about how the experimental designs comport.
-For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations.
-Roland response: I added what I hope are bridging sentences between all three sections.
--->
-
 Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
 In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
-<!--# Check this --> Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
+Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
 These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
@@
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Mention of viability is confusing because this issue isn't dealt with until the third section.
-Perhaps make the categorization based on frog survival?
--->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
 Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
-<!--# Add median Bd load, proportion below 5.8 copies threshold?
---> <!--# The Methods-Translocation section contains text that repeats similar results.
-Edit to reduce repetition.
---> <!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
--->
 
@@
 Therefore, frog resistance in non-natal habitats is maintained across generations.
+
+In addition to survival......
+
+<!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
+-->
 
@@
 First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
-Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008a; @ellison2014] (for more information, see SI X).
+Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
 For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
@@
 Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
 Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
@@
 
+<!--# How to insert >= symbol in header and table cells> -->
+
 ```{r}
-#| label: tbl-viability-earlylate
-#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of extinction over 50 years was 1 or < 1 (i.e., viability = 0 or 1, respectively)."
+#| label: tbl-survival-earlylate
+#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of survival was < 0.5 or >= 0.5."
 #| echo: FALSE
@@
   mutate(site_id = as.character(site_id),
-         viability = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), 0, 1),
+         survival = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), "< 0.5", ">= 0.5"),
          period = if_else(year < 2013, "early", "late")) %>% 
   group_by(period) %>% 
-  count(period, viability) %>% 
+  count(period, survival) %>% 
   mutate(n_period = sum(n), 
@@
   mutate(across(prop, round, 3)) %>% 
-  arrange(viability, period)
+  arrange(period, survival)
 
-  knitr::kable(viability_table, "pipe", align = "lcccr", col.names = c("Period", "Viability category", "Number of translocations", "Total translocations", "Proportion of translocations")) 
+  knitr::kable(viability_table, format = "pipe", align = "lcccc", col.names = c("Period", "Survival probability", "Number of translocations", "Total translocations", "Proportion of translocations")) 
 ```

### 2023-04-14 doc/manuscript/translocation.qmd
Added comments and text from Allie to Discussion

@@
 
-<!--# Bree comment: It seems to me that this paragraph might be better as a shorter high level compelling summary and then the subsequent sections of the Discussion go into each of the four lines of evidence in more detail.
-Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once.
-Mark here: I have broken this paragraph up in to two and tried to keep descriptions pretty short.
--->
-
 We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
@@
 
+<!--# Comment from Allie that will be helpful in making the cases that what we are seeing is evidence of strong selection/evolutionary rescue -->
+
+For the question about describing what we are seeing as evolutionary rescue - I feel comfortable and confident using that wording for two reasons:
+
+1\) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
+In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
+In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
+Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significantly outlier SNP that has an annotation.
+and
+
+2\) CORROBORATING STUDIES. For each gene mentioned above we can list many studies that have linked the genes to Bd/disease/wounds.
+In the case of GVINP1 for instance there are at least three Bd transcriptomic studies that have found this specific gene to be differentially expressed in amphibians (I currently only cite one in the paper but could add the other two: Grogan et al.
+2018, Ellison et al.
+2014a, Ellison et al.
+2014b).
+In fact, if you haven't taken a look at the Grogan paper it is really interesting because it is similar in that they are describing evolution in natural settings and they talk a lot about the GVINP1 gene.
+Even more remarkable that this gene was also linked so strongly to gill amoebic disease in salmon (explaining an astounding 20% of the resistance phenotype).
+
+<!--# End of comment -->
+
+Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
+Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
+uninfected *Xenopus tropicalis* [@rosenblum2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
+These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
+
+Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
+Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
+Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
+Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
+Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
+But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
+If it feels like we can't justify discussing it we can cut it and just focus on the other two.
+-->.
+This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
+Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
+
 Implications for recovery of Mountain yellow legged frogs
@@
 
--   <!--# Allie text -->
-
-    One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
+-   One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
     For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al.
@@
 -   Identifying the mechanisms of recovery in the face of disease is huge knowledge gap.
-
-<!--# Previous text from Allie below -->
-
-<!--# The next 2 paragraphs are new text from Allie -->
-
-Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
-Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
-These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
-For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
-uninfected *Xenopus tropicalis* [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
-These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
-
-Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
-Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
-Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
-Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
-But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
-If it feels like we can't justify discussing it we can cut it and just focus on the other two.
--->.
-This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
-Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
 

### 2023-04-14 doc/manuscript/translocation.qmd
Edits to first paragraph of Discussion, other minor changes

@@
 -->.
-In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overalapped for *F~ST~* and $\pi$ <!--# This is a different definition than the one given in the Methods: "4 * standard deviation".
---> <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
+In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overlapped for *F~ST~* and $\pi$ <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
 pi - diff -->.
@@
 -->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1 <!--# Allie comment: This gene is even more interesting now that I've done a lit review.
-More in the discussion!
--->).
+This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
@@
 
-<!--# Need summary of importance of genetic results, also need better integration of different selection results.
-Mark W: I took a crack at this to test my naive understanding of the genetic results.
-Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly.
--->
-
 <!--# Based on Fig.
@@
 Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria.
--->
-
-<!--# In reading through these results, I am not seeing a compelling theme that emerges from all of the different results, both within and between the SNP and splined window sections.
-The results show differences in various metrics between population types, but are these changes large or small?
-I lack the context to understand that.
-Similarly, is the magnitude of differences consistent with strong or weak selection?
 -->
@@
 
-Disease-induced declines are a central conservation challenge for wildlife, and identifying strategies to recover affected species is a critical need.
-One potential strategy, and the one we use in the current study, leverages disease-induced genetic changes in host populations and evolutionary rescue to recover species at the landscape scale.
+Disease-induced declines are a central challenge for the conservation of global biodiversity, but strategies to recover affected species are generally lacking.
+One potential strategy, and the one we use in the current study, leverages evolutionary rescue to recover species at the landscape scale.
 Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
-Here, we provide strong evidence that natural selection for resistance and the associated evolutionary rescue has facilitated recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
-The use of individuals from these resistant, rescued populations in a 15 year translocation effort plus associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
-These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how evolutionary rescue can be used to successfully recover a species in the presence of disease.
+Specifically, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd), and (ii) individuals from these resistant populations can be used to reestablish viable populations despite ongoing Bd infection.
+These results provide a rare, and perhaps novel, example of how evolutionary rescue can be used to successfully recover a species in the presence of disease.
 
@@
 Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
-For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
+For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions<!--# Only Class I and II are mentioned in Results -->.
 These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
@@
 Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
-Third, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
+Finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
 Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
@@
 In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
-Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significantly outlier SNP that has an annotation.
+Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significant outlier SNP that has an annotation.
 and
@@
 <!--# End of comment -->
-
-Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
-Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
-These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
-For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
-uninfected *Xenopus tropicalis* [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
-These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
-
-Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
-Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
-Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
-Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
-But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
-If it feels like we can't justify discussing it we can cut it and just focus on the other two.
--->.
-This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
-Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
 
@@
 
+### Selection Results
+
+Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
+Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
+uninfected *Xenopus tropicalis* [@rosenblum2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
+These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
+
+Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
+Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
+Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
+Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
+Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
+But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
+If it feels like we can't justify discussing it we can cut it and just focus on the other two.
+-->.
+This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
+Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
+
 ### Figures

### 2023-04-14 code/extinction_and_growth_rate_analysis.R
Update how we calculate uncertainty for translocated survival probabilities

@@ -190,7 +190,7 @@ alpha = rstan::extract(fit, "alpha")$alpha
 keep_lakes_new = c(keep_lakes)#, "average")
 plots = list()
 phi_omega = 2 # Inverse dispersion of recruitment probability
-omegas = c(0.15, 0.2, 0.25, 0.3, 0.35, 0.4)#c(0.05, 0.1, 0.15, 0.2, 0.25, 0.3)
+omegas = c(0.15, 0.2, 0.25, 0.3, 0.35, 0.4)
 extinction_at_fifty = list()
 
 # Specify whether we want to allow recruited adult survival probability to differ
@@ -219,93 +219,94 @@ for(omega in omegas){
 
 	    if(lake %in% c(keep_lakes_new)){
 
-	    cat("Working on", lake, "\n")
-	    
-	    extinction_array = array(NA, dim=c(sims, steps + 1))
-	    recruitment_array = array(NA, dim=c(sims, steps))
-	    omega_array = array(NA, dim=c(sims, steps))
-	    total_pop_array = array(NA, dim=c(sims, steps + 1))
-	    
-	    # Mu and phi parameters on beta distribution of adult survival
-	    if(lake != "average"){
+		    cat("Working on", lake, "\n")
+		    
+		    extinction_array = array(NA, dim=c(sims, steps + 1))
+		    recruitment_array = array(NA, dim=c(sims, steps))
+		    omega_array = array(NA, dim=c(sims, steps))
+		    total_pop_array = array(NA, dim=c(sims, steps + 1))
+		    
+		    # Mu and phi parameters on beta distribution of adult survival
+		    if(lake != "average"){
 
-	      muT =  median(1 / (1 + exp(-(beta0 + alpha[, l]))))
-	      lower_muT = quantile(1 / (1 + exp(-(beta0 + alpha[, l]))), c(0.05))
-	      upper_muT = quantile(1 / (1 + exp(-(beta0 + alpha[, l]))), c(0.95))
-	      
-	      if(run_type == "Recruit survival == Translocated survival"){
-	        muR = muT
-	      }
-	      else{
+		      muT =  median(1 / (1 + exp(-(beta0 + alpha[, l]))))
+		      lower_muT = quantile(1 / (1 + exp(-(beta0 + alpha[, l]))), c(0.05))
+		      upper_muT = quantile(1 / (1 + exp(-(beta0 + alpha[, l]))), c(0.95))
+		      
[…]

### 2023-04-14 code/survival_table_nontranslocated.R

@@ -53,14 +53,17 @@ convert_to_death_rates = function(surv_probs, deltat){
 
 }
 
-survival_table <- function(model, thin=1, upperlower=c(0.25, 0.75)) {
+survival_table <- function(model, thin=1, lowerupper=c(0.25, 0.75), translocated=FALSE) {
   # Take in MRMR model and return survival probabilities for pit-tagged 
-  # non-translocated individuals.
+  # individuals that are either translocated or naturall recruited
   #
   # Parameters
   # ----------
   # model : MRMR fitted model
   # thin : int, thin the bayesian chains
+  # lowerupper : vector, the lower and upper bounds for survival estimates
+  # translocated : bool, if TRUE only focus on translocated individuals, 
+  #               else focus on naturally recruited individuals
   #
   # Returns
   # -------
@@ -88,16 +91,32 @@ survival_table <- function(model, thin=1, upperlower=c(0.25, 0.75)) {
       get_numeric_indices(.data$name),
       primary_period = .data$index_2,
       pit_tag_id = dimnames(model$data$stan_d$Y)[[1]][.data$index_1]
-    ) %>%
-    filter(!(.data$pit_tag_id %in% as.character(model$data$translocations$pit_tag_id))) %>%
-    left_join(primary_period_dates) %>%
-    dplyr::transmute(
-      .data$.draw, .data$value, .data$primary_period, .data$pit_tag_id,
-      .data$date, .data$year
-    ) 
+    )
 
-  # Drop all augmented data and just extract survival probabiliites from
-  # pit tagged, non-translocated individuals.
+
+  if (translocated) {
+
+    # Get individuals that were translocated
+    survival_summary <- survival_summary %>% filter(.data$pit_tag_id %in% as.character(model$data$translocations$pit_tag_id)) %>%
+      left_join(primary_period_dates) %>%
+      dplyr::transmute(
+        .data$.draw, .data$value, .data$primary_period, .data$pit_tag_id,
+        .data$date, .data$year
+      ) 
+
+  } else {
+
+    # Get individuals that weren't translocated and were naturally recruited
+    survival_summary <- survival_summary %>% filter(!(.data$pit_tag_id %in% as.character(model$data$translocations$pit_tag_id))) %>%
+      left_join(primary_period_dates) %>%
+      dplyr::transmute(
+        .data$.draw, .data$value, .data$primary_period, .data$pit_tag_id,
+        .data$date, .data$year
+      ) 
+
+  }
+
[…]

### 2023-04-14 doc/manuscript/translocation.qmd
Address comments on population viability analyses

--- 2023-04-14 doc/manuscript/translocation.qmd
@@
 
+<!--# Bree comment: Here or elsewhere - it feels like we need a more specific bridge from the genomics section into the translocation section.
+I think we need to be explicit about how the populations we chose for the genomics section relate to the populations used in translocations.
+Even if they aren't matched in any way, we should reiterate something about how the experimental designs comport.
+For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations.
+Roland response: I added what I hope are bridging sentences between all three sections.
+-->
+
 Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
 In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
-Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
+<!--# Check this --> Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
 These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
@@
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-viability-earlylate SI) <!--# Mention of viability is confusing because this issue isn't dealt with until the third section.
+Perhaps make the categorization based on frog survival?
+-->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
 Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
+<!--# Add median Bd load, proportion below 5.8 copies threshold?
+--> <!--# The Methods-Translocation section contains text that repeats similar results.
+Edit to reduce repetition.
+--> <!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
+-->
 
@@
 
-In addition to survival......
-
-<!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
--->
-
 #### Population viability modeling
@@
 We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
-<!--# In regard to the previous sentence, is it worth stating that the inclusion of demographic stocahsticity and environmental variability will not change estimates of lambda?
-That might not be apparent.
---><!--# In the following sentences, I edited the text to present the results based on omega = 0.3 first and then the results with reduced omega of 0.15.
-This matches the order in which we present the results in the previous paragraph.
--->Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, all six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 (<!--# provide actual range here -->@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, all six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 (0 - 0.32; \@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
-When we decreased $\omega$ from 0.3 to our most conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.5 to 0.X <!--# give actual range here -->for three additional populations (@fig-viability B).
+When we decreased $\omega$ from 0.3 to our most conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 - 0.94 for three additional populations (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
+
 In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
@@
 ### Discussion
-
-<!--# If we switch the order of sections in the Results, we might want to mirror that order here too: translocations > selection results & evolutionary rescue > viability modeling -->
 
@@
 We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
-First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
-Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
-For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
-These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
+First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> populations, consistent with strong selection.
+Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008a; @ellison2014] (for more information, see SI X).
+For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering population and contains eight genes associated with either the MHC Class I or Class III regions.
+These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
 Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
@@
 
-The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistant phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
+The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistance phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
 Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
 Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
-Third, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
+Third and finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
 Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
@@
 Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
+Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
 Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ ("$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
-The parameter $\omega$ gives the probability of a juvenile successfully recruiting to an adult.
-<!--# Can the omega parameter be added to the list of other parameters: "..., probability of a first-year juvenile remaining as a juvenile pj1, and probability of a juvenile successfully recruiting to an adult omega." Listing omega separately seems unnecessary, at least to me.
--->
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ ("$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$., and the probability of a juvenile successfully recruiting to an adult $\omega$.
 
@@
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
+The red line shows where $\lambda = 1$.
 (B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
 The dashed line shows where $\omega = 0.3$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
-<!--# Remove period after plot labels, e.g., "A", not "A.".
--->](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability}
 
 ![Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
-(A) Elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch) <!--# Necessary to define F (when sigma and omega are not defined)?
-Maybe instead just cite "Table 2 SI"?) -->.
+(A) Elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
 (B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
-
-<!--# Remove period after plot labels -->
-
-<!--# Left panel color palette is 12-color color-blind friendly, but is not the most friendly color palette to look at.
-We can discuss.
-Roland says: Maybe try a palette that you like better and mention that order of bars (left to right) is the same as in the legend (top to bottom).
-In addition, if you think it is necessary, for those colors that will be hard for eyeballs like mine to distinguish, we can either tweak the colors or add diagnonal lines (ggpattern::geom_bar_pattern).
--->
 
@@
 ![Comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years.
-Points are median estimates and error bars give the 90% uncertainty intervals around the estimates.
-All estimates are derived from the mrmr package.
-<!--# Use 95% uncertainty intervals for consistency?
-Also, use dark blue instead of gray for naturally-recruited?
-Might show up better.
--->](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohort.
+Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
+The large error bars for 70550 reflect the large within and among individual variability in yearly survival probability at this site (as also reflected in one-year survival probabilities in [@fig-translocation-survival]).
+All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
 
@@
 
-<!--# How to insert >= symbol in header and table cells> -->
-
 ```{r}
-#| label: tbl-survival-earlylate
-#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of survival was < 0.5 or >= 0.5."
+#| label: tbl-viability-earlylate
+#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of extinction over 50 years was 1 or < 1 (i.e., viability = 0 or 1, respectively)."
 #| echo: FALSE
@@
   mutate(site_id = as.character(site_id),
-         survival = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), "< 0.5", ">= 0.5"),
+         viability = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), 0, 1),
          period = if_else(year < 2013, "early", "late")) %>% 
   group_by(period) %>% 
-  count(period, survival) %>% 
+  count(period, viability) %>% 
   mutate(n_period = sum(n), 
@@
   mutate(across(prop, round, 3)) %>% 
-  arrange(period, survival)
+  arrange(viability, period)
 
-  knitr::kable(viability_table, format = "pipe", align = "lcccc", col.names = c("Period", "Survival probability", "Number of translocations", "Total translocations", "Proportion of translocations")) 
+  knitr::kable(viability_table, "pipe", align = "lcccr", col.names = c("Period", "Viability category", "Number of translocations", "Total translocations", "Proportion of translocations")) 
 ```

### 2023-04-15 doc/manuscript/translocation.qmd
Rework first three paragraphs of discussion

@@
 
-<!--# Title should be less than 135 characters -->
+<!--# Title should be less than 135 characters: Mark -->
 
@@
 
-We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
-First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
-Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
-For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions<!--# Only Class I and II are mentioned in Results -->.
+While previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], a strong causal chain from selection, to an increase in host fitness in the presence of the pathogen, to subsequent host recoveries remains lacking in most wildlife-disease systems.
+To robustly conclude that evolutionary rescue is the primary mechanism of host recovery, ideally evidence for each link in the causal chain should be provided.
+Here, we provide evidence for each of these links.
+First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering populations, consistent with strong selection due to disease.
+Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014; @grogan2018] (for more information, see SI X).
+For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions<!--# Only Class I and II are mentioned in Results.
+Mark: Check with Allie, the supporting information text says Class I and III -->.
 These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
 Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
-<!--# For brevity in the discussion, I tried to a pull a few key bullets out of Allie's text. Allie and Bree: Edit and expand as you see fit. -->
 
-The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistant phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
-Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
-Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
-Finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
-Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.
+Second, repeated translocations and long-term monitoring of naturally recovered populations demonstrate that fitness benefits of robust phenotypes are maintained beyond the natal environment in the presence of Bd.
+Specifically, robust phenotypes in recovered populations have reduced infection intensities, below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
+This phenotype persists following translocations to new environments --- translocated populations maintain infection intensities below this threshold and infection intensity is not a significant predictor of adult survival following translocation.
+Moreover, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that robust phenotypes are heritable to some degree.
+Together, strong selection following disease-induced declines and heritable increases in fitness point to evolutionary rescue as an important driver of observed natural recoveries and host persistence in this system [@knapp2016].
+A key advance of our study is that we take these exciting results a step further, leveraging natural evolutionary rescue and translocations to promote MYL frog recovery at the landscape scale.
 

### 2023-04-18 code/translocation_survival_createdata.Rmd
Add code to summarize ves and cmr survey results

@@
 ---
-title: "Predictors of frog survival following translocations - create dataset"
+title: "Create datasets: frog survival, presence/absence of life stages indicative of reproductive success"
 author: "Roland Knapp"
@@
 
-The code in this notebook creates a dataset for use in analyses to identify predictors of frog survival following translocation (using translocation_survival_analysis.Rmd).
+The code in this notebook creates two datasets: (1) dataset for use in analyses to identify predictors of frog survival following translocation (using translocation_survival_analysis.Rmd), and (2) metrics related to reproductive success.
 The code is written in distinct chunks to allow review of the output from one chunk before running the next.
@@
 ```
+
+# DATASET 1
 
@@
 - survival: Probability of surviving one year after translocation.
+
+# DATASET 2
+
+### Retrieve database tables
+
+```{r}
+source("db_connect.R")
+
+visit <- dbReadTable(con,"visit") %>% 
+  filter(visit_status == "suitable")
+
+survey <- dbReadTable(con, "survey") %>% 
+  select(id, visit_id, survey_type, comment)
+
+visual_survey <- dbReadTable(con, "visual_survey")
+
+capture_survey <- dbReadTable(con, "capture_survey") %>% 
+  select(id, survey_id, pit_tag_ref, tag_new, species, capture_animal_state, length)
+
+relocate <- dbReadTable(con, "relocate")
+
+sites <- tibble(site_id = c(70134, 70688, 70370, 70114, 70175, 70279, 70413, 71570, 71968, 72008, 72093, 72264, 72390, 72442, 72694, 70414, 70449, 70034, 70505, 72092, 70550, 70556, 70619, 70628, 70641, 74976, 72973), siteid_alt = c(70134, 70134, 70370, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70413, 70414, 70449, 70449, 70505, 70505, 70550, 70556, 70619, 70628, 70641, 74976, 74976))
+
+source("db_disconnect.R")
+```
+
+### Summarize VES counts
+
+```{r}
+visit_ves <- visit %>% 
+  rename(visit_id = id) %>% 
+  rename(comment_visit = comment)
+
+survey_ves <- survey %>% 
+  rename(survey_id = id) %>% 
+  rename(comment_survey = comment) %>% 
+  select(survey_id, visit_id, survey_type, comment_survey)
+
+visual_survey_ves <- visual_survey %>% 
+  filter(species == "ramu" &
+         visual_life_stage %in% c("tadpole", "subadult") & 
+         visual_animal_state == "alive")
+
+relocate <- sites %>% 
+  select(siteid_alt) %>% 
+  distinct(siteid_alt) %>% 
+  inner_join(relocate, by = c("siteid_alt" = "release_siteid1")) %>% 
+  select(siteid_alt, release_date) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(release_date_min = min(release_date)) %>% 
+  mutate(release_year = year(release_date_min), .after = release_date_min)
+
+frog_counts <- sites %>% 
+  inner_join(visit_ves, by = "site_id") %>% 
+  inner_join(survey_ves, by = "visit_id") %>% 
+  filter(survey_type == "visual") %>% 
+  left_join(visual_survey_ves, by = "survey_id") %>% 
+  select(siteid_alt, visit_date, visual_life_stage, location, count) %>% 
+  mutate(visit_year = year(visit_date), .after = visit_date) %>% 
+  group_by(siteid_alt, visit_date, visit_year, visual_life_stage) %>% 
+  summarize(count_total = sum(count)) %>% 
+  group_by(siteid_alt, visit_year, visual_life_stage) %>% 
+  summarize(count_max = max(count_total)) %>% 
+  ungroup() %>% 
+  complete(siteid_alt, visit_year, visual_life_stage, fill = list(count_max = 0)) %>% 
+  filter(!is.na(visual_life_stage)) %>% 
+  inner_join(relocate, by = "siteid_alt") %>% 
+  select(-release_date_min) %>% 
+  filter(visit_year > release_year) # Exclude surveys/counts made in the year of translocation (before presence of tadpoles is possible)
+
+rm(visit_ves, survey_ves, visual_survey_ves)
+
+frog_counts_table1 <- frog_counts %>% 
+  distinct(siteid_alt, visit_year) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_survey = n())
+
+frog_counts_table2 <- frog_counts %>% 
+  filter(visual_life_stage == "tadpole" &
+           count_max > 0) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_tads_present = n())
+
+frog_counts_table3 <- frog_counts %>% 
+  filter(visual_life_stage == "subadult" &
+           count_max > 0) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_subs_present = n())
+
+frog_counts_table4 <- frog_counts %>% 
+  group_by(siteid_alt, visit_year) %>% 
+  summarize(count_tadsub = sum(count_max)) %>% 
+  ungroup() %>% 
+  filter(count_tadsub > 0) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_tadsub_present = n())
+
+frog_counts_table <- frog_counts_table1 %>% 
+  left_join(frog_counts_table2, by = "siteid_alt") %>% 
+  left_join(frog_counts_table3, by = "siteid_alt") %>% 
+  left_join(frog_counts_table4, by = "siteid_alt") %>% 
+  replace_na(list(years_tads_present = 0, years_subs_present = 0, years_tadsub_present = 0)) %>% 
+  mutate(pyears_tadsub_present = years_tadsub_present / years_survey)
+```
+
+### Summarize CMR captures
+
+```{r}
+visit_cmr <- visit %>% 
+  rename(visit_id = id) %>% 
+  rename(comment_visit = comment)
+
+survey_cmr <- survey %>% 
+  rename(survey_id = id) %>% 
+  rename(comment_survey = comment) %>% 
+  select(survey_id, visit_id, survey_type, comment_survey)
+
+capture_survey_cmr <- capture_survey %>% 
+  filter(species == "ramu" &
+         length >= 40 & 
+         tag_new == TRUE &
+         capture_animal_state != "dead")
+
+frog_captures <- sites %>% 
+  inner_join(visit_cmr, by = "site_id") %>% 
+  inner_join(survey_cmr, by = "visit_id") %>% 
+  filter(survey_type == "cmr" |
+         survey_type == "visual") %>% # Include CMR and VES records to ensure inclusion of all survey dates. Survey without RAMU detections typically recorded as VES. 
+  left_join(capture_survey_cmr, by = "survey_id") %>% 
+  select(siteid_alt, visit_date, survey_type, pit_tag_ref, tag_new) %>% 
+  mutate(visit_year = year(visit_date), .after = visit_date) %>% 
+  distinct(siteid_alt, visit_year, pit_tag_ref) %>% 
+  group_by(siteid_alt, visit_year) %>% 
+  summarize(recruits = sum(!is.na(pit_tag_ref))) %>%  # !is.na() converts pit_tag_ref records to TRUE/FALSE. sum() treats TRUE as 1/FALSE as 0, so sum is count of not NA values
+  ungroup() %>% 
+  inner_join(relocate, by = "siteid_alt") %>% 
+  select(-release_date_min) %>% 
+  filter(visit_year > (release_year + 2)) # Exclude counts made in first 3 years after translocation (before recruitment is possible)
+
+# 70628 is not represented in data set because it has only been 2 years since the 2020 translocation, so no records met the final filtering condition.
+
+rm(visit_cmr, survey_cmr, capture_survey_cmr)
+
+frog_captures_table1 <- frog_captures %>% 
+  distinct(siteid_alt, visit_year) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_survey = n())
+
+frog_captures_table2 <- frog_captures %>% 
+  filter(recruits > 0) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(years_recruits_present = n())
+
+frog_captures_table <- frog_captures_table1 %>% 
+  left_join(frog_captures_table2, by = "siteid_alt") %>% 
+  mutate(years_recruits_present = replace_na(years_recruits_present, 0),
+         pyears_recruits_present = years_recruits_present / years_survey)
+```

### 2023-04-18 doc/manuscript/translocation.qmd

@@
 Moreover, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that robust phenotypes are heritable to some degree.
-Together, strong selection following disease-induced declines and heritable increases in fitness point to evolutionary rescue as an important driver of observed natural recoveries and host persistence in this system [@knapp2016].
+Together, we show that strong selection following disease-induced declines and heritable increases in fitness point to evolutionary rescue as an important driver of observed natural recoveries and host persistence in this system [@knapp2016].
 A key advance of our study is that we take these exciting results a step further, leveraging natural evolutionary rescue and translocations to promote MYL frog recovery at the landscape scale.

### 2023-04-18 doc/manuscript/translocation.qmd
Minor edits prior to more extensive text changes

@@
 
-Roland A. Knapp^1,2^\*, Mark Q. Wilber^3^, Allison Q. Byrne^4,5^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
+Roland A. Knapp^1,2,^\*, Mark Q. Wilber^3,a^, Allison Q. Byrne^4,5,a^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
 
@@
 
-^7^NCX, San Francisco, CA, 94115
+^7^Planet, San Francisco, CA, 94107
 
@@
 \*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu
+
+^a^These authors contributed equally to this work.
 
@@
 
+<!--# To allow comparison with other translocation studies, add information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
+-->
+
 To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
@@
 Therefore, frog resistance in non-natal habitats is maintained across generations.
-
-In addition to survival......
-
-<!--# To allow comparison with other translocation studies, adding information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
--->
 

### 2023-04-19 code/translocation_survival_createdata.Rmd
Finalize code to summarize survey data

@@
 
-relocate <- sites %>% 
+relocate_ves <- sites %>% 
   select(siteid_alt) %>% 
@@
   ungroup() %>% 
-  complete(siteid_alt, visit_year, visual_life_stage, fill = list(count_max = 0)) %>% 
+  complete(visual_life_stage, # cross all possible life_stage values with the unique pairs of (siteid_alt, visit_year) that exist in the data 
+           nesting(siteid_alt, visit_year),
+           fill = list(count_max = 0)) %>% 
+  select(siteid_alt, visit_year, visual_life_stage, count_max) %>% 
   filter(!is.na(visual_life_stage)) %>% 
-  inner_join(relocate, by = "siteid_alt") %>% 
+  inner_join(relocate_ves, by = "siteid_alt") %>% 
   select(-release_date_min) %>% 
@@
   mutate(pyears_tadsub_present = years_tadsub_present / years_survey)
+
+write_csv(frog_counts_table, here::here("data", "clean", "frog_counts_table.csv"))
+
+rm(frog_counts_table1, frog_counts_table2, frog_counts_table3, frog_counts_table4)
 ```
@@
 
+relocate_cmr <- sites %>% 
+  select(siteid_alt) %>% 
+  distinct(siteid_alt) %>% 
+  inner_join(relocate, by = c("siteid_alt" = "release_siteid1")) %>% 
+  select(siteid_alt, release_date) %>% 
+  group_by(siteid_alt) %>% 
+  summarize(release_date_min = min(release_date)) %>% 
+  mutate(release_year = year(release_date_min), .after = release_date_min)
+
 frog_captures <- sites %>% 
@@
   ungroup() %>% 
-  inner_join(relocate, by = "siteid_alt") %>% 
+  inner_join(relocate_cmr, by = "siteid_alt") %>% 
   select(-release_date_min) %>% 
@@
 
-rm(visit_cmr, survey_cmr, capture_survey_cmr)
+rm(visit_cmr, survey_cmr, capture_survey_cmr, relocate_cmr)
 
@@
          pyears_recruits_present = years_recruits_present / years_survey)
+
+write_csv(frog_captures_table, here::here("data", "clean", "frog_captures_table.csv"))
+
+rm(frog_captures_table1, frog_captures_table2)
 ```

### 2023-04-19 doc/manuscript/translocation.qmd
Shorten Intro and Results and add frog count summary

@@
 Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Allie comment.
-I believe evolutionary rescue also encompasses the introduction of new variation through gene flow.
-See box 1 in ref 16 for a helpful discussion of genetic vs evolutionary rescue.
-Could delete "Originating from variation within a host population" here and just add more information about what make evolutionary rescue special - that is the increase in locally adapted phenotypes.
-Bree comment.
-Minor – there are a lot of clauses in this sentence.
-Wonder if setting this as " – as necessary for evolution rescue – " might make the sentence more readable-->.
-Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
-In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals <!--# Bree comment: Again minor, but seems like this could be slimmed (as the sentence repeats the idea of promoting recovery twice and the "provide sources of individuals" seems awkward).
-It could read something like "In addition to promoting recovery of remnant populations after disease-induced declines, natural selection can also lead to increased resistance and/or tolerance…." -->with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
-To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
-<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->
+"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
+Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022].
+However, linking these selective responses directly to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
+In addition to promoting recovery of remnant populations after disease-induced declines, the resistant/tolerant individuals <!--# How best to refer to resistance/tolerance?
+--> in recovering populations can also be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
+To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
 
 There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
-First, populations that survived <!--# Bree comment.
-Worth keeping tense consistent so that the four conditions feel parallel?
-This sentence could read "First, populations that survive following disease arrival evolve increased resistance" -->following disease arrival have evolved increased resistance.
-Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"?
-Resistance could convey complete prevention of disease, so I often struggle with which word to use -->and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
+First, populations that survive following disease emergence evolve increased resistance/tolerance.
+Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
+Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
 And fourth, reintroduced populations show long-term growth and persistence.
@@
 Endangered Species Act.
-In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] <!--# Bree comment: Very minor but there are lots of places with long structurally complex sentences that seem just as good as broken up shorter sentences -->, and the arrival of *B.
+In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B.
 dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
@@
 In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"?
-Mark here:  This transition reads fine to me.
--->.
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in recovering populations [@knapp2016].
 We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
@@
 Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
-
-<!--# Bree comment: All of this was written before we knew the genomics had something cool to say.
-But now that the genomics feel a bit more elevated, I think this paragraph should be more bold.
-For example, we might want more here or in the Discussion to make it clear how the results of each of these sections are super important on their own.
-And that combining them is really amazing.
-I think the current tone is appropriately factual, but at some point we need to be louder about the novelty involved and the incredible amount of work that this represents.
-This version feels like, we did a,b,c and so MYLF meet the conditions for feasibility (which makes it sound like we've only just gotten to the *starting* point of the study.
-I think we need to go further than saying they meet the four conditions of a specific conservation action.
-This is where beefing up the conclusions about evolutionary rescue seem appropriate.
-Also, genomic analyses typically come last in our papers as a dive into possible mechanistic explanations for observed patterns in the field.
-I think the order of the primary sections makes sense given the way you set up the argument about evolutionary rescue.
-That said, the order combined with the tone feels like it's contributing to the sense that the genomics are a box to be checked rather than contributing something exciting to the overall story.
-Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->
-
-<!--# Original pre-edit text from the above paragraph: Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
-Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
-These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
--->
 
@@
 
-To determine whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
-Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
-Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
-These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
-Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
+Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
+These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@knapp2011, @vredenburg2010].
 
-To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
-For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
-Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
-pi-diff.
--->.
-In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overlapped for *F~ST~* and $\pi$ <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent.
-pi - diff -->.
-Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.
+We used two complementary approaches to identify regions of the genome under selection.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
+Second, we used a splined window approach to identify larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
 
-<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->
+The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
+One of the 7 identified genes did not have an associated annotation.
+Frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
 
-Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
-Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
-The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
-A single gene in both groups did not have an associated annotation.
-For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion?
-:  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
--->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
-No GO terms were significantly over-represented in the set of 29 annotated outlier genes.
-
-Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph.
-To avoid confusion, use "windows" throughout?
---> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).
--->) and no GO terms were overrepresented within this set of genes.
-The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.
--->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+The splined window analysis identified 58 outlier regions for *F~ST~* and 33 outlier regions for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
+A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2).
+The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
+Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.
@@
 Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
-<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.
-Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove.
-Among susceptible species, survivors of epizootics bear alleles encoding these conformations."?
-What about MHC Class II characteristics linked to resistance by Bataille et al.
-(2015)?
-Savage and Zamudio (2011, 2016), Kosch et al.
-(2018)?-->
 
-Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
-While the functions of all genes and regions showing signatures of selection were not clearly interpretable, regions of the genome associated with resistant phenotypes in other amphibian systems such as the MHC complex, showed strong evidence of selection in recovering MYL frog populations.
+In summary, our genomics results show that recovering populations have signatures of strong selection on multiple regions of the genome.
+Although the functions of all associated genes were not clearly interpretable, regions of the genome under selection in recovering populations included those associated with resistance/tolerance in other amphibian systems, such as the MHC complex.
 
@@
 
-#### Frog translocations
+#### Frog population recovery
 
@@
 d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
+d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
+d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))
 
@@
   summarize(max(max_2period))
+
+frog_counts_gt0_nrow <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  nrow()
+
+frog_counts_gt0_median <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(median(pyears_tadsub_present)) %>% 
+  round(2)
+
+frog_counts_gt0_min <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(min(pyears_tadsub_present)) %>% 
+  round(2)
+
+frog_counts_gt0_max <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(max(pyears_tadsub_present))
+
+frog_captures_gt0_nrow <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  nrow()
+
+frog_captures_gt0_median <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(median(pyears_recruits_present)) %>% 
+  round(2)
+
+frog_captures_gt0_min <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(min(pyears_recruits_present)) %>% 
+  round(2)
+
+frog_captures_gt0_max <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(max(pyears_recruits_present))
 ```
 
-Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
-In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
-Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
-These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
-We refer to these reintroductions as "translocations" [@seddon2014].
-During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI).
-Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`) collected from 1 of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
+In this section, we describe the outcome of *R.
+sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
+Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) [@knapp2016] and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@joseph2018; @seddon2014].
+The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
+
 Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
@@
 
-<!--# To allow comparison with other translocation studies, add information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation.
--->
+Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or subadults.
+For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
+These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
+As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
+In summary, survey results suggest that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
 
-To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
-Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
+To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
@@
 Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
-In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
-Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
-<!--# Allie comment: Maybe tolerance is a better word here?
-Since they were infected.
-Or maybe we need to explicitly define what we mean with resistance vs tolerance.
-Bree comment: I noticed this also – in the Intro, you do say that you are lumping both resistance and tolerance into the phrase "resistance" but you are talking about multiple things (being able to avoid infection, being able to fight infection, being able to tolerate infection, etc).
-Maybe at some point we want to address specifically that the genomic changes we are observing could contribute to multiple facets of resistance / tolerance.
-And that we are lumping those together in an "evolved resistance" bin.
-Roland comment: We reworded this sentence, but probably have not fully dealt with resistance/tolerance issue.
---> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in three locations where longer CMR time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
+In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs trasnslocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
 Therefore, frog resistance in non-natal habitats is maintained across generations.
 
-#### Population viability modeling
+#### Long-term population viability
 
-Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
+Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
 Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
-To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
+To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
 Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
@@
 
-We then included demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into our model described above and used it to predict long-term population viability.
-Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, all six populations with the highest adult survival probabilities had 50-year extinction probabilities less than 0.5 (0 - 0.32; \@fig-viability B).
+We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; \@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
-When we decreased $\omega$ from 0.3 to our most conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 - 0.94 for three additional populations (@fig-viability B).
+When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 - 0.94 for three additional populations (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-Finally, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
+In addition, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013 (@tbl-survival-earlylate).
 Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
@@
 
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap SI) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 During each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
@@
 
+<!--# Add text describing calculations of VES and CMR-recruitment counts.
+For the latter, only included surveys conducted >= 3 years following translocation to account for duration of early life stages.
+-->
+
 ##### Predictors of post-translocation frog survival
 
-To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
+To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
 We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
@@
 Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations ([@fig-survival-postdens] : SI).
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), winter severity in the year following translocation (snow_t1), site elevation (elevation), donor population (donor_72996), sex (sex_male), frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first)).
+In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3\--16).
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
@@
 Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-<!--# Mention that in both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available.
--->
 
@@
 --> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful.
-Roland response:  -->](figures/allele_maps.jpg){#fig-allelefrequencies}
+-->](figures/allele_maps.jpg){#fig-allelefrequencies}
 
@@
 
-![Conditional effects of the important predictors of 1-year frog survival.
+![Results from a multilevel Bayesian model showing conditional effects of the important predictors of 1-year frog survival.
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
@@
 
-### Selection Results
+### Frog evolution in response to Bd
+
+#### Methods
+
+Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
+Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
+The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
+As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+
+We identified outlier SNPs using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
+In the main text we present results solely based on the stringent set.
+
+#### Results
+
+Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
+
+In the main manuscript, we describe the stringent set of variants.
+In addition, the liberal set of variants included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
 
@@
 Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
+
+Based on the SNP and splined-window analyses , no GO terms were significantly over-represented in the set of annotated genes associated with the outlier SNPs or regions.
+
+<!--# Mention that the patterns shown in the allele frequencies figure (panel A, B) may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors?
+-->
 
@@
 ![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-A variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
+In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
 
@@
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-Variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
@@
 #| label: tbl-survival-earlylate
-#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) for which the probability of survival was < 0.5 or >= 0.5."
+#| tbl-cap: "Association between the proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) and probability of survival (< 0.5 or >= 0.5)."
 #| echo: FALSE
@@
 
-Dataset 2: spline_window_shared_outliers.csv
+Dataset 2: spline_window_shared_outliers.csv <!--# The datasetcontains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
+Do we need to explain that one region is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows)?
+-->
 

### 2023-04-21 doc/manuscript/translocation.qmd
Major rewrite of Discussion text

@@
 
-<!--# Title should be less than 135 characters: Mark -->
+## Alternative title: Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
+
+<!--# Title should be less than 135 characters -->
 
@@
 Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.
--->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
 This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
-In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
+In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and four occur in the extended MHC Class III region (HSP70, LSM2, VARS <!--# Check on number of genes, add missing gene here -->) [@ohta2006].
+Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
-In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs trasnslocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
 Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
@@
 
-Disease-induced declines are a central challenge for the conservation of global biodiversity, but strategies to recover affected species are generally lacking.
-One potential strategy, and the one we use in the current study, leverages evolutionary rescue to recover species at the landscape scale.
-Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
-Specifically, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd), and (ii) individuals from these resistant populations can be used to reestablish viable populations despite ongoing Bd infection.
-These results provide a rare, and perhaps novel, example of how evolutionary rescue can be used to successfully recover a species in the presence of disease.
+Disease-induced population declines are impacting global biodiversity [@daszak2000], and strategies to recover affected species are lacking.
+Recovery in this context is likely to be particularly challenging because typically recovery will need to occur in the presence of ongoing disease.
+One potential strategy, and the one we employ in the current study, utilizes populations of resistant/tolerant individuals as a source of animals with which to reestablish extirpated populations in the presence of disease.
+Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
+In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to recover populations.
+Here, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
+Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014], obtaining strong evidence in wild animal populations for each of the associated causal links (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
+Nonetheless, in the following summary of our results, we use evolutionary rescue as an informative, but tentative framework.
 
-While previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], a strong causal chain from selection, to an increase in host fitness in the presence of the pathogen, to subsequent host recoveries remains lacking in most wildlife-disease systems.
-To robustly conclude that evolutionary rescue is the primary mechanism of host recovery, ideally evidence for each link in the causal chain should be provided.
-Here, we provide evidence for each of these links.
-First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering populations, consistent with strong selection due to disease.
-Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014; @grogan2018] (for more information, see SI X).
-For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions<!--# Only Class I and II are mentioned in Results.
-Mark: Check with Allie, the supporting information text says Class I and III -->.
-These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
-Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
+Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
+Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
+In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter - before translocation].
+These differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
+The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
 
-Second, repeated translocations and long-term monitoring of naturally recovered populations demonstrate that fitness benefits of robust phenotypes are maintained beyond the natal environment in the presence of Bd.
-Specifically, robust phenotypes in recovered populations have reduced infection intensities, below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
+Results from our genomic analyses strongly suggest that natural selection for adaptive alleles and resulting increased resistance/tolerance is at least partially responsible for the reduced susceptibility of frogs in recovering populations.
+We identified substantial differences in allele frequencies and larger genomic regions between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
+Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@grogan2018, reviewed in @zamudio2020].
+For example, the major histocompatibility complex (MHC) plays an important role in immunity.
+In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
+These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd \[e.g., @savage2011; @bataille2015\].
+Similarly, the region with the strongest indication of directional selection contains the gene, GVINP1.
+Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014].
+This gene was also strongly linked to disease in salmon, explaining a remarkable 20% of the resistance phenotype [@robledo2020; @robledo2018].
+Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection across a wide geographic area.
+We provide additional details on these and other genes identified in our study as under selection in the SI.
+
+Reintroductions are an increasingly used method to address escalating loss of wildlife species and populations.
+However, recent reviews suggest that such efforts generality have a low success rate (Griffiths Pavajeau 2008, Dodd 2005).
+Of the 24 translocations of MYL frogs that we conducted as part of the current study, x% showed evidence of successful reproduction, x% of recruitment of new adults, and x% showed high annual survival and a high probability of persistence over 50 years.
+Following translocation, Bd loads were relatively unchanged.
+In addition, survival/, low-moderate Bd loads extend beyond the natal environment, maintained between generations.
+Success was even higher for more recently conducted translocations.
+The xx% success rate for these more recent translocations is particularly remarkable given that population establishment is occurring in the presence of Bd, a pathogen to which MYL frogs are highly susceptible.
+
+several lines of evidence indicate that the selection-derived resistance/tolerance and increased fitness in the presence of Bd is associated with population recovery.
+Results from frog translocations conducted as part of the current study indicate that following translocation Bd loads in frogs changed little and post-translocation survival was relatively high at most sites.
+These results are in contrast to the outcome of translocations we conducted previously of naive frogs into Bd-positive habitats (conducted prior to our current understanding of the importance of Bd exposure history on frog susceptibilty).
+Following those translocations, frogs developed high Bd loads and experienced high mortality (details in SI <!--# These data are unpublished, so if we want to refer to them we'll have to include the data in the SI.
+Is this necessary to support this result?
+I don't see any obvious alternatives.
+-->).
+In addition, in the current translocations increased fitness maintained between generations.
+
+Results from frog translocations conducted as part of the current study indicate that higher fitness in the presence of Bd observed in frogs from recovering populations extends beyond the natal environment.
+Specifically, following translocation of frogs from natal to recipient sites, Finally, in translocated populations the higher fitness is also maintained in subsequent generations (@fig-compare_surv_probs).
+Collectively, these results indicate that the higher fitness observed in recovering Bd-positive populations is not simply a site effect, but also a frog effects and not only the sites, suggesting that the observed resistance and resulting h
+
+These are that the increased fitness repeated translocations and long-term assessment of naturally-recovering populations demonstrate that fitness benefits of robust phenotypes are maintained beyond the natal environment in the presence of Bd.
+Specifically, robust phenotypes in recovered populations have reduced Bd loads, below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
 This phenotype persists following translocations to new environments --- translocated populations maintain infection intensities below this threshold and infection intensity is not a significant predictor of adult survival following translocation.
@@
 
-<!--# Comment from Allie that will be helpful in making the cases that what we are seeing is evidence of strong selection/evolutionary rescue -->
+<!--# Comment from Allie that will be helpful in making the case that what we are seeing is evidence of strong selection/evolutionary rescue -->
 
@@
     For example, if the local biotic (e.g., community interactions) and abiotic (e.g., temperature) environment strongly affect recovery dynamics and there is strong heterogeneity in these conditions across habitat patches on a landscape, then recovery in a natal habitat will be less likely to translate into recovery in a non-natal habitat.
-
--   The MYL frog system is unique in that low productivity due to short summers provide a homogenizing effect across habitat patches, potentially minimizing the effect of the local environment on landscape-level recovery compared to what we might expect in higher productivity systems (for example, tropical amphibian assemblages) <!--# Roland, I am just brainstorming here and this might not make sense.
-We can delete all of this if necessary.
---> .
-    That being said, even in this low productivity system we still observed a strong site and environmental effect on translocated adult survival.
-    <!--# This is an interesting topic, but despite low productivity MYL frog habitats are highly variable: ponds, streams, meadows, alpine lakes, forested lakes, etc.
-I think we should drop this bullet.
-Mark here:  -->
 
@@
 --> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful.
--->](figures/allele_maps.jpg){#fig-allelefrequencies}
+-->](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
@@
 Capitalize plot identifiers (A, B, C, D).
--->](figures/spline_manhattan_figure.jpg){#fig-spline-manhattan}
+-->](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
 
@@
 Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
 
@@
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
 
@@
 The dashed line shows where $\omega = 0.3$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability}
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript1.jpg){#fig-viability}
 
@@
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
-(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp1.jpg){#fig-viability-supp}
 
@@
 
+### Translocations
+
+<!--# Necessary? -->
+
+Description of translocation of Bd-naive populations to Bd-positive sites
+
+-   Spur Basin to Pinchot, Upper, and Ruskin Basins
+
+-   Marmot to Square Lake
+
+Description of translocation of Bd-naive populations to Bd-negative sites
+
+-   Marmot to Ameletus Lake
+
+-   Upper Center Basin to South Center Basin
+
 ### Figures
@@
 Label/identify both the liberal and stringent outlier sets?
--->](figures/map_qqplot_manhattan.jpg){#fig-selectionresults}
+-->](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
 
@@
 The large error bars for 70550 reflect the large within and among individual variability in yearly survival probability at this site (as also reflected in one-year survival probabilities in [@fig-translocation-survival]).
-All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
+All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
 

### 2023-04-22 doc/manuscript/translocation.qmd
Add final paragraphs to Discussion, edits elsewhere

@@
 
-Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations.
-In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007].
+Emerging infectious diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
 Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
-Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
-At a time when novel infectious diseases are emerging at an unprecedented rate [@fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.
+Unfortunately, given the rarity of wildlife recoveries in the presence of disease, we know little about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
+At a time when novel infectious diseases are emerging at an unprecedented rate [@daszak2000; @fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.
 
@@
 "Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
-Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022].
-However, linking these selective responses directly to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
-In addition to promoting recovery of remnant populations after disease-induced declines, the resistant/tolerant individuals <!--# How best to refer to resistance/tolerance?
---> in recovering populations can also be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
+Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often result in increased host resistance and/or tolerance.
+However, linking these selective responses to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
+In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in recovering populations can, in theory, be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
@@
 Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
+Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and across generations, allowing for the establishment of reintroduced populations.
 And fourth, reintroduced populations show long-term growth and persistence.
@@
 The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
-Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
+Once the most common amphibian in the high elevation portion of the California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
 Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
@@
 Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
+Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
 Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
 
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in recovering populations [@knapp2016].
+In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
 We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
-These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
+These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the multi-year extent of the CMR data.
+Collectively, these results provide one of the first examples of how evolution and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
+This example exhibits several of the key characteristics of evolutionary rescue, and has potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
 
@@
 To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
-Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
-These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@knapp2011, @vredenburg2010].
+Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].
 
-We used two complementary approaches to identify regions of the genome under selection.
+We conducted a principal component analysis (PCA) of the genomic data to describe the association between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
-Second, we used a splined window approach to identify larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
 
-The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A: SI), and clusters reflected the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish between (i) allelic associations of individuals sampled in naive versus recovering populations, and (ii) allelic associations resulting from population structure and genetic drift.
+
+Results from the SNP and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults B, C: SI, SI Dataset 1).
 One of the 7 identified genes did not have an associated annotation.
@@
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
+Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
-This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
-In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and four occur in the extended MHC Class III region (HSP70, LSM2, VARS <!--# Check on number of genes, add missing gene here -->) [@ohta2006].
-Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-
-In summary, our genomics results show that recovering populations have signatures of strong selection on multiple regions of the genome.
-Although the functions of all associated genes were not clearly interpretable, regions of the genome under selection in recovering populations included those associated with resistance/tolerance in other amphibian systems, such as the MHC complex.
-
-<!--# Based on Fig.
-2, there are numerous outlier windows.
-How were the specific windows mentioned in above and in Fig.
-2 C, D chosen?
-Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria.
--->
+This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+In *Xenopus*, five of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes (@fig-synteny-plot SI).
 
@@
 sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
-Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) [@knapp2016] and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@joseph2018; @seddon2014].
+Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@seddon2014].
 The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
+Following translocation, adult survival and recruitment of new adults were estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
 
-Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+Estimates of frog survival during the 1 year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
 As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
@@
 In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
-A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
+A sensitivity analysis of long-run growth rate $\lambda$ to yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$, recruitment probability $\omega$, and female fecundity *F* showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
 Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
@@
 We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
-Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; \@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
-When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 - 0.94 for three additional populations (@fig-viability B).
+When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-In addition, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013 (@tbl-survival-earlylate).
-Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
+In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
+Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
 
@@
 Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
-In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to recover populations.
-Here, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
+In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
+The results of our study provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
 Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014], obtaining strong evidence in wild animal populations for each of the associated causal links (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
-Nonetheless, in the following summary of our results, we use evolutionary rescue as an informative, but tentative framework.
+Nonetheless, in the following narrative, we use evolutionary rescue as an informative, albeit tentative framework.
 
@@
 In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter - before translocation].
-These differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
+The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
 The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
@@
 We identified substantial differences in allele frequencies and larger genomic regions between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
-Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@grogan2018, reviewed in @zamudio2020].
+Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
 For example, the major histocompatibility complex (MHC) plays an important role in immunity.
 In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
-These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd \[e.g., @savage2011; @bataille2015\].
+These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
 Similarly, the region with the strongest indication of directional selection contains the gene, GVINP1.
@@
 This gene was also strongly linked to disease in salmon, explaining a remarkable 20% of the resistance phenotype [@robledo2020; @robledo2018].
-Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection across a wide geographic area.
-We provide additional details on these and other genes identified in our study as under selection in the SI.
+Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
+In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Retain or delete? -->.
+Collectively, these results suggest that the documented natural recovery of MYL frogs following Bd-caused declines [@knapp2016] is consistent with evolutionary rescue.
 
-Reintroductions are an increasingly used method to address escalating loss of wildlife species and populations.
-However, recent reviews suggest that such efforts generality have a low success rate (Griffiths Pavajeau 2008, Dodd 2005).
-Of the 24 translocations of MYL frogs that we conducted as part of the current study, x% showed evidence of successful reproduction, x% of recruitment of new adults, and x% showed high annual survival and a high probability of persistence over 50 years.
-Following translocation, Bd loads were relatively unchanged.
-In addition, survival/, low-moderate Bd loads extend beyond the natal environment, maintained between generations.
-Success was even higher for more recently conducted translocations.
-The xx% success rate for these more recent translocations is particularly remarkable given that population establishment is occurring in the presence of Bd, a pathogen to which MYL frogs are highly susceptible.
+Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
+Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
+Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
+Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
+The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
+In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
+Results from the population viability model were also encouraging.
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
+In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
 
-several lines of evidence indicate that the selection-derived resistance/tolerance and increased fitness in the presence of Bd is associated with population recovery.
-Results from frog translocations conducted as part of the current study indicate that following translocation Bd loads in frogs changed little and post-translocation survival was relatively high at most sites.
-These results are in contrast to the outcome of translocations we conducted previously of naive frogs into Bd-positive habitats (conducted prior to our current understanding of the importance of Bd exposure history on frog susceptibilty).
-Following those translocations, frogs developed high Bd loads and experienced high mortality (details in SI <!--# These data are unpublished, so if we want to refer to them we'll have to include the data in the SI.
-Is this necessary to support this result?
-I don't see any obvious alternatives.
--->).
-In addition, in the current translocations increased fitness maintained between generations.
+Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
+In an earlier study of 2 of our 12 translocated populations, Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment [@joseph2018: sites referred to as "Alpine" and "Subalpine" are referenced as "70550" and "70505" in the current study].
+Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
+In addition to these important but relatively subtle effects on adults, MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and experience high mortality.
+This high susceptibility of metamorphs is documented for numerous species of anurans, and may result from the poorly developed immune system characteristic of this early life stage [@humphries2022].
+In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
+Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
 
-Results from frog translocations conducted as part of the current study indicate that higher fitness in the presence of Bd observed in frogs from recovering populations extends beyond the natal environment.
-Specifically, following translocation of frogs from natal to recipient sites, Finally, in translocated populations the higher fitness is also maintained in subsequent generations (@fig-compare_surv_probs).
-Collectively, these results indicate that the higher fitness observed in recovering Bd-positive populations is not simply a site effect, but also a frog effects and not only the sites, suggesting that the observed resistance and resulting h
-
-These are that the increased fitness repeated translocations and long-term assessment of naturally-recovering populations demonstrate that fitness benefits of robust phenotypes are maintained beyond the natal environment in the presence of Bd.
-Specifically, robust phenotypes in recovered populations have reduced Bd loads, below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
-This phenotype persists following translocations to new environments --- translocated populations maintain infection intensities below this threshold and infection intensity is not a significant predictor of adult survival following translocation.
-Moreover, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that robust phenotypes are heritable to some degree.
-Together, we show that strong selection following disease-induced declines and heritable increases in fitness point to evolutionary rescue as an important driver of observed natural recoveries and host persistence in this system [@knapp2016].
-A key advance of our study is that we take these exciting results a step further, leveraging natural evolutionary rescue and translocations to promote MYL frog recovery at the landscape scale.
-
-<!--# Comment from Allie that will be helpful in making the case that what we are seeing is evidence of strong selection/evolutionary rescue -->
-
-For the question about describing what we are seeing as evolutionary rescue - I feel comfortable and confident using that wording for two reasons:
-
-1\) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
-In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
-In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
-Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significant outlier SNP that has an annotation.
-and
-
-2\) CORROBORATING STUDIES. For each gene mentioned above we can list many studies that have linked the genes to Bd/disease/wounds.
-In the case of GVINP1 for instance there are at least three Bd transcriptomic studies that have found this specific gene to be differentially expressed in amphibians (I currently only cite one in the paper but could add the other two: Grogan et al.
-2018, Ellison et al.
-2014a, Ellison et al.
-2014b).
-In fact, if you haven't taken a look at the Grogan paper it is really interesting because it is similar in that they are describing evolution in natural settings and they talk a lot about the GVINP1 gene.
-Even more remarkable that this gene was also linked so strongly to gill amoebic disease in salmon (explaining an astounding 20% of the resistance phenotype).
-
-<!--# End of comment -->
-
-Implications for recovery of Mountain yellow legged frogs
-
--   \[Elaborate on this a bit more\]
-
--   One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
-    For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al.
-2012) however many studies have found that immune overactivation is linked to susceptibility for Bd.
-    This could be different for populations with different histories of Bd exposure.
-    Perhaps Rana muscosa has a more recent history with Bd (maybe with another lineage) than for instance Atelopus.
-    All of this has led to me think this whole question deserves another paper integrating the Atelopus and MYLF datasets, so maybe we save all this for that?
-    Open to whatever you all think is relevant to this story.
-
--   Despite relatively high success rate in MYL frogs, it is apparent that not all such recovery efforts will succeed.
-    In addition to ongoing effects of Bd, several deterministic and stochastic factors likely also play an important role: (1) effects of elevation, winter severity, and donor population on frog survival, and (2) variability in recruitment driven by Bd-caused mortality of subadults.
-
--   Now have a means of reestablishing MYL frog populations across their historical range, including in areas in which natural recolonization is precluded by dispersal limitation.
-    Nonetheless, recovery across this large landscape will require substantial resources over many decades.
-
-The recovery of other amphibian species impacted by Bd will continue to be challenging.
-
--   For example, despite hundreds of declining amphibian species, most attempts at treatments and recovery in the presence of Bd have failed \[citations\].
-
--   *In-situ* treatments such as the application of fungicides, probiotics, or habitat manipulations (Bosch), have shown some case-by-case successes, but generally lack the necessary effectiveness and scalability to promote large-scale recovery.
-
--   Natural recoveries are occurring in the presence of Bd in other amphibian species (Newell, Tobin, Scheele, Voyles) and assisting natural recoveries may provide the most effective means for large-scale recovery efforts.
-
--   However, recoveries in other systems are not necessarily driven by the evolution of host defense (Brannley) and underlying mechanisms of recovery in other amphibian-Bd systems may be less translatable from a single population to the landscape-level than we observe in MYL frogs.
-    For example, if the local biotic (e.g., community interactions) and abiotic (e.g., temperature) environment strongly affect recovery dynamics and there is strong heterogeneity in these conditions across habitat patches on a landscape, then recovery in a natal habitat will be less likely to translate into recovery in a non-natal habitat.
-
--   Thus, the local environment will be a key consideration when assessing the feasibility of assisted recovery in other amphibian systems.
-
-What does this study tell us more broadly about disease-induced recovery in other systems and the conservation of biodiversity?
-
--   Identifying the mechanisms of recovery in the face of disease is huge knowledge gap.
-
-<!--# Previous text from Roland below -->
-
-Could include mention here of why MYL frogs show evidence of evolutionary rescue in response to Bd exposure when so few other amphibians do.
-Text from Introduction describing the conditions under which evolutionary rescue is most likely?
-
-3.  Translocations (including viability results?):
-    -   Recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
-Hundreds of declining species, but mostly failed attempts at recovery.
-Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
-In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
-    -   Observed reduced susceptibility is likely due to evolutionary rescue, and this outcome is likely facilitating both natural recovery (Knapp et al.
-2016) and success of translocation efforts described in this study.
-MYL frogs provide a rare example of highly susceptible species recovering in the presence of pathogen/disease.
-Recovery would likely not be possible without evolutionary rescue, as suggested by numerous failed attempts to reestablish populations of Bd-impacted amphibians.
-Although evolution of resistance in donor populations facilitates the long-term viability of translocated populations, survival is still highly dependent on environment independent of Bd.
-    -   IUCN guidance for reintroduction/translocation success is the presence of reproduction.
-Based on this standard, X of our 12 translocation efforts would be deemed successful.
-We used a higher standard - 1 year survival and probability of extinction over 50 years.
-More relevant standard because of the bottleneck to population establishment posed by subadult survival which despite evolutionary rescue is strongly influenced by disease.
-    -   This success in the face of a highly virulent pathogen is all the more surprising given the generally low rate of translocation/reintroduction success (several review papers).
-Despite relatively high success rate in MYL frogs, it is apparent that not all such recovery efforts will succeed.
-In addition to ongoing effects of Bd, several deterministic and stochastic factors likely also play an important role: (1) effects of elevation, winter severity, and donor population on frog survival, and (2) variability in recruitment driven by Bd-caused mortality of subadults.
-    -   Now have a means of reestablishing MYL frog populations across their historical range, including in areas in which natural recolonization is precluded by dispersal limitation.
-Nonetheless, recovery across this large landscape will require substantial resources over many decades.
-    -   Big picture: May also have implications for the recovery of the many Bd-impacted amphibians worldwide.
-Remains to be seen how many other amphibians respond to Bd exposure via evolutionary rescue.
+The recent emergence of Bd worldwide has contributed to the decline of more than 500 amphibian species, nearly 100 of which now are extinct in the wild [@scheele2019].
+This unprecedented impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
+Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of using animals from these recovering populations to reestablish extirpated populations.
+As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have acquired via evolution the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
+Despite the hopeful example of successful reestablishment of MYL frogs via evolutionary rescue and reintroductions, the challenge of recovering hundreds of Bd-impacted amphibians worldwide is a daunting prospect.
+Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
+The results of this study provide a hopeful starting point for that endeavor, and other future efforts worldwide.
 
@@
 
-![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the stringent set of 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
 The inset map locates the study area in California.
-<!--# Allie: Add compass rose and format distance bar to match that in Yosemite map.
-Relocate pie charts to avoid the perception that all recovering populations are west of naive populations.
-In legend, provide interpretation of "Homozygous reference", "Homozygous alternate", and "Heterozygous".
-How about identifying each population as R.
-muscosa or R.
-sierrae using squares and dots, respectively?
-The current version of the figure ignores the species identities, but including them might help with interpretation of results.
-Try moving inset map to lower left portion of panel - more room there to enlarge inset?
-Remove distance scale from inset map.
---> <!--# Bree comment: Minor but the little set-in map in the central panel may be too small to be useful.
--->](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
-![Results from the splined window analyses showing outlier windows for $\pi_{diff}$ (A) and *F~ST~* (B).
-Outlier windows for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier windows for both $\pi_{diff}$ and *F~ST~* are shown in red.
+\newpage
+
+![Results from the splined window analysis showing outlier windows for $\pi_{diff}$ (A) and *F~ST~* (B).
+Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
 (C) Magnified Contig19 from (A) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 10 annotated genes known to occur in the extended MHC Class I and II regions.
-<!--# As I commented in the Results text, numerous outlier regions were identified, so how and why were the regions highlighted in C and D chosen?
-Legend text that describes the outlier categories needs additional editing to clarify the categories.
-My attempt feels only partially successful.
-Capitalize plot identifiers (A, B, C, D).
--->](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.
+](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
 
-![Median 1-year survival of each cohort of translocated frogs at the 12 recipient sites included in this study.
+\newpage
+
+![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
 Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Frog survival was estimated using CMR methods.
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-Error bars show the 2.5% and 97.5% posterior quantiles (i.e., 95% uncertainty intervals).
+Error bars show the 95% uncertainty intervals.
 When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
 
-![Results from a multilevel Bayesian model showing conditional effects of the important predictors of 1-year frog survival.
+\newpage
+
+![Results showing conditional effects of the important predictors of 1-year frog survival.
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
@@
 
-![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated frogs.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line).
+\newpage
+
+![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.
@@
 
+\newpage
+
 ![Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
-(A) Elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+(A) Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
 (B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp1.jpg){#fig-viability-supp}
+
+\newpage
 
@@
 
-Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
+```{=html}
+<!--# Comment from Allie that will be helpful in making the case that what we are seeing is evidence of strong selection/evolutionary rescue:
 
+For the question about describing what we are seeing as evolutionary rescue - I feel comfortable and confident using that wording for two reasons:
+
+1) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
+In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
+In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
+Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significant outlier SNP that has an annotation.
+and
+
+2) CORROBORATING STUDIES. For each gene mentioned above we can list many studies that have linked the genes to Bd/disease/wounds.
+In the case of GVINP1 for instance there are at least three Bd transcriptomic studies that have found this specific gene to be differentially expressed in amphibians (I currently only cite one in the paper but could add the other two: Grogan et al.
+2018, Ellison et al.
+2014a, Ellison et al.
+2014b).
+In fact, if you haven't taken a look at the Grogan paper it is really interesting because it is similar in that they are describing evolution in natural settings and they talk a lot about the GVINP1 gene.
+Even more remarkable that this gene was also linked so strongly to gill amoebic disease in salmon (explaining an astounding 20% of the resistance phenotype).
+-->
+```
 In the main manuscript, we describe the stringent set of variants.
@@
 
-<!--# Necessary? -->
+Brief description of recipient populations, perhaps in table format?
+Includes mostly individual relatively isolated lakes, but also some lake-pond-stream complexes and one stream-meadow-pond complex.
 
-Description of translocation of Bd-naive populations to Bd-positive sites
-
--   Spur Basin to Pinchot, Upper, and Ruskin Basins
-
--   Marmot to Square Lake
-
-Description of translocation of Bd-naive populations to Bd-negative sites
-
--   Marmot to Ameletus Lake
-
--   Upper Center Basin to South Center Basin
+\newpage
 
@@
 ![Results from the exome capture selection study.
-(A) Map of sampling locations in the Sierra Nevada study area showing Bd-naive populations in red and recovering populations in blue.
-Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) are shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada is shown in light gray.
-(B) PCA calculated from binary SNPs showing genomic relationship of samples.
-Numeric labels and colors match those in (A).
-(C) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
-Dashed line shows Bonferroni corrected p-value for threshold of 0.05 (liberal) and 0.01 (stringent) outliers <!--# Add line for 0.05, if that distinction would be useful.
--->.
-(D) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
+(A) PCA calculated from binary SNPs showing genomic relationship of samples.
+Numeric labels and colors match those in @fig-allelefrequencies A.
+(B) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+Dashed line shows p-value that identifies outliers.
+(C) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
 SNPs are sorted by genomic position and contigs are sorted by size.
-Red line shows Bonferroni corrected p-value for threshold of 0.01.
-<!--# Show line for threshold of 0.05 also?
--->Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.
-<!--# Map (panel A) is duplicated here and in the mapped allele frequency figure.
-I think it can be removed from this figure.
-Capitalize panel labels.
-Label/identify both the liberal and stringent outlier sets?
--->](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+Red line shows p-value that identifies outliers.
+Outlier SNPs above this threshold are highlighted and labeled.
+Bold labels indicate at least one non-synonymous SNP is present in that gene.\
+](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+
+![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
+The plot was created with SimpleSynteny @veltri2016 using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
+Asterices indicate the location of SNP outliers in TCF19 and VARS1 genes.
+Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="1500"}
+
+\newpage
 
@@
 In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="450"}
+<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="350"}
 
@@
 All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
+
+\newpage
 
@@
 
+\newpage
+
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
@@
 
+\
+
 ### Datasets
@@
 
-Dataset 2: spline_window_shared_outliers.csv <!--# The datasetcontains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
-Do we need to explain that one region is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows)?
+Dataset 2: spline_window_shared_outliers.csv <!--# The dataset contains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
+Do we need to explain that one region is a repeat because there is a 2>1 match for one of the overlaps (i.e., one Fst outlier window overlaps with 2 pi outlier windows)?
 -->
 
+\
+
 ### References

### 2023-04-23 doc/manuscript/translocation.qmd
Extensive edits to supp info, methods, add abstract

@@
 
-<!--# No more than 250 words -->
+<!--# Limit = 250 words. -->
 
-### Significance Statement
+Infectious diseases are emerging at an unprecedented rate, with devastating consequences for global biodiversity.
+The recovery of impacted taxa is hindered by the ongoing presence of disease, and few broadly applicable conservation strategies exist to overcome this obstacle.
+Rapid evolution in response to disease may produce resistant/tolerant individuals capable of persisting despite the presence of disease ("evolutionary rescue"), and although untested, such hosts could be used to reestablish extirpated populations.
+We evaluated this scenario using the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the emergence of a lethal fungal pathogen.
+Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
+We show that MYL frogs have undergone substantial evolutionary change following disease outbreaks, and that these changes are associated with increased resistance/tolerance to Bd infection in naturally recovering populations.
+Large-scale reintroduction of frogs from these populations resulted in the establishment of reproducing populations at most recipient sites despite ongoing disease, and results from viability modeling suggests that established populations have a low probability of extinction over 50 years.
+Collectively, these results provide one of the first examples of how evolution and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted taxa.
+This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
 
-<!--# Less than 120 words. Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty.  -->
+### Significance
+
+<!--# Limit = 120 words. Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty.  -->
+
+Diseases are becoming more common in wildlife and threaten many species with extinction.
+TO BE WRITTEN
 
@@
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
+
+<!--# Drop the following paragraph? -->
 
@@
 Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+We then describe the reintroduction of these resistant/tolerant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
 These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the multi-year extent of the CMR data.
@@
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
+Interestingly, this region had a low $\pi_{diff}$ value, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
 This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
 In *Xenopus*, five of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes (@fig-synteny-plot SI).
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 
-Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or subadults.
+Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
 For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
@@
 
+<!--# To meet word count limits, we will likely need to move some of the following methods to the SI.  -->
+
 ##### Sampling and Sequencing
@@
 We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four Bd-naive populations, and 29 from five recovering populations.
-These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-selectionresults A).
+These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
 Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges (@fig-selectionresults A).
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
 DNA was extracted following Qiagen DNEasy manufacturer's protocols.
@@
 Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *Rana muscosa* transcriptome (GenBank accession GKCT00000000).
+Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
 Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
@@
 To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
-Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
-<!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
+Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual SNP variants and a splined window analysis to evaluate larger genomic regions.<!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
 For example, for us genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
 The first approach A, did X, Y, and Z but had C limitations.
-The second approach B, did X, Y, and Z and D limitations."--> GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
-For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
-GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?
+The second approach B, did X, Y, and Z and D limitations."--> For the SNP analysis, we first used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic<!--# What is the null with this test (or what are the two models being compared with the LRT)?
 Probably obvious for genomics folks, but not totally clear to me.
@@
 persistent) is significant for describing variability in a SNP?
---> for the set of 148,307 high quality SNPs across 127 contigs.
-We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
-<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs.
-persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.
+--> for the resulting set of 148,307 high quality SNPs across 127 contigs using GEMMA.
+GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
+We identified outlier SNPs using a Bonferroni- corrected significance level of 0.01.
+Outliers are SNPs showing different allele frequencies between naive versus recovering populations.
+We visualized the results using a Manhattan plot and qqplot.
+We developed a more liberal set of outlier SNPs using a Bonferroni- corrected significance level of 0.05 and used this set solely for the GO analysis (see below).
 
-To detect regions of the genome under selection <!--# Improve wording to indicate that this is the second analysis that we conducted to detect regions of the genome under selection.
-Also, can we contrast this with the previous approach?
-What is this analysis doing that the other one didn't?
--->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-
-We then used the package WINGEN R package [@bishop2023] <!--# Correct citation?
-No citation was provided in the original text from Allie so I guessed at what a suitable citation would be.
---> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for F~ST~ or above/below the mean for $\pi_{diff}$.
-Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
-We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.
-
-Using the liberal set of variant outliers, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] .
-First, we retrieved the Blast hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 29 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-For the stringent set of SNP outliers, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+For each outlier SNP, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
 To do this, we first extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
@@
 
+In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
+First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We then used the WINGEN R package [@bishop2023] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
+This method also calculates a W-statistic allowing for outlier identification.
+We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for F~ST~ or above/below the mean for $\pi_{diff}$.
+Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.
+
+Using the liberal set of variant outliers, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+First, we retrieved the Blast hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 29 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+
 #### Frog population recovery
@@
 
-Translocations of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species (@fig-yosemap).
-We identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], as well as landscape genetic information [@poorten2017].
-Results indicated that these populations contained several hundred *R.
-sierrae* adults and hundreds or thousands of tadpoles, and that frog abundances were recovering from previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd infection intensity ("load", expressed as median log~10~(load) by site: `r round(bdload_min, 2)`--`r round(bdload_max, 2)`) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 ITS copies per swab)[@vredenburg2010; @joseph2018].
+For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], as well as landscape genetic information [@poorten2017].
+The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
+These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
+As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
-CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and subadult abundance.
+CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
 During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
@@
 During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and subadults.
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
 These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
@@
 Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
-
-To quantify changes in Bd load, we used an analysis workflow similar to that described above.
+We used an analysis workflow similar to that described above.
 Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
@@
 Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect. That is the model that is most analogous to a paired t-test. -->
+<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. -->
 
@@
 We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
-Although we observed yearly variability in adult survival *within* a population, the magnitude of this variability was small compared to *among* population variability ([@fig-translocation-survival]).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among population variability ([@fig-translocation-survival]).
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
@@
 
-### Acknowledgments and funding sources
+### Acknowledgments
+
+We thank the following for important contributions to this study: E.
+Hegeman and A.
+Lindauer--project and data management, field and laboratory work; A.
+Barbella and K.
+Rose--laboratory work; numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
+Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
+This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National Science Foundation Grants EF-0723563 (to C.
+Briggs), DEB-1557190 (to C.
+Briggs), and DBI-2120084 (to C.
+Richards-Zawacki).
 
@@
 
+\
 <!--# Bree comment: At some point it might be good to talk about what figures go in main text and which stay in supplement (or if we might make some multi-panel figures so that we can fit more in the main text).
@@
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
-The inset map locates the study area in California.
-](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
@@
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.
-](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
 
@@
 
+We compared frog genomes sampled in naive versus recovering populations.
 Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
@@
 
-We identified outlier SNPs using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
-In the main text we present results solely based on the stringent set.
-
 #### Results
 
-```{=html}
-<!--# Comment from Allie that will be helpful in making the case that what we are seeing is evidence of strong selection/evolutionary rescue:
+In the Results section of the main paper, we describe the stringent set of outlier SNPS (identified using a Bonferroni-corrected p-value of 0.01).
+The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
 
-For the question about describing what we are seeing as evolutionary rescue - I feel comfortable and confident using that wording for two reasons:
+Multiple genes in our stringent set of outliers have known associations with amphibian Bd resistance/tolerance.
+Of particular interest is a \~940kb region on contig 8 that we identified as under selection, and that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+Numerous studies have linked MHC genes to frog resistance/tolerance against Bd infection [e.g., @savage2011; @bataille2015].
+In addition to the genes in this region that we describe in the Results section of the main paper, other genes also deserve mention.
+For example, we identified the MHC Class III gene, heat shock protein 70 (HSP70), as occurring in this region (@fig-synteny-plot).
+Similarly, @byrne2020 identified an HSP70-binding protein as a candidate gene under selection in *Atelopus varius* populations recovering after Bd-caused declines.
+HSP70 was also highly expressed in Bd-infected versus uninfected *Xenopus tropicalis* [@rosenblum2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3 (@fig-spline-manhattan C), a gene associated with inflammation response that is highly expressed in amputated *Xenopus* tails [@fukazawa2009].
 
-1) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
-In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
-In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
-Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significant outlier SNP that has an annotation.
-and
+Additionally, we identified a region on contig 19 that had the highest $\pi_{diff}$ of any outlier window, indicative of directional selection.
+This region contained the gene, interferon-induced very large GTPase 1-like (GVINP1).
+Interferons boost immune system processes and previous studies of frogs have identified interferon-related genes as associated with Bd resistance/tolerance [@byrne2020] or have shown increased expression of this class of genes when frogs were exposed to Bd in the lab [@rosenblum2009; @ellison2014].
+Variants in GVINP1 also explained approximately 20% of the resistance to amoebic gill disease in Atlantic salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
+Furthermore, in *Litoria verreauxii alpina*, the related gene<!--# Correct?
+Or is this refering to GVINP1?
+-->, IFN-induced very large GTPase, showed lower expression in frogs from less susceptible populations versus more susceptible populations [@grogan2018].
 
-2) CORROBORATING STUDIES. For each gene mentioned above we can list many studies that have linked the genes to Bd/disease/wounds.
-In the case of GVINP1 for instance there are at least three Bd transcriptomic studies that have found this specific gene to be differentially expressed in amphibians (I currently only cite one in the paper but could add the other two: Grogan et al.
-2018, Ellison et al.
-2014a, Ellison et al.
-2014b).
-In fact, if you haven't taken a look at the Grogan paper it is really interesting because it is similar in that they are describing evolution in natural settings and they talk a lot about the GVINP1 gene.
-Even more remarkable that this gene was also linked so strongly to gill amoebic disease in salmon (explaining an astounding 20% of the resistance phenotype).
--->
-```
-In the main manuscript, we describe the stringent set of variants.
-In addition, the liberal set of variants included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
-
-Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
-Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
-These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
-For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs.
-uninfected *Xenopus tropicalis* [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
-These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
-
-Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@byrne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008; @ellison2014].
-Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
-Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
-Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above.
-But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection.
-If it feels like we can't justify discussing it we can cut it and just focus on the other two.
--->.
-This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
-Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.
-
-Based on the SNP and splined-window analyses , no GO terms were significantly over-represented in the set of annotated genes associated with the outlier SNPs or regions.
+Finally, we identified a region characterized by high *F~ST~* and low $\pi_{diff}$ and that contained the complement genes C6 and C7.
+This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
 
@@
 
-### Translocations
-
-Brief description of recipient populations, perhaps in table format?
+<!--# Brief description of recipient populations, perhaps in table format?
 Includes mostly individual relatively isolated lakes, but also some lake-pond-stream complexes and one stream-meadow-pond complex.
+-->
 
@@
 ### Figures
+
+\
 
@@
 Asterices indicate the location of SNP outliers in TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="1500"}
+Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="600"}
 
@@
 
-<!--# How to insert >= symbol in header and table cells> -->
+\
+
+<!--# How to insert >= symbol in header and table cells? Move second sentence of caption to footnote.-->
 
@@
 #| label: tbl-survival-earlylate
-#| tbl-cap: "Association between the proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) and probability of survival (< 0.5 or >= 0.5)."
+#| tbl-cap: "Association between the proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) and probability of survival (< 0.5 or >= 0.5).
+Based on the viability analysis, survival probabilities < 0.5 and >= 0.5 produced 50-year extinction probabilities of 1 and < 0.5, respectively."
 #| echo: FALSE
@@
 ### References
+
+\

### 2023-04-28 doc/manuscript/translocation.qmd
Update translocation.qmd

@@
 editor: visual
+editor_options:
+  markdown:
+    wrap: sentence
 ---
@@
 Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and across generations, allowing for the establishment of reintroduced populations.
+Third, the fitness benefits gained through selection are maintained beyond the natal environment and across generations, allowing for the establishment of reintroduced populations.
 And fourth, reintroduced populations show long-term growth and persistence.
-To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].
+To date, few of these conditions are documented, and reintroducing hosts from recovering populations is largely untested [@scheele2021; @mendelson2019].
 
-The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
+The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
 Once the most common amphibian in the high elevation portion of the California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
@@
 
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
@@
 This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
 Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
@@
 sierrae*.
-Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
+Finally, Bd loads were fairly consistent before versus after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
 
@@
 Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
+<!--# Consider condensing the above paragraph and providing details in SI.
+Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon." -->
 
@@
 Disease-induced population declines are impacting global biodiversity [@daszak2000], and strategies to recover affected species are lacking.
-Recovery in this context is likely to be particularly challenging because typically recovery will need to occur in the presence of ongoing disease.
-One potential strategy, and the one we employ in the current study, utilizes populations of resistant/tolerant individuals as a source of animals with which to reestablish extirpated populations in the presence of disease.
+Here, we used populations of resistant/tolerant individuals to reestablish extirpated populations, despite ongoing disease.
 Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
@@
 Nonetheless, in the following narrative, we use evolutionary rescue as an informative, albeit tentative framework.
+<!-- I am wondering whether it is worth digging a bit deeper into the "difficult and perhaps impossible" task alluded to above.
+e.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data -->
 
@@
 Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
-For example, the major histocompatibility complex (MHC) plays an important role in immunity.
+For example, the MHC plays an important role in immunity.
 In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
@@
 In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Retain or delete? -->.
-Collectively, these results suggest that the documented natural recovery of MYL frogs following Bd-caused declines [@knapp2016] is consistent with evolutionary rescue.
 
@@
 First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
 We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ ("$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$., and the probability of a juvenile successfully recruiting to an adult $\omega$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$., and the probability of a juvenile successfully recruiting to an adult $\omega$.
 
@@
 
-![Results from the splined window analysis showing outlier windows for $\pi_{diff}$ (A) and *F~ST~* (B).
+![Results from the splined window analysis showing outlier windows for the difference in nucleotide diversity $\pi_{diff}$ (A) and *F~ST~* (B).
 Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
@@
 
-![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
-The red line shows where $\lambda = 1$.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript1.jpg){#fig-viability}
+![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate \$\\lambda\$ for different values of yearly adult survival probability \$\\sigma\_{A_R}\$ and probability of recruitment \$\\omega\$, given the parameterized, deterministic model.
+Colored points show the predicted \$\\lambda\$ values for the twelve translocated populations when probability of recruitment into the adult class is \$\\omega = 0.3\$ (indicated by the dashed line).
+The red line shows where \$\\lambda = 1\$.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in \$\\omega\$, and different mean values of \$\\omega\$.
+The dashed line shows where \$\\omega = 0.3\$, for comparison to (A).
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript1.jpg){#fig-viability} <!-- Consider a diverging color ramp for panel A, e.g., where lambda=1 is white, lambda>1 is blue, lambda<1 is red -->
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
+|:----------------------------|:-----------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-05-05 doc/manuscript/PNAS-template-empty.tex
Convert qmd to latex PNAS format

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b]{Roland A. Knapp}
\author[c,1]{Mark Q. Wilber} 
\author[d,e,1]{Allison Q. Byrne}
\author[f,g]{Maxwell B. Joseph}
\author[a,b]{Thomas C. Smith}
\author[d,e]{Andrew P. Rothstein}
\author[h]{Robert L. Grasso}
\author[d,e]{Erica Bree Rosenblum}

\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
\affil[g]{Planet, San Francisco, CA, 94107}
\affil[h]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}

% Please give the surname of the lead author for the running footer
\leadauthor{Knapp}

% Please add a significance statement to explain the relevance of your work
\significancestatement{}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{}
\authordeclaration{}
\equalauthors{\textsuperscript{1}M.W. contributed equally to this work with A.B.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{}

\begin{abstract}
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% Introduction

% Results

% Discussion

\matmethods{}

\showmatmethods{} % Display the Materials and Methods section

\acknow{}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{translocation}

% Figures

\end{document}

### 2023-05-05 doc/manuscript/translocation.qmd
Convert qmd to latex PNAS format

--- 2023-04-23 doc/manuscript/translocation.qmd
@@
 csl: pnas.csl
-format: docx
+# format: docx
+format :
+  latex:
+    cite-method: natbib
 editor: visual
@@
 Exact PCR Duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0, @van2020).
+Variants were then called following GATK best practices (v.4.2.0.0, [@van2020]).
 Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
@@
 Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
-For details of survey and swab analysis methods, see @joseph2018.
+For details of survey and swab analysis methods, see [@joseph2018].
 
@@
 Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
-Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
+Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013; @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
 Both approaches utilized the loo package [@vehtari2022].
@@
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
-The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies}
 
@@
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan}
 
@@
 Error bars show the 95% uncertainty intervals.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival}
 
@@
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
 
@@
 For example, we identified the MHC Class III gene, heat shock protein 70 (HSP70), as occurring in this region (@fig-synteny-plot).
-Similarly, @byrne2020 identified an HSP70-binding protein as a candidate gene under selection in *Atelopus varius* populations recovering after Bd-caused declines.
+Similarly, [@byrne2020] identified an HSP70-binding protein as a candidate gene under selection in *Atelopus varius* populations recovering after Bd-caused declines.
 HSP70 was also highly expressed in Bd-infected versus uninfected *Xenopus tropicalis* [@rosenblum2009].
@@
 Bold labels indicate at least one non-synonymous SNP is present in that gene.\
-](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+](figures/pca_qq_manhattan.png){#fig-selectionresults}
 
 ![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
-The plot was created with SimpleSynteny @veltri2016 using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
+The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
 Asterices indicate the location of SNP outliers in TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="600"}
+Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot}
 
@@
 In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="350"}
+<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap}
 
 ![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
+In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens}
 
 ![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}
+Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter}
 
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
 
@@
 The large error bars for 70550 reflect the large within and among individual variability in yearly survival probability at this site (as also reflected in one-year survival probabilities in [@fig-translocation-survival]).
-All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
+All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs}
 

### 2023-05-05 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -1,0 +1,236 @@
+import subprocess
+import sys
+import re
+
+"""
+Convert Quarto to PNAS format for translocation manucript
+
+Generates two files
+
+1. transoclation_pnas.tex : Main text in PNAS format
+2. transoclation_pnas_SI.tex : SI in PNAS format
+"""
+
+# Files for converting
+args = ['translocation.qmd', 'translocation.tex']
+
+### Build the main manuscript ####
+
+# Convert from qmd to latex with pandoc
+subprocess.call(['quarto', 'render', args[0], '--to', "latex"])
+
+# Read in the latex manuscript
+with open(args[1], 'r') as fin:
+    latex_manuscript = fin.readlines()
+latex_manuscript = "".join(latex_manuscript)
+
+with open("PNAS-template-empty.tex") as fin:
+	pnas_template = fin.readlines()
+pnas_template = "".join(pnas_template)
+
+# Fill in each section of manuscript
+
+## Title ## 
+
+title_start = latex_manuscript.find("\\subsection{Title")
+title_end = latex_manuscript[title_start:].find("}") + title_start
+title = latex_manuscript[title_start:title_end].replace("\n", " ").split("\\subsection{Title: ")[1]
+
+# Insert title
+pnas_ms = pnas_template.replace("\\title{}", "\\title{" + title + "}")
+
+## Abstract ##
+
+abkey1 = "\\hypertarget{abstract}{%\n\\subsubsection{Abstract}\\label{abstract}}"
+abkey2 = "\\hypertarget{significance}"
+abstract_start = latex_manuscript.find(abkey1)
+abstract_end = latex_manuscript.find(abkey2)
+abstract = latex_manuscript[abstract_start:abstract_end].split(abkey1)[1].strip("\n")
+
+pnas_ms = pnas_ms.replace("\\begin{abstract}\n\\end{abstract}", 
+						  "\\begin{abstract}\n" + abstract + "\n\\end{abstract}")
+
+
+## Significance statement ##
+
+sigkey = "\\hypertarget{significance}{%\n\\subsubsection{Significance}\\label{significance}}"
+intkey1 = "\\hypertarget{introduction}{%\n\\subsubsection{Introduction}\\label{introduction}}"
+sig_start = latex_manuscript.find(sigkey)
[…]

### 2023-05-08 doc/manuscript/PNAS-template-supp-info-empty.tex

\documentclass[9pt,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnassupportinginfo}
\usepackage{longtable}

\title{}
\author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
\correspondingauthor{Corresponding Author name. Roland A. Knapp \\ E-mail: roland.knapp@ucsb.edu}

\begin{document}

\maketitle

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
\SItext

% Figures

%%% Add this line AFTER all your figures and tables
\FloatBarrier

\dataset{gemma\_outliers.csv}{Description needed}

\dataset{spline\_window\_shared\_outliers.csv}{Description needed}

\bibliography{translocation}

\end{document}

### 2023-05-08 doc/manuscript/translocation_pnas.tex

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b]{Roland A. Knapp}
\author[c,1]{Mark Q. Wilber} 
\author[d,e,1]{Allison Q. Byrne}
\author[f,g]{Maxwell B. Joseph}
\author[a,b]{Thomas C. Smith}
\author[d,e]{Andrew P. Rothstein}
\author[h]{Robert L. Grasso}
\author[d,e]{Erica Bree Rosenblum}

\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
\affil[g]{Planet, San Francisco, CA, 94107}
\affil[h]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}

% Please give the surname of the lead author for the running footer
\leadauthor{Knapp}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Diseases are becoming more common in wildlife and threaten many species
with extinction. TO BE WRITTEN}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{}
\authordeclaration{}
\equalauthors{\textsuperscript{1}M.W. contributed equally to this work with A.B.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{}

\begin{abstract}
Infectious diseases are emerging at an unprecedented rate, with
devastating consequences for global biodiversity. The recovery of
impacted taxa is hindered by the ongoing presence of disease, and few
broadly applicable conservation strategies exist to overcome this
obstacle. Rapid evolution in response to disease may produce
resistant/tolerant individuals capable of persisting despite the
presence of disease (``evolutionary rescue''), and although untested,
such hosts could be used to reestablish extirpated populations. We
evaluated this scenario using the once-common mountain yellow-legged
(MYL) frog, which is threatened with extinction by the emergence of a
lethal fungal pathogen. Although most MYL frog populations are
extirpated following disease outbreaks, some persist and eventually
begin to recover. We show that MYL frogs have undergone substantial
evolutionary change following disease outbreaks, and that these changes
are associated with increased resistance/tolerance to Bd infection in
naturally recovering populations. Large-scale reintroduction of frogs
from these populations resulted in the establishment of reproducing
populations at most recipient sites despite ongoing disease, and results
from viability modeling suggests that established populations have a low
probability of extinction over 50 years. Collectively, these results
provide one of the first examples of how evolution and the
reintroduction of resistant/tolerant individuals can allow the
landscape-scale recovery of disease-impacted taxa. This example has
potentially broad implications for the many taxa worldwide that are
threatened with extinction by emerging diseases.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\dropcap{E}merging infectious diseases pose a severe threat to wildlife
populations \citep{daszak2000}, and have caused dramatic declines and
even extinctions in mammals, birds, amphibians, echinoderms, and other
taxa \citep{hewson2014, samuel2015, scheele2019, cunningham2021}.
Importantly, some populations previously decimated by disease have
subsequently recovered to varying degrees
\citep{newell2013, voyles2018, knapp2016} and might provide critical
insights for species conservation efforts. Unfortunately, given the
rarity of wildlife recoveries in the presence of disease, we know little
about the mechanisms driving population recovery
\citep{brannelly2021, russell2020} and how these mechanisms can be
leveraged to re-establish host populations across large spatial scales
\citep{mendelson2019}. At a time when novel infectious diseases are
emerging at an unprecedented rate \citep{daszak2000, fisher2012}, this
knowledge gap has far-reaching consequences for global biodiversity.

Host evolution following disease-induced declines is one mechanism by
which hosts can recover \citep{carlson2014, searle2020}. ``Evolutionary
rescue'' describes rapid evolutionary change that increases the
frequency of adaptive alleles and thereby restores positive population
growth and prevents extinction \citep{carlson2014}. Evolutionary changes
following disease-induced declines are known from a diversity of
wildlife populations
\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
often result in increased host resistance and/or tolerance. However,
linking these selective responses to population recoveries - as
necessary for evolutionary rescue - is difficult and remains rare. In
addition to promoting recovery of disease-impacted populations, the
resistant/tolerant individuals in recovering populations can, in theory,
be used to reestablish populations that were extirpated following
disease emergence \citep[e.g.,][]{mendelson2019}. To date, the use of
resistant/tolerant individuals for species recovery is rare \citep[but
see][]{joseph2018}, and studies that evaluate its potential utility are
critically needed.

There are four conditions that will often determine the feasibility of
using individuals from populations that are recovering following
disease-induced declines to reestablish extirpated populations. First,
populations that survive following disease emergence evolve increased
resistance/tolerance. Second, increased resistance/tolerance contributes
to higher individual fitness in the presence of disease and allows
positive population growth. Third, the fitness benefits gained through
selection for increased resistance are maintained beyond the natal
environment and across generations, allowing for the establishment of
reintroduced populations. And fourth, reintroduced populations show
long-term growth and persistence. To date, few of these conditions are
documented for any single species impacted by disease, and as a result
the reintroduction of resistant hosts as a means to recover species of
conservation concern remains largely an untested idea
\citep{scheele2021, mendelson2019}.

The mountain yellow-legged (MYL) frog, comprised of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the unprecedented global declines of amphibians caused by
the recently-emerged amphibian chytrid fungus
\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
most common amphibian in the high elevation portion of the California's
Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
disappeared from more than 90\% of its historical range during the past
century \citep{vredenburg2007}. Due to the severity of its decline and
the increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000}, and the arrival of \emph{B.
dendrobatidis} (Bd) in the 1960s and its subsequent spread
\citep{vredenburg2019} caused additional widespread population
extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
declines are fundamentally different from the fish-caused declines
because removal of fish populations is feasible and results in the rapid
recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
appears to persist in habitats even in the absence of amphibian hosts
\citep{walker2007}, and therefore represents a long-term alteration of
invaded ecosystems that amphibians will need to overcome to reestablish
populations.

Despite the catastrophic impact of Bd on MYL frogs in which most
Bd-naive populations are driven to extirpation following Bd arrival
\citep{vredenburg2010}, some populations have persisted after epizootics
\citep{briggs2010} and are now recovering \citep{knapp2016}. Frogs in
these recovering populations show reduced susceptibility to Bd
infection, consistent with selection for more resistant/tolerant
genotypes \citep{knapp2016}. Furthermore, exploratory efforts to
reintroduce frogs collected from recovering populations to sites from
which frogs were previously extirpated indicate that population
establishment is possible despite ongoing Bd infection
\citep{joseph2018}.

In the current study, we combine results from genomic analyses, a
15-year landscape-scale frog reintroduction effort, and detailed
population models to provide an unprecedented example of how evolution
in response to disease allows the successful reestablishment of
extirpated MYL frog populations. Specifically, we used exome capture
methods to show that MYL frogs have undergone substantial evolutionary
change associated with Bd exposure and that may provide a genetic basis
for the observed increased resistance/tolerance to Bd infection observed
in naturally recovering populations \citep{knapp2016}. We then describe
the reintroduction of these resistant/tolerant frogs to 24 sites and the
dynamics and drivers of population establishment using results from
long-term capture-mark-recapture (CMR) surveys. These data are also used
to parameterize a population model that extends our inferences of
population establishment well beyond the multi-year extent of the CMR
data. Collectively, these results provide one of the first examples of
how evolution and the reintroduction of resistant/tolerant animals can
allow the large-scale recovery of disease-impacted species. This example
exhibits several of the key characteristics of evolutionary rescue, and
has potentially broad implications for the many taxa worldwide that are
threatened with extinction by emerging diseases.

\section*{Results}

\subsection*{Frog evolution in response to Bd}

To determine whether frog populations show an evolutionary response to
Bd, we compared genomes of frogs in populations with different histories
of Bd exposure. Specifically, we compared frog genomes sampled in four
populations that have not yet experienced a Bd-caused epizootic
(``naive'') \citep{zhou2015} versus in five populations that experienced
a Bd epizootic during the past several decades and have since recovered
to varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
\citep{knapp2016, vredenburg2010}. These recovering populations persist
in an enzootic state \citep{briggs2010}, characterized by high Bd
prevalence and, in adults, moderate Bd infection intensities (``load'')
that are typically well below the level expected to cause mortality
\citep{vredenburg2010}.

We conducted a principal component analysis (PCA) of the genomic data to
describe the association between sampled populations, and then used two
complementary approaches to identify regions of the genome under
selection. First, we used a multivariate linear mixed model to evaluate
associations between population type (i.e., naive versus recovering) and
specific single nucleotide polymorphisms (SNPs). Second, we used a
splined window analysis to identify larger genomic regions showing
differences between population types in \emph{F\textsubscript{ST}} and
nucleotide diversity (\(\pi_{diff}\) = \(\pi_{naive}\) -
\(\pi_{recovering}\)).

Individual frogs clustered into 3 separate groups in principal component
space (Figure S1A), and clusters reflected
the strong signature of isolation by distance that is characteristic of
MYL frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
contained at least one population from both the naive and recovering
groups, allowing us to distinguish between (i) allelic associations of
individuals sampled in naive versus recovering populations, and (ii)
allelic associations resulting from population structure and genetic
drift.

Results from the SNP and splined window analyses show that recovering
populations have signatures of strong selection on multiple regions of
the genome. The SNP analysis identified 11 outlier SNPs from 7 distinct
genes across 4 contigs (Figure S1B,C, SI
Dataset 1). One of the 7 identified genes did not have an associated
annotation. Frequency differences between the naive and recovering
populations ranged from 41\% to 86\%. Most of these SNPs showed
frequency differences in only a subset of the sampled populations
(Figure~\ref{fig-allelefrequencies} A, B), but the outlier SNP in the
RIN3 gene showed consistent differences in frequencies across all
populations (Figure~\ref{fig-allelefrequencies} C). This is suggestive
of strong selection at this locus associated with exposure of frogs to
Bd.

The splined window analysis identified 58 outlier regions for
\emph{F\textsubscript{ST}} and 33 outlier regions for \(\pi_{diff}\)
(Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
outliers for both metrics and 2 of these shared regions also contained
outlier SNPs. A total of 33 annotated genes were found in the 9 shared
regions (SI Dataset 2). The largest \(\pi_{diff}\), indicative of
directional selection, occurred in a 163kb region on Contig19, 12.9Mb
upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C).
This region contains approximately 500 SNPs and one annotated gene
called ``interferon-induced very large GTPase 1-like'' (GVINP1).
Additionally, a shared outlier region on Contig1 contained two
complement factor genes (C6 and C7). Interestingly, this region had a
low \(\pi_{diff}\) value, consistent with balancing selection. Finally,
one shared outlier region on Contig8 contained one outlier SNP (TCF19)
and was within 360kb of another outlier SNP (VARS)
(Figure~\ref{fig-spline-manhattan} D, Figure S2).
This region (854kb from the beginning of the outlier window to the VARS
SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,
CCHCR1, TCF19) and three occur in the extended MHC Class III region
(HSP70, LSM2, VARS) \citep{ohta2006}. Therefore, this region under
selection is part of the extended MHC Class I and III complex and shows
synteny with other amphibian genomes.

\subsection*{Frog population recovery}

In this section, we describe the outcome of \emph{R. sierrae}
reintroductions conducted in Yosemite National Park as part of a 15-year
effort to reestablish populations of this endangered species
(2006--2020). Each of the 24 reintroductions involved collection of
adult frogs from 1 of 3 recovering, Bd-positive ``donor'' populations
(Figure S3) and translocating them to 1 of 12 nearby
recipient sites \citep{seddon2014}. The donor populations included 2 of
the 5 recovering populations in the frog evolution study described above
(Figure~\ref{fig-allelefrequencies} A: population 1 and 4). Following
translocation, adult survival and recruitment of new adults were
estimated from CMR surveys and counts of tadpoles and juveniles were
obtained from visual encounter surveys (VES).

Estimates of frog survival during the 1 year following translocation
indicate that survival was highly variable between recipient sites, but
relatively constant within recipient sites (for the subset of sites that
received multiple translocations;
Figure~\ref{fig-translocation-survival}). As such, the average survival
of frogs in the first translocation to a site was an important predictor
of survival of frogs in subsequent translocations to that site
(Figure S4). This suggests an important
effect of site characteristics and/or (site * frog) interactions on frog
survival. In addition, 1-year survival was higher for frogs translocated
later in the study period than earlier
(Table S1), which we suggest resulted
primarily from our improved ability over time to choose recipient sites
with higher habitat quality for \emph{R. sierrae}. Finally, Bd loads
changed little between immediately prior to translocation versus during
the one year after translocation (Figure S5), and loads were nearly always well below the level indicative of
severe chytridiomycosis and associated frog mortality (5.8 ITS copies on
a log\textsubscript{10} scale) \citep{vredenburg2010, joseph2018}.

Of the 12 translocated populations, 9 (0.75) showed evidence of
successful reproduction in subsequent years, as indicated by the
presence of tadpoles and/or juveniles. For these 9 populations, one or
both life stages were detected in nearly all survey-years following
translocation (proportion of survey-years: median = 0.9, range =
0.29--1). These same populations were also those in which recruitment of
new adults (i.e., progeny of translocated individuals) was detected. As
with early life stages, recruits were detected in the majority of
post-translocation survey-years (proportion of survey-years: median =
0.79, range = 0.12--1). In summary, survey results suggest that
translocations resulted in the establishment of reproducing MYL frog
populations at most recipient sites despite the ongoing presence of Bd.

To determine whether Bd load had a negative effect on adult survival, as
would be expected if frogs were highly susceptible to Bd infection, we
developed linear and multilevel models. The best model of 1-year frog
survival identified several important predictors, but Bd load at the
time of translocation was not among them (bd\_load;
Figure S6). Important predictors included
winter severity in the year following translocation (snow\_t1), site
elevation (elevation), and donor population (donor\_72996;
Figure~\ref{fig-cond-effects}). Males had somewhat higher survival than
females, but this effect was only marginally important (sex\_male;
Figure~\ref{fig-cond-effects}, Figure S6).
Collectively, these results are consistent with frogs in recovering
populations having evolved sufficient resistance/tolerance to minimize
the adverse effects of Bd infection. In addition, the moderate Bd loads
that characterized frogs following translocation
(Figure S5) and relatively high survival of
frogs in numerous translocated cohorts
(Figure~\ref{fig-translocation-survival}) suggests that frogs
translocated from recovering populations can maintain the benefits of
resistance/tolerance in non-natal habitats. Finally, in 3 locations
where longer CMR time series allowed us to assess the survival of new
adults recruited to the population, naturally recruited adults had
equivalent or higher survival probabilities than the originally
translocated adults (Figure S7). Therefore,
frog resistance in non-natal habitats is maintained across generations.

\subsection*{Long-term population viability}

Results from frog translocations indicated that adult survival was often
relatively high and that most populations showed evidence of successful
reproduction and recruitment (described above). Although suggestive of
population establishment, a decade or more of surveys may be necessary
to confirm that populations are in fact self-sustaining
\citep{joseph2018}. To extend our inferences of population establishment
beyond those possible from the 2-16 years of site-specific CMR data, we
developed a population viability model. Specifically, to test whether
the observed yearly adult survival probabilities in translocated
populations were sufficient to promote long-term viability, we built a
stage-structured matrix model that captured known frog demography and
included demographic and environmental stochasticity. We parameterized
the model using CMR data from translocated populations and known life
history values in this system (Table S2). Given
observed yearly adult survival probabilities (from analysis of CMR data)
and a baseline yearly recruitment probability \(\omega\) of 0.3 (i.e.,
the probability that, following survival, juveniles successfully recruit
to the adult class), our stage-structured model predicted that six of
twelve translocated populations should experience a long-run growth rate
\(\lambda\) greater than 1 in the presence of Bd
(Figure~\ref{fig-viability} A; median predicted \(\lambda\) ranges from
1.15-1.36 for these six populations). These six populations all had
observed yearly adult survival greater than 0.5. When we reduced the
probability of successful recruitment \(\omega\) to 0.1 (a highly
conservative estimate), long-run growth rate of five of these six
populations was still greater than 1 (Figure~\ref{fig-viability} A). In
contrast, for the other six translocated populations, where yearly adult
survival was substantially less than 0.5, our model predicted that it
was unlikely that long-run growth rate in these populations was greater
than 1, even when the probability of successful recruitment to the adult
life stage was near 0.4 (Figure~\ref{fig-viability} A). A sensitivity
analysis of long-run growth rate \(\lambda\) to yearly adult survival
probability of naturally recruited individuals \(\sigma_{A_R}\),
recruitment probability \(\omega\), and female fecundity \emph{F} showed
that \(\lambda\) was most sensitive to small changes in \(\sigma_{A_R}\)
(Figure~\ref{fig-viability-supp} A). Note that we assumed
\(\sigma_{A_R}\) was equivalent to the survival probability of
translocated adults \(\sigma_{A_T}\), except for three populations where
we had \textgreater{} 5 years of CMR data and recovering populations
that together allowed us to estimate \(\sigma_{A_R}\) and
\(\sigma_{A_T}\) separately (Figure S7).

We then incorporated demographic stochasticity (i.e., probabilistic
births and deaths of individuals in a population) and environmental
variability in the probability of juvenile recruitment to the adult
class \(\omega\) (the transition that we expect to be the most subject
to environmental stochasticity in the presence of Bd) into the model
described above and used it to predict long-term population viability.
Our model predicted that, following a single introduction of 40 adult
individuals into a population and using the baseline value for
\(\omega\) of 0.3, the six populations with the highest adult survival
probabilities had 50-year extinction probabilities of 0 to 0.32;
Figure~\ref{fig-viability} B). This indicates strong potential for
long-term persistence in the presence of Bd and environmental
variability in recruitment (Figure~\ref{fig-viability} B). When we
decreased \(\omega\) from 0.3 to a conservative estimate of 0.15, 50
year extinction probability remained less than 0.5 for three populations
and was 0.79 to 0.94 for three additional populations
(Figure~\ref{fig-viability} B). In contrast, for the six populations
where \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was
always predicted to be 1. To test the validity of our model predictions,
we demonstrated that our stochastic model could describe the general
recovery trajectory of our translocated population with the longest
survey history (Figure~\ref{fig-viability-supp} B; population 70550,
surveyed for 16 years).

In summary, our model demonstrates that given observed yearly adult
survival probabilities of translocated frogs, 50\% of our translocated
populations have a high probability of population growth and long-term
viability in the presence of Bd. This is likely a conservative estimate
because there is evidence that naturally-recruited adults have higher
survival probability than translocated adults
(Figure S7), but we considered these
probabilities to be equal in all but three of our populations where we
had sufficient data to distinguish these different probabilities. In
addition, 5 of the 7 populations translocated after 2013 were predicted
to have long-term viability, compared to only 1 of 5 populations
translocated prior to 2013 (Table S1). Thus,
there is some trend suggesting that our ability to select high quality
habitats that promote long-term viability of translocated populations
improved over time.

\section*{Discussion}

Disease-induced population declines are impacting global biodiversity
\citep{daszak2000}, and strategies to recover affected species are
lacking. Recovery in this context is likely to be particularly
challenging because typically recovery will need to occur in the
presence of ongoing disease. One potential strategy, and the one we
employ in the current study, utilizes populations of resistant/tolerant
individuals as a source of animals with which to reestablish extirpated
populations in the presence of disease. Although previously proposed
\citep{mendelson2019}, this strategy has rarely been successfully
implemented (see \citep{joseph2018} for a small-scale example). In
particular, although numerous studies have documented signatures of
selection in hosts following disease-induced declines
\citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
any that used individuals bearing the resulting adaptive alleles to
reestablish extirpated populations. The results of our study provide
evidence that (i) MYL frogs have experienced strong selection following
exposure to a recently-emerged lethal fungal disease (Bd) and that some
of the genomic regions under selection are associated with genes related
to disease resistance/tolerance, and (ii) reintroduction of individuals
from these resistant/tolerant populations allows the successful
reestablishment of viable frog populations despite ongoing Bd infection.
Although these results are generally consistent with expectations under
evolutionary rescue \citep{carlson2014}, obtaining strong evidence in
wild animal populations for each of the associated causal links (i.e.,
selection for adaptive alleles \(\rightarrow\) increase in host fitness
in the presence of the pathogen \(\rightarrow\) positive population
growth and recovery) is difficult and perhaps impossible. Nonetheless,
in the following narrative, we use evolutionary rescue as an
informative, albeit tentative framework.

Previous field studies in MYL frogs show that frog-Bd dynamics and frog
survival in the presence of Bd are fundamentally different between naive
and recovering populations. Following the arrival and establishment of
Bd in previously-naive populations, adult frogs develop high Bd loads
that lead to mass die-offs \citep{vredenburg2010}. In contrast, in
recovering populations adult frogs typically have low-to-moderate and
relatively constant Bd loads and mass die-offs are not observed
\citep[see also Figure S5 - before
translocation]{briggs2010, knapp2011}. The differences in Bd load of
frogs from naive and recovering populations are also observed in
controlled laboratory studies \citep[Figure 4]{knapp2016}. The observed
reduced susceptibility of frogs from recovering populations could in
theory be due to several factors including natural selection for more
resistant/tolerant genotypes and/or acquired immunity, but until now
evidence to evaluate the role of evolution was lacking.

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
substantial differences in allele frequencies and larger genomic regions
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the major
histocompatibility complex (MHC) plays an important role in immunity. In
our study, we identified a region that shows evidence of selection in
recovering populations and contains eight genes associated with either
the MHC Class I or Class III regions. These results corroborate numerous
previous studies linking MHC genes to amphibian resistance/tolerance
against Bd \citep[e.g.,][]{savage2011, bataille2015}. Similarly, the
region with the strongest indication of directional selection contains
the gene, GVINP1. Several previous studies have found this gene to be
differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018, ellison2014}. This gene was also strongly
linked to disease in salmon, explaining a remarkable 20\% of the
resistance phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3
gene showed a consistent pattern of allele frequency differences across
all nine of the frog populations sampled in this study, indicating
consistent selection in populations distributed across a wide geographic
area. In the SI, we provide additional details on these and other genes
identified in our study as under selection . Collectively, these results
suggest that the documented natural recovery of MYL frogs following
Bd-caused declines \citep{knapp2016} is consistent with evolutionary
rescue.

Reintroduction of resistant/tolerant MYL frogs was remarkably successful
in reestablishing viable populations in the presence of Bd. Of the 12
translocated populations, approximately 80\% showed evidence of both
successful reproduction and recruitment of new adults. Bd loads measured
before and after translocation were similar and generally remained well
below the level at which frog mortality occurs. Annual survival for 12
of the 24 translocated cohorts exceeded 50\%, and \textgreater{} 70\% of
translocated cohorts had survival above this 50\% level when the
earliest translocations are excluded (i.e., translocations conducted
when methods were still being refined). The fact that the relatively low
Bd loads and correspondingly high frog survival in the presence of Bd
was maintained between the natal/donor sites and the recipient sites
indicates that these characteristics of naturally-recovering populations
were not solely an effect of the natal/donor site or site * frog
interactions, but instead were due at least in part to
resistance/tolerance inherent in the frogs. In addition, the relatively
high survival of translocated frogs was maintained in their progeny, as
expected if resistance/tolerance has a genetic basis. Results from the
population viability model were also encouraging. In particular,
translocated populations with \textgreater{} 50\% survival in the first
year post-translocation were predicted to have a low probability of
extinction over 50 years (probability of extinction = 0 - 0.32). In
light of the generally low success rate of amphibian reintroduction
efforts \citep{dodd2005}, our success in reestablishing MYL frog
populations via translocation of resistant/tolerant individuals is
striking, and even more so given that this success occurred despite
ongoing infection of frogs with Bd.

Despite the resistance/tolerance of adult MYL frogs in recovering
populations, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations, Bd
infection and load had detectable effects on the survival of adults and
may have influenced population establishment \citep[sites referred to as
``Alpine'' and ``Subalpine'' are referenced as ``70550'' and ``70505''
in the current study]{joseph2018}. Applying similar analyses to all 12
of our translocated populations would likely provide a broader
perspective of the ongoing effect of Bd. In addition to these important
but relatively subtle effects on adults, MYL frogs immediately following
metamorphosis (``metamorphs'') are highly susceptible to Bd infection
\citep{ellison2018} and experience high mortality. This high
susceptibility of metamorphs is documented for numerous species of
anurans, and may result from the poorly developed immune system
characteristic of this early life stage \citep{humphries2022}. In
naturally recovering and translocated MYL frog populations, we suggest
that the high mortality of metamorphs is an important limitation on
subsequent recruitment of new adults. Therefore, although adult MYL
frogs appear relatively resistant/tolerant, Bd infection continues to
have important limiting effects on recovering populations \citep[see
also][]{hollanders2022}.

The recent emergence of Bd worldwide has contributed to the decline of
more than 500 amphibian species, nearly 100 of which now are extinct in
the wild \citep{scheele2019}. This unprecedented impact on global
amphibian biodiversity is compounded by the lack of any effective and
broadly applicable strategies to reverse these impacts
\citep{garner2016, knapp2022}. Importantly, in addition to the natural
recovery documented for MYL frogs \citep{knapp2016}, other species are
also showing evidence of post-epizootic recovery in the presence of Bd
\citep{scheele2017, voyles2018} and suggest the possibility of using
animals from these recovering populations to reestablish extirpated
populations. As with MYL frogs, the feasibility and long-term success of
such efforts will depend on the availability of robust donor populations
containing individuals that have acquired via evolution the adaptive
alleles necessary to allow frog survival and population growth in the
presence of Bd. Despite the hopeful example of successful
reestablishment of MYL frogs via evolutionary rescue and
reintroductions, the challenge of recovering hundreds of Bd-impacted
amphibians worldwide is a daunting prospect. Although we now have a
proven strategy to reestablish extirpated MYL frog populations, recovery
across their large historical range will require substantial resources
over many decades. The results of this study provide a hopeful starting
point for that endeavor, and other future efforts worldwide.

\matmethods{
\hypertarget{frog-evolution-in-response-to-bd-1}{%
\section*{Frog evolution in response to
Bd}\label{frog-evolution-in-response-to-bd-1}}

\hypertarget{sampling-and-sequencing}{%
\subsection*{Sampling and Sequencing}\label{sampling-and-sequencing}}

We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from four
Bd-naive populations, and 29 from five recovering populations. These
populations were located in the southern Sierra Nevada, from northern
Yosemite National Park to northern Sequoia National Park
(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
frogs per population. To minimize potential confounding effects caused
by variation in frog genotypes across latitude
\citep[e.g.,][]{rothstein2020}, we selected sampling sites such that
both population types were represented across similar latitudinal
ranges. DNA was extracted following Qiagen DNEasy manufacturer's
protocols. We sequenced the samples using an exome capture assay as
described in Byrne et al.~(in review). Briefly, genomic libraries were
prepared and captured using a custom Nimblegen capture pool. Capture
baits were designed based on the coding regions of the \emph{R. muscosa}
transcriptome (GenBank accession GKCT00000000). Captured libraries were
then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the
Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.

\hypertarget{data-pre-processing-and-cleaning}{%
\subsection*{Data Pre-processing and
Cleaning}\label{data-pre-processing-and-cleaning}}

Raw reads were filtered for adapters and contaminants using fastp
\citep{chen2018} and aligned to the \emph{Rana muscosa} genome
\citep[NCBI SRA: SRS6533475, reference =][]{hon2020} with repetitive
elements masked using bwa (``mem'' mode) \citep{li2013}. Exact PCR
Duplicates were marked using Picard. Variants were then called following
GATK best practices (v.4.2.0.0, \citep{van2020}). Briefly, raw variants
were called for each sample using HaplotypeCaller and combined using
CombineGVCFs. Next, genotypes were jointly called using GenotypeGVCFs.
Variants were then hard filtered using gatk VariantFiltration using the
following parameters to remove low-quality sites: QD \textless{} 2.0, FS
\textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
MQRankSum \textgreater{} 3.0, MQRankSum \textless{} -3.0, ReadPosRankSum
\textgreater{} 3.0, ReadPosRankSum \textless{} -3.0. This initial filter
resulted in 1,595,206 variant sites across 53 individuals.

We then further filtered our dataset at the individual and variant
level. First, we trimmed our variants to only include those with minor
allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
depth of 5, a minimum genotype quality of 20, and a maximum missing
proportion of 0.5. This filter resulted in 427,038 sites, of which
353,172 were single nucleotide polymorphisms (SNPs) and 73,866 were
insertions or deletions (INDELS). Finally, we trimmed samples with an
average depth across filtered sites \textless{} 7x. Our final dataset
included 50 samples (23 from naive and 27 from recovering populations)
with an average depth of 16.7x (range = 7.4x -- 26.1x).

\hypertarget{data-analysis}{%
\subsection*{Data Analysis}\label{data-analysis}}

To visualize the genomic relationship of our populations we created a
PCA using the glPCA function in the adegenet R package
\citep{jombart2008}. Next, to detect regions of the genome under
selection between naive and recovering populations, we used a
multivariate linear mixed model to evaluate individual SNP variants and
a splined window analysis to evaluate larger genomic regions. For the
SNP analysis, we first used a stringent data filter to include only SNPs
with \textless{} 5\% missing data (missing for no more than 2
individuals), and then calculated the likelihood ratio statistic for the
resulting set of 148,307 high quality SNPs across 127 contigs using
GEMMA. GEMMA calculates and incorporates a relatedness matrix for input
samples, allowing us to account for relatedness and population structure
when calculating likelihood ratio statistics. We identified outlier SNPs
using a Bonferroni- corrected significance level of 0.01. Outliers are
SNPs showing different allele frequencies between naive versus
recovering populations. We visualized the results using a Manhattan plot
and qqplot. We developed a more liberal set of outlier SNPs using a
Bonferroni- corrected significance level of 0.05 and used this set
solely for the GO analysis (see below).

For each outlier SNP, we determined whether the variant was synonymous
(protein sequence the same for each variant) or non-synonymous (protein
sequence differs between variants), and where in the gene it was
located. To do this, we first extracted the reference genome sequence
surrounding the variant using bedtools ``getfasta'' \citep{quinlan2010}.
Next, we re-annotated each sequence using blast to get the predicted
gene location based on the closest annotated reference
\citep{altschul1997}. We then translated each variant to amino acids and
aligned this translation to that of the gene annotation to ensure proper
frame of reference using Geneious \citep{kearse2012}. After ensuring
proper translation, we characterized variants as within or outside the
coding sequence of the gene and as either synonymous or non-synonymous.

In the splined window analysis, we identified outlier regions using
\emph{F\textsubscript{ST}} and differences in nucleotide diversity
(\(\pi_{diff}\)) between naive and recovering populations. First, we
calculated per-site \emph{F\textsubscript{ST}} between the naive and
recovering individuals for all bi-allelic SNPs in the 30 largest contigs
(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
calculated per-site \(\pi\) separately for individuals from the naive
and recovering populations using VCFtools, then calculated
\(\pi_{diff}\) for each population
(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
size-sorted chromosome number and adjusted the SNP position based on the
relative position in the genome. We then used the WINGEN R package
\citep{bishop2023} to conduct a splined discrete window analysis for
\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
where non-overlapping window boundaries should occur by identifying
inflection points in the fitted spline, therefore balancing false
positive and false negative results that occur using other window-based
methods \citep{beissinger2015} . This method also calculates a
W-statistic allowing for outlier identification. We identified outliers
as those with a W-statistic greater than 4 standard deviations above the
mean for F\textsubscript{ST} or above/below the mean for \(\pi_{diff}\).
Outlier regions were then identified as those that were outliers in both
analyses, meaning that they showed 1) high differentiation between naive
and recovering populations, and 2) differential patterns of nucleotide
diversity in the same region. Finally, we extracted gene transcripts
mapped within each region and retrieved annotation for that region using
blast.

Using the liberal set of variant outliers, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the Blast hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
29 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows.

\hypertarget{frog-population-recovery-1}{%
\section*{Frog population recovery}\label{frog-population-recovery-1}}

\hypertarget{field-methods}{%
\subsection*{Field methods}\label{field-methods}}

For the 24 translocations we conducted, we identified donor populations
from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
collected using several years of visual encounter surveys (VES) and skin
swab collections \citep{knapp2016}, as well as landscape genetic
information \citep{poorten2017}. The populations that we selected
contained several hundred \emph{R. sierrae} adults and hundreds or
thousands of tadpoles. These relatively high abundances were the result
of recent increases following previous Bd-caused declines
\citep{knapp2016}. As is typical for recovering MYL frog populations, Bd
prevalence in the donor populations was high (0.69--0.96) and Bd load
(median log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or
more orders of magnitude below the level at which frog mortality is
expected (log\textsubscript{10}(load) \(\approx\) 5.78
copies)\citep{vredenburg2010, joseph2018}. Recipient sites to which
frogs were translocated were chosen based on previous \emph{R. sierrae}
presence (determined from VES and/or museum records) or characteristics
that suggested high quality habitat for this species \citep{knapp2005}.

We conducted 1--4 translocations per site (Figure S3)
and each translocated cohort included 18 to 99 frogs (median = 30).
During each translocation, adult frogs were collected from the donor
population and measured, weighed, swabbed, and PIT tagged. Frogs were
transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately
once per month during the summer active season (median number of visits
per summer = 2; range = 1--10) and assessed using diurnal CMR surveys
and VES. CMR surveys allowed quantification of adult survival, adult
population size, and recruitment of new adults, and VES provided
estimates of tadpole and juvenile abundance. During 2006-2012, CMR
surveys were conducted on a single day (primary period) per site visit,
during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag
(or tagged if they were untagged), measured, weighed, swabbed, and
released at the capture location. During 2013-2022, we generally used a
robust design in which all habitats were searched during several
consecutive days (median number of secondary periods per primary period
= 3; range = 3--7), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was
zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were
captured during more than one secondary period were measured, weighed,
and swabbed during the first capture, and during subsequent captures
were only identified and released. Frog skin swabs were analyzed using
standard Bd DNA extraction and qPCR methods. For details of survey and
swab analysis methods, see \citep{joseph2018}.

During each site visit, we conducted VES either immediately before CMR
surveys or during the first day of CMR surveys. VES was conducted by
walking the entire water body perimeter, first 100 m of each inlet and
outlet stream, and any fringing ponds and wetlands, and counting all
\emph{R. sierrae} tadpoles and juveniles. These \emph{R. sierrae} life
stages have high detectablility, and counts are highly repeatable and
provide estimates of relative abundance \citep{knapp2000}.

\hypertarget{estimation-of-frog-survival-and-abundance}{%
\subsection*{Estimation of frog survival and
abundance}\label{estimation-of-frog-survival-and-abundance}}

We estimated survival of translocated frogs, recruitment of new frogs
into the adult population, and adult population size using a Bayesian
open-population Jolly-Seber mark-recapture model with known additions to
the population (i.e., translocated cohorts), as implemented by the mrmr
package \citep{joseph2019} and using R Statistical Software
\citep[v4.4.4,][]{rsoftware2022}. The model tracks the states of
\emph{M} individuals that comprise a superpopulation made up of real and
pseudo-individuals \citep[see][ for details]{joseph2018}. The possible
states of individuals include ``not recruited'', ``alive'', and
``dead''. The possible observations of individuals include ``detected''
and ``not detected''. We assume that individuals that are in the ``not
recruited'' or ``dead'' states are never detected (i.e., there are no
mistakes in the individual PIT tag records). For all models, we used
mrmr defaults for priors, number of chains (4), and warmup and
post-warmup iterations (2000 for each). We evaluated convergence of the
Markov chain Monte Carlo (MCMC) algorithm using trace plots and
Gelman-Rubin statistics (Rhat). Code used to create CMR datasets and fit
models for each translocated population is available from the following
GitHub repository:
\href{https://github.com/SNARL1/cmr-analysis.}{https://github.com/SNARL1/cmr-analysis}.

\hypertarget{predictors-of-post-translocation-frog-survival}{%
\subsection*{Predictors of post-translocation frog
survival}\label{predictors-of-post-translocation-frog-survival}}

To identify important predictors of frog survival following
translocation, we developed linear and multilevel models in a Bayesian
framework. We performed all analyses with the rstanarm package
\citep{rstanarm2022} and R Statistical Software
\citep[v4.4.4,][]{rsoftware2022}, and used default, weakly informative
priors. Starting models included all relevant population-level effects
and, when data contained possible hierarchical structure, we included
group-level effects in the full model. Within each analysis, we
evaluated model fit using visualizations of leave-one-out (``LOO'')
probability integral transformations \citep{gelman2013, gabry2019}, and
compared fits of models with and without group-level effects, and with
differing group-level effects using LOO cross-validation
\citep{vehtari2016}. Both approaches utilized the loo package
\citep{vehtari2022}.

We conducted two analyses. In the first, we addressed the more general
question of whether survival of frogs in the first translocation to a
site predicted the survival of frogs in subsequent translocations.
Group-level effects included site\_id or translocation\_id. The second
analysis focused specifically on identifying predictors of
post-translocation frog survival using a wide range of variables
describing characteristics of frogs, sites, and translocations
(predictor variables: Bd load (bd\_load), winter severity in the year
following translocation (snow\_t1), site elevation (elevation), donor
population (donor\_72996), sex (sex\_male), frog size (length), winter
severity in the year of translocation (snow\_t), day of year on which a
translocation was conducted (day), and translocation order
(order\_first)). In both analyses we used 1-year survival as a single
metric of survival that was available for all translocated populations
regardless of how many years of post-translocation data were available
(range = 3-\/-16). Group-level effects included site\_id,
translocation\_id, or translocation\_id nested within site\_id. In both
analyses, the response variable, probability of 1-year frog survival,
was rounded to integer values to produce a binomial outcome, and all
models used a bernoulli family. For all models, we used default priors,
four chains, and 5000 iterations each for warmup and post-warmup. We
checked convergence of the MCMC algorithm using trace plots and Rhat.
Code to create the datasets and fit models is available from the
following GitHub repository:
\url{https://github.com/SNARL1/translocation}.

\hypertarget{changes-in-bd-load-following-translocation}{%
\subsection*{Changes in Bd load following
translocation}\label{changes-in-bd-load-following-translocation}}

To assess the magnitude of changes in Bd load on frogs following
translocation, we compared Bd loads measured before versus after
translocation. Before-translocation loads were quantified using skin
swabs collected from all to-be-translocated frogs at the donor site on
the day before or the day of the translocation. After-translocation Bd
loads were based on all swabs collected from translocated frogs at the
recipient site in the year of and the year following translocation.
Individual frogs and their associated Bd loads were included in the
dataset only if frogs were captured at the recipient site at least once
during the year 0-1 period. We used an analysis workflow similar to that
described above. Briefly, the response variable was before-translocation
load and the predictor variable was after-translocation load. We
compared models with and without group-level effects, and group-level
effects included pit\_tag\_id, pit\_tag\_id nested within site\_id, or
pit\_tag\_id nested within translocation\_id. All models used a negative
binomial family. Priors, chains, iterations, and checks were as
described above. Code to create the dataset and fit models is available
from the following GitHub repository:
\url{https://github.com/SNARL1/translocation}.

\hypertarget{population-viability-modeling}{%
\section*{Population viability
modeling}\label{population-viability-modeling}}

\hypertarget{model-description}{%
\subsection*{Model description}\label{model-description}}

To determine the implications of observed 1-year adult survival on the
long-term viability of populations established via translocation, we
developed a population model for MYL frogs. Our central question was:
How does the magnitude and variation in observed adult survival
probability across translocated populations affect the long-term
persistence probability of populations? We developed a model that
tracked seven state variables of a frog population: density of
translocated adults (\(A_T\)), density of adults naturally recruited
into the population (\(A_R\)), density of first-year tadpoles (\(L_1\)),
density of second-year tadpoles (\(L_2\)), density of third-year
tadpoles (\(L_3\)), density of first-year juveniles (\(J_1\)), and
density of second-year juveniles (\(J_2\)). We divided adults into two
classes \(A_T\) and \(A_R\) because there is evidence that the survival
probability of translocated adults and naturally recruited adults
differs (Figure S7).

We modeled the dynamics of these seven state variables using a
discrete-time, stage-structured model where a time step is one year. The
dynamics are given by

\begin{figure*}\[
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t)\numberthis \label{eqn:matrix} 
\]\end{figure*}

The parameters in this model are yearly survival probability
\(\sigma_{\cdot}\) (``\(\cdot\)'' indicates a particular state
variable), probability that a female frog reproduces in a given year
\(p_F\), number of eggs produced by a female frog in a year that
successfully hatch \(F\), probability of a first-year tadpole remaining
as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
as a tadpole \(p_{L2}\), and probability of a first-year juvenile
remaining as a juvenile \$p\_\{J1\}\$., and the probability of a
juvenile successfully recruiting to an adult \(\omega\).

In this model we ignore density-dependent recruitment because we were
interested in the growth of the population from an initial
reintroduction and whether this growth was sufficient to prevent
extinction over 50 years following the introduction. We also did not
directly consider the dynamics of Bd in this model. We made this
decision because (i) translocated populations are infected with Bd at
high prevalence \citep{joseph2018}, and (ii) host density does not seem
to play a significant role in multi-year Bd infection dynamics in this
system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
instead assuming all host vital rates are in the presence of high Bd
prevalence significantly simplifies the model without much loss of
realism.

\hypertarget{model-analysis}{%
\subsection*{Model analysis}\label{model-analysis}}

\subsubsection*{Incorporating yearly variability in vital rates}

We extracted yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the CMR model used to estimate population-level parameters of
translocated populations (see ``Estimation of adult survival and
population size'' above). Although we observed yearly variability in
adult survival within a population, the magnitude of this variability
was small compared to among population variability
(Figure~\ref{fig-translocation-survival}). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of successful recruitment
\(\omega\) (additional details provided in ``Estimating model
parameters'' below).

\subsubsection*{Estimating model parameters}

The baseline parameter values for the model and how they were estimated
are given in Table S2. Parameters
\(\sigma_{A_T}\) and \(\sigma_{A_R}\) were extracted directly from our
CMR model described previously in this study. For populations where we
had a sufficient number of PIT-tagged, naturally-recruited adults, we
observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably
different, with \(\sigma_{A_R} > \sigma_{A_T}\)
(Figure S7). For populations lacking
sufficient numbers of naturally-recruited adults, we were unable to
directly estimate \(\sigma_{A_R}\), and instead set
\(\sigma_{A_R} = \sigma_{A_T}\).

\subsubsection*{Model analysis and simulation}

We performed four analyses on our model. First, we considered a
deterministic version of our model with no yearly heterogeneity in
probability of successful recruitment \(\omega\), and calculated the
predicted long-run growth rate \(\lambda\) of a population for different
values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
\(\omega = 0.3\) and calculated the predicted growth rate of our 12
populations.

Second, we performed a local elasticity analysis on \(\lambda\) with
respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
determine how small changes in these parameters could influence the
long-run deterministic growth rate of populations.

Third, we defined a version of the model with demographic and
environmental stochasticity, where environmental stochasticity was
represented by among-year variability in \(\omega\). We used this model
to simulate a one-time introduction of 40 translocated adult frogs. We
ran this simulation 1000 times for each population and computed the
probability of a population becoming extinct after 50 years given the
observed parameter values and environmental stochasticity in \(\omega\).
We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
and drew values of \(\omega\) each year from a beta distribution with a
dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
\(\phi = 2\) the beta distribution is uniform between 0 and 1).

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
Approximate Bayesian Computing and rejection sampling
\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
1000 trajectories) that minimized the sum of squared errors between the
observed and predicted data. The \(\omega\) values associated with these
``best'' trajectories represented an approximate posterior distribution
\citep{beaumont2010}. Using these best fit trajectories, we assessed
whether our model could qualitatively describe the patterns of recovery
in the observed data for population 70550.}

\showmatmethods{} % Display the Materials and Methods section

\acknow{We thank the following for important contributions to this study: E.
Hegeman and A. Lindauer--project and data management, field and
laboratory work; A. Barbella and K. Rose--laboratory work; numerous
summer technicians--field work; staff at Sequoia-Kings Canyon and
Yosemite National Parks, Inyo and Sierra National Forests, California
Department of Fish and Wildlife, U.S. Fish and Wildlife Service,
University of California-Santa Barbara Institutional Animal Care and Use
Committee, and Sierra Nevada Aquatic Research Laboratory--research
permits, logistical support, and/or field assistance. This project was
supported by grants from the National Park Service (to R.A.K.), Yosemite
Conservancy (to R.A.K.); and National Science Foundation Grants
EF-0723563 (to C. Briggs), DEB-1557190 (to C. Briggs), and DBI-2120084
(to C. Richards-Zawacki).}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{translocation}

\clearpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/allele_maps.png}

}

\caption{\label{fig-allelefrequencies}For each of the naive and
recovering MYL frog populations, pie charts showing allele frequencies
for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B)
TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3. Charts are superimposed
on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon,
and Sequoia National Parks (from north to south) shown in dark gray, and
the range boundary for MYL frogs in this portion of the Sierra Nevada
shown in light gray. The inset map locates the study area in
California.}

\end{figure*}

\newpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/splinewindow_manhattan.png}

}

\caption{\label{fig-spline-manhattan}Results from the splined window
analysis showing outlier windows for \(\pi_{diff}\) (A) and
\emph{F\textsubscript{ST}} (B). Outlier regions for either
\(\pi_{diff}\) or \emph{F\textsubscript{ST}} are shown in blue and
outlier regions for both \(\pi_{diff}\) and \emph{F\textsubscript{ST}}
are shown in red. (C) Magnified Contig19 from (A) showing two adjacent
outlier windows for \(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier
SNP (indicated with a dashed vertical blue line). (D) Magnified Contig8
from from (B) showing the \emph{F\textsubscript{ST}} outlier window that
includes the outlier SNPs TCF19 and VARS. This region of the genome
contains 8 annotated genes known to occur in the extended MHC Class I
and II regions.}

\end{figure*}

\newpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/translocation_survival_bysiteid.png}

}

\caption{\label{fig-translocation-survival}Median 1-year survival for
each cohort of translocated frogs at the 12 recipient sites. Sites are
arranged along the x-axis using the average of the median 1-year
survival per translocation at each site. Dot colors indicate the donor
population from which frogs in each translocated cohort were collected.
Error bars show the 95\% uncertainty intervals. When multiple
translocations were conducted to a site, points and error bars are
slightly offset to avoid overlap.}

\end{figure*}

\newpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/cond_effects_plot.png}

}

\caption{\label{fig-cond-effects}Results showing conditional effects of
the important predictors of 1-year frog survival. (A) winter severity in
the year following translocation, (B) elevation of recipient site, (C)
donor population, and (D) sex. In (A) and (B), blue lines are medians
and gray ribbons are 95\% uncertainty intervals. In (C) and (D), box
plots show medians, first and third quartiles, largest and smallest
values within 1.5x interquartile range, and values outside the 1.5x
interquartile range.}

\end{figure*}

\newpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_manuscript1.jpg}

}

\caption{\label{fig-viability}A population viability analysis of
translocated MYL frogs that quantifies the long-run growth rate and
extinction probability of frog populations: (A) Predicted long-run
growth rate \(\lambda\) for different values of yearly adult survival
probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
given the parameterized, deterministic model. Colored points show the
predicted \(\lambda\) values for the twelve translocated populations
when probability of recruitment into the adult class is \(\omega = 0.3\)
(indicated by the dashed line). The red line shows where
\(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
12 translocated populations given demographic stochasticity,
environmental variability in \(\omega\), and different mean values of
\(\omega\). The dashed line shows where \(\omega = 0.3\), for comparison
to (A). There are 6 lines at extinction probability = 1, 5 of which
(70414--70619) are hidden beneath the line for 70505.}

\end{figure*}

\newpage

\begin{figure*}

{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp1.jpg}

}

\caption{\label{fig-viability-supp}Sensitivity analysis and empirical
validation of the stage-structured MYL frog model. (A) Elasticity of
\(\lambda\) with changes in 3 parameters: \(\sigma_{A_R}\) (yearly
survival probability of naturally recruited adults), \(\omega\) (yearly
probability of successful recruitment from juveniles to adults), and
\emph{F (}number of eggs produced by a female frog in a year that
successfully hatch). Elasticity is calculated at the default parameter
values for each population and \(\omega\) = 0.3. (B) 100 simulated
trajectories (black lines) that most closely matched the observed
abundance trajectory of adult amphibians at site 70550 (light orange).}

\end{figure*}

\end{document}

### 2023-05-08 doc/manuscript/translocation_pnas_SI.tex

\documentclass[9pt,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnassupportinginfo}
\usepackage{longtable}

\title{Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
\author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
\correspondingauthor{Corresponding Author name. Roland A. Knapp \\ E-mail: roland.knapp@ucsb.edu}

\begin{document}

\maketitle

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
\SItext
\hypertarget{frog-evolution-in-response-to-bd-2}{%
\section{Frog evolution in response to
Bd}\label{frog-evolution-in-response-to-bd-2}}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

We compared frog genomes sampled in naive versus recovering populations.
Bd-exposure histories of these populations are based on a decade or more
of data from visual encounter surveys and Bd surveillance using skin
swabbing \citep[e.g.,][]{knapp2016}. Comparing populations with
different infection histories allowed larger sample sizes and
replication across the landscape. The alternative approach of comparing
samples from the same populations before and after Bd exposure isn't
feasible in this system because Bd arrived in most MYL frog populations
decades ago and population persistence/recovery is rare and
unpredictable. As a result, samples from recovering populations
collected from before and after Bd exposure are not available and are
unlikely to be available in the future.

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

In the Results section of the main paper, we describe the stringent set
of outlier SNPS (identified using a Bonferroni-corrected p-value of
0.01). The liberal set, identified using a Bonferroni-corrected p-value
of 0.05, included 38 outliers (35 SNPs and 3 structural variants) from
30 distinct genes across 16 contigs.

Multiple genes in our stringent set of outliers have known associations
with amphibian Bd resistance/tolerance. Of particular interest is a
\textasciitilde940kb region on contig 8 that we identified as under
selection, and that contains 8 genes associated with either the MHC
Class I or Class III regions. The order of these genes in the MYL frog
genome is conserved from \emph{Xenopus} \citep{ohta2006} and is
consistent with MHC Class I/III linkage reported in the Komodo dragon
\citep{reed2021}. Numerous studies have linked MHC genes to frog
resistance/tolerance against Bd infection
\citep[e.g.,][]{savage2011, bataille2015}. In addition to the genes in
this region that we describe in the Results section of the main paper,
other genes also deserve mention. For example, we identified the MHC
Class III gene, heat shock protein 70 (HSP70), as occurring in this
region (Figure~\ref{fig-synteny-plot}). Similarly, \citep{byrne2020}
identified an HSP70-binding protein as a candidate gene under selection
in \emph{Atelopus varius} populations recovering after Bd-caused
declines. HSP70 was also highly expressed in Bd-infected versus
uninfected \emph{Xenopus tropicalis} \citep{rosenblum2009}. This MHC
gene cluster lies on the same chromosome as the outlier SNP in RIN3
(Fig. 2 C), a gene associated with
inflammation response that is highly expressed in amputated
\emph{Xenopus} tails \citep{fukazawa2009}.

Additionally, we identified a region on contig 19 that had the highest
\(\pi_{diff}\) of any outlier window, indicative of directional
selection. This region contained the gene, interferon-induced very large
GTPase 1-like (GVINP1). Interferons boost immune system processes and
previous studies of frogs have identified interferon-related genes as
associated with Bd resistance/tolerance \citep{byrne2020} or have shown
increased expression of this class of genes when frogs were exposed to
Bd in the lab \citep{rosenblum2009, ellison2014}. Variants in GVINP1
also explained approximately 20\% of the resistance to amoebic gill
disease in Atlantic salmon, a finding confirmed via expression analysis
\citep{robledo2018, robledo2020}. Furthermore, in \emph{Litoria
verreauxii alpina}, the related gene, IFN-induced very large GTPase,
showed lower expression in frogs from less susceptible populations
versus more susceptible populations \citep{grogan2018}.

Finally, we identified a region characterized by high
\emph{F\textsubscript{ST}} and low \(\pi_{diff}\) and that contained the
complement genes C6 and C7. This could indicate that balancing selection
is acting in this region of the genome to favor a diverse set of
alleles, as is known for C6 in humans \citep{soejima2005}.

\newpage

\hfill\break

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/pca_qq_manhattan.png}

}

\caption{\label{fig-selectionresults}Results from the exome capture
selection study. (A) PCA calculated from binary SNPs showing genomic
relationship of samples. Numeric labels and colors match those in
Fig. 1 A. (B) qqplot showing observed and
expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by
genomic position and contigs are sorted by size. Red line shows p-value
that identifies outliers. Outlier SNPs above this threshold are
highlighted and labeled. Bold labels indicate at least one
non-synonymous SNP is present in that gene.\\
}

\end{figure}\clearpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/synteny_figure.png}

}

\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
order in \emph{Xenopus tropicalis} and \emph{Rana mucosa} for the
outlier region containing MHC Class I Classical and MHC Class III gene
regions. The plot was created with SimpleSynteny \citep{veltri2016}
using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
accession GCA\_000004195.4) and \emph{Rana mucosa} Contig19. Asterices
indicate the location of SNP outliers in TCF19 and VARS1 genes. Gap
sizes for each contig representation are labeled.}

\end{figure}\clearpage

\newpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/map_translocation_points.png}

}

\caption{\label{fig-yosemap}Map showing the locations of translocated
and donor MYL frog populations in Yosemite National Park (park boundary
indicated by black polygon). Symbol shapes indicate the donor population
used for each translocation site. To obscure the exact locations of
populations, random noise was added to all point coordinates. Inset map
shows the location of Yosemite within California. In both maps,
elevation is indicated by the colored hillshade layer (dark green =
lowest elevation, white = highest elevation). }

\end{figure}\clearpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/mcmc_areas_m2b.png}

}

\caption{\label{fig-transsurvival-postdens}For the best model describing
the average 1-year survival of frogs in the first translocation to a
site as a predictor of survival of frogs in all subsequent
translocations to that site, estimated posterior density curves and
shaded 95\% uncertainty intervals for the intercept and single predictor
variable. In the Bayesian framework in which the model was developed, a
variable is considered an important predictor if the associated
uncertainty interval does not overlap zero (indicated by the dashed red
line).}

\end{figure}\clearpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/bdload_beforeafter.png}

}

\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
the 12 recipient sites, Bd loads for the period immediately prior to
translocation versus during the 1-year period after translocation. Box
plot details are as in Fig. 4.}

\end{figure}\clearpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/mcmc_areas_m1d.png}

}

\caption{\label{fig-survival-postdens}For the best model identifying
predictors of 1-year frog survival following translocation, estimated
posterior density curves and shaded 95\% uncertainty intervals for the
intercept and all predictor variables. In the Bayesian framework in
which the model was developed, variables are considered important
predictors if the associated uncertainty interval does not overlap zero
(indicated by the dashed red line).}

\end{figure}\clearpage

\begin{figure}

{\centering \includegraphics[width=0.8\textwidth]{figures/compare_surv_probs.jpg}

}

\caption{\label{fig-compare_surv_probs}Comparison of the average yearly
adult survival probabilities for adults translocated to each of three
sites versus adults that were naturally recruited at each site (as a
result of reproduction by translocated frogs). In contrast to
Fig. 3, these are not survival
probabilities from the first year following translocation, but instead
represent averaged survival probabilities across multiple years and
cohort. Points are median estimates and error bars give the 95\%
uncertainty intervals around the estimates, accounting for yearly
variation in survival probabilities. The large error bars for 70550
reflect the large within and among individual variability in yearly
survival probability at this site (as also reflected in one-year
survival probabilities in Fig. 3). All
estimates are derived from the mrmr package.}

\end{figure}\clearpage

\newpage

\hypertarget{tables}{%
\subsubsection{Tables}\label{tables}}

\hfill\break

\hypertarget{tbl-survival-earlylate}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0654}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2430}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2804}}@{}}
\caption{\label{tbl-survival-earlylate}Association between the
proportion of populations translocated early versus late in the study
period (\textless{} 2013 or \textgreater= 2013, respectively) and
probability of survival (\textless{} 0.5 or \textgreater= 0.5). Based on
the viability analysis, survival probabilities \textless{} 0.5 and
\textgreater= 0.5 produced 50-year extinction probabilities of 1 and
\textless{} 0.5, respectively.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Period
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Survival probability
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Number of translocations
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Total translocations
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Proportion of translocations
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Period
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Survival probability
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Number of translocations
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Total translocations
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Proportion of translocations
\end{minipage} \\
\midrule()
\endhead
early & \textless{} 0.5 & 4 & 5 & 0.800 \\
early & \textgreater= 0.5 & 1 & 5 & 0.200 \\
late & \textless{} 0.5 & 2 & 7 & 0.286 \\
late & \textgreater= 0.5 & 5 & 7 & 0.714 \\
\bottomrule()
\end{longtable}

\newpage

\hypertarget{tbl-param_values}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4545}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4121}}@{}}
\caption{\label{tbl-param_values}Description and values of parameters
used in the model. All survival probabilities are in the presence of the
fungal pathogen Bd.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endhead
\(\sigma_{L_1}\), Yearly survival probability of year 1 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_2}\), Yearly survival probability of year 2 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_3}\), Yearly survival probability of year 3 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles & 0.25
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{A_R}\), Yearly survival probability of naturally recruited
adults & Varies by population & Estimated from mark-recapture studies \\
\(\sigma_{A_T}\), Yearly survival probability of translocated adults &
Varies by population & Estimated from mark-recapture studies \\
\(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
1 & Estimated from field data, observations, natural history
knowledge \\
\(p_{L_2}\), Probability of a year 2 tadpoles remaining as a tadpoles &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_{J_1}\), Probability of a year 1 juvenile remaining as a juvenile &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_F\), Probability of a adult female reproducing in a year & 0.5 &
Could be as high at 1, based on field observations \\
\(F\), Number of surviving eggs produced by an adult female & 100 & From
observations of captive frogs \\
\(\omega\), Probability of juvenile successfully recruiting to an adult
& Varies yearly & Explored different values of this parameter \\
\bottomrule()
\end{longtable}

\hfill\break

%%% Add this line AFTER all your figures and tables
\FloatBarrier

\dataset{gemma\_outliers.csv}{Description needed}

\dataset{spline\_window\_shared\_outliers.csv}{Description needed}

\bibliography{translocation}

\end{document}

### 2023-05-08 doc/manuscript/translocation.qmd
quarto autoformat

--- 2023-05-08 doc/manuscript/translocation.qmd
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.\
-](figures/pca_qq_manhattan.png){#fig-selectionresults}
+Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:----------------------------|:-----------------|:-------------------------|
+|:-----------------------------|:---------------|:--------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-05-10 doc/manuscript/translocation.qmd
Incorporate coauthor comments into Results and Discussion

@@
 
-## Title: Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
-
-## Alternative title: Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
+## Title: Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
 
@@
 We conducted a principal component analysis (PCA) of the genomic data to describe the association between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
 Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
@@
 Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
+This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
 
@@
 
-In this section, we describe the outcome of *R.
-sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
-Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@seddon2014].
+To determine whether *R.
+sierrae* from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@seddon2014].
 The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
@@
 sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter SI, see SI text for additional details) [@joseph2018; @vredenburg2010].
 
@@
 We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
+
 Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
@@
 Results from our genomic analyses strongly suggest that natural selection for adaptive alleles and resulting increased resistance/tolerance is at least partially responsible for the reduced susceptibility of frogs in recovering populations.
-We identified substantial differences in allele frequencies and larger genomic regions between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
+We identified specific alleles and genomic regions showing signatures of selection between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
 Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
@@
 These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest indication of directional selection contains the gene, GVINP1.
+Similarly, the region with the strongest indication of directional selection (as measured by $pi_{diff}$) contains the gene GVINP1.
 Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014].
@@
 Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
-In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Retain or delete? -->.
+In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Reference complement genes. See Allie's comment in the SI for more on this.  -->.
 
@@
 
-The recent emergence of Bd worldwide has contributed to the decline of more than 500 amphibian species, nearly 100 of which now are extinct in the wild [@scheele2019].
+The recent emergence of Bd worldwide has contributed to the decline of hundreds of amphibian species, some of which are now extinct in the wild [@scheele2019].
 This unprecedented impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
@@
 The results of this study provide a hopeful starting point for that endeavor, and other future efforts worldwide.
+
+Evolutionary rescue may be increasingly critical for the survival of wildlife populations in a rapidly changing world.
+The possibility - and importance - of evolutionary rescue has been discussed for a wide variety of taxa [e.g., @mills2018; @oziolor2019], and in the specific context of disease-impacted systems [@searle2020].
+However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
+Nonetheless, we provide a compelling example from the wild of how in the face of a novel stressor evolutionary rescue can produce individuals that harbor adaptive alleles and that these individuals can be used to reestablish extirpated imperiled populations.
+We expect that such "assisted evolutionary rescue" will be an essential tool in wildlife conservation in an era of global change.
+<!--# Work on this last paragraph, especially the last few sentences. -->
 
@@
 
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI for details].
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. -->
+<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. --> <!--# Add  -->
 
@@
 
-\
-<!--# Bree comment: At some point it might be good to talk about what figures go in main text and which stay in supplement (or if we might make some multi-panel figures so that we can fit more in the main text).
-One thing I'm struck by in the figures is that each of the three main sections (genetics, translocations, PVA) are each so information rich but also somewhat disconnected.
-Not sure if we want to create some kind of synthetic or hybrid conceptual/data figure that pulls the main point of each analysis out and puts it in visual/conceptual relationship with the others.
-The results are so cool and we don't want readers to gloss over the details or have to work too hard to put it all back together again.
--->
-
 ![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
@@
 The dashed line shows where \$\\omega = 0.3\$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript1.jpg){#fig-viability} <!-- Consider a diverging color ramp for panel A, e.g., where lambda=1 is white, lambda>1 is blue, lambda<1 is red -->
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability}
 
@@
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
-(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp1.jpg){#fig-viability-supp}
+(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
 
@@
 
+### Frog population recovery
+
+#### Methods
+
+Swab extracts were analyzed singly instead of in triplicate [@kriger2006].
+For analysis of swabs collected during 2005--2014, we used standards developed from known concentrations of zoospores [@boyle2004], and after 2014, we used standards based on single ITS1 PCR amplicons [@longo2013].
+Based on paired comparisons between samples analyzed using both types of standards, Bd in the study area has an average of 60 ITS1 copies per zoospore.
+To express all qPCR results as the number of ITS1 copies, starting quantities obtained using the zoospore standard (measured as "zoospore equivalents") were multiplied by 60.
+In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
+
 \newpage
@@
 ![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
+Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
+Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale; see SI text for details).
 Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter}
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:-----------------------------|:---------------|:--------------------------|
+|:--------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-05-10 doc/manuscript/convert_qmd_to_pnas_latex.py
Update gitignore and pnas conversion file

@@ -137,7 +137,7 @@ pnas_ms = pnas_ms.replace("\\acknow{}", "\\acknow{" + acknowledge + "}")
 suppmatkey = "\\hypertarget{supporting-information}{%\n\\subsection{Supporting Information}\\label{supporting-information}}"
 fig_start = latex_manuscript.find(figkey)
 fig_end = latex_manuscript.find(suppmatkey)
-figures = latex_manuscript[fig_start:fig_end].split(figkey)[1].strip("\n").strip("\\newpage").strip("hfill\\break")
+figures = latex_manuscript[fig_start:fig_end].split(figkey)[1].strip("\n")#.strip("\\newpage")#.strip("hfill\\break")
 pnas_ms = pnas_ms.replace("% Figures\n", "\\clearpage\n" + figures)
 
 pnas_ms = pnas_ms.replace("\\begin{figure", "\\begin{figure*")
@@ -152,6 +152,7 @@ supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
 			"Figure~\\ref{fig-selectionresults} B, C: SI": "Figure S1B,C",
 			"Figure~\\ref{fig-synteny-plot} SI": "Figure S2",
 			"Figure~\\ref{fig-yosemap} SI": "Figure S3",
+			"Figure~\\ref{fig-yosemap}\nSI": "Figure S3",
 			"Figure~\\ref{fig-transsurvival-postdens} SI": "Figure S4",
 			"Figure~\\ref{fig-bdload-beforeafter} SI": "Figure S5",
 			"Figure~\\ref{fig-bdload-beforeafter}\n SI": "Figure S5",
@@ -231,6 +232,6 @@ with open("translocation_pnas_SI.tex", "w") as fout:
 	fout.writelines(pnas_supp)
 
 # Remove transcloation.tex
-subprocess.call(['rm', args[1]])
+# subprocess.call(['rm', args[1]])
 
 

### 2023-05-10 doc/manuscript/translocation.qmd
Reduce viability methods and address Max's comments

@@
 When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
-In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
-A sensitivity analysis of long-run growth rate $\lambda$ to yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$, recruitment probability $\omega$, and female fecundity *F* showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
-Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
 <!--# Consider condensing the above paragraph and providing details in SI.
-Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon." -->
+Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon."  Mark here:  Addressed both of these comments.
+-->
 
-We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
+Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
 Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
-When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
+When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$., and the probability of a juvenile successfully recruiting to an adult $\omega$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.
 
@@
 
-###### Incorporating yearly variability in vital rates
-
-We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among population variability ([@fig-translocation-survival]).
-Thus, we did not include yearly within-population variability in adult survival in this analysis.
-However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
-
-###### Estimating model parameters
-
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values : SI.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
-For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-###### Model analysis and simulation
-
-We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
-We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
-
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
-
-Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
-We used this model to simulate a one-time introduction of 40 translocated adult frogs.
-We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
-We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
-
-Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
-We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
-We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
-Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
+First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to growth or decline in the long-run.
+Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
+Third, we included demographic and environmental stochasticity into our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
 
@@
 
+### Population viability modeling
+
+#### Methods
+
+##### Incorporating yearly variability in vital rates
+
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among population variability ([@fig-translocation-survival]).
+Thus, we did not include yearly within-population variability in adult survival in this analysis.
+However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
+Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
+
+##### Estimating model parameters
+
+The baseline parameter values for the model and how they were estimated are given in @tbl-param_values : SI.
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
+For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+
+##### Model analysis and simulation
+
+We performed four analyses on our model.
+First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
+We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
+
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
+
+Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
+We used this model to simulate a one-time introduction of 40 translocated adult frogs.
+We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
+We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+
+Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
+We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
+We simulated our model for 16 years, repeating the simulation 50,000 times.
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
+Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
+The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
+Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+
 \newpage
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults}
 

### 2023-05-10 doc/manuscript/translocation_pnas.tex
Reduce viability methods and address Max's comments

@@
 
-\title{Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
+\title{Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
 
@@
 positive population growth. Third, the fitness benefits gained through
-selection for increased resistance are maintained beyond the natal
-environment and across generations, allowing for the establishment of
-reintroduced populations. And fourth, reintroduced populations show
-long-term growth and persistence. To date, few of these conditions are
-documented for any single species impacted by disease, and as a result
-the reintroduction of resistant hosts as a means to recover species of
-conservation concern remains largely an untested idea
+selection are maintained beyond the natal environment and across
+generations, allowing for the establishment of reintroduced populations.
+And fourth, reintroduced populations show long-term growth and
+persistence. To date, few of these conditions are documented, and
+reintroducing hosts from recovering populations is largely untested
 \citep{scheele2021, mendelson2019}.
 
-The mountain yellow-legged (MYL) frog, comprised of the sister species
+The mountain yellow-legged (MYL) frog, composed of the sister species
 \emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
@@
 Despite the catastrophic impact of Bd on MYL frogs in which most
-Bd-naive populations are driven to extirpation following Bd arrival
+Bd-naive populations are extirpated following Bd arrival
 \citep{vredenburg2010}, some populations have persisted after epizootics
@@
 associations between population type (i.e., naive versus recovering) and
-specific single nucleotide polymorphisms (SNPs). Second, we used a
-splined window analysis to identify larger genomic regions showing
-differences between population types in \emph{F\textsubscript{ST}} and
-nucleotide diversity (\(\pi_{diff}\) = \(\pi_{naive}\) -
-\(\pi_{recovering}\)).
+specific variants, including single nucleotide polymorphisms (SNPs) and
+insertions/deletions (INDELS). Second, we used a splined window analysis
+to identify larger genomic regions showing differences between
+population types in \emph{F\textsubscript{ST}} and nucleotide diversity
+(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).
 
@@
 populations (Figure~\ref{fig-allelefrequencies} C). This is suggestive
-of strong selection at this locus associated with exposure of frogs to
-Bd.
+of strong - and parallel - selection at this locus associated with
+exposure of frogs to Bd across multiple populations.
 
@@
 SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
-these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,
-CCHCR1, TCF19) and three occur in the extended MHC Class III region
-(HSP70, LSM2, VARS) \citep{ohta2006}. Therefore, this region under
-selection is part of the extended MHC Class I and III complex and shows
-synteny with other amphibian genomes.
+these genes occur in the extended major histocompatibility complex (MHC)
+Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
+extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
+Therefore, this region under selection is part of the extended MHC Class
+I and III complex and shows synteny with other amphibian genomes.
 
@@
 
-In this section, we describe the outcome of \emph{R. sierrae}
-reintroductions conducted in Yosemite National Park as part of a 15-year
-effort to reestablish populations of this endangered species
-(2006--2020). Each of the 24 reintroductions involved collection of
-adult frogs from 1 of 3 recovering, Bd-positive ``donor'' populations
-(Figure S3) and translocating them to 1 of 12 nearby
-recipient sites \citep{seddon2014}. The donor populations included 2 of
-the 5 recovering populations in the frog evolution study described above
+To determine whether \emph{R. sierrae} from recovering populations could
+be used to reestablish extirpated populations, we conducted 24
+reintroductions in Yosemite National Park (2006--2020). Each of the
+reintroductions involved collection of adult frogs from 1 of 3
+recovering, Bd-positive ``donor'' populations (Figure S3) and translocating them to 1 of 12 nearby recipient sites
+\citep{seddon2014}. The donor populations included 2 of the 5 recovering
+populations in the frog evolution study described above
 (Figure~\ref{fig-allelefrequencies} A: population 1 and 4). Following
@@
 with higher habitat quality for \emph{R. sierrae}. Finally, Bd loads
-changed little between immediately prior to translocation versus during
-the one year after translocation (Figure S5), and loads were nearly always well below the level indicative of
-severe chytridiomycosis and associated frog mortality (5.8 ITS copies on
-a log\textsubscript{10} scale) \citep{vredenburg2010, joseph2018}.
+were fairly consistent before versus after translocation and loads were
+nearly always well below the level indicative of severe chytridiomycosis
+and associated frog mortality (Figure S5,
+see SI text for additional details) \citep{joseph2018, vredenburg2010}.
 
@@
 the model using CMR data from translocated populations and known life
-history values in this system (Table S2). Given
-observed yearly adult survival probabilities (from analysis of CMR data)
-and a baseline yearly recruitment probability \(\omega\) of 0.3 (i.e.,
-the probability that, following survival, juveniles successfully recruit
-to the adult class), our stage-structured model predicted that six of
-twelve translocated populations should experience a long-run growth rate
-\(\lambda\) greater than 1 in the presence of Bd
+history values in this system (Table S2).
+
+Given observed yearly adult survival probabilities (from analysis of CMR
+data) and a baseline yearly recruitment probability \(\omega\) of 0.3
+(i.e., the probability that, following survival, juveniles successfully
+recruit to the adult class), our stage-structured model predicted that
+six of twelve translocated populations should experience a long-run
+growth rate \(\lambda\) greater than 1 in the presence of Bd
 (Figure~\ref{fig-viability} A; median predicted \(\lambda\) ranges from
@@
 conservative estimate), long-run growth rate of five of these six
-populations was still greater than 1 (Figure~\ref{fig-viability} A). In
-contrast, for the other six translocated populations, where yearly adult
-survival was substantially less than 0.5, our model predicted that it
-was unlikely that long-run growth rate in these populations was greater
-than 1, even when the probability of successful recruitment to the adult
-life stage was near 0.4 (Figure~\ref{fig-viability} A). A sensitivity
-analysis of long-run growth rate \(\lambda\) to yearly adult survival
-probability of naturally recruited individuals \(\sigma_{A_R}\),
-recruitment probability \(\omega\), and female fecundity \emph{F} showed
-that \(\lambda\) was most sensitive to small changes in \(\sigma_{A_R}\)
-(Figure~\ref{fig-viability-supp} A). Note that we assumed
-\(\sigma_{A_R}\) was equivalent to the survival probability of
-translocated adults \(\sigma_{A_T}\), except for three populations where
-we had \textgreater{} 5 years of CMR data and recovering populations
-that together allowed us to estimate \(\sigma_{A_R}\) and
-\(\sigma_{A_T}\) separately (Figure S7).
-
-We then incorporated demographic stochasticity (i.e., probabilistic
-births and deaths of individuals in a population) and environmental
-variability in the probability of juvenile recruitment to the adult
-class \(\omega\) (the transition that we expect to be the most subject
-to environmental stochasticity in the presence of Bd) into the model
-described above and used it to predict long-term population viability.
-Our model predicted that, following a single introduction of 40 adult
-individuals into a population and using the baseline value for
-\(\omega\) of 0.3, the six populations with the highest adult survival
-probabilities had 50-year extinction probabilities of 0 to 0.32;
+populations was still greater than 1 (Figure~\ref{fig-viability} A).
+
+Even when incorporating demographic and environmental stochasticity in
+the probability of juvenile recruitment to the adult class \(\omega\)
+(the transition that we expect to be the most subject to environmental
+stochasticity in the presence of Bd), populations with high adult
+survival are likely to persist over a 50 year time horizon. Our model
+predicted that, following a single introduction of 40 adult individuals
+into a population and using the baseline value for \(\omega\) of 0.3,
+the six populations with the highest adult survival probabilities had
+50-year extinction probabilities of 0 to 0.32;
 Figure~\ref{fig-viability} B). This indicates strong potential for
@@
 \citep{daszak2000}, and strategies to recover affected species are
-lacking. Recovery in this context is likely to be particularly
-challenging because typically recovery will need to occur in the
-presence of ongoing disease. One potential strategy, and the one we
-employ in the current study, utilizes populations of resistant/tolerant
-individuals as a source of animals with which to reestablish extirpated
-populations in the presence of disease. Although previously proposed
-\citep{mendelson2019}, this strategy has rarely been successfully
-implemented (see \citep{joseph2018} for a small-scale example). In
-particular, although numerous studies have documented signatures of
-selection in hosts following disease-induced declines
+lacking. Here, we used populations of resistant/tolerant individuals to
+reestablish extirpated populations, despite ongoing disease. Although
+previously proposed \citep{mendelson2019}, this strategy has rarely been
+successfully implemented (see \citep{joseph2018} for a small-scale
+example). In particular, although numerous studies have documented
+signatures of selection in hosts following disease-induced declines
 \citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
@@
 susceptibility of frogs in recovering populations. We identified
-substantial differences in allele frequencies and larger genomic regions
+specific alleles and genomic regions showing signatures of selection
 between adjacent Bd-naive and recovering MYL frog populations,
@@
 immunological mechanisms known to contribute to disease resistance,
-including in amphibians \citep{zamudio2020}. For example, the major
-histocompatibility complex (MHC) plays an important role in immunity. In
-our study, we identified a region that shows evidence of selection in
-recovering populations and contains eight genes associated with either
-the MHC Class I or Class III regions. These results corroborate numerous
-previous studies linking MHC genes to amphibian resistance/tolerance
-against Bd \citep[e.g.,][]{savage2011, bataille2015}. Similarly, the
-region with the strongest indication of directional selection contains
-the gene, GVINP1. Several previous studies have found this gene to be
-differentially expressed during Bd infection
+including in amphibians \citep{zamudio2020}. For example, the MHC plays
+an important role in immunity. In our study, we identified a region that
+shows evidence of selection in recovering populations and contains eight
+genes associated with either the MHC Class I or Class III regions. These
+results corroborate numerous previous studies linking MHC genes to
+amphibian resistance/tolerance against Bd
+\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
+the strongest indication of directional selection (as measured by
+\(pi_{diff}\)) contains the gene GVINP1. Several previous studies have
+found this gene to be differentially expressed during Bd infection
 \citep[e.g.,][]{grogan2018, ellison2014}. This gene was also strongly
@@
 area. In the SI, we provide additional details on these and other genes
-identified in our study as under selection . Collectively, these results
-suggest that the documented natural recovery of MYL frogs following
-Bd-caused declines \citep{knapp2016} is consistent with evolutionary
-rescue.
+identified in our study as under selection .
 
@@
 The recent emergence of Bd worldwide has contributed to the decline of
-more than 500 amphibian species, nearly 100 of which now are extinct in
-the wild \citep{scheele2019}. This unprecedented impact on global
-amphibian biodiversity is compounded by the lack of any effective and
-broadly applicable strategies to reverse these impacts
+hundreds of amphibian species, some of which are now extinct in the wild
+\citep{scheele2019}. This unprecedented impact on global amphibian
+biodiversity is compounded by the lack of any effective and broadly
+applicable strategies to reverse these impacts
 \citep{garner2016, knapp2022}. Importantly, in addition to the natural
@@
 point for that endeavor, and other future efforts worldwide.
+
+Evolutionary rescue may be increasingly critical for the survival of
+wildlife populations in a rapidly changing world. The possibility - and
+importance - of evolutionary rescue has been discussed for a wide
+variety of taxa \citep[e.g.,][]{mills2018, oziolor2019}, and in the
+specific context of disease-impacted systems \citep{searle2020}.
+However, obtaining evidence in wild animal populations for all the
+causal links in this process (i.e., selection for adaptive alleles
+\(\rightarrow\) increase in host fitness in the presence of the pathogen
+\(\rightarrow\) positive population growth and recovery) is difficult
+and perhaps impossible. Nonetheless, we provide a compelling example
+from the wild of how in the face of a novel stressor evolutionary rescue
+can produce individuals that harbor adaptive alleles and that these
+individuals can be used to reestablish extirpated imperiled populations.
+We expect that such ``assisted evolutionary rescue'' will be an
+essential tool in wildlife conservation in an era of global change.
 
@@
 (98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
-calculated per-site \(\pi\) separately for individuals from the naive
-and recovering populations using VCFtools, then calculated
-\(\pi_{diff}\) for each population
+calculated per-site nucleotide diversity \(\pi\) separately for
+individuals from the naive and recovering populations using VCFtools,
+then calculated \(\pi_{diff}\) for each population
 (\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
@@
 
-To assess the magnitude of changes in Bd load on frogs following
-translocation, we compared Bd loads measured before versus after
-translocation. Before-translocation loads were quantified using skin
-swabs collected from all to-be-translocated frogs at the donor site on
-the day before or the day of the translocation. After-translocation Bd
-loads were based on all swabs collected from translocated frogs at the
-recipient site in the year of and the year following translocation.
-Individual frogs and their associated Bd loads were included in the
-dataset only if frogs were captured at the recipient site at least once
-during the year 0-1 period. We used an analysis workflow similar to that
-described above. Briefly, the response variable was before-translocation
-load and the predictor variable was after-translocation load. We
-compared models with and without group-level effects, and group-level
-effects included pit\_tag\_id, pit\_tag\_id nested within site\_id, or
-pit\_tag\_id nested within translocation\_id. All models used a negative
-binomial family. Priors, chains, iterations, and checks were as
-described above. Code to create the dataset and fit models is available
-from the following GitHub repository:
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
+\citep[see SI for details]{boyle2004}. To assess the magnitude of
+changes in Bd load on frogs following translocation, we compared Bd
+loads measured before versus after translocation. Before-translocation
+loads were quantified using skin swabs collected from all
+to-be-translocated frogs at the donor site on the day before or the day
+of the translocation. After-translocation Bd loads were based on all
+swabs collected from translocated frogs at the recipient site in the
+year of and the year following translocation. Individual frogs and their
+associated Bd loads were included in the dataset only if frogs were
+captured at the recipient site at least once during the year 0-1 period.
+We used an analysis workflow similar to that described above. Briefly,
+the response variable was before-translocation load and the predictor
+variable was after-translocation load. We compared models with and
+without group-level effects, and group-level effects included
+pit\_tag\_id, pit\_tag\_id nested within site\_id, or pit\_tag\_id
+nested within translocation\_id. All models used a negative binomial
+family. Priors, chains, iterations, and checks were as described above.
+Code to create the dataset and fit models is available from the
+following GitHub repository:
 \url{https://github.com/SNARL1/translocation}.
@@
 The parameters in this model are yearly survival probability
-\(\sigma_{\cdot}\) (``\(\cdot\)'' indicates a particular state
-variable), probability that a female frog reproduces in a given year
-\(p_F\), number of eggs produced by a female frog in a year that
+\(\sigma_{\cdot}\) (the subscript ``\(\cdot\)'' indicates a particular
+state variable), probability that a female frog reproduces in a given
+year \(p_F\), number of eggs produced by a female frog in a year that
 successfully hatch \(F\), probability of a first-year tadpole remaining
@@
 as a tadpole \(p_{L2}\), and probability of a first-year juvenile
-remaining as a juvenile \$p\_\{J1\}\$., and the probability of a
-juvenile successfully recruiting to an adult \(\omega\).
+remaining as a juvenile \(p_{J1}\), and the probability of a juvenile
+successfully recruiting to an adult \(\omega\).
 
@@
 
-\subsubsection*{Incorporating yearly variability in vital rates}
-
-We extracted yearly survival probabilities for translocated adults
-\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
-the CMR model used to estimate population-level parameters of
-translocated populations (see ``Estimation of adult survival and
-population size'' above). Although we observed yearly variability in
-adult survival within a population, the magnitude of this variability
-was small compared to among population variability
-(Figure~\ref{fig-translocation-survival}). Thus, we did not include
-yearly within-population variability in adult survival in this analysis.
-However, within a population there was substantial yearly variability in
-the successful recruitment of adults, greater than what we would expect
-from Poisson variability around a mean value. Therefore, we allowed for
-yearly variability in the probability of successful recruitment
-\(\omega\) (additional details provided in ``Estimating model
-parameters'' below).
-
-\subsubsection*{Estimating model parameters}
-
-The baseline parameter values for the model and how they were estimated
-are given in Table S2. Parameters
-\(\sigma_{A_T}\) and \(\sigma_{A_R}\) were extracted directly from our
-CMR model described previously in this study. For populations where we
-had a sufficient number of PIT-tagged, naturally-recruited adults, we
-observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably
-different, with \(\sigma_{A_R} > \sigma_{A_T}\)
-(Figure S7). For populations lacking
-sufficient numbers of naturally-recruited adults, we were unable to
-directly estimate \(\sigma_{A_R}\), and instead set
-\(\sigma_{A_R} = \sigma_{A_T}\).
-
-\subsubsection*{Model analysis and simulation}
-
-We performed four analyses on our model. First, we considered a
-deterministic version of our model with no yearly heterogeneity in
-probability of successful recruitment \(\omega\), and calculated the
-predicted long-run growth rate \(\lambda\) of a population for different
-values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
-\(\omega = 0.3\) and calculated the predicted growth rate of our 12
-populations.
-
-Second, we performed a local elasticity analysis on \(\lambda\) with
-respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
-determine how small changes in these parameters could influence the
-long-run deterministic growth rate of populations.
-
-Third, we defined a version of the model with demographic and
-environmental stochasticity, where environmental stochasticity was
-represented by among-year variability in \(\omega\). We used this model
-to simulate a one-time introduction of 40 translocated adult frogs. We
-ran this simulation 1000 times for each population and computed the
-probability of a population becoming extinct after 50 years given the
-observed parameter values and environmental stochasticity in \(\omega\).
-We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
-and drew values of \(\omega\) each year from a beta distribution with a
-dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
-\(\phi = 2\) the beta distribution is uniform between 0 and 1).
-
-Finally, we assessed whether our stochastic model could reproduce
-observed trajectories of population recovery. We focused on population
-70550 because this was our longest CMR time series for a translocated
-population and because this population shows evidence of substantial
-post-translocation increases in adult abundance associated with
-population establishment and recovery. We simulated our model for 16
-years, repeating the simulation 50,000 times. For each run and each
-year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
-equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
-Approximate Bayesian Computing and rejection sampling
-\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
-1000 trajectories) that minimized the sum of squared errors between the
-observed and predicted data. The \(\omega\) values associated with these
-``best'' trajectories represented an approximate posterior distribution
-\citep{beaumont2010}. Using these best fit trajectories, we assessed
-whether our model could qualitatively describe the patterns of recovery
-in the observed data for population 70550.}
+After parameterizing our model with CMR-estimated adult frog survival
+probabilities and other known vital rates (Table S2), we performed four analyses (described in detail in the SI).
+First, we compute the long-run growth rate \(\lambda\) for each of our
+12 translocated populations to determine if the populations were
+deterministically predicted to growth or decline in the long-run.
+Second, we compute the elasticity of \(\lambda\) to model parameters, to
+quantify how much changes in adult survival, for example, affected the
+long-run growth rate. Third, we included demographic and environmental
+stochasticity into our model and simulated the 50-year viability (i.e.,
+1 - extinction probability) of populations given an introduction of 40
+adult individuals into an unoccupied habitat. Finally, we fit our model
+to our longest translocation trajectory to confirm that our model could
+reasonably reproduce the observed recovery trajectories of MYL frogs
+following reintroductions.}
 
@@
 \clearpage
-
 \begin{figure*}
@@
 \caption{\label{fig-spline-manhattan}Results from the splined window
-analysis showing outlier windows for \(\pi_{diff}\) (A) and
-\emph{F\textsubscript{ST}} (B). Outlier regions for either
-\(\pi_{diff}\) or \emph{F\textsubscript{ST}} are shown in blue and
-outlier regions for both \(\pi_{diff}\) and \emph{F\textsubscript{ST}}
-are shown in red. (C) Magnified Contig19 from (A) showing two adjacent
-outlier windows for \(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier
-SNP (indicated with a dashed vertical blue line). (D) Magnified Contig8
-from from (B) showing the \emph{F\textsubscript{ST}} outlier window that
-includes the outlier SNPs TCF19 and VARS. This region of the genome
-contains 8 annotated genes known to occur in the extended MHC Class I
-and II regions.}
+analysis showing outlier windows for the difference in nucleotide
+diversity \(\pi_{diff}\) (A) and \emph{F\textsubscript{ST}} (B). Outlier
+regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
+shown in blue and outlier regions for both \(\pi_{diff}\) and
+\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
+(A) showing two adjacent outlier windows for \(\pi_{diff}\) 12.9Mb
+upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
+line). (D) Magnified Contig8 from from (B) showing the
+\emph{F\textsubscript{ST}} outlier window that includes the outlier SNPs
+TCF19 and VARS. This region of the genome contains 8 annotated genes
+known to occur in the extended MHC Class I and II regions.}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_manuscript1.jpg}
+{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
 
@@
 extinction probability of frog populations: (A) Predicted long-run
-growth rate \(\lambda\) for different values of yearly adult survival
-probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
-given the parameterized, deterministic model. Colored points show the
-predicted \(\lambda\) values for the twelve translocated populations
-when probability of recruitment into the adult class is \(\omega = 0.3\)
-(indicated by the dashed line). The red line shows where
-\(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
-12 translocated populations given demographic stochasticity,
-environmental variability in \(\omega\), and different mean values of
-\(\omega\). The dashed line shows where \(\omega = 0.3\), for comparison
-to (A). There are 6 lines at extinction probability = 1, 5 of which
-(70414--70619) are hidden beneath the line for 70505.}
+growth rate \$\textbackslash lambda\$ for different values of yearly
+adult survival probability \$\textbackslash sigma\_\{A\_R\}\$ and
+probability of recruitment \$\textbackslash omega\$, given the
+parameterized, deterministic model. Colored points show the predicted
+\$\textbackslash lambda\$ values for the twelve translocated populations
+when probability of recruitment into the adult class is
+\$\textbackslash omega = 0.3\$ (indicated by the dashed line). The red
+line shows where \$\textbackslash lambda = 1\$. (B) Predicted 50-year
+extinction probabilities of the 12 translocated populations given
+demographic stochasticity, environmental variability in
+\$\textbackslash omega\$, and different mean values of
+\$\textbackslash omega\$. The dashed line shows where
+\$\textbackslash omega = 0.3\$, for comparison to (A). There are 6 lines
+at extinction probability = 1, 5 of which (70414--70619) are hidden
+beneath the line for 70505.}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp1.jpg}
+{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp.jpg}
 
@@
 
+\newpage
 \end{document}

### 2023-05-10 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -165,6 +165,7 @@ supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
 			"Table~\\ref{tbl-survival-earlylate} SI": "Table S1",
 			"Table~\\ref{tbl-survival-earlylate}": "Table S1",
 			"Table~\\ref{tbl-param_values} : SI": "Table S2",
+			"Table~\\ref{tbl-param_values}\n: SI": "Table S2",
 			"Table~\\ref{tbl-param_values} SI": "Table S2"
 			}
 

### 2023-05-10 doc/manuscript/translocation.qmd
Edits to Discussion in response to coauthor comments

--- 2023-05-10 doc/manuscript/translocation.qmd
@@
 A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables) -->
+
+<!--# render to pdf using `quarto render doc/manuscript/translocation.qmd --to pdf` -->
 
@@
 
-Disease-induced population declines are impacting global biodiversity [@daszak2000], and strategies to recover affected species are lacking.
+Disease-induced population declines are decimating global biodiversity [@daszak2000], but strategies to recover affected species are generally lacking.
 Here, we used populations of resistant/tolerant individuals to reestablish extirpated populations, despite ongoing disease.
@@
 In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
-The results of our study provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
-Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014], obtaining strong evidence in wild animal populations for each of the associated causal links (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
-Nonetheless, in the following narrative, we use evolutionary rescue as an informative, albeit tentative framework.
-<!-- I am wondering whether it is worth digging a bit deeper into the "difficult and perhaps impossible" task alluded to above.
+The results of our study provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
+<!--# Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014]...
+--> <!-- I am wondering whether it is worth digging a bit deeper into the "difficult and perhaps impossible" task alluded to above.
 e.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data -->
@@
 These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest indication of directional selection (as measured by $pi_{diff}$) contains the gene GVINP1.
+Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the gene GVINP1.
 Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014].
-This gene was also strongly linked to disease in salmon, explaining a remarkable 20% of the resistance phenotype [@robledo2020; @robledo2018].
+This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [@robledo2020; @robledo2018].
 Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
-In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Reference complement genes. See Allie's comment in the SI for more on this.  -->.
+This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
+In the SI, we provide additional details on these and other genes identified in our study as under selection<!--# Reference complement genes. See Allie's comment in the SI for more on this.  -->.
 
@@
 Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
-In an earlier study of 2 of our 12 translocated populations, Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment [@joseph2018: sites referred to as "Alpine" and "Subalpine" are referenced as "70550" and "70505" in the current study].
+In an earlier study of 2 of our 12 translocated populations, Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in [@joseph2018] are identified as "70550" and "70505" in the current study).
 Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
@@
 This unprecedented impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
-Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of using animals from these recovering populations to reestablish extirpated populations.
+Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
 As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have acquired via evolution the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
-Despite the hopeful example of successful reestablishment of MYL frogs via evolutionary rescue and reintroductions, the challenge of recovering hundreds of Bd-impacted amphibians worldwide is a daunting prospect.
+Despite the hopeful example of successful reestablishment of MYL frogs via evolutionary rescue and reintroductions, the challenge of recovering hundreds of Bd-impacted amphibian species worldwide is a daunting prospect.
 Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
@@
 Evolutionary rescue may be increasingly critical for the survival of wildlife populations in a rapidly changing world.
-The possibility - and importance - of evolutionary rescue has been discussed for a wide variety of taxa [e.g., @mills2018; @oziolor2019], and in the specific context of disease-impacted systems [@searle2020].
-However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
-Nonetheless, we provide a compelling example from the wild of how in the face of a novel stressor evolutionary rescue can produce individuals that harbor adaptive alleles and that these individuals can be used to reestablish extirpated imperiled populations.
-We expect that such "assisted evolutionary rescue" will be an essential tool in wildlife conservation in an era of global change.
-<!--# Work on this last paragraph, especially the last few sentences. -->
+The importance of evolutionary rescue is known for a wide variety of taxa [e.g., @carlson2014; @mills2018; @oziolor2019], including in the specific context of disease-impacted systems [@searle2020].
+However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the stressor $\rightarrow$ positive population growth and recovery) is difficult in most systems.
+Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
+We expect that such "assisted evolutionary rescue" will be an essential tool in wildlife conservation in an era of accelerating global change.
 
@@
 
-To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
-Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual SNP variants and a splined window analysis to evaluate larger genomic regions.<!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.
-For example, for us genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection.
-The first approach A, did X, Y, and Z but had C limitations.
-The second approach B, did X, Y, and Z and D limitations."--> For the SNP analysis, we first used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic<!--# What is the null with this test (or what are the two models being compared with the LRT)?
-Probably obvious for genomics folks, but not totally clear to me.
-I am guessing we are testing whether the `population` coefficient (naive vs.
-persistent) is significant for describing variability in a SNP?
---> for the resulting set of 148,307 high quality SNPs across 127 contigs using GEMMA.
+To visualize the genomic relationship of our populations we conducted a principal components analysis (PCA) using the glPCA function in the adegenet R package [@jombart2008].
+To detect regions of the genome under selection between naive and recovering populations, we used two approches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA.
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified outlier SNPs using a Bonferroni- corrected significance level of 0.01.
-Outliers are SNPs showing different allele frequencies between naive versus recovering populations.
+We identified "outlier" variants using a Bonferroni- corrected significance level of 0.01.
+Outliers are variants showing different allele frequencies between naive versus recovering populations.
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier SNPs using a Bonferroni- corrected significance level of 0.05 and used this set solely for the GO analysis (see below).
+We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below).
 
-For each outlier SNP, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using bedtools "getfasta" [@quinlan2010].
+For each outlier variant, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
 Next, we re-annotated each sequence using blast to get the predicted gene location based on the closest annotated reference [@altschul1997].
@@
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+Code used for genomic analyses and to figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
+<!--# Make sure link is live before manuscript submission. -->
 
@@
 
-![For each of the naive and recovering MYL frog populations, pie charts showing allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+![Evidence for selection in recovering MYL frog populations at the landscape scale.
+For each of the naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
-The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies}
+The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
@@
 
-![Results from the splined window analysis showing outlier windows for the difference in nucleotide diversity $\pi_{diff}$ (A) and *F~ST~* (B).
+![Evidence for selection on genomic regions in recovering MYL frog populations.
+Results are from the splined window analysis and show outlier windows for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
 Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
@@
 
-![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
+![Evidence for frog survival at the recipient sites following translocation.
+Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
 Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
@@
 Error bars show the 95% uncertainty intervals.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
 
@@
 
-![Results showing conditional effects of the important predictors of 1-year frog survival.
+![Evidence for reduced impacts of Bd on MYL frogs as indicated by the lack of an effect of Bd load on survival.
+Results show conditional effects of the important predictors of 1-year frog survival.
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
 
@@
 
-![A population viability analysis of translocated MYL frogs that quantifies the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate \$\\lambda\$ for different values of yearly adult survival probability \$\\sigma\_{A_R}\$ and probability of recruitment \$\\omega\$, given the parameterized, deterministic model.
-Colored points show the predicted \$\\lambda\$ values for the twelve translocated populations when probability of recruitment into the adult class is \$\\omega = 0.3\$ (indicated by the dashed line).
-The red line shows where \$\\lambda = 1\$.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in \$\\omega\$, and different mean values of \$\\omega\$.
-The dashed line shows where \$\\omega = 0.3\$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability}
+![Predicted growth and persistence of translocated populations.
+Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability \$\sigma\_{A_R}\$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
+The red line shows where $\lambda = 1$.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.3$, for comparison to (A).
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}
 
@@
 
-![Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
+![<!--# Evidence sentence -->Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
 (A) Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
 Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
-(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}
+(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
 
@@
 HSP70 was also highly expressed in Bd-infected versus uninfected *Xenopus tropicalis* [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3 (@fig-spline-manhattan C), a gene associated with inflammation response that is highly expressed in amputated *Xenopus* tails [@fukazawa2009].
+This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3 (@fig-spline-manhattan C, @fig-allelefrequencies A).
 
@@
 In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap}
+<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="250"}
 
 ![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens}
+In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
 
@@
 ![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens}
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 

### 2023-05-10 doc/manuscript/translocation_pnas_SI.tex

@@
 
-\title{Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
+\title{Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
 \author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
@@
 
+\hypertarget{frog-population-recovery-2}{%
+\section{Frog population
+recovery}\label{frog-population-recovery-2}}
+
+\hypertarget{methods-1}{%
+\subsection{Methods}\label{methods-1}}
+
+Swab extracts were analyzed singly instead of in triplicate
+\citep{kriger2006}. For analysis of swabs collected during 2005--2014,
+we used standards developed from known concentrations of zoospores
+\citep{boyle2004}, and after 2014, we used standards based on single
+ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
+between samples analyzed using both types of standards, Bd in the study
+area has an average of 60 ITS1 copies per zoospore. To express all qPCR
+results as the number of ITS1 copies, starting quantities obtained using
+the zoospore standard (measured as ``zoospore equivalents'') were
+multiplied by 60. In addition, all qPCR quantities (regardless of
+standard) were multiplied by 80 to account for the fact that DNA
+extracts from swabs were diluted 80-fold during extraction and PCR
+\citep{vredenburg2010}.
+
+\hypertarget{population-viability-modeling-1}{%
+\section{Population viability
+modeling}\label{population-viability-modeling-1}}
+
+\hypertarget{methods-2}{%
+\subsection{Methods}\label{methods-2}}
+
+\hypertarget{incorporating-yearly-variability-in-vital-rates}{%
+\subparagraph{Incorporating yearly variability in vital
+rates}\label{incorporating-yearly-variability-in-vital-rates}}
+
+We extracted yearly survival probabilities for translocated adults
+\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
+the CMR model used to estimate population-level parameters of
+translocated populations (see ``Estimation of adult survival and
+population size'' above). Although we observed yearly variability in
+adult survival within a population, the magnitude of this variability
+was small compared to among population variability
+(Fig. 3). Thus, we did not include
+yearly within-population variability in adult survival in this analysis.
+However, within a population there was substantial yearly variability in
+the successful recruitment of adults, greater than what we would expect
+from Poisson variability around a mean value. Therefore, we allowed for
+yearly variability in the probability of successful recruitment
+\(\omega\) (additional details provided in ``Estimating model
+parameters'' below).
+
+\hypertarget{estimating-model-parameters}{%
+\subparagraph{Estimating model
+parameters}\label{estimating-model-parameters}}
+
+The baseline parameter values for the model and how they were estimated
+are given in Table~\ref{tbl-param_values} : SI. Parameters
+\(\sigma_{A_T}\) and \(\sigma_{A_R}\) were extracted directly from our
+CMR model described previously in this study. For populations where we
+had a sufficient number of PIT-tagged, naturally-recruited adults, we
+observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably
+different, with \(\sigma_{A_R} > \sigma_{A_T}\)
+(Figure~\ref{fig-compare_surv_probs} : SI). For populations lacking
+sufficient numbers of naturally-recruited adults, we were unable to
+directly estimate \(\sigma_{A_R}\), and instead set
+\(\sigma_{A_R} = \sigma_{A_T}\).
+
+\hypertarget{model-analysis-and-simulation}{%
+\subparagraph{Model analysis and
+simulation}\label{model-analysis-and-simulation}}
+
+We performed four analyses on our model. First, we considered a
+deterministic version of our model with no yearly heterogeneity in
+probability of successful recruitment \(\omega\), and calculated the
+predicted long-run growth rate \(\lambda\) of a population for different
+values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
+\(\omega = 0.3\) and calculated the predicted growth rate of our 12
+populations.
+
+Second, we performed a local elasticity analysis on \(\lambda\) with
+respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
+determine how small changes in these parameters could influence the
+long-run deterministic growth rate of populations.
+
+Third, we defined a version of the model with demographic and
+environmental stochasticity, where environmental stochasticity was
+represented by among-year variability in \(\omega\). We used this model
+to simulate a one-time introduction of 40 translocated adult frogs. We
+ran this simulation 1000 times for each population and computed the
+probability of a population becoming extinct after 50 years given the
+observed parameter values and environmental stochasticity in \(\omega\).
+We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
+and drew values of \(\omega\) each year from a beta distribution with a
+dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
+\(\phi = 2\) the beta distribution is uniform between 0 and 1).
+
+Finally, we assessed whether our stochastic model could reproduce
+observed trajectories of population recovery. We focused on population
+70550 because this was our longest CMR time series for a translocated
+population and because this population shows evidence of substantial
+post-translocation increases in adult abundance associated with
+population establishment and recovery. We simulated our model for 16
+years, repeating the simulation 50,000 times. For each run and each
+year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
+equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
+Approximate Bayesian Computing and rejection sampling
+\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
+1000 trajectories) that minimized the sum of squared errors between the
+observed and predicted data. The \(\omega\) values associated with these
+``best'' trajectories represented an approximate posterior distribution
+\citep{beaumont2010}. Using these best fit trajectories, we assessed
+whether our model could qualitatively describe the patterns of recovery
+in the observed data for population 70550.
+
 \newpage
@@
 highlighted and labeled. Bold labels indicate at least one
-non-synonymous SNP is present in that gene.\\
-}
+non-synonymous SNP is present in that gene.}
 
@@
 the 12 recipient sites, Bd loads for the period immediately prior to
-translocation versus during the 1-year period after translocation. Box
-plot details are as in Fig. 4.}
+translocation versus during the 1-year period after translocation. Bd
+loads are expressed as the number of ITS1 copies per skin swab, as
+estimated by qPCR of the Bd ITS1 region. Loads indicative of severe
+disease are \textgreater{} 5.8 ITS copies (on a log\textsubscript{10}
+scale; see SI text for details). Box plot details are as in
+Fig. 4.}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4545}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1333}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4121}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2639}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
 \caption{\label{tbl-param_values}Description and values of parameters

### 2023-05-12 doc/manuscript/translocation.qmd
Edits to Abstract and Intro in response to coauthor comments

@@
 
-Infectious diseases are emerging at an unprecedented rate, with devastating consequences for global biodiversity.
-The recovery of impacted taxa is hindered by the ongoing presence of disease, and few broadly applicable conservation strategies exist to overcome this obstacle.
-Rapid evolution in response to disease may produce resistant/tolerant individuals capable of persisting despite the presence of disease ("evolutionary rescue"), and although untested, such hosts could be used to reestablish extirpated populations.
-We evaluated this scenario using the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the emergence of a lethal fungal pathogen.
+Escalating anthropogenic change is devastating global biodiversity.
+To avoid extinction, affected populations will often need to adapt to the altered environment.
+This process of rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth is termed "evolutionary rescue".
+In addition to allowing recovery of extant populations, individuals from rescued populations could also be used to reestablish extirpated populations.
+We evaluated this scenario using the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
 Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
 We show that MYL frogs have undergone substantial evolutionary change following disease outbreaks, and that these changes are associated with increased resistance/tolerance to Bd infection in naturally recovering populations.
-Large-scale reintroduction of frogs from these populations resulted in the establishment of reproducing populations at most recipient sites despite ongoing disease, and results from viability modeling suggests that established populations have a low probability of extinction over 50 years.
-Collectively, these results provide one of the first examples of how evolution and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted taxa.
-This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
+Large-scale reintroduction of frogs from these populations resulted in the establishment of reproducing populations at most recipient sites despite ongoing disease.
+In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
+Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
+This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.
 
@@
 
-Diseases are becoming more common in wildlife and threaten many species with extinction.
-TO BE WRITTEN
+Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
+One possibility is evolutionary modification that restores positive growth to declining populations ("evolutionary rescue").
+Individuals from these rescued populations could also be used to reestablish exinct populations.
+We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse its impacts.
+We found that frogs have evolved increased resistance to infection, facilitating population recovery.
+In addition, introduction of frogs from these populations allowed reestablishment despite the ongoing presence of disease.
+Such reintroductions could play an important role in biodiversity conservation amidst unprecedented global change.
 
@@
 
-Emerging infectious diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
-Unfortunately, given the rarity of wildlife recoveries in the presence of disease, we know little about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
-At a time when novel infectious diseases are emerging at an unprecedented rate [@daszak2000; @fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.
+Human activities are intensifying threats to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
+To persist, affected taxa will often need to adapt to their altered environment or face decline and extinction.
+"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth [@carlson2014].
+In addition to preventing extinction of declining populations, in theory, this process can also result in the seeding of nearby empty habitats and allow population reestablishment and recovery.
+Therefore, understanding the mechanisms supporting population persistence, including evolutionary rescue, is critical for the conservation of the ever-growing number of threatened taxa.
 
-Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
-"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
-Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often result in increased host resistance and/or tolerance.
-However, linking these selective responses to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
-In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in recovering populations can, in theory, be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
+One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
+Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
+Amid this global loss of biodiverity, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
+Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see \[@medzhitov2012\] for definitions).
+However, linking these evolutionary responses to population recoveries - as necessary for evolutionary rescue - is difficult and few clear examples exist.
+In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence \[e.g., @mendelson2019; @brannelly2021\].
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
 
-<!--# Drop the following paragraph? -->
-
-There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
-First, populations that survive following disease emergence evolve increased resistance/tolerance.
-Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
-Third, the fitness benefits gained through selection are maintained beyond the natal environment and across generations, allowing for the establishment of reintroduced populations.
-And fourth, reintroduced populations show long-term growth and persistence.
-To date, few of these conditions are documented, and reintroducing hosts from recovering populations is largely untested [@scheele2021; @mendelson2019].
-
 The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
-Once the most common amphibian in the high elevation portion of the California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
+Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
 Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
@@
 
-Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
+Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
@@
 We then describe the reintroduction of these resistant/tolerant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
-These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the multi-year extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolution and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
-This example exhibits several of the key characteristics of evolutionary rescue, and has potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.
+These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
+Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
 

### 2023-05-12 doc/manuscript/translocation.qmd
Make three panel model figure and update main text

@@
 Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see \[@medzhitov2012\] for definitions).
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see [@medzhitov2012] for definitions).
 However, linking these evolutionary responses to population recoveries - as necessary for evolutionary rescue - is difficult and few clear examples exist.
-In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence \[e.g., @mendelson2019; @brannelly2021\].
+In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019; @brannelly2021].
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
@@
 In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
 
@@
 
-![Predicted growth and persistence of translocated populations.
-Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability \$\sigma\_{A_R}\$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
+![Predicted and observed growth and persistence of translocated populations.
+Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
@@
 The dashed line shows where $\omega = 0.3$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}
-
-\newpage
-
-![<!--# Evidence sentence -->Sensitivity analysis and empirical validation of the stage-structured MYL frog model.
-(A) Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.
-(B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
+(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}
 
@@
 
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
 
@@
 All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
+
+![Sensitivity analysis of the stage-structured MYL frog model.
+Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
 

### 2023-05-12 doc/manuscript/translocation_pnas.tex
Make three panel model figure and update main text

@@
 % Please add a significance statement to explain the relevance of your work
-\significancestatement{Diseases are becoming more common in wildlife and threaten many species
-with extinction. TO BE WRITTEN}
+\significancestatement{Understanding how species persist despite accelerating global change is
+critical for the conservation of biodiversity. One possibility is
+evolutionary modification that restores positive growth to declining
+populations (``evolutionary rescue''). Individuals from these rescued
+populations could also be used to reestablish exinct populations. We
+evaluated both possibilities in a once-common frog species driven to the
+brink of extinction by a disease that has devastated global amphibian
+biodiversity and for which few options exist to reverse its impacts. We
+found that frogs have evolved increased resistance to infection,
+facilitating population recovery. In addition, introduction of frogs
+from these populations allowed reestablishment despite the ongoing
+presence of disease. Such reintroductions could play an important role
+in biodiversity conservation amidst unprecedented global change.}
 
@@
 \begin{abstract}
-Infectious diseases are emerging at an unprecedented rate, with
-devastating consequences for global biodiversity. The recovery of
-impacted taxa is hindered by the ongoing presence of disease, and few
-broadly applicable conservation strategies exist to overcome this
-obstacle. Rapid evolution in response to disease may produce
-resistant/tolerant individuals capable of persisting despite the
-presence of disease (``evolutionary rescue''), and although untested,
-such hosts could be used to reestablish extirpated populations. We
-evaluated this scenario using the once-common mountain yellow-legged
-(MYL) frog, which is threatened with extinction by the emergence of a
-lethal fungal pathogen. Although most MYL frog populations are
-extirpated following disease outbreaks, some persist and eventually
-begin to recover. We show that MYL frogs have undergone substantial
-evolutionary change following disease outbreaks, and that these changes
-are associated with increased resistance/tolerance to Bd infection in
-naturally recovering populations. Large-scale reintroduction of frogs
-from these populations resulted in the establishment of reproducing
-populations at most recipient sites despite ongoing disease, and results
-from viability modeling suggests that established populations have a low
-probability of extinction over 50 years. Collectively, these results
-provide one of the first examples of how evolution and the
-reintroduction of resistant/tolerant individuals can allow the
-landscape-scale recovery of disease-impacted taxa. This example has
-potentially broad implications for the many taxa worldwide that are
-threatened with extinction by emerging diseases.
+Escalating anthropogenic change is devastating global biodiversity. To
+avoid extinction, affected populations will often need to adapt to the
+altered environment. This process of rapid evolutionary change that
+increases the frequency of adaptive alleles and thereby restores
+positive population growth is termed ``evolutionary rescue''. In
+addition to allowing recovery of extant populations, individuals from
+rescued populations could also be used to reestablish extirpated
+populations. We evaluated this scenario using the once-common mountain
+yellow-legged (MYL) frog, which is threatened with extinction by the
+human-facilitated emergence of a lethal fungal pathogen
+(\emph{Batrachochytrium dendrobatidis}; ``Bd''). Although most MYL frog
+populations are extirpated following disease outbreaks, some persist and
+eventually begin to recover. We show that MYL frogs have undergone
+substantial evolutionary change following disease outbreaks, and that
+these changes are associated with increased resistance/tolerance to Bd
+infection in naturally recovering populations. Large-scale
+reintroduction of frogs from these populations resulted in the
+establishment of reproducing populations at most recipient sites despite
+ongoing disease. In addition, results from viability modeling suggest
+that established populations have a low probability of extinction over
+50 years. Collectively, these results provide one of the first examples
+of how evolutionary rescue and the reintroduction of resistant/tolerant
+individuals can allow the landscape-scale recovery of disease-impacted
+species. This example has potentially broad implications for the many
+taxa worldwide that are threatened with extinction by anthropogenic
+stressors.
 \end{abstract}
@@
 
-\dropcap{E}merging infectious diseases pose a severe threat to wildlife
-populations \citep{daszak2000}, and have caused dramatic declines and
-even extinctions in mammals, birds, amphibians, echinoderms, and other
-taxa \citep{hewson2014, samuel2015, scheele2019, cunningham2021}.
-Importantly, some populations previously decimated by disease have
-subsequently recovered to varying degrees
-\citep{newell2013, voyles2018, knapp2016} and might provide critical
-insights for species conservation efforts. Unfortunately, given the
-rarity of wildlife recoveries in the presence of disease, we know little
-about the mechanisms driving population recovery
-\citep{brannelly2021, russell2020} and how these mechanisms can be
-leveraged to re-establish host populations across large spatial scales
-\citep{mendelson2019}. At a time when novel infectious diseases are
-emerging at an unprecedented rate \citep{daszak2000, fisher2012}, this
-knowledge gap has far-reaching consequences for global biodiversity.
-
-Host evolution following disease-induced declines is one mechanism by
-which hosts can recover \citep{carlson2014, searle2020}. ``Evolutionary
-rescue'' describes rapid evolutionary change that increases the
-frequency of adaptive alleles and thereby restores positive population
-growth and prevents extinction \citep{carlson2014}. Evolutionary changes
-following disease-induced declines are known from a diversity of
-wildlife populations
+\dropcap{H}uman activities are intensifying threats to global biodiversity
+\citep{ceballos2015}, with important implications for ecosystem
+resilience and human welfare \citep{naeem2009}. To persist, affected
+taxa will often need to adapt to their altered environment or face
+decline and extinction. ``Evolutionary rescue'' describes rapid
+evolutionary change that increases the frequency of adaptive alleles and
+thereby restores positive population growth \citep{carlson2014}. In
+addition to preventing extinction of declining populations, in theory,
+this process can also result in the seeding of nearby empty habitats and
+allow population reestablishment and recovery. Therefore, understanding
+the mechanisms supporting population persistence, including evolutionary
+rescue, is critical for the conservation of the ever-growing number of
+threatened taxa.
+
+One consequence of human alteration of the biosphere is an increase in
+emerging infectious diseases \citep{jones2008, fisher2012}. Such
+diseases pose a severe threat to wildlife populations
+\citep{daszak2000}, and have caused dramatic declines and extinctions in
+a wide range of taxa
+\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
+global loss of biodiverity, some populations previously decimated by
+disease have subsequently recovered to varying degrees
+\citep{newell2013, voyles2018, knapp2016}. Although we know relatively
+little about the underlying mechanisms
+\citep{brannelly2021, russell2020}, host evolution is one possibility
+\citep{carlson2014, searle2020}. This process is known from a diversity
+of wildlife populations
 \citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
-often result in increased host resistance and/or tolerance. However,
-linking these selective responses to population recoveries - as
-necessary for evolutionary rescue - is difficult and remains rare. In
+often results in increased host resistance and/or tolerance (see
+\citep{medzhitov2012} for definitions). However, linking these
+evolutionary responses to population recoveries - as necessary for
+evolutionary rescue - is difficult and few clear examples exist. In
 addition to promoting recovery of disease-impacted populations, the
-resistant/tolerant individuals in recovering populations can, in theory,
-be used to reestablish populations that were extirpated following
-disease emergence \citep[e.g.,][]{mendelson2019}. To date, the use of
-resistant/tolerant individuals for species recovery is rare \citep[but
-see][]{joseph2018}, and studies that evaluate its potential utility are
-critically needed.
-
-There are four conditions that will often determine the feasibility of
-using individuals from populations that are recovering following
-disease-induced declines to reestablish extirpated populations. First,
-populations that survive following disease emergence evolve increased
-resistance/tolerance. Second, increased resistance/tolerance contributes
-to higher individual fitness in the presence of disease and allows
-positive population growth. Third, the fitness benefits gained through
-selection are maintained beyond the natal environment and across
-generations, allowing for the establishment of reintroduced populations.
-And fourth, reintroduced populations show long-term growth and
-persistence. To date, few of these conditions are documented, and
-reintroducing hosts from recovering populations is largely untested
-\citep{scheele2021, mendelson2019}.
+resistant/tolerant individuals in such populations could potentially be
+used to reestablish populations that were extirpated following disease
+emergence \citep[e.g.,][]{mendelson2019, brannelly2021}. To date, the
+use of resistant/tolerant individuals for species recovery is rare
+\citep[but see][]{joseph2018}, and studies that evaluate its potential
+utility are critically needed.
 
@@
 \citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
-most common amphibian in the high elevation portion of the California's
+most common amphibian in the high elevation portion of California's
 Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
@@
 
-Despite the catastrophic impact of Bd on MYL frogs in which most
+Despite the catastrophic impact of Bd on MYL frogs, wherein most
 Bd-naive populations are extirpated following Bd arrival
@@
 to parameterize a population model that extends our inferences of
-population establishment well beyond the multi-year extent of the CMR
+population establishment well beyond the temporal extent of the CMR
 data. Collectively, these results provide one of the first examples of
-how evolution and the reintroduction of resistant/tolerant animals can
-allow the large-scale recovery of disease-impacted species. This example
-exhibits several of the key characteristics of evolutionary rescue, and
-has potentially broad implications for the many taxa worldwide that are
-threatened with extinction by emerging diseases.
+how evolutionary rescue and the reintroduction of resistant/tolerant
+animals can allow the large-scale recovery of disease-impacted species.
 
@@
 
-Even when incorporating demographic and environmental stochasticity in
-the probability of juvenile recruitment to the adult class \(\omega\)
-(the transition that we expect to be the most subject to environmental
-stochasticity in the presence of Bd), populations with high adult
-survival are likely to persist over a 50 year time horizon. Our model
-predicted that, following a single introduction of 40 adult individuals
-into a population and using the baseline value for \(\omega\) of 0.3,
-the six populations with the highest adult survival probabilities had
-50-year extinction probabilities of 0 to 0.32;
+Even when incorporating demographic stochasticity and environmental
+stochasticity in the probability of juvenile recruitment to the adult
+class \(\omega\) (the transition that we expect to be the most subject
+to environmental variability in the presence of Bd), populations with
+high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult
+individuals into a population and using the baseline value for
+\(\omega\) of 0.3, the six populations with the highest adult survival
+probabilities had 50-year extinction probabilities of 0 to 0.32;
 Figure~\ref{fig-viability} B). This indicates strong potential for
@@
 variability in recruitment (Figure~\ref{fig-viability} B). When we
-decreased \(\omega\) from 0.3 to a conservative estimate of 0.15, 50
-year extinction probability remained less than 0.5 for three populations
-and was 0.79 to 0.94 for three additional populations
+decreased \(\omega\) from 0.3 to a conservative estimate of 0.15,
+50-year extinction probability remained less than 0.5 for three
+populations and was 0.79 to 0.94 for three additional populations
 (Figure~\ref{fig-viability} B). In contrast, for the six populations
@@
 recovery trajectory of our translocated population with the longest
-survey history (Figure~\ref{fig-viability-supp} B; population 70550,
-surveyed for 16 years).
+survey history (Figure~\ref{fig-viability} C; population 70550, surveyed
+for 16 years).
 
@@
 
-Disease-induced population declines are impacting global biodiversity
-\citep{daszak2000}, and strategies to recover affected species are
-lacking. Here, we used populations of resistant/tolerant individuals to
-reestablish extirpated populations, despite ongoing disease. Although
-previously proposed \citep{mendelson2019}, this strategy has rarely been
-successfully implemented (see \citep{joseph2018} for a small-scale
-example). In particular, although numerous studies have documented
-signatures of selection in hosts following disease-induced declines
+Disease-induced population declines are decimating global biodiversity
+\citep{daszak2000}, but strategies to recover affected species are
+generally lacking. Here, we used populations of resistant/tolerant
+individuals to reestablish extirpated populations, despite ongoing
+disease. Although previously proposed \citep{mendelson2019}, this
+strategy has rarely been successfully implemented (see
+\citep{joseph2018} for a small-scale example). In particular, although
+numerous studies have documented signatures of selection in hosts
+following disease-induced declines
 \citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
@@
 evidence that (i) MYL frogs have experienced strong selection following
-exposure to a recently-emerged lethal fungal disease (Bd) and that some
-of the genomic regions under selection are associated with genes related
-to disease resistance/tolerance, and (ii) reintroduction of individuals
+exposure to a recently-emerged lethal fungal disease and that some of
+the genomic regions under selection are associated with genes related to
+disease resistance/tolerance, and (ii) reintroduction of individuals
 from these resistant/tolerant populations allows the successful
 reestablishment of viable frog populations despite ongoing Bd infection.
-Although these results are generally consistent with expectations under
-evolutionary rescue \citep{carlson2014}, obtaining strong evidence in
-wild animal populations for each of the associated causal links (i.e.,
-selection for adaptive alleles \(\rightarrow\) increase in host fitness
-in the presence of the pathogen \(\rightarrow\) positive population
-growth and recovery) is difficult and perhaps impossible. Nonetheless,
-in the following narrative, we use evolutionary rescue as an
-informative, albeit tentative framework.
 
@@
 the strongest indication of directional selection (as measured by
-\(pi_{diff}\)) contains the gene GVINP1. Several previous studies have
+\(\pi_{diff}\)) contains the gene GVINP1. Several previous studies have
 found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018, ellison2014}. This gene was also strongly
-linked to disease in salmon, explaining a remarkable 20\% of the
-resistance phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3
-gene showed a consistent pattern of allele frequency differences across
-all nine of the frog populations sampled in this study, indicating
+\citep[e.g.,][]{grogan2018, ellison2014}. This gene is also strongly
+linked to disease in salmon, explaining a notable 20\% of the resistance
+phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3 gene
+showed a consistent pattern of allele frequency differences across all
+nine of the frog populations sampled in this study, indicating
 consistent selection in populations distributed across a wide geographic
-area. In the SI, we provide additional details on these and other genes
-identified in our study as under selection .
+area. This gene is associated with inflammation response and in
+\emph{Xenopus} is highly expressed in wound stumps following amputation
+of tadpole tails \citep{fukazawa2009}. In the SI, we provide additional
+details on these and other genes identified in our study as under
+selection.
 
@@
 infection and load had detectable effects on the survival of adults and
-may have influenced population establishment \citep[sites referred to as
-``Alpine'' and ``Subalpine'' are referenced as ``70550'' and ``70505''
-in the current study]{joseph2018}. Applying similar analyses to all 12
-of our translocated populations would likely provide a broader
+may have influenced population establishment (sites referred to as
+``Alpine'' and ``Subalpine'' in \citep{joseph2018} are identified as
+``70550'' and ``70505'' in the current study). Applying similar analyses
+to all 12 of our translocated populations would likely provide a broader
 perspective of the ongoing effect of Bd. In addition to these important
@@
 also showing evidence of post-epizootic recovery in the presence of Bd
-\citep{scheele2017, voyles2018} and suggest the possibility of using
-animals from these recovering populations to reestablish extirpated
-populations. As with MYL frogs, the feasibility and long-term success of
-such efforts will depend on the availability of robust donor populations
-containing individuals that have acquired via evolution the adaptive
-alleles necessary to allow frog survival and population growth in the
-presence of Bd. Despite the hopeful example of successful
+\citep{scheele2017, voyles2018} and suggest the possibility of also
+using animals from these recovering populations to reestablish
+extirpated populations. As with MYL frogs, the feasibility and long-term
+success of such efforts will depend on the availability of robust donor
+populations containing individuals that have acquired via evolution the
+adaptive alleles necessary to allow frog survival and population growth
+in the presence of Bd. Despite the hopeful example of successful
 reestablishment of MYL frogs via evolutionary rescue and
 reintroductions, the challenge of recovering hundreds of Bd-impacted
-amphibians worldwide is a daunting prospect. Although we now have a
-proven strategy to reestablish extirpated MYL frog populations, recovery
-across their large historical range will require substantial resources
-over many decades. The results of this study provide a hopeful starting
-point for that endeavor, and other future efforts worldwide.
+amphibian species worldwide is a daunting prospect. Although we now have
+a proven strategy to reestablish extirpated MYL frog populations,
+recovery across their large historical range will require substantial
+resources over many decades. The results of this study provide a hopeful
+starting point for that endeavor, and other future efforts worldwide.
 
 Evolutionary rescue may be increasingly critical for the survival of
-wildlife populations in a rapidly changing world. The possibility - and
-importance - of evolutionary rescue has been discussed for a wide
-variety of taxa \citep[e.g.,][]{mills2018, oziolor2019}, and in the
+wildlife populations in a rapidly changing world. The importance of
+evolutionary rescue is known for a wide variety of taxa
+\citep[e.g.,][]{carlson2014, mills2018, oziolor2019}, including in the
 specific context of disease-impacted systems \citep{searle2020}.
@@
 causal links in this process (i.e., selection for adaptive alleles
-\(\rightarrow\) increase in host fitness in the presence of the pathogen
-\(\rightarrow\) positive population growth and recovery) is difficult
-and perhaps impossible. Nonetheless, we provide a compelling example
-from the wild of how in the face of a novel stressor evolutionary rescue
-can produce individuals that harbor adaptive alleles and that these
-individuals can be used to reestablish extirpated imperiled populations.
-We expect that such ``assisted evolutionary rescue'' will be an
-essential tool in wildlife conservation in an era of global change.
+\(\rightarrow\) increase in host fitness in the presence of the stressor
+\(\rightarrow\) positive population growth and recovery) is difficult in
+most systems. Nonetheless, we provide a compelling example from the wild
+not only that evolutionary rescue can produce individuals that harbor
+adaptive alleles and allow population recovery, but importantly that
+these individuals can be used in translocations to reestablish
+extirpated populations. We expect that such ``assisted evolutionary
+rescue'' will be an essential tool in wildlife conservation in an era of
+accelerating global change.
 
@@
 
-To visualize the genomic relationship of our populations we created a
-PCA using the glPCA function in the adegenet R package
-\citep{jombart2008}. Next, to detect regions of the genome under
-selection between naive and recovering populations, we used a
-multivariate linear mixed model to evaluate individual SNP variants and
-a splined window analysis to evaluate larger genomic regions. For the
-SNP analysis, we first used a stringent data filter to include only SNPs
-with \textless{} 5\% missing data (missing for no more than 2
-individuals), and then calculated the likelihood ratio statistic for the
-resulting set of 148,307 high quality SNPs across 127 contigs using
-GEMMA. GEMMA calculates and incorporates a relatedness matrix for input
-samples, allowing us to account for relatedness and population structure
-when calculating likelihood ratio statistics. We identified outlier SNPs
-using a Bonferroni- corrected significance level of 0.01. Outliers are
-SNPs showing different allele frequencies between naive versus
-recovering populations. We visualized the results using a Manhattan plot
-and qqplot. We developed a more liberal set of outlier SNPs using a
-Bonferroni- corrected significance level of 0.05 and used this set
-solely for the GO analysis (see below).
-
-For each outlier SNP, we determined whether the variant was synonymous
-(protein sequence the same for each variant) or non-synonymous (protein
-sequence differs between variants), and where in the gene it was
-located. To do this, we first extracted the reference genome sequence
-surrounding the variant using bedtools ``getfasta'' \citep{quinlan2010}.
-Next, we re-annotated each sequence using blast to get the predicted
-gene location based on the closest annotated reference
-\citep{altschul1997}. We then translated each variant to amino acids and
-aligned this translation to that of the gene annotation to ensure proper
-frame of reference using Geneious \citep{kearse2012}. After ensuring
-proper translation, we characterized variants as within or outside the
-coding sequence of the gene and as either synonymous or non-synonymous.
+To visualize the genomic relationship of our populations we conducted a
+principal components analysis (PCA) using the glPCA function in the
+adegenet R package \citep{jombart2008}. To detect regions of the genome
+under selection between naive and recovering populations, we used two
+approches: (1) a multivariate linear mixed model to evaluate individual
+variants (SNPs and INDELs), and (2) a splined window analysis to
+evaluate larger genomic regions. For the variant analysis, we first used
+a stringent data filter to include only variants with \textless{} 5\%
+missing data (missing for no more than 2 individuals), and then
+calculated the likelihood ratio statistic for the resulting set of
+148,307 high quality variants across 127 contigs using GEMMA. GEMMA
+calculates and incorporates a relatedness matrix for input samples,
+allowing us to account for relatedness and population structure when
+calculating likelihood ratio statistics. We identified ``outlier''
+variants using a Bonferroni- corrected significance level of 0.01.
+Outliers are variants showing different allele frequencies between naive
+versus recovering populations. We visualized the results using a
+Manhattan plot and qqplot. We developed a more liberal set of outlier
+variants using a Bonferroni- corrected significance level of 0.05 and
+used this set solely for the gene ontology (GO) analysis (see below).
+
+For each outlier variant, we determined whether the variant was
+synonymous (protein sequence the same for each variant) or
+non-synonymous (protein sequence differs between variants), and where in
+the gene it was located. To do this, we first extracted the reference
+genome sequence surrounding the variant using the bedtools ``getfasta''
+function \citep{quinlan2010}. Next, we re-annotated each sequence using
+blast to get the predicted gene location based on the closest annotated
+reference \citep{altschul1997}. We then translated each variant to amino
+acids and aligned this translation to that of the gene annotation to
+ensure proper frame of reference using Geneious \citep{kearse2012}.
+After ensuring proper translation, we characterized variants as within
+or outside the coding sequence of the gene and as either synonymous or
+non-synonymous.
 
@@
 located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
-splined windows.
+splined windows. Code used for genomic analyses and to figures is
+available from the following GitHub repository:
+\url{https://github.com/allie128/mylf-selection}.
 
@@
 
-\caption{\label{fig-allelefrequencies}For each of the naive and
-recovering MYL frog populations, pie charts showing allele frequencies
-for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B)
-TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3. Charts are superimposed
-on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon,
-and Sequoia National Parks (from north to south) shown in dark gray, and
-the range boundary for MYL frogs in this portion of the Sierra Nevada
-shown in light gray. The inset map locates the study area in
-California.}
+\caption{\label{fig-allelefrequencies}Evidence for selection in
+recovering MYL frog populations at the landscape scale. For each of the
+naive and recovering MYL frog populations, pie charts show allele
+frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
+LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with
+Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
+shown in dark gray, and the range boundary for MYL frogs in this portion
+of the Sierra Nevada shown in light gray. The inset map locates the
+study area in California.}
 
@@
 
-\caption{\label{fig-spline-manhattan}Results from the splined window
-analysis showing outlier windows for the difference in nucleotide
-diversity \(\pi_{diff}\) (A) and \emph{F\textsubscript{ST}} (B). Outlier
-regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
-shown in blue and outlier regions for both \(\pi_{diff}\) and
+\caption{\label{fig-spline-manhattan}Evidence for selection on genomic
+regions in recovering MYL frog populations. Results are from the splined
+window analysis and show outlier windows for the difference in (A)
+nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
+Outlier regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}}
+are shown in blue and outlier regions for both \(\pi_{diff}\) and
 \emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
@@
 
-\caption{\label{fig-translocation-survival}Median 1-year survival for
+\caption{\label{fig-translocation-survival}Evidence for frog survival at
+the recipient sites following translocation. Median 1-year survival for
 each cohort of translocated frogs at the 12 recipient sites. Sites are
@@
 
-\caption{\label{fig-cond-effects}Results showing conditional effects of
-the important predictors of 1-year frog survival. (A) winter severity in
-the year following translocation, (B) elevation of recipient site, (C)
-donor population, and (D) sex. In (A) and (B), blue lines are medians
-and gray ribbons are 95\% uncertainty intervals. In (C) and (D), box
-plots show medians, first and third quartiles, largest and smallest
-values within 1.5x interquartile range, and values outside the 1.5x
-interquartile range.}
+\caption{\label{fig-cond-effects}Evidence for reduced impacts of Bd on
+MYL frogs as indicated by the lack of an effect of Bd load on survival.
+Results show conditional effects of the important predictors of 1-year
+frog survival. (A) winter severity in the year following translocation,
+(B) elevation of recipient site, (C) donor population, and (D) sex. In
+(A) and (B), blue lines are medians and gray ribbons are 95\%
+uncertainty intervals. In (C) and (D), box plots show medians, first and
+third quartiles, largest and smallest values within 1.5x interquartile
+range, and values outside the 1.5x interquartile range.}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
 
@@
 
-\caption{\label{fig-viability}A population viability analysis of
-translocated MYL frogs that quantifies the long-run growth rate and
+\caption{\label{fig-viability}Predicted and observed growth and
+persistence of translocated populations. Population viability analysis
+of translocated MYL frogs quantified the long-run growth rate and
 extinction probability of frog populations: (A) Predicted long-run
-growth rate \$\textbackslash lambda\$ for different values of yearly
-adult survival probability \$\textbackslash sigma\_\{A\_R\}\$ and
-probability of recruitment \$\textbackslash omega\$, given the
-parameterized, deterministic model. Colored points show the predicted
-\$\textbackslash lambda\$ values for the twelve translocated populations
-when probability of recruitment into the adult class is
-\$\textbackslash omega = 0.3\$ (indicated by the dashed line). The red
-line shows where \$\textbackslash lambda = 1\$. (B) Predicted 50-year
-extinction probabilities of the 12 translocated populations given
-demographic stochasticity, environmental variability in
-\$\textbackslash omega\$, and different mean values of
-\$\textbackslash omega\$. The dashed line shows where
-\$\textbackslash omega = 0.3\$, for comparison to (A). There are 6 lines
-at extinction probability = 1, 5 of which (70414--70619) are hidden
-beneath the line for 70505.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp.jpg}
-
-}
-
-\caption{\label{fig-viability-supp}Sensitivity analysis and empirical
-validation of the stage-structured MYL frog model. (A) Elasticity of
-\(\lambda\) with changes in 3 parameters: \(\sigma_{A_R}\) (yearly
-survival probability of naturally recruited adults), \(\omega\) (yearly
-probability of successful recruitment from juveniles to adults), and
-\emph{F (}number of eggs produced by a female frog in a year that
-successfully hatch). Elasticity is calculated at the default parameter
-values for each population and \(\omega\) = 0.3. (B) 100 simulated
-trajectories (black lines) that most closely matched the observed
-abundance trajectory of adult amphibians at site 70550 (light orange).}
+growth rate \(\lambda\) for different values of yearly adult survival
+probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
+given the parameterized, deterministic model. Colored points show the
+predicted \(\lambda\) values for the twelve translocated populations
+when probability of recruitment into the adult class is \(\omega = 0.3\)
+(indicated by the dashed line). The red line shows where
+\(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
+12 translocated populations given demographic stochasticity,
+environmental variability in \(\omega\), and different mean values of
+\(\omega\). The dashed line shows where \(\omega = 0.3\), for comparison
+to (A). There are 6 lines at extinction probability = 1, 5 of which
+(70414--70619) are hidden beneath the line for 70505. (C) 100 simulated
+trajectories (black lines) from the population viability model that most
+closely matched the observed abundance trajectory of adult amphibians at
+site 70550 (light orange).}
 

### 2023-05-12 doc/manuscript/translocation_pnas_SI.tex
Make three panel model figure and update main text

@@
 gene cluster lies on the same chromosome as the outlier SNP in RIN3
-(Fig. 2 C), a gene associated with
-inflammation response that is highly expressed in amputated
-\emph{Xenopus} tails \citep{fukazawa2009}.
+(Fig. 2 C, Fig. 1
+A).
 
@@
 determine how small changes in these parameters could influence the
-long-run deterministic growth rate of populations.
+long-run deterministic growth rate of populations
+(Figure~\ref{fig-viability-supp}).
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp.jpg}
+
+}
+
+\caption{\label{fig-viability-supp}Sensitivity analysis of the
+stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
+in 3 parameters: \(\sigma_{A_R}\) (yearly survival probability of
+naturally recruited adults), \(\omega\) (yearly probability of
+successful recruitment from juveniles to adults), and \emph{F (}number
+of eggs produced by a female frog in a year that successfully hatch).
+Elasticity is calculated at the default parameter values for each
+population and \(\omega\) = 0.3.}
+
+\end{figure}\clearpage
+
 \newpage

### 2023-05-12 code/extinction_and_growth_rate_analysis.R

@@ -101,7 +101,7 @@ ggsave(file.path("..", "out", "lambda_estimates.pdf"), width=8, height=5)
 #############################################################
 
 
-muR_range = seq(0.0, 0.9, len=50)
+muR_range = seq(0.01, 0.9, len=50)
 omega_range = seq(0.01, .4, len=50)
 lambdas = array(NA, dim=c(length(muR_range), length(omega_range)))
 
@@ -125,12 +125,15 @@ ptile = ggplot(dat) + geom_tile(aes(x=muR, y=omega, fill=lambda)) +
 																												y=rep(0.3, length(surv_med)), 
 																													color=lake_id), size=2) +
 		   				scale_color_manual(values=colors) +
-							scale_fill_gradient("\u03bb", low = "white", high = "black") +
+							# scale_fill_gradient("\u03bb", low = "white", high = "black") +
+							scale_fill_gradient2(midpoint=1) + #guide=guide_legend(title.position="bottom")) +
 							theme_classic() + xlab(bquote("Adult survival probability (\u03c3"~.[AR]~")")) +
 							ylab("Prob. of recruitment (\u03c9)") + 
-							annotate("text", x=0.25, y=0.05, label="Population declines") +
-							annotate("text", x=0.7, y=0.35, label="Population grows", color="white") +
-							guides(color="none")
+							annotate("text", x=0.25, y=0.05, label="Population declines", size=3) +
+							annotate("text", x=0.7, y=0.35, label="Population grows", size=3) +
+							guides(color="none", fill=guide_colorbar(title="\u03bb")) + theme(legend.position="bottom", 
+																					 legend.text=element_text(size=7), 
+																					 legend.title=element_text(size=7))
 
 
 #############################################################
@@ -345,8 +348,9 @@ pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=omega, y=e
 							geom_line(aes(color=lake_id_surv)) + 
 							geom_point(aes(color=lake_id_surv)) + 
 							scale_color_manual(values=cdat$col) +
-							theme_classic() + xlab("Probability of recruitment (\u03c9)") + ylab("Extinction probability in 50 years") +
-							guides(color=guide_legend(title=bquote("Lake, \u03c3"~.[AR])))
[…]

### 2023-05-12 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -24,6 +24,11 @@ with open(args[1], 'r') as fin:
     latex_manuscript = fin.readlines()
 latex_manuscript = "".join(latex_manuscript)
 
+# Remove all the references to width and height
+latex_manuscript = re.sub("includegraphics\[.*\]", 
+						  "includegraphics", 
+						  latex_manuscript)
+
 with open("PNAS-template-empty.tex") as fin:
 	pnas_template = fin.readlines()
 pnas_template = "".join(pnas_template)

### 2023-05-12 code/translocation_survival_analysis.Rmd
Add comment to load_packages code chunk
Edit axis label to fit better in multipanel figure

@@
 
-source("code/classification_summary.R")
+source("code/classification_summary.R") # When running code chunks directly from within .Rmd (instead of via Makefile), remove "code/".
 ```
@@
         # geom_point(aes(y = .epred)) +
-          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival")
+          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year survival")
 
@@
           geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
-          labs(x = "Elevation of recipient site (meters)", y = "P(1-year frog survival)") +
+          labs(x = "Elevation of recipient site (meters)", y = "P(1-year survival)") +
           annotate(geom = "text", x = 2500, y = 0.94, label = "B", size = 4) +
@@
           geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
-          labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year frog survival)") +
+          labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year survival)") +
           annotate(geom = "text", x = 157, y = 0.92, label = "A", size = 4) +
@@
     geom_boxplot(aes(y = .epred, x = sex), fill = "lightgray") +
-    labs(x = "Sex", y = "P(1-year frog survival)") +
+    labs(x = "Sex", y = "P(1-year survival)") +
     annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "D", size = 4) +
@@
     geom_boxplot(aes(y = .epred, x = donor), fill = "lightgray") +
-    labs(x = "Donor population", y = "P(1-year frog survival)") +
+    labs(x = "Donor population", y = "P(1-year survival)") +
     annotate(geom = "text", x = 3.3, y = 0.94, hjust = 0, label = "C", size = 4) +
@@
           geom_line(aes(y = .epred, group = paste(sex, .draw)), linewidth = 0.4, alpha = 0.5) +
-          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
+          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year survival") +
           scale_color_manual(values = pal) + 
@@
           geom_line(aes(y = median, color = sex), linewidth = 1, alpha = 1) +
-          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year frog survival") +
+          labs(x = "Elevation of recipient site (meters)", y = "Probability of 1-year survival") +
           scale_color_manual(values = pal) + 

### 2023-05-12 doc/manuscript/translocation.qmd
Edits throughout text in response to coauthor comments

@@
 Roland A. Knapp^1,2,^\*, Mark Q. Wilber^3,a^, Allison Q. Byrne^4,5,a^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
-
-<!--# Exact author order to be determined -->
 
@@
 Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-One possibility is evolutionary modification that restores positive growth to declining populations ("evolutionary rescue").
+One possibility is rapid evolution that restores positive growth to declining populations ("evolutionary rescue").
 Individuals from these rescued populations could also be used to reestablish exinct populations.
@@
 In addition, introduction of frogs from these populations allowed reestablishment despite the ongoing presence of disease.
-Such reintroductions could play an important role in biodiversity conservation amidst unprecedented global change.
+Such reintroductions could play an important role in biodiversity conservation in our rapidly changing world.
 
@@
 Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amid this global loss of biodiverity, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
+Amid this global loss of biodiverity, some populations decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
 Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
@@
 Endangered Species Act.
-In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B.
-dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
+In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
+The arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
 These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
@@
 
-To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
 Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
@@
 
-We conducted a principal component analysis (PCA) of the genomic data to describe the association between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
@@
 Frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B, SI), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
@@
 Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
-Therefore, frog resistance in non-natal habitats is maintained across generations.
+Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.
 
@@
 Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the gene GVINP1.
-Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014].
+Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018].
 This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [@robledo2020; @robledo2018].
+We also identified a region, characterized by high *F~ST~* and low $\pi_{diff}$, that contained the complement genes C6 and C7.
+This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
 Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
 This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
-In the SI, we provide additional details on these and other genes identified in our study as under selection<!--# Reference complement genes. See Allie's comment in the SI for more on this.  -->.
+In the SI, we provide additional details on these and other genes identified in our study as under selection.
 
@@
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below).
+We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI).
 
@@
 Barbella and K.
-Rose--laboratory work; numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
+Rose--laboratory work; C.
+Kamoroff and numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
 Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
@@
 ![Evidence for reduced impacts of Bd on MYL frogs as indicated by the lack of an effect of Bd load on survival.
-Results show conditional effects of the important predictors of 1-year frog survival.
+Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
@@
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}
+(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
 
@@
 In the Results section of the main paper, we describe the stringent set of outlier SNPS (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
+The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
+Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
 
+In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
+Nonetheless, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6, 7, 8, and 9; @fig-allelefrequencies, @fig-selectionresults A SI).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
+<!--# Allie, check text for accuracy -->
+
+<!--#  CUT?
 Multiple genes in our stringent set of outliers have known associations with amphibian Bd resistance/tolerance.
-Of particular interest is a \~940kb region on contig 8 that we identified as under selection, and that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
+Of particular interest is a ~940kb region on contig 8 that we identified as under selection, and that contains 8 genes associated with either the MHC Class I or Class III regions.
+The order of these genes in the MYL frog genome is conserved from Xenopus [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
 Numerous studies have linked MHC genes to frog resistance/tolerance against Bd infection [e.g., @savage2011; @bataille2015].
@@
 For example, we identified the MHC Class III gene, heat shock protein 70 (HSP70), as occurring in this region (@fig-synteny-plot).
-Similarly, [@byrne2020] identified an HSP70-binding protein as a candidate gene under selection in *Atelopus varius* populations recovering after Bd-caused declines.
-HSP70 was also highly expressed in Bd-infected versus uninfected *Xenopus tropicalis* [@rosenblum2009].
+Similarly, [@byrne2020] identified an HSP70-binding protein as a candidate gene under selection in Atelopus varius populations recovering after Bd-caused declines.
+HSP70 was also highly expressed in Bd-infected versus uninfected Xenopus tropicalis [@rosenblum2009].
 This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3 (@fig-spline-manhattan C, @fig-allelefrequencies A).
-
 Additionally, we identified a region on contig 19 that had the highest $\pi_{diff}$ of any outlier window, indicative of directional selection.
@@
 Variants in GVINP1 also explained approximately 20% of the resistance to amoebic gill disease in Atlantic salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, in *Litoria verreauxii alpina*, the related gene<!--# Correct?
-Or is this refering to GVINP1?
--->, IFN-induced very large GTPase, showed lower expression in frogs from less susceptible populations versus more susceptible populations [@grogan2018].
-
-Finally, we identified a region characterized by high *F~ST~* and low $\pi_{diff}$ and that contained the complement genes C6 and C7.
+Furthermore, in Litoria verreauxii alpina, the related gene, IFN-induced very large GTPase, showed lower expression in frogs from less susceptible populations versus more susceptible populations [@grogan2018].Finally, we identified a region characterized by high FST and low $\pi_{diff}$ and that contained the complement genes C6 and C7.
 This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-
-<!--# Mention that the patterns shown in the allele frequencies figure (panel A, B) may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors?
--->
-
-<!--# Brief description of recipient populations, perhaps in table format?
-Includes mostly individual relatively isolated lakes, but also some lake-pond-stream complexes and one stream-meadow-pond complex.
 -->
@@
 
-![Results from the exome capture selection study.
+![Evidence for selection of individual variants in recovering MYL frogs.
 (A) PCA calculated from binary SNPs showing genomic relationship of samples.
 Numeric labels and colors match those in @fig-allelefrequencies A.
-(B) qqplot showing observed and expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+(B) qqplot showing observed and expected p-values for 148,307 SNPs and INDELS as calculated in GEMMA.
 Dashed line shows p-value that identifies outliers.
@@
 
-![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
+![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
 Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
@@
 
-![For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
+![Evidence that Bd load in not an important predictor of post-translocation frog survival.
+For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
 
-![Comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
+![Evidence that frog resistance/tolerance in non-natal habitats is maintained across generations.
+Comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
 In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohort.
@@
 
-Dataset 1: gemma_outliers.csv
+Dataset S1.
+Set of liberal SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.05), identified via GEMMA.
+file: gemma_outliers_all.csv
 
-Dataset 2: spline_window_shared_outliers.csv <!--# The dataset contains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
+Dataset S2.
+Set of strict SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.01), identified via GEMMA.
+Additional information includes variant location within the gene (predicted_gene_loc) and whether the variant is synonymous or nonsynonymous (predicted_effect_AA).
+file: gemma_outliers_strict_freq.csv
+
+Dataset S3.
+Description of overlapping *F~ST~* and $\pi_{diff}$ outlier windows, as identified in the splined window analysis.
+file: spline_window_shared_outliers.csv <!--# The dataset contains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
 Do we need to explain that one region is a repeat because there is a 2>1 match for one of the overlaps (i.e., one Fst outlier window overlaps with 2 pi outlier windows)?
@@
 
-\
+Dataset S4.
+Annotation information for all genes within each of the overlapping outlier windows in dataset S3.
+file: spline_window_gene_details
 

### 2023-05-13 doc/manuscript/translocation.qmd
Minor clarifying edits to Intro

@@
 To avoid extinction, affected populations will often need to adapt to the altered environment.
-This process of rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth is termed "evolutionary rescue".
+The process of rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth is termed "evolutionary rescue".
 In addition to allowing recovery of extant populations, individuals from rescued populations could also be used to reestablish extirpated populations.
-We evaluated this scenario using the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
+We evaluated this scenario in the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
 Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
@@
 Individuals from these rescued populations could also be used to reestablish exinct populations.
-We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse its impacts.
+We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse disease impacts.
 We found that frogs have evolved increased resistance to infection, facilitating population recovery.
@@
 "Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth [@carlson2014].
-In addition to preventing extinction of declining populations, in theory, this process can also result in the seeding of nearby empty habitats and allow population reestablishment and recovery.
-Therefore, understanding the mechanisms supporting population persistence, including evolutionary rescue, is critical for the conservation of the ever-growing number of threatened taxa.
+In addition to preventing extinction of declining populations, this process could also result in the seeding of nearby empty habitats and allow population reestablishment and recovery.
+Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of threatened taxa, descriptions of its operation in the wild are lacking.
 
@@
 Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amid this global loss of biodiverity, some populations decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
-Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
+Amid this global loss of biodiverity, some populations decimated by disease have subsequently recovered to varying extents [@newell2013; @voyles2018; @knapp2016].
+Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
 This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see [@medzhitov2012] for definitions).
-However, linking these evolutionary responses to population recoveries - as necessary for evolutionary rescue - is difficult and few clear examples exist.
-In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019; @brannelly2021].
+However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
+In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
@@
 The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
-Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
+Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
 Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
@@
 In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
-The arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
+The arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
 These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
@@
 
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then describe the reintroduction of these resistant/tolerant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
+In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
+We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
 These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
@@
 Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].
+These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].
 
 We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
 Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
 
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A: SI), and clusters reflected the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish between (i) allelic associations of individuals sampled in naive versus recovering populations, and (ii) allelic associations resulting from population structure and genetic drift.
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
 
-Results from the SNP and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults B, C: SI, SI Dataset 1).
+Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI).
 One of the 7 identified genes did not have an associated annotation.
 Frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B, SI), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
 
-The splined window analysis identified 58 outlier regions for *F~ST~* and 33 outlier regions for $\pi_{diff}$ (@fig-spline-manhattan A, B).
+The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
 Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
-A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2).
+A total of 33 annotated genes were found in the 9 shared regions (SI).
 The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
@@
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a low $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
+Interestingly, this region had a low $\pi_{diff}$ value <!--# Negative value shown in 2A? -->, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
 This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
@@
 To visualize the genomic relationship of our populations we conducted a principal components analysis (PCA) using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome under selection between naive and recovering populations, we used two approches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
 For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA.
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified "outlier" variants using a Bonferroni- corrected significance level of 0.01.
-Outliers are variants showing different allele frequencies between naive versus recovering populations.
+We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni- corrected significance level of 0.01.
 We visualized the results using a Manhattan plot and qqplot.
@@
 ![Evidence for selection in recovering MYL frog populations at the landscape scale.
-For each of the naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+For each of the 9 naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
@@
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and II regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
 
@@
 
-\
-
-![Evidence for selection of individual variants in recovering MYL frogs.
-(A) PCA calculated from binary SNPs showing genomic relationship of samples.
+![Evidence for selection of individual variants in recovering MYL frog populations.
+(A) PCA calculated from binary SNPs showing the genomic relationship of samples.
 Numeric labels and colors match those in @fig-allelefrequencies A.
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults}
+Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+
+<!--# Remove x = 0 and y = 0 lines in PC plot -->
 

### 2023-05-13 doc/manuscript/translocation.qmd
Convert manuscript to PNAS after small edits

@@
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a low $\pi_{diff}$ value <!--# Negative value shown in 2A? -->, consistent with balancing selection.
+Interestingly, this region had a low $\pi_{diff}$ value<!--# Negative value shown in 2A? -->, consistent with balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
@@
 When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
-In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
+Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis).
 
@@
 We checked convergence of the MCMC algorithm using trace plots and Rhat.
-Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+Code to create the datasets and fit models is available from the following GitHub repository: [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).
 
@@
 Priors, chains, iterations, and checks were as described above.
-Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
+Code to create the dataset and fit models is available from the following GitHub repository: [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).
 <!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. --> <!--# Add  -->
@@
 After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
-First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to growth or decline in the long-run.
+First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
 Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
@@
 Briggs), DEB-1557190 (to C.
-Briggs), and DBI-2120084 (to C.
+Briggs), DEB-2133401 (to M.Q.W), and DBI-2120084 (to C.
 Richards-Zawacki).
@@
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+Code to replicate the analyses can be found at [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).
 

### 2023-05-13 doc/manuscript/translocation_pnas.tex
Convert manuscript to PNAS after small edits

@@
 \significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. One possibility is
-evolutionary modification that restores positive growth to declining
-populations (``evolutionary rescue''). Individuals from these rescued
-populations could also be used to reestablish exinct populations. We
-evaluated both possibilities in a once-common frog species driven to the
-brink of extinction by a disease that has devastated global amphibian
-biodiversity and for which few options exist to reverse its impacts. We
-found that frogs have evolved increased resistance to infection,
+critical for the conservation of biodiversity. One possibility is rapid
+evolution that restores positive growth to declining populations
+(``evolutionary rescue''). Individuals from these rescued populations
+could also be used to reestablish exinct populations. We evaluated both
+possibilities in a once-common frog species driven to the brink of
+extinction by a disease that has devastated global amphibian
+biodiversity and for which few options exist to reverse disease impacts.
+We found that frogs have evolved increased resistance to infection,
 facilitating population recovery. In addition, introduction of frogs
@@
 presence of disease. Such reintroductions could play an important role
-in biodiversity conservation amidst unprecedented global change.}
+in biodiversity conservation in our rapidly changing world.}
 
@@
 avoid extinction, affected populations will often need to adapt to the
-altered environment. This process of rapid evolutionary change that
+altered environment. The process of rapid evolutionary change that
 increases the frequency of adaptive alleles and thereby restores
@@
 rescued populations could also be used to reestablish extirpated
-populations. We evaluated this scenario using the once-common mountain
+populations. We evaluated this scenario in the once-common mountain
 yellow-legged (MYL) frog, which is threatened with extinction by the
@@
 thereby restores positive population growth \citep{carlson2014}. In
-addition to preventing extinction of declining populations, in theory,
-this process can also result in the seeding of nearby empty habitats and
-allow population reestablishment and recovery. Therefore, understanding
-the mechanisms supporting population persistence, including evolutionary
-rescue, is critical for the conservation of the ever-growing number of
-threatened taxa.
+addition to preventing extinction of declining populations, this process
+could also result in the seeding of nearby empty habitats and allow
+population reestablishment and recovery. Although this overall scenario
+is potentially of critical importance for the conservation of an
+ever-growing number of threatened taxa, descriptions of its operation in
+the wild are lacking.
 
@@
 \citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
-global loss of biodiverity, some populations previously decimated by
-disease have subsequently recovered to varying degrees
+global loss of biodiverity, some populations decimated by disease have
+subsequently recovered to varying extents
 \citep{newell2013, voyles2018, knapp2016}. Although we know relatively
-little about the underlying mechanisms
+little about the mechanisms underlying recovery
 \citep{brannelly2021, russell2020}, host evolution is one possibility
@@
 \citep{medzhitov2012} for definitions). However, linking these
-evolutionary responses to population recoveries - as necessary for
-evolutionary rescue - is difficult and few clear examples exist. In
-addition to promoting recovery of disease-impacted populations, the
-resistant/tolerant individuals in such populations could potentially be
-used to reestablish populations that were extirpated following disease
-emergence \citep[e.g.,][]{mendelson2019, brannelly2021}. To date, the
-use of resistant/tolerant individuals for species recovery is rare
-\citep[but see][]{joseph2018}, and studies that evaluate its potential
-utility are critically needed.
+evolutionary responses to population recoveries is difficult and few
+clear examples exist. In addition to promoting recovery of
+disease-impacted populations, the resistant/tolerant individuals in such
+populations could potentially be used to reestablish populations that
+were extirpated following disease emergence
+\citep{mendelson2019, brannelly2021}. To date, the use of
+resistant/tolerant individuals for species recovery is rare \citep[but
+see][]{joseph2018}, and studies that evaluate its potential utility are
+critically needed.
 
@@
 most common amphibian in the high elevation portion of California's
-Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
-disappeared from more than 90\% of its historical range during the past
-century \citep{vredenburg2007}. Due to the severity of its decline and
-the increasing probability of extinction, both species are now listed as
+Sierra Nevada mountains \citep[USA,][]{grinnell1924}, during the past
+century this frog has disappeared from more than 90\% of its historical
+range \citep{vredenburg2007}. Due to the severity of its decline and the
+increasing probability of extinction, both species are now listed as
 ``endangered'' under the U.S. Endangered Species Act. In the Sierra
@@
 trout into the extensive historically-fishless region
-\citep{bradford1989, knapp2000}, and the arrival of \emph{B.
-dendrobatidis} (Bd) in the 1960s and its subsequent spread
-\citep{vredenburg2019} caused additional widespread population
+\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
+of \emph{B. dendrobatidis} (Bd) in the 1960s and its subsequent spread
+\citep{vredenburg2019} caused additional large-scale population
 extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
@@
 In the current study, we combine results from genomic analyses, a
-15-year landscape-scale frog reintroduction effort, and detailed
-population models to provide an unprecedented example of how evolution
-in response to disease allows the successful reestablishment of
-extirpated MYL frog populations. Specifically, we used exome capture
-methods to show that MYL frogs have undergone substantial evolutionary
-change associated with Bd exposure and that may provide a genetic basis
-for the observed increased resistance/tolerance to Bd infection observed
-in naturally recovering populations \citep{knapp2016}. We then describe
-the reintroduction of these resistant/tolerant frogs to 24 sites and the
-dynamics and drivers of population establishment using results from
-long-term capture-mark-recapture (CMR) surveys. These data are also used
-to parameterize a population model that extends our inferences of
+15-year landscape-scale frog reintroduction effort, and a detailed
+population model to provide an unprecedented example of how evolution in
+response to disease allows the successful reestablishment of extirpated
+MYL frog populations. Specifically, we used exome capture methods to
+show that MYL frogs have undergone substantial evolutionary change
+associated with Bd exposure and that might provide a genetic basis for
+the increased resistance/tolerance to Bd infection observed in naturally
+recovering populations \citep{knapp2016}. We then use the reintroduction
+of resistant/tolerant frogs to 24 sites and results from long-term
+capture-mark-recapture (CMR) surveys to describe the dynamics and
+drivers of population establishment. These data are also used to
+parameterize a population model that extends our inferences of
 population establishment well beyond the temporal extent of the CMR
@@
 
-To determine whether frog populations show an evolutionary response to
-Bd, we compared genomes of frogs in populations with different histories
-of Bd exposure. Specifically, we compared frog genomes sampled in four
-populations that have not yet experienced a Bd-caused epizootic
+To determine whether MYL frog populations show an evolutionary response
+to Bd, we compared frog genomes between populations with contrasting
+histories of Bd exposure. Specifically, we compared frog genomes sampled
+in four populations that have not yet experienced a Bd-caused epizootic
 (``naive'') \citep{zhou2015} versus in five populations that experienced
@@
 to varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
-\citep{knapp2016, vredenburg2010}. These recovering populations persist
-in an enzootic state \citep{briggs2010}, characterized by high Bd
+\citep{knapp2016, vredenburg2010}. These recovering populations exist in
+an enzootic state \citep{briggs2010}, characterized by high Bd
 prevalence and, in adults, moderate Bd infection intensities (``load'')
@@
 We conducted a principal component analysis (PCA) of the genomic data to
-describe the association between sampled populations, and then used two
-complementary approaches to identify regions of the genome under
+describe the relationships between sampled populations, and then used
+two complementary approaches to identify regions of the genome under
 selection. First, we used a multivariate linear mixed model to evaluate
 associations between population type (i.e., naive versus recovering) and
-specific variants, including single nucleotide polymorphisms (SNPs) and
-insertions/deletions (INDELS). Second, we used a splined window analysis
-to identify larger genomic regions showing differences between
+individual variants, including single nucleotide polymorphisms (SNPs)
+and insertions/deletions (INDELS). Second, we used a splined window
+analysis to identify larger genomic regions showing differences between
 population types in \emph{F\textsubscript{ST}} and nucleotide diversity
@@
 Individual frogs clustered into 3 separate groups in principal component
-space (Figure S1A), and clusters reflected
-the strong signature of isolation by distance that is characteristic of
-MYL frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
+space (Figure S1A), and clusters reflected the
+strong signature of isolation by distance that is characteristic of MYL
+frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
 contained at least one population from both the naive and recovering
-groups, allowing us to distinguish between (i) allelic associations of
-individuals sampled in naive versus recovering populations, and (ii)
-allelic associations resulting from population structure and genetic
-drift.
-
-Results from the SNP and splined window analyses show that recovering
-populations have signatures of strong selection on multiple regions of
-the genome. The SNP analysis identified 11 outlier SNPs from 7 distinct
-genes across 4 contigs (Figure S1B,C, SI
-Dataset 1). One of the 7 identified genes did not have an associated
+groups, allowing us to distinguish allelic associations of individuals
+sampled in the 2 population types versus allelic associations resulting
+from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show
+that recovering populations have signatures of strong selection on
+multiple regions of the genome. The analysis of individual variants
+identified 11 ``outlier'' SNPs (i.e., showing significantly different
+allele frequencies between naive versus recovering populations) from 7
+distinct genes across 4 contigs (Figure S1B,C). One of the 7 identified genes did not have an associated
 annotation. Frequency differences between the naive and recovering
@@
 frequency differences in only a subset of the sampled populations
-(Figure~\ref{fig-allelefrequencies} A, B), but the outlier SNP in the
-RIN3 gene showed consistent differences in frequencies across all
-populations (Figure~\ref{fig-allelefrequencies} C). This is suggestive
-of strong - and parallel - selection at this locus associated with
-exposure of frogs to Bd across multiple populations.
-
-The splined window analysis identified 58 outlier regions for
-\emph{F\textsubscript{ST}} and 33 outlier regions for \(\pi_{diff}\)
+(Figure~\ref{fig-allelefrequencies} A, B), but the SNP in the RIN3 gene
+showed consistent differences in frequencies across all populations
+(Figure~\ref{fig-allelefrequencies} C). This is suggestive of strong -
+and parallel - selection at this locus associated with exposure of frogs
+to Bd across multiple populations.
+
+The splined window analysis identified 33 outlier regions for
+\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
 (Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
@@
 outlier SNPs. A total of 33 annotated genes were found in the 9 shared
-regions (SI Dataset 2). The largest \(\pi_{diff}\), indicative of
-directional selection, occurred in a 163kb region on Contig19, 12.9Mb
-upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C).
-This region contains approximately 500 SNPs and one annotated gene
-called ``interferon-induced very large GTPase 1-like'' (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two
-complement factor genes (C6 and C7). Interestingly, this region had a
-low \(\pi_{diff}\) value, consistent with balancing selection. Finally,
-one shared outlier region on Contig8 contained one outlier SNP (TCF19)
-and was within 360kb of another outlier SNP (VARS)
-(Figure~\ref{fig-spline-manhattan} D, Figure S2).
-This region (854kb from the beginning of the outlier window to the VARS
-SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
-these genes occur in the extended major histocompatibility complex (MHC)
-Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
-extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
-Therefore, this region under selection is part of the extended MHC Class
-I and III complex and shows synteny with other amphibian genomes.
+regions (SI). The largest \(\pi_{diff}\), indicative of directional
+selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
+the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C). This region
+contains approximately 500 SNPs and one annotated gene called
+``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
+a shared outlier region on Contig1 contained two complement factor genes
+(C6 and C7). Interestingly, this region had a low \(\pi_{diff}\) value,
+consistent with balancing selection. Finally, one shared outlier region
+on Contig8 contained one outlier SNP (TCF19) and was within 360kb of
+another outlier SNP (VARS) (Figure~\ref{fig-spline-manhattan} D,
+Figure S2). This region (854kb from the beginning of
+the outlier window to the VARS SNP) contained a total of 8 annotated
+genes. In \emph{Xenopus}, five of these genes occur in the extended
+major histocompatibility complex (MHC) Class I region (FLOT1, TUBB,
+MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III
+region (HSP70, LSM2, VARS) \citep{ohta2006}. Therefore, this region
+under selection is part of the extended MHC Class I and III complex and
+shows synteny with other amphibian genomes.
 
@@
 translocated adults (Figure S7). Therefore,
-frog resistance in non-natal habitats is maintained across generations.
+frog resistance/tolerance in non-natal habitats is maintained across
+generations.
 
@@
 (Figure~\ref{fig-viability} B). In contrast, for the six populations
-where \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was
-always predicted to be 1. To test the validity of our model predictions,
-we demonstrated that our stochastic model could describe the general
-recovery trajectory of our translocated population with the longest
-survey history (Figure~\ref{fig-viability} C; population 70550, surveyed
-for 16 years).
+where yearly adult survival probability \(\sigma_{A_R} < 0.5\),
+extinction probability over 50 years was always predicted to be 1. To
+test the validity of our model predictions, we demonstrated that our
+stochastic model could describe the general recovery trajectory of our
+translocated population with the longest survey history
+(Figure~\ref{fig-viability} C; population 70550, surveyed for 16 years).
 
@@
 found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018, ellison2014}. This gene is also strongly
-linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3 gene
-showed a consistent pattern of allele frequency differences across all
-nine of the frog populations sampled in this study, indicating
-consistent selection in populations distributed across a wide geographic
-area. This gene is associated with inflammation response and in
-\emph{Xenopus} is highly expressed in wound stumps following amputation
-of tadpole tails \citep{fukazawa2009}. In the SI, we provide additional
-details on these and other genes identified in our study as under
-selection.
+\citep[e.g.,][]{grogan2018, ellison2014} and in populations differing in
+Bd susceptibility \citep{grogan2018}. This gene is also strongly linked
+to disease in salmon, explaining a notable 20\% of the resistance
+phenotype \citep{robledo2020, robledo2018}. We also identified a region,
+characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
+that contained the complement genes C6 and C7. This could indicate that
+balancing selection is acting in this region of the genome to favor a
+diverse set of alleles, as is known for C6 in humans
+\citep{soejima2005}. Finally, the RIN3 gene showed a consistent pattern
+of allele frequency differences across all nine of the frog populations
+sampled in this study, indicating consistent selection in populations
+distributed across a wide geographic area. This gene is associated with
+inflammation response and in \emph{Xenopus} is highly expressed in wound
+stumps following amputation of tadpole tails \citep{fukazawa2009}. In
+the SI, we provide additional details on these and other genes
+identified in our study as under selection.
 
@@
 under selection between naive and recovering populations, we used two
-approches: (1) a multivariate linear mixed model to evaluate individual
+approaches: (1) a multivariate linear mixed model to evaluate individual
 variants (SNPs and INDELs), and (2) a splined window analysis to
@@
 allowing us to account for relatedness and population structure when
-calculating likelihood ratio statistics. We identified ``outlier''
-variants using a Bonferroni- corrected significance level of 0.01.
-Outliers are variants showing different allele frequencies between naive
-versus recovering populations. We visualized the results using a
-Manhattan plot and qqplot. We developed a more liberal set of outlier
-variants using a Bonferroni- corrected significance level of 0.05 and
-used this set solely for the gene ontology (GO) analysis (see below).
+calculating likelihood ratio statistics. We identified variants showing
+different allele frequencies between naive versus recovering populations
+(``outliers'') using a Bonferroni- corrected significance level of 0.01.
+We visualized the results using a Manhattan plot and qqplot. We
+developed a more liberal set of outlier variants using a Bonferroni-
+corrected significance level of 0.05 and used this set solely for the
+gene ontology (GO) analysis (see below and SI).
 
@@
 models for each translocated population is available from the following
-GitHub repository:
-\href{https://github.com/SNARL1/cmr-analysis.}{https://github.com/SNARL1/cmr-analysis}.
+GitHub repository: \url{https://github.com/SNARL1/cmr-analysis}.
 
@@
 12 translocated populations to determine if the populations were
-deterministically predicted to growth or decline in the long-run.
-Second, we compute the elasticity of \(\lambda\) to model parameters, to
+deterministically predicted to grow or decline in the long-run. Second,
+we compute the elasticity of \(\lambda\) to model parameters, to
 quantify how much changes in adult survival, for example, affected the
@@
 Hegeman and A. Lindauer--project and data management, field and
-laboratory work; A. Barbella and K. Rose--laboratory work; numerous
-summer technicians--field work; staff at Sequoia-Kings Canyon and
-Yosemite National Parks, Inyo and Sierra National Forests, California
-Department of Fish and Wildlife, U.S. Fish and Wildlife Service,
-University of California-Santa Barbara Institutional Animal Care and Use
-Committee, and Sierra Nevada Aquatic Research Laboratory--research
-permits, logistical support, and/or field assistance. This project was
-supported by grants from the National Park Service (to R.A.K.), Yosemite
-Conservancy (to R.A.K.); and National Science Foundation Grants
-EF-0723563 (to C. Briggs), DEB-1557190 (to C. Briggs), and DBI-2120084
-(to C. Richards-Zawacki).}
+laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
+and numerous summer technicians--field work; staff at Sequoia-Kings
+Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
+California Department of Fish and Wildlife, U.S. Fish and Wildlife
+Service, University of California-Santa Barbara Institutional Animal
+Care and Use Committee, and Sierra Nevada Aquatic Research
+Laboratory--research permits, logistical support, and/or field
+assistance. This project was supported by grants from the National Park
+Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National
+Science Foundation Grants EF-0723563 (to C. Briggs), DEB-1557190 (to C.
+Briggs), DEB-2133401 (to M.Q.W), and DBI-2120084 (to C.
+Richards-Zawacki).}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/allele_maps.png}
+{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
 
@@
 recovering MYL frog populations at the landscape scale. For each of the
-naive and recovering MYL frog populations, pie charts show allele
+9 naive and recovering MYL frog populations, pie charts show allele
 frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/splinewindow_manhattan.png}
+{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}
 
@@
 TCF19 and VARS. This region of the genome contains 8 annotated genes
-known to occur in the extended MHC Class I and II regions.}
+known to occur in the extended MHC Class I and III regions.}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/translocation_survival_bysiteid.png}
+{\centering \includegraphics[width=\textwidth]{figures/translocation_survival_bysiteid.png}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/cond_effects_plot.png}
+{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
 
@@
 Results show conditional effects of the important predictors of 1-year
-frog survival. (A) winter severity in the year following translocation,
-(B) elevation of recipient site, (C) donor population, and (D) sex. In
-(A) and (B), blue lines are medians and gray ribbons are 95\%
-uncertainty intervals. In (C) and (D), box plots show medians, first and
-third quartiles, largest and smallest values within 1.5x interquartile
-range, and values outside the 1.5x interquartile range.}
+frog survival (survival is expressed as a probability). (A) winter
+severity in the year following translocation, (B) elevation of recipient
+site, (C) donor population, and (D) sex. In (A) and (B), blue lines are
+medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
+box plots show medians, first and third quartiles, largest and smallest
+values within 1.5x interquartile range, and values outside the 1.5x
+interquartile range.}
 

### 2023-05-15 doc/manuscript/translocation.qmd
Minor edits throughout document

--- 2023-05-13 doc/manuscript/translocation.qmd
@@
 
-**Preprint Server:**
+**Preprint Server:** bioRxiv
 
@@
 
-<!--# Limit = 120 words. Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty.  -->
+<!--# Limit = 120 words.   -->
 
@@
 These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].
+(See SI for additional details regarding the study design.)
 
@@
 
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+muscosa* versus *R.
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
 Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
@@
 Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI).
-One of the 7 identified genes did not have an associated annotation.
-Frequency differences between the naive and recovering populations ranged from 41% to 86%.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
+One of the 7 identified genes did not have an associated annotation (LOC108802036<!--# Allie - check for accuracy -->).
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
 Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+(See SI for additional description of these results.)
 
 The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more outlier SNPs (described above).
 A total of 33 annotated genes were found in the 9 shared regions (SI).
@@
 Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a low $\pi_{diff}$ value <!--# Negative value shown in 2A? -->, consistent with balancing selection.
+Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
@@
 
-To determine whether *R.
-sierrae* from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@seddon2014].
+To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap) [@seddon2014].
 The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
@@
 Estimates of frog survival during the 1 year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
 sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter SI, see SI text for additional details) [@joseph2018; @vredenburg2010].
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter, see SI text for additional details) [@joseph2018; @vredenburg2010].
 
@@
 To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
 Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
+Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens).
 Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
-In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
-Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
 Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.
@@
 Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
+We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).
 
@@
 When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
-<!--# Consider condensing the above paragraph and providing details in SI.
-Next paragraph topic sentence could also provide a more direct statement of key results, e.g., something like "Even when incorporating demographic and environmental stochasticity, populations with high adult survival are likely to persist over a 50 year time horizon."  Mark here:  Addressed both of these comments.
--->
 
-Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
 When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
@@
 In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
+This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
 In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
@@
 Disease-induced population declines are decimating global biodiversity [@daszak2000], but strategies to recover affected species are generally lacking.
-Here, we used populations of resistant/tolerant individuals to reestablish extirpated populations, despite ongoing disease.
-Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
+Here, we tested the possibility that populations of resistant/tolerant individuals could be used to reestablish extirpated populations.
+Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
 In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
-The results of our study provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
-<!--# Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014]...
---> <!-- I am wondering whether it is worth digging a bit deeper into the "difficult and perhaps impossible" task alluded to above.
-e.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data -->
+The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (2) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing disease.
 
@@
 Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
-In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter - before translocation].
+In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter].
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
-The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
+The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors, including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
 
@@
 This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
-In the SI, we provide additional details on these and other genes identified in our study as under selection.
 
@@
 Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
-In an earlier study of 2 of our 12 translocated populations, Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in [@joseph2018] are identified as "70550" and "70505" in the current study).
+In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in [@joseph2018] are identified as "70550" and "70505" in the current study).
 Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
-In addition to these important but relatively subtle effects on adults, MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and experience high mortality.
+In addition to these important but relatively subtle effects of Bd on adults, MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and experience high mortality.
 This high susceptibility of metamorphs is documented for numerous species of anurans, and may result from the poorly developed immune system characteristic of this early life stage [@humphries2022].
@@
 Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
-The results of this study provide a hopeful starting point for that endeavor, and other future efforts worldwide.
+The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
 
@@
 Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
-We expect that such "assisted evolutionary rescue" will be an essential tool in wildlife conservation in an era of accelerating global change.
+We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.<!--# Dig deeper into the "difficult" task alluded to above?
+E.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data?
+-->
 
@@
 
-<!--# To meet word count limits, we will likely need to move some of the following methods to the SI.  -->
-
 ##### Sampling and Sequencing
 
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four Bd-naive populations, and 29 from five recovering populations.
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four naive populations, and 29 from five recovering populations.
 These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
@@
 
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome [NCBI SRA: SRS6533475, reference = @hon2020] with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR Duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0, [@van2020]).
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Exact PCR duplicates were marked using Picard.
+Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
 Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
@@
 First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were single nucleotide polymorphisms (SNPs) and 73,866 were insertions or deletions (INDELS).
+This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
 Finally, we trimmed samples with an average depth across filtered sites \< 7x.
-Our final dataset included 50 samples (23 from naive and 27 from recovering populations) with an average depth of 16.7x (range = 7.4x -- 26.1x).
+Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
 
@@
 
-To visualize the genomic relationship of our populations we conducted a principal components analysis (PCA) using the glPCA function in the adegenet R package [@jombart2008].
+To visualize the genomic relationship of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
 To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA.
+For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
@@
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI).
+We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI: dataset S1, S2).
 
-For each outlier variant, we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
 To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using blast to get the predicted gene location based on the closest annotated reference [@altschul1997].
+Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
 We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
@@
 We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-We then used the WINGEN R package [@bishop2023] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
+We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
 This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for F~ST~ or above/below the mean for $\pi_{diff}$.
-Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.
+We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
+Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI: datasets S3, S4).
 
 Using the liberal set of variant outliers, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the Blast hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 29 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 30 <!--# Allie - Check for accuracy --> unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-Code used for genomic analyses and to figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-<!--# Make sure link is live before manuscript submission. -->
+Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
+<!--# Make sure that code have been uploaded to this link before manuscript submission. -->
 
@@
 
-For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], as well as landscape genetic information [@poorten2017].
+For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
 The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
@@
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap SI) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
-During each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
 Frogs were transported to the recipient site either on foot or via helicopter.
 Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
-CMR surveys allowed quantification of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
+CMR surveys allowed esimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
 During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
@@
 Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
-Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013; @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
-Both approaches utilized the loo package [@vehtari2022].
+We evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
 
@@
 Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), winter severity in the year following translocation (snow_t1), site elevation (elevation), donor population (donor_72996), sex (sex_male), frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first)).
-In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3\--16).
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), sex (sex_male), frog size (length), site elevation (elevation), winter severity in the year of translocation (snow_t), winter severity in the year following translocation (snow_t1), donor population (donor_72996), day of year on which a translocation was conducted (day), and translocation order (order_first)).
+In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
@@
 Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. --> <!--# Add  -->
+<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. -->
 
@@
 We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] : SI).
+We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs]).
 
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
+(Additional details regarding incorporating yearly variability in vital rates and estimating model parameters are provided in the SI.)
 
@@
 
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
-First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to growth or decline in the long-run.
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI).
+First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
 Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
-Third, we included demographic and environmental stochasticity into our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
+Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
+(Additional details are provided in the SI.) <!--# Add link to Mark's GitHub repo? -->
 
@@
 Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
-This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National Science Foundation Grants EF-0723563 (to C.
-Briggs), DEB-1557190 (to C.
-Briggs), and DBI-2120084 (to C.
+This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
+Briggs; DEB-1557190, to C.
+Briggs; and DBI-2120084, to C.
 Richards-Zawacki).
@@
 
-![Evidence for selection in recovering MYL frog populations at the landscape scale.
-For each of the 9 naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+For each of the 9 naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
@@
 ![Evidence for selection on genomic regions in recovering MYL frog populations.
-Results are from the splined window analysis and show outlier windows for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
+In (B), outlier regions are shown above the single dashed red line.
 Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from (A) showing two adjacent outlier windows for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier window that includes the outlier SNPs TCF19 and VARS.
+(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
 This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
@@
 Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
+Error bars show the 95% uncertainty intervals.
 Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-Error bars show the 95% uncertainty intervals.
 When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
@@
 
-![Predicted and observed growth and persistence of translocated populations.
-Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
+![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
 The dashed line shows where $\omega = 0.3$, for comparison to (A).
@@
 In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-Nonetheless, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6, 7, 8, and 9; @fig-allelefrequencies, @fig-selectionresults A SI).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A SI).
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 <!--# Allie, check text for accuracy -->
-
-<!--#  CUT?
-Multiple genes in our stringent set of outliers have known associations with amphibian Bd resistance/tolerance.
-Of particular interest is a ~940kb region on contig 8 that we identified as under selection, and that contains 8 genes associated with either the MHC Class I or Class III regions.
-The order of these genes in the MYL frog genome is conserved from Xenopus [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
-Numerous studies have linked MHC genes to frog resistance/tolerance against Bd infection [e.g., @savage2011; @bataille2015].
-In addition to the genes in this region that we describe in the Results section of the main paper, other genes also deserve mention.
-For example, we identified the MHC Class III gene, heat shock protein 70 (HSP70), as occurring in this region (@fig-synteny-plot).
-Similarly, [@byrne2020] identified an HSP70-binding protein as a candidate gene under selection in Atelopus varius populations recovering after Bd-caused declines.
-HSP70 was also highly expressed in Bd-infected versus uninfected Xenopus tropicalis [@rosenblum2009].
-This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3 (@fig-spline-manhattan C, @fig-allelefrequencies A).
-Additionally, we identified a region on contig 19 that had the highest $\pi_{diff}$ of any outlier window, indicative of directional selection.
-This region contained the gene, interferon-induced very large GTPase 1-like (GVINP1).
-Interferons boost immune system processes and previous studies of frogs have identified interferon-related genes as associated with Bd resistance/tolerance [@byrne2020] or have shown increased expression of this class of genes when frogs were exposed to Bd in the lab [@rosenblum2009; @ellison2014].
-Variants in GVINP1 also explained approximately 20% of the resistance to amoebic gill disease in Atlantic salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
-Furthermore, in Litoria verreauxii alpina, the related gene, IFN-induced very large GTPase, showed lower expression in frogs from less susceptible populations versus more susceptible populations [@grogan2018].Finally, we identified a region characterized by high FST and low $\pi_{diff}$ and that contained the complement genes C6 and C7.
-This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
--->
 
@@
 
-![Evidence for selection of individual variants in recovering MYL frog populations.
+![Evidence for selection on individual variants in recovering MYL frog populations.
 (A) PCA calculated from binary SNPs showing the genomic relationship of samples.
 Numeric labels and colors match those in @fig-allelefrequencies A.
-(B) qqplot showing observed and expected p-values for 148,307 SNPs and INDELS as calculated in GEMMA.
+Populations 1-7 are *R.
+sierrae* and populations 8 and 9 are *R.
+muscosa*.
+(B) qqplot showing observed and expected p-values for 148,307 SNPs and INDELS.
 Dashed line shows p-value that identifies outliers.
-(C) Manhattan plot showing p-values for each SNP as calculated by GEMMA.
+(C) Manhattan plot showing p-value for each SNP.
 SNPs are sorted by genomic position and contigs are sorted by size.
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate at least one non-synonymous SNP is present in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
 
-<!--# Remove x = 0 and y = 0 lines in PC plot -->
-
-![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
+![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical <!--# Elsewhere we refer only to "MHC Class I".
+Is "Classical" necessary here?
+--> and MHC Class III gene regions.
 The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
-Asterices indicate the location of SNP outliers in TCF19 and VARS1 genes.
+Asterices indicate the location of SNP outliers in the TCF19 and VARS1 genes.
 Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot}
@@
 In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-<!--# Make sure all points are inside YOSE polygon -->](figures/map_translocation_points.png){#fig-yosemap width="250"}
+](figures/map_translocation_points.png){#fig-yosemap width="250"}
 
@@
 Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale; see SI text for details).
-Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter}
+Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
 
-![Evidence that Bd load in not an important predictor of post-translocation frog survival.
+![Evidence that Bd load is not an important predictor of post-translocation frog survival.
 For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="450"}
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="350"}
 
 ![Evidence that frog resistance/tolerance in non-natal habitats is maintained across generations.
-Comparison of the average yearly adult survival probabilities for adults translocated to each of three sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohort.
+Comparison of the average yearly adult survival probabilities for adults translocated to each of 3 sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
@@
 
-\
+\newpage
 
@@
 Description of overlapping *F~ST~* and $\pi_{diff}$ outlier windows, as identified in the splined window analysis.
-file: spline_window_shared_outliers.csv <!--# The dataset contains descriptions of 10 shared regions, not the 9 regions mentioned in the Results.
-Do we need to explain that one region is a repeat because there is a 2>1 match for one of the overlaps (i.e., one Fst outlier window overlaps with 2 pi outlier windows)?
--->
+The outlier window on chromosome 19 (column name = chr_num_sorted) is included twice because one *F~ST~* outlier window overlapped with two $\pi_{diff}$ outlier windows.
+file: spline_window_shared_outliers.csv <!--# Allie - Check for accuracy  -->
 
@@
 ### References
-
-\

### 2023-05-15 doc/manuscript/PNAS-template-supp-info-empty.tex

@@
 
-\dataset{gemma\_outliers.csv}{Description needed}
+\dataset{gemma\_outliers\_all.csv}{Set of liberal SNP and INDEL outlier variants (Bonferroni corrected p-value $<$ 0.05), identified via GEMMA.}
 
-\dataset{spline\_window\_shared\_outliers.csv}{Description needed}
+\dataset{gemma\_outliers\_strict\_freq.csv}{Set of strict SNP and INDEL outlier variants (Bonferroni corrected p-value $<$ 0.01), identified via GEMMA.
+Additional information includes variant location within the gene (predicted\_gene\_loc) and whether the variant is synonymous or nonsynonymous (predicted\_effect\_AA).}
+
+\dataset{spline\_window\_shared\_outliers.csv}{Description of overlapping \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) outlier windows, as identified in the splined window
+analysis.}
+
+\dataset{spline\_window\_gene\_details.csv}{Annotation information for all genes within each of the
+overlapping outlier windows in dataset S3}
 

### 2023-05-15 doc/manuscript/translocation_pnas_SI.tex

@@
 0.01). The liberal set, identified using a Bonferroni-corrected p-value
-of 0.05, included 38 outliers (35 SNPs and 3 structural variants) from
-30 distinct genes across 16 contigs.
+of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
+genes across 16 contigs. Our analysis found no significantly
+over-represented GO terms in this set of 30 genes, or in the set of 38
+outlier regions identified with the splined window analysis.
 
-Multiple genes in our stringent set of outliers have known associations
-with amphibian Bd resistance/tolerance. Of particular interest is a
-\textasciitilde940kb region on contig 8 that we identified as under
-selection, and that contains 8 genes associated with either the MHC
-Class I or Class III regions. The order of these genes in the MYL frog
-genome is conserved from \emph{Xenopus} \citep{ohta2006} and is
-consistent with MHC Class I/III linkage reported in the Komodo dragon
-\citep{reed2021}. Numerous studies have linked MHC genes to frog
-resistance/tolerance against Bd infection
-\citep[e.g.,][]{savage2011, bataille2015}. In addition to the genes in
-this region that we describe in the Results section of the main paper,
-other genes also deserve mention. For example, we identified the MHC
-Class III gene, heat shock protein 70 (HSP70), as occurring in this
-region (Figure~\ref{fig-synteny-plot}). Similarly, \citep{byrne2020}
-identified an HSP70-binding protein as a candidate gene under selection
-in \emph{Atelopus varius} populations recovering after Bd-caused
-declines. HSP70 was also highly expressed in Bd-infected versus
-uninfected \emph{Xenopus tropicalis} \citep{rosenblum2009}. This MHC
-gene cluster lies on the same chromosome as the outlier SNP in RIN3
-(Fig. 2 C, Fig. 1
-A).
-
-Additionally, we identified a region on contig 19 that had the highest
-\(\pi_{diff}\) of any outlier window, indicative of directional
-selection. This region contained the gene, interferon-induced very large
-GTPase 1-like (GVINP1). Interferons boost immune system processes and
-previous studies of frogs have identified interferon-related genes as
-associated with Bd resistance/tolerance \citep{byrne2020} or have shown
-increased expression of this class of genes when frogs were exposed to
-Bd in the lab \citep{rosenblum2009, ellison2014}. Variants in GVINP1
-also explained approximately 20\% of the resistance to amoebic gill
-disease in Atlantic salmon, a finding confirmed via expression analysis
-\citep{robledo2018, robledo2020}. Furthermore, in \emph{Litoria
-verreauxii alpina}, the related gene, IFN-induced very large GTPase,
-showed lower expression in frogs from less susceptible populations
-versus more susceptible populations \citep{grogan2018}.
-
-Finally, we identified a region characterized by high
-\emph{F\textsubscript{ST}} and low \(\pi_{diff}\) and that contained the
-complement genes C6 and C7. This could indicate that balancing selection
-is acting in this region of the genome to favor a diverse set of
-alleles, as is known for C6 in humans \citep{soejima2005}.
+In the comparison of individual outlier variants between naive and
+recovering populations, we detected consistent allele frequency changes
+across all genetic clusters for 1 of 7 of the associated genes (RIN3;
+Fig. 1). Nonetheless, for the other 6 genes
+we still found a statistically significant signal of selection in 2 of
+the 3 genetic clusters (containing populations 6, 7, 8, and 9;
+Fig. 1, Figure~\ref{fig-selectionresults} A
+SI). Therefore, although some outlier variant associations have a more
+limited geographic extent than RIN3, they still describe parallel
+evolutionary changes following Bd exposure and are therefore interesting
+candidates for future studies.
 
@@
 whether our model could qualitatively describe the patterns of recovery
-in the observed data for population 70550.
+in the observed data for population 70550. Code to replicate the
+analyses can be found at \url{https://github.com/SNARL1/translocation}.
 
@@
 
-\hfill\break
-
 \begin{figure}
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/pca_qq_manhattan.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/pca_qq_manhattan.png}
 
@@
 
-\caption{\label{fig-selectionresults}Results from the exome capture
-selection study. (A) PCA calculated from binary SNPs showing genomic
-relationship of samples. Numeric labels and colors match those in
+\caption{\label{fig-selectionresults}Evidence for selection of
+individual variants in recovering MYL frog populations. (A) PCA
+calculated from binary SNPs showing the genomic relationship of samples.
+Numeric labels and colors match those in
 Fig. 1 A. (B) qqplot showing observed and
-expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
+expected p-values for 148,307 SNPs and INDELS as calculated in GEMMA.
 Dashed line shows p-value that identifies outliers. (C) Manhattan plot
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/synteny_figure.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/synteny_figure.png}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/map_translocation_points.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/map_translocation_points.png}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/mcmc_areas_m2b.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/mcmc_areas_m2b.png}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/bdload_beforeafter.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/bdload_beforeafter.png}
 
@@
 
-\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
-the 12 recipient sites, Bd loads for the period immediately prior to
-translocation versus during the 1-year period after translocation. Bd
-loads are expressed as the number of ITS1 copies per skin swab, as
-estimated by qPCR of the Bd ITS1 region. Loads indicative of severe
-disease are \textgreater{} 5.8 ITS copies (on a log\textsubscript{10}
-scale; see SI text for details). Box plot details are as in
-Fig. 4.}
+\caption{\label{fig-bdload-beforeafter}Evidence that frogs translocated
+from recovering populations can maintain the benefits of
+resistance/tolerance in non-natal habitats. For frogs translocated to
+each of the 12 recipient sites, Bd loads for the period immediately
+prior to translocation versus during the 1-year period after
+translocation. Bd loads are expressed as the number of ITS1 copies per
+skin swab, as estimated by qPCR of the Bd ITS1 region. Loads indicative
+of severe disease are \textgreater{} 5.8 ITS copies (on a
+log\textsubscript{10} scale; see SI text for details). Box plot details
+are as in Fig. 4.}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/mcmc_areas_m1d.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/mcmc_areas_m1d.png}
 
@@
 
-\caption{\label{fig-survival-postdens}For the best model identifying
-predictors of 1-year frog survival following translocation, estimated
-posterior density curves and shaded 95\% uncertainty intervals for the
-intercept and all predictor variables. In the Bayesian framework in
-which the model was developed, variables are considered important
-predictors if the associated uncertainty interval does not overlap zero
-(indicated by the dashed red line).}
+\caption{\label{fig-survival-postdens}Evidence that Bd load in not an
+important predictor of post-translocation frog survival. For the best
+model identifying predictors of 1-year frog survival following
+translocation, estimated posterior density curves and shaded 95\%
+uncertainty intervals for the intercept and all predictor variables. In
+the Bayesian framework in which the model was developed, variables are
+considered important predictors if the associated uncertainty interval
+does not overlap zero (indicated by the dashed red line).}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/compare_surv_probs.jpg}
+{\centering \includegraphics[width=0.85\textwidth]{figures/compare_surv_probs.jpg}
 
@@
 
-\caption{\label{fig-compare_surv_probs}Comparison of the average yearly
-adult survival probabilities for adults translocated to each of three
-sites versus adults that were naturally recruited at each site (as a
-result of reproduction by translocated frogs). In contrast to
+\caption{\label{fig-compare_surv_probs}Evidence that frog
+resistance/tolerance in non-natal habitats is maintained across
+generations. Comparison of the average yearly adult survival
+probabilities for adults translocated to each of three sites versus
+adults that were naturally recruited at each site (as a result of
+reproduction by translocated frogs). In contrast to
 Fig. 3, these are not survival
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_supp.jpg}
+{\centering \includegraphics[width=0.85\textwidth]{figures/pop_viability_figures_for_supp.jpg}
 

### 2023-05-15 doc/manuscript/translocation.qmd
Additional quarto autoformats

@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).
+Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: <https://github.com/SNARL1/cmr-analysis>.
 
@@
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-(Additional details are provided in the SI.) <!--# Add link to Mark's GitHub repo? -->
+<!--# Add link to Mark's GitHub repo? -->
 
@@
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
 
@@
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).
-](figures/map_translocation_points.png){#fig-yosemap width="250"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="250"}
 

### 2023-05-17 code/maps.qmd
Increase symbol and text size

--- 2023-03-21 code/maps.qmd
@@
 library(grid)
-library(readr)
+library(tidyverse)
+library(RPostgreSQL)
 ```
 
-## Map of exome capture sample sites
+## Map of sample sites for exome capture analysis
 
@@
   tm_shape(translocation_points) + 
-  tm_symbols(size = 0.7, 
+  tm_symbols(size = 0.9, 
              shape = "donor_id",
@@
             legend.bg.color = "white",
-            legend.title.size = 1.3,
-            legend.text.size = 1.0) +
+            legend.title.size = 1.5,
+            legend.text.size = 1.2) +
   tm_compass(type = "4star",
              position = c(0.80, 0.11),
-             size = 3) +
+             size = 5) +
   tm_scale_bar(breaks = c(0, 10, 20), 
-               text.size = 1.0, 
+               text.size = 1.2, 
                position = c("right", "bottom"))

### 2023-05-17 code/translocation_survival_analysis_supp.Rmd
Edit axis labels

@@
     geom_boxplot() + 
-    labs(x = "Period", y = "Bd load (log10(bd_load + 1))") +
+    labs(x = "Period (relative to translocation)", y = (expression(Bd~load~(log["10"]~copies+1)))) +
     facet_grid(. ~ site_id1) +

### 2023-05-17 doc/manuscript/translocation.qmd
Minor edits in response to TS & AB review

@@
 
-<!--# Intended outlet is PNAS.
-Preferred length of research reports is 6 pages, but up to 12 pages is allowed.
-A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables) -->
-
 <!--# render to pdf using `quarto render doc/manuscript/translocation.qmd --to pdf` -->
@@
 
-<!--# Title should be less than 135 characters -->
-
 ### Authors:
 
-Roland A. Knapp^1,2,^\*, Mark Q. Wilber^3,a^, Allison Q. Byrne^4,5,a^, Maxwell B. Joseph^6,7^, Thomas C. Smith^1,2^, Andrew P. Rothstein^4,5^, Rob L. Grasso^8^, Erica Bree Rosenblum^4,5^
+Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e^, Rob L. Grasso^h^, Erica Bree Rosenblum^d,e^
 
@@
 
-^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
+^a^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
 
-^2^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
+^b^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
 
-^3^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
+^c^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
 
-^4^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
+^d^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
 
-^5^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
+^e^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
 
-^6^Earth Lab, University of Colorado, Boulder, CO, 80303
+^f^Earth Lab, University of Colorado, Boulder, CO, 80303
 
-^7^Planet, San Francisco, CA, 94107
+^g^Current affiliation: Planet, San Francisco, CA, 94107
 
-^8^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
+^h^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
 
@@
 
-^a^These authors contributed equally to this work.
+^1^These authors contributed equally to this work.
 
-**Preprint Server:** bioRxiv
+<!--#  Preprint Server: bioRxiv  -->
 
 **Classification:** Biological Sciences - Ecology
+
+**Key words:** evolutionary rescue, amphibians, population recovery, reintroduction, disease, *Batrachochytrium dendrobatidis*
 
@@
 One possibility is rapid evolution that restores positive growth to declining populations ("evolutionary rescue").
-Individuals from these rescued populations could also be used to reestablish exinct populations.
+Individuals from these rescued populations could also be used to reestablish extinct populations.
 We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse disease impacts.
@@
 Human activities are intensifying threats to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
-To persist, affected taxa will often need to adapt to their altered environment or face decline and extinction.
-"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth [@carlson2014].
-In addition to preventing extinction of declining populations, this process could also result in the seeding of nearby empty habitats and allow population reestablishment and recovery.
-Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of threatened taxa, descriptions of its operation in the wild are lacking.
+To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
+The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
+In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
+Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of threatened taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
 
 One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
-Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amid this global loss of biodiverity, some populations decimated by disease have subsequently recovered to varying extents [@newell2013; @voyles2018; @knapp2016].
+Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
+Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@newell2013; @voyles2018; @knapp2016].
 Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
@@
 In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
 We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
@@
 Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].
-(See SI for additional details regarding the study design.)
+These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
+(See SI Appendix for additional details regarding the study design.)
 
@@
 The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
-One of the 7 identified genes did not have an associated annotation (LOC108802036<!--# Allie - check for accuracy -->).
+One of the 7 identified genes (LOC108802036) did not have an associated annotation.
 For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
@@
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-(See SI for additional description of these results.)
+(See SI Appendix for additional description of these results.)
 
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more outlier SNPs (described above).
-A total of 33 annotated genes were found in the 9 shared regions (SI).
+A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
 The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
@@
 
-Estimates of frog survival during the 1 year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
 As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
 This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
 sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter, see SI text for additional details) [@joseph2018; @vredenburg2010].
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter, see SI Appendix text for additional details) [@joseph2018; @vredenburg2010].
 
@@
 Here, we tested the possibility that populations of resistant/tolerant individuals could be used to reestablish extirpated populations.
-Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
+Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild (see [@joseph2018] for a small-scale example).
 In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
@@
 These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the gene GVINP1.
+Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
 Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018].
@@
 This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
+Finally, based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
 This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
+Also in this analysis, the outlier variant with the lowest p-value (uncharacterized gene LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2) in another frog genome [@gan2018], and together with GVINP1 further indicates the importance of interferon-related genes in this system.
 
@@
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI: dataset S1, S2).
+We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
 
@@
 Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI: datasets S3, S4).
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
 
@@
 First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 <!--# Allie - Check for accuracy --> unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
 Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-<!--# Make sure that code have been uploaded to this link before manuscript submission. -->
+<!--# Make sure that code has been uploaded to this link before manuscript submission. -->
 
@@
 Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
-CMR surveys allowed esimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
+CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
 During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI for details].
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-(Additional details regarding incorporating yearly variability in vital rates and estimating model parameters are provided in the SI.)
+(Additional details regarding incorporating yearly variability in vital rates and estimating model parameters are provided in the SI Appendix.)
 
@@
 
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI).
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI Appendix).
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
@@
 Briggs; DEB-1557190, to C.
-Briggs; and DBI-2120084, to C.
+Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
 Richards-Zawacki).
@@
 ![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
-For each of the 9 naive and recovering MYL frog populations, pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs in this portion of the Sierra Nevada shown in light gray.
-The inset map locates the study area in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
+The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
@@
 In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A SI).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
-<!--# Allie, check text for accuracy -->
 
@@
 
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values : SI.
+The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
 Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs] : SI).
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
 For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
@@
 
-![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical <!--# Elsewhere we refer only to "MHC Class I".
-Is "Classical" necessary here?
---> and MHC Class III gene regions.
+![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
 The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
@@
 Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale; see SI text for details).
+Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale; see SI Appendix text for details).
 Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
@@
 The outlier window on chromosome 19 (column name = chr_num_sorted) is included twice because one *F~ST~* outlier window overlapped with two $\pi_{diff}$ outlier windows.
-file: spline_window_shared_outliers.csv <!--# Allie - Check for accuracy  -->
+file: spline_window_shared_outliers.csv
 

### 2023-05-18 code/translocation_survival_analysis.Rmd
Remove unnecessary comments

@@
   select(-plot_survival) %>%
-  ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor)) +  # reorder site_id by siteid_order)
+  ggplot(aes(x = reorder(site_id, siteid_order), y = median_survival, color = donor)) +  # reorder site_id by siteid_order
       geom_linerange(aes(ymin = lo_survival, ymax = hi_survival), color = "black", width = 0.5, position = position_dodge2(width=0.5)) +
@@
 -   Strong effect of site/translocation on frog survival.
--   Survival is highly variable across sites, but relatively invariant within a site.
+-   Survival is highly variable across sites, but relatively constant within a site.
 -   Indicates need to include site_id and/or translocation_id as group-level effect(s).
@@
 
-# Confirm our prior is flat
+# Confirm that prior is flat
 prior_summary(m1_flat)
@@
 predictions_m1d <- posterior_epred(m1d_reduce, newdata = m1d_group_means)
-# predictions_m1d <- posterior_predict(m1d_reduce, newdata = m1d_group_means) # original code that produced incorrect uncertainty estimates
 

### 2023-05-18 doc/manuscript/translocation.qmd
Minor edits in response to coauthor RG comments

@@
 
-Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e^, Rob L. Grasso^h^, Erica Bree Rosenblum^d,e^
+Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e,h^, Robert L. Grasso^i^, Erica Bree Rosenblum^d,e^
 
@@
 
-^h^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
+^h^Current affiliation: Ginkgo Bioworks, Boston, MA, 02210
+
+^i^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
 
@@
 ^1^These authors contributed equally to this work.
-
-<!--#  Preprint Server: bioRxiv  -->
 
@@
 Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
-In addition to these important but relatively subtle effects of Bd on adults, MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and experience high mortality.
-This high susceptibility of metamorphs is documented for numerous species of anurans, and may result from the poorly developed immune system characteristic of this early life stage [@humphries2022].
+In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
+MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
+This high susceptibility of metamorphs is documented for numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
 In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
@@
 VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
-These *R. sierrae* life stages have high detectablility, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
+These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
 

### 2023-05-18 code/translocation_survival_analysis.Rmd
Minor edit to remove deprecated code

@@
   group_by(site_id) %>% 
-  mutate(survival_trans1 = mean(survival)) %>%    # use mean or median?
+  mutate(survival_trans1 = mean(survival)) %>%
   ungroup(site_id) %>% 
   distinct(site_id, survival_trans1) %>% 
-  mutate(across(c(survival_trans1), round, 3)) %>% 
+  mutate(survival_trans1 = round(survival_trans1, 3)) %>% 
   inner_join(d2, by = c("site_id")) %>% 

### 2023-05-18 code/translocation_survival_analysis_supp.Rmd
Remove unnecessary comments

@@
           "Posterior distributions with medians & 90% uncertainty intervals")
-
-# mean(bayes_R2(m1a))   # bayes_R2 not available for negative binomials models in stanreg. 
 ```
@@
           "Posterior distributions with medians & 90% uncertainty intervals")
-
-# mean(bayes_R2(m1b))
 ```
@@
           "Posterior distributions with medians & 90% uncertainty intervals")
-
-# mean(bayes_R2(m1c))
 ```
@@
           "Posterior distributions with medians & 90% uncertainty intervals")
-
-# mean(bayes_R2(m1d))
 ```

### 2023-05-18 code/translocation_survival_createdata.Rmd
Minor edits and remove unnecessary comments

@@
 translocation_sites %>% write_csv(here::here("data", "raw", "translocation_sites.csv"))
-# translocation_sites <- read_csv(here::here("data", "raw", "translocation_sites.csv"))
 ```
@@
 frog_characteristics %>% write_csv(here::here("data", "raw", "frog_characteristics.csv"))
-# frog_characteristics <- read_csv(here::here("data", "raw", "frog_characteristics.csv"))
 ```
@@
 frog_captures %>% write_csv(here::here("data", "raw", "frog_captures.csv"))
-# frog_captures <- read_csv(here::here("data", "raw", "frog_captures.csv"))
 ```
@@
 frog_translocation %>% write_csv(here::here("data", "raw", "frog_translocation.csv"))
-# frog_translocation <- read_csv(here::here("data", "raw", "frog_translocation.csv"))
 ```
@@
 frog_survival %>% write_csv(here::here("data", "raw", "frog_survival.csv"))
-# frog_survival <- read_csv(here::here("data", "raw", "frog_survival.csv"))
 ```
@@
 ```{r check-sex}
-frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1) # make sure that only one sex recorded per tag at recipient site
+frog_captures %>% distinct(pit_tag_ref, sex) %>% count(pit_tag_ref, sex) %>% filter(n > 1) # ensure only one sex per tag at recipient site
 
@@
 frog_translocation %>% write_csv(here::here("data", "clean", "frog_translocation_final.csv"))
-# frog_translocation <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
 ```

### 2023-05-18 code/translocation_survival_createdata_supp.Rmd
Minor edits and remove unnecessary comments

@@
   inner_join(bd_add, by = c("visit_date", "pit_tag_ref")) %>% 
-  mutate(period = if_else(survey_type == "cmr", "after", "before")) %>% # use 0/1 instead?
+  mutate(period = if_else(survey_type == "cmr", "after", "before")) %>% 
   relocate(period, .after = survey_type) %>% 

### 2023-05-18 code/translocation_survival_createdata.Rmd
Minor edits to comments

@@
   select(-release_date_min) %>% 
-  filter(visit_year > release_year) # Exclude surveys/counts made in the year of translocation (before presence of tadpoles is possible)
+  filter(visit_year > release_year) # Exclude surveys/counts made in the year of first translocation (before presence of tadpoles is possible)
 
@@
   select(-release_date_min) %>% 
-  filter(visit_year > (release_year + 2)) # Exclude counts made in first 3 years after translocation (before recruitment is possible)
+  filter(visit_year > (release_year + 2)) # Exclude counts made in first 3 years after initial translocation (before recruitment is possible)
 

### 2023-05-18 doc/manuscript/translocation.qmd
Add description of reproductive success calcs

@@
 
+##### Frog counts and reproductive success
+
+For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and captures of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
+To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
+This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
+Similarly, to calculcate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
+This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).
+
 ##### Estimation of frog survival and abundance

### 2023-05-19 doc/manuscript/translocation_bioarxiv.tex
Adjust manuscript formatting

--- 2023-05-13 doc/manuscript/translocation_pnas.tex
@@
 
-\templatetype{pnasresearcharticle} % Choose template 
-% {pnasresearcharticle} = Template for a two-column research article
+\templatetype{pnasmathematics} % Choose template 
+% {pnasmathematics} = Template for a two-column research article
 % {pnasmathematics} %= Template for a one-column mathematics article
@@
 (``evolutionary rescue''). Individuals from these rescued populations
-could also be used to reestablish exinct populations. We evaluated both
+could also be used to reestablish extinct populations. We evaluated both
 possibilities in a once-common frog species driven to the brink of
@@
 \dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+% \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
@@
 taxa will often need to adapt to their altered environment or face
-decline and extinction. ``Evolutionary rescue'' describes rapid
-evolutionary change that increases the frequency of adaptive alleles and
-thereby restores positive population growth \citep{carlson2014}. In
-addition to preventing extinction of declining populations, this process
-could also result in the seeding of nearby empty habitats and allow
-population reestablishment and recovery. Although this overall scenario
-is potentially of critical importance for the conservation of an
-ever-growing number of threatened taxa, descriptions of its operation in
-the wild are lacking.
+decline and possible extinction. The process whereby rapid evolutionary
+change increases the frequency of adaptive alleles and restores positive
+population growth is known as ``evolutionary rescue''
+\citep{carlson2014}. In addition to preventing extinction of declining
+populations, individuals from rescued populations could also disperse,
+or be reintroduced, to empty habitats and promote larger-scale recovery.
+Although this overall scenario is potentially of critical importance for
+the conservation of an ever-growing number of threatened taxa,
+descriptions of its occurrence and effectiveness in the wild are
+lacking.
 
@@
 \citep{daszak2000}, and have caused dramatic declines and extinctions in
-a wide range of taxa
-\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
-global loss of biodiverity, some populations decimated by disease have
-subsequently recovered to varying extents
+a wide range of taxa, including echinoderms, mammals, birds, and
+amphibians \citep{hewson2014, samuel2015, scheele2019, cunningham2021}.
+Amid this global loss of biodiversity, some populations decimated by
+disease have subsequently recovered to varying extents
 \citep{newell2013, voyles2018, knapp2016}. Although we know relatively
@@
 show that MYL frogs have undergone substantial evolutionary change
-associated with Bd exposure and that might provide a genetic basis for
-the increased resistance/tolerance to Bd infection observed in naturally
-recovering populations \citep{knapp2016}. We then use the reintroduction
-of resistant/tolerant frogs to 24 sites and results from long-term
-capture-mark-recapture (CMR) surveys to describe the dynamics and
-drivers of population establishment. These data are also used to
+associated with Bd exposure, changes that might provide a genetic basis
+for the increased resistance/tolerance to Bd infection observed in
+naturally recovering populations \citep{knapp2016}. We then use the
+reintroduction of resistant/tolerant frogs to 24 sites and results from
+long-term capture-mark-recapture (CMR) surveys to describe the dynamics
+and drivers of population establishment. These data are also used to
 parameterize a population model that extends our inferences of
@@
 an enzootic state \citep{briggs2010}, characterized by high Bd
-prevalence and, in adults, moderate Bd infection intensities (``load'')
-that are typically well below the level expected to cause mortality
-\citep{vredenburg2010}.
+prevalence and, in adults, moderate Bd infection intensity (``load'')
+that is typically well below the level expected to cause mortality
+\citep{vredenburg2010}. (See SI Appendix for additional details
+regarding the study design.)
 
@@
 space (Figure S1A), and clusters reflected the
+species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and the
 strong signature of isolation by distance that is characteristic of MYL
@@
 allele frequencies between naive versus recovering populations) from 7
-distinct genes across 4 contigs (Figure S1B,C). One of the 7 identified genes did not have an associated
-annotation. Frequency differences between the naive and recovering
-populations ranged from 41\% to 86\%. Most of these SNPs showed
-frequency differences in only a subset of the sampled populations
-(Figure~\ref{fig-allelefrequencies} A, B), but the SNP in the RIN3 gene
-showed consistent differences in frequencies across all populations
-(Figure~\ref{fig-allelefrequencies} C). This is suggestive of strong -
-and parallel - selection at this locus associated with exposure of frogs
-to Bd across multiple populations.
+distinct genes across 4 contigs (Figure S1B,C). One of the 7 identified genes (LOC108802036) did not have an
+associated annotation. For the outlier SNPs, frequency differences
+between the naive and recovering populations ranged from 41\% to 86\%.
+Most of these SNPs showed frequency differences in only a subset of the
+sampled populations (Figure~\ref{fig-allelefrequencies} A, B), but the
+SNP in the RIN3 gene showed consistent differences in frequencies across
+all populations (Figure~\ref{fig-allelefrequencies} C). This is
+suggestive of strong - and parallel - selection at this locus associated
+with exposure of frogs to Bd across multiple populations. (See SI
+Appendix for additional description of these results.)
 
@@
 (Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
-outliers for both metrics and 2 of these shared regions also contained
-outlier SNPs. A total of 33 annotated genes were found in the 9 shared
-regions (SI). The largest \(\pi_{diff}\), indicative of directional
-selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
-the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C). This region
-contains approximately 500 SNPs and one annotated gene called
-``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
-a shared outlier region on Contig1 contained two complement factor genes
-(C6 and C7). Interestingly, this region had a low \(\pi_{diff}\) value,
-consistent with balancing selection. Finally, one shared outlier region
-on Contig8 contained one outlier SNP (TCF19) and was within 360kb of
-another outlier SNP (VARS) (Figure~\ref{fig-spline-manhattan} D,
-Figure S2). This region (854kb from the beginning of
-the outlier window to the VARS SNP) contained a total of 8 annotated
-genes. In \emph{Xenopus}, five of these genes occur in the extended
-major histocompatibility complex (MHC) Class I region (FLOT1, TUBB,
-MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III
-region (HSP70, LSM2, VARS) \citep{ohta2006}. Therefore, this region
-under selection is part of the extended MHC Class I and III complex and
-shows synteny with other amphibian genomes.
+outliers for both metrics (``shared regions'') and 2 of these shared
+regions also contained one or more outlier SNPs (described above). A
+total of 33 annotated genes were found in the 9 shared regions (SI
+Appendix: Datasets S3, S4). The largest \(\pi_{diff}\), indicative of
+directional selection, occurred in a 163kb region on Contig19, 12.9Mb
+upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C).
+This region contains approximately 500 SNPs and one annotated gene
+called ``interferon-induced very large GTPase 1-like'' (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two
+complement factor genes (C6 and C7). Interestingly, this region had a
+large negative \(\pi_{diff}\) value, consistent with balancing
+selection. Finally, one shared outlier region on Contig8 contained one
+outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS)
+(Figure~\ref{fig-spline-manhattan} D, Figure S2).
+This region (854kb from the beginning of the outlier window to the VARS
+SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
+these genes occur in the extended major histocompatibility complex (MHC)
+Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
+extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
+Therefore, this region under selection is part of the extended MHC Class
+I and III complex and shows synteny with other amphibian genomes.
 
@@
 
-To determine whether \emph{R. sierrae} from recovering populations could
-be used to reestablish extirpated populations, we conducted 24
-reintroductions in Yosemite National Park (2006--2020). Each of the
-reintroductions involved collection of adult frogs from 1 of 3
-recovering, Bd-positive ``donor'' populations (Figure S3) and translocating them to 1 of 12 nearby recipient sites
-\citep{seddon2014}. The donor populations included 2 of the 5 recovering
-populations in the frog evolution study described above
-(Figure~\ref{fig-allelefrequencies} A: population 1 and 4). Following
-translocation, adult survival and recruitment of new adults were
-estimated from CMR surveys and counts of tadpoles and juveniles were
-obtained from visual encounter surveys (VES).
-
-Estimates of frog survival during the 1 year following translocation
-indicate that survival was highly variable between recipient sites, but
-relatively constant within recipient sites (for the subset of sites that
-received multiple translocations;
+To determine whether frogs from recovering populations could be used to
+reestablish extirpated populations, we conducted 24 reintroductions in
+Yosemite National Park (2006--2020). Each of the reintroductions
+involved collection of adult frogs from 1 of 3 recovering, Bd-positive
+``donor'' populations and translocating them to 1 of 12 nearby recipient
+sites (Figure S3) \citep{seddon2014}. The donor
+populations included 2 of the 5 recovering populations in the frog
+evolution study described above (Figure~\ref{fig-allelefrequencies} A:
+population 1 and 4). Following translocation, adult survival and
+recruitment of new adults were estimated from CMR surveys and counts of
+tadpoles and juveniles were obtained from visual encounter surveys
+(VES).
+
+Estimates of frog survival during the 1-year period following
+translocation indicate that survival was highly variable between
+recipient sites, but relatively constant within recipient sites (for the
+subset of sites that received multiple translocations;
 Figure~\ref{fig-translocation-survival}). As such, the average survival
@@
 (Table S1), which we suggest resulted
-primarily from our improved ability over time to choose recipient sites
-with higher habitat quality for \emph{R. sierrae}. Finally, Bd loads
-were fairly consistent before versus after translocation and loads were
+primarily from our improved ability to choose recipient sites with
+higher habitat quality for \emph{R. sierrae}. Finally, Bd loads were
+fairly consistent before versus after translocation and loads were
 nearly always well below the level indicative of severe chytridiomycosis
-and associated frog mortality (Figure S5,
-see SI text for additional details) \citep{joseph2018, vredenburg2010}.
+and associated frog mortality (Figure S5, see
+SI Appendix text for additional details)
+\citep{joseph2018, vredenburg2010}.
 
@@
 
-Even when incorporating demographic stochasticity and environmental
-stochasticity in the probability of juvenile recruitment to the adult
-class \(\omega\) (the transition that we expect to be the most subject
-to environmental variability in the presence of Bd), populations with
-high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult
-individuals into a population and using the baseline value for
-\(\omega\) of 0.3, the six populations with the highest adult survival
-probabilities had 50-year extinction probabilities of 0 to 0.32;
-Figure~\ref{fig-viability} B). This indicates strong potential for
-long-term persistence in the presence of Bd and environmental
-variability in recruitment (Figure~\ref{fig-viability} B). When we
-decreased \(\omega\) from 0.3 to a conservative estimate of 0.15,
-50-year extinction probability remained less than 0.5 for three
-populations and was 0.79 to 0.94 for three additional populations
+Even when incorporating (i) demographic stochasticity and (ii)
+environmental stochasticity in the probability of juvenile recruitment
+to the adult class \(\omega\) (the transition that we expect to be the
+most subject to environmental variability in the presence of Bd),
+populations with high adult survival are likely to persist over a 50
+year time horizon. Our model predicted that, following a single
+introduction of 40 adult individuals into a population and using the
+baseline value for \(\omega\) of 0.3, the six populations with the
+highest adult survival probabilities had 50-year extinction
+probabilities of 0 to 0.32 (Figure~\ref{fig-viability} B). This
+indicates strong potential for long-term persistence in the presence of
+Bd and environmental variability in recruitment. When we decreased
+\(\omega\) from 0.3 to a conservative estimate of 0.15, 50-year
+extinction probability remained less than 0.5 for three populations and
+was 0.79 to 0.94 for three additional populations
 (Figure~\ref{fig-viability} B). In contrast, for the six populations
@@
 \citep{daszak2000}, but strategies to recover affected species are
-generally lacking. Here, we used populations of resistant/tolerant
-individuals to reestablish extirpated populations, despite ongoing
-disease. Although previously proposed \citep{mendelson2019}, this
-strategy has rarely been successfully implemented (see
-\citep{joseph2018} for a small-scale example). In particular, although
-numerous studies have documented signatures of selection in hosts
-following disease-induced declines
+generally lacking. Here, we tested the possibility that populations of
+resistant/tolerant individuals could be used to reestablish extirpated
+populations. Although previously proposed
+\citep{mendelson2019, scheele2021}, this strategy has rarely been used
+in the wild (see \citep{joseph2018} for a small-scale example). In
+particular, although numerous studies have documented signatures of
+selection in hosts following disease-induced declines
 \citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
@@
 reestablish extirpated populations. The results of our study provide
-evidence that (i) MYL frogs have experienced strong selection following
+evidence that (1) MYL frogs have experienced strong selection following
 exposure to a recently-emerged lethal fungal disease and that some of
 the genomic regions under selection are associated with genes related to
-disease resistance/tolerance, and (ii) reintroduction of individuals
-from these resistant/tolerant populations allows the successful
-reestablishment of viable frog populations despite ongoing Bd infection.
+disease resistance/tolerance, and (2) reintroduction of individuals from
+these resistant/tolerant populations allows the successful
+reestablishment of viable frog populations despite ongoing disease.
 
@@
 relatively constant Bd loads and mass die-offs are not observed
-\citep[see also Figure S5 - before
-translocation]{briggs2010, knapp2011}. The differences in Bd load of
-frogs from naive and recovering populations are also observed in
-controlled laboratory studies \citep[Figure 4]{knapp2016}. The observed
-reduced susceptibility of frogs from recovering populations could in
-theory be due to several factors including natural selection for more
-resistant/tolerant genotypes and/or acquired immunity, but until now
-evidence to evaluate the role of evolution was lacking.
+\citep[see also
+Figure S5]{briggs2010, knapp2011}. The
+differences in Bd load of frogs from naive and recovering populations
+are also observed in controlled laboratory studies \citep[Figure
+4]{knapp2016}. The observed reduced susceptibility of frogs from
+recovering populations could in theory be due to several factors,
+including natural selection for more resistant/tolerant genotypes and/or
+acquired immunity, but until now evidence to evaluate the role of
+evolution was lacking.
 
@@
 the strongest indication of directional selection (as measured by
-\(\pi_{diff}\)) contains the gene GVINP1. Several previous studies have
-found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018, ellison2014} and in populations differing in
-Bd susceptibility \citep{grogan2018}. This gene is also strongly linked
-to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. We also identified a region,
-characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
-that contained the complement genes C6 and C7. This could indicate that
-balancing selection is acting in this region of the genome to favor a
-diverse set of alleles, as is known for C6 in humans
-\citep{soejima2005}. Finally, the RIN3 gene showed a consistent pattern
-of allele frequency differences across all nine of the frog populations
-sampled in this study, indicating consistent selection in populations
-distributed across a wide geographic area. This gene is associated with
-inflammation response and in \emph{Xenopus} is highly expressed in wound
-stumps following amputation of tadpole tails \citep{fukazawa2009}. In
-the SI, we provide additional details on these and other genes
-identified in our study as under selection.
+\(\pi_{diff}\)) contains the interferon-related gene GVINP1. Several
+previous studies have found this gene to be differentially expressed
+during Bd infection \citep[e.g.,][]{grogan2018, ellison2014} and in
+populations differing in Bd susceptibility \citep{grogan2018}. This gene
+is also strongly linked to disease in salmon, explaining a notable 20\%
+of the resistance phenotype \citep{robledo2020, robledo2018}. We also
+identified a region, characterized by high \emph{F\textsubscript{ST}}
+and low \(\pi_{diff}\), that contained the complement genes C6 and C7.
+This could indicate that balancing selection is acting in this region of
+the genome to favor a diverse set of alleles, as is known for C6 in
+humans \citep{soejima2005}. Finally, based on the analysis of individual
+outlier variants, the RIN3 gene showed a consistent pattern of allele
+frequency differences across all nine of the frog populations sampled in
+this study, indicating consistent selection in populations distributed
+across a wide geographic area. This gene is associated with inflammation
+response and in \emph{Xenopus} is highly expressed in wound stumps
+following amputation of tadpole tails \citep{fukazawa2009}. Also in this
+analysis, the outlier variant with the lowest p-value (uncharacterized
+gene LOC108802036) is located adjacent to a type I interferon gene
+(Np-IFNi2) in another frog genome \citep{gan2018}, and together with
+GVINP1 further indicates the importance of interferon-related genes in
+this system.
 
@@
 populations, individual and population-level impacts of Bd are still
-evident. In an earlier study of 2 of our 12 translocated populations, Bd
-infection and load had detectable effects on the survival of adults and
-may have influenced population establishment (sites referred to as
-``Alpine'' and ``Subalpine'' in \citep{joseph2018} are identified as
-``70550'' and ``70505'' in the current study). Applying similar analyses
-to all 12 of our translocated populations would likely provide a broader
-perspective of the ongoing effect of Bd. In addition to these important
-but relatively subtle effects on adults, MYL frogs immediately following
-metamorphosis (``metamorphs'') are highly susceptible to Bd infection
-\citep{ellison2018} and experience high mortality. This high
-susceptibility of metamorphs is documented for numerous species of
-anurans, and may result from the poorly developed immune system
-characteristic of this early life stage \citep{humphries2022}. In
-naturally recovering and translocated MYL frog populations, we suggest
-that the high mortality of metamorphs is an important limitation on
-subsequent recruitment of new adults. Therefore, although adult MYL
-frogs appear relatively resistant/tolerant, Bd infection continues to
-have important limiting effects on recovering populations \citep[see
-also][]{hollanders2022}.
+evident. In an earlier study of 2 of our 12 translocated populations
+\citep{joseph2018}, Bd infection and load had detectable effects on the
+survival of adults and may have influenced population establishment
+(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
+are identified as ``70550'' and ``70505'' in the current study).
+Applying similar analyses to all 12 of our translocated populations
+would likely provide a broader perspective of the ongoing effect of Bd.
+In addition to these important but relatively subtle effects of Bd on
+adults, the impacts on younger life stages are more apparent. MYL frogs
+immediately following metamorphosis (``metamorphs'') are highly
+susceptible to Bd infection \citep{ellison2018} and as a result
+experience high mortality. This high susceptibility of metamorphs is
+documented for numerous species of anurans, and may result from the
+poorly developed immune system characteristic of this life stage
+\citep{humphries2022}. In naturally recovering and translocated MYL frog
+populations, we suggest that the high mortality of metamorphs is an
+important limitation on subsequent recruitment of new adults. Therefore,
+although adult MYL frogs appear relatively resistant/tolerant, Bd
+infection continues to have important limiting effects on recovering
+populations \citep[see also][]{hollanders2022}.
 
@@
 resources over many decades. The results of this study provide a hopeful
-starting point for that endeavor, and other future efforts worldwide.
+starting point for that endeavor and other future efforts worldwide.
 
@@
 these individuals can be used in translocations to reestablish
-extirpated populations. We expect that such ``assisted evolutionary
-rescue'' will be an essential tool in wildlife conservation in an era of
-accelerating global change.
+extirpated populations. We expect that such evolutionary rescue-based
+translocations will be an essential tool in wildlife conservation in an
+era of accelerating global change.
 
@@
 We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
-\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from four
-Bd-naive populations, and 29 from five recovering populations. These
-populations were located in the southern Sierra Nevada, from northern
-Yosemite National Park to northern Sequoia National Park
+\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from four naive
+populations, and 29 from five recovering populations. These populations
+were located in the southern Sierra Nevada, from northern Yosemite
+National Park to northern Sequoia National Park
 (Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
@@
 Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome
-\citep[NCBI SRA: SRS6533475, reference =][]{hon2020} with repetitive
-elements masked using bwa (``mem'' mode) \citep{li2013}. Exact PCR
-Duplicates were marked using Picard. Variants were then called following
-GATK best practices (v.4.2.0.0, \citep{van2020}). Briefly, raw variants
-were called for each sample using HaplotypeCaller and combined using
-CombineGVCFs. Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the
-following parameters to remove low-quality sites: QD \textless{} 2.0, FS
+\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
+SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
+bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
+using Picard. Variants were then called following GATK best practices
+(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
+sample using HaplotypeCaller and combined using CombineGVCFs. Next,
+genotypes were jointly called using GenotypeGVCFs. Variants were then
+hard filtered using gatk VariantFiltration using the following
+parameters to remove low-quality sites: QD \textless{} 2.0, FS
 \textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
@@
 proportion of 0.5. This filter resulted in 427,038 sites, of which
-353,172 were single nucleotide polymorphisms (SNPs) and 73,866 were
-insertions or deletions (INDELS). Finally, we trimmed samples with an
-average depth across filtered sites \textless{} 7x. Our final dataset
-included 50 samples (23 from naive and 27 from recovering populations)
-with an average depth of 16.7x (range = 7.4x -- 26.1x).
+353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
+with an average depth across filtered sites \textless{} 7x. Our final
+dataset included 50 samples, 23 from naive and 27 from recovering
+populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
 
@@
 To visualize the genomic relationship of our populations we conducted a
-principal components analysis (PCA) using the glPCA function in the
-adegenet R package \citep{jombart2008}. To detect regions of the genome
-under selection between naive and recovering populations, we used two
-approaches: (1) a multivariate linear mixed model to evaluate individual
-variants (SNPs and INDELs), and (2) a splined window analysis to
-evaluate larger genomic regions. For the variant analysis, we first used
-a stringent data filter to include only variants with \textless{} 5\%
-missing data (missing for no more than 2 individuals), and then
-calculated the likelihood ratio statistic for the resulting set of
-148,307 high quality variants across 127 contigs using GEMMA. GEMMA
+PCA using the glPCA function in the adegenet R package
+\citep{jombart2008}. To detect regions of the genome under selection
+between naive and recovering populations, we used two approaches: (1) a
+multivariate linear mixed model to evaluate individual variants (SNPs
+and INDELs), and (2) a splined window analysis to evaluate larger
+genomic regions. For the variant analysis, we first used a stringent
+data filter to include only variants with \textless{} 5\% missing data
+(missing for no more than 2 individuals), and then calculated the
+likelihood ratio statistic for the resulting set of 148,307 high quality
+variants across 127 contigs using GEMMA \citep{zhou2014}. GEMMA
 calculates and incorporates a relatedness matrix for input samples,
@@
 corrected significance level of 0.05 and used this set solely for the
-gene ontology (GO) analysis (see below and SI).
-
-For each outlier variant, we determined whether the variant was
-synonymous (protein sequence the same for each variant) or
-non-synonymous (protein sequence differs between variants), and where in
-the gene it was located. To do this, we first extracted the reference
-genome sequence surrounding the variant using the bedtools ``getfasta''
-function \citep{quinlan2010}. Next, we re-annotated each sequence using
-blast to get the predicted gene location based on the closest annotated
-reference \citep{altschul1997}. We then translated each variant to amino
-acids and aligned this translation to that of the gene annotation to
-ensure proper frame of reference using Geneious \citep{kearse2012}.
-After ensuring proper translation, we characterized variants as within
-or outside the coding sequence of the gene and as either synonymous or
-non-synonymous.
+gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+
+In the analysis of individual variants, for each outlier variant we
+determined whether the variant was synonymous (protein sequence the same
+for each variant) or non-synonymous (protein sequence differs between
+variants), and where in the gene it was located. To do this, we first
+extracted the reference genome sequence surrounding the variant using
+the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
+re-annotated each sequence using BLAST to get the predicted gene
+location based on the closest annotated reference \citep{altschul1997}.
+We then translated each variant to amino acids and aligned this
+translation to that of the gene annotation to ensure proper frame of
+reference using Geneious \citep{kearse2012}. After ensuring proper
+translation, we characterized variants as within or outside the coding
+sequence of the gene and as either synonymous or non-synonymous.
 
@@
 size-sorted chromosome number and adjusted the SNP position based on the
-relative position in the genome. We then used the WINGEN R package
-\citep{bishop2023} to conduct a splined discrete window analysis for
+relative position in the genome. We then used the GenWin R package
+\citep{beissinger2015} to conduct a splined discrete window analysis for
 \emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
@@
 positive and false negative results that occur using other window-based
-methods \citep{beissinger2015} . This method also calculates a
+methods \citep{beissinger2015}. This method also calculates a
 W-statistic allowing for outlier identification. We identified outliers
 as those with a W-statistic greater than 4 standard deviations above the
-mean for F\textsubscript{ST} or above/below the mean for \(\pi_{diff}\).
-Outlier regions were then identified as those that were outliers in both
-analyses, meaning that they showed 1) high differentiation between naive
-and recovering populations, and 2) differential patterns of nucleotide
-diversity in the same region. Finally, we extracted gene transcripts
-mapped within each region and retrieved annotation for that region using
-blast.
+mean for \emph{F\textsubscript{ST}} or above/below the mean for
+\(\pi_{diff}\). Shared outliers were then identified as those that were
+outliers in both analyses, meaning that they showed (1) high
+differentiation between naive and recovering populations, and (2)
+differential patterns of nucleotide diversity in the same region.
+Finally, we extracted gene transcripts mapped within each region and
+retrieved annotation for that region using BLAST (SI Appendix: Datasets
+S3, S4).
 
@@
 overrepresentation test (Fisher's exact test) using Blast2GO
-\citep{conesa2005}. First, we retrieved the Blast hits and mapped GO
+\citep{conesa2005}. First, we retrieved the BLAST hits and mapped GO
 terms for each gene in our targeted transcriptome. Next, we compared the
-29 unique outlier genes from the liberal set of outlier variants to the
+30 unique outlier genes from the liberal set of outlier variants to the
 complete set of genes in our target transcriptome to determine if any GO
@@
 located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
-splined windows. Code used for genomic analyses and to figures is
+splined windows. Code used for genomic analyses and to create figures is
 available from the following GitHub repository:
@@
 collected using several years of visual encounter surveys (VES) and skin
-swab collections \citep{knapp2016}, as well as landscape genetic
-information \citep{poorten2017}. The populations that we selected
-contained several hundred \emph{R. sierrae} adults and hundreds or
-thousands of tadpoles. These relatively high abundances were the result
-of recent increases following previous Bd-caused declines
-\citep{knapp2016}. As is typical for recovering MYL frog populations, Bd
-prevalence in the donor populations was high (0.69--0.96) and Bd load
-(median log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or
-more orders of magnitude below the level at which frog mortality is
-expected (log\textsubscript{10}(load) \(\approx\) 5.78
+swab collections \citep{knapp2016}, and results from population genetic
+analyses \citep{poorten2017}. The populations that we selected contained
+several hundred \emph{R. sierrae} adults and hundreds or thousands of
+tadpoles. These relatively high abundances were the result of recent
+increases following previous Bd-caused declines \citep{knapp2016}. As is
+typical for recovering MYL frog populations, Bd prevalence in the donor
+populations was high (0.69--0.96) and Bd load (median
+log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or more
+orders of magnitude below the level at which frog mortality is expected
+(log\textsubscript{10}(load) \(\approx\) 5.78
 copies)\citep{vredenburg2010, joseph2018}. Recipient sites to which
@@
 
-We conducted 1--4 translocations per site (Figure S3)
-and each translocated cohort included 18 to 99 frogs (median = 30).
-During each translocation, adult frogs were collected from the donor
-population and measured, weighed, swabbed, and PIT tagged. Frogs were
-transported to the recipient site either on foot or via helicopter.
+We conducted 1--4 translocations per site (Figure S3) and
+each translocated cohort included 18 to 99 frogs (median = 30). In
+preparation for each translocation, adult frogs were collected from the
+donor population and measured, weighed, swabbed, and PIT tagged. Frogs
+were transported to the recipient site either on foot or via helicopter.
 Following release, translocated populations were visited approximately
@@
 per summer = 2; range = 1--10) and assessed using diurnal CMR surveys
-and VES. CMR surveys allowed quantification of adult survival, adult
+and VES. CMR surveys allowed estimation of adult survival, adult
 population size, and recruitment of new adults, and VES provided
@@
 \emph{R. sierrae} tadpoles and juveniles. These \emph{R. sierrae} life
-stages have high detectablility, and counts are highly repeatable and
+stages have high detectability, and counts are highly repeatable and
 provide estimates of relative abundance \citep{knapp2000}.
+
+\hypertarget{frog-counts-and-reproductive-success}{%
+\subsection*{Frog counts and reproductive
+success}\label{frog-counts-and-reproductive-success}}
+
+For each of the translocated populations, we used the presence of
+tadpoles and/or juveniles from VES and captures of new recruits (i.e,
+untagged adults) in CMR surveys to provide two measures of successful
+reproduction. To calculate the proportion of years in which
+tadpoles/juveniles were present at a site, we excluded surveys conducted
+in the year of the initial translocation to that site. This exclusion
+accounted for the fact that all translocations were conducted after the
+breeding period and reproduction would therefore not have occurred until
+the following year. Similarly, to calculcate the proportion of years in
+which new recruits were present at a site, we excluded surveys conducted
+during the 3 years following the initial translocation. This accounted
+for the multi-year tadpole and juvenile stages in MYL frogs
+(Table S2).
 
@@
 and, when data contained possible hierarchical structure, we included
-group-level effects in the full model. Within each analysis, we
-evaluated model fit using visualizations of leave-one-out (``LOO'')
-probability integral transformations \citep{gelman2013, gabry2019}, and
-compared fits of models with and without group-level effects, and with
-differing group-level effects using LOO cross-validation
-\citep{vehtari2016}. Both approaches utilized the loo package
-\citep{vehtari2022}.
+group-level effects in the full model. We evaluated model fit using
+leave-one-out (``LOO'') cross-validation
+\citep{gelman2013, gabry2019, vehtari2016}, as implemented by the loo
+package \citep{vehtari2022}.
 
@@
 describing characteristics of frogs, sites, and translocations
-(predictor variables: Bd load (bd\_load), winter severity in the year
-following translocation (snow\_t1), site elevation (elevation), donor
-population (donor\_72996), sex (sex\_male), frog size (length), winter
-severity in the year of translocation (snow\_t), day of year on which a
-translocation was conducted (day), and translocation order
+(predictor variables: Bd load (bd\_load), sex (sex\_male), frog size
+(length), site elevation (elevation), winter severity in the year of
+translocation (snow\_t), winter severity in the year following
+translocation (snow\_t1), donor population (donor\_72996), day of year
+on which a translocation was conducted (day), and translocation order
 (order\_first)). In both analyses we used 1-year survival as a single
@@
 regardless of how many years of post-translocation data were available
-(range = 3-\/-16). Group-level effects included site\_id,
+(range = 3--16). Group-level effects included site\_id,
 translocation\_id, or translocation\_id nested within site\_id. In both
@@
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see SI for details]{boyle2004}. To assess the magnitude of
-changes in Bd load on frogs following translocation, we compared Bd
+\citep[see SI Appendix for details]{boyle2004}. To assess the magnitude
+of changes in Bd load on frogs following translocation, we compared Bd
 loads measured before versus after translocation. Before-translocation
@@
 prevalence significantly simplifies the model without much loss of
-realism.
+realism. (Additional details regarding incorporating yearly variability
+in vital rates and estimating model parameters are provided in the SI
+Appendix.)
 
@@
 After parameterizing our model with CMR-estimated adult frog survival
-probabilities and other known vital rates (Table S2), we performed four analyses (described in detail in the SI).
-First, we compute the long-run growth rate \(\lambda\) for each of our
-12 translocated populations to determine if the populations were
-deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to model parameters, to
-quantify how much changes in adult survival, for example, affected the
-long-run growth rate. Third, we included demographic and environmental
-stochasticity into our model and simulated the 50-year viability (i.e.,
-1 - extinction probability) of populations given an introduction of 40
-adult individuals into an unoccupied habitat. Finally, we fit our model
-to our longest translocation trajectory to confirm that our model could
-reasonably reproduce the observed recovery trajectories of MYL frogs
-following reintroductions.}
+probabilities and other known vital rates
+(Table S2), we performed four analyses (described in
+detail in the SI Appendix). First, we compute the long-run growth rate
+\(\lambda\) for each of our 12 translocated populations to determine if
+the populations were deterministically predicted to grow or decline in
+the long-run. Second, we compute the elasticity of \(\lambda\) to model
+parameters, to quantify how much changes in adult survival, for example,
+affected the long-run growth rate. Third, we included demographic and
+environmental stochasticity in our model and simulated the 50-year
+viability (i.e., 1 - extinction probability) of populations given an
+introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to
+confirm that our model could reasonably reproduce the observed recovery
+trajectories of MYL frogs following reintroductions.}
 
@@
 assistance. This project was supported by grants from the National Park
-Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National
-Science Foundation Grants EF-0723563 (to C. Briggs), DEB-1557190 (to C.
-Briggs), DEB-2133401 (to M.Q.W), and DBI-2120084 (to C.
-Richards-Zawacki).}
+Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
+Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
 
@@
 
-\caption{\label{fig-allelefrequencies}Evidence for selection in
-recovering MYL frog populations at the landscape scale. For each of the
-9 naive and recovering MYL frog populations, pie charts show allele
+\caption{\label{fig-allelefrequencies}Evidence for selection on
+individual variants in recovering MYL frog populations at the landscape
+scale. For each of the 9 naive and recovering MYL frog populations
+(indicated by numbered points), adjacent pie charts show allele
 frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
-LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
+LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
 Charts are superimposed on a map of the Sierra Nevada study area, with
 Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
-shown in dark gray, and the range boundary for MYL frogs in this portion
-of the Sierra Nevada shown in light gray. The inset map locates the
-study area in California.}
+shown in dark gray, and the range boundary for MYL frogs shown in light
+gray. The inset map locates the study area and range boundary in
+California.}
 
@@
 \caption{\label{fig-spline-manhattan}Evidence for selection on genomic
-regions in recovering MYL frog populations. Results are from the splined
-window analysis and show outlier windows for the difference in (A)
-nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
-Outlier regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}}
-are shown in blue and outlier regions for both \(\pi_{diff}\) and
+regions in recovering MYL frog populations. Manhattan plot of the
+results from the splined window analysis showing outlier regions for the
+difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
+\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
+upper red dashed line and below the lower red dashed line. In (B),
+outlier regions are shown above the single dashed red line. Outlier
+regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
+shown in blue and outlier regions for both \(\pi_{diff}\) and
 \emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
-(A) showing two adjacent outlier windows for \(\pi_{diff}\) 12.9Mb
+(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
 upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
 line). (D) Magnified Contig8 from from (B) showing the
-\emph{F\textsubscript{ST}} outlier window that includes the outlier SNPs
+\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
 TCF19 and VARS. This region of the genome contains 8 annotated genes
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/translocation_survival_bysiteid.png}
+{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
 
@@
 the recipient sites following translocation. Median 1-year survival for
-each cohort of translocated frogs at the 12 recipient sites. Sites are
-arranged along the x-axis using the average of the median 1-year
-survival per translocation at each site. Dot colors indicate the donor
-population from which frogs in each translocated cohort were collected.
-Error bars show the 95\% uncertainty intervals. When multiple
-translocations were conducted to a site, points and error bars are
-slightly offset to avoid overlap.}
+each cohort of translocated frogs at the 12 recipient sites. Error bars
+show the 95\% uncertainty intervals. Sites are arranged along the x-axis
+using the average of the median 1-year survival per translocation at
+each site. Dot colors indicate the donor population from which frogs in
+each translocated cohort were collected. When multiple translocations
+were conducted to a site, points and error bars are slightly offset to
+avoid overlap.}
 
@@
 
-\caption{\label{fig-viability}Predicted and observed growth and
-persistence of translocated populations. Population viability analysis
-of translocated MYL frogs quantified the long-run growth rate and
+\caption{\label{fig-viability}Population viability analysis of
+translocated MYL frogs quantified the long-run growth rate and
 extinction probability of frog populations: (A) Predicted long-run
@@
 \(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
-12 translocated populations given demographic stochasticity,
+12 translocated populations, given demographic stochasticity,
 environmental variability in \(\omega\), and different mean values of

### 2023-05-19 doc/manuscript/translocation_pnas_SI.tex
Adjust manuscript formatting

@@
 across all genetic clusters for 1 of 7 of the associated genes (RIN3;
-Fig. 1). Nonetheless, for the other 6 genes
-we still found a statistically significant signal of selection in 2 of
-the 3 genetic clusters (containing populations 6, 7, 8, and 9;
-Fig. 1, Figure~\ref{fig-selectionresults} A
-SI). Therefore, although some outlier variant associations have a more
+Fig. 1). However, for the other 6 genes we
+still found a statistically significant signal of selection in 2 of the
+3 genetic clusters (containing populations 6--9;
+Fig. 1, Figure~\ref{fig-selectionresults}
+A). Therefore, although some outlier variant associations have a more
 limited geographic extent than RIN3, they still describe parallel
@@
 The baseline parameter values for the model and how they were estimated
-are given in Table~\ref{tbl-param_values} : SI. Parameters
-\(\sigma_{A_T}\) and \(\sigma_{A_R}\) were extracted directly from our
-CMR model described previously in this study. For populations where we
-had a sufficient number of PIT-tagged, naturally-recruited adults, we
-observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably
-different, with \(\sigma_{A_R} > \sigma_{A_T}\)
-(Figure~\ref{fig-compare_surv_probs} : SI). For populations lacking
+are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
+and \(\sigma_{A_R}\) were extracted directly from our CMR model
+described previously in this study. For populations where we had a
+sufficient number of PIT-tagged, naturally-recruited adults, we observed
+that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different,
+with \(\sigma_{A_R} > \sigma_{A_T}\)
+(Figure~\ref{fig-compare_surv_probs}). For populations lacking
 sufficient numbers of naturally-recruited adults, we were unable to
@@
 
-\caption{\label{fig-selectionresults}Evidence for selection of
+\caption{\label{fig-selectionresults}Evidence for selection on
 individual variants in recovering MYL frog populations. (A) PCA
@@
 Numeric labels and colors match those in
-Fig. 1 A. (B) qqplot showing observed and
-expected p-values for 148,307 SNPs and INDELS as calculated in GEMMA.
+Fig. 1 A. Populations 1-7 are \emph{R.
+sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
+showing observed and expected p-values for 148,307 SNPs and INDELS.
 Dashed line shows p-value that identifies outliers. (C) Manhattan plot
-showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by
-genomic position and contigs are sorted by size. Red line shows p-value
-that identifies outliers. Outlier SNPs above this threshold are
-highlighted and labeled. Bold labels indicate at least one
-non-synonymous SNP is present in that gene.}
+showing p-value for each SNP. SNPs are sorted by genomic position and
+contigs are sorted by size. Red line shows p-value that identifies
+outliers. Outlier SNPs above this threshold are highlighted and labeled.
+Bold labels indicate the presence of at least one non-synonymous SNP in
+that gene.}
 
@@
 accession GCA\_000004195.4) and \emph{Rana mucosa} Contig19. Asterices
-indicate the location of SNP outliers in TCF19 and VARS1 genes. Gap
+indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
 sizes for each contig representation are labeled.}
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/map_translocation_points.png}
+{\centering \includegraphics[width=0.60\textwidth]{figures/map_translocation_points.png}
 
@@
 elevation is indicated by the colored hillshade layer (dark green =
-lowest elevation, white = highest elevation). }
+lowest elevation, white = highest elevation).}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/mcmc_areas_m2b.png}
+{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m2b.png}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/bdload_beforeafter.png}
+{\centering \includegraphics[width=0.60\textwidth]{figures/bdload_beforeafter.png}
 
@@
 of severe disease are \textgreater{} 5.8 ITS copies (on a
-log\textsubscript{10} scale; see SI text for details). Box plot details
-are as in Fig. 4.}
+log\textsubscript{10} scale; see SI Appendix text for details). Box plot
+details are as in Fig. 4.}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/mcmc_areas_m1d.png}
+{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m1d.png}
 
@@
 
-\caption{\label{fig-survival-postdens}Evidence that Bd load in not an
+\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
 important predictor of post-translocation frog survival. For the best
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/compare_surv_probs.jpg}
+{\centering \includegraphics[width=0.60\textwidth]{figures/compare_surv_probs.jpg}
 
@@
 generations. Comparison of the average yearly adult survival
-probabilities for adults translocated to each of three sites versus
-adults that were naturally recruited at each site (as a result of
-reproduction by translocated frogs). In contrast to
+probabilities for adults translocated to each of 3 sites versus adults
+that were naturally recruited at each site (as a result of reproduction
+by translocated frogs). In contrast to
 Fig. 3, these are not survival
@@
 represent averaged survival probabilities across multiple years and
-cohort. Points are median estimates and error bars give the 95\%
+cohorts. Points are median estimates and error bars give the 95\%
 uncertainty intervals around the estimates, accounting for yearly
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/pop_viability_figures_for_supp.jpg}
+{\centering \includegraphics[width=0.60\textwidth]{figures/pop_viability_figures_for_supp.jpg}
 
@@
 proportion of populations translocated early versus late in the study
-period (\textless{} 2013 or \textgreater= 2013, respectively) and
-probability of survival (\textless{} 0.5 or \textgreater= 0.5). Based on
+period (\textless{} 2013 or $\geq$ 2013, respectively) and
+probability of survival (\textless{} 0.5 or $\geq$ 0.5). Based on
 the viability analysis, survival probabilities \textless{} 0.5 and
-\textgreater= 0.5 produced 50-year extinction probabilities of 1 and
+$\geq$ 0.5 produced 50-year extinction probabilities of 1 and
 \textless{} 0.5, respectively.}\tabularnewline
@@
 early & \textless{} 0.5 & 4 & 5 & 0.800 \\
-early & \textgreater= 0.5 & 1 & 5 & 0.200 \\
+early & $\geq$ 0.5 & 1 & 5 & 0.200 \\
 late & \textless{} 0.5 & 2 & 7 & 0.286 \\
-late & \textgreater= 0.5 & 5 & 7 & 0.714 \\
+late & $\geq$ 0.5 & 5 & 7 & 0.714 \\
 \bottomrule()
@@
 
-\hfill\break
+\newpage
 

### 2023-05-19 code/extinction_and_growth_rate_analysis.R

@@ -350,7 +350,7 @@ pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=omega, y=e
 							scale_color_manual(values=cdat$col) +
 							theme_classic() + xlab("Probability of recruitment (\u03c9)") + ylab("Extinction prob. in 50 years") +
 							guides(color=guide_legend(title=bquote("Lake, \u03c3"~.[AR]), title.position="top", ncol=3)) + 
-							theme(legend.position="bottom", legend.text=element_text(size=6), legend.title=element_text(size=8))
+							theme(legend.position="bottom", legend.text=element_text(size=7), legend.title=element_text(size=8))
 
 # Examine simulations
 # matplot(t(pop_trajectories[["70550"]][1:100, ]), type="l")

### 2023-05-19 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -149,29 +149,41 @@ pnas_ms = pnas_ms.replace("\\begin{figure", "\\begin{figure*")
 pnas_ms = pnas_ms.replace("\\end{figure", "\\end{figure*")
 pnas_ms = pnas_ms.replace("\\includegraphics", "\\includegraphics[width=\\textwidth]")
 
+# Set specific figure sizes 
+
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/translocation_survival_bysiteid.png}",
+						  "\\includegraphics[width=0.5\\textwidth]{figures/translocation_survival_bysiteid.png}")
 
 # Clean up references to SI material
 
 supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
+			"Figure~\\ref{fig-selectionresults} A": "Figure S1A",
 			"Figure~\\ref{fig-selectionresults} B, C: SI": "Figure S1B,C",
+			"Figure~\\ref{fig-selectionresults} B, C": "Figure S1B,C",
+			"Figure~\\ref{fig-selectionresults} B,\nC": "Figure S1B,C",
 			"Figure~\\ref{fig-selectionresults} B, C;\nSI": "Figure S1B,C",
 			"Figure~\\ref{fig-selectionresults} A": "Figure S1A",
 			"Figure~\\ref{fig-synteny-plot} SI": "Figure S2",
 			"Figure~\\ref{fig-synteny-plot}": "Figure S2",
 			"Figure~\\ref{fig-yosemap} SI": "Figure S3",
 			"Figure~\\ref{fig-yosemap}\nSI": "Figure S3",
+			"Figure~\\ref{fig-yosemap}": "Figure S3",
 			"Figure~\\ref{fig-transsurvival-postdens} SI": "Figure S4",
+			"Figure~\\ref{fig-transsurvival-postdens}": "Figure S4",
 			"Figure~\\ref{fig-bdload-beforeafter} SI": "Figure S5",
 			"Figure~\\ref{fig-bdload-beforeafter}\n SI": "Figure S5",
 			"Figure~\\ref{fig-bdload-beforeafter}\nSI": "Figure S5",
 			"Figure~\\ref{fig-bdload-beforeafter}": "Figure S5",
 			"Figure~\\ref{fig-survival-postdens} SI": "Figure S6",
+			"Figure~\\ref{fig-survival-postdens}": "Figure S6",
 			"Figure~\\ref{fig-compare_surv_probs} : SI": "Figure S7",
[…]

### 2023-05-19 doc/manuscript/translocation.qmd
Revise Abstract based on comments from coauthor EBR

@@
 
-Escalating anthropogenic change is devastating global biodiversity.
-To avoid extinction, affected populations will often need to adapt to the altered environment.
-The process of rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth is termed "evolutionary rescue".
-In addition to allowing recovery of extant populations, individuals from rescued populations could also be used to reestablish extirpated populations.
-We evaluated this scenario in the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
+Vast alteration of the biosphere by humans is causing a sixth mass extinction.
+Adaptation to modified environments often is the only means for species to persist, but the extent to which such "evolutionary rescue" can prevent or reverse biodiversity loss is largely unknown.
+Using results from genomic analyses, a 15-year reintroduction effort, and population modeling, we provide a compelling example of evolution reversing the declines of an imperiled amphibian and allowing the reestablishment of extirpated populations.
+The once-common mountain yellow-legged (MYL) frog is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
 Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
-We show that MYL frogs have undergone substantial evolutionary change following disease outbreaks, and that these changes are associated with increased resistance/tolerance to Bd infection in naturally recovering populations.
-Large-scale reintroduction of frogs from these populations resulted in the establishment of reproducing populations at most recipient sites despite ongoing disease.
+Using an exome capture approach, we show that MYL frogs have undergone substantial evolutionary change following disease outbreaks.
+Moreover, in naturally recovering populations observed changes in immune function loci are associated with increased resistance/tolerance to Bd infection.
+Large-scale reintroduction of frogs from rescued populations resulted in the establishment of reproducing populations despite ongoing disease.
 In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
+Collectively, these results provide a rare example of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
 This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.
@@
 In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
-Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of threatened taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
+Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of imperiled taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
 
@@
 Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
-We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.<!--# Dig deeper into the "difficult" task alluded to above?
-E.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data?
--->
+We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.
 
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: <https://github.com/SNARL1/cmr-analysis>.
-
-<!--# Add text describing calculations of VES and CMR-recruitment counts.
-For the latter, only included surveys conducted >= 3 years following translocation to account for duration of early life stages.
--->
 
@@
 We checked convergence of the MCMC algorithm using trace plots and Rhat.
-Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
 
@@
 Priors, chains, iterations, and checks were as described above.
-Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
-<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. -->
+
+Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.
 
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-(Additional details regarding incorporating yearly variability in vital rates and estimating model parameters are provided in the SI Appendix.)
+Additional details are provided in the SI Appendix.
 
@@
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-<!--# Add link to Mark's GitHub repo? -->
+Additional details are provided in the SI Appendix.
+Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
 
@@
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
-Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
 

### 2023-05-19 doc/manuscript/translocation_bioarxiv.tex
Update PNAS version and add biorxiv version

@@
 \begin{abstract}
-Escalating anthropogenic change is devastating global biodiversity. To
-avoid extinction, affected populations will often need to adapt to the
-altered environment. The process of rapid evolutionary change that
-increases the frequency of adaptive alleles and thereby restores
-positive population growth is termed ``evolutionary rescue''. In
-addition to allowing recovery of extant populations, individuals from
-rescued populations could also be used to reestablish extirpated
-populations. We evaluated this scenario in the once-common mountain
-yellow-legged (MYL) frog, which is threatened with extinction by the
+Vast alteration of the biosphere by humans is causing a sixth mass
+extinction. Adaptation to modified environments often is the only means
+for species to persist, but the extent to which such ``evolutionary
+rescue'' can prevent or reverse biodiversity loss is largely unknown.
+Using results from genomic analyses, a 15-year reintroduction effort,
+and population modeling, we provide a compelling example of evolution
+reversing the declines of an imperiled amphibian and allowing the
+reestablishment of extirpated populations. The once-common mountain
+yellow-legged (MYL) frog is threatened with extinction by the
 human-facilitated emergence of a lethal fungal pathogen
@@
 populations are extirpated following disease outbreaks, some persist and
-eventually begin to recover. We show that MYL frogs have undergone
-substantial evolutionary change following disease outbreaks, and that
-these changes are associated with increased resistance/tolerance to Bd
-infection in naturally recovering populations. Large-scale
-reintroduction of frogs from these populations resulted in the
-establishment of reproducing populations at most recipient sites despite
-ongoing disease. In addition, results from viability modeling suggest
-that established populations have a low probability of extinction over
-50 years. Collectively, these results provide one of the first examples
-of how evolutionary rescue and the reintroduction of resistant/tolerant
-individuals can allow the landscape-scale recovery of disease-impacted
-species. This example has potentially broad implications for the many
-taxa worldwide that are threatened with extinction by anthropogenic
-stressors.
+eventually begin to recover. Using an exome capture approach, we show
+that MYL frogs have undergone substantial evolutionary change following
+disease outbreaks. Moreover, in naturally recovering populations
+observed changes in immune function loci are associated with increased
+resistance/tolerance to Bd infection. Large-scale reintroduction of
+frogs from rescued populations resulted in the establishment of
+reproducing populations despite ongoing disease. In addition, results
+from viability modeling suggest that established populations have a low
+probability of extinction over 50 years. Collectively, these results
+provide a rare example of how evolutionary rescue and the reintroduction
+of resistant/tolerant individuals can allow the landscape-scale recovery
+of disease-impacted species. This example has potentially broad
+implications for the many taxa worldwide that are threatened with
+extinction by anthropogenic stressors.
 \end{abstract}
@@
 Although this overall scenario is potentially of critical importance for
-the conservation of an ever-growing number of threatened taxa,
+the conservation of an ever-growing number of imperiled taxa,
 descriptions of its occurrence and effectiveness in the wild are
@@
 Markov chain Monte Carlo (MCMC) algorithm using trace plots and
-Gelman-Rubin statistics (Rhat). Code used to create CMR datasets and fit
-models for each translocated population is available from the following
-GitHub repository: \url{https://github.com/SNARL1/cmr-analysis}.
+Gelman-Rubin statistics (Rhat).
 
@@
 checked convergence of the MCMC algorithm using trace plots and Rhat.
-Code to create the datasets and fit models is available from the
-following GitHub repository:
-\url{https://github.com/SNARL1/translocation}.
 
@@
 family. Priors, chains, iterations, and checks were as described above.
-Code to create the dataset and fit models is available from the
-following GitHub repository:
-\url{https://github.com/SNARL1/translocation}.
+
+Code to replicate all of the analyses described above is available from
+the following GitHub repositories:
+\url{https://github.com/SNARL1/translocation};
+\url{https://github.com/SNARL1/cmr-analysis}.
 
@@
 discrete-time, stage-structured model where a time step is one year. The
-dynamics are given by
+dynamics are given by equation \ref{eqn:matrix}.
 
@@
 prevalence significantly simplifies the model without much loss of
-realism. (Additional details regarding incorporating yearly variability
-in vital rates and estimating model parameters are provided in the SI
-Appendix.)
+realism. Additional details are provided in the SI Appendix.
 
@@
 confirm that our model could reasonably reproduce the observed recovery
-trajectories of MYL frogs following reintroductions.}
+trajectories of MYL frogs following reintroductions. Additional details
+are provided in the SI Appendix. Code to replicate the analyses can be
+found at \url{https://github.com/SNARL1/translocation}.}
 

### 2023-05-19 doc/manuscript/translocation_pnas.tex
Update PNAS version and add biorxiv version

@@
 taxa will often need to adapt to their altered environment or face
-decline and extinction. ``Evolutionary rescue'' describes rapid
-evolutionary change that increases the frequency of adaptive alleles and
-thereby restores positive population growth \citep{carlson2014}. In
-addition to preventing extinction of declining populations, this process
-could also result in the seeding of nearby empty habitats and allow
-population reestablishment and recovery. Although this overall scenario
-is potentially of critical importance for the conservation of an
-ever-growing number of threatened taxa, descriptions of its operation in
-the wild are lacking.
+decline and possible extinction. The process whereby rapid evolutionary
+change increases the frequency of adaptive alleles and restores positive
+population growth is known as ``evolutionary rescue''
+\citep{carlson2014}. In addition to preventing extinction of declining
+populations, individuals from rescued populations could also disperse,
+or be reintroduced, to empty habitats and promote larger-scale recovery.
+Although this overall scenario is potentially of critical importance for
+the conservation of an ever-growing number of imperiled taxa,
+descriptions of its occurrence and effectiveness in the wild are
+lacking.
 
@@
 prevalence significantly simplifies the model without much loss of
-realism.
+realism. Additional details are provided in the SI Appendix.
 
@@
 After parameterizing our model with CMR-estimated adult frog survival
-probabilities and other known vital rates (Table S2), we performed four analyses (described in detail in the SI).
-First, we compute the long-run growth rate \(\lambda\) for each of our
-12 translocated populations to determine if the populations were
-deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to model parameters, to
-quantify how much changes in adult survival, for example, affected the
-long-run growth rate. Third, we included demographic and environmental
-stochasticity into our model and simulated the 50-year viability (i.e.,
-1 - extinction probability) of populations given an introduction of 40
-adult individuals into an unoccupied habitat. Finally, we fit our model
-to our longest translocation trajectory to confirm that our model could
-reasonably reproduce the observed recovery trajectories of MYL frogs
-following reintroductions.}
+probabilities and other known vital rates
+(Table S2), we performed four analyses (described in
+detail in the SI Appendix). First, we compute the long-run growth rate
+\(\lambda\) for each of our 12 translocated populations to determine if
+the populations were deterministically predicted to grow or decline in
+the long-run. Second, we compute the elasticity of \(\lambda\) to model
+parameters, to quantify how much changes in adult survival, for example,
+affected the long-run growth rate. Third, we included demographic and
+environmental stochasticity in our model and simulated the 50-year
+viability (i.e., 1 - extinction probability) of populations given an
+introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to
+confirm that our model could reasonably reproduce the observed recovery
+trajectories of MYL frogs following reintroductions. Additional details
+are provided in the SI Appendix. Code to replicate the analyses can be
+found at \url{https://github.com/SNARL1/translocation}.}
 

### 2023-05-19 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -119,6 +119,7 @@ methods = latex_manuscript[methods_start:methods_end].split(matmethkey1)[1].stri
 # Update equation to span two columns
 methods = methods.replace("\\[\n\\begin{bmatrix}", "\\begin{figure}\\[\n\\begin{bmatrix}")
 methods = methods.replace("\\end{bmatrix}(t) \n\\]", "\\end{bmatrix}(t)\\numberthis \\label{eqn:matrix} \n\\]\\end{figure}")
+methods = methods.replace("The\ndynamics are given by", "The\ndynamics are given by equation \\ref{eqn:matrix}.")
 
 methods = methods.replace("\\paragraph{", "\\section*{")
 methods = methods.replace("\\subparagraph{", "\\subsection*{")
@@ -154,6 +155,8 @@ pnas_ms = pnas_ms.replace("\\includegraphics", "\\includegraphics[width=\\textwi
 pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/translocation_survival_bysiteid.png}",
 						  "\\includegraphics[width=0.5\\textwidth]{figures/translocation_survival_bysiteid.png}")
 
+
+pnas_ms = pnas_ms.replace("The dynamics are given by", "The dynamics are given by ")
 # Clean up references to SI material
 
 supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
@@ -233,7 +236,7 @@ figures_supp = figures_supp.replace("\\includegraphics[width=0.85\\textwidth]{fi
 							  "\\includegraphics[width=0.60\\textwidth]{figures/mcmc_areas_m2b.png}")
 
 figures_supp = figures_supp.replace("\\includegraphics[width=0.85\\textwidth]{figures/bdload_beforeafter.png}", 
-							  "\\includegraphics[width=0.60\\textwidth]{figures/bdload_beforeafter.png}")
+							  "\\includegraphics[width=0.7\\textwidth]{figures/bdload_beforeafter.png}")
 
 figures_supp = figures_supp.replace("\\includegraphics[width=0.85\\textwidth]{figures/mcmc_areas_m1d.png}", 
 							  "\\includegraphics[width=0.60\\textwidth]{figures/mcmc_areas_m1d.png}")

### 2023-05-19 doc/manuscript/translocation.qmd
Minor edits throughout based on final review

@@
 Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@newell2013; @voyles2018; @knapp2016].
+Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
 Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see [@medzhitov2012] for definitions).
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
 However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
@@
 Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
 One of the 7 identified genes (LOC108802036) did not have an associated annotation.
@@
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-(See SI Appendix for additional description of these results.)
 
 The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more outlier SNPs (described above).
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
 A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
 The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
@@
 To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap) [@seddon2014].
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@seddon2014] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
 The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
-Following translocation, adult survival and recruitment of new adults were estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
+Following translocation, adult survival and recruitment of new adults was estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
 
@@
 sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter, see SI Appendix text for additional details) [@joseph2018; @vredenburg2010].
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
+Additional details on methods used to estimate Bd loads are provided in the SI Appendix.
 
@@
 As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
-In summary, survey results suggest that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
+In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
 
 To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
-Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
+Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
+Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
 Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
@@
 Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
@@
 Here, we tested the possibility that populations of resistant/tolerant individuals could be used to reestablish extirpated populations.
-Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild (see [@joseph2018] for a small-scale example).
+Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild [see @joseph2018 for a small-scale example].
 In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
-The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (2) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing disease.
+The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection contain genes related to disease resistance/tolerance, and (2) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing disease.
 
@@
 In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter].
-The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
+The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016].
 The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors, including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
@@
 We also identified a region, characterized by high *F~ST~* and low $\pi_{diff}$, that contained the complement genes C6 and C7.
-This could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
+The complement system plays an important role in innate immunity [@rieraromo2016], and our results could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
 Finally, based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
 This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
-Also in this analysis, the outlier variant with the lowest p-value (uncharacterized gene LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2) in another frog genome [@gan2018], and together with GVINP1 further indicates the importance of interferon-related genes in this system.
+Also in this analysis, the outlier variant with the lowest p-value (uncharacterized gene LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2) in another frog genome [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
 
@@
 Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
-In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in [@joseph2018] are identified as "70550" and "70505" in the current study).
+In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
 Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
@@
 MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
-This high susceptibility of metamorphs is documented for numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
+This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
 In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
@@
 
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from four naive populations, and 29 from five recovering populations.
-These populations were located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
+These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
 Samples were collected from 5-6 frogs per population.
@@
 Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
+Raw sequencing reads are available from NCBI SRA (PRJNA870451).
 
@@
 
-To visualize the genomic relationship of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
+To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
 To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
@@
 GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni- corrected significance level of 0.01.
+We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni- corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
 
@@
 
-Using the liberal set of variant outliers, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
 First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
@@
 Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-<!--# Make sure that code has been uploaded to this link before manuscript submission. -->
 
@@
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
@@
 When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
-Frog skin swabs were analyzed using standard Bd DNA extraction and qPCR methods.
-For details of survey and swab analysis methods, see [@joseph2018].
 
@@
 
-For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and captures of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
+For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
 To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
@@
 Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), sex (sex_male), frog size (length), site elevation (elevation), winter severity in the year of translocation (snow_t), winter severity in the year following translocation (snow_t1), donor population (donor_72996), day of year on which a translocation was conducted (day), and translocation order (order_first)).
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
 In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
@@
 
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI Appendix).
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
@@
 The red line shows where $\lambda = 1$.
+Note that the point for 70413 is mostly hidden behind other points.
 (B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
@@
 
-In the Results section of the main paper, we describe the stringent set of outlier SNPS (identified using a Bonferroni-corrected p-value of 0.01).
+In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
 The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
@@
 
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
+In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
 However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
@@
 
-Swab extracts were analyzed singly instead of in triplicate [@kriger2006].
+Swab extracts were analyzed using standard Bd DNA extraction and qPCR methods [@boyle2004], and extracts were analyzed singly instead of in triplicate [@kriger2006].
 For analysis of swabs collected during 2005--2014, we used standards developed from known concentrations of zoospores [@boyle2004], and after 2014, we used standards based on single ITS1 PCR amplicons [@longo2013].
@@
 
-We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Estimation of adult survival and population size" above).
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among population variability ([@fig-translocation-survival]).
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
@@
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study.
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
 For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
@@
 Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale; see SI Appendix text for details).
-Box plot details are as in @fig-cond-effects.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
+Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
+Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
 
@@
 For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="350"}
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
+Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="350"}
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-The large error bars for 70550 reflect the large within and among individual variability in yearly survival probability at this site (as also reflected in one-year survival probabilities in [@fig-translocation-survival]).
-All estimates are derived from the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
+All estimates were derived using the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
 

### 2023-05-19 doc/manuscript/PNAS-template-empty.tex
Adjust manuscript formatting
Address formatting comments on compiled manuscript

@@
 \author[a,b]{Thomas C. Smith}
-\author[d,e]{Andrew P. Rothstein}
-\author[h]{Robert L. Grasso}
+\author[d,e,h]{Andrew P. Rothstein}
+\author[i]{Robert L. Grasso}
 \author[d,e]{Erica Bree Rosenblum}
@@
 \affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
-\affil[g]{Planet, San Francisco, CA, 94107}
-\affil[h]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}
+\affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
+\affil[h]{Current affiliation: Ginkgo Bioworks, Boston, MA, 02210}
+\affil[i]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}
 

### 2023-05-19 doc/manuscript/translocation_bioarxiv.tex
Address formatting comments on compiled manuscript

@@
 % Please add a significance statement to explain the relevance of your work
-\significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. One possibility is rapid
-evolution that restores positive growth to declining populations
-(``evolutionary rescue''). Individuals from these rescued populations
-could also be used to reestablish extinct populations. We evaluated both
-possibilities in a once-common frog species driven to the brink of
-extinction by a disease that has devastated global amphibian
-biodiversity and for which few options exist to reverse disease impacts.
-We found that frogs have evolved increased resistance to infection,
-facilitating population recovery. In addition, introduction of frogs
-from these populations allowed reestablishment despite the ongoing
-presence of disease. Such reintroductions could play an important role
-in biodiversity conservation in our rapidly changing world.}
+% \significancestatement{Understanding how species persist despite accelerating global change is
+% critical for the conservation of biodiversity. One possibility is rapid
+% evolution that restores positive growth to declining populations
+% (``evolutionary rescue''). Individuals from these rescued populations
+% could also be used to reestablish extinct populations. We evaluated both
+% possibilities in a once-common frog species driven to the brink of
+% extinction by a disease that has devastated global amphibian
+% biodiversity and for which few options exist to reverse disease impacts.
+% We found that frogs have evolved increased resistance to infection,
+% facilitating population recovery. In addition, introduction of frogs
+% from these populations allowed reestablishment despite the ongoing
+% presence of disease. Such reintroductions could play an important role
+% in biodiversity conservation in our rapidly changing world.}
 
@@
 disease have subsequently recovered to varying extents
-\citep{newell2013, voyles2018, knapp2016}. Although we know relatively
+\citep{scheele2017, voyles2018, knapp2016}. Although we know relatively
 little about the mechanisms underlying recovery
@@
 \citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
-often results in increased host resistance and/or tolerance (see
-\citep{medzhitov2012} for definitions). However, linking these
-evolutionary responses to population recoveries is difficult and few
-clear examples exist. In addition to promoting recovery of
-disease-impacted populations, the resistant/tolerant individuals in such
-populations could potentially be used to reestablish populations that
-were extirpated following disease emergence
+often results in increased host resistance and/or tolerance \citep[see][
+for definitions]{medzhitov2012}. However, linking these evolutionary
+responses to population recoveries is difficult and few clear examples
+exist. In addition to promoting recovery of disease-impacted
+populations, the resistant/tolerant individuals in such populations
+could potentially be used to reestablish populations that were
+extirpated following disease emergence
 \citep{mendelson2019, brannelly2021}. To date, the use of
@@
 allele frequencies between naive versus recovering populations) from 7
-distinct genes across 4 contigs (Figure S1B,C). One of the 7 identified genes (LOC108802036) did not have an
-associated annotation. For the outlier SNPs, frequency differences
-between the naive and recovering populations ranged from 41\% to 86\%.
-Most of these SNPs showed frequency differences in only a subset of the
-sampled populations (Figure~\ref{fig-allelefrequencies} A, B), but the
-SNP in the RIN3 gene showed consistent differences in frequencies across
-all populations (Figure~\ref{fig-allelefrequencies} C). This is
-suggestive of strong - and parallel - selection at this locus associated
-with exposure of frogs to Bd across multiple populations. (See SI
-Appendix for additional description of these results.)
+distinct genes across 4 contigs (Figure S1B, C;
+SI Appendix: Datasets S1, S2). One of the 7 identified genes
+(LOC108802036) did not have an associated annotation. For the outlier
+SNPs, frequency differences between the naive and recovering populations
+ranged from 41\% to 86\%. Most of these SNPs showed frequency
+differences in only a subset of the sampled populations
+(Figure~\ref{fig-allelefrequencies}A, B), but the SNP in the RIN3 gene
+showed consistent differences in frequencies across all populations
+(Figure~\ref{fig-allelefrequencies}C). This is suggestive of strong -
+and parallel - selection at this locus associated with exposure of frogs
+to Bd across multiple populations.
 
@@
 \(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
-(Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
+(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
 outliers for both metrics (``shared regions'') and 2 of these shared
-regions also contained one or more outlier SNPs (described above). A
-total of 33 annotated genes were found in the 9 shared regions (SI
-Appendix: Datasets S3, S4). The largest \(\pi_{diff}\), indicative of
+regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions (SI
+Appendix: Datasets S3, S4). Given this large number of genes, here we
+focus on those with the strongest signal of selection and/or
+immune-related functions. The largest \(\pi_{diff}\), indicative of
 directional selection, occurred in a 163kb region on Contig19, 12.9Mb
-upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C).
+upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C).
 This region contains approximately 500 SNPs and one annotated gene
@@
 outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS)
-(Figure~\ref{fig-spline-manhattan} D, Figure S2).
+(Figure~\ref{fig-spline-manhattan}D, Figure S2).
 This region (854kb from the beginning of the outlier window to the VARS
@@
 involved collection of adult frogs from 1 of 3 recovering, Bd-positive
-``donor'' populations and translocating them to 1 of 12 nearby recipient
-sites (Figure S3) \citep{seddon2014}. The donor
+``donor'' populations and translocating them \citep{seddon2014} to 1 of
+12 nearby recipient sites (Figure S3). The donor
 populations included 2 of the 5 recovering populations in the frog
-evolution study described above (Figure~\ref{fig-allelefrequencies} A:
+evolution study described above (Figure~\ref{fig-allelefrequencies}A:
 population 1 and 4). Following translocation, adult survival and
-recruitment of new adults were estimated from CMR surveys and counts of
+recruitment of new adults was estimated from CMR surveys and counts of
 tadpoles and juveniles were obtained from visual encounter surveys
@@
 nearly always well below the level indicative of severe chytridiomycosis
-and associated frog mortality (Figure S5, see
-SI Appendix text for additional details)
-\citep{joseph2018, vredenburg2010}.
+and associated frog mortality (Figure S5)
+\citep{joseph2018, vredenburg2010}. Additional details on methods used
+to estimate Bd loads are provided in the SI Appendix.
 
@@
 post-translocation survey-years (proportion of survey-years: median =
-0.79, range = 0.12--1). In summary, survey results suggest that
+0.79, range = 0.12--1). In summary, survey results indicate that
 translocations resulted in the establishment of reproducing MYL frog
@@
 survival identified several important predictors, but Bd load at the
-time of translocation was not among them (bd\_load;
-Figure S6). Important predictors included
-winter severity in the year following translocation (snow\_t1), site
-elevation (elevation), and donor population (donor\_72996;
-Figure~\ref{fig-cond-effects}). Males had somewhat higher survival than
-females, but this effect was only marginally important (sex\_male;
-Figure~\ref{fig-cond-effects}, Figure S6).
-Collectively, these results are consistent with frogs in recovering
-populations having evolved sufficient resistance/tolerance to minimize
-the adverse effects of Bd infection. In addition, the moderate Bd loads
-that characterized frogs following translocation
-(Figure S5) and relatively high survival of
-frogs in numerous translocated cohorts
+time of translocation was not among them
+(Figure S6). Important predictors included
+winter severity in the year following translocation, site elevation, and
+donor population (Figure~\ref{fig-cond-effects}). Males had somewhat
+higher survival than females, but this effect was only marginally
+important (Figure~\ref{fig-cond-effects},
+Figure S6). Collectively, these results are
+consistent with frogs in recovering populations having evolved
+sufficient resistance/tolerance to minimize the adverse effects of Bd
+infection. In addition, the moderate Bd loads that characterized frogs
+following translocation (Figure S5) and
+relatively high survival of frogs in numerous translocated cohorts
 (Figure~\ref{fig-translocation-survival}) suggests that frogs
@@
 growth rate \(\lambda\) greater than 1 in the presence of Bd
-(Figure~\ref{fig-viability} A; median predicted \(\lambda\) ranges from
+(Figure~\ref{fig-viability}A; median predicted \(\lambda\) ranges from
 1.15-1.36 for these six populations). These six populations all had
@@
 conservative estimate), long-run growth rate of five of these six
-populations was still greater than 1 (Figure~\ref{fig-viability} A).
+populations was still greater than 1 (Figure~\ref{fig-viability}A).
 
@@
 baseline value for \(\omega\) of 0.3, the six populations with the
-highest adult survival probabilities had 50-year extinction
-probabilities of 0 to 0.32 (Figure~\ref{fig-viability} B). This
-indicates strong potential for long-term persistence in the presence of
-Bd and environmental variability in recruitment. When we decreased
-\(\omega\) from 0.3 to a conservative estimate of 0.15, 50-year
-extinction probability remained less than 0.5 for three populations and
-was 0.79 to 0.94 for three additional populations
-(Figure~\ref{fig-viability} B). In contrast, for the six populations
-where yearly adult survival probability \(\sigma_{A_R} < 0.5\),
-extinction probability over 50 years was always predicted to be 1. To
-test the validity of our model predictions, we demonstrated that our
-stochastic model could describe the general recovery trajectory of our
-translocated population with the longest survey history
-(Figure~\ref{fig-viability} C; population 70550, surveyed for 16 years).
+highest adult survival probabilities (\(\sigma_{A_R} > 0.5\)) had
+50-year extinction probabilities of 0 to 0.32
+(Figure~\ref{fig-viability}B). This indicates strong potential for
+long-term persistence in the presence of Bd and environmental
+variability in recruitment. When we decreased \(\omega\) from 0.3 to a
+conservative estimate of 0.15, 50-year extinction probability remained
+less than 0.5 for three populations and was 0.79 to 0.94 for three
+additional populations (Figure~\ref{fig-viability}B). In contrast, for
+the six populations where yearly adult survival probability
+\(\sigma_{A_R} < 0.5\), extinction probability over 50 years was always
+predicted to be 1. To test the validity of our model predictions, we
+demonstrated that our stochastic model could describe the general
+recovery trajectory of our translocated population with the longest
+survey history (Figure~\ref{fig-viability}C; population 70550, surveyed
+for 16 years).
 
@@
 \citep{mendelson2019, scheele2021}, this strategy has rarely been used
-in the wild (see \citep{joseph2018} for a small-scale example). In
+in the wild \citep[see][ for a small-scale example]{joseph2018}. In
 particular, although numerous studies have documented signatures of
@@
 exposure to a recently-emerged lethal fungal disease and that some of
-the genomic regions under selection are associated with genes related to
-disease resistance/tolerance, and (2) reintroduction of individuals from
-these resistant/tolerant populations allows the successful
-reestablishment of viable frog populations despite ongoing disease.
+the genomic regions under selection contain genes related to disease
+resistance/tolerance, and (2) reintroduction of individuals from these
+resistant/tolerant populations allows the successful reestablishment of
+viable frog populations despite ongoing disease.
 
@@
 differences in Bd load of frogs from naive and recovering populations
-are also observed in controlled laboratory studies \citep[Figure
-4]{knapp2016}. The observed reduced susceptibility of frogs from
+are also observed in controlled laboratory studies \citep[see Figure 4
+in][]{knapp2016}. The observed reduced susceptibility of frogs from
 recovering populations could in theory be due to several factors,
@@
 and low \(\pi_{diff}\), that contained the complement genes C6 and C7.
-This could indicate that balancing selection is acting in this region of
-the genome to favor a diverse set of alleles, as is known for C6 in
-humans \citep{soejima2005}. Finally, based on the analysis of individual
-outlier variants, the RIN3 gene showed a consistent pattern of allele
-frequency differences across all nine of the frog populations sampled in
-this study, indicating consistent selection in populations distributed
-across a wide geographic area. This gene is associated with inflammation
-response and in \emph{Xenopus} is highly expressed in wound stumps
-following amputation of tadpole tails \citep{fukazawa2009}. Also in this
-analysis, the outlier variant with the lowest p-value (uncharacterized
-gene LOC108802036) is located adjacent to a type I interferon gene
-(Np-IFNi2) in another frog genome \citep{gan2018}, and together with
-GVINP1 further indicates the importance of interferon-related genes in
-this system.
+The complement system plays an important role in innate immunity
+\citep{rieraromo2016}, and our results could indicate that balancing
+selection is acting in this region of the genome to favor a diverse set
+of alleles, as is known for C6 in humans \citep{soejima2005}. Finally,
+based on the analysis of individual outlier variants, the RIN3 gene
+showed a consistent pattern of allele frequency differences across all
+nine of the frog populations sampled in this study, indicating
+consistent selection in populations distributed across a wide geographic
+area. This gene is associated with inflammation response and in
+\emph{Xenopus} is highly expressed in wound stumps following amputation
+of tadpole tails \citep{fukazawa2009}. Also in this analysis, the
+outlier variant with the lowest p-value (uncharacterized gene
+LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2)
+in another frog genome \citep{gan2018}, and together with GVINP1 further
+suggests the importance of interferon-related genes in this system.
 
@@
 experience high mortality. This high susceptibility of metamorphs is
-documented for numerous species of anurans, and may result from the
+documented in numerous species of anurans, and may result from the
 poorly developed immune system characteristic of this life stage
@@
 We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
-\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from four naive
-populations, and 29 from five recovering populations. These populations
-were located in the southern Sierra Nevada, from northern Yosemite
-National Park to northern Sequoia National Park
+\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
+populations, and 29 from 5 recovering populations. These populations are
+located in the southern Sierra Nevada, from northern Yosemite National
+Park to northern Sequoia National Park
 (Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
@@
 then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the
-Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
+Vincent J. Coates Genomics Sequencing Lab at UC Berkeley. Raw sequencing
+reads are available from NCBI SRA (PRJNA870451).
 
@@
 
-To visualize the genomic relationship of our populations we conducted a
+To visualize the genomic relationships of our populations we conducted a
 PCA using the glPCA function in the adegenet R package
@@
 different allele frequencies between naive versus recovering populations
-(``outliers'') using a Bonferroni- corrected significance level of 0.01.
+(``outliers'') using a Bonferroni-corrected significance level of 0.01.
 We visualized the results using a Manhattan plot and qqplot. We
-developed a more liberal set of outlier variants using a Bonferroni-
-corrected significance level of 0.05 and used this set solely for the
-gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+developed a more liberal set of outlier variants using a
+Bonferroni-corrected significance level of 0.05 and used this set solely
+for the gene ontology (GO) analysis (see below and SI Appendix: Dataset
+S1, S2).
 
@@
 
-Using the liberal set of variant outliers, we performed a statistical
+Using the liberal set of outlier variants, we performed a statistical
 overrepresentation test (Fisher's exact test) using Blast2GO
@@
 
-We conducted 1--4 translocations per site (Figure S3) and
-each translocated cohort included 18 to 99 frogs (median = 30). In
-preparation for each translocation, adult frogs were collected from the
-donor population and measured, weighed, swabbed, and PIT tagged. Frogs
-were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately
-once per month during the summer active season (median number of visits
-per summer = 2; range = 1--10) and assessed using diurnal CMR surveys
-and VES. CMR surveys allowed estimation of adult survival, adult
-population size, and recruitment of new adults, and VES provided
-estimates of tadpole and juvenile abundance. During 2006-2012, CMR
-surveys were conducted on a single day (primary period) per site visit,
-during which all habitats were searched repeatedly for adult frogs.
-Frogs were captured using handheld nets, identified via their PIT tag
-(or tagged if they were untagged), measured, weighed, swabbed, and
-released at the capture location. During 2013-2022, we generally used a
-robust design in which all habitats were searched during several
-consecutive days (median number of secondary periods per primary period
-= 3; range = 3--7), and frogs were processed as described above.
-However, when the number of frogs detected on the first survey day was
-zero or near zero, we typically conducted only a single-day CMR survey.
-When using a robust design, within a primary period, frogs that were
-captured during more than one secondary period were measured, weighed,
-and swabbed during the first capture, and during subsequent captures
-were only identified and released. Frog skin swabs were analyzed using
-standard Bd DNA extraction and qPCR methods. For details of survey and
-swab analysis methods, see \citep{joseph2018}.
+We conducted 1--4 translocations per site (Figure S3,
+Figure~\ref{fig-translocation-survival}) and each translocated cohort
+included 18 to 99 frogs (median = 30). In preparation for each
+translocation, adult frogs were collected from the donor population and
+measured, weighed, swabbed, and PIT tagged. Frogs were transported to
+the recipient site either on foot or via helicopter. Following release,
+translocated populations were visited approximately once per month
+during the summer active season (median number of visits per summer = 2;
+range = 1--10) and assessed using diurnal CMR surveys and VES. CMR
+surveys allowed estimation of adult survival, adult population size, and
+recruitment of new adults, and VES provided estimates of tadpole and
+juvenile abundance. During 2006-2012, CMR surveys were conducted on a
+single day (primary period) per site visit, during which all habitats
+were searched repeatedly for adult frogs. Frogs were captured using
+handheld nets, identified via their PIT tag (or tagged if they were
+untagged), measured, weighed, swabbed, and released at the capture
+location. During 2013-2022, we generally used a robust design in which
+all habitats were searched during several consecutive days (median
+number of secondary periods per primary period = 3; range = 3--7), and
+frogs were processed as described above. However, when the number of
+frogs detected on the first survey day was zero or near zero, we
+typically conducted only a single-day CMR survey. When using a robust
+design, within a primary period, frogs that were captured during more
+than one secondary period were measured, weighed, and swabbed during the
+first capture, and during subsequent captures were only identified and
+released.
 
@@
 For each of the translocated populations, we used the presence of
-tadpoles and/or juveniles from VES and captures of new recruits (i.e,
+tadpoles and/or juveniles from VES and counts of new recruits (i.e,
 untagged adults) in CMR surveys to provide two measures of successful
@@
 describing characteristics of frogs, sites, and translocations
-(predictor variables: Bd load (bd\_load), sex (sex\_male), frog size
-(length), site elevation (elevation), winter severity in the year of
-translocation (snow\_t), winter severity in the year following
-translocation (snow\_t1), donor population (donor\_72996), day of year
-on which a translocation was conducted (day), and translocation order
-(order\_first)). In both analyses we used 1-year survival as a single
-metric of survival that was available for all translocated populations
-regardless of how many years of post-translocation data were available
-(range = 3--16). Group-level effects included site\_id,
-translocation\_id, or translocation\_id nested within site\_id. In both
-analyses, the response variable, probability of 1-year frog survival,
-was rounded to integer values to produce a binomial outcome, and all
-models used a bernoulli family. For all models, we used default priors,
-four chains, and 5000 iterations each for warmup and post-warmup. We
-checked convergence of the MCMC algorithm using trace plots and Rhat.
+(predictor variables: Bd load, sex, frog size, site elevation, winter
+severity in the year of translocation, winter severity in the year
+following translocation, donor population, day of year on which a
+translocation was conducted, and translocation order). In both analyses
+we used 1-year survival as a single metric of survival that was
+available for all translocated populations regardless of how many years
+of post-translocation data were available (range = 3--16). Group-level
+effects included site\_id, translocation\_id, or translocation\_id
+nested within site\_id. In both analyses, the response variable,
+probability of 1-year frog survival, was rounded to integer values to
+produce a binomial outcome, and all models used a bernoulli family. For
+all models, we used default priors, four chains, and 5000 iterations
+each for warmup and post-warmup. We checked convergence of the MCMC
+algorithm using trace plots and Rhat.
 
@@
 probabilities and other known vital rates
-(Table S2), we performed four analyses (described in
-detail in the SI Appendix). First, we compute the long-run growth rate
-\(\lambda\) for each of our 12 translocated populations to determine if
-the populations were deterministically predicted to grow or decline in
-the long-run. Second, we compute the elasticity of \(\lambda\) to model
-parameters, to quantify how much changes in adult survival, for example,
-affected the long-run growth rate. Third, we included demographic and
-environmental stochasticity in our model and simulated the 50-year
-viability (i.e., 1 - extinction probability) of populations given an
-introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to
-confirm that our model could reasonably reproduce the observed recovery
-trajectories of MYL frogs following reintroductions. Additional details
-are provided in the SI Appendix. Code to replicate the analyses can be
-found at \url{https://github.com/SNARL1/translocation}.}
+(Table S2), we performed four analyses. First, we
+compute the long-run growth rate \(\lambda\) for each of our 12
+translocated populations to determine if the populations were
+deterministically predicted to grow or decline in the long-run. Second,
+we compute the elasticity of \(\lambda\) to model parameters, to
+quantify how much changes in adult survival, for example, affected the
+long-run growth rate. Third, we included demographic and environmental
+stochasticity in our model and simulated the 50-year viability (i.e., 1
+- extinction probability) of populations given an introduction of 40
+adult individuals into an unoccupied habitat. Finally, we fit our model
+to our longest translocation trajectory to confirm that our model could
+reasonably reproduce the observed recovery trajectories of MYL frogs
+following reintroductions. Additional details are provided in the SI
+Appendix. Code to replicate the analyses can be found at
+\url{https://github.com/SNARL1/translocation}.
+
+}
 
@@
 (indicated by the dashed line). The red line shows where
-\(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
-12 translocated populations, given demographic stochasticity,
-environmental variability in \(\omega\), and different mean values of
-\(\omega\). The dashed line shows where \(\omega = 0.3\), for comparison
-to (A). There are 6 lines at extinction probability = 1, 5 of which
-(70414--70619) are hidden beneath the line for 70505. (C) 100 simulated
-trajectories (black lines) from the population viability model that most
-closely matched the observed abundance trajectory of adult amphibians at
-site 70550 (light orange).}
+\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
+other points. (B) Predicted 50-year extinction probabilities of the 12
+translocated populations, given demographic stochasticity, environmental
+variability in \(\omega\), and different mean values of \(\omega\). The
+dashed line shows where \(\omega = 0.3\), for comparison to (A). There
+are 6 lines at extinction probability = 1, 5 of which (70414--70619) are
+hidden beneath the line for 70505. (C) 100 simulated trajectories (black
+lines) from the population viability model that most closely matched the
+observed abundance trajectory of adult amphibians at site 70550 (light
+orange).}
 

### 2023-05-19 doc/manuscript/translocation_bioarxiv_SI.tex
Address formatting comments on compiled manuscript

--- 2023-05-19 doc/manuscript/translocation_pnas_SI.tex
@@
 In the Results section of the main paper, we describe the stringent set
-of outlier SNPS (identified using a Bonferroni-corrected p-value of
+of outlier SNPs (identified using a Bonferroni-corrected p-value of
 0.01). The liberal set, identified using a Bonferroni-corrected p-value
@@
 recovering populations, we detected consistent allele frequency changes
-across all genetic clusters for 1 of 7 of the associated genes (RIN3;
-Fig. 1). However, for the other 6 genes we
-still found a statistically significant signal of selection in 2 of the
-3 genetic clusters (containing populations 6--9;
+across all genetic clusters for only 1 of 7 of the associated genes
+(RIN3; Fig. 1). However, for the other 6
+genes we still found a statistically significant signal of selection in
+2 of the 3 genetic clusters (containing populations 6--9;
 Fig. 1, Figure~\ref{fig-selectionresults}
@@
 
-Swab extracts were analyzed singly instead of in triplicate
-\citep{kriger2006}. For analysis of swabs collected during 2005--2014,
-we used standards developed from known concentrations of zoospores
-\citep{boyle2004}, and after 2014, we used standards based on single
-ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
+Swab extracts were analyzed using standard Bd DNA extraction and qPCR
+methods \citep{boyle2004}, and extracts were analyzed singly instead of
+in triplicate \citep{kriger2006}. For analysis of swabs collected during
+2005--2014, we used standards developed from known concentrations of
+zoospores \citep{boyle2004}, and after 2014, we used standards based on
+single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
 between samples analyzed using both types of standards, Bd in the study
@@
 the CMR model used to estimate population-level parameters of
-translocated populations (see ``Estimation of adult survival and
-population size'' above). Although we observed yearly variability in
-adult survival within a population, the magnitude of this variability
-was small compared to among population variability
+translocated populations (see ``Materials and Methods - Estimation of
+frog survival and abundance'' in main paper). Although we observed
+yearly variability in adult survival within a population, the magnitude
+of this variability was small compared to among-population variability
 (Fig. 3). Thus, we did not include
@@
 are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
-and \(\sigma_{A_R}\) were extracted directly from our CMR model
-described previously in this study. For populations where we had a
-sufficient number of PIT-tagged, naturally-recruited adults, we observed
-that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different,
-with \(\sigma_{A_R} > \sigma_{A_T}\)
-(Figure~\ref{fig-compare_surv_probs}). For populations lacking
-sufficient numbers of naturally-recruited adults, we were unable to
-directly estimate \(\sigma_{A_R}\), and instead set
-\(\sigma_{A_R} = \sigma_{A_T}\).
+and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
+``Materials and Methods - Estimation of frog survival and abundance'' in
+main paper). For populations where we had a sufficient number of
+PIT-tagged, naturally-recruited adults, we observed that
+\(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different, with
+\(\sigma_{A_R} > \sigma_{A_T}\) (Figure~\ref{fig-compare_surv_probs}).
+For populations lacking sufficient numbers of naturally-recruited
+adults, we were unable to directly estimate \(\sigma_{A_R}\), and
+instead set \(\sigma_{A_R} = \sigma_{A_T}\).
 
@@
 whether our model could qualitatively describe the patterns of recovery
-in the observed data for population 70550. Code to replicate the
-analyses can be found at \url{https://github.com/SNARL1/translocation}.
+in the observed data for population 70550.
 
@@
 Numeric labels and colors match those in
-Fig. 1 A. Populations 1-7 are \emph{R.
+Fig. 1A. Populations 1-7 are \emph{R.
 sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
@@
 
-{\centering \includegraphics[width=0.60\textwidth]{figures/bdload_beforeafter.png}
+{\centering \includegraphics[width=0.7\textwidth]{figures/bdload_beforeafter.png}
 
@@
 translocation. Bd loads are expressed as the number of ITS1 copies per
-skin swab, as estimated by qPCR of the Bd ITS1 region. Loads indicative
-of severe disease are \textgreater{} 5.8 ITS copies (on a
-log\textsubscript{10} scale; see SI Appendix text for details). Box plot
-details are as in Fig. 4.}
+skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show
+medians, first and third quartiles, largest and smallest values within
+1.5x interquartile range, and values outside the 1.5x interquartile
+range. Loads indicative of severe disease are \textgreater{} 5.8 ITS
+copies (on a log\textsubscript{10} scale).}
 
@@
 considered important predictors if the associated uncertainty interval
-does not overlap zero (indicated by the dashed red line).}
+does not overlap zero (indicated by the dashed red line). Predictor
+variables shown on the y-axis are defined as follows: length = frog
+size, snow\_t = winter severity in the year of translocation (measured
+on April 1), snow\_t1 = winter severity in the year following
+translocation (measured on April 1), day = day of year on which a
+translocation was conducted, bd\_load = Bd load, elevation = site
+elevation, order\_first = within-site translocation order, sex\_male =
+frog sex, donor\_70567 and donor\_72996 = donor population.}
 
@@
 uncertainty intervals around the estimates, accounting for yearly
-variation in survival probabilities. The large error bars for 70550
-reflect the large within and among individual variability in yearly
-survival probability at this site (as also reflected in one-year
-survival probabilities in Fig. 3). All
-estimates are derived from the mrmr package.}
+variation in survival probabilities. All estimates were derived using
+the mrmr package.}
 

### 2023-05-19 doc/manuscript/translocation_pnas.tex
Address formatting comments on compiled manuscript

@@
 dynamics are given by equation \ref{eqn:matrix}.
+
+The parameters in this model are yearly survival probability
+\(\sigma_{\cdot}\) (the subscript ``\(\cdot\)'' indicates a particular
+state variable), probability that a female frog reproduces in a given
+year \(p_F\), number of eggs produced by a female frog in a year that
+successfully hatch \(F\), probability of a first-year tadpole remaining
+as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
+as a tadpole \(p_{L2}\), and probability of a first-year juvenile
+remaining as a juvenile \(p_{J1}\), and the probability of a juvenile
+successfully recruiting to an adult \(\omega\).
+
+In this model we ignore density-dependent recruitment because we were
+interested in the growth of the population from an initial
+reintroduction and whether this growth was sufficient to prevent
+extinction over 50 years following the introduction. We also did not
+directly consider the dynamics of Bd in this model. We made this
+decision because (i) translocated populations are infected with Bd at
+high prevalence \citep{joseph2018}, and (ii) host density does not seem
+to play a significant role in multi-year Bd infection dynamics in this
+system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
+instead assuming all host vital rates are in the presence of high Bd
+prevalence significantly simplifies the model without much loss of
+realism. Additional details are provided in the SI Appendix.
+
+\hypertarget{model-analysis}{%
+\subsection*{Model analysis}\label{model-analysis}}
+
+After parameterizing our model with CMR-estimated adult frog survival
+probabilities and other known vital rates
+(Table S2), we performed four analyses. First, we
+compute the long-run growth rate \(\lambda\) for each of our 12
+translocated populations to determine if the populations were
+deterministically predicted to grow or decline in the long-run. Second,
+we compute the elasticity of \(\lambda\) to model parameters, to
+quantify how much changes in adult survival, for example, affected the
+long-run growth rate. Third, we included demographic and environmental
+stochasticity in our model and simulated the 50-year viability (i.e., 1
+- extinction probability) of populations given an introduction of 40
+adult individuals into an unoccupied habitat. Finally, we fit our model
+to our longest translocation trajectory to confirm that our model could
+reasonably reproduce the observed recovery trajectories of MYL frogs
+following reintroductions. Additional details are provided in the SI
+Appendix. Code to replicate the analyses can be found at
+\url{https://github.com/SNARL1/translocation}.
+
+}
+
+\showmatmethods{} % Display the Materials and Methods section
+
+\acknow{We thank the following for important contributions to this study: E.
+Hegeman and A. Lindauer--project and data management, field and
+laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
+and numerous summer technicians--field work; staff at Sequoia-Kings
+Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
+California Department of Fish and Wildlife, U.S. Fish and Wildlife
+Service, University of California-Santa Barbara Institutional Animal
+Care and Use Committee, and Sierra Nevada Aquatic Research
+Laboratory--research permits, logistical support, and/or field
+assistance. This project was supported by grants from the National Park
+Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
+Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
+
+\showacknow{} % Display the acknowledgments section
+
+% Bibliography
+\bibliography{translocation}
+
+\clearpage
 
@@
 
-The parameters in this model are yearly survival probability
-\(\sigma_{\cdot}\) (the subscript ``\(\cdot\)'' indicates a particular
-state variable), probability that a female frog reproduces in a given
-year \(p_F\), number of eggs produced by a female frog in a year that
-successfully hatch \(F\), probability of a first-year tadpole remaining
-as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
-as a tadpole \(p_{L2}\), and probability of a first-year juvenile
-remaining as a juvenile \(p_{J1}\), and the probability of a juvenile
-successfully recruiting to an adult \(\omega\).
-
-In this model we ignore density-dependent recruitment because we were
-interested in the growth of the population from an initial
-reintroduction and whether this growth was sufficient to prevent
-extinction over 50 years following the introduction. We also did not
-directly consider the dynamics of Bd in this model. We made this
-decision because (i) translocated populations are infected with Bd at
-high prevalence \citep{joseph2018}, and (ii) host density does not seem
-to play a significant role in multi-year Bd infection dynamics in this
-system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
-instead assuming all host vital rates are in the presence of high Bd
-prevalence significantly simplifies the model without much loss of
-realism. Additional details are provided in the SI Appendix.
-
-\hypertarget{model-analysis}{%
-\subsection*{Model analysis}\label{model-analysis}}
-
-After parameterizing our model with CMR-estimated adult frog survival
-probabilities and other known vital rates
-(Table S2), we performed four analyses (described in
-detail in the SI Appendix). First, we compute the long-run growth rate
-\(\lambda\) for each of our 12 translocated populations to determine if
-the populations were deterministically predicted to grow or decline in
-the long-run. Second, we compute the elasticity of \(\lambda\) to model
-parameters, to quantify how much changes in adult survival, for example,
-affected the long-run growth rate. Third, we included demographic and
-environmental stochasticity in our model and simulated the 50-year
-viability (i.e., 1 - extinction probability) of populations given an
-introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to
-confirm that our model could reasonably reproduce the observed recovery
-trajectories of MYL frogs following reintroductions. Additional details
-are provided in the SI Appendix. Code to replicate the analyses can be
-found at \url{https://github.com/SNARL1/translocation}.}
-
-\showmatmethods{} % Display the Materials and Methods section
-
-\acknow{We thank the following for important contributions to this study: E.
-Hegeman and A. Lindauer--project and data management, field and
-laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
-and numerous summer technicians--field work; staff at Sequoia-Kings
-Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
-California Department of Fish and Wildlife, U.S. Fish and Wildlife
-Service, University of California-Santa Barbara Institutional Animal
-Care and Use Committee, and Sierra Nevada Aquatic Research
-Laboratory--research permits, logistical support, and/or field
-assistance. This project was supported by grants from the National Park
-Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
-Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
-DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
-
-\showacknow{} % Display the acknowledgments section
-
-% Bibliography
-\bibliography{translocation}
-
 \clearpage

### 2023-05-19 code/extinction_and_growth_rate_analysis.R

@@ -92,7 +92,7 @@ p1 = ggplot(data=lake_growth_rates_dt) + geom_point(aes(x=lake_id, y=med, color=
 		   						    scale_color_manual(values=colors) +
 		   						    geom_hline(aes(yintercept=1), linetype="dashed") +
 		   						    theme_classic() + ylab("Long-run growth rate, lambda") +
-		   						    xlab("Lake ID")
+		   						    xlab("Site ID")
 
 ggsave(file.path("..", "out", "lambda_estimates.pdf"), width=8, height=5)
 
@@ -349,7 +349,7 @@ pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=omega, y=e
 							geom_point(aes(color=lake_id_surv)) + 
 							scale_color_manual(values=cdat$col) +
 							theme_classic() + xlab("Probability of recruitment (\u03c9)") + ylab("Extinction prob. in 50 years") +
-							guides(color=guide_legend(title=bquote("Lake, \u03c3"~.[AR]), title.position="top", ncol=3)) + 
+							guides(color=guide_legend(title=bquote("Site, \u03c3"~.[AR]), title.position="top", ncol=3)) + 
 							theme(legend.position="bottom", legend.text=element_text(size=7), legend.title=element_text(size=8))
 
 # Examine simulations
@@ -483,8 +483,8 @@ ptraj = ggplot() + geom_line(data=pred_traj, aes(x=year, y=abund, group=sim, col
 #########################################
 
 myplot = (ptile + pext + ptraj) + plot_annotation(tag_levels="A", tag_suffix="")
-ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.jpg"), width=11, height=5)
+ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.jpg"), width=11, height=5, dpi=300)
 
 myplot = (p2) + plot_annotation(tag_levels="A", tag_suffix="")
-ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=5, height=4)
+ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=5, height=4, dpi=300)
 

### 2023-05-19 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -127,7 +127,7 @@ methods = methods.replace("Incorporating yearly variability in vital rates", "\\
 methods = methods.replace("Estimating model parameters", "\\subsubsection*{Estimating model parameters}")
 methods = methods.replace("Model analysis and simulation", "\\subsubsection*{Model analysis and simulation}")
 
-pnas_ms = pnas_ms.replace("\\matmethods{}", "\n\n\\matmethods{\n" + methods + "}\n")
+pnas_ms = pnas_ms.replace("\\matmethods{}", "\n\n\\matmethods{\n" + methods + "\n\n}\n")
 
 ## Acknowledgements ##
 
@@ -161,10 +161,10 @@ pnas_ms = pnas_ms.replace("The dynamics are given by", "The dynamics are given b
 
 supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
 			"Figure~\\ref{fig-selectionresults} A": "Figure S1A",
-			"Figure~\\ref{fig-selectionresults} B, C: SI": "Figure S1B,C",
-			"Figure~\\ref{fig-selectionresults} B, C": "Figure S1B,C",
-			"Figure~\\ref{fig-selectionresults} B,\nC": "Figure S1B,C",
-			"Figure~\\ref{fig-selectionresults} B, C;\nSI": "Figure S1B,C",
+			"Figure~\\ref{fig-selectionresults} B, C: SI": "Figure S1B, C",
+			"Figure~\\ref{fig-selectionresults} B, C": "Figure S1B, C",
+			"Figure~\\ref{fig-selectionresults} B,\nC": "Figure S1B, C",
+			"Figure~\\ref{fig-selectionresults} B, C;\nSI": "Figure S1B, C",
 			"Figure~\\ref{fig-selectionresults} A": "Figure S1A",
 			"Figure~\\ref{fig-synteny-plot} SI": "Figure S2",
 			"Figure~\\ref{fig-synteny-plot}": "Figure S2",
@@ -194,13 +194,24 @@ supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S1A",
 for key, value in supp_map.items():
 	pnas_ms = pnas_ms.replace(key, value)
 
+# Adjust figure buffering
+letters = ["A", "B", "C", "D"]
+
+for letter in letters:
+	pnas_ms = re.sub("} " + letter, 
+					  "}" + letter, 
+					 pnas_ms)
+
[…]

### 2023-05-20 doc/manuscript/translocation.qmd
Minor edits to Results

@@
 To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
 These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
@@
 Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
-Additional details on methods used to estimate Bd loads are provided in the SI Appendix.
 

### 2023-05-20 doc/manuscript/translocation_bioarxiv.tex
Recompile PNAS version with new changes

@@
 histories of Bd exposure. Specifically, we compared frog genomes sampled
-in four populations that have not yet experienced a Bd-caused epizootic
-(``naive'') \citep{zhou2015} versus in five populations that experienced
-a Bd epizootic during the past several decades and have since recovered
-to varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
+in 4 populations that have not yet experienced a Bd-caused epizootic
+(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
+Bd epizootic during the past several decades and have since recovered to
+varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
 \citep{knapp2016, vredenburg2010}. These recovering populations exist in
@@
 and associated frog mortality (Figure S5)
-\citep{joseph2018, vredenburg2010}. Additional details on methods used
-to estimate Bd loads are provided in the SI Appendix.
+\citep{joseph2018, vredenburg2010}.
 

### 2023-05-20 doc/manuscript/translocation.qmd
Delete comments and fix typo

@@
 ---
-
-<!--# render to pdf using `quarto render doc/manuscript/translocation.qmd --to pdf` -->
 
@@
 
-<!--# Limit = 250 words. -->
-
 Vast alteration of the biosphere by humans is causing a sixth mass extinction.
@@
 ### Significance
-
-<!--# Limit = 120 words.   -->
 
@@
 Results from our genomic analyses strongly suggest that natural selection for adaptive alleles and resulting increased resistance/tolerance is at least partially responsible for the reduced susceptibility of frogs in recovering populations.
-We identified specific alleles and genomic regions showing signatures of selection between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
+We identified specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
 Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
@@
 
-\
-
-<!--# How to insert >= symbol in header and table cells? Move second sentence of caption to footnote.-->
-
 ```{r}

### 2023-11-18 doc/manuscript/translocation.qmd
Rearranged sections and revised last paragraph of Intro

--- 2023-05-20 doc/manuscript/translocation.qmd
@@
 ---
+
+<!--# Drop "evolutionary rescue" from title? -->
 
@@
 
+<!--# Revise to reflect reordering of sections -->
+
 Vast alteration of the biosphere by humans is causing a sixth mass extinction.
@@
 Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
-However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions; We do not distinguish between resistance and tolerance in this paper because although they have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, frog population reestablishment].
+<!--# The previous statement re: resistance/tolerance breaks the flow of the paragraph, and makes the next sentence feel clumsy.
+-->However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
 In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
@@
 
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
-These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
+<!--# I've reworded this paragraph somewhat to reflect the revised order of sections and to better highlight the key outcomes of recovery, persistence, and evolution.
+Consider rewording to provide explicit objectives of each of the sections (plus summary of results).
+E.g., for genomics section objective was to provide insights into the evolutionary response by MYL frogs to Bd infection across their range in the Sierra Nevada, and thereby provide a mechanism by which frogs reintroduced into sites in Yosemite from which they were extirpated by Bd could recover (instead of being driving to extirpation again).-->
+
+In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
+Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
+We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
+Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
+These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.
 
 ### Results
-
-#### Frog evolution in response to Bd
-
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
-(See SI Appendix for additional details regarding the study design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
-One of the 7 identified genes (LOC108802036) did not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
-This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 
+#### Frog evolution in response to Bd
+
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
+Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
+(See SI Appendix for additional details regarding the study design.) Bd-exposure histories of the nine study populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+muscosa* versus *R.
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
+One of the 7 identified genes (LOC108802036) did not have an associated annotation.
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+
+The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
+The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
+Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
+This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
+
 ### Discussion
@@
 ### Materials and Methods
-
-#### Frog evolution in response to Bd
-
-##### Sampling and Sequencing
-
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
-These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
-Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
-DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
-Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
-Raw sequencing reads are available from NCBI SRA (PRJNA870451).
-
-##### Data Pre-processing and Cleaning
-
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
-Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
-This initial filter resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant level.
-First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x.
-Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-##### Data Analysis
-
-To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
-We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
-After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
-
-Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
 
@@
 
+#### Frog evolution in response to Bd
+
+##### Sampling and Sequencing
+
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
+These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
+Samples were collected from 5-6 frogs per population.
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
+DNA was extracted following Qiagen DNEasy manufacturer's protocols.
+We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
+Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
+Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
+Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
+Raw sequencing reads are available from NCBI SRA (PRJNA870451).
+
+##### Data Pre-processing and Cleaning
+
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Exact PCR duplicates were marked using Picard.
+Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
+Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
+Next, genotypes were jointly called using GenotypeGVCFs.
+Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
+This initial filter resulted in 1,595,206 variant sites across 53 individuals.
+
+We then further filtered our dataset at the individual and variant level.
+First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
+This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
+Finally, we trimmed samples with an average depth across filtered sites \< 7x.
+Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+
+##### Data Analysis
+
+To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
+To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
+GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
+We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
+We visualized the results using a Manhattan plot and qqplot.
+We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+
+In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
+Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
+We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
+After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
+
+In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
+First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
+This method also calculates a W-statistic allowing for outlier identification.
+We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
+Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
+
+Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
+
 ### Acknowledgments
@@
 ### Figures
-
-![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
-For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
-The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
-
-\newpage
-
-![Evidence for selection on genomic regions in recovering MYL frog populations.
-Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
-In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
-In (B), outlier regions are shown above the single dashed red line.
-Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
-
-\newpage
 
@@
 
+![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
+The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+
+\newpage
+
+![Evidence for selection on genomic regions in recovering MYL frog populations.
+Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
+In (B), outlier regions are shown above the single dashed red line.
+Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
+(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+
+\newpage
+
 ## Supporting Information
-
-### Frog evolution in response to Bd
-
-#### Methods
-
-We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-
-#### Results
-
-In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 
@@
 \newpage
+
+### Frog evolution in response to Bd
+
+#### Methods
+
+We compared frog genomes sampled in naive versus recovering populations.
+Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
+The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
+As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+
+#### Results
+
+In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
+The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
+Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
+
+In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:--------------------------|:------------------|:-------------------------|
+|:---------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-11-21 code/maps.qmd
Add site identifiers to Yosemite map

@@
 ```{r}
+translocation_points <- translocation_points %>% 
+  mutate(
+    label = case_when(
+      site_id == 70567 ~ 4, 
+      site_id == 72996 ~ 1
+    ))
+
 translocation_points_map <- tm_shape(hillshade_yose_crop) + 
@@
   tm_shape(translocation_points) + 
-  tm_symbols(size = 0.9, 
+  tm_symbols(size = 3, 
              shape = "donor_id",
@@
             legend.text.size = 1.2) +
+  tm_text(text = "label") +
   tm_compass(type = "4star",

### 2023-11-21 doc/manuscript/translocation.qmd
Add site identifiers to Yosemite map

@@
 csl: pnas.csl
+format: pdf
 # format: docx
-format :
-  latex:
-    cite-method: natbib
+# format :
+#  latex:
+#    cite-method: natbib
 editor: visual
@@
 ![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by black polygon).
-Symbol shapes indicate the donor population used for each translocation site.
+Symbol shapes indicate the donor population used for each translocation site, and numbers inside the donor symbols refer to sites identified in @fig-allelefrequencies.
 To obscure the exact locations of populations, random noise was added to all point coordinates.
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="250"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}
 

### 2023-11-21 doc/manuscript/translocation.qmd
Describe objectives and edit translocation results

@@
 
-<!--# I've reworded this paragraph somewhat to reflect the revised order of sections and to better highlight the key outcomes of recovery, persistence, and evolution.
-Consider rewording to provide explicit objectives of each of the sections (plus summary of results).
-E.g., for genomics section objective was to provide insights into the evolutionary response by MYL frogs to Bd infection across their range in the Sierra Nevada, and thereby provide a mechanism by which frogs reintroduced into sites in Yosemite from which they were extirpated by Bd could recover (instead of being driving to extirpation again).-->
+<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->
 
-In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
+In the current study, we had three primary objectives.
+First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
+Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+
+<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
 Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
@@
 These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
-Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.
+Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
 
@@
 
+<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival.
+-->Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
+For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
+These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
+As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
+In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
+
 Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
@@
 Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
-
-Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
-For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
-These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
-As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
-In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
 
@@
 
-Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
+Results from frog translocations indicated that most populations showed evidence of successful reproduction and recruitment and that adult survival was often relatively high (described above).
 Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
@@
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
+<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->
 

### 2023-11-21 doc/manuscript/translocation.qmd
WIP detailed model specification

--- 2023-11-18 doc/manuscript/translocation.qmd
@@
 format :
-  latex:
+  # latex:
+  pdf:
     cite-method: natbib
+    include-in-header:
+      text: |
+        \usepackage{blkarray}
 editor: visual
@@
 ---
-
-<!--# Drop "evolutionary rescue" from title? -->
 
@@
 
-<!--# Revise to reflect reordering of sections -->
-
 Vast alteration of the biosphere by humans is causing a sixth mass extinction.
@@
 Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions; We do not distinguish between resistance and tolerance in this paper because although they have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, frog population reestablishment].
-<!--# The previous statement re: resistance/tolerance breaks the flow of the paragraph, and makes the next sentence feel clumsy.
--->However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
+However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
 In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
@@
 
-<!--# I've reworded this paragraph somewhat to reflect the revised order of sections and to better highlight the key outcomes of recovery, persistence, and evolution.
-Consider rewording to provide explicit objectives of each of the sections (plus summary of results).
-E.g., for genomics section objective was to provide insights into the evolutionary response by MYL frogs to Bd infection across their range in the Sierra Nevada, and thereby provide a mechanism by which frogs reintroduced into sites in Yosemite from which they were extirpated by Bd could recover (instead of being driving to extirpation again).-->
-
-In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
-Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
-We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
-Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
-These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
-Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.
+In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
+Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
+We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
+These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
+Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
 
 ### Results
+
+#### Frog evolution in response to Bd
+
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
+Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
+(See SI Appendix for additional details regarding the study design.)
+
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+muscosa* versus *R.
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
+One of the 7 identified genes (LOC108802036) did not have an associated annotation.
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+
+The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
+The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
+Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
+This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 
-#### Frog evolution in response to Bd
-
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
-(See SI Appendix for additional details regarding the study design.) Bd-exposure histories of the nine study populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
-One of the 7 identified genes (LOC108802036) did not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
-This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-
 ### Discussion
@@
 ### Materials and Methods
+
+#### Frog evolution in response to Bd
+
+##### Sampling and Sequencing
+
+We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
+These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
+Samples were collected from 5-6 frogs per population.
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
+DNA was extracted following Qiagen DNEasy manufacturer's protocols.
+We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
+Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
+Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
+Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
+Raw sequencing reads are available from NCBI SRA (PRJNA870451).
+
+##### Data Pre-processing and Cleaning
+
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Exact PCR duplicates were marked using Picard.
+Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
+Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
+Next, genotypes were jointly called using GenotypeGVCFs.
+Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
+This initial filter resulted in 1,595,206 variant sites across 53 individuals.
+
+We then further filtered our dataset at the individual and variant level.
+First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
+This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
+Finally, we trimmed samples with an average depth across filtered sites \< 7x.
+Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+
+##### Data Analysis
+
+To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
+To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
+GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
+We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
+We visualized the results using a Manhattan plot and qqplot.
+We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+
+In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
+To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
+Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
+We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
+After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
+
+In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
+First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
+Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
+This method also calculates a W-statistic allowing for outlier identification.
+We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
+Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
+
+Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
 
@@
 
-#### Frog evolution in response to Bd
-
-##### Sampling and Sequencing
-
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
-These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
-Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
-DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
-Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
-Raw sequencing reads are available from NCBI SRA (PRJNA870451).
-
-##### Data Pre-processing and Cleaning
-
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
-Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
-This initial filter resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant level.
-First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x.
-Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-##### Data Analysis
-
-To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
-We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
-After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
-
-Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-
 ### Acknowledgments
@@
 ### Figures
+
+![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
+The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+
+\newpage
+
+![Evidence for selection on genomic regions in recovering MYL frog populations.
+Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
+In (B), outlier regions are shown above the single dashed red line.
+Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
+(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+
+\newpage
 
@@
 
-![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
-For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
-The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+## Supporting Information
 
-\newpage
+### Frog evolution in response to Bd
 
-![Evidence for selection on genomic regions in recovering MYL frog populations.
-Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
-In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
-In (B), outlier regions are shown above the single dashed red line.
-Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+#### Methods
 
-\newpage
+We compared frog genomes sampled in naive versus recovering populations.
+Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
+Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
+The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
+As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
 
-## Supporting Information
+#### Results
+
+In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
+The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
+Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
+
+In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 
@@
 
-#### Methods
+#### Swab Methods
 
@@
 In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
+
+#### Mark-recapture model structure
+
+We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
+For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
+Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
+Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
+
+We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
+The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
+Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
+Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
+We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
+Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
+We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
+This assumption is justified by the short time intervals between surveys within primary periods.
+
+Live individuals are detected with probability $p_j$, which is modeled as:
+
+$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
+
+where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
+Not recruited and dead individuals are never captured. 
+We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
+
+$$
+\Omega_j =
+\begin{blockarray}{ccc}
+  \text{Not detected} & \text{Detected} \\
+\begin{block}{(cc)c}
+  1 & 0 & \text{Not recruited} \\
+  1 - p_j & p_j & \text{Alive} \\
+  1 & 0 & \text{Dead} \\
+\end{block}
+\end{blockarray}
+$$
+
+TODO:
+
+- State model description
+- State transition matrix for natural recruits
+- State transition matrix for introduced individuals
+- Parameter model description
+- Priors
+- Specification of the joint distribution (possible diagram of the DAG)
 
@@
 \newpage
-
-### Frog evolution in response to Bd
-
-#### Methods
-
-We compared frog genomes sampled in naive versus recovering populations.
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-
-#### Results
-
-In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:---------------------------|:------------------|:-------------------------|
+|:-----------------------------|:---------------|:--------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-11-22 doc/manuscript/translocation.qmd
Try analysis with omega on juvenile y1 survival probability

@@
 format :
-  # latex:
-  pdf:
+  latex:
     cite-method: natbib
-    include-in-header:
-      text: |
-        \usepackage{blkarray}
 editor: visual
@@
 
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.XX-1.XX for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
+As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
 
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
-When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in juvenile survival and recruitment (for year 1 and year 2 juveniles, the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average percent reduction in juvenile survival was less than 50% (given baseline year 1 juvenile survival probability of 0.25; @fig-viability B).
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
+When we decreased baseline year 1 juvenile survival to 0.09, 50-year extinction probability was of less than 50% for four of these populations (X-X) and greater than 50% (X-X) for two of these populations (@fig-viability B).
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of baseline year 1 survival probability.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
   \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
+  0 & 0 & 0 & \sigma_{J1} p_{J1}(1 - \omega) & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) (1 - \omega) & \sigma_{J2} (1 - \omega) & \sigma_{A_R} & 0 \\
   0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$.
+The parameter $(1 - \omega)$ gives the percent reduction in baseline juvenile survival and recruitment due environmental stochasticity.
 
@@
 
-#### Swab Methods
+#### Methods
 
@@
 In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
-
-#### Mark-recapture model structure
-
-We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
-For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
-Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
-Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
-
-We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
-The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
-Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
-Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
-We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
-Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
-We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
-This assumption is justified by the short time intervals between surveys within primary periods.
-
-Live individuals are detected with probability $p_j$, which is modeled as:
-
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
-
-where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
-Not recruited and dead individuals are never captured. 
-We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
-
-$$
-\Omega_j =
-\begin{blockarray}{ccc}
-  \text{Not detected} & \text{Detected} \\
-\begin{block}{(cc)c}
-  1 & 0 & \text{Not recruited} \\
-  1 - p_j & p_j & \text{Alive} \\
-  1 & 0 & \text{Dead} \\
-\end{block}
-\end{blockarray}
-$$
-
-TODO:
-
-- State model description
-- State transition matrix for natural recruits
-- State transition matrix for introduced individuals
-- Parameter model description
-- Priors
-- Specification of the joint distribution (possible diagram of the DAG)
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:-----------------------------|:---------------|:--------------------------|
+|:----------------------------|:-----------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-11-22 code/extinction_and_growth_rate_analysis.R

@@ -2,6 +2,7 @@ library(data.table)
 source("population_viability_functions.R")
 library(ggplot2)
 library(patchwork)
+library(parallel)
 
 #### Data loading and preliminary analysis ######
 
@@ -102,35 +103,35 @@ ggsave(file.path("..", "out", "lambda_estimates.pdf"), width=8, height=5)
 
 
 muR_range = seq(0.01, 0.9, len=50)
-omega_range = seq(0.01, .4, len=50)
-lambdas = array(NA, dim=c(length(muR_range), length(omega_range)))
+juv_surv_range = seq(0.01, .25, len=50)
+lambdas = array(NA, dim=c(length(muR_range), length(juv_surv_range)))
 
 for(i in 1:length(muR_range)){
-	for(j in 1:length(omega_range)){
+	for(j in 1:length(juv_surv_range)){
 
 		muR = muR_range[i]
-		omega = omega_range[j]
-		params = assign_parameters(muR, muR, new_params=list(omega=omega))
+		juv_surv = juv_surv_range[j]
+		params = assign_parameters(muR, muR, new_params=list(sigma_J1=juv_surv))
 		lambdas[i, j] = get_lambda(params)
 
 	}
 }
 
-dat = expand.grid(muR=muR_range, omega=omega_range)
+dat = expand.grid(muR=muR_range, juv_surv=juv_surv_range)
 dat$lambda = as.vector(lambdas)
-ptile = ggplot(dat) + geom_tile(aes(x=muR, y=omega, fill=lambda)) + 
-							geom_hline(aes(yintercept=0.3), linetype="dashed") +
-							geom_contour(aes(x=muR, y=omega, z=lambda), breaks=c(1), color="red", size=1) +
+ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) + 
+							geom_hline(aes(yintercept=0.08), linetype="dashed") +
+							geom_contour(aes(x=muR, y=juv_surv, z=lambda), breaks=c(1), color="red", size=1) +
 							geom_point(data=lake_growth_rates_dt, aes(x=surv_med, 
-																												y=rep(0.3, length(surv_med)), 
+																												y=rep(0.08, length(surv_med)), 
 																													color=lake_id), size=2) +
 		   				scale_color_manual(values=colors) +
 							# scale_fill_gradient("\u03bb", low = "white", high = "black") +
 							scale_fill_gradient2(midpoint=1) + #guide=guide_legend(title.position="bottom")) +
[…]

### 2023-11-22 code/population_viability_functions.R

@@ -25,12 +25,12 @@ assign_parameters = function(adult_survR, adult_survT, new_params=list()){
 			  sigma_L2=0.7, 
 			  sigma_L3=0.7, 
 			  sigma_J1=0.25, 
-			  sigma_J2=0.5, 
+			  sigma_J2=0.5,
 			  sigma_AR=adult_survR,
 			  sigma_AT=adult_survT,
 			  p_L1=1, p_L2=0.25, p_J1=0.25,
 			  p_F=0.5, F=100, 
-			  omega=0.3,
+			  omega=1,
 			  reproduction="poisson")
 
 	# Set any new parameters
@@ -57,7 +57,7 @@ build_transition_matrix = function(p){
 	L1 = c(0, p$sigma_L1*p$p_L1, 0, p$sigma_L1*(1 - p$p_L1), 0, 0, 0)
 	L2 = c(0, 0, p$sigma_L2*p$p_L2, p$sigma_L2*(1 - p$p_L2), 0, 0, 0)
 	L3 = c(0, 0, 0, p$sigma_L3, 0, 0, 0)
-	J1 = c(0, 0, 0, 0, p$sigma_J1*p$p_J1, p$sigma_J1*(1 - p$p_J1)*p$omega, 0)
+	J1 = c(0, 0, 0, 0, p$sigma_J1*p$p_J1*p$omega, p$sigma_J1*(1 - p$p_J1)*p$omega, 0)
 	J2 = c(0, 0, 0, 0, 0, p$sigma_J2*p$omega, 0)
 	AR = c(0, 0, 0, 0, 0, p$sigma_AR, 0)
 	AT = c(0, 0, 0, 0, 0, 0, p$sigma_AT)
@@ -166,7 +166,6 @@ stochastic_step = function(state_vars, p){
 
 	# How many are reproducing?
 	n = length(state_vars)
-
 
 	# Account for overflow
 	if(state_vars[(n - 1)] < 10000){
@@ -314,7 +313,7 @@ build_projection_matrix = function(params){
 
 # We also need the derivatives with respect to parameters
 get_sensitivity = function(params){
-	# Calculates sensitivites of lambda to sigma_AR, omega, F
[…]

### 2023-11-28 doc/manuscript/translocation.qmd
Finish CMR spec

--- 2023-11-21 doc/manuscript/translocation.qmd
@@
 
-TODO:
+The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$. 
+For unintroduced individuals which naturally recruit, this matrix is given by:
 
-- State model description
-- State transition matrix for natural recruits
-- State transition matrix for introduced individuals
-- Parameter model description
-- Priors
-- Specification of the joint distribution (possible diagram of the DAG)
+$$
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+$$
+where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
+
+For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:
+
+$$
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+$$
+where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
+
+We allow recruitment probabilities to vary in time via random effects, such that:
+
+$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
+
+where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.
+
+Survival probabilities also vary in time, and as a function of known covariates:
+
+$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
+
+where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.
+
+To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters. 
+The recruitment parameter priors were specified as follows:
+
+$$\alpha^{(\lambda)} \sim N(0,1),$$
+$$\sigma^{(\lambda)} \sim N_+(0,1),$$
+$$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$
+for periods $t=1, ..., T$. 
+Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
+
+Survival parameter priors were specified similarly as:
+
+$$\beta^{(\phi)} \sim N(0, 1),$$
+$$\sigma^{(\phi)} \sim N_+(0, 1),$$
+$$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$
+for time $t=1, ..., T$.
+
+The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.
+
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm (@zucchini2009hidden, @joseph2018). 
+All of the code to specify and fit the model in Stan is available in the open source mrmr package (@joseph2019).
+
+The joint distribution of the resulting model can be written as follows:
+
+$$
+\begin{multline*}
+[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{multline*}
+$$
+
+where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
 

### 2023-11-28 doc/manuscript/translocation.qmd
fix rendering of multiline eq
Change headers, fix references

@@
 
-#### Swab Methods
+#### Laboratory Methods
 
@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm (@zucchini2009hidden, @joseph2018). 
-All of the code to specify and fit the model in Stan is available in the open source mrmr package (@joseph2019).
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009hidden, @joseph2018]. 
+All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
 
@@
 
-$$
 \begin{multline*}
@@
 \end{multline*}
-$$
 

### 2023-11-29 code/extinction_and_growth_rate_analysis.R
Remove omega and focus on y1 juv survival

@@ -75,7 +75,7 @@ for(l in 1:nrow(all_surv)){
 		if(nm == "med")
 			med_surv_vals[l] = muR
 
-		params = assign_parameters(muR, muT)
+		params = assign_parameters(muR, muT, new_params=list(sigma_J1=0.09))
 		tlam = get_lambda(params)
 		lambdas_mlu[j] = tlam
 	}
@@ -103,7 +103,7 @@ ggsave(file.path("..", "out", "lambda_estimates.pdf"), width=8, height=5)
 
 
 muR_range = seq(0.01, 0.9, len=50)
-juv_surv_range = seq(0.01, .25, len=50)
+juv_surv_range = seq(0.01, .25, len=50) # Range for year 1 juvenile survival
 lambdas = array(NA, dim=c(length(muR_range), length(juv_surv_range)))
 
 for(i in 1:length(muR_range)){
@@ -121,7 +121,7 @@ dat = expand.grid(muR=muR_range, juv_surv=juv_surv_range)
 dat$lambda = as.vector(lambdas)
 ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) + 
 							geom_hline(aes(yintercept=0.08), linetype="dashed") +
-							geom_contour(aes(x=muR, y=juv_surv, z=lambda), breaks=c(1), color="red", size=1) +
+							geom_contour(aes(x=muR, y=juv_surv, z=lambda), breaks=c(1), color="red", linewidth=1) +
 							geom_point(data=lake_growth_rates_dt, aes(x=surv_med, 
 																												y=rep(0.08, length(surv_med)), 
 																													color=lake_id), size=2) +
@@ -191,23 +191,23 @@ steps = 50 # years
 beta0 = rstan::extract(fit, "beta0")$beta0
 alpha = rstan::extract(fit, "alpha")$alpha
 keep_lakes_new = c(keep_lakes)#, "average")
-phi_omega = 2 # Inverse dispersion of juvenile dispersion and recruitment probability
-omegas = seq(0, 1, by=0.05) 
+phi_sigma_J1 = 2 # Inverse dispersion of juvenile dispersion and recruitment probability
+sigma_J1s = seq(0, 0.25, by=0.025) 
 extinction_at_fifty = list()
 
 # Specify whether we want to allow recruited adult survival probability to differ
 # from translocated adult survival probability
 run_type = "Recruit survival != Translocated survival"
 
[…]

### 2023-11-29 code/population_viability_functions.R

@@ -30,7 +30,6 @@ assign_parameters = function(adult_survR, adult_survT, new_params=list()){
 			  sigma_AT=adult_survT,
 			  p_L1=1, p_L2=0.25, p_J1=0.25,
 			  p_F=0.5, F=100, 
-			  omega=1,
 			  reproduction="poisson")
 
 	# Set any new parameters
@@ -57,8 +56,8 @@ build_transition_matrix = function(p){
 	L1 = c(0, p$sigma_L1*p$p_L1, 0, p$sigma_L1*(1 - p$p_L1), 0, 0, 0)
 	L2 = c(0, 0, p$sigma_L2*p$p_L2, p$sigma_L2*(1 - p$p_L2), 0, 0, 0)
 	L3 = c(0, 0, 0, p$sigma_L3, 0, 0, 0)
-	J1 = c(0, 0, 0, 0, p$sigma_J1*p$p_J1*p$omega, p$sigma_J1*(1 - p$p_J1)*p$omega, 0)
-	J2 = c(0, 0, 0, 0, 0, p$sigma_J2*p$omega, 0)
+	J1 = c(0, 0, 0, 0, p$sigma_J1*p$p_J1, p$sigma_J1*(1 - p$p_J1), 0)
+	J2 = c(0, 0, 0, 0, 0, p$sigma_J2, 0)
 	AR = c(0, 0, 0, 0, 0, p$sigma_AR, 0)
 	AT = c(0, 0, 0, 0, 0, 0, p$sigma_AT)
 
@@ -218,8 +217,8 @@ build_fecundity_matrix = function(p){
 
 }
 
-stochastic_simulation = function(steps, initial_values, muR, muT, phi, phi_omega, 
-								 new_params=list(), omega_traj=NA){
+stochastic_simulation = function(steps, initial_values, muR, muT, phi, phi_sigma_J1, 
+								 new_params=list(), sigma_J1_traj=NA){
 	# Stochastic simulation of projection matrix
 	#
 	# Parameters
@@ -234,13 +233,13 @@ stochastic_simulation = function(steps, initial_values, muR, muT, phi, phi_omega
 	#		Average recruited adult survival probability
 	# muT : float
 	#		Average translocated adult survival probability
-	# mu_omega : float
+	# mu_sigma_J1 : float
 	# 	Mean successful recruitment probability
-	# phi_omega : float
[…]

### 2023-11-30 doc/manuscript/translocation.qmd
Update text to account for viability model changes

--- 2023-11-22 doc/manuscript/translocation.qmd
@@
 
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.XX-1.XX for these six populations).
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
@@
 
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in juvenile survival and recruitment (for year 1 and year 2 juveniles, the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average percent reduction in juvenile survival was less than 50% (given baseline year 1 juvenile survival probability of 0.25; @fig-viability B).
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-When we decreased baseline year 1 juvenile survival to 0.09, 50-year extinction probability was of less than 50% for four of these populations (X-X) and greater than 50% (X-X) for two of these populations (@fig-viability B).
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of baseline year 1 survival probability.
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
 Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
@@
   \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} p_{J1}(1 - \omega) & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) (1 - \omega) & \sigma_{J2} (1 - \omega) & \sigma_{A_R} & 0 \\
+  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
   0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
@@
 The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$.
-The parameter $(1 - \omega)$ gives the percent reduction in baseline juvenile survival and recruitment due environmental stochasticity.
+First year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced environmental stochasticity.
 
@@
 Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
-Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
+Third, we included demographic and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
@@
 
-![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
+![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.
 Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
-The dashed line shows where $\omega = 0.3$, for comparison to (A).
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of \$\\sigma\_{J_1}\$.
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
 (C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
@@
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
+Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in "Estimating model parameters" below).
+We also could have included environmental stochasticity in year 2 juvenile survival and recruitment \$\\sigma\_{J_2}\$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$
 
@@
 We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
-We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
+First, we considered a deterministic version of our model with no yearly heterogeneity in year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
+We then fixed $\sigma_{J_1} = 0.09$ and calculated the predicted growth rate of our 12 populations.
 
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters \$\\sigma\_{J_1}\$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
 
-Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
+Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\sigma_{J_1}$.
 We used this model to simulate a one-time introduction of 40 translocated adult frogs.
-We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
-We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\sigma_{J_1}$.
+We varied the mean recruitment probability $\sigma_{J_1}$ from 0 and 0.25 and drew values of $\sigma_{J_1}$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\sigma_{J_1} = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+Using different values of $\phi$ does not qualitatively change the existence of distinct extinction dynamics between populations with $\sigma_{A_R} < 0.5$ and those with $\sigma_{A_R} > 0.5$.
+However, increasing yearly variability in $\sigma_{J_1}$ increases extinction risk for all populations.
+For example, if we set $\phi = 0.001$, such that in a given year essentially either all year 1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
+Because we do not pit tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
+However, based on qualitative and semi-quantitative field observations over 25 years, a value of $\sigma_{J_1} = 0.25$ in the presence of Bd is probably a reasonable estimate for many populations.
+Thus, we expect our model predictions to be conservative with regards to population recovery.
 
@@
 We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
+For each run and each year, we drew $\sigma_{J_1}$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
 Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
+The yearly $\sigma_{J_1}$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
+Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year 1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year 2 juveniles).
+Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:----------------------------|:-----------------|:-------------------------|
+|:---------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
@@
 | $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
 | $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
@@
 | $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                                |
-| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
 

### 2023-11-30 code/extinction_and_growth_rate_analysis.R

@@ -334,7 +334,7 @@ cdat = cdat[order(surv_med)]
 extinction_at_fifty_dt$lake_id_surv = factor(extinction_at_fifty_dt$lake_id_surv, levels=cdat$lake_id_surv)
 
 pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=sigma_J1, y=ext_prob)) + 
-							#geom_vline(aes(xintercept=1 - 0.5), linetype="dashed") +
+							geom_vline(aes(xintercept=0.1), linetype="dashed") +
 							geom_line(aes(color=lake_id_surv)) + 
 							geom_point(aes(color=lake_id_surv)) + 
 							scale_color_manual(values=cdat$col) +

### 2023-12-02 doc/manuscript/translocation.qmd
Changes to text in response to reviewer comments

--- 2023-11-28 doc/manuscript/translocation.qmd
@@
 csl: pnas.csl
+format: pdf
 # format: docx
-format :
+# format :
 #  latex:
-  pdf:
-    cite-method: natbib
-    include-in-header:
-      text: |
-        \usepackage{blkarray}
+#    cite-method: natbib
 editor: visual
@@
 Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions.
 However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
@@
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
+(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., species recovery.) <!--# Feels clumsy to include this paranthetically, but I have not yet come up with a better alternative.
+-->
 
@@
 Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
-Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
+Frogs in these recovering populations show reduced susceptibility to Bd infection, with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
+This reduced susceptibility is most consistent with natural selection for more resistant/tolerant genotypes and/or acquired immunity [@grogan2018; @knapp2016], and additional investigation of the specific role of evolution is critically needed.
+Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].
 
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
-These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
+<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->
+
+In the current study, we had three primary objectives.
+First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
+Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+
+<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
+Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
+We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
+Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
+These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
 
 ### Results
-
-#### Frog evolution in response to Bd
-
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
-(See SI Appendix for additional details regarding the study design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
-One of the 7 identified genes (LOC108802036) did not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
-This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 
+<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->
+
+Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
+For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
+These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
+As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
+In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
+
 Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
@@
 Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
+
+#### Frog evolution in response to Bd
+
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
+Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
+(See SI Appendix for additional details regarding the study design.) Bd-exposure histories of the nine study populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+muscosa* versus *R.
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
+One of the 7 identified genes (LOC108802036) did not have an associated annotation.
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+
+The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
+The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
+Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
+This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
-The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
+The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
 In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
@@
 In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
-MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
+MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality.
 This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
@@
 
-#### Frog evolution in response to Bd
-
-##### Sampling and Sequencing
-
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
-These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
-Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
-DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
-Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
-Raw sequencing reads are available from NCBI SRA (PRJNA870451).
-
-##### Data Pre-processing and Cleaning
-
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
-Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
-This initial filter resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant level.
-First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x.
-Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-##### Data Analysis
-
-To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
-We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
-After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
-We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
-
-Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-
 #### Frog population recovery
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
+Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
 CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
@@
 This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
-Similarly, to calculcate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
+Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
 This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).
@@
 
+![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
+The inset map locates the study area and range boundary in California.
+For cross-referencing of numbered points to populations/sites mentioned in the **Frog Population Recovery** section (above), site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+
+\newpage
+
+![Evidence for selection on genomic regions in recovering MYL frog populations.
+Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
+In (B), outlier regions are shown above the single dashed red line.
+Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
+(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+
+\newpage
+
 ## Supporting Information
-
-### Frog evolution in response to Bd
-
-#### Methods
-
-We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-
-#### Results
-
-In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 
@@
 In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
-
-#### Mark-recapture model structure
-
-We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
-For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
-Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
-Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
-
-We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
-The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
-Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
-Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
-We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
-Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
-We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
-This assumption is justified by the short time intervals between surveys within primary periods.
-
-Live individuals are detected with probability $p_j$, which is modeled as:
-
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
-
-where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
-Not recruited and dead individuals are never captured. 
-We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
-
-$$
-\Omega_j =
-\begin{blockarray}{ccc}
-  \text{Not detected} & \text{Detected} \\
-\begin{block}{(cc)c}
-  1 & 0 & \text{Not recruited} \\
-  1 - p_j & p_j & \text{Alive} \\
-  1 & 0 & \text{Dead} \\
-\end{block}
-\end{blockarray}
-$$
-
-The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$. 
-For unintroduced individuals which naturally recruit, this matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 1 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$
-where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
-
-For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 1 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$
-where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
-
-We allow recruitment probabilities to vary in time via random effects, such that:
-
-$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
-
-where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.
-
-Survival probabilities also vary in time, and as a function of known covariates:
-
-$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
-
-where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.
-
-To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters. 
-The recruitment parameter priors were specified as follows:
-
-$$\alpha^{(\lambda)} \sim N(0,1),$$
-$$\sigma^{(\lambda)} \sim N_+(0,1),$$
-$$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$
-for periods $t=1, ..., T$. 
-Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
-
-Survival parameter priors were specified similarly as:
-
-$$\beta^{(\phi)} \sim N(0, 1),$$
-$$\sigma^{(\phi)} \sim N_+(0, 1),$$
-$$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$
-for time $t=1, ..., T$.
-
-The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.
-
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm (@zucchini2009hidden, @joseph2018). 
-All of the code to specify and fit the model in Stan is available in the open source mrmr package (@joseph2019).
-
-The joint distribution of the resulting model can be written as follows:
-
-$$
-\begin{multline*}
-[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
-\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
-\end{multline*}
-$$
-
-where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
 
@@
 
+### Frog evolution in response to Bd
+
+#### Methods
+
+We compared frog genomes sampled in naive versus recovering populations.
+Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
+The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
+As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+
+#### Results
+
+In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
+The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
+Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
+
+In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
+
 ### Figures
@@
 
-![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by black polygon).
-Symbol shapes indicate the donor population used for each translocation site.
+![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon).
+Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
 To obscure the exact locations of populations, random noise was added to all point coordinates.
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="250"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}
 
@@
 
-![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats <!--# Consider edits to this to respond to Reviewer 2 - Comment 4.
+In Discussion?-->.
 For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
@@
 Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
+Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
+Samples sizes are provided just above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="700"}
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                         |
-|:-----------------------------|:---------------|:--------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |

### 2023-12-02 doc/manuscript/translocation.qmd
Address review feedback

--- 2023-11-28 doc/manuscript/translocation.qmd
@@
 Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012]. 
 Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
@@
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
-Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
+Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
@@
 We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
-This assumption is justified by the short time intervals between surveys within primary periods.
+This assumption is justified by the short time intervals between surveys within primary periods, in cases where primary periods contain multiple secondary periods.
 
@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009hidden, @joseph2018]. 
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018]. 
 All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
@@
 
-We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
+We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mark-recapture models.
 Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).

### 2023-12-02 doc/manuscript/translocation.qmd

--- 2023-11-30 doc/manuscript/translocation.qmd
@@
 
-#### Methods
+#### Laboratory Methods
 
@@
 
+#### Mark-recapture model structure
+
+We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
+For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
+Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012]. 
+Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
+
+We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
+The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
+Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
+Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
+We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
+Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
+We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
+This assumption is justified by the short time intervals between surveys within primary periods, in cases where primary periods contain multiple secondary periods.
+
+Live individuals are detected with probability $p_j$, which is modeled as:
+
+$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
+
+where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
+Not recruited and dead individuals are never captured. 
+We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
+
+$$
+\Omega_j =
+\begin{blockarray}{ccc}
+  \text{Not detected} & \text{Detected} \\
+\begin{block}{(cc)c}
+  1 & 0 & \text{Not recruited} \\
+  1 - p_j & p_j & \text{Alive} \\
+  1 & 0 & \text{Dead} \\
+\end{block}
+\end{blockarray}
+$$
+
+The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$. 
+For unintroduced individuals which naturally recruit, this matrix is given by:
+
+$$
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+$$
+where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
+
+For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:
+
+$$
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+$$
+where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
+
+We allow recruitment probabilities to vary in time via random effects, such that:
+
+$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
+
+where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.
+
+Survival probabilities also vary in time, and as a function of known covariates:
+
+$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
+
+where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.
+
+To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters. 
+The recruitment parameter priors were specified as follows:
+
+$$\alpha^{(\lambda)} \sim N(0,1),$$
+$$\sigma^{(\lambda)} \sim N_+(0,1),$$
+$$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$
+for periods $t=1, ..., T$. 
+Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
+
+Survival parameter priors were specified similarly as:
+
+$$\beta^{(\phi)} \sim N(0, 1),$$
+$$\sigma^{(\phi)} \sim N_+(0, 1),$$
+$$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$
+for time $t=1, ..., T$.
+
+The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.
+
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018]. 
+All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
+
+The joint distribution of the resulting model can be written as follows:
+
+\begin{multline*}
+[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{multline*}
+
+where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
+
 ### Population viability modeling

### 2023-12-03 doc/manuscript/translocation.qmd

@@
 Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012]. 
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
 Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
@@
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
-Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
+Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
@@
 We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
-This assumption is justified by the short time intervals between surveys within primary periods, in cases where primary periods contain multiple secondary periods.
+This assumption is justified by the short time intervals between surveys within primary periods.
 
@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018]. 
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009hidden, @joseph2018]. 
 All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
@@
 
-We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mark-recapture models.
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
 Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).

### 2023-12-03 doc/manuscript/translocation.qmd
Minor edits to references and citations

@@
   pdf:
-    cite-method: natbib
+#    cite-method: natbib
     include-in-header:
@@
 Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012].
 Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
@@
 \end{blockarray},
-$$
-where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
+$$ where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
 
@@
 \end{blockarray},
-$$
-where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
+$$ where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
 
@@
 
-$$\alpha^{(\lambda)} \sim N(0,1),$$
-$$\sigma^{(\lambda)} \sim N_+(0,1),$$
-$$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$
-for periods $t=1, ..., T$. 
+$$\alpha^{(\lambda)} \sim N(0,1),$$ $$\sigma^{(\lambda)} \sim N_+(0,1),$$ $$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$ for periods $t=1, ..., T$.
 Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
@@
 
-$$\beta^{(\phi)} \sim N(0, 1),$$
-$$\sigma^{(\phi)} \sim N_+(0, 1),$$
-$$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$
-for time $t=1, ..., T$.
+$$\beta^{(\phi)} \sim N(0, 1),$$ $$\sigma^{(\phi)} \sim N_+(0, 1),$$ $$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$ for time $t=1, ..., T$.
 
@@
 
+```{=tex}
 \begin{multline*}
@@
 \end{multline*}
-
+```
 where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:---------------------------|:------------------|:-------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2023-12-03 doc/manuscript/translocation.qmd
Edited Zucchini & Dorazio reference

@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009hidden, @joseph2018].
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018].
 All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
+|:--------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2023-12-03 doc/manuscript/translocation.qmd

--- 2023-12-03 doc/manuscript/translocation.qmd
@@
 Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
+
+This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012]. 
 Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009hidden, @joseph2018]. 
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018].
+
 All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].

### 2023-12-03 doc/manuscript/translocation.qmd
Describe stan_glmer models

@@
 
-### Population viability modeling
-
-#### Methods
-
-##### Incorporating yearly variability in vital rates
+#### Population viability modeling
+
+##### Methods
+
+###### Incorporating yearly variability in vital rates
 
@@
 
-##### Estimating model parameters
+###### Estimating model parameters
 
@@
 
-##### Model analysis and simulation
+###### Model analysis and simulation
 
@@
 
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters \$\\sigma\_{J_1}\$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
 
@@
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+
+#### Among-site survival modeling
+
+We used Bayesian generalized linear mixed models to investigate predictors of survival among sites. 
+The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation. 
+We generated these point estimates by rounding the posterior median of 1-year post introduction survival for each individual, and modeled the data using a Bernoulli distribution:
+
+$$y_i \sim \text{Bernoulli}(p_i),$$
+
+where $p_i$ is the probability of survival.
+
+We modeled variation in probabilities as follows:
+
+$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$
+
+where $\alpha$ is an intercept, $X_i$ is a length $K$ row vector of predictors, $\pmb \beta$ is a column vector of predictor effects, and $\pmb \nu$ a vector of group level random effects. 
+Here $g[i]$ refers to the group $g$ containing individual $i$, and we estimate an adjustment for each of the $G$ groups ($\nu_1, ..., \nu_G$).
+
+These models were fit using the `stan_glmer()` function in the rstanarm package, with default priors described below [@rstanarm2022].
+These priors are vague, but include data dependent scaling as follows to account for different input variable scales.
+However, as we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical. 
+Specifically, we have:
+
+$$\alpha \sim \text{Normal}(0, 2.5),$$
+and
+
+$$\beta_k \sim \text{Normal}(0, 5),$$
+
+for $k=1, ..., K$ where $K$ is the number of predictor variables.
+
+The default prior for group level adjustments $\nu_1, ..., \nu_G$ in rstanarm is a zero-mean Gaussian, where the covariance matrix is constructed from a correlation matrix with an LKJ prior, and a vector of variance parameters -- the decomposition of variance prior with unit regularization, concentration, shape, and scale parameters [@lewandowski2009].
+
+We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan, with four parallel chains, each run for 10,000 iterations, discarding the first half of each chain as warm-up draws [@rstanarm2022].
+We used Rhat statistics and trace plots to verify convergence. 
+We considered models with different subsets of fixed and random effects, and used approximate leave-one-out cross validation to identify the best model [@vehtari2016].
 

### 2023-12-05 doc/manuscript/translocation.qmd
Adjust header hierarchy
Move survival modeling methods section

@@
 
+#### Among-site survival modeling
+
+We used Bayesian generalized linear mixed models to investigate predictors of survival among sites.
+The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation.
+We generated these point estimates by rounding the posterior median of 1-year post introduction survival for each individual, and modeled the data using a Bernoulli distribution:
+
+$$y_i \sim \text{Bernoulli}(p_i),$$
+
+where $p_i$ is the probability of survival.
+
+We modeled variation in probabilities as follows:
+
+$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$
+
+where $\alpha$ is an intercept, $X_i$ is a length $K$ row vector of predictors, $\pmb \beta$ is a column vector of predictor effects, and $\pmb \nu$ a vector of group level random effects.
+Here $g[i]$ refers to the group $g$ containing individual $i$, and we estimate an adjustment for each of the $G$ groups ($\nu_1, ..., \nu_G$).
+
+These models were fit using the `stan_glmer()` function in the rstanarm package, with default priors described below [@rstanarm2022].
+These priors are vague, but include data dependent scaling as follows to account for different input variable scales.
+However, as we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical.
+Specifically, we have:
+
+$$\alpha \sim \text{Normal}(0, 2.5),$$ and
+
+$$\beta_k \sim \text{Normal}(0, 5),$$
+
+for $k=1, ..., K$ where $K$ is the number of predictor variables.
+
+The default prior for group level adjustments $\nu_1, ..., \nu_G$ in rstanarm is a zero-mean Gaussian, where the covariance matrix is constructed from a correlation matrix with an LKJ prior, and a vector of variance parameters -- the decomposition of variance prior with unit regularization, concentration, shape, and scale parameters [@lewandowski2009].
+
+We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan, with four parallel chains, each run for 10,000 iterations, discarding the first half of each chain as warm-up draws [@rstanarm2022].
+We used Rhat statistics and trace plots to verify convergence.
+We considered models with different subsets of fixed and random effects, and used approximate leave-one-out cross validation to identify the best model [@vehtari2016].
+
 #### Population viability modeling
 
-##### Methods
-
-###### Incorporating yearly variability in vital rates
+##### Incorporating yearly variability in vital rates
 
@@
 
-###### Estimating model parameters
+##### Estimating model parameters
 
@@
 
-###### Model analysis and simulation
+##### Model analysis and simulation
 
@@
 Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
-
-#### Among-site survival modeling
-
-We used Bayesian generalized linear mixed models to investigate predictors of survival among sites. 
-The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation. 
-We generated these point estimates by rounding the posterior median of 1-year post introduction survival for each individual, and modeled the data using a Bernoulli distribution:
-
-$$y_i \sim \text{Bernoulli}(p_i),$$
-
-where $p_i$ is the probability of survival.
-
-We modeled variation in probabilities as follows:
-
-$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$
-
-where $\alpha$ is an intercept, $X_i$ is a length $K$ row vector of predictors, $\pmb \beta$ is a column vector of predictor effects, and $\pmb \nu$ a vector of group level random effects. 
-Here $g[i]$ refers to the group $g$ containing individual $i$, and we estimate an adjustment for each of the $G$ groups ($\nu_1, ..., \nu_G$).
-
-These models were fit using the `stan_glmer()` function in the rstanarm package, with default priors described below [@rstanarm2022].
-These priors are vague, but include data dependent scaling as follows to account for different input variable scales.
-However, as we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical. 
-Specifically, we have:
-
-$$\alpha \sim \text{Normal}(0, 2.5),$$
-and
-
-$$\beta_k \sim \text{Normal}(0, 5),$$
-
-for $k=1, ..., K$ where $K$ is the number of predictor variables.
-
-The default prior for group level adjustments $\nu_1, ..., \nu_G$ in rstanarm is a zero-mean Gaussian, where the covariance matrix is constructed from a correlation matrix with an LKJ prior, and a vector of variance parameters -- the decomposition of variance prior with unit regularization, concentration, shape, and scale parameters [@lewandowski2009].
-
-We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan, with four parallel chains, each run for 10,000 iterations, discarding the first half of each chain as warm-up draws [@rstanarm2022].
-We used Rhat statistics and trace plots to verify convergence. 
-We considered models with different subsets of fixed and random effects, and used approximate leave-one-out cross validation to identify the best model [@vehtari2016].
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:---------------------------|:------------------|:-------------------------|
+|:--------------------------|:------------------|:-------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2023-12-05 doc/manuscript/translocation.qmd
Edits primarily in response to Reviewer 2 comments

--- 2023-12-03 doc/manuscript/translocation.qmd
@@
 csl: pnas.csl
+format: pdf
 # format: docx
-format :
+# format :
 #  latex:
-  pdf:
 #    cite-method: natbib
-    include-in-header:
-      text: |
-        \usepackage{blkarray}
 editor: visual
@@
 ### Abstract
+
+<!--# Revise to reflect reordering of sections -->
 
@@
 To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed. 
+(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., describing recovery of MYL frogs in the presence of Bd.) <!--# Feels clumsy to include this parenthetically, but I have not yet come up with a better alternative.
+-->
 
@@
 Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
-Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
+Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
+This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], and is consistent with several non-mutually exclusive mechanisms.
+These include natural selection for more resistant/tolerant genotypes [@savage2016], acquired immunity [@grogan2018], and inherent between-populations differences in frog susceptibility that pre-date Bd exposure.
+Of these possibilities, evolution of resistance/tolerance would have the most general relevance for amphibian recovery from the effects of Bd, and as such additional investigation of this potential mechanism is critically needed.
+Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].
 
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
-These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
+<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->
+
+In the current study, we had three primary objectives.
+First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
+Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+
+<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
+Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
+We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
+Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
+These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
 
 ### Results
-
-#### Frog evolution in response to Bd
-
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
-(See SI Appendix for additional details regarding the study design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
-One of the 7 identified genes (LOC108802036) did not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
-This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 
-Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
-This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
+<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->
 
@@
 
+Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
+This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
+sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
+
 To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
@@
 Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
-Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
+Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize Bd load and/or the adverse effects of Bd infection.
 In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
@@
 
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
+When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
 
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
+When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
 Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
+
+#### Frog evolution in response to Bd
+
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
+Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
+Bd-exposure histories of the nine study populations are based on 10-20 years of visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low; wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
+In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally < 100), high Bd prevalence (often > 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
+Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load. 
+(See SI Appendix for additional details regarding the study design.)  
+<!--In response to Editor - Comment 3, mention that there is no dispersal between extant populations due to presence of fish and fish-induced habitat fragmentation?-->
+
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
+
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+muscosa* versus *R.
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
+One of the 7 identified genes (LOC108802036) did not have an associated annotation.
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
+This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+
+The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
+Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
+The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
+Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
+Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
+This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
+Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
@@
 Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
-The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
+The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
 In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
 Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
+<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->
 
@@
 ### Materials and Methods
+
+#### Frog population recovery
+
+##### Field methods
+
+For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
+The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
+These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
+As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
+Recipient sites to which frogs were translocated were chosen based on previous *R.
+sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
+At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics. 
+In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks).
+*R.
+sierrae* were absent from all recipient sites prior to the first translocation.
+
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
+Frogs were transported to the recipient site either on foot or via helicopter.
+Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
+CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
+During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
+Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
+During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
+However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
+When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
+
+During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
+VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
+These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
+
+##### Frog counts and reproductive success
+
+For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
+To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
+This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
+Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
+This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).
+
+##### Estimation of frog survival and abundance
+
+We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
+The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
+The possible states of individuals include "not recruited", "alive", and "dead".
+The possible observations of individuals include "detected" and "not detected".
+We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
+We also assume that new recruits were a result of within-site reproduction and not immigration from adjacent populations because no *R.
+sierrae* populations were present within several kilometers of the translocation sites.
+For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
+We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
+
+##### Predictors of post-translocation frog survival
+
+To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
+We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
+Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
+We evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
+
+We conducted two analyses.
+In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
+Group-level effects included site_id or translocation_id.
+The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
+In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
+Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
+In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
+For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
+We checked convergence of the MCMC algorithm using trace plots and Rhat.
+
+##### Changes in Bd load following translocation
+
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
+To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
+Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
+After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
+Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
+We used an analysis workflow similar to that described above.
+Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
+We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
+All models used a negative binomial family.
+Priors, chains, iterations, and checks were as described above.
+
+Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.
+
+#### Population viability modeling
+
+##### Model description
+
+To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for MYL frogs.
+Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
+We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
+We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs]).
+
+We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
+The dynamics are given by
+
+$$
+\begin{bmatrix}
+L_1 \\
+L_2 \\
+L_3 \\
+J_1 \\
+J_2 \\
+A_R \\ 
+A_T
+\end{bmatrix}(t + 1) = 
+\begin{bmatrix}
+  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
+  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
+  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
+  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
+  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
+  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
+\end{bmatrix} \begin{bmatrix}
+L_1 \\
+L_2 \\
+L_3 \\
+J_1 \\
+J_2 \\
+A_R \\
+A_T \\
+\end{bmatrix}(t) 
+$$
+
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.
+
+In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
+We also did not directly consider the dynamics of Bd in this model.
+We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
+Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
+Additional details are provided in the SI Appendix.
+
+##### Model analysis
+
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
+First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
+Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
+Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
+Additional details are provided in the SI Appendix.
+Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
 
@@
 
-#### Frog population recovery
-
-##### Field methods
-
-For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
-The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
-These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
-Recipient sites to which frogs were translocated were chosen based on previous *R.
-sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
-
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
-In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
-Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
-CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
-Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
-However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
-When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
-
-During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
-These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
-
-##### Frog counts and reproductive success
-
-For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
-To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
-This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
-Similarly, to calculcate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
-This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).
-
-##### Estimation of frog survival and abundance
-
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
-The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
-The possible states of individuals include "not recruited", "alive", and "dead".
-The possible observations of individuals include "detected" and "not detected".
-We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
-For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
-We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-
-##### Predictors of post-translocation frog survival
-
-To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
-We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
-Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
-We evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
-
-We conducted two analyses.
-In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
-Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
-Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
-In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
-For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked convergence of the MCMC algorithm using trace plots and Rhat.
-
-##### Changes in Bd load following translocation
-
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
-To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
-Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
-After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
-Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
-We used an analysis workflow similar to that described above.
-Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
-We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
-All models used a negative binomial family.
-Priors, chains, iterations, and checks were as described above.
-
-Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.
-
-#### Population viability modeling
-
-##### Model description
-
-To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for MYL frogs.
-Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs]).
-
-We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
-The dynamics are given by
-
-$$
-\begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\ 
-A_T
-\end{bmatrix}(t + 1) = 
-\begin{bmatrix}
-  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
-  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
-  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
-  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
-  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
-\end{bmatrix} \begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\
-A_T \\
-\end{bmatrix}(t) 
-$$
-
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$.
-First year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced environmental stochasticity.
-
-In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
-We also did not directly consider the dynamics of Bd in this model.
-We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
-Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-Additional details are provided in the SI Appendix.
-
-##### Model analysis
-
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
-First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
-Third, we included demographic and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-Additional details are provided in the SI Appendix.
-Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
-
 ### Acknowledgments
@@
 
-![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
+![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.
 Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of \$\\sigma\_{J_1}\$.
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$.
+The dashed line shows where $\omega = 0.3$, for comparison to (A).
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.
 (C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
@@
 
+![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
+Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
+The inset map locates the study area and range boundary in California.
+For cross-referencing of numbered points to populations/sites mentioned in the **Frog Population Recovery** section (above), site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+
+\newpage
+
+![Evidence for selection on genomic regions in recovering MYL frog populations.
+Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
+In (B), outlier regions are shown above the single dashed red line.
+Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
+(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
+(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+
+\newpage
+
 ## Supporting Information
+
+### Frog population recovery
+
+#### Methods
+
+Swab extracts were analyzed using standard Bd DNA extraction and qPCR methods [@boyle2004], and extracts were analyzed singly instead of in triplicate [@kriger2006].
+For analysis of swabs collected during 2005--2014, we used standards developed from known concentrations of zoospores [@boyle2004], and after 2014, we used standards based on single ITS1 PCR amplicons [@longo2013].
+Based on paired comparisons between samples analyzed using both types of standards, Bd in the study area has an average of 60 ITS1 copies per zoospore.
+To express all qPCR results as the number of ITS1 copies, starting quantities obtained using the zoospore standard (measured as "zoospore equivalents") were multiplied by 60.
+In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
+
+### Population viability modeling
+
+#### Methods
+
+##### Incorporating yearly variability in vital rates
+
+We extracted yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the CMR model used to estimate population-level parameters of translocated populations (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).
+Thus, we did not include yearly within-population variability in adult survival in this analysis.
+However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
+Therefore, we allowed for yearly variability in the probability of successful recruitment $\omega$ (additional details provided in "Estimating model parameters" below).
+
+##### Estimating model parameters
+
+The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
+For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+
+##### Model analysis and simulation
+
+We performed four analyses on our model.
+First, we considered a deterministic version of our model with no yearly heterogeneity in probability of successful recruitment $\omega$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\omega$.
+We then fixed $\omega = 0.3$ and calculated the predicted growth rate of our 12 populations.
+
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
+
+Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\omega$.
+We used this model to simulate a one-time introduction of 40 translocated adult frogs.
+We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\omega$.
+We varied the mean recruitment probability $\omega$ from 0.15 and 0.4 and drew values of $\omega$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\omega = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
+
+Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
+We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
+We simulated our model for 16 years, repeating the simulation 50,000 times.
+For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
+Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
+The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
+Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
+
+\newpage
 
@@
 We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
 Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
@@
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
-
-### Frog population recovery
-
-#### Laboratory Methods
-
-Swab extracts were analyzed using standard Bd DNA extraction and qPCR methods [@boyle2004], and extracts were analyzed singly instead of in triplicate [@kriger2006].
-For analysis of swabs collected during 2005--2014, we used standards developed from known concentrations of zoospores [@boyle2004], and after 2014, we used standards based on single ITS1 PCR amplicons [@longo2013].
-Based on paired comparisons between samples analyzed using both types of standards, Bd in the study area has an average of 60 ITS1 copies per zoospore.
-To express all qPCR results as the number of ITS1 copies, starting quantities obtained using the zoospore standard (measured as "zoospore equivalents") were multiplied by 60.
-In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
-
-#### Mark-recapture model structure
-
-We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
-For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
-Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012]. 
-Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
-
-We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
-The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
-Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
-Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
-We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
-Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
-We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
-This assumption is justified by the short time intervals between surveys within primary periods, in cases where primary periods contain multiple secondary periods.
-
-Live individuals are detected with probability $p_j$, which is modeled as:
-
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
-
-where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector.
-Not recruited and dead individuals are never captured.
-We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
-
-$$
-\Omega_j =
-\begin{blockarray}{ccc}
-  \text{Not detected} & \text{Detected} \\
-\begin{block}{(cc)c}
-  1 & 0 & \text{Not recruited} \\
-  1 - p_j & p_j & \text{Alive} \\
-  1 & 0 & \text{Dead} \\
-\end{block}
-\end{blockarray}
-$$
-
-The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$.
-For unintroduced individuals which naturally recruit, this matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 1 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$ where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
-
-For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 1 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$ where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
-
-We allow recruitment probabilities to vary in time via random effects, such that:
-
-$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
-
-where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.
-
-Survival probabilities also vary in time, and as a function of known covariates:
-
-$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
-
-where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.
-
-To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters.
-The recruitment parameter priors were specified as follows:
-
-$$\alpha^{(\lambda)} \sim N(0,1),$$ $$\sigma^{(\lambda)} \sim N_+(0,1),$$ $$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$ for periods $t=1, ..., T$.
-Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
-
-Survival parameter priors were specified similarly as:
-
-$$\beta^{(\phi)} \sim N(0, 1),$$ $$\sigma^{(\phi)} \sim N_+(0, 1),$$ $$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$ for time $t=1, ..., T$.
-
-The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.
-
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018].
-
-All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
-
-The joint distribution of the resulting model can be written as follows:
-
-```{=tex}
-\begin{multline*}
-[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
-\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
-\end{multline*}
-```
-where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
-
-### Population viability modeling
-
-#### Methods
-
-##### Incorporating yearly variability in vital rates
-
-We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mark-recapture models.
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).
-Thus, we did not include yearly within-population variability in adult survival in this analysis.
-However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in "Estimating model parameters" below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment \$\\sigma\_{J_2}\$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$
-
-##### Estimating model parameters
-
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
-For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-##### Model analysis and simulation
-
-We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
-We then fixed $\sigma_{J_1} = 0.09$ and calculated the predicted growth rate of our 12 populations.
-
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters \$\\sigma\_{J_1}\$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
-
-Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\sigma_{J_1}$.
-We used this model to simulate a one-time introduction of 40 translocated adult frogs.
-We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\sigma_{J_1}$.
-We varied the mean recruitment probability $\sigma_{J_1}$ from 0 and 0.25 and drew values of $\sigma_{J_1}$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\sigma_{J_1} = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
-Using different values of $\phi$ does not qualitatively change the existence of distinct extinction dynamics between populations with $\sigma_{A_R} < 0.5$ and those with $\sigma_{A_R} > 0.5$.
-However, increasing yearly variability in $\sigma_{J_1}$ increases extinction risk for all populations.
-For example, if we set $\phi = 0.001$, such that in a given year essentially either all year 1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
-Because we do not pit tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
-However, based on qualitative and semi-quantitative field observations over 25 years, a value of $\sigma_{J_1} = 0.25$ in the presence of Bd is probably a reasonable estimate for many populations.
-Thus, we expect our model predictions to be conservative with regards to population recovery.
-
-Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
-We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
-We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\sigma_{J_1}$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The yearly $\sigma_{J_1}$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
-Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
-
-\newpage
 
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year 1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year 2 juveniles).
-Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
+Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch).
+Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
 
@@
 | $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
 | $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
@@
 | $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
+| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
 

### 2023-12-06 code/maps.qmd

--- 2023-05-17 code/maps.qmd
@@
 # For printing on screen: xmin = 244000, ymax = 4242500, xmax = 309000, ymin = 4142000
-bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 308000, ymin = 4151000), 
+bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 311000, ymin = 4151000), 
                 crs = st_crs(yose_boundary)) %>%  
@@
 ```{r}
+#translocation_points <- translocation_points %>% 
+# mutate(
+#    label = case_when(
+#      site_id == 70567 ~ 4, 
+#      site_id == 72996 ~ 1
+#    ))
+
 translocation_points_map <- tm_shape(hillshade_yose_crop) + 
@@
   tm_shape(yose_boundary) + 
-  tm_borders(col = "black", lwd = 2) +
+  tm_borders(col = "#737373", lwd = 2) +
   tm_shape(translocation_points) + 
-  tm_symbols(size = 0.9, 
+  tm_symbols(size = 4, 
              shape = "donor_id",
@@
              legend.shape.show = FALSE, 
-             palette = c("#0000FF", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
+             palette = c("#87CEFA", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
              title.col = "Population type") +
@@
             legend.text.size = 1.2) +
+  tm_text(text = "site_id",
+          auto.placement = TRUE,
+          size = 1.1) +
   tm_compass(type = "4star",

### 2023-12-06 code/translocation_survival_analysis_supp.Rmd

@@
 ```{r}
-d1 %>% 
-  mutate(period = factor(period, levels = c("before", "after"))) %>% 
-  ggplot(aes(x = period, y = log10(bd_load + 1))) +
-          geom_boxplot()
+smplsize <- d1 %>% 
+  count(site_id1, period)
 
@@
   mutate(period = factor(period, levels = c("before", "after"))) %>% 
+  filter(interval < 2) %>% 
   ggplot(aes(x = period, y = log10(bd_load + 1))) +
@@
     labs(x = "Period (relative to translocation)", y = (expression(Bd~load~(log["10"]~copies+1)))) +
+    coord_cartesian(ylim = c(-0.5, NA)) +
     facet_grid(. ~ site_id1) +
     theme_bw() +
-    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
           strip.text.x = element_text(size = 8),
-          panel.grid = element_blank())
+          panel.grid = element_blank()) +
+    geom_text(data = smplsize, aes(x = period, y = -0.4, label = n), size = 2.2, vjust = 1)
+
 ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 6, height = 3)

### 2023-12-06 doc/manuscript/translocation.qmd

--- 2023-12-05 doc/manuscript/translocation.qmd
@@
 csl: pnas.csl
+format: pdf
 # format: docx
@@
 
+<!--# Drop "evolutionary rescue" from title? -->
+
 ## Title: Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
@@
 Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
-Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
-
-In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
-Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
-We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
-These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
-Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
+Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
+This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], and is consistent with several non-mutually exclusive mechanisms.
+These include natural selection for more resistant/tolerant genotypes [@savage2016], acquired immunity [@grogan2018], and inherent between-populations differences in frog susceptibility that pre-date Bd exposure.
+Of these possibilities, evolution of resistance/tolerance would have the most general relevance for amphibian recovery from the effects of Bd, and as such additional investigation of this potential mechanism is critically needed.
+Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].
+
+<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->
+
+In the current study, we had three primary objectives.
+First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
+Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+
+<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
+Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
+We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
+Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
+These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
 
 ### Results
+
+#### Frog population recovery
+
+```{r}
+#| include: false
+#| code-fold: true
+
+library(readr)
+library(dplyr)
+library(magrittr)
+library(stringr)
+library(lubridate)
+d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
+d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
+d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))
+
+transno_min <- d1 %>% 
+  distinct(site_id, date) %>% 
+  group_by(site_id) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(min(n))
+
+transno_max <- d1 %>% 
+  distinct(site_id, date) %>% 
+  group_by(site_id) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(max(n))
+
+transfrogs_max <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(max(n))
+
+transfrogs_min <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(min(n))
+
+transfrogs_median <- d1 %>% 
+  group_by(site_id, date) %>% 
+  count() %>% 
+  ungroup() %>% 
+  summarize(median(n))
+
+bdload_min <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  group_by(site_id) %>% 
+  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
+  ungroup() %>% 
+  summarize(min(bdload_ave))
+
+bdload_max <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  group_by(site_id) %>% 
+  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
+  ungroup() %>% 
+  summarize(max(bdload_ave))
+
+bdprev_max <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
+  group_by(site_id) %>% 
+  summarize(prev = sum(bdload01)/n()) %>% 
+  ungroup() %>% select(prev) %>% 
+  slice_max(prev, n = 1, with_ties = FALSE)
+
+bdprev_min <- d1 %>% 
+  filter(!is.na(bd_load)) %>% 
+  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
+  group_by(site_id) %>% 
+  summarize(prev = sum(bdload01)/n()) %>% 
+  ungroup() %>% select(prev) %>% 
+  slice_min(prev, n = 1, with_ties = FALSE)
+
+files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
+surveys <- read_csv(files, id = "path") %>% 
+  filter(survey_date < "2023-01-01") %>% 
+  select(-path)
+rm(files)
+
+surveys_yr_median <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(median(n))
+
+surveys_yr_max <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(max(n))
+
+surveys_yr_min <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
+  group_by(site_id) %>% 
+  count(yr) %>% ungroup() %>% 
+  summarize(min(n))
+
+robust_year_median <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(median(max_2period))
+
+robust_year_min <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(min(max_2period))
+
+robust_year_max <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr" & secondary_period > 1) %>% 
+  group_by(site_id) %>% 
+  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
+  summarize(max(max_2period))
+
+frog_counts_gt0_nrow <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  nrow()
+
+frog_counts_gt0_median <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(median(pyears_tadsub_present)) %>% 
+  round(2)
+
+frog_counts_gt0_min <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(min(pyears_tadsub_present)) %>% 
+  round(2)
+
+frog_counts_gt0_max <- d2 %>% 
+  filter(pyears_tadsub_present > 0) %>% 
+  summarize(max(pyears_tadsub_present))
+
+frog_captures_gt0_nrow <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  nrow()
+
+frog_captures_gt0_median <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(median(pyears_recruits_present)) %>% 
+  round(2)
+
+frog_captures_gt0_min <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(min(pyears_recruits_present)) %>% 
+  round(2)
+
+frog_captures_gt0_max <- d3 %>% 
+  filter(pyears_recruits_present > 0) %>% 
+  summarize(max(pyears_recruits_present))
+```
+
+To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@seddon2014] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
+The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
+Following translocation, adult survival and recruitment of new adults was estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
+
+<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->
+
+Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
+For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
+These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
+As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
+In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
+
+Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
+This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
+sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
+Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
+
+To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
+Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
+Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
+Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize Bd load and/or the adverse effects of Bd infection.
+In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
+Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
+Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.
+
+#### Long-term population viability
+
+Results from frog translocations indicated that most populations showed evidence of successful reproduction and recruitment and that adult survival was often relatively high (described above).
+Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
+To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
+Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).
+
+Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+These six populations all had observed yearly adult survival greater than 0.5.
+As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
+
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
+
+In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
+This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
+In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
+Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
 
@@
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-These recovering populations exist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensity ("load") that is typically well below the level expected to cause mortality [@vredenburg2010].
+Bd-exposure histories of the nine study populations are based on 10-20 years of visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low; wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
+In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally < 100), high Bd prevalence (often > 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
+Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load. 
 (See SI Appendix for additional details regarding the study design.)  
+<!--In response to Editor - Comment 3, mention that there is no dispersal between extant populations due to presence of fish and fish-induced habitat fragmentation?-->
 
@@
 Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-
-#### Frog population recovery
-
-```{r}
-#| include: false
-#| code-fold: true
-
-library(readr)
-library(dplyr)
-library(magrittr)
-library(stringr)
-library(lubridate)
-d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
-d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
-d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))
-
-transno_min <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-transno_max <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_max <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_min <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-transfrogs_median <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(median(n))
-
-bdload_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(min(bdload_ave))
-
-bdload_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(max(bdload_ave))
-
-bdprev_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_max(prev, n = 1, with_ties = FALSE)
-
-bdprev_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_min(prev, n = 1, with_ties = FALSE)
-
-files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
-surveys <- read_csv(files, id = "path") %>% 
-  filter(survey_date < "2023-01-01") %>% 
-  select(-path)
-rm(files)
-
-surveys_yr_median <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(median(n))
-
-surveys_yr_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(max(n))
-
-surveys_yr_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(min(n))
-
-robust_year_median <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(median(max_2period))
-
-robust_year_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(min(max_2period))
-
-robust_year_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(max(max_2period))
-
-frog_counts_gt0_nrow <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  nrow()
-
-frog_counts_gt0_median <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(median(pyears_tadsub_present)) %>% 
-  round(2)
-
-frog_counts_gt0_min <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(min(pyears_tadsub_present)) %>% 
-  round(2)
-
-frog_counts_gt0_max <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(max(pyears_tadsub_present))
-
-frog_captures_gt0_nrow <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  nrow()
-
-frog_captures_gt0_median <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(median(pyears_recruits_present)) %>% 
-  round(2)
-
-frog_captures_gt0_min <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(min(pyears_recruits_present)) %>% 
-  round(2)
-
-frog_captures_gt0_max <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(max(pyears_recruits_present))
-```
-
-To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@seddon2014] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
-The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
-Following translocation, adult survival and recruitment of new adults was estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
-
-Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
-This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae*.
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
-
-Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
-For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
-These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
-As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
-In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
-
-To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
-Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
-Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
-In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
-Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
-Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.
-
-#### Long-term population viability
-
-Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
-Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
-To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
-Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).
-
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
-These six populations all had observed yearly adult survival greater than 0.5.
-As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
-
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
-
-In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
-Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
 
@@
 sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
+At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics. 
+In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks).
+*R.
+sierrae* were absent from all recipient sites prior to the first translocation.
 
@@
 We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
+We also assume that new recruits were a result of within-site reproduction and not immigration from adjacent populations because no *R.
+sierrae* populations were present within several kilometers of the translocation sites.
 For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
@@
 
+![Evidence for frog survival at the recipient sites following translocation.
+Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
+Error bars show the 95% uncertainty intervals.
+Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
+Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
+
+\newpage
+
+![Evidence for reduced impacts of Bd on MYL frogs as indicated by the lack of an effect of Bd load on survival.
+Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
+(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
+In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
+
+\newpage
+
+![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
+The red line shows where $\lambda = 1$.
+Note that the point for 70413 is mostly hidden behind other points.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of \$\\sigma\_{J_1}\$.
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
+(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
+
+\newpage
+
 ![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
@@
 Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
-The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+The inset map locates the study area and range boundary in California.
+For cross-referencing of numbered points to populations/sites mentioned in the **Frog Population Recovery** section (above), site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
 
@@
 
-![Evidence for frog survival at the recipient sites following translocation.
-Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
-Error bars show the 95% uncertainty intervals.
-Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
-
-\newpage
-
-![Evidence for reduced impacts of Bd on MYL frogs as indicated by the lack of an effect of Bd load on survival.
-Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
-(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
-In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
-
-\newpage
-
-![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
-The red line shows where $\lambda = 1$.
-Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of \$\\sigma\_{J_1}\$.
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
-
-\newpage
-
 ## Supporting Information
-
-### Frog evolution in response to Bd
-
-#### Methods
-
-We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
-
-#### Results
-
-In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
 

### 2023-12-06 doc/manuscript/translocation.qmd
Quarto auto-formats

@@
 Bd-exposure histories of the nine study populations are based on 10-20 years of visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low; wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
-In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally < 100), high Bd prevalence (often > 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
+Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions \[during which Bd loads remain very low; wilber2022\], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
+In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally \< 100), high Bd prevalence (often \> 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
 Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-(See SI Appendix for additional details regarding the study design.)  
+(See SI Appendix for additional details regarding the study design.)\
 <!--In response to Editor - Comment 3, mention that there is no dispersal between extant populations due to presence of fish and fish-induced habitat fragmentation?-->
@@
 In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks).
-*R.
-sierrae* were absent from all recipient sites prior to the first translocation.
+*R. sierrae* were absent from all recipient sites prior to the first translocation.
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------|:------------------|:-------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2023-12-06 doc/manuscript/translocation.qmd
Format YAML and edit title

@@
 format: pdf
-# format: docx
-format :
+# format:
 #    latex:
-  pdf:
 #    cite-method: natbib
@@
 
-<!--# Drop "evolutionary rescue" from title? -->
-
-## Title: Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease
+## Title: Reintroduction of frogs with reduced susceptibility allows recovery in the presence of a lethal fungal disease
 
@@
 This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
-However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
 In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].

### 2023-12-07 doc/manuscript/translocation.qmd
Revise Introduction section

@@
 
-## Title: Reintroduction of frogs with reduced susceptibility allows recovery in the presence of a lethal fungal disease
+## Title: Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
 
@@
 
+<!--# Revise to reflect reduced emphasis on evolutionary rescue. -->
+
 Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
@@
 
-Human activities are intensifying threats to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
-To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
-The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
-In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
-Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of imperiled taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
-
+<!--# Remove uncited references in .bib -->
+
+Human activities are increasingly impacting to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
 One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
 Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
-Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
-In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
-To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
-(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., describing recovery of MYL frogs in the presence of Bd.) <!--# Feels clumsy to include this parenthetically, but I have not yet come up with a better alternative.
--->
+Amphibians are experiencing particularly devastating impacts of disease due to the recent emergence and global spread of the highly virulent amphibian chytrid fungus, *Batrachochytrium dendrobatidis* (Bd) [@luedtke2023; @scheele2019].
+By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
+
+Alarmingly, evidence of natural recovery in the many Bd-impacted amphibian populations is scarce [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggestive of generally low resilience to disease effects.
+Following pathogen arrival in a host population, the ability to either limit pathogen burden (resistance) or limit the harm caused by a particular burden (tolerance) are key mechanisms to reduce disease impacts [@råberg2009] and facilitate population persistence and recovery [@brannelly2021].
+Host immunity and evolution both play important roles in the development of resistance and tolerance, and utilizing these factors would seem a promising approach to developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
+However, several aspects of the amphibian-Bd system present difficult obstacles, including (1) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (2) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
+These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
+Indeed, in a review of the effectiveness of mitigation methods, including reintroductions, Garner et al.
+[-@garner2016] came to the gloomy (but accurate) conclusion that, "Despite decades of research into amphibian-chytrid host--pathogen biology, no effective method to reduce the impact of chytridiomycosis (i.e., the disease caused by Bd) has emerged and been tested broadly in the field." If true, the future of many amphibian species threatened by Bd appears bleak.
 
@@
 
+<!--# Add cross-references to the new Figure 1 in this and the preceding paragraph. -->
+
 Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
-This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], and is consistent with several non-mutually exclusive mechanisms.
-These include natural selection for more resistant/tolerant genotypes [@savage2016], acquired immunity [@grogan2018], and inherent between-populations differences in frog susceptibility that pre-date Bd exposure.
-Of these possibilities, evolution of resistance/tolerance would have the most general relevance for amphibian recovery from the effects of Bd, and as such additional investigation of this potential mechanism is critically needed.
-Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].
-
-<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->
+Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
+This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection (and not simply an effect of factors external to individual frogs, e.g., environmental conditions) <!--# Need to strengthen this point.-->.
+(We acknowledge that frogs from naive and recovering populations could also differ in their tolerance of Bd infection, but no data are currently available to evaluate this possibility.
+Therefore, we focus on resistance throughout this paper.) The observed resistance of MYL frogs could be the result of several non-mutually exclusive mechanisms, including natural selection for more resistant genotypes [@savage2016], acquired immunity [@grogan2018b], and/or inherent between-population differences in frog susceptibility that pre-date Bd exposure.
+The possibility that MYL frog resistance is the result of evolution, combined with the observed natural recovery of frogs is consistent with "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014].
+This intriguing possibility also suggets an opportunity to accelerate the current rate of recovery by utilizing resistant frogs from recovering populations in reintroductions to vacant habitats, actions that could potentially allow the reestablishment of extirpated frog populations despite the ongoing presence of Bd [@joseph2018].
 
 In the current study, we had three primary objectives.
-First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
-Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
-Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
-Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
+First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data from reintroduced populations, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we investigated whether MYL frogs show evidence of an evolutionary response to Bd exposre, and whether these genomic changes are associated with resistance and could provide a plausible mechanism whereby frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
+To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations and determined whether any of these regions contain genes associated with resistance.
+Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance to disease.
+Collectively, these results provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of the myriad species of wildlife endangered by escalating impacts from emerging infectious diseases.
+
+```{=html}
+<!--# Unused text from previous version of Introduction
+
+Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
+Although we know relatively little about the mechanisms underlying recovery from disease-caused declines [@brannelly2021; @russell2020], evolution of increased host resistance and/or tolerance [see @medzhitov2012 for definitions] is one possibility [@carlson2014; @searle2020].
+To reverse the escalating impacts of Bd on global amphibian biodiversity, the development of effective methods to recover impacted species is of the greatest importance.
+
+To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
+The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
+In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
+Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of imperiled taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
+
+This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
+In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
+To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
+(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., describing recovery of MYL frogs in the presence of Bd.) \--\>
 
@@
 Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
-
+```
 ### Results
@@
 Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
-Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018].
+Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
 This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [@robledo2020; @robledo2018].
@@
 
+\newpage
+
 ### Figures
@@
 
-![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana mucosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
-The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana mucosa* Contig19.
+\newpage
+
+![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana muscosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
+The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana muscosa* Contig19.
 Asterices indicate the location of SNP outliers in the TCF19 and VARS1 genes.
 Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot}
-
-\newpage
 
@@
 Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
-Samples sizes are provided just above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="700"}
+Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
 
@@
   mutate(n_period = sum(n), 
-         prop = (n / n_period)) %>% 
-  mutate(across(prop, round, 3)) %>% 
+         prop = round(n / n_period, 3)) %>% 
   arrange(period, survival)

### 2023-12-09 doc/manuscript/translocation.qmd
Edit Introduction and beginning of Results

@@
 bibliography: translocation.bib
+#  main: translocation.bib
+#  supporting: supporting_info.bib
+#filters: 
+#  - multibib
+#validate-yaml: false
 csl: pnas.csl
@@
 
-Human activities are increasingly impacting to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
+Human activities are increasingly impacting global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
 One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
@@
 By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
-
-Alarmingly, evidence of natural recovery in the many Bd-impacted amphibian populations is scarce [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggestive of generally low resilience to disease effects.
-Following pathogen arrival in a host population, the ability to either limit pathogen burden (resistance) or limit the harm caused by a particular burden (tolerance) are key mechanisms to reduce disease impacts [@råberg2009] and facilitate population persistence and recovery [@brannelly2021].
+These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding the key principles and mechanisms for maintaining species in the presence of emerging diseases [@russell2020].
+
+Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggesting that disease-caused declines will be difficult to reverse.
+This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
+Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@råberg2009] and facilitate population persistence and recovery [@brannelly2021].
 Host immunity and evolution both play important roles in the development of resistance and tolerance, and utilizing these factors would seem a promising approach to developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
@@
 These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
-Indeed, in a review of the effectiveness of mitigation methods, including reintroductions, Garner et al.
-[-@garner2016] came to the gloomy (but accurate) conclusion that, "Despite decades of research into amphibian-chytrid host--pathogen biology, no effective method to reduce the impact of chytridiomycosis (i.e., the disease caused by Bd) has emerged and been tested broadly in the field." If true, the future of many amphibian species threatened by Bd appears bleak.
-
-The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
+If true, the future of many amphibian species threatened by Bd appears bleak.
+
+The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by Bd [@scheele2019].
 Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
@@
 In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
-The arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
-These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
+The arrival of Bd in the 1960s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
+These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible [@knapp1998] and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
 In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
 
-<!--# Add cross-references to the new Figure 1 in this and the preceding paragraph. -->
-
-Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
-This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection (and not simply an effect of factors external to individual frogs, e.g., environmental conditions) <!--# Need to strengthen this point.-->.
-(We acknowledge that frogs from naive and recovering populations could also differ in their tolerance of Bd infection, but no data are currently available to evaluate this possibility.
-Therefore, we focus on resistance throughout this paper.) The observed resistance of MYL frogs could be the result of several non-mutually exclusive mechanisms, including natural selection for more resistant genotypes [@savage2016], acquired immunity [@grogan2018b], and/or inherent between-population differences in frog susceptibility that pre-date Bd exposure.
-The possibility that MYL frog resistance is the result of evolution, combined with the observed natural recovery of frogs is consistent with "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014].
-This intriguing possibility also suggets an opportunity to accelerate the current rate of recovery by utilizing resistant frogs from recovering populations in reintroductions to vacant habitats, actions that could potentially allow the reestablishment of extirpated frog populations despite the ongoing presence of Bd [@joseph2018].
+Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
+Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
+This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection (and not simply an effect of factors external to individual frogs, e.g., environmental conditions).
+In addition to frogs from recovering populations having higher resistance to Bd infection than those from naive populations, they could also have higher tolerance, but no data are currently available to evaluate this possibility.
+Therefore, we focus on resistance throughout this paper.
+The observed resistance of MYL frogs could be the result of several non-mutually exclusive mechanisms, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
+The evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
+This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by utilizing resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
 
 In the current study, we had three primary objectives.
-First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
-Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data from reintroduced populations, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
-Third, we investigated whether MYL frogs show evidence of an evolutionary response to Bd exposre, and whether these genomic changes are associated with resistance and could provide a plausible mechanism whereby frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
-To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations and determined whether any of these regions contain genes associated with resistance.
-Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance to disease.
-Collectively, these results provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of the myriad species of wildlife endangered by escalating impacts from emerging infectious diseases.
-
-```{=html}
-<!--# Unused text from previous version of Introduction
-
-Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
-Although we know relatively little about the mechanisms underlying recovery from disease-caused declines [@brannelly2021; @russell2020], evolution of increased host resistance and/or tolerance [see @medzhitov2012 for definitions] is one possibility [@carlson2014; @searle2020].
-To reverse the escalating impacts of Bd on global amphibian biodiversity, the development of effective methods to recover impacted species is of the greatest importance.
-
-To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
-The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
-In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
-Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of imperiled taxa, descriptions of its occurrence and effectiveness in the wild are lacking.
-
-This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
-In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
-To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
-(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., describing recovery of MYL frogs in the presence of Bd.) \--\>
-
-<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
-Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
-We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
-Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
-These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
-Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->
-```
+First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+Third, we investigated whether MYL frogs show evidence of an evolutionary response to Bd exposure, and whether these genomic changes are associated with resistance.
+To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model) and determined whether any of these regions contain genes associated with resistance.
+Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that MYL frogs have undergone substantial evolutionary change following Bd exposure and that some of the genomic regions under selection contain genes related to disease resistance and/or tolerance.
+Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and myriad other species of wildlife worldwide that are endangered by escalating impacts from emerging infectious diseases.
+
 ### Results
@@
 
+surveyyrs <- surveys %>% 
+  mutate(yr = year(ymd(survey_date))) %>%
+  filter(survey_type == "cmr") %>% 
+  distinct(site_id, yr) %>% 
+  count(site_id)
+
 frog_counts_gt0_nrow <- d2 %>% 
@@
 To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@seddon2014] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
-The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@berger2020] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
+The donor populations included 2 of the 5 recovering populations included in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
 Following translocation, adult survival and recruitment of new adults was estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
@@
 
+<!--# RESTART REVIEW HERE -->
+
+The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
+Given that it can take years or even decades to determine the self-sustainability of a reintroduced population [for an example in MYL frogs, see @joseph2018], the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
+Therefore, we conducted capture-mark-recapture (CMR) surveys of each reintroduced population to estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period following translocation (see **Materials and Methods - Frog population recovery - Field methods** for details).
 Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
-This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
-Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
+This suggests an important effect of site characteristics on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate) <!--# RESTART REVIEW HERE: Revise this table for clarity - see Reviewer 3 comment.
+-->, which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
+sierrae* <!--# Where did I mention what habitat factors are involved?
+-->.
+Finally, Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
 
@@
 Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
-Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.
+Therefore, frog resistance in non-natal habitats is maintained across generations.
 
@@
 When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
+Across all translocation sites, the duration of CMR time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
 
@@
 
+### References
+
+\newpage
+
 ### Figures
+
+![For *R.
+sierrae*, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
+Rectangles and hexagons represent outcomes and processes, respectives.
+Blue text indicates components that are included in the current study.
+The timeline shows the general sequence of the components, with the dotted portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model}
+
+\newpage
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------|:------------------|:-------------------------|
+|:-------------------------|:------------------|:--------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2023-12-10 doc/manuscript/translocation.qmd
Edits throughout Results and figure legends

@@
 
-The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by Bd [@scheele2019].
+The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global declines of amphibians caused by Bd [@scheele2019].
 Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
@@
 In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
-The arrival of Bd in the 1960s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
-These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible [@knapp1998] and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
+The arrival of Bd in the mid-1900s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
+These Bd-caused declines are fundamentally different from the fish-caused declines because fish eradication is feasible [@knapp1998] and results in the rapid recovery of frog populations [@knapp2007; @vredenburg2004].
 In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
@@
 The observed resistance of MYL frogs could be the result of several non-mutually exclusive mechanisms, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
-The evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
+The possible evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
 This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by utilizing resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
@@
 In the current study, we had three primary objectives.
-First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
+First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
 Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
@@
 To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model) and determined whether any of these regions contain genes associated with resistance.
-Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that MYL frogs have undergone substantial evolutionary change following Bd exposure and that some of the genomic regions under selection contain genes related to disease resistance and/or tolerance.
-Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and myriad other species of wildlife worldwide that are endangered by escalating impacts from emerging infectious diseases.
+Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that MYL frogs have undergone substantial evolutionary change following Bd exposure and that some of the genomic regions under selection contain genes related to disease resistance.
+Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and myriad other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
 
@@
 
+surveydates <- surveys %>% 
+  filter(survey_type == "cmr") %>% 
+  mutate(moday = format(as.Date(survey_date), "%b-%d"))
+
+translocationdates <- surveys %>% 
+  filter(survey_type == "translocation") %>% 
+  mutate(moday = format(as.Date(survey_date), "%b-%d"))
+
 frog_counts_gt0_nrow <- d2 %>% 
@@
 
-To determine whether frogs from recovering populations could be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them [@berger2020] to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
+To determine whether frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
 The donor populations included 2 of the 5 recovering populations included in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
-Following translocation, adult survival and recruitment of new adults was estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
-
-<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->
+Following translocation, adult survival and recruitment of new adults was estimated from capture-mark-recapture (CMR) surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
 
@@
 
-<!--# RESTART REVIEW HERE -->
-
 The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
 Given that it can take years or even decades to determine the self-sustainability of a reintroduced population [for an example in MYL frogs, see @joseph2018], the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
-Therefore, we conducted capture-mark-recapture (CMR) surveys of each reintroduced population to estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period following translocation (see **Materials and Methods - Frog population recovery - Field methods** for details).
-Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-This suggests an important effect of site characteristics on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate) <!--# RESTART REVIEW HERE: Revise this table for clarity - see Reviewer 3 comment.
--->, which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae* <!--# Where did I mention what habitat factors are involved?
--->.
-Finally, Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].
-
-To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
+Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
+Estimates of 1-year frog survival (from the site-specific mrmr models; see **Materials and Methods - Frog population recovery - Field methods**) indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier.
+Five of the 7 populations translocated after 2013 had estimated survival $\ge$ 0.5, compared to only 1 of 5 populations translocated prior to 2013.
+We suggest this resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
+sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
+Finally, Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
+Although it is possible that the observed relatively small change in load is a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
+Collectively, these results indicate that frogs show relatively high survival at the majority of recipient sites, site characteristics strongly affect survival, and Bd loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
+
+To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed a meta-analysis of the drivers of frog survival across all of our recipient sites.
+This analysis used linear and multilevel models, and included a diversity of frog, habitat, weather, and translocation characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
 Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
-Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
-Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize Bd load and/or the adverse effects of Bd infection.
-In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
-Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
-Therefore, frog resistance in non-natal habitats is maintained across generations.
+Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
+Together, these results are consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
+In addition, the moderate and relatively unchanged Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
+Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+This suggests that frog resistance in non-natal habitats is maintained across generations.
 
@@
 Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
-To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
-Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+To extend our inferences of population establishment beyond those possible from the 3--16 years of site-specific CMR data, we developed a population viability model.
+Specifically, to test whether the observed yearly adult survival probabilities in translocated populations (provided in legend of @fig-viability B) were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
 We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).
 
-Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
-
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
+
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
 Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
@@
 This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
+In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013.
 Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
+<!--# Given that we make a similar argument related to frog survival, the information in these last 2 sentences seems relatively obvious. Should we drop them? -->
 
@@
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-Bd-exposure histories of the nine study populations are based on 10-20 years of visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions \[during which Bd loads remain very low; wilber2022\], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
-In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally \< 100), high Bd prevalence (often \> 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
+Bd-exposure histories of the nine study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
+In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 100), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
 Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-(See SI Appendix for additional details regarding the study design.)\
-<!--In response to Editor - Comment 3, mention that there is no dispersal between extant populations due to presence of fish and fish-induced habitat fragmentation?-->
+Finally, there is no potential for frog dispersal between the nine study populations due to presence of introduced (predatory) fish and fish-induced habitat fragmentation.
+(See **Supporting Information** for additional details regarding the study design.)
 
@@
 
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A), and clusters reflected the species split (i.e., *R.
+Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A SI), and clusters reflected the species split (i.e., *R.
 muscosa* versus *R.
@@
 Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; SI Appendix: Datasets S1, S2).
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; Datasets S1, S2).
 One of the 7 identified genes (LOC108802036) did not have an associated annotation.
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI Appendix: Datasets S3, S4).
+A total of 33 annotated genes were found in the 9 shared regions (Datasets S3, S4).
 Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
@@
 Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
-Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
+Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
 The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
+Following release, translocated populations were visited approximately once per month during the summer active season (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
 CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
-We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
-Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
-We evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
-
-We conducted two analyses.
-In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
-Group-level effects included site_id or translocation_id.
-The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
+Included predictor variables described characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
+We used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
+Survival was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
-In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
-For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked convergence of the MCMC algorithm using trace plots and Rhat.
+We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022].
+For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
+We checked convergence of the MCMC algorithm using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
 
@@
 After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
-Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
+Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the first year after the translocation.
 We used an analysis workflow similar to that described above.
@@
 
-##### Sampling and Sequencing
+##### Sampling and sequencing
 
@@
 DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in Byrne et al. (in review).
+We sequenced the samples using an exome capture assay as described in [@byrne2023].
 Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
@@
 
-##### Data Pre-processing and Cleaning
+##### Data pre-processing and cleaning
 
@@
 
-##### Data Analysis
+##### Data analysis
 
@@
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals using VCFtools (v.0.1.16,) [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
+Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.
 Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
@@
 sierrae*, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
-Rectangles and hexagons represent outcomes and processes, respectives.
+Rectangles and hexagons represent outcomes and processes, respectively.
 Blue text indicates components that are included in the current study.
@@
 
-![Evidence for frog survival at the recipient sites following translocation.
-Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites.
+![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated from the mrmr model for each site.
 Error bars show the 95% uncertainty intervals.
@@
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="300"}
-
-\newpage
-
-![Evidence for reduced impacts of Bd on MYL frogs as indicated by the lack of an effect of Bd load on survival.
-Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+
+\newpage
+
+![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="450"}
-
-\newpage
-
-![Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="550"}
+
+\newpage
+
+![Results from population viability analysis: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
@@
 Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of \$\\sigma\_{J_1}\$.
+(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
@@
 The inset map locates the study area and range boundary in California.
-For cross-referencing of numbered points to populations/sites mentioned in the **Frog Population Recovery** section (above), site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="450"}
+For cross-referencing of numbered points to populations/sites mentioned in the text and figures, site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="600"}
 
@@
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="450"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="550"}
 
@@
 ### Figures
+
+![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon).
+Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Inset map shows the location of Yosemite within California.
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}
+
+![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
+Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
+Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
+Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
+Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="580"}
+
+![Evidence that Bd load is not an important predictor of post-translocation frog survival.
+Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables in the best model.
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
+Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="400"}
+
+![Comparison of average yearly adult survival probabilities for adults translocated to each of 3 sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
+In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
+Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
+All estimates were derived using the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
+
+<!--# Move some of these methods to a new section in Results - Frog population recovery. Reviewer 3 - Comment 5. -->
+
+![Sensitivity analysis of the stage-structured MYL frog model.
+Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year 1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year 2 juveniles).
+Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
 
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="450"}
+Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="550"}
 
@@
 
-![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon).
-Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}
-
-![For the best model describing the average 1-year survival of frogs in the first translocation to a site as a predictor of survival of frogs in all subsequent translocations to that site, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and single predictor variable.
-In the Bayesian framework in which the model was developed, a variable is considered an important predictor if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).](figures/mcmc_areas_m2b.png){#fig-transsurvival-postdens width="450"}
-
-![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats <!--# Consider edits to this to respond to Reviewer 2 - Comment 4.
-In Discussion?-->.
-For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
-Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}
-
-![Evidence that Bd load is not an important predictor of post-translocation frog survival.
-For the best model identifying predictors of 1-year frog survival following translocation, estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
-Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="350"}
-
-![Evidence that frog resistance/tolerance in non-natal habitats is maintained across generations.
-Comparison of the average yearly adult survival probabilities for adults translocated to each of 3 sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
-Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="450"}
-
-![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year 1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year 2 juveniles).
-Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
+\newpage
+
+![Violin plots showing individual heterozygosity for the Bd-naive and recovering sites included in the frog genomics study.
+Individual data points are shown on the plot with their corresponding site number (from @fig-allelefrequencies).
+A Wilcoxon signed-rank test showed no significant difference between the two groups.](figures/het_by_status_violin_plot.png){#fig-violinplot-heterozy width="500"}
+
+\newpage
+
+![For each study population (from @fig-allelefrequencies A), boxplots showing nucleotide diversity (pi) within 100kb sliding windows along the genome.
+The x-axis has been log~10~-transformed for display purposes.
+Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks as described in @fig-selectionresults A).
+Within the genetic clusters, mean pi is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).
+<!--# But these are tiny differences.
+Significance due to large samples sizes?
+Would it be better to describe the results qualitatively: Differences between sites were small, and did not show any consistent patterns between naive and recovering populations.
+-->](figures/boxplot_pi_by_site.png){#fig-boxplot-pi width="500"}
 
@@
 ### Tables
-
-```{r}
-#| label: tbl-survival-earlylate
-#| tbl-cap: "Association between the proportion of populations translocated early versus late in the study period (< 2013 or >= 2013, respectively) and probability of survival (< 0.5 or >= 0.5).
-Based on the viability analysis, survival probabilities < 0.5 and >= 0.5 produced 50-year extinction probabilities of 1 and < 0.5, respectively."
-#| echo: FALSE
-#| message: FALSE
-
-viability_table <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
-  distinct(site_id, year) %>% 
-  group_by(site_id) %>% 
-  filter(year == min(year)) %>% 
-  mutate(site_id = as.character(site_id),
-         survival = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), "< 0.5", ">= 0.5"),
-         period = if_else(year < 2013, "early", "late")) %>% 
-  group_by(period) %>% 
-  count(period, survival) %>% 
-  mutate(n_period = sum(n), 
-         prop = round(n / n_period, 3)) %>% 
-  arrange(period, survival)
-
-  knitr::kable(viability_table, format = "pipe", align = "lcccc", col.names = c("Period", "Survival probability", "Number of translocations", "Total translocations", "Proportion of translocations")) 
-```
-
-\newpage
 

### 2023-12-10 doc/manuscript/translocation.qmd
Add comment to genomics results
Improve between-section linkages in Results

@@
 In the current study, we had three primary objectives.
-First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
-Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
+First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort (@fig-recovery-model).
+Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
 Third, we investigated whether MYL frogs show evidence of an evolutionary response to Bd exposure, and whether these genomic changes are associated with resistance.
@@
 
-To determine whether frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+To determine whether MYL frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The donor populations included 2 of the 5 recovering populations included in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
-Following translocation, adult survival and recruitment of new adults was estimated from capture-mark-recapture (CMR) surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).
+The donor populations included 2 of the 5 recovering populations included in the frog evolution study summarized above (@fig-allelefrequencies A: population 1 and 4).
+Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
+The number survey years per site was a consequence of when translocations were conducted, and ranged from 2--16 years.
 
@@
 
+Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
+Although it is possible that the observed relatively small change in load is a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
+
 The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
@@
 Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
-Estimates of 1-year frog survival (from the site-specific mrmr models; see **Materials and Methods - Frog population recovery - Field methods**) indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier.
-Five of the 7 populations translocated after 2013 had estimated survival $\ge$ 0.5, compared to only 1 of 5 populations translocated prior to 2013.
+These estimates were made using site-specific models analyzed using the mrmr package (see **Materials and Methods - Frog population recovery - Field methods**).
+We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in a cross-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).
+
+Estimates of 1-year frog survival indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
+These patterns indicate an important effect of site characteristics on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in the study period than earlier: 5 of the 7 populations translocated after 2013 had estimated survival $\ge$ 0.5, compared to only 1 of 5 populations translocated prior to 2013.
 We suggest this resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
 sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
-Finally, Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
-Although it is possible that the observed relatively small change in load is a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
-Collectively, these results indicate that frogs show relatively high survival at the majority of recipient sites, site characteristics strongly affect survival, and Bd loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
-
-To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed a meta-analysis of the drivers of frog survival across all of our recipient sites.
-This analysis used linear and multilevel models, and included a diversity of frog, habitat, weather, and translocation characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
+This increased survival has direct implications for population viability (see **Results - Long-term population viability**).
+
+The goal of our meta-analysis was to identify important predictors of 1-year frog survival.
+We were particularly interested in whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
+This analysis was conducted in a Bayesian framework and included a diversity of frog, site, weather, and translocation characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
-Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
+Instead, important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
 Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
-Together, these results are consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
-In addition, the moderate and relatively unchanged Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
-This suggests that frog resistance in non-natal habitats is maintained across generations.
+The absence of Bd load as an important predictor of frog survival is consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
+
+In summary, results from our frog translocation study indicate that translocations resulted in (1) relatively high 1-year survival of translocated adults, reproduction, and recruitment at the majority of recipient sites, (2) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (3) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
+Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+This suggests that frog resistance is also maintained across generations.
+All of the conditions described above are supportive of population establishment and long-term population growth.
 
@@
 This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013.
-Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.
-<!--# Given that we make a similar argument related to frog survival, the information in these last 2 sentences seems relatively obvious. Should we drop them? -->
 
 #### Frog evolution in response to Bd
+
+Results from the preceding sections indicate the critical role of frog resistance in post-translocation frog survival, population growth, and population viability.
+As such, identifying the mechanisms underlying this resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
+Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs, the role of disease-mediated selection in these processes remains unknown.
 
@@
 Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
+<!--# Need to describe and reference Figures 14 SI and 15 SI. --> <!--# Add a few summary sentences to conclude this section.  -->
 
@@
 We checked convergence of the MCMC algorithm using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
+
+<!--# Add brief description of methods used in comparison of survival by translocated vs.
+naturally recruited individuals.
+Primary focus should be on describing that data were derived from mrmr model results, and on how those data were extracted.
+-->
 

### 2023-12-11 doc/manuscript/translocation.qmd
Provide the meta-analytic view, and minor edits in methods section

--- 2023-12-10 doc/manuscript/translocation.qmd
@@
 Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
+<!--# Need to describe and reference Figures 14 SI and 15 SI. -->
 
@@
 
-To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
-Included predictor variables described characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
-Survival was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
+To identify important predictors of frog survival following translocation, we used multilevel Bayesian models.
+We included predictor variables to describe characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
+We used 1-year post-translocation survival estimates from mark-recapture models as the response, which was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
+Survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
@@
 For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked convergence of the MCMC algorithm using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
+We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
 
@@
 
+Finally, while it would theoretically be possible to estimate survival covariate effects in a joint mark-recapture model that integrates capture histories across all sites, this was impractical due to computational requirements of the mark-recapture models (namely, run time and memory).
+In modeling survival estimates from site-specific mark-recapture analyses, we are in effect conducting an among-site meta-analysis to better understand how site, cohort, and individual level characteristics explain variation in survival after translocation.
+
 #### Population viability modeling

### 2023-12-11 doc/manuscript/translocation.qmd
Edit Discussion to address reviewer comments

--- 2023-12-10 doc/manuscript/translocation.qmd
@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The donor populations included 2 of the 5 recovering populations included in the frog evolution study summarized above (@fig-allelefrequencies A: population 1 and 4).
+The donor populations included 2 of the 5 recovering populations included in the frog evolution study summarized above (@fig-allelefrequencies A: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
@@
 
-Results from frog translocations indicated that most populations showed evidence of successful reproduction and recruitment and that adult survival was often relatively high (described above).
+Results from the frog translocation study indicated that most populations showed evidence of successful reproduction and recruitment, and that adult survival was often relatively high (described above).
 Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
@@
 Finally, there is no potential for frog dispersal between the nine study populations due to presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-(See **Supporting Information** for additional details regarding the study design.)
+(See **Supporting Information** for additional details regarding the study design.) <!--# Make sure that it is clear that 2 of our 3 donor populations were included in the genomics study.  -->
 
@@
 
-Disease-induced population declines are decimating global biodiversity [@daszak2000], but strategies to recover affected species are generally lacking.
-Here, we tested the possibility that populations of resistant/tolerant individuals could be used to reestablish extirpated populations.
+<!--# I kept the order of paragraphs from the original submission (evolution first, followed by translocations).
+This provides a more direct path to the argument that the MYL frog-Bd system is a likely example of evolutionary rescue.
+Does this order work?
+-->
+
+Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
+Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
 Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild [see @joseph2018 for a small-scale example].
-In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
-The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection contain genes related to disease resistance/tolerance, and (2) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing disease.
+Given the demonstrated resistance of adult frogs against Bd infection, and the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether the resulting adaptive alleles are associated with increased frog resistance.
+The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection contain genes related to disease resistance, and (2) reintroduction of individuals from these resistant populations allows the successful reestablishment of viable frog populations.
+In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
 
@@
 In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter].
-The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016].
-The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors, including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.
-
-Results from our genomic analyses strongly suggest that natural selection for adaptive alleles and resulting increased resistance/tolerance is at least partially responsible for the reduced susceptibility of frogs in recovering populations.
+The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
+This resistance could in theory be due to several factors, including natural selection for more resistant genotypes, acquired immunity, and/or inherent between-population differences that pre-date Bd exposure, but until now evidence to evaluate the role of evolution was lacking.
+
+Results from our genomic analyses strongly suggest that natural selection for adaptive alleles is at least partially responsible for the increased resistance of frogs in recovering populations.
 We identified specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
@@
 In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
-These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
+These results corroborate numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
 Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
-Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
+Several previous studies of amphibians have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
 This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [@robledo2020; @robledo2018].
@@
 Finally, based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
-This gene is associated with inflammation response and in *Xenopus* is highly expressed in wound stumps following amputation of tadpole tails [@fukazawa2009].
-Also in this analysis, the outlier variant with the lowest p-value (uncharacterized gene LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2) in another frog genome [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
-
-Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
+This gene is associated with inflammation response and in *Xenopus* is highly expressed in wounds [@fukazawa2009].
+In addition, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
+In the genome of another frog species, this gene is located adjacent to a type I interferon gene (Np-IFNi2) [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
+
+Reintroduction of resistant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
 Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
-Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
-Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
-The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
-In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
+Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
+The fact that the relatively low Bd loads and correspondingly high frog survival was maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by resistance inherent in the frogs.
+Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and translocated populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the translocated versus donor populations and pathogen pressure should follow a similar pattern.
+In addition to the maintenance of Bd load and frog survival between natal and translocation sites, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
+
 Results from the population viability model were also encouraging.
 In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
-In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
-<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->
-
-Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
+The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the 2-16 years covered by our post-translocation surveys.
+These long-term forecasts are important, given that reintroduced MYL frog populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
+Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is critically important to the process of adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
+Specifically, the combined results from our reintroduction study and viability modelling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
+In addition, given the repeatability of frog survival at a site, 1-year frog survival also serves as an effective proxy of site quality (i.e., the ability of a site to support high frog survival and a viable frog population over the long term).
+This proxy of site quality is important in the MYL frog system because accurately predicting the ability of a site to support a viable frog population *a priori* remains difficult, even after conducting 24 translocations over 16 years.
+
+Despite the demonstated resistance of adult MYL frogs against Bd infection, individual and population-level impacts of Bd are still evident.
 In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
@@
 In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
-MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality.
+MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality [@rachowicz2006].
 This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
 In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
-Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
+Therefore, although adult MYL frogs appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
 
 The recent emergence of Bd worldwide has contributed to the decline of hundreds of amphibian species, some of which are now extinct in the wild [@scheele2019].
-This unprecedented impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
+This extraordinary impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
 Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
 As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have acquired via evolution the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
-Despite the hopeful example of successful reestablishment of MYL frogs via evolutionary rescue and reintroductions, the challenge of recovering hundreds of Bd-impacted amphibian species worldwide is a daunting prospect.
+Despite the hopeful example of successful reestablishment of MYL frogs despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
 Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
 The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
-
-Evolutionary rescue may be increasingly critical for the survival of wildlife populations in a rapidly changing world.
-The importance of evolutionary rescue is known for a wide variety of taxa [e.g., @carlson2014; @mills2018; @oziolor2019], including in the specific context of disease-impacted systems [@searle2020].
-However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the stressor $\rightarrow$ positive population growth and recovery) is difficult in most systems.
-Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
-We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.
+<!--# It would be great to describe how reintroduction of the product of evolution - individuals with alleles that are adaptive in a novel environment - can allow recovery at a broader scale that natural recovery alone.
+Not sure we have space for this.
+-->
+
+In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
+Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an example of evolutionary rescue will likely always be uncertain.
+Regardless, we provide a compelling example from the wild that suggests that evolution can produce individuals that harbor adaptive alleles and allow population recovery in a novel environment, and show conclusively that individuals from these recovering populations can be used in translocations to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
+We expect that such translocations will be an essential tool in wildlife conservation in an era of accelerating global change.
 
@@
 Included predictor variables described characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
+We used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 2--16).
 Survival was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
@@
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals using VCFtools (v.0.1.16,) [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
+To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
 Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.

### 2023-12-11 doc/manuscript/translocation.qmd

@@
 
-To identify important predictors of frog survival following translocation, we developed linear and multilevel models in a Bayesian framework.
-Included predictor variables described characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 2--16).
-Survival was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
+To identify important predictors of frog survival following translocation, we used multilevel Bayesian models.
+We included predictor variables to describe characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
+We used 1-year post-translocation survival estimates from CMR models as the response, which was available for all translocated populations regardless of how many years of post-translocation data were available (range = 2--16).
+Survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
 Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.

### 2023-12-11 doc/manuscript/translocation.qmd
Minor changes to Frog recovery methods
Additional minor edits to Frog recovery methods

@@
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
-The number survey years per site was a consequence of when translocations were conducted, and ranged from 2--16 years.
+Across all translocation sites, the duration of survey time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
 
@@
 For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
-The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
+The populations that we selected contained hundreds to thousands of both *R. sierrae* adults and tadpoles.
 These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
@@
 At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics.
-In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks).
+In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwinter habitats such as submerged boulders and overhanging banks.
 *R. sierrae* were absent from all recipient sites prior to the first translocation.
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, translocated populations were visited approximately once per month during the summer active season (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
+Following release, we visited translocated populations approximately once per month during the summer active season and conducted diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
 CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
+During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we search all habitats repeatedly for adult frogs.
 Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
@@
 When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
-Across all translocation sites, the duration of CMR time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
 
@@
 For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
-We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
+We evaluated Markov chain Monte Carlo (MCMC) convergence using trace plots and Gelman-Rubin statistics (Rhat).
 

### 2023-12-11 doc/manuscript/translocation.qmd
More edits to Frog recovery methods

@@
 For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
-We evaluated Markov chain Monte Carlo (MCMC) convergence using trace plots and Gelman-Rubin statistics (Rhat).
+We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
+<!--# I thought Max edited this sentence to get rid of "algorithm" but I must be mis-remembering. I checked his explain-meta-analysis branch at the above sentence is repeated there.  -->
+
+<!--# Mark: Add brief description here of methods used in comparison of survival by translocated vs.
+naturally recruited individuals.
+Primary focus should be on describing that fact that data were derived from mrmr model results, and on how those data were extracted.
+-->
 
@@
 
-<!--# Add brief description of methods used in comparison of survival by translocated vs.
-naturally recruited individuals.
-Primary focus should be on describing that data were derived from mrmr model results, and on how those data were extracted.
--->
-
 ##### Changes in Bd load following translocation
@@
 Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the first year after the translocation.
-We used an analysis workflow similar to that described above.
-Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
-We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
-All models used a negative binomial family.
-Priors, chains, iterations, and checks were as described above.
 

### 2023-12-11 doc/manuscript/translocation.qmd
Address issue #57 regarding naturally-recruited survival

@@
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper and Appendi).
 For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
 For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in *Mark-recapture model structure*), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
+.
+This yielded posterior distributions for survival probabilities between primary periods.
+However, because the time interval between primary periods differed, the survival probabilities between different consecutive primary periods were not directly comparable.
+To address this, we converted the survival probabilities between each consecutive primary period to per day death rates, propagating the uncertainty from the posterior distributions.
+We then took a weighted average of these death rates, weighted by the time interval between primary periods, to get the average per day death rate over the entire CMR survey.
+We converted this per day death rate $d$ back to a yearly survival probability using $\exp(-d \times 365 \text{ days})$.
+We used the same procedure for $\sigma_{A_T}$ such that our estimates of average yearly survival probability were comparable between $\sigma_{A_R}$ and $\sigma_{A_T}$.
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
+All estimates were derived using the mrmr package and the methods for calculation is described in Population viability modeling -- Estimating model parameters in the SI.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
 

### 2023-12-11 doc/manuscript/translocation.qmd
Edits primarily to Supporting Information

--- 2023-12-11 doc/manuscript/translocation.qmd
@@
 Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
-These estimates were made using site-specific models analyzed using the mrmr package (see **Materials and Methods - Frog population recovery - Field methods**).
-We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in a cross-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).
+These estimates were made using site-specific models analyzed using the mrmr package.
+We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in an among-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).
 
@@
 We were particularly interested in whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
-This analysis was conducted in a Bayesian framework and included a diversity of frog, site, weather, and translocation characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
+This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
@@
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI; see **SI Population viability modeling - Estimating model parameters** for details).
 This suggests that frog resistance is also maintained across generations.
@@
 Finally, there is no potential for frog dispersal between the nine study populations due to presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-(See **Supporting Information** for additional details regarding the study design.) <!--# Make sure that it is clear that 2 of our 3 donor populations were included in the genomics study.  -->
+(See **Supporting Information - Frog evolution in response to Bd - Methods** for additional details regarding the study design.) <!--# Make sure that it is clear that 2 of our 3 donor populations were included in the genomics study.
+-->
 
@@
 This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
+(See **Supporting Information - Frog evolution in response to Bd - Results** for additional details.)
 
@@
 
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022] (see **Supporting Infomation - Frog population recovery - Mark-recapture model structure** for details).
 The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-<!--# I thought Max edited this sentence to get rid of "algorithm" but I must be mis-remembering. I checked his explain-meta-analysis branch at the above sentence is repeated there.  -->
+<!--# I thought Max edited this sentence to get rid of "algorithm" but I must be mis-remembering. I checked his explain-meta-analysis branch and the above sentence is repeated there.  -->
 
@@
 naturally recruited individuals.
-Primary focus should be on describing that fact that data were derived from mrmr model results, and on how those data were extracted.
+Primary focus should be on describing the fact that data were derived from mrmr model results, and on how those data were extracted.
 -->
@@
 We included predictor variables to describe characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year post-translocation survival estimates from CMR models as the response, which was available for all translocated populations regardless of how many years of post-translocation data were available (range = 2--16).
+We used 1-year post-translocation survival estimates from CMR models as the response, which was available for all translocated populations regardless of how many years of post-translocation data were available.
 Survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
@@
 We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
+(See **Supporting Information - Frog population recovery - Among-site survival modeling** for details.)
 
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see **Supporting Information-Frog population recovery - Laboratory methods** for details].
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile \$p\_{J1}\$.
-First year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced environmental stochasticity.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
+First-year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced environmental stochasticity.
 
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-Additional details are provided in the SI Appendix.
+Additional details are provided in **Supporting Information - Population viability modeling**.
 
@@
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-Additional details are provided in the SI Appendix.
+Additional details are provided in **Supporting Information - Population viability modeling**.
 Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
@@
 DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in [@byrne2023].
+We sequenced the samples using an exome capture assay as described in @byrne2023.
 Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
@@
 
-To visualize the genomic relationships of our populations we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
+To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
 To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
@@
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and SI Appendix: Dataset S1, S2).
+We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and **Supporting Information**: Dataset S1, S2).
 
@@
 Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (SI Appendix: Datasets S3, S4).
+Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (**Supporting Information**: Datasets S3, S4).
 
@@
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
+To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals <!--# At the beginning of this section, we state that 53 individuals were sampled.
+I assume the difference reflects the fact that some individuals were excluded by the filtering process?
+Clarify where 50 came from.
+-->using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
 Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.
@@
 
-We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
+We estimated survival and recruitment for each site using open population mark-recapture models based on @joseph2018.
 For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
@@
 The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation.
-We generated these point estimates by rounding the posterior median of 1-year post introduction survival for each individual, and modeled the data using a Bernoulli distribution:
+We generated these point estimates by rounding the posterior median of 1-year post-introduction survival for each individual, and modeled the data using a Bernoulli distribution:
 
@@
 We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mark-recapture models.
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability ([@fig-translocation-survival]).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (@fig-translocation-survival).
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in "Estimating model parameters" below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment \$\\sigma\_{J_2}\$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$
+Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
+We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$.
 
@@
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see "Materials and Methods - Estimation of frog survival and abundance" in main paper).
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ ([@fig-compare_surv_probs]).
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Estimation of frog survival and abundance** in main paper).
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs SI).
 For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
@@
 
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
+Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp SI).
 
@@
 In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9; @fig-allelefrequencies, @fig-selectionresults A).
+However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9 <!--# Populations 5-9?
+-->; @fig-allelefrequencies, @fig-selectionresults A SI).
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.

### 2023-12-11 doc/manuscript/translocation.qmd
Minor fix to viability methods

@@
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Estimation of frog survival and abundance** in main paper).
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Estimation of frog survival and abundance** and **Supporting Information - Frog population recovery - Estimation of frog survival and abundance** for details).
 For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs SI).
 For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
+
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in *Mark-recapture model structure*), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
+.
+This yielded posterior distributions for survival probabilities between primary periods.
+However, because the time interval between primary periods differed, the survival probabilities between different consecutive primary periods were not directly comparable.
+To address this, we converted the survival probabilities between each consecutive primary period to per day death rates, propagating the uncertainty from the posterior distributions.
+We then took a weighted average of these death rates, weighted by the time interval between primary periods, to get the average per day death rate over the entire CMR survey.
+We converted this per day death rate $d$ back to a yearly survival probability using $\exp(-d \times 365 \text{ days})$.
+We used the same procedure for $\sigma_{A_T}$ such that our estimates of average yearly survival probability were comparable between $\sigma_{A_R}$ and $\sigma_{A_T}$.
 

### 2023-12-11 doc/manuscript/translocation.qmd
Fix typo

@@
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI; see **SI Population viability modeling - Estimating model parameters** for details).
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI; see **Supporting Information - Population viability modeling - Estimating model parameters** for details).
 This suggests that frog resistance is also maintained across generations.

### 2023-12-12 code/extinction_and_growth_rate_analysis.R
Increase font size for Fig. 4 legends. Issue #59

@@ -133,8 +133,8 @@ ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) +
 							annotate("text", x=0.25, y=0.05, label="Population declines", size=3) +
 							annotate("text", x=0.7, y=0.2, label="Population grows", size=3) +
 							guides(color="none", fill=guide_colorbar(title="\u03bb")) + theme(legend.position="bottom", 
-																					 legend.text=element_text(size=7), 
-																					 legend.title=element_text(size=7))
+																					 legend.text=element_text(size=10), 
+																					 legend.title=element_text(size=10))
 
 
 #############################################################
@@ -340,7 +340,8 @@ pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=sigma_J1, 
 							scale_color_manual(values=cdat$col) +
 							theme_classic() + xlab(bquote("Mean year 1 juvenile survival, \u03c3"~.[J1])) + ylab("Extinction prob. in 50 years") +
 							guides(color=guide_legend(title=bquote("Site, \u03c3"~.[AR]), title.position="top", ncol=3)) + 
-							theme(legend.position="bottom", legend.text=element_text(size=7), legend.title=element_text(size=8))
+							theme(legend.position="bottom", legend.text=element_text(size=9.5), legend.title=element_text(size=10))
+
 
 
 # Save comparison between translocated and survival probabilities from all individuals
@@ -460,7 +461,7 @@ ptraj = ggplot() + geom_line(data=pred_traj, aes(x=year, y=abund, group=sim, col
 										theme_classic() + xlab("Time (year)") + ylab("Adult abundance") +
 										scale_x_continuous(breaks=seq(2006, 2021, 2)) + 
 										theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5),
-											   legend.position="bottom", legend.text=element_text(size=8), legend.title=element_blank())
+											   legend.position="bottom", legend.text=element_text(size=10), legend.title=element_blank())
 
 # Plot and compare observed and predicted recruitment
 # matplot(t(recruitment_trajectories[['70550']][pred_ss_dt$index[1:100], 1:15]), type="l", ylab="Recruitment", xlab="Time (years)")

### 2023-12-12 doc/manuscript/translocation.qmd
Minor text clean-up

@@
 
-<!--# Revise to reflect reordering of sections -->
-
-Vast alteration of the biosphere by humans is causing a sixth mass extinction.
-Adaptation to modified environments often is the only means for species to persist, but the extent to which such "evolutionary rescue" can prevent or reverse biodiversity loss is largely unknown.
-Using results from genomic analyses, a 15-year reintroduction effort, and population modeling, we provide a compelling example of evolution reversing the declines of an imperiled amphibian and allowing the reestablishment of extirpated populations.
-The once-common mountain yellow-legged (MYL) frog is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
-Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
-Using an exome capture approach, we show that MYL frogs have undergone substantial evolutionary change following disease outbreaks.
-Moreover, in naturally recovering populations observed changes in immune function loci are associated with increased resistance/tolerance to Bd infection.
-Large-scale reintroduction of frogs from rescued populations resulted in the establishment of reproducing populations despite ongoing disease.
-In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
-Collectively, these results provide a rare example of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
-This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.
+Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in emerging infectious diseases.
+The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
+Given that there are no broadly applicable methods to reverse these impacts in the wild, the future of many amphibians appears grim.
+The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
+Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
+Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes and/or acquired immunity.
+We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
+In addition, results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
+To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
+We show that frogs have undergone substantial evolutionary change in response to Bd exposure and that changes in immune function loci are consistent with increased resistance.
+These results provide a rare example of how the reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species.
+This example has broad implications for the many taxa worldwide that are threatened with extinction by disease.
 
@@
 
-<!--# Revise to reflect reduced emphasis on evolutionary rescue. -->
-
 Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-One possibility is rapid evolution that restores positive growth to declining populations ("evolutionary rescue").
-Individuals from these rescued populations could also be used to reestablish extinct populations.
-We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse disease impacts.
-We found that frogs have evolved increased resistance to infection, facilitating population recovery.
-In addition, introduction of frogs from these populations allowed reestablishment despite the ongoing presence of disease.
-Such reintroductions could play an important role in biodiversity conservation in our rapidly changing world.
+Emerging infectious diseases can have particularly devastating impacts, and few options typically exist to reverse these effects.
+We used large-scale reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to the brink of extinction by a disease that has recently devastated global amphibian biodiversity.
+Introduction of resistant frogs allowed reestablishment of populations in the presence of disease.
+In addition, resistance appears to have a genetic basis and to have evolved recently following exposure of frogs to Bd.
+The evolution of resistance and reintroductions of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
 
 ### Introduction
-
-<!--# Remove uncited references in .bib -->
 
@@
 <!--# I thought Max edited this sentence to get rid of "algorithm" but I must be mis-remembering. I checked his explain-meta-analysis branch and the above sentence is repeated there.  -->
-
-<!--# Mark: Add brief description here of methods used in comparison of survival by translocated vs.
-naturally recruited individuals.
-Primary focus should be on describing the fact that data were derived from mrmr model results, and on how those data were extracted.
--->
 
@@
 I assume the difference reflects the fact that some individuals were excluded by the filtering process?
-Clarify where 50 came from.
+Clarify where 50 comes from.
 -->using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.

### 2023-12-12 doc/manuscript/translocation.qmd
Resize small figures, add revised/new figures

@@
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="600"}
+(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="650"}
 
@@
 Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
-Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="580"}
+Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
 
@@
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
-Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="400"}
+Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="500"}
 
@@
 Asterices indicate the location of SNP outliers in the TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot}
+Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="650"}
 
@@
 -->](figures/boxplot_pi_by_site.png){#fig-boxplot-pi width="500"}
+
+![TO BE ADDED](figures/perwindow_persite_pi.png){width="650"}
 

### 2023-12-13 doc/manuscript/translocation.qmd
Final edits to draft revision

@@
 
-**Key words:** evolutionary rescue, amphibians, population recovery, reintroduction, disease, *Batrachochytrium dendrobatidis*
+**Key words:** evolution, resistance, amphibians, population recovery, reintroduction, disease, *Batrachochytrium dendrobatidis*
 
@@
 In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
+This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
 
@@
 
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
+To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog exomes (i.e., coding region of genome) between populations with contrasting histories of Bd exposure.
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
 Bd-exposure histories of the nine study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
+Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
 In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 100), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
@@
 We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS).
+First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
 Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
@@
 muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
+sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
 Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
 
-Results from the individual variant and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C; Datasets S1, S2).
+Results from the individual variant and splined window analyses show that recovering populations have signatures of selection on multiple regions of the genome.
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI; SI Datasets S1, S2).
 One of the 7 identified genes (LOC108802036) did not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
+For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 0.41 to 0.86.
 Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of strong - and parallel - selection at this locus associated with exposure of frogs to Bd across multiple populations.
-(See **Supporting Information - Frog evolution in response to Bd - Results** for additional details.)
+This is suggestive of parallel selection at this locus associated with exposure of frogs to Bd across multiple populations.
+The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that suggest parallel evolutionary changes following Bd exposure.
 
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (Datasets S3, S4).
+A total of 33 annotated genes (<!--# Is this correct, i.e., that we identified 33 outlier regions, and in the subset of 9 regions we found 33 annotated genes?
+I just want to make sure there isn't a typo in here.
+Seems a weird coincidence to have 33 of both.) And the file spline_window_shared_outliers.csv lists 35 genes.
+-->were found in the 9 shared regions (Datasets S3, S4).
 Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
@@
 Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot).
+Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
 This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
@@
 Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-<!--# Need to describe and reference Figures 14 SI and 15 SI. --> <!--# Add a few summary sentences to conclude this section.  -->
+
+In summary, these results indicate that the exomes of frogs from our naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
+The regions putatively under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
+
+<!--# Allie, I opted not to add the additional text and figures that you provided to this Results section (i.e., related to heterozygosity and pi).
+Instead, I mention the analyses in the "Materials and Methods - Frog evolution - Data analyis" section (along with a comment for you that provides a justification for my decision), and provide the analysis description and figures in "Supporting Information - Frog evolution".
+Let me know if that works for you.
+-->
 
@@
 
-<!--# I kept the order of paragraphs from the original submission (evolution first, followed by translocations).
-This provides a more direct path to the argument that the MYL frog-Bd system is a likely example of evolutionary rescue.
+<!--# In the Discussion, I retained the general order of paragraphs from the original submission (evolution first, followed by translocations).
+This aligns with the actual trajectory that we are proposing in Figure 1 (Bd epizootic - selection in response to Bd - frog recovery.
 Does this order work?
@@
 Given the demonstrated resistance of adult frogs against Bd infection, and the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether the resulting adaptive alleles are associated with increased frog resistance.
-The results of our study provide evidence that (1) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease and that some of the genomic regions under selection contain genes related to disease resistance, and (2) reintroduction of individuals from these resistant populations allows the successful reestablishment of viable frog populations.
+The results of our study show that (1) reintroduction of resistant MYL frogs from naturally recovering populations allows the successful reestablishment of viable frog populations, and (2) that frogs in recovering populations have likely experienced selection following exposure to Bd and that some of the genomic regions identified as putatively under selection contain genes related to disease resistance.
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
@@
 Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
-In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter].
+In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter SI].
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
@@
 
-Results from our genomic analyses strongly suggest that natural selection for adaptive alleles is at least partially responsible for the increased resistance of frogs in recovering populations.
-We identified specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
+Results from our genomic analyses suggest that natural selection for adaptive alleles is at least partially responsible for the increased resistance of frogs in recovering populations.
+We identified specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with selection following Bd exposure.
 Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
@@
 This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x.
+Finally, we trimmed samples with an average depth across filtered sites \< 7x (n = 3).
 Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
@@
 Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
+We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome (for more efficient data processing and to better contextualize the strength of the outlier signals in different regions of the genome).
 We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
+This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the spline fitted to *F~ST~* and $\pi_{diff}$ values along the genome, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
 This method also calculates a W-statistic allowing for outlier identification.
 We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
+These standard deviations represent strict criteria to select only the top \~ 0.3% of windows.
 Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
@@
 We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals <!--# At the beginning of this section, we state that 53 individuals were sampled.
-I assume the difference reflects the fact that some individuals were excluded by the filtering process?
-Clarify where 50 comes from.
--->using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
-Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.
+
 Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
+
+The joint processes of Bd-caused frog population declines and selection in response to Bd exposure could both affect genetic diversity of recovering populations.
+Comparisons between naive and recovering populations of (i) individual-level heterozygosity, and (ii) population-level $\pi$ are provided in **Supporting Information - Frog evolution in response to Bd**.
+<!--# Allie, instead of mentioning these results and referencing the associated figures in the main Results (as you suggested), I added them to the "SI - Frog evolution in response to Bd section".
+I did so for 2 reasons.
+First, the results are somewhat tangential to the main focus of this paper (frog recovery), and second, citing some of the figures where you suggested (in the Results above) would mean referring to results before those results were described (e.g., sliding window analysis and associated pi boxplots and barcharts).
+Doing so would have been confusing.
+-->
+
+###
 
@@
 
-\newpage
+See combined references at end of document (temporary placeholder).
+
+\
 
@@
 
-![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated from the mrmr model for each site.
+![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated for each site from the mrmr model.
 Error bars show the 95% uncertainty intervals.
@@
 
-![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability).
+![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability), as estimated from the model used in the among-site meta-analysis.
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
@@
 
-\newpage
-
 ### Frog evolution in response to Bd
@@
 #### Methods
+
+##### Study design
 
@@
 
-#### Results
+##### GO analysis
 
@@
 The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Our analysis found no significantly over-represented GO terms in this set of 30 genes, or in the set of 38 outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and recovering populations, we detected consistent allele frequency changes across all genetic clusters for only 1 of 7 of the associated genes (RIN3; @fig-allelefrequencies).
-However, for the other 6 genes we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 6--9 <!--# Populations 5-9?
--->; @fig-allelefrequencies, @fig-selectionresults A SI).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe parallel evolutionary changes following Bd exposure and are therefore interesting candidates for future studies.
+Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
+Next, we compared the 30 unique outlier genes from the liberal set of 38 outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+We then repeated this process for the set of 33 genes located in the overlapping F~ST~ and splined windows.
+
+##### Genetic diversity
+
+To characterize general patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity using VCFtools [@danecek2011], and compared average heterozygosity using a Wilcoxon rank-sum test<!--# This analysis is problematic because it doesn't account for the lack of independence between points (i.e., individuals grouped within populations).
+In addition, there is a pattern of R.
+muscosa populations in both groups (7, 8, 9) having low Fst, so species identity is another grouping that contributes to a lack of independence.
+Given that we don't have a large enough sample size to run an analysis that properly accounts for this lack of independence, I'd argue for dropping this analysis and providing only a visual description of the differences.
+-->.
+Additionally, we used VCFtools to calculate per-population $\pi$ in 100kb sliding windows along the genome, and in each of the 10 <!--# In the main manuscript, we state that 9 windows were identified.
+Why the difference?
+-->outlier windows identified in the splined window analysis.
+We compared per-population average $\pi$ values using Wilcoxon rank-sum tests.
+<!--# Allie, did I describe this correctly?
+--> <!--# The displayed per-population pi values show tiny differences between naive and persistent populations.
+Is the significance shown by the Wilcoxon test due to large sample sizes?
+Given that the interest here is in describing any meaningful differences between naive and persistent populations, it doesn't accomplish much to highlight tiny differences that almost certainly have no biological relevance.
+In addition, all of the lack of independence issues that I highlighted above are probably relevant here too.
+Would it be better to describe the results qualitatively?
+E.g., Differences between sites were small and did not show any consistent patterns between naive and recovering populations.
+-->
+
+#### Results
+
+##### GO analysis
+
+We found no significantly over-represented GO terms in the liberal set of outlier regions, or in the set of 33 genes located in the overlapping F~ST~ and splined windows.
+Therefore.....<!--# Allie, check wording for accuracy, add one concluding sentence about what this result means. -->
+
+##### Genetic diversity
+
+Average (i) individual-level heterozygosity and (ii) population-level nucleotide diversity ($\pi$) were similar between naive and recovering populations (@fig-violinplot-heterozy SI; @fig-boxplot-pi SI).
+The analysis of $\pi$ across the 10 <!--# 9?
+-->outlier windows indicates that within each of the windows there is considerable variation in $\pi$ between populations and between population clusters (@fig-barchart-pi-by-windowsite SI).
+<!--# Allie, please check that my text accurately describes the results. -->
 
@@
 ### Figures
+
+\newpage
 
@@
 
+\newpage
+
 ![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
@@
 
+\newpage
+
 ![Evidence that Bd load is not an important predictor of post-translocation frog survival.
-Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables in the best model.
+Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model in the between-site meta-analysis.
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
 Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="500"}
+
+\newpage
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package and the methods for calculation is described in Population viability modeling -- Estimating model parameters in the SI.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
-
-<!--# Move some of these methods to a new section in Results - Frog population recovery. Reviewer 3 - Comment 5. -->
+All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling -- Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
+
+\newpage
 
@@
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
+
+\newpage
 
@@
 
-![Violin plots showing individual heterozygosity for the Bd-naive and recovering sites included in the frog genomics study.
+![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog genomics study.
 Individual data points are shown on the plot with their corresponding site number (from @fig-allelefrequencies).
@@
 
-![For each study population (from @fig-allelefrequencies A), boxplots showing nucleotide diversity (pi) within 100kb sliding windows along the genome.
-The x-axis has been log~10~-transformed for display purposes.
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks as described in @fig-selectionresults A).
-Within the genetic clusters, mean pi is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).
-<!--# But these are tiny differences.
-Significance due to large samples sizes?
-Would it be better to describe the results qualitatively: Differences between sites were small, and did not show any consistent patterns between naive and recovering populations.
+![For each study population in the frog evolution study (from @fig-allelefrequencies A), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
+The y-axis has been log~10~-transformed for display purposes.
+Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).
+Within the genetic clusters, mean $\pi$ is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).<!--# Change x-axis label to "Population" and x-axis tick labels to population #s only.
 -->](figures/boxplot_pi_by_site.png){#fig-boxplot-pi width="500"}
 
-![TO BE ADDED](figures/perwindow_persite_pi.png){width="650"}
+\newpage
+
+![For each study population in the frog evolution study, average $\pi$ values by population calculated for each of the 10 outlier windows <!--# I thought it was 9?
+-->identified in the splined window analysis (A-J).
+Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI.
+Gene annotations for the outlier regions are as follows: (A) C6, C7, (B) DDX10, ZBTB24, (C) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (D) FLOT1, TUBB2B, TUBB, MDC1, CCHCR1, TCF19*, TRIM29, HSP70, LSM2, VARS,* (E) GCC2, CFAP251, PEG10, (F) ERO1A, GVINP1, (G) ERO1A, GVINP1, (H) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (I) C6H5ORF22, PKS6, BSPRY, MPV17, and (J) CAD, ATRAID, GPN1.
+Note that windows F and G are in the same gene region.
+\* by gene names indicate that an outlier SNP is also present in this gene <!--# No asterices are associated with any of the gene names listed above -->.
+(K) Average $\pi$ values of naive and recovering groups for each of the outlier windows shown in A-J.
+<!--# These are mean values so they should be displayed using boxplots.
+Not providing that context can be misleading.
+Given the variation between populations for each window, I don't think (K) is useful even if boxplots were used.
+I think plots A-J tell the story and K doesn't add anything.
+--> <!--# Put plot identifiers A-K inside the upper-right corner of each plot.
+Cleaner and less ambiguous than having identifiers between plots.
+Alternatively, the identifier for plot K could be placed inside the upper-left corner of that plot.-->](figures/perwindow_persite_pi.png){#fig-barchart-pi-by-windowsite width="650"}
 

### 2023-12-13 doc/manuscript/translocation.qmd
Adjust figure name

--- 2023-12-12 doc/manuscript/translocation.qmd
@@
 We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs]).
+We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] SI).
 

### 2023-12-19 doc/manuscript/translocation.qmd
Minor edits throughout main manuscript

@@
 To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
-We show that frogs have undergone substantial evolutionary change in response to Bd exposure and that changes in immune function loci are consistent with increased resistance.
-These results provide a rare example of how the reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species.
-This example has broad implications for the many taxa worldwide that are threatened with extinction by disease.
+We found substantial differences between these categories, including changes in immune function loci that may confer increased resistance, consistent with evolutionary changes in response to Bd exposure.
+These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species.
+This example has broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
 
@@
 Introduction of resistant frogs allowed reestablishment of populations in the presence of disease.
-In addition, resistance appears to have a genetic basis and to have evolved recently following exposure of frogs to Bd.
+In addition, resistance appears to have a genetic basis and is consistent with evolution following exposure of frogs to Bd.
 The evolution of resistance and reintroductions of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
@@
 By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
-These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding the key principles and mechanisms for maintaining species in the presence of emerging diseases [@russell2020].
+These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding mechanisms of species persistence in the presence of emerging diseases [@russell2020].
 
@@
 Host immunity and evolution both play important roles in the development of resistance and tolerance, and utilizing these factors would seem a promising approach to developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
-However, several aspects of the amphibian-Bd system present difficult obstacles, including (1) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (2) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
+However, several aspects of the amphibian-Bd system present difficult obstacles, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
 These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
@@
 To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model) and determined whether any of these regions contain genes associated with resistance.
-Our results indicate (1) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (2) that the recovering populations are likely to persist over a 50-year period, and (3) that MYL frogs have undergone substantial evolutionary change following Bd exposure and that some of the genomic regions under selection contain genes related to disease resistance.
-Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and myriad other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
 
@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The donor populations included 2 of the 5 recovering populations included in the frog evolution study summarized above (@fig-allelefrequencies A: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
+The donor populations included 2 of the 5 recovering populations included in the frog evolution study referenced above (@fig-allelefrequencies: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
@@
 
-Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
+Of the `r nrow(d2)` reintroduced populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
 For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
@@
 Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
-Although it is possible that the observed relatively small change in load is a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
+Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
 
@@
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
-Instead, important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
+Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, @fig-survival-postdens SI).
 Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
@@
 
-In summary, results from our frog translocation study indicate that translocations resulted in (1) relatively high 1-year survival of translocated adults, reproduction, and recruitment at the majority of recipient sites, (2) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (3) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
+In summary, results from our frog translocation study indicate that translocations resulted in (i) relatively high 1-year survival of translocated adults, as well as reproduction and recruitment, at the majority of recipient sites, (ii) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (iii) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI; see **Supporting Information - Population viability modeling - Estimating model parameters** for details).
-This suggests that frog resistance is also maintained across generations.
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+This suggests that frog resistance is maintained across generations.
 All of the conditions described above are supportive of population establishment and long-term population growth.
@@
 Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
-To extend our inferences of population establishment beyond those possible from the 3--16 years of site-specific CMR data, we developed a population viability model.
+To extend our inferences of population establishment beyond those possible from the site-specific CMR data, we developed a population viability model.
 Specifically, to test whether the observed yearly adult survival probabilities in translocated populations (provided in legend of @fig-viability B) were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).
+We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
 
@@
 Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 <!--# Why not use 0.09 like in A panel?
+-->(@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year 1 juvenile survival between 0 and 0.25.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
 As such, identifying the mechanisms underlying this resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
-Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs, the role of disease-mediated selection in these processes remains unknown.
+Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs the role of disease-mediated selection in these processes remains unknown.
 
@@
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-Bd-exposure histories of the nine study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
+Bd-exposure histories of the 9 study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
 Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
@@
 Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-Finally, there is no potential for frog dispersal between the nine study populations due to presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-(See **Supporting Information - Frog evolution in response to Bd - Methods** for additional details regarding the study design.) <!--# Make sure that it is clear that 2 of our 3 donor populations were included in the genomics study.
--->
-
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
+Finally, there is no potential for frog dispersal between the 9 study populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
+(See **Supporting Information - Frog evolution in response to Bd - Methods** for additional details regarding the study design.)
+
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions putatively under selection).
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
@@
 The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI; SI Datasets S1, S2).
-One of the 7 identified genes (LOC108802036) did not have an associated annotation.
+One of the 7 identified genes (LOC108802036) does not have an associated annotation.
 For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 0.41 to 0.86.
 Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of parallel selection at this locus associated with exposure of frogs to Bd across multiple populations.
+This is suggestive of parallel selection at this locus across multiple populations.
 The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes (<!--# Is this correct, i.e., that we identified 33 outlier regions, and in the subset of 9 regions we found 33 annotated genes?
+A total of 33 annotated genes <!--# Is this correct, i.e., that we identified 33 outlier regions, and in the subset of 9 regions we found 33 annotated genes?
 I just want to make sure there isn't a typo in here.
@@
 
+No gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
+In addition, although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd-Methods** for details).
+
 In summary, these results indicate that the exomes of frogs from our naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
 The regions putatively under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
-
-<!--# Allie, I opted not to add the additional text and figures that you provided to this Results section (i.e., related to heterozygosity and pi).
-Instead, I mention the analyses in the "Materials and Methods - Frog evolution - Data analyis" section (along with a comment for you that provides a justification for my decision), and provide the analysis description and figures in "Supporting Information - Frog evolution".
-Let me know if that works for you.
--->
 
@@
 Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
-Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
+<!--# Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
 Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild [see @joseph2018 for a small-scale example].
-Given the demonstrated resistance of adult frogs against Bd infection, and the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether the resulting adaptive alleles are associated with increased frog resistance.
-The results of our study show that (1) reintroduction of resistant MYL frogs from naturally recovering populations allows the successful reestablishment of viable frog populations, and (2) that frogs in recovering populations have likely experienced selection following exposure to Bd and that some of the genomic regions identified as putatively under selection contain genes related to disease resistance.
+Given the demonstrated resistance of adult frogs against Bd infection, and the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether the resulting adaptive alleles are associated with increased frog resistance.--> Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that the recovering populations are likely to persist over a 50-year period, (iii) that there are substantial genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic regions putatively under selection contain genes related to disease resistance.
+Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
@@
 In the genome of another frog species, this gene is located adjacent to a type I interferon gene (Np-IFNi2) [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
+Collectively, the genes associated with these genomic differences may confer at least some degree of resistance against Bd infection, an attribute that may be critically important to population reestablishment and recovery in the presence of Bd.
+<!--# Added this transition sentence. -->
 
@@
 Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
-The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the 2-16 years covered by our post-translocation surveys.
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10<!--# Why not use 0.09?-->).
+The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the years covered by our post-translocation surveys.
 These long-term forecasts are important, given that reintroduced MYL frog populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
 Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is critically important to the process of adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
-Specifically, the combined results from our reintroduction study and viability modelling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
+Specifically, the combined results from our reintroduction study and viability modeling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
 In addition, given the repeatability of frog survival at a site, 1-year frog survival also serves as an effective proxy of site quality (i.e., the ability of a site to support high frog survival and a viable frog population over the long term).
@@
 
-Despite the demonstated resistance of adult MYL frogs against Bd infection, individual and population-level impacts of Bd are still evident.
+Despite the demonstrated resistance of adult MYL frogs against Bd infection, individual and population-level impacts of Bd are still evident.
 In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
@@
 Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
-As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have acquired via evolution the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
+As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
 Despite the hopeful example of successful reestablishment of MYL frogs despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
@@
 Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an example of evolutionary rescue will likely always be uncertain.
-Regardless, we provide a compelling example from the wild that suggests that evolution can produce individuals that harbor adaptive alleles and allow population recovery in a novel environment, and show conclusively that individuals from these recovering populations can be used in translocations to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
-We expect that such translocations will be an essential tool in wildlife conservation in an era of accelerating global change.
+Regardless, we provide an example from the wild that suggests that evolution can produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
+We expect that similar species recovery actions will be an essential tool in wildlife conservation in an era of accelerating global change.
 
@@
 
-For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
+For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
 The populations that we selected contained hundreds to thousands of both *R. sierrae* adults and tadpoles.
@@
 sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
-At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics.
+At the beginning of this study, we had a relatively limited understanding of the factors that affect habitat quality.
 In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwinter habitats such as submerged boulders and overhanging banks.
@@
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap SI) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
@@
 Following release, we visited translocated populations approximately once per month during the summer active season and conducted diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
-CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we search all habitats repeatedly for adult frogs.
+CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size, and VES provided estimates of tadpole and juvenile abundance.
+During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we searched all habitats repeatedly for adult frogs.
 Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
@@
 To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
-This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
+This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not occur until the following year.
 Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
-This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).
+This accounted for the multi-year tadpole and juvenile stages in MYL frogs (@tbl-param_values SI).
 
@@
 
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022] (see **Supporting Infomation - Frog population recovery - Mark-recapture model structure** for details).
-The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022] (see **Supporting Information - Frog population recovery - CMR model structure** for details).
+Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
 The possible states of individuals include "not recruited", "alive", and "dead".
@@
 We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
-We also assume that new recruits were a result of within-site reproduction and not immigration from adjacent populations because no *R.
-sierrae* populations were present within several kilometers of the translocation sites.
+We also assume that new recruits were the result of within-site reproduction and not immigration from adjacent populations.
+This assumption is justified by the fact that no *R. sierrae* populations were present within several kilometers of the translocation sites.
 For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
@@
 
-To identify important predictors of frog survival following translocation, we used multilevel Bayesian models.
-We included predictor variables to describe characteristics of frogs, sites, and translocations (predictor variables: Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year post-translocation survival estimates from CMR models as the response, which was available for all translocated populations regardless of how many years of post-translocation data were available.
-Survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
-Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
+To identify important predictors of frog survival following translocation, we used multilevel Bayesian models [@gelman2013; @gabry2019].
+Included predictor variables describe characteristics of sites, translocated cohorts, and individuals (Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
+We used 1-year post-translocation survival estimates from CMR models as the response.
+Estimated survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
+Group-level (random) effects included site_id, translocation_id, or translocation_id nested within site_id.
 We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022].
 For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out ("LOO") cross-validation [@gelman2013; @gabry2019; @vehtari2016], as implemented by the loo package [@vehtari2022].
+We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out cross-validation [@vehtari2016], as implemented by the loo package [@vehtari2022].
 (See **Supporting Information - Frog population recovery - Among-site survival modeling** for details.)
@@
 After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
-Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the first year after the translocation.
+Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the 1-year period following translocation.
 
@@
 The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
-First-year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced environmental stochasticity.
+First-year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced by environmental stochasticity.
 
@@
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
+Second, we compute the elasticity of $\lambda$ to 4 key model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp)<!--# Added this to ensure that we cite this figure somewhere in the main text -->.
 Third, we included demographic and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
@@
 Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [e.g., @rothstein2020], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
+To minimize potential confounding effects caused by variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
 DNA was extracted following Qiagen DNEasy manufacturer's protocols.
@@
 To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome under selection between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+To detect regions of the genome that differed between naive and recovering populations, i.e., regions putatively under selection, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
 For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
@@
 These standard deviations represent strict criteria to select only the top \~ 0.3% of windows.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (1) high differentiation between naive and recovering populations, and (2) differential patterns of nucleotide diversity in the same region.
+Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (i) high differentiation between naive and recovering populations, and (ii) differential patterns of nucleotide diversity in the same region.
 Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (**Supporting Information**: Datasets S3, S4).
 
-Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-
 Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-
-The joint processes of Bd-caused frog population declines and selection in response to Bd exposure could both affect genetic diversity of recovering populations.
-Comparisons between naive and recovering populations of (i) individual-level heterozygosity, and (ii) population-level $\pi$ are provided in **Supporting Information - Frog evolution in response to Bd**.
-<!--# Allie, instead of mentioning these results and referencing the associated figures in the main Results (as you suggested), I added them to the "SI - Frog evolution in response to Bd section".
-I did so for 2 reasons.
-First, the results are somewhat tangential to the main focus of this paper (frog recovery), and second, citing some of the figures where you suggested (in the Results above) would mean referring to results before those results were described (e.g., sliding window analysis and associated pi boxplots and barcharts).
-Doing so would have been confusing.
--->
-
-###
 
@@
 Richards-Zawacki).
+<!--# Do we need to acknowledge anybody at SD Zoo for providing R. muscosa genome? -->
 
@@
 
-![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated for each site from the mrmr model.
+![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated for each site from the mrmr CMR model.
 Error bars show the 95% uncertainty intervals.
@@
 
-![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability), as estimated from the model used in the among-site meta-analysis.
-(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
+![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability), as estimated from the model used in the among-site meta-analysis: (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
@@
 The inset map locates the study area and range boundary in California.
-For cross-referencing of numbered points to populations/sites mentioned in the text and figures, site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="600"}
+Sites 1 and 4 (site id = 72996 and 70567, respectively) were also used as sources of frogs in the frog population recovery study.](figures/allele_maps.png){#fig-allelefrequencies width="600"}
 
@@
 
-#### Mark-recapture model structure
-
-We estimated survival and recruitment for each site using open population mark-recapture models based on @joseph2018.
+#### CMR model structure
+
+We estimated survival and recruitment for each site using open population CMR models based on @joseph2018.
 For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
@@
 
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009, @joseph2018].
+We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009; @joseph2018].
 
@@
 
-Finally, while it would theoretically be possible to estimate survival covariate effects in a joint mark-recapture model that integrates capture histories across all sites, this was impractical due to computational requirements of the mark-recapture models (namely, run time and memory).
-In modeling survival estimates from site-specific mark-recapture analyses, we are in effect conducting an among-site meta-analysis to better understand how site, cohort, and individual level characteristics explain variation in survival after translocation.
+Finally, while it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (namely, run time and memory).
+In modeling survival estimates from site-specific CMR analyses, we are in effect conducting an among-site meta-analysis to better understand how site, cohort, and individual level characteristics explain variation in survival after translocation.
 
@@
 
-We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mark-recapture models.
+We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from CMR models.
 Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (@fig-translocation-survival).
@@
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in *Mark-recapture model structure*), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in *CMR model structure*), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
 .
@@
 The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
+Using this liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
 First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
 Next, we compared the 30 unique outlier genes from the liberal set of 38 outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the overlapping F~ST~ and splined windows.
+We then repeated this process for the set of 33 genes located in the *F~ST~* and $\pi_{diff}$ splined windows.
 
@@
 
-We found no significantly over-represented GO terms in the liberal set of outlier regions, or in the set of 33 genes located in the overlapping F~ST~ and splined windows.
+We found no significantly over-represented GO terms in the liberal set of outlier regions, or in the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
 Therefore.....<!--# Allie, check wording for accuracy, add one concluding sentence about what this result means. -->
@@
 | $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                                             |
-| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                                             |
+| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies                                                        |
+| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from CMR studies                                                        |
 | $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge                |

### 2023-12-19 doc/manuscript/translocation.qmd
Edits throughout Supporting Information section

@@
 
-To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog exomes (i.e., coding region of genome) between populations with contrasting histories of Bd exposure.
+To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of genome) between populations with contrasting histories of Bd exposure.
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
@@
 muscosa* versus *R.
-sierrae*) and the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
+sierrae*) and the strong signature of isolation-by-distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
 Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
@@
 No gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-In addition, although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd-Methods** for details).
-
-In summary, these results indicate that the exomes of frogs from our naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
+In addition, although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd** for details).
+
+In summary, our genomic results indicate that the exomes of frogs from naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
 The regions putatively under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
@@
 For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
-Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
-
+Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.
 This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012].
@@
 
-For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:
+For introduced individuals, which have deterministic recruitment (i.e., they recruit when introduced), the state transition matrix is given by:
 
@@
 
+The objective of this analysis is to describe the influence of site, cohort, and individual level characteristics on post-translocation frog survival.
+By modeling survival estimates obtained from site-specific mrmr CMR analyses, we are in effect conducting an among-site meta-analysis.
+Although it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (namely, run time and memory).
+
 We used Bayesian generalized linear mixed models to investigate predictors of survival among sites.
 The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation.
-We generated these point estimates by rounding the posterior median of 1-year post-introduction survival for each individual, and modeled the data using a Bernoulli distribution:
+We generated these point estimates by rounding the posterior median of 1-year post-introduction survival for each individual (from site-specific mrmr CMR models) and modeled the data using a Bernoulli distribution:
 
@@
 These models were fit using the `stan_glmer()` function in the rstanarm package, with default priors described below [@rstanarm2022].
-These priors are vague, but include data dependent scaling as follows to account for different input variable scales.
-However, as we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical.
+These priors are vague, but include data-dependent scaling as follows to account for different input variable scales.
+However, because we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical.
 Specifically, we have:
@@
 
-Finally, while it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (namely, run time and memory).
-In modeling survival estimates from site-specific CMR analyses, we are in effect conducting an among-site meta-analysis to better understand how site, cohort, and individual level characteristics explain variation in survival after translocation.
-
 #### Population viability modeling
@@
 
-We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from CMR models.
+We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mrmr CMR models.
 Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (@fig-translocation-survival).
@@
 Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$.
+We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information-Population viability modeling-Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp)
 
@@
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in *CMR model structure*), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information-Frog population recovery-CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
 .
@@
 
-#### Methods
-
-##### Study design
-
-We compared frog genomes sampled in naive versus recovering populations.
+#### Study design
+
+To gain insights into the role of evolution in the development of resistance by MYL frogs, we compared frog genomes sampled in naive versus recovering populations.
 Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
@@
 
-##### GO analysis
-
-In the Results section of the main paper, we describe the stringent set of outlier SNPs (identified using a Bonferroni-corrected p-value of 0.01).
-The liberal set, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-Using this liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
-First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-Next, we compared the 30 unique outlier genes from the liberal set of 38 outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-We then repeated this process for the set of 33 genes located in the *F~ST~* and $\pi_{diff}$ splined windows.
-
-##### Genetic diversity
+Our study design, in which we compared frog genomes in naive and recovering populations, required sampling populations across the range of MYL frogs in the southern Sierra Nevada.
+This is due to the fact that very few MYL frog populations remain in the naive state, and those that do are scattered across a wide latitudinal range, from Yosemite National Park in the north to southern Kings Canyon National Park in south.
+To minimize potential confounding effects caused by known variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges (@fig-allelefrequencies).
+
+#### GO analysis
+
+In **Results-Frog evolution in response to Bd**, we describe the stringent set of outlier variants (identified using a Bonferroni-corrected p-value of 0.01).
+A liberal set of outlier variants, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
+We used this liberal set to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
+To do this, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
+We then conducted a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] to compare the 30 unique outlier genes to the complete set of genes in our target transcriptome.
+We repeated this process for the set of 33 genes located in the *F~ST~* and $\pi_{diff}$ splined windows.
+
+#### Genetic diversity
 
@@
 
-#### Results
-
-##### GO analysis
-
-We found no significantly over-represented GO terms in the liberal set of outlier regions, or in the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-Therefore.....<!--# Allie, check wording for accuracy, add one concluding sentence about what this result means. -->
-
-##### Genetic diversity
-
 Average (i) individual-level heterozygosity and (ii) population-level nucleotide diversity ($\pi$) were similar between naive and recovering populations (@fig-violinplot-heterozy SI; @fig-boxplot-pi SI).
@@
 
-![Evidence that Bd load is not an important predictor of post-translocation frog survival.
-Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model in the between-site meta-analysis.
+![Results from the among-site meta-analysis, showing that Bd load is not an important predictor of post-translocation frog survival.
+Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model.
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
@@
 
-![Evidence for selection on individual variants in recovering MYL frog populations.
+![Results from the analysis of individual variants, showing putative signatures of selection in recovering MYL frog populations.
 (A) PCA calculated from binary SNPs showing the genomic relationship of samples.
@@
 
-![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog genomics study.
-Individual data points are shown on the plot with their corresponding site number (from @fig-allelefrequencies).
+![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog evolution study.
+Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).
 A Wilcoxon signed-rank test showed no significant difference between the two groups.](figures/het_by_status_violin_plot.png){#fig-violinplot-heterozy width="500"}
@@
 
-![For each study population in the frog evolution study (from @fig-allelefrequencies A), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
+![For each study population in the frog evolution study (from @fig-allelefrequencies), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
 The y-axis has been log~10~-transformed for display purposes.

### 2023-12-20 doc/manuscript/translocation.qmd
Adjust viability figure and edit viability text

--- 2023-12-19 doc/manuscript/translocation.qmd
@@
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity of $\lambda$ to 4 key model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp)<!--# Added this to ensure that we cite this figure somewhere in the main text -->.
-Third, we included demographic and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
+Second, we compute the elasticity of $\lambda$ to four key model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp SI).
+This also helped us determine where in the model environmental variation in juvenile survival and recruitment would have the largest effects on population dynamics.
+Third, we included demographic stochasticity and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
@@
 Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$.
+We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
 

### 2023-12-20 code/extinction_and_growth_rate_analysis.R

@@ -120,10 +120,10 @@ for(i in 1:length(muR_range)){
 dat = expand.grid(muR=muR_range, juv_surv=juv_surv_range)
 dat$lambda = as.vector(lambdas)
 ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) + 
-							geom_hline(aes(yintercept=0.08), linetype="dashed") +
+							geom_hline(aes(yintercept=0.09), linetype="dashed") +
 							geom_contour(aes(x=muR, y=juv_surv, z=lambda), breaks=c(1), color="red", linewidth=1) +
 							geom_point(data=lake_growth_rates_dt, aes(x=surv_med, 
-																												y=rep(0.08, length(surv_med)), 
+																												y=rep(0.09, length(surv_med)), 
 																													color=lake_id), size=2) +
 		   				scale_color_manual(values=colors) +
 							# scale_fill_gradient("\u03bb", low = "white", high = "black") +
@@ -448,7 +448,6 @@ ind = which.min(pred_ss)
 best_sigma_J1s = reshape2::melt(sigma_J1_trajectories[['70550']][pred_ss_dt$index[1:250], 1:16])
 ggplot(best_sigma_J1s) + geom_histogram(aes(x=value), bins=15) + facet_wrap(~Var2)
 colMeans(sigma_J1_trajectories[['70550']][pred_ss_dt$index[1:50], 1:16])
-
 
 # Plot and compare observed and predicted abundance
 pred_traj = data.table(reshape2::melt(pop_trajectories[['70550']][pred_ss_dt$index[1:100], 1:16]))

### 2023-12-20 doc/manuscript/translocation.qmd
Update PNAS converstion script

@@
 csl: pnas.csl
-format: pdf
-# format:
-#   latex: 
-#      cite-method: natbib
+# format: pdf
+format:
+   latex: 
+      cite-method: natbib
 include-in-header:
@@
 This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
-Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@råberg2009] and facilitate population persistence and recovery [@brannelly2021].
+Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@raberg2009] and facilitate population persistence and recovery [@brannelly2021].
 Host immunity and evolution both play important roles in the development of resistance and tolerance, and utilizing these factors would seem a promising approach to developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].

### 2023-12-20 doc/manuscript/translocation_bioarxiv.tex
Update PNAS converstion script

@@
 
-\title{Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
+\title{Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease}
 
@@
 % Please add a significance statement to explain the relevance of your work
-% \significancestatement{Understanding how species persist despite accelerating global change is
-% critical for the conservation of biodiversity. One possibility is rapid
-% evolution that restores positive growth to declining populations
-% (``evolutionary rescue''). Individuals from these rescued populations
-% could also be used to reestablish extinct populations. We evaluated both
-% possibilities in a once-common frog species driven to the brink of
-% extinction by a disease that has devastated global amphibian
-% biodiversity and for which few options exist to reverse disease impacts.
-% We found that frogs have evolved increased resistance to infection,
-% facilitating population recovery. In addition, introduction of frogs
-% from these populations allowed reestablishment despite the ongoing
-% presence of disease. Such reintroductions could play an important role
-% in biodiversity conservation in our rapidly changing world.}
+\significancestatement{Understanding how species persist despite accelerating global change is
+critical for the conservation of biodiversity. Emerging infectious
+diseases can have particularly devastating impacts, and few options
+typically exist to reverse these effects. We used large-scale
+reintroductions of disease-resistant individuals in an effort to recover
+a once-common frog species driven to the brink of extinction by a
+disease that has recently devastated global amphibian biodiversity.
+Introduction of resistant frogs allowed reestablishment of populations
+in the presence of disease. In addition, resistance appears to have a
+genetic basis and is consistent with evolution following exposure of
+frogs to Bd. The evolution of resistance and reintroductions of
+resistant individuals could play an important role in biodiversity
+conservation in our rapidly changing world.}
 
@@
 Vast alteration of the biosphere by humans is causing a sixth mass
-extinction. Adaptation to modified environments often is the only means
-for species to persist, but the extent to which such ``evolutionary
-rescue'' can prevent or reverse biodiversity loss is largely unknown.
-Using results from genomic analyses, a 15-year reintroduction effort,
-and population modeling, we provide a compelling example of evolution
-reversing the declines of an imperiled amphibian and allowing the
-reestablishment of extirpated populations. The once-common mountain
-yellow-legged (MYL) frog is threatened with extinction by the
-human-facilitated emergence of a lethal fungal pathogen
-(\emph{Batrachochytrium dendrobatidis}; ``Bd''). Although most MYL frog
-populations are extirpated following disease outbreaks, some persist and
-eventually begin to recover. Using an exome capture approach, we show
-that MYL frogs have undergone substantial evolutionary change following
-disease outbreaks. Moreover, in naturally recovering populations
-observed changes in immune function loci are associated with increased
-resistance/tolerance to Bd infection. Large-scale reintroduction of
-frogs from rescued populations resulted in the establishment of
-reproducing populations despite ongoing disease. In addition, results
-from viability modeling suggest that established populations have a low
-probability of extinction over 50 years. Collectively, these results
-provide a rare example of how evolutionary rescue and the reintroduction
-of resistant/tolerant individuals can allow the landscape-scale recovery
-of disease-impacted species. This example has potentially broad
-implications for the many taxa worldwide that are threatened with
-extinction by anthropogenic stressors.
+extinction, driven in part by an increase in emerging infectious
+diseases. The emergence of the lethal fungal pathogen
+(\emph{Batrachochytrium dendrobatidis}; ``Bd'') has devastated global
+amphibian biodiversity, with hundreds of species experiencing declines
+or extinctions. Given that there are no broadly applicable methods to
+reverse these impacts in the wild, the future of many amphibians appears
+grim. The once-common mountain yellow-legged (MYL) frog is emblematic of
+amphibians threatened by Bd. Although most MYL frog populations are
+extirpated following disease outbreaks, some persist and eventually
+recover. Frogs in these recovering populations have increased resistance
+against Bd infection, consistent with evolution of resistant genotypes
+and/or acquired immunity. We conducted a 15-year landscape-scale
+reintroduction study and show that frogs collected from recovering
+populations and reintroduced to vacant habitats can reestablish
+populations despite the presence of Bd. In addition, results from
+viability modeling suggest that many reintroduced populations have a low
+probability of extinction over 50 years. To better understand the role
+of evolution in frog resistance, we compared the genomes of MYL frogs
+from Bd-naive and recovering populations. We found substantial
+differences between these categories, including changes in immune
+function loci that may confer increased resistance, consistent with
+evolutionary changes in response to Bd exposure. These results provide a
+rare example of how reintroduction of resistant individuals can allow
+the landscape-scale recovery of disease-impacted species. This example
+has broad implications for the many taxa worldwide that are threatened
+with extinction by novel pathogens.
 \end{abstract}
@@
 
-\dropcap{H}uman activities are intensifying threats to global biodiversity
+\dropcap{H}uman activities are increasingly impacting global biodiversity
 \citep{ceballos2015}, with important implications for ecosystem
-resilience and human welfare \citep{naeem2009}. To persist, affected
-taxa will often need to adapt to their altered environment or face
-decline and possible extinction. The process whereby rapid evolutionary
-change increases the frequency of adaptive alleles and restores positive
-population growth is known as ``evolutionary rescue''
-\citep{carlson2014}. In addition to preventing extinction of declining
-populations, individuals from rescued populations could also disperse,
-or be reintroduced, to empty habitats and promote larger-scale recovery.
-Although this overall scenario is potentially of critical importance for
-the conservation of an ever-growing number of imperiled taxa,
-descriptions of its occurrence and effectiveness in the wild are
-lacking.
-
-One consequence of human alteration of the biosphere is an increase in
-emerging infectious diseases \citep{jones2008, fisher2012}. Such
-diseases pose a severe threat to wildlife populations
-\citep{daszak2000}, and have caused dramatic declines and extinctions in
-a wide range of taxa, including echinoderms, mammals, birds, and
-amphibians \citep{hewson2014, samuel2015, scheele2019, cunningham2021}.
-Amid this global loss of biodiversity, some populations decimated by
-disease have subsequently recovered to varying extents
-\citep{scheele2017, voyles2018, knapp2016}. Although we know relatively
-little about the mechanisms underlying recovery
-\citep{brannelly2021, russell2020}, host evolution is one possibility
-\citep{carlson2014, searle2020}. This process is known from a diversity
-of wildlife populations
-\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
-often results in increased host resistance and/or tolerance \citep[see][
-for definitions]{medzhitov2012}. However, linking these evolutionary
-responses to population recoveries is difficult and few clear examples
-exist. In addition to promoting recovery of disease-impacted
-populations, the resistant/tolerant individuals in such populations
-could potentially be used to reestablish populations that were
-extirpated following disease emergence
-\citep{mendelson2019, brannelly2021}. To date, the use of
-resistant/tolerant individuals for species recovery is rare \citep[but
-see][]{joseph2018}, and studies that evaluate its potential utility are
-critically needed.
+resilience and human welfare \citep{naeem2009}. One consequence of human
+alteration of the biosphere is an increase in emerging infectious
+diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
+threat to wildlife populations \citep{daszak2000}, and have caused
+dramatic declines and extinctions in a wide range of taxa, including
+echinoderms, mammals, birds, and amphibians
+\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
+are experiencing particularly devastating impacts of disease due to the
+recent emergence and global spread of the highly virulent amphibian
+chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
+\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
+have experienced Bd-caused declines, and numerous susceptible taxa are
+extinct in the wild \citep{scheele2019}. These impacts to global
+biodiversity may be unprecedented, and highlight the importance of
+understanding mechanisms of species persistence in the presence of
+emerging diseases \citep{russell2020}.
+
+Evidence of natural recovery in the many Bd-impacted amphibian
+populations is surprisingly rare \citep[for notable exceptions,
+see][]{scheele2017, voyles2018, knapp2016}, suggesting that
+disease-caused declines will be difficult to reverse. This apparent low
+resilience to disease effects may be due to the limited ability of many
+amphibians to develop Bd resistance and/or tolerance, which in turn,
+could also lessen the effectiveness of potential Bd mitigation
+strategies. Following pathogen arrival in a host population, resistance
+(ability to limit pathogen burden) and tolerance (ability to limit the
+harm caused by a particular burden) are key mechanisms to reduce disease
+impacts \citep{raberg2009} and facilitate population persistence and
+recovery \citep{brannelly2021}. Host immunity and evolution both play
+important roles in the development of resistance and tolerance, and
+utilizing these factors would seem a promising approach to developing
+effective strategies to mitigate disease impacts in the wild
+\citep{garner2016, woodhams2011}. However, several aspects of the
+amphibian-Bd system present difficult obstacles, including (i) the
+general inability of amphibians to mount an effective immune response
+against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
+(ii) the apparent rarity of evolution of more resistant/tolerant
+genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
+suggest that reintroduction of amphibians into sites to reestablish
+populations extirpated by Bd will often result not in population
+recovery, but instead in the rapid reinfection and mortality of the
+introduced animals and/or their progeny
+\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
+future of many amphibian species threatened by Bd appears bleak.
 
@@
 \emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
-emblematic of the unprecedented global declines of amphibians caused by
-the recently-emerged amphibian chytrid fungus
-\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
-most common amphibian in the high elevation portion of California's
-Sierra Nevada mountains \citep[USA,][]{grinnell1924}, during the past
-century this frog has disappeared from more than 90\% of its historical
-range \citep{vredenburg2007}. Due to the severity of its decline and the
+emblematic of the global declines of amphibians caused by Bd
+\citep{scheele2019}. Once the most common amphibian in the high
+elevation portion of California's Sierra Nevada mountains
+\citep[USA,][]{grinnell1924}, during the past century this frog has
+disappeared from more than 90\% of its historical range
+\citep{vredenburg2007}. Due to the severity of its decline and the
 increasing probability of extinction, both species are now listed as
@@
 \citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
-of \emph{B. dendrobatidis} (Bd) in the 1960s and its subsequent spread
-\citep{vredenburg2019} caused additional large-scale population
-extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
-declines are fundamentally different from the fish-caused declines
-because removal of fish populations is feasible and results in the rapid
-recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
-appears to persist in habitats even in the absence of amphibian hosts
-\citep{walker2007}, and therefore represents a long-term alteration of
-invaded ecosystems that amphibians will need to overcome to reestablish
-populations.
+of Bd in the mid-1900s and its subsequent spread \citep{vredenburg2019}
+caused additional large-scale population extirpations
+\citep{vredenburg2010, rachowicz2006}. These Bd-caused declines are
+fundamentally different from the fish-caused declines because fish
+eradication is feasible \citep{knapp1998} and results in the rapid
+recovery of frog populations \citep{knapp2007, vredenburg2004}. In
+contrast, Bd appears to persist in habitats even in the absence of
+amphibian hosts \citep{walker2007}, and therefore represents a long-term
+alteration of invaded ecosystems that amphibians will need to overcome
+to reestablish populations.
 
@@
 \citep{vredenburg2010}, some populations have persisted after epizootics
-\citep{briggs2010} and are now recovering \citep{knapp2016}. Frogs in
-these recovering populations show reduced susceptibility to Bd
-infection, consistent with selection for more resistant/tolerant
-genotypes \citep{knapp2016}. Furthermore, exploratory efforts to
-reintroduce frogs collected from recovering populations to sites from
-which frogs were previously extirpated indicate that population
-establishment is possible despite ongoing Bd infection
-\citep{joseph2018}.
-
-In the current study, we combine results from genomic analyses, a
-15-year landscape-scale frog reintroduction effort, and a detailed
-population model to provide an unprecedented example of how evolution in
-response to disease allows the successful reestablishment of extirpated
-MYL frog populations. Specifically, we used exome capture methods to
-show that MYL frogs have undergone substantial evolutionary change
-associated with Bd exposure, changes that might provide a genetic basis
-for the increased resistance/tolerance to Bd infection observed in
-naturally recovering populations \citep{knapp2016}. We then use the
-reintroduction of resistant/tolerant frogs to 24 sites and results from
-long-term capture-mark-recapture (CMR) surveys to describe the dynamics
-and drivers of population establishment. These data are also used to
-parameterize a population model that extends our inferences of
-population establishment well beyond the temporal extent of the CMR
-data. Collectively, these results provide one of the first examples of
-how evolutionary rescue and the reintroduction of resistant/tolerant
-animals can allow the large-scale recovery of disease-impacted species.
+\citep[during which Bd infection intensity on frogs is very
+high,][]{briggs2010} and are now recovering
+(Figure~\ref{fig-recovery-model}) \citep{knapp2016}. Frogs in these
+recovering populations show reduced susceptibility to Bd infection
+\citep{knapp2016}, with infection intensity (``load'') on adults
+consistently in the low-to-moderate range
+\citep{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
+is evident even under controlled laboratory conditions
+\citep{knapp2016}, indicative of host resistance against Bd infection
+(and not simply an effect of factors external to individual frogs, e.g.,
+environmental conditions). In addition to frogs from recovering
+populations having higher resistance to Bd infection than those from
+naive populations, they could also have higher tolerance, but no data
+are currently available to evaluate this possibility. Therefore, we
+focus on resistance throughout this paper. The observed resistance of
+MYL frogs could be the result of several non-mutually exclusive
+mechanisms, including natural selection for more resistant genotypes
+\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
+and/or inherent between-population differences that pre-date Bd
+exposure. The possible evolution of MYL frog resistance and subsequent
+population recovery is consistent with that expected under
+``evolutionary rescue'', whereby rapid evolutionary change increases the
+frequency of adaptive alleles and restores positive population growth
+\citep{carlson2014, searle2020}. This intriguing possibility also
+suggests an opportunity to expand recovery beyond the spatial scale
+possible under natural recovery by utilizing resistant frogs from
+recovering populations in reintroductions to vacant habitats
+(Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.
+
+In the current study, we had three primary objectives. First, to
+determine whether the reintroduction of resistant MYL frogs obtained
+from populations recovering from Bd-caused declines allows the
+successful reestablishment of extirpated populations despite ongoing
+disease, we conducted a 15-year landscape-scale frog reintroduction
+effort (Figure~\ref{fig-recovery-model}). Second, to extend our
+inferences of population recovery well beyond the temporal extent of our
+reintroduction study, we developed a model to estimate the probability
+of persistence for the reintroduced populations over a multi-decadal
+period (Figure~\ref{fig-recovery-model}). Third, we investigated whether
+MYL frogs show evidence of an evolutionary response to Bd exposure, and
+whether these genomic changes are associated with resistance. To
+accomplish this objective, we used exome capture methods to identify
+genomic regions with consistent differences between frogs from naive
+versus recovering populations (Figure~\ref{fig-recovery-model}) and
+determined whether any of these regions contain genes associated with
+resistance.
 
@@
 
-\subsection*{Frog evolution in response to Bd}
-
-To determine whether MYL frog populations show an evolutionary response
-to Bd, we compared frog genomes between populations with contrasting
-histories of Bd exposure. Specifically, we compared frog genomes sampled
-in 4 populations that have not yet experienced a Bd-caused epizootic
-(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
-Bd epizootic during the past several decades and have since recovered to
-varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
-\citep{knapp2016, vredenburg2010}. These recovering populations exist in
-an enzootic state \citep{briggs2010}, characterized by high Bd
-prevalence and, in adults, moderate Bd infection intensity (``load'')
-that is typically well below the level expected to cause mortality
-\citep{vredenburg2010}. (See SI Appendix for additional details
-regarding the study design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to
-describe the relationships between sampled populations, and then used
-two complementary approaches to identify regions of the genome under
-selection. First, we used a multivariate linear mixed model to evaluate
-associations between population type (i.e., naive versus recovering) and
-individual variants, including single nucleotide polymorphisms (SNPs)
-and insertions/deletions (INDELS). Second, we used a splined window
-analysis to identify larger genomic regions showing differences between
-population types in \emph{F\textsubscript{ST}} and nucleotide diversity
-(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).
-
-Individual frogs clustered into 3 separate groups in principal component
-space (Figure S1A), and clusters reflected the
-species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and the
-strong signature of isolation by distance that is characteristic of MYL
-frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
-contained at least one population from both the naive and recovering
-groups, allowing us to distinguish allelic associations of individuals
-sampled in the 2 population types versus allelic associations resulting
-from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show
-that recovering populations have signatures of strong selection on
-multiple regions of the genome. The analysis of individual variants
-identified 11 ``outlier'' SNPs (i.e., showing significantly different
-allele frequencies between naive versus recovering populations) from 7
-distinct genes across 4 contigs (Figure S1B, C;
-SI Appendix: Datasets S1, S2). One of the 7 identified genes
-(LOC108802036) did not have an associated annotation. For the outlier
-SNPs, frequency differences between the naive and recovering populations
-ranged from 41\% to 86\%. Most of these SNPs showed frequency
-differences in only a subset of the sampled populations
-(Figure~\ref{fig-allelefrequencies}A, B), but the SNP in the RIN3 gene
-showed consistent differences in frequencies across all populations
-(Figure~\ref{fig-allelefrequencies}C). This is suggestive of strong -
-and parallel - selection at this locus associated with exposure of frogs
-to Bd across multiple populations.
-
-The splined window analysis identified 33 outlier regions for
-\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
-(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
-outliers for both metrics (``shared regions'') and 2 of these shared
-regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions (SI
-Appendix: Datasets S3, S4). Given this large number of genes, here we
-focus on those with the strongest signal of selection and/or
-immune-related functions. The largest \(\pi_{diff}\), indicative of
-directional selection, occurred in a 163kb region on Contig19, 12.9Mb
-upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C).
-This region contains approximately 500 SNPs and one annotated gene
-called ``interferon-induced very large GTPase 1-like'' (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two
-complement factor genes (C6 and C7). Interestingly, this region had a
-large negative \(\pi_{diff}\) value, consistent with balancing
-selection. Finally, one shared outlier region on Contig8 contained one
-outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS)
-(Figure~\ref{fig-spline-manhattan}D, Figure S2).
-This region (854kb from the beginning of the outlier window to the VARS
-SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
-these genes occur in the extended major histocompatibility complex (MHC)
-Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
-extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
-Therefore, this region under selection is part of the extended MHC Class
-I and III complex and shows synteny with other amphibian genomes.
-
 \subsection*{Frog population recovery}
 
-To determine whether frogs from recovering populations could be used to
-reestablish extirpated populations, we conducted 24 reintroductions in
-Yosemite National Park (2006--2020). Each of the reintroductions
+To determine whether MYL frogs from recovering populations can be used
+to reestablish extirpated populations, we conducted 24 reintroductions
+in Yosemite National Park (2006--2020). Each of the reintroductions
 involved collection of adult frogs from 1 of 3 recovering, Bd-positive
-``donor'' populations and translocating them \citep{seddon2014} to 1 of
-12 nearby recipient sites (Figure S3). The donor
-populations included 2 of the 5 recovering populations in the frog
-evolution study described above (Figure~\ref{fig-allelefrequencies}A:
-population 1 and 4). Following translocation, adult survival and
-recruitment of new adults was estimated from CMR surveys and counts of
-tadpoles and juveniles were obtained from visual encounter surveys
-(VES).
-
-Estimates of frog survival during the 1-year period following
-translocation indicate that survival was highly variable between
-recipient sites, but relatively constant within recipient sites (for the
-subset of sites that received multiple translocations;
-Figure~\ref{fig-translocation-survival}). As such, the average survival
-of frogs in the first translocation to a site was an important predictor
-of survival of frogs in subsequent translocations to that site
-(Figure S4). This suggests an important
-effect of site characteristics and/or (site * frog) interactions on frog
-survival. In addition, 1-year survival was higher for frogs translocated
-later in the study period than earlier
-(Table S1), which we suggest resulted
-primarily from our improved ability to choose recipient sites with
-higher habitat quality for \emph{R. sierrae}. Finally, Bd loads were
-fairly consistent before versus after translocation and loads were
-nearly always well below the level indicative of severe chytridiomycosis
-and associated frog mortality (Figure S5)
-\citep{joseph2018, vredenburg2010}.
-
-Of the 12 translocated populations, 9 (0.75) showed evidence of
+``donor'' populations and translocating them to 1 of 12 nearby recipient
+sites (Figure S1). The donor populations included 2 of
+the 5 recovering populations included in the frog evolution study
+referenced above (Figure~\ref{fig-allelefrequencies}: population 1 and
+4), and these donor populations contributed frogs for 20 of the 24
+translocations. Following translocation, we estimated adult survival and
+recruitment of new adults from capture-mark-recapture (CMR) surveys and
+obtained counts of tadpoles and juveniles from visual encounter surveys
+(VES). Across all translocation sites, the duration of survey time
+series was 1--16 years (median = 5).
+
+Of the 12 reintroduced populations, 9 (0.75) showed evidence of
 successful reproduction in subsequent years, as indicated by the
@@
 
-To determine whether Bd load had a negative effect on adult survival, as
-would be expected if frogs were highly susceptible to Bd infection, we
-developed linear and multilevel models. The best model of 1-year frog
-survival identified several important predictors, but Bd load at the
-time of translocation was not among them
-(Figure S6). Important predictors included
-winter severity in the year following translocation, site elevation, and
-donor population (Figure~\ref{fig-cond-effects}). Males had somewhat
-higher survival than females, but this effect was only marginally
-important (Figure~\ref{fig-cond-effects},
-Figure S6). Collectively, these results are
-consistent with frogs in recovering populations having evolved
-sufficient resistance/tolerance to minimize the adverse effects of Bd
-infection. In addition, the moderate Bd loads that characterized frogs
-following translocation (Figure S5) and
-relatively high survival of frogs in numerous translocated cohorts
-(Figure~\ref{fig-translocation-survival}) suggests that frogs
+Bd loads were fairly consistent before versus after translocation, and
+loads were nearly always well below the level indicative of severe
+chytridiomycosis (i.e., the disease caused by Bd) and associated frog
+mortality (Figure S2)
+\citep{joseph2018, vredenburg2010}. Although it is possible that the
+observed relatively small changes in load are a consequence of
+individuals with high Bd loads dying and therefore being unavailable for
+sampling during the post-translocation period, the fact that there was
+little difference in pre- versus post-translocation Bd loads even in
+those populations that had very high frog survival (70556, 74976;
+Figure~\ref{fig-translocation-survival},
+Figure S2) suggests a true lack of
+substantial change in Bd load.
+
+The ultimate measure of reintroduction success is the establishment of a
+self-sustaining population. Given that it can take years or even decades
+to determine the self-sustainability of a reintroduced population
+\citep[for an example in MYL frogs, see][]{joseph2018}, the use of
+proxies is essential for providing shorter-term insights into
+reintroduction success and the factors driving it. Results from our CMR
+surveys allowed us to accurately estimate frog survival, including over
+the entire CMR time series for each site and during only the 1-year
+period immediately following translocation. These estimates were made
+using site-specific models analyzed using the mrmr package. We use these
+estimates to describe general patterns of frog survival in all
+translocated cohorts, and in an among-site meta-analysis of frog
+survival to identify important predictors of 1-year frog survival (e.g.,
+Bd load).
+
+Estimates of 1-year frog survival indicate that survival was highly
+variable between recipient sites, but relatively constant within
+recipient sites (for the subset of sites that received multiple
+translocations; Figure~\ref{fig-translocation-survival}). These patterns
+indicate an important effect of site characteristics on frog survival.
+In addition, 1-year survival was higher for frogs translocated later in
+the study period than earlier: 5 of the 7 populations translocated after
+2013 had estimated survival \(\ge\) 0.5, compared to only 1 of 5
+populations translocated prior to 2013. We suggest this resulted
+primarily from our improved ability to choose recipient sites with
+higher habitat quality for \emph{R. sierrae} (see \textbf{Materials and
+Methods - Frog population recovery - Field methods} for details). This
+increased survival has direct implications for population viability (see
+\textbf{Results - Long-term population viability}).
+
+The goal of our meta-analysis was to identify important predictors of
+1-year frog survival. We were particularly interested in whether Bd load
+had a negative effect on adult survival, as would be expected if frogs
+were highly susceptible to Bd infection. This analysis was conducted in
+a Bayesian framework and included a diversity of site, cohort, and
+individual level characteristics as predictors and 1-year frog survival
+(Figure~\ref{fig-translocation-survival}) as the response variable. The
+best model of 1-year frog survival identified several important
+predictors, but Bd load at the time of translocation was not among them
+(Figure S3). Instead, important predictors
+included winter severity in the year following translocation (snow\_t1),
+site elevation, and donor population (Figure~\ref{fig-cond-effects},
+Figure S3). Males had somewhat higher
+survival than females, but this effect was small
+(Figure~\ref{fig-cond-effects}, Figure S3).
+The absence of Bd load as an important predictor of frog survival is
+consistent with frogs in recovering populations having sufficient
+resistance to suppress Bd loads below harmful levels.
+
+In summary, results from our frog translocation study indicate that
+translocations resulted in (i) relatively high 1-year survival of
+translocated adults, as well as reproduction and recruitment, at the
+majority of recipient sites, (ii) 1-year survival of adults is
+influenced by site characteristics, weather conditions, and donor
+population (but not Bd load), and (iii) based on the relatively small
+changes in Bd load after translocation, loads appear more strongly
+influenced by frog characteristics (e.g., resistance) than site
+characteristics. Together, these results indicate that frogs
 translocated from recovering populations can maintain the benefits of
-resistance/tolerance in non-natal habitats. Finally, in 3 locations
-where longer CMR time series allowed us to assess the survival of new
-adults recruited to the population, naturally recruited adults had
-equivalent or higher survival probabilities than the originally
-translocated adults (Figure S7). Therefore,
-frog resistance/tolerance in non-natal habitats is maintained across
-generations.
+resistance in non-natal habitats. In addition, in 3 locations where
+longer CMR time series allowed us to assess the survival of new adults
+recruited to the population, naturally-recruited adults had equivalent
+or higher survival probabilities than the originally translocated adults
+(Figure S4). This suggests that frog
+resistance is maintained across generations. All of the conditions
+described above are supportive of population establishment and long-term
+population growth.
 
@@
 
-Results from frog translocations indicated that adult survival was often
-relatively high and that most populations showed evidence of successful
-reproduction and recruitment (described above). Although suggestive of
-population establishment, a decade or more of surveys may be necessary
-to confirm that populations are in fact self-sustaining
-\citep{joseph2018}. To extend our inferences of population establishment
-beyond those possible from the 2-16 years of site-specific CMR data, we
-developed a population viability model. Specifically, to test whether
-the observed yearly adult survival probabilities in translocated
-populations were sufficient to promote long-term viability, we built a
-stage-structured matrix model that captured known frog demography and
-included demographic and environmental stochasticity. We parameterized
-the model using CMR data from translocated populations and known life
-history values in this system (Table S2).
-
-Given observed yearly adult survival probabilities (from analysis of CMR
-data) and a baseline yearly recruitment probability \(\omega\) of 0.3
-(i.e., the probability that, following survival, juveniles successfully
-recruit to the adult class), our stage-structured model predicted that
-six of twelve translocated populations should experience a long-run
-growth rate \(\lambda\) greater than 1 in the presence of Bd
+Results from the frog translocation study indicated that most
+populations showed evidence of successful reproduction and recruitment,
+and that adult survival was often relatively high (described above).
+Although suggestive of population establishment, a decade or more of
+surveys may be necessary to confirm that populations are in fact
+self-sustaining \citep{joseph2018}. To extend our inferences of
+population establishment beyond those possible from the site-specific
+CMR data, we developed a population viability model. Specifically, to
+test whether the observed yearly adult survival probabilities in
+translocated populations (provided in legend of
+Figure~\ref{fig-viability}B) were sufficient for long-term viability,
+we built a stage-structured matrix model that captured known frog
+demography and included demographic and environmental stochasticity. We
+parameterized the model using CMR data from translocated populations and
+known life history values in this system (Table S2
+SI).
+
+Given observed yearly adult survival probabilities of translocated frogs
+(from site-specific mrmr CMR models) and a yearly survival probability
+of the year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at
+least six of twelve translocated populations should experience a
+long-run growth rate \(\lambda\) greater than 1 in the presence of Bd
 (Figure~\ref{fig-viability}A; median predicted \(\lambda\) ranges from
-1.15-1.36 for these six populations). These six populations all had
-observed yearly adult survival greater than 0.5. When we reduced the
-probability of successful recruitment \(\omega\) to 0.1 (a highly
-conservative estimate), long-run growth rate of five of these six
-populations was still greater than 1 (Figure~\ref{fig-viability}A).
+1.19-1.40 for these six populations). These six populations all had
+observed yearly adult survival greater than 0.5. As year-1 juvenile
+survival probability increased above 0.2, the deterministic long-run
+growth rate of eight of twelve population was greater than 1
+(Figure~\ref{fig-viability}A).
 
 Even when incorporating (i) demographic stochasticity and (ii)
-environmental stochasticity in the probability of juvenile recruitment
-to the adult class \(\omega\) (the transition that we expect to be the
-most subject to environmental variability in the presence of Bd),
-populations with high adult survival are likely to persist over a 50
-year time horizon. Our model predicted that, following a single
-introduction of 40 adult individuals into a population and using the
-baseline value for \(\omega\) of 0.3, the six populations with the
-highest adult survival probabilities (\(\sigma_{A_R} > 0.5\)) had
-50-year extinction probabilities of 0 to 0.32
-(Figure~\ref{fig-viability}B). This indicates strong potential for
-long-term persistence in the presence of Bd and environmental
-variability in recruitment. When we decreased \(\omega\) from 0.3 to a
-conservative estimate of 0.15, 50-year extinction probability remained
-less than 0.5 for three populations and was 0.79 to 0.94 for three
-additional populations (Figure~\ref{fig-viability}B). In contrast, for
+environmental stochasticity in year-1 juvenile survival and recruitment
+(the transition that we expect to be the most subject to environmental
+variability in the presence of Bd), populations with high adult survival
+are likely to persist over a 50 year time horizon. Our model predicted
+that, following a single introduction of 40 adult individuals into a
+population, the six populations with the highest adult survival
+probabilities (\(\sigma_{A_R} > 0.5\)) had 50-year extinction
+probabilities of less than 0.5 when the average year 1 juvenile survival
+was greater than 0.10 (Figure~\ref{fig-viability}B). This indicates
+strong potential for long-term persistence in the presence of Bd and
+environmental variability in survival and recruitment. In contrast, for
 the six populations where yearly adult survival probability
 \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was always
-predicted to be 1. To test the validity of our model predictions, we
-demonstrated that our stochastic model could describe the general
-recovery trajectory of our translocated population with the longest
-survey history (Figure~\ref{fig-viability}C; population 70550, surveyed
-for 16 years).
+predicted to be \textgreater{} 50\% regardless of the value of mean year
+1 juvenile survival between 0 and 0.25. To test the validity of our
+model predictions, we demonstrated that our stochastic model could
+describe the general recovery trajectory of our translocated population
+with the longest survey history (Figure~\ref{fig-viability}C;
+population 70550, surveyed for 16 years).
 
@@
 survival probability than translocated adults
-(Figure S7), but we considered these
+(Figure S4), but we considered these
 probabilities to be equal in all but three of our populations where we
-had sufficient data to distinguish these different probabilities. In
-addition, 5 of the 7 populations translocated after 2013 were predicted
-to have long-term viability, compared to only 1 of 5 populations
-translocated prior to 2013 (Table S1). Thus,
-there is some trend suggesting that our ability to select high quality
-habitats that promote long-term viability of translocated populations
-improved over time.
+had sufficient data to distinguish these different probabilities.
+
+\subsection*{Frog evolution in response to Bd}
+
+Results from the preceding sections indicate the critical role of frog
+resistance in post-translocation frog survival, population growth, and
+population viability. As such, identifying the mechanisms underlying
+this resistance would fill a key gap in our understanding of the factors
+that promote population resilience in the presence of disease. Although
+natural selection for more resistant frog genotypes, and evolutionary
+rescue, may be foundational to the ability of frogs to recover despite
+ongoing Bd infection, for MYL frogs the role of disease-mediated
+selection in these processes remains unknown.
+
+To determine whether MYL frog populations show an evolutionary response
+to Bd, we compared frog exomes (i.e., coding region of genome) between
+populations with contrasting histories of Bd exposure. Specifically, we
+compared frog genomes sampled in 4 populations that have not yet
+experienced a Bd-caused epizootic (``naive'') \citep{zhou2015} versus in
+5 populations that experienced a Bd epizootic during the past several
+decades and have since recovered to varying degrees (``recovering'';
+Figure~\ref{fig-allelefrequencies}) \citep{knapp2016, vredenburg2010}.
+Bd-exposure histories of the 9 study populations are based on 10-20
+years of VES and Bd surveillance using skin swabbing
+\citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive populations are
+characterized by large numbers of adults (i.e, typically 1000s), Bd
+prevalence that is generally 0\% except during occasional Bd failed
+invasions \citep[during which Bd loads remain very low,][]{wilber2022},
+and no history of Bd epizootics since we first surveyed these
+populations in the late 1990s and early 2000s \citep{zhou2015}. In
+contrast, recovering populations exist in an enzootic state
+\citep{briggs2010}, characterized by smaller numbers of adults
+(generally \textless{} 100), high Bd prevalence \citep[often
+\textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
+that are typically well below the level expected to cause mortality
+\citep{vredenburg2010}. Naive and recovering populations can be
+identified unambiguously using these differences in Bd prevalence and
+load. Finally, there is no potential for frog dispersal between the 9
+study populations due to intervening distances and topography, as well
+as the presence of introduced (predatory) fish and fish-induced habitat
+fragmentation. (See \textbf{Supporting Information - Frog evolution in
+response to Bd - Methods} for additional details regarding the study
+design.)
+
+We conducted a principal component analysis (PCA) of the genomic data to
+describe the relationships between sampled populations, and then used
+two complementary approaches to identify regions of the genome that
+differed between naive and recovering populations (i.e., regions
+putatively under selection). First, we used a multivariate linear mixed
+model to evaluate associations between population type (i.e., naive
+versus recovering) and individual variants, including single nucleotide
+polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting
+for population structure. Second, we used a splined window analysis to
+identify larger genomic regions showing differences between population
+types in \emph{F\textsubscript{ST}} and nucleotide diversity
+(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).
+
+Individual frogs clustered into 3 separate groups in principal component
+space (Figure S6A SI), and clusters reflected
+the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
+the strong signature of isolation by distance that is characteristic of
+MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
+each cluster contained at least one population from both the naive and
+recovering groups, allowing us to distinguish allelic associations of
+individuals sampled in the 2 population types versus allelic
+associations resulting from population structure and genetic drift.
+
+Results from the individual variant and splined window analyses show
+that recovering populations have signatures of selection on multiple
+regions of the genome. The analysis of individual variants identified 11
+``outlier'' SNPs (i.e., showing significantly different allele
+frequencies between naive versus recovering populations) from 7 distinct
+genes across 4 contigs (Figure S6B, C SI; SI
+Datasets S1, S2). One of the 7 identified genes (LOC108802036) does not
+have an associated annotation. For the outlier SNPs, frequency
+differences between the naive and recovering populations ranged from
+0.41 to 0.86. Most of these SNPs showed frequency differences in only a
+subset of the sampled populations (Figure~\ref{fig-allelefrequencies}A,
+B), but the SNP in the RIN3 gene showed consistent differences in
+frequencies across all populations (Figure~\ref{fig-allelefrequencies}
+C). This is suggestive of parallel selection at this locus across
+multiple populations. The other 6 outlier variants showed less
+consistent frequency differences across the study populations, but for
+these we still found a statistically significant signal of selection in
+2 of the 3 genetic clusters (containing populations 5--9;
+Figure~\ref{fig-allelefrequencies}, Figure S6A
+SI). Therefore, although some outlier variant associations have a more
+limited geographic extent than RIN3, they still describe results that
+suggest parallel evolutionary changes following Bd exposure.
+
+The splined window analysis identified 33 outlier regions for
+\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
+(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
+outliers for both metrics (``shared regions'') and 2 of these shared
+regions also contained one or more of the outlier SNPs described above.
+A total of 33 annotated genes were found in the 9 shared regions
+(Datasets S3, S4). Given this large number of genes, here we focus on
+those with the strongest signal of selection and/or immune-related
+functions. The largest \(\pi_{diff}\), indicative of directional
+selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
+the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C). This region
+contains approximately 500 SNPs and one annotated gene called
+``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
+a shared outlier region on Contig1 contained two complement factor genes
+(C6 and C7). Interestingly, this region had a large negative
+\(\pi_{diff}\) value, consistent with balancing selection. Finally, one
+shared outlier region on Contig8 contained one outlier SNP (TCF19) and
+was within 360kb of another outlier SNP (VARS)
+(Figure~\ref{fig-spline-manhattan}D, Figure S7).
+This region (854kb from the beginning of the outlier window to the VARS
+SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
+these genes occur in the extended major histocompatibility complex (MHC)
+Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
+extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
+Therefore, this region under selection is part of the extended MHC Class
+I and III complex and shows synteny with other amphibian genomes.
+
+No gene ontology (GO) biological functions, molecular functions, or
+cellular processes were over-represented in either the outlier variants
+or the 33 genes located in the overlapping \emph{F\textsubscript{ST}}
+and \(\pi_{diff}\) splined windows. In addition, although the joint
+processes of Bd-caused population declines and selection in response to
+Bd exposure could affect genetic diversity of recovering populations, we
+found no consistent differences in individual-level heterozygosity or
+population-level \(\pi\) between naive and recovering populations (see
+\textbf{Supporting Information-Frog evolution in response to Bd-Methods}
+for details).
+
+In summary, these results indicate that the exomes of frogs from our
+naive and recovering study populations show substantial differences,
+consistent with parallel evolutionary changes following Bd exposure. The
+regions putatively under selection contain several
+immunologically-relevant genes and gene families that are directly
+linked to disease resistance in other taxa.
 
@@
 Disease-induced population declines are decimating global biodiversity
-\citep{daszak2000}, but strategies to recover affected species are
-generally lacking. Here, we tested the possibility that populations of
-resistant/tolerant individuals could be used to reestablish extirpated
-populations. Although previously proposed
-\citep{mendelson2019, scheele2021}, this strategy has rarely been used
-in the wild \citep[see][ for a small-scale example]{joseph2018}. In
-particular, although numerous studies have documented signatures of
-selection in hosts following disease-induced declines
-\citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
-any that used individuals bearing the resulting adaptive alleles to
-reestablish extirpated populations. The results of our study provide
-evidence that (1) MYL frogs have experienced strong selection following
-exposure to a recently-emerged lethal fungal disease and that some of
-the genomic regions under selection contain genes related to disease
-resistance/tolerance, and (2) reintroduction of individuals from these
-resistant/tolerant populations allows the successful reestablishment of
-viable frog populations despite ongoing disease.
+\citep{daszak2000}, but broadly-applicable strategies to recover
+affected species are generally lacking \citep[e.g.,][]{garner2016}. Our
+results indicate (i) the capacity of reintroduced populations to become
+established and eventually recover despite ongoing disease, (ii) that
+the recovering populations are likely to persist over a 50-year period,
+(iii) that there are substantial genomic differences between naive and
+recovering MYL frog populations, consistent with evolutionary change in
+frogs following Bd exposure, and (iv) that some of the genomic regions
+putatively under selection contain genes related to disease resistance.
+Collectively, these results (Figure~\ref{fig-recovery-model}) provide a
+rare example of amphibian recovery in the presence of a highly virulent
+pathogen, and have important implications for the conservation and
+recovery of amphibians and other taxa worldwide that are endangered by
+escalating impacts from emerging infectious diseases. In light of the
+generally low success rate of amphibian reintroduction efforts
+\citep{dodd2005}, our success in reestablishing MYL frog populations via
+translocation of resistant individuals is striking, and even more so
+given that this success occurred despite ongoing infection of frogs with
+Bd.
 
@@
 relatively constant Bd loads and mass die-offs are not observed
-\citep[see also
-Figure S5]{briggs2010, knapp2011}. The
-differences in Bd load of frogs from naive and recovering populations
-are also observed in controlled laboratory studies \citep[see Figure 4
-in][]{knapp2016}. The observed reduced susceptibility of frogs from
-recovering populations could in theory be due to several factors,
-including natural selection for more resistant/tolerant genotypes and/or
-acquired immunity, but until now evidence to evaluate the role of
+\citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
+naive and recovering populations are also observed in controlled
+laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
+indicate that frogs from recovering populations exhibit resistance
+against Bd infection. This resistance could in theory be due to several
+factors, including natural selection for more resistant genotypes,
+acquired immunity, and/or inherent between-population differences that
+pre-date Bd exposure, but until now evidence to evaluate the role of
 evolution was lacking.
 
-Results from our genomic analyses strongly suggest that natural
-selection for adaptive alleles and resulting increased
-resistance/tolerance is at least partially responsible for the reduced
-susceptibility of frogs in recovering populations. We identified
-specific alleles and genomic regions showing signatures of selection
-between adjacent Bd-naive and recovering MYL frog populations,
-consistent with strong selection following Bd exposure. Importantly,
-some regions under selection are associated with cellular and
-immunological mechanisms known to contribute to disease resistance,
-including in amphibians \citep{zamudio2020}. For example, the MHC plays
-an important role in immunity. In our study, we identified a region that
-shows evidence of selection in recovering populations and contains eight
-genes associated with either the MHC Class I or Class III regions. These
-results corroborate numerous previous studies linking MHC genes to
-amphibian resistance/tolerance against Bd
-\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
-the strongest indication of directional selection (as measured by
-\(\pi_{diff}\)) contains the interferon-related gene GVINP1. Several
-previous studies have found this gene to be differentially expressed
-during Bd infection \citep[e.g.,][]{grogan2018, ellison2014} and in
-populations differing in Bd susceptibility \citep{grogan2018}. This gene
-is also strongly linked to disease in salmon, explaining a notable 20\%
-of the resistance phenotype \citep{robledo2020, robledo2018}. We also
-identified a region, characterized by high \emph{F\textsubscript{ST}}
-and low \(\pi_{diff}\), that contained the complement genes C6 and C7.
-The complement system plays an important role in innate immunity
-\citep{rieraromo2016}, and our results could indicate that balancing
-selection is acting in this region of the genome to favor a diverse set
-of alleles, as is known for C6 in humans \citep{soejima2005}. Finally,
-based on the analysis of individual outlier variants, the RIN3 gene
-showed a consistent pattern of allele frequency differences across all
-nine of the frog populations sampled in this study, indicating
-consistent selection in populations distributed across a wide geographic
-area. This gene is associated with inflammation response and in
-\emph{Xenopus} is highly expressed in wound stumps following amputation
-of tadpole tails \citep{fukazawa2009}. Also in this analysis, the
-outlier variant with the lowest p-value (uncharacterized gene
-LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2)
-in another frog genome \citep{gan2018}, and together with GVINP1 further
-suggests the importance of interferon-related genes in this system.
-
-Reintroduction of resistant/tolerant MYL frogs was remarkably successful
-in reestablishing viable populations in the presence of Bd. Of the 12
+Results from our genomic analyses suggest that natural selection for
+adaptive alleles is at least partially responsible for the increased
+resistance of frogs in recovering populations. We identified specific
+alleles and genomic regions showing signatures of selection between
+adjacent naive and recovering MYL frog populations, consistent with
+selection following Bd exposure. Importantly, some regions under
+selection are associated with cellular and immunological mechanisms
+known to contribute to disease resistance, including in amphibians
+\citep{zamudio2020}. For example, the MHC plays an important role in
+immunity. In our study, we identified a region that shows evidence of
+selection in recovering populations and contains eight genes associated
+with either the MHC Class I or Class III regions. These results
+corroborate numerous previous studies linking MHC genes to amphibian
+resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
+Similarly, the region with the strongest indication of directional
+selection (as measured by \(\pi_{diff}\)) contains the
+interferon-related gene GVINP1. Several previous studies of amphibians
+have found this gene to be differentially expressed during Bd infection
+\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
+in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
+linked to disease in salmon, explaining a notable 20\% of the resistance
+phenotype \citep{robledo2020, robledo2018}. We also identified a region,
+characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
+that contained the complement genes C6 and C7. The complement system
+plays an important role in innate immunity \citep{rieraromo2016}, and
+our results could indicate that balancing selection is acting in this
+region of the genome to favor a diverse set of alleles, as is known for
+C6 in humans \citep{soejima2005}. Finally, based on the analysis of
+individual outlier variants, the RIN3 gene showed a consistent pattern
+of allele frequency differences across all nine of the frog populations
+sampled in this study, indicating consistent selection in populations
+distributed across a wide geographic area. This gene is associated with
+inflammation response and in \emph{Xenopus} is highly expressed in
+wounds \citep{fukazawa2009}. In addition, the outlier variant with the
+lowest p-value was the uncharacterized gene LOC108802036. In the genome
+of another frog species, this gene is located adjacent to a type I
+interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
+further suggests the importance of interferon-related genes in this
+system. Collectively, the genes associated with these genomic
+differences may confer at least some degree of resistance against Bd
+infection, an attribute that may be critically important to population
+reestablishment and recovery in the presence of Bd.
+
+Reintroduction of resistant MYL frogs was remarkably successful in
+reestablishing viable populations in the presence of Bd. Of the 12
 translocated populations, approximately 80\% showed evidence of both
-successful reproduction and recruitment of new adults. Bd loads measured
-before and after translocation were similar and generally remained well
-below the level at which frog mortality occurs. Annual survival for 12
-of the 24 translocated cohorts exceeded 50\%, and \textgreater{} 70\% of
-translocated cohorts had survival above this 50\% level when the
+successful reproduction and recruitment of new adults. Year-1 survival
+for 12 of the 24 translocated cohorts exceeded 50\%, and \textgreater{}
+70\% of translocated cohorts had survival above this 50\% level when the
 earliest translocations are excluded (i.e., translocations conducted
-when methods were still being refined). The fact that the relatively low
-Bd loads and correspondingly high frog survival in the presence of Bd
-was maintained between the natal/donor sites and the recipient sites
-indicates that these characteristics of naturally-recovering populations
-were not solely an effect of the natal/donor site or site * frog
-interactions, but instead were due at least in part to
-resistance/tolerance inherent in the frogs. In addition, the relatively
-high survival of translocated frogs was maintained in their progeny, as
-expected if resistance/tolerance has a genetic basis. Results from the
-population viability model were also encouraging. In particular,
-translocated populations with \textgreater{} 50\% survival in the first
-year post-translocation were predicted to have a low probability of
-extinction over 50 years (probability of extinction = 0 - 0.32). In
-light of the generally low success rate of amphibian reintroduction
-efforts \citep{dodd2005}, our success in reestablishing MYL frog
-populations via translocation of resistant/tolerant individuals is
-striking, and even more so given that this success occurred despite
-ongoing infection of frogs with Bd.
-
-Despite the resistance/tolerance of adult MYL frogs in recovering
-populations, individual and population-level impacts of Bd are still
+when methods were still being refined; see \textbf{Materials and Methods
+- Frog population recovery - Field methods} for a brief description of
+these refinements). The fact that the relatively low Bd loads and
+correspondingly high frog survival was maintained when frogs were moved
+from donor populations to recipient sites indicates that these
+characteristics of naturally-recovering populations were not solely an
+effect of site characteristics, but were also strongly influenced by
+resistance inherent in the frogs. Although it could be argued that the
+relatively invariant Bd loads before versus after translocation are a
+consequence of similar pathogen pressure in the donor and translocated
+populations, this is at odds with the fact that in the first year after
+translocation frog densities are typically 1-2 orders of magnitude lower
+in the translocated versus donor populations and pathogen pressure
+should follow a similar pattern. In addition to the maintenance of Bd
+load and frog survival between natal and translocation sites, the
+relatively high survival of translocated frogs was maintained in their
+progeny, as expected if resistance has a genetic basis.
+
+Results from the population viability model were also encouraging. In
+particular, translocated populations with \textgreater{} 50\% survival
+in the first year post-translocation were predicted to have a low
+probability of extinction over 50 years (probability of extinction \(<\)
+0.5 when year 1 juvenile survival probability was greater than 0.10).
+The viability model highlighted the important role of frog survival in
+affecting long-term population viability, and allowed us to extend the
+temporal scale of our study beyond the years covered by our
+post-translocation surveys. These long-term forecasts are important,
+given that reintroduced MYL frog populations may often take decades to
+achieve our ultimate goal of self-sustainability \citep{joseph2018}.
+Making well-supported projections about the long-term outcome of
+reintroduction efforts from shorter-term information is critically
+important to the process of adaptive management of species
+reintroduction programs \citep{seddon2007}, including the one we are
+carrying out for MYL frogs. Specifically, the combined results from our
+reintroduction study and viability modeling indicate that survival of
+frogs in the first year following translocation is an effective proxy of
+longer-term survival and population viability. In addition, given the
+repeatability of frog survival at a site, 1-year frog survival also
+serves as an effective proxy of site quality (i.e., the ability of a
+site to support high frog survival and a viable frog population over the
+long term). This proxy of site quality is important in the MYL frog
+system because accurately predicting the ability of a site to support a
+viable frog population \emph{a priori} remains difficult, even after
+conducting 24 translocations over 16 years.
+
+Despite the demonstrated resistance of adult MYL frogs against Bd
+infection, individual and population-level impacts of Bd are still
 evident. In an earlier study of 2 of our 12 translocated populations
@@
 immediately following metamorphosis (``metamorphs'') are highly
-susceptible to Bd infection \citep{ellison2018} and as a result
-experience high mortality. This high susceptibility of metamorphs is
-documented in numerous species of anurans, and may result from the
-poorly developed immune system characteristic of this life stage
-\citep{humphries2022}. In naturally recovering and translocated MYL frog
-populations, we suggest that the high mortality of metamorphs is an
-important limitation on subsequent recruitment of new adults. Therefore,
-although adult MYL frogs appear relatively resistant/tolerant, Bd
-infection continues to have important limiting effects on recovering
-populations \citep[see also][]{hollanders2022}.
+susceptible to Bd infection \citep{ellison2019} and as a result
+experience high mortality \citep{rachowicz2006}. This high
+susceptibility of metamorphs is documented in numerous species of
+anurans, and may result from the poorly developed immune system
+characteristic of this life stage \citep{humphries2022}. In naturally
+recovering and translocated MYL frog populations, we suggest that the
+high mortality of metamorphs is an important limitation on subsequent
+recruitment of new adults. Therefore, although adult MYL frogs appear
+relatively resistant, Bd infection continues to have important limiting
+effects on recovering populations \citep[see also][]{hollanders2022}.
 
@@
 hundreds of amphibian species, some of which are now extinct in the wild
-\citep{scheele2019}. This unprecedented impact on global amphibian
+\citep{scheele2019}. This extraordinary impact on global amphibian
 biodiversity is compounded by the lack of any effective and broadly
@@
 success of such efforts will depend on the availability of robust donor
-populations containing individuals that have acquired via evolution the
-adaptive alleles necessary to allow frog survival and population growth
-in the presence of Bd. Despite the hopeful example of successful
-reestablishment of MYL frogs via evolutionary rescue and
-reintroductions, the challenge of recovering hundreds of Bd-impacted
-amphibian species worldwide is a daunting prospect. Although we now have
-a proven strategy to reestablish extirpated MYL frog populations,
-recovery across their large historical range will require substantial
-resources over many decades. The results of this study provide a hopeful
-starting point for that endeavor and other future efforts worldwide.
-
-Evolutionary rescue may be increasingly critical for the survival of
-wildlife populations in a rapidly changing world. The importance of
-evolutionary rescue is known for a wide variety of taxa
-\citep[e.g.,][]{carlson2014, mills2018, oziolor2019}, including in the
-specific context of disease-impacted systems \citep{searle2020}.
-However, obtaining evidence in wild animal populations for all the
-causal links in this process (i.e., selection for adaptive alleles
-\(\rightarrow\) increase in host fitness in the presence of the stressor
-\(\rightarrow\) positive population growth and recovery) is difficult in
-most systems. Nonetheless, we provide a compelling example from the wild
-not only that evolutionary rescue can produce individuals that harbor
-adaptive alleles and allow population recovery, but importantly that
-these individuals can be used in translocations to reestablish
-extirpated populations. We expect that such evolutionary rescue-based
-translocations will be an essential tool in wildlife conservation in an
-era of accelerating global change.
+populations containing individuals that have the adaptive alleles
+necessary to allow frog survival and population growth in the presence
+of Bd. Despite the hopeful example of successful reestablishment of MYL
+frogs despite ongoing Bd infection, the challenge of recovering hundreds
+of Bd-impacted amphibian species globally is a daunting prospect.
+Although we now have a proven strategy to reestablish extirpated MYL
+frog populations, recovery across their large historical range will
+require substantial resources over many decades. The results of this
+study provide a hopeful starting point for that endeavor and other
+future efforts worldwide.
+
+In our rapidly changing world, evolution is likely to play in important
+role in facilitating the resilience of wildlife populations. Whether the
+documented disease resistance in MYL frogs and concurrent recovery of
+decimated populations provides an example of evolutionary rescue will
+likely always be uncertain. Regardless, we provide an example from the
+wild that suggests that evolution can produce individuals that harbor
+adaptive alleles and allow population recovery in a novel (i.e.,
+Bd-positive) environment, and show conclusively that individuals from
+these recovering populations can be used to reestablish extirpated
+populations and expand the scale of natural recovery
+(Figure~\ref{fig-recovery-model}). We expect that similar species
+recovery actions will be an essential tool in wildlife conservation in
+an era of accelerating global change.
 
 \matmethods{
-\hypertarget{frog-evolution-in-response-to-bd-1}{%
-\section*{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-1}}
-
-\hypertarget{sampling-and-sequencing}{%
-\subsection*{Sampling and Sequencing}\label{sampling-and-sequencing}}
-
-We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
-\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
-populations, and 29 from 5 recovering populations. These populations are
-located in the southern Sierra Nevada, from northern Yosemite National
-Park to northern Sequoia National Park
-(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
-frogs per population. To minimize potential confounding effects caused
-by variation in frog genotypes across latitude
-\citep[e.g.,][]{rothstein2020}, we selected sampling sites such that
-both population types were represented across similar latitudinal
-ranges. DNA was extracted following Qiagen DNEasy manufacturer's
-protocols. We sequenced the samples using an exome capture assay as
-described in Byrne et al.~(in review). Briefly, genomic libraries were
-prepared and captured using a custom Nimblegen capture pool. Capture
-baits were designed based on the coding regions of the \emph{R. muscosa}
-transcriptome (GenBank accession GKCT00000000). Captured libraries were
-then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the
-Vincent J. Coates Genomics Sequencing Lab at UC Berkeley. Raw sequencing
-reads are available from NCBI SRA (PRJNA870451).
-
-\hypertarget{data-pre-processing-and-cleaning}{%
-\subsection*{Data Pre-processing and
-Cleaning}\label{data-pre-processing-and-cleaning}}
-
-Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
-SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
-bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
-using Picard. Variants were then called following GATK best practices
-(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
-sample using HaplotypeCaller and combined using CombineGVCFs. Next,
-genotypes were jointly called using GenotypeGVCFs. Variants were then
-hard filtered using gatk VariantFiltration using the following
-parameters to remove low-quality sites: QD \textless{} 2.0, FS
-\textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
-MQRankSum \textgreater{} 3.0, MQRankSum \textless{} -3.0, ReadPosRankSum
-\textgreater{} 3.0, ReadPosRankSum \textless{} -3.0. This initial filter
-resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant
-level. First, we trimmed our variants to only include those with minor
-allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
-depth of 5, a minimum genotype quality of 20, and a maximum missing
-proportion of 0.5. This filter resulted in 427,038 sites, of which
-353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
-with an average depth across filtered sites \textless{} 7x. Our final
-dataset included 50 samples, 23 from naive and 27 from recovering
-populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-\hypertarget{data-analysis}{%
-\subsection*{Data Analysis}\label{data-analysis}}
-
-To visualize the genomic relationships of our populations we conducted a
-PCA using the glPCA function in the adegenet R package
-\citep{jombart2008}. To detect regions of the genome under selection
-between naive and recovering populations, we used two approaches: (1) a
-multivariate linear mixed model to evaluate individual variants (SNPs
-and INDELs), and (2) a splined window analysis to evaluate larger
-genomic regions. For the variant analysis, we first used a stringent
-data filter to include only variants with \textless{} 5\% missing data
-(missing for no more than 2 individuals), and then calculated the
-likelihood ratio statistic for the resulting set of 148,307 high quality
-variants across 127 contigs using GEMMA \citep{zhou2014}. GEMMA
-calculates and incorporates a relatedness matrix for input samples,
-allowing us to account for relatedness and population structure when
-calculating likelihood ratio statistics. We identified variants showing
-different allele frequencies between naive versus recovering populations
-(``outliers'') using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot. We
-developed a more liberal set of outlier variants using a
-Bonferroni-corrected significance level of 0.05 and used this set solely
-for the gene ontology (GO) analysis (see below and SI Appendix: Dataset
-S1, S2).
-
-In the analysis of individual variants, for each outlier variant we
-determined whether the variant was synonymous (protein sequence the same
-for each variant) or non-synonymous (protein sequence differs between
-variants), and where in the gene it was located. To do this, we first
-extracted the reference genome sequence surrounding the variant using
-the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
-re-annotated each sequence using BLAST to get the predicted gene
-location based on the closest annotated reference \citep{altschul1997}.
-We then translated each variant to amino acids and aligned this
-translation to that of the gene annotation to ensure proper frame of
-reference using Geneious \citep{kearse2012}. After ensuring proper
-translation, we characterized variants as within or outside the coding
-sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using
-\emph{F\textsubscript{ST}} and differences in nucleotide diversity
-(\(\pi_{diff}\)) between naive and recovering populations. First, we
-calculated per-site \emph{F\textsubscript{ST}} between the naive and
-recovering individuals for all bi-allelic SNPs in the 30 largest contigs
-(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
-calculated per-site nucleotide diversity \(\pi\) separately for
-individuals from the naive and recovering populations using VCFtools,
-then calculated \(\pi_{diff}\) for each population
-(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
-values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
-size-sorted chromosome number and adjusted the SNP position based on the
-relative position in the genome. We then used the GenWin R package
-\citep{beissinger2015} to conduct a splined discrete window analysis for
-\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
-where non-overlapping window boundaries should occur by identifying
-inflection points in the fitted spline, therefore balancing false
-positive and false negative results that occur using other window-based
-methods \citep{beissinger2015}. This method also calculates a
-W-statistic allowing for outlier identification. We identified outliers
-as those with a W-statistic greater than 4 standard deviations above the
-mean for \emph{F\textsubscript{ST}} or above/below the mean for
-\(\pi_{diff}\). Shared outliers were then identified as those that were
-outliers in both analyses, meaning that they showed (1) high
-differentiation between naive and recovering populations, and (2)
-differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and
-retrieved annotation for that region using BLAST (SI Appendix: Datasets
-S3, S4).
-
-Using the liberal set of outlier variants, we performed a statistical
-overrepresentation test (Fisher's exact test) using Blast2GO
-\citep{conesa2005}. First, we retrieved the BLAST hits and mapped GO
-terms for each gene in our targeted transcriptome. Next, we compared the
-30 unique outlier genes from the liberal set of outlier variants to the
-complete set of genes in our target transcriptome to determine if any GO
-biological functions, molecular functions, or cellular processes were
-overrepresented. We then repeated this process for the set of 33 genes
-located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
-splined windows. Code used for genomic analyses and to create figures is
-available from the following GitHub repository:
-\url{https://github.com/allie128/mylf-selection}.
-
 \hypertarget{frog-population-recovery-1}{%
@@
 from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
-collected using several years of visual encounter surveys (VES) and skin
-swab collections \citep{knapp2016}, and results from population genetic
-analyses \citep{poorten2017}. The populations that we selected contained
-several hundred \emph{R. sierrae} adults and hundreds or thousands of
-tadpoles. These relatively high abundances were the result of recent
-increases following previous Bd-caused declines \citep{knapp2016}. As is
-typical for recovering MYL frog populations, Bd prevalence in the donor
-populations was high (0.69--0.96) and Bd load (median
-log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or more
-orders of magnitude below the level at which frog mortality is expected
-(log\textsubscript{10}(load) \(\approx\) 5.78
-copies)\citep{vredenburg2010, joseph2018}. Recipient sites to which
-frogs were translocated were chosen based on previous \emph{R. sierrae}
-presence (determined from VES and/or museum records) or characteristics
-that suggested high quality habitat for this species \citep{knapp2005}.
-
-We conducted 1--4 translocations per site (Figure S3,
-Figure~\ref{fig-translocation-survival}) and each translocated cohort
-included 18 to 99 frogs (median = 30). In preparation for each
-translocation, adult frogs were collected from the donor population and
-measured, weighed, swabbed, and PIT tagged. Frogs were transported to
-the recipient site either on foot or via helicopter. Following release,
-translocated populations were visited approximately once per month
-during the summer active season (median number of visits per summer = 2;
-range = 1--10) and assessed using diurnal CMR surveys and VES. CMR
-surveys allowed estimation of adult survival, adult population size, and
-recruitment of new adults, and VES provided estimates of tadpole and
-juvenile abundance. During 2006-2012, CMR surveys were conducted on a
-single day (primary period) per site visit, during which all habitats
-were searched repeatedly for adult frogs. Frogs were captured using
-handheld nets, identified via their PIT tag (or tagged if they were
-untagged), measured, weighed, swabbed, and released at the capture
-location. During 2013-2022, we generally used a robust design in which
-all habitats were searched during several consecutive days (median
-number of secondary periods per primary period = 3; range = 3--7), and
-frogs were processed as described above. However, when the number of
-frogs detected on the first survey day was zero or near zero, we
-typically conducted only a single-day CMR survey. When using a robust
-design, within a primary period, frogs that were captured during more
-than one secondary period were measured, weighed, and swabbed during the
-first capture, and during subsequent captures were only identified and
-released.
+collected using several years of VES and skin swab collections
+\citep{knapp2016}, and results from population genetic analyses
+\citep{poorten2017}. The populations that we selected contained hundreds
+to thousands of both \emph{R. sierrae} adults and tadpoles. These
+relatively high abundances were the result of recent increases following
+previous Bd-caused declines \citep{knapp2016}. As is typical for
+recovering MYL frog populations, Bd prevalence in the donor populations
+was high (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
+3.06--3.78 ITS copies) was two or more orders of magnitude below the
+level at which frog mortality is expected (log\textsubscript{10}(load)
+\(\approx\) 5.78 copies)\citep{vredenburg2010, joseph2018}. Recipient
+sites to which frogs were translocated were chosen based on previous
+\emph{R. sierrae} presence (determined from VES and/or museum records)
+or characteristics that suggested high quality habitat for this species
+\citep{knapp2005}. At the beginning of this study, we had a relatively
+limited understanding of the factors that affect habitat quality. In
+subsequent years, we improved our site selection process by
+incorporating new information about important habitat features, in
+particular, overwinter habitats such as submerged boulders and
+overhanging banks. \emph{R. sierrae} were absent from all recipient
+sites prior to the first translocation.
+
+We conducted 1--4 translocations per site
+(Figure~\ref{fig-translocation-survival}, Figure S1)
+and each translocated cohort included 18 to 99 frogs (median = 30). In
+preparation for each translocation, adult frogs were collected from the
+donor population and measured, weighed, swabbed, and PIT tagged. Frogs
+were transported to the recipient site either on foot or via helicopter.
+Following release, we visited translocated populations approximately
+once per month during the summer active season and conducted diurnal CMR
+surveys and VES (summer active season is generally July-August but can
+start as early as May and end as late as September; range of survey
+dates = May-25 to Sep-29, range of translocation dates = Jun-28 to
+Sep-02; median number of visits per summer = 2, range = 1--10). CMR
+surveys allowed estimation of adult survival, recruitment of new adults,
+and adult population size, and VES provided estimates of tadpole and
+juvenile abundance. During 2006-2012, we conducted CMR surveys on a
+single day (primary period) per site visit, during which we searched all
+habitats repeatedly for adult frogs. Frogs were captured using handheld
+nets, identified via their PIT tag (or tagged if they were untagged),
+measured, weighed, swabbed, and released at the capture location. During
+2013-2022, we generally used a robust design in which all habitats were
+searched during several consecutive days (median number of secondary
+periods per primary period = 3; range = 3--7), and frogs were processed
+as described above. However, when the number of frogs detected on the
+first survey day was zero or near zero, we typically conducted only a
+single-day CMR survey. When using a robust design, within a primary
+period, frogs that were captured during more than one secondary period
+were measured, weighed, and swabbed during the first capture, and during
+subsequent captures were only identified and released.
 
@@
 accounted for the fact that all translocations were conducted after the
-breeding period and reproduction would therefore not have occurred until
-the following year. Similarly, to calculcate the proportion of years in
-which new recruits were present at a site, we excluded surveys conducted
+breeding period and reproduction would therefore not occur until the
+following year. Similarly, to calculate the proportion of years in which
+new recruits were present at a site, we excluded surveys conducted
 during the 3 years following the initial translocation. This accounted
 for the multi-year tadpole and juvenile stages in MYL frogs
-(Table S2).
+(Table S2 SI).
 
@@
 
-We estimated survival of translocated frogs, recruitment of new frogs
-into the adult population, and adult population size using a Bayesian
-open-population Jolly-Seber mark-recapture model with known additions to
-the population (i.e., translocated cohorts), as implemented by the mrmr
-package \citep{joseph2019} and using R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022}. The model tracks the states of
-\emph{M} individuals that comprise a superpopulation made up of real and
-pseudo-individuals \citep[see][ for details]{joseph2018}. The possible
-states of individuals include ``not recruited'', ``alive'', and
-``dead''. The possible observations of individuals include ``detected''
-and ``not detected''. We assume that individuals that are in the ``not
-recruited'' or ``dead'' states are never detected (i.e., there are no
-mistakes in the individual PIT tag records). For all models, we used
-mrmr defaults for priors, number of chains (4), and warmup and
-post-warmup iterations (2000 for each). We evaluated convergence of the
-Markov chain Monte Carlo (MCMC) algorithm using trace plots and
-Gelman-Rubin statistics (Rhat).
+For each translocation site, we estimated survival of translocated
+frogs, recruitment of new frogs into the adult population, and adult
+population size using a site-specific Bayesian open-population
+Jolly-Seber CMR model with known additions to the population (i.e.,
+translocated cohorts), as implemented by the mrmr package
+\citep{joseph2019} and using R Statistical Software
+\citep[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
+Frog population recovery - CMR model structure} for details). Briefly,
+the model tracks the states of \emph{M} individuals that comprise a
+superpopulation made up of real and pseudo-individuals \citep[see][ for
+details]{joseph2018}. The possible states of individuals include ``not
+recruited'', ``alive'', and ``dead''. The possible observations of
+individuals include ``detected'' and ``not detected''. We assume that
+individuals that are in the ``not recruited'' or ``dead'' states are
+never detected (i.e., there are no mistakes in the individual PIT tag
+records). We also assume that new recruits were the result of
+within-site reproduction and not immigration from adjacent populations.
+This assumption is justified by the fact that no \emph{R. sierrae}
+populations were present within several kilometers of the translocation
+sites. For all models, we used mrmr defaults for priors, number of
+chains (4), and warmup and post-warmup iterations (2000 for each). We
+evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm
+using trace plots and Gelman-Rubin statistics (Rhat).
 
@@
 To identify important predictors of frog survival following
-translocation, we developed linear and multilevel models in a Bayesian
-framework. We performed all analyses with the rstanarm package
+translocation, we used multilevel Bayesian models
+\citep{gelman2013, gabry2019}. Included predictor variables describe
+characteristics of sites, translocated cohorts, and individuals (Bd
+load, sex, frog size, site elevation, winter severity in the year of
+translocation, winter severity in the year following translocation,
+donor population, day of year on which a translocation was conducted,
+and translocation order). We used 1-year post-translocation survival
+estimates from CMR models as the response. Estimated survival was
+rounded to integer values to produce a binary outcome, and modeled with
+a Bernoulli distribution. Group-level (random) effects included
+site\_id, translocation\_id, or translocation\_id nested within
+site\_id. We performed all analyses with the rstanarm package
 \citep{rstanarm2022} and R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022}, and used default, weakly informative
-priors. Starting models included all relevant population-level effects
-and, when data contained possible hierarchical structure, we included
-group-level effects in the full model. We evaluated model fit using
-leave-one-out (``LOO'') cross-validation
-\citep{gelman2013, gabry2019, vehtari2016}, as implemented by the loo
-package \citep{vehtari2022}.
-
-We conducted two analyses. In the first, we addressed the more general
-question of whether survival of frogs in the first translocation to a
-site predicted the survival of frogs in subsequent translocations.
-Group-level effects included site\_id or translocation\_id. The second
-analysis focused specifically on identifying predictors of
-post-translocation frog survival using a wide range of variables
-describing characteristics of frogs, sites, and translocations
-(predictor variables: Bd load, sex, frog size, site elevation, winter
-severity in the year of translocation, winter severity in the year
-following translocation, donor population, day of year on which a
-translocation was conducted, and translocation order). In both analyses
-we used 1-year survival as a single metric of survival that was
-available for all translocated populations regardless of how many years
-of post-translocation data were available (range = 3--16). Group-level
-effects included site\_id, translocation\_id, or translocation\_id
-nested within site\_id. In both analyses, the response variable,
-probability of 1-year frog survival, was rounded to integer values to
-produce a binomial outcome, and all models used a bernoulli family. For
-all models, we used default priors, four chains, and 5000 iterations
-each for warmup and post-warmup. We checked convergence of the MCMC
-algorithm using trace plots and Rhat.
+\citep[v4.4.4,][]{rsoftware2022}. For all models, we used default,
+weakly informative priors, four chains, and 5000 iterations each for
+warmup and post-warmup. We checked MCMC convergence using trace plots
+and Rhat, and evaluated model fit using leave-one-out cross-validation
+\citep{vehtari2016}, as implemented by the loo package
+\citep{vehtari2022}. (See \textbf{Supporting Information - Frog
+population recovery - Among-site survival modeling} for details.)
 
@@
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see SI Appendix for details]{boyle2004}. To assess the magnitude
-of changes in Bd load on frogs following translocation, we compared Bd
+\citep[see \textbf{Supporting Information-Frog population recovery -
+Laboratory methods} for details]{boyle2004}. To assess the magnitude of
+changes in Bd load on frogs following translocation, we compared Bd
 loads measured before versus after translocation. Before-translocation
@@
 associated Bd loads were included in the dataset only if frogs were
-captured at the recipient site at least once during the year 0-1 period.
-We used an analysis workflow similar to that described above. Briefly,
-the response variable was before-translocation load and the predictor
-variable was after-translocation load. We compared models with and
-without group-level effects, and group-level effects included
-pit\_tag\_id, pit\_tag\_id nested within site\_id, or pit\_tag\_id
-nested within translocation\_id. All models used a negative binomial
-family. Priors, chains, iterations, and checks were as described above.
+captured at the recipient site at least once during the 1-year period
+following translocation.
 
@@
 probability of translocated adults and naturally recruited adults
-differs (Figure S7).
+differs (Figure S4).
 
@@
   0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
+  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
   0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
@@
 as a tadpole \(p_{L2}\), and probability of a first-year juvenile
-remaining as a juvenile \(p_{J1}\), and the probability of a juvenile
-successfully recruiting to an adult \(\omega\).
+remaining as a juvenile \(p_{J1}\). First-year juvenile survival and
+recruitment \(\sigma_{J1}\) is the parameter that we think is most
+influenced by environmental stochasticity.
 
@@
 prevalence significantly simplifies the model without much loss of
-realism. Additional details are provided in the SI Appendix.
+realism. Additional details are provided in \textbf{Supporting
+Information - Population viability modeling}.
 
@@
 deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to model parameters, to
-quantify how much changes in adult survival, for example, affected the
-long-run growth rate. Third, we included demographic and environmental
-stochasticity in our model and simulated the 50-year viability (i.e., 1
-- extinction probability) of populations given an introduction of 40
-adult individuals into an unoccupied habitat. Finally, we fit our model
-to our longest translocation trajectory to confirm that our model could
-reasonably reproduce the observed recovery trajectories of MYL frogs
-following reintroductions. Additional details are provided in the SI
-Appendix. Code to replicate the analyses can be found at
+we compute the elasticity of \(\lambda\) to four key model parameters to
+quantify how much changes in these parameters affected the long-run
+growth rate (Figure S5). This also helped us
+determine where in the model environmental variation in juvenile
+survival and recruitment would have the largest effects on population
+dynamics. Third, we included demographic stochasticity and environmental
+stochasticity in \(\sigma_{J_1}\) in our model and simulated the 50-year
+viability (i.e., 1 - extinction probability) of populations given an
+introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to
+confirm that our model could reasonably reproduce the observed recovery
+trajectories of MYL frogs following reintroductions. Additional details
+are provided in \textbf{Supporting Information - Population viability
+modeling}. Code to replicate the analyses can be found at
 \url{https://github.com/SNARL1/translocation}.
+
+\hypertarget{frog-evolution-in-response-to-bd-1}{%
+\section*{Frog evolution in response to
+Bd}\label{frog-evolution-in-response-to-bd-1}}
+
+\hypertarget{sampling-and-sequencing}{%
+\subsection*{Sampling and sequencing}\label{sampling-and-sequencing}}
+
+We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
+\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
+populations, and 29 from 5 recovering populations. These populations are
+located in the southern Sierra Nevada, from northern Yosemite National
+Park to northern Sequoia National Park
+(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
+frogs per population. To minimize potential confounding effects caused
+by variation in frog genotypes across latitude \citep{byrne2023}, we
+selected sampling sites such that both population types were represented
+across similar latitudinal ranges. DNA was extracted following Qiagen
+DNEasy manufacturer's protocols. We sequenced the samples using an exome
+capture assay as described in \citep{byrne2023}. Briefly, genomic
+libraries were prepared and captured using a custom Nimblegen capture
+pool. Capture baits were designed based on the coding regions of the
+\emph{R. muscosa} transcriptome (GenBank accession GKCT00000000).
+Captured libraries were then pooled and sequenced on a NovaSeq 6000
+150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at
+UC Berkeley. Raw sequencing reads are available from NCBI SRA
+(PRJNA870451).
+
+\hypertarget{data-pre-processing-and-cleaning}{%
+\subsection*{Data pre-processing and
+cleaning}\label{data-pre-processing-and-cleaning}}
+
+Raw reads were filtered for adapters and contaminants using fastp
+\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
+SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
+bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
+using Picard. Variants were then called following GATK best practices
+(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
+sample using HaplotypeCaller and combined using CombineGVCFs. Next,
+genotypes were jointly called using GenotypeGVCFs. Variants were then
+hard filtered using gatk VariantFiltration using the following
+parameters to remove low-quality sites: QD \textless{} 2.0, FS
+\textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
+MQRankSum \textgreater{} 3.0, MQRankSum \textless{} -3.0, ReadPosRankSum
+\textgreater{} 3.0, ReadPosRankSum \textless{} -3.0. This initial filter
+resulted in 1,595,206 variant sites across 53 individuals.
+
+We then further filtered our dataset at the individual and variant
+level. First, we trimmed our variants to only include those with minor
+allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
+depth of 5, a minimum genotype quality of 20, and a maximum missing
+proportion of 0.5. This filter resulted in 427,038 sites, of which
+353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
+with an average depth across filtered sites \textless{} 7x (n = 3). Our
+final dataset included 50 samples, 23 from naive and 27 from recovering
+populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+
+\hypertarget{data-analysis}{%
+\subsection*{Data analysis}\label{data-analysis}}
+
+To visualize the genomic relationships of our populations, we conducted
+a PCA using the glPCA function in the adegenet R package
+\citep{jombart2008}. To detect regions of the genome that differed
+between naive and recovering populations, i.e., regions putatively under
+selection, we used two approaches: (1) a multivariate linear mixed model
+to evaluate individual variants (SNPs and INDELs), and (2) a splined
+window analysis to evaluate larger genomic regions. For the variant
+analysis, we first used a stringent data filter to include only variants
+with \textless{} 5\% missing data (missing for no more than 2
+individuals), and then calculated the likelihood ratio statistic for the
+resulting set of 148,307 high quality variants across 127 contigs using
+GEMMA \citep{zhou2014}. GEMMA calculates and incorporates a relatedness
+matrix for input samples, allowing us to account for relatedness and
+population structure when calculating likelihood ratio statistics. We
+identified variants showing different allele frequencies between naive
+versus recovering populations (``outliers'') using a
+Bonferroni-corrected significance level of 0.01. We visualized the
+results using a Manhattan plot and qqplot. We developed a more liberal
+set of outlier variants using a Bonferroni-corrected significance level
+of 0.05 and used this set solely for the gene ontology (GO) analysis
+(see below and \textbf{Supporting Information}: Dataset S1, S2).
+
+In the analysis of individual variants, for each outlier variant we
+determined whether the variant was synonymous (protein sequence the same
+for each variant) or non-synonymous (protein sequence differs between
+variants), and where in the gene it was located. To do this, we first
+extracted the reference genome sequence surrounding the variant using
+the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
+re-annotated each sequence using BLAST to get the predicted gene
+location based on the closest annotated reference \citep{altschul1997}.
+We then translated each variant to amino acids and aligned this
+translation to that of the gene annotation to ensure proper frame of
+reference using Geneious \citep{kearse2012}. After ensuring proper
+translation, we characterized variants as within or outside the coding
+sequence of the gene and as either synonymous or non-synonymous.
+
+In the splined window analysis, we identified outlier regions using
+\emph{F\textsubscript{ST}} and differences in nucleotide diversity
+(\(\pi_{diff}\)) between naive and recovering populations. First, we
+calculated per-site \emph{F\textsubscript{ST}} between the naive and
+recovering individuals for all bi-allelic SNPs in the 30 largest contigs
+(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
+calculated per-site nucleotide diversity \(\pi\) separately for
+individuals from the naive and recovering populations using VCFtools,
+then calculated \(\pi_{diff}\) for each population
+(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
+values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
+size-sorted chromosome number and adjusted the SNP position based on the
+relative position in the genome (for more efficient data processing and
+to better contextualize the strength of the outlier signals in different
+regions of the genome). We then used the GenWin R package
+\citep{beissinger2015} to conduct a splined discrete window analysis for
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
+where non-overlapping window boundaries should occur by identifying
+inflection points in the spline fitted to \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) values along the genome, therefore balancing false
+positive and false negative results that occur using other window-based
+methods \citep{beissinger2015}. This method also calculates a
+W-statistic allowing for outlier identification. We identified outliers
+as those with a W-statistic greater than 4 standard deviations above the
+mean for \emph{F\textsubscript{ST}} or above/below the mean for
+\(\pi_{diff}\). These standard deviations represent strict criteria to
+select only the top \textasciitilde{} 0.3\% of windows. Shared outliers
+were then identified as those that were outliers in both analyses,
+meaning that they showed (i) high differentiation between naive and
+recovering populations, and (ii) differential patterns of nucleotide
+diversity in the same region. Finally, we extracted gene transcripts
+mapped within each region and retrieved annotation for that region using
+BLAST (\textbf{Supporting Information}: Datasets S3, S4).
+
+Code used for genomic analyses and to create figures is available from
+the following GitHub repository:
+\url{https://github.com/allie128/mylf-selection}.
 
@@
 
+{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
+
+}
+
+\caption{\label{fig-recovery-model}For \emph{R. sierrae}, a conceptual
+model depicting the Bd-caused decline and subsequent natural recovery
+(black text), facilitated recovery via reintroductions, and the linkages
+between these two pathways. Rectangles and hexagons represent outcomes
+and processes, respectively. Blue text indicates components that are
+included in the current study. The timeline shows the general sequence
+of the components, with the dotted portion indicating a projection into
+the future.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
+
+}
+
+\caption{\label{fig-translocation-survival}Median 1-year survival for
+each cohort of translocated frogs at the 12 recipient sites, as
+estimated for each site from the mrmr CMR model. Error bars show the
+95\% uncertainty intervals. Sites are arranged along the x-axis using
+the average of the median 1-year survival per translocation at each
+site. Dot colors indicate the donor population from which frogs in each
+translocated cohort were collected. When multiple translocations were
+conducted to a site, points and error bars are slightly offset to avoid
+overlap.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
+
+}
+
+\caption{\label{fig-cond-effects}Results show conditional effects of the
+important predictors of 1-year frog survival (survival is expressed as a
+probability), as estimated from the model used in the among-site
+meta-analysis: (A) winter severity in the year following translocation,
+(B) elevation of recipient site, (C) donor population, and (D) sex. In
+(A) and (B), blue lines are medians and gray ribbons are 95\%
+uncertainty intervals. In (C) and (D), box plots show medians, first and
+third quartiles, largest and smallest values within 1.5x interquartile
+range, and values outside the 1.5x interquartile range.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+
+}
+
+\caption{\label{fig-viability}Results from population viability
+analysis: (A) Predicted long-run growth rate \(\lambda\) for different
+values of yearly adult survival probability \(\sigma_{A_R}\) and year 1
+juvenile survival and recruitment probability \(\sigma_{J_1}\), given
+the parameterized, deterministic model. Colored points show the
+predicted \(\lambda\) values for the twelve translocated populations
+when year 1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
+(indicated by the dashed line). The red line shows where
+\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
+other points. (B) Predicted 50-year extinction probabilities of the 12
+translocated populations, given demographic stochasticity, environmental
+variability in \(\sigma_{J_1}\), and different mean values of
+\(\sigma_{J_1}\). There are 6 lines at extinction probability = 1, 5 of
+which (70414--70619) are hidden beneath the line for 70505 when
+\(\sigma_{J_1} < 0.10\). (C) 100 simulated trajectories (black lines)
+from the population viability model that most closely matched the
+observed abundance trajectory of adult amphibians at site 70550 (light
+orange).}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
 {\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
@@
 gray. The inset map locates the study area and range boundary in
-California.}
+California. Sites 1 and 4 (site id = 72996 and 70567, respectively) were
+also used as sources of frogs in the frog population recovery study.}
 
@@
 \newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
-
-}
-
-\caption{\label{fig-translocation-survival}Evidence for frog survival at
-the recipient sites following translocation. Median 1-year survival for
-each cohort of translocated frogs at the 12 recipient sites. Error bars
-show the 95\% uncertainty intervals. Sites are arranged along the x-axis
-using the average of the median 1-year survival per translocation at
-each site. Dot colors indicate the donor population from which frogs in
-each translocated cohort were collected. When multiple translocations
-were conducted to a site, points and error bars are slightly offset to
-avoid overlap.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
-
-}
-
-\caption{\label{fig-cond-effects}Evidence for reduced impacts of Bd on
-MYL frogs as indicated by the lack of an effect of Bd load on survival.
-Results show conditional effects of the important predictors of 1-year
-frog survival (survival is expressed as a probability). (A) winter
-severity in the year following translocation, (B) elevation of recipient
-site, (C) donor population, and (D) sex. In (A) and (B), blue lines are
-medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
-box plots show medians, first and third quartiles, largest and smallest
-values within 1.5x interquartile range, and values outside the 1.5x
-interquartile range.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
-
-}
-
-\caption{\label{fig-viability}Population viability analysis of
-translocated MYL frogs quantified the long-run growth rate and
-extinction probability of frog populations: (A) Predicted long-run
-growth rate \(\lambda\) for different values of yearly adult survival
-probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
-given the parameterized, deterministic model. Colored points show the
-predicted \(\lambda\) values for the twelve translocated populations
-when probability of recruitment into the adult class is \(\omega = 0.3\)
-(indicated by the dashed line). The red line shows where
-\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
-other points. (B) Predicted 50-year extinction probabilities of the 12
-translocated populations, given demographic stochasticity, environmental
-variability in \(\omega\), and different mean values of \(\omega\). The
-dashed line shows where \(\omega = 0.3\), for comparison to (A). There
-are 6 lines at extinction probability = 1, 5 of which (70414--70619) are
-hidden beneath the line for 70505. (C) 100 simulated trajectories (black
-lines) from the population viability model that most closely matched the
-observed abundance trajectory of adult amphibians at site 70550 (light
-orange).}
-
-\end{figure*}
-
-\newpage
 \end{document}

### 2023-12-20 doc/manuscript/translocation.qmd

--- 2023-12-19 doc/manuscript/translocation.qmd
@@
 Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information-Population viability modeling-Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp)
+We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information-Population viability modeling-Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
 

### 2023-12-20 doc/manuscript/translocation_bioarxiv_SI.tex

@@
 \usepackage{longtable}
+\usepackage{blkarray}
 
-\title{Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
+\title{Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease}
 \author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
@@
 
-%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
+%% Adds the main heading for the text. Comment out this line if you do not have any supporting information text.
 \SItext
-\hypertarget{frog-evolution-in-response-to-bd-2}{%
-\section{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-2}}
-
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more
-of data from visual encounter surveys and Bd surveillance using skin
-swabbing \citep[e.g.,][]{knapp2016}. Comparing populations with
-different infection histories allowed larger sample sizes and
-replication across the landscape. The alternative approach of comparing
-samples from the same populations before and after Bd exposure isn't
-feasible in this system because Bd arrived in most MYL frog populations
-decades ago and population persistence/recovery is rare and
-unpredictable. As a result, samples from recovering populations
-collected from before and after Bd exposure are not available and are
-unlikely to be available in the future.
-
-\hypertarget{results-1}{%
-\subsection{Results}\label{results-1}}
-
-In the Results section of the main paper, we describe the stringent set
-of outlier SNPs (identified using a Bonferroni-corrected p-value of
-0.01). The liberal set, identified using a Bonferroni-corrected p-value
-of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
-genes across 16 contigs. Our analysis found no significantly
-over-represented GO terms in this set of 30 genes, or in the set of 38
-outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and
-recovering populations, we detected consistent allele frequency changes
-across all genetic clusters for only 1 of 7 of the associated genes
-(RIN3; Fig. 1). However, for the other 6
-genes we still found a statistically significant signal of selection in
-2 of the 3 genetic clusters (containing populations 6--9;
-Fig. 1, Figure~\ref{fig-selectionresults}
-A). Therefore, although some outlier variant associations have a more
-limited geographic extent than RIN3, they still describe parallel
-evolutionary changes following Bd exposure and are therefore interesting
-candidates for future studies.
-
 \hypertarget{frog-population-recovery-2}{%
@@
 
-\hypertarget{methods-1}{%
-\subsection{Methods}\label{methods-1}}
+\hypertarget{laboratory-methods}{%
+\subsection{Laboratory Methods}\label{laboratory-methods}}
 
@@
 
+\hypertarget{cmr-model-structure}{%
+\subsection{CMR model structure}\label{cmr-model-structure}}
+
+We estimated survival and recruitment for each site using open
+population CMR models based on \citep{joseph2018}. For each individual
+\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
+the individual was not detected, and \(o_{i, j}=2\) if the individual
+was detected. Capture histories of \(M\) individuals are modeled, though
+only \(N_s\) individuals were captured.
+
+This parameter expanded data augmentation allows us to capture the
+possibility that undetected individuals may have recruited into the
+adult population \citep{royle2012}. Here, \(M\) was chosen to be three
+times the number of observed individuals (\(3N_s\)) to be considerably
+greater than our prior guess of \(N_s\).
+
+We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
+period \(t = 1,..., n_t\). The four states that we consider are:
+\(u_{i, t} = 1\) for individuals that have not recruited,
+\(u_{i, t} =2\) for live adults, and \(u_{i, t} = 3\) for dead adults.
+Each survey \(j=1, ..., n_j\) occurs in one of the \(n_t\) primary
+periods, and we denote the primary period in which survey \(j\) takes
+place as \(t_j\). Each primary period \(t\) occurs within one year, but
+within a year there can be multiple primary periods. We set the year
+containing the first primary period to \(y_{t = 1} = 1\), and generally
+\(y_t\) represents the year containing primary period \(t\). Years
+increment by one until the final year of the mark recapture efforts,
+which we denote \(n_y\): \(y \in \{1, 2, ..., n_y\}\). We assume that
+within a primary period, the state of each individual does not change
+(i.e., individuals do not recruit into the adult population, gain or
+lose Bd infection, or die). This assumption is justified by the short
+time intervals between surveys within primary periods, in cases where
+primary periods contain multiple secondary periods.
+
+Live individuals are detected with probability \(p_j\), which is modeled
+as:
+
+\[p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),\]
+
+where \(X_j^{(p)}\) is a known row vector and \(\beta^{(p)}\) an unknown
+parameter vector. Not recruited and dead individuals are never captured.
+We bundle these assumptions about the observation probabilities for
+survey \(j\) into an emission matrix \(\Omega_j\):
+
+\[
+\Omega_j =
+\begin{blockarray}{ccc}
+  \text{Not detected} & \text{Detected} \\
+\begin{block}{(cc)c}
+  1 & 0 & \text{Not recruited} \\
+  1 - p_j & p_j & \text{Alive} \\
+  1 & 0 & \text{Dead} \\
+\end{block}
+\end{blockarray}
+\]
+
+The state transition matrix \(\Psi_{t, i}\) contains the probabilities
+of individual \(i\) transitioning from state \(u_{i, t}\) (rows) to
+\(u_{i, t+1}\) (columns) between primary period \(t\) and \(t+1\). For
+unintroduced individuals which naturally recruit, this matrix is given
+by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(\lambda_t\) is the probability of recruiting in time \(t\)
+and \(\phi_t\) is the probability of survival in time \(t\).
+
+For introduced individuals which have deterministic recruitment (they
+recruit when introduced), the state transition matrix is given by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(I_{t, i}\) is a known indicator function for whether
+individual \(i\) was introduced in primary period \(t\).
+
+We allow recruitment probabilities to vary in time via random effects,
+such that:
+
+\[\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),\]
+
+where \(\alpha^{(\lambda)}\) is an intercept parameter and
+\(\epsilon^{(\lambda)}_t\) is an adjustment for time \(t\).
+
+Survival probabilities also vary in time, and as a function of known
+covariates:
+
+\[\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),\]
+
+where \(X^{(\phi)}_t\) is a row vector of known covariates,
+\(\beta^{(\phi)}\) is a column vector of unknown coefficients, and
+\(\epsilon^{(\phi)}_t\) is an adjustment for time \(t\).
+
+To complete the specification of the Bayesian model, we need to specify
+priors for all unknown parameters. The recruitment parameter priors were
+specified as follows:
+
+\[\alpha^{(\lambda)} \sim N(0,1),\]
+\[\sigma^{(\lambda)} \sim N_+(0,1),\]
+\[\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),\] for periods
+\(t=1, ..., T\). Here \(N\) represents the normal distribution and
+\(N_+\) the half normal distribution with positive support.
+
+Survival parameter priors were specified similarly as:
+
+\[\beta^{(\phi)} \sim N(0, 1),\] \[\sigma^{(\phi)} \sim N_+(0, 1),\]
+\[\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),\] for time
+\(t=1, ..., T\).
+
+The detection model coefficient vector also received a standard normal
+prior \(\beta^{(p)} \sim N(0, 1)\).
+
+We computed the likelihood of each individual capture history using the
+forward algorithm, and we estimated the latent states using the
+forward-backward algorithm \citep{zucchini2009, joseph2018}.
+
+All of the code to specify and fit the model in Stan is available in the
+open source mrmr package \citep{joseph2019}.
+
+The joint distribution of the resulting model can be written as follows:
+
+\begin{multline*}
+[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{multline*}
+
+where \(\pmb Y\) is an \(M \times T\) detection matrix, and \(Y_i\) the
+capture history of individual \(i\).
+
+\hypertarget{among-site-survival-modeling}{%
+\subsection{Among-site survival
+modeling}\label{among-site-survival-modeling}}
+
+We used Bayesian generalized linear mixed models to investigate
+predictors of survival among sites. The response \(y_i\) is binary,
+representing a point estimate of whether individual \(i\) survived in
+the year following translocation. We generated these point estimates by
+rounding the posterior median of 1-year post-introduction survival for
+each individual, and modeled the data using a Bernoulli distribution:
+
+\[y_i \sim \text{Bernoulli}(p_i),\]
+
+where \(p_i\) is the probability of survival.
+
+We modeled variation in probabilities as follows:
+
+\[\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},\]
+
+where \(\alpha\) is an intercept, \(X_i\) is a length \(K\) row vector
+of predictors, \(\pmb \beta\) is a column vector of predictor effects,
+and \(\pmb \nu\) a vector of group level random effects. Here \(g[i]\)
+refers to the group \(g\) containing individual \(i\), and we estimate
+an adjustment for each of the \(G\) groups (\(\nu_1, ..., \nu_G\)).
+
+These models were fit using the \texttt{stan\_glmer()} function in the
+rstanarm package, with default priors described below
+\citep{rstanarm2022}. These priors are vague, but include data dependent
+scaling as follows to account for different input variable scales.
+However, as we standardized all predictor variables similarly to have
+equal variance (by centering and dividing by twice the sample standard
+deviation), the resulting priors are identical. Specifically, we have:
+
+\[\alpha \sim \text{Normal}(0, 2.5),\] and
+
+\[\beta_k \sim \text{Normal}(0, 5),\]
+
+for \(k=1, ..., K\) where \(K\) is the number of predictor variables.
+
+The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
+rstanarm is a zero-mean Gaussian, where the covariance matrix is
+constructed from a correlation matrix with an LKJ prior, and a vector of
+variance parameters -- the decomposition of variance prior with unit
+regularization, concentration, shape, and scale parameters
+\citep{lewandowski2009}.
+
+We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
+with four parallel chains, each run for 10,000 iterations, discarding
+the first half of each chain as warm-up draws \citep{rstanarm2022}. We
+used Rhat statistics and trace plots to verify convergence. We
+considered models with different subsets of fixed and random effects,
+and used approximate leave-one-out cross validation to identify the best
+model \citep{vehtari2016}.
+
+Finally, while it would theoretically be possible to estimate survival
+covariate effects in a joint CMR model that integrates capture histories
+across all sites, this was impractical due to computational requirements
+of the CMR models (namely, run time and memory). In modeling survival
+estimates from site-specific CMR analyses, we are in effect conducting
+an among-site meta-analysis to better understand how site, cohort, and
+individual level characteristics explain variation in survival after
+translocation.
+
 \hypertarget{population-viability-modeling-1}{%
-\section{Population viability
+\subsection{Population viability
 modeling}\label{population-viability-modeling-1}}
-
-\hypertarget{methods-2}{%
-\subsection{Methods}\label{methods-2}}
 
@@
 
-We extracted yearly survival probabilities for translocated adults
+We computed yearly survival probabilities for translocated adults
 \(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
-the CMR model used to estimate population-level parameters of
-translocated populations (see ``Materials and Methods - Estimation of
-frog survival and abundance'' in main paper). Although we observed
-yearly variability in adult survival within a population, the magnitude
-of this variability was small compared to among-population variability
+the posterior distribution of individual state trajectories derived from
+CMR models. Although we observed yearly variability in adult survival
+within a population, the magnitude of this variability was small
+compared to among-population variability
 (Fig. 3). Thus, we did not include
@@
 from Poisson variability around a mean value. Therefore, we allowed for
-yearly variability in the probability of successful recruitment
-\(\omega\) (additional details provided in ``Estimating model
-parameters'' below).
+yearly variability in the probability of year 1 juvenile survival and
+recruitment (additional details provided in \textbf{Estimating model
+parameters} below). We also could have included environmental
+stochasticity in year 2 juvenile survival and recruitment
+\(\sigma_{J_2}\), but our elasticity analysis showed that this parameter
+had little effect on host growth rate relative to \(\sigma_{J_1}\)
+(Figure~\ref{fig-viability-supp}).
 
@@
 and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
-``Materials and Methods - Estimation of frog survival and abundance'' in
-main paper). For populations where we had a sufficient number of
-PIT-tagged, naturally-recruited adults, we observed that
-\(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different, with
-\(\sigma_{A_R} > \sigma_{A_T}\) (Figure~\ref{fig-compare_surv_probs}).
-For populations lacking sufficient numbers of naturally-recruited
+\textbf{Materials and Methods - Estimation of frog survival and
+abundance} and \textbf{Supporting Information - Frog population recovery
+- Estimation of frog survival and abundance} for details). For
+populations where we had a sufficient number of PIT-tagged,
+naturally-recruited adults, we observed that \(\sigma_{A_T}\) and
+\(\sigma_{A_R}\) could be notably different, with
+\(\sigma_{A_R} > \sigma_{A_T}\) (Figure~\ref{fig-compare_surv_probs}). For populations lacking sufficient numbers of naturally-recruited
 adults, we were unable to directly estimate \(\sigma_{A_R}\), and
 instead set \(\sigma_{A_R} = \sigma_{A_T}\).
+
+To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
+predicted true states for naturally-recruited individuals (1=not
+recruited, 2=alive, 3=dead, as described in \emph{CMR model structure}),
+then calculated the posterior probability of individuals surviving
+between consecutive primary periods, conditional on being alive in the
+first primary period (e.g., given a value of 2 (alive) in the first
+primary period, how often was the value still 2 (alive) in the next
+primary period compared to 3 (dead)?) . This yielded posterior
+distributions for survival probabilities between primary periods.
+However, because the time interval between primary periods differed, the
+survival probabilities between different consecutive primary periods
+were not directly comparable. To address this, we converted the survival
+probabilities between each consecutive primary period to per day death
+rates, propagating the uncertainty from the posterior distributions. We
+then took a weighted average of these death rates, weighted by the time
+interval between primary periods, to get the average per day death rate
+over the entire CMR survey. We converted this per day death rate \(d\)
+back to a yearly survival probability using
+\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
+\(\sigma_{A_T}\) such that our estimates of average yearly survival
+probability were comparable between \(\sigma_{A_R}\) and
+\(\sigma_{A_T}\).
 
@@
 We performed four analyses on our model. First, we considered a
-deterministic version of our model with no yearly heterogeneity in
-probability of successful recruitment \(\omega\), and calculated the
-predicted long-run growth rate \(\lambda\) of a population for different
-values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
-\(\omega = 0.3\) and calculated the predicted growth rate of our 12
-populations.
+deterministic version of our model with no yearly heterogeneity in year
+1 juvenile survival and recruitment probability \(\sigma_{J_1}\), and
+calculated the predicted long-run growth rate \(\lambda\) of a
+population for different values of \(\sigma_{A_R}\) and
+\(\sigma_{J_1}\). We then fixed \(\sigma_{J_1} = 0.09\) and calculated
+the predicted growth rate of our 12 populations.
 
 Second, we performed a local elasticity analysis on \(\lambda\) with
-respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
-determine how small changes in these parameters could influence the
-long-run deterministic growth rate of populations
-(Figure~\ref{fig-viability-supp}).
+respect to parameters \(\sigma_{J_1}\), \(\sigma_{J_2}\),
+\(\sigma_{A_R}\), and \(F\) to determine how small changes in these
+parameters could influence the long-run deterministic growth rate of
+populations (Figure~\ref{fig-viability-supp}).
 
@@
 environmental stochasticity, where environmental stochasticity was
-represented by among-year variability in \(\omega\). We used this model
-to simulate a one-time introduction of 40 translocated adult frogs. We
-ran this simulation 1000 times for each population and computed the
-probability of a population becoming extinct after 50 years given the
-observed parameter values and environmental stochasticity in \(\omega\).
-We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
-and drew values of \(\omega\) each year from a beta distribution with a
-dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
-\(\phi = 2\) the beta distribution is uniform between 0 and 1).
+represented by among-year variability in \(\sigma_{J_1}\). We used this
+model to simulate a one-time introduction of 40 translocated adult
+frogs. We ran this simulation 1000 times for each population and
+computed the probability of a population becoming extinct after 50 years
+given the observed parameter values and environmental stochasticity in
+\(\sigma_{J_1}\). We varied the mean recruitment probability
+\(\sigma_{J_1}\) from 0 and 0.25 and drew values of \(\sigma_{J_1}\)
+each year from a beta distribution with a dispersion parameter of
+\(\phi = 2\) (when \(\sigma_{J_1} = 0.5\) and \(\phi = 2\) the beta
+distribution is uniform between 0 and 1). Using different values of
+\(\phi\) does not qualitatively change the existence of distinct
+extinction dynamics between populations with \(\sigma_{A_R} < 0.5\) and
+those with \(\sigma_{A_R} > 0.5\). However, increasing yearly
+variability in \(\sigma_{J_1}\) increases extinction risk for all
+populations. For example, if we set \(\phi = 0.001\), such that in a
+given year essentially either all year 1 juveniles survive or all of
+them die, populations with \(\sigma_{A_R} > 0.5\) need to have
+\(\sigma_{J_1}\) greater than 0.2 to have a 50-year extinction
+probability of less 50\%. Because we do not pit tag juveniles, we do not
+have CMR estimates for \(\sigma_{J_1}\) or \(\phi\). However, based on
+qualitative and semi-quantitative field observations over 25 years, a
+value of \(\sigma_{J_1} = 0.25\) in the presence of Bd is probably a
+reasonable estimate for many populations. Thus, we expect our model
+predictions to be conservative with regards to population recovery.
 
@@
 years, repeating the simulation 50,000 times. For each run and each
-year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
-equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
-Approximate Bayesian Computing and rejection sampling
+year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
+1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
+Using Approximate Bayesian Computing and rejection sampling
 \citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
 1000 trajectories) that minimized the sum of squared errors between the
-observed and predicted data. The \(\omega\) values associated with these
-``best'' trajectories represented an approximate posterior distribution
-\citep{beaumont2010}. Using these best fit trajectories, we assessed
-whether our model could qualitatively describe the patterns of recovery
-in the observed data for population 70550.
+observed and predicted data. The yearly \(\sigma_{J_1}\) values
+associated with these ``best'' trajectories represented an approximate
+posterior distribution \citep{beaumont2010}. Using these best fit
+trajectories, we assessed whether our model could qualitatively describe
+the patterns of recovery in the observed data for population 70550.
+
+\hypertarget{frog-evolution-in-response-to-bd-2}{%
+\section{Frog evolution in response to
+Bd}\label{frog-evolution-in-response-to-bd-2}}
+
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+\hypertarget{study-design}{%
+\subparagraph{Study design}\label{study-design}}
+
+We compared frog genomes sampled in naive versus recovering populations.
+Comparing populations with different infection histories allowed larger
+sample sizes and replication across the landscape. The alternative
+approach of comparing samples from the same populations before and after
+Bd exposure isn't feasible in this system because Bd arrived in most MYL
+frog populations decades ago and population persistence/recovery is rare
+and unpredictable. As a result, samples from recovering populations
+collected from before and after Bd exposure are not available and are
+unlikely to be available in the future.
+
+\hypertarget{go-analysis}{%
+\subparagraph{GO analysis}\label{go-analysis}}
+
+In the Results section of the main paper, we describe the stringent set
+of outlier SNPs (identified using a Bonferroni-corrected p-value of
+0.01). The liberal set, identified using a Bonferroni-corrected p-value
+of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
+genes across 16 contigs. Using this liberal set of outlier variants, we
+performed a statistical overrepresentation test (Fisher's exact test)
+using Blast2GO \citep{conesa2005}. First, we retrieved the BLAST hits
+and mapped GO terms for each gene in our targeted transcriptome. Next,
+we compared the 30 unique outlier genes from the liberal set of 38
+outlier variants to the complete set of genes in our target
+transcriptome to determine if any GO biological functions, molecular
+functions, or cellular processes were overrepresented. We then repeated
+this process for the set of 33 genes located in the
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+
+\hypertarget{genetic-diversity}{%
+\subparagraph{Genetic diversity}\label{genetic-diversity}}
+
+To characterize general patterns of genetic diversity between naive and
+persistent populations, we calculated individual heterozygosity using
+VCFtools \citep{danecek2011}, and compared average heterozygosity using
+a Wilcoxon rank-sum test. Additionally, we used VCFtools to calculate
+per-population \(\pi\) in 100kb sliding windows along the genome, and in
+each of the 10 outlier windows identified in the splined window
+analysis. We compared per-population average \(\pi\) values using
+Wilcoxon rank-sum tests.
+
+\hypertarget{results-1}{%
+\subsection{Results}\label{results-1}}
+
+\hypertarget{go-analysis-1}{%
+\subparagraph{GO analysis}\label{go-analysis-1}}
+
+We found no significantly over-represented GO terms in the liberal set
+of outlier regions, or in the set of 33 genes located in the overlapping
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+Therefore\ldots..
+
+\hypertarget{genetic-diversity-1}{%
+\subparagraph{Genetic diversity}\label{genetic-diversity-1}}
+
+Average (i) individual-level heterozygosity and (ii) population-level
+nucleotide diversity (\(\pi\)) were similar between naive and recovering
+populations (Figure~\ref{fig-violinplot-heterozy};
+Figure~\ref{fig-boxplot-pi}). The analysis of \(\pi\) across the 10
+outlier windows indicates that within each of the windows there is
+considerable variation in \(\pi\) between populations and between
+population clusters (Figure~\ref{fig-barchart-pi-by-windowsite}).
+
+\newpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.60\textwidth]{figures/map_translocation_points.png}
+
+}
+
+\caption{\label{fig-yosemap}Map showing the locations of translocated
+and donor MYL frog populations in Yosemite National Park (park boundary
+indicated by gray polygon). Symbol shapes indicate the donor population
+used for each translocation site, and 5-digit numbers identify each
+donor and translocation site. To obscure the exact locations of
+populations, random noise was added to all point coordinates. Inset map
+shows the location of Yosemite within California. In both maps,
+elevation is indicated by the colored hillshade layer (dark green =
+lowest elevation, white = highest elevation).}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\textwidth]{figures/bdload_beforeafter.png}
+
+}
+
+\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
+the 12 recipient sites, Bd loads for the period immediately prior to
+translocation versus during the 1-year period after translocation. Bd
+loads are expressed as the number of ITS1 copies per skin swab, as
+estimated by qPCR of the Bd ITS1 region. Box plots show medians, first
+and third quartiles, largest and smallest values within 1.5x
+interquartile range, and values outside the 1.5x interquartile range.
+Loads indicative of severe disease are \textgreater{} 5.8 ITS copies (on
+a log\textsubscript{10} scale). Samples sizes are provided immediately
+above the x-axis.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m1d.png}
+
+}
+
+\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
+important predictor of post-translocation frog survival. Depicted
+distributions are the estimated posterior density curves and shaded 95\%
+uncertainty intervals for the intercept and all predictor variables from
+the best model in the between-site meta-analysis. In the Bayesian
+framework in which the model was developed, variables are considered
+important predictors if the associated uncertainty interval does not
+overlap zero (indicated by the dashed red line). Predictor variables
+shown on the y-axis are defined as follows: length = frog size, snow\_t
+= winter severity in the year of translocation (measured on April 1),
+snow\_t1 = winter severity in the year following translocation (measured
+on April 1), day = day of year on which a translocation was conducted,
+bd\_load = Bd load, elevation = site elevation, order\_first =
+within-site translocation order, sex\_male = frog sex, donor\_70567 and
+donor\_72996 = donor population.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.60\textwidth]{figures/compare_surv_probs.jpg}
+
+}
+
+\caption{\label{fig-compare_surv_probs}Comparison of average yearly
+adult survival probabilities for adults translocated to each of 3 sites
+versus adults that were naturally recruited at each site (as a result of
+reproduction by translocated frogs). In contrast to
+Fig. 3, these are not survival
+probabilities from the first year following translocation, but instead
+represent averaged survival probabilities across multiple years and
+cohorts. Points are median estimates and error bars give the 95\%
+uncertainty intervals around the estimates, accounting for yearly
+variation in survival probabilities. All estimates were derived using
+the mrmr package, and the methods for calculation are described in
+\textbf{Supporting Information - Population viability modeling --
+Estimating model parameters}.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.60\textwidth]{figures/pop_viability_figures_for_supp.jpg}
+
+}
+
+\caption{\label{fig-viability-supp}Sensitivity analysis of the
+stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
+in four parameters: \(\sigma_{A_R}\) (yearly survival probability of
+naturally recruited adults), \emph{F (}number of eggs produced by a
+female frog in a year that successfully hatch), \(\sigma_{J_1}\) (yearly
+probability of survival of year 1 juveniles that also affects
+recruitment), and \(\sigma_{J_2}\) (yearly probability of survival and
+recruitment of year 2 juveniles). Elasticity is calculated at the
+default parameter values for each population and \(\sigma_{J_1}\) =
+0.09.}
+
+\end{figure}\clearpage
 
@@
 
+\newpage
+
 \begin{figure}
@@
 \caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
-order in \emph{Xenopus tropicalis} and \emph{Rana mucosa} for the
+order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
 outlier region containing MHC Class I Classical and MHC Class III gene
@@
 using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
-accession GCA\_000004195.4) and \emph{Rana mucosa} Contig19. Asterices
+accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
 indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
@@
 
-{\centering \includegraphics[width=0.60\textwidth]{figures/map_translocation_points.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/het_by_status_violin_plot.png}
 
@@
 
-\caption{\label{fig-yosemap}Map showing the locations of translocated
-and donor MYL frog populations in Yosemite National Park (park boundary
-indicated by black polygon). Symbol shapes indicate the donor population
-used for each translocation site. To obscure the exact locations of
-populations, random noise was added to all point coordinates. Inset map
-shows the location of Yosemite within California. In both maps,
-elevation is indicated by the colored hillshade layer (dark green =
-lowest elevation, white = highest elevation).}
+\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
+heterozygosity for the Bd-naive and recovering populations included in
+the frog genomics study. Individual data points are shown on the plot
+with their corresponding site number (from
+Fig. 1). A Wilcoxon signed-rank test showed
+no significant difference between the two groups.}
 
@@
 
+\newpage
+
 \begin{figure}
 
-{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m2b.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_site.png}
 
@@
 
-\caption{\label{fig-transsurvival-postdens}For the best model describing
-the average 1-year survival of frogs in the first translocation to a
-site as a predictor of survival of frogs in all subsequent
-translocations to that site, estimated posterior density curves and
-shaded 95\% uncertainty intervals for the intercept and single predictor
-variable. In the Bayesian framework in which the model was developed, a
-variable is considered an important predictor if the associated
-uncertainty interval does not overlap zero (indicated by the dashed red
-line).}
+\caption{\label{fig-boxplot-pi}For each study population in the frog
+evolution study (from Fig. 1A), boxplots
+showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
+along the genome. The y-axis has been log\textsubscript{10}-transformed
+for display purposes. Sites are color-coded as ``Bd-naive'' or
+``recovering'', and vertical dashed lines show the genetic breaks
+between frog populations as described in
+Figure~\ref{fig-selectionresults} A). Within the genetic clusters,
+mean \(\pi\) is significantly different between sites (p \textless0.001
+Wilcoxon rank-sum test).}
 
@@
 
+\newpage
+
 \begin{figure}
 
-{\centering \includegraphics[width=0.7\textwidth]{figures/bdload_beforeafter.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/perwindow_persite_pi.png}
 
@@
 
-\caption{\label{fig-bdload-beforeafter}Evidence that frogs translocated
-from recovering populations can maintain the benefits of
-resistance/tolerance in non-natal habitats. For frogs translocated to
-each of the 12 recipient sites, Bd loads for the period immediately
-prior to translocation versus during the 1-year period after
-translocation. Bd loads are expressed as the number of ITS1 copies per
-skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show
-medians, first and third quartiles, largest and smallest values within
-1.5x interquartile range, and values outside the 1.5x interquartile
-range. Loads indicative of severe disease are \textgreater{} 5.8 ITS
-copies (on a log\textsubscript{10} scale).}
-
-\end{figure}\clearpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m1d.png}
-
-}
-
-\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
-important predictor of post-translocation frog survival. For the best
-model identifying predictors of 1-year frog survival following
-translocation, estimated posterior density curves and shaded 95\%
-uncertainty intervals for the intercept and all predictor variables. In
-the Bayesian framework in which the model was developed, variables are
-considered important predictors if the associated uncertainty interval
-does not overlap zero (indicated by the dashed red line). Predictor
-variables shown on the y-axis are defined as follows: length = frog
-size, snow\_t = winter severity in the year of translocation (measured
-on April 1), snow\_t1 = winter severity in the year following
-translocation (measured on April 1), day = day of year on which a
-translocation was conducted, bd\_load = Bd load, elevation = site
-elevation, order\_first = within-site translocation order, sex\_male =
-frog sex, donor\_70567 and donor\_72996 = donor population.}
-
-\end{figure}\clearpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.60\textwidth]{figures/compare_surv_probs.jpg}
-
-}
-
-\caption{\label{fig-compare_surv_probs}Evidence that frog
-resistance/tolerance in non-natal habitats is maintained across
-generations. Comparison of the average yearly adult survival
-probabilities for adults translocated to each of 3 sites versus adults
-that were naturally recruited at each site (as a result of reproduction
-by translocated frogs). In contrast to
-Fig. 3, these are not survival
-probabilities from the first year following translocation, but instead
-represent averaged survival probabilities across multiple years and
-cohorts. Points are median estimates and error bars give the 95\%
-uncertainty intervals around the estimates, accounting for yearly
-variation in survival probabilities. All estimates were derived using
-the mrmr package.}
-
-\end{figure}\clearpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.60\textwidth]{figures/pop_viability_figures_for_supp.jpg}
-
-}
-
-\caption{\label{fig-viability-supp}Sensitivity analysis of the
-stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
-in 3 parameters: \(\sigma_{A_R}\) (yearly survival probability of
-naturally recruited adults), \(\omega\) (yearly probability of
-successful recruitment from juveniles to adults), and \emph{F (}number
-of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each
-population and \(\omega\) = 0.3.}
+\caption{\label{fig-barchart-pi-by-windowsite}For each study population
+in the frog evolution study, average \(\pi\) values by population
+calculated for each of the 10 outlier windows identified in the splined
+window analysis (A-J). Sites are color-coded as ``Bd-naive'' or
+``recovering'', and vertical dashed lines show the genetic breaks
+between frog populations as described in
+Figure~\ref{fig-selectionresults} A. Gene annotations for the outlier
+regions are as follows: (A) C6, C7, (B) DDX10, ZBTB24, (C) SULTR-like,
+TRAF3IP2, VGLL3, EXOC1, (D) FLOT1, TUBB2B, TUBB, MDC1, CCHCR1,
+TCF19\emph{, TRIM29, HSP70, LSM2, VARS,} (E) GCC2, CFAP251, PEG10, (F)
+ERO1A, GVINP1, (G) ERO1A, GVINP1, (H) PPP1R12A, TSPAN4, PAWR, MFRP, MAX,
+PPP6R3, (I) C6H5ORF22, PKS6, BSPRY, MPV17, and (J) CAD, ATRAID, GPN1.
+Note that windows F and G are in the same gene region. * by gene names
+indicate that an outlier SNP is also present in this gene . (K) Average
+\(\pi\) values of naive and recovering groups for each of the outlier
+windows shown in A-J. }
 
@@
 
-\hfill\break
-
-\hypertarget{tbl-survival-earlylate}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0654}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2430}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2804}}@{}}
-\caption{\label{tbl-survival-earlylate}Association between the
-proportion of populations translocated early versus late in the study
-period (\textless{} 2013 or $\geq$ 2013, respectively) and
-probability of survival (\textless{} 0.5 or $\geq$ 0.5). Based on
-the viability analysis, survival probabilities \textless{} 0.5 and
-$\geq$ 0.5 produced 50-year extinction probabilities of 1 and
-\textless{} 0.5, respectively.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Survival probability
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Number of translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Total translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Proportion of translocations
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Survival probability
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Number of translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Total translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Proportion of translocations
-\end{minipage} \\
-\midrule()
-\endhead
-early & \textless{} 0.5 & 4 & 5 & 0.800 \\
-early & $\geq$ 0.5 & 1 & 5 & 0.200 \\
-late & \textless{} 0.5 & 2 & 7 & 0.286 \\
-late & $\geq$ 0.5 & 5 & 7 & 0.714 \\
-\bottomrule()
-\end{longtable}
-
-\newpage
-
 \hypertarget{tbl-param_values}{}
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2639}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}@{}}
 \caption{\label{tbl-param_values}Description and values of parameters
@@
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles & 0.25
-& Estimated from field data, observations, natural history knowledge \\
+\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles &
+Varies yearly & Varies. Bounds estimated from field data, observations,
+natural history knowledge \\
 \(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
@@
 \(\sigma_{A_R}\), Yearly survival probability of naturally recruited
-adults & Varies by population & Estimated from mark-recapture studies \\
+adults & Varies by population & Estimated from CMR studies \\
 \(\sigma_{A_T}\), Yearly survival probability of translocated adults &
-Varies by population & Estimated from mark-recapture studies \\
+Varies by population & Estimated from CMR studies \\
 \(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
@@
 observations of captive frogs \\
-\(\omega\), Probability of juvenile successfully recruiting to an adult
-& Varies yearly & Explored different values of this parameter \\
 \bottomrule()

### 2023-12-20 doc/manuscript/translocation_pnas.tex

@@
 \significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. One possibility is rapid
-evolution that restores positive growth to declining populations
-(``evolutionary rescue''). Individuals from these rescued populations
-could also be used to reestablish extinct populations. We evaluated both
-possibilities in a once-common frog species driven to the brink of
-extinction by a disease that has devastated global amphibian
-biodiversity and for which few options exist to reverse disease impacts.
-We found that frogs have evolved increased resistance to infection,
-facilitating population recovery. In addition, introduction of frogs
-from these populations allowed reestablishment despite the ongoing
-presence of disease. Such reintroductions could play an important role
-in biodiversity conservation in our rapidly changing world.}
+critical for the conservation of biodiversity. Emerging infectious
+diseases can have particularly devastating impacts, and few options
+typically exist to reverse these effects. We used large-scale
+reintroductions of disease-resistant individuals in an effort to recover
+a once-common frog species driven to the brink of extinction by a
+disease that has recently devastated global amphibian biodiversity.
+Introduction of resistant frogs allowed reestablishment of populations
+in the presence of disease. In addition, resistance appears to have a
+genetic basis and is consistent with evolution following exposure of
+frogs to Bd. The evolution of resistance and reintroductions of
+resistant individuals could play an important role in biodiversity
+conservation in our rapidly changing world.}
 
@@
 dynamics are given by equation \ref{eqn:matrix}.
-
-The parameters in this model are yearly survival probability
-\(\sigma_{\cdot}\) (the subscript ``\(\cdot\)'' indicates a particular
-state variable), probability that a female frog reproduces in a given
-year \(p_F\), number of eggs produced by a female frog in a year that
-successfully hatch \(F\), probability of a first-year tadpole remaining
-as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
-as a tadpole \(p_{L2}\), and probability of a first-year juvenile
-remaining as a juvenile \(p_{J1}\), and the probability of a juvenile
-successfully recruiting to an adult \(\omega\).
-
-In this model we ignore density-dependent recruitment because we were
-interested in the growth of the population from an initial
-reintroduction and whether this growth was sufficient to prevent
-extinction over 50 years following the introduction. We also did not
-directly consider the dynamics of Bd in this model. We made this
-decision because (i) translocated populations are infected with Bd at
-high prevalence \citep{joseph2018}, and (ii) host density does not seem
-to play a significant role in multi-year Bd infection dynamics in this
-system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
-instead assuming all host vital rates are in the presence of high Bd
-prevalence significantly simplifies the model without much loss of
-realism. Additional details are provided in the SI Appendix.
-
-\hypertarget{model-analysis}{%
-\subsection*{Model analysis}\label{model-analysis}}
-
-After parameterizing our model with CMR-estimated adult frog survival
-probabilities and other known vital rates
-(Table S2), we performed four analyses. First, we
-compute the long-run growth rate \(\lambda\) for each of our 12
-translocated populations to determine if the populations were
-deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to model parameters, to
-quantify how much changes in adult survival, for example, affected the
-long-run growth rate. Third, we included demographic and environmental
-stochasticity in our model and simulated the 50-year viability (i.e., 1
-- extinction probability) of populations given an introduction of 40
-adult individuals into an unoccupied habitat. Finally, we fit our model
-to our longest translocation trajectory to confirm that our model could
-reasonably reproduce the observed recovery trajectories of MYL frogs
-following reintroductions. Additional details are provided in the SI
-Appendix. Code to replicate the analyses can be found at
-\url{https://github.com/SNARL1/translocation}.
-
-}
-
-\showmatmethods{} % Display the Materials and Methods section
-
-\acknow{We thank the following for important contributions to this study: E.
-Hegeman and A. Lindauer--project and data management, field and
-laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
-and numerous summer technicians--field work; staff at Sequoia-Kings
-Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
-California Department of Fish and Wildlife, U.S. Fish and Wildlife
-Service, University of California-Santa Barbara Institutional Animal
-Care and Use Committee, and Sierra Nevada Aquatic Research
-Laboratory--research permits, logistical support, and/or field
-assistance. This project was supported by grants from the National Park
-Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
-Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
-DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
-
-\showacknow{} % Display the acknowledgments section
-
-% Bibliography
-\bibliography{translocation}
-
-\clearpage
 
@@
 
+The parameters in this model are yearly survival probability
+\(\sigma_{\cdot}\) (the subscript ``\(\cdot\)'' indicates a particular
+state variable), probability that a female frog reproduces in a given
+year \(p_F\), number of eggs produced by a female frog in a year that
+successfully hatch \(F\), probability of a first-year tadpole remaining
+as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
+as a tadpole \(p_{L2}\), and probability of a first-year juvenile
+remaining as a juvenile \(p_{J1}\). First-year juvenile survival and
+recruitment \(\sigma_{J1}\) is the parameter that we think is most
+influenced by environmental stochasticity.
+
+In this model we ignore density-dependent recruitment because we were
+interested in the growth of the population from an initial
+reintroduction and whether this growth was sufficient to prevent
+extinction over 50 years following the introduction. We also did not
+directly consider the dynamics of Bd in this model. We made this
+decision because (i) translocated populations are infected with Bd at
+high prevalence \citep{joseph2018}, and (ii) host density does not seem
+to play a significant role in multi-year Bd infection dynamics in this
+system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
+instead assuming all host vital rates are in the presence of high Bd
+prevalence significantly simplifies the model without much loss of
+realism. Additional details are provided in \textbf{Supporting
+Information - Population viability modeling}.
+
+\hypertarget{model-analysis}{%
+\subsection*{Model analysis}\label{model-analysis}}
+
+After parameterizing our model with CMR-estimated adult frog survival
+probabilities and other known vital rates
+(Table S2), we performed four analyses. First, we
+compute the long-run growth rate \(\lambda\) for each of our 12
+translocated populations to determine if the populations were
+deterministically predicted to grow or decline in the long-run. Second,
+we compute the elasticity of \(\lambda\) to four key model parameters to
+quantify how much changes in these parameters affected the long-run
+growth rate (Figure S5). This also helped us
+determine where in the model environmental variation in juvenile
+survival and recruitment would have the largest effects on population
+dynamics. Third, we included demographic stochasticity and environmental
+stochasticity in \(\sigma_{J_1}\) in our model and simulated the 50-year
+viability (i.e., 1 - extinction probability) of populations given an
+introduction of 40 adult individuals into an unoccupied habitat.
+Finally, we fit our model to our longest translocation trajectory to
+confirm that our model could reasonably reproduce the observed recovery
+trajectories of MYL frogs following reintroductions. Additional details
+are provided in \textbf{Supporting Information - Population viability
+modeling}. Code to replicate the analyses can be found at
+\url{https://github.com/SNARL1/translocation}.
+
+\hypertarget{frog-evolution-in-response-to-bd-1}{%
+\section*{Frog evolution in response to
+Bd}\label{frog-evolution-in-response-to-bd-1}}
+
+\hypertarget{sampling-and-sequencing}{%
+\subsection*{Sampling and sequencing}\label{sampling-and-sequencing}}
+
+We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
+\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
+populations, and 29 from 5 recovering populations. These populations are
+located in the southern Sierra Nevada, from northern Yosemite National
+Park to northern Sequoia National Park
+(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
+frogs per population. To minimize potential confounding effects caused
+by variation in frog genotypes across latitude \citep{byrne2023}, we
+selected sampling sites such that both population types were represented
+across similar latitudinal ranges. DNA was extracted following Qiagen
+DNEasy manufacturer's protocols. We sequenced the samples using an exome
+capture assay as described in \citep{byrne2023}. Briefly, genomic
+libraries were prepared and captured using a custom Nimblegen capture
+pool. Capture baits were designed based on the coding regions of the
+\emph{R. muscosa} transcriptome (GenBank accession GKCT00000000).
+Captured libraries were then pooled and sequenced on a NovaSeq 6000
+150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at
+UC Berkeley. Raw sequencing reads are available from NCBI SRA
+(PRJNA870451).
+
+\hypertarget{data-pre-processing-and-cleaning}{%
+\subsection*{Data pre-processing and
+cleaning}\label{data-pre-processing-and-cleaning}}
+
+Raw reads were filtered for adapters and contaminants using fastp
+\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
+SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
+bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
+using Picard. Variants were then called following GATK best practices
+(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
+sample using HaplotypeCaller and combined using CombineGVCFs. Next,
+genotypes were jointly called using GenotypeGVCFs. Variants were then
+hard filtered using gatk VariantFiltration using the following
+parameters to remove low-quality sites: QD \textless{} 2.0, FS
+\textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
+MQRankSum \textgreater{} 3.0, MQRankSum \textless{} -3.0, ReadPosRankSum
+\textgreater{} 3.0, ReadPosRankSum \textless{} -3.0. This initial filter
+resulted in 1,595,206 variant sites across 53 individuals.
+
+We then further filtered our dataset at the individual and variant
+level. First, we trimmed our variants to only include those with minor
+allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
+depth of 5, a minimum genotype quality of 20, and a maximum missing
+proportion of 0.5. This filter resulted in 427,038 sites, of which
+353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
+with an average depth across filtered sites \textless{} 7x (n = 3). Our
+final dataset included 50 samples, 23 from naive and 27 from recovering
+populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+
+\hypertarget{data-analysis}{%
+\subsection*{Data analysis}\label{data-analysis}}
+
+To visualize the genomic relationships of our populations, we conducted
+a PCA using the glPCA function in the adegenet R package
+\citep{jombart2008}. To detect regions of the genome that differed
+between naive and recovering populations, i.e., regions putatively under
+selection, we used two approaches: (1) a multivariate linear mixed model
+to evaluate individual variants (SNPs and INDELs), and (2) a splined
+window analysis to evaluate larger genomic regions. For the variant
+analysis, we first used a stringent data filter to include only variants
+with \textless{} 5\% missing data (missing for no more than 2
+individuals), and then calculated the likelihood ratio statistic for the
+resulting set of 148,307 high quality variants across 127 contigs using
+GEMMA \citep{zhou2014}. GEMMA calculates and incorporates a relatedness
+matrix for input samples, allowing us to account for relatedness and
+population structure when calculating likelihood ratio statistics. We
+identified variants showing different allele frequencies between naive
+versus recovering populations (``outliers'') using a
+Bonferroni-corrected significance level of 0.01. We visualized the
+results using a Manhattan plot and qqplot. We developed a more liberal
+set of outlier variants using a Bonferroni-corrected significance level
+of 0.05 and used this set solely for the gene ontology (GO) analysis
+(see below and \textbf{Supporting Information}: Dataset S1, S2).
+
+In the analysis of individual variants, for each outlier variant we
+determined whether the variant was synonymous (protein sequence the same
+for each variant) or non-synonymous (protein sequence differs between
+variants), and where in the gene it was located. To do this, we first
+extracted the reference genome sequence surrounding the variant using
+the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
+re-annotated each sequence using BLAST to get the predicted gene
+location based on the closest annotated reference \citep{altschul1997}.
+We then translated each variant to amino acids and aligned this
+translation to that of the gene annotation to ensure proper frame of
+reference using Geneious \citep{kearse2012}. After ensuring proper
+translation, we characterized variants as within or outside the coding
+sequence of the gene and as either synonymous or non-synonymous.
+
+In the splined window analysis, we identified outlier regions using
+\emph{F\textsubscript{ST}} and differences in nucleotide diversity
+(\(\pi_{diff}\)) between naive and recovering populations. First, we
+calculated per-site \emph{F\textsubscript{ST}} between the naive and
+recovering individuals for all bi-allelic SNPs in the 30 largest contigs
+(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
+calculated per-site nucleotide diversity \(\pi\) separately for
+individuals from the naive and recovering populations using VCFtools,
+then calculated \(\pi_{diff}\) for each population
+(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
+values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
+size-sorted chromosome number and adjusted the SNP position based on the
+relative position in the genome (for more efficient data processing and
+to better contextualize the strength of the outlier signals in different
+regions of the genome). We then used the GenWin R package
+\citep{beissinger2015} to conduct a splined discrete window analysis for
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
+where non-overlapping window boundaries should occur by identifying
+inflection points in the spline fitted to \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) values along the genome, therefore balancing false
+positive and false negative results that occur using other window-based
+methods \citep{beissinger2015}. This method also calculates a
+W-statistic allowing for outlier identification. We identified outliers
+as those with a W-statistic greater than 4 standard deviations above the
+mean for \emph{F\textsubscript{ST}} or above/below the mean for
+\(\pi_{diff}\). These standard deviations represent strict criteria to
+select only the top \textasciitilde{} 0.3\% of windows. Shared outliers
+were then identified as those that were outliers in both analyses,
+meaning that they showed (i) high differentiation between naive and
+recovering populations, and (ii) differential patterns of nucleotide
+diversity in the same region. Finally, we extracted gene transcripts
+mapped within each region and retrieved annotation for that region using
+BLAST (\textbf{Supporting Information}: Datasets S3, S4).
+
+Code used for genomic analyses and to create figures is available from
+the following GitHub repository:
+\url{https://github.com/allie128/mylf-selection}.
+
+}
+
+\showmatmethods{} % Display the Materials and Methods section
+
+\acknow{We thank the following for important contributions to this study: E.
+Hegeman and A. Lindauer--project and data management, field and
+laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
+and numerous summer technicians--field work; staff at Sequoia-Kings
+Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
+California Department of Fish and Wildlife, U.S. Fish and Wildlife
+Service, University of California-Santa Barbara Institutional Animal
+Care and Use Committee, and Sierra Nevada Aquatic Research
+Laboratory--research permits, logistical support, and/or field
+assistance. This project was supported by grants from the National Park
+Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
+Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
+
+\showacknow{} % Display the acknowledgments section
+
+% Bibliography
+\bibliography{translocation}
+
 \clearpage
+\begin{figure*}
+
+{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
+
+}
+
+\caption{\label{fig-recovery-model}For \emph{R. sierrae}, a conceptual
+model depicting the Bd-caused decline and subsequent natural recovery
+(black text), facilitated recovery via reintroductions, and the linkages
+between these two pathways. Rectangles and hexagons represent outcomes
+and processes, respectively. Blue text indicates components that are
+included in the current study. The timeline shows the general sequence
+of the components, with the dotted portion indicating a projection into
+the future.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
+
+}
+
+\caption{\label{fig-translocation-survival}Median 1-year survival for
+each cohort of translocated frogs at the 12 recipient sites, as
+estimated for each site from the mrmr CMR model. Error bars show the
+95\% uncertainty intervals. Sites are arranged along the x-axis using
+the average of the median 1-year survival per translocation at each
+site. Dot colors indicate the donor population from which frogs in each
+translocated cohort were collected. When multiple translocations were
+conducted to a site, points and error bars are slightly offset to avoid
+overlap.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
+
+}
+
+\caption{\label{fig-cond-effects}Results show conditional effects of the
+important predictors of 1-year frog survival (survival is expressed as a
+probability), as estimated from the model used in the among-site
+meta-analysis: (A) winter severity in the year following translocation,
+(B) elevation of recipient site, (C) donor population, and (D) sex. In
+(A) and (B), blue lines are medians and gray ribbons are 95\%
+uncertainty intervals. In (C) and (D), box plots show medians, first and
+third quartiles, largest and smallest values within 1.5x interquartile
+range, and values outside the 1.5x interquartile range.}
+
+\end{figure*}
+
+\newpage
+
+\begin{figure*}
+
+{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+
+}
+
+\caption{\label{fig-viability}Results from population viability
+analysis: (A) Predicted long-run growth rate \(\lambda\) for different
+values of yearly adult survival probability \(\sigma_{A_R}\) and year 1
+juvenile survival and recruitment probability \(\sigma_{J_1}\), given
+the parameterized, deterministic model. Colored points show the
+predicted \(\lambda\) values for the twelve translocated populations
+when year 1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
+(indicated by the dashed line). The red line shows where
+\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
+other points. (B) Predicted 50-year extinction probabilities of the 12
+translocated populations, given demographic stochasticity, environmental
+variability in \(\sigma_{J_1}\), and different mean values of
+\(\sigma_{J_1}\). There are 6 lines at extinction probability = 1, 5 of
+which (70414--70619) are hidden beneath the line for 70505 when
+\(\sigma_{J_1} < 0.10\). (C) 100 simulated trajectories (black lines)
+from the population viability model that most closely matched the
+observed abundance trajectory of adult amphibians at site 70550 (light
+orange).}
+
+\end{figure*}
+
+\newpage
+
 \begin{figure*}

### 2023-12-20 doc/manuscript/translocation.qmd
Add edits including from Bree

@@
 ---
-bibliography: translocation.bib
-#  main: translocation.bib
-#  supporting: supporting_info.bib
-#filters: 
-#  - multibib
-#validate-yaml: false
-csl: pnas.csl
-# format: pdf
+#format: pdf
 format:
@@
     wrap: sentence
+bibliography: translocation.bib
+csl: pnas.csl
 ---
@@
 Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
-Third, we investigated whether MYL frogs show evidence of an evolutionary response to Bd exposure, and whether these genomic changes are associated with resistance.
-To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model) and determined whether any of these regions contain genes associated with resistance.
+Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether these genomic changes are associated with resistance.
+To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model).
 
@@
 
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions putatively under selection).
+We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions under selection).
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
@@
 This is suggestive of parallel selection at this locus across multiple populations.
-The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
+The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5–9; @fig-allelefrequencies, @fig-selectionresults A SI).
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that suggest parallel evolutionary changes following Bd exposure.
@@
 
-No gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
-In addition, although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd** for details).
+Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd** for details).
+Thus, despite localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
+In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information-Frog evolution in response to Bd** for details).
 
 In summary, our genomic results indicate that the exomes of frogs from naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
-The regions putatively under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
+The regions under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
 
@@
 
-<!--# In the Discussion, I retained the general order of paragraphs from the original submission (evolution first, followed by translocations).
-This aligns with the actual trajectory that we are proposing in Figure 1 (Bd epizootic - selection in response to Bd - frog recovery.
-Does this order work?
--->
-
 Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
-<!--# Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
-Although previously proposed [@mendelson2019; @scheele2021], this strategy has rarely been used in the wild [see @joseph2018 for a small-scale example].
-Given the demonstrated resistance of adult frogs against Bd infection, and the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether the resulting adaptive alleles are associated with increased frog resistance.--> Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that the recovering populations are likely to persist over a 50-year period, (iii) that there are substantial genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic regions putatively under selection contain genes related to disease resistance.
-Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of a highly virulent pathogen, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
+Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
+Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that the recovering populations are likely to persist over a 50-year period, (iii) that there are substantial genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic regions under selection contain genes related to disease resistance.
+Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
 
+In the following discussion, we follow the sequence of frog recovery described in @fig-recovery-model to structure our key points.
 Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
@@
 Results from our genomic analyses suggest that natural selection for adaptive alleles is at least partially responsible for the increased resistance of frogs in recovering populations.
-We identified specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with selection following Bd exposure.
-Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
+We identified multiple specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with selection following Bd exposure.
+These analysis are based on samples collected from virtually all of the MYL frog populations remaining in a naive state, as well as adjacent recovering populations.
+This produced genetic clusters that contained at least one naive and one recovering population, allowing us to detect selection without the confounding effects of population structure.
+Finally, we did not find a reduction in overall genetic variation in the recovering populations, suggesting that despite localized selection in the genome, these populations retain adequate genetic diversity for long-term persistence.
+
+Importantly, some regions that we identified as under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
 For example, the MHC plays an important role in immunity.
@@
 The complement system plays an important role in innate immunity [@rieraromo2016], and our results could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-Finally, based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
+Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
 This gene is associated with inflammation response and in *Xenopus* is highly expressed in wounds [@fukazawa2009].
-In addition, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
+Finally, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
 In the genome of another frog species, this gene is located adjacent to a type I interferon gene (Np-IFNi2) [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
 Collectively, the genes associated with these genomic differences may confer at least some degree of resistance against Bd infection, an attribute that may be critically important to population reestablishment and recovery in the presence of Bd.
-<!--# Added this transition sentence. -->
 
@@
 In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
-Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an example of evolutionary rescue will likely always be uncertain.
+Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an airtight example of evolutionary rescue will likely always be uncertain (given that can never have a perfect understanding of the past).
 Regardless, we provide an example from the wild that suggests that evolution can produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
@@
 To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome that differed between naive and recovering populations, i.e., regions putatively under selection, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+To detect regions of the genome that differed between naive and recovering populations, i.e., regions under selection, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
 For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
@@
 
-![For *R.
-sierrae*, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
+![For MYL frogs, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
 Rectangles and hexagons represent outcomes and processes, respectively.

### 2023-12-21 doc/manuscript/translocation.qmd
Minor edits to Intro and Results

@@
 We were particularly interested in whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
-This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
+This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual-level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
 The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
@@
 Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 <!--# Why not use 0.09 like in A panel?
--->(@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 (@fig-viability B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year 1 juvenile survival between 0 and 0.25.
+In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
 To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
@@
 
-To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of genome) between populations with contrasting histories of Bd exposure.
+To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of a genome) between populations with contrasting histories of Bd exposure.
 Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
@@
 
-![Results show conditional effects of the important predictors of 1-year frog survival (survival is expressed as a probability), as estimated from the model used in the among-site meta-analysis: (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
+![Results from the among-site meta-analysis showing conditional effects of the important predictors of 1-year frog survival (expressed as a probability): (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.

### 2023-12-21 doc/manuscript/translocation_bioarxiv.tex
Minor edits to Intro and Results

@@
 
-\dropcap{H}uman activities are increasingly impacting global biodiversity
-\citep{ceballos2015}, with important implications for ecosystem
-resilience and human welfare \citep{naeem2009}. One consequence of human
-alteration of the biosphere is an increase in emerging infectious
-diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
-threat to wildlife populations \citep{daszak2000}, and have caused
-dramatic declines and extinctions in a wide range of taxa, including
-echinoderms, mammals, birds, and amphibians
-\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
-are experiencing particularly devastating impacts of disease due to the
-recent emergence and global spread of the highly virulent amphibian
-chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
-\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
-have experienced Bd-caused declines, and numerous susceptible taxa are
-extinct in the wild \citep{scheele2019}. These impacts to global
+\dropcap{H}uman activities are increasingly impacting global biodiversity (1),
+with important implications for ecosystem resilience and human welfare
+(2). One consequence of human alteration of the biosphere is an increase
+in emerging infectious diseases (3, 4). Such diseases pose a severe
+threat to wildlife populations (5), and have caused dramatic declines
+and extinctions in a wide range of taxa, including echinoderms, mammals,
+birds, and amphibians (6--9). Amphibians are experiencing particularly
+devastating impacts of disease due to the recent emergence and global
+spread of the highly virulent amphibian chytrid fungus,
+\emph{Batrachochytrium dendrobatidis} (Bd) (8, 10). By one estimate,
+hundreds of species have experienced Bd-caused declines, and numerous
+susceptible taxa are extinct in the wild (8). These impacts to global
 biodiversity may be unprecedented, and highlight the importance of
 understanding mechanisms of species persistence in the presence of
-emerging diseases \citep{russell2020}.
+emerging diseases (11).
 
 Evidence of natural recovery in the many Bd-impacted amphibian
-populations is surprisingly rare \citep[for notable exceptions,
-see][]{scheele2017, voyles2018, knapp2016}, suggesting that
-disease-caused declines will be difficult to reverse. This apparent low
-resilience to disease effects may be due to the limited ability of many
-amphibians to develop Bd resistance and/or tolerance, which in turn,
-could also lessen the effectiveness of potential Bd mitigation
-strategies. Following pathogen arrival in a host population, resistance
-(ability to limit pathogen burden) and tolerance (ability to limit the
-harm caused by a particular burden) are key mechanisms to reduce disease
-impacts \citep{raberg2009} and facilitate population persistence and
-recovery \citep{brannelly2021}. Host immunity and evolution both play
+populations is surprisingly rare (for notable exceptions, see 12, 13,
+14), suggesting that disease-caused declines will be difficult to
+reverse. This apparent low resilience to disease effects may be due to
+the limited ability of many amphibians to develop Bd resistance and/or
+tolerance, which in turn, could also lessen the effectiveness of
+potential Bd mitigation strategies. Following pathogen arrival in a host
+population, resistance (ability to limit pathogen burden) and tolerance
+(ability to limit the harm caused by a particular burden) are key
+mechanisms to reduce disease impacts (15) and facilitate population
+persistence and recovery (16). Host immunity and evolution both play
 important roles in the development of resistance and tolerance, and
 utilizing these factors would seem a promising approach to developing
-effective strategies to mitigate disease impacts in the wild
-\citep{garner2016, woodhams2011}. However, several aspects of the
-amphibian-Bd system present difficult obstacles, including (i) the
-general inability of amphibians to mount an effective immune response
-against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
-(ii) the apparent rarity of evolution of more resistant/tolerant
-genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
-suggest that reintroduction of amphibians into sites to reestablish
-populations extirpated by Bd will often result not in population
-recovery, but instead in the rapid reinfection and mortality of the
-introduced animals and/or their progeny
-\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
-future of many amphibian species threatened by Bd appears bleak.
+effective strategies to mitigate disease impacts in the wild (17, 18).
+However, several aspects of the amphibian-Bd system present difficult
+obstacles, including (i) the general inability of amphibians to mount an
+effective immune response against Bd infection (19--21), and (ii) the
+apparent rarity of evolution of more resistant/tolerant genotypes (but
+see 22, 23). These factors suggest that reintroduction of amphibians
+into sites to reestablish populations extirpated by Bd will often result
+not in population recovery, but instead in the rapid reinfection and
+mortality of the introduced animals and/or their progeny (24--27). If
+true, the future of many amphibian species threatened by Bd appears
+bleak.
 
 The mountain yellow-legged (MYL) frog, composed of the sister species
-\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
-emblematic of the global declines of amphibians caused by Bd
-\citep{scheele2019}. Once the most common amphibian in the high
-elevation portion of California's Sierra Nevada mountains
-\citep[USA,][]{grinnell1924}, during the past century this frog has
-disappeared from more than 90\% of its historical range
-\citep{vredenburg2007}. Due to the severity of its decline and the
-increasing probability of extinction, both species are now listed as
-``endangered'' under the U.S. Endangered Species Act. In the Sierra
-Nevada, this decline was initiated by the introduction of non-native
-trout into the extensive historically-fishless region
-\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
-of Bd in the mid-1900s and its subsequent spread \citep{vredenburg2019}
-caused additional large-scale population extirpations
-\citep{vredenburg2010, rachowicz2006}. These Bd-caused declines are
-fundamentally different from the fish-caused declines because fish
-eradication is feasible \citep{knapp1998} and results in the rapid
-recovery of frog populations \citep{knapp2007, vredenburg2004}. In
-contrast, Bd appears to persist in habitats even in the absence of
-amphibian hosts \citep{walker2007}, and therefore represents a long-term
-alteration of invaded ecosystems that amphibians will need to overcome
-to reestablish populations.
+\emph{Rana muscosa} and \emph{Rana sierrae} (28), is emblematic of the
+global declines of amphibians caused by Bd (8). Once the most common
+amphibian in the high elevation portion of California's Sierra Nevada
+mountains (USA, 29), during the past century this frog has disappeared
+from more than 90\% of its historical range (28). Due to the severity of
+its decline and the increasing probability of extinction, both species
+are now listed as ``endangered'' under the U.S. Endangered Species Act.
+In the Sierra Nevada, this decline was initiated by the introduction of
+non-native trout into the extensive historically-fishless region (30,
+31) starting in the late 1800s. The arrival of Bd in the mid-1900s and
+its subsequent spread (32) caused additional large-scale population
+extirpations (33, 34). These Bd-caused declines are fundamentally
+different from the fish-caused declines because fish eradication is
+feasible (35) and results in the rapid recovery of frog populations (36,
+37). In contrast, Bd appears to persist in habitats even in the absence
+of amphibian hosts (38), and therefore represents a long-term alteration
+of invaded ecosystems that amphibians will need to overcome to
+reestablish populations.
 
 Despite the catastrophic impact of Bd on MYL frogs, wherein most
-Bd-naive populations are extirpated following Bd arrival
-\citep{vredenburg2010}, some populations have persisted after epizootics
-\citep[during which Bd infection intensity on frogs is very
-high,][]{briggs2010} and are now recovering
-(Figure~\ref{fig-recovery-model}) \citep{knapp2016}. Frogs in these
-recovering populations show reduced susceptibility to Bd infection
-\citep{knapp2016}, with infection intensity (``load'') on adults
-consistently in the low-to-moderate range
-\citep{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
-is evident even under controlled laboratory conditions
-\citep{knapp2016}, indicative of host resistance against Bd infection
-(and not simply an effect of factors external to individual frogs, e.g.,
-environmental conditions). In addition to frogs from recovering
-populations having higher resistance to Bd infection than those from
-naive populations, they could also have higher tolerance, but no data
-are currently available to evaluate this possibility. Therefore, we
-focus on resistance throughout this paper. The observed resistance of
-MYL frogs could be the result of several non-mutually exclusive
-mechanisms, including natural selection for more resistant genotypes
-\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
-and/or inherent between-population differences that pre-date Bd
-exposure. The possible evolution of MYL frog resistance and subsequent
-population recovery is consistent with that expected under
-``evolutionary rescue'', whereby rapid evolutionary change increases the
-frequency of adaptive alleles and restores positive population growth
-\citep{carlson2014, searle2020}. This intriguing possibility also
-suggests an opportunity to expand recovery beyond the spatial scale
-possible under natural recovery by utilizing resistant frogs from
-recovering populations in reintroductions to vacant habitats
-(Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.
+Bd-naive populations are extirpated following Bd arrival (33), some
+populations have persisted after epizootics (during which Bd infection
+intensity on frogs is very high, 39) and are now recovering
+(Figure~\ref{fig-recovery-model}) (14). Frogs in these recovering
+populations show reduced susceptibility to Bd infection (14), with
+infection intensity (``load'') on adults consistently in the
+low-to-moderate range (39--41). This reduced susceptibility is evident
+even under controlled laboratory conditions (14), indicative of host
+resistance against Bd infection (and not simply an effect of factors
+external to individual frogs, e.g., environmental conditions). In
+addition to frogs from recovering populations having higher resistance
+to Bd infection than those from naive populations, they could also have
+higher tolerance, but no data are currently available to evaluate this
+possibility. Therefore, we focus on resistance throughout this paper.
+The observed resistance of MYL frogs could be the result of several
+non-mutually exclusive mechanisms, including natural selection for more
+resistant genotypes (22, 23), acquired immunity (21), and/or inherent
+between-population differences that pre-date Bd exposure. The possible
+evolution of MYL frog resistance and subsequent population recovery is
+consistent with that expected under ``evolutionary rescue'', whereby
+rapid evolutionary change increases the frequency of adaptive alleles
+and restores positive population growth (42, 43). This intriguing
+possibility also suggests an opportunity to expand recovery beyond the
+spatial scale possible under natural recovery by utilizing resistant
+frogs from recovering populations in reintroductions to vacant habitats
+(Figure~\ref{fig-recovery-model}) (41, 44).
 
@@
 of persistence for the reintroduced populations over a multi-decadal
-period (Figure~\ref{fig-recovery-model}). Third, we investigated whether
-MYL frogs show evidence of an evolutionary response to Bd exposure, and
+period (Figure~\ref{fig-recovery-model}). Third, given the importance of
+resistance for frog survival, population establishment, and long-term
+viability (this study), we also conducted a genomic study to determine
+whether MYL frogs show an evolutionary response to Bd exposure, and
 whether these genomic changes are associated with resistance. To
@@
 genomic regions with consistent differences between frogs from naive
-versus recovering populations (Figure~\ref{fig-recovery-model}) and
-determined whether any of these regions contain genes associated with
-resistance.
+versus recovering populations (Figure~\ref{fig-recovery-model}).
 
@@
 chytridiomycosis (i.e., the disease caused by Bd) and associated frog
-mortality (Figure S2)
-\citep{joseph2018, vredenburg2010}. Although it is possible that the
-observed relatively small changes in load are a consequence of
-individuals with high Bd loads dying and therefore being unavailable for
-sampling during the post-translocation period, the fact that there was
-little difference in pre- versus post-translocation Bd loads even in
-those populations that had very high frog survival (70556, 74976;
-Figure~\ref{fig-translocation-survival},
+mortality (Figure S2) (33, 41). Although it
+is possible that the observed relatively small changes in load are a
+consequence of individuals with high Bd loads dying and therefore being
+unavailable for sampling during the post-translocation period, the fact
+that there was little difference in pre- versus post-translocation Bd
+loads even in those populations that had very high frog survival (70556,
+74976; Figure~\ref{fig-translocation-survival},
 Figure S2) suggests a true lack of
@@
 self-sustaining population. Given that it can take years or even decades
-to determine the self-sustainability of a reintroduced population
-\citep[for an example in MYL frogs, see][]{joseph2018}, the use of
-proxies is essential for providing shorter-term insights into
-reintroduction success and the factors driving it. Results from our CMR
-surveys allowed us to accurately estimate frog survival, including over
-the entire CMR time series for each site and during only the 1-year
-period immediately following translocation. These estimates were made
-using site-specific models analyzed using the mrmr package. We use these
-estimates to describe general patterns of frog survival in all
-translocated cohorts, and in an among-site meta-analysis of frog
-survival to identify important predictors of 1-year frog survival (e.g.,
-Bd load).
+to determine the self-sustainability of a reintroduced population (for
+an example in MYL frogs, see 41), the use of proxies is essential for
+providing shorter-term insights into reintroduction success and the
+factors driving it. Results from our CMR surveys allowed us to
+accurately estimate frog survival, including over the entire CMR time
+series for each site and during only the 1-year period immediately
+following translocation. These estimates were made using site-specific
+models analyzed using the mrmr package. We use these estimates to
+describe general patterns of frog survival in all translocated cohorts,
+and in an among-site meta-analysis of frog survival to identify
+important predictors of 1-year frog survival (e.g., Bd load).
 
@@
 a Bayesian framework and included a diversity of site, cohort, and
-individual level characteristics as predictors and 1-year frog survival
+individual- level characteristics as predictors and 1-year frog survival
 (Figure~\ref{fig-translocation-survival}) as the response variable. The
@@
 surveys may be necessary to confirm that populations are in fact
-self-sustaining \citep{joseph2018}. To extend our inferences of
-population establishment beyond those possible from the site-specific
-CMR data, we developed a population viability model. Specifically, to
-test whether the observed yearly adult survival probabilities in
-translocated populations (provided in legend of
-Figure~\ref{fig-viability}B) were sufficient for long-term viability,
-we built a stage-structured matrix model that captured known frog
-demography and included demographic and environmental stochasticity. We
-parameterized the model using CMR data from translocated populations and
-known life history values in this system (Table S2
-SI).
+self-sustaining (41). To extend our inferences of population
+establishment beyond those possible from the site-specific CMR data, we
+developed a population viability model. Specifically, to test whether
+the observed yearly adult survival probabilities in translocated
+populations (provided in legend of Figure~\ref{fig-viability}B) were
+sufficient for long-term viability, we built a stage-structured matrix
+model that captured known frog demography and included demographic and
+environmental stochasticity. We parameterized the model using CMR data
+from translocated populations and known life history values in this
+system (Table S2 SI).
 
@@
 probabilities (\(\sigma_{A_R} > 0.5\)) had 50-year extinction
-probabilities of less than 0.5 when the average year 1 juvenile survival
+probabilities of less than 0.5 when the average year-1 juvenile survival
 was greater than 0.10 (Figure~\ref{fig-viability}B). This indicates
@@
 \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was always
-predicted to be \textgreater{} 50\% regardless of the value of mean year
-1 juvenile survival between 0 and 0.25. To test the validity of our
+predicted to be \textgreater{} 50\% regardless of the value of mean
+year-1 juvenile survival between 0 and 0.25. To test the validity of our
 model predictions, we demonstrated that our stochastic model could
@@
 
-To determine whether MYL frog populations show an evolutionary response
-to Bd, we compared frog exomes (i.e., coding region of genome) between
-populations with contrasting histories of Bd exposure. Specifically, we
-compared frog genomes sampled in 4 populations that have not yet
-experienced a Bd-caused epizootic (``naive'') \citep{zhou2015} versus in
-5 populations that experienced a Bd epizootic during the past several
-decades and have since recovered to varying degrees (``recovering'';
-Figure~\ref{fig-allelefrequencies}) \citep{knapp2016, vredenburg2010}.
+To determine whether MYL frog populations show genomic patterns
+consistent with an evolutionary response to Bd, we compared frog exomes
+(i.e., coding region of a genome) between populations with contrasting
+histories of Bd exposure. Specifically, we compared frog genomes sampled
+in 4 populations that have not yet experienced a Bd-caused epizootic
+(``naive'') (45) versus in 5 populations that experienced a Bd epizootic
+during the past several decades and have since recovered to varying
+degrees (``recovering''; Figure~\ref{fig-allelefrequencies}) (14, 33).
 Bd-exposure histories of the 9 study populations are based on 10-20
-years of VES and Bd surveillance using skin swabbing
-\citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive populations are
-characterized by large numbers of adults (i.e, typically 1000s), Bd
-prevalence that is generally 0\% except during occasional Bd failed
-invasions \citep[during which Bd loads remain very low,][]{wilber2022},
-and no history of Bd epizootics since we first surveyed these
-populations in the late 1990s and early 2000s \citep{zhou2015}. In
-contrast, recovering populations exist in an enzootic state
-\citep{briggs2010}, characterized by smaller numbers of adults
-(generally \textless{} 100), high Bd prevalence \citep[often
-\textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
-that are typically well below the level expected to cause mortality
-\citep{vredenburg2010}. Naive and recovering populations can be
-identified unambiguously using these differences in Bd prevalence and
-load. Finally, there is no potential for frog dispersal between the 9
-study populations due to intervening distances and topography, as well
-as the presence of introduced (predatory) fish and fish-induced habitat
+years of VES and Bd surveillance using skin swabbing (e.g., 14, 45, 46).
+Naive populations are characterized by large numbers of adults (i.e,
+typically 1000s), Bd prevalence that is generally 0\% except during
+occasional Bd failed invasions (during which Bd loads remain very low,
+46), and no history of Bd epizootics since we first surveyed these
+populations in the late 1990s and early 2000s (45). In contrast,
+recovering populations exist in an enzootic state (39), characterized by
+smaller numbers of adults (generally \textless{} 100), high Bd
+prevalence (often \textgreater{} 80\%, 40), and, in adults, moderate Bd
+loads that are typically well below the level expected to cause
+mortality (33). Naive and recovering populations can be identified
+unambiguously using these differences in Bd prevalence and load.
+Finally, there is no potential for frog dispersal between the 9 study
+populations due to intervening distances and topography, as well as the
+presence of introduced (predatory) fish and fish-induced habitat
 fragmentation. (See \textbf{Supporting Information - Frog evolution in
@@
 two complementary approaches to identify regions of the genome that
-differed between naive and recovering populations (i.e., regions
-putatively under selection). First, we used a multivariate linear mixed
-model to evaluate associations between population type (i.e., naive
-versus recovering) and individual variants, including single nucleotide
-polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting
-for population structure. Second, we used a splined window analysis to
-identify larger genomic regions showing differences between population
-types in \emph{F\textsubscript{ST}} and nucleotide diversity
-(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).
+differed between naive and recovering populations (i.e., regions under
+selection). First, we used a multivariate linear mixed model to evaluate
+associations between population type (i.e., naive versus recovering) and
+individual variants, including single nucleotide polymorphisms (SNPs)
+and insertions/deletions (INDELS), while accounting for population
+structure. Second, we used a splined window analysis to identify larger
+genomic regions showing differences between population types in
+\emph{F\textsubscript{ST}} and nucleotide diversity (\(\pi_{diff}\) =
+\(\pi_{naive}\) - \(\pi_{recovering}\)).
 
@@
 the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
-the strong signature of isolation by distance that is characteristic of
-MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
-each cluster contained at least one population from both the naive and
-recovering groups, allowing us to distinguish allelic associations of
-individuals sampled in the 2 population types versus allelic
-associations resulting from population structure and genetic drift.
+the strong signature of isolation-by-distance that is characteristic of
+MYL frogs (47--49). Importantly, each cluster contained at least one
+population from both the naive and recovering groups, allowing us to
+distinguish allelic associations of individuals sampled in the 2
+population types versus allelic associations resulting from population
+structure and genetic drift.
 
@@
 Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
-extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
-Therefore, this region under selection is part of the extended MHC Class
-I and III complex and shows synteny with other amphibian genomes.
-
-No gene ontology (GO) biological functions, molecular functions, or
-cellular processes were over-represented in either the outlier variants
-or the 33 genes located in the overlapping \emph{F\textsubscript{ST}}
-and \(\pi_{diff}\) splined windows. In addition, although the joint
-processes of Bd-caused population declines and selection in response to
-Bd exposure could affect genetic diversity of recovering populations, we
-found no consistent differences in individual-level heterozygosity or
-population-level \(\pi\) between naive and recovering populations (see
-\textbf{Supporting Information-Frog evolution in response to Bd-Methods}
-for details).
-
-In summary, these results indicate that the exomes of frogs from our
+extended MHC Class III region (HSP70, LSM2, VARS) (50). Therefore, this
+region under selection is part of the extended MHC Class I and III
+complex and shows synteny with other amphibian genomes.
+
+Although the joint processes of Bd-caused population declines and
+selection in response to Bd exposure could affect genetic diversity of
+recovering populations, we found no consistent differences in
+individual-level heterozygosity or population-level \(\pi\) between
+naive and recovering populations (see \textbf{Supporting
+Information-Frog evolution in response to Bd} for details). Thus,
+despite localized selection in particular regions of the genome, we did
+not find evidence for reduced genetic diversity across the genome in
+recovering populations. In addition, no gene ontology (GO) biological
+functions, molecular functions, or cellular processes were
+over-represented in either the outlier variants or the 33 genes located
+in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined
+windows (see \textbf{Supporting Information-Frog evolution in response
+to Bd} for details).
+
+In summary, our genomic results indicate that the exomes of frogs from
 naive and recovering study populations show substantial differences,
 consistent with parallel evolutionary changes following Bd exposure. The
-regions putatively under selection contain several
-immunologically-relevant genes and gene families that are directly
-linked to disease resistance in other taxa.
+regions under selection contain several immunologically-relevant genes
+and gene families that are directly linked to disease resistance in
+other taxa.
 
@@
 Disease-induced population declines are decimating global biodiversity
-\citep{daszak2000}, but broadly-applicable strategies to recover
-affected species are generally lacking \citep[e.g.,][]{garner2016}. Our
-results indicate (i) the capacity of reintroduced populations to become
-established and eventually recover despite ongoing disease, (ii) that
-the recovering populations are likely to persist over a 50-year period,
-(iii) that there are substantial genomic differences between naive and
-recovering MYL frog populations, consistent with evolutionary change in
-frogs following Bd exposure, and (iv) that some of the genomic regions
-putatively under selection contain genes related to disease resistance.
-Collectively, these results (Figure~\ref{fig-recovery-model}) provide a
-rare example of amphibian recovery in the presence of a highly virulent
-pathogen, and have important implications for the conservation and
-recovery of amphibians and other taxa worldwide that are endangered by
-escalating impacts from emerging infectious diseases. In light of the
-generally low success rate of amphibian reintroduction efforts
-\citep{dodd2005}, our success in reestablishing MYL frog populations via
+(5), but broadly-applicable strategies to recover affected species are
+generally lacking (e.g., 17). Here, we tested the possibility that
+populations of resistant individuals from naturally recovering
+populations can be used to reestablish extirpated populations of the
+endangered MYL frog in the presence of a highly virulent fungal pathogen
+(Bd). Our results indicate (i) the capacity of reintroduced populations
+to become established and eventually recover despite ongoing disease,
+(ii) that the recovering populations are likely to persist over a
+50-year period, (iii) that there are substantial genomic differences
+between naive and recovering MYL frog populations, consistent with
+evolutionary change in frogs following Bd exposure, and (iv) that some
+of the genomic regions under selection contain genes related to disease
+resistance. Collectively, these results
+(Figure~\ref{fig-recovery-model}) provide a rare example of amphibian
+recovery in the presence of Bd, and have important implications for the
+conservation and recovery of amphibians and other taxa worldwide that
+are endangered by escalating impacts from emerging infectious diseases.
+In light of the generally low success rate of amphibian reintroduction
+efforts (51), our success in reestablishing MYL frog populations via
 translocation of resistant individuals is striking, and even more so
@@
 
-Previous field studies in MYL frogs show that frog-Bd dynamics and frog
-survival in the presence of Bd are fundamentally different between naive
-and recovering populations. Following the arrival and establishment of
-Bd in previously-naive populations, adult frogs develop high Bd loads
-that lead to mass die-offs \citep{vredenburg2010}. In contrast, in
+In the following discussion, we follow the sequence of frog recovery
+described in Figure~\ref{fig-recovery-model} to structure our key
+points. Previous field studies in MYL frogs show that frog-Bd dynamics
+and frog survival in the presence of Bd are fundamentally different
+between naive and recovering populations. Following the arrival and
+establishment of Bd in previously-naive populations, adult frogs develop
+high Bd loads that lead to mass die-offs (33). In contrast, in
 recovering populations adult frogs typically have low-to-moderate and
-relatively constant Bd loads and mass die-offs are not observed
-\citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
-naive and recovering populations are also observed in controlled
-laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
-indicate that frogs from recovering populations exhibit resistance
-against Bd infection. This resistance could in theory be due to several
-factors, including natural selection for more resistant genotypes,
-acquired immunity, and/or inherent between-population differences that
-pre-date Bd exposure, but until now evidence to evaluate the role of
-evolution was lacking.
+relatively constant Bd loads and mass die-offs are not observed (39, 40,
+see also Figure S2). The differences in Bd
+load of frogs from naive and recovering populations are also observed in
+controlled laboratory studies (see Figure 4 in 14), and clearly indicate
+that frogs from recovering populations exhibit resistance against Bd
+infection. This resistance could in theory be due to several factors,
+including natural selection for more resistant genotypes, acquired
+immunity, and/or inherent between-population differences that pre-date
+Bd exposure, but until now evidence to evaluate the role of evolution
+was lacking.
 
@@
 adaptive alleles is at least partially responsible for the increased
-resistance of frogs in recovering populations. We identified specific
-alleles and genomic regions showing signatures of selection between
-adjacent naive and recovering MYL frog populations, consistent with
-selection following Bd exposure. Importantly, some regions under
-selection are associated with cellular and immunological mechanisms
-known to contribute to disease resistance, including in amphibians
-\citep{zamudio2020}. For example, the MHC plays an important role in
-immunity. In our study, we identified a region that shows evidence of
-selection in recovering populations and contains eight genes associated
-with either the MHC Class I or Class III regions. These results
-corroborate numerous previous studies linking MHC genes to amphibian
-resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
-Similarly, the region with the strongest indication of directional
-selection (as measured by \(\pi_{diff}\)) contains the
+resistance of frogs in recovering populations. We identified multiple
+specific alleles and genomic regions showing signatures of selection
+between adjacent naive and recovering MYL frog populations, consistent
+with selection following Bd exposure. These analysis are based on
+samples collected from virtually all of the MYL frog populations
+remaining in a naive state, as well as adjacent recovering populations.
+This produced genetic clusters that contained at least one naive and one
+recovering population, allowing us to detect selection without the
+confounding effects of population structure. Finally, we did not find a
+reduction in overall genetic variation in the recovering populations,
+suggesting that despite localized selection in the genome, these
+populations retain adequate genetic diversity for long-term persistence.
+
+Importantly, some regions that we identified as under selection are
+associated with cellular and immunological mechanisms known to
+contribute to disease resistance, including in amphibians (52). For
+example, the MHC plays an important role in immunity. In our study, we
+identified a region that shows evidence of selection in recovering
+populations and contains eight genes associated with either the MHC
+Class I or Class III regions. These results corroborate numerous
+previous studies linking MHC genes to amphibian resistance against Bd
+(e.g., 53, 54). Similarly, the region with the strongest indication of
+directional selection (as measured by \(\pi_{diff}\)) contains the
 interferon-related gene GVINP1. Several previous studies of amphibians
 have found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
-in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
-linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. We also identified a region,
-characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
-that contained the complement genes C6 and C7. The complement system
-plays an important role in innate immunity \citep{rieraromo2016}, and
+(e.g., 23, 55) and in populations differing in Bd susceptibility (23).
+This gene is also strongly linked to disease in salmon, explaining a
+notable 20\% of the resistance phenotype (56, 57). We also identified a
+region, characterized by high \emph{F\textsubscript{ST}} and low
+\(\pi_{diff}\), that contained the complement genes C6 and C7. The
+complement system plays an important role in innate immunity (58), and
 our results could indicate that balancing selection is acting in this
 region of the genome to favor a diverse set of alleles, as is known for
-C6 in humans \citep{soejima2005}. Finally, based on the analysis of
-individual outlier variants, the RIN3 gene showed a consistent pattern
-of allele frequency differences across all nine of the frog populations
-sampled in this study, indicating consistent selection in populations
-distributed across a wide geographic area. This gene is associated with
-inflammation response and in \emph{Xenopus} is highly expressed in
-wounds \citep{fukazawa2009}. In addition, the outlier variant with the
-lowest p-value was the uncharacterized gene LOC108802036. In the genome
-of another frog species, this gene is located adjacent to a type I
-interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
-further suggests the importance of interferon-related genes in this
-system. Collectively, the genes associated with these genomic
-differences may confer at least some degree of resistance against Bd
-infection, an attribute that may be critically important to population
-reestablishment and recovery in the presence of Bd.
+C6 in humans (59). Based on the analysis of individual outlier variants,
+the RIN3 gene showed a consistent pattern of allele frequency
+differences across all nine of the frog populations sampled in this
+study, indicating consistent selection in populations distributed across
+a wide geographic area. This gene is associated with inflammation
+response and in \emph{Xenopus} is highly expressed in wounds (60).
+Finally, the outlier variant with the lowest p-value was the
+uncharacterized gene LOC108802036. In the genome of another frog
+species, this gene is located adjacent to a type I interferon gene
+(Np-IFNi2) (61), and together with GVINP1 further suggests the
+importance of interferon-related genes in this system. Collectively, the
+genes associated with these genomic differences may confer at least some
+degree of resistance against Bd infection, an attribute that may be
+critically important to population reestablishment and recovery in the
+presence of Bd.
 
@@
 given that reintroduced MYL frog populations may often take decades to
-achieve our ultimate goal of self-sustainability \citep{joseph2018}.
-Making well-supported projections about the long-term outcome of
-reintroduction efforts from shorter-term information is critically
-important to the process of adaptive management of species
-reintroduction programs \citep{seddon2007}, including the one we are
-carrying out for MYL frogs. Specifically, the combined results from our
-reintroduction study and viability modeling indicate that survival of
-frogs in the first year following translocation is an effective proxy of
-longer-term survival and population viability. In addition, given the
-repeatability of frog survival at a site, 1-year frog survival also
-serves as an effective proxy of site quality (i.e., the ability of a
-site to support high frog survival and a viable frog population over the
-long term). This proxy of site quality is important in the MYL frog
-system because accurately predicting the ability of a site to support a
-viable frog population \emph{a priori} remains difficult, even after
-conducting 24 translocations over 16 years.
+achieve our ultimate goal of self-sustainability (41). Making
+well-supported projections about the long-term outcome of reintroduction
+efforts from shorter-term information is critically important to the
+process of adaptive management of species reintroduction programs (62),
+including the one we are carrying out for MYL frogs. Specifically, the
+combined results from our reintroduction study and viability modeling
+indicate that survival of frogs in the first year following
+translocation is an effective proxy of longer-term survival and
+population viability. In addition, given the repeatability of frog
+survival at a site, 1-year frog survival also serves as an effective
+proxy of site quality (i.e., the ability of a site to support high frog
+survival and a viable frog population over the long term). This proxy of
+site quality is important in the MYL frog system because accurately
+predicting the ability of a site to support a viable frog population
+\emph{a priori} remains difficult, even after conducting 24
+translocations over 16 years.
 
@@
 evident. In an earlier study of 2 of our 12 translocated populations
-\citep{joseph2018}, Bd infection and load had detectable effects on the
-survival of adults and may have influenced population establishment
-(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
-are identified as ``70550'' and ``70505'' in the current study).
-Applying similar analyses to all 12 of our translocated populations
-would likely provide a broader perspective of the ongoing effect of Bd.
-In addition to these important but relatively subtle effects of Bd on
-adults, the impacts on younger life stages are more apparent. MYL frogs
-immediately following metamorphosis (``metamorphs'') are highly
-susceptible to Bd infection \citep{ellison2019} and as a result
-experience high mortality \citep{rachowicz2006}. This high
+(41), Bd infection and load had detectable effects on the survival of
+adults and may have influenced population establishment (sites referred
+to as ``Alpine'' and ``Subalpine'' in (41) are identified as ``70550''
+and ``70505'' in the current study). Applying similar analyses to all 12
+of our translocated populations would likely provide a broader
+perspective of the ongoing effect of Bd. In addition to these important
+but relatively subtle effects of Bd on adults, the impacts on younger
+life stages are more apparent. MYL frogs immediately following
+metamorphosis (``metamorphs'') are highly susceptible to Bd infection
+(63) and as a result experience high mortality (34). This high
 susceptibility of metamorphs is documented in numerous species of
 anurans, and may result from the poorly developed immune system
-characteristic of this life stage \citep{humphries2022}. In naturally
-recovering and translocated MYL frog populations, we suggest that the
-high mortality of metamorphs is an important limitation on subsequent
-recruitment of new adults. Therefore, although adult MYL frogs appear
-relatively resistant, Bd infection continues to have important limiting
-effects on recovering populations \citep[see also][]{hollanders2022}.
+characteristic of this life stage (64). In naturally recovering and
+translocated MYL frog populations, we suggest that the high mortality of
+metamorphs is an important limitation on subsequent recruitment of new
+adults. Therefore, although adult MYL frogs appear relatively resistant,
+Bd infection continues to have important limiting effects on recovering
+populations (see also 65).
 
@@
 hundreds of amphibian species, some of which are now extinct in the wild
-\citep{scheele2019}. This extraordinary impact on global amphibian
-biodiversity is compounded by the lack of any effective and broadly
-applicable strategies to reverse these impacts
-\citep{garner2016, knapp2022}. Importantly, in addition to the natural
-recovery documented for MYL frogs \citep{knapp2016}, other species are
+(8). This extraordinary impact on global amphibian biodiversity is
+compounded by the lack of any effective and broadly applicable
+strategies to reverse these impacts (17, 25). Importantly, in addition
+to the natural recovery documented for MYL frogs (14), other species are
 also showing evidence of post-epizootic recovery in the presence of Bd
-\citep{scheele2017, voyles2018} and suggest the possibility of also
-using animals from these recovering populations to reestablish
-extirpated populations. As with MYL frogs, the feasibility and long-term
-success of such efforts will depend on the availability of robust donor
-populations containing individuals that have the adaptive alleles
-necessary to allow frog survival and population growth in the presence
-of Bd. Despite the hopeful example of successful reestablishment of MYL
-frogs despite ongoing Bd infection, the challenge of recovering hundreds
-of Bd-impacted amphibian species globally is a daunting prospect.
-Although we now have a proven strategy to reestablish extirpated MYL
-frog populations, recovery across their large historical range will
-require substantial resources over many decades. The results of this
-study provide a hopeful starting point for that endeavor and other
-future efforts worldwide.
+(12, 13) and suggest the possibility of also using animals from these
+recovering populations to reestablish extirpated populations. As with
+MYL frogs, the feasibility and long-term success of such efforts will
+depend on the availability of robust donor populations containing
+individuals that have the adaptive alleles necessary to allow frog
+survival and population growth in the presence of Bd. Despite the
+hopeful example of successful reestablishment of MYL frogs despite
+ongoing Bd infection, the challenge of recovering hundreds of
+Bd-impacted amphibian species globally is a daunting prospect. Although
+we now have a proven strategy to reestablish extirpated MYL frog
+populations, recovery across their large historical range will require
+substantial resources over many decades. The results of this study
+provide a hopeful starting point for that endeavor and other future
+efforts worldwide.
 
@@
 documented disease resistance in MYL frogs and concurrent recovery of
-decimated populations provides an example of evolutionary rescue will
-likely always be uncertain. Regardless, we provide an example from the
-wild that suggests that evolution can produce individuals that harbor
-adaptive alleles and allow population recovery in a novel (i.e.,
+decimated populations provides an airtight example of evolutionary
+rescue will likely always be uncertain (given that can never have a
+perfect understanding of the past). Regardless, we provide an example
+from the wild that suggests that evolution can produce individuals that
+harbor adaptive alleles and allow population recovery in a novel (i.e.,
 Bd-positive) environment, and show conclusively that individuals from
@@
 from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
-collected using several years of VES and skin swab collections
-\citep{knapp2016}, and results from population genetic analyses
-\citep{poorten2017}. The populations that we selected contained hundreds
-to thousands of both \emph{R. sierrae} adults and tadpoles. These
-relatively high abundances were the result of recent increases following
-previous Bd-caused declines \citep{knapp2016}. As is typical for
-recovering MYL frog populations, Bd prevalence in the donor populations
-was high (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
-3.06--3.78 ITS copies) was two or more orders of magnitude below the
-level at which frog mortality is expected (log\textsubscript{10}(load)
-\(\approx\) 5.78 copies)\citep{vredenburg2010, joseph2018}. Recipient
+collected using several years of VES and skin swab collections (14), and
+results from population genetic analyses (48). The populations that we
+selected contained hundreds to thousands of both \emph{R. sierrae}
+adults and tadpoles. These relatively high abundances were the result of
+recent increases following previous Bd-caused declines (14). As is
+typical for recovering MYL frog populations, Bd prevalence in the donor
+populations was high (0.69--0.96) and Bd load (median
+log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or more
+orders of magnitude below the level at which frog mortality is expected
+(log\textsubscript{10}(load) \(\approx\) 5.78 copies)(33, 41). Recipient
 sites to which frogs were translocated were chosen based on previous
@@
 or characteristics that suggested high quality habitat for this species
-\citep{knapp2005}. At the beginning of this study, we had a relatively
-limited understanding of the factors that affect habitat quality. In
-subsequent years, we improved our site selection process by
-incorporating new information about important habitat features, in
-particular, overwinter habitats such as submerged boulders and
-overhanging banks. \emph{R. sierrae} were absent from all recipient
-sites prior to the first translocation.
+(66). At the beginning of this study, we had a relatively limited
+understanding of the factors that affect habitat quality. In subsequent
+years, we improved our site selection process by incorporating new
+information about important habitat features, in particular, overwinter
+habitats such as submerged boulders and overhanging banks. \emph{R.
+sierrae} were absent from all recipient sites prior to the first
+translocation.
 
@@
 stages have high detectability, and counts are highly repeatable and
-provide estimates of relative abundance \citep{knapp2000}.
+provide estimates of relative abundance (31).
 
@@
 Jolly-Seber CMR model with known additions to the population (i.e.,
-translocated cohorts), as implemented by the mrmr package
-\citep{joseph2019} and using R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
-Frog population recovery - CMR model structure} for details). Briefly,
+translocated cohorts), as implemented by the mrmr package (67) and using
+R Statistical Software (v4.4.4, 68) (see \textbf{Supporting Information
+- Frog population recovery - CMR model structure} for details). Briefly,
 the model tracks the states of \emph{M} individuals that comprise a
-superpopulation made up of real and pseudo-individuals \citep[see][ for
-details]{joseph2018}. The possible states of individuals include ``not
-recruited'', ``alive'', and ``dead''. The possible observations of
-individuals include ``detected'' and ``not detected''. We assume that
-individuals that are in the ``not recruited'' or ``dead'' states are
-never detected (i.e., there are no mistakes in the individual PIT tag
-records). We also assume that new recruits were the result of
-within-site reproduction and not immigration from adjacent populations.
-This assumption is justified by the fact that no \emph{R. sierrae}
-populations were present within several kilometers of the translocation
-sites. For all models, we used mrmr defaults for priors, number of
-chains (4), and warmup and post-warmup iterations (2000 for each). We
-evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm
-using trace plots and Gelman-Rubin statistics (Rhat).
+superpopulation made up of real and pseudo-individuals (see 41 for
+details). The possible states of individuals include ``not recruited'',
+``alive'', and ``dead''. The possible observations of individuals
+include ``detected'' and ``not detected''. We assume that individuals
+that are in the ``not recruited'' or ``dead'' states are never detected
+(i.e., there are no mistakes in the individual PIT tag records). We also
+assume that new recruits were the result of within-site reproduction and
+not immigration from adjacent populations. This assumption is justified
+by the fact that no \emph{R. sierrae} populations were present within
+several kilometers of the translocation sites. For all models, we used
+mrmr defaults for priors, number of chains (4), and warmup and
+post-warmup iterations (2000 for each). We evaluated convergence of the
+Markov chain Monte Carlo (MCMC) algorithm using trace plots and
+Gelman-Rubin statistics (Rhat).
 
@@
 To identify important predictors of frog survival following
-translocation, we used multilevel Bayesian models
-\citep{gelman2013, gabry2019}. Included predictor variables describe
-characteristics of sites, translocated cohorts, and individuals (Bd
-load, sex, frog size, site elevation, winter severity in the year of
-translocation, winter severity in the year following translocation,
-donor population, day of year on which a translocation was conducted,
-and translocation order). We used 1-year post-translocation survival
-estimates from CMR models as the response. Estimated survival was
-rounded to integer values to produce a binary outcome, and modeled with
-a Bernoulli distribution. Group-level (random) effects included
-site\_id, translocation\_id, or translocation\_id nested within
-site\_id. We performed all analyses with the rstanarm package
-\citep{rstanarm2022} and R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022}. For all models, we used default,
-weakly informative priors, four chains, and 5000 iterations each for
-warmup and post-warmup. We checked MCMC convergence using trace plots
-and Rhat, and evaluated model fit using leave-one-out cross-validation
-\citep{vehtari2016}, as implemented by the loo package
-\citep{vehtari2022}. (See \textbf{Supporting Information - Frog
-population recovery - Among-site survival modeling} for details.)
+translocation, we used multilevel Bayesian models (69, 70). Included
+predictor variables describe characteristics of sites, translocated
+cohorts, and individuals (Bd load, sex, frog size, site elevation,
+winter severity in the year of translocation, winter severity in the
+year following translocation, donor population, day of year on which a
+translocation was conducted, and translocation order). We used 1-year
+post-translocation survival estimates from CMR models as the response.
+Estimated survival was rounded to integer values to produce a binary
+outcome, and modeled with a Bernoulli distribution. Group-level (random)
+effects included site\_id, translocation\_id, or translocation\_id
+nested within site\_id. We performed all analyses with the rstanarm
+package (71) and R Statistical Software (v4.4.4, 68). For all models, we
+used default, weakly informative priors, four chains, and 5000
+iterations each for warmup and post-warmup. We checked MCMC convergence
+using trace plots and Rhat, and evaluated model fit using leave-one-out
+cross-validation (72), as implemented by the loo package (73). (See
+\textbf{Supporting Information - Frog population recovery - Among-site
+survival modeling} for details.)
 
@@
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see \textbf{Supporting Information-Frog population recovery -
-Laboratory methods} for details]{boyle2004}. To assess the magnitude of
-changes in Bd load on frogs following translocation, we compared Bd
-loads measured before versus after translocation. Before-translocation
-loads were quantified using skin swabs collected from all
-to-be-translocated frogs at the donor site on the day before or the day
-of the translocation. After-translocation Bd loads were based on all
-swabs collected from translocated frogs at the recipient site in the
-year of and the year following translocation. Individual frogs and their
-associated Bd loads were included in the dataset only if frogs were
-captured at the recipient site at least once during the 1-year period
-following translocation.
+(74, see \textbf{Supporting Information-Frog population recovery -
+Laboratory methods} for details). To assess the magnitude of changes in
+Bd load on frogs following translocation, we compared Bd loads measured
+before versus after translocation. Before-translocation loads were
+quantified using skin swabs collected from all to-be-translocated frogs
+at the donor site on the day before or the day of the translocation.
+After-translocation Bd loads were based on all swabs collected from
+translocated frogs at the recipient site in the year of and the year
+following translocation. Individual frogs and their associated Bd loads
+were included in the dataset only if frogs were captured at the
+recipient site at least once during the 1-year period following
+translocation.
 
@@
 decision because (i) translocated populations are infected with Bd at
-high prevalence \citep{joseph2018}, and (ii) host density does not seem
-to play a significant role in multi-year Bd infection dynamics in this
-system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
-instead assuming all host vital rates are in the presence of high Bd
-prevalence significantly simplifies the model without much loss of
-realism. Additional details are provided in \textbf{Supporting
-Information - Population viability modeling}.
+high prevalence (41), and (ii) host density does not seem to play a
+significant role in multi-year Bd infection dynamics in this system
+(46). Thus, ignoring Bd infection dynamics and instead assuming all host
+vital rates are in the presence of high Bd prevalence significantly
+simplifies the model without much loss of realism. Additional details
+are provided in \textbf{Supporting Information - Population viability
+modeling}.
 
@@
 
-We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
-\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
-populations, and 29 from 5 recovering populations. These populations are
-located in the southern Sierra Nevada, from northern Yosemite National
-Park to northern Sequoia National Park
-(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
-frogs per population. To minimize potential confounding effects caused
-by variation in frog genotypes across latitude \citep{byrne2023}, we
-selected sampling sites such that both population types were represented
-across similar latitudinal ranges. DNA was extracted following Qiagen
-DNEasy manufacturer's protocols. We sequenced the samples using an exome
-capture assay as described in \citep{byrne2023}. Briefly, genomic
-libraries were prepared and captured using a custom Nimblegen capture
-pool. Capture baits were designed based on the coding regions of the
-\emph{R. muscosa} transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000
-150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at
-UC Berkeley. Raw sequencing reads are available from NCBI SRA
-(PRJNA870451).
+We collected DNA samples via buccal swabbing (75) from 53 \emph{Rana
+muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive populations,
+and 29 from 5 recovering populations. These populations are located in
+the southern Sierra Nevada, from northern Yosemite National Park to
+northern Sequoia National Park (Figure~\ref{fig-allelefrequencies}).
+Samples were collected from 5-6 frogs per population. To minimize
+potential confounding effects caused by variation in frog genotypes
+across latitude (49), we selected sampling sites such that both
+population types were represented across similar latitudinal ranges. DNA
+was extracted following Qiagen DNEasy manufacturer's protocols. We
+sequenced the samples using an exome capture assay as described in (49).
+Briefly, genomic libraries were prepared and captured using a custom
+Nimblegen capture pool. Capture baits were designed based on the coding
+regions of the \emph{R. muscosa} transcriptome (GenBank accession
+GKCT00000000). Captured libraries were then pooled and sequenced on a
+NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics
+Sequencing Lab at UC Berkeley. Raw sequencing reads are available from
+NCBI SRA (PRJNA870451).
 
@@
 
-Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
-SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
-bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
-using Picard. Variants were then called following GATK best practices
-(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
-sample using HaplotypeCaller and combined using CombineGVCFs. Next,
-genotypes were jointly called using GenotypeGVCFs. Variants were then
-hard filtered using gatk VariantFiltration using the following
-parameters to remove low-quality sites: QD \textless{} 2.0, FS
+Raw reads were filtered for adapters and contaminants using fastp (76)
+and aligned to the \emph{Rana muscosa} genome (NCBI SRA: SRS6533475
+(77)), with repetitive elements masked using bwa (``mem'' mode) (78).
+Exact PCR duplicates were marked using Picard. Variants were then called
+following GATK best practices (v.4.2.0.0 (79)). Briefly, raw variants
+were called for each sample using HaplotypeCaller and combined using
+CombineGVCFs. Next, genotypes were jointly called using GenotypeGVCFs.
+Variants were then hard filtered using gatk VariantFiltration using the
+following parameters to remove low-quality sites: QD \textless{} 2.0, FS
 \textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
@@
 To visualize the genomic relationships of our populations, we conducted
-a PCA using the glPCA function in the adegenet R package
-\citep{jombart2008}. To detect regions of the genome that differed
-between naive and recovering populations, i.e., regions putatively under
-selection, we used two approaches: (1) a multivariate linear mixed model
-to evaluate individual variants (SNPs and INDELs), and (2) a splined
-window analysis to evaluate larger genomic regions. For the variant
-analysis, we first used a stringent data filter to include only variants
-with \textless{} 5\% missing data (missing for no more than 2
-individuals), and then calculated the likelihood ratio statistic for the
-resulting set of 148,307 high quality variants across 127 contigs using
-GEMMA \citep{zhou2014}. GEMMA calculates and incorporates a relatedness
-matrix for input samples, allowing us to account for relatedness and
-population structure when calculating likelihood ratio statistics. We
-identified variants showing different allele frequencies between naive
-versus recovering populations (``outliers'') using a
-Bonferroni-corrected significance level of 0.01. We visualized the
-results using a Manhattan plot and qqplot. We developed a more liberal
-set of outlier variants using a Bonferroni-corrected significance level
-of 0.05 and used this set solely for the gene ontology (GO) analysis
-(see below and \textbf{Supporting Information}: Dataset S1, S2).
+a PCA using the glPCA function in the adegenet R package (80). To detect
+regions of the genome that differed between naive and recovering
+populations, i.e., regions under selection, we used two approaches: (1)
+a multivariate linear mixed model to evaluate individual variants (SNPs
+and INDELs), and (2) a splined window analysis to evaluate larger
+genomic regions. For the variant analysis, we first used a stringent
+data filter to include only variants with \textless{} 5\% missing data
+(missing for no more than 2 individuals), and then calculated the
+likelihood ratio statistic for the resulting set of 148,307 high quality
+variants across 127 contigs using GEMMA (81). GEMMA calculates and
+incorporates a relatedness matrix for input samples, allowing us to
+account for relatedness and population structure when calculating
+likelihood ratio statistics. We identified variants showing different
+allele frequencies between naive versus recovering populations
+(``outliers'') using a Bonferroni-corrected significance level of 0.01.
+We visualized the results using a Manhattan plot and qqplot. We
+developed a more liberal set of outlier variants using a
+Bonferroni-corrected significance level of 0.05 and used this set solely
+for the gene ontology (GO) analysis (see below and \textbf{Supporting
+Information}: Dataset S1, S2).
 
@@
 extracted the reference genome sequence surrounding the variant using
-the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
-re-annotated each sequence using BLAST to get the predicted gene
-location based on the closest annotated reference \citep{altschul1997}.
-We then translated each variant to amino acids and aligned this
-translation to that of the gene annotation to ensure proper frame of
-reference using Geneious \citep{kearse2012}. After ensuring proper
-translation, we characterized variants as within or outside the coding
-sequence of the gene and as either synonymous or non-synonymous.
+the bedtools ``getfasta'' function (82). Next, we re-annotated each
+sequence using BLAST to get the predicted gene location based on the
+closest annotated reference (83). We then translated each variant to
+amino acids and aligned this translation to that of the gene annotation
+to ensure proper frame of reference using Geneious (84). After ensuring
+proper translation, we characterized variants as within or outside the
+coding sequence of the gene and as either synonymous or non-synonymous.
 
@@
 recovering individuals for all bi-allelic SNPs in the 30 largest contigs
-(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
-calculated per-site nucleotide diversity \(\pi\) separately for
-individuals from the naive and recovering populations using VCFtools,
-then calculated \(\pi_{diff}\) for each population
+(98\% of all SNPs) using VCFtools (85). Next, we calculated per-site
+nucleotide diversity \(\pi\) separately for individuals from the naive
+and recovering populations using VCFtools, then calculated
+\(\pi_{diff}\) for each population
 (\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
@@
 to better contextualize the strength of the outlier signals in different
-regions of the genome). We then used the GenWin R package
-\citep{beissinger2015} to conduct a splined discrete window analysis for
+regions of the genome). We then used the GenWin R package (86) to
+conduct a splined discrete window analysis for
 \emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
@@
 positive and false negative results that occur using other window-based
-methods \citep{beissinger2015}. This method also calculates a
-W-statistic allowing for outlier identification. We identified outliers
-as those with a W-statistic greater than 4 standard deviations above the
-mean for \emph{F\textsubscript{ST}} or above/below the mean for
-\(\pi_{diff}\). These standard deviations represent strict criteria to
-select only the top \textasciitilde{} 0.3\% of windows. Shared outliers
-were then identified as those that were outliers in both analyses,
-meaning that they showed (i) high differentiation between naive and
-recovering populations, and (ii) differential patterns of nucleotide
-diversity in the same region. Finally, we extracted gene transcripts
-mapped within each region and retrieved annotation for that region using
-BLAST (\textbf{Supporting Information}: Datasets S3, S4).
+methods (86). This method also calculates a W-statistic allowing for
+outlier identification. We identified outliers as those with a
+W-statistic greater than 4 standard deviations above the mean for
+\emph{F\textsubscript{ST}} or above/below the mean for \(\pi_{diff}\).
+These standard deviations represent strict criteria to select only the
+top \textasciitilde{} 0.3\% of windows. Shared outliers were then
+identified as those that were outliers in both analyses, meaning that
+they showed (i) high differentiation between naive and recovering
+populations, and (ii) differential patterns of nucleotide diversity in
+the same region. Finally, we extracted gene transcripts mapped within
+each region and retrieved annotation for that region using BLAST
+(\textbf{Supporting Information}: Datasets S3, S4).
 
@@
 
-\caption{\label{fig-recovery-model}For \emph{R. sierrae}, a conceptual
-model depicting the Bd-caused decline and subsequent natural recovery
-(black text), facilitated recovery via reintroductions, and the linkages
+\caption{\label{fig-recovery-model}For MYL frogs, a conceptual model
+depicting the Bd-caused decline and subsequent natural recovery (black
+text), facilitated recovery via reintroductions, and the linkages
 between these two pathways. Rectangles and hexagons represent outcomes
@@
 
-\caption{\label{fig-cond-effects}Results show conditional effects of the
-important predictors of 1-year frog survival (survival is expressed as a
-probability), as estimated from the model used in the among-site
-meta-analysis: (A) winter severity in the year following translocation,
-(B) elevation of recipient site, (C) donor population, and (D) sex. In
-(A) and (B), blue lines are medians and gray ribbons are 95\%
-uncertainty intervals. In (C) and (D), box plots show medians, first and
-third quartiles, largest and smallest values within 1.5x interquartile
-range, and values outside the 1.5x interquartile range.}
+\caption{\label{fig-cond-effects}Results from the among-site
+meta-analysis showing conditional effects of the important predictors of
+1-year frog survival (expressed as a probability): (A) winter severity
+in the year following translocation, (B) elevation of recipient site,
+(C) donor population, and (D) sex. In (A) and (B), blue lines are
+medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
+box plots show medians, first and third quartiles, largest and smallest
+values within 1.5x interquartile range, and values outside the 1.5x
+interquartile range.}
 

### 2023-12-21 doc/manuscript/translocation_bioarxiv_SI.tex
Minor edits to Intro and Results

@@
 Swab extracts were analyzed using standard Bd DNA extraction and qPCR
-methods \citep{boyle2004}, and extracts were analyzed singly instead of
-in triplicate \citep{kriger2006}. For analysis of swabs collected during
-2005--2014, we used standards developed from known concentrations of
-zoospores \citep{boyle2004}, and after 2014, we used standards based on
-single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
-between samples analyzed using both types of standards, Bd in the study
-area has an average of 60 ITS1 copies per zoospore. To express all qPCR
-results as the number of ITS1 copies, starting quantities obtained using
-the zoospore standard (measured as ``zoospore equivalents'') were
-multiplied by 60. In addition, all qPCR quantities (regardless of
-standard) were multiplied by 80 to account for the fact that DNA
-extracts from swabs were diluted 80-fold during extraction and PCR
-\citep{vredenburg2010}.
+methods (74), and extracts were analyzed singly instead of in triplicate
+(87). For analysis of swabs collected during 2005--2014, we used
+standards developed from known concentrations of zoospores (74), and
+after 2014, we used standards based on single ITS1 PCR amplicons (88).
+Based on paired comparisons between samples analyzed using both types of
+standards, Bd in the study area has an average of 60 ITS1 copies per
+zoospore. To express all qPCR results as the number of ITS1 copies,
+starting quantities obtained using the zoospore standard (measured as
+``zoospore equivalents'') were multiplied by 60. In addition, all qPCR
+quantities (regardless of standard) were multiplied by 80 to account for
+the fact that DNA extracts from swabs were diluted 80-fold during
+extraction and PCR (33).
 
@@
 We estimated survival and recruitment for each site using open
-population CMR models based on \citep{joseph2018}. For each individual
-\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
-the individual was not detected, and \(o_{i, j}=2\) if the individual
-was detected. Capture histories of \(M\) individuals are modeled, though
-only \(N_s\) individuals were captured.
-
-This parameter expanded data augmentation allows us to capture the
-possibility that undetected individuals may have recruited into the
-adult population \citep{royle2012}. Here, \(M\) was chosen to be three
-times the number of observed individuals (\(3N_s\)) to be considerably
-greater than our prior guess of \(N_s\).
+population CMR models based on (41). For each individual \(i=1, ..., M\)
+on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
+not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
+histories of \(M\) individuals are modeled, although only \(N_s\)
+individuals were captured. This parameter expanded data augmentation
+allows us to capture the possibility that undetected individuals may
+have recruited into the adult population (89). Here, \(M\) was chosen to
+be three times the number of observed individuals (\(3N_s\)) to be
+considerably greater than our prior guess of \(N_s\).
 
@@
 
-For introduced individuals which have deterministic recruitment (they
-recruit when introduced), the state transition matrix is given by:
+For introduced individuals, which have deterministic recruitment (i.e.,
+they recruit when introduced), the state transition matrix is given by:
 
@@
 forward algorithm, and we estimated the latent states using the
-forward-backward algorithm \citep{zucchini2009, joseph2018}.
+forward-backward algorithm (41, 90).
 
 All of the code to specify and fit the model in Stan is available in the
-open source mrmr package \citep{joseph2019}.
+open source mrmr package (67).
 
@@
 
+The objective of this analysis is to describe the influence of site,
+cohort, and individual level characteristics on post-translocation frog
+survival. By modeling survival estimates obtained from site-specific
+mrmr CMR analyses, we are in effect conducting an among-site
+meta-analysis. Although it would theoretically be possible to estimate
+survival covariate effects in a joint CMR model that integrates capture
+histories across all sites, this was impractical due to computational
+requirements of the CMR models (namely, run time and memory).
+
 We used Bayesian generalized linear mixed models to investigate
@@
 rounding the posterior median of 1-year post-introduction survival for
-each individual, and modeled the data using a Bernoulli distribution:
+each individual (from site-specific mrmr CMR models) and modeled the
+data using a Bernoulli distribution:
 
@@
 These models were fit using the \texttt{stan\_glmer()} function in the
-rstanarm package, with default priors described below
-\citep{rstanarm2022}. These priors are vague, but include data dependent
-scaling as follows to account for different input variable scales.
-However, as we standardized all predictor variables similarly to have
-equal variance (by centering and dividing by twice the sample standard
-deviation), the resulting priors are identical. Specifically, we have:
+rstanarm package, with default priors described below (71). These priors
+are vague, but include data-dependent scaling as follows to account for
+different input variable scales. However, because we standardized all
+predictor variables similarly to have equal variance (by centering and
+dividing by twice the sample standard deviation), the resulting priors
+are identical. Specifically, we have:
 
@@
 variance parameters -- the decomposition of variance prior with unit
-regularization, concentration, shape, and scale parameters
-\citep{lewandowski2009}.
+regularization, concentration, shape, and scale parameters (91).
 
@@
 with four parallel chains, each run for 10,000 iterations, discarding
-the first half of each chain as warm-up draws \citep{rstanarm2022}. We
-used Rhat statistics and trace plots to verify convergence. We
-considered models with different subsets of fixed and random effects,
-and used approximate leave-one-out cross validation to identify the best
-model \citep{vehtari2016}.
-
-Finally, while it would theoretically be possible to estimate survival
-covariate effects in a joint CMR model that integrates capture histories
-across all sites, this was impractical due to computational requirements
-of the CMR models (namely, run time and memory). In modeling survival
-estimates from site-specific CMR analyses, we are in effect conducting
-an among-site meta-analysis to better understand how site, cohort, and
-individual level characteristics explain variation in survival after
-translocation.
+the first half of each chain as warm-up draws (71). We used Rhat
+statistics and trace plots to verify convergence. We considered models
+with different subsets of fixed and random effects, and used approximate
+leave-one-out cross validation to identify the best model (72).
 
@@
 the posterior distribution of individual state trajectories derived from
-CMR models. Although we observed yearly variability in adult survival
-within a population, the magnitude of this variability was small
-compared to among-population variability
+mrmr CMR models. Although we observed yearly variability in adult
+survival within a population, the magnitude of this variability was
+small compared to among-population variability
 (Fig. 3). Thus, we did not include
@@
 stochasticity in year 2 juvenile survival and recruitment
-\(\sigma_{J_2}\), but our elasticity analysis showed that this parameter
-had little effect on host growth rate relative to \(\sigma_{J_1}\)
-(Figure~\ref{fig-viability-supp}).
+\(\sigma_{J_2}\), but our elasticity analysis (\textbf{Supporting
+Information-Population viability modeling-Model analysis and
+simulation}) showed that this parameter had little effect on host growth
+rate relative to \(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp}).
 
@@
 predicted true states for naturally-recruited individuals (1=not
-recruited, 2=alive, 3=dead, as described in \emph{CMR model structure}),
-then calculated the posterior probability of individuals surviving
-between consecutive primary periods, conditional on being alive in the
-first primary period (e.g., given a value of 2 (alive) in the first
-primary period, how often was the value still 2 (alive) in the next
-primary period compared to 3 (dead)?) . This yielded posterior
-distributions for survival probabilities between primary periods.
-However, because the time interval between primary periods differed, the
-survival probabilities between different consecutive primary periods
-were not directly comparable. To address this, we converted the survival
+recruited, 2=alive, 3=dead, as described in \textbf{Supporting
+Information-Frog population recovery-CMR model structure}), then
+calculated the posterior probability of individuals surviving between
+consecutive primary periods, conditional on being alive in the first
+primary period (e.g., given a value of 2 (alive) in the first primary
+period, how often was the value still 2 (alive) in the next primary
+period compared to 3 (dead)?) . This yielded posterior distributions for
+survival probabilities between primary periods. However, because the
+time interval between primary periods differed, the survival
+probabilities between different consecutive primary periods were not
+directly comparable. To address this, we converted the survival
 probabilities between each consecutive primary period to per day death
@@
 1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
-Using Approximate Bayesian Computing and rejection sampling
-\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
-1000 trajectories) that minimized the sum of squared errors between the
-observed and predicted data. The yearly \(\sigma_{J_1}\) values
-associated with these ``best'' trajectories represented an approximate
-posterior distribution \citep{beaumont2010}. Using these best fit
-trajectories, we assessed whether our model could qualitatively describe
-the patterns of recovery in the observed data for population 70550.
+Using Approximate Bayesian Computing and rejection sampling (92), we
+identified the top 2\% of trajectories (i.e., 1000 trajectories) that
+minimized the sum of squared errors between the observed and predicted
+data. The yearly \(\sigma_{J_1}\) values associated with these ``best''
+trajectories represented an approximate posterior distribution (93).
+Using these best fit trajectories, we assessed whether our model could
+qualitatively describe the patterns of recovery in the observed data for
+population 70550.
 
@@
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\hypertarget{study-design}{%
+\subsection{Study design}\label{study-design}}
 
-\hypertarget{study-design}{%
-\subparagraph{Study design}\label{study-design}}
+To gain insights into the role of evolution in the development of
+resistance by MYL frogs, we compared frog genomes sampled in naive
+versus recovering populations. Comparing populations with different
+infection histories allowed larger sample sizes and replication across
+the landscape. The alternative approach of comparing samples from the
+same populations before and after Bd exposure isn't feasible in this
+system because Bd arrived in most MYL frog populations decades ago and
+population persistence/recovery is rare and unpredictable. As a result,
+samples from recovering populations collected from before and after Bd
+exposure are not available and are unlikely to be available in the
+future.
 
-We compared frog genomes sampled in naive versus recovering populations.
-Comparing populations with different infection histories allowed larger
-sample sizes and replication across the landscape. The alternative
-approach of comparing samples from the same populations before and after
-Bd exposure isn't feasible in this system because Bd arrived in most MYL
-frog populations decades ago and population persistence/recovery is rare
-and unpredictable. As a result, samples from recovering populations
-collected from before and after Bd exposure are not available and are
-unlikely to be available in the future.
+Our study design, in which we compared frog genomes in naive and
+recovering populations, required sampling populations across the range
+of MYL frogs in the southern Sierra Nevada. This is due to the fact that
+very few MYL frog populations remain in the naive state, and those that
+do are scattered across a wide latitudinal range, from Yosemite National
+Park in the north to southern Kings Canyon National Park in south. To
+minimize potential confounding effects caused by known variation in frog
+genotypes across latitude (49), we selected sampling sites such that
+both population types were represented across similar latitudinal ranges
+(Fig. 1).
 
 \hypertarget{go-analysis}{%
-\subparagraph{GO analysis}\label{go-analysis}}
+\subsection{GO analysis}\label{go-analysis}}
 
-In the Results section of the main paper, we describe the stringent set
-of outlier SNPs (identified using a Bonferroni-corrected p-value of
-0.01). The liberal set, identified using a Bonferroni-corrected p-value
-of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
-genes across 16 contigs. Using this liberal set of outlier variants, we
-performed a statistical overrepresentation test (Fisher's exact test)
-using Blast2GO \citep{conesa2005}. First, we retrieved the BLAST hits
-and mapped GO terms for each gene in our targeted transcriptome. Next,
-we compared the 30 unique outlier genes from the liberal set of 38
-outlier variants to the complete set of genes in our target
-transcriptome to determine if any GO biological functions, molecular
-functions, or cellular processes were overrepresented. We then repeated
-this process for the set of 33 genes located in the
-\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+In \textbf{Results-Frog evolution in response to Bd}, we describe the
+stringent set of outlier variants (identified using a
+Bonferroni-corrected p-value of 0.01). A liberal set of outlier
+variants, identified using a Bonferroni-corrected p-value of 0.05,
+included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
+across 16 contigs. We used this liberal set to determine if any GO
+biological functions, molecular functions, or cellular processes were
+overrepresented. To do this, we retrieved the BLAST hits and mapped GO
+terms for each gene in our targeted transcriptome. We then conducted a
+statistical overrepresentation test (Fisher's exact test) using Blast2GO
+(94) to compare the 30 unique outlier genes to the complete set of genes
+in our target transcriptome. We repeated this process for the set of 33
+genes located in the \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
+splined windows.
 
 \hypertarget{genetic-diversity}{%
-\subparagraph{Genetic diversity}\label{genetic-diversity}}
+\subsection{Genetic diversity}\label{genetic-diversity}}
 
@@
 persistent populations, we calculated individual heterozygosity using
-VCFtools \citep{danecek2011}, and compared average heterozygosity using
-a Wilcoxon rank-sum test. Additionally, we used VCFtools to calculate
+VCFtools (85), and compared average heterozygosity using a Wilcoxon
+rank-sum test. Additionally, we used VCFtools to calculate
 per-population \(\pi\) in 100kb sliding windows along the genome, and in
@@
 Wilcoxon rank-sum tests.
-
-\hypertarget{results-1}{%
-\subsection{Results}\label{results-1}}
-
-\hypertarget{go-analysis-1}{%
-\subparagraph{GO analysis}\label{go-analysis-1}}
-
-We found no significantly over-represented GO terms in the liberal set
-of outlier regions, or in the set of 33 genes located in the overlapping
-\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
-Therefore\ldots..
-
-\hypertarget{genetic-diversity-1}{%
-\subparagraph{Genetic diversity}\label{genetic-diversity-1}}
 
@@
 
-\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
-important predictor of post-translocation frog survival. Depicted
-distributions are the estimated posterior density curves and shaded 95\%
-uncertainty intervals for the intercept and all predictor variables from
-the best model in the between-site meta-analysis. In the Bayesian
-framework in which the model was developed, variables are considered
-important predictors if the associated uncertainty interval does not
-overlap zero (indicated by the dashed red line). Predictor variables
-shown on the y-axis are defined as follows: length = frog size, snow\_t
-= winter severity in the year of translocation (measured on April 1),
-snow\_t1 = winter severity in the year following translocation (measured
-on April 1), day = day of year on which a translocation was conducted,
-bd\_load = Bd load, elevation = site elevation, order\_first =
-within-site translocation order, sex\_male = frog sex, donor\_70567 and
-donor\_72996 = donor population.}
+\caption{\label{fig-survival-postdens}Results from the among-site
+meta-analysis, showing that Bd load is not an important predictor of
+post-translocation frog survival. Depicted distributions are the
+estimated posterior density curves and shaded 95\% uncertainty intervals
+for the intercept and all predictor variables from the best model. In
+the Bayesian framework in which the model was developed, variables are
+considered important predictors if the associated uncertainty interval
+does not overlap zero (indicated by the dashed red line). Predictor
+variables shown on the y-axis are defined as follows: length = frog
+size, snow\_t = winter severity in the year of translocation (measured
+on April 1), snow\_t1 = winter severity in the year following
+translocation (measured on April 1), day = day of year on which a
+translocation was conducted, bd\_load = Bd load, elevation = site
+elevation, order\_first = within-site translocation order, sex\_male =
+frog sex, donor\_70567 and donor\_72996 = donor population.}
 
@@
 
-\caption{\label{fig-selectionresults}Evidence for selection on
-individual variants in recovering MYL frog populations. (A) PCA
-calculated from binary SNPs showing the genomic relationship of samples.
-Numeric labels and colors match those in
-Fig. 1A. Populations 1-7 are \emph{R.
-sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
-showing observed and expected p-values for 148,307 SNPs and INDELS.
-Dashed line shows p-value that identifies outliers. (C) Manhattan plot
-showing p-value for each SNP. SNPs are sorted by genomic position and
-contigs are sorted by size. Red line shows p-value that identifies
-outliers. Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in
-that gene.}
+\caption{\label{fig-selectionresults}Results from the analysis of
+individual variants, showing putative signatures of selection in
+recovering MYL frog populations. (A) PCA calculated from binary SNPs
+showing the genomic relationship of samples. Numeric labels and colors
+match those in Fig. 1A. Populations 1-7 are
+\emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
+qqplot showing observed and expected p-values for 148,307 SNPs and
+INDELS. Dashed line shows p-value that identifies outliers. (C)
+Manhattan plot showing p-value for each SNP. SNPs are sorted by genomic
+position and contigs are sorted by size. Red line shows p-value that
+identifies outliers. Outlier SNPs above this threshold are highlighted
+and labeled. Bold labels indicate the presence of at least one
+non-synonymous SNP in that gene.}
 
@@
 outlier region containing MHC Class I Classical and MHC Class III gene
-regions. The plot was created with SimpleSynteny \citep{veltri2016}
-using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
-accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
-indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
-sizes for each contig representation are labeled.}
+regions. The plot was created with SimpleSynteny (95) using
+\emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank accession
+GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices indicate
+the location of SNP outliers in the TCF19 and VARS1 genes. Gap sizes for
+each contig representation are labeled.}
 
@@
 heterozygosity for the Bd-naive and recovering populations included in
-the frog genomics study. Individual data points are shown on the plot
-with their corresponding site number (from
+the frog evolution study. Individual data points are represented by
+their corresponding site number (from
 Fig. 1). A Wilcoxon signed-rank test showed
@@
 \caption{\label{fig-boxplot-pi}For each study population in the frog
-evolution study (from Fig. 1A), boxplots
+evolution study (from Fig. 1), boxplots
 showing nucleotide diversity (\(\pi\)) within 100kb sliding windows

### 2023-12-21 doc/manuscript/translocation.qmd
Update genomic figures and associated text

@@
 Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-Emerging infectious diseases can have particularly devastating impacts, and few options typically exist to reverse these effects.
-We used large-scale reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to the brink of extinction by a disease that has recently devastated global amphibian biodiversity.
-Introduction of resistant frogs allowed reestablishment of populations in the presence of disease.
-In addition, resistance appears to have a genetic basis and is consistent with evolution following exposure of frogs to Bd.
+Emerging infectious diseases can have particularly devastating impacts, and few options exist to reverse these effects.
+We used large-scale reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a disease that has decimated amphibian biodiversity.
+Introduction of resistant frogs allowed reestablishment of viable populations in the presence of disease.
+In addition, resistance may be at least partially the result of natural selection at specific immune function genes, which show evidence for selection in recovering populations.
 The evolution of resistance and reintroductions of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes <!--# Is this correct, i.e., that we identified 33 outlier regions, and in the subset of 9 regions we found 33 annotated genes?
-I just want to make sure there isn't a typo in here.
-Seems a weird coincidence to have 33 of both.) And the file spline_window_shared_outliers.csv lists 35 genes.
--->were found in the 9 shared regions (Datasets S3, S4).
+A total of 35 annotated genes were found in the 9 shared regions (Datasets S3, S4).
 Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
@@
 
-To characterize general patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity using VCFtools [@danecek2011], and compared average heterozygosity using a Wilcoxon rank-sum test<!--# This analysis is problematic because it doesn't account for the lack of independence between points (i.e., individuals grouped within populations).
-In addition, there is a pattern of R.
-muscosa populations in both groups (7, 8, 9) having low Fst, so species identity is another grouping that contributes to a lack of independence.
-Given that we don't have a large enough sample size to run an analysis that properly accounts for this lack of independence, I'd argue for dropping this analysis and providing only a visual description of the differences.
--->.
-Additionally, we used VCFtools to calculate per-population $\pi$ in 100kb sliding windows along the genome, and in each of the 10 <!--# In the main manuscript, we state that 9 windows were identified.
-Why the difference?
--->outlier windows identified in the splined window analysis.
-We compared per-population average $\pi$ values using Wilcoxon rank-sum tests.
-<!--# Allie, did I describe this correctly?
---> <!--# The displayed per-population pi values show tiny differences between naive and persistent populations.
-Is the significance shown by the Wilcoxon test due to large sample sizes?
-Given that the interest here is in describing any meaningful differences between naive and persistent populations, it doesn't accomplish much to highlight tiny differences that almost certainly have no biological relevance.
-In addition, all of the lack of independence issues that I highlighted above are probably relevant here too.
-Would it be better to describe the results qualitatively?
-E.g., Differences between sites were small and did not show any consistent patterns between naive and recovering populations.
--->
-
-Average (i) individual-level heterozygosity and (ii) population-level nucleotide diversity ($\pi$) were similar between naive and recovering populations (@fig-violinplot-heterozy SI; @fig-boxplot-pi SI).
-The analysis of $\pi$ across the 10 <!--# 9?
--->outlier windows indicates that within each of the windows there is considerable variation in $\pi$ between populations and between population clusters (@fig-barchart-pi-by-windowsite SI).
-<!--# Allie, please check that my text accurately describes the results. -->
+To characterize general patterns of genetic diversity between naive and recovering populations, we conducted three analyses.
+First, we calculated heterozygosity for each sampled frog using VCFtools [@danecek2011].
+Second, to characterize genome-wide patterns of diversity, we used VCFtools to calculate nucleotide diversity ($\pi$) in 100kb sliding windows along the genome for each population.
+Third, we calculated average $\pi$ per population within each of the 9 outlier windows identified in the splined window analysis.
+
+Average individual-level heterozygosity and genome-wide population-level $\pi$ were similar between the naive and recovering groups (@fig-violinplot-heterozy SI; @fig-boxplot-genomewide-pi-by-pop SI).
+Within each of the 9 outlier windows, average $\pi$ shows considerable variation between populations (@fig-boxplot-pi-by-windowpop SI) and no obvious patterns between the naive and recovering groups.
 
@@
 Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).
-A Wilcoxon signed-rank test showed no significant difference between the two groups.](figures/het_by_status_violin_plot.png){#fig-violinplot-heterozy width="500"}
+](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="500"}
 
@@
 Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).
-Within the genetic clusters, mean $\pi$ is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).<!--# Change x-axis label to "Population" and x-axis tick labels to population #s only.
--->](figures/boxplot_pi_by_site.png){#fig-boxplot-pi width="500"}
-
-\newpage
-
-![For each study population in the frog evolution study, average $\pi$ values by population calculated for each of the 10 outlier windows <!--# I thought it was 9?
--->identified in the splined window analysis (A-J).
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI.
-Gene annotations for the outlier regions are as follows: (A) C6, C7, (B) DDX10, ZBTB24, (C) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (D) FLOT1, TUBB2B, TUBB, MDC1, CCHCR1, TCF19*, TRIM29, HSP70, LSM2, VARS,* (E) GCC2, CFAP251, PEG10, (F) ERO1A, GVINP1, (G) ERO1A, GVINP1, (H) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (I) C6H5ORF22, PKS6, BSPRY, MPV17, and (J) CAD, ATRAID, GPN1.
-Note that windows F and G are in the same gene region.
-\* by gene names indicate that an outlier SNP is also present in this gene <!--# No asterices are associated with any of the gene names listed above -->.
-(K) Average $\pi$ values of naive and recovering groups for each of the outlier windows shown in A-J.
-<!--# These are mean values so they should be displayed using boxplots.
-Not providing that context can be misleading.
-Given the variation between populations for each window, I don't think (K) is useful even if boxplots were used.
-I think plots A-J tell the story and K doesn't add anything.
---> <!--# Put plot identifiers A-K inside the upper-right corner of each plot.
-Cleaner and less ambiguous than having identifiers between plots.
-Alternatively, the identifier for plot K could be placed inside the upper-left corner of that plot.-->](figures/perwindow_persite_pi.png){#fig-barchart-pi-by-windowsite width="650"}
+Within the genetic clusters, mean $\pi$ is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
+
+\newpage
+
+![For each of the 9 outlier windows identified by the overlapping splined windows for FST and \$\\pi\$, boxplots showing average \$\\pi\$ for naive and recovering populations.
+The corresponding gene annotations for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_windowpop.png){#fig-boxplot-pi-by-windowpop width="500"}
 

### 2023-12-21 doc/manuscript/translocation_bioarxiv.tex
Compile PDF versions of manuscript

--- 2023-12-20 doc/manuscript/translocation_bioarxiv.tex
@@
 diseases can have particularly devastating impacts, and few options
-typically exist to reverse these effects. We used large-scale
-reintroductions of disease-resistant individuals in an effort to recover
-a once-common frog species driven to the brink of extinction by a
-disease that has recently devastated global amphibian biodiversity.
-Introduction of resistant frogs allowed reestablishment of populations
-in the presence of disease. In addition, resistance appears to have a
-genetic basis and is consistent with evolution following exposure of
-frogs to Bd. The evolution of resistance and reintroductions of
-resistant individuals could play an important role in biodiversity
-conservation in our rapidly changing world.}
+exist to reverse these effects. We used large-scale reintroductions of
+disease-resistant individuals in an effort to recover a once-common frog
+species driven to near-extinction by a disease that has decimated
+amphibian biodiversity. Introduction of resistant frogs allowed
+reestablishment of viable populations in the presence of disease. In
+addition, resistance may be at least partially the result of natural
+selection at specific immune function genes, which show evidence for
+selection in recovering populations. The evolution of resistance and
+reintroductions of resistant individuals could play an important role in
+biodiversity conservation in our rapidly changing world.}
 
@@
 a Bayesian framework and included a diversity of site, cohort, and
-individual level characteristics as predictors and 1-year frog survival
+individual-level characteristics as predictors and 1-year frog survival
 (Figure~\ref{fig-translocation-survival}) as the response variable. The
@@
 
-To determine whether MYL frog populations show an evolutionary response
-to Bd, we compared frog exomes (i.e., coding region of genome) between
-populations with contrasting histories of Bd exposure. Specifically, we
-compared frog genomes sampled in 4 populations that have not yet
-experienced a Bd-caused epizootic (``naive'') \citep{zhou2015} versus in
-5 populations that experienced a Bd epizootic during the past several
-decades and have since recovered to varying degrees (``recovering'';
-Figure~\ref{fig-allelefrequencies}) \citep{knapp2016, vredenburg2010}.
-Bd-exposure histories of the 9 study populations are based on 10-20
-years of VES and Bd surveillance using skin swabbing
-\citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive populations are
-characterized by large numbers of adults (i.e, typically 1000s), Bd
-prevalence that is generally 0\% except during occasional Bd failed
-invasions \citep[during which Bd loads remain very low,][]{wilber2022},
-and no history of Bd epizootics since we first surveyed these
-populations in the late 1990s and early 2000s \citep{zhou2015}. In
-contrast, recovering populations exist in an enzootic state
-\citep{briggs2010}, characterized by smaller numbers of adults
-(generally \textless{} 100), high Bd prevalence \citep[often
+To determine whether MYL frog populations show genomic patterns
+consistent with an evolutionary response to Bd, we compared frog exomes
+(i.e., coding region of a genome) between populations with contrasting
+histories of Bd exposure. Specifically, we compared frog genomes sampled
+in 4 populations that have not yet experienced a Bd-caused epizootic
+(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
+Bd epizootic during the past several decades and have since recovered to
+varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
+\citep{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
+populations are based on 10-20 years of VES and Bd surveillance using
+skin swabbing \citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
+populations are characterized by large numbers of adults (i.e, typically
+1000s), Bd prevalence that is generally 0\% except during occasional Bd
+failed invasions \citep[during which Bd loads remain very
+low,][]{wilber2022}, and no history of Bd epizootics since we first
+surveyed these populations in the late 1990s and early 2000s
+\citep{zhou2015}. In contrast, recovering populations exist in an
+enzootic state \citep{briggs2010}, characterized by smaller numbers of
+adults (generally \textless{} 100), high Bd prevalence \citep[often
 \textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
@@
 the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
-the strong signature of isolation by distance that is characteristic of
+the strong signature of isolation-by-distance that is characteristic of
 MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
@@
 regions also contained one or more of the outlier SNPs described above.
-A total of 33 annotated genes were found in the 9 shared regions
+A total of 35 annotated genes were found in the 9 shared regions
 (Datasets S3, S4). Given this large number of genes, here we focus on
@@
 
-No gene ontology (GO) biological functions, molecular functions, or
-cellular processes were over-represented in either the outlier variants
-or the 33 genes located in the overlapping \emph{F\textsubscript{ST}}
-and \(\pi_{diff}\) splined windows. In addition, although the joint
-processes of Bd-caused population declines and selection in response to
-Bd exposure could affect genetic diversity of recovering populations, we
-found no consistent differences in individual-level heterozygosity or
-population-level \(\pi\) between naive and recovering populations (see
-\textbf{Supporting Information-Frog evolution in response to Bd-Methods}
-for details).
-
-In summary, these results indicate that the exomes of frogs from our
+Although the joint processes of Bd-caused population declines and
+selection in response to Bd exposure could affect genetic diversity of
+recovering populations, we found no consistent differences in
+individual-level heterozygosity or population-level \(\pi\) between
+naive and recovering populations (see \textbf{Supporting
+Information-Frog evolution in response to Bd} for details). Thus,
+despite localized selection in particular regions of the genome, we did
+not find evidence for reduced genetic diversity across the genome in
+recovering populations. In addition, no gene ontology (GO) biological
+functions, molecular functions, or cellular processes were
+over-represented in either the outlier variants or the 33 genes located
+in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined
+windows (see \textbf{Supporting Information-Frog evolution in response
+to Bd} for details).
+
+In summary, our genomic results indicate that the exomes of frogs from
 naive and recovering study populations show substantial differences,
 consistent with parallel evolutionary changes following Bd exposure. The
-regions putatively under selection contain several
-immunologically-relevant genes and gene families that are directly
-linked to disease resistance in other taxa.
+regions under selection contain several immunologically-relevant genes
+and gene families that are directly linked to disease resistance in
+other taxa.
 
@@
 \citep{daszak2000}, but broadly-applicable strategies to recover
-affected species are generally lacking \citep[e.g.,][]{garner2016}. Our
-results indicate (i) the capacity of reintroduced populations to become
-established and eventually recover despite ongoing disease, (ii) that
-the recovering populations are likely to persist over a 50-year period,
-(iii) that there are substantial genomic differences between naive and
-recovering MYL frog populations, consistent with evolutionary change in
-frogs following Bd exposure, and (iv) that some of the genomic regions
-putatively under selection contain genes related to disease resistance.
-Collectively, these results (Figure~\ref{fig-recovery-model}) provide a
-rare example of amphibian recovery in the presence of a highly virulent
-pathogen, and have important implications for the conservation and
-recovery of amphibians and other taxa worldwide that are endangered by
-escalating impacts from emerging infectious diseases. In light of the
-generally low success rate of amphibian reintroduction efforts
-\citep{dodd2005}, our success in reestablishing MYL frog populations via
-translocation of resistant individuals is striking, and even more so
-given that this success occurred despite ongoing infection of frogs with
-Bd.
-
-Previous field studies in MYL frogs show that frog-Bd dynamics and frog
-survival in the presence of Bd are fundamentally different between naive
-and recovering populations. Following the arrival and establishment of
-Bd in previously-naive populations, adult frogs develop high Bd loads
-that lead to mass die-offs \citep{vredenburg2010}. In contrast, in
-recovering populations adult frogs typically have low-to-moderate and
-relatively constant Bd loads and mass die-offs are not observed
-\citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
+affected species are generally lacking \citep[e.g.,][]{garner2016}.
+Here, we tested the possibility that populations of resistant
+individuals from naturally recovering populations can be used to
+reestablish extirpated populations of the endangered MYL frog in the
+presence of a highly virulent fungal pathogen (Bd). Our results indicate
+(i) the capacity of reintroduced populations to become established and
+eventually recover despite ongoing disease, (ii) that the recovering
+populations are likely to persist over a 50-year period, (iii) that
+there are substantial genomic differences between naive and recovering
+MYL frog populations, consistent with evolutionary change in frogs
+following Bd exposure, and (iv) that some of the genomic regions under
+selection contain genes related to disease resistance. Collectively,
+these results (Figure~\ref{fig-recovery-model}) provide a rare example
+of amphibian recovery in the presence of Bd, and have important
+implications for the conservation and recovery of amphibians and other
+taxa worldwide that are endangered by escalating impacts from emerging
+infectious diseases. In light of the generally low success rate of
+amphibian reintroduction efforts \citep{dodd2005}, our success in
+reestablishing MYL frog populations via translocation of resistant
+individuals is striking, and even more so given that this success
+occurred despite ongoing infection of frogs with Bd.
+
+In the following discussion, we follow the sequence of frog recovery
+described in Figure~\ref{fig-recovery-model} to structure our key
+points. Previous field studies in MYL frogs show that frog-Bd dynamics
+and frog survival in the presence of Bd are fundamentally different
+between naive and recovering populations. Following the arrival and
+establishment of Bd in previously-naive populations, adult frogs develop
+high Bd loads that lead to mass die-offs \citep{vredenburg2010}. In
+contrast, in recovering populations adult frogs typically have
+low-to-moderate and relatively constant Bd loads and mass die-offs are
+not observed \citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
 naive and recovering populations are also observed in controlled
@@
 adaptive alleles is at least partially responsible for the increased
-resistance of frogs in recovering populations. We identified specific
-alleles and genomic regions showing signatures of selection between
-adjacent naive and recovering MYL frog populations, consistent with
-selection following Bd exposure. Importantly, some regions under
-selection are associated with cellular and immunological mechanisms
-known to contribute to disease resistance, including in amphibians
+resistance of frogs in recovering populations. We identified multiple
+specific alleles and genomic regions showing signatures of selection
+between adjacent naive and recovering MYL frog populations, consistent
+with selection following Bd exposure. These analysis are based on
+samples collected from virtually all of the MYL frog populations
+remaining in a naive state, as well as adjacent recovering populations.
+This produced genetic clusters that contained at least one naive and one
+recovering population, allowing us to detect selection without the
+confounding effects of population structure. Finally, we did not find a
+reduction in overall genetic variation in the recovering populations,
+suggesting that despite localized selection in the genome, these
+populations retain adequate genetic diversity for long-term persistence.
+
+Importantly, some regions that we identified as under selection are
+associated with cellular and immunological mechanisms known to
+contribute to disease resistance, including in amphibians
 \citep{zamudio2020}. For example, the MHC plays an important role in
@@
 region of the genome to favor a diverse set of alleles, as is known for
-C6 in humans \citep{soejima2005}. Finally, based on the analysis of
-individual outlier variants, the RIN3 gene showed a consistent pattern
-of allele frequency differences across all nine of the frog populations
-sampled in this study, indicating consistent selection in populations
-distributed across a wide geographic area. This gene is associated with
-inflammation response and in \emph{Xenopus} is highly expressed in
-wounds \citep{fukazawa2009}. In addition, the outlier variant with the
-lowest p-value was the uncharacterized gene LOC108802036. In the genome
-of another frog species, this gene is located adjacent to a type I
+C6 in humans \citep{soejima2005}. Based on the analysis of individual
+outlier variants, the RIN3 gene showed a consistent pattern of allele
+frequency differences across all nine of the frog populations sampled in
+this study, indicating consistent selection in populations distributed
+across a wide geographic area. This gene is associated with inflammation
+response and in \emph{Xenopus} is highly expressed in wounds
+\citep{fukazawa2009}. Finally, the outlier variant with the lowest
+p-value was the uncharacterized gene LOC108802036. In the genome of
+another frog species, this gene is located adjacent to a type I
 interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
@@
 \citep{jombart2008}. To detect regions of the genome that differed
-between naive and recovering populations, i.e., regions putatively under
-selection, we used two approaches: (1) a multivariate linear mixed model
-to evaluate individual variants (SNPs and INDELs), and (2) a splined
-window analysis to evaluate larger genomic regions. For the variant
-analysis, we first used a stringent data filter to include only variants
-with \textless{} 5\% missing data (missing for no more than 2
-individuals), and then calculated the likelihood ratio statistic for the
-resulting set of 148,307 high quality variants across 127 contigs using
-GEMMA \citep{zhou2014}. GEMMA calculates and incorporates a relatedness
-matrix for input samples, allowing us to account for relatedness and
-population structure when calculating likelihood ratio statistics. We
-identified variants showing different allele frequencies between naive
-versus recovering populations (``outliers'') using a
-Bonferroni-corrected significance level of 0.01. We visualized the
-results using a Manhattan plot and qqplot. We developed a more liberal
-set of outlier variants using a Bonferroni-corrected significance level
-of 0.05 and used this set solely for the gene ontology (GO) analysis
-(see below and \textbf{Supporting Information}: Dataset S1, S2).
+between naive and recovering populations, i.e., regions under selection,
+we used two approaches: (1) a multivariate linear mixed model to
+evaluate individual variants (SNPs and INDELs), and (2) a splined window
+analysis to evaluate larger genomic regions. For the variant analysis,
+we first used a stringent data filter to include only variants with
+\textless{} 5\% missing data (missing for no more than 2 individuals),
+and then calculated the likelihood ratio statistic for the resulting set
+of 148,307 high quality variants across 127 contigs using GEMMA
+\citep{zhou2014}. GEMMA calculates and incorporates a relatedness matrix
+for input samples, allowing us to account for relatedness and population
+structure when calculating likelihood ratio statistics. We identified
+variants showing different allele frequencies between naive versus
+recovering populations (``outliers'') using a Bonferroni-corrected
+significance level of 0.01. We visualized the results using a Manhattan
+plot and qqplot. We developed a more liberal set of outlier variants
+using a Bonferroni-corrected significance level of 0.05 and used this
+set solely for the gene ontology (GO) analysis (see below and
+\textbf{Supporting Information}: Dataset S1, S2).
 

### 2023-12-21 doc/manuscript/translocation_bioarxiv_SI.tex
Compile PDF versions of manuscript

@@
 Swab extracts were analyzed using standard Bd DNA extraction and qPCR
-methods (74), and extracts were analyzed singly instead of in triplicate
-(87). For analysis of swabs collected during 2005--2014, we used
-standards developed from known concentrations of zoospores (74), and
-after 2014, we used standards based on single ITS1 PCR amplicons (88).
-Based on paired comparisons between samples analyzed using both types of
-standards, Bd in the study area has an average of 60 ITS1 copies per
-zoospore. To express all qPCR results as the number of ITS1 copies,
-starting quantities obtained using the zoospore standard (measured as
-``zoospore equivalents'') were multiplied by 60. In addition, all qPCR
-quantities (regardless of standard) were multiplied by 80 to account for
-the fact that DNA extracts from swabs were diluted 80-fold during
-extraction and PCR (33).
+methods \citep{boyle2004}, and extracts were analyzed singly instead of
+in triplicate \citep{kriger2006}. For analysis of swabs collected during
+2005--2014, we used standards developed from known concentrations of
+zoospores \citep{boyle2004}, and after 2014, we used standards based on
+single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
+between samples analyzed using both types of standards, Bd in the study
+area has an average of 60 ITS1 copies per zoospore. To express all qPCR
+results as the number of ITS1 copies, starting quantities obtained using
+the zoospore standard (measured as ``zoospore equivalents'') were
+multiplied by 60. In addition, all qPCR quantities (regardless of
+standard) were multiplied by 80 to account for the fact that DNA
+extracts from swabs were diluted 80-fold during extraction and PCR
+\citep{vredenburg2010}.
 
@@
 We estimated survival and recruitment for each site using open
-population CMR models based on (41). For each individual \(i=1, ..., M\)
-on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
-not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
-histories of \(M\) individuals are modeled, although only \(N_s\)
-individuals were captured. This parameter expanded data augmentation
-allows us to capture the possibility that undetected individuals may
-have recruited into the adult population (89). Here, \(M\) was chosen to
-be three times the number of observed individuals (\(3N_s\)) to be
-considerably greater than our prior guess of \(N_s\).
+population CMR models based on \citep{joseph2018}. For each individual
+\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
+the individual was not detected, and \(o_{i, j}=2\) if the individual
+was detected. Capture histories of \(M\) individuals are modeled,
+although only \(N_s\) individuals were captured. This parameter expanded
+data augmentation allows us to capture the possibility that undetected
+individuals may have recruited into the adult population
+\citep{royle2012}. Here, \(M\) was chosen to be three times the number
+of observed individuals (\(3N_s\)) to be considerably greater than our
+prior guess of \(N_s\).
 
@@
 forward algorithm, and we estimated the latent states using the
-forward-backward algorithm (41, 90).
+forward-backward algorithm \citep{zucchini2009, joseph2018}.
 
 All of the code to specify and fit the model in Stan is available in the
-open source mrmr package (67).
+open source mrmr package \citep{joseph2019}.
 
@@
 These models were fit using the \texttt{stan\_glmer()} function in the
-rstanarm package, with default priors described below (71). These priors
-are vague, but include data-dependent scaling as follows to account for
-different input variable scales. However, because we standardized all
-predictor variables similarly to have equal variance (by centering and
-dividing by twice the sample standard deviation), the resulting priors
-are identical. Specifically, we have:
+rstanarm package, with default priors described below
+\citep{rstanarm2022}. These priors are vague, but include data-dependent
+scaling as follows to account for different input variable scales.
+However, because we standardized all predictor variables similarly to
+have equal variance (by centering and dividing by twice the sample
+standard deviation), the resulting priors are identical. Specifically,
+we have:
 
@@
 variance parameters -- the decomposition of variance prior with unit
-regularization, concentration, shape, and scale parameters (91).
+regularization, concentration, shape, and scale parameters
+\citep{lewandowski2009}.
 
@@
 with four parallel chains, each run for 10,000 iterations, discarding
-the first half of each chain as warm-up draws (71). We used Rhat
-statistics and trace plots to verify convergence. We considered models
-with different subsets of fixed and random effects, and used approximate
-leave-one-out cross validation to identify the best model (72).
+the first half of each chain as warm-up draws \citep{rstanarm2022}. We
+used Rhat statistics and trace plots to verify convergence. We
+considered models with different subsets of fixed and random effects,
+and used approximate leave-one-out cross validation to identify the best
+model \citep{vehtari2016}.
 
@@
 1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
-Using Approximate Bayesian Computing and rejection sampling (92), we
-identified the top 2\% of trajectories (i.e., 1000 trajectories) that
-minimized the sum of squared errors between the observed and predicted
-data. The yearly \(\sigma_{J_1}\) values associated with these ``best''
-trajectories represented an approximate posterior distribution (93).
-Using these best fit trajectories, we assessed whether our model could
-qualitatively describe the patterns of recovery in the observed data for
-population 70550.
+Using Approximate Bayesian Computing and rejection sampling
+\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
+1000 trajectories) that minimized the sum of squared errors between the
+observed and predicted data. The yearly \(\sigma_{J_1}\) values
+associated with these ``best'' trajectories represented an approximate
+posterior distribution \citep{beaumont2010}. Using these best fit
+trajectories, we assessed whether our model could qualitatively describe
+the patterns of recovery in the observed data for population 70550.
 
@@
 minimize potential confounding effects caused by known variation in frog
-genotypes across latitude (49), we selected sampling sites such that
-both population types were represented across similar latitudinal ranges
-(Fig. 1).
+genotypes across latitude \citep{byrne2023}, we selected sampling sites
+such that both population types were represented across similar
+latitudinal ranges (Fig. 1).
 
@@
 statistical overrepresentation test (Fisher's exact test) using Blast2GO
-(94) to compare the 30 unique outlier genes to the complete set of genes
-in our target transcriptome. We repeated this process for the set of 33
-genes located in the \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
-splined windows.
+\citep{conesa2005} to compare the 30 unique outlier genes to the
+complete set of genes in our target transcriptome. We repeated this
+process for the set of 33 genes located in the
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
 
@@
 To characterize general patterns of genetic diversity between naive and
-persistent populations, we calculated individual heterozygosity using
-VCFtools (85), and compared average heterozygosity using a Wilcoxon
-rank-sum test. Additionally, we used VCFtools to calculate
-per-population \(\pi\) in 100kb sliding windows along the genome, and in
-each of the 10 outlier windows identified in the splined window
-analysis. We compared per-population average \(\pi\) values using
-Wilcoxon rank-sum tests.
+recovering populations, we conducted three analyses. First, we
+calculated heterozygosity for each sampled frog using VCFtools
+\citep{danecek2011}. Second, to characterize genome-wide patterns of
+diversity, we used VCFtools to calculate nucleotide diversity (\(\pi\))
+in 100kb sliding windows along the genome for each population. Third, we
+calculated average \(\pi\) per population within each of the 9 outlier
+windows identified in the splined window analysis.
 
-Average (i) individual-level heterozygosity and (ii) population-level
-nucleotide diversity (\(\pi\)) were similar between naive and recovering
-populations (Figure~\ref{fig-violinplot-heterozy};
-Figure~\ref{fig-boxplot-pi}). The analysis of \(\pi\) across the 10
-outlier windows indicates that within each of the windows there is
-considerable variation in \(\pi\) between populations and between
-population clusters (Figure~\ref{fig-barchart-pi-by-windowsite}).
+Average individual-level heterozygosity and genome-wide population-level
+\(\pi\) were similar between the naive and recovering groups
+(Figure~\ref{fig-violinplot-heterozy};
+Figure~\ref{fig-boxplot-genomewide-pi-by-pop}). Within each of the 9
+outlier windows, average \(\pi\) shows considerable variation between
+populations (Figure~\ref{fig-boxplot-pi-by-windowpop}) and no obvious
+patterns between the naive and recovering groups.
 
@@
 outlier region containing MHC Class I Classical and MHC Class III gene
-regions. The plot was created with SimpleSynteny (95) using
-\emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank accession
-GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices indicate
-the location of SNP outliers in the TCF19 and VARS1 genes. Gap sizes for
-each contig representation are labeled.}
+regions. The plot was created with SimpleSynteny \citep{veltri2016}
+using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
+accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
+indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
+sizes for each contig representation are labeled.}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/het_by_status_violin_plot.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/violin_plot_heterozy_by_group.png}
 
@@
 their corresponding site number (from
-Fig. 1). A Wilcoxon signed-rank test showed
-no significant difference between the two groups.}
+Fig. 1).}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_site.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_genomewide_pi_by_pop.png}
 
@@
 
-\caption{\label{fig-boxplot-pi}For each study population in the frog
-evolution study (from Fig. 1), boxplots
-showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
-along the genome. The y-axis has been log\textsubscript{10}-transformed
-for display purposes. Sites are color-coded as ``Bd-naive'' or
-``recovering'', and vertical dashed lines show the genetic breaks
-between frog populations as described in
-Figure~\ref{fig-selectionresults} A). Within the genetic clusters,
-mean \(\pi\) is significantly different between sites (p \textless0.001
-Wilcoxon rank-sum test).}
+\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each study
+population in the frog evolution study (from
+Fig. 1), boxplots showing nucleotide
+diversity (\(\pi\)) within 100kb sliding windows along the genome. The
+y-axis has been log\textsubscript{10}-transformed for display purposes.
+Sites are color-coded as ``Bd-naive'' or ``recovering'', and vertical
+dashed lines show the genetic breaks between frog populations as
+described in Figure~\ref{fig-selectionresults} A). Within the genetic
+clusters, mean \(\pi\) is significantly different between sites (p
+\textless0.001 Wilcoxon rank-sum test).}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/perwindow_persite_pi.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_windowpop.png}
 
@@
 
-\caption{\label{fig-barchart-pi-by-windowsite}For each study population
-in the frog evolution study, average \(\pi\) values by population
-calculated for each of the 10 outlier windows identified in the splined
-window analysis (A-J). Sites are color-coded as ``Bd-naive'' or
-``recovering'', and vertical dashed lines show the genetic breaks
-between frog populations as described in
-Figure~\ref{fig-selectionresults} A. Gene annotations for the outlier
-regions are as follows: (A) C6, C7, (B) DDX10, ZBTB24, (C) SULTR-like,
-TRAF3IP2, VGLL3, EXOC1, (D) FLOT1, TUBB2B, TUBB, MDC1, CCHCR1,
-TCF19\emph{, TRIM29, HSP70, LSM2, VARS,} (E) GCC2, CFAP251, PEG10, (F)
-ERO1A, GVINP1, (G) ERO1A, GVINP1, (H) PPP1R12A, TSPAN4, PAWR, MFRP, MAX,
-PPP6R3, (I) C6H5ORF22, PKS6, BSPRY, MPV17, and (J) CAD, ATRAID, GPN1.
-Note that windows F and G are in the same gene region. * by gene names
-indicate that an outlier SNP is also present in this gene . (K) Average
-\(\pi\) values of naive and recovering groups for each of the outlier
-windows shown in A-J. }
+\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
+windows identified by the overlapping splined windows for FST and
+\$\textbackslash pi\$, boxplots showing average \$\textbackslash pi\$
+for naive and recovering populations. The corresponding gene annotations
+for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
+SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
+TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
+(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
+BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
 

### 2023-12-22 doc/manuscript/translocation.qmd
Edits from final review of manuscript

@@
 In addition, resistance may be at least partially the result of natural selection at specific immune function genes, which show evidence for selection in recovering populations.
-The evolution of resistance and reintroductions of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
+The evolution of resistance and reintroduction of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
 
@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The donor populations included 2 of the 5 recovering populations included in the frog evolution study referenced above (@fig-allelefrequencies: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
+The donor populations included 2 of the 5 recovering populations used in the frog evolution study referenced above (@fig-allelefrequencies: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
@@
 To extend our inferences of population establishment beyond those possible from the site-specific CMR data, we developed a population viability model.
-Specifically, to test whether the observed yearly adult survival probabilities in translocated populations (provided in legend of @fig-viability B) were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
+Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
 We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
 
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of @fig-viability B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
@@
 Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
-In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 100), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
+In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 500), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
 Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
 Finally, there is no potential for frog dispersal between the 9 study populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-(See **Supporting Information - Frog evolution in response to Bd - Methods** for additional details regarding the study design.)
+(See **Supporting Information - Frog evolution in response to Bd - Study design** for additional details regarding the study design.)
 
@@
 Results from the individual variant and splined window analyses show that recovering populations have signatures of selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI; SI Datasets S1, S2).
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI).
 One of the 7 identified genes (LOC108802036) does not have an associated annotation.
@@
 Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions (Datasets S3, S4).
+A total of 35 annotated genes were found in the 9 shared regions.
 Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
@@
 
-Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information-Frog evolution in response to Bd** for details).
+Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog evolution in response to Bd - Genetic diversity** for details).
 Thus, despite localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
-In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information-Frog evolution in response to Bd** for details).
-
-In summary, our genomic results indicate that the exomes of frogs from naive and recovering study populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
+In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog evolution in response to Bd** **- GO analysis** for details).
+
+In summary, our genomic results indicate that the exomes of frogs from naive and recovering populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
 The regions under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
@@
 Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
-In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
+In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.
 
@@
 These analysis are based on samples collected from virtually all of the MYL frog populations remaining in a naive state, as well as adjacent recovering populations.
-This produced genetic clusters that contained at least one naive and one recovering population, allowing us to detect selection without the confounding effects of population structure.
-Finally, we did not find a reduction in overall genetic variation in the recovering populations, suggesting that despite localized selection in the genome, these populations retain adequate genetic diversity for long-term persistence.
-
-Importantly, some regions that we identified as under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
+This study design produced genetic clusters that each contained at least one naive and one recovering population, allowing us to detect selection without the confounding effects of population structure.
+In addition, we did not find a reduction in overall genetic variation in the recovering populations, suggesting that despite localized selection in the genome, these populations retain adequate genetic diversity for long-term persistence.
+
+Importantly, some genomic regions that we identified as under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
 For example, the MHC plays an important role in immunity.
@@
 Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
-This gene is associated with inflammation response and in *Xenopus* is highly expressed in wounds [@fukazawa2009].
+This gene is associated with immune response and in *Xenopus* is expressed during appendage regeneration [@fukazawa2009].
 Finally, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
@@
 Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10<!--# Why not use 0.09?-->).
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year-1 juvenile survival probability was greater than 0.10<!--# Why not use 0.09?-->).
 The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the years covered by our post-translocation surveys.
@@
 This extraordinary impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
-Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
+Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other amphibian species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
 As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
@@
 For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
-The populations that we selected contained hundreds to thousands of both *R. sierrae* adults and tadpoles.
+The populations that we selected contained hundreds of *R. sierrae* adults and thousands of tadpoles.
 These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see **Supporting Information-Frog population recovery - Laboratory methods** for details].
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see **Supporting Information - Frog population recovery - Laboratory methods** for details].
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-Additional details are provided in **Supporting Information - Population viability modeling**.
+Additional details are provided in **Supporting Information - Population viability modeling - Incorporating yearly variability in survival rates** and **Estimating model parameters**.
 
@@
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-Additional details are provided in **Supporting Information - Population viability modeling**.
+Additional details are provided in **Supporting Information - Population viability modeling - Model analysis and simulation**.
 Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
@@
 We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and **Supporting Information**: Dataset S1, S2).
+We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and **Supporting Information - Frog evolution in response to Bd - GO analysis;** Dataset S1, S2).
 
@@
 
-![Results from population viability analysis: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when year 1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
+![Results from population viability analysis: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.
@@
 
-#### Laboratory Methods
+#### Laboratory methods
 
@@
 The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$.
-For unintroduced individuals which naturally recruit, this matrix is given by:
+For non-introduced (i.e., naturally recruited) individuals, this matrix is given by:
 
@@
 
-To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters.
+To complete the specification of the Bayesian model, we specify priors for all unknown parameters.
 The recruitment parameter priors were specified as follows:
@@
 
-#### Population viability modeling
-
-##### Incorporating yearly variability in vital rates
+### Population viability modeling
+
+#### Incorporating yearly variability in vital rates
 
@@
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of year 1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year 2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information-Population viability modeling-Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
-
-##### Estimating model parameters
+Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
+We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information - Population viability modeling - Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
+
+#### Estimating model parameters
 
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Estimation of frog survival and abundance** and **Supporting Information - Frog population recovery - Estimation of frog survival and abundance** for details).
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Frog populaton recovery - Estimation of frog survival and abundance** and **Supporting Information - Frog population recovery - CMR model structure** for details).
 For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs SI).
@@
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information-Frog population recovery-CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
 .
@@
 
-##### Model analysis and simulation
+#### Model analysis and simulation
 
 We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in year 1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
+First, we considered a deterministic version of our model with no yearly heterogeneity in year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
 We then fixed $\sigma_{J_1} = 0.09$ and calculated the predicted growth rate of our 12 populations.
@@
 However, increasing yearly variability in $\sigma_{J_1}$ increases extinction risk for all populations.
-For example, if we set $\phi = 0.001$, such that in a given year essentially either all year 1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
+For example, if we set $\phi = 0.001$, such that in a given year essentially either all year-1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
 Because we do not pit tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
@@
 
-To gain insights into the role of evolution in the development of resistance by MYL frogs, we compared frog genomes sampled in naive versus recovering populations.
+To gain insights into the role of evolution in the development of resistance by MYL frogs, we compared frog exomes sampled in naive versus recovering populations.
 Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
 The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
+As a result, samples from recovering populations collected before and after Bd exposure are not available and are unlikely to be available in the future.
 
@@
 We then conducted a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] to compare the 30 unique outlier genes to the complete set of genes in our target transcriptome.
-We repeated this process for the set of 33 genes located in the *F~ST~* and $\pi_{diff}$ splined windows.
+We repeated this process for the set of 35 genes located in the 9 shared regions of the *F~ST~* and $\pi_{diff}$ splined windows.
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling -- Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
+All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling - Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
 
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year 1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year 2 juveniles).
+Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles).
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
@@
 (A) PCA calculated from binary SNPs showing the genomic relationship of samples.
-Numeric labels and colors match those in @fig-allelefrequencies A.
+Numeric labels and colors match those in @fig-allelefrequencies.
 Populations 1-7 are *R.
@@
 |:-------------------------|:------------------|:--------------------------|
-| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
+| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
+| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
+| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
+| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
 | $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies                                                        |
 | $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from CMR studies                                                        |
-| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge                |
-| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
-| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
+| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge                |
+| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
+| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
 | $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                                |

### 2023-12-22 doc/manuscript/translocation.qmd
Adjust numbering in conversion script

--- 2023-12-21 doc/manuscript/translocation.qmd
@@
 
-![For each of the 9 outlier windows identified by the overlapping splined windows for FST and \$\\pi\$, boxplots showing average \$\\pi\$ for naive and recovering populations.
+![For each of the 9 outlier windows identified by the overlapping splined windows for FST and $\pi$, boxplots showing average $\pi$ for naive and recovering populations.
 The corresponding gene annotations for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_windowpop.png){#fig-boxplot-pi-by-windowpop width="500"}

### 2023-12-22 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -163,8 +163,11 @@ pnas_ms = pnas_ms.replace("The dynamics are given by", "The dynamics are given b
 # Clean up references to SI material
 
 supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S6A",
+			"Figure~\\ref{fig-selectionresults} A SI": "Figure S6A",
+			"Figure~\\ref{fig-selectionresults} A\nSI": "Figure S6A",
 			"Figure~\\ref{fig-selectionresults} A": "Figure S6A",
 			"Figure~\\ref{fig-selectionresults} B, C: SI": "Figure S6B, C",
+			"Figure~\\ref{fig-selectionresults} B, C SI": "Figure S6B, C",
 			"Figure~\\ref{fig-selectionresults} B, C": "Figure S6B, C",
 			"Figure~\\ref{fig-selectionresults} B,\nC": "Figure S6B, C",
 			"Figure~\\ref{fig-selectionresults} B, C;\nSI": "Figure S6B, C",
@@ -186,13 +189,14 @@ supp_map = {"Figure~\\ref{fig-selectionresults} A: SI": "Figure S6A",
 			"Figure~\\ref{fig-compare_surv_probs} SI": "Figure S4",
 			"Figure~\\ref{fig-compare_surv_probs}": "Figure S4",
 			"Figure~\\ref{fig-viability-supp} SI": "Figure S5",
-			"Table~\\ref{tbl-survival-earlylate} : SI": "Table S1",
-			"Table~\\ref{tbl-survival-earlylate} SI": "Table S1",
-			"Table~\\ref{tbl-survival-earlylate}": "Table S1",
-			"Table~\\ref{tbl-param_values} : SI": "Table S2",
-			"Table~\\ref{tbl-param_values}": "Table S2",
-			"Table~\\ref{tbl-param_values}\n: SI": "Table S2",
-			"Table~\\ref{tbl-param_values} SI": "Table S2"
+			# "Table~\\ref{tbl-survival-earlylate} : SI": "Table S1",
+			# "Table~\\ref{tbl-survival-earlylate} SI": "Table S1",
+			# "Table~\\ref{tbl-survival-earlylate}": "Table S1",
+			"Table~\\ref{tbl-param_values} : SI": "Table S1",
+			"Table~\\ref{tbl-param_values}\n: SI": "Table S1",
[…]

### 2023-12-22 doc/manuscript/convert_qmd_to_pnas_latex.py
Remove temp files after conversion

@@ -314,7 +314,7 @@ with open("translocation_bioarxiv.tex", "w") as fout:
 	fout.writelines(bioarxiv_ms)
 
 # Remove transcloation.tex
-# subprocess.call(['rm', args[1]])
+subprocess.call(['rm', args[1]])
 subprocess.call(['cp', 'translocation_pnas_SI.tex', 'translocation_bioarxiv_SI.tex'])
 
 

### 2023-12-22 doc/manuscript/convert_qmd_to_pnas_latex.py
Fix spaces around datasets

@@ -126,9 +126,10 @@ methods = methods.replace("The\ndynamics are given by", "The\ndynamics are given
 
 methods = methods.replace("\\paragraph{", "\\section*{")
 methods = methods.replace("\\subparagraph{", "\\subsection*{")
-methods = methods.replace("Incorporating yearly variability in vital rates", "\\subsubsection*{Incorporating yearly variability in vital rates}")
-methods = methods.replace("Estimating model parameters", "\\subsubsection*{Estimating model parameters}")
-methods = methods.replace("Model analysis and simulation", "\\subsubsection*{Model analysis and simulation}")
+# methods = methods.replace("Incorporating yearly variability in vital rates", "\\subsubsection*{Incorporating yearly variability in vital rates}")
+# methods = methods.replace("Estimating model parameters", "\\subsubsection*{Estimating model parameters}")
+# methods = methods.replace("Model analysis and simulation", "\\subsubsection*{Model analysis and simulation}")
+methods = methods.replace("Dataset S1, S2", " Dataset S1, S2")
 
 pnas_ms = pnas_ms.replace("\\matmethods{}", "\n\n\\matmethods{\n" + methods + "\n\n}\n")
 
@@ -314,7 +315,7 @@ with open("translocation_bioarxiv.tex", "w") as fout:
 	fout.writelines(bioarxiv_ms)
 
 # Remove transcloation.tex
-subprocess.call(['rm', args[1]])
+# subprocess.call(['rm', args[1]])
 subprocess.call(['cp', 'translocation_pnas_SI.tex', 'translocation_bioarxiv_SI.tex'])
 
 

### 2023-12-23 doc/manuscript/translocation.qmd
Final version for resubmission

@@
 The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
-Given that there are no broadly applicable methods to reverse these impacts in the wild, the future of many amphibians appears grim.
+With no broadly applicable methods available to reverse these impacts in the wild, the future of many amphibians appears grim.
 The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
@@
 Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
-Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), we also conducted a genomic study to determine whether MYL frogs show an evolutionary response to Bd exposure, and whether these genomic changes are associated with resistance.
-To accomplish this objective, we used exome capture methods to identify genomic regions with consistent differences between frogs from naive versus recovering populations (@fig-recovery-model).
+Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), we conducted a genomic study using exome capture methods to determine whether MYL frogs in recovering populations show evidence of selection and whether these genomic changes are associated with resistance (@fig-recovery-model).
 
@@
 Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
-Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976; @fig-translocation-survival, @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
+Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
 
@@
 ![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog evolution study.
-Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).
-](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="500"}
+Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="500"}
 
@@
 Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).
-Within the genetic clusters, mean $\pi$ is significantly different between sites (p \<0.001 Wilcoxon rank-sum test).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
+](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
 

### 2023-12-23 doc/manuscript/translocation.tex
Compile versions for journal and preprint server

--- 2023-12-21 doc/manuscript/translocation_bioarxiv.tex
@@
-\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
-% Use the lineno option to display guide line numbers if required.
-
-\templatetype{pnasmathematics} % Choose template 
-% {pnasmathematics} = Template for a two-column research article
-% {pnasmathematics} %= Template for a one-column mathematics article
-% {pnasinvited} %= Template for a PNAS invited submission
-
-\title{Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease}
-
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,b]{Roland A. Knapp}
-\author[c,1]{Mark Q. Wilber} 
-\author[d,e,1]{Allison Q. Byrne}
-\author[f,g]{Maxwell B. Joseph}
-\author[a,b]{Thomas C. Smith}
-\author[d,e,h]{Andrew P. Rothstein}
-\author[i]{Robert L. Grasso}
-\author[d,e]{Erica Bree Rosenblum}
-
-\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
-\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
-\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
-\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
-\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
-\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
-\affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
-\affil[h]{Current affiliation: Ginkgo Bioworks, Boston, MA, 02210}
-\affil[i]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}
-
-% Please give the surname of the lead author for the running footer
-\leadauthor{Knapp}
-
-% Please add a significance statement to explain the relevance of your work
-\significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. Emerging infectious
-diseases can have particularly devastating impacts, and few options
-exist to reverse these effects. We used large-scale reintroductions of
-disease-resistant individuals in an effort to recover a once-common frog
-species driven to near-extinction by a disease that has decimated
-amphibian biodiversity. Introduction of resistant frogs allowed
-reestablishment of viable populations in the presence of disease. In
-addition, resistance may be at least partially the result of natural
-selection at specific immune function genes, which show evidence for
-selection in recovering populations. The evolution of resistance and
-reintroductions of resistant individuals could play an important role in
-biodiversity conservation in our rapidly changing world.}
-
-% Please include corresponding author, author contribution and author declaration information
-\authorcontributions{}
-\authordeclaration{}
-\equalauthors{\textsuperscript{1}M.W. contributed equally to this work with A.B.}
-\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu}
-
-% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{}
-
-\begin{abstract}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
+
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{blkarray}
+\KOMAoption{captions}{tableheading}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\author{}
+\date{}
+
+\begin{document}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+
+\hypertarget{title-reintroduction-of-resistant-frogs-facilitates-landscape-scale-recovery-in-the-presence-of-a-lethal-fungal-disease}{%
+\subsection{Title: Reintroduction of resistant frogs facilitates
+landscape-scale recovery in the presence of a lethal fungal
+disease}\label{title-reintroduction-of-resistant-frogs-facilitates-landscape-scale-recovery-in-the-presence-of-a-lethal-fungal-disease}}
+
+\hypertarget{authors}{%
+\subsubsection{Authors:}\label{authors}}
+
+Roland A. Knapp\textsuperscript{a,b,}*, Mark Q.
+Wilber\textsuperscript{c,1}, Allison Q. Byrne\textsuperscript{d,e,1},
+Maxwell B. Joseph\textsuperscript{f,g}, Thomas C.
+Smith\textsuperscript{a,b}, Andrew P. Rothstein\textsuperscript{d,e,h},
+Robert L. Grasso\textsuperscript{i}, Erica Bree
+Rosenblum\textsuperscript{d,e}
+
+\hypertarget{affiliations}{%
+\subsubsection{Affiliations:}\label{affiliations}}
+
+\textsuperscript{a}Sierra Nevada Aquatic Research Laboratory, University
+of California, Mammoth Lakes, CA, 93546
+
+\textsuperscript{b}Earth Research Institute, University of California,
+Santa Barbara, CA, 93106-3060
+
+\textsuperscript{c}School of Natural Resources, University of Tennessee
+Institute of Agriculture, Knoxville, TN, 37996
+
+\textsuperscript{d}Department of Environmental Science, Policy, and
+Management, University of California - Berkeley, Berkeley, CA,
+94720-3114
+
+\textsuperscript{e}Museum of Vertebrate Zoology, University of
+California - Berkeley, Berkeley, CA, 94720-3160
+
+\textsuperscript{f}Earth Lab, University of Colorado, Boulder, CO, 80303
+
+\textsuperscript{g}Current affiliation: Planet, San Francisco, CA, 94107
+
+\textsuperscript{h}Current affiliation: Ginkgo Bioworks, Boston, MA,
+02210
+
+\textsuperscript{i}Resources Management and Science, Yosemite National
+Park, El Portal, CA, 95318
+
+*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016
+Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034,
+email: roland.knapp@ucsb.edu
+
+\textsuperscript{1}These authors contributed equally to this work.
+
+\textbf{Classification:} Biological Sciences - Ecology
+
+\textbf{Key words:} evolution, resistance, amphibians, population
+recovery, reintroduction, disease, \emph{Batrachochytrium dendrobatidis}
+
+\hypertarget{abstract}{%
+\subsubsection{Abstract}\label{abstract}}
+
 Vast alteration of the biosphere by humans is causing a sixth mass
@@
 amphibian biodiversity, with hundreds of species experiencing declines
-or extinctions. Given that there are no broadly applicable methods to
-reverse these impacts in the wild, the future of many amphibians appears
-grim. The once-common mountain yellow-legged (MYL) frog is emblematic of
+or extinctions. With no broadly applicable methods available to reverse
+these impacts in the wild, the future of many amphibians appears grim.
+The once-common mountain yellow-legged (MYL) frog is emblematic of
 amphibians threatened by Bd. Although most MYL frog populations are
@@
 with extinction by novel pathogens.
-\end{abstract}
-
-\dates{This manuscript was compiled on \today}
-% \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\begin{document}
-
-\maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-\dropcap{H}uman activities are increasingly impacting global biodiversity
+
+\hypertarget{significance}{%
+\subsubsection{Significance}\label{significance}}
+
+Understanding how species persist despite accelerating global change is
+critical for the conservation of biodiversity. Emerging infectious
+diseases can have particularly devastating impacts, and few options
+exist to reverse these effects. We used large-scale reintroductions of
+disease-resistant individuals in an effort to recover a once-common frog
+species driven to near-extinction by a disease that has decimated
+amphibian biodiversity. Introduction of resistant frogs allowed
+reestablishment of viable populations in the presence of disease. In
+addition, resistance may be at least partially the result of natural
+selection at specific immune function genes, which show evidence for
+selection in recovering populations. The evolution of resistance and
+reintroduction of resistant individuals could play an important role in
+biodiversity conservation in our rapidly changing world.
+
+\hypertarget{introduction}{%
+\subsubsection{Introduction}\label{introduction}}
+
+Human activities are increasingly impacting global biodiversity
 \citep{ceballos2015}, with important implications for ecosystem
@@
 resistance for frog survival, population establishment, and long-term
-viability (this study), we also conducted a genomic study to determine
-whether MYL frogs show an evolutionary response to Bd exposure, and
-whether these genomic changes are associated with resistance. To
-accomplish this objective, we used exome capture methods to identify
-genomic regions with consistent differences between frogs from naive
-versus recovering populations (Figure~\ref{fig-recovery-model}).
-
-\section*{Results}
-
-\subsection*{Frog population recovery}
+viability (this study), we conducted a genomic study using exome capture
+methods to determine whether MYL frogs in recovering populations show
+evidence of selection and whether these genomic changes are associated
+with resistance (Figure~\ref{fig-recovery-model}).
+
+\hypertarget{results}{%
+\subsubsection{Results}\label{results}}
+
+\hypertarget{frog-population-recovery}{%
+\paragraph{Frog population recovery}\label{frog-population-recovery}}
 
@@
 ``donor'' populations and translocating them to 1 of 12 nearby recipient
-sites (Figure S1). The donor populations included 2 of
-the 5 recovering populations included in the frog evolution study
-referenced above (Figure~\ref{fig-allelefrequencies}: population 1 and
-4), and these donor populations contributed frogs for 20 of the 24
+sites (Figure~\ref{fig-yosemap} SI). The donor populations included 2 of
+the 5 recovering populations used in the frog evolution study referenced
+above (Figure~\ref{fig-allelefrequencies}: population 1 and 4), and
+these donor populations contributed frogs for 20 of the 24
 translocations. Following translocation, we estimated adult survival and
@@
 chytridiomycosis (i.e., the disease caused by Bd) and associated frog
-mortality (Figure S2)
+mortality (Figure~\ref{fig-bdload-beforeafter} SI)
 \citep{joseph2018, vredenburg2010}. Although it is possible that the
@@
 little difference in pre- versus post-translocation Bd loads even in
-those populations that had very high frog survival (70556, 74976;
-Figure~\ref{fig-translocation-survival},
-Figure S2) suggests a true lack of
+those populations that had very high frog survival (70556, 74976 - see
+below; Figure~\ref{fig-bdload-beforeafter} SI) suggests a true lack of
 substantial change in Bd load.
@@
 predictors, but Bd load at the time of translocation was not among them
-(Figure S3). Instead, important predictors
+(Figure~\ref{fig-survival-postdens} SI). Instead, important predictors
 included winter severity in the year following translocation (snow\_t1),
 site elevation, and donor population (Figure~\ref{fig-cond-effects},
-Figure S3). Males had somewhat higher
+Figure~\ref{fig-survival-postdens} SI). Males had somewhat higher
 survival than females, but this effect was small
-(Figure~\ref{fig-cond-effects}, Figure S3).
+(Figure~\ref{fig-cond-effects}, Figure~\ref{fig-survival-postdens} SI).
 The absence of Bd load as an important predictor of frog survival is
@@
 or higher survival probabilities than the originally translocated adults
-(Figure S4). This suggests that frog
+(Figure~\ref{fig-compare_surv_probs} SI). This suggests that frog
 resistance is maintained across generations. All of the conditions
@@
 
-\subsection*{Long-term population viability}
+\hypertarget{long-term-population-viability}{%
+\paragraph{Long-term population
+viability}\label{long-term-population-viability}}
 
@@
 test whether the observed yearly adult survival probabilities in
-translocated populations (provided in legend of
-Figure~\ref{fig-viability}B) were sufficient for long-term viability,
-we built a stage-structured matrix model that captured known frog
+translocated populations were sufficient for long-term viability, we
+built a stage-structured matrix model that captured known frog
 demography and included demographic and environmental stochasticity. We
 parameterized the model using CMR data from translocated populations and
-known life history values in this system (Table S2
+known life history values in this system (Table~\ref{tbl-param_values}
 SI).
@@
 Given observed yearly adult survival probabilities of translocated frogs
-(from site-specific mrmr CMR models) and a yearly survival probability
-of the year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at
-least six of twelve translocated populations should experience a
-long-run growth rate \(\lambda\) greater than 1 in the presence of Bd
-(Figure~\ref{fig-viability}A; median predicted \(\lambda\) ranges from
+(from site-specific mrmr CMR models; provided in legend of
+Figure~\ref{fig-viability} B) and a yearly survival probability of the
+year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at least six
+of twelve translocated populations should experience a long-run growth
+rate \(\lambda\) greater than 1 in the presence of Bd
+(Figure~\ref{fig-viability} A; median predicted \(\lambda\) ranges from
 1.19-1.40 for these six populations). These six populations all had
@@
 growth rate of eight of twelve population was greater than 1
-(Figure~\ref{fig-viability}A).
+(Figure~\ref{fig-viability} A).
 
@@
 probabilities of less than 0.5 when the average year-1 juvenile survival
-was greater than 0.10 (Figure~\ref{fig-viability}B). This indicates
+was greater than 0.10 (Figure~\ref{fig-viability} B). This indicates
 strong potential for long-term persistence in the presence of Bd and
@@
 describe the general recovery trajectory of our translocated population
-with the longest survey history (Figure~\ref{fig-viability}C;
+with the longest survey history (Figure~\ref{fig-viability} C;
 population 70550, surveyed for 16 years).
@@
 survival probability than translocated adults
-(Figure S4), but we considered these
+(Figure~\ref{fig-compare_surv_probs} SI), but we considered these
 probabilities to be equal in all but three of our populations where we
@@
 
-\subsection*{Frog evolution in response to Bd}
+\hypertarget{frog-evolution-in-response-to-bd}{%
+\paragraph{Frog evolution in response to
+Bd}\label{frog-evolution-in-response-to-bd}}
 
@@
 enzootic state \citep{briggs2010}, characterized by smaller numbers of
-adults (generally \textless{} 100), high Bd prevalence \citep[often
+adults (generally \textless{} 500), high Bd prevalence \citep[often
 \textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
@@
 fragmentation. (See \textbf{Supporting Information - Frog evolution in
-response to Bd - Methods} for additional details regarding the study
-design.)
+response to Bd - Study design} for additional details regarding the
+study design.)
 
@@
 Individual frogs clustered into 3 separate groups in principal component
-space (Figure S6A SI), and clusters reflected
+space (Figure~\ref{fig-selectionresults} A SI), and clusters reflected
 the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
@@
 frequencies between naive versus recovering populations) from 7 distinct
-genes across 4 contigs (Figure S6B, C SI; SI
-Datasets S1, S2). One of the 7 identified genes (LOC108802036) does not
-have an associated annotation. For the outlier SNPs, frequency
-differences between the naive and recovering populations ranged from
-0.41 to 0.86. Most of these SNPs showed frequency differences in only a
-subset of the sampled populations (Figure~\ref{fig-allelefrequencies}A,
-B), but the SNP in the RIN3 gene showed consistent differences in
-frequencies across all populations (Figure~\ref{fig-allelefrequencies}
-C). This is suggestive of parallel selection at this locus across
-multiple populations. The other 6 outlier variants showed less
-consistent frequency differences across the study populations, but for
-these we still found a statistically significant signal of selection in
-2 of the 3 genetic clusters (containing populations 5--9;
-Figure~\ref{fig-allelefrequencies}, Figure S6A
+genes across 4 contigs (Figure~\ref{fig-selectionresults} B, C SI). One
+of the 7 identified genes (LOC108802036) does not have an associated
+annotation. For the outlier SNPs, frequency differences between the
+naive and recovering populations ranged from 0.41 to 0.86. Most of these
+SNPs showed frequency differences in only a subset of the sampled
+populations (Figure~\ref{fig-allelefrequencies} A, B), but the SNP in
+the RIN3 gene showed consistent differences in frequencies across all
+populations (Figure~\ref{fig-allelefrequencies} C). This is suggestive
+of parallel selection at this locus across multiple populations. The
+other 6 outlier variants showed less consistent frequency differences
+across the study populations, but for these we still found a
+statistically significant signal of selection in 2 of the 3 genetic
+clusters (containing populations 5--9;
+Figure~\ref{fig-allelefrequencies}, Figure~\ref{fig-selectionresults} A
 SI). Therefore, although some outlier variant associations have a more
@@
 \(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
-(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
+(Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
 outliers for both metrics (``shared regions'') and 2 of these shared
 regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions
-(Datasets S3, S4). Given this large number of genes, here we focus on
-those with the strongest signal of selection and/or immune-related
-functions. The largest \(\pi_{diff}\), indicative of directional
-selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
-the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C). This region
-contains approximately 500 SNPs and one annotated gene called
+A total of 35 annotated genes were found in the 9 shared regions. Given
+this large number of genes, here we focus on those with the strongest
+signal of selection and/or immune-related functions. The largest
+\(\pi_{diff}\), indicative of directional selection, occurred in a 163kb
+region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP
+(Figure~\ref{fig-spline-manhattan} C). This region contains
+approximately 500 SNPs and one annotated gene called
 ``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
@@
 was within 360kb of another outlier SNP (VARS)
-(Figure~\ref{fig-spline-manhattan}D, Figure S7).
+(Figure~\ref{fig-spline-manhattan} D, Figure~\ref{fig-synteny-plot} SI).
 This region (854kb from the beginning of the outlier window to the VARS
@@
 individual-level heterozygosity or population-level \(\pi\) between
-naive and recovering populations (see \textbf{Supporting
-Information-Frog evolution in response to Bd} for details). Thus,
-despite localized selection in particular regions of the genome, we did
-not find evidence for reduced genetic diversity across the genome in
-recovering populations. In addition, no gene ontology (GO) biological
+naive and recovering populations (see \textbf{Supporting Information -
+Frog evolution in response to Bd - Genetic diversity} for details).
+Thus, despite localized selection in particular regions of the genome,
+we did not find evidence for reduced genetic diversity across the genome
+in recovering populations. In addition, no gene ontology (GO) biological
 functions, molecular functions, or cellular processes were
-over-represented in either the outlier variants or the 33 genes located
+over-represented in either the outlier variants or the 35 genes located
 in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined
-windows (see \textbf{Supporting Information-Frog evolution in response
-to Bd} for details).
+windows (see \textbf{Supporting Information - Frog evolution in response
+to Bd} \textbf{- GO analysis} for details).
 
 In summary, our genomic results indicate that the exomes of frogs from
-naive and recovering study populations show substantial differences,
+naive and recovering populations show substantial differences,
 consistent with parallel evolutionary changes following Bd exposure. The
@@
 
-\section*{Discussion}
+\hypertarget{discussion}{%
+\subsubsection{Discussion}\label{discussion}}
 
@@
 reestablishing MYL frog populations via translocation of resistant
-individuals is striking, and even more so given that this success
-occurred despite ongoing infection of frogs with Bd.
+individuals is striking, and even more so given that MYL frogs were
+driven to near-extinction by Bd.
 
@@
 low-to-moderate and relatively constant Bd loads and mass die-offs are
-not observed \citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
+not observed \citep[see also Figure~\ref{fig-bdload-beforeafter}
+SI]{briggs2010, knapp2011}. The differences in Bd load of frogs from
 naive and recovering populations are also observed in controlled
@@
 remaining in a naive state, as well as adjacent recovering populations.
-This produced genetic clusters that contained at least one naive and one
-recovering population, allowing us to detect selection without the
-confounding effects of population structure. Finally, we did not find a
-reduction in overall genetic variation in the recovering populations,
-suggesting that despite localized selection in the genome, these
-populations retain adequate genetic diversity for long-term persistence.
-
-Importantly, some regions that we identified as under selection are
-associated with cellular and immunological mechanisms known to
+This study design produced genetic clusters that each contained at least
+one naive and one recovering population, allowing us to detect selection
+without the confounding effects of population structure. In addition, we
+did not find a reduction in overall genetic variation in the recovering
+populations, suggesting that despite localized selection in the genome,
+these populations retain adequate genetic diversity for long-term
+persistence.
+
+Importantly, some genomic regions that we identified as under selection
+are associated with cellular and immunological mechanisms known to
 contribute to disease resistance, including in amphibians
@@
 this study, indicating consistent selection in populations distributed
-across a wide geographic area. This gene is associated with inflammation
-response and in \emph{Xenopus} is highly expressed in wounds
-\citep{fukazawa2009}. Finally, the outlier variant with the lowest
-p-value was the uncharacterized gene LOC108802036. In the genome of
-another frog species, this gene is located adjacent to a type I
+across a wide geographic area. This gene is associated with immune
+response and in \emph{Xenopus} is expressed during appendage
+regeneration \citep{fukazawa2009}. Finally, the outlier variant with the
+lowest p-value was the uncharacterized gene LOC108802036. In the genome
+of another frog species, this gene is located adjacent to a type I
 interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
@@
 probability of extinction over 50 years (probability of extinction \(<\)
-0.5 when year 1 juvenile survival probability was greater than 0.10).
+0.5 when year-1 juvenile survival probability was greater than 0.10).
 The viability model highlighted the important role of frog survival in
@@
 survival of adults and may have influenced population establishment
-(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
+(sites referred to as ``Alpine'' and ``Subalpine'' in \citet{joseph2018}
 are identified as ``70550'' and ``70505'' in the current study).
@@
 \citep{garner2016, knapp2022}. Importantly, in addition to the natural
-recovery documented for MYL frogs \citep{knapp2016}, other species are
-also showing evidence of post-epizootic recovery in the presence of Bd
-\citep{scheele2017, voyles2018} and suggest the possibility of also
-using animals from these recovering populations to reestablish
-extirpated populations. As with MYL frogs, the feasibility and long-term
-success of such efforts will depend on the availability of robust donor
-populations containing individuals that have the adaptive alleles
-necessary to allow frog survival and population growth in the presence
-of Bd. Despite the hopeful example of successful reestablishment of MYL
-frogs despite ongoing Bd infection, the challenge of recovering hundreds
-of Bd-impacted amphibian species globally is a daunting prospect.
-Although we now have a proven strategy to reestablish extirpated MYL
-frog populations, recovery across their large historical range will
-require substantial resources over many decades. The results of this
-study provide a hopeful starting point for that endeavor and other
-future efforts worldwide.
+recovery documented for MYL frogs \citep{knapp2016}, other amphibian
+species are also showing evidence of post-epizootic recovery in the
+presence of Bd \citep{scheele2017, voyles2018} and suggest the
+possibility of also using animals from these recovering populations to
+reestablish extirpated populations. As with MYL frogs, the feasibility
+and long-term success of such efforts will depend on the availability of
+robust donor populations containing individuals that have the adaptive
+alleles necessary to allow frog survival and population growth in the
+presence of Bd. Despite the hopeful example of successful
+reestablishment of MYL frogs despite ongoing Bd infection, the challenge
+of recovering hundreds of Bd-impacted amphibian species globally is a
+daunting prospect. Although we now have a proven strategy to reestablish
+extirpated MYL frog populations, recovery across their large historical
+range will require substantial resources over many decades. The results
+of this study provide a hopeful starting point for that endeavor and
+other future efforts worldwide.
 
@@
 
-\matmethods{
+\hypertarget{materials-and-methods}{%
+\subsubsection{Materials and Methods}\label{materials-and-methods}}
+
 \hypertarget{frog-population-recovery-1}{%
-\section*{Frog population recovery}\label{frog-population-recovery-1}}
+\paragraph{Frog population recovery}\label{frog-population-recovery-1}}
 
 \hypertarget{field-methods}{%
-\subsection*{Field methods}\label{field-methods}}
+\subparagraph{Field methods}\label{field-methods}}
 
@@
 \citep{poorten2017}. The populations that we selected contained hundreds
-to thousands of both \emph{R. sierrae} adults and tadpoles. These
-relatively high abundances were the result of recent increases following
-previous Bd-caused declines \citep{knapp2016}. As is typical for
-recovering MYL frog populations, Bd prevalence in the donor populations
-was high (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
+of \emph{R. sierrae} adults and thousands of tadpoles. These relatively
+high abundances were the result of recent increases following previous
+Bd-caused declines \citep{knapp2016}. As is typical for recovering MYL
+frog populations, Bd prevalence in the donor populations was high
+(0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
 3.06--3.78 ITS copies) was two or more orders of magnitude below the
@@
 We conducted 1--4 translocations per site
-(Figure~\ref{fig-translocation-survival}, Figure S1)
+(Figure~\ref{fig-translocation-survival}, Figure~\ref{fig-yosemap} SI)
 and each translocated cohort included 18 to 99 frogs (median = 30). In
@@
 \hypertarget{frog-counts-and-reproductive-success}{%
-\subsection*{Frog counts and reproductive
+\subparagraph{Frog counts and reproductive
 success}\label{frog-counts-and-reproductive-success}}
@@
 for the multi-year tadpole and juvenile stages in MYL frogs
-(Table S2 SI).
+(Table~\ref{tbl-param_values} SI).
 
 \hypertarget{estimation-of-frog-survival-and-abundance}{%
-\subsection*{Estimation of frog survival and
+\subparagraph{Estimation of frog survival and
 abundance}\label{estimation-of-frog-survival-and-abundance}}
@@
 \hypertarget{predictors-of-post-translocation-frog-survival}{%
-\subsection*{Predictors of post-translocation frog
+\subparagraph{Predictors of post-translocation frog
 survival}\label{predictors-of-post-translocation-frog-survival}}
@@
 \hypertarget{changes-in-bd-load-following-translocation}{%
-\subsection*{Changes in Bd load following
+\subparagraph{Changes in Bd load following
 translocation}\label{changes-in-bd-load-following-translocation}}
@@
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see \textbf{Supporting Information-Frog population recovery -
+\citep[see \textbf{Supporting Information - Frog population recovery -
 Laboratory methods} for details]{boyle2004}. To assess the magnitude of
@@
 \hypertarget{population-viability-modeling}{%
-\section*{Population viability
+\paragraph{Population viability
 modeling}\label{population-viability-modeling}}
@@
 \hypertarget{model-description}{%
-\subsection*{Model description}\label{model-description}}
+\subparagraph{Model description}\label{model-description}}
 
@@
 probability of translocated adults and naturally recruited adults
-differs (Figure S4).
+differs (Figure~\ref{fig-compare_surv_probs} SI).
 
@@
 discrete-time, stage-structured model where a time step is one year. The
-dynamics are given by equation \ref{eqn:matrix}.
-
-\begin{figure*}\[
+dynamics are given by
+
+\[
 \begin{bmatrix}
@@
 A_T \\
-\end{bmatrix}(t)\numberthis \label{eqn:matrix} 
-\]\end{figure*}
+\end{bmatrix}(t) 
+\]
 
@@
 realism. Additional details are provided in \textbf{Supporting
-Information - Population viability modeling}.
+Information - Population viability modeling - Incorporating yearly
+variability in survival rates} and \textbf{Estimating model parameters}.
 
 \hypertarget{model-analysis}{%
-\subsection*{Model analysis}\label{model-analysis}}
+\subparagraph{Model analysis}\label{model-analysis}}
 
@@
 probabilities and other known vital rates
-(Table S2), we performed four analyses. First, we
+(Table~\ref{tbl-param_values}), we performed four analyses. First, we
 compute the long-run growth rate \(\lambda\) for each of our 12
@@
 quantify how much changes in these parameters affected the long-run
-growth rate (Figure S5). This also helped us
+growth rate (Figure~\ref{fig-viability-supp} SI). This also helped us
 determine where in the model environmental variation in juvenile
@@
 are provided in \textbf{Supporting Information - Population viability
-modeling}. Code to replicate the analyses can be found at
-\url{https://github.com/SNARL1/translocation}.
+modeling - Model analysis and simulation}. Code to replicate the
+analyses can be found at \url{https://github.com/SNARL1/translocation}.
 
 \hypertarget{frog-evolution-in-response-to-bd-1}{%
-\section*{Frog evolution in response to
+\paragraph{Frog evolution in response to
 Bd}\label{frog-evolution-in-response-to-bd-1}}
@@
 \hypertarget{sampling-and-sequencing}{%
-\subsection*{Sampling and sequencing}\label{sampling-and-sequencing}}
+\subparagraph{Sampling and sequencing}\label{sampling-and-sequencing}}
 
@@
 DNEasy manufacturer's protocols. We sequenced the samples using an exome
-capture assay as described in \citep{byrne2023}. Briefly, genomic
+capture assay as described in \citet{byrne2023}. Briefly, genomic
 libraries were prepared and captured using a custom Nimblegen capture
@@
 \hypertarget{data-pre-processing-and-cleaning}{%
-\subsection*{Data pre-processing and
+\subparagraph{Data pre-processing and
 cleaning}\label{data-pre-processing-and-cleaning}}
@@
 \hypertarget{data-analysis}{%
-\subsection*{Data analysis}\label{data-analysis}}
+\subparagraph{Data analysis}\label{data-analysis}}
 
@@
 set solely for the gene ontology (GO) analysis (see below and
-\textbf{Supporting Information}: Dataset S1, S2).
+\textbf{Supporting Information - Frog evolution in response to Bd - GO
+analysis;} Dataset S1, S2).
 
@@
 
-}
-
-\showmatmethods{} % Display the Materials and Methods section
-
-\acknow{We thank the following for important contributions to this study: E.
+\hypertarget{acknowledgments}{%
+\subsubsection{Acknowledgments}\label{acknowledgments}}
+
+We thank the following for important contributions to this study: E.
 Hegeman and A. Lindauer--project and data management, field and
@@
 Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
-DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
-
-\showacknow{} % Display the acknowledgments section
-
-% Bibliography
-\bibliography{translocation}
-
-\clearpage
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).
+
+\hypertarget{references}{%
+\subsubsection{References}\label{references}}
+
+See combined references at end of document (temporary placeholder).
+
+\hfill\break
+
+\hypertarget{figures}{%
+\subsubsection{Figures}\label{figures}}
+
+\begin{figure}
+
+{\centering \includegraphics{figures/conceptual_model.png}
 
@@
 
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
+\begin{figure}
+
+{\centering \includegraphics[width=4.6875in,height=\textheight]{figures/translocation_survival_bysiteid.png}
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
+\begin{figure}
+
+{\centering \includegraphics[width=5.72917in,height=\textheight]{figures/cond_effects_plot.png}
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+\begin{figure}
+
+{\centering \includegraphics[width=6.77083in,height=\textheight]{figures/pop_viability_figures_for_manuscript.jpg}
 
@@
 analysis: (A) Predicted long-run growth rate \(\lambda\) for different
-values of yearly adult survival probability \(\sigma_{A_R}\) and year 1
+values of yearly adult survival probability \(\sigma_{A_R}\) and year-1
 juvenile survival and recruitment probability \(\sigma_{J_1}\), given
@@
 predicted \(\lambda\) values for the twelve translocated populations
-when year 1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
+when year-1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
 (indicated by the dashed line). The red line shows where
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
+\begin{figure}
+
+{\centering \includegraphics[width=6.25in,height=\textheight]{figures/allele_maps.png}
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}
+\begin{figure}
+
+{\centering \includegraphics[width=5.72917in,height=\textheight]{figures/splinewindow_manhattan.png}
 
@@
 
-\end{figure*}
+\end{figure}
 
 \newpage
+
+\hypertarget{supporting-information}{%
+\subsection{Supporting Information}\label{supporting-information}}
+
+\hypertarget{frog-population-recovery-2}{%
+\subsubsection{Frog population
+recovery}\label{frog-population-recovery-2}}
+
+\hypertarget{laboratory-methods}{%
+\paragraph{Laboratory methods}\label{laboratory-methods}}
+
+Swab extracts were analyzed using standard Bd DNA extraction and qPCR
+methods \citep{boyle2004}, and extracts were analyzed singly instead of
+in triplicate \citep{kriger2006}. For analysis of swabs collected during
+2005--2014, we used standards developed from known concentrations of
+zoospores \citep{boyle2004}, and after 2014, we used standards based on
+single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
+between samples analyzed using both types of standards, Bd in the study
+area has an average of 60 ITS1 copies per zoospore. To express all qPCR
+results as the number of ITS1 copies, starting quantities obtained using
+the zoospore standard (measured as ``zoospore equivalents'') were
+multiplied by 60. In addition, all qPCR quantities (regardless of
+standard) were multiplied by 80 to account for the fact that DNA
+extracts from swabs were diluted 80-fold during extraction and PCR
+\citep{vredenburg2010}.
+
+\hypertarget{cmr-model-structure}{%
+\paragraph{CMR model structure}\label{cmr-model-structure}}
+
+We estimated survival and recruitment for each site using open
+population CMR models based on \citet{joseph2018}. For each individual
+\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
+the individual was not detected, and \(o_{i, j}=2\) if the individual
+was detected. Capture histories of \(M\) individuals are modeled,
+although only \(N_s\) individuals were captured. This parameter expanded
+data augmentation allows us to capture the possibility that undetected
+individuals may have recruited into the adult population
+\citep{royle2012}. Here, \(M\) was chosen to be three times the number
+of observed individuals (\(3N_s\)) to be considerably greater than our
+prior guess of \(N_s\).
+
+We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
+period \(t = 1,..., n_t\). The four states that we consider are:
+\(u_{i, t} = 1\) for individuals that have not recruited,
+\(u_{i, t} =2\) for live adults, and \(u_{i, t} = 3\) for dead adults.
+Each survey \(j=1, ..., n_j\) occurs in one of the \(n_t\) primary
+periods, and we denote the primary period in which survey \(j\) takes
+place as \(t_j\). Each primary period \(t\) occurs within one year, but
+within a year there can be multiple primary periods. We set the year
+containing the first primary period to \(y_{t = 1} = 1\), and generally
+\(y_t\) represents the year containing primary period \(t\). Years
+increment by one until the final year of the mark recapture efforts,
+which we denote \(n_y\): \(y \in \{1, 2, ..., n_y\}\). We assume that
+within a primary period, the state of each individual does not change
+(i.e., individuals do not recruit into the adult population, gain or
+lose Bd infection, or die). This assumption is justified by the short
+time intervals between surveys within primary periods, in cases where
+primary periods contain multiple secondary periods.
+
+Live individuals are detected with probability \(p_j\), which is modeled
+as:
+
+\[p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),\]
+
+where \(X_j^{(p)}\) is a known row vector and \(\beta^{(p)}\) an unknown
+parameter vector. Not recruited and dead individuals are never captured.
+We bundle these assumptions about the observation probabilities for
+survey \(j\) into an emission matrix \(\Omega_j\):
+
+\[
+\Omega_j =
+\begin{blockarray}{ccc}
+  \text{Not detected} & \text{Detected} \\
+\begin{block}{(cc)c}
+  1 & 0 & \text{Not recruited} \\
+  1 - p_j & p_j & \text{Alive} \\
+  1 & 0 & \text{Dead} \\
+\end{block}
+\end{blockarray}
+\]
+
+The state transition matrix \(\Psi_{t, i}\) contains the probabilities
+of individual \(i\) transitioning from state \(u_{i, t}\) (rows) to
+\(u_{i, t+1}\) (columns) between primary period \(t\) and \(t+1\). For
+non-introduced (i.e., naturally recruited) individuals, this matrix is
+given by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(\lambda_t\) is the probability of recruiting in time \(t\)
+and \(\phi_t\) is the probability of survival in time \(t\).
+
+For introduced individuals, which have deterministic recruitment (i.e.,
+they recruit when introduced), the state transition matrix is given by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 1 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(I_{t, i}\) is a known indicator function for whether
+individual \(i\) was introduced in primary period \(t\).
+
+We allow recruitment probabilities to vary in time via random effects,
+such that:
+
+\[\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),\]
+
+where \(\alpha^{(\lambda)}\) is an intercept parameter and
+\(\epsilon^{(\lambda)}_t\) is an adjustment for time \(t\).
+
+Survival probabilities also vary in time, and as a function of known
+covariates:
+
+\[\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),\]
+
+where \(X^{(\phi)}_t\) is a row vector of known covariates,
+\(\beta^{(\phi)}\) is a column vector of unknown coefficients, and
+\(\epsilon^{(\phi)}_t\) is an adjustment for time \(t\).
+
+To complete the specification of the Bayesian model, we specify priors
+for all unknown parameters. The recruitment parameter priors were
+specified as follows:
+
+\[\alpha^{(\lambda)} \sim N(0,1),\]
+\[\sigma^{(\lambda)} \sim N_+(0,1),\]
+\[\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),\] for periods
+\(t=1, ..., T\). Here \(N\) represents the normal distribution and
+\(N_+\) the half normal distribution with positive support.
+
+Survival parameter priors were specified similarly as:
+
+\[\beta^{(\phi)} \sim N(0, 1),\] \[\sigma^{(\phi)} \sim N_+(0, 1),\]
+\[\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),\] for time
+\(t=1, ..., T\).
+
+The detection model coefficient vector also received a standard normal
+prior \(\beta^{(p)} \sim N(0, 1)\).
+
+We computed the likelihood of each individual capture history using the
+forward algorithm, and we estimated the latent states using the
+forward-backward algorithm \citep{zucchini2009, joseph2018}.
+
+All of the code to specify and fit the model in Stan is available in the
+open source mrmr package \citep{joseph2019}.
+
+The joint distribution of the resulting model can be written as follows:
+
+\begin{multline*}
+[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{multline*}
+
+where \(\pmb Y\) is an \(M \times T\) detection matrix, and \(Y_i\) the
+capture history of individual \(i\).
+
+\hypertarget{among-site-survival-modeling}{%
+\paragraph{Among-site survival
+modeling}\label{among-site-survival-modeling}}
+
+The objective of this analysis is to describe the influence of site,
+cohort, and individual level characteristics on post-translocation frog
+survival. By modeling survival estimates obtained from site-specific
+mrmr CMR analyses, we are in effect conducting an among-site
+meta-analysis. Although it would theoretically be possible to estimate
+survival covariate effects in a joint CMR model that integrates capture
+histories across all sites, this was impractical due to computational
+requirements of the CMR models (namely, run time and memory).
+
+We used Bayesian generalized linear mixed models to investigate
+predictors of survival among sites. The response \(y_i\) is binary,
+representing a point estimate of whether individual \(i\) survived in
+the year following translocation. We generated these point estimates by
+rounding the posterior median of 1-year post-introduction survival for
+each individual (from site-specific mrmr CMR models) and modeled the
+data using a Bernoulli distribution:
+
+\[y_i \sim \text{Bernoulli}(p_i),\]
+
+where \(p_i\) is the probability of survival.
+
+We modeled variation in probabilities as follows:
+
+\[\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},\]
+
+where \(\alpha\) is an intercept, \(X_i\) is a length \(K\) row vector
+of predictors, \(\pmb \beta\) is a column vector of predictor effects,
+and \(\pmb \nu\) a vector of group level random effects. Here \(g[i]\)
+refers to the group \(g\) containing individual \(i\), and we estimate
+an adjustment for each of the \(G\) groups (\(\nu_1, ..., \nu_G\)).
+
+These models were fit using the \texttt{stan\_glmer()} function in the
+rstanarm package, with default priors described below
+\citep{rstanarm2022}. These priors are vague, but include data-dependent
+scaling as follows to account for different input variable scales.
+However, because we standardized all predictor variables similarly to
+have equal variance (by centering and dividing by twice the sample
+standard deviation), the resulting priors are identical. Specifically,
+we have:
+
+\[\alpha \sim \text{Normal}(0, 2.5),\] and
+
+\[\beta_k \sim \text{Normal}(0, 5),\]
+
+for \(k=1, ..., K\) where \(K\) is the number of predictor variables.
+
+The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
+rstanarm is a zero-mean Gaussian, where the covariance matrix is
+constructed from a correlation matrix with an LKJ prior, and a vector of
+variance parameters -- the decomposition of variance prior with unit
+regularization, concentration, shape, and scale parameters
+\citep{lewandowski2009}.
+
+We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
+with four parallel chains, each run for 10,000 iterations, discarding
+the first half of each chain as warm-up draws \citep{rstanarm2022}. We
+used Rhat statistics and trace plots to verify convergence. We
+considered models with different subsets of fixed and random effects,
+and used approximate leave-one-out cross validation to identify the best
+model \citep{vehtari2016}.
+
+\hypertarget{population-viability-modeling-1}{%
+\subsubsection{Population viability
+modeling}\label{population-viability-modeling-1}}
+
+\hypertarget{incorporating-yearly-variability-in-vital-rates}{%
+\paragraph{Incorporating yearly variability in vital
+rates}\label{incorporating-yearly-variability-in-vital-rates}}
+
+We computed yearly survival probabilities for translocated adults
+\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
+the posterior distribution of individual state trajectories derived from
+mrmr CMR models. Although we observed yearly variability in adult
+survival within a population, the magnitude of this variability was
+small compared to among-population variability
+(Figure~\ref{fig-translocation-survival}). Thus, we did not include
+yearly within-population variability in adult survival in this analysis.
+However, within a population there was substantial yearly variability in
+the successful recruitment of adults, greater than what we would expect
+from Poisson variability around a mean value. Therefore, we allowed for
+yearly variability in the probability of year-1 juvenile survival and
+recruitment (additional details provided in \textbf{Estimating model
+parameters} below). We also could have included environmental
+stochasticity in year-2 juvenile survival and recruitment
+\(\sigma_{J_2}\), but our elasticity analysis (\textbf{Supporting
+Information - Population viability modeling - Model analysis and
+simulation}) showed that this parameter had little effect on host growth
+rate relative to \(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp} SI).
+
+\hypertarget{estimating-model-parameters}{%
+\paragraph{Estimating model
+parameters}\label{estimating-model-parameters}}
+
+The baseline parameter values for the model and how they were estimated
+are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
+and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
+\textbf{Materials and Methods - Frog populaton recovery - Estimation of
+frog survival and abundance} and \textbf{Supporting Information - Frog
+population recovery - CMR model structure} for details). For populations
+where we had a sufficient number of PIT-tagged, naturally-recruited
+adults, we observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be
+notably different, with \(\sigma_{A_R} > \sigma_{A_T}\)
+(Figure~\ref{fig-compare_surv_probs} SI). For populations lacking
+sufficient numbers of naturally-recruited adults, we were unable to
+directly estimate \(\sigma_{A_R}\), and instead set
+\(\sigma_{A_R} = \sigma_{A_T}\).
+
+To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
+predicted true states for naturally-recruited individuals (1=not
+recruited, 2=alive, 3=dead, as described in \textbf{Supporting
+Information - Frog population recovery - CMR model structure}), then
+calculated the posterior probability of individuals surviving between
+consecutive primary periods, conditional on being alive in the first
+primary period (e.g., given a value of 2 (alive) in the first primary
+period, how often was the value still 2 (alive) in the next primary
+period compared to 3 (dead)?) . This yielded posterior distributions for
+survival probabilities between primary periods. However, because the
+time interval between primary periods differed, the survival
+probabilities between different consecutive primary periods were not
+directly comparable. To address this, we converted the survival
+probabilities between each consecutive primary period to per day death
+rates, propagating the uncertainty from the posterior distributions. We
+then took a weighted average of these death rates, weighted by the time
+interval between primary periods, to get the average per day death rate
+over the entire CMR survey. We converted this per day death rate \(d\)
+back to a yearly survival probability using
+\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
+\(\sigma_{A_T}\) such that our estimates of average yearly survival
+probability were comparable between \(\sigma_{A_R}\) and
+\(\sigma_{A_T}\).
+
+\hypertarget{model-analysis-and-simulation}{%
+\paragraph{Model analysis and
+simulation}\label{model-analysis-and-simulation}}
+
+We performed four analyses on our model. First, we considered a
+deterministic version of our model with no yearly heterogeneity in
+year-1 juvenile survival and recruitment probability \(\sigma_{J_1}\),
+and calculated the predicted long-run growth rate \(\lambda\) of a
+population for different values of \(\sigma_{A_R}\) and
+\(\sigma_{J_1}\). We then fixed \(\sigma_{J_1} = 0.09\) and calculated
+the predicted growth rate of our 12 populations.
+
+Second, we performed a local elasticity analysis on \(\lambda\) with
+respect to parameters \(\sigma_{J_1}\), \(\sigma_{J_2}\),
+\(\sigma_{A_R}\), and \(F\) to determine how small changes in these
+parameters could influence the long-run deterministic growth rate of
+populations (Figure~\ref{fig-viability-supp} SI).
+
+Third, we defined a version of the model with demographic and
+environmental stochasticity, where environmental stochasticity was
+represented by among-year variability in \(\sigma_{J_1}\). We used this
+model to simulate a one-time introduction of 40 translocated adult
+frogs. We ran this simulation 1000 times for each population and
+computed the probability of a population becoming extinct after 50 years
+given the observed parameter values and environmental stochasticity in
+\(\sigma_{J_1}\). We varied the mean recruitment probability
+\(\sigma_{J_1}\) from 0 and 0.25 and drew values of \(\sigma_{J_1}\)
+each year from a beta distribution with a dispersion parameter of
+\(\phi = 2\) (when \(\sigma_{J_1} = 0.5\) and \(\phi = 2\) the beta
+distribution is uniform between 0 and 1). Using different values of
+\(\phi\) does not qualitatively change the existence of distinct
+extinction dynamics between populations with \(\sigma_{A_R} < 0.5\) and
+those with \(\sigma_{A_R} > 0.5\). However, increasing yearly
+variability in \(\sigma_{J_1}\) increases extinction risk for all
+populations. For example, if we set \(\phi = 0.001\), such that in a
+given year essentially either all year-1 juveniles survive or all of
+them die, populations with \(\sigma_{A_R} > 0.5\) need to have
+\(\sigma_{J_1}\) greater than 0.2 to have a 50-year extinction
+probability of less 50\%. Because we do not pit tag juveniles, we do not
+have CMR estimates for \(\sigma_{J_1}\) or \(\phi\). However, based on
+qualitative and semi-quantitative field observations over 25 years, a
+value of \(\sigma_{J_1} = 0.25\) in the presence of Bd is probably a
+reasonable estimate for many populations. Thus, we expect our model
+predictions to be conservative with regards to population recovery.
+
+Finally, we assessed whether our stochastic model could reproduce
+observed trajectories of population recovery. We focused on population
+70550 because this was our longest CMR time series for a translocated
+population and because this population shows evidence of substantial
+post-translocation increases in adult abundance associated with
+population establishment and recovery. We simulated our model for 16
+years, repeating the simulation 50,000 times. For each run and each
+year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
+1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
+Using Approximate Bayesian Computing and rejection sampling
+\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
+1000 trajectories) that minimized the sum of squared errors between the
+observed and predicted data. The yearly \(\sigma_{J_1}\) values
+associated with these ``best'' trajectories represented an approximate
+posterior distribution \citep{beaumont2010}. Using these best fit
+trajectories, we assessed whether our model could qualitatively describe
+the patterns of recovery in the observed data for population 70550.
+
+\hypertarget{frog-evolution-in-response-to-bd-2}{%
+\subsubsection{Frog evolution in response to
+Bd}\label{frog-evolution-in-response-to-bd-2}}
+
+\hypertarget{study-design}{%
+\paragraph{Study design}\label{study-design}}
+
+To gain insights into the role of evolution in the development of
+resistance by MYL frogs, we compared frog exomes sampled in naive versus
+recovering populations. Comparing populations with different infection
+histories allowed larger sample sizes and replication across the
+landscape. The alternative approach of comparing samples from the same
+populations before and after Bd exposure isn't feasible in this system
+because Bd arrived in most MYL frog populations decades ago and
+population persistence/recovery is rare and unpredictable. As a result,
+samples from recovering populations collected before and after Bd
+exposure are not available and are unlikely to be available in the
+future.
+
+Our study design, in which we compared frog genomes in naive and
+recovering populations, required sampling populations across the range
+of MYL frogs in the southern Sierra Nevada. This is due to the fact that
+very few MYL frog populations remain in the naive state, and those that
+do are scattered across a wide latitudinal range, from Yosemite National
+Park in the north to southern Kings Canyon National Park in south. To
+minimize potential confounding effects caused by known variation in frog
+genotypes across latitude \citep{byrne2023}, we selected sampling sites
+such that both population types were represented across similar
+latitudinal ranges (Figure~\ref{fig-allelefrequencies}).
+
+\hypertarget{go-analysis}{%
+\paragraph{GO analysis}\label{go-analysis}}
+
+In \textbf{Results-Frog evolution in response to Bd}, we describe the
+stringent set of outlier variants (identified using a
+Bonferroni-corrected p-value of 0.01). A liberal set of outlier
+variants, identified using a Bonferroni-corrected p-value of 0.05,
+included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
+across 16 contigs. We used this liberal set to determine if any GO
+biological functions, molecular functions, or cellular processes were
+overrepresented. To do this, we retrieved the BLAST hits and mapped GO
+terms for each gene in our targeted transcriptome. We then conducted a
+statistical overrepresentation test (Fisher's exact test) using Blast2GO
+\citep{conesa2005} to compare the 30 unique outlier genes to the
+complete set of genes in our target transcriptome. We repeated this
+process for the set of 35 genes located in the 9 shared regions of the
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+
+\hypertarget{genetic-diversity}{%
+\paragraph{Genetic diversity}\label{genetic-diversity}}
+
+To characterize general patterns of genetic diversity between naive and
+recovering populations, we conducted three analyses. First, we
+calculated heterozygosity for each sampled frog using VCFtools
+\citep{danecek2011}. Second, to characterize genome-wide patterns of
+diversity, we used VCFtools to calculate nucleotide diversity (\(\pi\))
+in 100kb sliding windows along the genome for each population. Third, we
+calculated average \(\pi\) per population within each of the 9 outlier
+windows identified in the splined window analysis.
+
+Average individual-level heterozygosity and genome-wide population-level
+\(\pi\) were similar between the naive and recovering groups
+(Figure~\ref{fig-violinplot-heterozy} SI;
+Figure~\ref{fig-boxplot-genomewide-pi-by-pop} SI). Within each of the 9
+outlier windows, average \(\pi\) shows considerable variation between
+populations (Figure~\ref{fig-boxplot-pi-by-windowpop} SI) and no obvious
+patterns between the naive and recovering groups.
+
+\newpage
+
+\hypertarget{figures-1}{%
+\subsubsection{Figures}\label{figures-1}}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=4.16667in,height=\textheight]{figures/map_translocation_points.png}
+
+}
+
+\caption{\label{fig-yosemap}Map showing the locations of translocated
+and donor MYL frog populations in Yosemite National Park (park boundary
+indicated by gray polygon). Symbol shapes indicate the donor population
+used for each translocation site, and 5-digit numbers identify each
+donor and translocation site. To obscure the exact locations of
+populations, random noise was added to all point coordinates. Inset map
+shows the location of Yosemite within California. In both maps,
+elevation is indicated by the colored hillshade layer (dark green =
+lowest elevation, white = highest elevation).}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=6.5625in,height=\textheight]{figures/bdload_beforeafter.png}
+
+}
+
+\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
+the 12 recipient sites, Bd loads for the period immediately prior to
+translocation versus during the 1-year period after translocation. Bd
+loads are expressed as the number of ITS1 copies per skin swab, as
+estimated by qPCR of the Bd ITS1 region. Box plots show medians, first
+and third quartiles, largest and smallest values within 1.5x
+interquartile range, and values outside the 1.5x interquartile range.
+Loads indicative of severe disease are \textgreater{} 5.8 ITS copies (on
+a log\textsubscript{10} scale). Samples sizes are provided immediately
+above the x-axis.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.20833in,height=\textheight]{figures/mcmc_areas_m1d.png}
+
+}
+
+\caption{\label{fig-survival-postdens}Results from the among-site
+meta-analysis, showing that Bd load is not an important predictor of
+post-translocation frog survival. Depicted distributions are the
+estimated posterior density curves and shaded 95\% uncertainty intervals
+for the intercept and all predictor variables from the best model. In
+the Bayesian framework in which the model was developed, variables are
+considered important predictors if the associated uncertainty interval
+does not overlap zero (indicated by the dashed red line). Predictor
+variables shown on the y-axis are defined as follows: length = frog
+size, snow\_t = winter severity in the year of translocation (measured
+on April 1), snow\_t1 = winter severity in the year following
+translocation (measured on April 1), day = day of year on which a
+translocation was conducted, bd\_load = Bd load, elevation = site
+elevation, order\_first = within-site translocation order, sex\_male =
+frog sex, donor\_70567 and donor\_72996 = donor population.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=6.04167in,height=\textheight]{figures/compare_surv_probs.jpg}
+
+}
+
+\caption{\label{fig-compare_surv_probs}Comparison of average yearly
+adult survival probabilities for adults translocated to each of 3 sites
+versus adults that were naturally recruited at each site (as a result of
+reproduction by translocated frogs). In contrast to
+Figure~\ref{fig-translocation-survival}, these are not survival
+probabilities from the first year following translocation, but instead
+represent averaged survival probabilities across multiple years and
+cohorts. Points are median estimates and error bars give the 95\%
+uncertainty intervals around the estimates, accounting for yearly
+variation in survival probabilities. All estimates were derived using
+the mrmr package, and the methods for calculation are described in
+\textbf{Supporting Information - Population viability modeling -
+Estimating model parameters}.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.72917in,height=\textheight]{figures/pop_viability_figures_for_supp.jpg}
+
+}
+
+\caption{\label{fig-viability-supp}Sensitivity analysis of the
+stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
+in four parameters: \(\sigma_{A_R}\) (yearly survival probability of
+naturally recruited adults), \emph{F (}number of eggs produced by a
+female frog in a year that successfully hatch), \(\sigma_{J_1}\) (yearly
+probability of survival of year-1 juveniles that also affects
+recruitment), and \(\sigma_{J_2}\) (yearly probability of survival and
+recruitment of year-2 juveniles). Elasticity is calculated at the
+default parameter values for each population and \(\sigma_{J_1}\) =
+0.09.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.72917in,height=\textheight]{figures/pca_qq_manhattan.png}
+
+}
+
+\caption{\label{fig-selectionresults}Results from the analysis of
+individual variants, showing putative signatures of selection in
+recovering MYL frog populations. (A) PCA calculated from binary SNPs
+showing the genomic relationship of samples. Numeric labels and colors
+match those in Figure~\ref{fig-allelefrequencies}. Populations 1-7 are
+\emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
+qqplot showing observed and expected p-values for 148,307 SNPs and
+INDELS. Dashed line shows p-value that identifies outliers. (C)
+Manhattan plot showing p-value for each SNP. SNPs are sorted by genomic
+position and contigs are sorted by size. Red line shows p-value that
+identifies outliers. Outlier SNPs above this threshold are highlighted
+and labeled. Bold labels indicate the presence of at least one
+non-synonymous SNP in that gene.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=6.77083in,height=\textheight]{figures/synteny_figure.png}
+
+}
+
+\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
+order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
+outlier region containing MHC Class I Classical and MHC Class III gene
+regions. The plot was created with SimpleSynteny \citep{veltri2016}
+using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
+accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
+indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
+sizes for each contig representation are labeled.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.20833in,height=\textheight]{figures/violin_plot_heterozy_by_group.png}
+
+}
+
+\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
+heterozygosity for the Bd-naive and recovering populations included in
+the frog evolution study. Individual data points are represented by
+their corresponding site number (from
+Figure~\ref{fig-allelefrequencies}).}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.72917in,height=\textheight]{figures/boxplot_genomewide_pi_by_pop.png}
+
+}
+
+\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each study
+population in the frog evolution study (from
+Figure~\ref{fig-allelefrequencies}), boxplots showing nucleotide
+diversity (\(\pi\)) within 100kb sliding windows along the genome. The
+y-axis has been log\textsubscript{10}-transformed for display purposes.
+Sites are color-coded as ``Bd-naive'' or ``recovering'', and vertical
+dashed lines show the genetic breaks between frog populations as
+described in Figure~\ref{fig-selectionresults} A SI).}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=5.20833in,height=\textheight]{figures/boxplot_pi_by_windowpop.png}
+
+}
+
+\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
+windows identified by the overlapping splined windows for FST and
+\(\pi\), boxplots showing average \(\pi\) for naive and recovering
+populations. The corresponding gene annotations for each window are as
+follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3,
+EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5)
+GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR,
+MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD,
+ATRAID, GPN1.}
+
+\end{figure}
+
+\newpage
+
+\hypertarget{tables}{%
+\subsubsection{Tables}\label{tables}}
+
+\hypertarget{tbl-param_values}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2639}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}@{}}
+\caption{\label{tbl-param_values}Description and values of parameters
+used in the model. All survival probabilities are in the presence of the
+fungal pathogen Bd.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Parameter}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Value}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Source}
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Parameter}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Value}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Source}
+\end{minipage} \\
+\midrule()
+\endhead
+\(\sigma_{L_1}\), Yearly survival probability of year-1 tadpoles & 0.7 &
+Estimated from field data, observations, natural history knowledge \\
+\(\sigma_{L_2}\), Yearly survival probability of year-2 tadpoles & 0.7 &
+Estimated from field data, observations, natural history knowledge \\
+\(\sigma_{L_3}\), Yearly survival probability of year-3 tadpoles & 0.7 &
+Estimated from field data, observations, natural history knowledge \\
+\(\sigma_{J_1}\), Yearly survival probability of year-1 juveniles &
+Varies yearly & Varies. Bounds estimated from field data, observations,
+natural history knowledge \\
+\(\sigma_{J_2}\), Yearly survival probability of year-2 juveniles & 0.5
+& Estimated from field data, observations, natural history knowledge \\
+\(\sigma_{A_R}\), Yearly survival probability of naturally recruited
+adults & Varies by population & Estimated from CMR studies \\
+\(\sigma_{A_T}\), Yearly survival probability of translocated adults &
+Varies by population & Estimated from CMR studies \\
+\(p_{L_1}\), Probability of a year-1 tadpoles remaining as a tadpoles &
+1 & Estimated from field data, observations, natural history
+knowledge \\
+\(p_{L_2}\), Probability of a year-2 tadpoles remaining as a tadpoles &
+0.25 & Estimated from field data, observations, natural history
+knowledge \\
+\(p_{J_1}\), Probability of a year-1 juvenile remaining as a juvenile &
+0.25 & Estimated from field data, observations, natural history
+knowledge \\
+\(p_F\), Probability of a adult female reproducing in a year & 0.5 &
+Could be as high at 1, based on field observations \\
+\(F\), Number of surviving eggs produced by an adult female & 100 & From
+observations of captive frogs \\
+\bottomrule()
+\end{longtable}
+
+\newpage
+
+\hypertarget{datasets}{%
+\subsubsection{Datasets}\label{datasets}}
+
+Dataset S1. Set of liberal SNP and INDEL outlier variants (Bonferroni
+corrected p-value \textless{} 0.05), identified via GEMMA. file:
+gemma\_outliers\_all.csv
+
+Dataset S2. Set of strict SNP and INDEL outlier variants (Bonferroni
+corrected p-value \textless{} 0.01), identified via GEMMA. Additional
+information includes variant location within the gene
+(predicted\_gene\_loc) and whether the variant is synonymous or
+nonsynonymous (predicted\_effect\_AA). file:
+gemma\_outliers\_strict\_freq.csv
+
+Dataset S3. Description of overlapping \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) outlier windows, as identified in the splined window
+analysis. The outlier window on chromosome 19 (column name =
+chr\_num\_sorted) is included twice because one
+\emph{F\textsubscript{ST}} outlier window overlapped with two
+\(\pi_{diff}\) outlier windows. file:
+spline\_window\_shared\_outliers.csv
+
+Dataset S4. Annotation information for all genes within each of the
+overlapping outlier windows in dataset S3. file:
+spline\_window\_gene\_details
+
+\hypertarget{references-1}{%
+\subsubsection{References}\label{references-1}}
+
+  \bibliography{translocation.bib}
+
 \end{document}

### 2023-12-23 doc/manuscript/translocation_bioarxiv.tex
Compile versions for journal and preprint server

@@
 selection in recovering populations. The evolution of resistance and
-reintroductions of resistant individuals could play an important role in
+reintroduction of resistant individuals could play an important role in
 biodiversity conservation in our rapidly changing world.}
@@
 resistance for frog survival, population establishment, and long-term
-viability (this study), we also conducted a genomic study to determine
-whether MYL frogs show an evolutionary response to Bd exposure, and
-whether these genomic changes are associated with resistance. To
-accomplish this objective, we used exome capture methods to identify
-genomic regions with consistent differences between frogs from naive
-versus recovering populations (Figure~\ref{fig-recovery-model}).
+viability (this study), we conducted a genomic study using exome capture
+methods to determine whether MYL frogs in recovering populations show
+evidence of selection and whether these genomic changes are associated
+with resistance (Figure~\ref{fig-recovery-model}).
 
@@
 sites (Figure S1). The donor populations included 2 of
-the 5 recovering populations included in the frog evolution study
-referenced above (Figure~\ref{fig-allelefrequencies}: population 1 and
-4), and these donor populations contributed frogs for 20 of the 24
+the 5 recovering populations used in the frog evolution study referenced
+above (Figure~\ref{fig-allelefrequencies}: population 1 and 4), and
+these donor populations contributed frogs for 20 of the 24
 translocations. Following translocation, we estimated adult survival and
@@
 little difference in pre- versus post-translocation Bd loads even in
-those populations that had very high frog survival (70556, 74976;
-Figure~\ref{fig-translocation-survival},
-Figure S2) suggests a true lack of
+those populations that had very high frog survival (70556, 74976 - see
+below; Figure S2) suggests a true lack of
 substantial change in Bd load.
@@
 test whether the observed yearly adult survival probabilities in
-translocated populations (provided in legend of
-Figure~\ref{fig-viability}B) were sufficient for long-term viability,
-we built a stage-structured matrix model that captured known frog
+translocated populations were sufficient for long-term viability, we
+built a stage-structured matrix model that captured known frog
 demography and included demographic and environmental stochasticity. We
 parameterized the model using CMR data from translocated populations and
-known life history values in this system (Table S2
-SI).
+known life history values in this system (Table S1).
 
 Given observed yearly adult survival probabilities of translocated frogs
-(from site-specific mrmr CMR models) and a yearly survival probability
-of the year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at
-least six of twelve translocated populations should experience a
-long-run growth rate \(\lambda\) greater than 1 in the presence of Bd
+(from site-specific mrmr CMR models; provided in legend of
+Figure~\ref{fig-viability}B) and a yearly survival probability of the
+year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at least six
+of twelve translocated populations should experience a long-run growth
+rate \(\lambda\) greater than 1 in the presence of Bd
 (Figure~\ref{fig-viability}A; median predicted \(\lambda\) ranges from
@@
 Individual frogs clustered into 3 separate groups in principal component
-space (Figure S6A SI), and clusters reflected
+space (Figure S6A), and clusters reflected
 the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
@@
 frequencies between naive versus recovering populations) from 7 distinct
-genes across 4 contigs (Figure S6B, C SI; SI
-Datasets S1, S2). One of the 7 identified genes (LOC108802036) does not
-have an associated annotation. For the outlier SNPs, frequency
-differences between the naive and recovering populations ranged from
-0.41 to 0.86. Most of these SNPs showed frequency differences in only a
-subset of the sampled populations (Figure~\ref{fig-allelefrequencies}A,
-B), but the SNP in the RIN3 gene showed consistent differences in
-frequencies across all populations (Figure~\ref{fig-allelefrequencies}
-C). This is suggestive of parallel selection at this locus across
-multiple populations. The other 6 outlier variants showed less
-consistent frequency differences across the study populations, but for
-these we still found a statistically significant signal of selection in
-2 of the 3 genetic clusters (containing populations 5--9;
-Figure~\ref{fig-allelefrequencies}, Figure S6A
-SI). Therefore, although some outlier variant associations have a more
+genes across 4 contigs (Figure S6B, C). One
+of the 7 identified genes (LOC108802036) does not have an associated
+annotation. For the outlier SNPs, frequency differences between the
+naive and recovering populations ranged from 0.41 to 0.86. Most of these
+SNPs showed frequency differences in only a subset of the sampled
+populations (Figure~\ref{fig-allelefrequencies}A, B), but the SNP in
+the RIN3 gene showed consistent differences in frequencies across all
+populations (Figure~\ref{fig-allelefrequencies}C). This is suggestive
+of parallel selection at this locus across multiple populations. The
+other 6 outlier variants showed less consistent frequency differences
+across the study populations, but for these we still found a
+statistically significant signal of selection in 2 of the 3 genetic
+clusters (containing populations 5--9;
+Figure~\ref{fig-allelefrequencies}, Figure S6A). Therefore, although some outlier variant associations have a more
 limited geographic extent than RIN3, they still describe results that
@@
 regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions
-(Datasets S3, S4). Given this large number of genes, here we focus on
-those with the strongest signal of selection and/or immune-related
-functions. The largest \(\pi_{diff}\), indicative of directional
-selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
-the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C). This region
-contains approximately 500 SNPs and one annotated gene called
+A total of 35 annotated genes were found in the 9 shared regions. Given
+this large number of genes, here we focus on those with the strongest
+signal of selection and/or immune-related functions. The largest
+\(\pi_{diff}\), indicative of directional selection, occurred in a 163kb
+region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP
+(Figure~\ref{fig-spline-manhattan}C). This region contains
+approximately 500 SNPs and one annotated gene called
 ``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
@@
 for the multi-year tadpole and juvenile stages in MYL frogs
-(Table S2 SI).
+(Table S1).
 
@@
 realism. Additional details are provided in \textbf{Supporting
-Information - Population viability modeling}.
+Information - Population viability modeling - Incorporating yearly
+variability in survival rates} and \textbf{Estimating model parameters}.
 
@@
 probabilities and other known vital rates
-(Table S2), we performed four analyses. First, we
+(Table S1), we performed four analyses. First, we
 compute the long-run growth rate \(\lambda\) for each of our 12
@@
 are provided in \textbf{Supporting Information - Population viability
-modeling}. Code to replicate the analyses can be found at
-\url{https://github.com/SNARL1/translocation}.
+modeling - Model analysis and simulation}. Code to replicate the
+analyses can be found at \url{https://github.com/SNARL1/translocation}.
 
@@
 set solely for the gene ontology (GO) analysis (see below and
-\textbf{Supporting Information}: Dataset S1, S2).
+\textbf{Supporting Information - Frog evolution in response to Bd - GO
+analysis;}  Dataset S1, S2).
 

### 2023-12-23 doc/manuscript/translocation_bioarxiv_SI.tex
Compile versions for journal and preprint server

@@
 \hypertarget{laboratory-methods}{%
-\subsection{Laboratory Methods}\label{laboratory-methods}}
+\subsection{Laboratory methods}\label{laboratory-methods}}
 
@@
 \(u_{i, t+1}\) (columns) between primary period \(t\) and \(t+1\). For
-unintroduced individuals which naturally recruit, this matrix is given
-by:
+non-introduced (i.e., naturally recruited) individuals, this matrix is
+given by:
 
@@
 
-To complete the specification of the Bayesian model, we need to specify
-priors for all unknown parameters. The recruitment parameter priors were
+To complete the specification of the Bayesian model, we specify priors
+for all unknown parameters. The recruitment parameter priors were
 specified as follows:
@@
 \hypertarget{population-viability-modeling-1}{%
-\subsection{Population viability
+\section{Population viability
 modeling}\label{population-viability-modeling-1}}
@@
 \hypertarget{incorporating-yearly-variability-in-vital-rates}{%
-\subparagraph{Incorporating yearly variability in vital
+\subsection{Incorporating yearly variability in vital
 rates}\label{incorporating-yearly-variability-in-vital-rates}}
@@
 small compared to among-population variability
-(Fig. 3). Thus, we did not include
+(Fig. 2). Thus, we did not include
 yearly within-population variability in adult survival in this analysis.
@@
 from Poisson variability around a mean value. Therefore, we allowed for
-yearly variability in the probability of year 1 juvenile survival and
+yearly variability in the probability of year-1 juvenile survival and
 recruitment (additional details provided in \textbf{Estimating model
 parameters} below). We also could have included environmental
-stochasticity in year 2 juvenile survival and recruitment
+stochasticity in year-2 juvenile survival and recruitment
 \(\sigma_{J_2}\), but our elasticity analysis (\textbf{Supporting
-Information-Population viability modeling-Model analysis and
+Information - Population viability modeling - Model analysis and
 simulation}) showed that this parameter had little effect on host growth
@@
 \hypertarget{estimating-model-parameters}{%
-\subparagraph{Estimating model
+\subsection{Estimating model
 parameters}\label{estimating-model-parameters}}
@@
 and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
-\textbf{Materials and Methods - Estimation of frog survival and
-abundance} and \textbf{Supporting Information - Frog population recovery
-- Estimation of frog survival and abundance} for details). For
-populations where we had a sufficient number of PIT-tagged,
-naturally-recruited adults, we observed that \(\sigma_{A_T}\) and
-\(\sigma_{A_R}\) could be notably different, with
-\(\sigma_{A_R} > \sigma_{A_T}\) (Figure~\ref{fig-compare_surv_probs}). For populations lacking sufficient numbers of naturally-recruited
-adults, we were unable to directly estimate \(\sigma_{A_R}\), and
-instead set \(\sigma_{A_R} = \sigma_{A_T}\).
+\textbf{Materials and Methods - Frog populaton recovery - Estimation of
+frog survival and abundance} and \textbf{Supporting Information - Frog
+population recovery - CMR model structure} for details). For populations
+where we had a sufficient number of PIT-tagged, naturally-recruited
+adults, we observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be
+notably different, with \(\sigma_{A_R} > \sigma_{A_T}\)
+(Figure~\ref{fig-compare_surv_probs}). For populations lacking
+sufficient numbers of naturally-recruited adults, we were unable to
+directly estimate \(\sigma_{A_R}\), and instead set
+\(\sigma_{A_R} = \sigma_{A_T}\).
 
@@
 recruited, 2=alive, 3=dead, as described in \textbf{Supporting
-Information-Frog population recovery-CMR model structure}), then
+Information - Frog population recovery - CMR model structure}), then
 calculated the posterior probability of individuals surviving between
@@
 \hypertarget{model-analysis-and-simulation}{%
-\subparagraph{Model analysis and
+\subsection{Model analysis and
 simulation}\label{model-analysis-and-simulation}}
@@
 We performed four analyses on our model. First, we considered a
-deterministic version of our model with no yearly heterogeneity in year
-1 juvenile survival and recruitment probability \(\sigma_{J_1}\), and
-calculated the predicted long-run growth rate \(\lambda\) of a
+deterministic version of our model with no yearly heterogeneity in
+year-1 juvenile survival and recruitment probability \(\sigma_{J_1}\),
+and calculated the predicted long-run growth rate \(\lambda\) of a
 population for different values of \(\sigma_{A_R}\) and
@@
 populations. For example, if we set \(\phi = 0.001\), such that in a
-given year essentially either all year 1 juveniles survive or all of
+given year essentially either all year-1 juveniles survive or all of
 them die, populations with \(\sigma_{A_R} > 0.5\) need to have
@@
 To gain insights into the role of evolution in the development of
-resistance by MYL frogs, we compared frog genomes sampled in naive
-versus recovering populations. Comparing populations with different
-infection histories allowed larger sample sizes and replication across
-the landscape. The alternative approach of comparing samples from the
-same populations before and after Bd exposure isn't feasible in this
-system because Bd arrived in most MYL frog populations decades ago and
+resistance by MYL frogs, we compared frog exomes sampled in naive versus
+recovering populations. Comparing populations with different infection
+histories allowed larger sample sizes and replication across the
+landscape. The alternative approach of comparing samples from the same
+populations before and after Bd exposure isn't feasible in this system
+because Bd arrived in most MYL frog populations decades ago and
 population persistence/recovery is rare and unpredictable. As a result,
-samples from recovering populations collected from before and after Bd
+samples from recovering populations collected before and after Bd
 exposure are not available and are unlikely to be available in the
@@
 such that both population types were represented across similar
-latitudinal ranges (Fig. 1).
+latitudinal ranges (Fig. 5).
 
@@
 complete set of genes in our target transcriptome. We repeated this
-process for the set of 33 genes located in the
+process for the set of 35 genes located in the 9 shared regions of the
 \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
@@
 reproduction by translocated frogs). In contrast to
-Fig. 3, these are not survival
+Fig. 2, these are not survival
 probabilities from the first year following translocation, but instead
@@
 the mrmr package, and the methods for calculation are described in
-\textbf{Supporting Information - Population viability modeling --
+\textbf{Supporting Information - Population viability modeling -
 Estimating model parameters}.}
@@
 female frog in a year that successfully hatch), \(\sigma_{J_1}\) (yearly
-probability of survival of year 1 juveniles that also affects
+probability of survival of year-1 juveniles that also affects
 recruitment), and \(\sigma_{J_2}\) (yearly probability of survival and
-recruitment of year 2 juveniles). Elasticity is calculated at the
+recruitment of year-2 juveniles). Elasticity is calculated at the
 default parameter values for each population and \(\sigma_{J_1}\) =
@@
 showing the genomic relationship of samples. Numeric labels and colors
-match those in Fig. 1A. Populations 1-7 are
+match those in Fig. 5. Populations 1-7 are
 \emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
@@
 their corresponding site number (from
-Fig. 1).}
+Fig. 5).}
 
@@
 population in the frog evolution study (from
-Fig. 1), boxplots showing nucleotide
+Fig. 5), boxplots showing nucleotide
 diversity (\(\pi\)) within 100kb sliding windows along the genome. The
@@
 dashed lines show the genetic breaks between frog populations as
-described in Figure~\ref{fig-selectionresults} A). Within the genetic
-clusters, mean \(\pi\) is significantly different between sites (p
-\textless0.001 Wilcoxon rank-sum test).}
+described in Figure~\ref{fig-selectionresults} A).}
 
@@
 windows identified by the overlapping splined windows for FST and
-\$\textbackslash pi\$, boxplots showing average \$\textbackslash pi\$
-for naive and recovering populations. The corresponding gene annotations
-for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
-SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
-TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
-(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
-BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
+\(\pi\), boxplots showing average \(\pi\) for naive and recovering
+populations. The corresponding gene annotations for each window are as
+follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3,
+EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5)
+GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR,
+MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD,
+ATRAID, GPN1.}
 
@@
 \endhead
-\(\sigma_{L_1}\), Yearly survival probability of year 1 tadpoles & 0.7 &
+\(\sigma_{L_1}\), Yearly survival probability of year-1 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{L_2}\), Yearly survival probability of year 2 tadpoles & 0.7 &
+\(\sigma_{L_2}\), Yearly survival probability of year-2 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{L_3}\), Yearly survival probability of year 3 tadpoles & 0.7 &
+\(\sigma_{L_3}\), Yearly survival probability of year-3 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles &
+\(\sigma_{J_1}\), Yearly survival probability of year-1 juveniles &
 Varies yearly & Varies. Bounds estimated from field data, observations,
 natural history knowledge \\
-\(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
+\(\sigma_{J_2}\), Yearly survival probability of year-2 juveniles & 0.5
 & Estimated from field data, observations, natural history knowledge \\
@@
 Varies by population & Estimated from CMR studies \\
-\(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
+\(p_{L_1}\), Probability of a year-1 tadpoles remaining as a tadpoles &
 1 & Estimated from field data, observations, natural history
 knowledge \\
-\(p_{L_2}\), Probability of a year 2 tadpoles remaining as a tadpoles &
+\(p_{L_2}\), Probability of a year-2 tadpoles remaining as a tadpoles &
 0.25 & Estimated from field data, observations, natural history
 knowledge \\
-\(p_{J_1}\), Probability of a year 1 juvenile remaining as a juvenile &
+\(p_{J_1}\), Probability of a year-1 juvenile remaining as a juvenile &
 0.25 & Estimated from field data, observations, natural history

### 2024-01-05 doc/manuscript/translocation_pnas.tex
Update PNAS template

@@
 % {pnasinvited} %= Template for a PNAS invited submission
+
+\begin{document}
 
@@
 \affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
-\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
+\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}\
 \affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
@@
 % At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{}
+\keywords{evolution | resistance | amphibians | population recovery | reintroduction | disease | \emph{Batrachochytrium dendrobatidis}}
 
@@
 
-\begin{document}
-
 \maketitle
@@
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+
+% \firstpage{12}
 

### 2024-01-05 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -315,7 +315,7 @@ with open("translocation_bioarxiv.tex", "w") as fout:
 	fout.writelines(bioarxiv_ms)
 
 # Remove transcloation.tex
-# subprocess.call(['rm', args[1]])
+subprocess.call(['rm', args[1]])
 subprocess.call(['cp', 'translocation_pnas_SI.tex', 'translocation_bioarxiv_SI.tex'])
 
 

### 2024-01-05 doc/manuscript/convert_qmd_to_pnas_latex.py
Drop table header

@@ -288,6 +288,8 @@ for key, value in main_map.items():
 
 pnas_supp = re.sub("Fig. 5 A", "Fig. 5A", pnas_supp)
 
+# Drop Table header
+pnas_supp = pnas_supp.replace("\\hypertarget{tables}{%\n\\subsubsection{Tables}\\label{tables}}", "")
 
 # Remove verbatim
 verb_start = latex_manuscript.find("\\begin{verbatim}")

### 2024-01-05 doc/manuscript/translocation.qmd

@@
 
+**Competing interests:** This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
+Briggs; DEB-1557190, to C.
+Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
+Richards-Zawacki).
+The authors have no competing interests to disclose related to organizational affiliations, stock ownership, patent filings, public statements and positions, and non-financial or additional information.
+
 ### Abstract
@@
 Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year-1 juvenile survival probability was greater than 0.10<!--# Why not use 0.09?-->).
+In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year-1 juvenile survival probability was greater than 0.10).
 The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the years covered by our post-translocation surveys.
@@
 The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
-<!--# It would be great to describe how reintroduction of the product of evolution - individuals with alleles that are adaptive in a novel environment - can allow recovery at a broader scale that natural recovery alone.
-Not sure we have space for this.
--->
 
@@
 We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-<!--# I thought Max edited this sentence to get rid of "algorithm" but I must be mis-remembering. I checked his explain-meta-analysis branch and the above sentence is repeated there.  -->
 
@@
 The y-axis has been log~10~-transformed for display purposes.
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).
-](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
+Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
 

### 2024-01-06 doc/manuscript/translocation.qmd
Minor edits to manuscript, figures, and latex files

@@
 
-**Key words:** evolution, resistance, amphibians, population recovery, reintroduction, disease, *Batrachochytrium dendrobatidis*
+**Key words:** amphibians, disease, evolution, reintroduction, population recovery
+
+**Author contributions**: All authors designed research; R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., A.P.R., and R.L.G.
+performed research; M.B.J.
+contributed new analytic tools; R.A.K., M.Q.W., A.Q.B., A.P.R., and E.B.R.
+analyzed data; and R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., and E.B.R.
+wrote paper.
 
@@
 Richards-Zawacki).
-The authors have no competing interests to disclose related to organizational affiliations, stock ownership, patent filings, public statements and positions, and non-financial or additional information.
+The authors declare no additional competing interests.
 
@@
 Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-Emerging infectious diseases can have particularly devastating impacts, and few options exist to reverse these effects.
-We used large-scale reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a disease that has decimated amphibian biodiversity.
-Introduction of resistant frogs allowed reestablishment of viable populations in the presence of disease.
-In addition, resistance may be at least partially the result of natural selection at specific immune function genes, which show evidence for selection in recovering populations.
+We used reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a novel disease that has decimated amphibians worldwide.
+Reintroductions allowed establishment of viable populations in the presence of disease.
+In addition, resistance may be at least partially the result of natural selection at specific immune function genes.
 The evolution of resistance and reintroduction of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
@@
 
-In summary, results from our frog translocation study indicate that translocations resulted in (i) relatively high 1-year survival of translocated adults, as well as reproduction and recruitment, at the majority of recipient sites, (ii) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (iii) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
+In summary, results from our frog translocation study indicate that (i) translocations produced relatively high 1-year survival of translocated adults, as well as reproduction and recruitment, at the majority of recipient sites, (ii) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (iii) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
@@
 Richards-Zawacki).
-<!--# Do we need to acknowledge anybody at SD Zoo for providing R. muscosa genome? -->
 

### 2024-01-06 doc/manuscript/translocation_pnas.tex
Minor edits to manuscript, figures, and latex files

@@
 \affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
-\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}\
+\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
 \affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
@@
 \dropcap{H}uman activities are increasingly impacting global biodiversity
-\citep{ceballos2015}, with important implications for ecosystem
-resilience and human welfare \citep{naeem2009}. One consequence of human
+\cite{ceballos2015}, with important implications for ecosystem
+resilience and human welfare \cite{naeem2009}. One consequence of human
 alteration of the biosphere is an increase in emerging infectious
-diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
-threat to wildlife populations \citep{daszak2000}, and have caused
+diseases \cite{jones2008, fisher2012}. Such diseases pose a severe
+threat to wildlife populations \cite{daszak2000}, and have caused
 dramatic declines and extinctions in a wide range of taxa, including
 echinoderms, mammals, birds, and amphibians
-\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
+\cite{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
 are experiencing particularly devastating impacts of disease due to the
@@
 chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
-\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
+\cite{luedtke2023, scheele2019}. By one estimate, hundreds of species
 have experienced Bd-caused declines, and numerous susceptible taxa are
-extinct in the wild \citep{scheele2019}. These impacts to global
+extinct in the wild \cite{scheele2019}. These impacts to global
 biodiversity may be unprecedented, and highlight the importance of
 understanding mechanisms of species persistence in the presence of
-emerging diseases \citep{russell2020}.
+emerging diseases \cite{russell2020}.
 
 Evidence of natural recovery in the many Bd-impacted amphibian
-populations is surprisingly rare \citep[for notable exceptions,
+populations is surprisingly rare \cite[for notable exceptions,
 see][]{scheele2017, voyles2018, knapp2016}, suggesting that
@@
 harm caused by a particular burden) are key mechanisms to reduce disease
-impacts \citep{raberg2009} and facilitate population persistence and
-recovery \citep{brannelly2021}. Host immunity and evolution both play
+impacts \cite{raberg2009} and facilitate population persistence and
+recovery \cite{brannelly2021}. Host immunity and evolution both play
 important roles in the development of resistance and tolerance, and
@@
 effective strategies to mitigate disease impacts in the wild
-\citep{garner2016, woodhams2011}. However, several aspects of the
+\cite{garner2016, woodhams2011}. However, several aspects of the
 amphibian-Bd system present difficult obstacles, including (i) the
 general inability of amphibians to mount an effective immune response
-against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
+against Bd infection \cite{rosenblum2012, fites2013, grogan2018a}, and
 (ii) the apparent rarity of evolution of more resistant/tolerant
-genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
+genotypes \cite[but see][]{savage2016, grogan2018b}. These factors
 suggest that reintroduction of amphibians into sites to reestablish
@@
 introduced animals and/or their progeny
-\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
+\cite{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
 future of many amphibian species threatened by Bd appears bleak.
@@
 The mountain yellow-legged (MYL) frog, composed of the sister species
-\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
+\emph{Rana muscosa} and \emph{Rana sierrae} \cite{vredenburg2007}, is
 emblematic of the global declines of amphibians caused by Bd
-\citep{scheele2019}. Once the most common amphibian in the high
+\cite{scheele2019}. Once the most common amphibian in the high
 elevation portion of California's Sierra Nevada mountains
-\citep[USA,][]{grinnell1924}, during the past century this frog has
+\cite[USA,][]{grinnell1924}, during the past century this frog has
 disappeared from more than 90\% of its historical range
-\citep{vredenburg2007}. Due to the severity of its decline and the
+\cite{vredenburg2007}. Due to the severity of its decline and the
 increasing probability of extinction, both species are now listed as
@@
 trout into the extensive historically-fishless region
-\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
-of Bd in the mid-1900s and its subsequent spread \citep{vredenburg2019}
+\cite{bradford1989, knapp2000} starting in the late 1800s. The arrival
+of Bd in the mid-1900s and its subsequent spread \cite{vredenburg2019}
 caused additional large-scale population extirpations
-\citep{vredenburg2010, rachowicz2006}. These Bd-caused declines are
+\cite{vredenburg2010, rachowicz2006}. These Bd-caused declines are
 fundamentally different from the fish-caused declines because fish
-eradication is feasible \citep{knapp1998} and results in the rapid
-recovery of frog populations \citep{knapp2007, vredenburg2004}. In
+eradication is feasible \cite{knapp1998} and results in the rapid
+recovery of frog populations \cite{knapp2007, vredenburg2004}. In
 contrast, Bd appears to persist in habitats even in the absence of
-amphibian hosts \citep{walker2007}, and therefore represents a long-term
+amphibian hosts \cite{walker2007}, and therefore represents a long-term
 alteration of invaded ecosystems that amphibians will need to overcome
@@
 Bd-naive populations are extirpated following Bd arrival
-\citep{vredenburg2010}, some populations have persisted after epizootics
-\citep[during which Bd infection intensity on frogs is very
+\cite{vredenburg2010}, some populations have persisted after epizootics
+\cite[during which Bd infection intensity on frogs is very
 high,][]{briggs2010} and are now recovering
-(Figure~\ref{fig-recovery-model}) \citep{knapp2016}. Frogs in these
+(Figure~\ref{fig-recovery-model}) \cite{knapp2016}. Frogs in these
 recovering populations show reduced susceptibility to Bd infection
-\citep{knapp2016}, with infection intensity (``load'') on adults
+\cite{knapp2016}, with infection intensity (``load'') on adults
 consistently in the low-to-moderate range
-\citep{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
+\cite{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
 is evident even under controlled laboratory conditions
-\citep{knapp2016}, indicative of host resistance against Bd infection
+\cite{knapp2016}, indicative of host resistance against Bd infection
 (and not simply an effect of factors external to individual frogs, e.g.,
@@
 mechanisms, including natural selection for more resistant genotypes
-\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
+\cite{savage2016, grogan2018b}, acquired immunity \cite{grogan2018a},
 and/or inherent between-population differences that pre-date Bd
@@
 frequency of adaptive alleles and restores positive population growth
-\citep{carlson2014, searle2020}. This intriguing possibility also
+\cite{carlson2014, searle2020}. This intriguing possibility also
 suggests an opportunity to expand recovery beyond the spatial scale
@@
 recovering populations in reintroductions to vacant habitats
-(Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.
+(Figure~\ref{fig-recovery-model}) \cite{joseph2018, mendelson2019}.
 
@@
 mortality (Figure S2)
-\citep{joseph2018, vredenburg2010}. Although it is possible that the
+\cite{joseph2018, vredenburg2010}. Although it is possible that the
 observed relatively small changes in load are a consequence of
@@
 to determine the self-sustainability of a reintroduced population
-\citep[for an example in MYL frogs, see][]{joseph2018}, the use of
+\cite[for an example in MYL frogs, see][]{joseph2018}, the use of
 proxies is essential for providing shorter-term insights into
@@
 surveys may be necessary to confirm that populations are in fact
-self-sustaining \citep{joseph2018}. To extend our inferences of
+self-sustaining \cite{joseph2018}. To extend our inferences of
 population establishment beyond those possible from the site-specific
@@
 in 4 populations that have not yet experienced a Bd-caused epizootic
-(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
+(``naive'') \cite{zhou2015} versus in 5 populations that experienced a
 Bd epizootic during the past several decades and have since recovered to
 varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
-\citep{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
+\cite{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
 populations are based on 10-20 years of VES and Bd surveillance using
-skin swabbing \citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
+skin swabbing \cite[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
 populations are characterized by large numbers of adults (i.e, typically
 1000s), Bd prevalence that is generally 0\% except during occasional Bd
-failed invasions \citep[during which Bd loads remain very
+failed invasions \cite[during which Bd loads remain very
 low,][]{wilber2022}, and no history of Bd epizootics since we first
 surveyed these populations in the late 1990s and early 2000s
-\citep{zhou2015}. In contrast, recovering populations exist in an
-enzootic state \citep{briggs2010}, characterized by smaller numbers of
-adults (generally \textless{} 500), high Bd prevalence \citep[often
+\cite{zhou2015}. In contrast, recovering populations exist in an
+enzootic state \cite{briggs2010}, characterized by smaller numbers of
+adults (generally \textless{} 500), high Bd prevalence \cite[often
 \textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
 that are typically well below the level expected to cause mortality
-\citep{vredenburg2010}. Naive and recovering populations can be
+\cite{vredenburg2010}. Naive and recovering populations can be
 identified unambiguously using these differences in Bd prevalence and
@@
 the strong signature of isolation-by-distance that is characteristic of
-MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
+MYL frogs \cite{rothstein2020, poorten2017, byrne2023}. Importantly,
 each cluster contained at least one population from both the naive and
@@
 Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
-extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
+extended MHC Class III region (HSP70, LSM2, VARS) \cite{ohta2006}.
 Therefore, this region under selection is part of the extended MHC Class
@@
 Disease-induced population declines are decimating global biodiversity
-\citep{daszak2000}, but broadly-applicable strategies to recover
-affected species are generally lacking \citep[e.g.,][]{garner2016}.
+\cite{daszak2000}, but broadly-applicable strategies to recover
+affected species are generally lacking \cite[e.g.,][]{garner2016}.
 Here, we tested the possibility that populations of resistant
@@
 infectious diseases. In light of the generally low success rate of
-amphibian reintroduction efforts \citep{dodd2005}, our success in
+amphibian reintroduction efforts \cite{dodd2005}, our success in
 reestablishing MYL frog populations via translocation of resistant
@@
 establishment of Bd in previously-naive populations, adult frogs develop
-high Bd loads that lead to mass die-offs \citep{vredenburg2010}. In
+high Bd loads that lead to mass die-offs \cite{vredenburg2010}. In
 contrast, in recovering populations adult frogs typically have
 low-to-moderate and relatively constant Bd loads and mass die-offs are
-not observed \citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
+not observed \cite[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
 naive and recovering populations are also observed in controlled
-laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
+laboratory studies \cite[see Figure 4 in][]{knapp2016}, and clearly
 indicate that frogs from recovering populations exhibit resistance
@@
 contribute to disease resistance, including in amphibians
-\citep{zamudio2020}. For example, the MHC plays an important role in
+\cite{zamudio2020}. For example, the MHC plays an important role in
 immunity. In our study, we identified a region that shows evidence of
@@
 corroborate numerous previous studies linking MHC genes to amphibian
-resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
+resistance against Bd \cite[e.g.,][]{savage2011, bataille2015}.
 Similarly, the region with the strongest indication of directional
@@
 have found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
-in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
+\cite[e.g.,][]{grogan2018b, ellison2014} and in populations differing
+in Bd susceptibility \cite{grogan2018b}. This gene is also strongly
 linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. We also identified a region,
+phenotype \cite{robledo2020, robledo2018}. We also identified a region,
 characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
 that contained the complement genes C6 and C7. The complement system
-plays an important role in innate immunity \citep{rieraromo2016}, and
+plays an important role in innate immunity \cite{rieraromo2016}, and
 our results could indicate that balancing selection is acting in this
 region of the genome to favor a diverse set of alleles, as is known for
-C6 in humans \citep{soejima2005}. Based on the analysis of individual
+C6 in humans \cite{soejima2005}. Based on the analysis of individual
 outlier variants, the RIN3 gene showed a consistent pattern of allele
@@
 response and in \emph{Xenopus} is expressed during appendage
-regeneration \citep{fukazawa2009}. Finally, the outlier variant with the
+regeneration \cite{fukazawa2009}. Finally, the outlier variant with the
 lowest p-value was the uncharacterized gene LOC108802036. In the genome
 of another frog species, this gene is located adjacent to a type I
-interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
+interferon gene (Np-IFNi2) \cite{gan2018}, and together with GVINP1
 further suggests the importance of interferon-related genes in this
@@
 given that reintroduced MYL frog populations may often take decades to
-achieve our ultimate goal of self-sustainability \citep{joseph2018}.
+achieve our ultimate goal of self-sustainability \cite{joseph2018}.
 Making well-supported projections about the long-term outcome of
@@
 important to the process of adaptive management of species
-reintroduction programs \citep{seddon2007}, including the one we are
+reintroduction programs \cite{seddon2007}, including the one we are
 carrying out for MYL frogs. Specifically, the combined results from our
@@
 evident. In an earlier study of 2 of our 12 translocated populations
-\citep{joseph2018}, Bd infection and load had detectable effects on the
+\cite{joseph2018}, Bd infection and load had detectable effects on the
 survival of adults and may have influenced population establishment
-(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
+(sites referred to as ``Alpine'' and ``Subalpine'' in \cite{joseph2018}
 are identified as ``70550'' and ``70505'' in the current study).
@@
 immediately following metamorphosis (``metamorphs'') are highly
-susceptible to Bd infection \citep{ellison2019} and as a result
-experience high mortality \citep{rachowicz2006}. This high
+susceptible to Bd infection \cite{ellison2019} and as a result
+experience high mortality \cite{rachowicz2006}. This high
 susceptibility of metamorphs is documented in numerous species of
 anurans, and may result from the poorly developed immune system
-characteristic of this life stage \citep{humphries2022}. In naturally
+characteristic of this life stage \cite{humphries2022}. In naturally
 recovering and translocated MYL frog populations, we suggest that the
@@
 relatively resistant, Bd infection continues to have important limiting
-effects on recovering populations \citep[see also][]{hollanders2022}.
+effects on recovering populations \cite[see also][]{hollanders2022}.
 
@@
 hundreds of amphibian species, some of which are now extinct in the wild
-\citep{scheele2019}. This extraordinary impact on global amphibian
+\cite{scheele2019}. This extraordinary impact on global amphibian
 biodiversity is compounded by the lack of any effective and broadly
 applicable strategies to reverse these impacts
-\citep{garner2016, knapp2022}. Importantly, in addition to the natural
-recovery documented for MYL frogs \citep{knapp2016}, other amphibian
+\cite{garner2016, knapp2022}. Importantly, in addition to the natural
+recovery documented for MYL frogs \cite{knapp2016}, other amphibian
 species are also showing evidence of post-epizootic recovery in the
-presence of Bd \citep{scheele2017, voyles2018} and suggest the
+presence of Bd \cite{scheele2017, voyles2018} and suggest the
 possibility of also using animals from these recovering populations to
@@
 collected using several years of VES and skin swab collections
-\citep{knapp2016}, and results from population genetic analyses
-\citep{poorten2017}. The populations that we selected contained hundreds
+\cite{knapp2016}, and results from population genetic analyses
+\cite{poorten2017}. The populations that we selected contained hundreds
 of \emph{R. sierrae} adults and thousands of tadpoles. These relatively
 high abundances were the result of recent increases following previous
-Bd-caused declines \citep{knapp2016}. As is typical for recovering MYL
+Bd-caused declines \cite{knapp2016}. As is typical for recovering MYL
 frog populations, Bd prevalence in the donor populations was high
@@
 level at which frog mortality is expected (log\textsubscript{10}(load)
-\(\approx\) 5.78 copies)\citep{vredenburg2010, joseph2018}. Recipient
+\(\approx\) 5.78 copies)\cite{vredenburg2010, joseph2018}. Recipient
 sites to which frogs were translocated were chosen based on previous
@@
 or characteristics that suggested high quality habitat for this species
-\citep{knapp2005}. At the beginning of this study, we had a relatively
+\cite{knapp2005}. At the beginning of this study, we had a relatively
 limited understanding of the factors that affect habitat quality. In
@@
 stages have high detectability, and counts are highly repeatable and
-provide estimates of relative abundance \citep{knapp2000}.
+provide estimates of relative abundance \cite{knapp2000}.
 
@@
 translocated cohorts), as implemented by the mrmr package
-\citep{joseph2019} and using R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
+\cite{joseph2019} and using R Statistical Software
+\cite[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
 Frog population recovery - CMR model structure} for details). Briefly,
 the model tracks the states of \emph{M} individuals that comprise a
-superpopulation made up of real and pseudo-individuals \citep[see][ for
+superpopulation made up of real and pseudo-individuals \cite[see][ for
 details]{joseph2018}. The possible states of individuals include ``not
@@
 translocation, we used multilevel Bayesian models
-\citep{gelman2013, gabry2019}. Included predictor variables describe
+\cite{gelman2013, gabry2019}. Included predictor variables describe
 characteristics of sites, translocated cohorts, and individuals (Bd
@@
 site\_id. We performed all analyses with the rstanarm package
-\citep{rstanarm2022} and R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022}. For all models, we used default,
+\cite{rstanarm2022} and R Statistical Software
+\cite[v4.4.4,][]{rsoftware2022}. For all models, we used default,
 weakly informative priors, four chains, and 5000 iterations each for
@@
 and Rhat, and evaluated model fit using leave-one-out cross-validation
-\citep{vehtari2016}, as implemented by the loo package
-\citep{vehtari2022}. (See \textbf{Supporting Information - Frog
+\cite{vehtari2016}, as implemented by the loo package
+\cite{vehtari2022}. (See \textbf{Supporting Information - Frog
 population recovery - Among-site survival modeling} for details.)
@@
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see \textbf{Supporting Information - Frog population recovery -
+\cite[see \textbf{Supporting Information - Frog population recovery -
 Laboratory methods} for details]{boyle2004}. To assess the magnitude of
@@
 decision because (i) translocated populations are infected with Bd at
-high prevalence \citep{joseph2018}, and (ii) host density does not seem
+high prevalence \cite{joseph2018}, and (ii) host density does not seem
 to play a significant role in multi-year Bd infection dynamics in this
-system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
+system \cite{wilber2022}. Thus, ignoring Bd infection dynamics and
 instead assuming all host vital rates are in the presence of high Bd
@@
 
-We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
+We collected DNA samples via buccal swabbing \cite{broquet2007} from 53
 \emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
@@
 frogs per population. To minimize potential confounding effects caused
-by variation in frog genotypes across latitude \citep{byrne2023}, we
+by variation in frog genotypes across latitude \cite{byrne2023}, we
 selected sampling sites such that both population types were represented
@@
 DNEasy manufacturer's protocols. We sequenced the samples using an exome
-capture assay as described in \citep{byrne2023}. Briefly, genomic
+capture assay as described in \cite{byrne2023}. Briefly, genomic
 libraries were prepared and captured using a custom Nimblegen capture
@@
 Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
-SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
-bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
+\cite{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
+SRA: SRS6533475 \cite{hon2020}), with repetitive elements masked using
+bwa (``mem'' mode) \cite{li2013}. Exact PCR duplicates were marked
 using Picard. Variants were then called following GATK best practices
-(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
+(v.4.2.0.0 \cite{van2020}). Briefly, raw variants were called for each
 sample using HaplotypeCaller and combined using CombineGVCFs. Next,
@@
 a PCA using the glPCA function in the adegenet R package
-\citep{jombart2008}. To detect regions of the genome that differed
+\cite{jombart2008}. To detect regions of the genome that differed
 between naive and recovering populations, i.e., regions under selection,
@@
 of 148,307 high quality variants across 127 contigs using GEMMA
-\citep{zhou2014}. GEMMA calculates and incorporates a relatedness matrix
+\cite{zhou2014}. GEMMA calculates and incorporates a relatedness matrix
 for input samples, allowing us to account for relatedness and population
@@
 extracted the reference genome sequence surrounding the variant using
-the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
+the bedtools ``getfasta'' function \cite{quinlan2010}. Next, we
 re-annotated each sequence using BLAST to get the predicted gene
-location based on the closest annotated reference \citep{altschul1997}.
+location based on the closest annotated reference \cite{altschul1997}.
 We then translated each variant to amino acids and aligned this
 translation to that of the gene annotation to ensure proper frame of
-reference using Geneious \citep{kearse2012}. After ensuring proper
+reference using Geneious \cite{kearse2012}. After ensuring proper
 translation, we characterized variants as within or outside the coding
@@
 recovering individuals for all bi-allelic SNPs in the 30 largest contigs
-(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
+(98\% of all SNPs) using VCFtools \cite{danecek2011}. Next, we
 calculated per-site nucleotide diversity \(\pi\) separately for
@@
 regions of the genome). We then used the GenWin R package
-\citep{beissinger2015} to conduct a splined discrete window analysis for
+\cite{beissinger2015} to conduct a splined discrete window analysis for
 \emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
@@
 positive and false negative results that occur using other window-based
-methods \citep{beissinger2015}. This method also calculates a
+methods \cite{beissinger2015}. This method also calculates a
 W-statistic allowing for outlier identification. We identified outliers

### 2024-04-05 doc/manuscript/translocation.qmd
Add initial responses to reviewer 2 comments

@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The donor populations included 2 of the 5 recovering populations used in the frog evolution study referenced above (@fig-allelefrequencies: population 1 and 4), and these donor populations contributed frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
+The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog evolution in response to Bd:** population 1 and 4).
+In addition, this subset of 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
@@
 This proxy of site quality is important in the MYL frog system because accurately predicting the ability of a site to support a viable frog population *a priori* remains difficult, even after conducting 24 translocations over 16 years.
+
+The results of the meta-analysis indicate the important influence of several variables on adult survival, and by extension on population viability.
+Although we assume that the effects of winter severity and frog sex are direct effects of the variables themselves, the elevation and donor population variables are likely serving as proxies for a range of factors.
+For example, elevation is positively correlated with several attributes that may influence frog survival, including the presence of boulder habitat that is important for frog overwintering, abundance of frog predators including garter snakes, and frog developmental rates.
+The latter can influence the duration of frog life stages that have differential susceptibility to chytridiomycosis (see below).
+The duration of the tadpole and juvenile stages increases with elevation, but the fact that population growth rate is relatively insensitive to tadpole and juvenile vital rates suggests that developmental rates of early frog life stages have little effect on the predicted long-term viability of reintroduced populations.
+Nonetheless, a clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
 
@@
 Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
-Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
+Samples sizes are provided immediately above the x-axis.
+The red dashed horizontal line indicates the approximate load above which severe disease is typically observed \[@vredenburg2010\].
+This threshold is commonly exceeded during epizootics but only rarely in recovering populations \[@briggs2010\].
+](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
 

### 2024-04-05 doc/manuscript/translocation.qmd
Incorporate reviewer 3 feedback

--- 2024-01-06 doc/manuscript/translocation.qmd
@@
 We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$.
-The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
+The states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
+Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and end as late as September).
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.
@@
   0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 1 & 1 & \text{Dead} \\
+  0 & 0 & 1 & \text{Dead} \\
 \end{block}

### 2024-04-05 doc/manuscript/translocation.qmd
rm end

@@
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and end as late as September).
+Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September).
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.

### 2024-04-05 doc/manuscript/translocation.qmd
Add elasticity for p_L2 and p_J1

@@
 We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$.
-The states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
+The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September).
+Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.
@@
   0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 0 & 1 & \text{Dead} \\
+  0 & 1 & 1 & \text{Dead} \\
 \end{block}
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles).
+Elasticity of $\lambda$ with changes in six parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year 2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year 1 juveniles mature to adults).
+We did not include $p_{L_1}$ as this parameter is generally 1 across populations (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}

### 2024-04-05 code/extinction_and_growth_rate_analysis.R

@@ -169,7 +169,7 @@ for(l in 1:nrow(all_surv)){
 }
 
 lake_elas_dt = data.table(do.call(rbind, lake_elas))
-colnames(lake_elas_dt) = c("\U03C3_AR", "F", "\U03C3_J1", "\U03C3_J2") #c("sigma_AR", "F", "sigma_J1", "sigma_J2")
+colnames(lake_elas_dt) = c("\U03C3_AR", "F", "\U03C3_J1", "\U03C3_J2", "p_L2", "p_J1") #c("sigma_AR", "F", "sigma_J1", "sigma_J2")
 lake_elas_dt$lake = names(lake_elas)
 
 lake_elas_dt_melt = melt(lake_elas_dt, id.vars="lake")
@@ -473,6 +473,6 @@ ptraj = ggplot() + geom_line(data=pred_traj, aes(x=year, y=abund, group=sim, col
 myplot = (ptile + pext + ptraj) + plot_annotation(tag_levels="A", tag_suffix="")
 ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.jpg"), width=11, height=5, dpi=300)
 
-myplot = (p2) + plot_annotation(tag_levels="A", tag_suffix="")
+myplot = (p2)# + plot_annotation(tag_levels="A", tag_suffix="")
 ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=5, height=4, dpi=300)
 

### 2024-04-05 code/population_viability_functions.R

@@ -345,10 +345,12 @@ get_sensitivity = function(params){
 	dAdF = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, params$p_F*params$sigma_AR, 0, 0, 0, 0, 0, 0, params$p_F*params$sigma_AT, 0, 0, 0, 0, 0, 0) 
 	dAdsj1 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, params$p_J1, (1 - params$p_J1), 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
 	dAdsj2 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
-	dmatA = t(rbind(dAdsar, dAdF, dAdsj1, dAdsj2))
+	dAdpl2 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, params$sigma_L2, -params$sigma_L2, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
+	dAdpj1 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, params$sigma_J1, -params$sigma_J1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
+	dmatA = t(rbind(dAdsar, dAdF, dAdsj1, dAdsj2, dAdpl2, dAdpj1))
 
 	# From Caswell 2019
[…]

### 2024-04-05 doc/manuscript/translocation.qmd
Update Figure S5 legend

@@
 Elasticity of $\lambda$ with changes in six parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year 2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year 1 juveniles mature to adults).
-We did not include $p_{L_1}$ as this parameter is generally 1 across populations (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
+We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study populations (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}

### 2024-04-05 doc/manuscript/translocation.qmd
Remove typo

--- 2024-04-05 doc/manuscript/translocation.qmd
@@
 Elasticity of $\lambda$ with changes in six parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year 2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year 1 juveniles mature to adults).
-We did not include $p_{L_1}$ as this parameter is generally 1 across populations (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
+We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}

### 2024-04-09 code/extinction_and_growth_rate_analysis.R
Add elasticity and sensitivity

@@ -169,16 +169,30 @@ for(l in 1:nrow(all_surv)){
 }
 
 lake_elas_dt = data.table(do.call(rbind, lake_elas))
-colnames(lake_elas_dt) = c("\U03C3_AR", "F", "\U03C3_J1", "\U03C3_J2", "p_L2", "p_J1") #c("sigma_AR", "F", "sigma_J1", "sigma_J2")
+colnames(lake_elas_dt) = c("\U03C3_AR", "F", "\U03C3_J1", "\U03C3_J2", "p_L2", "p_J1", "\U03C3_L1", "\U03C3_L2", "\U03C3_L3") #c("sigma_AR", "F", "sigma_J1", "sigma_J2")
 lake_elas_dt$lake = names(lake_elas)
 
 lake_elas_dt_melt = melt(lake_elas_dt, id.vars="lake")
 
+lake_sens_dt = data.table(do.call(rbind, lake_sens))
+colnames(lake_sens_dt) = c("\U03C3_AR", "F", "\U03C3_J1", "\U03C3_J2", "p_L2", "p_J1", "\U03C3_L1", "\U03C3_L2", "\U03C3_L3") #c("sigma_AR", "F", "sigma_J1", "sigma_J2")
+lake_sens_dt$lake = names(lake_sens)
+
+lake_sens_dt_melt = melt(lake_sens_dt, id.vars="lake")
+
 p2 = ggplot(lake_elas_dt_melt) + geom_bar(aes(x=variable, y=value, fill=lake), stat="identity", position="dodge") +
 														scale_fill_manual(values=colors) +
 														xlab("Parameter") + ylab("Elasticity of \U03BB") + theme_classic() +
-														guides(fill=guide_legend(title="Site"))
+														guides(fill="none") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 														# theme(legend.position = c(0.7, 0.7))
+
+p2b = ggplot(lake_sens_dt_melt) + geom_bar(aes(x=variable, y=value, fill=lake), stat="identity", position="dodge") +
+														scale_fill_manual(values=colors) +
+														xlab("Parameter") + ylab("Sensitivity of \U03BB") + theme_classic() +
+														guides(fill=guide_legend(title="Site")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+														# theme(legend.position = c(0.7, 0.7))
+
+sens_plot = (p2 + p2b) + plot_annotation(tag_levels="A", tag_suffix="")
 
[…]

### 2024-04-09 code/population_viability_functions.R

@@ -347,10 +347,13 @@ get_sensitivity = function(params){
 	dAdsj2 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
 	dAdpl2 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, params$sigma_L2, -params$sigma_L2, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
 	dAdpj1 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, params$sigma_J1, -params$sigma_J1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
-	dmatA = t(rbind(dAdsar, dAdF, dAdsj1, dAdsj2, dAdpl2, dAdpj1))
+	dAdsl1 = c(0, params$p_L1, 0, 1 - params$p_L1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
+	dAdsl2 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, params$p_L2, 1 - params$p_L2, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
[…]

### 2024-04-09 doc/manuscript/translocation.qmd
Update sensitivity text in ms

--- 2024-04-05 doc/manuscript/translocation.qmd
@@
 This is suggestive of parallel selection at this locus across multiple populations.
-The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5–9; @fig-allelefrequencies, @fig-selectionresults A SI).
+The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
 Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that suggest parallel evolutionary changes following Bd exposure.
@@
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity of $\lambda$ to four key model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp SI).
+Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp SI).
 This also helped us determine where in the model environmental variation in juvenile survival and recruitment would have the largest effects on population dynamics.
@@
 
-Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp SI).
+Second, we performed a local elasticity and sensitivity analysis on $\lambda$ with respect to parameters $\sigma_{L_1}$, $\sigma_{L_2}$, $\sigma_{L_3}$, $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, $F$, $p_{L_2}$, $p_{J_1}$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp SI).
+Note that $\lambda$ is equally or more sensitive to changes in adult survival ($\sigma_{A_R}$) and survival of year-1 juveniles ($\sigma_{J_1}$) than it is to other model parameters, providing additional justification for focusing on variation in $\sigma_{A_R}$ and $\sigma_{J_1}$ in our viability analyses.
 
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in six parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year 2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year 1 juveniles mature to adults).
-We did not include $p_{L_1}$ as this parameter is generally 1 across populations (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
-Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
+A.
+Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+B.
+Sensitivity of $\lambda$ with changes in the same nine parameters.
+We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
+Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
 

### 2024-04-09 code/extinction_and_growth_rate_analysis.R

@@ -488,5 +488,5 @@ myplot = (ptile + pext + ptraj) + plot_annotation(tag_levels="A", tag_suffix="")
 ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.jpg"), width=11, height=5, dpi=300)
 
 myplot = (p2 + p2b) + plot_annotation(tag_levels="A", tag_suffix="")
-ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=5, height=4, dpi=300)
+ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=9, height=4, dpi=300)
 

### 2024-04-09 doc/manuscript/translocation.qmd
Add description to sensitivity legend

@@
 We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$.
-The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
+The states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Each primary period $t$ occurs within one year, but within a year there can be multiple primary periods.
+Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September).
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.
@@
 Sensitivity of $\lambda$ with changes in the same nine parameters.
+Note that $F$ has different units than the other parameters, which is why its sensitivity is low.
+Sensitivities of all other parameters are comparable.
 We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.

### 2024-04-09 doc/manuscript/translocation.qmd
Add edits in response to reviewer 2 & 3 comments

--- 2024-04-05 doc/manuscript/translocation.qmd
@@
 
-Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in emerging infectious diseases.
-The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
+Revised version: Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
+The emergence of the lethal fungal pathogen (Batrachochytrium dendrobatidis; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
+With no broadly applicable methods available to reverse these impacts, the future of many amphibians appears grim.
+The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
+Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
+Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes.
+We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
+The results also indicate that the likelihood of establishment is influenced by site, population, and frog attributes.
+In addition, results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
+To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
+We found substantial differences between these categories, including changes in immune function loci that may confer increased resistance.
+These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
+
+Original version: Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in emerging infectious diseases.
+The emergence of the lethal fungal pathogen (Batrachochytrium dendrobatidis; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
 With no broadly applicable methods available to reverse these impacts in the wild, the future of many amphibians appears grim.
@@
 Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@raberg2009] and facilitate population persistence and recovery [@brannelly2021].
-Host immunity and evolution both play important roles in the development of resistance and tolerance, and utilizing these factors would seem a promising approach to developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
-However, several aspects of the amphibian-Bd system present difficult obstacles, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
+Host immunity and evolution both play important roles in the development of resistance and tolerance, and an enhanced understanding of these processes is important for developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
+However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
 These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
@@
 Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
-This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection (and not simply an effect of factors external to individual frogs, e.g., environmental conditions).
+This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection and not simply an effect of factors external to individual frogs (e.g., environmental conditions).
 In addition to frogs from recovering populations having higher resistance to Bd infection than those from naive populations, they could also have higher tolerance, but no data are currently available to evaluate this possibility.
 Therefore, we focus on resistance throughout this paper.
-The observed resistance of MYL frogs could be the result of several non-mutually exclusive mechanisms, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
+The observed resistance of MYL frogs could be the result of several mechanisms that are not mutually exclusive, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
 The possible evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
-This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by utilizing resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
+This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
 
 In the current study, we had three primary objectives.
-First, to determine whether the reintroduction of resistant MYL frogs obtained from populations recovering from Bd-caused declines allows the successful reestablishment of extirpated populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort (@fig-recovery-model).
-Second, to extend our inferences of population recovery well beyond the temporal extent of our reintroduction study, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
-Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), we conducted a genomic study using exome capture methods to determine whether MYL frogs in recovering populations show evidence of selection and whether these genomic changes are associated with resistance (@fig-recovery-model).
+First, determine whether the reintroduction of resistant MYL frogs, obtained from populations recovering from Bd-caused declines, allows the reestablishment of populations despite ongoing disease (@fig-recovery-model).
+We accomplished this using a 15-year large-scale reintroduction study.
+Second, extend our inferences of population recovery beyond the temporal extent of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
+We did this using a stage-structured matrix model.
+Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show evidence of selection and whether these genomic changes are associated with resistance (@fig-recovery-model).
+We achieved this with a genomic study that utilized exome capture methods.
 
@@
 We estimated survival and recruitment for each site using open population CMR models based on @joseph2018.
-For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
+For each individual <!--# in a superpopulation of real and psuedo-individuals, --> $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
 Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.
@@
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)?)
-.
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)) .
 This yielded posterior distributions for survival probabilities between primary periods.
@@
 Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).
-Samples sizes are provided immediately above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
+Samples sizes are provided immediately above the x-axis.
+The red dashed horizontal line indicates the approximate load above which severe disease is typically observed [@vredenburg2010].
+This threshold (5.8 ITS1 copies - on a log~10~ scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations [@briggs2010].](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
 
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in six parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year 2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year 1 juveniles mature to adults).
-We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles must spend at least a year as tadpoles before metamorphosing.
+Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles).
 Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}

### 2024-04-10 doc/manuscript/translocation.qmd
Minor edit to SI CMR methods suggested by Reviewer 3

--- 2024-04-09 doc/manuscript/translocation.qmd
@@
 We estimated survival and recruitment for each site using open population CMR models based on @joseph2018.
-For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
+For each individual in a superpopulation of real and psuedo-individuals, $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
 Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.

### 2024-04-10 doc/manuscript/translocation.qmd
Minor edit to viability section of Discussion

--- 2024-04-09 doc/manuscript/translocation.qmd
@@
 The latter can influence the duration of frog life stages that have differential susceptibility to chytridiomycosis (see below).
-The duration of the tadpole and juvenile stages increases with elevation, but the fact that population growth rate is relatively insensitive to tadpole and juvenile vital rates suggests that developmental rates of early frog life stages have little effect on the predicted long-term viability of reintroduced populations.
+The duration of the tadpole and juvenile stages increases with elevation, but the fact that long-run population growth rate is relatively insensitive to variation in these durations suggests that developmental rates of early frog life stages have little effect on the predicted long-term viability of reintroduced populations.
 Nonetheless, a clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.

### 2024-04-10 doc/manuscript/translocation.qmd

@@
 We estimated survival and recruitment for each site using open population CMR models based on @joseph2018.
-For each individual <!--# in a superpopulation of real and psuedo-individuals, --> $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
+For each individual in a superpopulation of real and psuedo-individuals, $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
 Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.
@@
 ![Sensitivity analysis of the stage-structured MYL frog model.
-Elasticity of $\lambda$ with changes in four parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), *F (*number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), and $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles).
-Elasticity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
+A.
+Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+B.
+Sensitivity of $\lambda$ with changes in the same nine parameters.
+Note that $F$ has different units than the other parameters, which is why its sensitivity is low.
+Sensitivities of all other parameters are comparable.
+We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
+Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
 

### 2024-04-10 code/translocation_survival_analysis_supp.Rmd

@@
     facet_grid(. ~ site_id1) +
+    geom_hline(yintercept = 5.78, color = "red", linetype = 2) +
     theme_bw() +

### 2024-04-10 doc/manuscript/translocation.qmd
Edits to Abstract

@@
 
-Revised version: Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
-The emergence of the lethal fungal pathogen (Batrachochytrium dendrobatidis; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
-With no broadly applicable methods available to reverse these impacts, the future of many amphibians appears grim.
+Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
+The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
+Lacking any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
 The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
@@
 We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
-The results also indicate that the likelihood of establishment is influenced by site, population, and frog attributes.
-In addition, results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
+In addition, the likelihood of establishment is influenced by site, population, and frog attributes.
+Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
 To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
-We found substantial differences between these categories, including changes in immune function loci that may confer increased resistance.
+We found numerous differences between these categories, including in immune function loci associated with disease resistance, consistent with evolutionary changes in response to Bd exposure.
 These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
-
-Original version: Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in emerging infectious diseases.
-The emergence of the lethal fungal pathogen (Batrachochytrium dendrobatidis; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
-With no broadly applicable methods available to reverse these impacts in the wild, the future of many amphibians appears grim.
-The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
-Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
-Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes and/or acquired immunity.
-We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
-In addition, results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
-To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
-We found substantial differences between these categories, including changes in immune function loci that may confer increased resistance, consistent with evolutionary changes in response to Bd exposure.
-These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species.
-This example has broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
 

### 2024-04-18 doc/manuscript/translocation.qmd
Add new figure, text edits, and latex files

@@
 ---
-#format: pdf
+# format: pdf
 format:
@@
 We accomplished this using a 15-year large-scale reintroduction study.
-Second, extend our inferences of population recovery beyond the temporal extent of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
+Second, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
 We did this using a stage-structured matrix model.
-Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show evidence of selection and whether these genomic changes are associated with resistance (@fig-recovery-model).
-We achieved this with a genomic study that utilized exome capture methods.
+Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show genomic patterns consistent with selection for resistance (@fig-recovery-model).
+We achieved this with a comparative study using exome capture methods applied to naive and recovering populations.
 
@@
 
-#### Frog evolution in response to Bd
-
-Results from the preceding sections indicate the critical role of frog resistance in post-translocation frog survival, population growth, and population viability.
-As such, identifying the mechanisms underlying this resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
+#### Frog genome evolution
+
+Results from the preceding sections indicate the critical role of intrinsic factors (such as frog resistance) in post-translocation frog survival, population growth, and population viability.
+As such, identifying the mechanisms underlying resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
 Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs the role of disease-mediated selection in these processes remains unknown.
@@
 Finally, there is no potential for frog dispersal between the 9 study populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
+
+It is important to note that our genomic analysis is necessarily limited by the small number of naive populations remaining on the landscape (nearly all of which we sampled as part of the current study) and their clumped distribution ([@fig-allelefrequencies]).
+These characteristics of naive populations precluded a replicated analyses among pairs of geographically proximate naive and recovering populations.
+Instead, to obtain adequate power, we conducted a pooled analysis across population types to identify genomic regions showing patterns consistent with selection.
+Our comparative approach cannot provide a direct assessment of the strength or timing of natural selection, but does allow us to identify specific regions of the genome that merit further study for their potential contribution to frog resistance in recovering populations.
 (See **Supporting Information - Frog evolution in response to Bd - Study design** for additional details regarding the study design.)
 
-We conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations, and then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions under selection).
+As a first step in our pooled analysis, we conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations.
+We then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions possibly under selection).
 First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
@@
 
-Results from the individual variant and splined window analyses show that recovering populations have signatures of selection on multiple regions of the genome.
+Results from the individual variant and splined window analyses show that recovering populations show patterns consistent with selection on multiple regions of the genome.
 The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI).
@@
 This is suggestive of parallel selection at this locus across multiple populations.
-The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found a statistically significant signal of selection in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that suggest parallel evolutionary changes following Bd exposure.
+The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found statistically significant differences in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes following Bd exposure.
+<!--# Add text citing new figure -->
 
@@
 Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
+The largest $\pi_{diff}$, suggestive of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
 This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
@@
 In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
+Therefore, this region is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
 
 Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog evolution in response to Bd - Genetic diversity** for details).
-Thus, despite localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
+Thus, despite patterns consistent with localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
 In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog evolution in response to Bd** **- GO analysis** for details).
 
-In summary, our genomic results indicate that the exomes of frogs from naive and recovering populations show substantial differences, consistent with parallel evolutionary changes following Bd exposure.
-The regions under selection contain several immunologically-relevant genes and gene families that are directly linked to disease resistance in other taxa.
+In summary, our genomic results indicate that frogs from naive and recovering populations show differences at several immune-function loci, consistent with selection following Bd exposure.
 
@@
 Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
-Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that the recovering populations are likely to persist over a 50-year period, (iii) that there are substantial genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic regions under selection contain genes related to disease resistance.
+Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the recovering populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
 Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
@@
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
-This resistance could in theory be due to several factors, including natural selection for more resistant genotypes, acquired immunity, and/or inherent between-population differences that pre-date Bd exposure, but until now evidence to evaluate the role of evolution was lacking.
-
-Results from our genomic analyses suggest that natural selection for adaptive alleles is at least partially responsible for the increased resistance of frogs in recovering populations.
-We identified multiple specific alleles and genomic regions showing signatures of selection between adjacent naive and recovering MYL frog populations, consistent with selection following Bd exposure.
-These analysis are based on samples collected from virtually all of the MYL frog populations remaining in a naive state, as well as adjacent recovering populations.
-This study design produced genetic clusters that each contained at least one naive and one recovering population, allowing us to detect selection without the confounding effects of population structure.
-In addition, we did not find a reduction in overall genetic variation in the recovering populations, suggesting that despite localized selection in the genome, these populations retain adequate genetic diversity for long-term persistence.
-
-Importantly, some genomic regions that we identified as under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
+This resistance could be due at least in part to natural selection for more resistant genotypes, but until now evidence to evaluate the possible role of evolution was lacking.
+
+Our genomics study used a pooled design that compared MYL frogs from naive and recovering populations distributed across a similar latitudinal range.
+The study design produced frog genetic clusters that each contained at least one naive and one recovering population, allowing us to infer selection while attempting to minimize the confounding effects of population structure.
+We identified multiple specific alleles and genomic regions showing differences between naive and recovering MYL frog populations, consistent with selection following Bd exposure.
+In addition, we did not find a reduction in overall genetic variation in recovering populations, suggesting that despite possible localized selection in the genome, some of these populations likely retain adequate genetic diversity for long-term persistence.
+
+Importantly, some genomic regions that we identified as possibly under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
 For example, the MHC plays an important role in immunity.
@@
 These results corroborate numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest indication of directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
+Similarly, the region with the strongest pattern of possible directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
 Several previous studies of amphibians have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
-This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [@robledo2020; @robledo2018].
+This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [i.e., gill damage and amoebic load as heritable host resistance traits, @robledo2020; @robledo2018].
 We also identified a region, characterized by high *F~ST~* and low $\pi_{diff}$, that contained the complement genes C6 and C7.
 The complement system plays an important role in innate immunity [@rieraromo2016], and our results could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
+Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, suggesting possible consistent selection in populations distributed across a wide geographic area.
 This gene is associated with immune response and in *Xenopus* is expressed during appendage regeneration [@fukazawa2009].
@@
 Although we assume that the effects of winter severity and frog sex are direct effects of the variables themselves, the elevation and donor population variables are likely serving as proxies for a range of factors.
-For example, elevation is positively correlated with several attributes that may influence frog survival, including the presence of boulder habitat that is important for frog overwintering, abundance of frog predators including garter snakes, and frog developmental rates.
+For example, elevation is correlated with several attributes that may influence frog survival, including the presence of boulder habitat (positive correlation) that is important for frog overwintering, abundance of frog predators including garter snakes (negative correlation), and frog developmental rates (negative correlation).
 The latter can influence the duration of frog life stages that have differential susceptibility to chytridiomycosis (see below).
@@
 In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
-Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an airtight example of evolutionary rescue will likely always be uncertain (given that can never have a perfect understanding of the past).
-Regardless, we provide an example from the wild that suggests that evolution can produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
+Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an airtight example of evolutionary rescue will likely always be uncertain (given that we can never have a perfect understanding of the past).
+Regardless, we provide an example from the wild that suggests that evolution might produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
 We expect that similar species recovery actions will be an essential tool in wildlife conservation in an era of accelerating global change.
@@
 
-#### Frog evolution in response to Bd
+#### Frog genome evolution
 
@@
 To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome that differed between naive and recovering populations, i.e., regions under selection, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
+To detect regions of the genome that differed between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
 For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
@@
 Hegeman and A.
-Lindauer--project and data management, field and laboratory work; A.
+Lindauer (project and data management, field and laboratory work); A.
 Barbella and K.
-Rose--laboratory work; C.
-Kamoroff and numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
-Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
+Rose-(laboratory work); C.
+Kamoroff and numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
+Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
 This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
@@
 Blue text indicates components that are included in the current study.
-The timeline shows the general sequence of the components, with the dotted portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model}
+We quantified frog survival and population viability following reintroductions directly using field studies and models, and evaluated "disease-mediated selection" indirectly using a comparison of frog exomes in naive and recovering populations.
+The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model}
 
@@
 
-![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale.
+![Genomic patterns consistent with selection on individual variants in recovering MYL frog populations at the landscape scale (based on a linear mixed model analysis).
 For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
@@
 
-![Evidence for selection on genomic regions in recovering MYL frog populations.
-Manhattan plot of the results from the splined window analysis showing outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
+![Patterns consistent with selection on genomic regions in recovering MYL frog populations based on a splined window analysis.
+Note that several key candidate genes (RIN3, TCF19, VARS) are outliers in both the individual variant analysis [@fig-allelefrequencies] and splined window analysis (this figure).
+Manhattan plots show outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
 In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
@@
 Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity analysis (**Supporting Information - Population viability modeling - Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
+We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
 
@@
 
-### Frog evolution in response to Bd
+### Frog genome evolution
 
@@
 Third, we calculated average $\pi$ per population within each of the 9 outlier windows identified in the splined window analysis.
+We compared average $\pi$ per population between naive and recovering population types, and also for each naive and recovering population in Yosemite.
+The latter comparison focused on Yosemite because these populations are the most relevant to the translocated populations described in the paper (see results below for additional details).
 
 Average individual-level heterozygosity and genome-wide population-level $\pi$ were similar between the naive and recovering groups (@fig-violinplot-heterozy SI; @fig-boxplot-genomewide-pi-by-pop SI).
-Within each of the 9 outlier windows, average $\pi$ shows considerable variation between populations (@fig-boxplot-pi-by-windowpop SI) and no obvious patterns between the naive and recovering groups.
+Within each of the 9 outlier windows, average $\pi$ shows considerable variation between populations and no obvious patterns between the naive and recovering populations (@fig-boxplot-pi-by-windowpop SI).
+For the four populations in Yosemite, within each of the 9 outlier windows, average $\pi$ per population was generally higher in the 3 recovering populations than in the 1 naive population (@fig-ave-pi-per-site-yose SI).
+ADD CONCLUDING SENTENCE OR TWO THAT SUMMARIZE THE SIGNIFICANCE OF THESE RESULTS?
 
@@
 We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="550"}
+Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="650"}
 
@@
 
-![For each of the 9 outlier windows identified by the overlapping splined windows for FST and $\pi$, boxplots showing average $\pi$ for naive and recovering populations.
+![For each of the 9 outlier windows identified by the overlapping splined windows for *F~ST~* and $\pi$, boxplots showing average $\pi$ for naive and recovering populations.
 The corresponding gene annotations for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_windowpop.png){#fig-boxplot-pi-by-windowpop width="500"}
+
+\newpage
+
+![For each study population located in Yosemite National Park (populations 1-4 shown in Site Map), average $\pi$ for the 9 outlier windows (panels A-I) identified in the splined window analysis.
+Populations 1 and 4 were the source of frogs for 20 of the 24 translocations and the donor population for the remaining four translocations is located adjacent to populations 3 and 4.
+Bars are color-coded by population type (naive, recovering), and gene annotations within each window are listed above each panel.
+Gene regions described in the paper are in panel A (Complement C6,C7), panel E (GVINP1), and panel F (extended MHC region that includes two SNPs identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by an asterisk).
+In the Site Map, the four populations included in panels A-I are color-coded by population type, and dashed lines delineate the genetic clusters as defined in [@fig-selectionresults SI, panel A].
+<!--# Update caption based on feedback from Allie and Bree -->](figures/ave_pi_per_site_yose.png){#fig-ave-pi-per-site-yose width="650"}
 

### 2024-04-18 doc/manuscript/translocation_bioarxiv.tex
Update PNAS template
Add new figure, text edits, and latex files

@@
 \significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. Emerging infectious
-diseases can have particularly devastating impacts, and few options
-exist to reverse these effects. We used large-scale reintroductions of
-disease-resistant individuals in an effort to recover a once-common frog
-species driven to near-extinction by a disease that has decimated
-amphibian biodiversity. Introduction of resistant frogs allowed
-reestablishment of viable populations in the presence of disease. In
-addition, resistance may be at least partially the result of natural
-selection at specific immune function genes, which show evidence for
-selection in recovering populations. The evolution of resistance and
-reintroduction of resistant individuals could play an important role in
-biodiversity conservation in our rapidly changing world.}
+critical for the conservation of biodiversity. We used reintroductions
+of disease-resistant individuals in an effort to recover a once-common
+frog species driven to near-extinction by a novel disease that has
+decimated amphibians worldwide. Reintroductions allowed establishment of
+viable populations in the presence of disease. In addition, resistance
+may be at least partially the result of natural selection at specific
+immune function genes. The evolution of resistance and reintroduction of
+resistant individuals could play an important role in biodiversity
+conservation in our rapidly changing world.}
 
@@
 Vast alteration of the biosphere by humans is causing a sixth mass
-extinction, driven in part by an increase in emerging infectious
-diseases. The emergence of the lethal fungal pathogen
-(\emph{Batrachochytrium dendrobatidis}; ``Bd'') has devastated global
-amphibian biodiversity, with hundreds of species experiencing declines
-or extinctions. With no broadly applicable methods available to reverse
-these impacts in the wild, the future of many amphibians appears grim.
-The once-common mountain yellow-legged (MYL) frog is emblematic of
-amphibians threatened by Bd. Although most MYL frog populations are
-extirpated following disease outbreaks, some persist and eventually
-recover. Frogs in these recovering populations have increased resistance
-against Bd infection, consistent with evolution of resistant genotypes
-and/or acquired immunity. We conducted a 15-year landscape-scale
+extinction, driven in part by an increase in infectious diseases. The
+emergence of the lethal fungal pathogen (\emph{Batrachochytrium
+dendrobatidis}; ``Bd'') has devastated global amphibian biodiversity,
+with hundreds of species experiencing declines or extinctions. Lacking
+any broadly applicable methods to reverse these impacts, the future of
+many amphibians appears grim. The once-common mountain yellow-legged
+(MYL) frog is emblematic of amphibians threatened by Bd. Although most
+MYL frog populations are extirpated following disease outbreaks, some
+persist and eventually recover. Frogs in these recovering populations
+have increased resistance against Bd infection, consistent with
+evolution of resistant genotypes. We conducted a 15-year landscape-scale
 reintroduction study and show that frogs collected from recovering
 populations and reintroduced to vacant habitats can reestablish
-populations despite the presence of Bd. In addition, results from
-viability modeling suggest that many reintroduced populations have a low
-probability of extinction over 50 years. To better understand the role
-of evolution in frog resistance, we compared the genomes of MYL frogs
-from Bd-naive and recovering populations. We found substantial
-differences between these categories, including changes in immune
-function loci that may confer increased resistance, consistent with
+populations despite the presence of Bd. In addition, the likelihood of
+establishment is influenced by site, population, and frog attributes.
+Results from viability modeling suggest that many reintroduced
+populations have a low probability of extinction over 50 years. To
+better understand the role of evolution in frog resistance, we compared
+the genomes of MYL frogs from Bd-naive and recovering populations. We
+found numerous differences between these categories, including in immune
+function loci associated with disease resistance, consistent with
 evolutionary changes in response to Bd exposure. These results provide a
 rare example of how reintroduction of resistant individuals can allow
-the landscape-scale recovery of disease-impacted species. This example
-has broad implications for the many taxa worldwide that are threatened
-with extinction by novel pathogens.
+the landscape-scale recovery of disease-impacted species, and have broad
+implications for the many taxa worldwide that are threatened with
+extinction by novel pathogens.
 \end{abstract}
@@
 % \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\begin{document}
 
@@
 recovery \citep{brannelly2021}. Host immunity and evolution both play
-important roles in the development of resistance and tolerance, and
-utilizing these factors would seem a promising approach to developing
+important roles in the development of resistance and tolerance, and an
+enhanced understanding of these processes is important for developing
 effective strategies to mitigate disease impacts in the wild
 \citep{garner2016, woodhams2011}. However, several aspects of the
-amphibian-Bd system present difficult obstacles, including (i) the
-general inability of amphibians to mount an effective immune response
-against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
-(ii) the apparent rarity of evolution of more resistant/tolerant
-genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
-suggest that reintroduction of amphibians into sites to reestablish
-populations extirpated by Bd will often result not in population
-recovery, but instead in the rapid reinfection and mortality of the
-introduced animals and/or their progeny
+amphibian-Bd system present difficult obstacles for the development of
+mitigation efforts, including (i) the general inability of amphibians to
+mount an effective immune response against Bd infection
+\citep{rosenblum2012, fites2013, grogan2018a}, and (ii) the apparent
+rarity of evolution of more resistant/tolerant genotypes \citep[but
+see][]{savage2016, grogan2018b}. These factors suggest that
+reintroduction of amphibians into sites to reestablish populations
+extirpated by Bd will often result not in population recovery, but
+instead in the rapid reinfection and mortality of the introduced animals
+and/or their progeny
 \citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
@@
 \citep{knapp2016}, indicative of host resistance against Bd infection
-(and not simply an effect of factors external to individual frogs, e.g.,
+and not simply an effect of factors external to individual frogs (e.g.,
 environmental conditions). In addition to frogs from recovering
@@
 focus on resistance throughout this paper. The observed resistance of
-MYL frogs could be the result of several non-mutually exclusive
-mechanisms, including natural selection for more resistant genotypes
-\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
-and/or inherent between-population differences that pre-date Bd
-exposure. The possible evolution of MYL frog resistance and subsequent
-population recovery is consistent with that expected under
+MYL frogs could be the result of several mechanisms that are not
+mutually exclusive, including natural selection for more resistant
+genotypes \citep{savage2016, grogan2018b}, acquired immunity
+\citep{grogan2018a}, and/or inherent between-population differences that
+pre-date Bd exposure. The possible evolution of MYL frog resistance and
+subsequent population recovery is consistent with that expected under
 ``evolutionary rescue'', whereby rapid evolutionary change increases the
@@
 suggests an opportunity to expand recovery beyond the spatial scale
-possible under natural recovery by utilizing resistant frogs from
-recovering populations in reintroductions to vacant habitats
+possible under natural recovery by using resistant frogs from recovering
+populations in reintroductions to vacant habitats
 (Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.
 
-In the current study, we had three primary objectives. First, to
-determine whether the reintroduction of resistant MYL frogs obtained
-from populations recovering from Bd-caused declines allows the
-successful reestablishment of extirpated populations despite ongoing
-disease, we conducted a 15-year landscape-scale frog reintroduction
-effort (Figure~\ref{fig-recovery-model}). Second, to extend our
-inferences of population recovery well beyond the temporal extent of our
-reintroduction study, we developed a model to estimate the probability
-of persistence for the reintroduced populations over a multi-decadal
-period (Figure~\ref{fig-recovery-model}). Third, given the importance of
-resistance for frog survival, population establishment, and long-term
-viability (this study), we conducted a genomic study using exome capture
-methods to determine whether MYL frogs in recovering populations show
-evidence of selection and whether these genomic changes are associated
-with resistance (Figure~\ref{fig-recovery-model}).
+In the current study, we had three primary objectives. First, determine
+whether the reintroduction of resistant MYL frogs, obtained from
+populations recovering from Bd-caused declines, allows the
+reestablishment of populations despite ongoing disease
+(Figure~\ref{fig-recovery-model}). We accomplished this using a 15-year
+large-scale reintroduction study. Second, extend our inferences of
+population recovery beyond the temporal extent of our reintroduction
+study to estimate the probability of persistence for the reintroduced
+populations over a multi-decadal period
+(Figure~\ref{fig-recovery-model}). We did this using a stage-structured
+matrix model. Third, given the importance of resistance for frog
+survival, population establishment, and long-term viability (this
+study), determine whether MYL frogs in recovering populations show
+genomic patterns consistent with selection for resistance
+(Figure~\ref{fig-recovery-model}). We achieved this with a comparative
+study using exome capture methods applied to naive and recovering
+populations.
 
@@
 ``donor'' populations and translocating them to 1 of 12 nearby recipient
-sites (Figure S1). The donor populations included 2 of
-the 5 recovering populations used in the frog evolution study referenced
-above (Figure~\ref{fig-allelefrequencies}: population 1 and 4), and
-these donor populations contributed frogs for 20 of the 24
-translocations. Following translocation, we estimated adult survival and
-recruitment of new adults from capture-mark-recapture (CMR) surveys and
-obtained counts of tadpoles and juveniles from visual encounter surveys
-(VES). Across all translocation sites, the duration of survey time
-series was 1--16 years (median = 5).
+sites (Figure S1). The 3 donor populations are among
+the largest MYL frog populations in Yosemite, and included 2 of the 5
+recovering populations used in the frog evolution study referenced above
+(see \textbf{Results - Frog evolution in response to Bd:} population 1
+and 4). In addition, this subset of 2 donor populations provided frogs
+for 20 of the 24 translocations. Following translocation, we estimated
+adult survival and recruitment of new adults from capture-mark-recapture
+(CMR) surveys and obtained counts of tadpoles and juveniles from visual
+encounter surveys (VES). Across all translocation sites, the duration of
+survey time series was 1--16 years (median = 5).
 
@@
 
-In summary, results from our frog translocation study indicate that
-translocations resulted in (i) relatively high 1-year survival of
-translocated adults, as well as reproduction and recruitment, at the
-majority of recipient sites, (ii) 1-year survival of adults is
-influenced by site characteristics, weather conditions, and donor
-population (but not Bd load), and (iii) based on the relatively small
-changes in Bd load after translocation, loads appear more strongly
-influenced by frog characteristics (e.g., resistance) than site
-characteristics. Together, these results indicate that frogs
-translocated from recovering populations can maintain the benefits of
-resistance in non-natal habitats. In addition, in 3 locations where
-longer CMR time series allowed us to assess the survival of new adults
-recruited to the population, naturally-recruited adults had equivalent
-or higher survival probabilities than the originally translocated adults
+In summary, results from our frog translocation study indicate that (i)
+translocations produced relatively high 1-year survival of translocated
+adults, as well as reproduction and recruitment, at the majority of
+recipient sites, (ii) 1-year survival of adults is influenced by site
+characteristics, weather conditions, and donor population (but not Bd
+load), and (iii) based on the relatively small changes in Bd load after
+translocation, loads appear more strongly influenced by frog
+characteristics (e.g., resistance) than site characteristics. Together,
+these results indicate that frogs translocated from recovering
+populations can maintain the benefits of resistance in non-natal
+habitats. In addition, in 3 locations where longer CMR time series
+allowed us to assess the survival of new adults recruited to the
+population, naturally-recruited adults had equivalent or higher survival
+probabilities than the originally translocated adults
 (Figure S4). This suggests that frog
@@
 
-\subsection*{Frog evolution in response to Bd}
-
-Results from the preceding sections indicate the critical role of frog
-resistance in post-translocation frog survival, population growth, and
-population viability. As such, identifying the mechanisms underlying
-this resistance would fill a key gap in our understanding of the factors
-that promote population resilience in the presence of disease. Although
-natural selection for more resistant frog genotypes, and evolutionary
-rescue, may be foundational to the ability of frogs to recover despite
-ongoing Bd infection, for MYL frogs the role of disease-mediated
-selection in these processes remains unknown.
+\hypertarget{frog-genome-evolution}{%
+\paragraph{Frog genome evolution}\label{frog-genome-evolution}}
+
+Results from the preceding sections indicate the critical role of
+intrinsic factors (such as frog resistance) in post-translocation frog
+survival, population growth, and population viability. As such,
+identifying the mechanisms underlying resistance would fill a key gap in
+our understanding of the factors that promote population resilience in
+the presence of disease. Although natural selection for more resistant
+frog genotypes, and evolutionary rescue, may be foundational to the
+ability of frogs to recover despite ongoing Bd infection, for MYL frogs
+the role of disease-mediated selection in these processes remains
+unknown.
 
@@
 as the presence of introduced (predatory) fish and fish-induced habitat
-fragmentation. (See \textbf{Supporting Information - Frog evolution in
-response to Bd - Study design} for additional details regarding the
-study design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to
-describe the relationships between sampled populations, and then used
-two complementary approaches to identify regions of the genome that
-differed between naive and recovering populations (i.e., regions under
+fragmentation.
+
+It is important to note that our genomic analysis is necessarily limited
+by the small number of naive populations remaining on the landscape
+(nearly all of which we sampled as part of the current study) and their
+clumped distribution Figure~\ref{fig-allelefrequencies}. These
+characteristics of naive populations precluded a replicated analyses
+among pairs of geographically proximate naive and recovering
+populations. Instead, to obtain adequate power, we conducted a pooled
+analysis across population types to identify genomic regions showing
+patterns consistent with selection. (See \textbf{Supporting Information
+- Frog evolution in response to Bd - Study design} for additional
+details regarding the study design.)
+
+As a first step in our pooled analysis, we conducted a principal
+component analysis (PCA) of the genomic data to describe the
+relationships between sampled populations. We then used two
+complementary approaches to identify regions of the genome that differed
+between naive and recovering populations (i.e., regions under
 selection). First, we used a multivariate linear mixed model to evaluate
@@
 Results from the individual variant and splined window analyses show
-that recovering populations have signatures of selection on multiple
-regions of the genome. The analysis of individual variants identified 11
-``outlier'' SNPs (i.e., showing significantly different allele
-frequencies between naive versus recovering populations) from 7 distinct
-genes across 4 contigs (Figure S6B, C). One
-of the 7 identified genes (LOC108802036) does not have an associated
-annotation. For the outlier SNPs, frequency differences between the
-naive and recovering populations ranged from 0.41 to 0.86. Most of these
-SNPs showed frequency differences in only a subset of the sampled
-populations (Figure~\ref{fig-allelefrequencies}A, B), but the SNP in
-the RIN3 gene showed consistent differences in frequencies across all
-populations (Figure~\ref{fig-allelefrequencies}C). This is suggestive
-of parallel selection at this locus across multiple populations. The
-other 6 outlier variants showed less consistent frequency differences
-across the study populations, but for these we still found a
-statistically significant signal of selection in 2 of the 3 genetic
-clusters (containing populations 5--9;
+that recovering populations show patterns consistent with selection on
+multiple regions of the genome. The analysis of individual variants
+identified 11 ``outlier'' SNPs (i.e., showing significantly different
+allele frequencies between naive versus recovering populations) from 7
+distinct genes across 4 contigs (Figure S6B, C
+SI). One of the 7 identified genes (LOC108802036) does not have an
+associated annotation. For the outlier SNPs, frequency differences
+between the naive and recovering populations ranged from 0.41 to 0.86.
+Most of these SNPs showed frequency differences in only a subset of the
+sampled populations (Figure~\ref{fig-allelefrequencies}A, B), but the
+SNP in the RIN3 gene showed consistent differences in frequencies across
+all populations (Figure~\ref{fig-allelefrequencies}C). This is
+suggestive of parallel selection at this locus across multiple
+populations. The other 6 outlier variants showed less consistent
+frequency differences across the study populations, but for these we
+still found a statistically significant signal of selection in 2 of the
+3 genetic clusters (containing populations 5--9;
 Figure~\ref{fig-allelefrequencies}, Figure S6A). Therefore, although some outlier variant associations have a more
@@
 
-In summary, our genomic results indicate that the exomes of frogs from
-naive and recovering populations show substantial differences,
-consistent with parallel evolutionary changes following Bd exposure. The
-regions under selection contain several immunologically-relevant genes
-and gene families that are directly linked to disease resistance in
-other taxa.
+In summary, our genomic results indicate that frogs from naive and
+recovering populations show differences at several immune-function loci,
+consistent with selection following Bd exposure.
 
@@
 
-Results from our genomic analyses suggest that natural selection for
-adaptive alleles is at least partially responsible for the increased
-resistance of frogs in recovering populations. We identified multiple
-specific alleles and genomic regions showing signatures of selection
-between adjacent naive and recovering MYL frog populations, consistent
-with selection following Bd exposure. These analysis are based on
-samples collected from virtually all of the MYL frog populations
-remaining in a naive state, as well as adjacent recovering populations.
-This study design produced genetic clusters that each contained at least
-one naive and one recovering population, allowing us to detect selection
-without the confounding effects of population structure. In addition, we
-did not find a reduction in overall genetic variation in the recovering
-populations, suggesting that despite localized selection in the genome,
-these populations retain adequate genetic diversity for long-term
-persistence.
-
-Importantly, some genomic regions that we identified as under selection
-are associated with cellular and immunological mechanisms known to
-contribute to disease resistance, including in amphibians
+Our genomics study used a pooled design that compared MYL frogs from
+naive and recovering populations distributed across a similar
+latitudinal range. The study design produced frog genetic clusters that
+each contained at least one naive and one recovering population,
+allowing us to infer selection while attempting to minimize the
+confounding effects of population structure. We identified multiple
+specific alleles and genomic regions showing differences between naive
+and recovering MYL frog populations, consistent with selection following
+Bd exposure. In addition, we did not find a reduction in overall genetic
+variation in recovering populations, suggesting that despite localized
+selection in the genome, some of these populations likely retain
+adequate genetic diversity for long-term persistence.
+
+Importantly, some genomic regions that we identified as possibly under
+selection are associated with cellular and immunological mechanisms
+known to contribute to disease resistance, including in amphibians
 \citep{zamudio2020}. For example, the MHC plays an important role in
@@
 linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. We also identified a region,
-characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
-that contained the complement genes C6 and C7. The complement system
-plays an important role in innate immunity \citep{rieraromo2016}, and
-our results could indicate that balancing selection is acting in this
-region of the genome to favor a diverse set of alleles, as is known for
-C6 in humans \citep{soejima2005}. Based on the analysis of individual
-outlier variants, the RIN3 gene showed a consistent pattern of allele
-frequency differences across all nine of the frog populations sampled in
-this study, indicating consistent selection in populations distributed
-across a wide geographic area. This gene is associated with immune
-response and in \emph{Xenopus} is expressed during appendage
-regeneration \citep{fukazawa2009}. Finally, the outlier variant with the
-lowest p-value was the uncharacterized gene LOC108802036. In the genome
-of another frog species, this gene is located adjacent to a type I
-interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
-further suggests the importance of interferon-related genes in this
-system. Collectively, the genes associated with these genomic
-differences may confer at least some degree of resistance against Bd
-infection, an attribute that may be critically important to population
-reestablishment and recovery in the presence of Bd.
+phenotype \citep[i.e., gill damage and amoebic load as heritable host
+resistance traits,][]{robledo2020, robledo2018}. We also identified a
+region, characterized by high \emph{F\textsubscript{ST}} and low
+\(\pi_{diff}\), that contained the complement genes C6 and C7. The
+complement system plays an important role in innate immunity
+\citep{rieraromo2016}, and our results could indicate that balancing
+selection is acting in this region of the genome to favor a diverse set
+of alleles, as is known for C6 in humans \citep{soejima2005}. Based on
+the analysis of individual outlier variants, the RIN3 gene showed a
+consistent pattern of allele frequency differences across all nine of
+the frog populations sampled in this study, indicating consistent
+selection in populations distributed across a wide geographic area. This
+gene is associated with immune response and in \emph{Xenopus} is
+expressed during appendage regeneration \citep{fukazawa2009}. Finally,
+the outlier variant with the lowest p-value was the uncharacterized gene
+LOC108802036. In the genome of another frog species, this gene is
+located adjacent to a type I interferon gene (Np-IFNi2) \citep{gan2018},
+and together with GVINP1 further suggests the importance of
+interferon-related genes in this system. Collectively, the genes
+associated with these genomic differences may confer at least some
+degree of resistance against Bd infection, an attribute that may be
+critically important to population reestablishment and recovery in the
+presence of Bd.
 
@@
 conducting 24 translocations over 16 years.
+
+The results of the meta-analysis indicate the important influence of
+several variables on adult survival, and by extension on population
+viability. Although we assume that the effects of winter severity and
+frog sex are direct effects of the variables themselves, the elevation
+and donor population variables are likely serving as proxies for a range
+of factors. For example, elevation is positively correlated with several
+attributes that may influence frog survival, including the presence of
+boulder habitat that is important for frog overwintering, abundance of
+frog predators including garter snakes, and frog developmental rates.
+The latter can influence the duration of frog life stages that have
+differential susceptibility to chytridiomycosis (see below). The
+duration of the tadpole and juvenile stages increases with elevation,
+but the fact that long-run population growth rate is relatively
+insensitive to variation in these durations suggests that developmental
+rates of early frog life stages have little effect on the predicted
+long-term viability of reintroduced populations. Nonetheless, a clearer
+understanding of the mechanistic components of the elevation and donor
+population variables will require additional research.
 
@@
 deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to four key model parameters to
-quantify how much changes in these parameters affected the long-run
-growth rate (Figure S5). This also helped us
-determine where in the model environmental variation in juvenile
-survival and recruitment would have the largest effects on population
-dynamics. Third, we included demographic stochasticity and environmental
-stochasticity in \(\sigma_{J_1}\) in our model and simulated the 50-year
-viability (i.e., 1 - extinction probability) of populations given an
-introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to
-confirm that our model could reasonably reproduce the observed recovery
-trajectories of MYL frogs following reintroductions. Additional details
-are provided in \textbf{Supporting Information - Population viability
-modeling - Model analysis and simulation}. Code to replicate the
-analyses can be found at \url{https://github.com/SNARL1/translocation}.
-
-\hypertarget{frog-evolution-in-response-to-bd-1}{%
-\section*{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-1}}
+we compute the elasticity and sensitivity of \(\lambda\) to nine model
+parameters to quantify how much changes in these parameters affected the
+long-run growth rate (Figure S5). This also
+helped us determine where in the model environmental variation in
+juvenile survival and recruitment would have the largest effects on
+population dynamics. Third, we included demographic stochasticity and
+environmental stochasticity in \(\sigma_{J_1}\) in our model and
+simulated the 50-year viability (i.e., 1 - extinction probability) of
+populations given an introduction of 40 adult individuals into an
+unoccupied habitat. Finally, we fit our model to our longest
+translocation trajectory to confirm that our model could reasonably
+reproduce the observed recovery trajectories of MYL frogs following
+reintroductions. Additional details are provided in \textbf{Supporting
+Information - Population viability modeling - Model analysis and
+simulation}. Code to replicate the analyses can be found at
+\url{https://github.com/SNARL1/translocation}.
+
+\hypertarget{frog-genome-evolution-1}{%
+\section*{Frog genome evolution}\label{frog-genome-evolution-1}}
 
@@
 and processes, respectively. Blue text indicates components that are
-included in the current study. The timeline shows the general sequence
-of the components, with the dotted portion indicating a projection into
-the future.}
+included in the current study. We quantified frog survival and
+population viability following reintroductions directly using field
+studies and models, and evaluated ``disease-mediated selection''
+indirectly using a comparison of frog exomes in naive and recovering
+populations. The timeline shows the general sequence of the components,
+with the dashed portion indicating a projection into the future.}
 
@@
 
-\caption{\label{fig-allelefrequencies}Evidence for selection on
-individual variants in recovering MYL frog populations at the landscape
-scale. For each of the 9 naive and recovering MYL frog populations
-(indicated by numbered points), adjacent pie charts show allele
-frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
-LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with
-Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
-shown in dark gray, and the range boundary for MYL frogs shown in light
-gray. The inset map locates the study area and range boundary in
-California. Sites 1 and 4 (site id = 72996 and 70567, respectively) were
-also used as sources of frogs in the frog population recovery study.}
+\caption{\label{fig-allelefrequencies}Genomic patterns consistent with
+selection on individual variants in recovering MYL frog populations at
+the landscape scale (based on a linear mixed model analysis). For each
+of the 9 naive and recovering MYL frog populations (indicated by
+numbered points), adjacent pie charts show allele frequencies for the 11
+outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS,
+MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map
+of the Sierra Nevada study area, with Yosemite, Kings Canyon, and
+Sequoia National Parks (from north to south) shown in dark gray, and the
+range boundary for MYL frogs shown in light gray. The inset map locates
+the study area and range boundary in California. Sites 1 and 4 (site id
+= 72996 and 70567, respectively) were also used as sources of frogs in
+the frog population recovery study.}
 
@@
 
-\caption{\label{fig-spline-manhattan}Evidence for selection on genomic
-regions in recovering MYL frog populations. Manhattan plot of the
-results from the splined window analysis showing outlier regions for the
-difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
-\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
-upper red dashed line and below the lower red dashed line. In (B),
-outlier regions are shown above the single dashed red line. Outlier
-regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
-shown in blue and outlier regions for both \(\pi_{diff}\) and
-\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
-(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
-upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
-line). (D) Magnified Contig8 from from (B) showing the
-\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
-TCF19 and VARS. This region of the genome contains 8 annotated genes
-known to occur in the extended MHC Class I and III regions.}
+\caption{\label{fig-spline-manhattan}Patterns consistent with selection
+on genomic regions in recovering MYL frog populations based on a splined
+window analysis. Note that several key candidate genes (RIN3, TCF19,
+VARS) are outliers in both the individual variant analysis
+Figure~\ref{fig-allelefrequencies} and splined window analysis (this
+figure). Manhattan plots show outlier regions for the difference in (A)
+nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
+In (A), outlier regions are shown above the upper red dashed line and
+below the lower red dashed line. In (B), outlier regions are shown above
+the single dashed red line. Outlier regions for either \(\pi_{diff}\) or
+\emph{F\textsubscript{ST}} are shown in blue and outlier regions for
+both \(\pi_{diff}\) and \emph{F\textsubscript{ST}} are shown in red. (C)
+Magnified Contig19 from (A) showing two adjacent outlier regions for
+\(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier SNP (indicated with a
+dashed vertical blue line). (D) Magnified Contig8 from from (B) showing
+the \emph{F\textsubscript{ST}} outlier region that includes the outlier
+SNPs TCF19 and VARS. This region of the genome contains 8 annotated
+genes known to occur in the extended MHC Class I and III regions.}
 

### 2024-04-18 doc/manuscript/translocation_bioarxiv_SI.tex
Update PNAS template
Add new figure, text edits, and latex files

@@
-\documentclass[9pt,twoside,lineno]{pnas-new}
+\documentclass[9pt,twoside,lineno]{pnas-new-SI}
 % Use the lineno option to display guide line numbers if required.
@@
 population CMR models based on \citep{joseph2018}. For each individual
-\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
-the individual was not detected, and \(o_{i, j}=2\) if the individual
-was detected. Capture histories of \(M\) individuals are modeled,
-although only \(N_s\) individuals were captured. This parameter expanded
-data augmentation allows us to capture the possibility that undetected
-individuals may have recruited into the adult population
-\citep{royle2012}. Here, \(M\) was chosen to be three times the number
-of observed individuals (\(3N_s\)) to be considerably greater than our
-prior guess of \(N_s\).
+in a superpopulation of real and psuedo-individuals, \(i=1, ..., M\) on
+each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
+not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
+histories of \(M\) individuals are modeled, although only \(N_s\)
+individuals were captured. This parameter expanded data augmentation
+allows us to capture the possibility that undetected individuals may
+have recruited into the adult population \citep{royle2012}. Here, \(M\)
+was chosen to be three times the number of observed individuals
+(\(3N_s\)) to be considerably greater than our prior guess of \(N_s\).
 
 We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
-period \(t = 1,..., n_t\). The four states that we consider are:
+period \(t = 1,..., n_t\). The states that we consider are:
 \(u_{i, t} = 1\) for individuals that have not recruited,
@@
 periods, and we denote the primary period in which survey \(j\) takes
-place as \(t_j\). Each primary period \(t\) occurs within one year, but
-within a year there can be multiple primary periods. We set the year
-containing the first primary period to \(y_{t = 1} = 1\), and generally
-\(y_t\) represents the year containing primary period \(t\). Years
-increment by one until the final year of the mark recapture efforts,
-which we denote \(n_y\): \(y \in \{1, 2, ..., n_y\}\). We assume that
-within a primary period, the state of each individual does not change
-(i.e., individuals do not recruit into the adult population, gain or
-lose Bd infection, or die). This assumption is justified by the short
-time intervals between surveys within primary periods, in cases where
-primary periods contain multiple secondary periods.
+place as \(t_j\). Multiple primary periods can occur during a summer
+active season (typically July-August, but in some years as early as May
+and as late as September). We set the year containing the first primary
+period to \(y_{t = 1} = 1\), and generally \(y_t\) represents the year
+containing primary period \(t\). Years increment by one until the final
+year of the mark recapture efforts, which we denote \(n_y\):
+\(y \in \{1, 2, ..., n_y\}\). We assume that within a primary period,
+the state of each individual does not change (i.e., individuals do not
+recruit into the adult population, gain or lose Bd infection, or die).
+This assumption is justified by the short time intervals between surveys
+within primary periods, in cases where primary periods contain multiple
+secondary periods.
 
@@
 period, how often was the value still 2 (alive) in the next primary
-period compared to 3 (dead)?) . This yielded posterior distributions for
+period compared to 3 (dead)) . This yielded posterior distributions for
 survival probabilities between primary periods. However, because the
@@
 
-Second, we performed a local elasticity analysis on \(\lambda\) with
-respect to parameters \(\sigma_{J_1}\), \(\sigma_{J_2}\),
-\(\sigma_{A_R}\), and \(F\) to determine how small changes in these
-parameters could influence the long-run deterministic growth rate of
-populations (Figure~\ref{fig-viability-supp}).
+Second, we performed a local elasticity and sensitivity analysis on
+\(\lambda\) with respect to parameters \(\sigma_{L_1}\),
+\(\sigma_{L_2}\), \(\sigma_{L_3}\), \(\sigma_{J_1}\), \(\sigma_{J_2}\),
+\(\sigma_{A_R}\), \(F\), \(p_{L_2}\), \(p_{J_1}\) to determine how small
+changes in these parameters could influence the long-run deterministic
+growth rate of populations (Figure~\ref{fig-viability-supp}). Note
+that \(\lambda\) is equally or more sensitive to changes in adult
+survival (\(\sigma_{A_R}\)) and survival of year-1 juveniles
+(\(\sigma_{J_1}\)) than it is to other model parameters, providing
+additional justification for focusing on variation in \(\sigma_{A_R}\)
+and \(\sigma_{J_1}\) in our viability analyses.
 
@@
 
-\hypertarget{frog-evolution-in-response-to-bd-2}{%
-\section{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-2}}
+\hypertarget{frog-genome-evolution-2}{%
+\section{Frog genome evolution}\label{frog-genome-evolution-2}}
 
@@
 interquartile range, and values outside the 1.5x interquartile range.
-Loads indicative of severe disease are \textgreater{} 5.8 ITS copies (on
-a log\textsubscript{10} scale). Samples sizes are provided immediately
-above the x-axis.}
+Samples sizes are provided immediately above the x-axis. The red dashed
+horizontal line indicates the approximate load above which severe
+disease is typically observed \citep{vredenburg2010}. This threshold
+(5.8 ITS1 copies - on a log\textsubscript{10} scale) is commonly
+exceeded during Bd epizootics, but is exceeded only rarely in recovering
+populations \citep{briggs2010}.}
 
@@
 \caption{\label{fig-viability-supp}Sensitivity analysis of the
-stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
-in four parameters: \(\sigma_{A_R}\) (yearly survival probability of
-naturally recruited adults), \emph{F (}number of eggs produced by a
-female frog in a year that successfully hatch), \(\sigma_{J_1}\) (yearly
-probability of survival of year-1 juveniles that also affects
-recruitment), and \(\sigma_{J_2}\) (yearly probability of survival and
-recruitment of year-2 juveniles). Elasticity is calculated at the
-default parameter values for each population and \(\sigma_{J_1}\) =
-0.09.}
+stage-structured MYL frog model. A. Elasticity of \(\lambda\) with
+changes in nine parameters: \(\sigma_{A_R}\) (yearly survival
+probability of naturally recruited adults), \(F\) (number of eggs
+produced by a female frog in a year that successfully hatch),
+\(\sigma_{J_1}\) (yearly probability of survival of year-1 juveniles
+that also affects recruitment), \(\sigma_{J_2}\) (yearly probability of
+survival and recruitment of year-2 juveniles), \(p_{L_2}\) (probability
+that year-2 tadpoles metamorph into juveniles), \(p_{J_1}\) (probability
+that year-1 juveniles mature to adults), \(\sigma_{L_1}\) (survival
+probability of year-1 tadpoles), \(\sigma_{L_2}\) (survival probability
+of year-2 tadpoles), and \(\sigma_{L_3}\) (survival probability of
+year-3 tadpoles). B. Sensitivity of \(\lambda\) with changes in the same
+nine parameters. Note that \(F\) has different units than the other
+parameters, which is why its sensitivity is low. Sensitivities of all
+other parameters are comparable. We did not include \(p_{L_1}\) as this
+parameter is generally 1 across the populations in this study (Table S1)
+-- tadpoles spend at least a year as tadpoles before metamorphosing.
+Elasticity and sensitivity is calculated at the default parameter values
+for each population and \(\sigma_{J_1}\) = 0.09.}
 
@@
 \caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
-windows identified by the overlapping splined windows for FST and
-\(\pi\), boxplots showing average \(\pi\) for naive and recovering
-populations. The corresponding gene annotations for each window are as
-follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3,
-EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5)
-GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR,
-MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD,
-ATRAID, GPN1.}
+windows identified by the overlapping splined windows for
+\emph{F\textsubscript{ST}} and \(\pi\), boxplots showing average \(\pi\)
+for naive and recovering populations. The corresponding gene annotations
+for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
+SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
+TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
+(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
+BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
 
@@
 
-\hypertarget{tables}{%
-\subsubsection{Tables}\label{tables}}
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/ave_pi_per_site_yose.png}
+
+}
+
+\caption{For each study population located in Yosemite National Park
+(populations 1-4 shown in Site Map), average \(\pi\) for the 9 outlier
+windows (panels A-I) identified in the splined window analysis.
+Populations within Yosemite are the focus because they are the most
+relevant to the translocated populations described in the paper: donor
+populations 1 and 4 were the source of frogs for 20 of the 24
+translocations and the donor population for the remaining four
+translocations is located adjacent to populations 3 and 4. Bars are
+color-coded by population category (naive, recovering), and gene
+annotations within each window are listed above each panel. Gene regions
+described in the paper are in panel A (Complement C6,C7), panel E
+(GVINP1), and panel F (extended MHC region that includes two SNPs
+identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by
+an *). In the Site Map, the four populations included in panels A-I are
+color-coded by population category, and dashed lines delineate the
+genetic clusters as defined in Figure~\ref{fig-selectionresults}.}
+
+\end{figure}\clearpage
+
+\newpage
 

### 2024-04-18 doc/manuscript/PNAS-template-empty.tex
Update PNAS template
Add new figure, text edits, and latex files
Add author contributions and declarations to template

@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{}
-\authordeclaration{}
+\authorcontributions{All authors designed research; R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., A.P.R., and R.L.G.
+performed research; M.B.J.
+contributed new analytic tools; R.A.K., M.Q.W., A.Q.B., A.P.R., and E.B.R.
+analyzed data; and R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., and E.B.R.
+wrote paper.}
+\authordeclaration{This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
+Briggs; DEB-1557190, to C.
+Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
+Richards-Zawacki).
+The authors declare no additional competing interests.}
 \equalauthors{\textsuperscript{1}M.W. contributed equally to this work with A.B.}
@@
 % At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{}
+\keywords{amphibians | disease | evolution | reintroduction | population recovery}
 

### 2024-04-18 doc/manuscript/translocation.qmd
Add author contributions and declarations to template

@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:-------------------------|:------------------|:--------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2024-04-18 doc/manuscript/translocation_bioarxiv.tex
Render to latex using revised template

@@
 % At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{evolution | resistance | amphibians | population recovery | reintroduction | disease | \emph{Batrachochytrium dendrobatidis}}
+\keywords{amphibians | disease | evolution | reintroduction | population recovery}
 
@@
 large-scale reintroduction study. Second, extend our inferences of
-population recovery beyond the temporal extent of our reintroduction
+population recovery beyond the temporal bounds of our reintroduction
 study to estimate the probability of persistence for the reintroduced
@@
 (nearly all of which we sampled as part of the current study) and their
-clumped distribution Figure~\ref{fig-allelefrequencies}. These
+clumped distribution (Figure~\ref{fig-allelefrequencies}). These
 characteristics of naive populations precluded a replicated analyses
@@
 analysis across population types to identify genomic regions showing
-patterns consistent with selection. (See \textbf{Supporting Information
-- Frog evolution in response to Bd - Study design} for additional
-details regarding the study design.)
+patterns consistent with selection. Our comparative approach cannot
+provide a direct assessment of the strength or timing of natural
+selection, but does allow us to identify specific regions of the genome
+that merit further study for their potential contribution to frog
+resistance in recovering populations. (See \textbf{Supporting
+Information - Frog evolution in response to Bd - Study design} for
+additional details regarding the study design.)
 
@@
 complementary approaches to identify regions of the genome that differed
-between naive and recovering populations (i.e., regions under
+between naive and recovering populations (i.e., regions possibly under
 selection). First, we used a multivariate linear mixed model to evaluate
@@
 frequency differences across the study populations, but for these we
-still found a statistically significant signal of selection in 2 of the
-3 genetic clusters (containing populations 5--9;
+still found statistically significant differences in 2 of the 3 genetic
+clusters (containing populations 5--9;
 Figure~\ref{fig-allelefrequencies}, Figure S6A). Therefore, although some outlier variant associations have a more
 limited geographic extent than RIN3, they still describe results that
-suggest parallel evolutionary changes following Bd exposure.
+are consistent with parallel evolutionary changes following Bd exposure.
 
@@
 signal of selection and/or immune-related functions. The largest
-\(\pi_{diff}\), indicative of directional selection, occurred in a 163kb
+\(\pi_{diff}\), suggestive of directional selection, occurred in a 163kb
 region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP
@@
 extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
-Therefore, this region under selection is part of the extended MHC Class
-I and III complex and shows synteny with other amphibian genomes.
+Therefore, this region is part of the extended MHC Class I and III
+complex and shows synteny with other amphibian genomes.
 
@@
 Frog evolution in response to Bd - Genetic diversity} for details).
-Thus, despite localized selection in particular regions of the genome,
-we did not find evidence for reduced genetic diversity across the genome
-in recovering populations. In addition, no gene ontology (GO) biological
-functions, molecular functions, or cellular processes were
-over-represented in either the outlier variants or the 35 genes located
-in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined
-windows (see \textbf{Supporting Information - Frog evolution in response
-to Bd} \textbf{- GO analysis} for details).
+Thus, despite patterns consistent with localized selection in particular
+regions of the genome, we did not find evidence for reduced genetic
+diversity across the genome in recovering populations. In addition, no
+gene ontology (GO) biological functions, molecular functions, or
+cellular processes were over-represented in either the outlier variants
+or the 35 genes located in the overlapping \emph{F\textsubscript{ST}}
+and \(\pi_{diff}\) splined windows (see \textbf{Supporting Information -
+Frog evolution in response to Bd} \textbf{- GO analysis} for details).
 
@@
 (i) the capacity of reintroduced populations to become established and
-eventually recover despite ongoing disease, (ii) that the recovering
-populations are likely to persist over a 50-year period, (iii) that
-there are substantial genomic differences between naive and recovering
+eventually recover despite ongoing disease, (ii) that 50\% of the
+recovering populations are likely to persist over a 50-year period,
+(iii) that there are genomic differences between naive and recovering
 MYL frog populations, consistent with evolutionary change in frogs
-following Bd exposure, and (iv) that some of the genomic regions under
-selection contain genes related to disease resistance. Collectively,
-these results (Figure~\ref{fig-recovery-model}) provide a rare example
-of amphibian recovery in the presence of Bd, and have important
-implications for the conservation and recovery of amphibians and other
-taxa worldwide that are endangered by escalating impacts from emerging
-infectious diseases. In light of the generally low success rate of
-amphibian reintroduction efforts \citep{dodd2005}, our success in
+following Bd exposure, and (iv) that some of the genomic differences are
+in regions that contain genes related to disease resistance.
+Collectively, these results (Figure~\ref{fig-recovery-model}) provide a
+rare example of amphibian recovery in the presence of Bd, and have
+important implications for the conservation and recovery of amphibians
+and other taxa worldwide that are endangered by escalating impacts from
+emerging infectious diseases. In light of the generally low success rate
+of amphibian reintroduction efforts \citep{dodd2005}, our success in
 reestablishing MYL frog populations via translocation of resistant
@@
 indicate that frogs from recovering populations exhibit resistance
-against Bd infection. This resistance could in theory be due to several
-factors, including natural selection for more resistant genotypes,
-acquired immunity, and/or inherent between-population differences that
-pre-date Bd exposure, but until now evidence to evaluate the role of
-evolution was lacking.
+against Bd infection. This resistance could be due at least in part to
+natural selection for more resistant genotypes, but until now evidence
+to evaluate the possible role of evolution was lacking.
 
@@
 Bd exposure. In addition, we did not find a reduction in overall genetic
-variation in recovering populations, suggesting that despite localized
-selection in the genome, some of these populations likely retain
-adequate genetic diversity for long-term persistence.
+variation in recovering populations, suggesting that despite possible
+localized selection in the genome, some of these populations likely
+retain adequate genetic diversity for long-term persistence.
 
@@
 resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
-Similarly, the region with the strongest indication of directional
+Similarly, the region with the strongest pattern of possible directional
 selection (as measured by \(\pi_{diff}\)) contains the
@@
 consistent pattern of allele frequency differences across all nine of
-the frog populations sampled in this study, indicating consistent
-selection in populations distributed across a wide geographic area. This
-gene is associated with immune response and in \emph{Xenopus} is
-expressed during appendage regeneration \citep{fukazawa2009}. Finally,
-the outlier variant with the lowest p-value was the uncharacterized gene
-LOC108802036. In the genome of another frog species, this gene is
-located adjacent to a type I interferon gene (Np-IFNi2) \citep{gan2018},
-and together with GVINP1 further suggests the importance of
-interferon-related genes in this system. Collectively, the genes
-associated with these genomic differences may confer at least some
-degree of resistance against Bd infection, an attribute that may be
+the frog populations sampled in this study, suggesting possible
+consistent selection in populations distributed across a wide geographic
+area. This gene is associated with immune response and in \emph{Xenopus}
+is expressed during appendage regeneration \citep{fukazawa2009}.
+Finally, the outlier variant with the lowest p-value was the
+uncharacterized gene LOC108802036. In the genome of another frog
+species, this gene is located adjacent to a type I interferon gene
+(Np-IFNi2) \citep{gan2018}, and together with GVINP1 further suggests
+the importance of interferon-related genes in this system. Collectively,
+the genes associated with these genomic differences may confer at least
+some degree of resistance against Bd infection, an attribute that may be
 critically important to population reestablishment and recovery in the
@@
 and donor population variables are likely serving as proxies for a range
-of factors. For example, elevation is positively correlated with several
-attributes that may influence frog survival, including the presence of
-boulder habitat that is important for frog overwintering, abundance of
-frog predators including garter snakes, and frog developmental rates.
-The latter can influence the duration of frog life stages that have
+of factors. For example, elevation is correlated with several attributes
+that may influence frog survival, including the presence of boulder
+habitat (positive correlation) that is important for frog overwintering,
+abundance of frog predators including garter snakes (negative
+correlation), and frog developmental rates (negative correlation). The
+latter can influence the duration of frog life stages that have
 differential susceptibility to chytridiomycosis (see below). The
@@
 decimated populations provides an airtight example of evolutionary
-rescue will likely always be uncertain (given that can never have a
+rescue will likely always be uncertain (given that we can never have a
 perfect understanding of the past). Regardless, we provide an example
-from the wild that suggests that evolution can produce individuals that
-harbor adaptive alleles and allow population recovery in a novel (i.e.,
-Bd-positive) environment, and show conclusively that individuals from
-these recovering populations can be used to reestablish extirpated
+from the wild that suggests that evolution might produce individuals
+that harbor adaptive alleles and allow population recovery in a novel
+(i.e., Bd-positive) environment, and show conclusively that individuals
+from these recovering populations can be used to reestablish extirpated
 populations and expand the scale of natural recovery
@@
 \citep{jombart2008}. To detect regions of the genome that differed
-between naive and recovering populations, i.e., regions under selection,
-we used two approaches: (1) a multivariate linear mixed model to
-evaluate individual variants (SNPs and INDELs), and (2) a splined window
-analysis to evaluate larger genomic regions. For the variant analysis,
-we first used a stringent data filter to include only variants with
-\textless{} 5\% missing data (missing for no more than 2 individuals),
-and then calculated the likelihood ratio statistic for the resulting set
-of 148,307 high quality variants across 127 contigs using GEMMA
-\citep{zhou2014}. GEMMA calculates and incorporates a relatedness matrix
-for input samples, allowing us to account for relatedness and population
-structure when calculating likelihood ratio statistics. We identified
-variants showing different allele frequencies between naive versus
-recovering populations (``outliers'') using a Bonferroni-corrected
-significance level of 0.01. We visualized the results using a Manhattan
-plot and qqplot. We developed a more liberal set of outlier variants
-using a Bonferroni-corrected significance level of 0.05 and used this
-set solely for the gene ontology (GO) analysis (see below and
-\textbf{Supporting Information - Frog evolution in response to Bd - GO
-analysis;}  Dataset S1, S2).
+between naive and recovering populations, we used two approaches: (1) a
+multivariate linear mixed model to evaluate individual variants (SNPs
+and INDELs), and (2) a splined window analysis to evaluate larger
+genomic regions. For the variant analysis, we first used a stringent
+data filter to include only variants with \textless{} 5\% missing data
+(missing for no more than 2 individuals), and then calculated the
+likelihood ratio statistic for the resulting set of 148,307 high quality
+variants across 127 contigs using GEMMA \citep{zhou2014}. GEMMA
+calculates and incorporates a relatedness matrix for input samples,
+allowing us to account for relatedness and population structure when
+calculating likelihood ratio statistics. We identified variants showing
+different allele frequencies between naive versus recovering populations
+(``outliers'') using a Bonferroni-corrected significance level of 0.01.
+We visualized the results using a Manhattan plot and qqplot. We
+developed a more liberal set of outlier variants using a
+Bonferroni-corrected significance level of 0.05 and used this set solely
+for the gene ontology (GO) analysis (see below and \textbf{Supporting
+Information - Frog evolution in response to Bd - GO analysis;}Dataset
+S1, S2).
 
@@
 \acknow{We thank the following for important contributions to this study: E.
-Hegeman and A. Lindauer--project and data management, field and
-laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
-and numerous summer technicians--field work; staff at Sequoia-Kings
+Hegeman and A. Lindauer (project and data management, field and
+laboratory work); A. Barbella and K. Rose-(laboratory work); C. Kamoroff
+and numerous summer technicians (field work); staff at Sequoia-Kings
 Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
@@
 Service, University of California-Santa Barbara Institutional Animal
-Care and Use Committee, and Sierra Nevada Aquatic Research
-Laboratory--research permits, logistical support, and/or field
-assistance. This project was supported by grants from the National Park
-Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
-Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
+Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory
+(research permits, logistical support, and/or field assistance). This
+project was supported by grants from the National Park Service (to
+R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science
+Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
 DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}

### 2024-04-18 doc/manuscript/translocation_pnas_SI.tex
Update PNAS version and add biorxiv version
Render to latex using revised template

@@
 
-%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
+%% Adds the main heading for the text. Comment out this line if you do not have any supporting information text.
 \SItext
-\hypertarget{frog-evolution-in-response-to-bd-2}{%
-\section{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-2}}
-
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-We compared frog genomes sampled in naive versus recovering populations.
-Bd-exposure histories of these populations are based on a decade or more
-of data from visual encounter surveys and Bd surveillance using skin
-swabbing \citep[e.g.,][]{knapp2016}. Comparing populations with
-different infection histories allowed larger sample sizes and
-replication across the landscape. The alternative approach of comparing
-samples from the same populations before and after Bd exposure isn't
-feasible in this system because Bd arrived in most MYL frog populations
-decades ago and population persistence/recovery is rare and
-unpredictable. As a result, samples from recovering populations
-collected from before and after Bd exposure are not available and are
-unlikely to be available in the future.
-
-\hypertarget{results-1}{%
-\subsection{Results}\label{results-1}}
-
-In the Results section of the main paper, we describe the stringent set
-of outlier SNPS (identified using a Bonferroni-corrected p-value of
-0.01). The liberal set, identified using a Bonferroni-corrected p-value
-of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
-genes across 16 contigs. Our analysis found no significantly
-over-represented GO terms in this set of 30 genes, or in the set of 38
-outlier regions identified with the splined window analysis.
-
-In the comparison of individual outlier variants between naive and
-recovering populations, we detected consistent allele frequency changes
-across all genetic clusters for 1 of 7 of the associated genes (RIN3;
-Fig. 1). However, for the other 6 genes we
-still found a statistically significant signal of selection in 2 of the
-3 genetic clusters (containing populations 6--9;
-Fig. 1, Figure~\ref{fig-selectionresults}
-A). Therefore, although some outlier variant associations have a more
-limited geographic extent than RIN3, they still describe parallel
-evolutionary changes following Bd exposure and are therefore interesting
-candidates for future studies.
-
 \hypertarget{frog-population-recovery-2}{%
@@
 
-\hypertarget{methods-1}{%
-\subsection{Methods}\label{methods-1}}
+\hypertarget{laboratory-methods}{%
+\subsection{Laboratory methods}\label{laboratory-methods}}
 
-Swab extracts were analyzed singly instead of in triplicate
-\citep{kriger2006}. For analysis of swabs collected during 2005--2014,
-we used standards developed from known concentrations of zoospores
-\citep{boyle2004}, and after 2014, we used standards based on single
-ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
+Swab extracts were analyzed using standard Bd DNA extraction and qPCR
+methods \citep{boyle2004}, and extracts were analyzed singly instead of
+in triplicate \citep{kriger2006}. For analysis of swabs collected during
+2005--2014, we used standards developed from known concentrations of
+zoospores \citep{boyle2004}, and after 2014, we used standards based on
+single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
 between samples analyzed using both types of standards, Bd in the study
@@
 
+\hypertarget{cmr-model-structure}{%
+\subsection{CMR model structure}\label{cmr-model-structure}}
+
+We estimated survival and recruitment for each site using open
+population CMR models based on \citep{joseph2018}. For each individual
+in a superpopulation of real and psuedo-individuals, \(i=1, ..., M\) on
+each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
+not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
+histories of \(M\) individuals are modeled, although only \(N_s\)
+individuals were captured. This parameter expanded data augmentation
+allows us to capture the possibility that undetected individuals may
+have recruited into the adult population \citep{royle2012}. Here, \(M\)
+was chosen to be three times the number of observed individuals
+(\(3N_s\)) to be considerably greater than our prior guess of \(N_s\).
+
+We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
+period \(t = 1,..., n_t\). The states that we consider are:
+\(u_{i, t} = 1\) for individuals that have not recruited,
+\(u_{i, t} =2\) for live adults, and \(u_{i, t} = 3\) for dead adults.
+Each survey \(j=1, ..., n_j\) occurs in one of the \(n_t\) primary
+periods, and we denote the primary period in which survey \(j\) takes
+place as \(t_j\). Multiple primary periods can occur during a summer
+active season (typically July-August, but in some years as early as May
+and as late as September). We set the year containing the first primary
+period to \(y_{t = 1} = 1\), and generally \(y_t\) represents the year
+containing primary period \(t\). Years increment by one until the final
+year of the mark recapture efforts, which we denote \(n_y\):
+\(y \in \{1, 2, ..., n_y\}\). We assume that within a primary period,
+the state of each individual does not change (i.e., individuals do not
+recruit into the adult population, gain or lose Bd infection, or die).
+This assumption is justified by the short time intervals between surveys
+within primary periods, in cases where primary periods contain multiple
+secondary periods.
+
+Live individuals are detected with probability \(p_j\), which is modeled
+as:
+
+\[p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),\]
+
+where \(X_j^{(p)}\) is a known row vector and \(\beta^{(p)}\) an unknown
+parameter vector. Not recruited and dead individuals are never captured.
+We bundle these assumptions about the observation probabilities for
+survey \(j\) into an emission matrix \(\Omega_j\):
+
+\[
+\Omega_j =
+\begin{blockarray}{ccc}
+  \text{Not detected} & \text{Detected} \\
+\begin{block}{(cc)c}
+  1 & 0 & \text{Not recruited} \\
+  1 - p_j & p_j & \text{Alive} \\
+  1 & 0 & \text{Dead} \\
+\end{block}
+\end{blockarray}
+\]
+
+The state transition matrix \(\Psi_{t, i}\) contains the probabilities
+of individual \(i\) transitioning from state \(u_{i, t}\) (rows) to
+\(u_{i, t+1}\) (columns) between primary period \(t\) and \(t+1\). For
+non-introduced (i.e., naturally recruited) individuals, this matrix is
+given by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 0 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(\lambda_t\) is the probability of recruiting in time \(t\)
+and \(\phi_t\) is the probability of survival in time \(t\).
+
+For introduced individuals, which have deterministic recruitment (i.e.,
+they recruit when introduced), the state transition matrix is given by:
+
+\[
+\Psi_{t, i} =
+\begin{blockarray}{cccc}
+  \text{Not recruited} & \text{Alive} & \text{Dead} \\
+\begin{block}{(ccc)c}
+  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
+  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
+  0 & 0 & 1 & \text{Dead} \\
+\end{block}
+\end{blockarray},
+\] where \(I_{t, i}\) is a known indicator function for whether
+individual \(i\) was introduced in primary period \(t\).
+
+We allow recruitment probabilities to vary in time via random effects,
+such that:
+
+\[\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),\]
+
+where \(\alpha^{(\lambda)}\) is an intercept parameter and
+\(\epsilon^{(\lambda)}_t\) is an adjustment for time \(t\).
+
+Survival probabilities also vary in time, and as a function of known
+covariates:
+
+\[\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),\]
+
+where \(X^{(\phi)}_t\) is a row vector of known covariates,
+\(\beta^{(\phi)}\) is a column vector of unknown coefficients, and
+\(\epsilon^{(\phi)}_t\) is an adjustment for time \(t\).
+
+To complete the specification of the Bayesian model, we specify priors
+for all unknown parameters. The recruitment parameter priors were
+specified as follows:
+
+\[\alpha^{(\lambda)} \sim N(0,1),\]
+\[\sigma^{(\lambda)} \sim N_+(0,1),\]
+\[\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),\] for periods
+\(t=1, ..., T\). Here \(N\) represents the normal distribution and
+\(N_+\) the half normal distribution with positive support.
+
+Survival parameter priors were specified similarly as:
+
+\[\beta^{(\phi)} \sim N(0, 1),\] \[\sigma^{(\phi)} \sim N_+(0, 1),\]
+\[\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),\] for time
+\(t=1, ..., T\).
+
+The detection model coefficient vector also received a standard normal
+prior \(\beta^{(p)} \sim N(0, 1)\).
+
+We computed the likelihood of each individual capture history using the
+forward algorithm, and we estimated the latent states using the
+forward-backward algorithm \citep{zucchini2009, joseph2018}.
+
+All of the code to specify and fit the model in Stan is available in the
+open source mrmr package \citep{joseph2019}.
+
+The joint distribution of the resulting model can be written as follows:
+
+\begin{multline*}
+[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{multline*}
+
+where \(\pmb Y\) is an \(M \times T\) detection matrix, and \(Y_i\) the
+capture history of individual \(i\).
+
+\hypertarget{among-site-survival-modeling}{%
+\subsection{Among-site survival
+modeling}\label{among-site-survival-modeling}}
+
+The objective of this analysis is to describe the influence of site,
+cohort, and individual level characteristics on post-translocation frog
+survival. By modeling survival estimates obtained from site-specific
+mrmr CMR analyses, we are in effect conducting an among-site
+meta-analysis. Although it would theoretically be possible to estimate
+survival covariate effects in a joint CMR model that integrates capture
+histories across all sites, this was impractical due to computational
+requirements of the CMR models (namely, run time and memory).
+
+We used Bayesian generalized linear mixed models to investigate
+predictors of survival among sites. The response \(y_i\) is binary,
+representing a point estimate of whether individual \(i\) survived in
+the year following translocation. We generated these point estimates by
+rounding the posterior median of 1-year post-introduction survival for
+each individual (from site-specific mrmr CMR models) and modeled the
+data using a Bernoulli distribution:
+
+\[y_i \sim \text{Bernoulli}(p_i),\]
+
+where \(p_i\) is the probability of survival.
+
+We modeled variation in probabilities as follows:
+
+\[\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},\]
+
+where \(\alpha\) is an intercept, \(X_i\) is a length \(K\) row vector
+of predictors, \(\pmb \beta\) is a column vector of predictor effects,
+and \(\pmb \nu\) a vector of group level random effects. Here \(g[i]\)
+refers to the group \(g\) containing individual \(i\), and we estimate
+an adjustment for each of the \(G\) groups (\(\nu_1, ..., \nu_G\)).
+
+These models were fit using the \texttt{stan\_glmer()} function in the
+rstanarm package, with default priors described below
+\citep{rstanarm2022}. These priors are vague, but include data-dependent
+scaling as follows to account for different input variable scales.
+However, because we standardized all predictor variables similarly to
+have equal variance (by centering and dividing by twice the sample
+standard deviation), the resulting priors are identical. Specifically,
+we have:
+
+\[\alpha \sim \text{Normal}(0, 2.5),\] and
+
+\[\beta_k \sim \text{Normal}(0, 5),\]
+
+for \(k=1, ..., K\) where \(K\) is the number of predictor variables.
+
+The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
+rstanarm is a zero-mean Gaussian, where the covariance matrix is
+constructed from a correlation matrix with an LKJ prior, and a vector of
+variance parameters -- the decomposition of variance prior with unit
+regularization, concentration, shape, and scale parameters
+\citep{lewandowski2009}.
+
+We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
+with four parallel chains, each run for 10,000 iterations, discarding
+the first half of each chain as warm-up draws \citep{rstanarm2022}. We
+used Rhat statistics and trace plots to verify convergence. We
+considered models with different subsets of fixed and random effects,
+and used approximate leave-one-out cross validation to identify the best
+model \citep{vehtari2016}.
+
 \hypertarget{population-viability-modeling-1}{%
@@
 
-\hypertarget{methods-2}{%
-\subsection{Methods}\label{methods-2}}
-
 \hypertarget{incorporating-yearly-variability-in-vital-rates}{%
-\subparagraph{Incorporating yearly variability in vital
+\subsection{Incorporating yearly variability in vital
 rates}\label{incorporating-yearly-variability-in-vital-rates}}
 
-We extracted yearly survival probabilities for translocated adults
+We computed yearly survival probabilities for translocated adults
 \(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
-the CMR model used to estimate population-level parameters of
-translocated populations (see ``Estimation of adult survival and
-population size'' above). Although we observed yearly variability in
-adult survival within a population, the magnitude of this variability
-was small compared to among population variability
-(Fig. 3). Thus, we did not include
+the posterior distribution of individual state trajectories derived from
+mrmr CMR models. Although we observed yearly variability in adult
+survival within a population, the magnitude of this variability was
+small compared to among-population variability
+(Fig. 2). Thus, we did not include
 yearly within-population variability in adult survival in this analysis.
@@
 from Poisson variability around a mean value. Therefore, we allowed for
-yearly variability in the probability of successful recruitment
-\(\omega\) (additional details provided in ``Estimating model
-parameters'' below).
+yearly variability in the probability of year-1 juvenile survival and
+recruitment (additional details provided in \textbf{Estimating model
+parameters} below). We also could have included environmental
+stochasticity in year-2 juvenile survival and recruitment
+\(\sigma_{J_2}\), but our elasticity and sensitivity analysis (see
+below, \textbf{Model analysis and simulation}) showed that this
+parameter had little effect on host growth rate relative to
+\(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp}).
 
 \hypertarget{estimating-model-parameters}{%
-\subparagraph{Estimating model
+\subsection{Estimating model
 parameters}\label{estimating-model-parameters}}
@@
 are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
-and \(\sigma_{A_R}\) were extracted directly from our CMR model
-described previously in this study. For populations where we had a
-sufficient number of PIT-tagged, naturally-recruited adults, we observed
-that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different,
-with \(\sigma_{A_R} > \sigma_{A_T}\)
+and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
+\textbf{Materials and Methods - Frog populaton recovery - Estimation of
+frog survival and abundance} and \textbf{Supporting Information - Frog
+population recovery - CMR model structure} for details). For populations
+where we had a sufficient number of PIT-tagged, naturally-recruited
+adults, we observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be
+notably different, with \(\sigma_{A_R} > \sigma_{A_T}\)
 (Figure~\ref{fig-compare_surv_probs}). For populations lacking
@@
 
+To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
+predicted true states for naturally-recruited individuals (1=not
+recruited, 2=alive, 3=dead, as described in \textbf{Supporting
+Information - Frog population recovery - CMR model structure}), then
+calculated the posterior probability of individuals surviving between
+consecutive primary periods, conditional on being alive in the first
+primary period (e.g., given a value of 2 (alive) in the first primary
+period, how often was the value still 2 (alive) in the next primary
+period compared to 3 (dead)) . This yielded posterior distributions for
+survival probabilities between primary periods. However, because the
+time interval between primary periods differed, the survival
+probabilities between different consecutive primary periods were not
+directly comparable. To address this, we converted the survival
+probabilities between each consecutive primary period to per day death
+rates, propagating the uncertainty from the posterior distributions. We
+then took a weighted average of these death rates, weighted by the time
+interval between primary periods, to get the average per day death rate
+over the entire CMR survey. We converted this per day death rate \(d\)
+back to a yearly survival probability using
+\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
+\(\sigma_{A_T}\) such that our estimates of average yearly survival
+probability were comparable between \(\sigma_{A_R}\) and
+\(\sigma_{A_T}\).
+
 \hypertarget{model-analysis-and-simulation}{%
-\subparagraph{Model analysis and
+\subsection{Model analysis and
 simulation}\label{model-analysis-and-simulation}}
@@
 deterministic version of our model with no yearly heterogeneity in
-probability of successful recruitment \(\omega\), and calculated the
-predicted long-run growth rate \(\lambda\) of a population for different
-values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
-\(\omega = 0.3\) and calculated the predicted growth rate of our 12
-populations.
+year-1 juvenile survival and recruitment probability \(\sigma_{J_1}\),
+and calculated the predicted long-run growth rate \(\lambda\) of a
+population for different values of \(\sigma_{A_R}\) and
+\(\sigma_{J_1}\). We then fixed \(\sigma_{J_1} = 0.09\) and calculated
+the predicted growth rate of our 12 populations.
 
-Second, we performed a local elasticity analysis on \(\lambda\) with
-respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
-determine how small changes in these parameters could influence the
-long-run deterministic growth rate of populations
-(Figure~\ref{fig-viability-supp}).
+Second, we performed a local elasticity and sensitivity analysis on
+\(\lambda\) with respect to parameters \(\sigma_{L_1}\),
+\(\sigma_{L_2}\), \(\sigma_{L_3}\), \(\sigma_{J_1}\), \(\sigma_{J_2}\),
+\(\sigma_{A_R}\), \(F\), \(p_{L_2}\), \(p_{J_1}\) to determine how small
+changes in these parameters could influence the long-run deterministic
+growth rate of populations (Figure~\ref{fig-viability-supp}). Note
+that \(\lambda\) is equally or more sensitive to changes in adult
+survival (\(\sigma_{A_R}\)) and survival of year-1 juveniles
+(\(\sigma_{J_1}\)) than it is to other model parameters, providing
+additional justification for focusing on variation in \(\sigma_{A_R}\)
+and \(\sigma_{J_1}\) in our viability analyses.
 
@@
 environmental stochasticity, where environmental stochasticity was
-represented by among-year variability in \(\omega\). We used this model
-to simulate a one-time introduction of 40 translocated adult frogs. We
-ran this simulation 1000 times for each population and computed the
-probability of a population becoming extinct after 50 years given the
-observed parameter values and environmental stochasticity in \(\omega\).
-We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
-and drew values of \(\omega\) each year from a beta distribution with a
-dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
-\(\phi = 2\) the beta distribution is uniform between 0 and 1).
+represented by among-year variability in \(\sigma_{J_1}\). We used this
+model to simulate a one-time introduction of 40 translocated adult
+frogs. We ran this simulation 1000 times for each population and
+computed the probability of a population becoming extinct after 50 years
+given the observed parameter values and environmental stochasticity in
+\(\sigma_{J_1}\). We varied the mean recruitment probability
+\(\sigma_{J_1}\) from 0 and 0.25 and drew values of \(\sigma_{J_1}\)
+each year from a beta distribution with a dispersion parameter of
+\(\phi = 2\) (when \(\sigma_{J_1} = 0.5\) and \(\phi = 2\) the beta
+distribution is uniform between 0 and 1). Using different values of
+\(\phi\) does not qualitatively change the existence of distinct
+extinction dynamics between populations with \(\sigma_{A_R} < 0.5\) and
+those with \(\sigma_{A_R} > 0.5\). However, increasing yearly
+variability in \(\sigma_{J_1}\) increases extinction risk for all
+populations. For example, if we set \(\phi = 0.001\), such that in a
+given year essentially either all year-1 juveniles survive or all of
+them die, populations with \(\sigma_{A_R} > 0.5\) need to have
+\(\sigma_{J_1}\) greater than 0.2 to have a 50-year extinction
+probability of less 50\%. Because we do not pit tag juveniles, we do not
+have CMR estimates for \(\sigma_{J_1}\) or \(\phi\). However, based on
+qualitative and semi-quantitative field observations over 25 years, a
+value of \(\sigma_{J_1} = 0.25\) in the presence of Bd is probably a
+reasonable estimate for many populations. Thus, we expect our model
+predictions to be conservative with regards to population recovery.
 
@@
 years, repeating the simulation 50,000 times. For each run and each
-year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
-equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
-Approximate Bayesian Computing and rejection sampling
+year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
+1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
+Using Approximate Bayesian Computing and rejection sampling
 \citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
 1000 trajectories) that minimized the sum of squared errors between the
-observed and predicted data. The \(\omega\) values associated with these
-``best'' trajectories represented an approximate posterior distribution
-\citep{beaumont2010}. Using these best fit trajectories, we assessed
-whether our model could qualitatively describe the patterns of recovery
-in the observed data for population 70550. Code to replicate the
-analyses can be found at \url{https://github.com/SNARL1/translocation}.
+observed and predicted data. The yearly \(\sigma_{J_1}\) values
+associated with these ``best'' trajectories represented an approximate
+posterior distribution \citep{beaumont2010}. Using these best fit
+trajectories, we assessed whether our model could qualitatively describe
+the patterns of recovery in the observed data for population 70550.
+
+\hypertarget{frog-genome-evolution-2}{%
+\section{Frog genome evolution}\label{frog-genome-evolution-2}}
+
+\hypertarget{study-design}{%
+\subsection{Study design}\label{study-design}}
+
+To gain insights into the role of evolution in the development of
+resistance by MYL frogs, we compared frog exomes sampled in naive versus
+recovering populations. Comparing populations with different infection
+histories allowed larger sample sizes and replication across the
+landscape. The alternative approach of comparing samples from the same
+populations before and after Bd exposure isn't feasible in this system
+because Bd arrived in most MYL frog populations decades ago and
+population persistence/recovery is rare and unpredictable. As a result,
+samples from recovering populations collected before and after Bd
+exposure are not available and are unlikely to be available in the
+future.
+
+Our study design, in which we compared frog genomes in naive and
+recovering populations, required sampling populations across the range
+of MYL frogs in the southern Sierra Nevada. This is due to the fact that
+very few MYL frog populations remain in the naive state, and those that
+do are scattered across a wide latitudinal range, from Yosemite National
+Park in the north to southern Kings Canyon National Park in south. To
+minimize potential confounding effects caused by known variation in frog
+genotypes across latitude \citep{byrne2023}, we selected sampling sites
+such that both population types were represented across similar
+latitudinal ranges (Fig. 5).
+
+\hypertarget{go-analysis}{%
+\subsection{GO analysis}\label{go-analysis}}
+
+In \textbf{Results-Frog evolution in response to Bd}, we describe the
+stringent set of outlier variants (identified using a
+Bonferroni-corrected p-value of 0.01). A liberal set of outlier
+variants, identified using a Bonferroni-corrected p-value of 0.05,
+included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
+across 16 contigs. We used this liberal set to determine if any GO
+biological functions, molecular functions, or cellular processes were
+overrepresented. To do this, we retrieved the BLAST hits and mapped GO
+terms for each gene in our targeted transcriptome. We then conducted a
+statistical overrepresentation test (Fisher's exact test) using Blast2GO
+\citep{conesa2005} to compare the 30 unique outlier genes to the
+complete set of genes in our target transcriptome. We repeated this
+process for the set of 35 genes located in the 9 shared regions of the
+\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+
+\hypertarget{genetic-diversity}{%
+\subsection{Genetic diversity}\label{genetic-diversity}}
+
+To characterize general patterns of genetic diversity between naive and
+recovering populations, we conducted three analyses. First, we
+calculated heterozygosity for each sampled frog using VCFtools
+\citep{danecek2011}. Second, to characterize genome-wide patterns of
+diversity, we used VCFtools to calculate nucleotide diversity (\(\pi\))
+in 100kb sliding windows along the genome for each population. Third, we
+calculated average \(\pi\) per population within each of the 9 outlier
+windows identified in the splined window analysis. We compared average
+\(\pi\) per population between naive and recovering population types,
+and also for each naive and recovering population in Yosemite. The
+latter comparison focused on Yosemite because these populations are the
+most relevant to the translocated populations described in the paper
+(see results below for additional details).
+
+Average individual-level heterozygosity and genome-wide population-level
+\(\pi\) were similar between the naive and recovering groups
+(Figure~\ref{fig-violinplot-heterozy};
+Figure~\ref{fig-boxplot-genomewide-pi-by-pop}). Within each of the 9
+outlier windows, average \(\pi\) shows considerable variation between
+populations and no obvious patterns between the naive and recovering
+populations (Figure~\ref{fig-boxplot-pi-by-windowpop}). For the four
+populations in Yosemite, within each of the 9 outlier windows, average
+\(\pi\) per population was generally higher in the 3 recovering
+populations than in the 1 naive population
+(Figure~\ref{fig-ave-pi-per-site-yose}). ADD CONCLUDING SENTENCE OR
+TWO THAT SUMMARIZE THEGNIFICANCE OF THESE RESULTS?
 
 \newpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\textwidth]{figures/pca_qq_manhattan.png}
-
-}
-
-\caption{\label{fig-selectionresults}Evidence for selection on
-individual variants in recovering MYL frog populations. (A) PCA
-calculated from binary SNPs showing the genomic relationship of samples.
-Numeric labels and colors match those in
-Fig. 1 A. Populations 1-7 are \emph{R.
-sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
-showing observed and expected p-values for 148,307 SNPs and INDELS.
-Dashed line shows p-value that identifies outliers. (C) Manhattan plot
-showing p-value for each SNP. SNPs are sorted by genomic position and
-contigs are sorted by size. Red line shows p-value that identifies
-outliers. Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in
-that gene.}
-
-\end{figure}\clearpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\textwidth]{figures/synteny_figure.png}
-
-}
-
-\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
-order in \emph{Xenopus tropicalis} and \emph{Rana mucosa} for the
-outlier region containing MHC Class I Classical and MHC Class III gene
-regions. The plot was created with SimpleSynteny \citep{veltri2016}
-using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
-accession GCA\_000004195.4) and \emph{Rana mucosa} Contig19. Asterices
-indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
-sizes for each contig representation are labeled.}
-
-\end{figure}\clearpage
 
@@
 and donor MYL frog populations in Yosemite National Park (park boundary
-indicated by black polygon). Symbol shapes indicate the donor population
-used for each translocation site. To obscure the exact locations of
+indicated by gray polygon). Symbol shapes indicate the donor population
+used for each translocation site, and 5-digit numbers identify each
+donor and translocation site. To obscure the exact locations of
 populations, random noise was added to all point coordinates. Inset map
@@
 
+\newpage
+
 \begin{figure}
 
-{\centering \includegraphics[width=0.60\textwidth]{figures/mcmc_areas_m2b.png}
+{\centering \includegraphics[width=0.7\textwidth]{figures/bdload_beforeafter.png}
 
@@
 
-\caption{\label{fig-transsurvival-postdens}For the best model describing
-the average 1-year survival of frogs in the first translocation to a
-site as a predictor of survival of frogs in all subsequent
-translocations to that site, estimated posterior density curves and
-shaded 95\% uncertainty intervals for the intercept and single predictor
-variable. In the Bayesian framework in which the model was developed, a
-variable is considered an important predictor if the associated
-uncertainty interval does not overlap zero (indicated by the dashed red
-line).}
+\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
+the 12 recipient sites, Bd loads for the period immediately prior to
+translocation versus during the 1-year period after translocation. Bd
+loads are expressed as the number of ITS1 copies per skin swab, as
+estimated by qPCR of the Bd ITS1 region. Box plots show medians, first
+and third quartiles, largest and smallest values within 1.5x
+interquartile range, and values outside the 1.5x interquartile range.
+Samples sizes are provided immediately above the x-axis. The red dashed
+horizontal line indicates the approximate load above which severe
+disease is typically observed \citep{vredenburg2010}. This threshold
+(5.8 ITS1 copies - on a log\textsubscript{10} scale) is commonly
+exceeded during Bd epizootics, but is exceeded only rarely in recovering
+populations \citep{briggs2010}.}
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.60\textwidth]{figures/bdload_beforeafter.png}
-
-}
-
-\caption{\label{fig-bdload-beforeafter}Evidence that frogs translocated
-from recovering populations can maintain the benefits of
-resistance/tolerance in non-natal habitats. For frogs translocated to
-each of the 12 recipient sites, Bd loads for the period immediately
-prior to translocation versus during the 1-year period after
-translocation. Bd loads are expressed as the number of ITS1 copies per
-skin swab, as estimated by qPCR of the Bd ITS1 region. Loads indicative
-of severe disease are \textgreater{} 5.8 ITS copies (on a
-log\textsubscript{10} scale; see SI Appendix text for details). Box plot
-details are as in Fig. 4.}
-
-\end{figure}\clearpage
+\newpage
 
@@
 
-\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
-important predictor of post-translocation frog survival. For the best
-model identifying predictors of 1-year frog survival following
-translocation, estimated posterior density curves and shaded 95\%
-uncertainty intervals for the intercept and all predictor variables. In
+\caption{\label{fig-survival-postdens}Results from the among-site
+meta-analysis, showing that Bd load is not an important predictor of
+post-translocation frog survival. Depicted distributions are the
+estimated posterior density curves and shaded 95\% uncertainty intervals
+for the intercept and all predictor variables from the best model. In
 the Bayesian framework in which the model was developed, variables are
 considered important predictors if the associated uncertainty interval
-does not overlap zero (indicated by the dashed red line).}
+does not overlap zero (indicated by the dashed red line). Predictor
+variables shown on the y-axis are defined as follows: length = frog
+size, snow\_t = winter severity in the year of translocation (measured
+on April 1), snow\_t1 = winter severity in the year following
+translocation (measured on April 1), day = day of year on which a
+translocation was conducted, bd\_load = Bd load, elevation = site
+elevation, order\_first = within-site translocation order, sex\_male =
+frog sex, donor\_70567 and donor\_72996 = donor population.}
 
 \end{figure}\clearpage
+
+\newpage
 
@@
 
-\caption{\label{fig-compare_surv_probs}Evidence that frog
-resistance/tolerance in non-natal habitats is maintained across
-generations. Comparison of the average yearly adult survival
-probabilities for adults translocated to each of 3 sites versus adults
-that were naturally recruited at each site (as a result of reproduction
-by translocated frogs). In contrast to
-Fig. 3, these are not survival
+\caption{\label{fig-compare_surv_probs}Comparison of average yearly
+adult survival probabilities for adults translocated to each of 3 sites
+versus adults that were naturally recruited at each site (as a result of
+reproduction by translocated frogs). In contrast to
+Fig. 2, these are not survival
 probabilities from the first year following translocation, but instead
@@
 uncertainty intervals around the estimates, accounting for yearly
-variation in survival probabilities. The large error bars for 70550
-reflect the large within and among individual variability in yearly
-survival probability at this site (as also reflected in one-year
-survival probabilities in Fig. 3). All
-estimates are derived from the mrmr package.}
+variation in survival probabilities. All estimates were derived using
+the mrmr package, and the methods for calculation are described in
+\textbf{Supporting Information - Population viability modeling -
+Estimating model parameters}.}
 
 \end{figure}\clearpage
+
+\newpage
 
@@
 \caption{\label{fig-viability-supp}Sensitivity analysis of the
-stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
-in 3 parameters: \(\sigma_{A_R}\) (yearly survival probability of
-naturally recruited adults), \(\omega\) (yearly probability of
-successful recruitment from juveniles to adults), and \emph{F (}number
-of eggs produced by a female frog in a year that successfully hatch).
-Elasticity is calculated at the default parameter values for each
-population and \(\omega\) = 0.3.}
+stage-structured MYL frog model. A. Elasticity of \(\lambda\) with
+changes in nine parameters: \(\sigma_{A_R}\) (yearly survival
+probability of naturally recruited adults), \(F\) (number of eggs
+produced by a female frog in a year that successfully hatch),
+\(\sigma_{J_1}\) (yearly probability of survival of year-1 juveniles
+that also affects recruitment), \(\sigma_{J_2}\) (yearly probability of
+survival and recruitment of year-2 juveniles), \(p_{L_2}\) (probability
+that year-2 tadpoles metamorph into juveniles), \(p_{J_1}\) (probability
+that year-1 juveniles mature to adults), \(\sigma_{L_1}\) (survival
+probability of year-1 tadpoles), \(\sigma_{L_2}\) (survival probability
+of year-2 tadpoles), and \(\sigma_{L_3}\) (survival probability of
+year-3 tadpoles). B. Sensitivity of \(\lambda\) with changes in the same
+nine parameters. Note that \(F\) has different units than the other
+parameters, which is why its sensitivity is low. Sensitivities of all
+other parameters are comparable. We did not include \(p_{L_1}\) as this
+parameter is generally 1 across the populations in this study (Table S1)
+-- tadpoles spend at least a year as tadpoles before metamorphosing.
+Elasticity and sensitivity is calculated at the default parameter values
+for each population and \(\sigma_{J_1}\) = 0.09.}
 
@@
 
-\hypertarget{tables}{%
-\subsubsection{Tables}\label{tables}}
+\begin{figure}
 
-\hfill\break
+{\centering \includegraphics[width=0.85\textwidth]{figures/pca_qq_manhattan.png}
 
-\hypertarget{tbl-survival-earlylate}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0654}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2430}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2056}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2804}}@{}}
-\caption{\label{tbl-survival-earlylate}Association between the
-proportion of populations translocated early versus late in the study
-period (\textless{} 2013 or $\geq$ 2013, respectively) and
-probability of survival (\textless{} 0.5 or $\geq$ 0.5). Based on
-the viability analysis, survival probabilities \textless{} 0.5 and
-$\geq$ 0.5 produced 50-year extinction probabilities of 1 and
-\textless{} 0.5, respectively.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Survival probability
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Number of translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Total translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Proportion of translocations
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Survival probability
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Number of translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Total translocations
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Proportion of translocations
-\end{minipage} \\
-\midrule()
-\endhead
-early & \textless{} 0.5 & 4 & 5 & 0.800 \\
-early & $\geq$ 0.5 & 1 & 5 & 0.200 \\
-late & \textless{} 0.5 & 2 & 7 & 0.286 \\
-late & $\geq$ 0.5 & 5 & 7 & 0.714 \\
-\bottomrule()
-\end{longtable}
+}
+
+\caption{\label{fig-selectionresults}Results from the analysis of
+individual variants, showing putative signatures of selection in
+recovering MYL frog populations. (A) PCA calculated from binary SNPs
+showing the genomic relationship of samples. Numeric labels and colors
+match those in Fig. 5. Populations 1-7 are
+\emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
+qqplot showing observed and expected p-values for 148,307 SNPs and
+INDELS. Dashed line shows p-value that identifies outliers. (C)
+Manhattan plot showing p-value for each SNP. SNPs are sorted by genomic
+position and contigs are sorted by size. Red line shows p-value that
+identifies outliers. Outlier SNPs above this threshold are highlighted
+and labeled. Bold labels indicate the presence of at least one
+non-synonymous SNP in that gene.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/synteny_figure.png}
+
+}
+
+\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
+order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
+outlier region containing MHC Class I Classical and MHC Class III gene
+regions. The plot was created with SimpleSynteny \citep{veltri2016}
+using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
+accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
+indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
+sizes for each contig representation are labeled.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/violin_plot_heterozy_by_group.png}
+
+}
+
+\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
+heterozygosity for the Bd-naive and recovering populations included in
+the frog evolution study. Individual data points are represented by
+their corresponding site number (from
+Fig. 5).}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_genomewide_pi_by_pop.png}
+
+}
+
+\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each study
+population in the frog evolution study (from
+Fig. 5), boxplots showing nucleotide
+diversity (\(\pi\)) within 100kb sliding windows along the genome. The
+y-axis has been log\textsubscript{10}-transformed for display purposes.
+Sites are color-coded as ``Bd-naive'' or ``recovering'', and vertical
+dashed lines show the genetic breaks between frog populations as
+described in Figure~\ref{fig-selectionresults} A).}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_windowpop.png}
+
+}
+
+\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
+windows identified by the overlapping splined windows for
+\emph{F\textsubscript{ST}} and \(\pi\), boxplots showing average \(\pi\)
+for naive and recovering populations. The corresponding gene annotations
+for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
+SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
+TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
+(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
+BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth]{figures/ave_pi_per_site_yose.png}
+
+}
+
+\caption{\label{fig-ave-pi-per-site-yose}For each study population
+located in Yosemite National Park (populations 1-4 shown in Site Map),
+average \(\pi\) for the 9 outlier windows (panels A-I) identified in the
+splined window analysis. Populations 1 and 4 were the source of frogs
+for 20 of the 24 translocations and the donor population for the
+remaining four translocations is located adjacent to populations 3 and
+4. Bars are color-coded by population type (naive, recovering), and gene
+annotations within each window are listed above each panel. Gene regions
+described in the paper are in panel A (Complement C6,C7), panel E
+(GVINP1), and panel F (extended MHC region that includes two SNPs
+identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by
+an asterisk). In the Site Map, the four populations included in panels
+A-I are color-coded by population type, and dashed lines delineate the
+genetic clusters as defined in Figure~\ref{fig-selectionresults}. }
+
+\end{figure}\clearpage
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2639}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1222}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4611}}@{}}
 \caption{\label{tbl-param_values}Description and values of parameters
@@
 \endhead
-\(\sigma_{L_1}\), Yearly survival probability of year 1 tadpoles & 0.7 &
+\(\sigma_{L_1}\), Yearly survival probability of year-1 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{L_2}\), Yearly survival probability of year 2 tadpoles & 0.7 &
+\(\sigma_{L_2}\), Yearly survival probability of year-2 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{L_3}\), Yearly survival probability of year 3 tadpoles & 0.7 &
+\(\sigma_{L_3}\), Yearly survival probability of year-3 tadpoles & 0.7 &
 Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles & 0.25
-& Estimated from field data, observations, natural history knowledge \\
-\(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
+\(\sigma_{J_1}\), Yearly survival probability of year-1 juveniles &
+Varies yearly & Varies. Bounds estimated from field data, observations,
+natural history knowledge \\
+\(\sigma_{J_2}\), Yearly survival probability of year-2 juveniles & 0.5
 & Estimated from field data, observations, natural history knowledge \\
 \(\sigma_{A_R}\), Yearly survival probability of naturally recruited
-adults & Varies by population & Estimated from mark-recapture studies \\
+adults & Varies by population & Estimated from CMR studies \\
 \(\sigma_{A_T}\), Yearly survival probability of translocated adults &
-Varies by population & Estimated from mark-recapture studies \\
-\(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
+Varies by population & Estimated from CMR studies \\
+\(p_{L_1}\), Probability of a year-1 tadpoles remaining as a tadpoles &
 1 & Estimated from field data, observations, natural history
 knowledge \\
-\(p_{L_2}\), Probability of a year 2 tadpoles remaining as a tadpoles &
+\(p_{L_2}\), Probability of a year-2 tadpoles remaining as a tadpoles &
 0.25 & Estimated from field data, observations, natural history
 knowledge \\
-\(p_{J_1}\), Probability of a year 1 juvenile remaining as a juvenile &
+\(p_{J_1}\), Probability of a year-1 juvenile remaining as a juvenile &
 0.25 & Estimated from field data, observations, natural history

### 2024-04-21 doc/manuscript/submission_files/research_report_2023-22711R.tex
Create final files for submission

--- 2024-04-18 doc/manuscript/translocation_pnas.tex
@@
 \emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
-emblematic of the global declines of amphibians caused by Bd
+emblematic of the global decline of amphibians caused by Bd
 \citep{scheele2019}. Once the most common amphibian in the high
@@
 
+\begin{figure*}
+{\centering \includegraphics[width=7in]{conceptual_model_7in.pdf}}
+\caption{\label{fig-recovery-model}For MYL frogs, a conceptual model
+depicting the Bd-caused decline and subsequent natural recovery (black
+text), facilitated recovery via reintroductions, and the linkages
+between these two pathways. Rectangles and hexagons represent outcomes
+and processes, respectively. Blue text indicates components that are
+included in the current study. We quantified frog survival and
+population viability following reintroductions directly using field
+studies and models, and evaluated ``disease-mediated selection''
+indirectly using a comparison of frog exomes in naive and recovering
+populations. The timeline shows the general sequence of the components,
+with the dashed portion indicating a projection into the future.}
+\end{figure*}
+
 \section*{Results}
@@
 recovering populations used in the frog evolution study referenced above
-(see \textbf{Results - Frog evolution in response to Bd:} population 1
-and 4). In addition, this subset of 2 donor populations provided frogs
-for 20 of the 24 translocations. Following translocation, we estimated
-adult survival and recruitment of new adults from capture-mark-recapture
-(CMR) surveys and obtained counts of tadpoles and juveniles from visual
-encounter surveys (VES). Across all translocation sites, the duration of
-survey time series was 1--16 years (median = 5).
+(see \textbf{Results - Frog genome evolution:} populations 1 and 4).
+These 2 donor populations provided frogs for 20 of the 24
+translocations. Following translocation, we estimated adult survival and
+recruitment of new adults from capture-mark-recapture (CMR) surveys and
+obtained counts of tadpoles and juveniles from visual encounter surveys
+(VES). Across all translocation sites, the duration of survey time
+series was 1--16 years (median = 5).
 
@@
 \textbf{Results - Long-term population viability}).
+
+\begin{figure}
+{\centering \includegraphics[width=3.4in]{translocation_survival_bysiteid_3pt42in.pdf}}
+\caption{\label{fig-translocation-survival}Median 1-year survival for
+each cohort of translocated frogs at the 12 recipient sites, as
+estimated for each site from the mrmr CMR model. Error bars show the
+95\% uncertainty intervals. Sites are arranged along the x-axis using
+the average of the median 1-year survival per translocation at each
+site. Dot colors indicate the donor population from which frogs in each
+translocated cohort were collected. When multiple translocations were
+conducted to a site, points and error bars are slightly offset to avoid
+overlap.}
+\end{figure}
 
@@
 
+\begin{SCfigure*}[\sidecaptionrelwidth][t!]
+{\centering \includegraphics[width=5.5in]{conditional_effects_plot_4pt5in.pdf}}
+\caption{\label{fig-cond-effects}Results from the among-site
+meta-analysis showing conditional effects of the important predictors of
+1-year frog survival (expressed as a probability): (A) winter severity
+in the year following translocation, (B) elevation of recipient site,
+(C) donor population, and (D) sex. In (A) and (B), blue lines are
+medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
+box plots show medians, first and third quartiles, largest and smallest
+values within 1.5x interquartile range, and values outside the 1.5x
+interquartile range.}
+\end{SCfigure*}
+
 In summary, results from our frog translocation study indicate that (i)
@@
 
-Results from the frog translocation study indicated that most
-populations showed evidence of successful reproduction and recruitment,
-and that adult survival was often relatively high (described above).
-Although suggestive of population establishment, a decade or more of
-surveys may be necessary to confirm that populations are in fact
-self-sustaining \citep{joseph2018}. To extend our inferences of
-population establishment beyond those possible from the site-specific
-CMR data, we developed a population viability model. Specifically, to
-test whether the observed yearly adult survival probabilities in
-translocated populations were sufficient for long-term viability, we
-built a stage-structured matrix model that captured known frog
-demography and included demographic and environmental stochasticity. We
-parameterized the model using CMR data from translocated populations and
-known life history values in this system (Table S1).
+Results from the frog translocation study are suggestive of population
+establishment. However, a decade or more of surveys may be necessary to
+confirm that populations are in fact self-sustaining \citep{joseph2018}.
+To extend our inferences of population establishment beyond those
+possible from the site-specific CMR data, we developed a population
+viability model. Specifically, to test whether the observed yearly adult
+survival probabilities in translocated populations were sufficient for
+long-term viability, we built a stage-structured matrix model that
+captured known frog demography and included demographic and
+environmental stochasticity. We parameterized the model using CMR data
+from translocated populations and known life history values in this
+system (Table S1).
 
@@
 identified unambiguously using these differences in Bd prevalence and
-load. Finally, there is no potential for frog dispersal between the 9
-study populations due to intervening distances and topography, as well
-as the presence of introduced (predatory) fish and fish-induced habitat
-fragmentation.
+load. There is no potential for frog dispersal between the 9 study
+populations or any other populations due to intervening distances and
+topography, as well as the presence of introduced (predatory) fish and
+fish-induced habitat fragmentation.
+
+\begin{figure*}
+{\centering \includegraphics[width=7in]{pop_viability_figures_for_manuscript_7in.pdf}}
+\caption{\label{fig-viability}Results from population viability
+analysis: (A) Predicted long-run growth rate \(\lambda\) for different
+values of yearly adult survival probability \(\sigma_{A_R}\) and year-1
+juvenile survival and recruitment probability \(\sigma_{J_1}\), given
+the parameterized, deterministic model. Colored points show the
+predicted \(\lambda\) values for the twelve translocated populations
+when year-1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
+(indicated by the dashed line). The red line shows where
+\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
+other points. (B) Predicted 50-year extinction probabilities of the 12
+translocated populations, given demographic stochasticity, environmental
+variability in \(\sigma_{J_1}\), and different mean values of
+\(\sigma_{J_1}\). There are 6 lines at extinction probability = 1, 5 of
+which (70414--70619) are hidden beneath the line for 70505 when
+\(\sigma_{J_1} < 0.10\). (C) 100 simulated trajectories (black lines)
+from the population viability model that most closely matched the
+observed abundance trajectory of adult amphibians at site 70550 (light
+orange).}
+\end{figure*}
 
@@
 clumped distribution (Figure~\ref{fig-allelefrequencies}). These
-characteristics of naive populations precluded a replicated analyses
+characteristics of naive populations precluded a replicated analysis
 among pairs of geographically proximate naive and recovering
@@
 resistance in recovering populations. (See \textbf{Supporting
-Information - Frog evolution in response to Bd - Study design} for
-additional details regarding the study design.)
+Information - Frog genome evolution - Study design} for additional
+details.)
 
@@
 allele frequencies between naive versus recovering populations) from 7
-distinct genes across 4 contigs (Figure S6B, C
-SI). One of the 7 identified genes (LOC108802036) does not have an
+distinct genes across 4 contigs (Figure S6B, C). One of the 7 identified genes (LOC108802036) does not have an
 associated annotation. For the outlier SNPs, frequency differences
@@
 limited geographic extent than RIN3, they still describe results that
-are consistent with parallel evolutionary changes following Bd exposure.
+are consistent with parallel evolutionary changes.
 
@@
 (Figure~\ref{fig-spline-manhattan}D, Figure S7).
-This region (854kb from the beginning of the outlier window to the VARS
+This region (854kb from the beginning of the outlier region to the VARS
 SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
@@
 Therefore, this region is part of the extended MHC Class I and III
-complex and shows synteny with other amphibian genomes.
+complex and shows synteny with other amphibian genomes. To provide
+details specifically on possible selection in the Yosemite populations
+described in \textbf{Results - Frog population recovery}, \(\pi\) values
+for the 9 shared outlier regions are shown in
+Figure S8 for each of the four Yosemite
+populations; two of these populations provided frogs for 20 of the 24
+frog translocations.
+
+\begin{SCfigure*}[\sidecaptionrelwidth][t!]
+{\centering \includegraphics[width=4.5in]{allele_maps_4pt5in.pdf}}
+\caption{\label{fig-allelefrequencies}Genomic patterns consistent with
+selection on individual variants in recovering MYL frog populations at
+the landscape scale (based on a linear mixed model analysis). For each
+of the 9 naive and recovering MYL frog populations (indicated by
+numbered points), adjacent pie charts show allele frequencies for the 11
+outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS,
+MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map
+of the Sierra Nevada study area, with Yosemite, Kings Canyon, and
+Sequoia National Parks (from north to south) shown in dark gray, and the
+range boundary for MYL frogs shown in light gray. The inset map locates
+the study area and range boundary in California. Sites 1 and 4 (site id
+= 72996 and 70567, respectively) were also used as sources of frogs for
+reintroductions described in \textbf{Results - Frog population
+recovery}.}
+\end{SCfigure*}
 
@@
 naive and recovering populations (see \textbf{Supporting Information -
-Frog evolution in response to Bd - Genetic diversity} for details).
-Thus, despite patterns consistent with localized selection in particular
-regions of the genome, we did not find evidence for reduced genetic
-diversity across the genome in recovering populations. In addition, no
-gene ontology (GO) biological functions, molecular functions, or
-cellular processes were over-represented in either the outlier variants
-or the 35 genes located in the overlapping \emph{F\textsubscript{ST}}
-and \(\pi_{diff}\) splined windows (see \textbf{Supporting Information -
-Frog evolution in response to Bd} \textbf{- GO analysis} for details).
+Frog genomic evolution - Genetic diversity} for details). Thus, despite
+patterns consistent with localized selection in particular regions of
+the genome, we did not find evidence for reduced genetic diversity
+across the genome in recovering populations. In addition, no gene
+ontology (GO) biological functions, molecular functions, or cellular
+processes were over-represented in either the outlier variants or the 35
+genes located in the overlapping \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) splined windows (see \textbf{Supporting Information -
+Frog genomic evolution} \textbf{- GO analysis} for details).
 
@@
 reestablish extirpated populations of the endangered MYL frog in the
-presence of a highly virulent fungal pathogen (Bd). Our results indicate
-(i) the capacity of reintroduced populations to become established and
+presence of a highly virulent pathogen (Bd). Our results indicate (i)
+the capacity of reintroduced populations to become established and
 eventually recover despite ongoing disease, (ii) that 50\% of the
-recovering populations are likely to persist over a 50-year period,
+reintroduced populations are likely to persist over a 50-year period,
 (iii) that there are genomic differences between naive and recovering
@@
 of amphibian reintroduction efforts \citep{dodd2005}, our success in
-reestablishing MYL frog populations via translocation of resistant
+reestablishing MYL frog populations via reintroduction of resistant
 individuals is striking, and even more so given that MYL frogs were
 driven to near-extinction by Bd.
+
+\begin{SCfigure*}[\sidecaptionrelwidth][t!]
+{\centering \includegraphics[width=4.5in]{splinewindow_manhattan_4pt5in.pdf}}
+\caption{\label{fig-spline-manhattan}Results from the splined window
+analysis showing patterns consistent with selection on genomic regions
+in recovering MYL frog populations. Note that several key candidate
+genes (RIN3, TCF19, VARS) are outliers in both the individual variant
+analysis (Figure~\ref{fig-allelefrequencies}) and splined window
+analysis (this figure). Manhattan plots show outlier regions for the
+difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
+\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
+upper red dashed line and below the lower red dashed line. In (B),
+outlier regions are shown above the single dashed red line. Outlier
+regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
+shown in blue and outlier regions for both \(\pi_{diff}\) and
+\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
+(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
+upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
+line). (D) Magnified Contig8 from from (B) showing the
+\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
+TCF19 and VARS. This region of the genome contains 8 annotated genes
+known to occur in the extended MHC Class I and III regions.}
+\end{SCfigure*}
 
@@
 \citep{zamudio2020}. For example, the MHC plays an important role in
-immunity. In our study, we identified a region that shows evidence of
-selection in recovering populations and contains eight genes associated
-with either the MHC Class I or Class III regions. These results
-corroborate numerous previous studies linking MHC genes to amphibian
-resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
-Similarly, the region with the strongest pattern of possible directional
-selection (as measured by \(\pi_{diff}\)) contains the
-interferon-related gene GVINP1. Several previous studies of amphibians
-have found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
-in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
-linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep[i.e., gill damage and amoebic load as heritable host
-resistance traits,][]{robledo2020, robledo2018}. We also identified a
-region, characterized by high \emph{F\textsubscript{ST}} and low
-\(\pi_{diff}\), that contained the complement genes C6 and C7. The
-complement system plays an important role in innate immunity
-\citep{rieraromo2016}, and our results could indicate that balancing
-selection is acting in this region of the genome to favor a diverse set
-of alleles, as is known for C6 in humans \citep{soejima2005}. Based on
-the analysis of individual outlier variants, the RIN3 gene showed a
-consistent pattern of allele frequency differences across all nine of
-the frog populations sampled in this study, suggesting possible
-consistent selection in populations distributed across a wide geographic
-area. This gene is associated with immune response and in \emph{Xenopus}
-is expressed during appendage regeneration \citep{fukazawa2009}.
-Finally, the outlier variant with the lowest p-value was the
-uncharacterized gene LOC108802036. In the genome of another frog
-species, this gene is located adjacent to a type I interferon gene
-(Np-IFNi2) \citep{gan2018}, and together with GVINP1 further suggests
-the importance of interferon-related genes in this system. Collectively,
-the genes associated with these genomic differences may confer at least
-some degree of resistance against Bd infection, an attribute that may be
-critically important to population reestablishment and recovery in the
-presence of Bd.
+immunity. We identified a region containing eight genes associated with
+the MHC Class I and III regions that showed patterns consistent with
+selection in recovering populations. These results align with numerous
+previous studies linking MHC genes to amphibian resistance against Bd
+\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
+the strongest pattern of possible directional selection (as measured by
+\(\pi_{diff}\)) contains the interferon-related gene GVINP1. Several
+previous studies of amphibians have found this gene to be differentially
+expressed during Bd infection \citep[e.g.,][]{grogan2018b, ellison2014}
+and in populations differing in Bd susceptibility \citep{grogan2018b}.
+This gene is also strongly linked to disease in salmon, explaining a
+notable 20\% of the resistance phenotype \citep[i.e., gill damage and
+amoebic load as heritable host resistance
+traits,][]{robledo2020, robledo2018}. We also identified a region,
+characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
+that contained the complement genes C6 and C7. The complement system
+plays an important role in innate immunity \citep{rieraromo2016}, and
+our results could indicate that balancing selection is acting in this
+region of the genome to favor a diverse set of alleles, as is known for
+C6 in humans \citep{soejima2005}. Based on the analysis of individual
+outlier variants, the RIN3 gene showed a consistent pattern of allele
+frequency differences across all nine of the frog populations sampled in
+this study, suggesting possible consistent selection in populations
+distributed across a wide geographic area. This gene is associated with
+immune response and in \emph{Xenopus} is expressed during appendage
+regeneration \citep{fukazawa2009}. Finally, the outlier variant with the
+lowest p-value was the uncharacterized gene LOC108802036. In the genome
+of another frog species, this gene is located adjacent to a type I
+interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
+further suggests the importance of interferon-related genes in this
+system. Collectively, the genes associated with these genomic
+differences may confer at least some degree of resistance against Bd
+infection, an attribute that may be critically important to population
+reestablishment and recovery in the presence of Bd.
 
@@
 effect of site characteristics, but were also strongly influenced by
-resistance inherent in the frogs. Although it could be argued that the
-relatively invariant Bd loads before versus after translocation are a
-consequence of similar pathogen pressure in the donor and translocated
-populations, this is at odds with the fact that in the first year after
-translocation frog densities are typically 1-2 orders of magnitude lower
-in the translocated versus donor populations and pathogen pressure
-should follow a similar pattern. In addition to the maintenance of Bd
-load and frog survival between natal and translocation sites, the
-relatively high survival of translocated frogs was maintained in their
-progeny, as expected if resistance has a genetic basis.
+intrinsic characteristics of frogs, including resistance. Although it
+could be argued that the relatively invariant Bd loads before versus
+after translocation are a consequence of similar pathogen pressure in
+the donor and reintroduced populations, this is at odds with the fact
+that in the first year after translocation frog densities are typically
+1-2 orders of magnitude lower in the reintroduced versus donor
+populations and pathogen pressure should follow a similar pattern. In
+addition to the maintenance of Bd load and frog survival between natal
+and recipient sites, the relatively high survival of translocated frogs
+was maintained in their progeny, as expected if resistance has a genetic
+basis.
 
@@
 Making well-supported projections about the long-term outcome of
-reintroduction efforts from shorter-term information is critically
-important to the process of adaptive management of species
-reintroduction programs \citep{seddon2007}, including the one we are
-carrying out for MYL frogs. Specifically, the combined results from our
-reintroduction study and viability modeling indicate that survival of
-frogs in the first year following translocation is an effective proxy of
-longer-term survival and population viability. In addition, given the
-repeatability of frog survival at a site, 1-year frog survival also
-serves as an effective proxy of site quality (i.e., the ability of a
-site to support high frog survival and a viable frog population over the
-long term). This proxy of site quality is important in the MYL frog
-system because accurately predicting the ability of a site to support a
-viable frog population \emph{a priori} remains difficult, even after
-conducting 24 translocations over 16 years.
-
-The results of the meta-analysis indicate the important influence of
-several variables on adult survival, and by extension on population
-viability. Although we assume that the effects of winter severity and
-frog sex are direct effects of the variables themselves, the elevation
-and donor population variables are likely serving as proxies for a range
-of factors. For example, elevation is correlated with several attributes
+reintroduction efforts from shorter-term information is crucial for
+adaptive management of species reintroduction programs
+\citep{seddon2007}, including the one we are carrying out for MYL frogs.
+Specifically, the combined results from our reintroduction study and
+viability modeling indicate that survival of frogs in the first year
+following translocation is an effective proxy of longer-term survival
+and population viability. In addition, given the repeatability of frog
+survival at a site, 1-year frog survival also serves as an effective
+proxy of site quality (i.e., the ability of a site to support high frog
+survival and a viable frog population over the long term). This proxy of
+site quality is important in the MYL frog system because accurately
+predicting the ability of a site to support a viable frog population
+\emph{a priori} remains difficult, even after conducting 24
+translocations over 16 years.
+
+The results of the meta-analysis indicate the influence of several
+variables on adult survival, and by extension on population viability.
+Although we assume that the effects of winter severity and frog sex are
+direct effects of the variables themselves, the elevation and donor
+population variables are likely serving as proxies for a range of
+factors. For example, elevation is correlated with several attributes
 that may influence frog survival, including the presence of boulder
@@
 duration of the tadpole and juvenile stages increases with elevation,
-but the fact that long-run population growth rate is relatively
-insensitive to variation in these durations suggests that developmental
-rates of early frog life stages have little effect on the predicted
-long-term viability of reintroduced populations. Nonetheless, a clearer
-understanding of the mechanistic components of the elevation and donor
-population variables will require additional research.
+but the fact that long-run population growth rate from our viability
+model is relatively insensitive to variation in these durations suggests
+that developmental rates of early frog life stages have minimal effect
+on the predicted long-term viability of reintroduced populations.
+Nonetheless, a clearer understanding of the mechanistic components of
+the elevation and donor population variables will require additional
+research.
 
@@
 anurans, and may result from the poorly developed immune system
-characteristic of this life stage \citep{humphries2022}. In naturally
-recovering and translocated MYL frog populations, we suggest that the
-high mortality of metamorphs is an important limitation on subsequent
-recruitment of new adults. Therefore, although adult MYL frogs appear
-relatively resistant, Bd infection continues to have important limiting
-effects on recovering populations \citep[see also][]{hollanders2022}.
+characteristic of this life stage \citep{humphries2022}. In both
+naturally recovering and reintroduced MYL frog populations, we suggest
+that the high mortality of metamorphs is an important limitation on
+subsequent recruitment of new adults. Therefore, although adult MYL
+frogs appear relatively resistant, Bd infection continues to have
+important limiting effects on recovering populations \citep[see
+also][]{hollanders2022}.
 
@@
 In our rapidly changing world, evolution is likely to play in important
-role in facilitating the resilience of wildlife populations. Whether the
-documented disease resistance in MYL frogs and concurrent recovery of
-decimated populations provides an airtight example of evolutionary
-rescue will likely always be uncertain (given that we can never have a
-perfect understanding of the past). Regardless, we provide an example
-from the wild that suggests that evolution might produce individuals
-that harbor adaptive alleles and allow population recovery in a novel
-(i.e., Bd-positive) environment, and show conclusively that individuals
-from these recovering populations can be used to reestablish extirpated
-populations and expand the scale of natural recovery
+role in facilitating the resilience of wildlife populations. We provide
+an example from the wild that suggests that evolution might produce
+individuals that harbor adaptive alleles and allow population recovery
+in a novel (i.e., Bd-positive) environment, and show conclusively that
+individuals from these recovering populations can be used to reestablish
+extirpated populations and expand the scale of natural recovery
 (Figure~\ref{fig-recovery-model}). We expect that similar species
@@
 incorporating new information about important habitat features, in
-particular, overwinter habitats such as submerged boulders and
+particular, overwintering habitats such as submerged boulders and
 overhanging banks. \emph{R. sierrae} were absent from all recipient
@@
 Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
-SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
-bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
-using Picard. Variants were then called following GATK best practices
+\citep{chen2018} and aligned to the \emph{R. muscosa} genome (NCBI SRA:
+SRS6533475 \citep{hon2020}), with repetitive elements masked using bwa
+(``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked using
+Picard. Variants were then called following GATK best practices
 (v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
@@
 with an average depth across filtered sites \textless{} 7x (n = 3). Our
-final dataset included 50 samples, 23 from naive and 27 from recovering
-populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+final dataset included 50 samples, 23 from naive populations and 27 from
+recovering populations, with an average depth of 16.7x (range = 7.4x --
+26.1x).
 
@@
 \(\pi_{diff}\). These standard deviations represent strict criteria to
-select only the top \textasciitilde{} 0.3\% of windows. Shared outliers
-were then identified as those that were outliers in both analyses,
-meaning that they showed (i) high differentiation between naive and
-recovering populations, and (ii) differential patterns of nucleotide
-diversity in the same region. Finally, we extracted gene transcripts
-mapped within each region and retrieved annotation for that region using
-BLAST (\textbf{Supporting Information}: Datasets S3, S4).
+select only the top approximately 0.3\% of windows. Shared outliers were
+then identified as those that were outliers in both analyses, meaning
+that they showed (i) high differentiation between naive and recovering
+populations, and (ii) differential patterns of nucleotide diversity in
+the same region. Finally, we extracted gene transcripts mapped within
+each region and retrieved annotation for that region using BLAST
+(\textbf{Supporting Information}: Datasets S3, S4).
 
@@
 Hegeman and A. Lindauer (project and data management, field and
-laboratory work); A. Barbella and K. Rose-(laboratory work); C. Kamoroff
-and numerous summer technicians (field work); staff at Sequoia-Kings
-Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
-California Department of Fish and Wildlife, U.S. Fish and Wildlife
-Service, University of California-Santa Barbara Institutional Animal
-Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory
-(research permits, logistical support, and/or field assistance). This
-project was supported by grants from the National Park Service (to
-R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science
-Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
-DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
+laboratory work); A. Barbella and K. Rose (laboratory work); numerous
+summer technicians (field work); staff at Sequoia-Kings Canyon and
+Yosemite National Parks, Inyo and Sierra National Forests, California
+Department of Fish and Wildlife, U.S. Fish and Wildlife Service,
+University of California-Santa Barbara Institutional Animal Care and Use
+Committee, and Sierra Nevada Aquatic Research Laboratory (research
+permits, logistical support, and/or field assistance). We also thank the
+editor and three anonymous reviewers whose input helped us improve the
+manuscript. This project was supported by grants from the National Park
+Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
+Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).
+
+\hypertarget{references}{%
+\subsubsection{References}\label{references}}}
 
@@
 \clearpage
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
-
-}
-
-\caption{\label{fig-recovery-model}For MYL frogs, a conceptual model
-depicting the Bd-caused decline and subsequent natural recovery (black
-text), facilitated recovery via reintroductions, and the linkages
-between these two pathways. Rectangles and hexagons represent outcomes
-and processes, respectively. Blue text indicates components that are
-included in the current study. We quantified frog survival and
-population viability following reintroductions directly using field
-studies and models, and evaluated ``disease-mediated selection''
-indirectly using a comparison of frog exomes in naive and recovering
-populations. The timeline shows the general sequence of the components,
-with the dashed portion indicating a projection into the future.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
-
-}
-
-\caption{\label{fig-translocation-survival}Median 1-year survival for
-each cohort of translocated frogs at the 12 recipient sites, as
-estimated for each site from the mrmr CMR model. Error bars show the
-95\% uncertainty intervals. Sites are arranged along the x-axis using
-the average of the median 1-year survival per translocation at each
-site. Dot colors indicate the donor population from which frogs in each
-translocated cohort were collected. When multiple translocations were
-conducted to a site, points and error bars are slightly offset to avoid
-overlap.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
-
-}
-
-\caption{\label{fig-cond-effects}Results from the among-site
-meta-analysis showing conditional effects of the important predictors of
-1-year frog survival (expressed as a probability): (A) winter severity
-in the year following translocation, (B) elevation of recipient site,
-(C) donor population, and (D) sex. In (A) and (B), blue lines are
-medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
-box plots show medians, first and third quartiles, largest and smallest
-values within 1.5x interquartile range, and values outside the 1.5x
-interquartile range.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
-
-}
-
-\caption{\label{fig-viability}Results from population viability
-analysis: (A) Predicted long-run growth rate \(\lambda\) for different
-values of yearly adult survival probability \(\sigma_{A_R}\) and year-1
-juvenile survival and recruitment probability \(\sigma_{J_1}\), given
-the parameterized, deterministic model. Colored points show the
-predicted \(\lambda\) values for the twelve translocated populations
-when year-1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
-(indicated by the dashed line). The red line shows where
-\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
-other points. (B) Predicted 50-year extinction probabilities of the 12
-translocated populations, given demographic stochasticity, environmental
-variability in \(\sigma_{J_1}\), and different mean values of
-\(\sigma_{J_1}\). There are 6 lines at extinction probability = 1, 5 of
-which (70414--70619) are hidden beneath the line for 70505 when
-\(\sigma_{J_1} < 0.10\). (C) 100 simulated trajectories (black lines)
-from the population viability model that most closely matched the
-observed abundance trajectory of adult amphibians at site 70550 (light
-orange).}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
-
-}
-
-\caption{\label{fig-allelefrequencies}Genomic patterns consistent with
-selection on individual variants in recovering MYL frog populations at
-the landscape scale (based on a linear mixed model analysis). For each
-of the 9 naive and recovering MYL frog populations (indicated by
-numbered points), adjacent pie charts show allele frequencies for the 11
-outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS,
-MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map
-of the Sierra Nevada study area, with Yosemite, Kings Canyon, and
-Sequoia National Parks (from north to south) shown in dark gray, and the
-range boundary for MYL frogs shown in light gray. The inset map locates
-the study area and range boundary in California. Sites 1 and 4 (site id
-= 72996 and 70567, respectively) were also used as sources of frogs in
-the frog population recovery study.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}
-
-}
-
-\caption{\label{fig-spline-manhattan}Patterns consistent with selection
-on genomic regions in recovering MYL frog populations based on a splined
-window analysis. Note that several key candidate genes (RIN3, TCF19,
-VARS) are outliers in both the individual variant analysis
-Figure~\ref{fig-allelefrequencies} and splined window analysis (this
-figure). Manhattan plots show outlier regions for the difference in (A)
-nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
-In (A), outlier regions are shown above the upper red dashed line and
-below the lower red dashed line. In (B), outlier regions are shown above
-the single dashed red line. Outlier regions for either \(\pi_{diff}\) or
-\emph{F\textsubscript{ST}} are shown in blue and outlier regions for
-both \(\pi_{diff}\) and \emph{F\textsubscript{ST}} are shown in red. (C)
-Magnified Contig19 from (A) showing two adjacent outlier regions for
-\(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier SNP (indicated with a
-dashed vertical blue line). (D) Magnified Contig8 from from (B) showing
-the \emph{F\textsubscript{ST}} outlier region that includes the outlier
-SNPs TCF19 and VARS. This region of the genome contains 8 annotated
-genes known to occur in the extended MHC Class I and III regions.}
-
-\end{figure*}
-
 \newpage

### 2024-04-21 doc/manuscript/translocation.qmd
Render to latex using revised template
Create final files for submission

--- 2024-04-18 doc/manuscript/translocation.qmd
@@
 ---
-# format: pdf
+#format: docx
 format:
@@
 
-The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global declines of amphibians caused by Bd [@scheele2019].
+The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
 Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog evolution in response to Bd:** population 1 and 4).
-In addition, this subset of 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
+The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
+These 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
@@
 
-Results from the frog translocation study indicated that most populations showed evidence of successful reproduction and recruitment, and that adult survival was often relatively high (described above).
-Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
+Results from the frog translocation study are suggestive of population establishment.
+However, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
 To extend our inferences of population establishment beyond those possible from the site-specific CMR data, we developed a population viability model.
@@
 Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-Finally, there is no potential for frog dispersal between the 9 study populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
+There is no potential for frog dispersal between the 9 study populations or any other populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
 
 It is important to note that our genomic analysis is necessarily limited by the small number of naive populations remaining on the landscape (nearly all of which we sampled as part of the current study) and their clumped distribution ([@fig-allelefrequencies]).
-These characteristics of naive populations precluded a replicated analyses among pairs of geographically proximate naive and recovering populations.
+These characteristics of naive populations precluded a replicated analysis among pairs of geographically proximate naive and recovering populations.
 Instead, to obtain adequate power, we conducted a pooled analysis across population types to identify genomic regions showing patterns consistent with selection.
 Our comparative approach cannot provide a direct assessment of the strength or timing of natural selection, but does allow us to identify specific regions of the genome that merit further study for their potential contribution to frog resistance in recovering populations.
-(See **Supporting Information - Frog evolution in response to Bd - Study design** for additional details regarding the study design.)
+(See **Supporting Information - Frog genome evolution - Study design** for additional details.)
 
@@
 Results from the individual variant and splined window analyses show that recovering populations show patterns consistent with selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C SI).
+The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
 One of the 7 identified genes (LOC108802036) does not have an associated annotation.
@@
 The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found statistically significant differences in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes following Bd exposure.
-<!--# Add text citing new figure -->
+Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes.
 
@@
 Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
-This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
+This region (854kb from the beginning of the outlier region to the VARS SNP) contained a total of 8 annotated genes.
 In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
 Therefore, this region is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-
-Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog evolution in response to Bd - Genetic diversity** for details).
+To provide details specifically on possible selection in the Yosemite populations described in **Results - Frog population recovery**, $\pi$ values for the 9 shared outlier regions are shown in @fig-ave-pi-per-popn-yose SI for each of the four Yosemite populations; two of these populations provided frogs for 20 of the 24 frog translocations.
+
+Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog genomic evolution - Genetic diversity** for details).
 Thus, despite patterns consistent with localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
-In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog evolution in response to Bd** **- GO analysis** for details).
+In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog genomic evolution** **- GO analysis** for details).
 
@@
 Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
-Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent fungal pathogen (Bd).
-Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the recovering populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
+Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent pathogen (Bd).
+Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the reintroduced populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
 Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
-In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.
+In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via reintroduction of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.
 
@@
 For example, the MHC plays an important role in immunity.
-In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
-These results corroborate numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
+We identified a region containing eight genes associated with the MHC Class I and III regions that showed patterns consistent with selection in recovering populations.
+These results align with numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
 Similarly, the region with the strongest pattern of possible directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
@@
 Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
-The fact that the relatively low Bd loads and correspondingly high frog survival was maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by resistance inherent in the frogs.
-Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and translocated populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the translocated versus donor populations and pathogen pressure should follow a similar pattern.
-In addition to the maintenance of Bd load and frog survival between natal and translocation sites, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
+The fact that the relatively low Bd loads and correspondingly high frog survival was maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
+Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and reintroduced populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the reintroduced versus donor populations and pathogen pressure should follow a similar pattern.
+In addition to the maintenance of Bd load and frog survival between natal and recipient sites, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
 
@@
 These long-term forecasts are important, given that reintroduced MYL frog populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
-Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is critically important to the process of adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
+Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is crucial for adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
 Specifically, the combined results from our reintroduction study and viability modeling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
@@
 
-The results of the meta-analysis indicate the important influence of several variables on adult survival, and by extension on population viability.
+The results of the meta-analysis indicate the influence of several variables on adult survival, and by extension on population viability.
 Although we assume that the effects of winter severity and frog sex are direct effects of the variables themselves, the elevation and donor population variables are likely serving as proxies for a range of factors.
@@
 The latter can influence the duration of frog life stages that have differential susceptibility to chytridiomycosis (see below).
-The duration of the tadpole and juvenile stages increases with elevation, but the fact that long-run population growth rate is relatively insensitive to variation in these durations suggests that developmental rates of early frog life stages have little effect on the predicted long-term viability of reintroduced populations.
+The duration of the tadpole and juvenile stages increases with elevation, but the fact that long-run population growth rate from our viability model is relatively insensitive to variation in these durations suggests that developmental rates of early frog life stages have minimal effect on the predicted long-term viability of reintroduced populations.
 Nonetheless, a clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
@@
 This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
-In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
+In both naturally recovering and reintroduced MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
 Therefore, although adult MYL frogs appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
@@
 In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
-Whether the documented disease resistance in MYL frogs and concurrent recovery of decimated populations provides an airtight example of evolutionary rescue will likely always be uncertain (given that we can never have a perfect understanding of the past).
-Regardless, we provide an example from the wild that suggests that evolution might produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
+We provide an example from the wild that suggests that evolution might produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
 We expect that similar species recovery actions will be an essential tool in wildlife conservation in an era of accelerating global change.
@@
 At the beginning of this study, we had a relatively limited understanding of the factors that affect habitat quality.
-In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwinter habitats such as submerged boulders and overhanging banks.
+In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwintering habitats such as submerged boulders and overhanging banks.
 *R. sierrae* were absent from all recipient sites prior to the first translocation.
@@
 
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
+Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *R.
+muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
 Exact PCR duplicates were marked using Picard.
@@
 Finally, we trimmed samples with an average depth across filtered sites \< 7x (n = 3).
-Our final dataset included 50 samples, 23 from naive and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+Our final dataset included 50 samples, 23 from naive populations and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
 
@@
 We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-These standard deviations represent strict criteria to select only the top \~ 0.3% of windows.
+These standard deviations represent strict criteria to select only the top approximately 0.3% of windows.
 Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (i) high differentiation between naive and recovering populations, and (ii) differential patterns of nucleotide diversity in the same region.
@@
 Barbella and K.
-Rose-(laboratory work); C.
-Kamoroff and numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
+Rose (laboratory work); numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
 Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
+We also thank the editor and three anonymous reviewers whose input helped us improve the manuscript.
 This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
@@
 
-See combined references at end of document (temporary placeholder).
-
-\
-
 ### Figures
@@
 We quantified frog survival and population viability following reintroductions directly using field studies and models, and evaluated "disease-mediated selection" indirectly using a comparison of frog exomes in naive and recovering populations.
-The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model}
+The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
 
@@
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="60%"}
 
@@
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="550"}
+In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
 
@@
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="650"}
+(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
 
@@
 The inset map locates the study area and range boundary in California.
-Sites 1 and 4 (site id = 72996 and 70567, respectively) were also used as sources of frogs in the frog population recovery study.](figures/allele_maps.png){#fig-allelefrequencies width="600"}
-
-\newpage
-
-![Patterns consistent with selection on genomic regions in recovering MYL frog populations based on a splined window analysis.
-Note that several key candidate genes (RIN3, TCF19, VARS) are outliers in both the individual variant analysis [@fig-allelefrequencies] and splined window analysis (this figure).
+Sites 1 and 4 (site id = 72996 and 70567, respectively) were also used as sources of frogs for reintroductions described in **Results - Frog population recovery**.](figures/allele_maps.png){#fig-allelefrequencies width="95%"}
+
+\newpage
+
+![Results from the splined window analysis showing patterns consistent with selection on genomic regions in recovering MYL frog populations.
+Note that several key candidate genes (RIN3, TCF19, VARS) are outliers in both the individual variant analysis (@fig-allelefrequencies) and splined window analysis (this figure).
 Manhattan plots show outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
@@
 (D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="550"}
+This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="95%"}
 
@@
 Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012].
+This parameter expanded data augmentation allows us to account for the possibility that undetected individuals may have recruited into the adult population [@royle2012].
 Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
@@
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)) .
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)).
 This yielded posterior distributions for survival probabilities between primary periods.
@@
 For example, if we set $\phi = 0.001$, such that in a given year essentially either all year-1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
-Because we do not pit tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
+Because we do not PIT tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
 However, based on qualitative and semi-quantitative field observations over 25 years, a value of $\sigma_{J_1} = 0.25$ in the presence of Bd is probably a reasonable estimate for many populations.
@@
 
-In **Results-Frog evolution in response to Bd**, we describe the stringent set of outlier variants (identified using a Bonferroni-corrected p-value of 0.01).
+In **Results-Frog genome evolution**, we describe the stringent set of outlier variants (identified using a Bonferroni-corrected p-value of 0.01).
 A liberal set of outlier variants, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
@@
 Second, to characterize genome-wide patterns of diversity, we used VCFtools to calculate nucleotide diversity ($\pi$) in 100kb sliding windows along the genome for each population.
-Third, we calculated average $\pi$ per population within each of the 9 outlier windows identified in the splined window analysis.
+Third, we calculated average $\pi$ per population within each of the 9 shared outlier regions identified in the splined window analysis.
 We compared average $\pi$ per population between naive and recovering population types, and also for each naive and recovering population in Yosemite.
-The latter comparison focused on Yosemite because these populations are the most relevant to the translocated populations described in the paper (see results below for additional details).
+The latter comparison focused on Yosemite because these populations are the most relevant to the translocated populations described in the paper.
 
 Average individual-level heterozygosity and genome-wide population-level $\pi$ were similar between the naive and recovering groups (@fig-violinplot-heterozy SI; @fig-boxplot-genomewide-pi-by-pop SI).
-Within each of the 9 outlier windows, average $\pi$ shows considerable variation between populations and no obvious patterns between the naive and recovering populations (@fig-boxplot-pi-by-windowpop SI).
-For the four populations in Yosemite, within each of the 9 outlier windows, average $\pi$ per population was generally higher in the 3 recovering populations than in the 1 naive population (@fig-ave-pi-per-site-yose SI).
-ADD CONCLUDING SENTENCE OR TWO THAT SUMMARIZE THE SIGNIFICANCE OF THESE RESULTS?
+Within each of the 9 shared outlier regions, $\pi$ shows considerable variation between populations and no obvious patterns between the naive and recovering populations (@fig-boxplot-pi-by-regionpop SI).
 
@@
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="60%"}
 
@@
 The red dashed horizontal line indicates the approximate load above which severe disease is typically observed [@vredenburg2010].
-This threshold (5.8 ITS1 copies - on a log~10~ scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations [@briggs2010].](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="630"}
+This threshold (5.8 ITS1 copies - on a log~10~ scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations [@briggs2010].](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
 
@@
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
-Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="500"}
+Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling - Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="580"}
+All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling - Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="60%"}
 
@@
 Sensitivities of all other parameters are comparable.
-We did not include $p_{L_1}$ as this parameter is generally 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="650"}
-
-\newpage
-
-![Results from the analysis of individual variants, showing putative signatures of selection in recovering MYL frog populations.
+We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
+Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
+
+\newpage
+
+![Results from the analysis of individual variants, showing possible signatures of selection in recovering MYL frog populations.
 (A) PCA calculated from binary SNPs showing the genomic relationship of samples.
@@
 Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="550"}
+Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="85%"}
 
@@
 The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana muscosa* Contig19.
-Asterices indicate the location of SNP outliers in the TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="650"}
+Asterisks indicate the location of SNP outliers in the TCF19 and VARS1 genes.
+Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="95%"}
+
+\newpage
+
+![For each study population located in Yosemite National Park (populations 1-4 shown in the map), $\pi$ for the 9 shared outlier regions identified in the splined window analysis.
+Populations 1 and 4 were the source of frogs for 20 of the 24 frog translocations, and the donor population for the remaining four translocations is located adjacent to populations 3 and 4 (see also @fig-yosemap).
+Bars are color-coded by population type (naive, recovering).
+The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.
+Region 4 is the extended MHC region that includes two SNPs identified by the GEMMA analysis in genes TCF19 and VARS1.
+In the map, the four populations included in panels 1-9 are color-coded by population type, and dashed lines delineate the genetic clusters as defined in [@fig-selectionresults]A.](figures/ave_pi_per_popn_yose.png){#fig-ave-pi-per-popn-yose width="85%"}
 
@@
 ![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog evolution study.
-Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="500"}
-
-\newpage
-
-![For each study population in the frog evolution study (from @fig-allelefrequencies), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
+Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="60%"}
+
+\newpage
+
+![For each population in the frog evolution study (@fig-allelefrequencies), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
 The y-axis has been log~10~-transformed for display purposes.
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in @fig-selectionresults A SI).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="550"}
-
-\newpage
-
-![For each of the 9 outlier windows identified by the overlapping splined windows for *F~ST~* and $\pi$, boxplots showing average $\pi$ for naive and recovering populations.
-The corresponding gene annotations for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_windowpop.png){#fig-boxplot-pi-by-windowpop width="500"}
-
-\newpage
-
-![For each study population located in Yosemite National Park (populations 1-4 shown in Site Map), average $\pi$ for the 9 outlier windows (panels A-I) identified in the splined window analysis.
-Populations 1 and 4 were the source of frogs for 20 of the 24 translocations and the donor population for the remaining four translocations is located adjacent to populations 3 and 4.
-Bars are color-coded by population type (naive, recovering), and gene annotations within each window are listed above each panel.
-Gene regions described in the paper are in panel A (Complement C6,C7), panel E (GVINP1), and panel F (extended MHC region that includes two SNPs identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by an asterisk).
-In the Site Map, the four populations included in panels A-I are color-coded by population type, and dashed lines delineate the genetic clusters as defined in [@fig-selectionresults SI, panel A].
-<!--# Update caption based on feedback from Allie and Bree -->](figures/ave_pi_per_site_yose.png){#fig-ave-pi-per-site-yose width="650"}
+Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in [@fig-selectionresults]A).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="70%"}
+
+\newpage
+
+![For each of the 9 shared outlier regions identified by the splined window analysis, boxplots showing $\pi$ values for naive and recovering populations.
+The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_regionpop.png){#fig-boxplot-pi-by-regionpop width="70%"}
 

### 2024-04-21 doc/manuscript/translocation_bioarxiv.tex
Create final files for submission

@@
 identified unambiguously using these differences in Bd prevalence and
-load. Finally, there is no potential for frog dispersal between the 9
-study populations due to intervening distances and topography, as well
-as the presence of introduced (predatory) fish and fish-induced habitat
-fragmentation.
+load. There is no potential for frog dispersal between the 9 study
+populations or any other populations due to intervening distances and
+topography, as well as the presence of introduced (predatory) fish and
+fish-induced habitat fragmentation.
 
@@
 Therefore, this region is part of the extended MHC Class I and III
-complex and shows synteny with other amphibian genomes.
+complex and shows synteny with other amphibian genomes. To provide
+details specifically on possible selection in the Yosemite populations
+described in \textbf{Results - Frog population recovery}, \(\pi\) values
+for the 9 shared outlier regions are shown in
+Figure S8 for each of the four Yosemite
+populations; two of these populations provided frogs for 20 of the 24
+frog translocations.
 
@@
 of amphibian reintroduction efforts \citep{dodd2005}, our success in
-reestablishing MYL frog populations via translocation of resistant
+reestablishing MYL frog populations via reintroduction of resistant
 individuals is striking, and even more so given that MYL frogs were
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
+{\centering \includegraphics[width=0.95\textwidth]{figures/conceptual_model.png}
 
@@
 
-{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
+{\centering \includegraphics[width=0.60\textwidth]{figures/translocation_survival_bysiteid.png}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
+{\centering \includegraphics[width=0.80\textwidth]{figures/cond_effects_plot.png}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+{\centering \includegraphics[width=0.95\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
+{\centering \includegraphics[width=0.95\textwidth]{figures/allele_maps.png}
 
@@
 the study area and range boundary in California. Sites 1 and 4 (site id
-= 72996 and 70567, respectively) were also used as sources of frogs in
-the frog population recovery study.}
+= 72996 and 70567, respectively) were also used as sources of frogs for
+reintroductions described in \textbf{Results - Frog population
+recovery}.}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}
+{\centering \includegraphics[width=0.95\textwidth]{figures/splinewindow_manhattan.png}
 
@@
 
-\caption{\label{fig-spline-manhattan}Patterns consistent with selection
-on genomic regions in recovering MYL frog populations based on a splined
-window analysis. Note that several key candidate genes (RIN3, TCF19,
-VARS) are outliers in both the individual variant analysis
-Figure~\ref{fig-allelefrequencies} and splined window analysis (this
-figure). Manhattan plots show outlier regions for the difference in (A)
-nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
-In (A), outlier regions are shown above the upper red dashed line and
-below the lower red dashed line. In (B), outlier regions are shown above
-the single dashed red line. Outlier regions for either \(\pi_{diff}\) or
-\emph{F\textsubscript{ST}} are shown in blue and outlier regions for
-both \(\pi_{diff}\) and \emph{F\textsubscript{ST}} are shown in red. (C)
-Magnified Contig19 from (A) showing two adjacent outlier regions for
-\(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier SNP (indicated with a
-dashed vertical blue line). (D) Magnified Contig8 from from (B) showing
-the \emph{F\textsubscript{ST}} outlier region that includes the outlier
-SNPs TCF19 and VARS. This region of the genome contains 8 annotated
-genes known to occur in the extended MHC Class I and III regions.}
+\caption{\label{fig-spline-manhattan}Results from the splined window
+analysis showing patterns consistent with selection on genomic regions
+in recovering MYL frog populations. Note that several key candidate
+genes (RIN3, TCF19, VARS) are outliers in both the individual variant
+analysis (Figure~\ref{fig-allelefrequencies}) and splined window
+analysis (this figure). Manhattan plots show outlier regions for the
+difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
+\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
+upper red dashed line and below the lower red dashed line. In (B),
+outlier regions are shown above the single dashed red line. Outlier
+regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
+shown in blue and outlier regions for both \(\pi_{diff}\) and
+\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
+(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
+upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
+line). (D) Magnified Contig8 from from (B) showing the
+\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
+TCF19 and VARS. This region of the genome contains 8 annotated genes
+known to occur in the extended MHC Class I and III regions.}
 

### 2024-04-21 doc/manuscript/translocation_bioarxiv_SI.tex
Create final files for submission

@@
 individuals were captured. This parameter expanded data augmentation
-allows us to capture the possibility that undetected individuals may
+allows us to account for the possibility that undetected individuals may
 have recruited into the adult population \citep{royle2012}. Here, \(M\)
@@
 small compared to among-population variability
-(Fig. 2). Thus, we did not include
+(Figure 2). Thus, we did not include
 yearly within-population variability in adult survival in this analysis.
@@
 stochasticity in year-2 juvenile survival and recruitment
-\(\sigma_{J_2}\), but our elasticity analysis (\textbf{Supporting
-Information - Population viability modeling - Model analysis and
-simulation}) showed that this parameter had little effect on host growth
-rate relative to \(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp}).
+\(\sigma_{J_2}\), but our elasticity and sensitivity analysis (see
+below, \textbf{Model analysis and simulation}) showed that this
+parameter had little effect on host growth rate relative to
+\(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp}).
 
@@
 period, how often was the value still 2 (alive) in the next primary
-period compared to 3 (dead)) . This yielded posterior distributions for
+period compared to 3 (dead)). This yielded posterior distributions for
 survival probabilities between primary periods. However, because the
@@
 \(\sigma_{J_1}\) greater than 0.2 to have a 50-year extinction
-probability of less 50\%. Because we do not pit tag juveniles, we do not
+probability of less 50\%. Because we do not PIT tag juveniles, we do not
 have CMR estimates for \(\sigma_{J_1}\) or \(\phi\). However, based on
@@
 such that both population types were represented across similar
-latitudinal ranges (Fig. 5).
+latitudinal ranges (Figure 5).
 
@@
 
-In \textbf{Results-Frog evolution in response to Bd}, we describe the
-stringent set of outlier variants (identified using a
-Bonferroni-corrected p-value of 0.01). A liberal set of outlier
-variants, identified using a Bonferroni-corrected p-value of 0.05,
-included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
-across 16 contigs. We used this liberal set to determine if any GO
-biological functions, molecular functions, or cellular processes were
-overrepresented. To do this, we retrieved the BLAST hits and mapped GO
-terms for each gene in our targeted transcriptome. We then conducted a
-statistical overrepresentation test (Fisher's exact test) using Blast2GO
-\citep{conesa2005} to compare the 30 unique outlier genes to the
-complete set of genes in our target transcriptome. We repeated this
-process for the set of 35 genes located in the 9 shared regions of the
-\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
+In \textbf{Results-Frog genome evolution}, we describe the stringent set
+of outlier variants (identified using a Bonferroni-corrected p-value of
+0.01). A liberal set of outlier variants, identified using a
+Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and
+3 INDELS) from 30 distinct genes across 16 contigs. We used this liberal
+set to determine if any GO biological functions, molecular functions, or
+cellular processes were overrepresented. To do this, we retrieved the
+BLAST hits and mapped GO terms for each gene in our targeted
+transcriptome. We then conducted a statistical overrepresentation test
+(Fisher's exact test) using Blast2GO \citep{conesa2005} to compare the
+30 unique outlier genes to the complete set of genes in our target
+transcriptome. We repeated this process for the set of 35 genes located
+in the 9 shared regions of the \emph{F\textsubscript{ST}} and
+\(\pi_{diff}\) splined windows.
 
@@
 in 100kb sliding windows along the genome for each population. Third, we
-calculated average \(\pi\) per population within each of the 9 outlier
-windows identified in the splined window analysis.
+calculated average \(\pi\) per population within each of the 9 shared
+outlier regions identified in the splined window analysis. We compared
+average \(\pi\) per population between naive and recovering population
+types, and also for each naive and recovering population in Yosemite.
+The latter comparison focused on Yosemite because these populations are
+the most relevant to the translocated populations described in the
+paper.
 
@@
 Figure~\ref{fig-boxplot-genomewide-pi-by-pop}). Within each of the 9
-outlier windows, average \(\pi\) shows considerable variation between
-populations (Figure~\ref{fig-boxplot-pi-by-windowpop}) and no obvious
-patterns between the naive and recovering groups.
+shared outlier regions, \(\pi\) shows considerable variation between
+populations and no obvious patterns between the naive and recovering
+populations (Figure~\ref{fig-boxplot-pi-by-regionpop}).
 
@@
 
-{\centering \includegraphics[width=0.7\textwidth]{figures/bdload_beforeafter.png}
+{\centering \includegraphics[width=0.80\textwidth]{figures/bdload_beforeafter.png}
 
@@
 reproduction by translocated frogs). In contrast to
-Fig. 2, these are not survival
+Figure 2, these are not survival
 probabilities from the first year following translocation, but instead
@@
 
-{\centering \includegraphics[width=0.60\textwidth]{figures/pop_viability_figures_for_supp.jpg}
+{\centering \includegraphics[width=0.95\textwidth]{figures/pop_viability_figures_for_supp.jpg}
 
@@
 other parameters are comparable. We did not include \(p_{L_1}\) as this
-parameter is generally 1 across the populations in this study (Table S1)
--- tadpoles spend at least a year as tadpoles before metamorphosing.
+parameter is 1 across the populations in this study (Table S1) --
+tadpoles spend at least a year as tadpoles before metamorphosing.
 Elasticity and sensitivity is calculated at the default parameter values
@@
 \caption{\label{fig-selectionresults}Results from the analysis of
-individual variants, showing putative signatures of selection in
+individual variants, showing possible signatures of selection in
 recovering MYL frog populations. (A) PCA calculated from binary SNPs
 showing the genomic relationship of samples. Numeric labels and colors
-match those in Fig. 5. Populations 1-7 are
+match those in Figure 5. Populations 1-7 are
 \emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/synteny_figure.png}
+{\centering \includegraphics[width=0.95\textwidth]{figures/synteny_figure.png}
 
@@
 using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
-accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
+accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterisks
 indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/violin_plot_heterozy_by_group.png}
+{\centering \includegraphics[width=0.85\textwidth]{figures/ave_pi_per_popn_yose.png}
+
+}
+
+\caption{\label{fig-ave-pi-per-popn-yose}For each study population
+located in Yosemite National Park (populations 1-4 shown in the map),
+\(\pi\) for the 9 shared outlier regions identified in the splined
+window analysis. Populations 1 and 4 were the source of frogs for 20 of
+the 24 frog translocations, and the donor population for the remaining
+four translocations is located adjacent to populations 3 and 4 (see also
+Figure~\ref{fig-yosemap}). Bars are color-coded by population type
+(naive, recovering). The corresponding gene annotations for each region
+are as follows (gene regions described in the \textbf{Results} and
+\textbf{Discussion} are indicated with an asterisk): (1) C6*, C7*, (2)
+DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB,
+MDC1, CCHCR1, TCF19*, HSP70, LSM2, VARS1*, (5) GCC2, CFAP251, PEG10, (6)
+ERO1A, GVINP1*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8)
+C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1. Region 4 is
+the extended MHC region that includes two SNPs identified by the GEMMA
+analysis in genes TCF19 and VARS1. In the map, the four populations
+included in panels 1-9 are color-coded by population type, and dashed
+lines delineate the genetic clusters as defined in
+Figure~\ref{fig-selectionresults}A.}
+
+\end{figure}\clearpage
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.60\textwidth]{figures/violin_plot_heterozy_by_group.png}
 
@@
 their corresponding site number (from
-Fig. 5).}
+Figure 5).}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_genomewide_pi_by_pop.png}
+{\centering \includegraphics[width=0.70\textwidth]{figures/boxplot_genomewide_pi_by_pop.png}
 
@@
 
-\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each study
-population in the frog evolution study (from
-Fig. 5), boxplots showing nucleotide
-diversity (\(\pi\)) within 100kb sliding windows along the genome. The
-y-axis has been log\textsubscript{10}-transformed for display purposes.
-Sites are color-coded as ``Bd-naive'' or ``recovering'', and vertical
-dashed lines show the genetic breaks between frog populations as
-described in Figure~\ref{fig-selectionresults} A).}
+\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each population in
+the frog evolution study (Figure 5), boxplots
+showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
+along the genome. The y-axis has been log\textsubscript{10}-transformed
+for display purposes. Sites are color-coded as ``Bd-naive'' or
+``recovering'', and vertical dashed lines show the genetic breaks
+between frog populations as described in
+Figure~\ref{fig-selectionresults}A).}
 
@@
 
-{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_windowpop.png}
+{\centering \includegraphics[width=0.70\textwidth]{figures/boxplot_pi_by_regionpop.png}
 
@@
 
-\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
-windows identified by the overlapping splined windows for
-\emph{F\textsubscript{ST}} and \(\pi\), boxplots showing average \(\pi\)
-for naive and recovering populations. The corresponding gene annotations
-for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
+\caption{\label{fig-boxplot-pi-by-regionpop}For each of the 9 shared
+outlier regions identified by the splined window analysis, boxplots
+showing \(\pi\) values for naive and recovering populations. The
+corresponding gene annotations for each region are as follows (gene
+regions described in the \textbf{Results} and \textbf{Discussion} are
+indicated with an asterisk): (1) C6*, C7*, (2) DDX10, ZBTB24, (3)
 SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
-TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
-(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
-BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
-
-\end{figure}\clearpage
-
-\newpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\textwidth]{figures/ave_pi_per_site_yose.png}
-
-}
-
-\caption{For each study population located in Yosemite National Park
-(populations 1-4 shown in Site Map), average \(\pi\) for the 9 outlier
-windows (panels A-I) identified in the splined window analysis.
-Populations within Yosemite are the focus because they are the most
-relevant to the translocated populations described in the paper: donor
-populations 1 and 4 were the source of frogs for 20 of the 24
-translocations and the donor population for the remaining four
-translocations is located adjacent to populations 3 and 4. Bars are
-color-coded by population category (naive, recovering), and gene
-annotations within each window are listed above each panel. Gene regions
-described in the paper are in panel A (Complement C6,C7), panel E
-(GVINP1), and panel F (extended MHC region that includes two SNPs
-identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by
-an *). In the Site Map, the four populations included in panels A-I are
-color-coded by population category, and dashed lines delineate the
-genetic clusters as defined in Figure~\ref{fig-selectionresults}.}
+TCF19*, HSP70, LSM2, VARS1*, (5) GCC2, CFAP251, PEG10, (6) ERO1A,
+GVINP1*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22,
+PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
 

### 2024-04-21 doc/manuscript/convert_qmd_to_pnas_latex.py

@@ -25,8 +25,8 @@ with open(args[1], 'r') as fin:
 latex_manuscript = "".join(latex_manuscript)
 
 # Remove all the references to width and height
-latex_manuscript = re.sub("includegraphics\[.*\]", 
-						  "includegraphics", 
+latex_manuscript = re.sub("includegraphics\[.*\]",
+						  "includegraphics",
 						  latex_manuscript)
 
 # Drop the References place holder
@@ -156,11 +156,26 @@ pnas_ms = pnas_ms.replace("\\includegraphics", "\\includegraphics[width=\\textwi
 
 # Set specific figure sizes 
 
-pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/translocation_survival_bysiteid.png}",
-						  "\\includegraphics[width=0.5\\textwidth]{figures/translocation_survival_bysiteid.png}")
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/conceptual_model.png}",
+					  "\\includegraphics[width=0.95\\textwidth]{figures/conceptual_model.png}")
 
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/translocation_survival_bysiteid.png}",
+					  "\\includegraphics[width=0.60\\textwidth]{figures/translocation_survival_bysiteid.png}")
+
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/cond_effects_plot.png}",
+					  "\\includegraphics[width=0.80\\textwidth]{figures/cond_effects_plot.png}")
+
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}",
+					  "\\includegraphics[width=0.95\\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}")
+
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/allele_maps.png}",
+					  "\\includegraphics[width=0.95\\textwidth]{figures/allele_maps.png}")
+
+pnas_ms = pnas_ms.replace("\\includegraphics[width=\\textwidth]{figures/splinewindow_manhattan.png}",
+					  "\\includegraphics[width=0.95\\textwidth]{figures/splinewindow_manhattan.png}")
 
[…]

### 2024-05-19 doc/manuscript/supplementary_information.qmd
Revise manuscript for Nature Communications

--- 2024-04-21 doc/manuscript/translocation.qmd
@@
 ---
-#format: docx
-format:
-   latex: 
-      cite-method: natbib
+format: pdf
+#format:
+#   latex: 
+#      cite-method: natbib
 include-in-header:
@@
     wrap: sentence
+crossref: 
+  fig-title: "**Supplementary Fig.**"
+  tbl-title: "**Supplementary Table**"
+  title-delim: "|"
 bibliography: translocation.bib
-csl: pnas.csl
+csl: nature.csl
 ---
 
-## Title: Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
+### Supplementary Information for
 
-### Authors:
+## Reintroduction of resistant frogs facilitates recovery in the presence of a lethal fungal disease
 
-Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e,h^, Robert L. Grasso^i^, Erica Bree Rosenblum^d,e^
+Roland A. Knapp\*, Mark Q. Wilber, Maxwell B. Joseph, Thomas C. Smith, Robert L. Grasso
 
-### Affiliations:
-
-^a^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
-
-^b^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
-
-^c^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
-
-^d^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
-
-^e^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
-
-^f^Earth Lab, University of Colorado, Boulder, CO, 80303
-
-^g^Current affiliation: Planet, San Francisco, CA, 94107
-
-^h^Current affiliation: Ginkgo Bioworks, Boston, MA, 02210
-
-^i^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
-
-\*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu
-
-^1^These authors contributed equally to this work.
-
-**Classification:** Biological Sciences - Ecology
-
-**Key words:** amphibians, disease, evolution, reintroduction, population recovery
-
-**Author contributions**: All authors designed research; R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., A.P.R., and R.L.G.
-performed research; M.B.J.
-contributed new analytic tools; R.A.K., M.Q.W., A.Q.B., A.P.R., and E.B.R.
-analyzed data; and R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., and E.B.R.
-wrote paper.
-
-**Competing interests:** This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
-Briggs; DEB-1557190, to C.
-Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
-Richards-Zawacki).
-The authors declare no additional competing interests.
-
-### Abstract
-
-Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
-The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
-Lacking any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
-The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
-Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
-Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes.
-We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
-In addition, the likelihood of establishment is influenced by site, population, and frog attributes.
-Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
-To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
-We found numerous differences between these categories, including in immune function loci associated with disease resistance, consistent with evolutionary changes in response to Bd exposure.
-These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
-
-### Significance
-
-Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-We used reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a novel disease that has decimated amphibians worldwide.
-Reintroductions allowed establishment of viable populations in the presence of disease.
-In addition, resistance may be at least partially the result of natural selection at specific immune function genes.
-The evolution of resistance and reintroduction of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
-
-### Introduction
-
-Human activities are increasingly impacting global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
-One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
-Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
-Amphibians are experiencing particularly devastating impacts of disease due to the recent emergence and global spread of the highly virulent amphibian chytrid fungus, *Batrachochytrium dendrobatidis* (Bd) [@luedtke2023; @scheele2019].
-By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
-These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding mechanisms of species persistence in the presence of emerging diseases [@russell2020].
-
-Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggesting that disease-caused declines will be difficult to reverse.
-This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
-Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@raberg2009] and facilitate population persistence and recovery [@brannelly2021].
-Host immunity and evolution both play important roles in the development of resistance and tolerance, and an enhanced understanding of these processes is important for developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
-However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
-These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
-If true, the future of many amphibian species threatened by Bd appears bleak.
-
-The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
-Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
-Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
-Endangered Species Act.
-In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
-The arrival of Bd in the mid-1900s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
-These Bd-caused declines are fundamentally different from the fish-caused declines because fish eradication is feasible [@knapp1998] and results in the rapid recovery of frog populations [@knapp2007; @vredenburg2004].
-In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.
-
-Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
-This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection and not simply an effect of factors external to individual frogs (e.g., environmental conditions).
-In addition to frogs from recovering populations having higher resistance to Bd infection than those from naive populations, they could also have higher tolerance, but no data are currently available to evaluate this possibility.
-Therefore, we focus on resistance throughout this paper.
-The observed resistance of MYL frogs could be the result of several mechanisms that are not mutually exclusive, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
-The possible evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
-This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
-
-In the current study, we had three primary objectives.
-First, determine whether the reintroduction of resistant MYL frogs, obtained from populations recovering from Bd-caused declines, allows the reestablishment of populations despite ongoing disease (@fig-recovery-model).
-We accomplished this using a 15-year large-scale reintroduction study.
-Second, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
-We did this using a stage-structured matrix model.
-Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show genomic patterns consistent with selection for resistance (@fig-recovery-model).
-We achieved this with a comparative study using exome capture methods applied to naive and recovering populations.
-
-### Results
-
-#### Frog population recovery
-
-```{r}
-#| include: false
-#| code-fold: true
-
-library(readr)
-library(dplyr)
-library(magrittr)
-library(stringr)
-library(lubridate)
-d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
-d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
-d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))
-
-transno_min <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-transno_max <- d1 %>% 
-  distinct(site_id, date) %>% 
-  group_by(site_id) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_max <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(max(n))
-
-transfrogs_min <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(min(n))
-
-transfrogs_median <- d1 %>% 
-  group_by(site_id, date) %>% 
-  count() %>% 
-  ungroup() %>% 
-  summarize(median(n))
-
-bdload_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(min(bdload_ave))
-
-bdload_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  group_by(site_id) %>% 
-  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
-  ungroup() %>% 
-  summarize(max(bdload_ave))
-
-bdprev_max <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_max(prev, n = 1, with_ties = FALSE)
-
-bdprev_min <- d1 %>% 
-  filter(!is.na(bd_load)) %>% 
-  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
-  group_by(site_id) %>% 
-  summarize(prev = sum(bdload01)/n()) %>% 
-  ungroup() %>% select(prev) %>% 
-  slice_min(prev, n = 1, with_ties = FALSE)
-
-files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
-surveys <- read_csv(files, id = "path") %>% 
-  filter(survey_date < "2023-01-01") %>% 
-  select(-path)
-rm(files)
-
-surveys_yr_median <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(median(n))
-
-surveys_yr_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(max(n))
-
-surveys_yr_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
-  group_by(site_id) %>% 
-  count(yr) %>% ungroup() %>% 
-  summarize(min(n))
-
-robust_year_median <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(median(max_2period))
-
-robust_year_min <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(min(max_2period))
-
-robust_year_max <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr" & secondary_period > 1) %>% 
-  group_by(site_id) %>% 
-  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
-  summarize(max(max_2period))
-
-surveyyrs <- surveys %>% 
-  mutate(yr = year(ymd(survey_date))) %>%
-  filter(survey_type == "cmr") %>% 
-  distinct(site_id, yr) %>% 
-  count(site_id)
-
-surveydates <- surveys %>% 
-  filter(survey_type == "cmr") %>% 
-  mutate(moday = format(as.Date(survey_date), "%b-%d"))
-
-translocationdates <- surveys %>% 
-  filter(survey_type == "translocation") %>% 
-  mutate(moday = format(as.Date(survey_date), "%b-%d"))
-
-frog_counts_gt0_nrow <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  nrow()
-
-frog_counts_gt0_median <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(median(pyears_tadsub_present)) %>% 
-  round(2)
-
-frog_counts_gt0_min <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(min(pyears_tadsub_present)) %>% 
-  round(2)
-
-frog_counts_gt0_max <- d2 %>% 
-  filter(pyears_tadsub_present > 0) %>% 
-  summarize(max(pyears_tadsub_present))
-
-frog_captures_gt0_nrow <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  nrow()
-
-frog_captures_gt0_median <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(median(pyears_recruits_present)) %>% 
-  round(2)
-
-frog_captures_gt0_min <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(min(pyears_recruits_present)) %>% 
-  round(2)
-
-frog_captures_gt0_max <- d3 %>% 
-  filter(pyears_recruits_present > 0) %>% 
-  summarize(max(pyears_recruits_present))
-```
-
-To determine whether MYL frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
-These 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
-Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
-Across all translocation sites, the duration of survey time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
-
-Of the `r nrow(d2)` reintroduced populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
-For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
-These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
-As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
-In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
-
-Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
-Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
-
-The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
-Given that it can take years or even decades to determine the self-sustainability of a reintroduced population [for an example in MYL frogs, see @joseph2018], the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
-Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
-These estimates were made using site-specific models analyzed using the mrmr package.
-We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in an among-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).
-
-Estimates of 1-year frog survival indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
-These patterns indicate an important effect of site characteristics on frog survival.
-In addition, 1-year survival was higher for frogs translocated later in the study period than earlier: 5 of the 7 populations translocated after 2013 had estimated survival $\ge$ 0.5, compared to only 1 of 5 populations translocated prior to 2013.
-We suggest this resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
-This increased survival has direct implications for population viability (see **Results - Long-term population viability**).
-
-The goal of our meta-analysis was to identify important predictors of 1-year frog survival.
-We were particularly interested in whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
-This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual-level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
-Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, @fig-survival-postdens SI).
-Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
-The absence of Bd load as an important predictor of frog survival is consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
-
-In summary, results from our frog translocation study indicate that (i) translocations produced relatively high 1-year survival of translocated adults, as well as reproduction and recruitment, at the majority of recipient sites, (ii) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (iii) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
-Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
-This suggests that frog resistance is maintained across generations.
-All of the conditions described above are supportive of population establishment and long-term population growth.
-
-#### Long-term population viability
-
-Results from the frog translocation study are suggestive of population establishment.
-However, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
-To extend our inferences of population establishment beyond those possible from the site-specific CMR data, we developed a population viability model.
-Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
-
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of @fig-viability B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
-These six populations all had observed yearly adult survival greater than 0.5.
-As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
-
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 (@fig-viability B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
-In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
-
-In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-
-#### Frog genome evolution
-
-Results from the preceding sections indicate the critical role of intrinsic factors (such as frog resistance) in post-translocation frog survival, population growth, and population viability.
-As such, identifying the mechanisms underlying resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
-Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs the role of disease-mediated selection in these processes remains unknown.
-
-To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of a genome) between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-Bd-exposure histories of the 9 study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
-In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 500), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
-Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-There is no potential for frog dispersal between the 9 study populations or any other populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-
-It is important to note that our genomic analysis is necessarily limited by the small number of naive populations remaining on the landscape (nearly all of which we sampled as part of the current study) and their clumped distribution ([@fig-allelefrequencies]).
-These characteristics of naive populations precluded a replicated analysis among pairs of geographically proximate naive and recovering populations.
-Instead, to obtain adequate power, we conducted a pooled analysis across population types to identify genomic regions showing patterns consistent with selection.
-Our comparative approach cannot provide a direct assessment of the strength or timing of natural selection, but does allow us to identify specific regions of the genome that merit further study for their potential contribution to frog resistance in recovering populations.
-(See **Supporting Information - Frog genome evolution - Study design** for additional details.)
-
-As a first step in our pooled analysis, we conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations.
-We then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions possibly under selection).
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A SI), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation-by-distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations show patterns consistent with selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
-One of the 7 identified genes (LOC108802036) does not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 0.41 to 0.86.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of parallel selection at this locus across multiple populations.
-The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found statistically significant differences in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions.
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, suggestive of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
-This region (854kb from the beginning of the outlier region to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-To provide details specifically on possible selection in the Yosemite populations described in **Results - Frog population recovery**, $\pi$ values for the 9 shared outlier regions are shown in @fig-ave-pi-per-popn-yose SI for each of the four Yosemite populations; two of these populations provided frogs for 20 of the 24 frog translocations.
-
-Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog genomic evolution - Genetic diversity** for details).
-Thus, despite patterns consistent with localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
-In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog genomic evolution** **- GO analysis** for details).
-
-In summary, our genomic results indicate that frogs from naive and recovering populations show differences at several immune-function loci, consistent with selection following Bd exposure.
-
-### Discussion
-
-Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
-Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent pathogen (Bd).
-Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the reintroduced populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
-Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
-In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via reintroduction of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.
-
-In the following discussion, we follow the sequence of frog recovery described in @fig-recovery-model to structure our key points.
-Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
-Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
-In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter SI].
-The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
-This resistance could be due at least in part to natural selection for more resistant genotypes, but until now evidence to evaluate the possible role of evolution was lacking.
-
-Our genomics study used a pooled design that compared MYL frogs from naive and recovering populations distributed across a similar latitudinal range.
-The study design produced frog genetic clusters that each contained at least one naive and one recovering population, allowing us to infer selection while attempting to minimize the confounding effects of population structure.
-We identified multiple specific alleles and genomic regions showing differences between naive and recovering MYL frog populations, consistent with selection following Bd exposure.
-In addition, we did not find a reduction in overall genetic variation in recovering populations, suggesting that despite possible localized selection in the genome, some of these populations likely retain adequate genetic diversity for long-term persistence.
-
-Importantly, some genomic regions that we identified as possibly under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
-For example, the MHC plays an important role in immunity.
-We identified a region containing eight genes associated with the MHC Class I and III regions that showed patterns consistent with selection in recovering populations.
-These results align with numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest pattern of possible directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
-Several previous studies of amphibians have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
-This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [i.e., gill damage and amoebic load as heritable host resistance traits, @robledo2020; @robledo2018].
-We also identified a region, characterized by high *F~ST~* and low $\pi_{diff}$, that contained the complement genes C6 and C7.
-The complement system plays an important role in innate immunity [@rieraromo2016], and our results could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, suggesting possible consistent selection in populations distributed across a wide geographic area.
-This gene is associated with immune response and in *Xenopus* is expressed during appendage regeneration [@fukazawa2009].
-Finally, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
-In the genome of another frog species, this gene is located adjacent to a type I interferon gene (Np-IFNi2) [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
-Collectively, the genes associated with these genomic differences may confer at least some degree of resistance against Bd infection, an attribute that may be critically important to population reestablishment and recovery in the presence of Bd.
-
-Reintroduction of resistant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
-Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
-Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
-The fact that the relatively low Bd loads and correspondingly high frog survival was maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
-Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and reintroduced populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the reintroduced versus donor populations and pathogen pressure should follow a similar pattern.
-In addition to the maintenance of Bd load and frog survival between natal and recipient sites, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
-
-Results from the population viability model were also encouraging.
-In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year-1 juvenile survival probability was greater than 0.10).
-The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the years covered by our post-translocation surveys.
-These long-term forecasts are important, given that reintroduced MYL frog populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
-Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is crucial for adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
-Specifically, the combined results from our reintroduction study and viability modeling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
-In addition, given the repeatability of frog survival at a site, 1-year frog survival also serves as an effective proxy of site quality (i.e., the ability of a site to support high frog survival and a viable frog population over the long term).
-This proxy of site quality is important in the MYL frog system because accurately predicting the ability of a site to support a viable frog population *a priori* remains difficult, even after conducting 24 translocations over 16 years.
-
-The results of the meta-analysis indicate the influence of several variables on adult survival, and by extension on population viability.
-Although we assume that the effects of winter severity and frog sex are direct effects of the variables themselves, the elevation and donor population variables are likely serving as proxies for a range of factors.
-For example, elevation is correlated with several attributes that may influence frog survival, including the presence of boulder habitat (positive correlation) that is important for frog overwintering, abundance of frog predators including garter snakes (negative correlation), and frog developmental rates (negative correlation).
-The latter can influence the duration of frog life stages that have differential susceptibility to chytridiomycosis (see below).
-The duration of the tadpole and juvenile stages increases with elevation, but the fact that long-run population growth rate from our viability model is relatively insensitive to variation in these durations suggests that developmental rates of early frog life stages have minimal effect on the predicted long-term viability of reintroduced populations.
-Nonetheless, a clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
-
-Despite the demonstrated resistance of adult MYL frogs against Bd infection, individual and population-level impacts of Bd are still evident.
-In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
-Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
-In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
-MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality [@rachowicz2006].
-This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
-In both naturally recovering and reintroduced MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
-Therefore, although adult MYL frogs appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
-
-The recent emergence of Bd worldwide has contributed to the decline of hundreds of amphibian species, some of which are now extinct in the wild [@scheele2019].
-This extraordinary impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
-Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other amphibian species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
-As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
-Despite the hopeful example of successful reestablishment of MYL frogs despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
-Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
-The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
-
-In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
-We provide an example from the wild that suggests that evolution might produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
-We expect that similar species recovery actions will be an essential tool in wildlife conservation in an era of accelerating global change.
-
-### Materials and Methods
-
-#### Frog population recovery
-
-##### Field methods
-
-For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
-The populations that we selected contained hundreds of *R. sierrae* adults and thousands of tadpoles.
-These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
-Recipient sites to which frogs were translocated were chosen based on previous *R.
-sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
-At the beginning of this study, we had a relatively limited understanding of the factors that affect habitat quality.
-In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwintering habitats such as submerged boulders and overhanging banks.
-*R. sierrae* were absent from all recipient sites prior to the first translocation.
-
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap SI) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
-In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
-Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, we visited translocated populations approximately once per month during the summer active season and conducted diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
-CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size, and VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we searched all habitats repeatedly for adult frogs.
-Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
-During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
-However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
-When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.
-
-During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
-VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
-These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
-
-##### Frog counts and reproductive success
-
-For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
-To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
-This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not occur until the following year.
-Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
-This accounted for the multi-year tadpole and juvenile stages in MYL frogs (@tbl-param_values SI).
-
-##### Estimation of frog survival and abundance
-
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022] (see **Supporting Information - Frog population recovery - CMR model structure** for details).
-Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
-The possible states of individuals include "not recruited", "alive", and "dead".
-The possible observations of individuals include "detected" and "not detected".
-We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
-We also assume that new recruits were the result of within-site reproduction and not immigration from adjacent populations.
-This assumption is justified by the fact that no *R. sierrae* populations were present within several kilometers of the translocation sites.
-For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
-We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
-
-##### Predictors of post-translocation frog survival
-
-To identify important predictors of frog survival following translocation, we used multilevel Bayesian models [@gelman2013; @gabry2019].
-Included predictor variables describe characteristics of sites, translocated cohorts, and individuals (Bd load, sex, frog size, site elevation, winter severity in the year of translocation, winter severity in the year following translocation, donor population, day of year on which a translocation was conducted, and translocation order).
-We used 1-year post-translocation survival estimates from CMR models as the response.
-Estimated survival was rounded to integer values to produce a binary outcome, and modeled with a Bernoulli distribution.
-Group-level (random) effects included site_id, translocation_id, or translocation_id nested within site_id.
-We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022].
-For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
-We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out cross-validation [@vehtari2016], as implemented by the loo package [@vehtari2022].
-(See **Supporting Information - Frog population recovery - Among-site survival modeling** for details.)
-
-##### Changes in Bd load following translocation
-
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see **Supporting Information - Frog population recovery - Laboratory methods** for details].
-To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
-Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
-After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
-Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the 1-year period following translocation.
-
-Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.
-
-#### Population viability modeling
-
-##### Model description
-
-To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for MYL frogs.
-Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] SI).
-
-We modeled the dynamics of these seven state variables using a discrete-time, stage-structured model where a time step is one year.
-The dynamics are given by
-
-$$
-\begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\ 
-A_T
-\end{bmatrix}(t + 1) = 
-\begin{bmatrix}
-  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
-  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
-  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
-  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
-  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
-  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
-\end{bmatrix} \begin{bmatrix}
-L_1 \\
-L_2 \\
-L_3 \\
-J_1 \\
-J_2 \\
-A_R \\
-A_T \\
-\end{bmatrix}(t) 
-$$
-
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
-First-year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced by environmental stochasticity.
-
-In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
-We also did not directly consider the dynamics of Bd in this model.
-We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
-Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-Additional details are provided in **Supporting Information - Population viability modeling - Incorporating yearly variability in survival rates** and **Estimating model parameters**.
-
-##### Model analysis
-
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
-First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp SI).
-This also helped us determine where in the model environmental variation in juvenile survival and recruitment would have the largest effects on population dynamics.
-Third, we included demographic stochasticity and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
-Additional details are provided in **Supporting Information - Population viability modeling - Model analysis and simulation**.
-Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
-
-#### Frog genome evolution
-
-##### Sampling and sequencing
-
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
-These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
-Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
-DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in @byrne2023.
-Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
-Raw sequencing reads are available from NCBI SRA (PRJNA870451).
-
-##### Data pre-processing and cleaning
-
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *R.
-muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
-Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
-This initial filter resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant level.
-First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x (n = 3).
-Our final dataset included 50 samples, 23 from naive populations and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-##### Data analysis
-
-To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome that differed between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and **Supporting Information - Frog evolution in response to Bd - GO analysis;** Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
-We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
-After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome (for more efficient data processing and to better contextualize the strength of the outlier signals in different regions of the genome).
-We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the spline fitted to *F~ST~* and $\pi_{diff}$ values along the genome, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-These standard deviations represent strict criteria to select only the top approximately 0.3% of windows.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (i) high differentiation between naive and recovering populations, and (ii) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (**Supporting Information**: Datasets S3, S4).
-
-Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
-
-### Acknowledgments
-
-We thank the following for important contributions to this study: E.
-Hegeman and A.
-Lindauer (project and data management, field and laboratory work); A.
-Barbella and K.
-Rose (laboratory work); numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
-Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
-We also thank the editor and three anonymous reviewers whose input helped us improve the manuscript.
-This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
-Briggs; DEB-1557190, to C.
-Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
-Richards-Zawacki).
-
-### References
-
-### Figures
-
-![For MYL frogs, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
-Rectangles and hexagons represent outcomes and processes, respectively.
-Blue text indicates components that are included in the current study.
-We quantified frog survival and population viability following reintroductions directly using field studies and models, and evaluated "disease-mediated selection" indirectly using a comparison of frog exomes in naive and recovering populations.
-The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
+\* **Author for correspondence:** roland.knapp\@ucsb.edu
 
@@
 
-![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated for each site from the mrmr CMR model.
-Error bars show the 95% uncertainty intervals.
-Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site.
-Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="60%"}
+## Supplementary Methods
 
-\newpage
-
-![Results from the among-site meta-analysis showing conditional effects of the important predictors of 1-year frog survival (expressed as a probability): (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
-In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
-
-\newpage
-
-![Results from population viability analysis: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
-The red line shows where $\lambda = 1$.
-Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
-
-\newpage
-
-![Genomic patterns consistent with selection on individual variants in recovering MYL frog populations at the landscape scale (based on a linear mixed model analysis).
-For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
-The inset map locates the study area and range boundary in California.
-Sites 1 and 4 (site id = 72996 and 70567, respectively) were also used as sources of frogs for reintroductions described in **Results - Frog population recovery**.](figures/allele_maps.png){#fig-allelefrequencies width="95%"}
-
-\newpage
-
-![Results from the splined window analysis showing patterns consistent with selection on genomic regions in recovering MYL frog populations.
-Note that several key candidate genes (RIN3, TCF19, VARS) are outliers in both the individual variant analysis (@fig-allelefrequencies) and splined window analysis (this figure).
-Manhattan plots show outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
-In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
-In (B), outlier regions are shown above the single dashed red line.
-Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="95%"}
-
-\newpage
-
-## Supporting Information
+\
 
@@
 \end{block}
-\end{blockarray},
+\end{blockarray}
 $$ where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
@@
 \end{block}
-\end{blockarray},
-$$ where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
+\end{blockarray}
+$$
+
+where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
 
@@
 
-### Frog genome evolution
+### References
 
-#### Study design
-
-To gain insights into the role of evolution in the development of resistance by MYL frogs, we compared frog exomes sampled in naive versus recovering populations.
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected before and after Bd exposure are not available and are unlikely to be available in the future.
-
-Our study design, in which we compared frog genomes in naive and recovering populations, required sampling populations across the range of MYL frogs in the southern Sierra Nevada.
-This is due to the fact that very few MYL frog populations remain in the naive state, and those that do are scattered across a wide latitudinal range, from Yosemite National Park in the north to southern Kings Canyon National Park in south.
-To minimize potential confounding effects caused by known variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges (@fig-allelefrequencies).
-
-#### GO analysis
-
-In **Results-Frog genome evolution**, we describe the stringent set of outlier variants (identified using a Bonferroni-corrected p-value of 0.01).
-A liberal set of outlier variants, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-We used this liberal set to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-To do this, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-We then conducted a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] to compare the 30 unique outlier genes to the complete set of genes in our target transcriptome.
-We repeated this process for the set of 35 genes located in the 9 shared regions of the *F~ST~* and $\pi_{diff}$ splined windows.
-
-#### Genetic diversity
-
-To characterize general patterns of genetic diversity between naive and recovering populations, we conducted three analyses.
-First, we calculated heterozygosity for each sampled frog using VCFtools [@danecek2011].
-Second, to characterize genome-wide patterns of diversity, we used VCFtools to calculate nucleotide diversity ($\pi$) in 100kb sliding windows along the genome for each population.
-Third, we calculated average $\pi$ per population within each of the 9 shared outlier regions identified in the splined window analysis.
-We compared average $\pi$ per population between naive and recovering population types, and also for each naive and recovering population in Yosemite.
-The latter comparison focused on Yosemite because these populations are the most relevant to the translocated populations described in the paper.
-
-Average individual-level heterozygosity and genome-wide population-level $\pi$ were similar between the naive and recovering groups (@fig-violinplot-heterozy SI; @fig-boxplot-genomewide-pi-by-pop SI).
-Within each of the 9 shared outlier regions, $\pi$ shows considerable variation between populations and no obvious patterns between the naive and recovering populations (@fig-boxplot-pi-by-regionpop SI).
+::: {#refs}
+:::
 
@@
 
-\newpage
-
-![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon).
-Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="60%"}
+![**Bd loads on frogs before and after translocation.** The "before" period is immediately prior to translocation and the "after" period is the 1-year period after translocation.
+Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
+Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
+Samples sizes are provided immediately above the x-axis.
+The red dashed horizontal line indicates the approximate load above which severe disease is typically observed @vredenburg2010.
+This threshold (5.8 ITS1 copies - on a log10 scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations @briggs2010.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
 
@@
 
-![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Samples sizes are provided immediately above the x-axis.
-The red dashed horizontal line indicates the approximate load above which severe disease is typically observed [@vredenburg2010].
-This threshold (5.8 ITS1 copies - on a log~10~ scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations [@briggs2010].](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
+![**Predictors of post-translocation frog survival in the among-site meta-analysis.** Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model.
+In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
+Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.
+Bd load is not an important predictor of post-translocation frog survival.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
 
@@
 
-![Results from the among-site meta-analysis, showing that Bd load is not an important predictor of post-translocation frog survival.
-Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
-Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
+![**Average yearly survival probabilities of translocated versus naturally recruited adults.** Naturally recruited individualos were the result of reproduction by translocated frogs at each site.
+In contrast to \@fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
+Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
+All estimates were derived using the mrmr package, and the methods for calculation are described in Supporting Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="60%"}
 
@@
 
-![Comparison of average yearly adult survival probabilities for adults translocated to each of 3 sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
-Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling - Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="60%"}
-
-\newpage
-
-![Sensitivity analysis of the stage-structured MYL frog model.
-A.
-Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+![**Sensitivity analysis of the stage-structured *R.
+sierrae* model.** A.
+Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
 B.
@@
 Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
-
-\newpage
-
-![Results from the analysis of individual variants, showing possible signatures of selection in recovering MYL frog populations.
-(A) PCA calculated from binary SNPs showing the genomic relationship of samples.
-Numeric labels and colors match those in @fig-allelefrequencies.
-Populations 1-7 are *R.
-sierrae* and populations 8 and 9 are *R.
-muscosa*.
-(B) qqplot showing observed and expected p-values for 148,307 SNPs and INDELS.
-Dashed line shows p-value that identifies outliers.
-(C) Manhattan plot showing p-value for each SNP.
-SNPs are sorted by genomic position and contigs are sorted by size.
-Red line shows p-value that identifies outliers.
-Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="85%"}
-
-\newpage
-
-![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana muscosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
-The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana muscosa* Contig19.
-Asterisks indicate the location of SNP outliers in the TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="95%"}
-
-\newpage
-
-![For each study population located in Yosemite National Park (populations 1-4 shown in the map), $\pi$ for the 9 shared outlier regions identified in the splined window analysis.
-Populations 1 and 4 were the source of frogs for 20 of the 24 frog translocations, and the donor population for the remaining four translocations is located adjacent to populations 3 and 4 (see also @fig-yosemap).
-Bars are color-coded by population type (naive, recovering).
-The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.
-Region 4 is the extended MHC region that includes two SNPs identified by the GEMMA analysis in genes TCF19 and VARS1.
-In the map, the four populations included in panels 1-9 are color-coded by population type, and dashed lines delineate the genetic clusters as defined in [@fig-selectionresults]A.](figures/ave_pi_per_popn_yose.png){#fig-ave-pi-per-popn-yose width="85%"}
-
-\newpage
-
-![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog evolution study.
-Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="60%"}
-
-\newpage
-
-![For each population in the frog evolution study (@fig-allelefrequencies), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
-The y-axis has been log~10~-transformed for display purposes.
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in [@fig-selectionresults]A).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="70%"}
-
-\newpage
-
-![For each of the 9 shared outlier regions identified by the splined window analysis, boxplots showing $\pi$ values for naive and recovering populations.
-The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_regionpop.png){#fig-boxplot-pi-by-regionpop width="70%"}
-
-\newpage
 
@@
 
-: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
-
-\newpage
-
-### Datasets
-
-Dataset S1.
-Set of liberal SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.05), identified via GEMMA.
-file: gemma_outliers_all.csv
-
-Dataset S2.
-Set of strict SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.01), identified via GEMMA.
-Additional information includes variant location within the gene (predicted_gene_loc) and whether the variant is synonymous or nonsynonymous (predicted_effect_AA).
-file: gemma_outliers_strict_freq.csv
-
-Dataset S3.
-Description of overlapping *F~ST~* and $\pi_{diff}$ outlier windows, as identified in the splined window analysis.
-The outlier window on chromosome 19 (column name = chr_num_sorted) is included twice because one *F~ST~* outlier window overlapped with two $\pi_{diff}$ outlier windows.
-file: spline_window_shared_outliers.csv
-
-Dataset S4.
-Annotation information for all genes within each of the overlapping outlier windows in dataset S3.
-file: spline_window_gene_details
-
-### References
+: **Description and values of parameters used in the population viability model.** <!--# MOVE NEXT SENTENCE TO FOOTNOTE.
+--> All survival probabilities are in the presence of the fungal pathogen Bd.
+{#tbl-param_values}

### 2024-05-19 doc/manuscript/translocation.qmd
Revise manuscript for Nature Communications

--- 2024-04-21 doc/manuscript/translocation.qmd
@@
     wrap: sentence
+crossref: 
+  fig-title: "**Fig.**"
+  title-delim: "|"
 bibliography: translocation.bib
-csl: pnas.csl
+csl: nature.csl
+link-bibliography: false
 ---
 
-## Title: Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
+## Reintroduction of resistant frogs facilitates recovery in the presence of a lethal fungal disease
 
-### Authors:
+Roland A.
+Knapp^1,2,^\*, Mark Q.
+Wilber^3^, <!--#Allison Q.
+Byrned,e,1-->, Maxwell B.
+Joseph^4,5^, Thomas C.
+Smith^1,2^ <!--#Andrew P.
+Rothsteind,e,h--> & Robert L.
+Grasso^6^, <!--#Erica Bree Rosenblumd,e-->
 
-Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e,h^, Robert L. Grasso^i^, Erica Bree Rosenblum^d,e^
+^1^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
 
-### Affiliations:
+^2^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
 
-^a^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546
+^3^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
 
-^b^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060
+<!--#dDepartment of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114-->
 
-^c^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996
+<!--#eMuseum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160 -->
 
-^d^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114
+^4^Earth Lab, University of Colorado, Boulder, CO, 80303
 
-^e^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160
+^5^Current affiliation: Planet, San Francisco, CA, 94107
 
-^f^Earth Lab, University of Colorado, Boulder, CO, 80303
+<!--#Current affiliation: Ginkgo Bioworks, Boston, MA, 02210-->
 
-^g^Current affiliation: Planet, San Francisco, CA, 94107
-
-^h^Current affiliation: Ginkgo Bioworks, Boston, MA, 02210
-
-^i^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
+^6^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
 
@@
 
-^1^These authors contributed equally to this work.
-
-**Classification:** Biological Sciences - Ecology
-
-**Key words:** amphibians, disease, evolution, reintroduction, population recovery
-
-**Author contributions**: All authors designed research; R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., A.P.R., and R.L.G.
-performed research; M.B.J.
-contributed new analytic tools; R.A.K., M.Q.W., A.Q.B., A.P.R., and E.B.R.
-analyzed data; and R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., and E.B.R.
-wrote paper.
-
-**Competing interests:** This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
-Briggs; DEB-1557190, to C.
-Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C.
-Richards-Zawacki).
-The authors declare no additional competing interests.
+<!--#1These authors contributed equally to this work.-->
 
@@
 Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
-The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
+The emergence of the lethal fungal pathogen *Batrachochytrium dendrobatidis* ("Bd") has devastated global amphibian biodiversity.
 Lacking any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
-The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
-Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
-Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes.
-We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
+The Sierra Nevada yellow-legged frog (*Rana sierrae*) is highly susceptible to Bd infection.
+Although most *R.
+sierrae* populations are extirpated following disease outbreaks, some persist and eventually recover and frogs in these recovering populations have increased resistance against infection.
+We conducted a 15-year reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
 In addition, the likelihood of establishment is influenced by site, population, and frog attributes.
 Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
-To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
-We found numerous differences between these categories, including in immune function loci associated with disease resistance, consistent with evolutionary changes in response to Bd exposure.
-These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.
-
-### Significance
-
-Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
-We used reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a novel disease that has decimated amphibians worldwide.
-Reintroductions allowed establishment of viable populations in the presence of disease.
-In addition, resistance may be at least partially the result of natural selection at specific immune function genes.
-The evolution of resistance and reintroduction of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.
+These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for amphibians and many other taxa that are threatened with extinction by novel pathogens.
 
@@
 
-Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggesting that disease-caused declines will be difficult to reverse.
+Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare (for notable exceptions, see [@scheele2017; @voyles2018; @knapp2016]), suggesting that disease-caused declines will be difficult to reverse.
 This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
@@
 Host immunity and evolution both play important roles in the development of resistance and tolerance, and an enhanced understanding of these processes is important for developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
-However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
+However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes (but see [@savage2016; @grogan2018b]).
 These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
@@
 
-The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
+The Sierra Nevada yellow-legged frog (*Rana sierrae*) [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
 Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
@@
 
-Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
+Despite the catastrophic impact of Bd on *R.
+sierrae*, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
 Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
@@
 Therefore, we focus on resistance throughout this paper.
-The observed resistance of MYL frogs could be the result of several mechanisms that are not mutually exclusive, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
-The possible evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
+The observed resistance of *R.
+sierrae* could be the result of several mechanisms that are not mutually exclusive, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
+The possible evolution of resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
 This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
@@
 In the current study, we had three primary objectives.
-First, determine whether the reintroduction of resistant MYL frogs, obtained from populations recovering from Bd-caused declines, allows the reestablishment of populations despite ongoing disease (@fig-recovery-model).
+First, determine whether the reintroduction of resistant *R.
+sierrae*, obtained from populations recovering from Bd-caused declines, allows the reestablishment of populations despite ongoing disease (@fig-recovery-model).
 We accomplished this using a 15-year large-scale reintroduction study.
-Second, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
+Second, identify important site, population, and individual-level predictors of frog survival using results from the reintroduction study.
+Third, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
 We did this using a stage-structured matrix model.
-Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show genomic patterns consistent with selection for resistance (@fig-recovery-model).
-We achieved this with a comparative study using exome capture methods applied to naive and recovering populations.
 
@@
 
-To determine whether MYL frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
-Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
-The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
+To determine whether *R.
+sierrae* from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
+Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
+The 3 donor populations are among the largest *R.
+sierrae* populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
 These 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
@@
 As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
-In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.
+In summary, survey results indicate that translocations resulted in the establishment of reproducing *R. sierrae* populations at most recipient sites despite the ongoing presence of Bd.
 
-Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
-Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.
+Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (Supplementary Fig.
+1) [@joseph2018; @vredenburg2010].
+Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; Supplementary Fig.
+1) suggests a true lack of substantial change in Bd load.
 
 The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
-Given that it can take years or even decades to determine the self-sustainability of a reintroduced population [for an example in MYL frogs, see @joseph2018], the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
+Given that it can take years or even decades to determine the self-sustainability of a reintroduced population (for an example in *R.
+sierrae*, see [@joseph2018]), the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
 Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
@@
 This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual-level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
-Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, @fig-survival-postdens SI).
-Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (Supplemental Fig. 2).
+Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, Supplemental Fig. 2).
+Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, Supplemental Fig. 2).
 The absence of Bd load as an important predictor of frog survival is consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
@@
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (Supplemental Fig.
+3).
 This suggests that frog resistance is maintained across generations.
@@
 Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
-We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
+We parameterized the model using CMR data from translocated populations and known life history values in this system (Supplementary Table 1).
 
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of @fig-viability B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).
+As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 ([@fig-viability]A).
 
 Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 (@fig-viability B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 ([@fig-viability]B).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
 In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history ([@fig-viability]C; population 70550, surveyed for 16 years).
 
 In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
-This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
-
-#### Frog genome evolution
-
-Results from the preceding sections indicate the critical role of intrinsic factors (such as frog resistance) in post-translocation frog survival, population growth, and population viability.
-As such, identifying the mechanisms underlying resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
-Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs the role of disease-mediated selection in these processes remains unknown.
-
-To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of a genome) between populations with contrasting histories of Bd exposure.
-Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
-Bd-exposure histories of the 9 study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
-Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
-In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 500), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
-Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
-There is no potential for frog dispersal between the 9 study populations or any other populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.
-
-It is important to note that our genomic analysis is necessarily limited by the small number of naive populations remaining on the landscape (nearly all of which we sampled as part of the current study) and their clumped distribution ([@fig-allelefrequencies]).
-These characteristics of naive populations precluded a replicated analysis among pairs of geographically proximate naive and recovering populations.
-Instead, to obtain adequate power, we conducted a pooled analysis across population types to identify genomic regions showing patterns consistent with selection.
-Our comparative approach cannot provide a direct assessment of the strength or timing of natural selection, but does allow us to identify specific regions of the genome that merit further study for their potential contribution to frog resistance in recovering populations.
-(See **Supporting Information - Frog genome evolution - Study design** for additional details.)
-
-As a first step in our pooled analysis, we conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations.
-We then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions possibly under selection).
-First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
-Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
-
-Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A SI), and clusters reflected the species split (i.e., *R.
-muscosa* versus *R.
-sierrae*) and the strong signature of isolation-by-distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
-Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show that recovering populations show patterns consistent with selection on multiple regions of the genome.
-The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
-One of the 7 identified genes (LOC108802036) does not have an associated annotation.
-For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 0.41 to 0.86.
-Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
-This is suggestive of parallel selection at this locus across multiple populations.
-The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found statistically significant differences in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
-Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes.
-
-The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
-Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions.
-Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
-The largest $\pi_{diff}$, suggestive of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
-This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
-Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
-Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
-Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
-This region (854kb from the beginning of the outlier region to the VARS SNP) contained a total of 8 annotated genes.
-In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
-Therefore, this region is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
-To provide details specifically on possible selection in the Yosemite populations described in **Results - Frog population recovery**, $\pi$ values for the 9 shared outlier regions are shown in @fig-ave-pi-per-popn-yose SI for each of the four Yosemite populations; two of these populations provided frogs for 20 of the 24 frog translocations.
-
-Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog genomic evolution - Genetic diversity** for details).
-Thus, despite patterns consistent with localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
-In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog genomic evolution** **- GO analysis** for details).
-
-In summary, our genomic results indicate that frogs from naive and recovering populations show differences at several immune-function loci, consistent with selection following Bd exposure.
+This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (Supplementary Fig.
+3), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
 
@@
 Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
-Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent pathogen (Bd).
-Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the reintroduced populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
-Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
-In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via reintroduction of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.
+Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered *R.
+sierrae* in the presence of a highly virulent pathogen (Bd).
+Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that post-translocation frog survival is influenced by site, population, and individual-level characteristics, and (iii) 50% of the reintroduced populations are likely to persist over a 50-year period Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
+In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing *R.
+sierrae* populations via reintroduction of resistant individuals is striking, and even more so given that *R.
+sierrae* were driven to near-extinction by Bd.
 
-In the following discussion, we follow the sequence of frog recovery described in @fig-recovery-model to structure our key points.
-Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
+Previous field studies in *R. sierrae* show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
 Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
-In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter SI].
-The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
+In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed ([@briggs2010; @knapp2011] see also Supplementary Fig.
+2).
+The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies (see Figure 4 in [@knapp2016]), and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
 This resistance could be due at least in part to natural selection for more resistant genotypes, but until now evidence to evaluate the possible role of evolution was lacking.
 
-Our genomics study used a pooled design that compared MYL frogs from naive and recovering populations distributed across a similar latitudinal range.
-The study design produced frog genetic clusters that each contained at least one naive and one recovering population, allowing us to infer selection while attempting to minimize the confounding effects of population structure.
-We identified multiple specific alleles and genomic regions showing differences between naive and recovering MYL frog populations, consistent with selection following Bd exposure.
-In addition, we did not find a reduction in overall genetic variation in recovering populations, suggesting that despite possible localized selection in the genome, some of these populations likely retain adequate genetic diversity for long-term persistence.
-
-Importantly, some genomic regions that we identified as possibly under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
-For example, the MHC plays an important role in immunity.
-We identified a region containing eight genes associated with the MHC Class I and III regions that showed patterns consistent with selection in recovering populations.
-These results align with numerous previous studies linking MHC genes to amphibian resistance against Bd [e.g., @savage2011; @bataille2015].
-Similarly, the region with the strongest pattern of possible directional selection (as measured by $\pi_{diff}$) contains the interferon-related gene GVINP1.
-Several previous studies of amphibians have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018b; @ellison2014] and in populations differing in Bd susceptibility [@grogan2018b].
-This gene is also strongly linked to disease in salmon, explaining a notable 20% of the resistance phenotype [i.e., gill damage and amoebic load as heritable host resistance traits, @robledo2020; @robledo2018].
-We also identified a region, characterized by high *F~ST~* and low $\pi_{diff}$, that contained the complement genes C6 and C7.
-The complement system plays an important role in innate immunity [@rieraromo2016], and our results could indicate that balancing selection is acting in this region of the genome to favor a diverse set of alleles, as is known for C6 in humans [@soejima2005].
-Based on the analysis of individual outlier variants, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, suggesting possible consistent selection in populations distributed across a wide geographic area.
-This gene is associated with immune response and in *Xenopus* is expressed during appendage regeneration [@fukazawa2009].
-Finally, the outlier variant with the lowest p-value was the uncharacterized gene LOC108802036.
-In the genome of another frog species, this gene is located adjacent to a type I interferon gene (Np-IFNi2) [@gan2018], and together with GVINP1 further suggests the importance of interferon-related genes in this system.
-Collectively, the genes associated with these genomic differences may confer at least some degree of resistance against Bd infection, an attribute that may be critically important to population reestablishment and recovery in the presence of Bd.
-
-Reintroduction of resistant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
+Reintroduction of resistant *R. sierrae* was remarkably successful in reestablishing viable populations in the presence of Bd.
 Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
@@
 The viability model highlighted the important role of frog survival in affecting long-term population viability, and allowed us to extend the temporal scale of our study beyond the years covered by our post-translocation surveys.
-These long-term forecasts are important, given that reintroduced MYL frog populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
-Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is crucial for adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for MYL frogs.
+These long-term forecasts are important, given that reintroduced *R. sierrae* populations may often take decades to achieve our ultimate goal of self-sustainability [@joseph2018].
+Making well-supported projections about the long-term outcome of reintroduction efforts from shorter-term information is crucial for adaptive management of species reintroduction programs [@seddon2007], including the one we are carrying out for *R.
+sierrae*.
 Specifically, the combined results from our reintroduction study and viability modeling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
 In addition, given the repeatability of frog survival at a site, 1-year frog survival also serves as an effective proxy of site quality (i.e., the ability of a site to support high frog survival and a viable frog population over the long term).
-This proxy of site quality is important in the MYL frog system because accurately predicting the ability of a site to support a viable frog population *a priori* remains difficult, even after conducting 24 translocations over 16 years.
+This proxy of site quality is important in the *R.
+sierrae* system because accurately predicting the ability of a site to support a viable frog population *a priori* remains difficult, even after conducting 24 translocations over 16 years.
 
@@
 
-Despite the demonstrated resistance of adult MYL frogs against Bd infection, individual and population-level impacts of Bd are still evident.
+<!--# Include paragraph describing implications for habitat quality on frog survival in the presence of Bd?
+This would blunt the view of PNAS editor that we put frogs in the "wrong" places.
+In the presence of Bd-caused mortality, habitats that once would have supported frog populations may no longer do so because the extra mortality changes population growth from positive to negative.
+As such, choosing the highest quality habitats may be important for maximizing the probability of population reestablishment.
+-->
+
+Despite the demonstrated resistance of adult *R. sierrae* against Bd infection, individual and population-level impacts of Bd are still evident.
 In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
@@
 In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
-MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality [@rachowicz2006].
+*R. sierrae* immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality [@rachowicz2006].
 This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
-In both naturally recovering and reintroduced MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
-Therefore, although adult MYL frogs appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
+In both naturally recovering and reintroduced *R. sierrae* populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
+Therefore, although adult *R. sierrae* appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
 
@@
 This extraordinary impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
-Importantly, in addition to the natural recovery documented for MYL frogs [@knapp2016], other amphibian species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
-As with MYL frogs, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
-Despite the hopeful example of successful reestablishment of MYL frogs despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
-Although we now have a proven strategy to reestablish extirpated MYL frog populations, recovery across their large historical range will require substantial resources over many decades.
+Importantly, in addition to the natural recovery documented for *R.
+sierrae* [@knapp2016], other amphibian species are also showing evidence of post-epizootic recovery in the presence of Bd [@scheele2017; @voyles2018] and suggest the possibility of also using animals from these recovering populations to reestablish extirpated populations.
+As with *R.
+sierrae*, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
+Despite the hopeful example of successful reestablishment of *R.
+sierrae* despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
+Although we now have a proven strategy to reestablish extirpated *R. sierrae* populations, recovery across their large historical range will require substantial resources over many decades.
 The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
 
-In our rapidly changing world, evolution is likely to play in important role in facilitating the resilience of wildlife populations.
-We provide an example from the wild that suggests that evolution might produce individuals that harbor adaptive alleles and allow population recovery in a novel (i.e., Bd-positive) environment, and show conclusively that individuals from these recovering populations can be used to reestablish extirpated populations and expand the scale of natural recovery (@fig-recovery-model).
-We expect that similar species recovery actions will be an essential tool in wildlife conservation in an era of accelerating global change.
-
-### Materials and Methods
+### Methods
 
@@
 These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
-As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
+As is typical for recovering *R.
+sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap SI) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
@@
 Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
-This accounted for the multi-year tadpole and juvenile stages in MYL frogs (@tbl-param_values SI).
+This accounted for the multi-year tadpole and juvenile stages in *R. sierrae* (Supplementary Table 1).
 
@@
 
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022] (see **Supporting Information - Frog population recovery - CMR model structure** for details).
-Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software (v4.4.4 [@rsoftware2022]) (see **Supporting Information - Frog population recovery - CMR model structure** for details).
+Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals (see [@joseph2018] for details).
 The possible states of individuals include "not recruited", "alive", and "dead".
@@
 Group-level (random) effects included site_id, translocation_id, or translocation_id nested within site_id.
-We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022].
+We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software (v4.4.4, [@rsoftware2022]).
 For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see **Supporting Information - Frog population recovery - Laboratory methods** for details].
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004] (see **Supporting Information - Frog population recovery - Laboratory methods** for details).
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 
-Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.
-
 #### Population viability modeling
@@
 
-To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for MYL frogs.
+To determine the implications of observed 1-year adult survival on the long-term viability of populations established via translocation, we developed a population model for *R. sierrae*.
 Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
 We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
-We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs ([@fig-compare_surv_probs] SI).
+We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs (Supplementary Fig. 3).
 
@@
 
-After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
+After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates (Supplementary Table 1), we performed four analyses.
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (@fig-viability-supp SI).
+Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (Supplemental Fig. 4).
 This also helped us determine where in the model environmental variation in juvenile survival and recruitment would have the largest effects on population dynamics.
 Third, we included demographic stochasticity and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
+Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of *R. sierrae* following reintroductions.
 Additional details are provided in **Supporting Information - Population viability modeling - Model analysis and simulation**.
@@
 
-#### Frog genome evolution
+### Data availability
 
-##### Sampling and sequencing
+The data used in this study are available at <https://github.com/SNARL1/translocation>.
+The CMR datasets from each reintroduction site are available at <https://github.com/SNARL1/cmr-analysis>.
 
-We collected DNA samples via buccal swabbing [@broquet2007] from 53 *Rana muscosa*/*Rana sierrae* individuals: 24 from 4 naive populations, and 29 from 5 recovering populations.
-These populations are located in the southern Sierra Nevada, from northern Yosemite National Park to northern Sequoia National Park (@fig-allelefrequencies).
-Samples were collected from 5-6 frogs per population.
-To minimize potential confounding effects caused by variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges.
-DNA was extracted following Qiagen DNEasy manufacturer's protocols.
-We sequenced the samples using an exome capture assay as described in @byrne2023.
-Briefly, genomic libraries were prepared and captured using a custom Nimblegen capture pool.
-Capture baits were designed based on the coding regions of the *R. muscosa* transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at UC Berkeley.
-Raw sequencing reads are available from NCBI SRA (PRJNA870451).
+### Code availability
 
-##### Data pre-processing and cleaning
+R code used to run our analyses is available at <https://github.com/SNARL1/translocation>.
+R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis>.
 
-Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *R.
-muscosa* genome (NCBI SRA: SRS6533475 [@hon2020]), with repetitive elements masked using bwa ("mem" mode) [@li2013].
-Exact PCR duplicates were marked using Picard.
-Variants were then called following GATK best practices (v.4.2.0.0 [@van2020]).
-Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
-Next, genotypes were jointly called using GenotypeGVCFs.
-Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
-This initial filter resulted in 1,595,206 variant sites across 53 individuals.
+### References
 
-We then further filtered our dataset at the individual and variant level.
-First, we trimmed our variants to only include those with minor allele frequency \> 0.03, a maximum depth of 250 and minimum depth of 5, a minimum genotype quality of 20, and a maximum missing proportion of 0.5.
-This filter resulted in 427,038 sites, of which 353,172 were SNPs and 73,866 were INDELS.
-Finally, we trimmed samples with an average depth across filtered sites \< 7x (n = 3).
-Our final dataset included 50 samples, 23 from naive populations and 27 from recovering populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
+::: {#refs}
+:::
 
-##### Data analysis
-
-To visualize the genomic relationships of our populations, we conducted a PCA using the glPCA function in the adegenet R package [@jombart2008].
-To detect regions of the genome that differed between naive and recovering populations, we used two approaches: (1) a multivariate linear mixed model to evaluate individual variants (SNPs and INDELs), and (2) a splined window analysis to evaluate larger genomic regions.
-For the variant analysis, we first used a stringent data filter to include only variants with \< 5% missing data (missing for no more than 2 individuals), and then calculated the likelihood ratio statistic for the resulting set of 148,307 high quality variants across 127 contigs using GEMMA [@zhou2014].
-GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics.
-We identified variants showing different allele frequencies between naive versus recovering populations ("outliers") using a Bonferroni-corrected significance level of 0.01.
-We visualized the results using a Manhattan plot and qqplot.
-We developed a more liberal set of outlier variants using a Bonferroni-corrected significance level of 0.05 and used this set solely for the gene ontology (GO) analysis (see below and **Supporting Information - Frog evolution in response to Bd - GO analysis;** Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we determined whether the variant was synonymous (protein sequence the same for each variant) or non-synonymous (protein sequence differs between variants), and where in the gene it was located.
-To do this, we first extracted the reference genome sequence surrounding the variant using the bedtools "getfasta" function [@quinlan2010].
-Next, we re-annotated each sequence using BLAST to get the predicted gene location based on the closest annotated reference [@altschul1997].
-We then translated each variant to amino acids and aligned this translation to that of the gene annotation to ensure proper frame of reference using Geneious [@kearse2012].
-After ensuring proper translation, we characterized variants as within or outside the coding sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using *F~ST~* and differences in nucleotide diversity ($\pi_{diff}$) between naive and recovering populations.
-First, we calculated per-site *F~ST~* between the naive and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
-Next, we calculated per-site nucleotide diversity $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
-We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome (for more efficient data processing and to better contextualize the strength of the outlier signals in different regions of the genome).
-We then used the GenWin R package [@beissinger2015] to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
-This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the spline fitted to *F~ST~* and $\pi_{diff}$ values along the genome, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015].
-This method also calculates a W-statistic allowing for outlier identification.
-We identified outliers as those with a W-statistic greater than 4 standard deviations above the mean for *F~ST~* or above/below the mean for $\pi_{diff}$.
-These standard deviations represent strict criteria to select only the top approximately 0.3% of windows.
-Shared outliers were then identified as those that were outliers in both analyses, meaning that they showed (i) high differentiation between naive and recovering populations, and (ii) differential patterns of nucleotide diversity in the same region.
-Finally, we extracted gene transcripts mapped within each region and retrieved annotation for that region using BLAST (**Supporting Information**: Datasets S3, S4).
-
-Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
+\newpage
 
@@
 Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
-We also thank the editor and three anonymous reviewers whose input helped us improve the manuscript.
 This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
@@
 
-### References
+### Author contributions
+
+R.A.K., M.Q.W., M.B.J., T.C.S., and R.L.G designed the research.
+R.A.K., M.B.J., T.C.S., and R.L.G. contributed to data collection; M.B.J. contributed new analytic tools; R.A.K. and M.Q.W. conducted the analyses.
+R.A.K. and M.Q.W. wrote the manuscript with input from M.B.J., T.C.S., and R.L.G.
+
+### **Competing interests**
+
+The authors declare no competing interests.
+
+### Additional information
+
+**Supplemental information** is available for this paper at
+
+**Correspondence** and requests for materials should be addressed to R.A.K.
+
+\newpage
 
@@
 
-![For MYL frogs, a conceptual model depicting the Bd-caused decline and subsequent natural recovery (black text), facilitated recovery via reintroductions, and the linkages between these two pathways.
+![**Conceptual model of Bd-caused decline and recoveryin *R.
+sierrae*.** In addition to the original Bd-caused decline, the model depicts (i) natural recovery and (ii) facilitated recovery via reintroductions, as well as the linkages between these two pathways.
 Rectangles and hexagons represent outcomes and processes, respectively.
 Blue text indicates components that are included in the current study.
-We quantified frog survival and population viability following reintroductions directly using field studies and models, and evaluated "disease-mediated selection" indirectly using a comparison of frog exomes in naive and recovering populations.
 The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
@@
 
-![Median 1-year survival for each cohort of translocated frogs at the 12 recipient sites, as estimated for each site from the mrmr CMR model.
+![**Locations of translocated and donor *R.
+sierrae* populations.** All populations are in Yosemite National Park (park boundary indicated by gray polygon).
+Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
+To obscure the exact locations of populations, random noise was added to all point coordinates.
+Inset map shows the location of Yosemite within California.
+In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="60%"}
+
+\newpage
+
+![**1-year survival for each cohort of translocated frogs**.
+Median cohort-level survival was estimated for each of the 12 recipient sites using mrmr CMR models.
 Error bars show the 95% uncertainty intervals.
@@
 
-![Results from the among-site meta-analysis showing conditional effects of the important predictors of 1-year frog survival (expressed as a probability): (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
+![**Conditional effects of predictors of 1-year frog survival**.
+Results are from the among-site meta-analysis and 1-year frog survival (expressed as a probability).
+(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
@@
 
-![Results from population viability analysis: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+![**Estimated population growth rates and extinction probabilites of reintroduced populations**.
+(A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
@@
 (C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
-
-\newpage
-
-![Genomic patterns consistent with selection on individual variants in recovering MYL frog populations at the landscape scale (based on a linear mixed model analysis).
-For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray.
-The inset map locates the study area and range boundary in California.
-Sites 1 and 4 (site id = 72996 and 70567, respectively) were also used as sources of frogs for reintroductions described in **Results - Frog population recovery**.](figures/allele_maps.png){#fig-allelefrequencies width="95%"}
-
-\newpage
-
-![Results from the splined window analysis showing patterns consistent with selection on genomic regions in recovering MYL frog populations.
-Note that several key candidate genes (RIN3, TCF19, VARS) are outliers in both the individual variant analysis (@fig-allelefrequencies) and splined window analysis (this figure).
-Manhattan plots show outlier regions for the difference in (A) nucleotide diversity $\pi_{diff}$ and (B) *F~ST~*.
-In (A), outlier regions are shown above the upper red dashed line and below the lower red dashed line.
-In (B), outlier regions are shown above the single dashed red line.
-Outlier regions for either $\pi_{diff}$ or *F~ST~* are shown in blue and outlier regions for both $\pi_{diff}$ and *F~ST~* are shown in red.
-(C) Magnified Contig19 from (A) showing two adjacent outlier regions for $\pi_{diff}$ 12.9Mb upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue line).
-(D) Magnified Contig8 from from (B) showing the *F~ST~* outlier region that includes the outlier SNPs TCF19 and VARS.
-This region of the genome contains 8 annotated genes known to occur in the extended MHC Class I and III regions.](figures/splinewindow_manhattan.png){#fig-spline-manhattan width="95%"}
-
-\newpage
-
-## Supporting Information
-
-### Frog population recovery
-
-#### Laboratory methods
-
-Swab extracts were analyzed using standard Bd DNA extraction and qPCR methods [@boyle2004], and extracts were analyzed singly instead of in triplicate [@kriger2006].
-For analysis of swabs collected during 2005--2014, we used standards developed from known concentrations of zoospores [@boyle2004], and after 2014, we used standards based on single ITS1 PCR amplicons [@longo2013].
-Based on paired comparisons between samples analyzed using both types of standards, Bd in the study area has an average of 60 ITS1 copies per zoospore.
-To express all qPCR results as the number of ITS1 copies, starting quantities obtained using the zoospore standard (measured as "zoospore equivalents") were multiplied by 60.
-In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010].
-
-#### CMR model structure
-
-We estimated survival and recruitment for each site using open population CMR models based on @joseph2018.
-For each individual in a superpopulation of real and psuedo-individuals, $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected.
-Capture histories of $M$ individuals are modeled, although only $N_s$ individuals were captured.
-This parameter expanded data augmentation allows us to account for the possibility that undetected individuals may have recruited into the adult population [@royle2012].
-Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.
-
-We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$.
-The states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults.
-Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September).
-We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.
-Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
-We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die).
-This assumption is justified by the short time intervals between surveys within primary periods, in cases where primary periods contain multiple secondary periods.
-
-Live individuals are detected with probability $p_j$, which is modeled as:
-
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
-
-where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector.
-Not recruited and dead individuals are never captured.
-We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:
-
-$$
-\Omega_j =
-\begin{blockarray}{ccc}
-  \text{Not detected} & \text{Detected} \\
-\begin{block}{(cc)c}
-  1 & 0 & \text{Not recruited} \\
-  1 - p_j & p_j & \text{Alive} \\
-  1 & 0 & \text{Dead} \\
-\end{block}
-\end{blockarray}
-$$
-
-The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$.
-For non-introduced (i.e., naturally recruited) individuals, this matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 0 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$ where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
-
-For introduced individuals, which have deterministic recruitment (i.e., they recruit when introduced), the state transition matrix is given by:
-
-$$
-\Psi_{t, i} =
-\begin{blockarray}{cccc}
-  \text{Not recruited} & \text{Alive} & \text{Dead} \\
-\begin{block}{(ccc)c}
-  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
-  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
-  0 & 0 & 1 & \text{Dead} \\
-\end{block}
-\end{blockarray},
-$$ where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.
-
-We allow recruitment probabilities to vary in time via random effects, such that:
-
-$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
-
-where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.
-
-Survival probabilities also vary in time, and as a function of known covariates:
-
-$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
-
-where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.
-
-To complete the specification of the Bayesian model, we specify priors for all unknown parameters.
-The recruitment parameter priors were specified as follows:
-
-$$\alpha^{(\lambda)} \sim N(0,1),$$ $$\sigma^{(\lambda)} \sim N_+(0,1),$$ $$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$ for periods $t=1, ..., T$.
-Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
-
-Survival parameter priors were specified similarly as:
-
-$$\beta^{(\phi)} \sim N(0, 1),$$ $$\sigma^{(\phi)} \sim N_+(0, 1),$$ $$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$ for time $t=1, ..., T$.
-
-The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.
-
-We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm [@zucchini2009; @joseph2018].
-
-All of the code to specify and fit the model in Stan is available in the open source mrmr package [@joseph2019].
-
-The joint distribution of the resulting model can be written as follows:
-
-```{=tex}
-\begin{multline*}
-[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
-\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
-\end{multline*}
-```
-where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
-
-#### Among-site survival modeling
-
-The objective of this analysis is to describe the influence of site, cohort, and individual level characteristics on post-translocation frog survival.
-By modeling survival estimates obtained from site-specific mrmr CMR analyses, we are in effect conducting an among-site meta-analysis.
-Although it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (namely, run time and memory).
-
-We used Bayesian generalized linear mixed models to investigate predictors of survival among sites.
-The response $y_i$ is binary, representing a point estimate of whether individual $i$ survived in the year following translocation.
-We generated these point estimates by rounding the posterior median of 1-year post-introduction survival for each individual (from site-specific mrmr CMR models) and modeled the data using a Bernoulli distribution:
-
-$$y_i \sim \text{Bernoulli}(p_i),$$
-
-where $p_i$ is the probability of survival.
-
-We modeled variation in probabilities as follows:
-
-$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$
-
-where $\alpha$ is an intercept, $X_i$ is a length $K$ row vector of predictors, $\pmb \beta$ is a column vector of predictor effects, and $\pmb \nu$ a vector of group level random effects.
-Here $g[i]$ refers to the group $g$ containing individual $i$, and we estimate an adjustment for each of the $G$ groups ($\nu_1, ..., \nu_G$).
-
-These models were fit using the `stan_glmer()` function in the rstanarm package, with default priors described below [@rstanarm2022].
-These priors are vague, but include data-dependent scaling as follows to account for different input variable scales.
-However, because we standardized all predictor variables similarly to have equal variance (by centering and dividing by twice the sample standard deviation), the resulting priors are identical.
-Specifically, we have:
-
-$$\alpha \sim \text{Normal}(0, 2.5),$$ and
-
-$$\beta_k \sim \text{Normal}(0, 5),$$
-
-for $k=1, ..., K$ where $K$ is the number of predictor variables.
-
-The default prior for group level adjustments $\nu_1, ..., \nu_G$ in rstanarm is a zero-mean Gaussian, where the covariance matrix is constructed from a correlation matrix with an LKJ prior, and a vector of variance parameters -- the decomposition of variance prior with unit regularization, concentration, shape, and scale parameters [@lewandowski2009].
-
-We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan, with four parallel chains, each run for 10,000 iterations, discarding the first half of each chain as warm-up draws [@rstanarm2022].
-We used Rhat statistics and trace plots to verify convergence.
-We considered models with different subsets of fixed and random effects, and used approximate leave-one-out cross validation to identify the best model [@vehtari2016].
-
-### Population viability modeling
-
-#### Incorporating yearly variability in vital rates
-
-We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mrmr CMR models.
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (@fig-translocation-survival).
-Thus, we did not include yearly within-population variability in adult survival in this analysis.
-However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
-
-#### Estimating model parameters
-
-The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Frog populaton recovery - Estimation of frog survival and abundance** and **Supporting Information - Frog population recovery - CMR model structure** for details).
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs SI).
-For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
-
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)).
-This yielded posterior distributions for survival probabilities between primary periods.
-However, because the time interval between primary periods differed, the survival probabilities between different consecutive primary periods were not directly comparable.
-To address this, we converted the survival probabilities between each consecutive primary period to per day death rates, propagating the uncertainty from the posterior distributions.
-We then took a weighted average of these death rates, weighted by the time interval between primary periods, to get the average per day death rate over the entire CMR survey.
-We converted this per day death rate $d$ back to a yearly survival probability using $\exp(-d \times 365 \text{ days})$.
-We used the same procedure for $\sigma_{A_T}$ such that our estimates of average yearly survival probability were comparable between $\sigma_{A_R}$ and $\sigma_{A_T}$.
-
-#### Model analysis and simulation
-
-We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
-We then fixed $\sigma_{J_1} = 0.09$ and calculated the predicted growth rate of our 12 populations.
-
-Second, we performed a local elasticity and sensitivity analysis on $\lambda$ with respect to parameters $\sigma_{L_1}$, $\sigma_{L_2}$, $\sigma_{L_3}$, $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, $F$, $p_{L_2}$, $p_{J_1}$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp SI).
-Note that $\lambda$ is equally or more sensitive to changes in adult survival ($\sigma_{A_R}$) and survival of year-1 juveniles ($\sigma_{J_1}$) than it is to other model parameters, providing additional justification for focusing on variation in $\sigma_{A_R}$ and $\sigma_{J_1}$ in our viability analyses.
-
-Third, we defined a version of the model with demographic and environmental stochasticity, where environmental stochasticity was represented by among-year variability in $\sigma_{J_1}$.
-We used this model to simulate a one-time introduction of 40 translocated adult frogs.
-We ran this simulation 1000 times for each population and computed the probability of a population becoming extinct after 50 years given the observed parameter values and environmental stochasticity in $\sigma_{J_1}$.
-We varied the mean recruitment probability $\sigma_{J_1}$ from 0 and 0.25 and drew values of $\sigma_{J_1}$ each year from a beta distribution with a dispersion parameter of $\phi = 2$ (when $\sigma_{J_1} = 0.5$ and $\phi = 2$ the beta distribution is uniform between 0 and 1).
-Using different values of $\phi$ does not qualitatively change the existence of distinct extinction dynamics between populations with $\sigma_{A_R} < 0.5$ and those with $\sigma_{A_R} > 0.5$.
-However, increasing yearly variability in $\sigma_{J_1}$ increases extinction risk for all populations.
-For example, if we set $\phi = 0.001$, such that in a given year essentially either all year-1 juveniles survive or all of them die, populations with $\sigma_{A_R} > 0.5$ need to have $\sigma_{J_1}$ greater than 0.2 to have a 50-year extinction probability of less 50%.
-Because we do not PIT tag juveniles, we do not have CMR estimates for $\sigma_{J_1}$ or $\phi$.
-However, based on qualitative and semi-quantitative field observations over 25 years, a value of $\sigma_{J_1} = 0.25$ in the presence of Bd is probably a reasonable estimate for many populations.
-Thus, we expect our model predictions to be conservative with regards to population recovery.
-
-Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
-We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
-We simulated our model for 16 years, repeating the simulation 50,000 times.
-For each run and each year, we drew $\sigma_{J_1}$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
-Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
-The yearly $\sigma_{J_1}$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
-Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
-
-### Frog genome evolution
-
-#### Study design
-
-To gain insights into the role of evolution in the development of resistance by MYL frogs, we compared frog exomes sampled in naive versus recovering populations.
-Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
-The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
-As a result, samples from recovering populations collected before and after Bd exposure are not available and are unlikely to be available in the future.
-
-Our study design, in which we compared frog genomes in naive and recovering populations, required sampling populations across the range of MYL frogs in the southern Sierra Nevada.
-This is due to the fact that very few MYL frog populations remain in the naive state, and those that do are scattered across a wide latitudinal range, from Yosemite National Park in the north to southern Kings Canyon National Park in south.
-To minimize potential confounding effects caused by known variation in frog genotypes across latitude [@byrne2023], we selected sampling sites such that both population types were represented across similar latitudinal ranges (@fig-allelefrequencies).
-
-#### GO analysis
-
-In **Results-Frog genome evolution**, we describe the stringent set of outlier variants (identified using a Bonferroni-corrected p-value of 0.01).
-A liberal set of outlier variants, identified using a Bonferroni-corrected p-value of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes across 16 contigs.
-We used this liberal set to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
-To do this, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
-We then conducted a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005] to compare the 30 unique outlier genes to the complete set of genes in our target transcriptome.
-We repeated this process for the set of 35 genes located in the 9 shared regions of the *F~ST~* and $\pi_{diff}$ splined windows.
-
-#### Genetic diversity
-
-To characterize general patterns of genetic diversity between naive and recovering populations, we conducted three analyses.
-First, we calculated heterozygosity for each sampled frog using VCFtools [@danecek2011].
-Second, to characterize genome-wide patterns of diversity, we used VCFtools to calculate nucleotide diversity ($\pi$) in 100kb sliding windows along the genome for each population.
-Third, we calculated average $\pi$ per population within each of the 9 shared outlier regions identified in the splined window analysis.
-We compared average $\pi$ per population between naive and recovering population types, and also for each naive and recovering population in Yosemite.
-The latter comparison focused on Yosemite because these populations are the most relevant to the translocated populations described in the paper.
-
-Average individual-level heterozygosity and genome-wide population-level $\pi$ were similar between the naive and recovering groups (@fig-violinplot-heterozy SI; @fig-boxplot-genomewide-pi-by-pop SI).
-Within each of the 9 shared outlier regions, $\pi$ shows considerable variation between populations and no obvious patterns between the naive and recovering populations (@fig-boxplot-pi-by-regionpop SI).
-
-\newpage
-
-### Figures
-
-\newpage
-
-![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon).
-Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
-To obscure the exact locations of populations, random noise was added to all point coordinates.
-Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="60%"}
-
-\newpage
-
-![For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.
-Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Samples sizes are provided immediately above the x-axis.
-The red dashed horizontal line indicates the approximate load above which severe disease is typically observed [@vredenburg2010].
-This threshold (5.8 ITS1 copies - on a log~10~ scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations [@briggs2010].](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
-
-\newpage
-
-![Results from the among-site meta-analysis, showing that Bd load is not an important predictor of post-translocation frog survival.
-Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model.
-In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
-Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
-
-\newpage
-
-![Comparison of average yearly adult survival probabilities for adults translocated to each of 3 sites versus adults that were naturally recruited at each site (as a result of reproduction by translocated frogs).
-In contrast to @fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
-Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in **Supporting Information - Population viability modeling - Estimating model parameters**.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="60%"}
-
-\newpage
-
-![Sensitivity analysis of the stage-structured MYL frog model.
-A.
-Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
-B.
-Sensitivity of $\lambda$ with changes in the same nine parameters.
-Note that $F$ has different units than the other parameters, which is why its sensitivity is low.
-Sensitivities of all other parameters are comparable.
-We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
-
-\newpage
-
-![Results from the analysis of individual variants, showing possible signatures of selection in recovering MYL frog populations.
-(A) PCA calculated from binary SNPs showing the genomic relationship of samples.
-Numeric labels and colors match those in @fig-allelefrequencies.
-Populations 1-7 are *R.
-sierrae* and populations 8 and 9 are *R.
-muscosa*.
-(B) qqplot showing observed and expected p-values for 148,307 SNPs and INDELS.
-Dashed line shows p-value that identifies outliers.
-(C) Manhattan plot showing p-value for each SNP.
-SNPs are sorted by genomic position and contigs are sorted by size.
-Red line shows p-value that identifies outliers.
-Outlier SNPs above this threshold are highlighted and labeled.
-Bold labels indicate the presence of at least one non-synonymous SNP in that gene.](figures/pca_qq_manhattan.png){#fig-selectionresults width="85%"}
-
-\newpage
-
-![Synteny plot showing conserved gene order in *Xenopus tropicalis* and *Rana muscosa* for the outlier region containing MHC Class I Classical and MHC Class III gene regions.
-The plot was created with SimpleSynteny [@veltri2016] using *Xenopus tropicalis* Chromosome 8 (NC_030684.2, genbank accession GCA_000004195.4) and *Rana muscosa* Contig19.
-Asterisks indicate the location of SNP outliers in the TCF19 and VARS1 genes.
-Gap sizes for each contig representation are labeled.](figures/synteny_figure.png){#fig-synteny-plot width="95%"}
-
-\newpage
-
-![For each study population located in Yosemite National Park (populations 1-4 shown in the map), $\pi$ for the 9 shared outlier regions identified in the splined window analysis.
-Populations 1 and 4 were the source of frogs for 20 of the 24 frog translocations, and the donor population for the remaining four translocations is located adjacent to populations 3 and 4 (see also @fig-yosemap).
-Bars are color-coded by population type (naive, recovering).
-The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.
-Region 4 is the extended MHC region that includes two SNPs identified by the GEMMA analysis in genes TCF19 and VARS1.
-In the map, the four populations included in panels 1-9 are color-coded by population type, and dashed lines delineate the genetic clusters as defined in [@fig-selectionresults]A.](figures/ave_pi_per_popn_yose.png){#fig-ave-pi-per-popn-yose width="85%"}
-
-\newpage
-
-![Violin plots showing individual heterozygosity for the Bd-naive and recovering populations included in the frog evolution study.
-Individual data points are represented by their corresponding site number (from @fig-allelefrequencies).](figures/violin_plot_heterozy_by_group.png){#fig-violinplot-heterozy width="60%"}
-
-\newpage
-
-![For each population in the frog evolution study (@fig-allelefrequencies), boxplots showing nucleotide diversity ($\pi$) within 100kb sliding windows along the genome.
-The y-axis has been log~10~-transformed for display purposes.
-Sites are color-coded as "Bd-naive" or "recovering", and vertical dashed lines show the genetic breaks between frog populations as described in [@fig-selectionresults]A).](figures/boxplot_genomewide_pi_by_pop.png){#fig-boxplot-genomewide-pi-by-pop width="70%"}
-
-\newpage
-
-![For each of the 9 shared outlier regions identified by the splined window analysis, boxplots showing $\pi$ values for naive and recovering populations.
-The corresponding gene annotations for each region are as follows (gene regions described in the **Results** and **Discussion** are indicated with an asterisk): (1) C6\*, C7\*, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19\*, HSP70, LSM2, VARS1\*, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1\*, (7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.](figures/boxplot_pi_by_regionpop.png){#fig-boxplot-pi-by-regionpop width="70%"}
-
-\newpage
-
-### Tables
-
-| **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:-------------------------|:------------------|:--------------------------|
-| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies                                                        |
-| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from CMR studies                                                        |
-| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge                |
-| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
-| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
-| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                                |
-| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                                |
-
-: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
-
-\newpage
-
-### Datasets
-
-Dataset S1.
-Set of liberal SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.05), identified via GEMMA.
-file: gemma_outliers_all.csv
-
-Dataset S2.
-Set of strict SNP and INDEL outlier variants (Bonferroni corrected p-value \< 0.01), identified via GEMMA.
-Additional information includes variant location within the gene (predicted_gene_loc) and whether the variant is synonymous or nonsynonymous (predicted_effect_AA).
-file: gemma_outliers_strict_freq.csv
-
-Dataset S3.
-Description of overlapping *F~ST~* and $\pi_{diff}$ outlier windows, as identified in the splined window analysis.
-The outlier window on chromosome 19 (column name = chr_num_sorted) is included twice because one *F~ST~* outlier window overlapped with two $\pi_{diff}$ outlier windows.
-file: spline_window_shared_outliers.csv
-
-Dataset S4.
-Annotation information for all genes within each of the overlapping outlier windows in dataset S3.
-file: spline_window_gene_details
-
-### References

### 2024-05-19 doc/manuscript/supplementary_information.qmd
Minor edits to fix typos and improve readability

@@
   fig-title: "**Supplementary Fig.**"
+  fig-prefix: "Supplementary Fig."
   tbl-title: "**Supplementary Table**"
+  tbl-prefix: "Supplementary Table"
   title-delim: "|"
@@
 csl: nature.csl
+link-bibliography: false
 ---
@@
 
-## Reintroduction of resistant frogs facilitates recovery in the presence of a lethal fungal disease
+## Reintroduction of resistant frogs facilitates recovery despite a lethal fungal disease
 
@@
 We computed yearly survival probabilities for translocated adults $\sigma_{A_T}$ and naturally recruited adults $\sigma_{A_R}$ from the posterior distribution of individual state trajectories derived from mrmr CMR models.
-Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (@fig-translocation-survival).
+Although we observed yearly variability in adult survival within a population, the magnitude of this variability was small compared to among-population variability (Fig. 3).
 Thus, we did not include yearly within-population variability in adult survival in this analysis.
@@
 Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp SI).
+We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp).
 
@@
 
-![**Predictors of post-translocation frog survival in the among-site meta-analysis.** Depicted distributions are the estimated posterior density curves and shaded 95% uncertainty intervals for the intercept and all predictor variables from the best model.
+![**Predictors of post-translocation frog survival in the among-site meta-analysis.** Depicted distributions for the intercept and all predictor variables are the estimated posterior density curves and shaded 95% uncertainty intervals from the best model.
 In the Bayesian framework in which the model was developed, variables are considered important predictors if the associated uncertainty interval does not overlap zero (indicated by the dashed red line).
 Predictor variables shown on the y-axis are defined as follows: length = frog size, snow_t = winter severity in the year of translocation (measured on April 1), snow_t1 = winter severity in the year following translocation (measured on April 1), day = day of year on which a translocation was conducted, bd_load = Bd load, elevation = site elevation, order_first = within-site translocation order, sex_male = frog sex, donor_70567 and donor_72996 = donor population.
-Bd load is not an important predictor of post-translocation frog survival.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
+Note that Bd load is not an important predictor of post-translocation frog survival.](figures/mcmc_areas_m1d.png){#fig-survival-postdens width="60%"}
 
@@
 
-![**Average yearly survival probabilities of translocated versus naturally recruited adults.** Naturally recruited individualos were the result of reproduction by translocated frogs at each site.
-In contrast to \@fig-translocation-survival, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
+![**Average yearly survival probabilities of translocated versus naturally recruited adults.** Naturally recruited individuals were the result of reproduction by translocated frogs at each site.
+In contrast to Fig.
+3, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts.
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
@@
 Sensitivities of all other parameters are comparable.
-We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study (Table S1) -- tadpoles spend at least a year as tadpoles before metamorphosing.
+We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study ([@tbl-param_values]) -- tadpoles spend at least a year as tadpoles before metamorphosing.
 Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}

### 2024-05-19 doc/manuscript/translocation.qmd
Minor edits to fix typos and improve readability

@@
 
-## Reintroduction of resistant frogs facilitates recovery in the presence of a lethal fungal disease
+## Reintroduction of resistant frogs facilitates landscape-scale recovery despite a lethal fungal disease
 
@@
 Wilber^3^, <!--#Allison Q.
-Byrned,e,1-->, Maxwell B.
+Byrned,e,1--> Maxwell B.
 Joseph^4,5^, Thomas C.
@@
 Rothsteind,e,h--> & Robert L.
-Grasso^6^, <!--#Erica Bree Rosenblumd,e-->
+Grasso^6^ <!--#Erica Bree Rosenblumd,e-->
 
@@
 The emergence of the lethal fungal pathogen *Batrachochytrium dendrobatidis* ("Bd") has devastated global amphibian biodiversity.
-Lacking any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
-The Sierra Nevada yellow-legged frog (*Rana sierrae*) is highly susceptible to Bd infection.
-Although most *R.
-sierrae* populations are extirpated following disease outbreaks, some persist and eventually recover and frogs in these recovering populations have increased resistance against infection.
+Given the lack of any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
+The Sierra Nevada yellow-legged frog (*Rana sierrae*) is highly susceptible to Bd infection and most *R. sierrae* populations are extirpated following disease outbreaks.
+However, some populations persist and eventually recover, and frogs in these recovering populations have increased resistance against infection.
 We conducted a 15-year reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
-In addition, the likelihood of establishment is influenced by site, population, and frog attributes.
+In addition, the likelihood of establishment is influenced by site, cohort, and frog attributes.
 Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
@@
 
-Human activities are increasingly impacting global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
+Human activities are devastating global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
 One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
@@
 By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
-These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding mechanisms of species persistence in the presence of emerging diseases [@russell2020].
+The escalating frequency of disease emergence and the strength of impacts on wild populations increasingly highlight the critical need for successful mitigation of disease in species conservation.
 
-Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare (for notable exceptions, see [@scheele2017; @voyles2018; @knapp2016]), suggesting that disease-caused declines will be difficult to reverse.
-This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
-Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@raberg2009] and facilitate population persistence and recovery [@brannelly2021].
-Host immunity and evolution both play important roles in the development of resistance and tolerance, and an enhanced understanding of these processes is important for developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
-However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes (but see [@savage2016; @grogan2018b]).
-These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
+Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms by which hosts reduce disease impacts [@raberg2009].
+The effectiveness of these mechanisms has direct consequences for the persistence and recovery of host populations in the presence of disease [@brannelly2021].
+Host immunity and evolution both play important roles in the development of resistance and tolerance, and are key targets for strategies aimed at mitigating disease impacts in the wild [@garner2016].
+
+Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare (for notable exceptions, see [@scheele2017; @voyles2018; @knapp2016]).
+This apparent low resilience to disease may be due to the limited ability of many amphibians to develop Bd resistance or tolerance via either immunity [@rosenblum2012; @fites2013; @grogan2018a] or evolution (but see [@savage2016; @grogan2018b]).
+In turn, this would also reduce the effectiveness of potential Bd mitigation strategies.
+For example, a limited ability by amphibians to develop resistance or tolerance suggests that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
 If true, the future of many amphibian species threatened by Bd appears bleak.
 
-The Sierra Nevada yellow-legged frog (*Rana sierrae*) [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
-Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
-Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
+The Sierra Nevada yellow-legged frog (*Rana sierrae*), is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
+Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains, USA [@grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
+Due to the severity of its decline and the increasing probability of extinction, *R. sierrae* is now listed as "endangered" under the U.S.
 Endangered Species Act.
-In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
+The decline of *R. sierrae* was initiated by the introduction of non-native trout into the extensive fishless region of the Sierra Nevada [@bradford1989; @knapp2000] starting in the late 1800s.
 The arrival of Bd in the mid-1900s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
@@
 Despite the catastrophic impact of Bd on *R.
-sierrae*, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
-Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
+sierrae*, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016] (@fig-recovery-model).
+During the epizootic phase, *R. sierrae* are highly susceptible to Bd infection as indicated by very high infection intensities ("load").
+In constrast, frogs in recovering populations have reduced susceptibility to Bd infection [@knapp2016], with loads on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
 This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection and not simply an effect of factors external to individual frogs (e.g., environmental conditions).
@@
 Therefore, we focus on resistance throughout this paper.
+
 The observed resistance of *R.
@@
 The possible evolution of resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
-This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].
+This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats [@joseph2018; @mendelson2019] (@fig-recovery-model).
 
@@
 We accomplished this using a 15-year large-scale reintroduction study.
-Second, identify important site, population, and individual-level predictors of frog survival using results from the reintroduction study.
-Third, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
+Second, identify important site, cohort, and individual-level predictors of frog survival using results from the reintroduction study.
+Third, estimate the probability of persistence for the reintroduced populations over a multi-decadal period, thereby extending our inferences of population recovery beyond the temporal bounds of our reintroduction study (@fig-recovery-model).
 We did this using a stage-structured matrix model.
@@
 Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap).
-The 3 donor populations are among the largest *R.
-sierrae* populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
-These 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
+The 3 donor populations are among the largest *R. sierrae* populations in Yosemite.
 Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
-Across all translocation sites, the duration of survey time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
+Across all translocation sites, the duration of CMR surveys and VES was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).
 
@@
 
-Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (Supplementary Fig.
-1) [@joseph2018; @vredenburg2010].
+Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality [@joseph2018; @vredenburg2010] (Supplementary Fig.
+1) .
 Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; Supplementary Fig.
@@
 We parameterized the model using CMR data from translocated populations and known life history values in this system (Supplementary Table 1).
+Results from a sensitivity analysis are provided in Supplementary Fig.
+4)
 
@@
 
-Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
+Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking (e.g., [@garner2016]).
 Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered *R.
 sierrae* in the presence of a highly virulent pathogen (Bd).
-Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that post-translocation frog survival is influenced by site, population, and individual-level characteristics, and (iii) 50% of the reintroduced populations are likely to persist over a 50-year period Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
+Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that post-translocation frog survival is influenced by site, cohort, and individual-level characteristics, and (iii) that 50% of the reintroduced populations are likely to persist over a 50-year period.
 In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing *R.
 sierrae* populations via reintroduction of resistant individuals is striking, and even more so given that *R.
-sierrae* were driven to near-extinction by Bd.
+sierrae* was driven to near-extinction by Bd.
+Collectively, these results provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
 
@@
 Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
-In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed ([@briggs2010; @knapp2011] see also Supplementary Fig.
+In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011] (see also Supplementary Fig.
 2).
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies (see Figure 4 in [@knapp2016]), and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
-This resistance could be due at least in part to natural selection for more resistant genotypes, but until now evidence to evaluate the possible role of evolution was lacking.
 
-Reintroduction of resistant *R. sierrae* was remarkably successful in reestablishing viable populations in the presence of Bd.
+Reintroduction of resistant *R. sierrae* was remarkably successful in reestablishing populations in the presence of Bd.
 Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
 Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
-The fact that the relatively low Bd loads and correspondingly high frog survival was maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
+The fact that the relatively low Bd loads and correspondingly high frog survival were both maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
 Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and reintroduced populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the reintroduced versus donor populations and pathogen pressure should follow a similar pattern.
@@
 The duration of the tadpole and juvenile stages increases with elevation, but the fact that long-run population growth rate from our viability model is relatively insensitive to variation in these durations suggests that developmental rates of early frog life stages have minimal effect on the predicted long-term viability of reintroduced populations.
-Nonetheless, a clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
+The mechanisms underlying the effect of donor population on the probability of post-translocation frog survival is intriguing, and could indicate population-specific fitness differences of frogs in the presence of Bd.
+A clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
 
@@
 In both naturally recovering and reintroduced *R. sierrae* populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
-Therefore, although adult *R. sierrae* appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].
+Therefore, although adult *R. sierrae* appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations (see also [@hollanders2022]).
 
@@
 As with *R.
-sierrae*, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have the adaptive alleles necessary to allow frog survival and population growth in the presence of Bd.
-Despite the hopeful example of successful reestablishment of *R.
+sierrae*, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have sufficiently high resistance to allow frog survival and population growth in the presence of Bd.
+Despite the promising example provided by our study of successful reestablishment of *R.
 sierrae* despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
@@
 
-For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
+For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected based on several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
 The populations that we selected contained hundreds of *R. sierrae* adults and thousands of tadpoles.
@@
 As is typical for recovering *R.
-sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
+sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected [@vredenburg2010; @joseph2018] (log~10~(load) $\approx$ 5.78 copies).
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 
-We conducted `r transno_min`--`r transno_max` translocations per site (@fig-translocation-survival, @fig-yosemap) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
+We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
 In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
@@
 Following release, we visited translocated populations approximately once per month during the summer active season and conducted diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
-CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size, and VES provided estimates of tadpole and juvenile abundance.
+CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size.
+VES provided estimates of tadpole and juvenile abundance.
 During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we searched all habitats repeatedly for adult frogs.
@@
 
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software (v4.4.4 [@rsoftware2022]) (see **Supporting Information - Frog population recovery - CMR model structure** for details).
-Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals (see [@joseph2018] for details).
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [@rsoftware2022] (v4.4.4) (see **Supporting Information - Frog population recovery - CMR model structure** for details).
+Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [@joseph2018].
 The possible states of individuals include "not recruited", "alive", and "dead".
@@
 Additional details are provided in **Supporting Information - Population viability modeling - Model analysis and simulation**.
-Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.
 
@@
 
-R code used to run our analyses is available at <https://github.com/SNARL1/translocation>.
+R code to replicate our analyses is available at <https://github.com/SNARL1/translocation>.
 R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis>.
@@
 
-![**Conceptual model of Bd-caused decline and recoveryin *R.
-sierrae*.** In addition to the original Bd-caused decline, the model depicts (i) natural recovery and (ii) facilitated recovery via reintroductions, as well as the linkages between these two pathways.
+![**Conceptual model of Bd-caused decline and recovery in *R.
+sierrae*.** For the recovery portion, the model depicts both natural recovery and facilitated recovery via reintroductions, as well as the linkages between these two pathways.
 Rectangles and hexagons represent outcomes and processes, respectively.
 Blue text indicates components that are included in the current study.
-The timeline shows the general sequence of the components, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
+The general temporal scale of the depicted scenario is provided by the timeline, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
 
@@
 ![**Conditional effects of predictors of 1-year frog survival**.
-Results are from the among-site meta-analysis and 1-year frog survival (expressed as a probability).
+Results are from the among-site meta-analysis and 1-year frog survival is expressed as a probability.
 (A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
@@
 
-![**Estimated population growth rates and extinction probabilites of reintroduced populations**.
+![**Estimated growth rates and extinction probabilities of reintroduced populations**.
 (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
@@
 Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 translocated populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
-There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505 when $\sigma_{J_1} < 0.10$.
+(B) Predicted 50-year extinction probabilities of the 12 reintroduced populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
+There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are partially or completely hidden beneath the line for 70505.
 (C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}

### 2024-05-21 doc/manuscript/cover_letter_nature_comm.qmd
Add cover letter

---
format: 
  pdf:
    fig-pos: "H"
    include-in-header: 
      text: '\pagenumbering{gobble}'
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

In this cover letter, I attempt to provide some context for our manuscript, "Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease".
It is my sincere hope that this context will serve as a more compelling argument for why this manuscript is worthy of consideration for publication in *Nature Communications* than the usual (and somewhat pedantic) arguments of how novel this research is or how important it is for a particular branch of science.

Amphibians are the most threatened vertebrate class, with more than 40% of species threatened with extinction.
Despite their almost 400 million year history on Earth and their record of having survived the last four mass extinction events, amphibians are uniquely vulnerable to many of the human pressures that characterize our current epoch: habitat destruction, introduction of nonnative species, emergence of novel infectious diseases, rapidly warming climate, etc., etc.
A description of the impacts of any one of these factors is enough to bring a sense of gloom to even the most optimistic among us.
In less than a generation, the human-driven emergence of the amphibian chytrid fungus (*Batrachochytrium dendrobatidis*; Bd) has devastated global amphibian biodiversity, with thousands of populations in decline or extirpated, and dozens of species now extinct in the wild.
Even those amphibians in the most protected habitats have fallen victim to this invisible killer.

During my own 30+ year career studying mountain ecosystems and their fauna, I've experienced those disappearances first hand and more times than I care to remember.
After restoring lakes to their original fishless condition and documenting the subsequent recolonization and skyrocketing abundance of Sierra Nevada yellow-legged frogs, I've watched helplessly as those same amphibians succumbed to Bd and their mountain home went silent.
Gone was the shower of frogs diving from the shore into water as they fled my approach.
Gone were the masses of tadpoles in lake shallows, once so dense that the water frothed with their wriggling tails.
Diminished too were the birds and snakes that fed on those frogs.
These protected habitats in Yosemite National Park were unraveling like a tapestry come unknotted.

Unwilling to simply walk away from this loss, my colleagues and I put everything we had into understanding this pathogen and its impact on amphibians.
We gained key insights from our decades of research and from sharing results with others around the world who were witnessing their own amphibian catastrophes.
The work was slow and arduous, with as many setbacks as successes, and all the while the threat of extinction loomed large.
A key inflection point for me came twenty years ago with the realization that some frog populations driven to near extinction by Bd were, almost miraculously, beginning to recover on their own.
The subsequent intensive study of those populations provided the knowledge that we needed to bolster this natural recovery using frog reintroductions.

The science behind that uniquely successful 15 year recovery effort is the focus of our manuscript, a story told through graphs, tables, and methodological minutiae.
But there is another way to understand the consequences of that work, and that is to sit on the shore of any one our study lakes in which we reintroduced frogs 10 or more years ago.
The silence that settled over the lake like a morning fog following the frogs' extinction is no more.
Instead, the lake shore is alive with frogs, one atop every rock and lining the sandy beaches.
On calm mornings, pairs of eyes bulge above the water's surface like unblinking periscopes.
The tadpoles too are back, their black thousands filling bays with ripples and splashing.
The lake is alive again.

Given the paucity of well-documented examples of frog recovery following Bd-caused declines, the scientific story that we describe in our manuscript is worth spreading widely.
Its implications for the world's declining amphibians are clear, and the example that our study provides is a badly needed roadmap for the recovery of global amphibian biodiversity.
We believe that *Nature Communications* has the broad audience and global reach that this story deserves.

Thanks for your careful review of our manuscript.

\
Sincerely,

*Roland A. Knapp*, Ph.D.

### 2024-05-21 doc/manuscript/supplementary_information.qmd
Minor edits to manuscript files

@@
 ---
-format: pdf
+format: 
+  pdf:
+    fig-pos: "H"
 #format:
@@
     \usepackage{blkarray}
+include-before-body: 
+  text: |
+    \captionsetup{format=plain, labelfont=bf, labelsep=period}
 editor: visual
@@
   tbl-prefix: "Supplementary Table"
-  title-delim: "|"
 bibliography: translocation.bib
@@
 
-## Reintroduction of resistant frogs facilitates recovery despite a lethal fungal disease
+## Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
 
-Roland A. Knapp\*, Mark Q. Wilber, Maxwell B. Joseph, Thomas C. Smith, Robert L. Grasso
+Roland A. Knapp\*, Mark Q. Wilber, Maxwell B. Joseph, Thomas C. Smith, & Robert L. Grasso
 
@@
 ![**Sensitivity analysis of the stage-structured *R.
-sierrae* model.** A.
-Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
-B.
-Sensitivity of $\lambda$ with changes in the same nine parameters.
+sierrae* model.** **A** Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+**B** Sensitivity of $\lambda$ with changes in the same nine parameters.
 Note that $F$ has different units than the other parameters, which is why its sensitivity is low.
@@
 
+\newpage
+
 ### Tables

### 2024-05-21 doc/manuscript/translocation.qmd
Minor edits to manuscript files

@@
     \usepackage{blkarray}
+include-before-body: 
+  text: |
+    \captionsetup{format=plain, labelfont=bf, labelsep=space}
 editor: visual
@@
   fig-title: "**Fig.**"
-  title-delim: "|"
+  fig-prefix: "Fig."
 bibliography: translocation.bib
@@
 
-## Reintroduction of resistant frogs facilitates landscape-scale recovery despite a lethal fungal disease
+## Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
 
-Roland A.
-Knapp^1,2,^\*, Mark Q.
-Wilber^3^, <!--#Allison Q.
-Byrned,e,1--> Maxwell B.
-Joseph^4,5^, Thomas C.
-Smith^1,2^ <!--#Andrew P.
-Rothsteind,e,h--> & Robert L.
-Grasso^6^ <!--#Erica Bree Rosenblumd,e-->
+Roland A. Knapp^1,2,^\*, Mark Q. Wilber^3^, Maxwell B. Joseph^4,5^, Thomas C. Smith^1,2^ & Robert L. Grasso^6^
 
@@
 
-<!--#dDepartment of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114-->
-
-<!--#eMuseum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160 -->
-
 ^4^Earth Lab, University of Colorado, Boulder, CO, 80303
@@
 
-<!--#Current affiliation: Ginkgo Bioworks, Boston, MA, 02210-->
-
 ^6^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
 
-\*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu
-
-<!--#1These authors contributed equally to this work.-->
+\*Corresponding author: email: roland.knapp\@ucsb.edu
 
@@
 Results are from the among-site meta-analysis and 1-year frog survival is expressed as a probability.
-(A) winter severity in the year following translocation, (B) elevation of recipient site, (C) donor population, and (D) sex.
+**A** winter severity in the year following translocation, **B** elevation of recipient site, **C** donor population, and **D** sex.
 In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
@@
 ![**Estimated growth rates and extinction probabilities of reintroduced populations**.
-(A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+**A** Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
@@
 Note that the point for 70413 is mostly hidden behind other points.
-(B) Predicted 50-year extinction probabilities of the 12 reintroduced populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
+**B** Predicted 50-year extinction probabilities of the 12 reintroduced populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are partially or completely hidden beneath the line for 70505.
-(C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
+**C** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}

### 2024-05-21 doc/manuscript/cover_letter_nature_comm.qmd
Final edits prior to submission

@@
   pdf:
-    fig-pos: "H"
+    fontsize: 12pt
     include-in-header: 
@@
 
-In this cover letter, I attempt to provide some context for our manuscript, "Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease".
-It is my sincere hope that this context will serve as a more compelling argument for why this manuscript is worthy of consideration for publication in *Nature Communications* than the usual (and somewhat pedantic) arguments of how novel this research is or how important it is for a particular branch of science.
+In this cover letter, I'd like to provide some context for our manuscript, "*Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease*".
+Although this cover letter takes a different approach than most cover letters (including those I have written for other papers), it is my sincere hope that the provided context will serve as a compelling argument for why this manuscript is worthy of consideration for publication in *Nature Communications*.
 
@@
 During my own 30+ year career studying mountain ecosystems and their fauna, I've experienced those disappearances first hand and more times than I care to remember.
-After restoring lakes to their original fishless condition and documenting the subsequent recolonization and skyrocketing abundance of Sierra Nevada yellow-legged frogs, I've watched helplessly as those same amphibians succumbed to Bd and their mountain home went silent.
+After restoring lakes to their original fishless condition and documenting the subsequent recolonization and skyrocketing abundance of Sierra Nevada yellow-legged frogs, I've watched helplessly as those same amphibians succumbed to Bd, and their mountain home went silent.
 Gone was the shower of frogs diving from the shore into water as they fled my approach.
@@
 
-Unwilling to simply walk away from this loss, my colleagues and I put everything we had into understanding this pathogen and its impact on amphibians.
+Unwilling to simply walk away from this loss, my colleagues and I put everything we had into understanding Bd and its impact on amphibians.
 We gained key insights from our decades of research and from sharing results with others around the world who were witnessing their own amphibian catastrophes.
@@
 The science behind that uniquely successful 15 year recovery effort is the focus of our manuscript, a story told through graphs, tables, and methodological minutiae.
-But there is another way to understand the consequences of that work, and that is to sit on the shore of any one our study lakes in which we reintroduced frogs 10 or more years ago.
+But there is another way to understand that work, and that is to sit on the shore of any one of our study lakes in which we reintroduced frogs ten or more years ago.
 The silence that settled over the lake like a morning fog following the frogs' extinction is no more.
-Instead, the lake shore is alive with frogs, one atop every rock and lining the sandy beaches.
+Instead, the lake shore is alive with frogs, atop rocks and lining the sandy beaches.
 On calm mornings, pairs of eyes bulge above the water's surface like unblinking periscopes.

### 2024-05-21 doc/manuscript/supplementary_information.qmd
Final edits prior to submission

@@
     \usepackage{blkarray}
+    \usepackage{lineno}\linenumbers
 include-before-body: 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
+|:--------------------------|:---------------|:-----------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2024-07-26 doc/manuscript/response_to_reviewers_original_submission.qmd
Initial commit of revised manuscript

---
title: "Response to Reviewer Comments"
subtitle: "MS # NCOMMS-24-29988"
format: html
editor: visual
---

Our point-by-point response to all comments from the three reviewers is provided below.
Reviewer comments are in black and our responses are in blue.

### Reviewer 1

This is a brilliant manuscript that advances our understanding and ability to mitigate the impact of the worst disease to impact vertebrates, chytridiomycosis.
The main findings are that reintroduced frogs sourced from populations of the Sierra Nevada yellow-legged frog (Rana sierrae) that are recovering from initial outbreaks and have increased resistance against chytridiomycosis are able to reestablish populations in the presence of the pathogen, Bd.
The authors also investigated factors affecting reintroduction success and found that site, cohort, and frog attributes were important.
They modelled the viability of reintroduced populations suggesting that many have a low probability of extinction over 50 years.

The work has major significance to the growing fields of chytrididomycosis, wildlife diseases, mitigation of threatening processes, conservation, reintroduction of species and related fields.
The work is original and the first to demonstrate reintroduction of resistant frogs as a viable long term solution for recovering amphibian populations from the impacts of chytridiomycosis.
The evidence presented supports the conclusions of the authors and the methodology is sound.
The figures are excellent.

Some suggestions for improvement of the manuscript:

I appreciate the arguments made in favour of frog resistance rather than site characteristics being the reason for translocation success.
Perhaps the strongest argument is that non resistant frogs were extirpated from these sites previously and would be expected to be extirpated again unless they had evolved resistance and this could be stated in the discussion.

[We added a sentence to the Discussion to make the point highlighted by the reviewer more strongly (Discussion - 2nd para, last sentence)]{style="color: blue"}

The manuscript could possibly be improved by recognising that although many species have failed to evolve resistance, this may occur over a longer time frame or alternatively, species could be assisted to evolve resistance through selective breeding programs etc (Kosch et al 2022).

[We added a sentence to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}

Similarly, the manuscript could perhaps expand on the barriers to natural recovery and spread of resistant populations and the importance of considering elevational disease gradients and connectivity when choosing reintroduction sites (Bell et al 2020).

[This in-depth topic would require the addition of a substantial amount of new text, and would be better addressed in a separate manuscript. As such, we have chosen not to expand our treatment of this topic in the current manuscript.]{style="color: blue"}

Perhaps the authors could make a point in the discussion regarding the importance of sustained effort and using adaptive management techniques.
This approach ensured they improved their ability to choose recipient sites with higher habitat quality and ultimately led to improved translocation success.

[We added a sentence to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}

Finally, perhaps the conclusions could be a little more hopeful given the study's results along with other current efforts at mitigating the threat of chytridiomycosis (Berger et al 2024, Waddle et al 2024).

[We agree with the reviewer that important progress is being made to mitigate the impacts of chytridiomycosis on the world's amphibians. Nonetheless, many challenges remain and we don't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer). In writing the final paragraph of the Discussion, we sought to provide a balanced view of the importance of our study and the many challenges that remain. We suggest that we have stuck a reasonable balance and did not make any changes in response to this comment.]{style="color: blue"}

Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation.
Extremely well done!

[We appreciate the reviewer's enthusiastic endorsement of our study.]{style="color: blue"}

### Reviewer 2

This is an extremely important paper – the authors have conducted extraordinary conservation work an amphibian species known to be highly susceptible to the lethal fungal pathogen (Batrachochytrium dendrobatidis; “Bd”).
Their efforts to reintroduce the Sierra Nevada yellow-legged frog (Rana sierrae) to locations where it had formally been extirpated using adults from naturally recovering populations is extremely promising.
To my knowledge, this paper constitutes the largest reintroduction effort to date for an amphibian challenged by Bd, both in terms of the number of reintroductions and their subsequent success.
I have few minor comments (below) and highly recommend publishing this paper.
It will certainly be highly cited and serve as a model for other amphibian systems impacted by the pathogen.
Additionally, I think the methods and associated data analyses are well explained and appropriate (see minor comments/suggestions below).

Minor comments/suggestions: (these are in no order)

1\.
Line 90-98.
It seems like the sentences that discuss the three primary objectives may not be complete sentences?
It seems like they might need another word: e.g., First, ‘we’ determined whether…Second, ‘we’ identified….
Etc.
Maybe authors mean to provide a list.
We had three primary objectives (1) determine whether...; (2) identify important site,...
(3) estimate the probability of persistence...but that is a lot to include in a single sentence, so I understand (and like) breaking it apart, but I am just not sure the sentence structure is correct - note: this is an extremely minor point.

[The sentences in question are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes them from objectives to descriptions of the three components of the study. We did not make any changes to these sentences, but are open to any suggestions to improve the structure of these sentences.]{style="color: blue"}

2\.
The map in Figure 2 is a little hard to understand.
First, the legend states that: “All populations are in Yosemite National Park (park boundary indicated by gray polygon)” – but I think the polygon is grey in the large map, but black in the inset.
I would make the park boundary black in both maps.
The legend also states: “In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).” This is a little difficult for me to see too – there is dark green in the inset corresponding the Central Valley, but in the larger map it looks like there is green on the top of the ridges, based on the perceived topography.
Maybe I’m seeing the colors incorrectly.

3\.
Methods and associated supplemental materials a.
Line 348-354.
This explanation is confusing because of the location of the '(primary period)' reference.
As written, it seems like the authors are saying that a single day is the primary period - only after you read the supplemental information do you realize that they are using a site visit as the primary period.
I think it’s good that the authors used the primary/secondary period terminology that is common for robust design CMR surveys, but I think it would be better and more correct to say:

‘During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period)....
During 2013-2022, we generally used a robust design where all habitats were searched during several consecutive days (median number....) during each visit...’

In addition to the median number of secondary periods per primary period, it would also be nice to know the mean (or median) number of visits (primary periods) per summer active season during the study.
At first, I thought that there was only 1 primary period per year, but the supplementary information clearly states: "Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September)." This would be a nice place to give the mean (or median) number of visits (primary periods) per summer active season.

b.  Population viability model description.
c.  The authors have 7 state variables which they define as ‘density’ of individuals in different life history states. Is this really density (individuals per unit area) or just the number of individuals.

ii. Additionally, are the authors modelling both males and females, or just females. Based on the model structure it looks like they are modelling females, as there is no adjustment for the proportion of females in the population. This should be clarified. If I am correct, then shouldn’t fecundity be expressed as the number of female eggs?
iii. I am also a little confused at what they consider ‘recruitment’ in this model. I understand the survival probability, indicated by sigma notation (σ), but none of the parameters listed in Supplementary Table 1 are termed recruitment. However, in both the text the term is used in multiple places, e.g., lines 106, 102, 458 and figure legends (Fig 5), but I’m sure it means different thing in these places. For this reason, I think the authors should clarify their use of the term, especially in the viability modelling. To me, recruitment is conditional on survival so I would tend to equate it with (1- 𝑝𝐽1) or the probability that given a juvenile survives it ‘recruits’ or transitions into an adult frog. I don’t think that is how the adults are using it in this model. I think they could refrain from using it at all in this section.

4.  Given the page limitations employed by the journal, the authors must make decisions about what information to supply.
    Personally, I would have liked to see a little more information on the variation in recruitment (number of new adults a population) and population growth (change in abundance over time) among recipient sites.
    These are estimated (line 106; Supplementary Material) but not reported from their Bayesian Jolly-Seber CMR model.
    Instead, the authors chose to use limited demographic estimates from reintroduced populations (adult survival only) in a population viability model to explore variation in population viability over 50 years.
    The only thing that varies across their 12 reintroduced populations in this exercise is adult survival; the other parameters are the same across populations (Supplementary Table 1).
    In essence, this exercise shows how population viability varies with adult survival.
    It’s really important to choose reintroduction sites where adult annual survival exceeds 0.50, preferably 0.60, and the authors make a case that survival 1-year post-translocation is a good proxy for habitat quality; however, part of that case is assuming that all other demographic parameters are the same among populations, which is probably not true.
    The elasticity and sensitivity analyses associated with the viability analyses do suggest that adult survival is certainly the most important parameter (almost always true in vertebrate systems).
    Still, the decision of what to present given limited space is up to the authors and just because it does not match what this one reviewer would have preferred does not detract from the quality and quantity of work demonstrated in this paper.
    This is a really good paper and definitely worthy of publication in Nature Communications.

5.  Lines 221-233.
    This is very true – congratulations!!

6.  Line 252.
    I suggest revising this sentence; frog survival at natal (donor) sites was never reported.
    If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.

### Reviewer 3

General comments: This is a long-term dataset, in many ways exceptionally long for in depth amphibian studies, and a valuable manuscript on a priority topic for biodiversity conservation.
Bd research has been criticised for missing direct conservation applications in the field, which makes this study particularly novel and insightful, despite focusing on a single species.
The results are very encouraging and give much cause of cautious optimism for what has often appeared to be an almost intractable problem for amphibian conservation but while I understand the specificity of your results it would be good to have at least a paragraph that looks at the potential implications for other species and important steps forward that could be concluded from your own work.
In my opinion the text is clear and very well-written and I do not have much to add in relation to that.
There is some repetition between introduction and discussion so please be aware of that.
As a general comment, you mention that the three donor populations are large, with several hundred adults and that you have documented the genetic structure but I suggest it would be useful to briefly discuss the potential implications of the reintroductions from a genetic perspective and any potential monitoring follow-up of that aspect and mitigation strategy if required.
There is a reasonable potential for creating a genetic bottleneck as a results of these reintroductions in the absence of some kind of genetic monitoring.
Also, it would be useful to briefly explain why are reintroductions important (and cost-effective) instead of simply waiting for these recovering populations to naturally expand over time.
The methods look sound but I have limited experience in this type of multilevel Bayesian modelling.
Detailed comments Title- Just a suggestion but for me the title would benefit from mentioning the species.
Your work obviously has wider implications but remains currently very species-specific.
Line 155- for me this aspect of the apparent high variability linked to donor population is important and I suggest should be highlighted both in the results and especially in the discussion as a separate point.
Would the reintroductions be feasible in terms of achieving long term population viability with the lowest survival based on donor population?
Line 266- I suggest more caution in this statement and perhaps combining with the following sentence, which seems a reasonable explanation but remains based on a relatively small number of sites and reintroduction events.
I would strongly expect specific weather patterns to also play an important role in any given year, even if these might be rarer events that you have not yet captured.
Line 288- for me the value in this call for additional research warrants some brief but specific details about what that would entail.
Otherwise, researchers naturally always call for additional research.
Lines 298-301- I think this is important and relevant in the wider amphibian conservation context, where juveniles at this crucial life stage are often the specifically missing dataset because they are harder to monitor.
This is in many ways a failure of the sector and the opposite compared to say birds or mammals, yet has become inbuilt in much amphibian research and is worth highlighting as it has important consequences for progressing practical conservation.
Lines 310-315 I suggest you could be more forward looking here and suggest some critical steps forward for interpreting the success of your results for advancing the conservation prospects for other similarly threatened amphibian species.
What are the crucial considerations that should inform their strategies and how would you go about it?
It might also be useful to also briefly mention the relevance of your results for reintroductions from the perspective of captive breeding and amphibian arks given that the current approaches would not facilitate any kind of resistance building (if at all possible).
Line 317- why would it though?
Now that you have developed a solid understanding of the process, could you not strategically reintroduce populations to facilitate ongoing natural colonisation once there are sufficient numbers to create some wider population resilience?
Line 337- this is an important methodological point and which requires a bit more substance in terms of how was absence determined, over what period, etc

### 2024-07-26 doc/manuscript/translocation.qmd
Final edits prior to submission
Initial commit of revised manuscript

@@
 Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
-The emergence of the lethal fungal pathogen *Batrachochytrium dendrobatidis* ("Bd") has devastated global amphibian biodiversity.
+The emergence of the lethal fungal pathogen *Batrachochytrium dendrobatidis* (Bd) has devastated global amphibian biodiversity.
 Given the lack of any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
@@
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies (see Figure 4 in [@knapp2016]), and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
+Finally, reintroductions of frogs collected from Bd-naive populations and reintroduced into Bd-positive sites have consistently failed due to the development of high Bd loads and resulting high frog mortality, suggesting the importance of frog resistance in population reestablishment in the presence of Bd.
 
-Reintroduction of resistant *R. sierrae* was remarkably successful in reestablishing populations in the presence of Bd.
+In the current study, reintroduction of resistant *R. sierrae* was remarkably successful in reestablishing populations in the presence of Bd.
 Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
@@
 sierrae* despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
+An adaptive management approach such as the one we utilized, in which the outcomes of recovery actions are rigorously assessed and this information is used to reduce uncertainties and inform future actions, will be essential in maximizing the effectiveness of this global recovery effort \[@canessa2019\].
+In addition, developing an improved understanding of the factors facilitating or limiting the development of frog resistance against Bd infection and the effectiveness of selective breeding programs and other genetic interventions to enhance frog resistance \[@kosch2022\] may be particularly important.
 Although we now have a proven strategy to reestablish extirpated *R. sierrae* populations, recovery across their large historical range will require substantial resources over many decades.

### 2024-07-26 doc/manuscript/response_to_reviewers_original_submission.qmd
Minor formatting edits

@@
 3\.
-Methods and associated supplemental materials a.
+Methods and associated supplemental materials
+
+a.
 Line 348-354.
@@
 b.  Population viability model description.
-c.
-The authors have 7 state variables which they define as ‘density’ of individuals in different life history states.
+
+-   The authors have 7 state variables which they define as ‘density’ of individuals in different life history states.
     Is this really density (individuals per unit area) or just the number of individuals.
 
-ii.
-Additionally, are the authors modelling both males and females, or just females.
+-   Additionally, are the authors modelling both males and females, or just females.
     Based on the model structure it looks like they are modelling females, as there is no adjustment for the proportion of females in the population.
@@
     If I am correct, then shouldn’t fecundity be expressed as the number of female eggs?
-iii.
-I am also a little confused at what they consider ‘recruitment’ in this model.
+
+-   I am also a little confused at what they consider ‘recruitment’ in this model.
     I understand the survival probability, indicated by sigma notation (σ), but none of the parameters listed in Supplementary Table 1 are termed recruitment.

### 2024-07-28 doc/manuscript/response_to_reviewers_original_submission.qmd
Format Reviewer 3 comments

@@
 
-Our point-by-point response to all comments from the three reviewers is provided below.
-Reviewer comments are in black and our responses are in blue.
+Our point-by-point response to all comments from the three reviewers is provided below. Reviewer comments are in black and our responses are in blue.
 
@@
 
-Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation.
-Extremely well done!
+Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation. Extremely well done!
 
@@
 
-3\.
-Methods and associated supplemental materials
+3\. Methods and associated supplemental materials
 
@@
 
-4.  Given the page limitations employed by the journal, the authors must make decisions about what information to supply.
+4.
+Given the page limitations employed by the journal, the authors must make decisions about what information to supply.
 Personally, I would have liked to see a little more information on the variation in recruitment (number of new adults a population) and population growth (change in abundance over time) among recipient sites.
@@
 
-5.  Lines 221-233.
-    This is very true – congratulations!!
+5.  Lines 221-233. This is very true – congratulations!!
 
-6.  Line 252.
+6.
+Line 252.
 I suggest revising this sentence; frog survival at natal (donor) sites was never reported.
-    If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.
+If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig.
+3) I would clarify that.
 
@@
 Bd research has been criticised for missing direct conservation applications in the field, which makes this study particularly novel and insightful, despite focusing on a single species.
-The results are very encouraging and give much cause of cautious optimism for what has often appeared to be an almost intractable problem for amphibian conservation but while I understand the specificity of your results it would be good to have at least a paragraph that looks at the potential implications for other species and important steps forward that could be concluded from your own work.
-In my opinion the text is clear and very well-written and I do not have much to add in relation to that.
-There is some repetition between introduction and discussion so please be aware of that.
-As a general comment, you mention that the three donor populations are large, with several hundred adults and that you have documented the genetic structure but I suggest it would be useful to briefly discuss the potential implications of the reintroductions from a genetic perspective and any potential monitoring follow-up of that aspect and mitigation strategy if required.
+
+-   The results are very encouraging and give much cause of cautious optimism for what has often appeared to be an almost intractable problem for amphibian conservation but while I understand the specificity of your results it would be good to have at least a paragraph that looks at the potential implications for other species and important steps forward that could be concluded from your own work.
+
+-   In my opinion the text is clear and very well-written and I do not have much to add in relation to that.
+
+-   There is some repetition between introduction and discussion so please be aware of that.
+
+-   As a general comment, you mention that the three donor populations are large, with several hundred adults and that you have documented the genetic structure but I suggest it would be useful to briefly discuss the potential implications of the reintroductions from a genetic perspective and any potential monitoring follow-up of that aspect and mitigation strategy if required.
 There is a reasonable potential for creating a genetic bottleneck as a results of these reintroductions in the absence of some kind of genetic monitoring.
-Also, it would be useful to briefly explain why are reintroductions important (and cost-effective) instead of simply waiting for these recovering populations to naturally expand over time.
-The methods look sound but I have limited experience in this type of multilevel Bayesian modelling.
-Detailed comments Title- Just a suggestion but for me the title would benefit from mentioning the species.
-Your work obviously has wider implications but remains currently very species-specific.
-Line 155- for me this aspect of the apparent high variability linked to donor population is important and I suggest should be highlighted both in the results and especially in the discussion as a separate point.
+
+-   Also, it would be useful to briefly explain why are reintroductions important (and cost-effective) instead of simply waiting for these recovering populations to naturally expand over time.
+
+-   The methods look sound but I have limited experience in this type of multilevel Bayesian modelling.
+
+Detailed comments
+
+-   Title- Just a suggestion but for me the title would benefit from mentioning the species. Your work obviously has wider implications but remains currently very species-specific.
+
+-   Line 155- for me this aspect of the apparent high variability linked to donor population is important and I suggest should be highlighted both in the results and especially in the discussion as a separate point.
 Would the reintroductions be feasible in terms of achieving long term population viability with the lowest survival based on donor population?
-Line 266- I suggest more caution in this statement and perhaps combining with the following sentence, which seems a reasonable explanation but remains based on a relatively small number of sites and reintroduction events.
+
+-   Line 266- I suggest more caution in this statement and perhaps combining with the following sentence, which seems a reasonable explanation but remains based on a relatively small number of sites and reintroduction events.
 I would strongly expect specific weather patterns to also play an important role in any given year, even if these might be rarer events that you have not yet captured.
-Line 288- for me the value in this call for additional research warrants some brief but specific details about what that would entail.
+
+-   Line 288- for me the value in this call for additional research warrants some brief but specific details about what that would entail.
 Otherwise, researchers naturally always call for additional research.
-Lines 298-301- I think this is important and relevant in the wider amphibian conservation context, where juveniles at this crucial life stage are often the specifically missing dataset because they are harder to monitor.
+
+-   Lines 298-301- I think this is important and relevant in the wider amphibian conservation context, where juveniles at this crucial life stage are often the specifically missing dataset because they are harder to monitor.
 This is in many ways a failure of the sector and the opposite compared to say birds or mammals, yet has become inbuilt in much amphibian research and is worth highlighting as it has important consequences for progressing practical conservation.
-Lines 310-315 I suggest you could be more forward looking here and suggest some critical steps forward for interpreting the success of your results for advancing the conservation prospects for other similarly threatened amphibian species.
+
+-   Lines 310-315 I suggest you could be more forward looking here and suggest some critical steps forward for interpreting the success of your results for advancing the conservation prospects for other similarly threatened amphibian species.
 What are the crucial considerations that should inform their strategies and how would you go about it?
 It might also be useful to also briefly mention the relevance of your results for reintroductions from the perspective of captive breeding and amphibian arks given that the current approaches would not facilitate any kind of resistance building (if at all possible).
-Line 317- why would it though?
+
+-   Line 317- why would it though?
 Now that you have developed a solid understanding of the process, could you not strategically reintroduce populations to facilitate ongoing natural colonisation once there are sufficient numbers to create some wider population resilience?
-Line 337- this is an important methodological point and which requires a bit more substance in terms of how was absence determined, over what period, etc
+
+-   Line 337- this is an important methodological point and which requires a bit more substance in terms of how was absence determined, over what period, etc

### 2024-07-29 doc/manuscript/response_to_reviewers_original_submission.qmd
First pass on reviewer 2 comments

--- 2024-07-26 doc/manuscript/response_to_reviewers_original_submission.qmd
@@
 
-This is an extremely important paper – the authors have conducted extraordinary conservation work an amphibian species known to be highly susceptible to the lethal fungal pathogen (Batrachochytrium dendrobatidis; “Bd”).
+This is an extremely important paper -- the authors have conducted extraordinary conservation work an amphibian species known to be highly susceptible to the lethal fungal pathogen (Batrachochytrium dendrobatidis; "Bd").
 Their efforts to reintroduce the Sierra Nevada yellow-legged frog (Rana sierrae) to locations where it had formally been extirpated using adults from naturally recovering populations is extremely promising.
@@
 It seems like the sentences that discuss the three primary objectives may not be complete sentences?
-It seems like they might need another word: e.g., First, ‘we’ determined whether…Second, ‘we’ identified….
+It seems like they might need another word: e.g., First, 'we' determined whether...Second, 'we' identified....
 Etc.
@@
 The map in Figure 2 is a little hard to understand.
-First, the legend states that: “All populations are in Yosemite National Park (park boundary indicated by gray polygon)” – but I think the polygon is grey in the large map, but black in the inset.
+First, the legend states that: "All populations are in Yosemite National Park (park boundary indicated by gray polygon)" -- but I think the polygon is grey in the large map, but black in the inset.
 I would make the park boundary black in both maps.
-The legend also states: “In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).” This is a little difficult for me to see too – there is dark green in the inset corresponding the Central Valley, but in the larger map it looks like there is green on the top of the ridges, based on the perceived topography.
-Maybe I’m seeing the colors incorrectly.
+The legend also states: "In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation)." This is a little difficult for me to see too -- there is dark green in the inset corresponding the Central Valley, but in the larger map it looks like there is green on the top of the ridges, based on the perceived topography.
+Maybe I'm seeing the colors incorrectly.
 
@@
 As written, it seems like the authors are saying that a single day is the primary period - only after you read the supplemental information do you realize that they are using a site visit as the primary period.
-I think it’s good that the authors used the primary/secondary period terminology that is common for robust design CMR surveys, but I think it would be better and more correct to say:
+I think it's good that the authors used the primary/secondary period terminology that is common for robust design CMR surveys, but I think it would be better and more correct to say:
 
-‘During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period)....
-During 2013-2022, we generally used a robust design where all habitats were searched during several consecutive days (median number....) during each visit...’
+'During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period)....
+During 2013-2022, we generally used a robust design where all habitats were searched during several consecutive days (median number....) during each visit...'
 
@@
 
--   The authors have 7 state variables which they define as ‘density’ of individuals in different life history states.
+-   The authors have 7 state variables which they define as 'density' of individuals in different life history states.
     Is this really density (individuals per unit area) or just the number of individuals.
+
+    -   [Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area.]{style="color: blue"}
 
@@
     This should be clarified.
-    If I am correct, then shouldn’t fecundity be expressed as the number of female eggs?
+    If I am correct, then shouldn't fecundity be expressed as the number of female eggs?
 
--   I am also a little confused at what they consider ‘recruitment’ in this model.
+    -   [Thank you for the careful reading of our model.
+We are modeling the entire population, not just females, and are assuming a 50:50 sex ratio.
+This was not explicit in our equations and we have thus updated our viability model equation to explicitly include a parameter for sex ratio ]{style="color: blue"}($\rho$).
+While addressing this comment we noticed that we were not consistent with modeling females vs.
+total population between the deterministic and stochastic models.
+We have adjusted the deterministic model to be exactly analogous to the stochastic model (i.e., both modeling total population).
+We updated the deterministic model results to reflect this change.
+While predicted median deterministic growth rates were lower with this change, it did not change any of our conclusions regarding deterministic viability.
+
+-   I am also a little confused at what they consider 'recruitment' in this model.
     I understand the survival probability, indicated by sigma notation (σ), but none of the parameters listed in Supplementary Table 1 are termed recruitment.
-    However, in both the text the term is used in multiple places, e.g., lines 106, 102, 458 and figure legends (Fig 5), but I’m sure it means different thing in these places.
+    However, in both the text the term is used in multiple places, e.g., lines 106, 102, 458 and figure legends (Fig 5), but I'm sure it means different thing in these places.
     For this reason, I think the authors should clarify their use of the term, especially in the viability modelling.
-    To me, recruitment is conditional on survival so I would tend to equate it with (1- 𝑝𝐽1) or the probability that given a juvenile survives it ‘recruits’ or transitions into an adult frog.
-    I don’t think that is how the adults are using it in this model.
+    To me, recruitment is conditional on survival so I would tend to equate it with (1- 𝑝𝐽1) or the probability that given a juvenile survives it 'recruits' or transitions into an adult frog.
+    I don't think that is how the adults are using it in this model.
     I think they could refrain from using it at all in this section.
+
+    -   Good question.
+        This was a point of some discussion within our group.
+        For a juvenile to recruit to an adult, it has to survive, and thus $\sigma_{J_1}$ encompasses survival and recruitment.
+        This is why we used the language of "survival and recruitment" to describe this parameter in the text.
+        However, we see how this could potentially be confusing.
+        As we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text.
 
@@
     In essence, this exercise shows how population viability varies with adult survival.
-    It’s really important to choose reintroduction sites where adult annual survival exceeds 0.50, preferably 0.60, and the authors make a case that survival 1-year post-translocation is a good proxy for habitat quality; however, part of that case is assuming that all other demographic parameters are the same among populations, which is probably not true.
+    It's really important to choose reintroduction sites where adult annual survival exceeds 0.50, preferably 0.60, and the authors make a case that survival 1-year post-translocation is a good proxy for habitat quality; however, part of that case is assuming that all other demographic parameters are the same among populations, which is probably not true.
     The elasticity and sensitivity analyses associated with the viability analyses do suggest that adult survival is certainly the most important parameter (almost always true in vertebrate systems).
@@
 
+    -   We appreciate the reviewer's thoughtful point.
+        It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$ which our sensitivity analyses indicate is likely the second most important parameter for viability following adult survival.
+        So we are actually exploring viability across environmental conditions that affect the two most important parameters for population growth.
+        While we certainly agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameter we explore are likely the most important for long-term viability.
+        As no changes were requested we, we did not make any.
+
 5.  Lines 221-233.
-    This is very true – congratulations!!
+    This is very true -- congratulations!!
+
+    Thanks
 
@@
     If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.
+
+    \[Not sure what lines they are referring to\]
 

### 2024-07-29 doc/manuscript/supplementary_information.qmd
First pass on reviewer 2 comments

@@
 However, within a population there was substantial yearly variability in the successful recruitment of adults, greater than what we would expect from Poisson variability around a mean value.
-Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival and recruitment (additional details provided in **Estimating model parameters** below).
-We also could have included environmental stochasticity in year-2 juvenile survival and recruitment $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp).
+Therefore, we allowed for yearly variability in the probability of year-1 juvenile survival (additional details provided in **Estimating model parameters** below).
+We also could have included environmental stochasticity in year-2 juvenile survival probability $\sigma_{J_2}$, but our elasticity and sensitivity analysis (see below, **Model analysis and simulation**) showed that this parameter had little effect on host growth rate relative to $\sigma_{J_1}$ (@fig-viability-supp).
 
@@
 We performed four analyses on our model.
-First, we considered a deterministic version of our model with no yearly heterogeneity in year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
-We then fixed $\sigma_{J_1} = 0.09$ and calculated the predicted growth rate of our 12 populations.
+First, we considered a deterministic version of our model with no yearly heterogeneity in year-1 juvenile survival probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
+We then fixed $\sigma_{J_1} = 0.1-$ and calculated the predicted growth rate of our 12 populations.
 
@@
 ![**Sensitivity analysis of the stage-structured *R.
-sierrae* model.** **A** Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival and recruitment of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+sierrae* model.** **A** Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
 **B** Sensitivity of $\lambda$ with changes in the same nine parameters.
@@
 We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study ([@tbl-param_values]) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity is calculated at the default parameter values for each population and $\sigma_{J_1}$ = 0.09.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
+Elasticity and sensitivity are calculated at the default parameter values for each population given in Table [@tbl-param_values].](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------|:---------------|:-----------------------------|
+|:-------------------------|:------------------|:---------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
@@
 | $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles           | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
+| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles.          | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
 | $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
@@
 | $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
-| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                                |
+| $p_F$, Probability of an adult female reproducing in a year               | 0.5                  | Could be as high at 1, based on field observations                                |
+| $\rho$, Proportion of adults that are female                              | 0.5                  | SOURCE                                                                            |
 | $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                                |

### 2024-07-29 code/extinction_and_growth_rate_analysis.R

@@ -75,7 +75,7 @@ for(l in 1:nrow(all_surv)){
 		if(nm == "med")
 			med_surv_vals[l] = muR
 
-		params = assign_parameters(muR, muT, new_params=list(sigma_J1=0.09))
+		params = assign_parameters(muR, muT, new_params=list(sigma_J1=0.1)) 
 		tlam = get_lambda(params)
 		lambdas_mlu[j] = tlam
 	}
@@ -120,10 +120,10 @@ for(i in 1:length(muR_range)){
 dat = expand.grid(muR=muR_range, juv_surv=juv_surv_range)
 dat$lambda = as.vector(lambdas)
 ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) + 
-							geom_hline(aes(yintercept=0.09), linetype="dashed") +
+							geom_hline(aes(yintercept=0.1), linetype="dashed") +
 							geom_contour(aes(x=muR, y=juv_surv, z=lambda), breaks=c(1), color="red", linewidth=1) +
 							geom_point(data=lake_growth_rates_dt, aes(x=surv_med, 
-																												y=rep(0.09, length(surv_med)), 
+																												y=rep(0.1, length(surv_med)), 
 																													color=lake_id), size=2) +
 		   				scale_color_manual(values=colors) +
 							# scale_fill_gradient("\u03bb", low = "white", high = "black") +
@@ -161,7 +161,7 @@ for(l in 1:nrow(all_surv)){
 		muR = recruited_surv[lake_id == lake][[paste0('surv_prob_', 'med')]]
 	}
 
-	params = assign_parameters(muR, muT)
+	params = assign_parameters(muR, muT, new_params=list(sigma_J1=0.25))
 	sens_elas = get_sensitivity(params)
 
 	lake_sens[[paste0(lake)]] = sens_elas$sens

### 2024-07-29 code/population_viability_functions.R

@@ -29,7 +29,8 @@ assign_parameters = function(adult_survR, adult_survT, new_params=list()){
 			  sigma_AR=adult_survR,
 			  sigma_AT=adult_survT,
 			  p_L1=1, p_L2=0.25, p_J1=0.25,
-			  p_F=0.5, F=100, 
+			  p_F=0.5, F=100,
+			  rho=0.5,
 			  reproduction="poisson")
 
 	# Set any new parameters
@@ -168,22 +169,21 @@ stochastic_step = function(state_vars, p){
 
 	# Account for overflow
 	if(state_vars[(n - 1)] < 10000){
-		reproducing_adultsR = rbinom(1, state_vars[(n - 1)], p$sigma_AR*p$p_F)
-		reproducing_adultsT = rbinom(1, state_vars[n], p$sigma_AT*p$p_F)
+		reproducing_adultsR = rbinom(1, state_vars[(n - 1)], p$sigma_AR*p$p_F*p$rho)
+		reproducing_adultsT = rbinom(1, state_vars[n], p$sigma_AT*p$p_F*p$rho)
 
-		# Only females reproduce so divide by 2
 		if(p$reproduction == "nbd"){
-			new_larvaeR = rnbinom(1, mu=(reproducing_adultsR / 2)*p$F, size=1)
-			new_larvaeT = rnbinom(1, mu=(reproducing_adultsT / 2)*p$F, size=1)
+			new_larvaeR = rnbinom(1, mu=(reproducing_adultsR) * p$F, size=1)
+			new_larvaeT = rnbinom(1, mu=(reproducing_adultsT) * p$F, size=1)
 		} else{
-			new_larvaeR = rpois(1, (reproducing_adultsR) / 2 * p$F)
-			new_larvaeT = rpois(1, (reproducing_adultsT) / 2 * p$F)
+			new_larvaeR = rpois(1, (reproducing_adultsR) * p$F)
+			new_larvaeT = rpois(1, (reproducing_adultsT) * p$F)
 		}
 	} else{
 
[…]

### 2024-07-29 doc/manuscript/response_to_reviewers_original_submission.qmd
Address Reviewer 3 comment

@@
 We are modeling the entire population, not just females, and are assuming a 50:50 sex ratio.
-This was not explicit in our equations and we have thus updated our viability model equation to explicitly include a parameter for sex ratio ]{style="color: blue"}($\rho$).
-While addressing this comment we noticed that we were not consistent with modeling females vs.
+This was not explicit in our equations and we have thus updated our viability model equation to explicitly include a parameter for sex ratio ($\rho$).
+We now describe this parameter in the Table in the supplementary material.
+While addressing this comment, we noticed that we were not consistent with modeling females vs.
 total population between the deterministic and stochastic models.
-We have adjusted the deterministic model to be exactly analogous to the stochastic model (i.e., both modeling total population).
-We updated the deterministic model results to reflect this change.
-While predicted median deterministic growth rates were lower with this change, it did not change any of our conclusions regarding deterministic viability.
+We have updated the deterministic model to be exactly analogous to the stochastic model (i.e., both are now modeling total population).
+Our conclusions regarding viability are unchanged.]{style="color: blue"}
 
@@
 
-    -   Good question.
+    -   [Good question.
 This was a point of some discussion within our group.
@@
 This is why we used the language of "survival and recruitment" to describe this parameter in the text.
-        However, we see how this could potentially be confusing.
-        As we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text.
+However, we see how this could be confusing.
+As we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text as suggested by the reviewer.]{style="color: blue"}
 
@@
 
-    -   We appreciate the reviewer's thoughtful point.
-        It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$ which our sensitivity analyses indicate is likely the second most important parameter for viability following adult survival.
+    -   [We appreciate the reviewer's thoughtful point.
+It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$ --- a parameter that our sensitivity analyses indicate is likely the second most important parameter for viability following adult survival.
 So we are actually exploring viability across environmental conditions that affect the two most important parameters for population growth.
-        While we certainly agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameter we explore are likely the most important for long-term viability.
-        As no changes were requested we, we did not make any.
+While we agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameters we explore are the most important for long-term viability.
+As no changes were requested we, we did not make any.]{style="color: blue"}
 
@@
 
-    Thanks
+    [Thanks]{style="color: blue"}
 
@@
 
-    \[Not sure what lines they are referring to\]
+    [ROLAND: I am not sure what exactly this is referring to as I don't have the line numbers for the submission.]{style="color: blue"}
 
@@
 The methods look sound but I have limited experience in this type of multilevel Bayesian modelling.
+
 Detailed comments Title- Just a suggestion but for me the title would benefit from mentioning the species.
@@
 Would the reintroductions be feasible in terms of achieving long term population viability with the lowest survival based on donor population?
+
+-   [Good question.
+Our population viability analyses show that long-term viability is significantly more likely when 1-year adult survival probability is \>0.5.
+We hesitate to try to separate the effect of recipient site and donor population on 1-year adult survival as these factors were not crossed in our study (an inevitable logistical constraint when working on real landscapes and taking into account host landscape genetics \[ROLAND, update as needed\]).
+That being said, there is strong support from the model that the lowest 1-year survival from donor population 72996 would not be sufficient for long-term viability (\< 0.5) while the lowest 1-year adult survival from 70459 and 70567 would be sufficient for long-term viability (\> 0.5).
+While it is tempting to conclude that donor populations 70459 and 70567 should be the focus of subsequent translocations, our design cannot disentangle site and donor effects.]{style="color: blue"}
+
 Line 266- I suggest more caution in this statement and perhaps combining with the following sentence, which seems a reasonable explanation but remains based on a relatively small number of sites and reintroduction events.

### 2024-07-29 doc/manuscript/supplementary_information.qmd

@@
 | $p_F$, Probability of an adult female reproducing in a year               | 0.5                  | Could be as high at 1, based on field observations                                |
-| $\rho$, Proportion of adults that are female                              | 0.5                  | SOURCE                                                                            |
+| $\rho$, Proportion of adults that are female                              | 0.5                  | Estimated from field data, observations, natural history knowledge                |
 | $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                                |

### 2024-07-29 doc/manuscript/translocation.qmd

@@
 
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.10, at least six of twelve translocated populations are likely to experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]A; median predicted $\lambda$ ranges from 1.07-1.28 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 ([@fig-viability]A).
+As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of seven of twelve populations was greater than 1 ([@fig-viability]A).
 
-Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
+Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
 Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 ([@fig-viability]B).
-This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
+This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in juvenile survival.
 In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
@@
 sierrae* despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
-An adaptive management approach such as the one we utilized, in which the outcomes of recovery actions are rigorously assessed and this information is used to reduce uncertainties and inform future actions, will be essential in maximizing the effectiveness of this global recovery effort \[@canessa2019\].
-In addition, developing an improved understanding of the factors facilitating or limiting the development of frog resistance against Bd infection and the effectiveness of selective breeding programs and other genetic interventions to enhance frog resistance \[@kosch2022\] may be particularly important.
+An adaptive management approach such as the one we utilized, in which the outcomes of recovery actions are rigorously assessed and this information is used to reduce uncertainties and inform future actions, will be essential in maximizing the effectiveness of this global recovery effort [@canessa2019].
+In addition, developing an improved understanding of the factors facilitating or limiting the development of frog resistance against Bd infection and the effectiveness of selective breeding programs and other genetic interventions to enhance frog resistance [@kosch2022] may be particularly important.
 Although we now have a proven strategy to reestablish extirpated *R. sierrae* populations, recovery across their large historical range will require substantial resources over many decades.
@@
 Our central question was: How does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations?
-We developed a model that tracked seven state variables of a frog population: density of translocated adults ($A_T$), density of adults naturally recruited into the population ($A_R$), density of first-year tadpoles ($L_1$), density of second-year tadpoles ($L_2$), density of third-year tadpoles ($L_3$), density of first-year juveniles ($J_1$), and density of second-year juveniles ($J_2$).
+We developed a model that tracked seven state variables of a frog population: abundance of translocated adults ($A_T$), abundance of adults naturally recruited into the population ($A_R$), abundance of first-year tadpoles ($L_1$), abundance of second-year tadpoles ($L_2$), abundance of third-year tadpoles ($L_3$), abundance of first-year juveniles ($J_1$), and abundance of second-year juveniles ($J_2$).
 We divided adults into two classes $A_T$ and $A_R$ because there is evidence that the survival probability of translocated adults and naturally recruited adults differs (Supplementary Fig. 3).
+The modeled population included both male and female frogs.
 
@@
 \begin{bmatrix}
-  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
+  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F \rho F & \sigma_{A_T} p_F \rho F \\
   \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
-First-year juvenile survival and recruitment $\sigma_{J1}$ is the parameter that we think is most influenced by environmental stochasticity.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year \$p_F\$, the percentage of adults that are female $\rho$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
+First-year juvenile survival $\sigma_{J1}$ is the parameter that we think is most influenced by environmental stochasticity.
 
@@
 Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (Supplemental Fig. 4).
-This also helped us determine where in the model environmental variation in juvenile survival and recruitment would have the largest effects on population dynamics.
+This also helped us determine where in the model environmental variation in juvenile survival would have the largest effects on population dynamics.
 Third, we included demographic stochasticity and environmental stochasticity in $\sigma_{J_1}$ in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
@@
 ![**Estimated growth rates and extinction probabilities of reintroduced populations**.
-**A** Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival and recruitment probability $\sigma_{J_1}$, given the parameterized, deterministic model.
-Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.09$ (indicated by the dashed line).
+**A** Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.10$ (indicated by the dashed line).
 The red line shows where $\lambda = 1$.

### 2024-07-31 doc/manuscript/response_to_reviewers_original_submission.qmd
Additional edits

@@
 In writing the final paragraph of the Discussion, we sought to provide a balanced view of the importance of our study and the many challenges that remain.
-We suggest that we have stuck a reasonable balance and did not make any changes in response to this comment.]{style="color: blue"}
+We suggest that we have stuck a reasonable balance and therefore we did not make any changes in response to this comment.]{style="color: blue"}
 
-Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation.
-Extremely well done!
+Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation. Extremely well done!
 
@@
 [The sentences in question are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes them from objectives to descriptions of the three components of the study.
-We did not make any changes to these sentences, but are open to any suggestions to improve the structure of these sentences.]{style="color: blue"}
+We did not make any changes in response to this comment, but we are open to any suggestions to improve the structure of these sentences.]{style="color: blue"}
 
@@
 
-5.  Lines 221-233.
-    This is very true -- congratulations!!
+5.  Lines 221-233. This is very true -- congratulations!!
 
@@
 
-6.  Line 252.
+6.
+Line 252.
 I suggest revising this sentence; frog survival at natal (donor) sites was never reported.
-    If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.
+If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig.
+3) I would clarify that.
 
-    [ROLAND: I am not sure what exactly this is referring to as I don't have the line numbers for the submission.]{style="color: blue"}
+    [As the reviewer suggested, we added a reference to Supplementary Fig. 3]{style="color: blue"}
 

### 2024-07-31 doc/manuscript/translocation.qmd
Additional edits

@@
 Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and reintroduced populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the reintroduced versus donor populations and pathogen pressure should follow a similar pattern.
-In addition to the maintenance of Bd load and frog survival between natal and recipient sites, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
+In addition to the maintenance of Bd load and frog survival between natal and recipient sites (Supplementary Fig.
+3), the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance has a genetic basis.
 
@@
 VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, we conducted CMR surveys on a single day (primary period) per site visit, during which we searched all habitats repeatedly for adult frogs.
+During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period) , during which we searched all habitats repeatedly for adult frogs.
 Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.

### 2024-08-02 code/maps.qmd
Change boundary color

@@
   tm_shape(yose_boundary) + 
-  tm_borders(col = "#737373", lwd = 2) +
+  tm_borders(col = "black", lwd = 2) +
   tm_shape(translocation_points) + 

### 2024-08-02 doc/manuscript/response_to_reviewers_original_submission.qmd
More edits in response to reviewer comments

@@
 
-Our point-by-point response to all comments from the three reviewers is provided below. Reviewer comments are in black and our responses are in blue.
+Our point-by-point response to all comments from the three reviewers is provided below.
+Reviewer comments are in black and our responses are in blue.
 
@@
 
+1.
 I appreciate the arguments made in favour of frog resistance rather than site characteristics being the reason for translocation success.
@@
 
+2.
 The manuscript could possibly be improved by recognising that although many species have failed to evolve resistance, this may occur over a longer time frame or alternatively, species could be assisted to evolve resistance through selective breeding programs etc (Kosch et al 2022).
@@
 
+3.
 Similarly, the manuscript could perhaps expand on the barriers to natural recovery and spread of resistant populations and the importance of considering elevational disease gradients and connectivity when choosing reintroduction sites (Bell et al 2020).
@@
 
+4.
 Perhaps the authors could make a point in the discussion regarding the importance of sustained effort and using adaptive management techniques.
@@
 
-[We added a sentence to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}
+[We added additional sentences to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}
 
+5.
 Finally, perhaps the conclusions could be a little more hopeful given the study's results along with other current efforts at mitigating the threat of chytridiomycosis (Berger et al 2024, Waddle et al 2024).
@@
 [We agree with the reviewer that important progress is being made to mitigate the impacts of chytridiomycosis on the world's amphibians.
-Nonetheless, many challenges remain and we don't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
-In writing the final paragraph of the Discussion, we sought to provide a balanced view of the importance of our study and the many challenges that remain.
-We suggest that we have stuck a reasonable balance and therefore we did not make any changes in response to this comment.]{style="color: blue"}
+Nonetheless, many challenges remain and in writing the final paragraph of the Discussion we didn't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
+However, we agree that the final paragraph could provide a bit more hopeful outlook.
+We made several changes to this paragraph to accomplish that.]{style="color: blue"}
 
-Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation. Extremely well done!
+Congratulations to the authors on a very impressive body of work and most importantly, wonderful results for amphibian conservation.
+Extremely well done!
 
@@
 
-3\. Methods and associated supplemental materials
+[Thank you for pointing out these difficulties in interpreting the map.
+In the revised version, the polygon indicating the park boundary in the main map and the inset map are both displayed in black.
+In addition, we added clarifying text to the figure legend to describe the colors and their meaning.]{style="color: blue"}
+
+3\.
+Methods and associated supplemental materials
 
@@
 
+[We updated this sentence as the reviewer suggested.]{style="color: blue"}
+
 In addition to the median number of secondary periods per primary period, it would also be nice to know the mean (or median) number of visits (primary periods) per summer active season during the study.
 At first, I thought that there was only 1 primary period per year, but the supplementary information clearly states: "Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September)." This would be a nice place to give the mean (or median) number of visits (primary periods) per summer active season.
+
+[In the reviewed version of the manuscript, we actually included the median, minimum, and maximum number of visits to each translocated population per summer active season (lines 345-346).
+However, it may not have been clear that a "visit" is equivalent to a "primary period".
+Therefore, we added some additional text to the sentence in question (revised manuscript: lines xxx-xxx) to clarify this point.]{style="color: blue"}
 
@@
 
-    -   [Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area.]{style="color: blue"}
+[Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area.]{style="color: blue"}
 
@@
 
-    -   [Thank you for the careful reading of our model.
+[Thank you for the careful reading of our model.
 We are modeling the entire population, not just females, and are assuming a 50:50 sex ratio.
@@
 
-    -   [Good question.
+[Good question.
 This was a point of some discussion within our group.
@@
 
-4.
-Given the page limitations employed by the journal, the authors must make decisions about what information to supply.
+-   Given the page limitations employed by the journal, the authors must make decisions about what information to supply.
 Personally, I would have liked to see a little more information on the variation in recruitment (number of new adults a population) and population growth (change in abundance over time) among recipient sites.
@@
 
-    -   [We appreciate the reviewer's thoughtful point.
+[We appreciate the reviewer's thoughtful point.
 It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$ --- a parameter that our sensitivity analyses indicate is likely the second most important parameter for viability following adult survival.
 So we are actually exploring viability across environmental conditions that affect the two most important parameters for population growth.
-While we agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameters we explore are the most important for long-term viability.
-As no changes were requested we, we did not make any.]{style="color: blue"}
+While we agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameters we explore in detail are the most important for long-term viability.
+The reviewer did not request any changes in response to this comment, and we did not make any.]{style="color: blue"}
 
-5.  Lines 221-233. This is very true -- congratulations!!
+-   Lines 221-233.
+    This is very true -- congratulations!!
 
@@
 
-6.
-Line 252.
+-   Line 252.
     I suggest revising this sentence; frog survival at natal (donor) sites was never reported.
-If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig.
-3) I would clarify that.
+    If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.
 
@@
 
-General comments: This is a long-term dataset, in many ways exceptionally long for in depth amphibian studies, and a valuable manuscript on a priority topic for biodiversity conservation.
+General comments:
+
+This is a long-term dataset, in many ways exceptionally long for in depth amphibian studies, and a valuable manuscript on a priority topic for biodiversity conservation.
 Bd research has been criticised for missing direct conservation applications in the field, which makes this study particularly novel and insightful, despite focusing on a single species.
+
 The results are very encouraging and give much cause of cautious optimism for what has often appeared to be an almost intractable problem for amphibian conservation but while I understand the specificity of your results it would be good to have at least a paragraph that looks at the potential implications for other species and important steps forward that could be concluded from your own work.
+
+[We added the suggested text to the final paragraph of the Discussion (lines xxx-xxx). See also our response to Reviewer 1 - Comment 4 above.]{style="color: blue"}
+
 In my opinion the text is clear and very well-written and I do not have much to add in relation to that.
 There is some repetition between introduction and discussion so please be aware of that.
+
+[We edited the Discussion to minimize any repetition between the Introduction and Discussion.
+Any remaining repetition reflects our interest in reinforcing points in the Discussion that were made initially in the Introduction.
+When specific points are repeated in both sections, we have ensured that the text differs between the two sections of the manuscript.]{style="color: blue"}
+
 As a general comment, you mention that the three donor populations are large, with several hundred adults and that you have documented the genetic structure but I suggest it would be useful to briefly discuss the potential implications of the reintroductions from a genetic perspective and any potential monitoring follow-up of that aspect and mitigation strategy if required.
 There is a reasonable potential for creating a genetic bottleneck as a results of these reintroductions in the absence of some kind of genetic monitoring.
+
+[We appreciate this important point. We added text to the last paragraph of the Discussion (lines xxx-xxx) to address this issue.]{style="color: blue"}
+
 Also, it would be useful to briefly explain why are reintroductions important (and cost-effective) instead of simply waiting for these recovering populations to naturally expand over time.
+
+[This is another very important issue, and we thank the reviewer for highlighting it.
+We added a new paragraph to the Discussion that summarizes the inadequacy of natural recovery and the importance of translocations in allowing reestablishment of populations at the landscape scale.]{style="color: blue"}
+
 The methods look sound but I have limited experience in this type of multilevel Bayesian modelling.
 
-Detailed comments Title- Just a suggestion but for me the title would benefit from mentioning the species.
+Detailed comments
+
+Title- Just a suggestion but for me the title would benefit from mentioning the species.
 Your work obviously has wider implications but remains currently very species-specific.
+
 Line 155- for me this aspect of the apparent high variability linked to donor population is important and I suggest should be highlighted both in the results and especially in the discussion as a separate point.
@@
 I would strongly expect specific weather patterns to also play an important role in any given year, even if these might be rarer events that you have not yet captured.
+
 Line 288- for me the value in this call for additional research warrants some brief but specific details about what that would entail.
 Otherwise, researchers naturally always call for additional research.
+
 Lines 298-301- I think this is important and relevant in the wider amphibian conservation context, where juveniles at this crucial life stage are often the specifically missing dataset because they are harder to monitor.
 This is in many ways a failure of the sector and the opposite compared to say birds or mammals, yet has become inbuilt in much amphibian research and is worth highlighting as it has important consequences for progressing practical conservation.
+
 Lines 310-315 I suggest you could be more forward looking here and suggest some critical steps forward for interpreting the success of your results for advancing the conservation prospects for other similarly threatened amphibian species.
@@
 It might also be useful to also briefly mention the relevance of your results for reintroductions from the perspective of captive breeding and amphibian arks given that the current approaches would not facilitate any kind of resistance building (if at all possible).
+
 Line 317- why would it though?
 Now that you have developed a solid understanding of the process, could you not strategically reintroduce populations to facilitate ongoing natural colonisation once there are sufficient numbers to create some wider population resilience?
+
 Line 337- this is an important methodological point and which requires a bit more substance in terms of how was absence determined, over what period, etc

### 2024-08-02 doc/manuscript/supplementary_information.qmd
More edits in response to reviewer comments

@@
 Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$.
-Multiple primary periods can occur during a summer active season (typically July-August, but in some years as early as May and as late as September).
+Multiple primary periods can occur during a summer active season (see Methods).
 We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$.
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:-------------------------|:------------------|:---------------------------|
+|:--------------------------------------------------------------------------|:---------------------|:----------------------------------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2024-08-02 doc/manuscript/translocation.qmd
More edits in response to reviewer comments

@@
 
-The recent emergence of Bd worldwide has contributed to the decline of hundreds of amphibian species, some of which are now extinct in the wild [@scheele2019].
-This extraordinary impact on global amphibian biodiversity is compounded by the lack of any effective and broadly applicable strategies to reverse these impacts [@garner2016; @knapp2022].
+Our successful use of translocations to reestablish *R.
+sierrae* populations despite ongoing Bd infection is particularly important because natural recovery alone appears insufficient to allow reestablishment of populations at the landscape scale.
+Naturally recovering populations in Yosemite are relatively rare and are often isolated from other such populations and from suitable but vacant habitats.
+This isolation is the consequence of both natural fragmentation due to rugged geography and the ongoing presence of introduced, predatory trout in the majority of lakes and streams [@bradford1993].
+Collectively, these factors often prevent naturally recovering *R. sierrae* populations from expanding via dispersal of frogs to nearby unoccupied habitats.
+Translocations allowed us to overcome these dispersal limitations and reestablish populations in sites that would otherwise remain unoccupied for the foreseeable future.
+Of particular note are the translocations that we conducted as part of the current study into areas from which *R.
+sierrae* were completely extirpated, but instead of single isolated habitats these areas contained habitats composed of interconnected fishless lakes, ponds, marshes, and/or streams.
+Successful reestablishment of one *R. sierrae* population in such areas has produced a source of frogs from which adults and juveniles are now dispersing into adjacent habitats.
+This process of translocation, establishment, and subsequent dispersal has significant potential to eventually allow the reestablishment of robust metapopulations that may be more resilient to ongoing and future natural and anthropogenic impacts than more fragmented populations (e.g., [@heard2015]).
+
+The promising example provided by our study of successful reestablishment of *R.
+sierrae* despite ongoing Bd infection may provide an important roadmap for ongoing and future efforts to recover the hundreds of Bd-impacted amphibian species globally.
 Importantly, in addition to the natural recovery documented for *R.
@@
 sierrae*, the feasibility and long-term success of such efforts will depend on the availability of robust donor populations containing individuals that have sufficiently high resistance to allow frog survival and population growth in the presence of Bd.
-Despite the promising example provided by our study of successful reestablishment of *R.
-sierrae* despite ongoing Bd infection, the challenge of recovering hundreds of Bd-impacted amphibian species globally is a daunting prospect.
-An adaptive management approach such as the one we utilized, in which the outcomes of recovery actions are rigorously assessed and this information is used to reduce uncertainties and inform future actions, will be essential in maximizing the effectiveness of this global recovery effort [@canessa2019].
-In addition, developing an improved understanding of the factors facilitating or limiting the development of frog resistance against Bd infection and the effectiveness of selective breeding programs and other genetic interventions to enhance frog resistance [@kosch2022] may be particularly important.
+We suggest that several components of recovery efforts could play important roles in maximizing their success.
+First, an adaptive management approach such as the one we utilized, in which the outcomes of recovery actions are rigorously assessed and this information is used to reduce uncertainties and inform future actions, will often be essential [@canessa2019].
+Second, given the apparent rarity in amphibians of evolution of resistance against Bd infection, developing an improved understanding of the factors facilitating or limiting the development of frog resistance and the effectiveness of selective breeding programs and other genetic interventions to enhance frog resistance [@kosch2022] may be particularly important.
+Finally, although we did not quantify the genetic structure of the donor and translocated populations of *R.
+sierrae* as part of the current study, conducting genetic monitoring may be important to identify sub-optimal genetic outcomes during establishment of translocated populations (e.g., genetic bottlenecks) and allow timely implementation of actions to mitigate such effects [@whiteley2015].
 Although we now have a proven strategy to reestablish extirpated *R. sierrae* populations, recovery across their large historical range will require substantial resources over many decades.
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, we visited translocated populations approximately once per month during the summer active season and conducted diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
+Following release, we visited translocated populations approximately once per month during the summer active season to conduct diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits/primary periods per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
 CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size.
 VES provided estimates of tadpole and juvenile abundance.
-During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period) , during which we searched all habitats repeatedly for adult frogs.
+During 2006-2012, we conducted CMR surveys on a single day per site visit (primary period), during which we searched all habitats repeatedly for adult frogs.
 Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
@@
 ![**Locations of translocated and donor *R.
-sierrae* populations.** All populations are in Yosemite National Park (park boundary indicated by gray polygon).
+sierrae* populations.** All populations are in Yosemite National Park (park boundary indicated by black polygon).
 Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site.
@@
 Inset map shows the location of Yosemite within California.
-In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="60%"}
+In both maps, topography is shown with a grayscale hillshade layer, and elevation is indicated by colors layered over the hillshade.
+In the absence of hillshading, dark green colors indicate the lowest elevations and white indicates the highest elevation.](figures/map_translocation_points.png){#fig-yosemap width="60%"}
 

### 2024-08-02 doc/manuscript/response_to_reviewers_original_submission.qmd
Edits in response to comments from Reviewer 3

@@
 
-[We added a sentence to the Discussion to make the point highlighted by the reviewer more strongly (Discussion - 2nd para, last sentence)]{style="color: blue"}
+[We added a sentence to the Discussion to make the point highlighted by the reviewer more strongly (lines 235-239)]{style="color: blue"}
 
@@
 
-[We added a sentence to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}
+[We added addition text to the Discussion to make the point suggested by the reviewer (lnes 346-350)]{style="color: blue"}
 
@@
 
-[This in-depth topic would require the addition of a substantial amount of new text, and would be better addressed in a separate manuscript.
-As such, we have chosen not to expand our treatment of this topic in the current manuscript.]{style="color: blue"}
+[This is an important point and we appreciate the reviewer highlighting it.
+We added a paragraph to the Discussion that addresses several of these points (lines 316-333; see also our response to the final "general comment" made by Reviewer 3).]{style="color: blue"}
 
@@
 
-[We added additional sentences to the Discussion to make the point suggested by the reviewer (Discussion, last paragraph)]{style="color: blue"}
+[We added additional sentences to the Discussion to make the point suggested by the reviewer (lines 344-346).]{style="color: blue"}
 
@@
 [We agree with the reviewer that important progress is being made to mitigate the impacts of chytridiomycosis on the world's amphibians.
-Nonetheless, many challenges remain and in writing the final paragraph of the Discussion we didn't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
-However, we agree that the final paragraph could provide a bit more hopeful outlook.
-We made several changes to this paragraph to accomplish that.]{style="color: blue"}
+Nonetheless, many challenges remain.
+In writing the final paragraph of the Discussion, we therefore didn't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
+However, we agree that the paragraph in question could provide an outlook that is a bit more hopeful.
+We made several changes to this paragraph to accomplish that (lines 334-354).]{style="color: blue"}
 
@@
 
-[The sentences in question are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes them from objectives to descriptions of the three components of the study.
+[The sentences in question (lines 90-98) are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes them from objectives to descriptions of the three components of the study.
 We did not make any changes in response to this comment, but we are open to any suggestions to improve the structure of these sentences.]{style="color: blue"}
@@
 [Thank you for pointing out these difficulties in interpreting the map.
-In the revised version, the polygon indicating the park boundary in the main map and the inset map are both displayed in black.
+In the revised version of Figure 2, the polygon indicating the park boundary in the main map and the inset map are both displayed in black.
 In addition, we added clarifying text to the figure legend to describe the colors and their meaning.]{style="color: blue"}
@@
 
-[We updated this sentence as the reviewer suggested.]{style="color: blue"}
+[We updated this sentence as the reviewer suggested (lines 389-390).]{style="color: blue"}
 
@@
 However, it may not have been clear that a "visit" is equivalent to a "primary period".
-Therefore, we added some additional text to the sentence in question (revised manuscript: lines xxx-xxx) to clarify this point.]{style="color: blue"}
+Therefore, we added some additional text to the sentence in question (revised manuscript: lines 386-387) to clarify this point.]{style="color: blue"}
 
@@
 
-[Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area.]{style="color: blue"}
+[Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area (lines 465-469).]{style="color: blue"}
 
@@
 We are modeling the entire population, not just females, and are assuming a 50:50 sex ratio.
-This was not explicit in our equations and we have thus updated our viability model equation to explicitly include a parameter for sex ratio ($\rho$).
-We now describe this parameter in the Table in the supplementary material.
+This was not explicit in our equations and we have thus updated our viability model equation to explicitly include a parameter for sex ratio ($\rho$; line 477).
+We also added a description of this parameter in Supplementary Table 1.
 While addressing this comment, we noticed that we were not consistent with modeling females vs.
@@
 However, we see how this could be confusing.
-As we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text as suggested by the reviewer.]{style="color: blue"}
+Becuase we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text as suggested by the reviewer.]{style="color: blue"}
 
@@
 
-    [Thanks]{style="color: blue"}
+    [We appreciate the reviewer's acknowledgement and agreement with this important point of ours.]{style="color: blue"}
 
@@
 
-    [As the reviewer suggested, we added a reference to Supplementary Fig. 3]{style="color: blue"}
+    [As the reviewer suggested, we added a reference to Supplementary Fig. 3 in this sentence (lines 255-258).]{style="color: blue"}
 
@@
 
-[We added the suggested text to the final paragraph of the Discussion (lines xxx-xxx). See also our response to Reviewer 1 - Comment 4 above.]{style="color: blue"}
+[We added the suggested text to one of the final paragraphs of the Discussion (lines 334-354). See also our response to Reviewer 1 - Comment 4 above.]{style="color: blue"}
 
@@
 
-[We appreciate this important point. We added text to the last paragraph of the Discussion (lines xxx-xxx) to address this issue.]{style="color: blue"}
+[We appreciate this important point. We added text to one of the last paragraphs of the Discussion (lines 350-354) to address this issue.]{style="color: blue"}
 
@@
 [This is another very important issue, and we thank the reviewer for highlighting it.
-We added a new paragraph to the Discussion that summarizes the inadequacy of natural recovery and the importance of translocations in allowing reestablishment of populations at the landscape scale.]{style="color: blue"}
+We added a new paragraph to the Discussion that summarizes the inadequacy of natural recovery and the importance of translocations in allowing reestablishment of populations at the landscape scale (lines 316-333).]{style="color: blue"}
 
@@
 
+[The title is currently at the recommended limit of 15 words.
+As such, although adding "(*Rana sierrae*)" to the title might be useful in some contexts, we opted to leave the title as is.]{style="color: blue"}
+
 Line 155- for me this aspect of the apparent high variability linked to donor population is important and I suggest should be highlighted both in the results and especially in the discussion as a separate point.
@@
 
--   [Good question.
-Our population viability analyses show that long-term viability is significantly more likely when 1-year adult survival probability is \>0.5.
-We hesitate to try to separate the effect of recipient site and donor population on 1-year adult survival as these factors were not crossed in our study (an inevitable logistical constraint when working on real landscapes and taking into account host landscape genetics \[ROLAND, update as needed\]).
-That being said, there is strong support from the model that the lowest 1-year survival from donor population 72996 would not be sufficient for long-term viability (\< 0.5) while the lowest 1-year adult survival from 70459 and 70567 would be sufficient for long-term viability (\> 0.5).
-While it is tempting to conclude that donor populations 70459 and 70567 should be the focus of subsequent translocations, our design cannot disentangle site and donor effects.]{style="color: blue"}
+[This is an important question, but we cannot address it directly using results from the current study.
+Our population viability analysis shows that long-term viability is significantly more likely when 1-year adult survival probability is \>0.5.
+In addition, there is strong support from the model that the lowest 1-year survival from donor population 72996 would not be sufficient for long-term viability (adult survival probability \< 0.5) and that the lowest 1-year adult survival from 70459 and 70567 would be sufficient for long-term viability (adult survival probability \> 0.5).
+Although it is tempting to conclude that donor populations 70459 and 70567 should be used in future translocations, our study design cannot disentangle the effects of recipient site and donor population on 1-year adult survival because these factors were not crossed (an inevitable logistical constraint when working on real landscapes and taking into account *R.
+sierrae* landscape genetics).
+We are currently conducting translocations in which frogs from several donor populations are added to a single site, and this design will allow us to separate these effects.
+Given our inability in the current study to estimate the independent effect of donor population on adult survival, we opted not to emphasize this issue, and instead will provide a detailed treatment of it in a subsequent paper that describes the results of the multi-donor translocations.
+However, we added a brief description of these multi-donor translocations to the Discussion (lines 291-297).]{style="color: blue"}
 
@@
 
+[We softened our assertion that 1-year frog survival serves as an effective proxy of longer-term survival and population viability by replacing the word "indicate" with "suggest" (lines 269-271).]{style="color: blue"}
+
 Line 288- for me the value in this call for additional research warrants some brief but specific details about what that would entail.
@@
 
+[We replaced the text in question with a specific description of the study we are currently conducting to quantify the effects of donor population (lines ]{style="color: blue"}[ 291-297]{style="color: blue"}[; see also our response to a comment by Reviewer 3 above).]{style="color: blue"}
+
 Lines 298-301- I think this is important and relevant in the wider amphibian conservation context, where juveniles at this crucial life stage are often the specifically missing dataset because they are harder to monitor.
 This is in many ways a failure of the sector and the opposite compared to say birds or mammals, yet has become inbuilt in much amphibian research and is worth highlighting as it has important consequences for progressing practical conservation.
+
+[We added a sentence to the end of this paragraph that emphasizes the critical need for better monitoring of this vulnerable life stage (lines 313-315)]{style="color: blue"}
 
@@
 
+[As described in our response to Reviewer 1-Comment 4, we revised this paragraph to include specific recommendations for ongoing and future efforts to recover Bd-impacted amphibians.
+The added text addresses some of the specific points made by Reviewer 3, including mention of methods such as captive breeding and genetic inventions to facilitate the development of frog resistance against Bd infection (lines 334-354).]{style="color: blue"}
+
 Line 317- why would it though?
@@
 
+[We added a paragraph to the Discussion to elaborate on the limitations of natural recovery and the need for active management via reintroductions to reestablish frogs at the landscape scale (lines 316-333).
+The specific scenario that the reviewer mentions is described in some detail in that paragraph (see also our response to the "general comment" from Reviewer 3 above regarding natural recovery versus recovery via reintroductions).]{style="color: blue"}
+
 Line 337- this is an important methodological point and which requires a bit more substance in terms of how was absence determined, over what period, etc
+
+[We added a sentence to this paragraph that briefly describes the methods we used to determine *R. sierrae* absence (lines 376-378).]{style="color: blue"}

### 2024-08-02 doc/manuscript/translocation.qmd
Edits in response to comments from Reviewer 3

@@
 sierrae*.
-Specifically, the combined results from our reintroduction study and viability modeling indicate that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
+Specifically, the combined results from our reintroduction study and viability modeling suggest that survival of frogs in the first year following translocation is an effective proxy of longer-term survival and population viability.
 In addition, given the repeatability of frog survival at a site, 1-year frog survival also serves as an effective proxy of site quality (i.e., the ability of a site to support high frog survival and a viable frog population over the long term).
@@
 The mechanisms underlying the effect of donor population on the probability of post-translocation frog survival is intriguing, and could indicate population-specific fitness differences of frogs in the presence of Bd.
-A clearer understanding of the mechanistic components of the elevation and donor population variables will require additional research.
+Although our study design did not allow us to completely separate the effects of donor population and recipient site on 1-year adult survival, we are currently doing so using a modified study design.
+Specifically, we are conducting translocations in which frogs from several donor populations are added concurrently to a single site.
+Survival of frogs from each donor population is estimated using CMR and the contribution of each donor population to subsequent generations is quantified using genetic methods.
 
@@
 Therefore, although adult *R. sierrae* appear relatively resistant, Bd infection continues to have important limiting effects on recovering populations (see also [@hollanders2022]).
+The ongoing high susceptibility of metamorphs to chytridiomycosis also highlights the critical need to develop better methods of monitoring this vulnerable life stage, the results of which could significantly advance the conservation and recovery of Bd-impacted amphibians.
 
@@
 sierrae* as part of the current study, conducting genetic monitoring may be important to identify sub-optimal genetic outcomes during establishment of translocated populations (e.g., genetic bottlenecks) and allow timely implementation of actions to mitigate such effects [@whiteley2015].
-Although we now have a proven strategy to reestablish extirpated *R. sierrae* populations, recovery across their large historical range will require substantial resources over many decades.
+
+In conclusion, we now have a proven strategy to reestablish extirpated *R. sierrae* populations,.
+However, recovery across their large historical range will require substantial resources over many decades.
 The results of this study provide a hopeful starting point for that endeavor and other future efforts worldwide.
@@
 *R. sierrae* were absent from all recipient sites prior to the first translocation.
+This determination was based on VES conducted at recipient sites over several years (see additional details below regarding frog detectability during VES).
 

### 2024-08-03 doc/manuscript/supplementary_information.qmd
Change plot labels from upper to lower case

@@
 ![**Sensitivity analysis of the stage-structured *R.
-sierrae* model.** **A** Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
-**B** Sensitivity of $\lambda$ with changes in the same nine parameters.
+sierrae* model.** **a** Elasticity of $\lambda$ with changes in nine parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $F$ (number of eggs produced by a female frog in a year that successfully hatch), $\sigma_{J_1}$ (yearly probability of survival of year-1 juveniles that also affects recruitment), $\sigma_{J_2}$ (yearly probability of survival of year-2 juveniles), $p\_{L_2}$ (probability that year-2 tadpoles metamorph into juveniles), $p_{J_1}$ (probability that year-1 juveniles mature to adults), $\sigma_{L_1}$ (survival probability of year-1 tadpoles), $\sigma_{L_2}$ (survival probability of year-2 tadpoles), and $\sigma_{L_3}$ (survival probability of year-3 tadpoles).
+**b** Sensitivity of $\lambda$ with changes in the same nine parameters.
 Note that $F$ has different units than the other parameters, which is why its sensitivity is low.

### 2024-08-03 doc/manuscript/translocation.qmd
Change plot labels from upper to lower case

@@
 Results are from the among-site meta-analysis and 1-year frog survival is expressed as a probability.
-**A** winter severity in the year following translocation, **B** elevation of recipient site, **C** donor population, and **D** sex.
-In (A) and (B), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (C) and (D), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
+**a** winter severity in the year following translocation, **b** elevation of recipient site, **c** donor population, and **d** sex.
+In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals.
+In (c) and (d), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
 
@@
 ![**Estimated growth rates and extinction probabilities of reintroduced populations**.
-**A** Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival probability $\sigma_{J_1}$, given the parameterized, deterministic model.
+**a** Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and year-1 juvenile survival probability $\sigma_{J_1}$, given the parameterized, deterministic model.
 Colored points show the predicted $\lambda$ values for the twelve translocated populations when year-1 juvenile survival probability is $\sigma_{J_1} = 0.10$ (indicated by the dashed line).
@@
 Note that the point for 70413 is mostly hidden behind other points.
-**B** Predicted 50-year extinction probabilities of the 12 reintroduced populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
+**b** Predicted 50-year extinction probabilities of the 12 reintroduced populations, given demographic stochasticity, environmental variability in $\sigma_{J_1}$, and different mean values of $\sigma_{J_1}$.
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are partially or completely hidden beneath the line for 70505.
-**C** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
+**c** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}

### 2024-08-03 doc/manuscript/response_to_reviewers_original_submission.qmd
Update figure formatting and review responses

@@
 
-[We added addition text to the Discussion to make the point suggested by the reviewer (lnes 346-350)]{style="color: blue"}
+[We added additional text to the Discussion to make the point suggested by the reviewer (lines 346-350)]{style="color: blue"}
 
@@
 However, we see how this could be confusing.
-Becuase we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text as suggested by the reviewer.]{style="color: blue"}
+Because we do not lose much biological interpretation by just referring to $\sigma_{J_1}$ as "survival" we have made this change throughout the text as suggested by the reviewer.]{style="color: blue"}
 
@@
 
-[We replaced the text in question with a specific description of the study we are currently conducting to quantify the effects of donor population (lines ]{style="color: blue"}[ 291-297]{style="color: blue"}[; see also our response to a comment by Reviewer 3 above).]{style="color: blue"}
+[We replaced the text in question with a specific description of the study we are currently conducting to quantify the effects of donor population (lines 291-297; see also our response to a comment by Reviewer 3 above).]{style="color: blue"}
 

### 2024-08-03 doc/manuscript/supplementary_information.qmd
Update figure formatting and review responses

@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in Supporting Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.jpg){#fig-compare_surv_probs width="60%"}
+All estimates were derived using the mrmr package, and the methods for calculation are described in Supporting Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
 
@@
 We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study ([@tbl-param_values]) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity are calculated at the default parameter values for each population given in Table [@tbl-param_values].](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="95%"}
+Elasticity and sensitivity are calculated at the default parameter values for each population given in Table [@tbl-param_values].](figures/pop_viability_figures_for_supp.png){#fig-viability-supp width="95%"}
 
@@
 | **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:--------------------------|:---------------|:-----------------------------|
+|:-------------------------|:-----------------|:---------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |

### 2024-08-03 doc/manuscript/translocation.qmd
Update figure formatting and review responses

@@
 During the epizootic phase, *R. sierrae* are highly susceptible to Bd infection as indicated by very high infection intensities ("load").
-In constrast, frogs in recovering populations have reduced susceptibility to Bd infection [@knapp2016], with loads on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
+In contrast, frogs in recovering populations have reduced susceptibility to Bd infection [@knapp2016], with loads on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
 This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection and not simply an effect of factors external to individual frogs (e.g., environmental conditions).
@@
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are partially or completely hidden beneath the line for 70505.
-**c** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="95%"}
+**c** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.png){#fig-viability width="95%"}

### 2024-08-03 code/extinction_and_growth_rate_analysis.R

@@ -484,9 +484,10 @@ ptraj = ggplot() + geom_line(data=pred_traj, aes(x=year, y=abund, group=sim, col
 ##### Join all of the plots together ####
 #########################################
 
-myplot = (ptile + pext + ptraj) + plot_annotation(tag_levels="A", tag_suffix="")
-ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.jpg"), width=11, height=5, dpi=300)
+myplot = ((ptile + pext + ptraj) + plot_annotation(tag_levels="a", tag_suffix="") & 
+					theme(plot.tag = element_text(face = 'bold'), plot.tag.position  = c(0.17, 1)))
+ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.png"), width=11, height=5, dpi=300)
 
-myplot = (p2 + p2b) + plot_annotation(tag_levels="A", tag_suffix="")
-ggsave(file.path("..", "out", "pop_viability_figures_for_supp.jpg"), width=9, height=4, dpi=300)
+myplot = (p2 + p2b) + plot_annotation(tag_levels="a", tag_suffix="")
+ggsave(file.path("..", "out", "pop_viability_figures_for_supp.png"), width=9, height=4, dpi=300)
 

### 2024-08-03 code/extract_survival_estimates.R

@@ -167,7 +167,7 @@ ggplot(all_dt) +
 				ylab("Yearly survival probability") + xlab("Lake ID") + 
 				theme_classic() + theme(legend.title=element_blank()) + ylim(c(0, 1))
 
-ggsave(file.path("..", "out", "compare_surv_probs.jpg"), width=6, height=4)
+ggsave(file.path("..", "out", "compare_surv_probs.png"), width=6, height=4)
 
 
 

### 2024-08-03 code/translocation_survival_analysis.Rmd
Change plot identifiers to lower case bold font

@@
 library(tidyverse)
-library(brms)
+#library(brms)
 library(rstanarm)
@@
             median = quantile(.epred, 0.5)) %>%
+  # ggplot(aes(x = elevation)) + 
+  #         geom_ribbon(aes(ymin = lower, ymax = upper), fill= "lightgray", alpha = 1) +
+  #         geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
+  #         labs(x = "Elevation of recipient site (meters)", y = "P(1-year survival)") +
+  #         annotate(geom = "text", x = 2500, y = 0.94, label = "b", size = 4, fontface = "bold") +
+  #         theme_classic()
   ggplot(aes(x = elevation)) + 
@@
           geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
+          coord_cartesian(ylim = c(0, 1.0)) +
           labs(x = "Elevation of recipient site (meters)", y = "P(1-year survival)") +
-          annotate(geom = "text", x = 2500, y = 0.94, label = "B", size = 4) +
+          annotate(geom = "text", x = 2400, y = 1, label = "b", size = 4, fontface = "bold") +
           theme_classic()
@@
             median = quantile(.epred, 0.5)) %>%
+  # ggplot(aes(x = snow_t1)) +
+  #         geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightgray", alpha = 1) +
+  #         geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
+  #         labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year survival)") +
+  #         annotate(geom = "text", x = 157, y = 0.92, label = "a", size = 4, fontface = "bold") +
+  #         theme_classic()
+
   ggplot(aes(x = snow_t1)) +
@@
           geom_line(aes(y = median), color = "blue", linewidth = 1, alpha = 1) +
+          coord_cartesian(ylim = c(0, 1.0)) +
           labs(x = "Winter severity in year t+1\n(% of average snow depth)", y = "P(1-year survival)") +
-          annotate(geom = "text", x = 157, y = 0.92, label = "A", size = 4) +
+          annotate(geom = "text", x = 7, y = 1, label = "a", size = 4, fontface = "bold") +
           theme_classic()
@@
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  # ggplot(aes(x = male, y = surv)) +
+  #   geom_boxplot(aes(y = .epred, x = sex), fill = "lightgray") +
+  #   labs(x = "Sex", y = "P(1-year survival)") +
+  #   annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "d", size = 4, fontface = "bold") +
+  #   theme_classic()
   ggplot(aes(x = male, y = surv)) +
     geom_boxplot(aes(y = .epred, x = sex), fill = "lightgray") +
+    coord_cartesian(ylim = c(0, 1.0)) +
+    scale_x_discrete(labels = c("female", "male")) +
     labs(x = "Sex", y = "P(1-year survival)") +
-    annotate(geom = "text", x = 0.4, y = 0.94, hjust = -1, label = "D", size = 4) +
+    annotate(geom = "text", x = 0.15, y = 1, hjust = -1, label = "d", size = 4, fontface = "bold") +
     theme_classic()
@@
   add_epred_draws(m1d_reduce, ndraws = 100, re_formula = NA) %>% 
+  # ggplot(aes(x = donor, y = surv)) +
+  #   geom_boxplot(aes(y = .epred, x = donor), fill = "lightgray") +
+  #   labs(x = "Donor population", y = "P(1-year survival)") +
+  #   annotate(geom = "text", x = 3.3, y = 0.94, hjust = 0, label = "c", size = 4, fontface = "bold") +
+  #   theme_classic()
   ggplot(aes(x = donor, y = surv)) +
     geom_boxplot(aes(y = .epred, x = donor), fill = "lightgray") +
+    coord_cartesian(ylim = c(0, 1.0)) +
     labs(x = "Donor population", y = "P(1-year survival)") +
-    annotate(geom = "text", x = 3.3, y = 0.94, hjust = 0, label = "C", size = 4) +
+    annotate(geom = "text", x = 0.5, y = 1, hjust = 0, label = "c", size = 4, fontface = "bold") +
     theme_classic()

### 2024-08-03 code/extinction_and_growth_rate_analysis.R
Label position change

@@ -488,6 +488,7 @@ myplot = ((ptile + pext + ptraj) + plot_annotation(tag_levels="a", tag_suffix=""
 					theme(plot.tag = element_text(face = 'bold'), plot.tag.position  = c(0.17, 1)))
 ggsave(file.path("..", "out", "pop_viability_figures_for_manuscript.png"), width=11, height=5, dpi=300)
 
-myplot = (p2 + p2b) + plot_annotation(tag_levels="a", tag_suffix="")
+myplot = ((p2 + p2b) + plot_annotation(tag_levels="a", tag_suffix="") & 
+					theme(plot.tag = element_text(face = 'bold'), plot.tag.position  = c(0.14, 1)))
 ggsave(file.path("..", "out", "pop_viability_figures_for_supp.png"), width=9, height=4, dpi=300)
 

### 2024-08-04 doc/manuscript/response_to_reviewers_original_submission.qmd
Final version for re-submission

@@
 
-Our point-by-point response to all comments from the three reviewers is provided below.
-Reviewer comments are in black and our responses are in blue.
+[Our point-by-point response to all comments from the three reviewers is provided below. Reviewer comments are in black and our responses are in blue.]{style="color: blue"}
 
@@
 
-[We added a sentence to the Discussion to make the point highlighted by the reviewer more strongly (lines 235-239)]{style="color: blue"}
+[We added a sentence to the Discussion to make the point highlighted by the reviewer more strongly (lines 236-240).]{style="color: blue"}
 
@@
 
-[We added additional text to the Discussion to make the point suggested by the reviewer (lines 346-350)]{style="color: blue"}
+[We added additional text to the Discussion to make the point suggested by the reviewer (lines 347-351).]{style="color: blue"}
 
@@
 
-[This is an important point and we appreciate the reviewer highlighting it.
-We added a paragraph to the Discussion that addresses several of these points (lines 316-333; see also our response to the final "general comment" made by Reviewer 3).]{style="color: blue"}
+[The reviewer highlights several points in this comment.
+We added a paragraph to the Discussion that addresses several of these points (lines 316-334; see also our response to the final "general comment" made by Reviewer 3).]{style="color: blue"}
 
@@
 
-[We added additional sentences to the Discussion to make the point suggested by the reviewer (lines 344-346).]{style="color: blue"}
+[We added additional sentences to the Discussion to make the point suggested by the reviewer (lines 345-347).]{style="color: blue"}
 
@@
 Nonetheless, many challenges remain.
-In writing the final paragraph of the Discussion, we therefore didn't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
+In writing the final paragraphs of the Discussion, we therefore didn't want to oversell the significance of our study and that of other recent studies (as cited by the reviewer).
 However, we agree that the paragraph in question could provide an outlook that is a bit more hopeful.
-We made several changes to this paragraph to accomplish that (lines 334-354).]{style="color: blue"}
+We made several changes to this paragraph to accomplish that (lines 335-355).]{style="color: blue"}
 
@@
 
-[The sentences in question (lines 90-98) are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes them from objectives to descriptions of the three components of the study.
+[The sentences in question (lines 90-98) are complete sentences, and the reviewer's suggestion to add "we" to each sentence changes the sentences from objectives to descriptions of the three components of the study.
 We did not make any changes in response to this comment, but we are open to any suggestions to improve the structure of these sentences.]{style="color: blue"}
@@
 [Thank you for pointing out these difficulties in interpreting the map.
-In the revised version of Figure 2, the polygon indicating the park boundary in the main map and the inset map are both displayed in black.
+In the revised version of Figure 2, the polygon indicating the park boundary is now displayed in black in both the main map and the inset map.
 In addition, we added clarifying text to the figure legend to describe the colors and their meaning.]{style="color: blue"}
@@
 
-[We updated this sentence as the reviewer suggested (lines 389-390).]{style="color: blue"}
+[We updated this sentence as the reviewer suggested (lines 390-392).]{style="color: blue"}
 
@@
 However, it may not have been clear that a "visit" is equivalent to a "primary period".
-Therefore, we added some additional text to the sentence in question (revised manuscript: lines 386-387) to clarify this point.]{style="color: blue"}
+Therefore, we added some additional text to the sentence in question to clarify this point (revised manuscript: lines 385-388).]{style="color: blue"}
 
@@
 
-[Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area (lines 465-469).]{style="color: blue"}
+[Good point. We are modeling abundance per population so we have replaced the word density with abundance to avoid confusion over the units of area (lines 468-472).]{style="color: blue"}
 
@@
 [We appreciate the reviewer's thoughtful point.
-It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$ --- a parameter that our sensitivity analyses indicate is likely the second most important parameter for viability following adult survival.
+It is worth noting that the viability analysis also explores variation in $\sigma_{J_1}$, a parameter that our sensitivity analyses indicate is likely the second-most important parameter for viability following adult survival.
 So we are actually exploring viability across environmental conditions that affect the two most important parameters for population growth.
-While we agree that other vital rates likely vary to some degree across population, we have good empirical and modeling evidence that the two parameters we explore in detail are the most important for long-term viability.
+While we agree that other vital rates likely vary to some degree across populations, we have good empirical and modeling evidence that the two parameters we explore in detail are the most important for long-term viability.
 The reviewer did not request any changes in response to this comment, and we did not make any.]{style="color: blue"}
 
--   Lines 221-233.
-    This is very true -- congratulations!!
+-   Lines 221-233. This is very true -- congratulations!!
 
@@
 I suggest revising this sentence; frog survival at natal (donor) sites was never reported.
-    If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig. 3) I would clarify that.
+If you are referring to the difference between translocated and naturally recruited adults at the 3 recipient locations (Supplementary Fig.
+3) I would clarify that.
 
-    [As the reviewer suggested, we added a reference to Supplementary Fig. 3 in this sentence (lines 255-258).]{style="color: blue"}
+[As the reviewer suggested, we added a reference to Supplementary Fig. 3 in this sentence (lines 256-258).]{style="color: blue"}
 
@@
 
-[We added the suggested text to one of the final paragraphs of the Discussion (lines 334-354). See also our response to Reviewer 1 - Comment 4 above.]{style="color: blue"}
+[We added the suggested text to one of the final paragraphs of the Discussion (lines 335-355). See also our response to Reviewer 1 - Comment 4 above.]{style="color: blue"}
 
@@
 
-[We appreciate this important point. We added text to one of the last paragraphs of the Discussion (lines 350-354) to address this issue.]{style="color: blue"}
+[We appreciate this important point. We added text to one of the last paragraphs of the Discussion to address this issue (lines 351-355).]{style="color: blue"}
 
@@
 [This is another very important issue, and we thank the reviewer for highlighting it.
-We added a new paragraph to the Discussion that summarizes the inadequacy of natural recovery and the importance of translocations in allowing reestablishment of populations at the landscape scale (lines 316-333).]{style="color: blue"}
+We added a new paragraph to the Discussion that summarizes the inadequacy of natural recovery and the importance of translocations in allowing reestablishment of populations at the landscape scale (lines 316-334).]{style="color: blue"}
 
@@
 [The title is currently at the recommended limit of 15 words.
-As such, although adding "(*Rana sierrae*)" to the title might be useful in some contexts, we opted to leave the title as is.]{style="color: blue"}
+As such, although adding "(*Rana sierrae*)" to the title might be useful in some contexts, we opted to leave the title unchanged.]{style="color: blue"}
 
@@
 
-[We softened our assertion that 1-year frog survival serves as an effective proxy of longer-term survival and population viability by replacing the word "indicate" with "suggest" (lines 269-271).]{style="color: blue"}
+[We softened our assertion that 1-year frog survival is an effective proxy of longer-term survival and population viability by replacing the word "indicate" with "suggest" (lines 269-271).]{style="color: blue"}
 
@@
 
-[We added a sentence to the end of this paragraph that emphasizes the critical need for better monitoring of this vulnerable life stage (lines 313-315)]{style="color: blue"}
+[We added a sentence to the end of this paragraph that emphasizes the critical need for better monitoring of this vulnerable life stage (lines 312-315).]{style="color: blue"}
 
@@
 
-[As described in our response to Reviewer 1-Comment 4, we revised this paragraph to include specific recommendations for ongoing and future efforts to recover Bd-impacted amphibians.
-The added text addresses some of the specific points made by Reviewer 3, including mention of methods such as captive breeding and genetic inventions to facilitate the development of frog resistance against Bd infection (lines 334-354).]{style="color: blue"}
+[As described in our response to Reviewer 1 - Comment 4, we revised this paragraph to include specific recommendations for ongoing and future efforts to recover Bd-impacted amphibians.
+The added text addresses some of the specific points made by Reviewer 3, including mention of methods such as captive breeding and genetic inventions to facilitate the development of frog resistance against Bd infection (lines 335-355).]{style="color: blue"}
 
@@
 
-[We added a paragraph to the Discussion to elaborate on the limitations of natural recovery and the need for active management via reintroductions to reestablish frogs at the landscape scale (lines 316-333).
+[We added a paragraph to the Discussion to elaborate on the limitations of natural recovery and the need for active management via reintroductions to reestablish frogs at the landscape scale (lines 316-334).
 The specific scenario that the reviewer mentions is described in some detail in that paragraph (see also our response to the "general comment" from Reviewer 3 above regarding natural recovery versus recovery via reintroductions).]{style="color: blue"}
@@
 
-[We added a sentence to this paragraph that briefly describes the methods we used to determine *R. sierrae* absence (lines 376-378).]{style="color: blue"}
+[We added a sentence to this paragraph that briefly describes the methods we used to determine *R. sierrae* absence (lines 377-379).]{style="color: blue"}

### 2024-08-04 doc/manuscript/supplementary_information.qmd
Final version for re-submission

@@
 csl: nature.csl
-link-bibliography: false
+link-bibliography: true
 ---
 
-### Supplementary Information for
+## Supplementary Information for:
 
-## Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
+# Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease
 
@@
 
-\newpage
-
-## Supplementary Methods
-
 \
 
-### Frog population recovery
+# Supplementary Methods
+
+## Frog population recovery
 
@@
 By modeling survival estimates obtained from site-specific mrmr CMR analyses, we are in effect conducting an among-site meta-analysis.
-Although it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (namely, run time and memory).
+Although it would theoretically be possible to estimate survival covariate effects in a joint CMR model that integrates capture histories across all sites, this was impractical due to computational requirements of the CMR models (i.e., run time and memory).
 
@@
 
-### Population viability modeling
+## Population viability modeling
 
@@
 The baseline parameter values for the model and how they were estimated are given in @tbl-param_values.
-Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Materials and Methods - Frog populaton recovery - Estimation of frog survival and abundance** and **Supporting Information - Frog population recovery - CMR model structure** for details).
-For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs SI).
+Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR models (see **Supplementary Methods - Frog population recovery - Estimation of frog survival and abundance** and **CMR model structure** for details).
+For populations where we had a sufficient number of PIT-tagged, naturally-recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$ (@fig-compare_surv_probs).
 For populations lacking sufficient numbers of naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.
 
-To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supporting Information - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)).
+To estimate the $\sigma_{A_R}$, we used the posterior distribution of predicted true states for naturally-recruited individuals (1=not recruited, 2=alive, 3=dead, as described in **Supplementary Methods - Frog population recovery - CMR model structure**), then calculated the posterior probability of individuals surviving between consecutive primary periods, conditional on being alive in the first primary period (e.g., given a value of 2 (alive) in the first primary period, how often was the value still 2 (alive) in the next primary period compared to 3 (dead)).
 This yielded posterior distributions for survival probabilities between primary periods.
@@
 First, we considered a deterministic version of our model with no yearly heterogeneity in year-1 juvenile survival probability $\sigma_{J_1}$, and calculated the predicted long-run growth rate $\lambda$ of a population for different values of $\sigma_{A_R}$ and $\sigma_{J_1}$.
-We then fixed $\sigma_{J_1} = 0.1-$ and calculated the predicted growth rate of our 12 populations.
+We then fixed $\sigma_{J_1}$ = 0.1 and calculated the predicted growth rate of our 12 populations.
 
-Second, we performed a local elasticity and sensitivity analysis on $\lambda$ with respect to parameters $\sigma_{L_1}$, $\sigma_{L_2}$, $\sigma_{L_3}$, $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, $F$, $p_{L_2}$, $p_{J_1}$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp SI).
+Second, we performed a local elasticity and sensitivity analysis on $\lambda$ with respect to parameters $\sigma_{L_1}$, $\sigma_{L_2}$, $\sigma_{L_3}$, $\sigma_{J_1}$, $\sigma_{J_2}$, $\sigma_{A_R}$, $F$, $p_{L_2}$, $p_{J_1}$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).
 Note that $\lambda$ is equally or more sensitive to changes in adult survival ($\sigma_{A_R}$) and survival of year-1 juveniles ($\sigma_{J_1}$) than it is to other model parameters, providing additional justification for focusing on variation in $\sigma_{A_R}$ and $\sigma_{J_1}$ in our viability analyses.
@@
 
-### References
+## Supplementary References
 
@@
 
-### Figures
+## Supplementary Figures
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in Supporting Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
+All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
 
@@
 Sensitivities of all other parameters are comparable.
-We did not include $p_{L_1}$ as this parameter is 1 across the populations in this study ([@tbl-param_values]) -- tadpoles spend at least a year as tadpoles before metamorphosing.
-Elasticity and sensitivity are calculated at the default parameter values for each population given in Table [@tbl-param_values].](figures/pop_viability_figures_for_supp.png){#fig-viability-supp width="95%"}
+Elasticity and sensitivity are calculated at the default parameter values for each population ([@tbl-param_values]).
+We did not include $p_{L_1}$ in this analysis because $p_{L_1}$ = 1 for all populations in this study -- tadpoles spend at least a year as tadpoles before metamorphosing ([@tbl-param_values]).](figures/pop_viability_figures_for_supp.png){#fig-viability-supp width="95%"}
 
@@
 
-### Tables
+## Supplementary Tables
 
-| **Parameter**                                                             | **Value**            | **Source**                                                                        |
-|:-------------------------|:-----------------|:---------------------------|
-| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge                |
-| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles.          | Varies yearly        | Varies. Bounds estimated from field data, observations, natural history knowledge |
-| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge                |
+| **Parameter^a^**                                                          | **Value**            | **Source**                                             |
+|:-------------------------|:------------------|:--------------------------|
+| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations                |
+| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles            | 0.7                  | Estimated from field data, observations                |
+| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles            | 0.7                  | Estimated from field data, observations                |
+| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles.          | Varies yearly        | Varies. Bounds estimated from field data, observations |
+| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles           | 0.5                  | Estimated from field data, observations                |
 | $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies                             |
 | $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from CMR studies                             |
-| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge                |
-| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
-| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge                |
+| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations                |
+| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations                |
+| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations                |
 | $p_F$, Probability of an adult female reproducing in a year               | 0.5                  | Could be as high at 1, based on field observations     |
-| $\rho$, Proportion of adults that are female                              | 0.5                  | Estimated from field data, observations, natural history knowledge                |
+| $\rho$, Proportion of adults that are female                              | 0.5                  | Estimated from field data                              |
 | $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                     |
 
-: **Description and values of parameters used in the population viability model.** <!--# MOVE NEXT SENTENCE TO FOOTNOTE.
---> All survival probabilities are in the presence of the fungal pathogen Bd.
-{#tbl-param_values}
+: **Description and values of parameters used in the population viability model.** {#tbl-param_values}
+
+^a^All survival probabilities are in the presence of the fungal pathogen Bd.

### 2024-08-04 doc/manuscript/translocation.qmd
Final version for re-submission

@@
 Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality [@joseph2018; @vredenburg2010] (Supplementary Fig.
-1) .
+1).
 Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; Supplementary Fig.
@@
 Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
-These estimates were made using site-specific models analyzed using the mrmr package.
+These estimates were made using site-specific models analyzed using the mrmr package [@joseph2019] (<https://snarl1.github.io/mrmr/index.html>).
 We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in an among-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).
@@
 We suggest this resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R.
-sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
+sierrae* (see **Methods - Frog population recovery - Field methods** for details).
 This increased survival has direct implications for population viability (see **Results - Long-term population viability**).
@@
 This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual-level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
-The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (Supplemental Fig. 2).
-Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, Supplemental Fig. 2).
-Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, Supplemental Fig. 2).
+The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (Supplementary Fig. 2).
+Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, Supplementary Fig. 2).
+Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, Supplementary Fig. 2).
 The absence of Bd load as an important predictor of frog survival is consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.
@@
 Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
-In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (Supplemental Fig.
+In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (Supplementary Fig.
 3).
@@
 Results from a sensitivity analysis are provided in Supplementary Fig.
-4)
+4.
 
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.10, at least six of twelve translocated populations are likely to experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]A; median predicted $\lambda$ ranges from 1.07-1.28 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]a) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.10, at least six of twelve translocated populations are likely to experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]a; median predicted $\lambda$ ranges from 1.07-1.28 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
-As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of seven of twelve populations was greater than 1 ([@fig-viability]A).
+As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of seven of twelve populations was greater than 1 ([@fig-viability]a).
 
 Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
-Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 ([@fig-viability]B).
+Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 ([@fig-viability]b).
 This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in juvenile survival.
 In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
-To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history ([@fig-viability]C; population 70550, surveyed for 16 years).
+To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history ([@fig-viability]c; population 70550, surveyed for 16 years).
 
@@
 The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies (see Figure 4 in [@knapp2016]), and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
-Finally, reintroductions of frogs collected from Bd-naive populations and reintroduced into Bd-positive sites have consistently failed due to the development of high Bd loads and resulting high frog mortality, suggesting the importance of frog resistance in population reestablishment in the presence of Bd.
+Finally, reintroductions of *R.
+sierrae* collected from Bd-naive populations and reintroduced into Bd-positive sites have consistently failed due to the development of high Bd loads and resulting high frog mortality, suggesting the importance of frog resistance in population reestablishment in the presence of Bd.
 
@@
 Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
-Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Materials and Methods - Frog population recovery - Field methods** for a brief description of these refinements).
-The fact that the relatively low Bd loads and correspondingly high frog survival were both maintained when frogs were moved from donor populations to recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
+Year-1 survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined; see **Methods - Frog population recovery - Field methods** for a brief description of these refinements).
+The fact that the relatively low Bd loads and correspondingly high frog survival were both maintained when frogs were moved from donor populations to recipient sites indicates that these qualities of naturally-recovering populations were not solely an effect of site characteristics, but were also strongly influenced by intrinsic characteristics of frogs, including resistance.
 Although it could be argued that the relatively invariant Bd loads before versus after translocation are a consequence of similar pathogen pressure in the donor and reintroduced populations, this is at odds with the fact that in the first year after translocation frog densities are typically 1-2 orders of magnitude lower in the reintroduced versus donor populations and pathogen pressure should follow a similar pattern.
@@
 Specifically, we are conducting translocations in which frogs from several donor populations are added concurrently to a single site.
-Survival of frogs from each donor population is estimated using CMR and the contribution of each donor population to subsequent generations is quantified using genetic methods.
-
-<!--# Include paragraph describing implications for habitat quality on frog survival in the presence of Bd?
-This would blunt the view of PNAS editor that we put frogs in the "wrong" places.
-In the presence of Bd-caused mortality, habitats that once would have supported frog populations may no longer do so because the extra mortality changes population growth from positive to negative.
-As such, choosing the highest quality habitats may be important for maximizing the probability of population reestablishment.
--->
+Survival of frogs from each donor population is estimated using CMR methods and the contribution of each donor population to subsequent generations is quantified using genetic methods.
 
@@
 sierrae* populations despite ongoing Bd infection is particularly important because natural recovery alone appears insufficient to allow reestablishment of populations at the landscape scale.
-Naturally recovering populations in Yosemite are relatively rare and are often isolated from other such populations and from suitable but vacant habitats.
+Naturally recovering *R. sierrae* populations in Yosemite are relatively rare and are often isolated from other such populations and from suitable but vacant habitats.
 This isolation is the consequence of both natural fragmentation due to rugged geography and the ongoing presence of introduced, predatory trout in the majority of lakes and streams [@bradford1993].
@@
 Successful reestablishment of one *R. sierrae* population in such areas has produced a source of frogs from which adults and juveniles are now dispersing into adjacent habitats.
-This process of translocation, establishment, and subsequent dispersal has significant potential to eventually allow the reestablishment of robust metapopulations that may be more resilient to ongoing and future natural and anthropogenic impacts than more fragmented populations (e.g., [@heard2015]).
+This process of translocation, establishment, and subsequent dispersal has significant potential to eventually allow the reestablishment of robust metapopulations that may be more resilient to current and future natural and anthropogenic impacts than more fragmented populations (e.g., [@heard2015]).
 
@@
 
-In conclusion, we now have a proven strategy to reestablish extirpated *R. sierrae* populations,.
+In conclusion, we now have a proven strategy to reestablish extirpated *R. sierrae* populations.
 However, recovery across their large historical range will require substantial resources over many decades.
@@
 As is typical for recovering *R.
-sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected [@vredenburg2010; @joseph2018] (log~10~(load) $\approx$ 5.78 copies).
+sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected [@vredenburg2010; @joseph2018] (log~10~(load) $\approx$ 5.78 ITS copies).
 Recipient sites to which frogs were translocated were chosen based on previous *R.
@@
 Frogs were transported to the recipient site either on foot or via helicopter.
-Following release, we visited translocated populations approximately once per month during the summer active season to conduct diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits/primary periods per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
+Following release, we visited translocated populations approximately once per month during the summer active season to conduct diurnal CMR surveys and VES (summer active season is generally July-August but can start as early as May and end as late as September; range of survey dates = May-25 to Sep-29, range of translocation dates = Jun-28 to Sep-02; median number of visits (i.e., primary periods; see below) per summer = `r round(surveys_yr_median)`, range = `r surveys_yr_min`--`r surveys_yr_max`).
 CMR surveys allowed estimation of adult survival, recruitment of new adults, and adult population size.
@@
 These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
+The above-described collection of data from live *R. sierrae* was approved under the University of California-Santa Barbara IACUC protocol #478.
 
@@
 
-For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [@rsoftware2022] (v4.4.4) (see **Supporting Information - Frog population recovery - CMR model structure** for details).
+For each translocation site, we estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a site-specific Bayesian open-population Jolly-Seber CMR model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [@rsoftware2022] (v4.4.4) (see **Supplementary Methods - Frog population recovery - CMR model structure** for details).
 Briefly, the model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [@joseph2018].
@@
 Group-level (random) effects included site_id, translocation_id, or translocation_id nested within site_id.
-We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software (v4.4.4, [@rsoftware2022]).
+We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [@rsoftware2022] (v4.4.4).
 For all models, we used default, weakly informative priors, four chains, and 5000 iterations each for warmup and post-warmup.
 We checked MCMC convergence using trace plots and Rhat, and evaluated model fit using leave-one-out cross-validation [@vehtari2016], as implemented by the loo package [@vehtari2022].
-(See **Supporting Information - Frog population recovery - Among-site survival modeling** for details.)
+(See **Supplementary Methods - Frog population recovery - Among-site survival modeling** for details.)
 
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004] (see **Supporting Information - Frog population recovery - Laboratory methods** for details).
+We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004] (see **Supplementary Information - Frog population recovery - Laboratory methods** for details).
 To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
@@
 
-The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year \$p_F\$, the percentage of adults that are female $\rho$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
+The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, the percentage of adults that are female $\rho$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$.
 First-year juvenile survival $\sigma_{J1}$ is the parameter that we think is most influenced by environmental stochasticity.
@@
 Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
-Additional details are provided in **Supporting Information - Population viability modeling - Incorporating yearly variability in survival rates** and **Estimating model parameters**.
+Additional details are provided in **Supplementary Methods - Population viability modeling - Incorporating yearly variability in survival rates** and **Estimating model parameters**.
 
@@
 First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
-Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (Supplemental Fig. 4).
+Second, we compute the elasticity and sensitivity of $\lambda$ to nine model parameters to quantify how much changes in these parameters affected the long-run growth rate (Supplementary Fig. 4).
 This also helped us determine where in the model environmental variation in juvenile survival would have the largest effects on population dynamics.
@@
 Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of *R. sierrae* following reintroductions.
-Additional details are provided in **Supporting Information - Population viability modeling - Model analysis and simulation**.
+Additional details are provided in **Supplementary Methods - Population viability modeling - Model analysis and simulation**.
 

### 2024-09-26 doc/manuscript/supplementary_information.qmd
Add equation numbers

@@
 
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$
+$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$${#eq-logitpj}
 
@@
 \end{blockarray}
-$$
+$${#eq-transmat}
 
@@
 \end{blockarray}
-$$ where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
+$${#eq-transmat2} where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
 
@@
 \end{blockarray}
-$$
+$${#eq-transmat3}
 
@@
 
-$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$
+$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$${#eq-logitlambda}
 
@@
 
-$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$
+$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$${#eq-logitphi}
 
@@
 
-$$\alpha^{(\lambda)} \sim N(0,1),$$ $$\sigma^{(\lambda)} \sim N_+(0,1),$$ $$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$ for periods $t=1, ..., T$.
+$$
+\begin{aligned}
+\alpha^{(\lambda)} &\sim N(0,1) \\
+\sigma^{(\lambda)} &\sim N_+(0,1) \\
+\epsilon^{(\lambda)}_t &\sim N(0, \sigma^{(\lambda)}),
+\end{aligned}
+$${#eq-priors1}
+for periods $t=1, ..., T$.
 Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
@@
 
-$$\beta^{(\phi)} \sim N(0, 1),$$ $$\sigma^{(\phi)} \sim N_+(0, 1),$$ $$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$ for time $t=1, ..., T$.
+$$
+\begin{aligned}
+\beta^{(\phi)} &\sim N(0, 1) \\
+\sigma^{(\phi)} &\sim N_+(0, 1) \\
+\epsilon^{(\phi)}_t &\sim N(0, \sigma^{(\phi)})
+\end{aligned}
+$${#eq-priors2}
+for time $t=1, ..., T$.
 
@@
 
-```{=tex}
-\begin{multline*}
-[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
-\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
-\end{multline*}
-```
+$$
+\begin{aligned}
+\hspace{0.001cm} [\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \mathbf{Y}] \propto &\prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
+&\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} | \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
+\end{aligned}
+$${#eq-likelihood}
+
 where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
@@
 
-$$y_i \sim \text{Bernoulli}(p_i),$$
+$$y_i \sim \text{Bernoulli}(p_i),$${#eq-bernoulli}
 
@@
 
-$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$
+$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$${#eq-logitmodel}
 
@@
 
-$$\alpha \sim \text{Normal}(0, 2.5),$$ and
-
-$$\beta_k \sim \text{Normal}(0, 5),$$
-
+$$
+\begin{aligned}
+\alpha &\sim \text{Normal}(0, 2.5) \\
+\beta_k &\sim \text{Normal}(0, 5),
+\end{aligned}
+$${#eq-priors3}
 for $k=1, ..., K$ where $K$ is the number of predictor variables.

### 2024-09-26 doc/manuscript/supplementary_information.qmd
Changes made in response to editorial checklist

--- 2024-08-04 doc/manuscript/supplementary_information.qmd
@@
 
-```{=tex}
 \begin{multline*}
@@
 \end{multline*}
-```
+
 where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.
@@
 | **Parameter^a^** | **Value** | **Source** |
-|:-------------------------|:------------------|:--------------------------|
+|:---|:---|:---|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles | 0.7 | Estimated from field data, observations |

### 2024-09-26 doc/manuscript/translocation.qmd
Add equation numbers
Changes made in response to editorial checklist

@@
 However, some populations persist and eventually recover, and frogs in these recovering populations have increased resistance against infection.
-We conducted a 15-year reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
+Here, we conduct a 15-year reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
 In addition, the likelihood of establishment is influenced by site, cohort, and frog attributes.
 Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
-These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for amphibians and many other taxa that are threatened with extinction by novel pathogens.
+These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for amphibians and other taxa that are threatened with extinction by novel pathogens.
 
@@
 Third, estimate the probability of persistence for the reintroduced populations over a multi-decadal period, thereby extending our inferences of population recovery beyond the temporal bounds of our reintroduction study (@fig-recovery-model).
-We did this using a stage-structured matrix model.
+We did this using a stage-structured matrix model ("population viability model").
+
+Here, we show that the majority of reintroduced populations showed evidence of successful reproduction and recruitment despite the presence of Bd.
+In addition, the likelihood of establishment is influenced by site, cohort, and frog attributes, but not by Bd load.
+Finally, results from the population viability model suggest that many reintroduced populations have a low probability of extinction over 50 years.
+These results indicate that reintroduction of resistant individuals can allow the recovery of species driven to near-extinction by novel infectious diseases.
 
@@
 
-### Acknowledgments
+### Acknowledgements
 

### 2024-09-26 doc/manuscript/supplementary_information.qmd
Auto-formats in Quarto

@@
 
-$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$${#eq-logitpj}
+$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$ {#eq-logitpj}
 
@@
 \end{blockarray}
-$${#eq-transmat}
+$$ {#eq-transmat}
 
@@
 \end{blockarray}
-$${#eq-transmat2} where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
+$$ {#eq-transmat2} where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.
 
@@
 \end{blockarray}
-$${#eq-transmat3}
+$$ {#eq-transmat3}
 
@@
 
-$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$${#eq-logitlambda}
+$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$ {#eq-logitlambda}
 
@@
 
-$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$${#eq-logitphi}
+$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$ {#eq-logitphi}
 
@@
 \end{aligned}
-$${#eq-priors1}
-for periods $t=1, ..., T$.
+$$ {#eq-priors1} for periods $t=1, ..., T$.
 Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.
@@
 \end{aligned}
-$${#eq-priors2}
-for time $t=1, ..., T$.
+$$ {#eq-priors2} for time $t=1, ..., T$.
 
@@
 \end{aligned}
-$${#eq-likelihood}
+$$ {#eq-likelihood}
 
@@
 
-$$y_i \sim \text{Bernoulli}(p_i),$${#eq-bernoulli}
+$$y_i \sim \text{Bernoulli}(p_i),$$ {#eq-bernoulli}
 
@@
 
-$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$${#eq-logitmodel}
+$$\text{logit}(p_i) = \alpha + X_i \pmb \beta + \pmb \nu_{g[1:N]},$$ {#eq-logitmodel}
 
@@
 \end{aligned}
-$${#eq-priors3}
-for $k=1, ..., K$ where $K$ is the number of predictor variables.
+$$ {#eq-priors3} for $k=1, ..., K$ where $K$ is the number of predictor variables.
 

### 2024-09-26 doc/manuscript/supplementary_information.qmd
Buffer equation number in main text. Auto table update

@@
 | **Parameter^a^**                                                          | **Value**            | **Source**                                             |
-|:---|:---|:---|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations                |

### 2024-09-27 code/extinction_and_growth_rate_analysis.R
Update figure equations for Nat Comm formatting

@@ -128,11 +128,11 @@ ptile = ggplot(dat) + geom_tile(aes(x=muR, y=juv_surv, fill=lambda)) +
 		   				scale_color_manual(values=colors) +
 							# scale_fill_gradient("\u03bb", low = "white", high = "black") +
 							scale_fill_gradient2(midpoint=1) + #guide=guide_legend(title.position="bottom")) +
-							theme_classic() + xlab(bquote("Adult survival probability (\u03c3"~.[AR]~")")) +
-							ylab(bquote("Juvenile (Y1) survival probability (\u03c3"~.[J1]~")")) + 
+							theme_classic() + xlab(expression(paste("Adult survival probability,", italic(sigma[A[R]])))) +
+							ylab(expression(paste("Juvenile (Y1) survival probability, ", italic(sigma[J[1]])))) +
 							annotate("text", x=0.25, y=0.05, label="Population declines", size=3) +
 							annotate("text", x=0.7, y=0.2, label="Population grows", size=3) +
-							guides(color="none", fill=guide_colorbar(title="\u03bb")) + theme(legend.position="bottom", 
+							guides(color="none", fill=guide_colorbar(title=expression(italic(lambda)))) + theme(legend.position="bottom", 
 																					 legend.text=element_text(size=10), 
 																					 legend.title=element_text(size=10))
 
@@ -180,16 +180,21 @@ lake_sens_dt$lake = names(lake_sens)
 
 lake_sens_dt_melt = melt(lake_sens_dt, id.vars="lake")
 
+
+xtick_labels = expression(italic(sigma[A[R]]), italic(F), italic(sigma[J[1]]), italic(sigma[J[2]]), 
+											    italic(p[L[2]]), italic(p[J[1]]), italic(sigma[L[1]]), italic(sigma[L[2]]), italic(sigma[L[3]]))
 p2 = ggplot(lake_elas_dt_melt) + geom_bar(aes(x=variable, y=value, fill=lake), stat="identity", position="dodge") +
 														scale_fill_manual(values=colors) +
-														xlab("Parameter") + ylab("Elasticity of \U03BB") + theme_classic() +
-														guides(fill="none") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+														xlab("Parameter") + ylab(expression(paste("Elasticity of ", italic(lambda)))) + theme_classic() +
[…]

### 2024-09-27 code/extinction_and_growth_rate_analysis.R
Change xlabel match other labels

@@ -357,7 +357,7 @@ pext = ggplot(extinction_at_fifty_dt[order(surv_med, lake_id)], aes(x=sigma_J1, 
 							geom_line(aes(color=lake_id_surv)) + 
 							geom_point(aes(color=lake_id_surv)) + 
 							scale_color_manual(values=cdat$col) +
-							theme_classic() + xlab(expression(paste("Mean year 1 juvenile survival, ", italic(sigma[J[1]])))) + ylab("Extinction prob. in 50 years") +
+							theme_classic() + xlab(expression(paste("Mean juvenile (Y1) survival probability, ", italic(sigma[J[1]])))) + ylab("Extinction prob. in 50 years") +
 							guides(color=guide_legend(title=expression(paste("Site, ", italic(sigma[A[R]]))), title.position="top", ncol=3)) + 
 							theme(legend.position="bottom", legend.text=element_text(size=9.2), legend.title=element_text(size=10))
 

### 2024-09-27 doc/manuscript/supplementary_information.qmd
Changes made in response to editorial checklist

--- 2024-09-26 doc/manuscript/supplementary_information.qmd
@@
 Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region.
-Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.
-Samples sizes are provided immediately above the x-axis.
+Box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots).
+Sample sizes are provided immediately above the x-axis.
 The red dashed horizontal line indicates the approximate load above which severe disease is typically observed @vredenburg2010.
-This threshold (5.8 ITS1 copies - on a log10 scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations @briggs2010.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
+This threshold (5.8 ITS1 copies - on a log10 scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations @briggs2010.
+Source data are provided as a Source Data file.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}
 
@@
 Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities.
-All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
+All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters.
+Source data are provided as a Source Data file.
+](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
 
@@
 | **Parameter^a^** | **Value** | **Source** |
-|:---|:---|:---|
+|:-----------------------|:-----------------------|:-----------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles | 0.7 | Estimated from field data, observations |

### 2024-09-27 doc/manuscript/translocation.qmd
Buffer equation number in main text. Auto table update
Changes made in response to editorial checklist

@@
 
-^6^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318
+^6^Resources Management and Science Division, Yosemite National Park, El Portal, CA, 95318
 
@@
 
+The collection of data from live *R.
+sierrae* described here complies with all ethical regulations and was approved under the University of California-Santa Barbara Institutional Animal Care and Use Committe (IACUC) protocol number 478.
 For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected based on several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
@@
 These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
-The above-described collection of data from live *R. sierrae* was approved under the University of California-Santa Barbara IACUC protocol #478.
 
@@
 \end{bmatrix}(t)
-$$
+$$ {#eq-matrix}
 
@@
 
-The data used in this study are available at <https://github.com/SNARL1/translocation>.
-The CMR datasets from each reintroduction site are available at <https://github.com/SNARL1/cmr-analysis>.
+The data used in this study are available at <https://github.com/SNARL1/translocation> [@knapp2024a].
+This includes the estimated survival of each reintroduced frog cohort, data used to identify predictors of 1-year frog survival, data used to estimate growth rates and extinction probabilities of reintroduced populations, Bd loads on frogs before and after translocation, and survival probabilities of translocated versus naturally recruited adults.
+For the relevant figures, source data are provided in a Source Data file.
+The CMR datasets from each reintroduction site that we used to calculate post-translocation frog survival are available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].
 
@@
 
-R code to replicate our analyses is available at <https://github.com/SNARL1/translocation>.
-R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis>.
+R code to replicate our analyses is available at <https://github.com/SNARL1/translocation> [@knapp2024a].
+R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].
 
@@
 Rose (laboratory work); numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S.
-Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
+Fish and Wildlife Service, Sierra Nevada Aquatic Research Laboratory (DOI: 10.21973/N3966F), and University of California-Santa Barbara Institutional Animal Care and Use Committee (research permits, logistical support, and/or field assistance).
 This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C.
@@
 **a** winter severity in the year following translocation, **b** elevation of recipient site, **c** donor population, and **d** sex.
-In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals.
-In (c) and (d), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
+In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals for the posterior distributions of the predicted effects.
+In (c) and (d), box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots) for the posterior distributions of the predicted effects.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
 

### 2024-09-27 doc/manuscript/supplementary_information.qmd

@@
 | **Parameter^a^**                                                          | **Value**            | **Source**                                             |
-|:-----------------------|:-----------------------|:-----------------------|
+|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles            | 0.7                  | Estimated from field data, observations                |

### 2024-09-27 doc/manuscript/supplementary_information.qmd
Minor edits

--- 2024-09-27 doc/manuscript/supplementary_information.qmd
@@
     \usepackage{blkarray}
-    \usepackage{lineno}\linenumbers
+#    \usepackage{lineno}\linenumbers # Removed in final version
 include-before-body: 
@@
 All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters.
-Source data are provided as a Source Data file.
-](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
+Source data are provided as a Source Data file.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}
 

### 2024-09-27 doc/manuscript/translocation.qmd
Minor edits
Replace links to GitHub repos with Zenodo DOI

@@
 Here, we show that the majority of reintroduced populations showed evidence of successful reproduction and recruitment despite the presence of Bd.
-In addition, the likelihood of establishment is influenced by site, cohort, and frog attributes, but not by Bd load.
+In addition, post-translocation frog survival is influenced by site, cohort, and frog attributes, but not by Bd load.
 Finally, results from the population viability model suggest that many reintroduced populations have a low probability of extinction over 50 years.
@@
 $$
+\begin{aligned}
 \begin{bmatrix}
@@
 A_T \\
-\end{bmatrix}(t)
+\end{bmatrix}(t) \\
+\hspace{0.0001cm}
+\end{aligned}
 $$ {#eq-matrix}
@@
 
-The data used in this study are available at <https://github.com/SNARL1/translocation> [@knapp2024a].
+The data used in this study are available at <https://doi.org/10.5281/zenodo.13851590> [@knapp2024a].
 This includes the estimated survival of each reintroduced frog cohort, data used to identify predictors of 1-year frog survival, data used to estimate growth rates and extinction probabilities of reintroduced populations, Bd loads on frogs before and after translocation, and survival probabilities of translocated versus naturally recruited adults.
 For the relevant figures, source data are provided in a Source Data file.
-The CMR datasets from each reintroduction site that we used to calculate post-translocation frog survival are available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].
+The CMR datasets from each reintroduction site that we used to calculate post-translocation frog survival are available at <https://doi.org/10.5281/zenodo.13851652> [@knapp2024b].
 
@@
 
-R code to replicate our analyses is available at <https://github.com/SNARL1/translocation> [@knapp2024a].
-R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].
+R code to replicate our analyses is available at <https://doi.org/10.5281/zenodo.13851590> [@knapp2024a].
+R code to analyze the CMR datasets is available at <https://doi.org/10.5281/zenodo.13851652> [@knapp2024b].
 
@@
 
-### Figures
+### Figures and Legends
 

### 2024-09-27 doc/manuscript/supplementary_information.qmd
Edits to render to latex

@@
 | **Parameter^a^** | **Value** | **Source** |
-|:-----------------------|:-----------------------|:-----------------------|
+|:---|:---|:---|
 | $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles | 0.7 | Estimated from field data, observations |

### 2024-09-27 doc/manuscript/translocation.tex
Edits to render to latex

--- 2023-12-21 doc/manuscript/translocation_bioarxiv.tex
@@
-\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
-% Use the lineno option to display guide line numbers if required.
-
-\templatetype{pnasmathematics} % Choose template 
-% {pnasmathematics} = Template for a two-column research article
-% {pnasmathematics} %= Template for a one-column mathematics article
-% {pnasinvited} %= Template for a PNAS invited submission
-
-\title{Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease}
-
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,b]{Roland A. Knapp}
-\author[c,1]{Mark Q. Wilber} 
-\author[d,e,1]{Allison Q. Byrne}
-\author[f,g]{Maxwell B. Joseph}
-\author[a,b]{Thomas C. Smith}
-\author[d,e,h]{Andrew P. Rothstein}
-\author[i]{Robert L. Grasso}
-\author[d,e]{Erica Bree Rosenblum}
-
-\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
-\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
-\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
-\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
-\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
-\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
-\affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
-\affil[h]{Current affiliation: Ginkgo Bioworks, Boston, MA, 02210}
-\affil[i]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}
-
-% Please give the surname of the lead author for the running footer
-\leadauthor{Knapp}
-
-% Please add a significance statement to explain the relevance of your work
-\significancestatement{Understanding how species persist despite accelerating global change is
-critical for the conservation of biodiversity. Emerging infectious
-diseases can have particularly devastating impacts, and few options
-exist to reverse these effects. We used large-scale reintroductions of
-disease-resistant individuals in an effort to recover a once-common frog
-species driven to near-extinction by a disease that has decimated
-amphibian biodiversity. Introduction of resistant frogs allowed
-reestablishment of viable populations in the presence of disease. In
-addition, resistance may be at least partially the result of natural
-selection at specific immune function genes, which show evidence for
-selection in recovering populations. The evolution of resistance and
-reintroductions of resistant individuals could play an important role in
-biodiversity conservation in our rapidly changing world.}
-
-% Please include corresponding author, author contribution and author declaration information
-\authorcontributions{}
-\authordeclaration{}
-\equalauthors{\textsuperscript{1}M.W. contributed equally to this work with A.B.}
-\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: roland.knapp@ucsb.edu}
-
-% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{}
-
-\begin{abstract}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
+
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{blkarray}
+\usepackage{lineno}\linenumbers
+\KOMAoption{captions}{tableheading}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{\textbf{Fig.}}
+\else
+  \newcommand\figurename{\textbf{Fig.}}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
+\usepackage{bookmark}
+
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\author{}
+\date{}
+
+\begin{document}
+
+\captionsetup{format=plain, labelfont=bf, labelsep=space}
+
+\subsection{Reintroduction of resistant frogs facilitates
+landscape-scale recovery in the presence of a lethal fungal
+disease}\label{reintroduction-of-resistant-frogs-facilitates-landscape-scale-recovery-in-the-presence-of-a-lethal-fungal-disease}
+
+Roland A. Knapp\textsuperscript{1,2,}*, Mark Q.
+Wilber\textsuperscript{3}, Maxwell B. Joseph\textsuperscript{4,5},
+Thomas C. Smith\textsuperscript{1,2} \& Robert L.
+Grasso\textsuperscript{6}
+
+\textsuperscript{1}Sierra Nevada Aquatic Research Laboratory, University
+of California, Mammoth Lakes, CA, 93546
+
+\textsuperscript{2}Earth Research Institute, University of California,
+Santa Barbara, CA, 93106-3060
+
+\textsuperscript{3}School of Natural Resources, University of Tennessee
+Institute of Agriculture, Knoxville, TN, 37996
+
+\textsuperscript{4}Earth Lab, University of Colorado, Boulder, CO, 80303
+
+\textsuperscript{5}Current affiliation: Planet, San Francisco, CA, 94107
+
+\textsuperscript{6}Resources Management and Science Division, Yosemite
+National Park, El Portal, CA, 95318
+
+*Corresponding author: email: roland.knapp@ucsb.edu
+
+\subsubsection{Abstract}\label{abstract}
+
 Vast alteration of the biosphere by humans is causing a sixth mass
-extinction, driven in part by an increase in emerging infectious
-diseases. The emergence of the lethal fungal pathogen
-(\emph{Batrachochytrium dendrobatidis}; ``Bd'') has devastated global
-amphibian biodiversity, with hundreds of species experiencing declines
-or extinctions. Given that there are no broadly applicable methods to
-reverse these impacts in the wild, the future of many amphibians appears
-grim. The once-common mountain yellow-legged (MYL) frog is emblematic of
-amphibians threatened by Bd. Although most MYL frog populations are
-extirpated following disease outbreaks, some persist and eventually
-recover. Frogs in these recovering populations have increased resistance
-against Bd infection, consistent with evolution of resistant genotypes
-and/or acquired immunity. We conducted a 15-year landscape-scale
-reintroduction study and show that frogs collected from recovering
-populations and reintroduced to vacant habitats can reestablish
-populations despite the presence of Bd. In addition, results from
-viability modeling suggest that many reintroduced populations have a low
-probability of extinction over 50 years. To better understand the role
-of evolution in frog resistance, we compared the genomes of MYL frogs
-from Bd-naive and recovering populations. We found substantial
-differences between these categories, including changes in immune
-function loci that may confer increased resistance, consistent with
-evolutionary changes in response to Bd exposure. These results provide a
-rare example of how reintroduction of resistant individuals can allow
-the landscape-scale recovery of disease-impacted species. This example
-has broad implications for the many taxa worldwide that are threatened
-with extinction by novel pathogens.
-\end{abstract}
-
-\dates{This manuscript was compiled on \today}
-% \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\begin{document}
-
-\maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-\dropcap{H}uman activities are increasingly impacting global biodiversity
+extinction, driven in part by an increase in infectious diseases. The
+emergence of the lethal fungal pathogen \emph{Batrachochytrium
+dendrobatidis} (Bd) has devastated global amphibian biodiversity. Given
+the lack of any broadly applicable methods to reverse these impacts, the
+future of many amphibians appears grim. The Sierra Nevada yellow-legged
+frog (\emph{Rana sierrae}) is highly susceptible to Bd infection and
+most \emph{R. sierrae} populations are extirpated following disease
+outbreaks. However, some populations persist and eventually recover, and
+frogs in these recovering populations have increased resistance against
+infection. Here, we conduct a 15-year reintroduction study and show that
+frogs collected from recovering populations and reintroduced to vacant
+habitats can reestablish populations despite the presence of Bd. In
+addition, the likelihood of establishment is influenced by site, cohort,
+and frog attributes. Results from viability modeling suggest that many
+reintroduced populations have a low probability of extinction over 50
+years. These results provide a rare example of how reintroduction of
+resistant individuals can allow the landscape-scale recovery of
+disease-impacted species, and have broad implications for amphibians and
+other taxa that are threatened with extinction by novel pathogens.
+
+\subsubsection{Introduction}\label{introduction}
+
+Human activities are devastating global biodiversity
 \citep{ceballos2015}, with important implications for ecosystem
@@
 have experienced Bd-caused declines, and numerous susceptible taxa are
-extinct in the wild \citep{scheele2019}. These impacts to global
-biodiversity may be unprecedented, and highlight the importance of
-understanding mechanisms of species persistence in the presence of
-emerging diseases \citep{russell2020}.
+extinct in the wild \citep{scheele2019}. The escalating frequency of
+disease emergence and the strength of impacts on wild populations
+increasingly highlight the critical need for successful mitigation of
+disease in species conservation.
+
+Following pathogen arrival in a host population, resistance (ability to
+limit pathogen burden) and tolerance (ability to limit the harm caused
+by a particular burden) are key mechanisms by which hosts reduce disease
+impacts \citep{raberg2009}. The effectiveness of these mechanisms has
+direct consequences for the persistence and recovery of host populations
+in the presence of disease \citep{brannelly2021}. Host immunity and
+evolution both play important roles in the development of resistance and
+tolerance, and are key targets for strategies aimed at mitigating
+disease impacts in the wild \citep{garner2016}.
 
 Evidence of natural recovery in the many Bd-impacted amphibian
-populations is surprisingly rare \citep[for notable exceptions,
-see][]{scheele2017, voyles2018, knapp2016}, suggesting that
-disease-caused declines will be difficult to reverse. This apparent low
-resilience to disease effects may be due to the limited ability of many
-amphibians to develop Bd resistance and/or tolerance, which in turn,
-could also lessen the effectiveness of potential Bd mitigation
-strategies. Following pathogen arrival in a host population, resistance
-(ability to limit pathogen burden) and tolerance (ability to limit the
-harm caused by a particular burden) are key mechanisms to reduce disease
-impacts \citep{raberg2009} and facilitate population persistence and
-recovery \citep{brannelly2021}. Host immunity and evolution both play
-important roles in the development of resistance and tolerance, and
-utilizing these factors would seem a promising approach to developing
-effective strategies to mitigate disease impacts in the wild
-\citep{garner2016, woodhams2011}. However, several aspects of the
-amphibian-Bd system present difficult obstacles, including (i) the
-general inability of amphibians to mount an effective immune response
-against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
-(ii) the apparent rarity of evolution of more resistant/tolerant
-genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
-suggest that reintroduction of amphibians into sites to reestablish
+populations is surprisingly rare (for notable exceptions, see
+\citep{scheele2017, voyles2018, knapp2016}). This apparent low
+resilience to disease may be due to the limited ability of many
+amphibians to develop Bd resistance or tolerance via either immunity
+\citep{rosenblum2012, fites2013, grogan2018a} or evolution (but see
+\citep{savage2016, grogan2018b}). In turn, this would also reduce the
+effectiveness of potential Bd mitigation strategies. For example, a
+limited ability by amphibians to develop resistance or tolerance
+suggests that reintroduction of amphibians into sites to reestablish
 populations extirpated by Bd will often result not in population
@@
 
-The mountain yellow-legged (MYL) frog, composed of the sister species
-\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
-emblematic of the global declines of amphibians caused by Bd
+The Sierra Nevada yellow-legged frog (\emph{Rana sierrae}), is
+emblematic of the global decline of amphibians caused by Bd
 \citep{scheele2019}. Once the most common amphibian in the high
-elevation portion of California's Sierra Nevada mountains
-\citep[USA,][]{grinnell1924}, during the past century this frog has
-disappeared from more than 90\% of its historical range
-\citep{vredenburg2007}. Due to the severity of its decline and the
-increasing probability of extinction, both species are now listed as
-``endangered'' under the U.S. Endangered Species Act. In the Sierra
-Nevada, this decline was initiated by the introduction of non-native
-trout into the extensive historically-fishless region
-\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
-of Bd in the mid-1900s and its subsequent spread \citep{vredenburg2019}
-caused additional large-scale population extirpations
-\citep{vredenburg2010, rachowicz2006}. These Bd-caused declines are
-fundamentally different from the fish-caused declines because fish
-eradication is feasible \citep{knapp1998} and results in the rapid
-recovery of frog populations \citep{knapp2007, vredenburg2004}. In
-contrast, Bd appears to persist in habitats even in the absence of
-amphibian hosts \citep{walker2007}, and therefore represents a long-term
-alteration of invaded ecosystems that amphibians will need to overcome
-to reestablish populations.
-
-Despite the catastrophic impact of Bd on MYL frogs, wherein most
+elevation portion of California's Sierra Nevada mountains, USA
+\citep{grinnell1924}, during the past century this frog has disappeared
+from more than 90\% of its historical range \citep{vredenburg2007}. Due
+to the severity of its decline and the increasing probability of
+extinction, \emph{R. sierrae} is now listed as ``endangered'' under the
+U.S. Endangered Species Act. The decline of \emph{R. sierrae} was
+initiated by the introduction of non-native trout into the extensive
+fishless region of the Sierra Nevada \citep{bradford1989, knapp2000}
+starting in the late 1800s. The arrival of Bd in the mid-1900s and its
+subsequent spread \citep{vredenburg2019} caused additional large-scale
+population extirpations \citep{vredenburg2010, rachowicz2006}. These
+Bd-caused declines are fundamentally different from the fish-caused
+declines because fish eradication is feasible \citep{knapp1998} and
+results in the rapid recovery of frog populations
+\citep{knapp2007, vredenburg2004}. In contrast, Bd appears to persist in
+habitats even in the absence of amphibian hosts \citep{walker2007}, and
+therefore represents a long-term alteration of invaded ecosystems that
+amphibians will need to overcome to reestablish populations.
+
+Despite the catastrophic impact of Bd on \emph{R. sierrae}, wherein most
 Bd-naive populations are extirpated following Bd arrival
 \citep{vredenburg2010}, some populations have persisted after epizootics
-\citep[during which Bd infection intensity on frogs is very
-high,][]{briggs2010} and are now recovering
-(Figure~\ref{fig-recovery-model}) \citep{knapp2016}. Frogs in these
-recovering populations show reduced susceptibility to Bd infection
-\citep{knapp2016}, with infection intensity (``load'') on adults
-consistently in the low-to-moderate range
-\citep{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
-is evident even under controlled laboratory conditions
-\citep{knapp2016}, indicative of host resistance against Bd infection
-(and not simply an effect of factors external to individual frogs, e.g.,
-environmental conditions). In addition to frogs from recovering
-populations having higher resistance to Bd infection than those from
-naive populations, they could also have higher tolerance, but no data
-are currently available to evaluate this possibility. Therefore, we
-focus on resistance throughout this paper. The observed resistance of
-MYL frogs could be the result of several non-mutually exclusive
-mechanisms, including natural selection for more resistant genotypes
-\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
-and/or inherent between-population differences that pre-date Bd
-exposure. The possible evolution of MYL frog resistance and subsequent
-population recovery is consistent with that expected under
-``evolutionary rescue'', whereby rapid evolutionary change increases the
-frequency of adaptive alleles and restores positive population growth
-\citep{carlson2014, searle2020}. This intriguing possibility also
-suggests an opportunity to expand recovery beyond the spatial scale
-possible under natural recovery by utilizing resistant frogs from
-recovering populations in reintroductions to vacant habitats
-(Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.
-
-In the current study, we had three primary objectives. First, to
-determine whether the reintroduction of resistant MYL frogs obtained
-from populations recovering from Bd-caused declines allows the
-successful reestablishment of extirpated populations despite ongoing
-disease, we conducted a 15-year landscape-scale frog reintroduction
-effort (Figure~\ref{fig-recovery-model}). Second, to extend our
-inferences of population recovery well beyond the temporal extent of our
-reintroduction study, we developed a model to estimate the probability
-of persistence for the reintroduced populations over a multi-decadal
-period (Figure~\ref{fig-recovery-model}). Third, given the importance of
-resistance for frog survival, population establishment, and long-term
-viability (this study), we also conducted a genomic study to determine
-whether MYL frogs show an evolutionary response to Bd exposure, and
-whether these genomic changes are associated with resistance. To
-accomplish this objective, we used exome capture methods to identify
-genomic regions with consistent differences between frogs from naive
-versus recovering populations (Figure~\ref{fig-recovery-model}).
-
-\section*{Results}
-
-\subsection*{Frog population recovery}
-
-To determine whether MYL frogs from recovering populations can be used
-to reestablish extirpated populations, we conducted 24 reintroductions
-in Yosemite National Park (2006--2020). Each of the reintroductions
-involved collection of adult frogs from 1 of 3 recovering, Bd-positive
-``donor'' populations and translocating them to 1 of 12 nearby recipient
-sites (Figure S1). The donor populations included 2 of
-the 5 recovering populations included in the frog evolution study
-referenced above (Figure~\ref{fig-allelefrequencies}: population 1 and
-4), and these donor populations contributed frogs for 20 of the 24
-translocations. Following translocation, we estimated adult survival and
+\citep{briggs2010} and are now recovering \citep{knapp2016}
+(Fig.~\ref{fig-recovery-model}). During the epizootic phase, \emph{R.
+sierrae} are highly susceptible to Bd infection as indicated by very
+high infection intensities (``load''). In contrast, frogs in recovering
+populations have reduced susceptibility to Bd infection
+\citep{knapp2016}, with loads on adults consistently in the
+low-to-moderate range \citep{briggs2010, knapp2011, joseph2018}. This
+reduced susceptibility is evident even under controlled laboratory
+conditions \citep{knapp2016}, indicative of host resistance against Bd
+infection and not simply an effect of factors external to individual
+frogs (e.g., environmental conditions). In addition to frogs from
+recovering populations having higher resistance to Bd infection than
+those from naive populations, they could also have higher tolerance, but
+no data are currently available to evaluate this possibility. Therefore,
+we focus on resistance throughout this paper.
+
+The observed resistance of \emph{R. sierrae} could be the result of
+several mechanisms that are not mutually exclusive, including natural
+selection for more resistant genotypes \citep{savage2016, grogan2018b},
+acquired immunity \citep{grogan2018a}, and/or inherent
+between-population differences that pre-date Bd exposure. The possible
+evolution of resistance and subsequent population recovery is consistent
+with that expected under ``evolutionary rescue'', whereby rapid
+evolutionary change increases the frequency of adaptive alleles and
+restores positive population growth \citep{carlson2014, searle2020}.
+This intriguing possibility also suggests an opportunity to expand
+recovery beyond the spatial scale possible under natural recovery by
+using resistant frogs from recovering populations in reintroductions to
+vacant habitats \citep{joseph2018, mendelson2019}
+(Fig.~\ref{fig-recovery-model}).
+
+In the current study, we had three primary objectives. First, determine
+whether the reintroduction of resistant \emph{R. sierrae}, obtained from
+populations recovering from Bd-caused declines, allows the
+reestablishment of populations despite ongoing disease
+(Fig.~\ref{fig-recovery-model}). We accomplished this using a 15-year
+large-scale reintroduction study. Second, identify important site,
+cohort, and individual-level predictors of frog survival using results
+from the reintroduction study. Third, estimate the probability of
+persistence for the reintroduced populations over a multi-decadal
+period, thereby extending our inferences of population recovery beyond
+the temporal bounds of our reintroduction study
+(Fig.~\ref{fig-recovery-model}). We did this using a stage-structured
+matrix model (``population viability model'').
+
+Here, we show that the majority of reintroduced populations showed
+evidence of successful reproduction and recruitment despite the presence
+of Bd. In addition, post-translocation frog survival is influenced by
+site, cohort, and frog attributes, but not by Bd load. Finally, results
+from the population viability model suggest that many reintroduced
+populations have a low probability of extinction over 50 years. These
+results indicate that reintroduction of resistant individuals can allow
+the recovery of species driven to near-extinction by novel infectious
+diseases.
+
+\subsubsection{Results}\label{results}
+
+\paragraph{Frog population recovery}\label{frog-population-recovery}
+
+To determine whether \emph{R. sierrae} from recovering populations can
+be used to reestablish extirpated populations, we conducted 24
+reintroductions in Yosemite National Park (2006--2020). Each of the
+reintroductions involved collection of adult frogs from 1 of 3
+recovering, Bd-positive ``donor'' populations and translocating them to
+1 of 12 nearby recipient sites (Fig.~\ref{fig-yosemap}). The 3 donor
+populations are among the largest \emph{R. sierrae} populations in
+Yosemite. Following translocation, we estimated adult survival and
 recruitment of new adults from capture-mark-recapture (CMR) surveys and
 obtained counts of tadpoles and juveniles from visual encounter surveys
-(VES). Across all translocation sites, the duration of survey time
-series was 1--16 years (median = 5).
+(VES). Across all translocation sites, the duration of CMR surveys and
+VES was 1--16 years (median = 5).
 
@@
 0.79, range = 0.12--1). In summary, survey results indicate that
-translocations resulted in the establishment of reproducing MYL frog
-populations at most recipient sites despite the ongoing presence of Bd.
+translocations resulted in the establishment of reproducing \emph{R.
+sierrae} populations at most recipient sites despite the ongoing
+presence of Bd.
 
@@
 chytridiomycosis (i.e., the disease caused by Bd) and associated frog
-mortality (Figure S2)
-\citep{joseph2018, vredenburg2010}. Although it is possible that the
-observed relatively small changes in load are a consequence of
-individuals with high Bd loads dying and therefore being unavailable for
-sampling during the post-translocation period, the fact that there was
-little difference in pre- versus post-translocation Bd loads even in
-those populations that had very high frog survival (70556, 74976;
-Figure~\ref{fig-translocation-survival},
-Figure S2) suggests a true lack of
-substantial change in Bd load.
+mortality \citep{joseph2018, vredenburg2010} (Supplementary Fig. 1).
+Although it is possible that the observed relatively small changes in
+load are a consequence of individuals with high Bd loads dying and
+therefore being unavailable for sampling during the post-translocation
+period, the fact that there was little difference in pre- versus
+post-translocation Bd loads even in those populations that had very high
+frog survival (70556, 74976 - see below; Supplementary Fig. 1) suggests
+a true lack of substantial change in Bd load.
 
@@
 self-sustaining population. Given that it can take years or even decades
-to determine the self-sustainability of a reintroduced population
-\citep[for an example in MYL frogs, see][]{joseph2018}, the use of
+to determine the self-sustainability of a reintroduced population (for
+an example in \emph{R. sierrae}, see \citep{joseph2018}), the use of
 proxies is essential for providing shorter-term insights into
@@
 period immediately following translocation. These estimates were made
-using site-specific models analyzed using the mrmr package. We use these
-estimates to describe general patterns of frog survival in all
+using site-specific models analyzed using the mrmr package
+\citep{joseph2019} (\url{https://snarl1.github.io/mrmr/index.html}). We
+use these estimates to describe general patterns of frog survival in all
 translocated cohorts, and in an among-site meta-analysis of frog
@@
 recipient sites (for the subset of sites that received multiple
-translocations; Figure~\ref{fig-translocation-survival}). These patterns
+translocations; Fig.~\ref{fig-translocation-survival}). These patterns
 indicate an important effect of site characteristics on frog survival.
@@
 primarily from our improved ability to choose recipient sites with
-higher habitat quality for \emph{R. sierrae} (see \textbf{Materials and
-Methods - Frog population recovery - Field methods} for details). This
-increased survival has direct implications for population viability (see
+higher habitat quality for \emph{R. sierrae} (see \textbf{Methods - Frog
+population recovery - Field methods} for details). This increased
+survival has direct implications for population viability (see
 \textbf{Results - Long-term population viability}).
@@
 individual-level characteristics as predictors and 1-year frog survival
-(Figure~\ref{fig-translocation-survival}) as the response variable. The
+(Fig.~\ref{fig-translocation-survival}) as the response variable. The
 best model of 1-year frog survival identified several important
 predictors, but Bd load at the time of translocation was not among them
-(Figure S3). Instead, important predictors
-included winter severity in the year following translocation (snow\_t1),
-site elevation, and donor population (Figure~\ref{fig-cond-effects},
-Figure S3). Males had somewhat higher
-survival than females, but this effect was small
-(Figure~\ref{fig-cond-effects}, Figure S3).
-The absence of Bd load as an important predictor of frog survival is
-consistent with frogs in recovering populations having sufficient
-resistance to suppress Bd loads below harmful levels.
-
-In summary, results from our frog translocation study indicate that
-translocations resulted in (i) relatively high 1-year survival of
-translocated adults, as well as reproduction and recruitment, at the
-majority of recipient sites, (ii) 1-year survival of adults is
-influenced by site characteristics, weather conditions, and donor
-population (but not Bd load), and (iii) based on the relatively small
-changes in Bd load after translocation, loads appear more strongly
-influenced by frog characteristics (e.g., resistance) than site
-characteristics. Together, these results indicate that frogs
-translocated from recovering populations can maintain the benefits of
-resistance in non-natal habitats. In addition, in 3 locations where
-longer CMR time series allowed us to assess the survival of new adults
-recruited to the population, naturally-recruited adults had equivalent
-or higher survival probabilities than the originally translocated adults
-(Figure S4). This suggests that frog
-resistance is maintained across generations. All of the conditions
-described above are supportive of population establishment and long-term
-population growth.
-
-\subsection*{Long-term population viability}
-
-Results from the frog translocation study indicated that most
-populations showed evidence of successful reproduction and recruitment,
-and that adult survival was often relatively high (described above).
-Although suggestive of population establishment, a decade or more of
-surveys may be necessary to confirm that populations are in fact
-self-sustaining \citep{joseph2018}. To extend our inferences of
-population establishment beyond those possible from the site-specific
-CMR data, we developed a population viability model. Specifically, to
-test whether the observed yearly adult survival probabilities in
-translocated populations (provided in legend of
-Figure~\ref{fig-viability}B) were sufficient for long-term viability,
-we built a stage-structured matrix model that captured known frog
-demography and included demographic and environmental stochasticity. We
-parameterized the model using CMR data from translocated populations and
-known life history values in this system (Table S2
-SI).
+(Supplementary Fig. 2). Instead, important predictors included winter
+severity in the year following translocation (snow\_t1), site elevation,
+and donor population (Fig.~\ref{fig-cond-effects}, Supplementary Fig.
+2). Males had somewhat higher survival than females, but this effect was
+small (Fig.~\ref{fig-cond-effects}, Supplementary Fig. 2). The absence
+of Bd load as an important predictor of frog survival is consistent with
+frogs in recovering populations having sufficient resistance to suppress
+Bd loads below harmful levels.
+
+In summary, results from our frog translocation study indicate that (i)
+translocations produced relatively high 1-year survival of translocated
+adults, as well as reproduction and recruitment, at the majority of
+recipient sites, (ii) 1-year survival of adults is influenced by site
+characteristics, weather conditions, and donor population (but not Bd
+load), and (iii) based on the relatively small changes in Bd load after
+translocation, loads appear more strongly influenced by frog
+characteristics (e.g., resistance) than site characteristics. Together,
+these results indicate that frogs translocated from recovering
+populations can maintain the benefits of resistance in non-natal
+habitats. In addition, in 3 locations where longer CMR time series
+allowed us to assess the survival of new adults recruited to the
+population, naturally-recruited adults had equivalent or higher survival
+probabilities than the originally translocated adults (Supplementary
+Fig. 3). This suggests that frog resistance is maintained across
+generations. All of the conditions described above are supportive of
+population establishment and long-term population growth.
+
+\paragraph{Long-term population
+viability}\label{long-term-population-viability}
+
+Results from the frog translocation study are suggestive of population
+establishment. However, a decade or more of surveys may be necessary to
+confirm that populations are in fact self-sustaining \citep{joseph2018}.
+To extend our inferences of population establishment beyond those
+possible from the site-specific CMR data, we developed a population
+viability model. Specifically, to test whether the observed yearly adult
+survival probabilities in translocated populations were sufficient for
+long-term viability, we built a stage-structured matrix model that
+captured known frog demography and included demographic and
+environmental stochasticity. We parameterized the model using CMR data
+from translocated populations and known life history values in this
+system (Supplementary Table 1). Results from a sensitivity analysis are
+provided in Supplementary Fig. 4.
 
 Given observed yearly adult survival probabilities of translocated frogs
-(from site-specific mrmr CMR models) and a yearly survival probability
-of the year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.09, at
-least six of twelve translocated populations should experience a
-long-run growth rate \(\lambda\) greater than 1 in the presence of Bd
-(Figure~\ref{fig-viability}A; median predicted \(\lambda\) ranges from
-1.19-1.40 for these six populations). These six populations all had
+(from site-specific mrmr CMR models; provided in legend of
+Fig.~\ref{fig-viability}a) and a yearly survival probability of the
+year-1 juvenile class (\(\sigma_{J_1}\)) greater than 0.10, at least six
+of twelve translocated populations are likely to experience a long-run
+growth rate \(\lambda\) greater than 1 in the presence of Bd
+(Fig.~\ref{fig-viability}a; median predicted \(\lambda\) ranges from
+1.07-1.28 for these six populations). These six populations all had
 observed yearly adult survival greater than 0.5. As year-1 juvenile
 survival probability increased above 0.2, the deterministic long-run
-growth rate of eight of twelve population was greater than 1
-(Figure~\ref{fig-viability}A).
+growth rate of seven of twelve populations was greater than 1
+(Fig.~\ref{fig-viability}a).
 
 Even when incorporating (i) demographic stochasticity and (ii)
-environmental stochasticity in year-1 juvenile survival and recruitment
-(the transition that we expect to be the most subject to environmental
-variability in the presence of Bd), populations with high adult survival
-are likely to persist over a 50 year time horizon. Our model predicted
-that, following a single introduction of 40 adult individuals into a
-population, the six populations with the highest adult survival
-probabilities (\(\sigma_{A_R} > 0.5\)) had 50-year extinction
-probabilities of less than 0.5 when the average year-1 juvenile survival
-was greater than 0.10 (Figure~\ref{fig-viability}B). This indicates
-strong potential for long-term persistence in the presence of Bd and
-environmental variability in survival and recruitment. In contrast, for
-the six populations where yearly adult survival probability
-\(\sigma_{A_R} < 0.5\), extinction probability over 50 years was always
-predicted to be \textgreater{} 50\% regardless of the value of mean
-year-1 juvenile survival between 0 and 0.25. To test the validity of our
-model predictions, we demonstrated that our stochastic model could
-describe the general recovery trajectory of our translocated population
-with the longest survey history (Figure~\ref{fig-viability}C;
-population 70550, surveyed for 16 years).
+environmental stochasticity in year-1 juvenile survival (the transition
+that we expect to be the most subject to environmental variability in
+the presence of Bd), populations with high adult survival are likely to
+persist over a 50 year time horizon. Our model predicted that, following
+a single introduction of 40 adult individuals into a population, the six
+populations with the highest adult survival probabilities
+(\(\sigma_{A_R} > 0.5\)) had 50-year extinction probabilities of less
+than 0.5 when the average year-1 juvenile survival was greater than 0.10
+(Fig.~\ref{fig-viability}b). This indicates strong potential for
+long-term persistence in the presence of Bd and environmental
+variability in juvenile survival. In contrast, for the six populations
+where yearly adult survival probability \(\sigma_{A_R} < 0.5\),
+extinction probability over 50 years was always predicted to be
+\textgreater{} 50\% regardless of the value of mean year-1 juvenile
+survival between 0 and 0.25. To test the validity of our model
+predictions, we demonstrated that our stochastic model could describe
+the general recovery trajectory of our translocated population with the
+longest survey history (Fig.~\ref{fig-viability}c; population 70550,
+surveyed for 16 years).
 
@@
 because there is evidence that naturally-recruited adults have higher
-survival probability than translocated adults
-(Figure S4), but we considered these
-probabilities to be equal in all but three of our populations where we
-had sufficient data to distinguish these different probabilities.
-
-\subsection*{Frog evolution in response to Bd}
-
-Results from the preceding sections indicate the critical role of frog
-resistance in post-translocation frog survival, population growth, and
-population viability. As such, identifying the mechanisms underlying
-this resistance would fill a key gap in our understanding of the factors
-that promote population resilience in the presence of disease. Although
-natural selection for more resistant frog genotypes, and evolutionary
-rescue, may be foundational to the ability of frogs to recover despite
-ongoing Bd infection, for MYL frogs the role of disease-mediated
-selection in these processes remains unknown.
-
-To determine whether MYL frog populations show genomic patterns
-consistent with an evolutionary response to Bd, we compared frog exomes
-(i.e., coding region of a genome) between populations with contrasting
-histories of Bd exposure. Specifically, we compared frog genomes sampled
-in 4 populations that have not yet experienced a Bd-caused epizootic
-(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
-Bd epizootic during the past several decades and have since recovered to
-varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
-\citep{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
-populations are based on 10-20 years of VES and Bd surveillance using
-skin swabbing \citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
-populations are characterized by large numbers of adults (i.e, typically
-1000s), Bd prevalence that is generally 0\% except during occasional Bd
-failed invasions \citep[during which Bd loads remain very
-low,][]{wilber2022}, and no history of Bd epizootics since we first
-surveyed these populations in the late 1990s and early 2000s
-\citep{zhou2015}. In contrast, recovering populations exist in an
-enzootic state \citep{briggs2010}, characterized by smaller numbers of
-adults (generally \textless{} 100), high Bd prevalence \citep[often
-\textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
-that are typically well below the level expected to cause mortality
-\citep{vredenburg2010}. Naive and recovering populations can be
-identified unambiguously using these differences in Bd prevalence and
-load. Finally, there is no potential for frog dispersal between the 9
-study populations due to intervening distances and topography, as well
-as the presence of introduced (predatory) fish and fish-induced habitat
-fragmentation. (See \textbf{Supporting Information - Frog evolution in
-response to Bd - Methods} for additional details regarding the study
-design.)
-
-We conducted a principal component analysis (PCA) of the genomic data to
-describe the relationships between sampled populations, and then used
-two complementary approaches to identify regions of the genome that
-differed between naive and recovering populations (i.e., regions under
-selection). First, we used a multivariate linear mixed model to evaluate
-associations between population type (i.e., naive versus recovering) and
-individual variants, including single nucleotide polymorphisms (SNPs)
-and insertions/deletions (INDELS), while accounting for population
-structure. Second, we used a splined window analysis to identify larger
-genomic regions showing differences between population types in
-\emph{F\textsubscript{ST}} and nucleotide diversity (\(\pi_{diff}\) =
-\(\pi_{naive}\) - \(\pi_{recovering}\)).
-
-Individual frogs clustered into 3 separate groups in principal component
-space (Figure S6A SI), and clusters reflected
-the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
-the strong signature of isolation-by-distance that is characteristic of
-MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
-each cluster contained at least one population from both the naive and
-recovering groups, allowing us to distinguish allelic associations of
-individuals sampled in the 2 population types versus allelic
-associations resulting from population structure and genetic drift.
-
-Results from the individual variant and splined window analyses show
-that recovering populations have signatures of selection on multiple
-regions of the genome. The analysis of individual variants identified 11
-``outlier'' SNPs (i.e., showing significantly different allele
-frequencies between naive versus recovering populations) from 7 distinct
-genes across 4 contigs (Figure S6B, C SI; SI
-Datasets S1, S2). One of the 7 identified genes (LOC108802036) does not
-have an associated annotation. For the outlier SNPs, frequency
-differences between the naive and recovering populations ranged from
-0.41 to 0.86. Most of these SNPs showed frequency differences in only a
-subset of the sampled populations (Figure~\ref{fig-allelefrequencies}A,
-B), but the SNP in the RIN3 gene showed consistent differences in
-frequencies across all populations (Figure~\ref{fig-allelefrequencies}
-C). This is suggestive of parallel selection at this locus across
-multiple populations. The other 6 outlier variants showed less
-consistent frequency differences across the study populations, but for
-these we still found a statistically significant signal of selection in
-2 of the 3 genetic clusters (containing populations 5--9;
-Figure~\ref{fig-allelefrequencies}, Figure S6A
-SI). Therefore, although some outlier variant associations have a more
-limited geographic extent than RIN3, they still describe results that
-suggest parallel evolutionary changes following Bd exposure.
-
-The splined window analysis identified 33 outlier regions for
-\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
-(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
-outliers for both metrics (``shared regions'') and 2 of these shared
-regions also contained one or more of the outlier SNPs described above.
-A total of 35 annotated genes were found in the 9 shared regions
-(Datasets S3, S4). Given this large number of genes, here we focus on
-those with the strongest signal of selection and/or immune-related
-functions. The largest \(\pi_{diff}\), indicative of directional
-selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
-the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan}C). This region
-contains approximately 500 SNPs and one annotated gene called
-``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
-a shared outlier region on Contig1 contained two complement factor genes
-(C6 and C7). Interestingly, this region had a large negative
-\(\pi_{diff}\) value, consistent with balancing selection. Finally, one
-shared outlier region on Contig8 contained one outlier SNP (TCF19) and
-was within 360kb of another outlier SNP (VARS)
-(Figure~\ref{fig-spline-manhattan}D, Figure S7).
-This region (854kb from the beginning of the outlier window to the VARS
-SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
-these genes occur in the extended major histocompatibility complex (MHC)
-Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
-extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
-Therefore, this region under selection is part of the extended MHC Class
-I and III complex and shows synteny with other amphibian genomes.
-
-Although the joint processes of Bd-caused population declines and
-selection in response to Bd exposure could affect genetic diversity of
-recovering populations, we found no consistent differences in
-individual-level heterozygosity or population-level \(\pi\) between
-naive and recovering populations (see \textbf{Supporting
-Information-Frog evolution in response to Bd} for details). Thus,
-despite localized selection in particular regions of the genome, we did
-not find evidence for reduced genetic diversity across the genome in
-recovering populations. In addition, no gene ontology (GO) biological
-functions, molecular functions, or cellular processes were
-over-represented in either the outlier variants or the 33 genes located
-in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined
-windows (see \textbf{Supporting Information-Frog evolution in response
-to Bd} for details).
-
-In summary, our genomic results indicate that the exomes of frogs from
-naive and recovering study populations show substantial differences,
-consistent with parallel evolutionary changes following Bd exposure. The
-regions under selection contain several immunologically-relevant genes
-and gene families that are directly linked to disease resistance in
-other taxa.
-
-\section*{Discussion}
+survival probability than translocated adults (Supplementary Fig. 3),
+but we considered these probabilities to be equal in all but three of
+our populations where we had sufficient data to distinguish these
+different probabilities.
+
+\subsubsection{Discussion}\label{discussion}
 
@@
 \citep{daszak2000}, but broadly-applicable strategies to recover
-affected species are generally lacking \citep[e.g.,][]{garner2016}.
-Here, we tested the possibility that populations of resistant
-individuals from naturally recovering populations can be used to
-reestablish extirpated populations of the endangered MYL frog in the
-presence of a highly virulent fungal pathogen (Bd). Our results indicate
-(i) the capacity of reintroduced populations to become established and
-eventually recover despite ongoing disease, (ii) that the recovering
-populations are likely to persist over a 50-year period, (iii) that
-there are substantial genomic differences between naive and recovering
-MYL frog populations, consistent with evolutionary change in frogs
-following Bd exposure, and (iv) that some of the genomic regions under
-selection contain genes related to disease resistance. Collectively,
-these results (Figure~\ref{fig-recovery-model}) provide a rare example
-of amphibian recovery in the presence of Bd, and have important
-implications for the conservation and recovery of amphibians and other
-taxa worldwide that are endangered by escalating impacts from emerging
-infectious diseases. In light of the generally low success rate of
+affected species are generally lacking (e.g., \citep{garner2016}). Here,
+we tested the possibility that populations of resistant individuals from
+naturally recovering populations can be used to reestablish extirpated
+populations of the endangered \emph{R. sierrae} in the presence of a
+highly virulent pathogen (Bd). Our results indicate (i) the capacity of
+reintroduced populations to become established and eventually recover
+despite ongoing disease, (ii) that post-translocation frog survival is
+influenced by site, cohort, and individual-level characteristics, and
+(iii) that 50\% of the reintroduced populations are likely to persist
+over a 50-year period. In light of the generally low success rate of
 amphibian reintroduction efforts \citep{dodd2005}, our success in
-reestablishing MYL frog populations via translocation of resistant
-individuals is striking, and even more so given that this success
-occurred despite ongoing infection of frogs with Bd.
-
-In the following discussion, we follow the sequence of frog recovery
-described in Figure~\ref{fig-recovery-model} to structure our key
-points. Previous field studies in MYL frogs show that frog-Bd dynamics
+reestablishing \emph{R. sierrae} populations via reintroduction of
+resistant individuals is striking, and even more so given that \emph{R.
+sierrae} was driven to near-extinction by Bd. Collectively, these
+results provide a rare example of amphibian recovery in the presence of
+Bd, and have important implications for the conservation and recovery of
+amphibians and other taxa worldwide that are endangered by escalating
+impacts from emerging infectious diseases.
+
+Previous field studies in \emph{R. sierrae} show that frog-Bd dynamics
 and frog survival in the presence of Bd are fundamentally different
@@
 low-to-moderate and relatively constant Bd loads and mass die-offs are
-not observed \citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
-naive and recovering populations are also observed in controlled
-laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
-indicate that frogs from recovering populations exhibit resistance
-against Bd infection. This resistance could in theory be due to several
-factors, including natural selection for more resistant genotypes,
-acquired immunity, and/or inherent between-population differences that
-pre-date Bd exposure, but until now evidence to evaluate the role of
-evolution was lacking.
-
-Results from our genomic analyses suggest that natural selection for
-adaptive alleles is at least partially responsible for the increased
-resistance of frogs in recovering populations. We identified multiple
-specific alleles and genomic regions showing signatures of selection
-between adjacent naive and recovering MYL frog populations, consistent
-with selection following Bd exposure. These analysis are based on
-samples collected from virtually all of the MYL frog populations
-remaining in a naive state, as well as adjacent recovering populations.
-This produced genetic clusters that contained at least one naive and one
-recovering population, allowing us to detect selection without the
-confounding effects of population structure. Finally, we did not find a
-reduction in overall genetic variation in the recovering populations,
-suggesting that despite localized selection in the genome, these
-populations retain adequate genetic diversity for long-term persistence.
-
-Importantly, some regions that we identified as under selection are
-associated with cellular and immunological mechanisms known to
-contribute to disease resistance, including in amphibians
-\citep{zamudio2020}. For example, the MHC plays an important role in
-immunity. In our study, we identified a region that shows evidence of
-selection in recovering populations and contains eight genes associated
-with either the MHC Class I or Class III regions. These results
-corroborate numerous previous studies linking MHC genes to amphibian
-resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
-Similarly, the region with the strongest indication of directional
-selection (as measured by \(\pi_{diff}\)) contains the
-interferon-related gene GVINP1. Several previous studies of amphibians
-have found this gene to be differentially expressed during Bd infection
-\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
-in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
-linked to disease in salmon, explaining a notable 20\% of the resistance
-phenotype \citep{robledo2020, robledo2018}. We also identified a region,
-characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
-that contained the complement genes C6 and C7. The complement system
-plays an important role in innate immunity \citep{rieraromo2016}, and
-our results could indicate that balancing selection is acting in this
-region of the genome to favor a diverse set of alleles, as is known for
-C6 in humans \citep{soejima2005}. Based on the analysis of individual
-outlier variants, the RIN3 gene showed a consistent pattern of allele
-frequency differences across all nine of the frog populations sampled in
-this study, indicating consistent selection in populations distributed
-across a wide geographic area. This gene is associated with inflammation
-response and in \emph{Xenopus} is highly expressed in wounds
-\citep{fukazawa2009}. Finally, the outlier variant with the lowest
-p-value was the uncharacterized gene LOC108802036. In the genome of
-another frog species, this gene is located adjacent to a type I
-interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
-further suggests the importance of interferon-related genes in this
-system. Collectively, the genes associated with these genomic
-differences may confer at least some degree of resistance against Bd
-infection, an attribute that may be critically important to population
-reestablishment and recovery in the presence of Bd.
-
-Reintroduction of resistant MYL frogs was remarkably successful in
-reestablishing viable populations in the presence of Bd. Of the 12
-translocated populations, approximately 80\% showed evidence of both
-successful reproduction and recruitment of new adults. Year-1 survival
-for 12 of the 24 translocated cohorts exceeded 50\%, and \textgreater{}
-70\% of translocated cohorts had survival above this 50\% level when the
-earliest translocations are excluded (i.e., translocations conducted
-when methods were still being refined; see \textbf{Materials and Methods
-- Frog population recovery - Field methods} for a brief description of
-these refinements). The fact that the relatively low Bd loads and
-correspondingly high frog survival was maintained when frogs were moved
-from donor populations to recipient sites indicates that these
-characteristics of naturally-recovering populations were not solely an
-effect of site characteristics, but were also strongly influenced by
-resistance inherent in the frogs. Although it could be argued that the
-relatively invariant Bd loads before versus after translocation are a
-consequence of similar pathogen pressure in the donor and translocated
-populations, this is at odds with the fact that in the first year after
-translocation frog densities are typically 1-2 orders of magnitude lower
-in the translocated versus donor populations and pathogen pressure
-should follow a similar pattern. In addition to the maintenance of Bd
-load and frog survival between natal and translocation sites, the
-relatively high survival of translocated frogs was maintained in their
-progeny, as expected if resistance has a genetic basis.
+not observed \citep{briggs2010, knapp2011} (see also Supplementary Fig.
+2). The differences in Bd load of frogs from naive and recovering
+populations are also observed in controlled laboratory studies (see
+Figure 4 in \citep{knapp2016}), and clearly indicate that frogs from
+recovering populations exhibit resistance against Bd infection. Finally,
+reintroductions of \emph{R. sierrae} collected from Bd-naive populations
+and reintroduced into Bd-positive sites have consistently failed due to
+the development of high Bd loads and resulting high frog mortality,
+suggesting the importance of frog resistance in population
+reestablishment in the presence of Bd.
+
+In the current study, reintroduction of resistant \emph{R. sierrae} was
+remarkably successful in reestablishing populations in the presence of
+Bd. Of the 12 translocated populations, approximately 80\% showed
+evidence of both successful reproduction and recruitment of new adults.
+Year-1 survival for 12 of the 24 translocated cohorts exceeded 50\%, and
+\textgreater{} 70\% of translocated cohorts had survival above this 50\%
+level when the earliest translocations are excluded (i.e.,
+translocations conducted when methods were still being refined; see
+\textbf{Methods - Frog population recovery - Field methods} for a brief
+description of these refinements). The fact that the relatively low Bd
+loads and correspondingly high frog survival were both maintained when
+frogs were moved from donor populations to recipient sites indicates
+that these qualities of naturally-recovering populations were not solely
+an effect of site characteristics, but were also strongly influenced by
+intrinsic characteristics of frogs, including resistance. Although it
+could be argued that the relatively invariant Bd loads before versus
+after translocation are a consequence of similar pathogen pressure in
+the donor and reintroduced populations, this is at odds with the fact
+that in the first year after translocation frog densities are typically
+1-2 orders of magnitude lower in the reintroduced versus donor
+populations and pathogen pressure should follow a similar pattern. In
+addition to the maintenance of Bd load and frog survival between natal
+and recipient sites (Supplementary Fig. 3), the relatively high survival
+of translocated frogs was maintained in their progeny, as expected if
+resistance has a genetic basis.
 
@@
 post-translocation surveys. These long-term forecasts are important,
-given that reintroduced MYL frog populations may often take decades to
-achieve our ultimate goal of self-sustainability \citep{joseph2018}.
-Making well-supported projections about the long-term outcome of
-reintroduction efforts from shorter-term information is critically
-important to the process of adaptive management of species
-reintroduction programs \citep{seddon2007}, including the one we are
-carrying out for MYL frogs. Specifically, the combined results from our
-reintroduction study and viability modeling indicate that survival of
+given that reintroduced \emph{R. sierrae} populations may often take
+decades to achieve our ultimate goal of self-sustainability
+\citep{joseph2018}. Making well-supported projections about the
+long-term outcome of reintroduction efforts from shorter-term
+information is crucial for adaptive management of species reintroduction
+programs \citep{seddon2007}, including the one we are carrying out for
+\emph{R. sierrae}. Specifically, the combined results from our
+reintroduction study and viability modeling suggest that survival of
 frogs in the first year following translocation is an effective proxy of
@@
 site to support high frog survival and a viable frog population over the
-long term). This proxy of site quality is important in the MYL frog
-system because accurately predicting the ability of a site to support a
-viable frog population \emph{a priori} remains difficult, even after
-conducting 24 translocations over 16 years.
-
-Despite the demonstrated resistance of adult MYL frogs against Bd
-infection, individual and population-level impacts of Bd are still
+long term). This proxy of site quality is important in the \emph{R.
+sierrae} system because accurately predicting the ability of a site to
+support a viable frog population \emph{a priori} remains difficult, even
+after conducting 24 translocations over 16 years.
+
+The results of the meta-analysis indicate the influence of several
+variables on adult survival, and by extension on population viability.
+Although we assume that the effects of winter severity and frog sex are
+direct effects of the variables themselves, the elevation and donor
+population variables are likely serving as proxies for a range of
+factors. For example, elevation is correlated with several attributes
+that may influence frog survival, including the presence of boulder
+habitat (positive correlation) that is important for frog overwintering,
+abundance of frog predators including garter snakes (negative
+correlation), and frog developmental rates (negative correlation). The
+latter can influence the duration of frog life stages that have
+differential susceptibility to chytridiomycosis (see below). The
+duration of the tadpole and juvenile stages increases with elevation,
+but the fact that long-run population growth rate from our viability
+model is relatively insensitive to variation in these durations suggests
+that developmental rates of early frog life stages have minimal effect
+on the predicted long-term viability of reintroduced populations. The
+mechanisms underlying the effect of donor population on the probability
+of post-translocation frog survival is intriguing, and could indicate
+population-specific fitness differences of frogs in the presence of Bd.
+Although our study design did not allow us to completely separate the
+effects of donor population and recipient site on 1-year adult survival,
+we are currently doing so using a modified study design. Specifically,
+we are conducting translocations in which frogs from several donor
+populations are added concurrently to a single site. Survival of frogs
+from each donor population is estimated using CMR methods and the
+contribution of each donor population to subsequent generations is
+quantified using genetic methods.
+
+Despite the demonstrated resistance of adult \emph{R. sierrae} against
+Bd infection, individual and population-level impacts of Bd are still
 evident. In an earlier study of 2 of our 12 translocated populations
@@
 In addition to these important but relatively subtle effects of Bd on
-adults, the impacts on younger life stages are more apparent. MYL frogs
-immediately following metamorphosis (``metamorphs'') are highly
+adults, the impacts on younger life stages are more apparent. \emph{R.
+sierrae} immediately following metamorphosis (``metamorphs'') are highly
 susceptible to Bd infection \citep{ellison2019} and as a result
@@
 anurans, and may result from the poorly developed immune system
-characteristic of this life stage \citep{humphries2022}. In naturally
-recovering and translocated MYL frog populations, we suggest that the
-high mortality of metamorphs is an important limitation on subsequent
-recruitment of new adults. Therefore, although adult MYL frogs appear
-relatively resistant, Bd infection continues to have important limiting
-effects on recovering populations \citep[see also][]{hollanders2022}.
-
-The recent emergence of Bd worldwide has contributed to the decline of
-hundreds of amphibian species, some of which are now extinct in the wild
-\citep{scheele2019}. This extraordinary impact on global amphibian
-biodiversity is compounded by the lack of any effective and broadly
-applicable strategies to reverse these impacts
-\citep{garner2016, knapp2022}. Importantly, in addition to the natural
-recovery documented for MYL frogs \citep{knapp2016}, other species are
-also showing evidence of post-epizootic recovery in the presence of Bd
+characteristic of this life stage \citep{humphries2022}. In both
+naturally recovering and reintroduced \emph{R. sierrae} populations, we
+suggest that the high mortality of metamorphs is an important limitation
+on subsequent recruitment of new adults. Therefore, although adult
+\emph{R. sierrae} appear relatively resistant, Bd infection continues to
+have important limiting effects on recovering populations (see also
+\citep{hollanders2022}). The ongoing high susceptibility of metamorphs
+to chytridiomycosis also highlights the critical need to develop better
+methods of monitoring this vulnerable life stage, the results of which
+could significantly advance the conservation and recovery of Bd-impacted
+amphibians.
+
+Our successful use of translocations to reestablish \emph{R. sierrae}
+populations despite ongoing Bd infection is particularly important
+because natural recovery alone appears insufficient to allow
+reestablishment of populations at the landscape scale. Naturally
+recovering \emph{R. sierrae} populations in Yosemite are relatively rare
+and are often isolated from other such populations and from suitable but
+vacant habitats. This isolation is the consequence of both natural
+fragmentation due to rugged geography and the ongoing presence of
+introduced, predatory trout in the majority of lakes and streams
+\citep{bradford1993}. Collectively, these factors often prevent
+naturally recovering \emph{R. sierrae} populations from expanding via
+dispersal of frogs to nearby unoccupied habitats. Translocations allowed
+us to overcome these dispersal limitations and reestablish populations
+in sites that would otherwise remain unoccupied for the foreseeable
+future. Of particular note are the translocations that we conducted as
+part of the current study into areas from which \emph{R. sierrae} were
+completely extirpated, but instead of single isolated habitats these
+areas contained habitats composed of interconnected fishless lakes,
+ponds, marshes, and/or streams. Successful reestablishment of one
+\emph{R. sierrae} population in such areas has produced a source of
+frogs from which adults and juveniles are now dispersing into adjacent
+habitats. This process of translocation, establishment, and subsequent
+dispersal has significant potential to eventually allow the
+reestablishment of robust metapopulations that may be more resilient to
+current and future natural and anthropogenic impacts than more
+fragmented populations (e.g., \citep{heard2015}).
+
+The promising example provided by our study of successful
+reestablishment of \emph{R. sierrae} despite ongoing Bd infection may
+provide an important roadmap for ongoing and future efforts to recover
+the hundreds of Bd-impacted amphibian species globally. Importantly, in
+addition to the natural recovery documented for \emph{R. sierrae}
+\citep{knapp2016}, other amphibian species are also showing evidence of
+post-epizootic recovery in the presence of Bd
 \citep{scheele2017, voyles2018} and suggest the possibility of also
 using animals from these recovering populations to reestablish
-extirpated populations. As with MYL frogs, the feasibility and long-term
-success of such efforts will depend on the availability of robust donor
-populations containing individuals that have the adaptive alleles
-necessary to allow frog survival and population growth in the presence
-of Bd. Despite the hopeful example of successful reestablishment of MYL
-frogs despite ongoing Bd infection, the challenge of recovering hundreds
-of Bd-impacted amphibian species globally is a daunting prospect.
-Although we now have a proven strategy to reestablish extirpated MYL
-frog populations, recovery across their large historical range will
-require substantial resources over many decades. The results of this
-study provide a hopeful starting point for that endeavor and other
-future efforts worldwide.
-
-In our rapidly changing world, evolution is likely to play in important
-role in facilitating the resilience of wildlife populations. Whether the
-documented disease resistance in MYL frogs and concurrent recovery of
-decimated populations provides an airtight example of evolutionary
-rescue will likely always be uncertain (given that can never have a
-perfect understanding of the past). Regardless, we provide an example
-from the wild that suggests that evolution can produce individuals that
-harbor adaptive alleles and allow population recovery in a novel (i.e.,
-Bd-positive) environment, and show conclusively that individuals from
-these recovering populations can be used to reestablish extirpated
-populations and expand the scale of natural recovery
-(Figure~\ref{fig-recovery-model}). We expect that similar species
-recovery actions will be an essential tool in wildlife conservation in
-an era of accelerating global change.
-
-\matmethods{
-\hypertarget{frog-population-recovery-1}{%
-\section*{Frog population recovery}\label{frog-population-recovery-1}}
-
-\hypertarget{field-methods}{%
-\subsection*{Field methods}\label{field-methods}}
-
-For the 24 translocations we conducted, we identified donor populations
-from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
-collected using several years of VES and skin swab collections
-\citep{knapp2016}, and results from population genetic analyses
-\citep{poorten2017}. The populations that we selected contained hundreds
-to thousands of both \emph{R. sierrae} adults and tadpoles. These
-relatively high abundances were the result of recent increases following
-previous Bd-caused declines \citep{knapp2016}. As is typical for
-recovering MYL frog populations, Bd prevalence in the donor populations
-was high (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
+extirpated populations. As with \emph{R. sierrae}, the feasibility and
+long-term success of such efforts will depend on the availability of
+robust donor populations containing individuals that have sufficiently
+high resistance to allow frog survival and population growth in the
+presence of Bd. We suggest that several components of recovery efforts
+could play important roles in maximizing their success. First, an
+adaptive management approach such as the one we utilized, in which the
+outcomes of recovery actions are rigorously assessed and this
+information is used to reduce uncertainties and inform future actions,
+will often be essential \citep{canessa2019}. Second, given the apparent
+rarity in amphibians of evolution of resistance against Bd infection,
+developing an improved understanding of the factors facilitating or
+limiting the development of frog resistance and the effectiveness of
+selective breeding programs and other genetic interventions to enhance
+frog resistance \citep{kosch2022} may be particularly important.
+Finally, although we did not quantify the genetic structure of the donor
+and translocated populations of \emph{R. sierrae} as part of the current
+study, conducting genetic monitoring may be important to identify
+sub-optimal genetic outcomes during establishment of translocated
+populations (e.g., genetic bottlenecks) and allow timely implementation
+of actions to mitigate such effects \citep{whiteley2015}.
+
+In conclusion, we now have a proven strategy to reestablish extirpated
+\emph{R. sierrae} populations. However, recovery across their large
+historical range will require substantial resources over many decades.
+The results of this study provide a hopeful starting point for that
+endeavor and other future efforts worldwide.
+
+\subsubsection{Methods}\label{methods}
+
+\paragraph{Frog population recovery}\label{frog-population-recovery-1}
+
+\subparagraph{Field methods}\label{field-methods}
+
+The collection of data from live \emph{R. sierrae} described here
+complies with all ethical regulations and was approved under the
+University of California-Santa Barbara Institutional Animal Care and Use
+Committe (IACUC) protocol number 478. For the 24 translocations we
+conducted, we identified donor populations from which adult frogs
+(\(\geq\) 40 mm snout-vent length) could be collected based on several
+years of VES and skin swab collections \citep{knapp2016}, and results
+from population genetic analyses \citep{poorten2017}. The populations
+that we selected contained hundreds of \emph{R. sierrae} adults and
+thousands of tadpoles. These relatively high abundances were the result
+of recent increases following previous Bd-caused declines
+\citep{knapp2016}. As is typical for recovering \emph{R. sierrae}
+populations, Bd prevalence in the donor populations was high
+(0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
 3.06--3.78 ITS copies) was two or more orders of magnitude below the
-level at which frog mortality is expected (log\textsubscript{10}(load)
-\(\approx\) 5.78 copies)\citep{vredenburg2010, joseph2018}. Recipient
-sites to which frogs were translocated were chosen based on previous
-\emph{R. sierrae} presence (determined from VES and/or museum records)
-or characteristics that suggested high quality habitat for this species
-\citep{knapp2005}. At the beginning of this study, we had a relatively
-limited understanding of the factors that affect habitat quality. In
-subsequent years, we improved our site selection process by
-incorporating new information about important habitat features, in
-particular, overwinter habitats such as submerged boulders and
-overhanging banks. \emph{R. sierrae} were absent from all recipient
-sites prior to the first translocation.
-
-We conducted 1--4 translocations per site
-(Figure~\ref{fig-translocation-survival}, Figure S1)
-and each translocated cohort included 18 to 99 frogs (median = 30). In
-preparation for each translocation, adult frogs were collected from the
-donor population and measured, weighed, swabbed, and PIT tagged. Frogs
-were transported to the recipient site either on foot or via helicopter.
-Following release, we visited translocated populations approximately
-once per month during the summer active season and conducted diurnal CMR
-surveys and VES (summer active season is generally July-August but can
-start as early as May and end as late as September; range of survey
-dates = May-25 to Sep-29, range of translocation dates = Jun-28 to
-Sep-02; median number of visits per summer = 2, range = 1--10). CMR
+level at which frog mortality is expected
+\citep{vredenburg2010, joseph2018} (log\textsubscript{10}(load)
+\(\approx\) 5.78 ITS copies). Recipient sites to which frogs were
+translocated were chosen based on previous \emph{R. sierrae} presence
+(determined from VES and/or museum records) or characteristics that
+suggested high quality habitat for this species \citep{knapp2005}. At
+the beginning of this study, we had a relatively limited understanding
+of the factors that affect habitat quality. In subsequent years, we
+improved our site selection process by incorporating new information
+about important habitat features, in particular, overwintering habitats
+such as submerged boulders and overhanging banks. \emph{R. sierrae} were
+absent from all recipient sites prior to the first translocation. This
+determination was based on VES conducted at recipient sites over several
+years (see additional details below regarding frog detectability during
+VES).
+
+We conducted 1--4 translocations per site (Fig.~\ref{fig-yosemap},
+Fig.~\ref{fig-translocation-survival}) and each translocated cohort
+included 18 to 99 frogs (median = 30). In preparation for each
+translocation, adult frogs were collected from the donor population and
+measured, weighed, swabbed, and PIT tagged. Frogs were transported to
+the recipient site either on foot or via helicopter. Following release,
+we visited translocated populations approximately once per month during
+the summer active season to conduct diurnal CMR surveys and VES (summer
+active season is generally July-August but can start as early as May and
+end as late as September; range of survey dates = May-25 to Sep-29,
+range of translocation dates = Jun-28 to Sep-02; median number of visits
+(i.e., primary periods; see below) per summer = 2, range = 1--10). CMR
 surveys allowed estimation of adult survival, recruitment of new adults,
-and adult population size, and VES provided estimates of tadpole and
+and adult population size. VES provided estimates of tadpole and
 juvenile abundance. During 2006-2012, we conducted CMR surveys on a
-single day (primary period) per site visit, during which we searched all
+single day per site visit (primary period), during which we searched all
 habitats repeatedly for adult frogs. Frogs were captured using handheld
@@
 
-\hypertarget{frog-counts-and-reproductive-success}{%
-\subsection*{Frog counts and reproductive
-success}\label{frog-counts-and-reproductive-success}}
+\subparagraph{Frog counts and reproductive
+success}\label{frog-counts-and-reproductive-success}
 
@@
 during the 3 years following the initial translocation. This accounted
-for the multi-year tadpole and juvenile stages in MYL frogs
-(Table S2 SI).
-
-\hypertarget{estimation-of-frog-survival-and-abundance}{%
-\subsection*{Estimation of frog survival and
-abundance}\label{estimation-of-frog-survival-and-abundance}}
+for the multi-year tadpole and juvenile stages in \emph{R. sierrae}
+(Supplementary Table 1).
+
+\subparagraph{Estimation of frog survival and
+abundance}\label{estimation-of-frog-survival-and-abundance}
 
@@
 \citep{joseph2019} and using R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
-Frog population recovery - CMR model structure} for details). Briefly,
-the model tracks the states of \emph{M} individuals that comprise a
-superpopulation made up of real and pseudo-individuals \citep[see][ for
-details]{joseph2018}. The possible states of individuals include ``not
+\citep{rsoftware2022} (v4.4.4) (see \textbf{Supplementary Methods - Frog
+population recovery - CMR model structure} for details). Briefly, the
+model tracks the states of \emph{M} individuals that comprise a
+superpopulation made up of real and pseudo-individuals
+\citep{joseph2018}. The possible states of individuals include ``not
 recruited'', ``alive'', and ``dead''. The possible observations of
@@
 
-\hypertarget{predictors-of-post-translocation-frog-survival}{%
-\subsection*{Predictors of post-translocation frog
-survival}\label{predictors-of-post-translocation-frog-survival}}
+\subparagraph{Predictors of post-translocation frog
+survival}\label{predictors-of-post-translocation-frog-survival}
 
@@
 site\_id. We performed all analyses with the rstanarm package
-\citep{rstanarm2022} and R Statistical Software
-\citep[v4.4.4,][]{rsoftware2022}. For all models, we used default,
-weakly informative priors, four chains, and 5000 iterations each for
-warmup and post-warmup. We checked MCMC convergence using trace plots
-and Rhat, and evaluated model fit using leave-one-out cross-validation
-\citep{vehtari2016}, as implemented by the loo package
-\citep{vehtari2022}. (See \textbf{Supporting Information - Frog
-population recovery - Among-site survival modeling} for details.)
-
-\hypertarget{changes-in-bd-load-following-translocation}{%
-\subsection*{Changes in Bd load following
-translocation}\label{changes-in-bd-load-following-translocation}}
+\citep{rstanarm2022} and R Statistical Software \citep{rsoftware2022}
+(v4.4.4). For all models, we used default, weakly informative priors,
+four chains, and 5000 iterations each for warmup and post-warmup. We
+checked MCMC convergence using trace plots and Rhat, and evaluated model
+fit using leave-one-out cross-validation \citep{vehtari2016}, as
+implemented by the loo package \citep{vehtari2022}. (See
+\textbf{Supplementary Methods - Frog population recovery - Among-site
+survival modeling} for details.)
+
+\subparagraph{Changes in Bd load following
+translocation}\label{changes-in-bd-load-following-translocation}
 
 We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep[see \textbf{Supporting Information-Frog population recovery -
-Laboratory methods} for details]{boyle2004}. To assess the magnitude of
-changes in Bd load on frogs following translocation, we compared Bd
-loads measured before versus after translocation. Before-translocation
-loads were quantified using skin swabs collected from all
-to-be-translocated frogs at the donor site on the day before or the day
-of the translocation. After-translocation Bd loads were based on all
-swabs collected from translocated frogs at the recipient site in the
+\citep{boyle2004} (see \textbf{Supplementary Information - Frog
+population recovery - Laboratory methods} for details). To assess the
+magnitude of changes in Bd load on frogs following translocation, we
+compared Bd loads measured before versus after translocation.
+Before-translocation loads were quantified using skin swabs collected
+from all to-be-translocated frogs at the donor site on the day before or
+the day of the translocation. After-translocation Bd loads were based on
+all swabs collected from translocated frogs at the recipient site in the
 year of and the year following translocation. Individual frogs and their
@@
 
-Code to replicate all of the analyses described above is available from
-the following GitHub repositories:
-\url{https://github.com/SNARL1/translocation};
-\url{https://github.com/SNARL1/cmr-analysis}.
-
-\hypertarget{population-viability-modeling}{%
-\section*{Population viability
-modeling}\label{population-viability-modeling}}
-
-\hypertarget{model-description}{%
-\subsection*{Model description}\label{model-description}}
+\paragraph{Population viability
+modeling}\label{population-viability-modeling}
+
+\subparagraph{Model description}\label{model-description}
 
@@
 long-term viability of populations established via translocation, we
-developed a population model for MYL frogs. Our central question was:
-How does the magnitude and variation in observed adult survival
+developed a population model for \emph{R. sierrae}. Our central question
+was: How does the magnitude and variation in observed adult survival
 probability across translocated populations affect the long-term
 persistence probability of populations? We developed a model that
-tracked seven state variables of a frog population: density of
-translocated adults (\(A_T\)), density of adults naturally recruited
-into the population (\(A_R\)), density of first-year tadpoles (\(L_1\)),
-density of second-year tadpoles (\(L_2\)), density of third-year
-tadpoles (\(L_3\)), density of first-year juveniles (\(J_1\)), and
-density of second-year juveniles (\(J_2\)). We divided adults into two
-classes \(A_T\) and \(A_R\) because there is evidence that the survival
-probability of translocated adults and naturally recruited adults
-differs (Figure S4).
+tracked seven state variables of a frog population: abundance of
+translocated adults (\(A_T\)), abundance of adults naturally recruited
+into the population (\(A_R\)), abundance of first-year tadpoles
+(\(L_1\)), abundance of second-year tadpoles (\(L_2\)), abundance of
+third-year tadpoles (\(L_3\)), abundance of first-year juveniles
+(\(J_1\)), and abundance of second-year juveniles (\(J_2\)). We divided
+adults into two classes \(A_T\) and \(A_R\) because there is evidence
+that the survival probability of translocated adults and naturally
+recruited adults differs (Supplementary Fig. 3). The modeled population
+included both male and female frogs.
 
@@
 discrete-time, stage-structured model where a time step is one year. The
-dynamics are given by equation \ref{eqn:matrix}.
-
-\begin{figure*}\[
+dynamics are given by
+
+\begin{equation}\phantomsection\label{eq-matrix}{
+\begin{aligned}
 \begin{bmatrix}
@@
 A_T \\
-\end{bmatrix}(t)\numberthis \label{eqn:matrix} 
-\]\end{figure*}
+\end{bmatrix}(t) \\
+\hspace{0.0001cm}
+\end{aligned}
+}\end{equation}
 
@@
 state variable), probability that a female frog reproduces in a given
-year \(p_F\), number of eggs produced by a female frog in a year that
-successfully hatch \(F\), probability of a first-year tadpole remaining
-as a tadpole \(p_{L1}\), probability of a second-year tadpole remaining
-as a tadpole \(p_{L2}\), and probability of a first-year juvenile
-remaining as a juvenile \(p_{J1}\). First-year juvenile survival and
-recruitment \(\sigma_{J1}\) is the parameter that we think is most
-influenced by environmental stochasticity.
+year \(p_F\), the percentage of adults that are female \(\rho\), number
+of eggs produced by a female frog in a year that successfully hatch
+\(F\), probability of a first-year tadpole remaining as a tadpole
+\(p_{L1}\), probability of a second-year tadpole remaining as a tadpole
+\(p_{L2}\), and probability of a first-year juvenile remaining as a
+juvenile \(p_{J1}\). First-year juvenile survival \(\sigma_{J1}\) is the
+parameter that we think is most influenced by environmental
+stochasticity.
 
@@
 prevalence significantly simplifies the model without much loss of
-realism. Additional details are provided in \textbf{Supporting
-Information - Population viability modeling}.
-
-\hypertarget{model-analysis}{%
-\subsection*{Model analysis}\label{model-analysis}}
+realism. Additional details are provided in \textbf{Supplementary
+Methods - Population viability modeling - Incorporating yearly
+variability in survival rates} and \textbf{Estimating model parameters}.
+
+\subparagraph{Model analysis}\label{model-analysis}
 
 After parameterizing our model with CMR-estimated adult frog survival
-probabilities and other known vital rates
-(Table S2), we performed four analyses. First, we
-compute the long-run growth rate \(\lambda\) for each of our 12
-translocated populations to determine if the populations were
-deterministically predicted to grow or decline in the long-run. Second,
-we compute the elasticity of \(\lambda\) to four key model parameters to
-quantify how much changes in these parameters affected the long-run
-growth rate (Figure S5). This also helped us
-determine where in the model environmental variation in juvenile
-survival and recruitment would have the largest effects on population
-dynamics. Third, we included demographic stochasticity and environmental
-stochasticity in \(\sigma_{J_1}\) in our model and simulated the 50-year
-viability (i.e., 1 - extinction probability) of populations given an
-introduction of 40 adult individuals into an unoccupied habitat.
-Finally, we fit our model to our longest translocation trajectory to
-confirm that our model could reasonably reproduce the observed recovery
-trajectories of MYL frogs following reintroductions. Additional details
-are provided in \textbf{Supporting Information - Population viability
-modeling}. Code to replicate the analyses can be found at
-\url{https://github.com/SNARL1/translocation}.
-
-\hypertarget{frog-evolution-in-response-to-bd-1}{%
-\section*{Frog evolution in response to
-Bd}\label{frog-evolution-in-response-to-bd-1}}
-
-\hypertarget{sampling-and-sequencing}{%
-\subsection*{Sampling and sequencing}\label{sampling-and-sequencing}}
-
-We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
-\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
-populations, and 29 from 5 recovering populations. These populations are
-located in the southern Sierra Nevada, from northern Yosemite National
-Park to northern Sequoia National Park
-(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
-frogs per population. To minimize potential confounding effects caused
-by variation in frog genotypes across latitude \citep{byrne2023}, we
-selected sampling sites such that both population types were represented
-across similar latitudinal ranges. DNA was extracted following Qiagen
-DNEasy manufacturer's protocols. We sequenced the samples using an exome
-capture assay as described in \citep{byrne2023}. Briefly, genomic
-libraries were prepared and captured using a custom Nimblegen capture
-pool. Capture baits were designed based on the coding regions of the
-\emph{R. muscosa} transcriptome (GenBank accession GKCT00000000).
-Captured libraries were then pooled and sequenced on a NovaSeq 6000
-150PE Flow Cell S1 at the Vincent J. Coates Genomics Sequencing Lab at
-UC Berkeley. Raw sequencing reads are available from NCBI SRA
-(PRJNA870451).
-
-\hypertarget{data-pre-processing-and-cleaning}{%
-\subsection*{Data pre-processing and
-cleaning}\label{data-pre-processing-and-cleaning}}
-
-Raw reads were filtered for adapters and contaminants using fastp
-\citep{chen2018} and aligned to the \emph{Rana muscosa} genome (NCBI
-SRA: SRS6533475 \citep{hon2020}), with repetitive elements masked using
-bwa (``mem'' mode) \citep{li2013}. Exact PCR duplicates were marked
-using Picard. Variants were then called following GATK best practices
-(v.4.2.0.0 \citep{van2020}). Briefly, raw variants were called for each
-sample using HaplotypeCaller and combined using CombineGVCFs. Next,
-genotypes were jointly called using GenotypeGVCFs. Variants were then
-hard filtered using gatk VariantFiltration using the following
-parameters to remove low-quality sites: QD \textless{} 2.0, FS
-\textgreater{} 40.0, SOR \textgreater{} 3.0, MQ \textless{} 50.0,
-MQRankSum \textgreater{} 3.0, MQRankSum \textless{} -3.0, ReadPosRankSum
-\textgreater{} 3.0, ReadPosRankSum \textless{} -3.0. This initial filter
-resulted in 1,595,206 variant sites across 53 individuals.
-
-We then further filtered our dataset at the individual and variant
-level. First, we trimmed our variants to only include those with minor
-allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
-depth of 5, a minimum genotype quality of 20, and a maximum missing
-proportion of 0.5. This filter resulted in 427,038 sites, of which
-353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
-with an average depth across filtered sites \textless{} 7x (n = 3). Our
-final dataset included 50 samples, 23 from naive and 27 from recovering
-populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).
-
-\hypertarget{data-analysis}{%
-\subsection*{Data analysis}\label{data-analysis}}
-
-To visualize the genomic relationships of our populations, we conducted
-a PCA using the glPCA function in the adegenet R package
-\citep{jombart2008}. To detect regions of the genome that differed
-between naive and recovering populations, i.e., regions under selection,
-we used two approaches: (1) a multivariate linear mixed model to
-evaluate individual variants (SNPs and INDELs), and (2) a splined window
-analysis to evaluate larger genomic regions. For the variant analysis,
-we first used a stringent data filter to include only variants with
-\textless{} 5\% missing data (missing for no more than 2 individuals),
-and then calculated the likelihood ratio statistic for the resulting set
-of 148,307 high quality variants across 127 contigs using GEMMA
-\citep{zhou2014}. GEMMA calculates and incorporates a relatedness matrix
-for input samples, allowing us to account for relatedness and population
-structure when calculating likelihood ratio statistics. We identified
-variants showing different allele frequencies between naive versus
-recovering populations (``outliers'') using a Bonferroni-corrected
-significance level of 0.01. We visualized the results using a Manhattan
-plot and qqplot. We developed a more liberal set of outlier variants
-using a Bonferroni-corrected significance level of 0.05 and used this
-set solely for the gene ontology (GO) analysis (see below and
-\textbf{Supporting Information}: Dataset S1, S2).
-
-In the analysis of individual variants, for each outlier variant we
-determined whether the variant was synonymous (protein sequence the same
-for each variant) or non-synonymous (protein sequence differs between
-variants), and where in the gene it was located. To do this, we first
-extracted the reference genome sequence surrounding the variant using
-the bedtools ``getfasta'' function \citep{quinlan2010}. Next, we
-re-annotated each sequence using BLAST to get the predicted gene
-location based on the closest annotated reference \citep{altschul1997}.
-We then translated each variant to amino acids and aligned this
-translation to that of the gene annotation to ensure proper frame of
-reference using Geneious \citep{kearse2012}. After ensuring proper
-translation, we characterized variants as within or outside the coding
-sequence of the gene and as either synonymous or non-synonymous.
-
-In the splined window analysis, we identified outlier regions using
-\emph{F\textsubscript{ST}} and differences in nucleotide diversity
-(\(\pi_{diff}\)) between naive and recovering populations. First, we
-calculated per-site \emph{F\textsubscript{ST}} between the naive and
-recovering individuals for all bi-allelic SNPs in the 30 largest contigs
-(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
-calculated per-site nucleotide diversity \(\pi\) separately for
-individuals from the naive and recovering populations using VCFtools,
-then calculated \(\pi_{diff}\) for each population
-(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
-values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
-size-sorted chromosome number and adjusted the SNP position based on the
-relative position in the genome (for more efficient data processing and
-to better contextualize the strength of the outlier signals in different
-regions of the genome). We then used the GenWin R package
-\citep{beissinger2015} to conduct a splined discrete window analysis for
-\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
-where non-overlapping window boundaries should occur by identifying
-inflection points in the spline fitted to \emph{F\textsubscript{ST}} and
-\(\pi_{diff}\) values along the genome, therefore balancing false
-positive and false negative results that occur using other window-based
-methods \citep{beissinger2015}. This method also calculates a
-W-statistic allowing for outlier identification. We identified outliers
-as those with a W-statistic greater than 4 standard deviations above the
-mean for \emph{F\textsubscript{ST}} or above/below the mean for
-\(\pi_{diff}\). These standard deviations represent strict criteria to
-select only the top \textasciitilde{} 0.3\% of windows. Shared outliers
-were then identified as those that were outliers in both analyses,
-meaning that they showed (i) high differentiation between naive and
-recovering populations, and (ii) differential patterns of nucleotide
-diversity in the same region. Finally, we extracted gene transcripts
-mapped within each region and retrieved annotation for that region using
-BLAST (\textbf{Supporting Information}: Datasets S3, S4).
-
-Code used for genomic analyses and to create figures is available from
-the following GitHub repository:
-\url{https://github.com/allie128/mylf-selection}.
-
-}
-
-\showmatmethods{} % Display the Materials and Methods section
-
-\acknow{We thank the following for important contributions to this study: E.
-Hegeman and A. Lindauer--project and data management, field and
-laboratory work; A. Barbella and K. Rose--laboratory work; C. Kamoroff
-and numerous summer technicians--field work; staff at Sequoia-Kings
-Canyon and Yosemite National Parks, Inyo and Sierra National Forests,
-California Department of Fish and Wildlife, U.S. Fish and Wildlife
-Service, University of California-Santa Barbara Institutional Animal
-Care and Use Committee, and Sierra Nevada Aquatic Research
-Laboratory--research permits, logistical support, and/or field
-assistance. This project was supported by grants from the National Park
+probabilities and other known vital rates (Supplementary Table 1), we
+performed four analyses. First, we compute the long-run growth rate
+\(\lambda\) for each of our 12 translocated populations to determine if
+the populations were deterministically predicted to grow or decline in
+the long-run. Second, we compute the elasticity and sensitivity of
+\(\lambda\) to nine model parameters to quantify how much changes in
+these parameters affected the long-run growth rate (Supplementary Fig.
+4). This also helped us determine where in the model environmental
+variation in juvenile survival would have the largest effects on
+population dynamics. Third, we included demographic stochasticity and
+environmental stochasticity in \(\sigma_{J_1}\) in our model and
+simulated the 50-year viability (i.e., 1 - extinction probability) of
+populations given an introduction of 40 adult individuals into an
+unoccupied habitat. Finally, we fit our model to our longest
+translocation trajectory to confirm that our model could reasonably
+reproduce the observed recovery trajectories of \emph{R. sierrae}
+following reintroductions. Additional details are provided in
+\textbf{Supplementary Methods - Population viability modeling - Model
+analysis and simulation}.
+
+\subsubsection{Data availability}\label{data-availability}
+
+The data used in this study are available at
+\url{https://doi.org/10.5281/zenodo.13851590} \citep{knapp2024a}. This
+includes the estimated survival of each reintroduced frog cohort, data
+used to identify predictors of 1-year frog survival, data used to
+estimate growth rates and extinction probabilities of reintroduced
+populations, Bd loads on frogs before and after translocation, and
+survival probabilities of translocated versus naturally recruited
+adults. For the relevant figures, source data are provided in a Source
+Data file. The CMR datasets from each reintroduction site that we used
+to calculate post-translocation frog survival are available at
+\url{https://doi.org/10.5281/zenodo.13851652} \citep{knapp2024b}.
+
+\subsubsection{Code availability}\label{code-availability}
+
+R code to replicate our analyses is available at
+\url{https://doi.org/10.5281/zenodo.13851590} \citep{knapp2024a}. R code
+to analyze the CMR datasets is available at
+\url{https://doi.org/10.5281/zenodo.13851652} \citep{knapp2024b}.
+
+\subsubsection{References}\label{references}
+
+\renewcommand{\bibsection}{}
+\bibliography{translocation.bib}
+
+\newpage
+
+\subsubsection{Acknowledgements}\label{acknowledgements}
+
+We thank the following for important contributions to this study: E.
+Hegeman and A. Lindauer (project and data management, field and
+laboratory work); A. Barbella and K. Rose (laboratory work); numerous
+summer technicians (field work); staff at Sequoia-Kings Canyon and
+Yosemite National Parks, Inyo and Sierra National Forests, California
+Department of Fish and Wildlife, U.S. Fish and Wildlife Service, Sierra
+Nevada Aquatic Research Laboratory (DOI: 10.21973/N3966F), and
+University of California-Santa Barbara Institutional Animal Care and Use
+Committee (research permits, logistical support, and/or field
+assistance). This project was supported by grants from the National Park
 Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National
 Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs;
-DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).}
-
-\showacknow{} % Display the acknowledgments section
-
-% Bibliography
-\bibliography{translocation}
-
-\clearpage
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/conceptual_model.png}
+DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).
+
+\subsubsection{Author contributions}\label{author-contributions}
+
+R.A.K., M.Q.W., M.B.J., T.C.S., and R.L.G designed the research. R.A.K.,
+M.B.J., T.C.S., and R.L.G. contributed to data collection; M.B.J.
+contributed new analytic tools; R.A.K. and M.Q.W. conducted the
+analyses. R.A.K. and M.Q.W. wrote the manuscript with input from M.B.J.,
+T.C.S., and R.L.G.
+
+\subsubsection{\texorpdfstring{\textbf{Competing
+interests}}{Competing interests}}\label{competing-interests}
+
+The authors declare no competing interests.
+
+\subsubsection{Additional information}\label{additional-information}
+
+\textbf{Supplemental information} is available for this paper at
+
+\textbf{Correspondence} and requests for materials should be addressed
+to R.A.K.
+
+\newpage
+
+\subsubsection{Figures and Legends}\label{figures-and-legends}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{figures/conceptual_model.png}
 
@@
 
-\caption{\label{fig-recovery-model}For MYL frogs, a conceptual model
-depicting the Bd-caused decline and subsequent natural recovery (black
-text), facilitated recovery via reintroductions, and the linkages
-between these two pathways. Rectangles and hexagons represent outcomes
-and processes, respectively. Blue text indicates components that are
-included in the current study. The timeline shows the general sequence
-of the components, with the dotted portion indicating a projection into
-the future.}
-
-\end{figure*}
+\caption{\label{fig-recovery-model}\textbf{Conceptual model of Bd-caused
+decline and recovery in \emph{R. sierrae}.} For the recovery portion,
+the model depicts both natural recovery and facilitated recovery via
+reintroductions, as well as the linkages between these two pathways.
+Rectangles and hexagons represent outcomes and processes, respectively.
+Blue text indicates components that are included in the current study.
+The general temporal scale of the depicted scenario is provided by the
+timeline, with the dashed portion indicating a projection into the
+future.}
+
+\end{figure}%
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=0.5\textwidth]{figures/translocation_survival_bysiteid.png}
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{figures/map_translocation_points.png}
 
@@
 
-\caption{\label{fig-translocation-survival}Median 1-year survival for
-each cohort of translocated frogs at the 12 recipient sites, as
-estimated for each site from the mrmr CMR model. Error bars show the
-95\% uncertainty intervals. Sites are arranged along the x-axis using
-the average of the median 1-year survival per translocation at each
-site. Dot colors indicate the donor population from which frogs in each
-translocated cohort were collected. When multiple translocations were
-conducted to a site, points and error bars are slightly offset to avoid
-overlap.}
-
-\end{figure*}
+\caption{\label{fig-yosemap}\textbf{Locations of translocated and donor
+\emph{R. sierrae} populations.} All populations are in Yosemite National
+Park (park boundary indicated by black polygon). Symbol shapes indicate
+the donor population used for each translocation site, and 5-digit
+numbers identify each donor and translocation site. To obscure the exact
+locations of populations, random noise was added to all point
+coordinates. Inset map shows the location of Yosemite within California.
+In both maps, topography is shown with a grayscale hillshade layer, and
+elevation is indicated by colors layered over the hillshade. In the
+absence of hillshading, dark green colors indicate the lowest elevations
+and white indicates the highest elevation.}
+
+\end{figure}%
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{figures/translocation_survival_bysiteid.png}
 
@@
 
-\caption{\label{fig-cond-effects}Results from the among-site
-meta-analysis showing conditional effects of the important predictors of
-1-year frog survival (expressed as a probability): (A) winter severity
-in the year following translocation, (B) elevation of recipient site,
-(C) donor population, and (D) sex. In (A) and (B), blue lines are
-medians and gray ribbons are 95\% uncertainty intervals. In (C) and (D),
-box plots show medians, first and third quartiles, largest and smallest
-values within 1.5x interquartile range, and values outside the 1.5x
-interquartile range.}
-
-\end{figure*}
+\caption{\label{fig-translocation-survival}\textbf{1-year survival for
+each cohort of translocated frogs}. Median cohort-level survival was
+estimated for each of the 12 recipient sites using mrmr CMR models.
+Error bars show the 95\% uncertainty intervals. Sites are arranged along
+the x-axis using the average of the median 1-year survival per
+translocation at each site. Dot colors indicate the donor population
+from which frogs in each translocated cohort were collected. When
+multiple translocations were conducted to a site, points and error bars
+are slightly offset to avoid overlap.}
+
+\end{figure}%
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{figures/cond_effects_plot.png}
 
@@
 
-\caption{\label{fig-viability}Results from population viability
-analysis: (A) Predicted long-run growth rate \(\lambda\) for different
-values of yearly adult survival probability \(\sigma_{A_R}\) and year 1
-juvenile survival and recruitment probability \(\sigma_{J_1}\), given
-the parameterized, deterministic model. Colored points show the
-predicted \(\lambda\) values for the twelve translocated populations
-when year 1 juvenile survival probability is \(\sigma_{J_1} = 0.09\)
-(indicated by the dashed line). The red line shows where
-\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
-other points. (B) Predicted 50-year extinction probabilities of the 12
-translocated populations, given demographic stochasticity, environmental
-variability in \(\sigma_{J_1}\), and different mean values of
-\(\sigma_{J_1}\). There are 6 lines at extinction probability = 1, 5 of
-which (70414--70619) are hidden beneath the line for 70505 when
-\(\sigma_{J_1} < 0.10\). (C) 100 simulated trajectories (black lines)
-from the population viability model that most closely matched the
-observed abundance trajectory of adult amphibians at site 70550 (light
-orange).}
-
-\end{figure*}
+\caption{\label{fig-cond-effects}\textbf{Conditional effects of
+predictors of 1-year frog survival}. Results are from the among-site
+meta-analysis and 1-year frog survival is expressed as a probability.
+\textbf{a} winter severity in the year following translocation,
+\textbf{b} elevation of recipient site, \textbf{c} donor population, and
+\textbf{d} sex. In (a) and (b), blue lines are medians and gray ribbons
+are 95\% uncertainty intervals for the posterior distributions of the
+predicted effects. In (c) and (d), box plots show medians (horizontal
+line), first and third quartiles (hinges), largest and smallest values
+within 1.5x interquartile range (whiskers), and values outside the 1.5x
+interquartile range (dots) for the posterior distributions of the
+predicted effects.}
+
+\end{figure}%
 
@@
 
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{figures/pop_viability_figures_for_manuscript.png}
 
@@
 
-\caption{\label{fig-allelefrequencies}Evidence for selection on
-individual variants in recovering MYL frog populations at the landscape
-scale. For each of the 9 naive and recovering MYL frog populations
-(indicated by numbered points), adjacent pie charts show allele
-frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
-LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
-Charts are superimposed on a map of the Sierra Nevada study area, with
-Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
-shown in dark gray, and the range boundary for MYL frogs shown in light
-gray. The inset map locates the study area and range boundary in
-California. Sites 1 and 4 (site id = 72996 and 70567, respectively) were
-also used as sources of frogs in the frog population recovery study.}
-
-\end{figure*}
-
-\newpage
-
-\begin{figure*}
-
-{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}
-
-}
-
-\caption{\label{fig-spline-manhattan}Evidence for selection on genomic
-regions in recovering MYL frog populations. Manhattan plot of the
-results from the splined window analysis showing outlier regions for the
-difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
-\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
-upper red dashed line and below the lower red dashed line. In (B),
-outlier regions are shown above the single dashed red line. Outlier
-regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
-shown in blue and outlier regions for both \(\pi_{diff}\) and
-\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
-(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
-upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
-line). (D) Magnified Contig8 from from (B) showing the
-\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
-TCF19 and VARS. This region of the genome contains 8 annotated genes
-known to occur in the extended MHC Class I and III regions.}
-
-\end{figure*}
-
-\newpage
+\caption{\label{fig-viability}\textbf{Estimated growth rates and
+extinction probabilities of reintroduced populations}. \textbf{a}
+Predicted long-run growth rate \(\lambda\) for different values of
+yearly adult survival probability \(\sigma_{A_R}\) and year-1 juvenile
+survival probability \(\sigma_{J_1}\), given the parameterized,
+deterministic model. Colored points show the predicted \(\lambda\)
+values for the twelve translocated populations when year-1 juvenile
+survival probability is \(\sigma_{J_1} = 0.10\) (indicated by the dashed
+line). The red line shows where \(\lambda = 1\). Note that the point for
+70413 is mostly hidden behind other points. \textbf{b} Predicted 50-year
+extinction probabilities of the 12 reintroduced populations, given
+demographic stochasticity, environmental variability in
+\(\sigma_{J_1}\), and different mean values of \(\sigma_{J_1}\). There
+are 6 lines at extinction probability = 1, 5 of which (70414--70619) are
+partially or completely hidden beneath the line for 70505. \textbf{c}
+100 simulated trajectories (black lines) from the population viability
+model that most closely matched the observed abundance trajectory of
+adult amphibians at site 70550 (light orange).}
+
+\end{figure}%
+
 \end{document}

### 2024-09-27 doc/manuscript/convert_qmd_to_pnas_latex.py
Remove unneeded file

@@ -1,354 +1,0 @@
-import subprocess
-import sys
-import re
-
-"""
-Convert Quarto to PNAS format for translocation manucript
-
-Generates two files
-
-1. transoclation_pnas.tex : Main text in PNAS format
-2. transoclation_pnas_SI.tex : SI in PNAS format
-"""
-
-# Files for converting
-args = ['translocation.qmd', 'translocation.tex']
-
-### Build the main manuscript ####
-
-# Convert from qmd to latex with pandoc
-subprocess.call(['quarto', 'render', args[0], '--to', "latex"])
-
-# Read in the latex manuscript
-with open(args[1], 'r') as fin:
-    latex_manuscript = fin.readlines()
-latex_manuscript = "".join(latex_manuscript)
-
-# Remove all the references to width and height
-latex_manuscript = re.sub("includegraphics\[.*\]",
-						  "includegraphics",
-						  latex_manuscript)
-
-# Drop the References place holder
-latex_manuscript = latex_manuscript.replace("\\hypertarget{references}{%\n\\subsubsection{References}\\label{references}}\n\nSee combined references at end of document (temporary placeholder).\n\n\\hfill\\break\n\n", "")
-
-with open("PNAS-template-empty.tex") as fin:
-	pnas_template = fin.readlines()
-pnas_template = "".join(pnas_template)
-
-# Fill in each section of manuscript
-
-## Title ## 
-
-title_start = latex_manuscript.find("\\subsection{Title")
-title_end = latex_manuscript[title_start:].find("}") + title_start
-title = latex_manuscript[title_start:title_end].replace("\n", " ").split("\\subsection{Title: ")[1]
-
-# Insert title
-pnas_ms = pnas_template.replace("\\title{}", "\\title{" + title + "}")
-
-## Abstract ##
-
-abkey1 = "\\hypertarget{abstract}{%\n\\subsubsection{Abstract}\\label{abstract}}"
-abkey2 = "\\hypertarget{significance}"
-abstract_start = latex_manuscript.find(abkey1)
-abstract_end = latex_manuscript.find(abkey2)
-abstract = latex_manuscript[abstract_start:abstract_end].split(abkey1)[1].strip("\n")
-
[…]

### 2024-09-27 doc/manuscript/submission_files/translocation.tex
Use default cite method instead of natbib

--- 2024-09-27 doc/manuscript/translocation.tex
@@
 \makeatother
+% journal abbreviations
+\usepackage{jabbrv}
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{plainnat}
 \usepackage{bookmark}
@@
 
-Human activities are devastating global biodiversity
-\citep{ceballos2015}, with important implications for ecosystem
-resilience and human welfare \citep{naeem2009}. One consequence of human
-alteration of the biosphere is an increase in emerging infectious
-diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
-threat to wildlife populations \citep{daszak2000}, and have caused
-dramatic declines and extinctions in a wide range of taxa, including
-echinoderms, mammals, birds, and amphibians
-\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
-are experiencing particularly devastating impacts of disease due to the
+Human activities are devastating global biodiversity\textsuperscript{1},
+with important implications for ecosystem resilience and human
+welfare\textsuperscript{2}. One consequence of human alteration of the
+biosphere is an increase in emerging infectious
+diseases\textsuperscript{3,4}. Such diseases pose a severe threat to
+wildlife populations\textsuperscript{5}, and have caused dramatic
+declines and extinctions in a wide range of taxa, including echinoderms,
+mammals, birds, and amphibians\textsuperscript{6--9}. Amphibians are
+experiencing particularly devastating impacts of disease due to the
 recent emergence and global spread of the highly virulent amphibian
-chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
-\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
-have experienced Bd-caused declines, and numerous susceptible taxa are
-extinct in the wild \citep{scheele2019}. The escalating frequency of
+chytrid fungus, \emph{Batrachochytrium dendrobatidis}
+(Bd)\textsuperscript{8,10}. By one estimate, hundreds of species have
+experienced Bd-caused declines, and numerous susceptible taxa are
+extinct in the wild\textsuperscript{8}. The escalating frequency of
 disease emergence and the strength of impacts on wild populations
@@
 by a particular burden) are key mechanisms by which hosts reduce disease
-impacts \citep{raberg2009}. The effectiveness of these mechanisms has
+impacts\textsuperscript{11}. The effectiveness of these mechanisms has
 direct consequences for the persistence and recovery of host populations
-in the presence of disease \citep{brannelly2021}. Host immunity and
+in the presence of disease\textsuperscript{12}. Host immunity and
 evolution both play important roles in the development of resistance and
 tolerance, and are key targets for strategies aimed at mitigating
-disease impacts in the wild \citep{garner2016}.
+disease impacts in the wild\textsuperscript{13}.
 
 Evidence of natural recovery in the many Bd-impacted amphibian
-populations is surprisingly rare (for notable exceptions, see
-\citep{scheele2017, voyles2018, knapp2016}). This apparent low
-resilience to disease may be due to the limited ability of many
-amphibians to develop Bd resistance or tolerance via either immunity
-\citep{rosenblum2012, fites2013, grogan2018a} or evolution (but see
-\citep{savage2016, grogan2018b}). In turn, this would also reduce the
-effectiveness of potential Bd mitigation strategies. For example, a
-limited ability by amphibians to develop resistance or tolerance
-suggests that reintroduction of amphibians into sites to reestablish
-populations extirpated by Bd will often result not in population
-recovery, but instead in the rapid reinfection and mortality of the
-introduced animals and/or their progeny
-\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
-future of many amphibian species threatened by Bd appears bleak.
+populations is surprisingly rare (for notable exceptions,
+see\textsuperscript{14--16}). This apparent low resilience to disease
+may be due to the limited ability of many amphibians to develop Bd
+resistance or tolerance via either immunity\textsuperscript{17--19} or
+evolution (but see\textsuperscript{20,21}). In turn, this would also
+reduce the effectiveness of potential Bd mitigation strategies. For
+example, a limited ability by amphibians to develop resistance or
+tolerance suggests that reintroduction of amphibians into sites to
+reestablish populations extirpated by Bd will often result not in
+population recovery, but instead in the rapid reinfection and mortality
+of the introduced animals and/or their progeny\textsuperscript{22--25}.
+If true, the future of many amphibian species threatened by Bd appears
+bleak.
 
 The Sierra Nevada yellow-legged frog (\emph{Rana sierrae}), is
-emblematic of the global decline of amphibians caused by Bd
-\citep{scheele2019}. Once the most common amphibian in the high
-elevation portion of California's Sierra Nevada mountains, USA
-\citep{grinnell1924}, during the past century this frog has disappeared
-from more than 90\% of its historical range \citep{vredenburg2007}. Due
-to the severity of its decline and the increasing probability of
-extinction, \emph{R. sierrae} is now listed as ``endangered'' under the
-U.S. Endangered Species Act. The decline of \emph{R. sierrae} was
-initiated by the introduction of non-native trout into the extensive
-fishless region of the Sierra Nevada \citep{bradford1989, knapp2000}
-starting in the late 1800s. The arrival of Bd in the mid-1900s and its
-subsequent spread \citep{vredenburg2019} caused additional large-scale
-population extirpations \citep{vredenburg2010, rachowicz2006}. These
-Bd-caused declines are fundamentally different from the fish-caused
-declines because fish eradication is feasible \citep{knapp1998} and
-results in the rapid recovery of frog populations
-\citep{knapp2007, vredenburg2004}. In contrast, Bd appears to persist in
-habitats even in the absence of amphibian hosts \citep{walker2007}, and
-therefore represents a long-term alteration of invaded ecosystems that
-amphibians will need to overcome to reestablish populations.
+emblematic of the global decline of amphibians caused by
+Bd\textsuperscript{8}. Once the most common amphibian in the high
+elevation portion of California's Sierra Nevada mountains,
+USA\textsuperscript{26}, during the past century this frog has
+disappeared from more than 90\% of its historical
+range\textsuperscript{27}. Due to the severity of its decline and the
+increasing probability of extinction, \emph{R. sierrae} is now listed as
+``endangered'' under the U.S. Endangered Species Act. The decline of
+\emph{R. sierrae} was initiated by the introduction of non-native trout
+into the extensive fishless region of the Sierra
+Nevada\textsuperscript{28,29} starting in the late 1800s. The arrival of
+Bd in the mid-1900s and its subsequent spread\textsuperscript{30} caused
+additional large-scale population extirpations\textsuperscript{31,32}.
+These Bd-caused declines are fundamentally different from the
+fish-caused declines because fish eradication is
+feasible\textsuperscript{33} and results in the rapid recovery of frog
+populations\textsuperscript{34,35}. In contrast, Bd appears to persist
+in habitats even in the absence of amphibian hosts\textsuperscript{36},
+and therefore represents a long-term alteration of invaded ecosystems
+that amphibians will need to overcome to reestablish populations.
 
 Despite the catastrophic impact of Bd on \emph{R. sierrae}, wherein most
-Bd-naive populations are extirpated following Bd arrival
-\citep{vredenburg2010}, some populations have persisted after epizootics
-\citep{briggs2010} and are now recovering \citep{knapp2016}
-(Fig.~\ref{fig-recovery-model}). During the epizootic phase, \emph{R.
-sierrae} are highly susceptible to Bd infection as indicated by very
-high infection intensities (``load''). In contrast, frogs in recovering
-populations have reduced susceptibility to Bd infection
-\citep{knapp2016}, with loads on adults consistently in the
-low-to-moderate range \citep{briggs2010, knapp2011, joseph2018}. This
-reduced susceptibility is evident even under controlled laboratory
-conditions \citep{knapp2016}, indicative of host resistance against Bd
+Bd-naive populations are extirpated following Bd
+arrival\textsuperscript{31}, some populations have persisted after
+epizootics\textsuperscript{37} and are now
+recovering\textsuperscript{16} (Fig.~\ref{fig-recovery-model}). During
+the epizootic phase, \emph{R. sierrae} are highly susceptible to Bd
+infection as indicated by very high infection intensities (``load''). In
+contrast, frogs in recovering populations have reduced susceptibility to
+Bd infection\textsuperscript{16}, with loads on adults consistently in
+the low-to-moderate range\textsuperscript{37--39}. This reduced
+susceptibility is evident even under controlled laboratory
+conditions\textsuperscript{16}, indicative of host resistance against Bd
 infection and not simply an effect of factors external to individual
@@
 several mechanisms that are not mutually exclusive, including natural
-selection for more resistant genotypes \citep{savage2016, grogan2018b},
-acquired immunity \citep{grogan2018a}, and/or inherent
-between-population differences that pre-date Bd exposure. The possible
-evolution of resistance and subsequent population recovery is consistent
-with that expected under ``evolutionary rescue'', whereby rapid
-evolutionary change increases the frequency of adaptive alleles and
-restores positive population growth \citep{carlson2014, searle2020}.
-This intriguing possibility also suggests an opportunity to expand
-recovery beyond the spatial scale possible under natural recovery by
-using resistant frogs from recovering populations in reintroductions to
-vacant habitats \citep{joseph2018, mendelson2019}
+selection for more resistant genotypes\textsuperscript{20,21}, acquired
+immunity\textsuperscript{19}, and/or inherent between-population
+differences that pre-date Bd exposure. The possible evolution of
+resistance and subsequent population recovery is consistent with that
+expected under ``evolutionary rescue'', whereby rapid evolutionary
+change increases the frequency of adaptive alleles and restores positive
+population growth\textsuperscript{40,41}. This intriguing possibility
+also suggests an opportunity to expand recovery beyond the spatial scale
+possible under natural recovery by using resistant frogs from recovering
+populations in reintroductions to vacant habitats\textsuperscript{39,42}
 (Fig.~\ref{fig-recovery-model}).
@@
 chytridiomycosis (i.e., the disease caused by Bd) and associated frog
-mortality \citep{joseph2018, vredenburg2010} (Supplementary Fig. 1).
-Although it is possible that the observed relatively small changes in
-load are a consequence of individuals with high Bd loads dying and
-therefore being unavailable for sampling during the post-translocation
-period, the fact that there was little difference in pre- versus
-post-translocation Bd loads even in those populations that had very high
-frog survival (70556, 74976 - see below; Supplementary Fig. 1) suggests
-a true lack of substantial change in Bd load.
+mortality\textsuperscript{31,39} (Supplementary Fig. 1). Although it is
+possible that the observed relatively small changes in load are a
+consequence of individuals with high Bd loads dying and therefore being
+unavailable for sampling during the post-translocation period, the fact
+that there was little difference in pre- versus post-translocation Bd
+loads even in those populations that had very high frog survival (70556,
+74976 - see below; Supplementary Fig. 1) suggests a true lack of
+substantial change in Bd load.
 
@@
 to determine the self-sustainability of a reintroduced population (for
-an example in \emph{R. sierrae}, see \citep{joseph2018}), the use of
+an example in \emph{R. sierrae}, see\textsuperscript{39}), the use of
 proxies is essential for providing shorter-term insights into
@@
 period immediately following translocation. These estimates were made
-using site-specific models analyzed using the mrmr package
-\citep{joseph2019} (\url{https://snarl1.github.io/mrmr/index.html}). We
-use these estimates to describe general patterns of frog survival in all
-translocated cohorts, and in an among-site meta-analysis of frog
-survival to identify important predictors of 1-year frog survival (e.g.,
-Bd load).
+using site-specific models analyzed using the mrmr
+package\textsuperscript{43}
+(\url{https://snarl1.github.io/mrmr/index.html}). We use these estimates
+to describe general patterns of frog survival in all translocated
+cohorts, and in an among-site meta-analysis of frog survival to identify
+important predictors of 1-year frog survival (e.g., Bd load).
 
@@
 establishment. However, a decade or more of surveys may be necessary to
-confirm that populations are in fact self-sustaining \citep{joseph2018}.
-To extend our inferences of population establishment beyond those
-possible from the site-specific CMR data, we developed a population
-viability model. Specifically, to test whether the observed yearly adult
-survival probabilities in translocated populations were sufficient for
-long-term viability, we built a stage-structured matrix model that
-captured known frog demography and included demographic and
-environmental stochasticity. We parameterized the model using CMR data
-from translocated populations and known life history values in this
-system (Supplementary Table 1). Results from a sensitivity analysis are
-provided in Supplementary Fig. 4.
+confirm that populations are in fact
+self-sustaining\textsuperscript{39}. To extend our inferences of
+population establishment beyond those possible from the site-specific
+CMR data, we developed a population viability model. Specifically, to
+test whether the observed yearly adult survival probabilities in
+translocated populations were sufficient for long-term viability, we
+built a stage-structured matrix model that captured known frog
+demography and included demographic and environmental stochasticity. We
+parameterized the model using CMR data from translocated populations and
+known life history values in this system (Supplementary Table 1).
+Results from a sensitivity analysis are provided in Supplementary Fig.
+4.
 
@@
 
-Disease-induced population declines are decimating global biodiversity
-\citep{daszak2000}, but broadly-applicable strategies to recover
-affected species are generally lacking (e.g., \citep{garner2016}). Here,
-we tested the possibility that populations of resistant individuals from
-naturally recovering populations can be used to reestablish extirpated
-populations of the endangered \emph{R. sierrae} in the presence of a
-highly virulent pathogen (Bd). Our results indicate (i) the capacity of
-reintroduced populations to become established and eventually recover
-despite ongoing disease, (ii) that post-translocation frog survival is
-influenced by site, cohort, and individual-level characteristics, and
-(iii) that 50\% of the reintroduced populations are likely to persist
-over a 50-year period. In light of the generally low success rate of
-amphibian reintroduction efforts \citep{dodd2005}, our success in
+Disease-induced population declines are decimating global
+biodiversity\textsuperscript{5}, but broadly-applicable strategies to
+recover affected species are generally lacking
+(e.g.,\textsuperscript{13}). Here, we tested the possibility that
+populations of resistant individuals from naturally recovering
+populations can be used to reestablish extirpated populations of the
+endangered \emph{R. sierrae} in the presence of a highly virulent
+pathogen (Bd). Our results indicate (i) the capacity of reintroduced
+populations to become established and eventually recover despite ongoing
+disease, (ii) that post-translocation frog survival is influenced by
+site, cohort, and individual-level characteristics, and (iii) that 50\%
+of the reintroduced populations are likely to persist over a 50-year
+period. In light of the generally low success rate of amphibian
+reintroduction efforts\textsuperscript{44}, our success in
 reestablishing \emph{R. sierrae} populations via reintroduction of
@@
 establishment of Bd in previously-naive populations, adult frogs develop
-high Bd loads that lead to mass die-offs \citep{vredenburg2010}. In
+high Bd loads that lead to mass die-offs\textsuperscript{31}. In
 contrast, in recovering populations adult frogs typically have
 low-to-moderate and relatively constant Bd loads and mass die-offs are
-not observed \citep{briggs2010, knapp2011} (see also Supplementary Fig.
-2). The differences in Bd load of frogs from naive and recovering
-populations are also observed in controlled laboratory studies (see
-Figure 4 in \citep{knapp2016}), and clearly indicate that frogs from
-recovering populations exhibit resistance against Bd infection. Finally,
+not observed\textsuperscript{37,38} (see also Supplementary Fig. 2). The
+differences in Bd load of frogs from naive and recovering populations
+are also observed in controlled laboratory studies (see Figure 4
+in\textsuperscript{16}), and clearly indicate that frogs from recovering
+populations exhibit resistance against Bd infection. Finally,
 reintroductions of \emph{R. sierrae} collected from Bd-naive populations
@@
 given that reintroduced \emph{R. sierrae} populations may often take
-decades to achieve our ultimate goal of self-sustainability
-\citep{joseph2018}. Making well-supported projections about the
-long-term outcome of reintroduction efforts from shorter-term
-information is crucial for adaptive management of species reintroduction
-programs \citep{seddon2007}, including the one we are carrying out for
-\emph{R. sierrae}. Specifically, the combined results from our
-reintroduction study and viability modeling suggest that survival of
-frogs in the first year following translocation is an effective proxy of
-longer-term survival and population viability. In addition, given the
-repeatability of frog survival at a site, 1-year frog survival also
-serves as an effective proxy of site quality (i.e., the ability of a
-site to support high frog survival and a viable frog population over the
-long term). This proxy of site quality is important in the \emph{R.
-sierrae} system because accurately predicting the ability of a site to
-support a viable frog population \emph{a priori} remains difficult, even
-after conducting 24 translocations over 16 years.
+decades to achieve our ultimate goal of
+self-sustainability\textsuperscript{39}. Making well-supported
+projections about the long-term outcome of reintroduction efforts from
+shorter-term information is crucial for adaptive management of species
+reintroduction programs\textsuperscript{45}, including the one we are
+carrying out for \emph{R. sierrae}. Specifically, the combined results
+from our reintroduction study and viability modeling suggest that
+survival of frogs in the first year following translocation is an
+effective proxy of longer-term survival and population viability. In
+addition, given the repeatability of frog survival at a site, 1-year
+frog survival also serves as an effective proxy of site quality (i.e.,
+the ability of a site to support high frog survival and a viable frog
+population over the long term). This proxy of site quality is important
+in the \emph{R. sierrae} system because accurately predicting the
+ability of a site to support a viable frog population \emph{a priori}
+remains difficult, even after conducting 24 translocations over 16
+years.
 
@@
 Bd infection, individual and population-level impacts of Bd are still
-evident. In an earlier study of 2 of our 12 translocated populations
-\citep{joseph2018}, Bd infection and load had detectable effects on the
-survival of adults and may have influenced population establishment
-(sites referred to as ``Alpine'' and ``Subalpine'' in \citet{joseph2018}
-are identified as ``70550'' and ``70505'' in the current study).
-Applying similar analyses to all 12 of our translocated populations
-would likely provide a broader perspective of the ongoing effect of Bd.
-In addition to these important but relatively subtle effects of Bd on
-adults, the impacts on younger life stages are more apparent. \emph{R.
-sierrae} immediately following metamorphosis (``metamorphs'') are highly
-susceptible to Bd infection \citep{ellison2019} and as a result
-experience high mortality \citep{rachowicz2006}. This high
-susceptibility of metamorphs is documented in numerous species of
-anurans, and may result from the poorly developed immune system
-characteristic of this life stage \citep{humphries2022}. In both
-naturally recovering and reintroduced \emph{R. sierrae} populations, we
-suggest that the high mortality of metamorphs is an important limitation
-on subsequent recruitment of new adults. Therefore, although adult
-\emph{R. sierrae} appear relatively resistant, Bd infection continues to
-have important limiting effects on recovering populations (see also
-\citep{hollanders2022}). The ongoing high susceptibility of metamorphs
-to chytridiomycosis also highlights the critical need to develop better
-methods of monitoring this vulnerable life stage, the results of which
-could significantly advance the conservation and recovery of Bd-impacted
-amphibians.
+evident. In an earlier study of 2 of our 12 translocated
+populations\textsuperscript{39}, Bd infection and load had detectable
+effects on the survival of adults and may have influenced population
+establishment (sites referred to as ``Alpine'' and ``Subalpine''
+in\textsuperscript{39} are identified as ``70550'' and ``70505'' in the
+current study). Applying similar analyses to all 12 of our translocated
+populations would likely provide a broader perspective of the ongoing
+effect of Bd. In addition to these important but relatively subtle
+effects of Bd on adults, the impacts on younger life stages are more
+apparent. \emph{R. sierrae} immediately following metamorphosis
+(``metamorphs'') are highly susceptible to Bd
+infection\textsuperscript{46} and as a result experience high
+mortality\textsuperscript{32}. This high susceptibility of metamorphs is
+documented in numerous species of anurans, and may result from the
+poorly developed immune system characteristic of this life
+stage\textsuperscript{47}. In both naturally recovering and reintroduced
+\emph{R. sierrae} populations, we suggest that the high mortality of
+metamorphs is an important limitation on subsequent recruitment of new
+adults. Therefore, although adult \emph{R. sierrae} appear relatively
+resistant, Bd infection continues to have important limiting effects on
+recovering populations (see also\textsuperscript{48}). The ongoing high
+susceptibility of metamorphs to chytridiomycosis also highlights the
+critical need to develop better methods of monitoring this vulnerable
+life stage, the results of which could significantly advance the
+conservation and recovery of Bd-impacted amphibians.
 
@@
 fragmentation due to rugged geography and the ongoing presence of
-introduced, predatory trout in the majority of lakes and streams
-\citep{bradford1993}. Collectively, these factors often prevent
+introduced, predatory trout in the majority of lakes and
+streams\textsuperscript{49}. Collectively, these factors often prevent
 naturally recovering \emph{R. sierrae} populations from expanding via
@@
 current and future natural and anthropogenic impacts than more
-fragmented populations (e.g., \citep{heard2015}).
+fragmented populations (e.g.,\textsuperscript{50}).
 
@@
 the hundreds of Bd-impacted amphibian species globally. Importantly, in
-addition to the natural recovery documented for \emph{R. sierrae}
-\citep{knapp2016}, other amphibian species are also showing evidence of
-post-epizootic recovery in the presence of Bd
-\citep{scheele2017, voyles2018} and suggest the possibility of also
-using animals from these recovering populations to reestablish
-extirpated populations. As with \emph{R. sierrae}, the feasibility and
-long-term success of such efforts will depend on the availability of
-robust donor populations containing individuals that have sufficiently
-high resistance to allow frog survival and population growth in the
-presence of Bd. We suggest that several components of recovery efforts
-could play important roles in maximizing their success. First, an
-adaptive management approach such as the one we utilized, in which the
-outcomes of recovery actions are rigorously assessed and this
-information is used to reduce uncertainties and inform future actions,
-will often be essential \citep{canessa2019}. Second, given the apparent
-rarity in amphibians of evolution of resistance against Bd infection,
-developing an improved understanding of the factors facilitating or
-limiting the development of frog resistance and the effectiveness of
-selective breeding programs and other genetic interventions to enhance
-frog resistance \citep{kosch2022} may be particularly important.
-Finally, although we did not quantify the genetic structure of the donor
-and translocated populations of \emph{R. sierrae} as part of the current
+addition to the natural recovery documented for \emph{R.
+sierrae}\textsuperscript{16}, other amphibian species are also showing
+evidence of post-epizootic recovery in the presence of
+Bd\textsuperscript{14,15} and suggest the possibility of also using
+animals from these recovering populations to reestablish extirpated
+populations. As with \emph{R. sierrae}, the feasibility and long-term
+success of such efforts will depend on the availability of robust donor
+populations containing individuals that have sufficiently high
+resistance to allow frog survival and population growth in the presence
+of Bd. We suggest that several components of recovery efforts could play
+important roles in maximizing their success. First, an adaptive
+management approach such as the one we utilized, in which the outcomes
+of recovery actions are rigorously assessed and this information is used
+to reduce uncertainties and inform future actions, will often be
+essential\textsuperscript{51}. Second, given the apparent rarity in
+amphibians of evolution of resistance against Bd infection, developing
+an improved understanding of the factors facilitating or limiting the
+development of frog resistance and the effectiveness of selective
+breeding programs and other genetic interventions to enhance frog
+resistance\textsuperscript{52} may be particularly important. Finally,
+although we did not quantify the genetic structure of the donor and
+translocated populations of \emph{R. sierrae} as part of the current
 study, conducting genetic monitoring may be important to identify
@@
 populations (e.g., genetic bottlenecks) and allow timely implementation
-of actions to mitigate such effects \citep{whiteley2015}.
+of actions to mitigate such effects\textsuperscript{53}.
 
@@
 (\(\geq\) 40 mm snout-vent length) could be collected based on several
-years of VES and skin swab collections \citep{knapp2016}, and results
-from population genetic analyses \citep{poorten2017}. The populations
+years of VES and skin swab collections\textsuperscript{16}, and results
+from population genetic analyses\textsuperscript{54}. The populations
 that we selected contained hundreds of \emph{R. sierrae} adults and
 thousands of tadpoles. These relatively high abundances were the result
-of recent increases following previous Bd-caused declines
-\citep{knapp2016}. As is typical for recovering \emph{R. sierrae}
-populations, Bd prevalence in the donor populations was high
+of recent increases following previous Bd-caused
+declines\textsuperscript{16}. As is typical for recovering \emph{R.
+sierrae} populations, Bd prevalence in the donor populations was high
 (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
 3.06--3.78 ITS copies) was two or more orders of magnitude below the
-level at which frog mortality is expected
-\citep{vredenburg2010, joseph2018} (log\textsubscript{10}(load)
-\(\approx\) 5.78 ITS copies). Recipient sites to which frogs were
-translocated were chosen based on previous \emph{R. sierrae} presence
-(determined from VES and/or museum records) or characteristics that
-suggested high quality habitat for this species \citep{knapp2005}. At
-the beginning of this study, we had a relatively limited understanding
-of the factors that affect habitat quality. In subsequent years, we
-improved our site selection process by incorporating new information
-about important habitat features, in particular, overwintering habitats
-such as submerged boulders and overhanging banks. \emph{R. sierrae} were
-absent from all recipient sites prior to the first translocation. This
-determination was based on VES conducted at recipient sites over several
-years (see additional details below regarding frog detectability during
-VES).
+level at which frog mortality is expected\textsuperscript{31,39}
+(log\textsubscript{10}(load) \(\approx\) 5.78 ITS copies). Recipient
+sites to which frogs were translocated were chosen based on previous
+\emph{R. sierrae} presence (determined from VES and/or museum records)
+or characteristics that suggested high quality habitat for this
+species\textsuperscript{55}. At the beginning of this study, we had a
+relatively limited understanding of the factors that affect habitat
+quality. In subsequent years, we improved our site selection process by
+incorporating new information about important habitat features, in
+particular, overwintering habitats such as submerged boulders and
+overhanging banks. \emph{R. sierrae} were absent from all recipient
+sites prior to the first translocation. This determination was based on
+VES conducted at recipient sites over several years (see additional
+details below regarding frog detectability during VES).
 
@@
 stages have high detectability, and counts are highly repeatable and
-provide estimates of relative abundance \citep{knapp2000}.
+provide estimates of relative abundance\textsuperscript{29}.
 
@@
 Jolly-Seber CMR model with known additions to the population (i.e.,
-translocated cohorts), as implemented by the mrmr package
-\citep{joseph2019} and using R Statistical Software
-\citep{rsoftware2022} (v4.4.4) (see \textbf{Supplementary Methods - Frog
-population recovery - CMR model structure} for details). Briefly, the
-model tracks the states of \emph{M} individuals that comprise a
-superpopulation made up of real and pseudo-individuals
-\citep{joseph2018}. The possible states of individuals include ``not
-recruited'', ``alive'', and ``dead''. The possible observations of
-individuals include ``detected'' and ``not detected''. We assume that
-individuals that are in the ``not recruited'' or ``dead'' states are
-never detected (i.e., there are no mistakes in the individual PIT tag
-records). We also assume that new recruits were the result of
-within-site reproduction and not immigration from adjacent populations.
-This assumption is justified by the fact that no \emph{R. sierrae}
-populations were present within several kilometers of the translocation
-sites. For all models, we used mrmr defaults for priors, number of
-chains (4), and warmup and post-warmup iterations (2000 for each). We
-evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm
-using trace plots and Gelman-Rubin statistics (Rhat).
+translocated cohorts), as implemented by the mrmr
+package\textsuperscript{43} and using R Statistical
+Software\textsuperscript{56} (v4.4.4) (see \textbf{Supplementary Methods
+- Frog population recovery - CMR model structure} for details). Briefly,
+the model tracks the states of \emph{M} individuals that comprise a
+superpopulation made up of real and
+pseudo-individuals\textsuperscript{39}. The possible states of
+individuals include ``not recruited'', ``alive'', and ``dead''. The
+possible observations of individuals include ``detected'' and ``not
+detected''. We assume that individuals that are in the ``not recruited''
+or ``dead'' states are never detected (i.e., there are no mistakes in
+the individual PIT tag records). We also assume that new recruits were
+the result of within-site reproduction and not immigration from adjacent
+populations. This assumption is justified by the fact that no \emph{R.
+sierrae} populations were present within several kilometers of the
+translocation sites. For all models, we used mrmr defaults for priors,
+number of chains (4), and warmup and post-warmup iterations (2000 for
+each). We evaluated convergence of the Markov chain Monte Carlo (MCMC)
+algorithm using trace plots and Gelman-Rubin statistics (Rhat).
 
@@
 To identify important predictors of frog survival following
-translocation, we used multilevel Bayesian models
-\citep{gelman2013, gabry2019}. Included predictor variables describe
+translocation, we used multilevel Bayesian
+models\textsuperscript{57,58}. Included predictor variables describe
 characteristics of sites, translocated cohorts, and individuals (Bd
@@
 site\_id, translocation\_id, or translocation\_id nested within
-site\_id. We performed all analyses with the rstanarm package
-\citep{rstanarm2022} and R Statistical Software \citep{rsoftware2022}
-(v4.4.4). For all models, we used default, weakly informative priors,
-four chains, and 5000 iterations each for warmup and post-warmup. We
-checked MCMC convergence using trace plots and Rhat, and evaluated model
-fit using leave-one-out cross-validation \citep{vehtari2016}, as
-implemented by the loo package \citep{vehtari2022}. (See
-\textbf{Supplementary Methods - Frog population recovery - Among-site
-survival modeling} for details.)
+site\_id. We performed all analyses with the rstanarm
+package\textsuperscript{59} and R Statistical
+Software\textsuperscript{56} (v4.4.4). For all models, we used default,
+weakly informative priors, four chains, and 5000 iterations each for
+warmup and post-warmup. We checked MCMC convergence using trace plots
+and Rhat, and evaluated model fit using leave-one-out
+cross-validation\textsuperscript{60}, as implemented by the loo
+package\textsuperscript{61}. (See \textbf{Supplementary Methods - Frog
+population recovery - Among-site survival modeling} for details.)
 
@@
 
-We analyzed skin swabs using standard Bd DNA extraction and qPCR methods
-\citep{boyle2004} (see \textbf{Supplementary Information - Frog
-population recovery - Laboratory methods} for details). To assess the
-magnitude of changes in Bd load on frogs following translocation, we
+We analyzed skin swabs using standard Bd DNA extraction and qPCR
+methods\textsuperscript{62} (see \textbf{Supplementary Information -
+Frog population recovery - Laboratory methods} for details). To assess
+the magnitude of changes in Bd load on frogs following translocation, we
 compared Bd loads measured before versus after translocation.
@@
 decision because (i) translocated populations are infected with Bd at
-high prevalence \citep{joseph2018}, and (ii) host density does not seem
+high prevalence\textsuperscript{39}, and (ii) host density does not seem
 to play a significant role in multi-year Bd infection dynamics in this
-system \citep{wilber2022}. Thus, ignoring Bd infection dynamics and
+system\textsuperscript{63}. Thus, ignoring Bd infection dynamics and
 instead assuming all host vital rates are in the presence of high Bd
@@
 The data used in this study are available at
-\url{https://doi.org/10.5281/zenodo.13851590} \citep{knapp2024a}. This
+\url{https://doi.org/10.5281/zenodo.13851590}\textsuperscript{64}. This
 includes the estimated survival of each reintroduced frog cohort, data
@@
 to calculate post-translocation frog survival are available at
-\url{https://doi.org/10.5281/zenodo.13851652} \citep{knapp2024b}.
+\url{https://doi.org/10.5281/zenodo.13851652}\textsuperscript{65}.
 
@@
 R code to replicate our analyses is available at
-\url{https://doi.org/10.5281/zenodo.13851590} \citep{knapp2024a}. R code
-to analyze the CMR datasets is available at
-\url{https://doi.org/10.5281/zenodo.13851652} \citep{knapp2024b}.
+\url{https://doi.org/10.5281/zenodo.13851590}\textsuperscript{64}. R
+code to analyze the CMR datasets is available at
+\url{https://doi.org/10.5281/zenodo.13851652}\textsuperscript{65}.
 
@@
 
-\renewcommand{\bibsection}{}
-\bibliography{translocation.bib}
+\phantomsection\label{refs}
+\begin{CSLReferences}{0}{0}
+\bibitem[\citeproctext]{ref-ceballos2015}
+\CSLLeftMargin{1. }%
+\CSLRightInline{Ceballos, G. \emph{et al.} Accelerated modern
+human-induced species losses: Entering the sixth mass extinction.
+\emph{Science Advances} \textbf{1}, (2015).}
+
+\bibitem[\citeproctext]{ref-naeem2009}
+\CSLLeftMargin{2. }%
+\CSLRightInline{Naeem, S., Bunker, D. E., Hector, A., Loreau, M. \&
+Perrings, C. \emph{Biodiversity, Ecosystem Functioning, and Human
+Wellbeing: An Ecological and Economic Perspective}. (Oxford University
+Press, Oxford, 2009).}
+
+\bibitem[\citeproctext]{ref-jones2008}
+\CSLLeftMargin{3. }%
+\CSLRightInline{Jones, K. E. \emph{et al.} Global trends in emerging
+infectious diseases. \emph{Nature} \textbf{451}, 990--993 (2008).}
+
+\bibitem[\citeproctext]{ref-fisher2012}
+\CSLLeftMargin{4. }%
+\CSLRightInline{Fisher, M. C. \emph{et al.} Emerging fungal threats to
+animal, plant and ecosystem health. \emph{Nature} \textbf{484}, 186--94
+(2012).}
+
+\bibitem[\citeproctext]{ref-daszak2000}
+\CSLLeftMargin{5. }%
+\CSLRightInline{Daszak, P., Cunningham, A. A. \& Hyatt, A. D. Emerging
+infectious diseases of wildlife -- threats to biodiversity and human
+health. \emph{Science} \textbf{287}, 443--449 (2000).}
+
+\bibitem[\citeproctext]{ref-hewson2014}
+\CSLLeftMargin{6. }%
+\CSLRightInline{Hewson, I. \emph{et al.} Densovirus associated with
+sea-star wasting disease and mass mortality. \emph{Proceedings of the
+National Academy of Sciences USA} \textbf{111}, 17278--17283 (2014).}
+
+\bibitem[\citeproctext]{ref-samuel2015}
+\CSLLeftMargin{7. }%
+\CSLRightInline{Samuel, M. D., Woodworth, B. L., Atkinson, C. T., Hart,
+P. J. \& LaPointe, D. A. Avian malaria in {H}awaiian forest birds:
+Infection and population impacts across species and elevations.
+\emph{Ecosphere} \textbf{6}, art104 (2015).}
+
+\bibitem[\citeproctext]{ref-scheele2019}
+\CSLLeftMargin{8. }%
+\CSLRightInline{Scheele, B. C. \emph{et al.} Amphibian fungal panzootic
+causes catastrophic and ongoing loss of biodiversity. \emph{Science}
+\textbf{363}, 1459--1463 (2019).}
+
+\bibitem[\citeproctext]{ref-cunningham2021}
+\CSLLeftMargin{9. }%
+\CSLRightInline{Cunningham, C. X. \emph{et al.} Quantifying 25 years of
+disease-caused declines in {T}asmanian devil populations: {h}ost density
+drives spatial pathogen spread. \emph{Ecology Letters} \textbf{24},
+958--969 (2021).}
+
+\bibitem[\citeproctext]{ref-luedtke2023}
+\CSLLeftMargin{10. }%
+\CSLRightInline{Luedtke, J. A. \emph{et al.} Ongoing declines for the
+world{'}s amphibians in the face of emerging threats. \emph{Nature}
+\textbf{622}, 308--314 (2023).}
+
+\bibitem[\citeproctext]{ref-raberg2009}
+\CSLLeftMargin{11. }%
+\CSLRightInline{Råberg, L., Graham, A. L. \& Read, A. F. Decomposing
+health: Tolerance and resistance to parasites in animals.
+\emph{Philosophical Transactions of the Royal Society B: Biological
+Sciences} \textbf{364}, 37--49 (2008).}
+
+\bibitem[\citeproctext]{ref-brannelly2021}
+\CSLLeftMargin{12. }%
+\CSLRightInline{Brannelly, L. A. \emph{et al.} Mechanisms underlying
+host persistence following amphibian disease emergence determine
+appropriate management strategies. \emph{Ecology Letters} \textbf{24},
+130--148 (2021).}
+
+\bibitem[\citeproctext]{ref-garner2016}
+\CSLLeftMargin{13. }%
+\CSLRightInline{Garner, T. W. J. \emph{et al.} Mitigating amphibian
+chytridiomycoses in nature. \emph{Philosophical Transactions of the
+Royal Society B: Biological Sciences} \textbf{371}, 20160207 (2016).}
+
+\bibitem[\citeproctext]{ref-scheele2017}
+\CSLLeftMargin{14. }%
+\CSLRightInline{Scheele, B. C. \emph{et al.} After the epidemic: Ongoing
+declines, stabilizations and recoveries in amphibians afflicted by
+chytridiomycosis. \emph{Biological Conservation} \textbf{206}, 37--46
+(2017).}
+
+\bibitem[\citeproctext]{ref-voyles2018}
+\CSLLeftMargin{15. }%
+\CSLRightInline{Voyles, J. \emph{et al.} Shifts in disease dynamics in a
+tropical amphibian assemblage are not due to pathogen attenuation.
+\emph{Science} \textbf{359}, 1517--1519 (2018).}
+
+\bibitem[\citeproctext]{ref-knapp2016}
+\CSLLeftMargin{16. }%
+\CSLRightInline{Knapp, R. A. \emph{et al.} Large-scale recovery of an
+endangered amphibian despite ongoing exposure to multiple stressors.
+\emph{Proceedings of the National Academy of Sciences USA} \textbf{113},
+11889--11894 (2016).}
+
+\bibitem[\citeproctext]{ref-rosenblum2012}
+\CSLLeftMargin{17. }%
+\CSLRightInline{Rosenblum, E. B., Poorten, T. J., Settles, M. \&
+Murdoch, G. K. Only skin deep: Shared genetic response to the deadly
+chytrid fungus in susceptible frog species. \emph{Molecular Ecology}
+\textbf{21}, 3110--3120 (2012).}
+
+\bibitem[\citeproctext]{ref-fites2013}
+\CSLLeftMargin{18. }%
+\CSLRightInline{Fites, J. S. \emph{et al.} The invasive chytrid fungus
+of amphibians paralyzes lymphocyte responses. \emph{Science}
+\textbf{342}, 366369 (2013).}
+
+\bibitem[\citeproctext]{ref-grogan2018a}
+\CSLLeftMargin{19. }%
+\CSLRightInline{Grogan, L. F. \emph{et al.} Review of the amphibian
+immune response to chytridiomycosis, and future directions.
+\emph{Frontiers in Immunology} \textbf{9}, 2536 (2018).}
+
+\bibitem[\citeproctext]{ref-savage2016}
+\CSLLeftMargin{20. }%
+\CSLRightInline{Savage, A. E. \& Zamudio, K. R. Adaptive tolerance to a
+pathogenic fungus drives major histocompatibility complex evolution in
+natural amphibian populations. \emph{Proceedings of the Royal Society B:
+Biological Sciences} \textbf{283}, 20153115 (2016).}
+
+\bibitem[\citeproctext]{ref-grogan2018b}
+\CSLLeftMargin{21. }%
+\CSLRightInline{Grogan, L. F. \emph{et al.} Evolution of resistance to
+chytridiomycosis is associated with a robust early immune response.
+\emph{Molecular Ecology} \textbf{27}, 919--934 (2018).}
+
+\bibitem[\citeproctext]{ref-hammond2021}
+\CSLLeftMargin{22. }%
+\CSLRightInline{Hammond, T. T. \emph{et al.} Overwinter behavior,
+movement, and survival in a recently reintroduced, endangered amphibian,
+\emph{{R}ana muscosa}. \emph{Journal for Nature Conservation}
+\textbf{64}, 126086 (2021).}
+
+\bibitem[\citeproctext]{ref-knapp2022}
+\CSLLeftMargin{23. }%
+\CSLRightInline{Knapp, R. A. \emph{et al.} Effectiveness of antifungal
+treatments during chytridiomycosis epizootics in populations of an
+endangered frog. \emph{PeerJ} \textbf{10}, e12712 (2022).}
+
+\bibitem[\citeproctext]{ref-stockwell2008}
+\CSLLeftMargin{24. }%
+\CSLRightInline{Stockwell, M., Clulow, S., Clulow, J. \& Mahony, M. The
+impact of the {Amphibian Chytrid Fungus} on a {Green and Golden Bell
+Frog} {\emph{{L}itoria aurea}} reintroduction program at the {Hunter
+Wetlands Centre Australia in the Hunter Region of NSW}. \emph{Australian
+Zoologist} \textbf{34}, 379--386 (2008).}
+
+\bibitem[\citeproctext]{ref-scheele2021}
+\CSLLeftMargin{25. }%
+\CSLRightInline{Scheele, B. C. \emph{et al.} Conservation translocations
+for amphibian species threatened by chytrid fungus: A review, conceptual
+framework, and recommendations. \emph{Conservation Science and Practice}
+\textbf{3}, e524 (2021).}
+
+\bibitem[\citeproctext]{ref-grinnell1924}
+\CSLLeftMargin{26. }%
+\CSLRightInline{Grinnell, J. \& Storer, T. I. \emph{{A}nimal {L}ife in
+the {Y}osemite}. (University of California Press, Berkeley, 1924).}
+
+\bibitem[\citeproctext]{ref-vredenburg2007}
+\CSLLeftMargin{27. }%
+\CSLRightInline{Vredenburg, V. T. \emph{et al.} Concordant molecular and
+phenotypic data delineate new taxonomy and conservation priorities for
+the endangered mountain yellow-legged frog. \emph{Journal of Zoology}
+\textbf{271}, 361--374 (2007).}
+
+\bibitem[\citeproctext]{ref-bradford1989}
+\CSLLeftMargin{28. }%
+\CSLRightInline{Bradford, D. F. Allotopic distribution of native frogs
+and introduced fishes in high {S}ierra {N}evada lakes of {C}alifornia:
+{i}mplication of the negative effect of fish introductions.
+\emph{Copeia} \textbf{1989}, 775--778 (1989).}
+
+\bibitem[\citeproctext]{ref-knapp2000}
+\CSLLeftMargin{29. }%
+\CSLRightInline{Knapp, R. A. \& Matthews, K. R. Non-native fish
+introductions and the decline of the mountain yellow-legged frog from
+within protected areas. \emph{Conservation Biology} \textbf{14},
+428--438 (2000).}
+
+\bibitem[\citeproctext]{ref-vredenburg2019}
+\CSLLeftMargin{30. }%
+\CSLRightInline{Vredenburg, V. T. \emph{et al.} Pathogen invasion
+history elucidates contemporary host pathogen dynamics. \emph{PLOS ONE}
+\textbf{14}, e0219981 (2019).}
+
+\bibitem[\citeproctext]{ref-vredenburg2010}
+\CSLLeftMargin{31. }%
+\CSLRightInline{Vredenburg, V. T., Knapp, R. A., Tunstall, T. S. \&
+Briggs, C. J. Dynamics of an emerging disease drive large-scale
+amphibian population extinctions. \emph{Proceedings of the National
+Academy of Sciences USA} \textbf{107}, 9689--9694 (2010).}
+
+\bibitem[\citeproctext]{ref-rachowicz2006}
+\CSLLeftMargin{32. }%
+\CSLRightInline{Rachowicz, L. J. \emph{et al.} Emerging infectious
+disease as a proximate cause of amphibian mass mortality. \emph{Ecology}
+\textbf{87}, 1671--1683 (2006).}
+
+\bibitem[\citeproctext]{ref-knapp1998}
+\CSLLeftMargin{33. }%
+\CSLRightInline{Knapp, R. A. \& Matthews, K. R. Eradication of nonnative
+fish by gill netting from a small mountain lake in {C}alifornia.
+\emph{Restoration Ecology} \textbf{6}, 207213 (1998).}
+
+\bibitem[\citeproctext]{ref-knapp2007}
+\CSLLeftMargin{34. }%
+\CSLRightInline{Knapp, R. A., Boiano, D. M. \& Vredenburg, V. T. Removal
+of nonnative fish results in population expansion of a declining
+amphibian (mountain yellow-legged frog, \emph{{R}ana muscosa}).
+\emph{Biological Conservation} \textbf{135}, 11--20 (2007).}
+
+\bibitem[\citeproctext]{ref-vredenburg2004}
+\CSLLeftMargin{35. }%
+\CSLRightInline{Vredenburg, V. T. Reversing introduced species effects:
+Experimental removal of introduced fish leads to rapid recovery of a
+declining frog. \emph{Proceedings of the National Academy of Sciences
+USA} \textbf{101}, 7646--7650 (2004).}
+
+\bibitem[\citeproctext]{ref-walker2007}
+\CSLLeftMargin{36. }%
+\CSLRightInline{Walker, S. \emph{et al.} Environmental detection of
+\emph{{B}atrachochytrium dendrobatidis} in a temperate climate.
+\emph{Diseases of Aquatic Organisms} \textbf{77}, 105--112 (2007).}
+
+\bibitem[\citeproctext]{ref-briggs2010}
+\CSLLeftMargin{37. }%
+\CSLRightInline{Briggs, C. J., Knapp, R. A. \& Vredenburg, V. T.
+Enzootic and epizootic dynamics of the chytrid fungal pathogen of
+amphibians. \emph{Proceedings of the National Academy of Sciences USA}
+\textbf{107}, 9695--9700 (2010).}
+
+\bibitem[\citeproctext]{ref-knapp2011}
+\CSLLeftMargin{38. }%
+\CSLRightInline{Knapp, R. A., Briggs, C. J., Smith, T. C. \& Maurer, J.
+R. Nowhere to hide: Impact of a temperature-sensitive amphibian pathogen
+along an elevation gradient in the temperate zone. \emph{Ecosphere}
+\textbf{2}, art93 (2011).}
+
+\bibitem[\citeproctext]{ref-joseph2018}
+\CSLLeftMargin{39. }%
+\CSLRightInline{Joseph, M. B. \& Knapp, R. A. Disease and climate
+effects on individuals drive post{-}reintroduction population dynamics
+of an endangered amphibian. \emph{Ecosphere} \textbf{9}, e02499 (2018).}
+
+\bibitem[\citeproctext]{ref-carlson2014}
+\CSLLeftMargin{40. }%
+\CSLRightInline{Carlson, S. M., Cunningham, C. J. \& Westley, P. A.
+Evolutionary rescue in a changing world. \emph{Trends in Ecology \&
+Evolution} \textbf{29}, 521--530 (2014).}
+
+\bibitem[\citeproctext]{ref-searle2020}
+\CSLLeftMargin{41. }%
+\CSLRightInline{Searle, C. L. \& Christie, M. R. Evolutionary rescue in
+host-pathogen systems. \emph{Evolution} \textbf{75}, 2948--2958 (2021).}
+
+\bibitem[\citeproctext]{ref-mendelson2019}
+\CSLLeftMargin{42. }%
+\CSLRightInline{Mendelson III, J. R., Whitfield, S. M. \& Sredl, M. J.
+{A recovery engine strategy for amphibian conservation in the context of
+disease}. \emph{Biological Conservation} \textbf{236}, 188--191 (2019).}
+
+\bibitem[\citeproctext]{ref-joseph2019}
+\CSLLeftMargin{43. }%
+\CSLRightInline{Joseph, M. B. {mrmr: mark recapture miscellany in R}.
+(2019).}
+
+\bibitem[\citeproctext]{ref-dodd2005}
+\CSLLeftMargin{44. }%
+\CSLRightInline{Dodd Jr, C. Amphibian conservation and population
+manipulation. in \emph{Status and conservation of {US} amphibians} (ed.
+Lannoo, M.) 265--270 (University of California Press, Berkeley, 2005).}
+
+\bibitem[\citeproctext]{ref-seddon2007}
+\CSLLeftMargin{45. }%
+\CSLRightInline{Seddon, P. J., Armstrong, D. P. \& Maloney, R. F.
+Developing the science of reintroduction biology. \emph{Conservation
+Biology} \textbf{21}, 303--312 (2007).}
+
+\bibitem[\citeproctext]{ref-ellison2019}
+\CSLLeftMargin{46. }%
+\CSLRightInline{Ellison, S., Knapp, R. A., Sparagon, W., Swei, A. \&
+Vredenburg, V. T. Reduced skin bacterial diversity correlates with
+increased pathogen infection intensity in an endangered amphibian host.
+\emph{Molecular Ecology} \textbf{28}, 127--140 (2019).}
+
+\bibitem[\citeproctext]{ref-humphries2022}
+\CSLLeftMargin{47. }%
+\CSLRightInline{Humphries, J. E. \emph{et al.} Do immune system changes
+at metamorphosis predict vulnerability to chytridiomycosis? {A}n update.
+\emph{{Developmental and Comparative Immunology}} \textbf{136}, 104510
+(2022).}
+
+\bibitem[\citeproctext]{ref-hollanders2022}
+\CSLLeftMargin{48. }%
+\CSLRightInline{Hollanders, M., Grogan, L. F., Nock, C. J., McCallum, H.
+I. \& Newell, D. A. {Recovered frog populations coexist with endemic
+\emph{Batrachochytrium dendrobatidis} despite load{-}dependent
+mortality}. \emph{Ecological Applications} \textbf{33}, (2022).}
+
+\bibitem[\citeproctext]{ref-bradford1993}
+\CSLLeftMargin{49. }%
+\CSLRightInline{Bradford, D. F., Tabatabai, F. \& Graber, D. M.
+Isolation of remaining populations of the native frog, {\emph{{R}ana
+muscosa}}, by introduced fishes in sequoia and kings canyon national
+parks, california. \emph{Conservation Biology} \textbf{7}, 882--888
+(1993).}
+
+\bibitem[\citeproctext]{ref-heard2015}
+\CSLLeftMargin{50. }%
+\CSLRightInline{Heard, G. W. \emph{et al.} Refugia and connectivity
+sustain amphibian metapopulations afflicted by disease. \emph{Ecology
+Letters} \textbf{18}, 853--863 (2015).}
+
+\bibitem[\citeproctext]{ref-canessa2019}
+\CSLLeftMargin{51. }%
+\CSLRightInline{Canessa, S. \emph{et al.} Adaptive management of species
+recovery programs: A real-world application for an endangered amphibian.
+\emph{Biological Conservation} \textbf{236}, 202--210 (2019).}
+
+\bibitem[\citeproctext]{ref-kosch2022}
+\CSLLeftMargin{52. }%
+\CSLRightInline{Kosch, T. A. \emph{et al.} Genetic approaches for
+increasing fitness in endangered species. \emph{Trends in Ecology \&
+Evolution} \textbf{37}, 332--345 (2022).}
+
+\bibitem[\citeproctext]{ref-whiteley2015}
+\CSLLeftMargin{53. }%
+\CSLRightInline{Whiteley, A. R., Fitzpatrick, S. W., Funk, W. C. \&
+Tallmon, D. A. Genetic rescue to the rescue. \emph{Trends in Ecology \&
+Evolution} \textbf{30}, 42--49 (2015).}
+
+\bibitem[\citeproctext]{ref-poorten2017}
+\CSLLeftMargin{54. }%
+\CSLRightInline{Poorten, T. J., Knapp, R. A. \& Rosenblum, E. B.
+{Population genetic structure of the endangered Sierra Nevada
+yellow-legged frog (\emph{{R}ana sierrae}) in Yosemite National Park
+based on multi-locus nuclear data from swab samples}. \emph{Conservation
+Genetics} \textbf{18}, 731--744 (2017).}
+
+\bibitem[\citeproctext]{ref-knapp2005}
+\CSLLeftMargin{55. }%
+\CSLRightInline{Knapp, R. A. Effects of nonnative fish and habitat
+characteristics on lentic herpetofauna in {Yosemite National Park},
+{USA}. \emph{Biological Conservation} \textbf{121}, 265--279 (2005).}
+
+\bibitem[\citeproctext]{ref-rsoftware2022}
+\CSLLeftMargin{56. }%
+\CSLRightInline{R Core Team. \emph{{R: A Language and Environment for
+Statistical Computing}}. (R Foundation for Statistical Computing,
+Vienna, Austria, 2022).}
+
+\bibitem[\citeproctext]{ref-gelman2013}
+\CSLLeftMargin{57. }%
+\CSLRightInline{Gelman, A. \emph{et al.} \emph{Bayesian Data Analysis}.
+(CRC Press, New York, 2013). doi:10.1201/b16018.}
+
+\bibitem[\citeproctext]{ref-gabry2019}
+\CSLLeftMargin{58. }%
+\CSLRightInline{Gabry, J., Simpson, D., Vehtari, A., Betancourt, M. \&
+Gelman, A. Visualization in {B}ayesian workflow. \emph{Journal of the
+Royal Statistical Society: Series A} \textbf{182}, 389--402 (2019).}
+
+\bibitem[\citeproctext]{ref-rstanarm2022}
+\CSLLeftMargin{59. }%
+\CSLRightInline{Goodrich, B., Gabry, J., Ali, I. \& Brilleman, S.
+{r}stanarm: Bayesian applied regression modeling via {Stan}. (2022).}
+
+\bibitem[\citeproctext]{ref-vehtari2016}
+\CSLLeftMargin{60. }%
+\CSLRightInline{Vehtari, A., Gelman, A. \& Gabry, J. Practical
+{B}ayesian model evaluation using leave-one-out cross-validation and
+{WAIC}. \emph{Statistics and Computing} \textbf{27}, 1413--1432 (2016).}
+
+\bibitem[\citeproctext]{ref-vehtari2022}
+\CSLLeftMargin{61. }%
+\CSLRightInline{Vehtari, A. \emph{et al.} {l}oo: Efficient leave-one-out
+cross-validation and WAIC for {B}ayesian models. (2022).}
+
+\bibitem[\citeproctext]{ref-boyle2004}
+\CSLLeftMargin{62. }%
+\CSLRightInline{Boyle, D., Boyle, D., Olsen, V., Morgan, J. \& Hyatt, A.
+Rapid quantitative detection of chytridiomycosis
+(\emph{{B}atrachochytrium dendrobatidis}) in amphibian samples using
+real-time {T}aqman {PCR} assay. \emph{Diseases of Aquatic Organisms}
+\textbf{60}, 141--148 (2004).}
+
+\bibitem[\citeproctext]{ref-wilber2022}
+\CSLLeftMargin{63. }%
+\CSLRightInline{Wilber, M. Q., Knapp, R. A., Smith, T. C. \& Briggs, C.
+J. Host density has limited effects on pathogen invasion,
+disease-induced declines and within-host infection dynamics across a
+landscape of disease. \emph{Journal of Animal Ecology} \textbf{91},
+2451--2464 (2022).}
+
+\bibitem[\citeproctext]{ref-knapp2024a}
+\CSLLeftMargin{64. }%
+\CSLRightInline{Knapp, R. A., Wilber, M. Q., Joseph, M. B., Smith, T. C.
+\& Grasso, R. L. Reintroduction of resistant frogs facilitates
+landscape-scale recovery in the presence of a lethal fungal disease.
+\emph{GitHub repository} https://doi.org/10.5281/zenodo.13851652 (2024)
+doi:10.5281/zenodo.13851652.}
+
+\bibitem[\citeproctext]{ref-knapp2024b}
+\CSLLeftMargin{65. }%
+\CSLRightInline{Knapp, R. A., Wilber, M. Q., Joseph, M. B., Smith, T. C.
+\& Grasso, R. L. Capture-mark-recapture datasets and analyses for:
+Reintroduction of resistant frogs facilitates landscape-scale recovery
+in the presence of a lethal fungal disease. \emph{GitHub repository}
+https://doi.org/10.5281/zenodo.13851590 (2024)
+doi:10.5281/zenodo.13851590.}
+
+\end{CSLReferences}
 

### 2024-09-27 doc/manuscript/translocation.tex
Use default cite method instead of natbib

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 

### 2024-09-28 doc/manuscript/submission_files/translocation.tex
Remove file path for all figures

@@
 \makeatother
-% journal abbreviations
-\usepackage{jabbrv}
 % definitions for citeproc citations
@@
 we are currently doing so using a modified study design. Specifically,
-we are conducting translocations in which frogs from several donor
+we are conducting reintroductions in which frogs from several donor
 populations are added concurrently to a single site. Survival of frogs
@@
 translocation sites. For all models, we used mrmr defaults for priors,
-number of chains (4), and warmup and post-warmup iterations (2000 for
+\_number of chains (4), and warmup and post-warmup iterations (2000 for
 each). We evaluated convergence of the Markov chain Monte Carlo (MCMC)
@@
 instead assuming all host vital rates are in the presence of high Bd
-prevalence significantly simplifies the model without much loss of
+prevalence significantly simplifies the model without\_ much loss of
 realism. Additional details are provided in \textbf{Supplementary
@@
 
-\subsubsection{Figures and Legends}\label{figures-and-legends}
+\subsubsection{Figures and Figure
+Legends}\label{figures-and-figure-legends}
 
@@
 
-\includegraphics[width=0.95\textwidth,height=\textheight]{figures/conceptual_model.png}
+\includegraphics{conceptual_model_180mm.pdf}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{figures/map_translocation_points.png}
+\includegraphics{map_translocation_points_88mm.pdf}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{figures/translocation_survival_bysiteid.png}
+\includegraphics{translocation_survival_bysiteid_88mm.pdf}
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{figures/cond_effects_plot.png}
+\includegraphics{conditional_effects_plot_88mm.pdf}
 
@@
 
-\includegraphics[width=0.95\textwidth,height=\textheight]{figures/pop_viability_figures_for_manuscript.png}
+\includegraphics{pop_viability_figures_for_manuscript_180mm.pdf}
 

### 2024-09-28 doc/manuscript/translocation.qmd
Edits to render to latex
Use default cite method instead of natbib
Minor edits

@@
 ---
-format: pdf
-#format:
-#   latex: 
+#format: pdf
+format:
+   latex 
 #      cite-method: natbib
@@
 Although our study design did not allow us to completely separate the effects of donor population and recipient site on 1-year adult survival, we are currently doing so using a modified study design.
-Specifically, we are conducting translocations in which frogs from several donor populations are added concurrently to a single site.
+Specifically, we are conducting reintroductions in which frogs from several donor populations are added concurrently to a single site.
 Survival of frogs from each donor population is estimated using CMR methods and the contribution of each donor population to subsequent generations is quantified using genetic methods.
@@
 
-### Figures and Legends
+### Figures and Figure Legends
 
@@
 Blue text indicates components that are included in the current study.
-The general temporal scale of the depicted scenario is provided by the timeline, with the dashed portion indicating a projection into the future.](figures/conceptual_model.png){#fig-recovery-model width="95%"}
+The general temporal scale of the depicted scenario is provided by the timeline, with the dashed portion indicating a projection into the future.](submission_files/figures/conceptual_model_180mm.pdf){#fig-recovery-model}
 
@@
 In both maps, topography is shown with a grayscale hillshade layer, and elevation is indicated by colors layered over the hillshade.
-In the absence of hillshading, dark green colors indicate the lowest elevations and white indicates the highest elevation.](figures/map_translocation_points.png){#fig-yosemap width="60%"}
+In the absence of hillshading, dark green colors indicate the lowest elevations and white indicates the highest elevation.](submission_files/figures/map_translocation_points_88mm.pdf){#fig-yosemap}
 
@@
 Dot colors indicate the donor population from which frogs in each translocated cohort were collected.
-When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="60%"}
+When multiple translocations were conducted to a site, points and error bars are slightly offset to avoid overlap.](submission_files/figures/translocation_survival_bysiteid_88mm.pdf){#fig-translocation-survival}
 
@@
 In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals for the posterior distributions of the predicted effects.
-In (c) and (d), box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots) for the posterior distributions of the predicted effects.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
+In (c) and (d), box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots) for the posterior distributions of the predicted effects.](submission_files/figures/conditional_effects_plot_88mm.pdf){#fig-cond-effects}
 
@@
 There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are partially or completely hidden beneath the line for 70505.
-**c** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.png){#fig-viability width="95%"}
+**c** 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](submission_files/figures/pop_viability_figures_for_manuscript_180mm.pdf){#fig-viability}

### 2024-09-28 doc/manuscript/translocation.tex
Minor edits

@@
 
-\includegraphics[width=0.95\textwidth,height=\textheight]{figures/conceptual_model.png}
+\includegraphics{submission_files/figures/conceptual_model_180mm.pdf}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{figures/map_translocation_points.png}
+\includegraphics{submission_files/figures/map_translocation_points_88mm.pdf}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{figures/translocation_survival_bysiteid.png}
+\includegraphics{submission_files/figures/translocation_survival_bysiteid_88mm.pdf}
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{figures/cond_effects_plot.png}
+\includegraphics{submission_files/figures/conditional_effects_plot_88mm.pdf}
 
@@
 
-\includegraphics[width=0.95\textwidth,height=\textheight]{figures/pop_viability_figures_for_manuscript.png}
+\includegraphics{submission_files/figures/pop_viability_figures_for_manuscript_180mm.pdf}
 

### 2024-10-27 doc/manuscript/translocation.qmd
Fix typos

@@
 
-Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]a) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.10, at least six of twelve translocated populations are likely to experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]a; median predicted $\lambda$ ranges from 1.07-1.28 for these six populations).
+Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of [@fig-viability]b) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.10, at least six of twelve translocated populations are likely to experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd ([@fig-viability]a; median predicted $\lambda$ ranges from 1.07-1.28 for these six populations).
 These six populations all had observed yearly adult survival greater than 0.5.
